### 2006-01-02 talk/chalmers/techpaper.tex
Copied the CCC paper (itself copied from architecture.txt) as the Chalmers Winter Meeting paper.  Changed the style, the title and the authors (mentioned all non-trivial svn authors in a footnote, for want of a better idea).

\documentclass[a4paper,11pt,english]{article}
\usepackage{babel}
\usepackage{ae}
\usepackage{aeguill}
\usepackage{shortvrb}
\usepackage[latin1]{inputenc}
\usepackage{tabularx}
\usepackage{longtable}
\setlength{\extrarowheight}{2pt}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{color}
\usepackage{multirow}
\usepackage{ifthen}
\usepackage[colorlinks=false,linkcolor=black,urlcolor=black]{hyperref}
\usepackage[DIV12]{typearea}
%% generator Docutils: http://docutils.sourceforge.net/
\newlength{\admonitionwidth}
\setlength{\admonitionwidth}{0.9\textwidth}
\newlength{\docinfowidth}
\setlength{\docinfowidth}{0.9\textwidth}
\newlength{\locallinewidth}
\newcommand{\optionlistlabel}[1]{\bf #1 \hfill}
\newenvironment{optionlist}[1]
{\begin{list}{}
  {\setlength{\labelwidth}{#1}
   \setlength{\rightmargin}{1cm}
   \setlength{\leftmargin}{\rightmargin}
   \addtolength{\leftmargin}{\labelwidth}
   \addtolength{\leftmargin}{\labelsep}
   \renewcommand{\makelabel}{\optionlistlabel}}
}{\end{list}}
\newlength{\lineblockindentation}
\setlength{\lineblockindentation}{2.5em}
\newenvironment{lineblock}[1]
{\begin{list}{}
  {\setlength{\partopsep}{\parskip}
   \addtolength{\partopsep}{\baselineskip}
   \topsep0pt\itemsep0.15\baselineskip\parsep0pt
   \leftmargin#1}
 \raggedright}
{\end{list}}
% begin: floats for footnotes tweaking.
\setlength{\floatsep}{0.5em}
\setlength{\textfloatsep}{\fill}
\addtolength{\textfloatsep}{3em}
\renewcommand{\textfraction}{0.5}
\renewcommand{\topfraction}{0.5}
\renewcommand{\bottomfraction}{0.5}
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}
% end floats for footnotes
% some commands, that could be overwritten in the style file.
\newcommand{\rubric}[1]{\subsection*{~\hfill {\it #1} \hfill ~}}
\newcommand{\titlereference}[1]{\textsl{#1}}
% end of "some commands"
\input{techpaper.sty}
\title{PyPy - Architecture Overview}
\hypersetup{
pdftitle={PyPy - Architecture Overview}
}
\raggedbottom
\begin{document}
\maketitle

\setlength{\locallinewidth}{\linewidth}
% command to produce latex file: 
% rst2latex.py --use-latex-citations --use-latex-footnotes --use-latex-toc --documentoptions=a4paper,11pt --hyperlink-color=0 techpaper.txt --use-latex-docinfo --stylesheet=techpaper.sty techpaper.tex 
\begin{abstract}
PyPy\footnotemark[1] is an implementation of the Python\footnotemark[2] programming language written in
Python itself, flexible and easy to experiment with.  Our long-term goals are
to target a large variety of platforms, small and large, by providing a
compiler toolsuite that can produce custom Python versions.  Platform, memory
and threading models are to become aspects of the translation process - as
opposed to encoding low level details into the language implementation itself.
Eventually, dynamic optimization techniques - implemented as another
translation aspect - should become robust against language changes.
\footnotetext[1]{
\href{http://codespeak.net/pypy}{http://codespeak.net/pypy}
}
\footnotetext[2]{
\href{http://docs.python.org/ref}{http://docs.python.org/ref}
}
\end{abstract}

%___________________________________________________________________________

\hypertarget{pypy-an-implementation-of-python-in-python}{}
\section{PyPy - an implementation of Python in Python}

It has become a tradition in the development of computer languages to
implement each language in itself.  This serves many purposes. By doing so,
you demonstrate the versatility of the language and its applicability for
large projects.  Writing compilers and interpreters are among the most
complex endeavours in software development.

An important aspect of implementing Python in Python is the high level of
abstraction and compactness of the language. This allows an implementation
that is, in some respects, easier to understand and play with than the one
written in C (referred to throughout the PyPy documentation and source as
``CPython''\footnotemark[3]).
\footnotetext[3]{
\href{http://www.python.org}{http://www.python.org}
}

Another central idea in PyPy is building the implementation in the form
of a number of independent modules with clearly defined and well tested API's. 
This eases reuse and allows experimenting with multiple implementations 
of specific features.

Later in the project we will introduce optimizations, following the
ideas of Psyco\footnotemark[4], a Just in Time Specializer, that should make PyPy
run Python programs faster than CPython. Extensions that increase the
expressive power are also planned. For instance, we will include the
ideas of Stackless\footnotemark[5], which moves the execution frames off the stack into
heap space, allowing for massive parallellism.
\footnotetext[4]{
\href{http://psyco.sourceforge.net}{http://psyco.sourceforge.net}
}
\footnotetext[5]{
\href{http://stackless.com}{http://stackless.com}
}

%___________________________________________________________________________

\hypertarget{pypy-meta-goals}{}
\section{PyPy - Meta Goals}

PyPy is not only about writing another Python interpreter.
Traditionally, interpreters are written in a target platform language
like C/Posix, Java or C{\#}.  Each such interpreter provides a ``mapping''
from application source code to the target environment.  One of the
goals of the ``all-encompassing'' environments, like the .NET framework
and to some extent the Java virtual machine, is to provide standardized
and higher level functionalities to language implementors.  This reduces
the burden of having to write and maintain many interpreters or
compilers.

PyPy is experimenting with a different approach.  We are not writing a
Python interpreter for a specific target platform.  We have written a
Python interpreter in Python, with as few references as possible to
low-level details.  (Because of the nature of Python, this is already
a complicated task, although not as complicated as writing it in - say
- C.)  Then we use this as a ``language specification'' and manipulate
it to produce the more traditional interpreters that we want.  In the
above sense, we are generating the concrete ``mappings'' of Python into
lower-level target platforms.

So far (autumn 2005), we have already succeeded in turning this ``language
specification'' into reasonably efficient C-level code that performs
basically the same job as CPython.  Memory management is inserted during
this \emph{translation} process.  It can be configured to use reference
counting or not; thus we have already achieved two very different
mappings of application Python code over C/Posix.  We have
also successfully translated our Python interpreter into LLVM\footnotemark[6] code,
and we are working on targeting higher-level environments like
Java and Squeak.

In some senses, PyPy project's central component is not its
interpreter implementation, but its configurable translator.
We think it provides a good way to avoid writing \texttt{n * m * o}
interpreters for \texttt{n} dynamic languages and \texttt{m} platforms
with \texttt{o} crucial design decisions.  PyPy aims at having any
one of these parameters changeable independently from each other:
\begin{itemize}
\item {} 
we can modify or replace the language we interpret and just regenerate
a concrete interpreter for each target;

\item {} 
we can write new translator back-ends to target new platforms;

\item {} 
we can tweak the translation process to produce low-level code based
on different models and tradeoffs.

\end{itemize}

By contrast, a standardized target environment - say .NET - 
enforces \texttt{m=1} as far as it is concerned.  This helps making \texttt{o} a
bit smaller by providing a higher-level base to build upon.  Still,
we believe that enforcing the use of one common environment 
is not necessary.  PyPy's goal is to give weight to this claim - at least 
as far as language implementation is concerned - showing an approach
to the \texttt{n * m * o} problem that does not rely on standardization.

This is the \emph{meta-goal}; a more concrete goal worth mentioning at this
point is that language specifications can be used to generate cool stuff
in addition to traditional interpreters - e.g. Just-In-Time compilers.
\footnotetext[6]{
\href{http://llvm.cs.uiuc.edu/}{http://llvm.cs.uiuc.edu/}
}

%___________________________________________________________________________

\hypertarget{higher-level-picture}{}
\section{Higher level picture}

As you would expect from a project implemented using ideas from the world
of Extreme Programming\footnotemark[7], the architecture of PyPy has evolved over time
and continues to evolve.  Nevertheless, the high level architecture is now
clear.  There are two independent basic subsystems: the Standard
Interpreter and the Translation Process.
\footnotetext[7]{
\href{http://www.extremeprogramming.com/}{http://www.extremeprogramming.com/}
}

%___________________________________________________________________________

\hypertarget{the-standard-interpreter}{}
\subsection{The Standard Interpreter}

The \emph{standard interpreter} is the subsystem implementing the Python language.
It is divided into two components:
\begin{itemize}
\item {} 
the bytecode interpreter which is responsible for interpreting 
code objects and implementing bytecodes,

\item {} 
the standard object space which implements creating, accessing and
modifying application level objects.

\end{itemize}

Note that the \emph{standard interpreter} can run fine on top of CPython if one
is willing to pay the performance penalty for double-interpretation.

The \emph{bytecode interpreter} is the part that interprets the compact
bytecode format produced from user Python sources by a preprocessing
phase, the \emph{bytecode compiler}.  The bytecode compiler itself is
implemented as a chain of flexible passes (tokenizer, lexer, parser,
abstract syntax tree builder, bytecode generator).  The bytecode
interpreter then does its work by delegating all actual manipulation of
user objects to the \emph{object space}.  The latter can be thought of as the
library of built-in types.  It defines the implementation of the user
objects, like integers and lists, as well as the operations between
them, like addition or truth-value-testing.

This division between bytecode interpreter and object space is very
important, as it gives a lot of flexibility. It is possible to use
different object spaces to get different behaviours of the Python
objects.  Using a special object space is also an important technique
for our translation process.

%___________________________________________________________________________

\hypertarget{the-translation-process}{}
\subsection{The Translation Process}

The \emph{translation process} aims at producing a different (low-level)
representation of our standard interpreter.  The \emph{translation process} 
is done in four steps:
\begin{itemize}
\item {} 
producing a \emph{flow graph} representation of the standard interpreter. 
A combination of the bytecode interpreter and a \emph{flow object space}
performs \emph{abstract interpretation} to record the flow of objects
and execution throughout a Python program into such a \emph{flow graph};

\item {} 
the \emph{annotator} which performs type inference on the flow graph;

\item {} 
the \emph{typer} which, based on the type annotations, turns the flow graph
into one using only low-level operations that fit the model of the
target platform;

\item {} 
the \emph{code generator} which translates the resulting flow graph into
another language, currently C, LLVM, Javascript (experimental).

\end{itemize}

A more complete description of the phases of this process is out of the scope
of the present introduction.  We will only give a short overview in the sequel.
\hypertarget{initialization-time}{}\hypertarget{translation-process-in-more-details}{}

%___________________________________________________________________________

\hypertarget{rpython-the-flow-object-space-and-translation}{}
\section{RPython, the Flow Object Space and translation}

One of PyPy's now achieved objectives is to enable translation of our
bytecode interpreter and standard object space into a lower-level language.
In order for our translation and type inference mechanisms to work
effectively, we need to restrict the dynamism of our interpreter-level
Python code at some point.  However, in the start-up phase, we are
completely free to use all kinds of powerful Python constructs, including
metaclasses and execution of dynamically constructed strings.  However,
when the initialization phase finishes, all code objects involved need to
adhere to a more static subset of Python:
Restricted Python, also known as RPython.

The Flow Object Space then, with the help of our bytecode interpreter,
works through those initialized RPython code objects.  The result of
this abstract interpretation is a flow graph: yet another
representation of a Python program, but one which is suitable for
applying translation and type inference techniques.  The nodes of the
graph are basic blocks consisting of Object Space operations, flowing
of values, and an exitswitch to one, two or multiple links which connect
each basic block to other basic blocks.

The flow graphs are fed as input into the Annotator.  The Annotator,
given entry point types, infers the types of values that flow through
the program variables.  This is the core of the definition of RPython:
RPython code is restricted in such a way that the
Annotator is able to infer consistent types.  How much
dynamism we allow in RPython depends on, and is restricted by, the Flow
Object Space and the Annotator implementation.  The more we can improve
this translation phase, the more dynamism we can allow.  In some cases,
however, it is more feasible and practical to just get rid
of some of the dynamism we use in our interpreter level code.  It is
mainly because of this trade-off situation that the definition of
RPython has shifted over time.  Although the Annotator is
pretty stable now and able to process the whole of PyPy, the RPython
definition will probably continue to shift marginally as we improve it.

The newest piece of this puzzle is the
\emph{Typer}, which inputs the high-level types inferred by the Annotator and
uses them to modify the flow graph in-place to replace its operations with
low-level ones, directly manipulating low-level values and data structures.

The actual low-level code is emitted by ``visiting'' the type-annotated
flow graph after the typer introduced low-level operations.  Currently we have
a C-producing backend, and an LLVM-producing backend.  The former also
accepts non-annotated or partially-annotated graphs, which allow us to
test it on a larger class of programs than what the Annotator can (or
ever will) fully process.
\begin{figure*}[htbp]\begin{center}
\includegraphics{translation-greyscale-small.pdf}
overview of the translation process
\end{center}\end{figure*}
The complete translation process is described in more detail in the
documentation section on the PyPy homepage\footnotemark[8].
\footnotetext[8]{
\href{http://codespeak.net/pypy}{http://codespeak.net/pypy}
}

%___________________________________________________________________________

\hypertarget{status-of-the-implementation-nov-2005}{}
\section{Status of the implementation (Nov 2005)}

With the pypy-0.8.0 release we have integrated our Abstract Syntax
Tree (AST) compiler with the rest of PyPy. The compiler gets
translated with the rest to a static self-contained version of the
standard interpreter.  Like with 0.7.0 this version is very compliant\footnotemark[9] to CPython 2.4.1 but you cannot run many existing programs on it
yet because we are still missing a number of C-modules like socket or
support for process creation.

The self-contained PyPy version (single-threaded and using the
Boehm-Demers-Weiser garbage collector\footnotemark[10]) now runs around 10-20
times slower than CPython, i.e. around 10 times faster than 0.7.0.
This is the result of optimizations, adding short cuts for some common
paths in our interpreter and adding relatively straight forward
optimising transforms to our tool chain, like inlining paired with
simple escape analysis to remove unnecessary heap allocations.  We
still have some way to go. However we expect that most of our speed
will come from the Just-In-Time compiler - work which we have barely
started yet.

With the 0.8.0 release the ``Thunk Object Space'' can also be
translated. This is a module that proxies the Standard Object Space,
adding lazy evaluation features to Python. It is a small scale
show-case for how our whole tool-chain supports flexibility from the
interpreter written in Python to the resulting self-contained
executable.

Our rather complete and Python 2.4-compliant interpreter consists 
of about 30,000-50,000 lines of code (depending on the way you
count code borrowed and adapted from other sources), with
another 14,000 lines of unit tests.  If we include the tools,
the parts related to code analysis and generation, and the
standard library modules ported from C, PyPy is now 138,000
lines of code and 32,000 lines of tests. Refer to 
the statistics web page\footnotemark[11] for more detailed information.
\footnotetext[9]{
\href{http://www.hpl.hp.com/personal/Hans_Boehm/gc/}{http://www.hpl.hp.com/personal/Hans{\_}Boehm/gc/}
}
\footnotetext[10]{
\href{http://codespeak.net/~hpk/pypy-testresult/}{http://codespeak.net/{\textasciitilde}hpk/pypy-testresult/}
}
\footnotetext[11]{
\href{http://codespeak.net/~hpk/pypy-stat/}{http://codespeak.net/{\textasciitilde}hpk/pypy-stat/}
}

%___________________________________________________________________________

\hypertarget{future-work-and-foreseen-possibilities}{}
\section{Future work and foreseen possibilities}

In 2006, the PyPy project aims to translate the standard Python
Interpreter to a JIT-compiler and also to support massive parallelism
(micro-threads) within the language.  These are not trivial tasks
especially if we want to retain and improve the modularity and
flexibility aspects of our implementation - like giving an independent
choice of memory or threading models for translation.  Moreover it is
likely that our Javascript and other higher level backends (in
contrast to our current low-level ones) will continue to evolve.

Apart from optimization-related translation choices PyPy is to enable
new possibilities regarding persistence, security and distribution
issues.  We intend to experiment with ortoghonal persistence for
Python objects, i.e. one that doesn't require application objects to
behave in a particular manner.  Security-wise we will look at
sandboxing or capabilities based schemes.  For distribution we already
experimented with allowing transparent migration of objects between
processes with the help of the existing (and translateable) Thunk
Object Space.  In general, all experiments are much easier to conduct
in PyPy and should provide a resulting standalone executable in
a shorter time than traditional approaches.

\end{document}

### 2006-05-30 talk/dls2006/paper.tex
add an initial latex version
on the upside: it compiles
on the downside: all the figures are in verbatim environments (this looks
entertainingly awful), most things that should be \cite{foo}s are just
left in as [foo]s

\documentclass{acm_proc_article-sp}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}

\begin{document}

\title{Still Missing a Cool Title}

% title ideas: Implementing Virtual Machines in Dynamic Languages?

\maketitle

\begin{abstract}
The PyPy project seeks to prove both on a research and a
practical level the feasibility of writing a virtual machine (VM)
for a dynamic language in a dynamic language -- in this case, Python.
The aim is to translate (i.e. compile) the VM to arbitrary target
environments, ranging in level from C/Posix to Smalltalk/Squeak via
Java and CLI/.NET, while still being of reasonable efficiency within
these environments.

A key tool to achieve this goal is the systematic reuse of the
(unmodified, dynamically typed) Python language as a system
programming language at various levels of our architecture and
translation process.  For each level, we design a corresponding type
system and apply a generic type inference engine -- for example, the
garbage collector is written in a style that manipulates
simulated pointer and address objects, and when translated to C
these operations become C-level pointer and address instructions.
\end{abstract}

\section{Introduction}

Despite the constant trend in the programming world towards
portability and reusability, there are some areas in which it is still
notoriously difficult to write flexible, portable, and reasonably
efficient programs.  The implementation of virtual machines is one
such area.  Building implementations of general programming languages,
in particular highly dynamic ones, using a classic direct coding
approach, is typically a long-winded effort and produces a result that
is quite [quite could be removed here?] tailored to a specific
platform and where architectural decisions (e.g. about GC) are spread
across the code in a pervasive and invasive way.

For this and other reasons, standard platforms emerge; nowadays, a
language implementer could cover most general platforms in use by
writing three versions of his virtual machine: for C/Posix, for Java,
and for CLI/.NET.  This is, at least, the current situation of the
Python programming language, where independent volunteers have
developed and are now maintaining Java and .NET versions of Python,
which follow the evolution of the "official" C version (CPython).

However, we believe that platform standardization does not have to be
a necessary component of this equation.  We are basically using the
standard "meta-programming" argument: if one could write the VM in a
very high level language, then the VM itself could be automatically
\textit{translated} to any lower-level platform.  Moreover by writing
the VM in such a language we would gain in flexibility in
architectural choices and expressiveness.

PyPy achieves this goal without giving up on the efficiency of the
compiled VMs.

The key factors enabling this result are not to be found in recent
advances in any particular research area -- we are not for example using
constraint-based type inference.  Instead, we are following a novel
overall architecture: it is split into many levels of stepwise
translation from the high-level source of the VM to the final target
platform.  Similar platforms can reuse many of these steps, while for
very different platforms we have the option to perform very different
translation steps.  Each step reuses a common type inference component
with a different, ad-hoc type system.

Experiments also suggest a more mundane reason why such an approach is
only practical today: a typical translation takes about half an hour
on a modern PC and consumes between 512MB and 1GB of RAM.

We shortly describe the architecture of PyPy in section
\ref{architecture}.  In section \ref{systemprog} we describe our
approach of varying the type systems at various levels of the
translation.  Section \ref{typeinference} gives an overview of the
type inference engine we developed (and can be read independently from
section 3.)  We present experimental results in section
\ref{experimentalresults} and future work directions in section
\ref{futurework}.  In section \ref{relatedwork} we compare with
related work, and finally we conclude in section \ref{conclusion}.

\section{Architecture}
\label{architecture}

There are two major components in PyPy:

\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environment.
\end{enumerate}

In particular, we have defined a subset of the Python language called
"restricted Python" or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough types to generate efficient code.
The foremost purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

Our current efforts, and the present paper, focus on this tool-suite.
We will not describe the Standard Interpreter component of PyPy in the
sequel, other than mention that it is written in RPython and can thus be
translated.  At close to 90,000 lines of code, it is the largest RPython
program that we have translated so far.  More information can be found
in [S].

\section{System programming with Python}
\label{systemprog}

\hypertarget{the-translation-process}{}
\subsection{The translation process}
\label{translationprocess}

The translation process starts from RPython source code and eventually
produces low-level code suitable for the target environment.  It can be
described as performing a series of step-wise transformations.  Each
step is based on control flow graph transformations and rewriting, and
on the ability to augment the program with further implementation code
written in Python and analysed with the suitable type system.

The front-end part of the translation process analyses the input
RPython program in two phases, as follows\footnote{Note that the two
phases are intermingled in time, because type inference proceeds from
an entry point function and follows all calls, and thus only gradually
discovers (the reachable parts of) the input program.}:

\begin{verbatim}
::

  [figure 0: flow graph and annotator, e.g. part of doc/image/translation.*
             then a stack of transformations
  ]
\end{verbatim}

\begin{enumerate}
\item We take as input RPython functions\footnote{The input to our
       translation chain are indeed loaded runtime function objects,
       not source code or ASTs.  This allows us to use unrestricted
       python for meta-programming purposes at load time, in a
       seemingly staged programming approach, in which the whole of
       the source program -- as Python program -- produces the RPython
       program input to the tool-chain as the object graph loaded in
       memory.  This includes both the relevant functions and prebuilt
       data.}, and convert them to control flow graphs -- a structure
       amenable to analysis.  These flow graphs contain polymorphic
       operations only: in Python, almost all operations are
       dynamically overloaded by type.

\item We perform type inference on the control flow graphs.  At this
      stage, types inferred are part of the type system which is the
      very definition of the RPython sub-language: they are roughly a
      subset of Python's built-in types, with some more precision to
      describe e.g. the items stored in container types.
      Occasionally, a single input function can produce several
      specialized versions, i.e. several similar but differently typed
      graphs.  This type inference process is described in more
      details in section \ref{typeinference}.
\end{enumerate}

At the end of the front-end analysis, the input RPython program is
represented as a forest of flow graphs with typed variables.  Following
this analysis are a number of transformation steps.  Each transformation
step modifies the graphs in-place, by altering their structure and/or
the operations they contain.  Each step inputs graphs typed in one type
system and leaves them typed in a possibly different type system, as we
will describe in the sequel.  Finally, a back-end turns the resulting
graphs into code suitable for the target environment, e.g. C source code
ready to be compiled.

\subsection{Transformations}

When the translation target is C or C-like environments, the first of
the transformation steps takes the RPython-typed flow graphs, still
containing polymorphic operations only, and produces flow graphs with
monomorphic C-like operations and C-like types.  In the simplest case,
this is the only transformation step: these graphs are directly fed to
the C back-end, which turns them into ANSI C source code.

But RPython comes with automatic memory management, and this first
transformation step produces flow graphs that also assume automatic
memory management.  Generating C code directly from there produces a
fully leaking program, unless we link it with an external garbage
collector (GC) like the Boehm conservative GC [Boehm], which is a
viable option.

We have two alternatives, each implemented as a transformation step.
The first one inserts naive reference counting throughout the whole
program's graphs, which without further optimizations gives exceedingly
bad performance (it should be noted that the CPython interpreter is also
based on reference counting, and experience suggests that it was not a
bad choice in this particular case).

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
and replaces all \texttt{malloc} operations with calls to a garbage
collector's innards.  It can inspect all the graphs to discover the
\texttt{struct} types in use by the program, and assign a unique type id to
each of them.  These type ids are collected in internal tables that
describe the layout of the structures, e.g. their sizes and the location
of the pointer fields.

We have implemented other transformations as well, e.g. performing
various optimizations, or turning the whole code into a
continuation-passing style (CPS) [I'm not sure our transformation
can be classified as classical CPS, although there are known similar techniques but the terminology is quite confused] that allows us to use coroutines
without giving up the ability to generate fully ANSI C code.  (This will
be the subject of another paper.)  [mention exception transformer too]

Finally, currently under development is a variant of the very first
transformation step, for use when targeting higher-level,
object-oriented (OO) environments.  It is currently being designed
together with back-ends for Smalltalk/Squeak\footnote{Our simple OO
type system is designed for \textit{statically-typed} OO environments,
including Java; the presence of Smalltalk as a back-end might be
misleading in that respect.} and CLI/.NET.  This first transformation
step, for C-like environments, is called the \textit{LLTyper}: it produces
C-level flow graphs, where the object-oriented features of RPython
(classes and instances) become manipulations of C structs with
explicit virtual table pointers.  By contrast, for OO environments the
transformation step is called the \textit{OOTyper}: it targets a simple
object-oriented type system, and preserves the classes and instances
of the original RPython program.  The LLTyper and OOTyper still have
much code in common, to convert the more Python-specific features like
its complex calling conventions.

More information about these transformations can be found in [T].

\subsection{System code}

A common pattern in all the transformation steps is to somehow lower the
level at which the graphs are currently expressed.  Because of this,
there are operations that were atomic in the input (higher-level) graphs
but that need to be decomposed into several operations in the target
(lower-level) graphs.  In some cases, the equivalent functionality
requires more than a couple of operations: a single operation must be
replaced by a call to whole new code -- functions and classes that serve
as helpers.  An example of this is the \texttt{malloc} operation for the GC
transformer.  Another example is the \texttt{list.append()} method, which is
atomic for Python or RPython programs, but needs to be replaced in
C-level code by a helper that possibly reallocates the array of items.

This means that in addition to transforming the existing graphs, each
transformation step also needs to insert new functions into the forest.
A key feature of our approach is that we can write such "system-level"
code -- relevant only to a particular transformation -- in plain Python
as well:

\begin{verbatim}
.. topic:: Figure 1 - a helper to implement \texttt{list.append()}

  ::

    def ll_append(lst, newitem):
        # Append an item to the end of the vector.
        index = lst.length         # get the 'length' field
        ll_resize(lst, index+1)    # call a helper not shown here
        itemsarray = lst.items     # get the 'items' field
        itemsarray[index] = item   # this behaves like a C array
\end{verbatim}

The idea is to feed these new Python functions into the front-end, using
this time the transformation's target (lower-level) type system during
the type inference.  In other words, we can write plain Python code that
manipulates objects that conform to the lower-level type system, and
have these functions automatically transformed into appropriately typed
graphs.

For example, \texttt{ll\textunderscore{}append()} in figure 1 is a Python function
that manipulates objects that behave like C structures and arrays.
This function is inserted by the LLTyper, as a helper to implement the
\texttt{list.append()} calls found in its RPython-level input graphs.
By going through the front-end reconfigured to use C-level types, the
above function becomes a graph with such C-level types\footnote{The
low-level type system specifies that the function should be
specialized by the C-level type of its input arguments, so it actually
turns into one graph per list type -- list of integers, list of
pointers, etc.  This behavior gives the programmer a feeling
comparable to C++ templates, without the declarations.}, which is then
indistinguishable from the other graphs of the forest produced by the
LLTyper.

In the example of the \texttt{malloc} operation, replaced by a call to GC
code, this GC code can invoke a complete collection of dead objects, and
can thus be arbitrarily complicated.  Still, our GC code is entirely
written in plain Python, and it manipulates "objects" that are still at
a lower level: pointer and address objects.  Even with the restriction
of having to use pointer-like and address-like objects, Python remains
more expressive than, say, C to write a GC.  [see also Jikes]

In the sequel, we will call \textit{system code} functions written in
Python that are meant to be analysed by the front-end.  For the
purpose of this article we will restrict this definition to helpers
introduced by transformations, as opposed to the original RPython
program, although the difference is not fundamental to the translation
process (and although our input RPython program, as seen in section
\ref{architecture}, is often itself a Python virtual machine!).

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared to other approaches [e.g. Squeak], we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}.

\subsection{Type systems}

The four levels that we considered so far are summarized in figure 2.

\begin{verbatim}
::

    [figure 2:    RPython
                  /     \
                 /       \
       LLTypeSystem     OOTypeSystem
               /
              /
      Raw addresses
    ]
\end{verbatim}

The RPython level is a subset of Python, so the types mostly follow
Python types, and the instances of these types are instances in the
normal Python sense; e.g. whereas Python has only got a single type
\texttt{list}, RPython has a parametric type \texttt{list(T)} for every RPython
type \texttt{T}, but instances of \texttt{list(T)} are just those Python lists
whose items are all instances of \texttt{T}.

The other type systems, however, do not correspond to built-in Python
types.  For each of them, we implemented:

\begin{enumerate}
\item the types, which we use to tag the variables of the graphs at
      the given level.  (Types are actually annotated, self-recursive
      formal terms, and would have been implemented simply as such if
      Python supported them directly.)

\item the Python objects that emulate instances of these types.  (More
      about them below.)
\end{enumerate}

We have defined well-typed operations between variables of these types,
plugging on the standard Python operators.  These operations are the
ones that the emulating instances implement.  As seen above, the types
can also be used by type inference when analysing system code like the
helpers of figure 1.

Now, clearly, the purpose of types like a "C-like struct" or a "C-like
array" is to be translated to a real \texttt{struct} or array declaration by
the C back-end.  What, then, is the purpose of emulating such things in
Python?  The answer is three-fold.  Firstly, if we have objects that
live within the Python interpreter, but faithfully emulate the behavior
of their C equivalent while performing additional safety checks, they
are an invaluable help for testing and debugging.  For example, we can
check the correctness of our hash table implementation, written in
Python in term of struct- and array-like objects, just by running it.
The same holds for the GC.

Secondly, and anecdotically, as the type inference process (section
\ref{typeinference}) is based on abstract interpretation, we can use
the following trick: the resulting type of most low-level operations
is deduced simply by example.  Sample C-level objects are
instantiated, used as arguments to a given operation, and produce a
sample result, whose C-level type must be the type of the result
variable in the graph.

The third reason is fundamental: we use these emulating objects to
\textit{represent} pre-built objects at that level.  For example, the GC
transformer instantiates the objects emulating C arrays for the internal
type id tables, and it fills them with the correct values.  These array
objects are then either used directly when testing the GC, or translated
by the C back-end into static pre-initialized arrays.

\section{Type inference}
\label{typeinference}

The various analyses used -- from type inference to lifetime analysis -
are generally formulated as [abstract interpretation].  While this
approach is known to be less efficient than more tailored algorithms
like constraint-based type inference, we gain in freedom,
controllability and simplicity.  This proved essential in our overall
approach: as described in section \ref{systemprog}, we need to perform
type inference with many different type systems, the details of which
have evolved along the road.

We mitigate the potential efficiency problem by wise choices and
compromises for the domain used; the foremost example of this is that
our RPython type inference performs almost no automatic specialization
of functions XXX class annotation vs concrete runtime class setsXXX.
We achieved enough precision for our purpose, though.

In the sequel, we give a more precise description of this process and
justify our claim that good performance and enough precision can be
achieved -- at least in some contexts -- without giving up the naive but
flexible approach.

\subsection{Building control flow graphs}
\label{flowobjspace}
\hypertarget{flowobjspace}{}

As described in the overview of \href{\#the-translation-process}{the
translation process}, the front-end of the translation tool-chain
works in two phases: it first builds control flow graphs from Python
functions, and then performs whole-program type inference on these
graphs.

Remember that building the control flow graphs is not done, as one might
first expect, by following a function at the syntactic level.  Instead,
the whole program is imported in a normal Python interpreter; the full
Python language is used at this point as a kind of preprocessor with
meta-programming capabilities.  Once the program is imported, the object
data in memory consists of Python function objects in bytecode format,
and any other kind of objects created at import-time, like class
objects, prebuilt instances of those, prebuilt tables, and so on.  Note
that these objects have typically no text representation any more; for
example, cyclic data structures may have been built at this point.  The
translation tool-chain first turns these function objects into in-memory
control flow graphs which contain direct references to the prebuilt data
objects, and then handles and transforms these graphs.

We found in-process debugging sufficient and did not implement dumping
of any intermediate step to disk.  Figure 3 shows the control flow graph
obtained for a simple function -- this is a screenshot from our graph
viewer, used for debugging; basic block placement is performed by
[Graphviz].

\begin{verbatim}
::

    [figure 3: insert a nice pygame screenshot]
\end{verbatim}

The actual transformation from function objects -- i.e. bytecode -- to
flow graph is performed by the Flow Object Space, a short but generic
plug-in component for the Python interpreter of PyPy.  The architecture
of our Python interpreter is shown in figure 4.

\begin{verbatim}
.. topic:: Figure 4 - the interpreter and object spaces

    +------------------------------------------------------+
    |   forest of bytecode objects from the application    |
    +------------------------------------------------------+
    |               Python bytecode interpreter            |
    +--------------------------------+---------------------+
    |      Standard Object Space     |  Flow Object Space  |
    +--------------------------------+---------------------+
\end{verbatim}

Note that the left column, i.e. the bytecode interpreter and the
Standard Object Space, form the full Python interpreter of PyPy.  It is
an RPython program, and the whole purpose of the translation process is
to accept this as \textit{input}, and translate it to an efficient form.  Its
architecture is not relevant to the way it is translated. XXX?

However, the bytecode interpreter plays a double role, at two different
levels.  The so-called Object Spaces are \textit{domains} in the abstract
interpretation terminology.  By design, we cleanly separated these
domains from the bytecode interpreter core; the latter is only
responsible for decoding the bytecodes of an application and emulating
the corresponding stack machine.  It treats all actual application-level
objects as black boxes, and dispatches all operations on them to the
Object Space.  The Standard Object Space is a concrete domain, in which
objects are the concrete Python objects of the various built-in types:
lists, dictionaries, and so on.  By opposition, the Flow Object Space is
really an abstract domain.  It handles objects that are placeholders.
Its lattice order is shown in figure 5.

\begin{verbatim}
::

    [figure 5:                Variable

                         /      |             \           \
                        /       |              \           \
                       /        |               \           \
      Constant(1) ... Constant(n) ... Constant([1,2,3])  ... Constant(<instance of class A>) ...
    ]
\end{verbatim}

This order is extremely simple, because most actual analysis is delayed
to the next phase, the type inference engine.  The objects are either
\textit{Variables}, which are pure placeholders for entierely unknown values,
or \textit{Constants} with a concrete Python object as value.  The order places
Variable as the top, and keeps all \textit{Constants} unordered.  Thus if two
different constants merge during abstract interpretation, we immediately
widen them to Variable.

In conjunction with the Flow Object Space, the bytecode interpreter of
PyPy thus performs abstract interpretation of Python bytecodes from
the application\footnote{Note that this process uses the
\textit{unmodified} bytecode interpreter.  This means that it is
independent of most language details.  Changes in syntax or in
bytecode format or opcode semantics only need to be implemented once,
in the bytecode interpreter.  In effect, the Flow Object Space enables
an interpreter for \textit{any} language to work as a front-end for
the rest of the tool-chain.}.  In this case, the bytecodes in
question come from the RPython application that we would like to
translate.

The Flow Object Space records all operations that the bytecode
interpreter "would like" to do between the placeholder objects.  It
records them into basic block objects that will eventually be part of
the control flow graph of the whole function.  The recorded operations
take Variables and Constants as argument, and produce new Variables as
results.  The Constants serve two purposes: they are a way to
introduce constant values into the flow graphs -- these values may be
arbitrarily complex objects, not just primitives -- and they allow
basic constant propagation\footnote{This is useful at this level for
some constructs of the bytecode interpreter, which can temporarily
wrap internal values and push them onto the regular value stack among
the other application-level objects.  We need to be able to unwrap
them again later. XXX compile-time computation in the helpers.}

In the flow graph, branching occurs when the bytecode interpreter tries
to inspect the truth value of placeholder objects, as it would in
response to conditional jump opcodes or other more complicated opcodes:
at this point, the Flow Object Space starts two new basic blocks and -
with a technique akin to continuations -- tricks the interpreter into
following both branches, one after the other.  Additionally, the
bytecode interpreter sends simple positional signals that allow the Flow
Object Space to detect when control paths merge, or when loops close.
In this way, abstract interpretation quickly terminates and the recorded
operations form a graph, which is the control flow graph of the original
bytecode.

Note that we produce flow graphs in Static Single Information (SSI)
form, an extension of Static Single Assignment ([SSA]): each variable is
only used in exactly one basic block.  All variables that are not dead
at the end of a basic block are explicitly carried over to the next
block and renamed.

While the Flow Object Space is quite a short piece of code -- its core
functionality holds in 300 lines -- the detail of the interactions
sketched above is not entierely straightforward; we refer the reader to
[D] for more information.

\subsection{The Annotator}

The type inference engine, which we call the \textit{annotator}, is the central
component of the front-end part of the translation process.  Given a
program considered as a family of control flow graphs, the annotator
assigns to each variable of each graph a so-called \textit{annotation}, which
describes the possible run-time objects that this variable can contain.
Following usual terminology, we will call such annotations \textit{types} -- not
to be confused with the Python notion of the concrete type of an object.
An annotation is a set of possible values, and such a set is not always
the set of all objects of a specific Python type.

Here is a simplified, static model of how the annotator works.  It can
be considered as taking as input a finite family of functions calling
each other, and working on the control flow graphs of each of these
functions as built by the \href{flowobjspace}{Flow Object Space}.
Additionally, for a particular "entry point" function, the annotator
is provided with user-specified types for the function's arguments.

The goal of the annotator is to find the most precise type that can be
given to each variable of all control flow graphs while respecting the
constraints imposed by the operations in which these variables are
involved.

More precisely, it is usually possible to deduce information about the
result variable of an operation given information about its arguments.
For example, we can say that the addition of two integers must be an
integer.  Most programming languages have this property.  However,
Python -- like many languages not specifically designed with type
inference in mind -- does not possess a type system that allows much
useful information to be derived about variables based on how they are
\textit{used}; only on how they were \textit{produced}.  For example, a number of very
different built-in types can be involved in an addition; the meaning of
the addition and the type of the result depends on the type of the input
arguments.  Merely knowing that a variable will be used in an addition
does not give much information per se.  For this reason, our annotator
works by flowing types forward, operation after operation, i.e. by
performing abstract interpretation of the flow graphs.  In a sense, it
is a more naive approach than the one taken by type systems specifically
designed to enable more advanced inference algorithms.  For example,
Hindley-Milner type inference works in an inside-out direction, by
starting from individual operations and propagating type constraints
outwards [H-M].

Naturally, simply propagating types forward requires the use of a fixed
point algorithm in the presence of loops in the flow graphs or in the
inter-procedural call graph.  Indeed, we flow types forward from the
beginning of the entry point function into each basic block, operation
after operation, and follow all calls recursively.  During this process,
each variable along the way gets a type.  In various cases, e.g. when we
close a loop, the previously assigned types can be found to be too
restrictive.  In this case, we generalise them to allow for a larger set
of possible run-time values, and schedule the block where they appear
for reflowing.  The more general type can generalise the types of the
results of the variables in the block, which in turn can generalise the
types that flow into the following blocks, and so on.  This process
continues until a fixed point is reached.

We can consider that all variables are initially assigned the "bottom"
type corresponding to the empty set of possible run-time values.  Types
can only ever be generalised, and the model is simple enough to show
that there is no infinite chain of generalisation, so that this process
necessarily terminates.

\subsection{RPython types}

As seen in section \ref{systemprog}, we use the annotator with more than one type
systems.  The more interesting and complex one is the RPython type
system, which describes how the input RPython program can be annotated.
The other type systems contain lower-level, C-like types that are mostly
unordered, thus forming more trivial lattices than the one formed by
RPython types.

The set $A$ of RPython types is defined as the following formal terms:

\begin{itemize}
\item $Bot$, $Top$ -- the minimum and maximum elements (corresponding
      to "impossible value" and "most general value");

\item $Int$, $NonNegInt$, $Bool$ -- integers, known-non-negative
      integers, booleans;

\item $Str$, $Char$ -- strings, characters (which are strings of
      length 1);

\item $Inst(class)$ -- instance of $class$ or a subclass thereof
      (there is one such term per $class$);

\item $List(v)$ -- list; $v$ is a variable summarising the items of
      the list (there is one such term per variable);

\item $Pbc(set)$ -- where the $set$ is a subset of the (finite) set of
      all prebuilt constant objects.  This set includes all the
      callables of the input program: functions, classes, and methods.

\item $None$ -- stands for the singleton \texttt{None} object of
      Python.

\item $NullableStr$, $NullableInst(class)$ -- a string or
      \texttt{None}; resp. an instance or \texttt{None}.
\end{itemize}

Figures 6 and 7 shows how these types are ordered to form a lattice.  We
mostly use its structure of [join-semilattice] only.

\begin{verbatim}
.. graphviz:: image/lattice1.dot

:Figure 6: the lattice of annotations.
\end{verbatim}

\begin{verbatim}
.. graphviz:: image/lattice2.dot

:Figure 7: The part about instances and nullable instances, assuming a
           simple class hierarchy with only two direct subclasses of
           \texttt{object}.
\end{verbatim}

All list terms for all variables are unordered.  The Pbcs form a
classical finite set-of-subsets lattice.  In addition, we have left
out a number of other annotations that are irrelevant for the basic
description of the annotator and straightforward to handle:
$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
complete list is described in [T].

The type system moreover comes with a family of rules, which for every
operation and every sensible combination of input types describes the
type of its result variable.  Let $V$ be the set of Variables that
appear in the user program's flow graphs.  Let $b$ be a map from $V$
to $A$; it is a "binding" that gives to each variable a type.  The
purpose of the annotator is to compute such a binding stepwise.

Let $x$, $y$ and $z$ be Variables.  We introduce the rule:

% XXX format this!
$$
\begin{array}{c}
 z = \mathrm{add}(x, y), \; b(x) =
    Int, \; Bool \leq b(y) \leq Int \\ \hline
 b' = b \hbox{\ with\ } (z
    \rightarrow Int) 
\end{array}
$$

which specify that if we see the addition operation applied to
Variables whose current binding is $Int$, a new binding $b'$ can be
produced: it is $b$ except on $z$, where we have $b'(z) = Int$.

The type inference engine can be seen as applying this kind of rules
repeatedly.  It does not apply them in random order, but follows a
forward-propagation order typical of abstract interpretation.

It is outside the scope of the present paper to describe the type
inference engine and the rules more formally.  The difficult points
involve mutable containers -- e.g. initially empty list that are filled
somewhere else -- and the discovery of instance attributes -- in Python,
classes do not declare upfront a fixed set of attributes for their
instances, let alone their types.  Both these examples require
sophisticated reflowing techniques that invalidate existing types in
already-annotated basic blocks, to account for the influence of more
general types coming indirectly from a possibly distant part of the
program.  The reader is referred to [D] for more information.

\subsection{Termination and complexity}

The lattice model clearly contains no infinite chain.  Moreover, it is
designed to convince oneself that the number of reflowings required in
practice is small.  For example, we are not trying to do range analysis
beyond detecting non-negatives -- the reason is that range analysis is
not an essential part of writing \textit{reasonably} efficient code.  Consider
that a translated RPython program runs hundreds of times faster than
when the same program is executed by the standard Python interpreter: in
this context, range analysis appears less critical.  It is a possible
optimization that we can introduce in a later, optional analysis and
transformation step.

The worst case behaviors that can appear in the model described above
involve the lattice of Pbcs, involving variables that could contain
e.g.  one function object among many.  An example of such behavior is
code manipulating a table of function objects: when an item is read
out of the table, its type is a large Pbc set: $Pbc(\{f1, f2, f3,
\ldots\})$.  But in this example, the whole set is available at once,
and not built incrementally by successive discoveries.  This seems to
be often the case in practice: it is not very common for programs to
manipulate objects that belong to a large but finite family -- and when
they do, the whole family tends to be available early on, requiring
few reflowing.

This means that \textit{in practice} the complete process requires a time that
is far lower than the worst case.  We have experimentally confirmed
this: annotating the whole PyPy interpreter (90,000 lines) takes on the
order of 5 to 10 minutes, and basic blocks are typically only reflown a
handful of times, providing a close-to-linear practical complexity.

We give formal termination and correctness proofs in [D], as well as
worst-case bounds and some experimental evidence of their practical
irrelevance.

\subsection{Precision}

Of course, this would be pointless if the annotation did not give
precise enough information for our needs.  We must describe a detail of
the abstract interpretation engine that is critical for precision: the
propagation of conditional types.  Consider the following source code
fragment::

    if isinstance(x, MyClass):
        f(x)
    else:
        g(x)

Although the type of \texttt{x} may be some parent class of
\texttt{MyClass}, it can be deduced to be of the more precise type
$Inst(MyClass)$ within the positive branch of the \texttt{if}.
(Remember that our graphs are in SSI form, which means that the
\texttt{x} inside each basic block is a different Variable with a
possibly different type as annotation.) XXX flow sensivity and SSA

This is implemented by introducing an extended family of types for
boolean values:

$$
Bool(v_1: (t_1, f_1), v_2: (t_2, f_2), ...)
$$

where the $v_n$ are variables and $t_n$ and $f_n$ are types.  The
result of a check, like \texttt{isintance()} above, is typically
annotated with such an extended $Bool$.  The meaning of the type is as
follows: if the run-time value of the boolean is True, then we know
that each variable $v_n$ has a type at most as general as $t_n$; and
if the boolean is False, then each variable $v_n$ has a type at most
as general as $f_n$.  This information is propagated from the check
operation to the exit of the block via such an extended $Bool$ type,
and the conditional exit logic in the type inference engine uses it to
trim the types it propagates into the next blocks (this is where the
\textit{meet} of the lattice is used).

With the help of the above technique, we achieve a reasonable precision
in small examples.  For larger examples, a different, non-local
technique is required: the specialization of functions. XXX use 'polymorphism'

As described in the introduction, the most important downside of our
approach is that automatic specialization is a potential
performance-killer.  We \textit{do} support specialization, however: we can
generate several independently-annotated copies of the flow graphs of
certain functions.  When annotating RPython programs, such
specialization does not happen automatically: we rely on hints provided
by the programmer in the source code, in the form of flags attached to
function objects.  As we had this trade-off in mind when we wrote the
Python interpreter of PyPy, we only had to add a dozen or so hints in
the end.

This does not mean that automatic specialization policies are difficult
to implement.  Indeed, the simpler lower-level type systems rely quite
heavily on them: this is because the system code helpers are often
generic and can receive arguments of various C-level types.  In this
case, because the types at this level are limited and mostly unordered,
specializing all functions on their input argument's types works well.

At the level of RPython, on the other hand, the range of specializations
that make sense is much wider.  We have used anything between
specialization by the type of an argument, to specialization by an
expected-to-be-constant argument value, to memoized functions that the
type inference engine will actually call during annotation and replace
by look-up tables, to complete overriding of the annotator's behavior in
extreme cases.  In this sense, the need for manual specialization turned
into an advantage, in term of simplicity and flexibility of implementing
and using new specialization schemes.

This conclusion can be generalized.  We experimented with a simple
approach to type inference that works well in practice, and that can
very flexibly accomodate changes in the type system and even completely
different type systems.  We think that the reasons for this success are
to be found on the one hand in the (reasonable) restrictions we put on
ourselves when designing the RPython language and writing the Python
interpreter of PyPy in RPython, and on the other hand in an ad-hoc type
system that is designed to produce enough precision (but not more) for
the purpose of the subsequent transformations to C-level code.

We should mention that restricting oneself to write RPython code instead
of Python is still a burden, and that we are not proposing in any way
that the Python language itself should evolve in this direction, nor
even that RPython usage should become widespread.  It is a tool designed
with a specific goal in mind, which is the ability to produce reasonably
efficient, stand-alone code in a large variety of environment.

\section{Experimental results}
\label{experimentalresults}

\subsection{Performance}

Our tool-chain is capable of translating the Python interpreter of
PyPy, written in RPython, producing right now either ANSI C code as
described before, or LLVM\footnote{the LLVM project is the realisation
of a portable assembler infrastructure, offering both a virtual
machine with JIT capabilities and static compilation. Currently we are
using the latter with its good high-level optimisations for PyPy.}
assembler which is then natively compiled with LLVM tools.

The tool-chain has been tested with and can sucessfully apply
transformations enabling various combinations of features. The
translated interpreters are benchmarked using pystone (a [Dhrystone]
derivative traditionally used by the Python community, although it is a
rather poor benchmark) and the classical [Richards] benchmark and
compared against [CPython] 2.4.3 results:

\begin{verbatim}
+------------------------------------+-------------------+-------------------+
|  Interpreter                       | Richards,         | Pystone,          |
|                                    | Time/iteration    | Iterations/second |
+====================================+===================+===================+
|  CPython 2.4.3                     |   789ms    (1.0x) |   40322    (1.0x) |
+------------------------------------+-------------------+-------------------+
|  pypy-c                            |  4269ms    (5.4x) |    7587    (5.3x) |
+------------------------------------+-------------------+-------------------+
|  pypy-c-thread                     |  4552ms    (5.8x) |    7122    (5.7x) |
+------------------------------------+-------------------+-------------------+
|  pypy-c-stackless                  |  XXX       (6.0x) |    XXX     (6.2x) |
+------------------------------------+-------------------+-------------------+
|  pypy-c-gcframework                |  6327ms    (8.0x) |    4960    (8.1x) |
+------------------------------------+-------------------+-------------------+
|  pypy-c-stackless-gcframework      |  XXX       (    ) |    XXX     (    ) |
+------------------------------------+-------------------+-------------------+
|  pypy-llvm-c                       |  3797ms    (4.8x) |    7763    (5.2x) |
+------------------------------------+-------------------+-------------------+
|  pypy-llvm-c-prof                  |  2772ms    (3.5x) |   10245    (3.9x) |
+------------------------------------+-------------------+-------------------+
\end{verbatim}

The numbers in parenthesis are slow-down factors compared to CPython.
These measures reflect PyPy revision 27815, compiled with GCC 3.4.4.
LLVM is version 1.8cvs (May 11, 2006).  The machine runs GNU/Linux SMP
on an Intel(R) Pentium(R) 4 CPU at 3.20GHz with 2GB of RAM and 1MB of
cache.  The rows correspond to variants of the translation process, as
follows:

pypy-c
    The simplest variant: translated to C code with no explicit memory
    management, and linked with the Boehm conservative GC [Boehm].

pypy-c-thread
    The same, with OS thread support enabled.  (For measurement purposes,
    thread support is kept separate because it has an impact on the GC
    performance.)

pypy-c-stackless
    The same as pypy-c, plus the "stackless transformation" step which
    modifies the flow graph of all functions in a way that allows them
    to save and restore their local state, as a way to enable coroutines.

pypy-c-gcframework
    In this variant, the "gc transformation" step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

pypy-c-stackless-gcframework
    This variant combines the "gc transformation" step with the
    "stackless transformation" step.  The overhead introduced by the
    stackless feature is balanced with the removal of the overhead of
    pushing and popping roots explicitly on an alternate stack: indeed,
    in this variant it is possible to ask the functions in the current C
    call chain to save their local state and return.  This has the
    side-effect of moving all roots to the heap, where the GC can find
    them.

pypy-llvm-c
    The same as pypy-c, but using the LLVM back-end instead of the C
    back-end.  The LLVM assembler-compiler gives the best results when -
    as we do here -- it optimizes its input and generates again C code,
    which is fed to GCC.

pypy-llvm-c-prof
    The same as pypy-llvm-c, but using GCC's profile-driven
    optimizations.

The speed difference with CPython 2.4.3 can be explained at two levels.
One is that CPython is hand-crafted C code that has been continuously
optimized for a decade now, whereas the Python interpreter of PyPy first
seeks flexibility and high abstraction levels.  The other, probably
dominant, factor is that various indices show that our approach places a
very high load on the GC and on the memory caches of the machine.  The
Boehm GC is known to be less efficient than more customized approach;
kernel-level profiling shows that pypy-c typically spends 30% of its
time in the Boehm library.  Our current, naively simple mark-and-sweep
GC is even quite worse.  The interaction with processor caches is also
hard to predict and account for; in general, we tend to produce
relatively large amounts of code and prebuilt data.

\subsection{Translation times}

A complete translation of the pypy-c variant takes about 39 minutes,
divided as follows:

\begin{verbatim}
+-------------------------------------------+------------------------------+
| Step                                      |   Time (minutes:seconds)     |
+===========================================+==============================+
| Front-end                                 |            9:01              |
| (flow graphs and type inference)          |                              |
+-------------------------------------------+------------------------------+
| LLTyper                                   |           10:38              |
| (from RPython-level to C-level graphs     |                              |
| and data)                                 |                              |
+-------------------------------------------+------------------------------+
| Various low-level optimizations           |            6:51              |
| (convert some heap allocations to local   |                              |
| variables, inlining, ...)                 |                              |
+-------------------------------------------+------------------------------+
| Database building                         |            8:39              |
| (this initial back-end step follows all   |                              |
| graphs and prebuilt data structures       |                              |
| recursively, assigns names, and orders    |                              |
| them suitably for code generation)        |                              |
+-------------------------------------------+------------------------------+
| Generating C source                       |            2:25              |
+-------------------------------------------+------------------------------+
| Compiling (\texttt{gcc -O2})              |            3:23              |
+-------------------------------------------+------------------------------+
\end{verbatim}

An interesting feature of this table is that type inference is not the
bottleneck.  Indeed, further transformation steps typically take longer
than type inference alone.  This is the case for the LLTyper step,
although it has a linear complexity on the size of its input (most
transformations do).

Other transformations like the "gc" and the "stackless" ones actually
take more time, particuarly when used in combination with each other (we
speculate it is because of the increase in size caused by the previous
transformations).  A translation of pypy-c-stackless, without counting
GCC time, takes 60 minutes; the same for pypy-c-stackless-gcframework
takes XXX minutes.

\section{Future work}
\label{futurework}

As described in section \ref{experimentalresults}, the performance of
the compiled Python interpreters is still not up to competing with the
well-established CPython.  We are always working to improve matters,
considering new optimizations and better GCs.  Also, the OOTyper and
back-ends for Smalltalk/Squeak and CLI/.NET are currently in progress.

\subsection{JIT Specializer}

So far, the PyPy tool-chain can only translate the Python interpreter of
PyPy into a program which is again an interpreter -- the same interpreter
translated to C, essentially, although we have already shown that some
aspects can be "weaved" in at translation time, like support for
coroutines.

To achieve high performance for dynamic languages such as Python, the
proven approach is to use dynamic compilation techniques, i.e. to write
JITs.  With direct techniques, this is however a major endeavour, and
increases the efforts to further evolve the language.

In the context of the PyPy project, we are now exploring -- as we planned
from the start -- the possibility to produce a JIT as a graph
transformation aspect from the Python interpreter.  This idea is based
on the theoretical possibiliy to turn interpreters into compilers by
partial evaluation [PE].  In our approach, this is done by analysing
the forest of flow graphs built from the Python interpreter, which is a
well-suited input for this kind of techniques.  We can currently perform
binding-time analysis [BTA] on these graphs, again with abstract
interpretation techniques reusing the type inference engine.  The next
step is to transform the graphs -- following the binding-time annotations
- into a compiler; more precisely, in partial evalution terminology, a
generating extension.  We can currently do this on trivial examples.

The resulting generating extension will be essentially similar to
[Psyco], which is the only (and hand-written) JIT available for Python so
far, based on run-time specialization.

\section{Related work}
\label{relatedwork}

Applying the expressiveness or at least syntax of very high-level and
dynamically typed languages to their implementation has been
investigated many times.

One typical approach is writing a static compiler.  The viability of
and effort required for such an approach depend usually on the binding
and dispatch semantics of the language.  Common Lisp native compilers,
usable interactively and taking functions or files as compilation
units are a well-known example of that approach.  Late binding for all
names and load semantics make such an approach very hard for Python,
if speed improvements are desired.

It is more relevant to consider and compare with projects using
dynamic and very-high level languages for interpreters and VM
implementations, and Just-In-Time compilers.

[Scheme48] was a Scheme implementation using a restricted Scheme, with
static type inference based on Hindley-Milner, this is viable for
Scheme as base language. Portability and simplicity were its major
goals.

[Squeak] is a Smalltalk implementation in Smalltalk. It uses SLang, a
very restricted subset of Smalltalk with few types and strict
conventions such that mostly direct translation to C is possible.
Both the VM and the object memory and garbage collector support are
explicitly written together in this style. Again simplicity and
portability were the major goals, not sophisticated manipulation and
analysis and waving in of features.

[Jikes RVM] is a Java VM and Just-In-Time compiler written in Java.
Bootstrapping happens by self-applying the compiler on a host VM, and
dumping a snapshot from memory of the resulting native code.

This approach enables directly high performance, at the price of
portability as usual with pure native code emitting
approaches. Modularity of features, when possible, is achieved with
normal software modularity. The indirection costs are taken care by
the inlining done by the compiler, sometimes even through explicit
ways to request for it. In particular this modular approach is used
for implementing a range of choices for GC support XXX ref.  This was
the inspiration for PyPy own GC framework, although much more tuning
and work went into Jikes RVM. PyPy own GC framework also exploits
inlining of helpers and barriers to recover performance.

Jikes RVM native JIT compilers can likely not easily be retargeted to
run on top and target another VM (for example a CLR runtime) instead
of hardware processors. Also Jikes RVM pays the complexity of writing
a JIT up-front, which also means that features and semantics of the
language are encoded in the JIT compiler code, meaning likely that
major changes would correspond to major surgery needed on it.

PyPy more indirect approach, together hopefully with our future work
on generating a JIT compiler, tries to overcome these limitations, at
the price of some more effort required to achieve very good
performance. It is too soon to compare completely the complexity (and
performance) trade-offs of these approaches.

XXX Jython, IronPython, UVM.

\section{Conclusion}
\label{conclusion}

XXX

nice interpreter not polluted by implementation details
here Python, but any interpreter works

architecture allows implementing features at the right level

dynamic language enables defining our own various type systems
[ref pluggable type systems]

practical VMs will result with a bit more efforts

\end{document}

### 2006-05-30 talk/dls2006/paper.tex
Manual merge of r27925 to r27930 of draft.txt.  At this point the two documents are in sync and only the .tex should be updated.

@@
 
-Applying the expressiveness or at least syntax of very high-level and
-dynamically typed languages to their implementation has been
-investigated many times.
-
-One typical approach is writing a static compiler.  The viability of
-and effort required for such an approach depend usually on the binding
+Applying the expressiveness - or at least the syntax - of very
+high-level and dynamically typed languages to their implementation has
+been investigated many times.
+
+One typical approach is writing a static compiler.  The viability of,
+and effort required for, such an approach depends usually on the binding
 and dispatch semantics of the language.  Common Lisp native compilers,
-usable interactively and taking functions or files as compilation
-units are a well-known example of that approach.  Late binding for all
-names and load semantics make such an approach very hard for Python,
-if speed improvements are desired.
-
-It is more relevant to consider and compare with projects using
-dynamic and very-high level languages for interpreters and VM
-implementations, and Just-In-Time compilers.
-
-[Scheme48] was a Scheme implementation using a restricted Scheme, with
-static type inference based on Hindley-Milner, this is viable for
-Scheme as base language. Portability and simplicity were its major
-goals.
-
-[Squeak] is a Smalltalk implementation in Smalltalk. It uses SLang, a
+usable interactively and taking single functions or files as compilation
+units, are a well-known example of that approach.  However, the late
+binding for all names and the load semantics make such an approach very
+hard for Python, if speed improvements are desired.
+
+In this context, it is more relevant to consider and compare with
+projects using dynamic and very-high-level languages for interpreters
+and VM implementations, and Just-In-Time compilers.
+
+Scheme48 was a Scheme implementation using a restricted Scheme, PreScheme
+[PreScheme], with static type inference based on Hindley-Milner.  This is
+viable for Scheme as base language.  Simplicity and portability across C
+platforms were its major goals.
+
+Squeak [Squeak] is a Smalltalk implementation in Smalltalk. It uses SLang, a
 very restricted subset of Smalltalk with few types and strict
-conventions such that mostly direct translation to C is possible.
-Both the VM and the object memory and garbage collector support are
+conventions, which can be mostly directly translated to C.
+The VM, the object memory and the garbage collector support are
 explicitly written together in this style. Again simplicity and
-portability were the major goals, not sophisticated manipulation and
-analysis and waving in of features.
-
-[Jikes RVM] is a Java VM and Just-In-Time compiler written in Java.
+portability were the major goals, as opposed to sophisticated manipulation
+and analysis or "weaving" in of features as transformation aspects.
+
+Jikes RVM [DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00] is a
+Java VM and Just-In-Time compiler written in Java.
 Bootstrapping happens by self-applying the compiler on a host VM, and
 dumping a snapshot from memory of the resulting native code.
-
-This approach enables directly high performance, at the price of
-portability as usual with pure native code emitting
+This approach directly enables high performance, at the price of
+portability - as usual with pure native code emitting
 approaches. Modularity of features, when possible, is achieved with
-normal software modularity. The indirection costs are taken care by
-the inlining done by the compiler, sometimes even through explicit
-ways to request for it. In particular this modular approach is used
-for implementing a range of choices for GC support XXX ref.  This was
-the inspiration for PyPy own GC framework, although much more tuning
-and work went into Jikes RVM. PyPy own GC framework also exploits
+normal software modularity. The indirection costs are taken care of by
+the compiler performing inlining (which is sometimes even explicitly
+requested).  In particular this modular approach is used
+for implementing a range of choices for GC support [Jikes-GC].  This was
+the inspiration for PyPy's own GC framework, although much more tuning
+and work went into Jikes RVM.  PyPy's own GC framework also exploits
 inlining of helpers and barriers to recover performance.
 
-Jikes RVM native JIT compilers can likely not easily be retargeted to
-run on top and target another VM (for example a CLR runtime) instead
-of hardware processors. Also Jikes RVM pays the complexity of writing
+Jikes RVM's native JIT compilers [Jikes-JIT]
+are not meant to be retargetted to run in other environments
+that hardware processors, for example in a CLR/.NET
+runtime. Also Jikes RVM pays the complexity of writing
 a JIT up-front, which also means that features and semantics of the
-language are encoded in the JIT compiler code, meaning likely that
-major changes would correspond to major surgery needed on it.
-
-PyPy more indirect approach, together hopefully with our future work
+language are encoded in the JIT compiler code.  Major changes of the
+language are likely to correspond to major surgery of the JIT.
+
+PyPy's more indirect approach, together hopefully with our future work
 on generating a JIT compiler, tries to overcome these limitations, at
-the price of some more effort required to achieve very good
-performance. It is too soon to compare completely the complexity (and
-performance) trade-offs of these approaches.
-
-XXX Jython, IronPython, UVM.
+the price of some more efforts required to achieve very good
+performance. It is too soon for a complete comparison of the complexity,
+performance and trade-offs of these approaches.
 
@@
 
-XXX
-
-nice interpreter not polluted by implementation details
-here Python, but any interpreter works
-
-architecture allows implementing features at the right level
-
-dynamic language enables defining our own various type systems
-[ref pluggable type systems]
-
-practical VMs will result with a bit more efforts
+The PyPy project aims at showing that dynamic languages are suitable and
+quite useful for writing virtual machines in.  We believe that we have
+achieved this objective.  The present paper gave an overview of the
+architecture that enabled this result.  Experiments suggest that
+practical virtual machines could reasonably follow in the near future,
+with faster-than-current virtual machines with JIT specialization
+techniques for the mid-term future.
+
+Targetting platforms that are very different from C/Posix is work in
+progress, but given that many of the initial components are shared with
+the existing stack of transformations leading to C, we are confident
+that this work will soon give results.  Moreover, we believe that these
+results will show a reasonable efficiency, because the back-ends for VMs
+like Squeak and .NET can take advantage of high-level input (as opposed
+to trying to translate, say, C-like code to Smalltalk).
+
+A desirable property of our approach is to allow a given language and VM
+to be specified only once, in the form of an interpreter.  Moreover, the
+interpreter can be kept simple (and thus keep its role as a
+specification): not only is it written in a high-level language, but it
+is not overloaded with low-level design choices and implementation
+details.  This makes language evolution and experimentation easier.
+More generally, this property is important because many interpreters for
+very difference languages can be written: the simpler these interpreters
+can be kept, the more we win from our investment in writing the
+tool-chain itself - a one-time effort.
+
+Dynamic languages enable the definition of multiple custom type systems,
+similar to \textit{pluggable type systems} in [Bracha] but with simple
+type inference instead of explicit annotations.  This proved a key
+feature in implementing our translation tool-chain, because it makes a
+many-levels approach convenient: each abstraction level can provide an
+implementation for some of the features that the higher levels
+considered primitive.  It offsets the need to define a minimal kernel of
+primitives and build everything on top of it; instead, we have been able
+to implement, so to speak, the right feature at the right level.
 

### 2006-05-30 talk/dls2006/paper.tex
insert XXX as needed

@@
 approach, is typically a long-winded effort and produces a result that
-is quite [quite could be removed here?] tailored to a specific
+is quite [XXX quite could be removed here?] tailored to a specific
 platform and where architectural decisions (e.g. about GC) are spread
@@
 various optimizations, or turning the whole code into a
-continuation-passing style (CPS) [I'm not sure our transformation
+continuation-passing style (CPS) [XXX I'm not sure our transformation
 can be classified as classical CPS, although there are known similar techniques but the terminology is quite confused] that allows us to use coroutines
 without giving up the ability to generate fully ANSI C code.  (This will
-be the subject of another paper.)  [mention exception transformer too]
+be the subject of another paper.)  [XXX mention exception transformer too]
 
@@
 of having to use pointer-like and address-like objects, Python remains
-more expressive than, say, C to write a GC.  [see also Jikes]
+more expressive than, say, C to write a GC.  [XXX see also Jikes]
 

### 2006-05-30 talk/dls2006/paper.tex
Filling in numbers, and trying to explain them.

@@
 PyPy, written in RPython, producing right now either ANSI C code as
-described before, or LLVM\footnote{the LLVM project is the realisation
+described before, or LLVM\footnote{The LLVM project is the realisation
 of a portable assembler infrastructure, offering both a virtual
@@
 +------------------------------------+-------------------+-------------------+
-|  pypy-c-stackless                  |  XXX       (6.0x) |    XXX     (6.2x) |
+|  pypy-c-stackless                  |  5121ms    (6.5x) |    6060    (6.7x) |
 +------------------------------------+-------------------+-------------------+
@@
 +------------------------------------+-------------------+-------------------+
-|  pypy-c-stackless-gcframework      |  XXX       (    ) |    XXX     (    ) |
+|  pypy-c-stackless-gcframework      |  8743ms   (11.0x) |    3755   (10.7x) |
 +------------------------------------+-------------------+-------------------+
@@
     "stackless transformation" step.  The overhead introduced by the
-    stackless feature is balanced with the removal of the overhead of
-    pushing and popping roots explicitly on an alternate stack: indeed,
-    in this variant it is possible to ask the functions in the current C
-    call chain to save their local state and return.  This has the
-    side-effect of moving all roots to the heap, where the GC can find
-    them.
+    stackless feature is theoretically balanced with the removal of the
+    overhead of pushing and popping roots explicitly on an alternate
+    stack: indeed, in this variant it is possible to ask the functions
+    in the current C call chain to save their local state and return.
+    This has the side-effect of moving all roots to the heap, where the
+    GC can find them.  (We hypothesize that the large slowdown is caused
+    by the extreme size of the executable in this case - 21MB, compared to
+    6MB for the basic pypy-c.  Making it smaller is work in progress.)
 
@@
 GCC time, takes 60 minutes; the same for pypy-c-stackless-gcframework
-takes XXX minutes.
+takes 129 minutes.
 

### 2006-05-30 talk/dls2006/paper.tex
Minor latexifications.

@@
 
-pypy-c
+{\bf pypy-c.}
     The simplest variant: translated to C code with no explicit memory
@@
 
-pypy-c-thread
+{\bf pypy-c-thread.}
     The same, with OS thread support enabled.  (For measurement purposes,
@@
 
-pypy-c-stackless
+{\bf pypy-c-stackless.}
     The same as pypy-c, plus the "stackless transformation" step which
@@
 
-pypy-c-gcframework
+{\bf pypy-c-gcframework.}
     In this variant, the "gc transformation" step inserts explicit
@@
 
-pypy-c-stackless-gcframework
+{\bf pypy-c-stackless-gcframework.}
     This variant combines the "gc transformation" step with the
@@
 
-pypy-llvm-c
+{\bf pypy-llvm-c.}
     The same as pypy-c, but using the LLVM back-end instead of the C
@@
 
-pypy-llvm-c-prof
+{\bf pypy-llvm-c-prof.}
     The same as pypy-llvm-c, but using GCC's profile-driven
@@
 Boehm GC is known to be less efficient than more customized approach;
-kernel-level profiling shows that pypy-c typically spends 30% of its
+kernel-level profiling shows that pypy-c typically spends 30\% of its
 time in the Boehm library.  Our current, naively simple mark-and-sweep

### 2006-05-31 talk/dls2006/paper.tex
Fix figure 1.

@@
 
-\section{System programming with Python}
+\section{System programming in Python}
 \label{systemprog}
@@
 code -- relevant only to a particular transformation -- in plain Python
-as well:
-
-\begin{verbatim}
-.. topic:: Figure 1 - a helper to implement \texttt{list.append()}
-
-  ::
-
-    def ll_append(lst, newitem):
-        # Append an item to the end of the vector.
-        index = lst.length         # get the 'length' field
-        ll_resize(lst, index+1)    # call a helper not shown here
-        itemsarray = lst.items     # get the 'items' field
-        itemsarray[index] = item   # this behaves like a C array
-\end{verbatim}
-
-The idea is to feed these new Python functions into the front-end, using
-this time the transformation's target (lower-level) type system during
-the type inference.  In other words, we can write plain Python code that
-manipulates objects that conform to the lower-level type system, and
-have these functions automatically transformed into appropriately typed
-graphs.
-
-For example, \texttt{ll\textunderscore{}append()} in figure 1 is a Python function
+as well.  The idea is to feed these new Python functions into the
+front-end, using this time the transformation's target (lower-level)
+type system during the type inference.  In other words, we can write
+plain Python code that manipulates objects that conform to the
+lower-level type system, and have these functions automatically
+transformed into appropriately typed graphs.
+
+For example, \texttt{ll\textunderscore{}append()} in figure \ref{llappend}
+is a Python function
 that manipulates objects that behave like C structures and arrays.
@@
 LLTyper.
+
+\begin{figure}
+\begin{verbatim}
+def ll_append(lst, newitem):
+  # Append an item to the end of the vector.
+  index = lst.length       # get the 'length' field
+  ll_resize(lst, index+1)  # call another helper
+  itemsarray = lst.items   # get the 'items' field
+  itemsarray[index] = item # behaves like a C array
+\end{verbatim}
+\caption{A helper to implement \texttt{list.append()}.}
+\label{llappend}
+\end{figure}
 
@@
     GC can find them.  (We hypothesize that the large slowdown is caused
-    by the extreme size of the executable in this case - 21MB, compared to
+    by the extreme size of the executable in this case -- 21MB, compared to
     6MB for the basic pypy-c.  Making it smaller is work in progress.)
@@
 
-Applying the expressiveness - or at least the syntax - of very
+Applying the expressiveness -- or at least the syntax -- of very
 high-level and dynamically typed languages to their implementation has
@@
 This approach directly enables high performance, at the price of
-portability - as usual with pure native code emitting
+portability -- as usual with pure native code emitting
 approaches. Modularity of features, when possible, is achieved with
@@
 can be kept, the more we win from our investment in writing the
-tool-chain itself - a one-time effort.
+tool-chain itself -- a one-time effort.
 

### 2006-05-31 talk/dls2006/paper.tex
latexify tables.

@@
 plug-in component for the Python interpreter of PyPy.  The architecture
-of our Python interpreter is shown in figure 4.
-
-\begin{verbatim}
-.. topic:: Figure 4 - the interpreter and object spaces
-
-    +------------------------------------------------------+
-    |   forest of bytecode objects from the application    |
-    +------------------------------------------------------+
-    |               Python bytecode interpreter            |
-    +--------------------------------+---------------------+
-    |      Standard Object Space     |  Flow Object Space  |
-    +--------------------------------+---------------------+
-\end{verbatim}
+of our Python interpreter is shown in figure \ref{interpobjspace}.
+
+\begin{figure}
+  \centering
+  \caption{the interpreter and object spaces}
+  \label{interpobjspace}
+  \begin{tabular}{|c|c|} \hline
+\multicolumn{2}{|c|}{forest of bytecode objects from the application}    \\ \hline
+\multicolumn{2}{|c|}{Python bytecode interpreter}            \\ \hline
+    Standard Object Space     &  Flow Object Space  \\ \hline
+  \end{tabular}
+\end{figure}
 
@@
 translated interpreters are benchmarked using pystone (a [Dhrystone]
-derivative traditionally used by the Python community, although it is a
-rather poor benchmark) and the classical [Richards] benchmark and
-compared against [CPython] 2.4.3 results:
-
-\begin{verbatim}
-+------------------------------------+-------------------+-------------------+
-|  Interpreter                       | Richards,         | Pystone,          |
-|                                    | Time/iteration    | Iterations/second |
-+====================================+===================+===================+
-|  CPython 2.4.3                     |   789ms    (1.0x) |   40322    (1.0x) |
-+------------------------------------+-------------------+-------------------+
-|  pypy-c                            |  4269ms    (5.4x) |    7587    (5.3x) |
-+------------------------------------+-------------------+-------------------+
-|  pypy-c-thread                     |  4552ms    (5.8x) |    7122    (5.7x) |
-+------------------------------------+-------------------+-------------------+
-|  pypy-c-stackless                  |  5121ms    (6.5x) |    6060    (6.7x) |
-+------------------------------------+-------------------+-------------------+
-|  pypy-c-gcframework                |  6327ms    (8.0x) |    4960    (8.1x) |
-+------------------------------------+-------------------+-------------------+
-|  pypy-c-stackless-gcframework      |  8743ms   (11.0x) |    3755   (10.7x) |
-+------------------------------------+-------------------+-------------------+
-|  pypy-llvm-c                       |  3797ms    (4.8x) |    7763    (5.2x) |
-+------------------------------------+-------------------+-------------------+
-|  pypy-llvm-c-prof                  |  2772ms    (3.5x) |   10245    (3.9x) |
-+------------------------------------+-------------------+-------------------+
-\end{verbatim}
+derivative traditionally used by the Python community, although it is
+a rather poor benchmark) and the classical [Richards] benchmark and
+compared against [CPython] 2.4.3 results and are summarized in table
+\ref{perfsumm}.
+
+\begin{table*}
+% the layout of this table isn't wonderful, but it's probably OK.
+\centering
+\caption{Summary of interpreter performance}
+\label{perfsumm}
+\begin{tabular}{|l|c|c|} \hline
+\textbf{Interpreter} & 
+\textbf{Richards, Time/iteration} &
+\textbf{Pystone, Iterations/second} \\ \hline
+CPython 2.4.3                &   789ms  (1.0x) & 40322  (1.0x) \\ \hline
+pypy-c                       &  4269ms  (5.4x) &  7587  (5.3x) \\ \hline
+pypy-c-thread                &  4552ms  (5.8x) &  7122  (5.7x) \\ \hline
+pypy-c-stackless             &  5121ms  (6.5x) &  6060  (6.7x) \\ \hline
+pypy-c-gcframework           &  6327ms  (8.0x) &  4960  (8.1x) \\ \hline
+pypy-c-stackless-gcframework &  8743ms (11.0x) &  3755 (10.7x) \\ \hline
+pypy-llvm-c                  &  3797ms  (4.8x) &  7763  (5.2x) \\ \hline
+pypy-llvm-c-prof             &  2772ms  (3.5x) & 10245  (3.9x) \\ \hline
+\end{tabular}
+\end{table*}
 
@@
 A complete translation of the pypy-c variant takes about 39 minutes,
-divided as follows:
-
-\begin{verbatim}
-+-------------------------------------------+------------------------------+
-| Step                                      |   Time (minutes:seconds)     |
-+===========================================+==============================+
-| Front-end                                 |            9:01              |
-| (flow graphs and type inference)          |                              |
-+-------------------------------------------+------------------------------+
-| LLTyper                                   |           10:38              |
-| (from RPython-level to C-level graphs     |                              |
-| and data)                                 |                              |
-+-------------------------------------------+------------------------------+
-| Various low-level optimizations           |            6:51              |
-| (convert some heap allocations to local   |                              |
-| variables, inlining, ...)                 |                              |
-+-------------------------------------------+------------------------------+
-| Database building                         |            8:39              |
-| (this initial back-end step follows all   |                              |
-| graphs and prebuilt data structures       |                              |
-| recursively, assigns names, and orders    |                              |
-| them suitably for code generation)        |                              |
-+-------------------------------------------+------------------------------+
-| Generating C source                       |            2:25              |
-+-------------------------------------------+------------------------------+
-| Compiling (\texttt{gcc -O2})              |            3:23              |
-+-------------------------------------------+------------------------------+
-\end{verbatim}
+divided as shown in table \ref{translationtimes}.
+
+\begin{table*}
+\centering
+\caption{Dividing overall translation time by stage}
+\label{translationtimes}
+\begin{tabular}{|p{0.65\linewidth}|c|} \hline
+\textbf{Step} & \textbf{Time (minutes:seconds)} \\ \hline
+Front-end                                 
+(flow graphs and type inference)          
+& 9:01 \\ \hline
+LLTyper                                   
+(from RPython-level to C-level graphs     
+and data)                                 
+& 10:38 \\ \hline
+Various low-level optimizations           
+(convert some heap allocations to local   
+variables, inlining, ...)                 
+& 6:51 \\ \hline
+Database building                         
+(this initial back-end step follows all   
+graphs and prebuilt data structures       
+recursively, assigns names, and orders    
+them suitably for code generation)        
+& 8:39 \\ \hline
+Generating C source                       
+& 2:25 \\ \hline
+Compiling (\texttt{gcc -O2})              
+& 3:23 \\ \hline
+\end{tabular}
+\end{table*}
 

### 2006-05-31 talk/dls2006/paper.tex
Next figure.

@@
 There are two major components in PyPy:
-
+%
 \begin{enumerate}
@@
 \end{enumerate}
-
+%
 In particular, we have defined a subset of the Python language called
@@
 \end{verbatim}
-
+%
 \begin{enumerate}
@@
 \end{enumerate}
-
+ %     
 At the end of the front-end analysis, the input RPython program is
@@
 
-For example, \texttt{ll\textunderscore{}append()} in figure \ref{llappend}
+For example, \texttt{ll\textunderscore{}append()} in figure \ref{fig_llappend}
 is a Python function
@@
 \caption{A helper to implement \texttt{list.append()}.}
-\label{llappend}
+\label{fig_llappend}
 \end{figure}
@@
 
-The four levels that we considered so far are summarized in figure 2.
-
+The four levels that we considered so far are summarized in figure
+\ref{fig_typesystems}.
+
+\begin{figure}
 \begin{verbatim}
-::
-
-    [figure 2:    RPython
-                  /     \
+                      RPython
                       /     \
@@
                   /
-              /
           Raw addresses
-    ]
 \end{verbatim}
+\caption{Type systems, informally ordered by abstraction level}
+\label{fig_typesystems}
+\end{figure}
 
@@
 types.  For each of them, we implemented:
-
+%
 \begin{enumerate}
@@
 \end{enumerate}
-
+%
 We have defined well-typed operations between variables of these types,
@@
 can also be used by type inference when analysing system code like the
-helpers of figure 1.
+helpers of figure \ref{fig_llappend}.
 
@@
 The set $A$ of RPython types is defined as the following formal terms:
-
+%
 \begin{itemize}
@@
 \end{itemize}
-
+%
 Figures 6 and 7 shows how these types are ordered to form a lattice.  We

### 2006-05-31 talk/dls2006/paper.tex
Verbatimized.
Adding a bit of text, and fixing e.g. => e.g.\

@@
 approach, is typically a long-winded effort and produces a result that
-is quite [XXX quite could be removed here?] tailored to a specific
-platform and where architectural decisions (e.g. about GC) are spread
+is tailored to a specific
+platform and where architectural decisions (e.g.\ about GC) are spread
 across the code in a pervasive and invasive way.
@@
       subset of Python's built-in types, with some more precision to
-      describe e.g. the items stored in container types.
+      describe e.g.\ the items stored in container types.
       Occasionally, a single input function can produce several
@@
 will describe in the sequel.  Finally, a back-end turns the resulting
-graphs into code suitable for the target environment, e.g. C source code
+graphs into code suitable for the target environment, e.g.\ C source code
 ready to be compiled.
@@
 each of them.  These type ids are collected in internal tables that
-describe the layout of the structures, e.g. their sizes and the location
+describe the layout of the structures, e.g.\ their sizes and the location
 of the pointer fields.
 
-We have implemented other transformations as well, e.g. performing
-various optimizations, or turning the whole code into a
-continuation-passing style (CPS) [XXX I'm not sure our transformation
-can be classified as classical CPS, although there are known similar techniques but the terminology is quite confused] that allows us to use coroutines
-without giving up the ability to generate fully ANSI C code.  (This will
-be the subject of another paper.)  [XXX mention exception transformer too]
+We have implemented other transformations as well, e.g.\ performing
+various optimizations, or turning the whole code into a style that
+allows us to use coroutines (still in ANSI C: it is a variant of
+continuation-passing that will be the subject of another paper.)
+Another example is the exception transformer, which transforms graphs
+that still contain implicit exception handling into a form suitable for
+C (currently based on a global flag to signal the presence of an
+exception, set and checked around selected function calls).
 
@@
 accesses, many string processing methods, a good subset of all Python
-built-in functions...  Compared to other approaches [e.g. Squeak], we
+built-in functions...  Compared to other approaches [e.g.\ Squeak], we
 do not try to minimize the number of primitives -- at least not at the
@@
 Python types, and the instances of these types are instances in the
-normal Python sense; e.g. whereas Python has only got a single type
+normal Python sense; e.g.\ whereas Python has only got a single type
 \texttt{list}, RPython has a parametric type \texttt{list(T)} for every RPython
@@
 after operation, and follow all calls recursively.  During this process,
-each variable along the way gets a type.  In various cases, e.g. when we
+each variable along the way gets a type.  In various cases, e.g.\ when we
 close a loop, the previously assigned types can be found to be too
@@
 inference engine and the rules more formally.  The difficult points
-involve mutable containers -- e.g. initially empty list that are filled
+involve mutable containers -- e.g.\ initially empty list that are filled
 somewhere else -- and the discovery of instance attributes -- in Python,
@@
 involve the lattice of Pbcs, involving variables that could contain
-e.g.  one function object among many.  An example of such behavior is
+e.g.\ one function object among many.  An example of such behavior is
 code manipulating a table of function objects: when an item is read
@@
 propagation of conditional types.  Consider the following source code
-fragment::
-
+fragment:
+
+\begin{verbatim}
     if isinstance(x, MyClass):
@@
         g(x)
+\end{verbatim}
 

### 2006-05-31 talk/dls2006/paper.tex
Another XXX.

@@
 compromises for the domain used; the foremost example of this is that
-our RPython type inference performs almost no automatic specialization
-of functions XXX class annotation vs concrete runtime class setsXXX.
-We achieved enough precision for our purpose, though.
+our RPython type inference performs almost no polymorphic specialization
+of functions.  Another example is that unlike some schemes which follow
+exact sets of concrete run-time classes, we only propagate a single
+superclass per variable: the most precise common parent class.  This
+gives enough precision for our code generation purposes.
 

### 2006-05-31 talk/dls2006/paper.tex
i.e. => i.e.\

@@
 for a dynamic language in a dynamic language -- in this case, Python.
-The aim is to translate (i.e. compile) the VM to arbitrary target
+The aim is to translate (i.e.\ compile) the VM to arbitrary target
 environments, ranging in level from C/Posix to Smalltalk/Squeak via
@@
       Occasionally, a single input function can produce several
-      specialized versions, i.e. several similar but differently typed
+      specialized versions, i.e.\ several similar but differently typed
       graphs.  This type inference process is described in more
@@
 
-The actual transformation from function objects -- i.e. bytecode -- to
+The actual transformation from function objects -- i.e.\ bytecode -- to
 flow graph is performed by the Flow Object Space, a short but generic
@@
 
-Note that the left column, i.e. the bytecode interpreter and the
+Note that the left column, i.e.\ the bytecode interpreter and the
 Standard Object Space, form the full Python interpreter of PyPy.  It is
@@
 does not give much information per se.  For this reason, our annotator
-works by flowing types forward, operation after operation, i.e. by
+works by flowing types forward, operation after operation, i.e.\ by
 performing abstract interpretation of the flow graphs.  In a sense, it
@@
 To achieve high performance for dynamic languages such as Python, the
-proven approach is to use dynamic compilation techniques, i.e. to write
+proven approach is to use dynamic compilation techniques, i.e.\ to write
 JITs.  With direct techniques, this is however a major endeavour, and

### 2006-05-31 talk/dls2006/paper.tex
use lattice{1,2}.pdf.  about zero thought went in to the amount of scaling to use and the captions; tweaking may be desirable.

@@
 %
-Figures 6 and 7 shows how these types are ordered to form a lattice.  We
-mostly use its structure of [join-semilattice] only.
-
-\begin{verbatim}
-.. graphviz:: image/lattice1.dot
-
-:Figure 6: the lattice of annotations.
-\end{verbatim}
-
-\begin{verbatim}
-.. graphviz:: image/lattice2.dot
-
-:Figure 7: The part about instances and nullable instances, assuming a
+Figures \ref{latticeoverall} and \ref{latticedetail} show how these
+types are ordered to form a lattice.  We mostly use its structure of
+[join-semilattice] only.
+
+\begin{figure*}
+\centering
+\includegraphics[scale=0.75]{image/lattice1.pdf}
+\caption{The overall lattice}
+\label{latticeoverall}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\includegraphics[scale=0.75]{image/lattice2.pdf}
+\caption{The part about instances and nullable instances, assuming a
   simple class hierarchy with only two direct subclasses of
-           \texttt{object}.
-\end{verbatim}
+  \texttt{object}.}
+\label{latticedetail}
+\end{figure*}
 

### 2006-05-31 talk/dls2006/paper.tex
a very quick pdf of "figure 5" done in omnigraffle.  it might be better to do it in dot instead, for consistency with the other lattices -- or remove it entirely, i'm not sure it's worth the space it takes up.
More XXXes.

@@
 an RPython program, and the whole purpose of the translation process is
-to accept this as \textit{input}, and translate it to an efficient form.  Its
-architecture is not relevant to the way it is translated. XXX?
+to accept this as \textit{input}, and translate it to an efficient form.
+The description of this particular input program is beyond the scope of
+the present paper; see [S].
 
@@
 really an abstract domain.  It handles objects that are placeholders.
-Its lattice order is shown in figure 5.
-
-\begin{verbatim}
-::
-
-    [figure 5:                Variable
-
-                         /      |             \           \
-                        /       |              \           \
-                       /        |               \           \
-      Constant(1) ... Constant(n) ... Constant([1,2,3])  ... Constant(<instance of class A>) ...
-    ]
-\end{verbatim}
+Its lattice order is shown in figure \ref{flowlattice}.
+
+\begin{figure*}
+\centering
+\includegraphics[scale=0.75]{image/flowlattice.pdf}
+\caption{The lattice order of the flow object space}
+\label{flowlattice}
+\end{figure*}
 
@@
 the other application-level objects.  We need to be able to unwrap
-them again later. XXX compile-time computation in the helpers.}
+them again later.  Moreover, we rely on this feature to perform
+compile-time computations, particularly in the generic system code
+helpers, which ask for and compute with the concrete types of the
+variables they receive.}
 
@@
 
-Let $x$, $y$ and $z$ be Variables.  We introduce the rule:
-
-% XXX format this!
+Let $x$, $y$ and $z$ be Variables.  We introduce the rules:
+%
 $$
@@
  z = \mathrm{add}(x, y), \; b(x) =
-    Int, \; Bool \leq b(y) \leq Int \\ \hline
+    Int, \; b(y) = Int \\ \hline
  b' = b \hbox{\ with\ } (z
@@
 $$
-
-which specify that if we see the addition operation applied to
+%
+$$
+\begin{array}{c}
+ z = \mathrm{add}(x, y), \; Bool \leq b(x), b(y) \leq
+    NonNegInt \\ \hline
+ b' = b \hbox{\ with\ } (z
+    \rightarrow NonNegInt) 
+\end{array}
+$$
+%
+The first rule specifies that if we see the addition operation applied to
 Variables whose current binding is $Int$, a new binding $b'$ can be
 produced: it is $b$ except on $z$, where we have $b'(z) = Int$.
+The second rule specifies that in a similar case, if both arguments are
+known to be non-negative, so is the result.  Extra rules control addition
+between other combinations of numbers, as well as strings, lists and tuples
+(for which addition is concatenation).  Similar sets of rules are
+introduced for each operation.
 

### 2006-05-31 talk/dls2006/paper.tex
add two authors
Yay.  Only one XXX left

@@
 
+\numberofauthors{2}
+\author{
+\alignauthor Armin Rigo\\
+       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
+       \affaddr{Institut für Informatik}\\ 
+       \affaddr{Universitätsstraße 1}\\
+       \affaddr{D-40225 Düsseldorf}\\
+       \affaddr{Deutschland}\\
+       \email{arigo@tunes.org}
+\alignauthor Samuele Pedroni\\
+       \affaddr{AB Strakt}\\
+       \affaddr{Norra Ågatan 10A}\\
+       \affaddr{416 64  Göteborg}\\
+       \affaddr{Sweden}\\
+       \email{pedronis@strakt.com}
+}
+\date{31 May 2006}
 \maketitle
@@
 really an abstract domain.  It handles objects that are placeholders.
-Its lattice order is shown in figure \ref{flowlattice}.
+Its lattice order (actually a join-semilattice only) is shown in
+figure \ref{flowlattice}.
 
@@
 types are ordered to form a lattice.  We mostly use its structure of
-[join-semilattice] only.
+join-semilattice only.
 
@@
 fragment:
-
+%
 \begin{verbatim}
@@
 \end{verbatim}
-
+%
 Although the type of \texttt{x} may be some parent class of
@@
 $Inst(MyClass)$ within the positive branch of the \texttt{if}.
-(Remember that our graphs are in SSI form, which means that the
+\footnote{Remember that our graphs are in SSI form, which means that the
 \texttt{x} inside each basic block is a different Variable with a
-possibly different type as annotation.) XXX flow sensivity and SSA
+possibly different type as annotation.  Given that our type inference
+is not flow-sensitive, SSI gives an advantage over SSA here.}
 
@@
 boolean values:
-
+%
 $$
@@
 $$
-
+%
 where the $v_n$ are variables and $t_n$ and $f_n$ are types.  The
@@
 in small examples.  For larger examples, a different, non-local
-technique is required: the specialization of functions. XXX use 'polymorphism'
+technique is required: the specialization of type-polymorphic functions.
 
@@
 approach is that automatic specialization is a potential
-performance-killer.  We \textit{do} support specialization, however: we can
-generate several independently-annotated copies of the flow graphs of
-certain functions.  When annotating RPython programs, such
-specialization does not happen automatically: we rely on hints provided
-by the programmer in the source code, in the form of flags attached to
-function objects.  As we had this trade-off in mind when we wrote the
-Python interpreter of PyPy, we only had to add a dozen or so hints in
-the end.
+performance-killer.  So we decided that in the normal case, all calls to
+a given function contribute to a single annotation of that function.
+Inside the function graph, we propagate the \textit{join} of the types
+of the actual parameters of the call sites.  We \textit{do} support
+specialization, however: we can generate several independently-annotated
+copies of the flow graphs of certain functions.  When annotating RPython
+programs, such specialization does not happen automatically: we rely on
+hints provided by the programmer in the source code, in the form of
+flags attached to function objects.  As we had this trade-off in mind
+when we wrote the Python interpreter of PyPy, we only had to add a dozen
+or so hints in the end.
 

### 2006-05-31 talk/dls2006/paper.tex
bibtex fixes, a little bibtex boilerplate and one \cite{}.
Added the typesystem graph.

@@
 \begin{figure}
-\begin{verbatim}
-                      RPython
-                      /     \
-           LLTypeSystem     OOTypeSystem
-                  /
-          Raw addresses
-\end{verbatim}
-\caption{Type systems, informally ordered by abstraction level}
+\centering
+\includegraphics[scale=0.667]{image/typesystems.pdf}
+\caption{type systems; the arrows are transformations.}
 \label{fig_typesystems}
@@
 Scheme48 was a Scheme implementation using a restricted Scheme, PreScheme
-[PreScheme], with static type inference based on Hindley-Milner.  This is
+\cite{kelsey-prescheme}, with static type inference based on Hindley-Milner.  This is
 viable for Scheme as base language.  Simplicity and portability across C
@@
 
+\bibliographystyle{abbrv}
+\bibliography{paper}
+
 \end{document}

### 2006-05-31 talk/dls2006/paper.tex
The \ss sign.
use eps figures

@@
        \affaddr{Institut für Informatik}\\ 
-       \affaddr{Universitätsstraße 1}\\
+       \affaddr{Universitätsstra{\ss}e 1}\\
        \affaddr{D-40225 Düsseldorf}\\
@@
 \centering
-\includegraphics[scale=0.667]{image/typesystems.pdf}
+\epsfig{file=image/typesystems.eps, scale=0.667}
 \caption{type systems; the arrows are transformations.}
@@
 compromises for the domain used; the foremost example of this is that
-our RPython type inference performs almost no polymorphic specialization
-of functions.  Another example is that unlike some schemes which follow
-exact sets of concrete run-time classes, we only propagate a single
-superclass per variable: the most precise common parent class.  This
-gives enough precision for our code generation purposes.
+our RPython type inference performs almost no automatic specialization
+of functions in the presence of polymorphic arguments.  Another
+example is that unlike some schemes which follow exact sets of
+concrete run-time classes, we only propagate a single superclass per
+variable: the most precise common parent class.  This gives enough
+precision for our code generation purposes.
 
@@
 \centering
-\includegraphics[scale=0.75]{image/flowlattice.pdf}
+\epsfig{file=image/flowlattice.eps, scale=0.65}
 \caption{The lattice order of the flow object space}
@@
 \centering
-\includegraphics[scale=0.75]{image/lattice1.pdf}
+\epsfig{file=image/lattice1.eps, scale=0.75}
 \caption{The overall lattice}
@@
 \centering
-\includegraphics[scale=0.75]{image/lattice2.pdf}
+\epsfig{file=image/lattice2.eps, scale=0.75}
 \caption{The part about instances and nullable instances, assuming a

### 2006-05-31 talk/dls2006/paper.tex
it seems we really want pdf
first round add doing real citations

@@
 program that we have translated so far.  More information can be found
-in [S].
+in \cite{A}.
 
@@
 fully leaking program, unless we link it with an external garbage
-collector (GC) like the Boehm conservative GC [Boehm], which is a
+collector (GC) like the Boehm conservative GC \cite{Boehm}, which is a
 viable option.
@@
 
-More information about these transformations can be found in [T].
+More information about these transformations can be found in \cite{T}.
 
@@
 \centering
-\epsfig{file=image/typesystems.eps, scale=0.667}
+\includegraphics[scale=0.667]{image/typesystems.pdf}
 \caption{type systems; the arrows are transformations.}
@@
 \centering
-\epsfig{file=image/flowlattice.eps, scale=0.65}
+\includegraphics[scale=0.75]{image/flowlattice.pdf}
 \caption{The lattice order of the flow object space}
@@
 
-Note that we produce flow graphs in Static Single Information (SSI)
-form, an extension of Static Single Assignment ([SSA]): each variable is
+Note that we produce flow graphs in Static Single Information (SSI, \cite{SSI})
+form, an extension of Static Single Assignment (\cite{SSA}): each variable is
 only used in exactly one basic block.  All variables that are not dead
@@
 sketched above is not entierely straightforward; we refer the reader to
-[D] for more information.
+\cite{D} for more information.
 
@@
 \centering
-\epsfig{file=image/lattice1.eps, scale=0.75}
+\includegraphics[scale=0.75]{image/lattice1.pdf}
 \caption{The overall lattice}
@@
 \centering
-\epsfig{file=image/lattice2.eps, scale=0.75}
+\includegraphics[scale=0.75]{image/lattice2.pdf}
 \caption{The part about instances and nullable instances, assuming a
@@
 
-Squeak [Squeak] is a Smalltalk implementation in Smalltalk. It uses SLang, a
+Squeak \cite{Squeak} is a Smalltalk implementation in Smalltalk. It uses SLang, a
 very restricted subset of Smalltalk with few types and strict
@@
 
-Jikes RVM [DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00] is a
+Jikes RVM \cite{DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00} is a
 Java VM and Just-In-Time compiler written in Java.
@@
 requested).  In particular this modular approach is used
-for implementing a range of choices for GC support [Jikes-GC].  This was
+for implementing a range of choices for GC support \cite{JikesGC}.  This was
 the inspiration for PyPy's own GC framework, although much more tuning
@@
 
-Jikes RVM's native JIT compilers [Jikes-JIT]
+Jikes RVM's native JIT compilers \cite{Jikes-JIT}
 are not meant to be retargetted to run in other environments
@@
 Dynamic languages enable the definition of multiple custom type systems,
-similar to \textit{pluggable type systems} in [Bracha] but with simple
+similar to \textit{pluggable type systems} in \cite{PlugType} but with simple
 type inference instead of explicit annotations.  This proved a key

### 2006-05-31 talk/dls2006/paper.tex
dhrystone 2.0
take care of last XXX

@@
 
-In the example of the \texttt{malloc} operation, replaced by a call to GC
-code, this GC code can invoke a complete collection of dead objects, and
-can thus be arbitrarily complicated.  Still, our GC code is entirely
-written in plain Python, and it manipulates "objects" that are still at
-a lower level: pointer and address objects.  Even with the restriction
-of having to use pointer-like and address-like objects, Python remains
-more expressive than, say, C to write a GC.  [XXX see also Jikes]
+In the example of the \texttt{malloc} operation, replaced by a call to
+GC code, this GC code can invoke a complete collection of dead
+objects, and can thus be arbitrarily complicated.  Still, our GC code
+is entirely written in plain Python, and it manipulates "objects" that
+are still at a lower level: pointer and address objects.  Even with
+the restriction of having to use pointer-like and address-like
+objects, Python remains more expressive than, say, C to write a GC
+(Jikes RVM's GC work \cite{JikesGC} was the inspiration to try to
+express GCs in Python, see \ref{relatedwork}).
 
@@
 transformations enabling various combinations of features. The
-translated interpreters are benchmarked using pystone (a [Dhrystone]
+translated interpreters are benchmarked using pystone (a [Dhrystone] 2.0
 derivative traditionally used by the Python community, although it is

### 2006-05-31 talk/dls2006/paper.tex
Move the footnote numbers after the "," or "." or ":".

@@
 The front-end part of the translation process analyses the input
-RPython program in two phases, as follows\footnote{Note that the two
+RPython program in two phases, as follows:\footnote{Note that the two
 phases are intermingled in time, because type inference proceeds from
 an entry point function and follows all calls, and thus only gradually
-discovers (the reachable parts of) the input program.}:
+discovers (the reachable parts of) the input program.}
 
@@
 \begin{enumerate}
-\item We take as input RPython functions\footnote{The input to our
+\item We take as input RPython functions,\footnote{The input to our
        translation chain are indeed loaded runtime function objects,
@@
        memory.  This includes both the relevant functions and prebuilt
-       data.}, and convert them to control flow graphs -- a structure
+       data.} and convert them to control flow graphs -- a structure
        amenable to analysis.  These flow graphs contain polymorphic
@@
 By going through the front-end reconfigured to use C-level types, the
-above function becomes a graph with such C-level types\footnote{The
+above function becomes a graph with such C-level types,\footnote{The
 low-level type system specifies that the function should be
@@
 pointers, etc.  This behavior gives the programmer a feeling
-comparable to C++ templates, without the declarations.}, which is then
+comparable to C++ templates, without the declarations.} which is then
 indistinguishable from the other graphs of the forest produced by the
@@
 Python -- with the only difference that they use (and have to be
-typeable in) the lower-level type system\footnote{This is not strictly
+typeable in) the lower-level type system.\footnote{This is not strictly
 true: the type systems are even allowed to co-exist in the same
@@
 turned into lower-level operations by the previous transformations in
-the chain, which leave the already-low-level operations untouched.}.
+the chain, which leave the already-low-level operations untouched.}
 
@@
 PyPy thus performs abstract interpretation of Python bytecodes from
-the application\footnote{Note that this process uses the
+the application.\footnote{Note that this process uses the
 \textit{unmodified} bytecode interpreter.  This means that it is
@@
 an interpreter for \textit{any} language to work as a front-end for
-the rest of the tool-chain.}.  In this case, the bytecodes in
+the rest of the tool-chain.}  In this case, the bytecodes in
 question come from the RPython application that we would like to
@@
 arbitrarily complex objects, not just primitives -- and they allow
-basic constant propagation\footnote{This is useful at this level for
+basic constant propagation.\footnote{This is useful at this level for
 some constructs of the bytecode interpreter, which can temporarily
@@
 \texttt{MyClass}, it can be deduced to be of the more precise type
-$Inst(MyClass)$ within the positive branch of the \texttt{if}.
-\footnote{Remember that our graphs are in SSI form, which means that the
+$Inst(MyClass)$ within the positive branch of the
+\texttt{if}.\footnote{Remember that our graphs are in SSI form,
+which means that the
 \texttt{x} inside each basic block is a different Variable with a

### 2006-05-31 talk/dls2006/paper.tex
A reference for partial evaluation.

@@
 on the theoretical possibiliy to turn interpreters into compilers by
-partial evaluation [PE].  In our approach, this is done by analysing
+partial evaluation\cite{Jones:1993:PartialEvaluation}.
+In our approach, this is done by analysing
 the forest of flow graphs built from the Python interpreter, which is a
 well-suited input for this kind of techniques.  We can currently perform
-binding-time analysis [BTA] on these graphs, again with abstract
+binding-time analysis on these graphs, again with abstract
 interpretation techniques reusing the type inference engine.  The next

### 2006-05-31 talk/dls2006/paper.tex
remove hyperlink and add real reference

@@
 \documentclass{acm_proc_article-sp}
-\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
 
@@
 
-\hypertarget{the-translation-process}{}
 \subsection{The translation process}
@@
 \label{flowobjspace}
-\hypertarget{flowobjspace}{}
-
-As described in the overview of \href{\#the-translation-process}{the
-translation process}, the front-end of the translation tool-chain
-works in two phases: it first builds control flow graphs from Python
-functions, and then performs whole-program type inference on these
-graphs.
+
+As described in the overview of the translation process (section
+\ref{translationprocess}), the front-end of the translation tool-chain works in
+two phases: it first builds control flow graphs from Python functions, and then
+performs whole-program type inference on these graphs.
 
@@
 each other, and working on the control flow graphs of each of these
-functions as built by the \href{flowobjspace}{Flow Object Space}.
+functions as built by the Flow Object Space (section \ref{flowobjspace}).
 Additionally, for a particular "entry point" function, the annotator

### 2006-05-31 talk/dls2006/paper.tex
add a missing "section"

@@
 (Jikes RVM's GC work \cite{JikesGC} was the inspiration to try to
-express GCs in Python, see \ref{relatedwork}).
+express GCs in Python, see section \ref{relatedwork}).
 

### 2006-05-31 talk/dls2006/paper.tex
more bib tweaks

@@
 fully leaking program, unless we link it with an external garbage
-collector (GC) like the Boehm conservative GC \cite{Boehm}, which is a
-viable option.
+collector (GC) like the Boehm conservative GC \cite{Boehm}\cite{boehm-softw},
+which is a viable option.
 
@@
 viewer, used for debugging; basic block placement is performed by
-[Graphviz].
+Graphviz \cite{graphviz}.
 

### 2006-05-31 talk/dls2006/paper.tex
references for Boehm already introduced before.
some more reference and tweaks

@@
 useful information to be derived about variables based on how they are
-\textit{used}; only on how they were \textit{produced}.  For example, a number of very
-different built-in types can be involved in an addition; the meaning of
-the addition and the type of the result depends on the type of the input
-arguments.  Merely knowing that a variable will be used in an addition
-does not give much information per se.  For this reason, our annotator
-works by flowing types forward, operation after operation, i.e.\ by
-performing abstract interpretation of the flow graphs.  In a sense, it
-is a more naive approach than the one taken by type systems specifically
-designed to enable more advanced inference algorithms.  For example,
+\textit{used}; only on how they were \textit{produced}.  For example,
+a number of very different built-in types can be involved in an
+addition; the meaning of the addition and the type of the result
+depends on the type of the input arguments.  Merely knowing that a
+variable will be used in an addition does not give much information
+per se.  For this reason, our annotator works by flowing types
+forward, operation after operation, i.e.\ by performing abstract
+interpretation of the flow graphs.  In a sense, it is a more naive
+approach than the one taken by type systems specifically designed to
+enable more advanced inference algorithms.  For example,
 Hindley-Milner type inference works in an inside-out direction, by
 starting from individual operations and propagating type constraints
-outwards [H-M].
+outwards \cite{DaMi}\cite{Miln}.
 
@@
 transformations enabling various combinations of features. The
-translated interpreters are benchmarked using pystone (a [Dhrystone] 2.0
-derivative traditionally used by the Python community, although it is
+translated interpreters are benchmarked using pystone (a Dhrystone 2.0
+\cite{dhry20} derivative traditionally used by the Python community, although it is
 a rather poor benchmark) and the classical [Richards] benchmark and
@@
     The simplest variant: translated to C code with no explicit memory
-    management, and linked with the Boehm conservative GC [Boehm].
+    management, and linked with the Boehm conservative GC.
 
@@
 The resulting generating extension will be essentially similar to
-[Psyco], which is the only (and hand-written) JIT available for Python so
-far, based on run-time specialization.
+Psyco \cite{psyco}, which is the only (and hand-written) JIT available for Python
+so far, based on run-time specialization.
 
@@
 
-Jikes RVM \cite{DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00} is a
+Jikes RVM \cite{jalapeno} is a
 Java VM and Just-In-Time compiler written in Java.

### 2006-05-31 talk/dls2006/paper.tex
[D|T] -> \cite

@@
 $Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
-complete list is described in [T].
+complete list is described in document \cite{T}.
 
@@
 inference engine and the rules more formally.  The difficult points
-involve mutable containers -- e.g.\ initially empty list that are filled
-somewhere else -- and the discovery of instance attributes -- in Python,
-classes do not declare upfront a fixed set of attributes for their
-instances, let alone their types.  Both these examples require
-sophisticated reflowing techniques that invalidate existing types in
-already-annotated basic blocks, to account for the influence of more
-general types coming indirectly from a possibly distant part of the
-program.  The reader is referred to [D] for more information.
+involve mutable containers -- e.g.\ initially empty list that are
+filled somewhere else -- and the discovery of instance attributes --
+in Python, classes do not declare upfront a fixed set of attributes
+for their instances, let alone their types.  Both these examples
+require sophisticated reflowing techniques that invalidate existing
+types in already-annotated basic blocks, to account for the influence
+of more general types coming indirectly from a possibly distant part
+of the program.  The reader is referred to technical report \cite{D}
+for more information.
 
@@
 
-We give formal termination and correctness proofs in [D], as well as
+We give formal termination and correctness proofs in \cite{D}, as well as
 worst-case bounds and some experimental evidence of their practical

### 2006-05-31 talk/dls2006/paper.tex
More URL fighting.
title
sort out quote marks (I love query-replace-regexp)

@@
 
-\title{Still Missing a Cool Title}
-
-% title ideas: Implementing Virtual Machines in Dynamic Languages?
+\title{PyPy's Approach to Virtual Machine Construction}
 
@@
 developed and are now maintaining Java and .NET versions of Python,
-which follow the evolution of the "official" C version (CPython).
+which follow the evolution of the ``official'' C version (CPython).
 
@@
 a necessary component of this equation.  We are basically using the
-standard "meta-programming" argument: if one could write the VM in a
+standard ``meta-programming'' argument: if one could write the VM in a
 very high level language, then the VM itself could be automatically
@@
 In particular, we have defined a subset of the Python language called
-"restricted Python" or RPython.  This sublanguage is not restricted
+``restricted Python'' or RPython.  This sublanguage is not restricted
 syntactically, but only in the way it manipulates objects of different
@@
 transformation step also needs to insert new functions into the forest.
-A key feature of our approach is that we can write such "system-level"
+A key feature of our approach is that we can write such ``system-level''
 code -- relevant only to a particular transformation -- in plain Python
@@
 objects, and can thus be arbitrarily complicated.  Still, our GC code
-is entirely written in plain Python, and it manipulates "objects" that
+is entirely written in plain Python, and it manipulates ``objects'' that
 are still at a lower level: pointer and address objects.  Even with
@@
 
-Now, clearly, the purpose of types like a "C-like struct" or a "C-like
-array" is to be translated to a real \texttt{struct} or array declaration by
+Now, clearly, the purpose of types like a ``C-like struct'' or a ``C-like
+array'' is to be translated to a real \texttt{struct} or array declaration by
 the C back-end.  What, then, is the purpose of emulating such things in
@@
 The description of this particular input program is beyond the scope of
-the present paper; see [S].
+the present paper; see \cite{S}.
 
@@
 The Flow Object Space records all operations that the bytecode
-interpreter "would like" to do between the placeholder objects.  It
+interpreter ``would like'' to do between the placeholder objects.  It
 records them into basic block objects that will eventually be part of
@@
 functions as built by the Flow Object Space (section \ref{flowobjspace}).
-Additionally, for a particular "entry point" function, the annotator
+Additionally, for a particular ``entry point'' function, the annotator
 is provided with user-specified types for the function's arguments.
@@
 
-We can consider that all variables are initially assigned the "bottom"
+We can consider that all variables are initially assigned the ``bottom''
 type corresponding to the empty set of possible run-time values.  Types
@@
 \item $Bot$, $Top$ -- the minimum and maximum elements (corresponding
-      to "impossible value" and "most general value");
+      to ``impossible value'' and ``most general value'');
 
@@
 appear in the user program's flow graphs.  Let $b$ be a map from $V$
-to $A$; it is a "binding" that gives to each variable a type.  The
+to $A$; it is a ``binding'' that gives to each variable a type.  The
 purpose of the annotator is to compute such a binding stepwise.
@@
 {\bf pypy-c-stackless.}
-    The same as pypy-c, plus the "stackless transformation" step which
+    The same as pypy-c, plus the ``stackless transformation'' step which
     modifies the flow graph of all functions in a way that allows them
@@
 {\bf pypy-c-gcframework.}
-    In this variant, the "gc transformation" step inserts explicit
+    In this variant, the ``gc transformation'' step inserts explicit
     memory management and a simple mark-and-sweep GC implementation.
@@
 {\bf pypy-c-stackless-gcframework.}
-    This variant combines the "gc transformation" step with the
-    "stackless transformation" step.  The overhead introduced by the
+    This variant combines the ``gc transformation'' step with the
+    ``stackless transformation'' step.  The overhead introduced by the
     stackless feature is theoretically balanced with the removal of the
@@
 
-Other transformations like the "gc" and the "stackless" ones actually
+Other transformations like the ``gc'' and the ``stackless'' ones actually
 take more time, particuarly when used in combination with each other (we
@@
 translated to C, essentially, although we have already shown that some
-aspects can be "weaved" in at translation time, like support for
+aspects can be ``weaved'' in at translation time, like support for
 coroutines.
@@
 portability were the major goals, as opposed to sophisticated manipulation
-and analysis or "weaving" in of features as transformation aspects.
+and analysis or ``weaving'' in of features as transformation aspects.
 

### 2006-05-31 talk/dls2006/paper.tex
Rephrase.

@@
        translation chain are indeed loaded runtime function objects,
-       not source code or ASTs.  This allows us to use unrestricted
-       python for meta-programming purposes at load time, in a
-       seemingly staged programming approach, in which the whole of
+       not source code or ASTs.  This enables a form of staged
+       programming in which we can use unrestricted
+       Python for meta-programming purposes at load time: the whole of
        the source program -- as Python program -- produces the RPython

### 2006-05-31 talk/dls2006/paper.tex
LL = low-level

@@
 misleading in that respect.} and CLI/.NET.  This first transformation
-step, for C-like environments, is called the \textit{LLTyper}: it produces
+step, for C-like environments, is called the \textit{LLTyper} or
+low-level typer: it produces
 C-level flow graphs, where the object-oriented features of RPython

### 2006-05-31 talk/dls2006/paper.tex
cpython 2.4.3 ref
no ref for Richards at the moment

@@
 translated interpreters are benchmarked using pystone (a Dhrystone 2.0
-\cite{dhry20} derivative traditionally used by the Python community, although it is
-a rather poor benchmark) and the classical [Richards] benchmark and
-compared against [CPython] 2.4.3 results and are summarized in table
-\ref{perfsumm}.
+\cite{dhry20} derivative traditionally used by the Python community,
+although it is a rather poor benchmark) and the classical Richards
+benchmark (ported to Python) and compared against CPython 2.4.3 \cite{cpy243}
+results and are summarized in table \ref{perfsumm}.
 

### 2006-05-31 talk/dls2006/paper.tex
Clarify this paragraph (thanks mwh)

@@
 %
-We have defined well-typed operations between variables of these types,
-plugging on the standard Python operators.  These operations are the
-ones that the emulating instances implement.  As seen above, the types
-can also be used by type inference when analysing system code like the
-helpers of figure \ref{fig_llappend}.
+We have defined well-typed operations between instances of these types,
+syntactically expressed with standard Python operators (e.g.\ if \texttt{x}
+is a C-level array, \texttt{x[n]} accesses its \texttt{n}th item).
+The emulating instances provide a concrete implementation of
+these operations that works in normal Python; the types involved in the
+operations are also known to the type inference engine when it
+analyses system code like the helpers of figure \ref{fig_llappend}.
 

### 2006-05-31 talk/dls2006/paper.tex
typos, wording tweaks, small fixes.  i'm done, i think.

@@
 type inference engine we developed (and can be read independently from
-section 3.)  We present experimental results in section
+section 3).  We present experimental results in section
 \ref{experimentalresults} and future work directions in section
@@
 \item the \textit{Translation Process}: a translation tool-suite whose goal is to
-compile subsets of Python to various environment.
+compile subsets of Python to various environments.
 \end{enumerate}
@@
 types.  The restrictions are a compromise between the expressivity and
-the need to statically infer enough types to generate efficient code.
+the need to statically infer enough type information to generate efficient code.
 The foremost purpose of the translation tool-suite is to compile such
@@
 We will not describe the Standard Interpreter component of PyPy in the
-sequel, other than mention that it is written in RPython and can thus be
+sequel, other than to mention that it is written in RPython and can thus be
 translated.  At close to 90,000 lines of code, it is the largest RPython
@@
 and replaces all \texttt{malloc} operations with calls to a garbage
-collector's innards.  It can inspect all the graphs to discover the
+collector's allocation routine.  It can inspect all the graphs to discover the
 \texttt{struct} types in use by the program, and assign a unique type id to
@@
 objects, Python remains more expressive than, say, C to write a GC
-(Jikes RVM's GC work \cite{JikesGC} was the inspiration to try to
+(the work on the Jikes RVM's GC \cite{JikesGC} was the inspiration to try to
 express GCs in Python, see section \ref{relatedwork}).
@@
 Python types, and the instances of these types are instances in the
-normal Python sense; e.g.\ whereas Python has only got a single type
+normal Python sense; for example where Python has only a single type
 \texttt{list}, RPython has a parametric type \texttt{list(T)} for every RPython
@@
 \begin{enumerate}
-\item the types, which we use to tag the variables of the graphs at
-      the given level.  (Types are actually annotated, self-recursive
+\item The types, which we use to tag the variables of the graphs at
+      the given level (types are actually annotated, self-recursive
       formal terms, and would have been implemented simply as such if
-      Python supported them directly.)
-
-\item the Python objects that emulate instances of these types.  (More
-      about them below.)
+      Python supported them directly).
+
+\item The Python objects that emulate instances of these types (more
+      about them below).
 \end{enumerate}
@@
 the C back-end.  What, then, is the purpose of emulating such things in
-Python?  The answer is three-fold.  Firstly, if we have objects that
+Python?  The answer is three-fold.  Firstly, having objects that
 live within the Python interpreter, but faithfully emulate the behavior
-of their C equivalent while performing additional safety checks, they
-are an invaluable help for testing and debugging.  For example, we can
+of their C equivalent while performing additional safety checks, is 
+an invaluable help for testing and debugging.  For example, we can
 check the correctness of our hash table implementation, written in
@@
 
-Secondly, and anecdotically, as the type inference process (section
+Secondly, and anecdotally, as the type inference process (section
 \ref{typeinference}) is based on abstract interpretation, we can use
@@
 type inference with many different type systems, the details of which
-have evolved along the road.
+have evolved over time.
 
@@
 to the next phase, the type inference engine.  The objects are either
-\textit{Variables}, which are pure placeholders for entierely unknown values,
+\textit{Variables}, which are pure placeholders for entirely unknown values,
 or \textit{Constants} with a concrete Python object as value.  The order places
@@
 
-Note that we produce flow graphs in Static Single Information (SSI, \cite{SSI})
-form, an extension of Static Single Assignment (\cite{SSA}): each variable is
+Note that we produce flow graphs in Static Single Information or SSI\cite{SSI}
+form, an extension of Static Single Assignment\cite{SSA}: each variable is
 only used in exactly one basic block.  All variables that are not dead
@@
 While the Flow Object Space is quite a short piece of code -- its core
-functionality holds in 300 lines -- the detail of the interactions
-sketched above is not entierely straightforward; we refer the reader to
+functionality takes only 300 lines -- the detail of the interactions
+sketched above is not entirely straightforward; we refer the reader to
 \cite{D} for more information.
@@
 
-The goal of the annotator is to find the most precise type that can be
+The goal of the annotator is to find the most precise of our types that can be
 given to each variable of all control flow graphs while respecting the
@@
 can only ever be generalised, and the model is simple enough to show
-that there is no infinite chain of generalisation, so that this process
+that there is no infinite chain of generalization, so that this process
 necessarily terminates.
@@
 As seen in section \ref{systemprog}, we use the annotator with more than one type
-systems.  The more interesting and complex one is the RPython type
+systems.  The most interesting and complex one is the RPython type
 system, which describes how the input RPython program can be annotated.
@@
 $Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
-complete list is described in document \cite{T}.
+complete list is described in \cite{T}.
 
@@
 code manipulating a table of function objects: when an item is read
-out of the table, its type is a large Pbc set: $Pbc(\{f1, f2, f3,
+out of the table, its type is a large Pbc set: $Pbc(\{f_1, f_2, f_3,
 \ldots\})$.  But in this example, the whole set is available at once,
@@
 Our tool-chain is capable of translating the Python interpreter of
-PyPy, written in RPython, producing right now either ANSI C code as
-described before, or LLVM\footnote{The LLVM project is the realisation
+PyPy, written in RPython, currently producing either ANSI C code as
+described before, or LLVM\footnote{The LLVM project is the realization
 of a portable assembler infrastructure, offering both a virtual
 machine with JIT capabilities and static compilation. Currently we are
-using the latter with its good high-level optimisations for PyPy.}
-assembler which is then natively compiled with LLVM tools.
+using the latter with its good high-level optimizations for PyPy.}
+assembler which is then compiled to native code with LLVM tools.
 
@@
 
-{\bf pypy-c.}
-    The simplest variant: translated to C code with no explicit memory
+{\bf pypy-c:}
+    The simplest variant; translated to C code with no explicit memory
     management, and linked with the Boehm conservative GC.
 
-{\bf pypy-c-thread.}
-    The same, with OS thread support enabled.  (For measurement purposes,
-    thread support is kept separate because it has an impact on the GC
-    performance.)
-
-{\bf pypy-c-stackless.}
+{\bf pypy-c-thread:}
+    The same, with OS thread support enabled (thread support is kept
+    separate for measurement purposes because it has an impact on the
+    GC performance).
+
+{\bf pypy-c-stackless:}
     The same as pypy-c, plus the ``stackless transformation'' step which
-    modifies the flow graph of all functions in a way that allows them
+    modifies the flow graph of all functions to allow them
     to save and restore their local state, as a way to enable coroutines.
 
-{\bf pypy-c-gcframework.}
+{\bf pypy-c-gcframework:}
     In this variant, the ``gc transformation'' step inserts explicit
@@
 
-{\bf pypy-c-stackless-gcframework.}
+{\bf pypy-c-stackless-gcframework:}
     This variant combines the ``gc transformation'' step with the
@@
 
-{\bf pypy-llvm-c.}
+{\bf pypy-llvm-c:}
     The same as pypy-c, but using the LLVM back-end instead of the C
@@
 
-{\bf pypy-llvm-c-prof.}
+{\bf pypy-llvm-c-prof:}
     The same as pypy-llvm-c, but using GCC's profile-driven
@@
 time in the Boehm library.  Our current, naively simple mark-and-sweep
-GC is even quite worse.  The interaction with processor caches is also
+GC is manages to be a bit worse.  The interaction with processor caches is also
 hard to predict and account for; in general, we tend to produce
@@
 JITs.  With direct techniques, this is however a major endeavour, and
-increases the efforts to further evolve the language.
+increases the effort involved in further evolution of the language.
 
 In the context of the PyPy project, we are now exploring -- as we planned
-from the start -- the possibility to produce a JIT as a graph
+from the start -- the possibility of producing a JIT as a graph
 transformation aspect from the Python interpreter.  This idea is based
@@
 step is to transform the graphs -- following the binding-time annotations
-- into a compiler; more precisely, in partial evalution terminology, a
+-- into a compiler; more precisely, in partial evalution terminology, a
 generating extension.  We can currently do this on trivial examples.
@@
 are not meant to be retargetted to run in other environments
-that hardware processors, for example in a CLR/.NET
+than hardware processors, for example in a CLR/.NET
 runtime. Also Jikes RVM pays the complexity of writing
@@
 that this work will soon give results.  Moreover, we believe that these
-results will show a reasonable efficiency, because the back-ends for VMs
+results will show reasonable efficiency, because the back-ends for VMs
 like Squeak and .NET can take advantage of high-level input (as opposed

### 2006-05-31 talk/dls2006/paper.tex
two small tweaks motivated by line lengths.

@@
 Although the type of \texttt{x} may be some parent class of
-\texttt{MyClass}, it can be deduced to be of the more precise type
-$Inst(MyClass)$ within the positive branch of the
-\texttt{if}.\footnote{Remember that our graphs are in SSI form,
-which means that the
-\texttt{x} inside each basic block is a different Variable with a
-possibly different type as annotation.  Given that our type inference
-is not flow-sensitive, SSI gives an advantage over SSA here.}
+\texttt{MyClass}, within the positive branch of the \texttt{if} it can
+be deduced to be of the more precise type $Inst(MyClass)$.\footnote{Remember
+ that our graphs are in SSI form, which means
+that the \texttt{x} inside each basic block is a different Variable
+with a possibly different type as annotation.  Given that our type
+inference is not flow-sensitive, SSI gives an advantage over SSA
+here.}
 
@@
 considering new optimizations and better GCs.  Also, the OOTyper and
-back-ends for Smalltalk/Squeak and CLI/.NET are currently in progress.
+back-ends for CLI/.NET and Smalltalk/Squeak are currently in progress.
 

### 2006-05-31 talk/dls2006/paper.tex
Some kind of translation overview graph.

@@
 The translation process starts from RPython source code and eventually
-produces low-level code suitable for the target environment.  It can be
-described as performing a series of step-wise transformations.  Each
-step is based on control flow graph transformations and rewriting, and
-on the ability to augment the program with further implementation code
-written in Python and analysed with the suitable type system.
+produces low-level code suitable for the target environment.  Its
+architecture is shown in figure \ref{fig_arch}.  It can be described as
+a front-end followed by a series of step-wise transformations.  Each
+transformation step is based on control flow graph transformations and
+rewriting, and on the ability to augment the program with further
+implementation code written in Python and analysed with the suitable
+type system.
 
@@
 discovers (the reachable parts of) the input program.}
-
-\begin{verbatim}
-::
-
-  [figure 0: flow graph and annotator, e.g. part of doc/image/translation.*
-             then a stack of transformations
-  ]
-\end{verbatim}
 %
 \begin{enumerate}
-\item We take as input RPython functions,\footnote{The input to our
+\item We take as input RPython function objects,\footnote{The input to our
        translation chain are indeed loaded runtime function objects,
@@
 ready to be compiled.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.667]{image/arch.pdf}
+\caption{Overview of the translation process.}
+\label{fig_arch}
+\end{figure}
 

### 2006-05-31 talk/dls2006/paper.tex
Added the screenshot.  Moved the figures around until they are reasonably placed.

@@
 \includegraphics[scale=0.667]{image/arch.pdf}
-\caption{Overview of the translation process.}
+\caption{overview of the translation process.}
 \label{fig_arch}
@@
 \end{verbatim}
-\caption{A helper to implement \texttt{list.append()}.}
+\caption{a helper to implement \texttt{list.append()}.}
 \label{fig_llappend}
@@
 accesses, many string processing methods, a good subset of all Python
-built-in functions...  Compared to other approaches [e.g.\ Squeak], we
+built-in functions...  Compared to other approaches
+(e.g.\ Squeak \cite{Squeak}), we
 do not try to minimize the number of primitives -- at least not at the
@@
 The various analyses used -- from type inference to lifetime analysis -
-are generally formulated as [abstract interpretation].  While this
+are generally formulated as \textit{abstract interpretation.}  While this
 approach is known to be less efficient than more tailored algorithms
-like constraint-based type inference, we gain in freedom,
+like constraint-based type inference \cite{Constraint}, we gain in freedom,
 controllability and simplicity.  This proved essential in our overall
@@
 We found in-process debugging sufficient and did not implement dumping
-of any intermediate step to disk.  Figure 3 shows the control flow graph
+of any intermediate step to disk.  Figure \ref{fig_screenshot}
+shows the control flow graph
 obtained for a simple function -- this is a screenshot from our graph
@@
 
-\begin{verbatim}
-::
-
-    [figure 3: insert a nice pygame screenshot]
-\end{verbatim}
+\begin{figure}
+\includegraphics[scale=0.59]{image/screenshot-bw.pdf}
+\caption{the control flow graph of a simple function, computed and
+displayed by our tool-chain.}
+\label{fig_screenshot}
+\end{figure}
 
@@
   \centering
-  \caption{the interpreter and object spaces}
   \label{interpobjspace}
   \begin{tabular}{|c|c|} \hline
-\multicolumn{2}{|c|}{forest of bytecode objects from the application}    \\ \hline
-\multicolumn{2}{|c|}{Python bytecode interpreter}            \\ \hline
-    Standard Object Space     &  Flow Object Space  \\ \hline
+\multicolumn{2}{|c|}{\vphantom{\Large X}forest of bytecode objects from the application}    \\ \hline
+\multicolumn{2}{|c|}{\vphantom{\Large X}Python bytecode interpreter}            \\ \hline
+    \vphantom{\Large X} Standard Object Space     &  Flow Object Space  \\ \hline
   \end{tabular}
+  \caption{the interpreter and object spaces.}
 \end{figure}
@@
 \includegraphics[scale=0.75]{image/flowlattice.pdf}
-\caption{The lattice order of the flow object space}
+\caption{the lattice order of the flow object space.}
 \label{flowlattice}
@@
 \subsection{The Annotator}
+
+\begin{figure*}
+\centering
+\includegraphics[scale=0.667]{image/lattice1.pdf}
+\caption{the overall lattice.}
+\label{latticeoverall}
+\end{figure*}
 
@@
 
-\begin{figure*}
+\begin{figure}
 \centering
-\includegraphics[scale=0.75]{image/lattice1.pdf}
-\caption{The overall lattice}
-\label{latticeoverall}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[scale=0.75]{image/lattice2.pdf}
-\caption{The part about instances and nullable instances, assuming a
+\includegraphics[scale=0.667]{image/lattice2.pdf}
+\caption{the part about instances and nullable instances, assuming a
   simple class hierarchy with only two direct subclasses of
@@
 \label{latticedetail}
-\end{figure*}
+\end{figure}
 
@@
 \centering
-\caption{Summary of interpreter performance}
 \label{perfsumm}
@@
 \end{tabular}
+\caption{summary of interpreter performance.}
 \end{table*}
@@
 \centering
-\caption{Dividing overall translation time by stage}
 \label{translationtimes}
@@
 \end{tabular}
+\caption{dividing overall translation time by stage.}
 \end{table*}

### 2006-05-31 talk/dls2006/paper.tex
Tentative categories.

@@
 
+\category{D.3.4}{Programming Languages}{Processors}[code generation,
+interpreters, run-time environments]
+\category{F.3.2}{Logics and Meanings of Programs}{Semantics of Programming
+Languages}[program analysis]
+
 \begin{abstract}
@@
   # Append an item to the end of the vector.
-  index = lst.length       # get the 'length' field
-  ll_resize(lst, index+1)  # call another helper
-  itemsarray = lst.items   # get the 'items' field
-  itemsarray[index] = item # behaves like a C array
+  index = lst.length       #get the 'length' field
+  ll_resize(lst, index+1)  #call another helper
+  itemsarray = lst.items   #get the 'items' field
+  itemsarray[index] = item #behaves like a C array
 \end{verbatim}

### 2006-06-01 talk/dls2006/paper.tex
Minor tweaks on the paper.

@@
        translation chain are indeed loaded runtime function objects,
-       not source code or ASTs.  This enables a form of staged
+       not source code nor ASTs.  This enables a form of staged
        programming in which we can use unrestricted
        Python for meta-programming purposes at load time: the whole of
-       the source program -- as Python program -- produces the RPython
-       program input to the tool-chain as the object graph loaded in
+       the source program -- as a Python program -- produces the RPython
+       program which is sent to the tool-chain in the form of its object
+       graph loaded in
        memory.  This includes both the relevant functions and prebuilt
@@
 The first one inserts naive reference counting throughout the whole
-program's graphs, which without further optimizations gives exceedingly
+program's graphs, which without further optimizations gives a somewhat
 bad performance (it should be noted that the CPython interpreter is also
@@
 and replaces all \texttt{malloc} operations with calls to a garbage
-collector's allocation routine.  It can inspect all the graphs to discover the
-\texttt{struct} types in use by the program, and assign a unique type id to
-each of them.  These type ids are collected in internal tables that
+collector's allocation routine.
+The transformation inspects all the graphs to discover the
+\texttt{struct} types in use by the program, and assigns a unique type id to
+each of them.  These type ids are then collected in internal tables that
 describe the layout of the structures, e.g.\ their sizes and the location
@@
 C (currently based on a global flag to signal the presence of an
-exception, set and checked around selected function calls).
+exception, which is set and checked around selected function calls).
 
@@
 explicit virtual table pointers.  By contrast, for OO environments the
-transformation step is called the \textit{OOTyper}: it targets a simple
+transformation step is called the \textit{OOTyper:} it targets a simple
 object-oriented type system, and preserves the classes and instances
@@
 plain Python code that manipulates objects that conform to the
-lower-level type system, and have these functions automatically
-transformed into appropriately typed graphs.
+lower-level type system, and turn these functions into graph that are
+directly typed in this lower-level type system.
 
@@
 (the work on the Jikes RVM's GC \cite{JikesGC} was the inspiration to try to
-express GCs in Python, see section \ref{relatedwork}).
+express GCs in Python; see section \ref{relatedwork}).
 
@@
 accesses, many string processing methods, a good subset of all Python
-built-in functions...  Compared to other approaches
+built-in functions...  Compared with other approaches
 (e.g.\ Squeak \cite{Squeak}), we
@@
 \item The types, which we use to tag the variables of the graphs at
-      the given level (types are actually annotated, self-recursive
+      the given level (types are actually annotated self-recursive
       formal terms, and would have been implemented simply as such if
@@
 operations are also known to the type inference engine when it
-analyses system code like the helpers of figure \ref{fig_llappend}.
+analyses system code like the helper of figure \ref{fig_llappend}.
 
@@
 \textit{represent} pre-built objects at that level.  For example, the GC
-transformer instantiates the objects emulating C arrays for the internal
-type id tables, and it fills them with the correct values.  These array
-objects are then either used directly when testing the GC, or translated
-by the C back-end into static pre-initialized arrays.
+transformer instantiates an object emulating a C array for the internal
+type id table, and fills it with the correct values.  This array
+object is then either used directly when testing the GC, or translated
+by the C back-end into a static pre-initialized array.
 
@@
 first expect, by following a function at the syntactic level.  Instead,
-the whole program is imported in a normal Python interpreter; the full
-Python language is used at this point as a kind of preprocessor with
+the whole program is imported in a normal Python interpreter -- the full
+Python language is used as a kind of preprocessor with
 meta-programming capabilities.  Once the program is imported, the object
 data in memory consists of Python function objects in bytecode format,
-and any other kind of objects created at import-time, like class
+as well as any other kind of objects created at import-time, like class
 objects, prebuilt instances of those, prebuilt tables, and so on.  Note
@@
 
-The actual transformation from function objects -- i.e.\ bytecode -- to
-flow graph is performed by the Flow Object Space, a short but generic
-plug-in component for the Python interpreter of PyPy.  The architecture
-of our Python interpreter is shown in figure \ref{interpobjspace}.
-
 \begin{figure}
   \centering
-  \label{interpobjspace}
   \begin{tabular}{|c|c|} \hline
@@
   \caption{the interpreter and object spaces.}
+  \label{fig_interpobjspace}
 \end{figure}
 
+The actual transformation from function objects -- i.e.\ bytecode -- to
+flow graph is performed by the Flow Object Space, a short but generic
+plug-in component for the Python interpreter of PyPy.  The architecture
+of our Python interpreter is shown in figure \ref{fig_interpobjspace}.
 Note that the left column, i.e.\ the bytecode interpreter and the
@@
 \centering
-\includegraphics[scale=0.75]{image/flowlattice.pdf}
+\includegraphics[scale=0.667]{image/flowlattice.pdf}
 \caption{the lattice order of the flow object space.}
@@
 at the end of a basic block are explicitly carried over to the next
-block and renamed.
+block and renamed, as can be seen in figure \ref{fig_interpobjspace}:
+each link carries a list of variables matching the formal input
+arguments of its target block.
 
@@
 of possible run-time values, and schedule the block where they appear
-for reflowing.  The more general type can generalise the types of the
-results of the variables in the block, which in turn can generalise the
+for reflowing.
+The newly generalised types can in turn generalise the types of other
+result variables in the block, which in turn can generalise the
 types that flow into the following blocks, and so on.  This process
@@
 \end{itemize}
-%
+
 Figures \ref{latticeoverall} and \ref{latticedetail} show how these
-types are ordered to form a lattice.  We mostly use its structure of
-join-semilattice only.
+types are ordered to form a lattice -- we generally use its structure of
+join-semilattice only, but we will see in section \ref{precision}
+a use case for the \textit{meet}.
+Not shown on the figures are the list terms, which are simply
+unordered with each other, and the Pbcs, which form a
+classical finite set-of-subsets lattice.  In addition, we have left
+out a number of other annotations that are irrelevant for the basic
+description of the annotator and straightforward to handle:
+$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
+complete list is described in \cite{T}.
 
@@
 
-All list terms for all variables are unordered.  The Pbcs form a
-classical finite set-of-subsets lattice.  In addition, we have left
-out a number of other annotations that are irrelevant for the basic
-description of the annotator and straightforward to handle:
-$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
-complete list is described in \cite{T}.
-
 The type system moreover comes with a family of rules, which for every
-operation and every sensible combination of input types describes the
+operation and every meaningful combination of input types describes the
 type of its result variable.  Let $V$ be the set of Variables that
@@
 inference engine and the rules more formally.  The difficult points
-involve mutable containers -- e.g.\ initially empty list that are
-filled somewhere else -- and the discovery of instance attributes --
+involve mutable containers, e.g.\ initially empty list that are
+filled somewhere else; the discovery of instance attributes --
 in Python, classes do not declare upfront a fixed set of attributes
-for their instances, let alone their types.  Both these examples
-require sophisticated reflowing techniques that invalidate existing
+for their instances, let alone their types; and the discovery of
+base class interfaces -- across a class hierarchy, some methods with
+the same name may be called polymorphically by the program, while
+others may be unrelated methods used internally by the subclasses.
+These examples require reflowing techniques that invalidate existing
 types in already-annotated basic blocks, to account for the influence
 of more general types coming indirectly from a possibly distant part
-of the program.  The reader is referred to technical report \cite{D}
+of the program.  The reader is referred to \cite{D}
 for more information.
@@
 We give formal termination and correctness proofs in \cite{D}, as well as
-worst-case bounds and some experimental evidence of their practical
-irrelevance.
+worst-case bounds and common-case estimates.
 
 \subsection{Precision}
+\label{precision}
 
@@
 specialization, however: we can generate several independently-annotated
-copies of the flow graphs of certain functions.  When annotating RPython
+copies of the flow graph of certain functions.  When annotating RPython
 programs, such specialization does not happen automatically: we rely on
@@
 flags attached to function objects.  As we had this trade-off in mind
-when we wrote the Python interpreter of PyPy, we only had to add a dozen
-or so hints in the end.
+when we wrote the Python interpreter of PyPy, we only had to add about
+a dozen hints in the end.
 
@@
 
-At the level of RPython, on the other hand, the range of specializations
-that make sense is much wider.  We have used anything between
-specialization by the type of an argument, to specialization by an
-expected-to-be-constant argument value, to memoized functions that the
+At the level of RPython, the range of specializations
+that make sense is actually much wider.  Although we only specialize a
+very small subset of the functions, we use criteria as diverse as
+specialization by the type of an argument, specialization by an
+expected-to-be-constant argument value, memoized functions that the
 type inference engine will actually call during annotation and replace
-by look-up tables, to complete overriding of the annotator's behavior in
+by look-up tables, and even complete overriding of the annotator's behavior in
 extreme cases.  In this sense, the need for manual specialization turned
@@
 
-This conclusion can be generalized.  We experimented with a simple
+This conclusion can be generalised.  We experimented with a simple
 approach to type inference that works well in practice, and that can
-very flexibly accomodate changes in the type system and even completely
-different type systems.  We think that the reasons for this success are
+very flexibly accomodate small and big changes in the type system.
+We think that the reasons for this success are
 to be found on the one hand in the (reasonable) restrictions we put on

### 2006-06-01 talk/dls2006/paper.tex
Final tweaks.

@@
 PyPy, written in RPython, currently producing either ANSI C code as
-described before, or LLVM\footnote{The LLVM project is the realization
+described before, or LLVM\footnote{
+The LLVM \cite{LLVM:CGO04} project is the realization
 of a portable assembler infrastructure, offering both a virtual
@@
 although it is a rather poor benchmark) and the classical Richards
-benchmark (ported to Python) and compared against CPython 2.4.3 \cite{cpy243}
-results and are summarized in table \ref{perfsumm}.
+benchmark (ported to Python) and compared against CPython 2.4.3 \cite{cpy243}.
+Results are summarized in table \ref{perfsumm}.
 
@@
 {\bf pypy-c:}
-    The simplest variant; translated to C code with no explicit memory
+    the simplest variant; translated to C code with no explicit memory
     management, and linked with the Boehm conservative GC.
@@
 {\bf pypy-c-thread:}
-    The same, with OS thread support enabled (thread support is kept
+    the same, with OS thread support enabled (thread support is kept
     separate for measurement purposes because it has an impact on the
@@
 {\bf pypy-c-stackless:}
-    The same as pypy-c, plus the ``stackless transformation'' step which
+    the same as pypy-c, plus the ``stackless transformation'' step which
     modifies the flow graph of all functions to allow them
@@
 {\bf pypy-c-gcframework:}
-    In this variant, the ``gc transformation'' step inserts explicit
+    in this variant, the ``gc transformation'' step inserts explicit
     memory management and a simple mark-and-sweep GC implementation.
@@
 {\bf pypy-c-stackless-gcframework:}
-    This variant combines the ``gc transformation'' step with the
+    this variant combines the ``gc transformation'' step with the
     ``stackless transformation'' step.  The overhead introduced by the
@@
 {\bf pypy-llvm-c:}
-    The same as pypy-c, but using the LLVM back-end instead of the C
+    the same as pypy-c, but using the LLVM back-end instead of the C
     back-end.  The LLVM assembler-compiler gives the best results when -
@@
 {\bf pypy-llvm-c-prof:}
-    The same as pypy-llvm-c, but using GCC's profile-driven
+    the same as pypy-llvm-c, but using GCC's profile-driven
     optimizations.
@@
 very high load on the GC and on the memory caches of the machine.  The
-Boehm GC is known to be less efficient than more customized approach;
+Boehm GC is known to be less efficient than a more customized approach;
 kernel-level profiling shows that pypy-c typically spends 30\% of its
 time in the Boehm library.  Our current, naively simple mark-and-sweep
-GC is manages to be a bit worse.  The interaction with processor caches is also
-hard to predict and account for; in general, we tend to produce
-relatively large amounts of code and prebuilt data.
+GC manages to be a bit worse.  The interaction with processor caches is
+also hard to predict and account for; in general, we tend to produce
+quite large amounts of code and prebuilt data.  The overall performance
+is still reasonable: some variants are within the same order of
+magnitude as CPython, while others trade a slow-down for functionalities
+not available in CPython.
 
@@
 As described in section \ref{experimentalresults}, the performance of
-the compiled Python interpreters is still not up to competing with the
+the compiled Python interpreters is not yet up to competing with the
 well-established CPython.  We are always working to improve matters,
-considering new optimizations and better GCs.  Also, the OOTyper and
-back-ends for CLI/.NET and Smalltalk/Squeak are currently in progress.
+considering new optimizations and better GCs.
+
+Current work also includes the OOTyper and back-ends for CLI/.NET and
+Smalltalk/Squeak, all in-progress.
 
@@
 proven approach is to use dynamic compilation techniques, i.e.\ to write
-JITs.  With direct techniques, this is however a major endeavour, and
+JITs.  With direct techniques, this is a major endeavour, and
 increases the effort involved in further evolution of the language.
@@
 that this work will soon give results.  Moreover, we believe that these
-results will show reasonable efficiency, because the back-ends for VMs
-like Squeak and .NET can take advantage of high-level input (as opposed
+results will show reasonable efficiency, because our back-ends for VMs
+like Squeak and .NET can take advantage of a high-level input (as opposed
 to trying to translate, say, C-like code to Smalltalk).

### 2006-07-12 talk/dls2006/paper.tex
simple fixes from the reviews

@@
 this is the only transformation step: these graphs are directly fed to
-the C back-end, which turns them into ANSI C source code.
+the C emitting back-end, which turns them into ANSI C source code.
 
@@
 they do, the whole family tends to be available early on, requiring
-few reflowing.
+few reflowings.
 

### 2006-07-12 talk/dls2006/paper.tex
pointer to code repository (we are again above 12 pages, something will have to give)
clarify interpreter line count and role.

@@
 the need to statically infer enough type information to generate efficient code.
-The foremost purpose of the translation tool-suite is to compile such
+The purpose of the translation tool-suite is to compile such
 RPython programs to a variety of different platforms.
@@
 sequel, other than to mention that it is written in RPython and can thus be
-translated.  At close to 90,000 lines of code, it is the largest RPython
-program that we have translated so far.  More information can be found
+translated.  At close to 90,000 lines of code\footnote{The core interpreter
+is more like 30,000 lines of code, the rest includes built-in modules and 30,000 lines
+of auto-generated unicode character database data.}, it is the largest RPython
+program that we have translated so far and is the main target driving
+the development of the tool-chain.  More information can be found
 in \cite{A}.
@@
 The numbers in parenthesis are slow-down factors compared to CPython.
-These measures reflect PyPy revision 27815, compiled with GCC 3.4.4.
-LLVM is version 1.8cvs (May 11, 2006).  The machine runs GNU/Linux SMP
-on an Intel(R) Pentium(R) 4 CPU at 3.20GHz with 2GB of RAM and 1MB of
-cache.  The rows correspond to variants of the translation process, as
-follows:
+These measures reflect PyPy revision 27815\footnote{PyPy is an open-source project
+under the MIT license, PyPy source repository using subversion lives at 
+http://codespeak.net/svn/pypy/dist/.}, 
+compiled with GCC 3.4.4.  LLVM is version 1.8cvs (May 11, 2006).  The
+machine runs GNU/Linux SMP on an Intel(R) Pentium(R) 4 CPU at 3.20GHz
+with 2GB of RAM and 1MB of cache.  The rows correspond to variants of
+the translation process, as follows:
 

### 2006-07-12 talk/dls2006/paper.tex
short discussion of RPython features vs Python, and an example of how RPython code is quite like idiomatic Python.

@@
 types.  The restrictions are a compromise between the expressivity and
-the need to statically infer enough type information to generate efficient code.
-The purpose of the translation tool-suite is to compile such
-RPython programs to a variety of different platforms.
+the need to statically infer enough type information to generate
+efficient code.  RPython still supports inheritance but limited to
+single inheritance with some mix-in support, dynamic dispatch, to some
+extent keywords arguments and varargs, first-class function and class
+values, limited use of bound methods, runtime 'isinstance' and type
+queries, but no runtime reflection, further bindings in class and
+global namespaces are assumed constant. RPython code can be run on a
+Python interpreter without severe semantics mismatches, figure
+\ref{fig_create_frame} shows some RPython code from the
+\textit{Standard Interpreter}, it is still quite idiomatic Python code
+using the built-in dictionary type and first-class class values
+instantiation.
+
+\begin{figure}
+\begin{verbatim}
+  def get_frame_class(self):
+    # select the appropriate kind of frame
+    if not frame_classes:
+      setup_frame_classes()   # lazily
+    choose = 0
+    if self.co_cellvars or self.co_freevars:
+      choose |= NESTED
+    if self.co_flags & CO_GENERATOR:
+      choose |= GENERATOR
+    Frame = frame_classes[choose]
+    return Frame
+
+  def create_frame(self, space, 
+        w_globals, closure=None):
+    return self.get_frame_class()(space, self, 
+              w_globals, closure)
+\end{verbatim}
+\caption{methods from Python bytecode class to instantiate frames.}
+\label{fig_create_frame}
+\end{figure}
+
+The purpose of the translation tool-suite is to compile such RPython
+programs to a variety of different platforms.
 

### 2006-07-12 talk/dls2006/paper.tex
give a better feeling of the expressiveness of helpers.

@@
 extent keywords arguments and varargs, first-class function and class
-values, limited use of bound methods, runtime 'isinstance' and type
+values, limited use of bound methods, runtime \texttt{isinstance} and type
 queries, but no runtime reflection, further bindings in class and
@@
 
+Because our helpers can be both run on top of CPython and translated
+we can reuse them both as runtime implementations and to render constant
+objects at translation time. This is both natural and useful (especially for complex
+data structures). Figure \ref{fig_insertclean} shows a function
+implementing insertion in a built-in dictionary for keys known to be
+absent, this is used at runtime after resizing and to fill the constant
+dictionaries at translation.
+
+\begin{figure}
+\begin{verbatim}
+def ll_dict_insertclean(d, key, value, hash):
+  entry = ll_dict_lookup_clean(d, hash)
+  ENTRY = lltype.typeOf(entry).TO
+  entry.value = value
+  entry.key = key
+  if hasattr(ENTRY, 'f_hash'): 
+    entry.f_hash = hash
+  if hasattr(ENTRY, 'f_valid'):    
+    entry.f_valid = True
+  if hasattr(ENTRY, 'f_everused'): 
+    entry.f_everused = True
+  d.num_items += 1
+  d.num_pristine_entries -= 1
+\end{verbatim}
+\label{fig_insertclean}
+\caption{built-in dict insertion helper.}
+\end{figure}
+
 \section{Type inference}
@@
 precision for our code generation purposes.
+
+It is also simple, in the context of abstract interpretation, to
+perform translation-time constant propagation, computations and
+implement conditional code in a natural way. For example, the code in
+figure \ref{fig_insertclean}, which is then specialized for the
+possible types of dictionary \texttt{d}, shows this kind of
+compile-time computation and introspection on low-level types,
+\texttt{ENTRY} is a C-like structure type and the
+\texttt{hasattr}-ibute checks introspect it at translation time. This
+kind of expressiveness increased our ability of reuse code even in
+low-level helpers.
 

### 2006-07-12 talk/dls2006/paper.tex
typo
adjustements in the intro

@@
 \begin{abstract}
-The PyPy project seeks to prove both on a research and a
-practical level the feasibility of writing a virtual machine (VM)
-for a dynamic language in a dynamic language -- in this case, Python.
-The aim is to translate (i.e.\ compile) the VM to arbitrary target
+The PyPy project seeks to prove both on a research and a practical
+level the feasibility of constructing a virtual machine (VM) for a
+dynamic language in a dynamic language -- in this case, Python.  The
+aim is to translate (i.e.\ compile) the VM to arbitrary target
 environments, ranging in level from C/Posix to Smalltalk/Squeak via
@@
 A key tool to achieve this goal is the systematic reuse of the
-(unmodified, dynamically typed) Python language as a system
+Python language as a system
 programming language at various levels of our architecture and
@@
 approach, is typically a long-winded effort and produces a result that
-is tailored to a specific
-platform and where architectural decisions (e.g.\ about GC) are spread
-across the code in a pervasive and invasive way.
+is tailored to a specific platform and where architectural decisions
+(e.g.\ about GC) are spread across the code in a pervasive and
+invasive way.
 
@@
 a necessary component of this equation.  We are basically using the
-standard ``meta-programming'' argument: if one could write the VM in a
-very high level language, then the VM itself could be automatically
-\textit{translated} to any lower-level platform.  Moreover by writing
-the VM in such a language we would gain in flexibility in
-architectural choices and expressiveness.
+standard ``meta-programming'' argument: if one could write the VM at a
+sufficiently high level of abstraction in a language supporting such
+level, then the VM itself could be automatically \textit{translated}
+to any lower-level platform.  Moreover by writing the VM in such a way
+we would gain in flexibility in architectural choices and
+expressiveness.
 
 PyPy achieves this goal without giving up on the efficiency of the
-compiled VMs.
+compiled VMs, or keeping it within reach of further reasonable
+optimization work.
 
 The key factors enabling this result are not to be found in recent
-advances in any particular research area -- we are not for example using
-constraint-based type inference.  Instead, we are following a novel
-overall architecture: it is split into many levels of stepwise
+advances in any particular research area -- we are not for example
+using constraint-based type inference.  Instead, we are following a
+novel overall architecture: it is split into many levels of stepwise
 translation from the high-level source of the VM to the final target
-platform.  Similar platforms can reuse many of these steps, while for
-very different platforms we have the option to perform very different
-translation steps.  Each step reuses a common type inference component
-with a different, ad-hoc type system.
+platform, each step adding support for features that were assumed
+primitive by the previous ones.  Similar platforms can reuse many of
+these steps, while for very different platforms we have the option to
+perform very different translation steps.  Each step reuses a common
+type inference component with a different, ad-hoc type system.
 
@@
 \texttt{hasattr}-ibute checks introspect it at translation time. This
-kind of expressiveness increased our ability of reuse code even in
+kind of expressiveness increased our ability to reuse code even in
 low-level helpers.

### 2006-07-12 talk/dls2006/paper.tex
Carl's suggestion
sabi's suggestion

@@
 PyPy achieves this goal without giving up on the efficiency of the
-compiled VMs, or keeping it within reach of further reasonable
+compiled VMs, thereby keeping it within reach of further reasonable
 optimization work.

### 2006-07-12 talk/dls2006/paper.tex
tweaking the RPython feature enumeration

@@
 the need to statically infer enough type information to generate
-efficient code.  RPython still supports inheritance but limited to
-single inheritance with some mix-in support, dynamic dispatch, to some
-extent keywords arguments and varargs, first-class function and class
-values, limited use of bound methods, runtime \texttt{isinstance} and type
-queries, but no runtime reflection, further bindings in class and
-global namespaces are assumed constant. RPython code can be run on a
-Python interpreter without severe semantics mismatches, figure
-\ref{fig_create_frame} shows some RPython code from the
-\textit{Standard Interpreter}, it is still quite idiomatic Python code
-using the built-in dictionary type and first-class class values
-instantiation.
+efficient code.  RPython still supports exceptions, inheritance but
+limited to single inheritance with some mix-in support, dynamic
+dispatch, to some extent keywords arguments and varargs, first-class
+function and class values, limited use of bound methods, runtime
+\texttt{isinstance} and type queries, but no runtime
+reflection. Bindings in class and global namespaces are assumed
+constant. RPython code can be run on a Python interpreter without
+severe semantics mismatches, figure \ref{fig_create_frame} shows some
+RPython code from the \textit{Standard Interpreter}, it is still quite
+idiomatic Python code using the built-in dictionary type and
+first-class class values instantiation.
 

### 2006-07-12 talk/dls2006/paper.tex
remove the lattice figures to get again down to 12 pages.
I'm done for today, please review my changes.

I still need to add a reference tough,
will resubmit a first time tomorrow.

@@
 really an abstract domain.  It handles objects that are placeholders.
-Its lattice order (actually a join-semilattice only) is shown in
-figure \ref{flowlattice}.
-
-\begin{figure*}
-\centering
-\includegraphics[scale=0.667]{image/flowlattice.pdf}
-\caption{the lattice order of the flow object space.}
-\label{flowlattice}
-\end{figure*}
-
-This order is extremely simple, because most actual analysis is delayed
-to the next phase, the type inference engine.  The objects are either
-\textit{Variables}, which are pure placeholders for entirely unknown values,
-or \textit{Constants} with a concrete Python object as value.  The order places
-Variable as the top, and keeps all \textit{Constants} unordered.  Thus if two
-different constants merge during abstract interpretation, we immediately
-widen them to Variable.
+
+Its lattice order (actually a join-semilattice only) is extremely
+simple, because most actual analysis is delayed to the next phase, the
+type inference engine.  The objects are either \textit{Variables},
+which are pure placeholders for entirely unknown values, or
+\textit{Constants} with a concrete Python object as value.  The order
+places Variable as the top, and keeps all \textit{Constants}
+unordered.  Thus if two different constants merge during abstract
+interpretation, we immediately widen them to Variable.
 
@@
 \subsection{The Annotator}
-
-\begin{figure*}
-\centering
-\includegraphics[scale=0.667]{image/lattice1.pdf}
-\caption{the overall lattice.}
-\label{latticeoverall}
-\end{figure*}
 
@@
 
-Figures \ref{latticeoverall} and \ref{latticedetail} show how these
-types are ordered to form a lattice -- we generally use its structure of
-join-semilattice only, but we will see in section \ref{precision}
-a use case for the \textit{meet}.
-Not shown on the figures are the list terms, which are simply
-unordered with each other, and the Pbcs, which form a
-classical finite set-of-subsets lattice.  In addition, we have left
-out a number of other annotations that are irrelevant for the basic
-description of the annotator and straightforward to handle:
-$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
-complete list is described in \cite{T}.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.667]{image/lattice2.pdf}
-\caption{the part about instances and nullable instances, assuming a
-  simple class hierarchy with only two direct subclasses of
-  \texttt{object}.}
-\label{latticedetail}
-\end{figure}
+These types are ordered to form a lattice -- we generally use its
+structure of join-semilattice only, but we will see in section
+\ref{precision} a use case for the \textit{meet}. Pbcs form a
+classical finite set-of-subsets lattice, class annotations are ordered
+by inheritance.  We have left out a number of other annotations that
+are irrelevant for the basic description of the annotator and
+straightforward to handle: $Dictionary$, $Tuple$, $Float$,
+$UnicodePoint$, $Iterator$, etc.  The complete list and order is
+described in \cite{T}.
 

### 2006-07-13 talk/dls2006/paper.tex
a few sentence refactorings and a couple of XXX comments

@@
 constant. RPython code can be run on a Python interpreter without
-severe semantics mismatches, figure \ref{fig_create_frame} shows some
-RPython code from the \textit{Standard Interpreter}, it is still quite
-idiomatic Python code using the built-in dictionary type and
-first-class class values instantiation.
+severe semantics mismatches. Figure \ref{fig_create_frame} shows some
+RPython code from the \textit{Standard Interpreter} and it can be seen
+that it is still quite idiomatic Python code using the built-in
+dictionary type and first-class class values instantiation.
 
@@
 exception, which is set and checked around selected function calls).
+
+%%% XXX can we be more optimistic about the OOTyper?  It's pretty
+%%% likely that pypy.net will work in some form by OOPSLA -- hi
+%%% Antonio :)
 
@@
 
+%%% XXX if we're looking to save space, i think this reason can go,
+%%% the others seem more important (to me, at least)
+
 Secondly, and anecdotally, as the type inference process (section
@@
 possible types of dictionary \texttt{d}, shows this kind of
-compile-time computation and introspection on low-level types,
+compile-time computation and introspection on low-level types --
 \texttt{ENTRY} is a C-like structure type and the
@@
 objects, and then handles and transforms these graphs.
+
+%%% XXX the first sentence seems to come a bit out of the blue.
 

### 2006-07-13 talk/dls2006/paper.tex
be more confident about, and give greater prominence to, the OOTyper

@@
 
-When the translation target is C or C-like environments, the first of
-the transformation steps takes the RPython-typed flow graphs, still
-containing polymorphic operations only, and produces flow graphs with
-monomorphic C-like operations and C-like types.  In the simplest case,
-this is the only transformation step: these graphs are directly fed to
-the C emitting back-end, which turns them into ANSI C source code.
-
-But RPython comes with automatic memory management, and this first
-transformation step produces flow graphs that also assume automatic
-memory management.  Generating C code directly from there produces a
-fully leaking program, unless we link it with an external garbage
-collector (GC) like the Boehm conservative GC \cite{Boehm}\cite{boehm-softw},
-which is a viable option.
-
-We have two alternatives, each implemented as a transformation step.
+The first of the transformation steps takes the RPython-typed flow
+graphs, still containing polymorphic operations only, and produces
+flow graphs with operations more familiar to the target environment.
+When the target is C or a C-like environment, this means monomorphic
+C-like operations and C-like types.  In the simplest case, this is the
+only transformation step: these graphs are directly fed to the C
+emitting back-end, which turns them into ANSI C source code.
+
+Not all targets of interest are C-like, though, and we have recently
+developed a variant of this step for use when targeting higher-level,
+object-oriented (OO) environments.  Its design was motivated by work
+on back-ends for Smalltalk/Squeak\footnote{Our simple OO type system
+  is designed for \textit{statically-typed} OO environments, including
+  Java; the presence of Smalltalk as a back-end might be misleading in
+  that respect.} and CLI/.NET.  When targeting a C-like environment,
+the first transformation step is called the \textit{LLTyper} or
+low-level typer: it produces C-level flow graphs, where the
+object-oriented features of RPython (classes and instances) become
+manipulations of C structs with explicit virtual table pointers.  By
+contrast, for OO environments this step is called the
+\textit{OOTyper:} it targets a simple object-oriented type system, and
+preserves the classes and instances of the original RPython program.
+The LLTyper and OOTyper still have much code in common, to convert the
+more Python-specific features like the more complex calling
+conventions.  In the sequel, we assume the use of the LLTyper, as
+there are as yet no back-ends for the OOTyper which are complete
+enough to translate the Standard Interpreter -- although the CLI
+backend should be able to within a few months.
+
+For an example of a transformation that is applied after this first
+step, consider that RPython has automatic memory management. Even with
+the LLTyper, the first transformation step produces flow graphs that
+also assume automatic memory management.  Generating C code directly
+from there produces a fully leaking program, unless we link it with an
+external garbage collector (GC) like the Boehm conservative GC
+\cite{Boehm}\cite{boehm-softw}, which is a viable option.
+
+We have two other alternatives, each implemented as a transformation step.
 The first one inserts naive reference counting throughout the whole
@@
 The other, and better, alternative is an exact GC, coupled with a
-transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
-and replaces all \texttt{malloc} operations with calls to a garbage
-collector's allocation routine.
-The transformation inspects all the graphs to discover the
-\texttt{struct} types in use by the program, and assigns a unique type id to
-each of them.  These type ids are then collected in internal tables that
-describe the layout of the structures, e.g.\ their sizes and the location
-of the pointer fields.
+transformation, the \textit{GC transformer}.  It takes as input
+C-level-typed graphs and replaces all \texttt{malloc} operations with
+calls to a garbage collector's allocation routine.  The transformation
+inspects all the graphs to discover the structure types in use
+by the program, and assigns a unique type id to each of them.  These
+type ids are then collected in internal tables that describe the
+layout of the structures, e.g.\ their sizes and the location of the
+pointer fields.
 
@@
 exception, which is set and checked around selected function calls).
-
-%%% XXX can we be more optimistic about the OOTyper?  It's pretty
-%%% likely that pypy.net will work in some form by OOPSLA -- hi
-%%% Antonio :)
-
-Finally, currently under development is a variant of the very first
-transformation step, for use when targeting higher-level,
-object-oriented (OO) environments.  It is currently being designed
-together with back-ends for Smalltalk/Squeak\footnote{Our simple OO
-type system is designed for \textit{statically-typed} OO environments,
-including Java; the presence of Smalltalk as a back-end might be
-misleading in that respect.} and CLI/.NET.  This first transformation
-step, for C-like environments, is called the \textit{LLTyper} or
-low-level typer: it produces
-C-level flow graphs, where the object-oriented features of RPython
-(classes and instances) become manipulations of C structs with
-explicit virtual table pointers.  By contrast, for OO environments the
-transformation step is called the \textit{OOTyper:} it targets a simple
-object-oriented type system, and preserves the classes and instances
-of the original RPython program.  The LLTyper and OOTyper still have
-much code in common, to convert the more Python-specific features like
-its complex calling conventions.
 
@@
 
-Current work also includes the OOTyper and back-ends for CLI/.NET and
-Smalltalk/Squeak, all in-progress.
+Current work also includes refininments of the OOTyper and work on
+back-ends that use it, including those targeting CLI/.NET and
+Smalltalk/Squeak.
 

### 2006-07-13 talk/dls2006/paper.tex
a paragraph about the jit goal in the intro.

@@
 
-We shortly describe the architecture of PyPy in section
+Recent projects developing meta-circular interpreters
+\cite{KleinVM}\cite{Jikes-JIT}\cite{jalapeno} incorporate a native
+compiler directly in their virtual machine design and high-level
+implementations. We believe that this is a less evolvable and
+maintainable encoding of the semantics of a language than a bytecode
+intepreter, especially with languages such as Python for which
+internal design simplicity was not a goal from the start or in its
+current evolution. Our on-going efforts (beyond the scope of this
+paper) aim instead at generating native compilers as a (non-trivial)
+step of our translation process (see sections \ref{relatedwork} and
+\ref{futurework}). This belief and goal inform our architecture.
+
+In the paper we shortly describe the architecture of PyPy in section
 \ref{architecture}.  In section \ref{systemprog} we describe our

### 2007-05-01 talk/badhonnef2007/talk.tex
talk about pypy and prolog, heavily stolen from the dls talk

\documentclass[utf8]{beamer}

% This file is a solution template for:

% - Talk at a conference/colloquium.
% - Talk length is about 20min.
% - Style is ornate.

\mode<presentation>
{
  \usetheme{Warsaw}
  % or ...

  \setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}

\usepackage[english]{babel}

\usepackage[utf8]{inputenc}
% or whatever

\usepackage{times}
\usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.

\title{A flexible Prolog interpreter in Python}

\author{Carl Friedrich Bolz}
% - Give the names in the same order as the appear in the paper.
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\institute[Heinrich-Heine-Universität Düsseldorf]
{
  Institut für Informatik\\
  Heinrich-Heine-Universität Düsseldorf
}

\date{24. Workshop der GI-Fachgruppe Programmiersprachen und Rechenkonzepte, 4. Mai 2007}
% - Either use conference name or its abbreviation.
% - Not really informative to the audience, more for people (including
%   yourself) who are reading the slides online

% If you have a file called "university-logo-filename.xxx", where xxx
% is a graphic format that can be processed by latex or pdflatex,
% resp., then you can add a logo as follows:

\pgfdeclareimage[height=0.5cm]{pypy-logo}{image/py-web.png}
\logo{\pgfuseimage{pypy-logo}}

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
%\AtBeginSubsection[]
%{
%  \begin{frame}<beamer>
%    \frametitle{Outline}
%    \tableofcontents[currentsection,currentsubsection]
%  \end{frame}
%}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
  \frametitle{Outline}
  \tableofcontents
  % You might wish to add the option [pausesections]
\end{frame}

% Structuring a talk is a difficult task and the following structure
% may not be suitable. Here are some rules that apply for this
% solution:

% - Exactly two or three sections (other than the summary).
% - At *most* three subsections per section.
% - Talk about 30s to 2min per frame. So there should be between about
%   15 and 30 frames, all told.

% - A conference audience is likely to know very little of what you
%   are going to talk about. So *simplify*!
% - In a 20min talk, getting the main ideas across is hard
%   enough. Leave out details, even if it means being less precise than
%   you think necessary.
% - If you omit details that are vital to the proof/implementation,
%   just say so once. Everybody will be happy with that.

\section{What is Pyrolog?}

\begin{frame}
  \frametitle{Pyrolog}

  \begin{itemize}
  \item
    Pyrolog is a Prolog interpreter written in RPython
  \item
    RPython is a subset of Python translatable to other languages
  \item
    translation part done with the help of the PyPy project
  \end{itemize}
\end{frame}

\section{The PyPy Approach to VM Construction}
\subsection{Overview}
\begin{frame}
  \frametitle{What is PyPy?}
  \begin{itemize}
  \item
    started as a Python VM implementation
    in RPython (a well-chosen subset of Python)
  \item
    includes a translation tool-chain
  \item
    is becoming a general environment for writing interpreters (JavaScript, Prolog started)
  \item
    Open source project (MIT license)
  \item
    received EU funding for 2.5 years
  \end{itemize}

\end{frame}

\subsection{Motivation}
\begin{frame}
  \frametitle{VMs are still hard}
  It is hard to achieve:

  \begin{itemize}
  \item
    flexibility
  \item
    maintainability
  \item
    performance (needs dynamic compilation techniques)
  \end{itemize}
  Especially with limited resources (like Open Source projects, research projects)
\end{frame}

\begin{frame}
  \frametitle{The Python case (i)}
  CPython (the reference implementation) is a straightforward, portable VM.

  \begin{itemize}
  \item
    Pervasive decisions: reference counting, single global lock ...
  \item
    No dynamic compilation
%  \pause
  \item
    Extensions:
    \begin{itemize}
    \item
      \alert{Stackless} (unlimited recursion, coroutines, serializable continuations)
    \item
      \alert{Psyco} (run-time specializer)
    \item
      \alert{Jython}, \alert{IronPython}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Python case (ii)}
  \begin{itemize}
  \item
    Extensions have problems
    \begin{itemize}
    \item
      need to keep track of CPython
    \item
      are hard to maintain
    \item
      Psyco very hard to port to other hardware architectures
    \end{itemize}
  \item
    The community wants Python to run everywhere:
    Jython (Java), IronPython (.NET).
    Lots of effort and duplication.

  \item
    At various points various incompatibilities between
    the implementations

  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Prolog case}
  \begin{itemize}
  \item
    problem mitigated by the fact that Prolog the language does not change
  \item
    a lot of implementations out there
  \item
    well-tuned mature C implementations (Sictsus, XSB, SWI, GNU-Prolog)
    \begin{itemize}
    \item
      have sometimes incompatible extensions to core Prolog
    \item
      interfacing with libraries is tedious
    \item
      changing the language to experiment is hard
    \item
      fixed implementation decisions (GC, how to generate code, etc.)
    \end{itemize}
  \item
    on CLR (P\#) and JVM (Prolog Café, tuProlog)
    \begin{itemize}
    \item
      interfacing with libraries of the platform mostly easy
    \item
      no extensions to core Prolog (like tabling, coroutines)
    \item
      slow, compared to good C implementations
    \end{itemize}
  \end{itemize}
\end{frame}

\subsection{Approach}
\begin{frame}
  \frametitle{PyPy's Approach}
    \alert{Goal:} generate VMs from a single high-level description of the
     language, in a retargettable way.
  \begin{itemize}
  \item
    Write an  interpreter for a dynamic language (Python, Prolog, JavaScript,
    whatever) in a high-level language (Python)
  \item
    Leave out low-level details
  \item
    Favour simplicity and flexibility
  \item
    Define a mapping to low-level targets
  \item
    Generate VMs from the interpreter
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Mapping to low-level targets}
  \begin{itemize}
  \item
    Mechanically translate the interpreter to multiple
    lower-level targets
    \begin{itemize}
    \item C-like
    \item Java
    \item .NET
    \end{itemize}
  \item
    Insert low-level aspects into the code as required by
    the target (Object layout, memory management)
    \begin{itemize}
    \item object layout
    \item memory management
    \end{itemize}  \item
    Optionally insert new pervasive features not expressed
    in the source
    \begin{itemize}
    \item continuations
    \item dynamic compilation
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Translation Steps}
  \begin{columns}[c]
  \begin{column}{5cm}
    \includegraphics[width=5cm]{image/arch.pdf}
  \end{column}
  \begin{column}{7cm}
  \begin{itemize}
    \item
      Generate flow graphs from the RPython program
    \item
      Peform global type inference on the flow graphs
    \item
      Transform flow graphs through several steps until they match the level of
      the target environment
    \item
      Weave in translation aspects in the process
  \end{itemize}
  \end{column}
  \end{columns}

\end{frame}

\begin{frame}
  \frametitle{Translation Aspects (i)}
  Features not present in the source can be
  added during translation.

  Example: memory management:
  \begin{itemize}
  \item
  Boehm garbage collector
  \item
  mark-n-sweep written in RPython, with additional features
  \item
  reference counting
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Translation Aspects (ii)}
  \begin{itemize}
  \item
    \alert{Stackless transformation}: continuation capture, implemented by
    saving the low-level frames' local variables into the heap and back
    \begin{itemize}
      \item allows arbitrarily deep stack usage
      \item uses the C stack as long as possible
      \item has the consequence of making RPython do tail call elimination
    \end{itemize}
  \item
    work in progress: turning an interpreter into a just-in-time compiler
    is a translation aspect too
  \end{itemize}
\end{frame}

\section{The Prolog interpreter}
\begin{frame}
  \frametitle{Prolog Interpreter Implementation}
  \begin{itemize}
  \item
    naive, very simple interpreter
    \begin{itemize}
    \item uses "structure copying"
    \item interprets Prolog terms directly, no bytecode
    \end{itemize}
  \item
    uses continuation passing style similar to BinProlog
  \item
    Prolog calls mapped to RPython calls
    \begin{itemize}
    \item possible because stackless allows arbitrary deep recursion
    \end{itemize}
  \item
    implements large parts of the ISO standard (some builtins missing)
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Builtins}
  \begin{itemize}
  \item
    builtins implemented in Python
  \item
    easy to add new ones to interface with libraries
  \item
    application-specific builtins
  \item
    examples:
    \begin{itemize}
    \item functions to download and analyze webpages
    \item an imperative hashmap
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Interpreter Facts}
  \begin{itemize}
  \item
    2500 lines of Python code in total
  \item
    700 of those are for builtins
  \item
    after translation to C: 14000 line of C code
  \item
    part of the PyPy distribution at:
    http://codespeak.net/pypy
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Performance (i)}
  \includegraphics[scale=0.4]{image/bench.png}
\end{frame}

\begin{frame}
  \frametitle{Performance (ii)}
  \begin{itemize}
  \item
    performance is quite bad compared to tuned C implementations
  \item
    performance is pretty good compared to Java and .NET implementations
  \item
    surprising, since those are often based on the WAM
  \item
    maybe it's hard to simulate the WAM on such a VM
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Title}
  \begin{itemize}
  \item
  \end{itemize}
\end{frame}

\section*{Summary}
\subsection*{Summary}
\begin{frame}
  \frametitle<presentation>{Summary}

  % Keep the summary *very short*.
  \begin{itemize}
  \item
    The construction of virtual machines gets easier when using high-level
    languages
  \item
    XXX
  \item
    XXX
  \end{itemize}
\end{frame}
\subsection*{Outlook}
\begin{frame}
  \frametitle<presentation>{Outlook}
  % The following outlook is optional.
  \begin{itemize}
  \item
  \end{itemize}
\end{frame}

\end{document}

### 2007-05-01 talk/badhonnef2007/talk.tex
reworking the talk

@@
   \item
-    RPython is a subset of Python translatable to other languages
+    RPython (``Restricted Python'') is a subset of Python translatable to other
+    languages
   \item
-    translation part done with the help of the PyPy project
+    RPython is designed to be significantly faster than regular Python
+  \item
+    translation tool-chain part of the PyPy project
   \end{itemize}
@@
   \item
-    started as a Python VM implementation
-    in RPython (a well-chosen subset of Python)
+    started as a Python VM implementation in Python
   \item
-    includes a translation tool-chain
+    Python itself is too dynamic to be translatable to other languages, need a subset
+  \item
+    includes a translation tool-chain for RPython
   \item
@@
   \item
-    Open source project (MIT license)
+    Open Source project (MIT license)
   \item
@@
   \frametitle{VMs are still hard}
-  It is hard to achieve:
+  Hard to reconcile:
 
@@
   \item
-    Pervasive decisions: reference counting, single global lock ...
+    Pervasive decisions: reference counting, global lock \dots
   \item
@@
 %  \pause
-  \item
-    Extensions:
+  \end{itemize}
+  \begin{block}{
+    Extensions:}
     \begin{itemize}
@@
     \end{itemize}
-  \end{itemize}
+  \end{block}
 \end{frame}
@@
   \frametitle{The Python case (ii)}
+  \begin{block}{Problems of Extensions:}
     \begin{itemize}
-  \item
-    Extensions have problems
+    \item hard to maintain: need to keep track of CPython
+    \item Psyco very hard to port to other hardware architectures
+    \item tedious to write them, Python semantics need to be re-implemented
+    \end{itemize}
+  \end{block}
   \begin{itemize}
-    \item
-      need to keep track of CPython
-    \item
-      are hard to maintain
-    \item
-      Psyco very hard to port to other hardware architectures
-    \end{itemize}
   \item
@@
 \begin{frame}
-  \frametitle{The Prolog case}
+  \frametitle{The Prolog case (i)}
   \begin{itemize}
@@
   \item
+    the core of Prolog is very simple (at least compared to Python)
+  \item
     a lot of implementations out there
   \item
-    well-tuned mature C implementations (Sictsus, XSB, SWI, GNU-Prolog)
+    well-tuned mature C implementations (Sicstus, XSB, SWI, GNU-Prolog)
+  \item
+    on CLR (P\#) and JVM (Prolog Café, tuProlog)
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{The Prolog case (ii)}
+  \begin{block}{mature C implementations}
     \begin{itemize}
-    \item
-      have sometimes incompatible extensions to core Prolog
     \item
@@
     \item
+      often extensions to core Prolog, incompatible between each other
+    \item
       fixed implementation decisions (GC, how to generate code, etc.)
     \end{itemize}
-  \item
-    on CLR (P\#) and JVM (Prolog Café, tuProlog)
+  \end{block}
+  \begin{block}{implementations on CLR and JVM}
     \begin{itemize}
@@
     \begin{itemize}
-    \item continuations
+    \item continuations, ``micro-threads''
     \item dynamic compilation
@@
 \begin{frame}
-  \frametitle{Translation Steps}
-  \begin{columns}[c]
-  \begin{column}{5cm}
-    \includegraphics[width=5cm]{image/arch.pdf}
-  \end{column}
-  \begin{column}{7cm}
+  \frametitle{Translation Aspects (i)}
+  Features not present in the source can be added during translation.
   \begin{itemize}
   \item
-      Generate flow graphs from the RPython program
+    \alert{memory management}: use different GC strategies (Boehm collector, custom mark-n-sweep)
   \item
-      Peform global type inference on the flow graphs
-    \item
-      Transform flow graphs through several steps until they match the level of
-      the target environment
-    \item
-      Weave in translation aspects in the process
-  \end{itemize}
-  \end{column}
-  \end{columns}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Translation Aspects (i)}
-  Features not present in the source can be
-  added during translation.
-
-  Example: memory management:
-  \begin{itemize}
-  \item
-  Boehm garbage collector
-  \item
-  mark-n-sweep written in RPython, with additional features
-  \item
-  reference counting
+    \alert{Stackless transformation}: allows program to control its stack (continuations, \dots)
   \end{itemize}
@@
   \frametitle{Translation Aspects (ii)}
+  A \alert{JIT compiler} as a translation aspect
   \begin{itemize}
   \item
-    \alert{Stackless transformation}: continuation capture, implemented by
-    saving the low-level frames' local variables into the heap and back
-    \begin{itemize}
-      \item allows arbitrarily deep stack usage
-      \item uses the C stack as long as possible
-      \item has the consequence of making RPython do tail call elimination
-    \end{itemize}
+    Transform the interpreter into a JIT compiler, using partial evaluation and
+    specialization techniques
   \item
-    work in progress: turning an interpreter into a just-in-time compiler
-    is a translation aspect too
+    Some hints in the interpreter source needed
+  \item
+    Current prototype applied to Python interpreter gives impressive speedups
   \end{itemize}
@@
 \section{The Prolog interpreter}
+\subsection{Interpreter}
 \begin{frame}
@@
   \item
-    uses continuation passing style similar to BinProlog
+    uses continuation passing style inspired by BinProlog
   \item
     Prolog calls mapped to RPython calls
-    \begin{itemize}
-    \item possible because stackless allows arbitrary deep recursion
-    \end{itemize}
   \item
@@
     part of the PyPy distribution at:
-    http://codespeak.net/pypy
+    \texttt{http://codespeak.net/pypy}
   \end{itemize}
@@
 
+\subsection{Performance}
 \begin{frame}
@@
   \item
-    maybe it's hard to simulate the WAM on such a VM
+    maybe the WAM model does not match these VMs very well?
   \end{itemize}
+\end{frame}
+
+\section*{Summary}
+\subsection*{Summary}
+\begin{frame}
+  \frametitle{Summary}
+
+  % Keep the summary *very short*.
+  \begin{itemize}
+  \item
+    Very simple Prolog interpreter in RPython can compete with interpreters on the JVM, CLR
+  \item
+    Interpreter implementation eased by use of a high-level language
+  \item
+    Low-level details abstracted away but re-inserted later
+  \end{itemize}
+\end{frame}
+
+\subsection*{Outlook}
+\begin{frame}
+  \frametitle{Outlook}
+  \begin{itemize}
+  \item
+    complete the set of builtins
+  \item
+    tight language integration between Prolog and Python
+  \item
+    apply the dynamic compiler generator to the Prolog interpreter
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Backup slides}
+  \dots
+\end{frame}
+
+\begin{frame}
+  \frametitle{Translation Steps}
+  \begin{columns}[c]
+  \begin{column}{5cm}
+    \includegraphics[width=5cm]{image/arch.pdf}
+  \end{column}
+  \begin{column}{7cm}
+  \begin{itemize}
+    \item
+      Generate flow graphs from the RPython program
+    \item
+      Peform global type inference on the flow graphs
+    \item
+      Transform flow graphs through several steps until they match the level of
+      the target environment
+    \item
+      Weave in translation aspects in the process
+  \end{itemize}
+  \end{column}
+  \end{columns}
 \end{frame}
@@
 
-\section*{Summary}
-\subsection*{Summary}
-\begin{frame}
-  \frametitle<presentation>{Summary}
-
-  % Keep the summary *very short*.
-  \begin{itemize}
-  \item
-    The construction of virtual machines gets easier when using high-level
-    languages
-  \item
-    XXX
-  \item
-    XXX
-  \end{itemize}
-\end{frame}
-\subsection*{Outlook}
-\begin{frame}
-  \frametitle<presentation>{Outlook}
-  % The following outlook is optional.
-  \begin{itemize}
-  \item
-  \end{itemize}
-\end{frame}
-
 \end{document}

### 2007-05-26 talk/dyla2007/dyla.tex
latexify what is there of the paper

\documentclass{llncs}

\usepackage{makeidx}
\usepackage{graphicx}

\begin{document}

\pagestyle{headings}

\title{How to Not write Virtual Machines for Dynamic Languages}

%\titlerunning{XXX}     % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used

\author{Carl Friedrich Bolz and Armin Rigo}

\authorrunning{Bolz and Rigo}   % abbreviated author list (for running head)

%%%% modified list of authors for the TOC (add the affiliations)
\tocauthor{Bolz and Rigo (D\"usseldorf)}

\institute{ Lehrstuhl Softwaretechnik und Programmiersprachen\\
Institut f\"{u}r Informatik, Universit\"at D\"usseldorf, Germany\\
  \email{ cfbolz@gmx.de, arigo@tunes.org}}

\maketitle

\begin{abstract}

Dynamic languages are traditionally implemented by writing a virtual
machine for them, centered around an interpreter and/or a built-in
compiler and providing the object model and memory management. When
a language becomes more successful, the limitations of such an implementation
lead to the emergence of alternative implementations that try to solve some
of the problems. Another reason for new implementations is the desire to have
the language integrate well with existing, well-tuned object-oriented virtual
machine like Java's. In this paper, we describe the mechanisms that lead to an
abundance of implementations, and explore some of the limitations of standard
VMs.  We propose a different complementary alternative: flexibly generating
virtual machines instead
of writing them by hand.  This approach has been validated by the PyPy
project, with which we can automatically insert many features into the
generated virtual machines - including good just-in-time compilers tuned
to the dynamic language at hand.
\footnote{This research is partially supported by the EU funded %research
 project:
IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

\end{abstract}

\section{Introduction}

Dynamic languages are traditionally implemented by writing a virtual
machine for them in a low-level language like C, or in a language that
can relatively easily be turned into C.  The machine implements an
object model supporting the high level dynamic language's objects.  It
typically provides features like automatic garbage collection.  Recent
languages like Python, Ruby, Perl and JavaScript have complicated
semantics which are most easily mapped to a simple interpreter operating
on syntax trees or bytecode; simpler languages like Lisp and Self
typically have more efficient implementations based on just-in-time code
generation.

The efforts required to build a new virtual machine are relatively
important.  This is particularly true for languages which are complex
and in constant evolution. Language implementation communities from an
open-source or academic context have only limited resources. Therefore they
cannot afford to have a highly complex implementation and often chose simpler
techniques even if that entails lower execution speed. Similarly fragmentation
(for example because of other implementations of the same language) is a
problem because it divides available resources. All these points also apply to
the implementation of domain-specific-languages where it is important to keep
implementation effort small.

For these reasons writing a virtual machine in C has many problems because it
forces the language implementer to deal with many low-level details. Limitations
of the C implementation lead to alternative implementations which draw
work-power from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines ("OO VM") such as the JVM or the CLR. This is often wanted by
the community anyway, since it leads to better re-usability of libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters in conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand for having it run on various platforms - high-level VMs as well as
C-level environments.

The argument we will make in the present paper is that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented by a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a good enough
translation toolchain we can then generate whole virtual machines from
this specification - either full custom VMs for C-level operating
systems, or layers on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-VM-fits-all standardization attempt.

The argument boils down to: VMs for dynamic languages should not be
written by hand!  The justification is based on the
PyPy project, which proves that the approach is
feasible in practice.  Just as importantly, it also brings new insights
and concrete benefits in term of flexibility and performance that go
beyond the state of the art.

In section XXX we will explore the way VMs are typically implemented
in C and some of the problems of this approach. In section XXX we will describe
our proposed meta-programming approach. XXX

\section{XXX XXX}

Dynamic languages are traditionally implemented by writing a virtual
machine for them in a low-level language like C, or in a language that
can relatively easily be turned into C.  The machine implements an
object model supporting the high level dynamic language's objects.  It
typically provides features like automatic garbage collection.  Recent
languages like Python, Ruby, Perl and JavaScript have complicated
semantics which are most easily mapped to a simple interpreter operating
on syntax trees or bytecode. Simpler languages like Lisp and Self
typically have more efficient implementations based on just-in-time code
generation.

There are many good reasons for choosing C as an implementation language. It
gives a decent base performance, is highly portable across different hardware
architectures and operating system if some care is taken and it allows the reuse
of many existing libraries.

On the other hand the use of C comes with a set of problems. Since C has no ways
to do automatic memory management and since the implemented languages are
garbage collected, the VM needs to contain a garbage collector. Implementing a
high-performance GC is a tedious task and so many implementations resort to
relatively simple algorithms like mark-and-sweep or reference counting. Also the
choice of garbage collector is fixed relatively early in the life of the project
and then manifests itself throughout the complete VM source code. Thus it
becomes extremely tedious to change it later.

%A similar problem is the threading model the VM uses
%  - simpler threading models
%    - don't make single-threaded apps slower due to locks
%    - global locks
%    - green threading

% - wrapping with C libraries simple but tedious

These limitations of a C-based implementation often lead to more
implementations of the language. For example in addition to the main C
implementation of Python ("CPython") there are a number of additional
implementations that want to fix some of the limitations of CPython. One of
them is "Stackless Python" which is a fork of CPython that adds micro-threading
capabilities to Python. One of the reasons for not incorporating it back into
CPython was that it was felt that they would make the implementation too
complex. Another implementation of the Python language is the Psyco project
which adds a JIT-compiler to CPython. As all other re-implementations, Psyco
contains the Python semantics and needs to be kept synchronized with CPython
manually.

% - free threading attempts for Python

As a language becomes more popular, yet another reason for different
implementations emerges: People want to use it together with Java and C\#
libraries and tools. Although it is in principle possible to do that by
bridging the C-VM with the respective VMs of these languages (XXX examples)
this is an incomplete solution since it prevents many use-cases.
Therefore in the end implementations of the dynamic language on top of the
object-oriented VMs are created to properly integrate with the complete
environment. Implementing a language on top of an OO VM is in many ways easier
than implementing it in C.

\subsection{Assessing the Advantages of Implementing a Language on Top of OO
VMs}

Let's take a look at the advantages that are usually cited for basing a
language implementation of a dynamic language on a standard object oriented
virtual machine, for example the JVM or the CLR. The central theme of the
benefits  of OO VMs is the ability to implement certain hard things only once
and share the benefits between all language implementations on top of the OO VM.

\emph{Better interoperability than the C level:} Since the VM offers a standard
object model and all the languages implemented on top of it are using it, it is
easier to integrate the languages that are running on top of the VM. This
allows reuse of libraries between all the implemented languages. This is
typically the most important reason for wanting an implementation on the VM in
the first place.

\emph{Cross-platform portability:} Only the underlying VM has to be ported to
various hardware architectures and operating systems. The languages implemented
on top can then be run without change in various environments.

\emph{Better tools:} XXX

\emph{Better implementation of low-level issues like garbage collection,
threading:} Since an OO VM is expected to be widely used and usually backed by
a company, it becomes worthwhile and possible to spend a lot of effort tuning
its garbage collector, threading model, exception support and other low-level
implementation details.

\emph{Better performance:} Similarly, object-oriented VMs usually come with a
highly tuned just-in-time compiler to make them perform well without requiring
ahead-of-time compilation to machine language. This in addition with the
previous point leads to much better performance of the languages running on top
of the VM.

\emph{Ease of implementation:} The implementation of a language on top of an OO
VM is easier because it starts of at a higher level than C. Usually a
high-level language like Java or C\# is used for the language implementation,
which both offer the language implementer a much higher level of abstraction
than when implementing in C.

\emph{A single unified implementation base:} The higher level of abstraction
mentioned in the last point also helps to reduce the number of implementations
of the language. Since the implementation on top of an OO VM is more abstract
it is also easier to modify and to extend. XXX somewhat vague point

At a closer look, some of these advantages are only partially true in practice.

\emph{Better performance:} So far it seems like performance of highly dynamic
languages is not actually significantly improved on OO VMs. Jython is around 5
times slower than CPython, for IronPython the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this that the VM's JIT compilers are not prepared to deal with the highly
dynamic behaviour and the complex dispatch semantics of dynamic languages. (XXX
expand)

\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a much
higher memory overhead to start with (XXX ref)

\emph{Cross-platform portability:} While this is true to some extend, the
situation with regard to portability is not significantly improved compared to
e.g.  C/Posix, which is relatively portable too. Also portability sometimes
comes at the price of performance, because even if the OO VM is running on a
particular hardware architecture it is not clear that the JIT is tuned for this
architecture too or working at all, which leads to significantly less
performance.

\emph{Ease of implementation:} This point is disputable. On the one hand, OO
VMs typically allow the language implementor to start at a higher level. On the
other hand they also enforce a specific object and execution model. This means
that the concepts of the implemented language need to be mapped to the
execution model of the underlying VM, which may be easy or not, depending very
much on the language in question. XXX give example where it works well.

An example where this mapping does not work too well is Prolog. While there
exist several implementations of Prolog on top of the JVM and also one on .NET,
they are not particular efficient, especially when compared to good Prolog VMs
in written in C. This is mostly because the Prolog execution model, which
involves backtracking and deep recursion does not fit the JVM and .NET very
well. Therefore the Prolog implementations on top of OO VMs resort to models
that is quite unnatural both for the OO VM and for Prolog.

On the other hand some of the benefits are real and very useful, the most
prominent being the easy interaction with the rest of the VM. Furthermore there
is better tool support and better GCs. XXX expand

\subsection{The Cost of Implementation-Proliferation}

The described proliferation of language implementations is a big problem for
language communities. Although most individual implementations exist for good
reasons, the sum of all of them and the need to keep them synchronized with the
reference implementations lead to a lot of duplicated work and division of
efforts. This is especially true for open source languages which tend to evolve
quickly. At any one point in time some of the implementations will lag behind
which makes writing code which can work on all of the implementations harder.

Implementing a language on top of a OO VM has many advantages, so some
people propose the solution of standardizing on one particular OO VM to not have
to maintain implementations for several of them. While this would in theory
alleviate the problem it is unlikely to happen. On the one hand many political
issues are involved in such a decision. On the other hand deciding on one single
object and execution model would not be an equally good fit for all languages.

In the next section we are exploring a different approach for implementing
dynamic languages that we hope is able to solve many of the problems of
implementing a language, in particular the problem of an explosion of the number
of implementations.

\section{Meta-Programming Is Good}

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically "around" a
description of the language in the form of an interpreter for it.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe -
mostly in term of flexibility.  But most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping multiple ones in sync.  The single
source can be used to generate either custom VMs for C-like
environments, or interpreters running on top of OO VMs.  It makes it
practical to experiment with large changes to the language and with
entirely new languages, like domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/Posix to the JVM to .NET.

\subsection{PyPy architecture}

We implemented this idea in the PyPy project.  The dynamic language for
which we wrote an interpreter is Python.  It is a language which,
because of its size and rather intricate semantics, is a good target for
our approach, in the following sense: its previous reimplementation
(Jython for the JVM and IronPython for .NET) have each proved to be very
time-consuming to maintain.  Our implementation is by construction
easier to maintain, and extremely portable (including to C/Posix, to the
JVM and to .NET).

In meta-programming terms, the PyPy architecture is as follows:

\begin{itemize}

\item
we use a very expressive \emph{object language} (RPython - an analyzable
subset of Python) as the language in which the complete Python
interpreter is written, together with the implementation of its
built-in types.  The language is still close to Python, e.g.  it is
object-oriented, provides rich built-in types and has automatic memory
management.  In other words, the source code of our complete Python
interpreter is mostly free of low-level details - no explicit memory
management, no pieces of C or C-level code.

\item
we use a very expressive metalanguage (namely regular Python) to
perform the analysis of RPython code (control flow and data flow
construction, type inference, etc.) and its successive
transformations.

\item
this meta-programming component of PyPy is called the \emph{translation
framework,} as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine, or integrates it into an existing OO VM.
The necessary support code - e.g. the garbage collector when
targeting C - is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java [JIKES]; as needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\ref{LLVM:CGO04} and .NET are described in [XXX].  These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.

\subsection{A single source}

Our approach - a single "meta-written" implementation - naturally leads
to language implementations that have various advantages over the
"hand-written" implementations.  First of all, it is a single-source
approach - we explicitly seek to solve the problem of proliferation of
implementations.  In the sequel we will show more precise evidence that
this can be done in a practical way with no major drawback.  By itself
this would already be a valid argument against the need for
standardization on a single OO VM.  But there are also other advantages
in generating language implementations. In our opinion these are very
significant, to the extent that it hints that meta-programming, though
not widely used in general-purpose programming, is an essential tool in
a language implementer's toolbox.

\subsection{Writing the interpreter is easier}

A first point is that it makes interpreters easy to write, update and
generally experiment with.  More expressiveness helps at all levels: our
Python interpreter is written in RPython as a relatively simple
interpreter, in some respects easier to understand than CPython.  We are
using its high level and flexibility to quickly experiment with features
or implementation techniques in ways that would, in a traditional
approach, require pervasive changes to the source code.  For example,
PyPy's Python interpreter can optionally provide lazily computed objects
- a 150-lines extension in PyPy that would require global changes in
CPython.  Further examples can be found in our technical reports; we
should notably mention an extension adding a state-of-the-art security
model for Python based on data flow tacking [XXX], and general
performance improvements found by extensive experimentation [XXX], some
of which were back-ported to CPython.

If we compare with hand-writing an implementation for a specific OO VM,
then the latter requires not only good knowledge of the OO VM in
question and its object model - it requires the language implementer to
fit the language into the imposed models.  For example, it is natural to
map classes and instances of a dynamic object-oriented language to the
OO VM's notion of classes and instances, but this might not be a simple
task at all if the models are substantially different and/or if the OO
VM is essentially less dynamic than the language to implement.  In our
approach, this efforts is done at two levels: in a first step, while
writing the interpreter, the implementer does not need to worry about
integration with an OO VM's object model.  Of course, the integration
effort does not simply vanish - indeed, a simple translation of such an
interpreter to a given OO VM would give an interpreter for the dynamic
language which is unable to communicate with the host VM (which might
already be interesting in specific cases, but not in general).
Integration comes as a second step, and occurs at a different level, by
introducing mappings between the relevant classes of the interpreter and
the corresponding classes of the OO VM.  As of yet we have no evidence
that this makes the total integration effort much lower or higher; the
point is that it has proven possible to take the single, OO
VM-independent source code of PyPy's Python interpreter, and produce
from it a version that runs in and integrates with the rest of the .NET
environment - by writing \emph{orthogonal} code.

At the level of the translation framework, the ability to change or
insert new whole-program transformations makes some aspects of the
interpreter easier to deal with.  By "aspect" we mean, in the original
AOP sense, a feature that is added to an object program by a
meta-program.  The most obvious example in our context is the garbage
collector for the target environments that lack it.  The hand-written C
source of CPython is littered with macro calls that increment and
decrement reference counters.  Our translation framework can insert such
macro calls automatically - in fact, we have a choice of GCs, and
reference counting is only one of them (not a particularly efficient
one, either).  Some GCs require different transformations of the code.
By contrast, supporting more than one GC in CPython is close to
impossible without forking the whole source code base.

In fact, there is a well-known example of a CPython fork: Stackless
Python [XXX], which adds support for coroutines.  This is very hard to
do in CPython because the interpreter is written in a highly recursive
style.  Stackless Python required large-scale changes; it is not merged
back into CPython due to the pervasive increase in complexity that it
requires.  In PyPy, though, an optional "stackless" transformation is
able to turn the Python interpreter - also written in a simple highly
recursive style - into an efficient variant of continuation-passing
style (CPS), enabling the usage of coroutines in the translated
interpreter.  For more details and other examples of translation-level
transformations, see [XXX].

\subsection{Getting good GCs and tools is possible}

XXX 
% ---- Bibliography ----
%\begin{small}
\bibliographystyle{abbrv}
\bibliography{dyla}
%\end{small}

\end{document}

### 2007-05-26 talk/dyla2007/dyla.tex
fix some of the references

@@
 this meta-programming component of PyPy is called the \emph{translation
-framework,} as it translates RPython source code (i.e. the full Python
+framework}, as it translates RPython source code (i.e. the full Python
 interpreter) into lower-level code.  Its purpose is to add aspects to
@@
 targeting C - is itself written in RPython in much the same spirit
-that the Jikes RVM's GCs are written in Java [JIKES]; as needed, it is
+that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
 translated together with the interpreter to form the final custom VM.
@@
 interpreter of PyPy and the results we achieved by translating it to C, LLVM
-\ref{LLVM:CGO04} and .NET are described in [XXX].  These results show that the
+\cite{LLVM} and .NET are described in [XXX].  These results show that the
 approach is practical and gives results whose performance is within the same
@@
 should notably mention an extension adding a state-of-the-art security
-model for Python based on data flow tacking [XXX], and general
-performance improvements found by extensive experimentation [XXX], some
+model for Python based on data flow tacking \cite{D12.1}, and general
+performance improvements found by extensive experimentation \cite{D06.1}, some
 of which were back-ported to CPython.
@@
 In fact, there is a well-known example of a CPython fork: Stackless
-Python [XXX], which adds support for coroutines.  This is very hard to
+Python \cite{stackless}, which adds support for coroutines.  This is very hard to
 do in CPython because the interpreter is written in a highly recursive
@@
 interpreter.  For more details and other examples of translation-level
-transformations, see [XXX].
+transformations, see \cite{D07.1}.
 

### 2007-05-26 talk/dyla2007/dyla.tex
use proper quotes and long dashes

@@
 project, with which we can automatically insert many features into the
-generated virtual machines - including good just-in-time compilers tuned
+generated virtual machines – including good just-in-time compilers tuned
 to the dynamic language at hand.
@@
 implementations in C is to build them on top of one of the newer object oriented
-virtual machines ("OO VM") such as the JVM or the CLR. This is often wanted by
+virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
 the community anyway, since it leads to better re-usability of libraries of
@@
 given programming language: as the language becomes popular, there will be a
-demand for having it run on various platforms - high-level VMs as well as
+demand for having it run on various platforms – high-level VMs as well as
 C-level environments.
@@
 translation toolchain we can then generate whole virtual machines from
-this specification - either full custom VMs for C-level operating
+this specification – either full custom VMs for C-level operating
 systems, or layers on top of various OO VMs.  In other words,
@@
 implementations of the language. For example in addition to the main C
-implementation of Python ("CPython") there are a number of additional
+implementation of Python (``CPython'') there are a number of additional
 implementations that want to fix some of the limitations of CPython. One of
-them is "Stackless Python" which is a fork of CPython that adds micro-threading
+them is ``Stackless Python'' which is a fork of CPython that adds micro-threading
 capabilities to Python. One of the reasons for not incorporating it back into
@@
 languages from a meta-level: virtual machines for such languages should
-not be written by hand, but generated automatically "around" a
+not be written by hand, but generated automatically ``around'' a
 description of the language in the form of an interpreter for it.  We
@@
 an implementer when he decides to target an existing object-oriented
-virtual machine.  It also gives other benefits that we will describe -
+virtual machine.  It also gives other benefits that we will describe –
 mostly in term of flexibility.  But most importantly, it lets a
@@
 \item
-we use a very expressive \emph{object language} (RPython - an analyzable
+we use a very expressive \emph{object language} (RPython – an analyzable
 subset of Python) as the language in which the complete Python
@@
 management.  In other words, the source code of our complete Python
-interpreter is mostly free of low-level details - no explicit memory
+interpreter is mostly free of low-level details – no explicit memory
 management, no pieces of C or C-level code.
@@
 standalone virtual machine, or integrates it into an existing OO VM.
-The necessary support code - e.g. the garbage collector when
-targeting C - is itself written in RPython in much the same spirit
+The necessary support code – e.g. the garbage collector when
+targeting C – is itself written in RPython in much the same spirit
 that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
@@
 
-Our approach - a single "meta-written" implementation - naturally leads
+Our approach – a single ``meta-written'' implementation – naturally leads
 to language implementations that have various advantages over the
-"hand-written" implementations.  First of all, it is a single-source
-approach - we explicitly seek to solve the problem of proliferation of
+``hand-written'' implementations.  First of all, it is a single-source
+approach – we explicitly seek to solve the problem of proliferation of
 implementations.  In the sequel we will show more precise evidence that
@@
 PyPy's Python interpreter can optionally provide lazily computed objects
-- a 150-lines extension in PyPy that would require global changes in
+– a 150-lines extension in PyPy that would require global changes in
 CPython.  Further examples can be found in our technical reports; we
@@
 then the latter requires not only good knowledge of the OO VM in
-question and its object model - it requires the language implementer to
+question and its object model – it requires the language implementer to
 fit the language into the imposed models.  For example, it is natural to
@@
 integration with an OO VM's object model.  Of course, the integration
-effort does not simply vanish - indeed, a simple translation of such an
+effort does not simply vanish – indeed, a simple translation of such an
 interpreter to a given OO VM would give an interpreter for the dynamic
@@
 from it a version that runs in and integrates with the rest of the .NET
-environment - by writing \emph{orthogonal} code.
+environment – by writing \emph{orthogonal} code.
 
@@
 insert new whole-program transformations makes some aspects of the
-interpreter easier to deal with.  By "aspect" we mean, in the original
+interpreter easier to deal with.  By ``aspect'' we mean, in the original
 AOP sense, a feature that is added to an object program by a
@@
 decrement reference counters.  Our translation framework can insert such
-macro calls automatically - in fact, we have a choice of GCs, and
+macro calls automatically – in fact, we have a choice of GCs, and
 reference counting is only one of them (not a particularly efficient
@@
 back into CPython due to the pervasive increase in complexity that it
-requires.  In PyPy, though, an optional "stackless" transformation is
-able to turn the Python interpreter - also written in a simple highly
-recursive style - into an efficient variant of continuation-passing
+requires.  In PyPy, though, an optional ``stackless'' transformation is
+able to turn the Python interpreter – also written in a simple highly
+recursive style – into an efficient variant of continuation-passing
 style (CPS), enabling the usage of coroutines in the translated
@@
 \subsection{Getting good GCs and tools is possible}
+XXX
 
+\section{Related Work}
+XXX
+
+\section{Conclusion}
 XXX 

### 2007-05-27 talk/dyla2007/dyla.tex
Details in section 2. Started reworking and completing section 3, intermediate check-in.

@@
 \emph{Ease of implementation:} The implementation of a language on top of an OO
-VM is easier because it starts of at a higher level than C. Usually a
+VM is easier because it starts at a higher level than C. Usually a
 high-level language like Java or C\# is used for the language implementation,
@@
 
-\emph{Better performance:} So far it seems like performance of highly dynamic
-languages is not actually significantly improved on OO VMs. Jython is around 5
-times slower than CPython, for IronPython the figures vary but it is mostly
+\emph{Better performance:} So far it seems that performance of highly dynamic
+languages is not actually significantly improved on OO VMs. Jython
+\footnote{Python on the Java VM} is around 5
+times slower than CPython, for IronPython\footnote{Python on .NET, which
+gives up on some features to improve performance}
+the figures vary but it is mostly
 within the same order of magnitude as CPython. The most important reason for
-this that the VM's JIT compilers are not prepared to deal with the highly
+this is that the VM's JIT compilers are not prepared to deal with the highly
 dynamic behaviour and the complex dispatch semantics of dynamic languages. (XXX
@@
 an implementer when he decides to target an existing object-oriented
-virtual machine.  It also gives other benefits that we will describe –
+virtual machine.  It also gives other benefits that we will describe --
 mostly in term of flexibility.  But most importantly, it lets a
@@
 \item
-we use a very expressive \emph{object language} (RPython – an analyzable
+we use a very expressive \emph{object language} (RPython -- an analyzable
 subset of Python) as the language in which the complete Python
@@
 management.  In other words, the source code of our complete Python
-interpreter is mostly free of low-level details – no explicit memory
+interpreter is mostly free of low-level details -- no explicit memory
 management, no pieces of C or C-level code.
@@
 standalone virtual machine, or integrates it into an existing OO VM.
-The necessary support code – e.g. the garbage collector when
-targeting C – is itself written in RPython in much the same spirit
+The necessary support code -- e.g. the garbage collector when
+targeting C -- is itself written in RPython in much the same spirit
 that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
@@
 
-Our approach – a single ``meta-written'' implementation – naturally leads
+Our approach -- a single ``meta-written'' implementation -- naturally leads
 to language implementations that have various advantages over the
 ``hand-written'' implementations.  First of all, it is a single-source
-approach – we explicitly seek to solve the problem of proliferation of
-implementations.  In the sequel we will show more precise evidence that
-this can be done in a practical way with no major drawback.  By itself
-this would already be a valid argument against the need for
-standardization on a single OO VM.  But there are also other advantages
-in generating language implementations. In our opinion these are very
-significant, to the extent that it hints that meta-programming, though
-not widely used in general-purpose programming, is an essential tool in
-a language implementer's toolbox.
+approach -- we explicitly seek to solve the problem of proliferation of
+implementations.
+
+Separating the implementation of a language in a high-level
+``description'' and a custom translation framework has also many
+advantages -- in our opinion significant enough to hint that
+meta-programming, though not widely used in general-purpose programming,
+is an essential tool in a language implementer's toolbox.
 
@@
 PyPy's Python interpreter can optionally provide lazily computed objects
-– a 150-lines extension in PyPy that would require global changes in
+-- a 150-lines extension in PyPy that would require global changes in
 CPython.  Further examples can be found in our technical reports; we
@@
 
-If we compare with hand-writing an implementation for a specific OO VM,
-then the latter requires not only good knowledge of the OO VM in
-question and its object model – it requires the language implementer to
-fit the language into the imposed models.  For example, it is natural to
-map classes and instances of a dynamic object-oriented language to the
-OO VM's notion of classes and instances, but this might not be a simple
-task at all if the models are substantially different and/or if the OO
-VM is essentially less dynamic than the language to implement.  In our
-approach, this efforts is done at two levels: in a first step, while
-writing the interpreter, the implementer does not need to worry about
-integration with an OO VM's object model.  Of course, the integration
-effort does not simply vanish – indeed, a simple translation of such an
-interpreter to a given OO VM would give an interpreter for the dynamic
-language which is unable to communicate with the host VM (which might
-already be interesting in specific cases, but not in general).
-Integration comes as a second step, and occurs at a different level, by
-introducing mappings between the relevant classes of the interpreter and
-the corresponding classes of the OO VM.  As of yet we have no evidence
-that this makes the total integration effort much lower or higher; the
-point is that it has proven possible to take the single, OO
-VM-independent source code of PyPy's Python interpreter, and produce
-from it a version that runs in and integrates with the rest of the .NET
-environment – by writing \emph{orthogonal} code.
+\subsection{The effort of writing a translation toolchain}
+
+Of course, the price to pay is the need for a translation toolchain
+capable of analyzing and transforming the high-level source code and
+generating lower-level output in various languages.  Of course, the
+translation toolchain, once written, can be reused to implement other
+languages as well.  Even so, we found that the required effort that must
+be put into the translation toolchain in the first place is still much
+lower than that of writing a complete, commercial-quality OO VM.  A
+reason appears to be that we could design our translation toolchain
+specifically for our needs, i.e. a language implementer's needs, instead
+of for general-purpose usage.
+
+...
 
@@
 decrement reference counters.  Our translation framework can insert such
-macro calls automatically – in fact, we have a choice of GCs, and
+macro calls automatically -- in fact, we have a choice of GCs, and
 reference counting is only one of them (not a particularly efficient
@@
 requires.  In PyPy, though, an optional ``stackless'' transformation is
-able to turn the Python interpreter – also written in a simple highly
-recursive style – into an efficient variant of continuation-passing
+able to turn the Python interpreter -- also written in a simple highly
+recursive style -- into an efficient variant of continuation-passing
 style (CPS), enabling the usage of coroutines in the translated
@@
 transformations, see \cite{D07.1}.
+
+\subsection{Integration with a host OO VM}
+
+A good example of this is to compare it against the task of hand-writing
+an implementation for a specific, choosen OO VM.  The latter requires
+not only good knowledge of the OO VM in question and its object model --
+it requires the language implementer to fit the language into the
+imposed models.  For example, it is natural to map classes and instances
+of a dynamic object-oriented language to the OO VM's notion of classes
+and instances, but this might not be a simple task at all if the models
+are substantially different and/or if the OO VM is essentially less
+dynamic than the language to implement.  In our approach, this task
+is done at two levels: in a first step, while writing the interpreter,
+the implementer does not need to worry about integration with an OO VM's
+object model.  Of course, the integration effort does not simply vanish
+-- indeed, a simple translation of such an interpreter to a given OO VM
+would give an interpreter for the dynamic language which is unable to
+communicate with the host VM (which might already be interesting in
+specific cases, but not in general).  Integration comes as a second
+step, and occurs at a different level, by introducing mappings between
+the relevant classes of the interpreter and the corresponding classes of
+the OO VM.  As of yet we have no evidence that this makes the total
+integration effort much lower or higher; the point is that it has proven
+possible to take the single, OO VM-independent source code of PyPy's
+Python interpreter, and produce from it a version that runs in and
+integrates with the rest of the .NET environment -- by writing
+\emph{orthogonal} code.
 

### 2007-05-27 talk/dyla2007/dyla.tex
removing some XXXs

@@
 
-In section XXX we will explore the way VMs are typically implemented
-in C and some of the problems of this approach. In section XXX we will describe
-our proposed meta-programming approach. XXX
+In section \ref{sect:approaches} we will explore the way VMs are typically
+implemented in C and on top of OO VMs and some of the problems of these
+approaches. In \ref{section sect:metaprogramming} we will describe our proposed
+meta-programming approach.
 
-\section{XXX XXX}
+\section{Approaches to Dynamic Language Implementation}
+\label{sect:appraches}
 
@@
 object-oriented VMs are created to properly integrate with the complete
-environment. Implementing a language on top of an OO VM is in many ways easier
-than implementing it in C.
+environment.
 
@@
 
-Let's take a look at the advantages that are usually cited for basing a
+XXX mention that OO languages don't have good metaprogramming support
+
+Implementing a language on top of an OO VM is in many ways easier than
+implementing it in C. Let's take a look at the advantages that are usually
+cited for basing a
 language implementation of a dynamic language on a standard object oriented
@@
 
-\emph{Better tools:} XXX
+\emph{Better tools:} Better IDEs, debuggers and profilers.
 
@@
 within the same order of magnitude as CPython. The most important reason for
-this is that the VM's JIT compilers are not prepared to deal with the highly
-dynamic behaviour and the complex dispatch semantics of dynamic languages. (XXX
-expand)
+this is that the VM's JIT compilers are optimized for specific usage patterns
+that are common in the main language of the OO VM. To get good speeds the
+language implementers would have to carefully produce code that matches these
+usage patterns, which is not a simple task.
 
@@
 \section{Meta-Programming Is Good}
+\label{sect:metaprogramming}
 

### 2007-05-27 talk/dyla2007/dyla.tex
More reorganization, and finished another subsection.

@@
 languages from a meta-level: virtual machines for such languages should
-not be written by hand, but generated automatically ``around'' a
-description of the language in the form of an interpreter for it.  We
+not be written by hand, but generated automatically ``around'' an
+interpreter playing the role of a high-level description of the language.  We
 argue that this approach gives many of the benefits usually expected by
@@
 
-Our approach -- a single ``meta-written'' implementation -- naturally leads
-to language implementations that have various advantages over the
+Our approach -- a single ``meta-written'' implementation -- naturally
+leads to language implementations that have various advantages over the
 ``hand-written'' implementations.  First of all, it is a single-source
 approach -- we explicitly seek to solve the problem of proliferation of
-implementations.
-
-Separating the implementation of a language in a high-level
-``description'' and a custom translation framework has also many
-advantages -- in our opinion significant enough to hint that
-meta-programming, though not widely used in general-purpose programming,
-is an essential tool in a language implementer's toolbox.
+implementations.  In the sequel, we will show that this goal can be
+achieved without giving up on the advantages of hand-written
+implementations for OO VMs.  Moreover, there are additional advantages
+-- in our opinion significant enough to hint that meta-programming,
+though not widely used in general-purpose programming, is an essential
+tool in a language implementer's toolbox.
 
@@
 
-\subsection{The effort of writing a translation toolchain}
-
-Of course, the price to pay is the need for a translation toolchain
-capable of analyzing and transforming the high-level source code and
-generating lower-level output in various languages.  Of course, the
-translation toolchain, once written, can be reused to implement other
-languages as well.  Even so, we found that the required effort that must
-be put into the translation toolchain in the first place is still much
-lower than that of writing a complete, commercial-quality OO VM.  A
-reason appears to be that we could design our translation toolchain
-specifically for our needs, i.e. a language implementer's needs, instead
-of for general-purpose usage.
-
-...
+\subsection{Separation of concerns}
 
@@
 AOP sense, a feature that is added to an object program by a
-meta-program.  The most obvious example in our context is the garbage
-collector for the target environments that lack it.  The hand-written C
-source of CPython is littered with macro calls that increment and
-decrement reference counters.  Our translation framework can insert such
-macro calls automatically -- in fact, we have a choice of GCs, and
-reference counting is only one of them (not a particularly efficient
-one, either).  Some GCs require different transformations of the code.
-By contrast, supporting more than one GC in CPython is close to
-impossible without forking the whole source code base.
+meta-program.  The most obvious example in our context is the insertion
+of a garbage collector (chosen among several available ones) for the
+target environments that lack it.  Another example is the translation of
+the interpreter into a form of continuation-passing style (CPS), which
+allows the translated interpreter to provide coroutines even though its
+source is written in a simple highly recursive style.  For more details
+and other examples of translation-level transformations, see
+\cite{D07.1}.
 
-In fact, there is a well-known example of a CPython fork: Stackless
-Python \cite{stackless}, which adds support for coroutines.  This is very hard to
-do in CPython because the interpreter is written in a highly recursive
-style.  Stackless Python required large-scale changes; it is not merged
-back into CPython due to the pervasive increase in complexity that it
-requires.  In PyPy, though, an optional ``stackless'' transformation is
-able to turn the Python interpreter -- also written in a simple highly
-recursive style -- into an efficient variant of continuation-passing
-style (CPS), enabling the usage of coroutines in the translated
-interpreter.  For more details and other examples of translation-level
-transformations, see \cite{D07.1}.
+A more subtle example of separation of concerns is the way our generated
+implementations can be integrated with a host OO VM.  As mentioned
+above, an implementer deciding to directly target a specific OO VM needs
+not only good knowledge of the OO VM in question and its object model --
+he must fit the language into the imposed models.  Instead, in our
+approach this task is done at two levels: in a first step, a stand-alone
+interpreter is written -- which, if translated to a given OO VM, would
+simply give an interpreter for the dynamic language which is unable to
+communicate with the host VM.  Integration comes as a second step, and
+occurs at a different level, by introducing mappings between the
+relevant classes of the interpreter and the corresponding classes of the
+OO VM.
 
-\subsection{Integration with a host OO VM}
+\subsection{The effort of writing a translation toolchain}
 
-A good example of this is to compare it against the task of hand-writing
-an implementation for a specific, choosen OO VM.  The latter requires
-not only good knowledge of the OO VM in question and its object model --
-it requires the language implementer to fit the language into the
-imposed models.  For example, it is natural to map classes and instances
-of a dynamic object-oriented language to the OO VM's notion of classes
-and instances, but this might not be a simple task at all if the models
-are substantially different and/or if the OO VM is essentially less
-dynamic than the language to implement.  In our approach, this task
-is done at two levels: in a first step, while writing the interpreter,
-the implementer does not need to worry about integration with an OO VM's
-object model.  Of course, the integration effort does not simply vanish
--- indeed, a simple translation of such an interpreter to a given OO VM
-would give an interpreter for the dynamic language which is unable to
-communicate with the host VM (which might already be interesting in
-specific cases, but not in general).  Integration comes as a second
-step, and occurs at a different level, by introducing mappings between
-the relevant classes of the interpreter and the corresponding classes of
-the OO VM.  As of yet we have no evidence that this makes the total
-integration effort much lower or higher; the point is that it has proven
-possible to take the single, OO VM-independent source code of PyPy's
-Python interpreter, and produce from it a version that runs in and
-integrates with the rest of the .NET environment -- by writing
-\emph{orthogonal} code.
+What are the efforts required to develop a translation toolchain capable
+of analyzing and transforming the high-level source code and generating
+lower-level output in various languages?
 
-\subsection{Getting good GCs and tools is possible}
-XXX
+Although it is able to generate, among other things, a complete custom
+VM for C-like environments, we found that the required effort that must
+be put into the translation toolchain was still much lower than that of
+writing a good-quality OO VM.  A reason seems to be that we could design
+our translation toolchain specifically for our needs, i.e. a language
+implementer's needs, instead of for general-purpose usage.  Of course,
+the translation toolchain, once written, can also be reused to implement
+other languages, and possibly tailored on a case-by-case basis to fit
+each implementer's need.  The process is incremental: we can add more
+features as needed instead of starting from a maximal up-front design,
+and gradually improve the quality of the tools, the garbage collectors,
+the various optimizations, etc.
+
+Let us expand on the topic of the garbage collector, which for C-like
+envrionments is inserted into the generated VM by a transformation step.
+We started by ignoring the issue and just using the conservative Boehm
+\cite{Boehm} collector for C.  Later, we experimented with a range of
+simple custom collectors - reference counting, mark-and-sweep, etc.
+Ultimately, though, more advanced GCs will be needed to get the best
+performance.  It seems that RPython, enhanced with support for direct
+address manipulations, is a good language for writing GCs, so it would
+be possible for a GC expert to write one for our translation framework.
+However, this is not the only way to obtain good GCs: we will soon
+investigate a more practical course of action, which is to reuse
+existing GCs.  A good candidate is the GCs written in the Jikes RVM
+\cite{JikesGC}.  As they are in Java, it should be relatively
+straightforward to add a translation step that turns one of them into
+RPython (or directly our RPython-level intermediate representation) and
+integrate it with the rest of the program being translated.
+
+In summary, developing a meta-programming translation toolchain requires
+some work, but it can be done incrementally, it can reuse existing code,
+and it gives a toolchain that is itself highly reusable and flexible in
+nature.
 

### 2007-05-27 talk/dyla2007/dyla.tex
some more references, an XXX, a reformulation

@@
 implemented in C and on top of OO VMs and some of the problems of these
-approaches. In \ref{section sect:metaprogramming} we will describe our proposed
+approaches, using various Python implementations as the main example. In
+\ref{section sect:metaprogramming} we will describe our proposed
 meta-programming approach.
@@
 \label{sect:appraches}
+
+XXX the next paragraph is also the start of the introduction right now!
 
@@
 
-These limitations of a C-based implementation often lead to more
-implementations of the language. For example in addition to the main C
-implementation of Python (``CPython'') there are a number of additional
-implementations that want to fix some of the limitations of CPython. One of
-them is ``Stackless Python'' which is a fork of CPython that adds micro-threading
+These limitations of a C-based implementation often lead to more implementations
+of the language. For example in addition to the main C implementation of Python
+(``CPython'' \cite{cpy251}) there are a number of additional implementations
+that want to fix some of the limitations of CPython. One of them is ``Stackless
+Python'' \cite{stackless} which is a fork of CPython that adds micro-threading
 capabilities to Python. One of the reasons for not incorporating it back into
 CPython was that it was felt that they would make the implementation too
-complex. Another implementation of the Python language is the Psyco project
-which adds a JIT-compiler to CPython. As all other re-implementations, Psyco
-contains the Python semantics and needs to be kept synchronized with CPython
-manually.
+complex. Another implementation of the Python language is the Psyco \cite{Psyco}
+project which adds a JIT-compiler to CPython. As all other re-implementations,
+Psyco contains the Python semantics and needs to be kept synchronized with
+CPython manually.
 
@@
 libraries and tools. Although it is in principle possible to do that by
-bridging the C-VM with the respective VMs of these languages (XXX examples)
+bridging the C-VM with the respective VMs of these languages (for example the
+``PythonNet'' project \cite{pythonnet} that brigdes CPython with a .NET VM)
 this is an incomplete solution since it prevents many use-cases.
@@
 other languages, and possibly tailored on a case-by-case basis to fit
-each implementer's need.  The process is incremental: we can add more
+the specific need of a language.  The process is incremental: we can add more
 features as needed instead of starting from a maximal up-front design,

### 2007-05-27 talk/dyla2007/dyla.tex
A short dynamic compilers section.

@@
 order of magnitude (within a factor of 2 and improving) of the hand-written,
-well-tuned CPython, the C reference implementation.
+well-tuned CPython, the C reference implementation.  These figures do not
+include the spectacular speed-ups obtained in some cases by the JIT compiler
+described in section \ref{subsect:dynamic_compilers}.
 
@@
 
-A more subtle example of separation of concerns is the way our generated
-implementations can be integrated with a host OO VM.  As mentioned
-above, an implementer deciding to directly target a specific OO VM needs
-not only good knowledge of the OO VM in question and its object model --
-he must fit the language into the imposed models.  Instead, in our
-approach this task is done at two levels: in a first step, a stand-alone
-interpreter is written -- which, if translated to a given OO VM, would
-simply give an interpreter for the dynamic language which is unable to
-communicate with the host VM.  Integration comes as a second step, and
-occurs at a different level, by introducing mappings between the
-relevant classes of the interpreter and the corresponding classes of the
-OO VM.
+%A more subtle example of separation of concerns is the way our generated
+%implementations can be integrated with a host OO VM.  As mentioned
+%above, an implementer deciding to directly target a specific OO VM needs
+%not only good knowledge of the OO VM in question and its object model --
+%he must fit the language into the imposed models.  Instead, in our
+%approach this task is done at two levels: in a first step, a stand-alone
+%interpreter is written -- which, if translated to a given OO VM, would
+%simply give an interpreter for the dynamic language which is unable to
+%communicate with the host VM.  Integration comes as a second step, and
+%occurs at a different level, by introducing mappings between the
+%relevant classes of the interpreter and the corresponding classes of the
+%OO VM.
 
@@
 be put into the translation toolchain was still much lower than that of
-writing a good-quality OO VM.  A reason seems to be that we could design
-our translation toolchain specifically for our needs, i.e. a language
-implementer's needs, instead of for general-purpose usage.  Of course,
-the translation toolchain, once written, can also be reused to implement
-other languages, and possibly tailored on a case-by-case basis to fit
-the specific need of a language.  The process is incremental: we can add more
-features as needed instead of starting from a maximal up-front design,
-and gradually improve the quality of the tools, the garbage collectors,
-the various optimizations, etc.
+writing a good-quality OO VM.  A reason is that a translation toolchain
+operates in a more static way, which allows it to leverage good C
+compilers.  It is self-supporting: pieces of the implementation can be
+written in RPython as well and translated along with the rest of the
+RPython source, and they can all be compiled and optimized by the C
+compiler.  In order to write an OO VM in this style you need to start by
+assuming an efficient dynamic compiler.
 
-Let us expand on the topic of the garbage collector, which for C-like
-envrionments is inserted into the generated VM by a transformation step.
-We started by ignoring the issue and just using the conservative Boehm
-\cite{Boehm} collector for C.  Later, we experimented with a range of
-simple custom collectors - reference counting, mark-and-sweep, etc.
-Ultimately, though, more advanced GCs will be needed to get the best
-performance.  It seems that RPython, enhanced with support for direct
-address manipulations, is a good language for writing GCs, so it would
-be possible for a GC expert to write one for our translation framework.
-However, this is not the only way to obtain good GCs: we will soon
-investigate a more practical course of action, which is to reuse
-existing GCs.  A good candidate is the GCs written in the Jikes RVM
-\cite{JikesGC}.  As they are in Java, it should be relatively
-straightforward to add a translation step that turns one of them into
-RPython (or directly our RPython-level intermediate representation) and
-integrate it with the rest of the program being translated.
+Of course, the translation toolchain, once written, can also be reused
+to implement other languages, and possibly tailored on a case-by-case
+basis to fit the specific needs of a language.  The process is
+incremental: we can add more features as needed instead of starting from
+a maximal up-front design, and gradually improve the quality of the
+tools, the garbage collectors, the various optimizations, etc.
+
+Writing a good garbage collector remains hard, though.  At least, it is
+easy to experiment with various kind of GCs, so we started by just using
+the conservative Boehm \cite{Boehm} collector for C and moved up to a
+range of simple custom collectors -- reference counting, mark-and-sweep,
+etc.  Ultimately, though, more advanced GCs will be needed to get the
+best performance.  It seems that RPython, enhanced with support for
+direct address manipulations, is a good language for writing GCs, so it
+would be possible for a GC expert to write one for our translation
+framework.  However, this is not the only way to obtain good GCs:
+existing GCs can also be reused.  Good candidates are the GCs written in
+the Jikes RVM \cite{JikesGC}.  As they are in Java, it should be
+relatively straightforward to add a translation step that turns one of
+them into RPython (or directly into our RPython-level intermediate
+representation) and integrate it with the rest of the program being
+translated.
 
 In summary, developing a meta-programming translation toolchain requires
-some work, but it can be done incrementally, it can reuse existing code,
-and it gives a toolchain that is itself highly reusable and flexible in
+work, but it can be done incrementally, it can reuse existing code, and
+it gives a toolchain that is itself highly reusable and flexible in
 nature.
+
+\subsection{Dynamic compilers}
+\label{subsect:dynamic_compilers}
+
+As mentioned above, the performance of the VMs generated by our
+translation framework are quite acceptable -- e.g. the Python VM
+generated via C code is much faster than Jython running on the best
+JVMs.  Of course, the JIT compilers in these JVMs are essential to
+achieve even this performance, which further proves the point that
+writing good OO VMs -- especially ones meant to support dynamic
+languages -- is a lot of work.
+
+The deeper problem with the otherwise highly-tuned JIT compilers of the
+OO VMs is that they are not a very good match for running dynamic
+languages.  It might be possible to tune a general-purpose JIT compiler
+enough, and write the dynamic language implementation accordingly, so
+that most of the bookkeeping work involved in running the dynamic
+language can be removed -- dispatching, boxing, unboxing...  However
+this has not been demonstrated yet.
+
+By far the fastest Python implementation, Psyco \cite{Psyco} contains a
+hand-written language-specific dynamic compiler.  PyPy's translation
+tool-chain is able to extend the generated VMs with an automatically
+generated dynamic compiler similar to Psyco, derived from the
+interpreter.  This is achieved by a pragmatic application of partial
+evaluation techniques guided by a few hints added to the source of the
+interpreter.  In other words, it is possible to produce a reasonably
+good language-specific JIT compiler and insert it into a VM, alongside
+with the necessary support code and the rest of the regular interpreter.
+
+This result was one of the major goals and motivations for the whole
+approach.  By construction, any code written in the dynamic language
+runs correctly under the JIT.  Some very simple Python examples run more
+than 100 times faster.  At the time of this writing this is still rather
+experimental, and the techniques involved are well beyond the scope of
+the present paper.  The reader is referred to \cite{D08.2} for more
+information.
 

### 2007-05-27 talk/dyla2007/dyla.tex
attempts to make the OO VM (dis)advantages more readable by itemizing them. more references

@@
 
+\begin{itemize}
+\item
 \emph{Better interoperability than the C level:} Since the VM offers a standard
@@
 
+\item
 \emph{Cross-platform portability:} Only the underlying VM has to be ported to
@@
 
+\item
 \emph{Better tools:} Better IDEs, debuggers and profilers.
 
+\item
 \emph{Better implementation of low-level issues like garbage collection,
@@
 
+\item
 \emph{Better performance:} Similarly, object-oriented VMs usually come with a
@@
 
+\item
 \emph{Ease of implementation:} The implementation of a language on top of an OO
@@
 
+\item
 \emph{A single unified implementation base:} The higher level of abstraction
@@
 it is also easier to modify and to extend. XXX somewhat vague point
+\end{itemize}
 
-At a closer look, some of these advantages are only partially true in practice.
+At a closer look, some of these advantages are only partially true in practice:
 
+\begin{itemize}
+\item
 \emph{Better performance:} So far it seems that performance of highly dynamic
-languages is not actually significantly improved on OO VMs. Jython
-\footnote{Python on the Java VM} is around 5
+languages is not actually significantly improved on OO VMs. 
+Jython\footnote{Python on the Java VM} is around 5
 times slower than CPython, for IronPython\footnote{Python on .NET, which
@@
 
+\item
 \emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a much
@@
 
+\item
 \emph{Cross-platform portability:} While this is true to some extend, the
@@
 architecture too or working at all, which leads to significantly less
-performance.
+speed.
 
+\item
 \emph{Ease of implementation:} This point is disputable. On the one hand, OO
@@
 An example where this mapping does not work too well is Prolog. While there
-exist several implementations of Prolog on top of the JVM and also one on .NET,
+exist several implementations of Prolog on top of the JVM \cite{prologcafe}
+\cite{DBLP:conf/jelia/Calejo04} and also one on .NET \cite{psharp},
 they are not particular efficient, especially when compared to good Prolog VMs
@@
 that is quite unnatural both for the OO VM and for Prolog.
+\end{itemize}
 
@@
 prominent being the easy interaction with the rest of the VM. Furthermore there
-is better tool support and better GCs. XXX expand
+is better tool support and better GCs. Also for languages where the execution
+model fits the OO VM well, many of the disadvantages disappear.
 
@@
 
-We implemented this idea in the PyPy project.  The dynamic language for
-which we wrote an interpreter is Python.  It is a language which,
+We implemented this idea in the PyPy project \cite{pypy}.  The dynamic language
+for which we wrote an interpreter is Python.  It is a language which,
 because of its size and rather intricate semantics, is a good target for
-our approach, in the following sense: its previous reimplementation
+our approach, in the following sense: its previous reimplementations
 (Jython for the JVM and IronPython for .NET) have each proved to be very
@@
 interpreter of PyPy and the results we achieved by translating it to C, LLVM
-\cite{LLVM} and .NET are described in [XXX].  These results show that the
+\cite{LLVM} and .NET are described in \cite{architecture} \cite{translationdoc}
+These results show that the
 approach is practical and gives results whose performance is within the same

### 2007-05-27 talk/dyla2007/dyla.tex
Detail.
add a citation of the psyco home page too, differentiate better between citations of the psyco paper and psyco-the-program

@@
 CPython was that it was felt that they would make the implementation too
-complex. Another implementation of the Python language is the Psyco \cite{Psyco}
+complex. Another implementation of the Python language is the Psyco
+\cite{psyco-software}
 project which adds a JIT-compiler to CPython. As all other re-implementations,
@@
 exist several implementations of Prolog on top of the JVM \cite{prologcafe}
-\cite{DBLP:conf/jelia/Calejo04} and also one on .NET \cite{psharp},
+\cite{InterProlog} and also one on .NET \cite{psharp},
 they are not particular efficient, especially when compared to good Prolog VMs
@@
 
-By far the fastest Python implementation, Psyco \cite{Psyco} contains a
+By far the fastest Python implementation, Psyco \cite{psyco-software} contains a
 hand-written language-specific dynamic compiler.  PyPy's translation
 tool-chain is able to extend the generated VMs with an automatically
-generated dynamic compiler similar to Psyco, derived from the
+generated dynamic compiler that uses techniques similar to those of Psyco
+\cite{Psyco-paper}, derived from the
 interpreter.  This is achieved by a pragmatic application of partial
@@
 This result was one of the major goals and motivations for the whole
-approach.  By construction, any code written in the dynamic language
-runs correctly under the JIT.  Some very simple Python examples run more
-than 100 times faster.  At the time of this writing this is still rather
+approach.  By construction, the JIT does not get out of sync when the
+language evolves, and any code written in the dynamic language runs
+correctly under the JIT.  Some very simple Python examples run more than
+100 times faster.  At the time of this writing this is still rather
 experimental, and the techniques involved are well beyond the scope of
@@
 information.
-
-\section{Related Work}
-XXX
 

### 2007-05-27 talk/dyla2007/dyla.tex
Conclusion.

@@
 
+In the sequel, we will focus on the relative advantages and
+inconvenients of the PyPy approach compared to the approach of
+hand-writing a language implementation on top of an OO VM.
+
 \subsection{A single source}
@@
 \section{Conclusion}
-XXX 
+
+Here are the two central points that we have asserted in the present
+paper:
+
+\begin{itemize}
+\item \emph{Do not write dynamic language implementations ``by hand''.}
+Writing them more abstractly, at a higher level, has mostly only
+advantages, among them avoiding a proliferation of implementations
+growing out of sync.  Writing interpreters both flexibly and efficiently
+is difficult, and meta-programming is a good way to achieve it.
+Moreover, this is not incompatible with targetting and benefiting from
+existing good object-oriented virtual machines like the Java and .NET
+ones.
+
+\item \emph{Do not write VMs ``by hand''.}
+Writing language-specific virtual machines is a time-consuming task for
+medium to large languages.  Unless large amounts of resources can be
+invested, the resulting VMs are bound to have limitations which lead to
+the emergence of many implementations, a fact that is taxing precisely
+for a community with limited resources.  (This is of course even more
+true for general-purpose VMs.)
+\end{itemize}
+
+As a better alternative, we advocate a more general usage of
+meta-programming:
+
+\begin{itemize}
+\item \emph{Let's write more meta-programming translation toolchains.}
+Aside from the advantages described in section
+\ref{sect:metaprogramming}, a translation toolchain need not be
+standardized for inter-operability, but can be tailored to the needs of
+each project.  Diversity is good; there is no need to attempt to
+standardize on a single OO VM.
+\end{itemize}
+
+The approach we outlined is actually just one in a very large, mostly
+unexplored design space; it is likely that some of the choices made in
+PyPy will turn out to be suboptimal.  We are hoping that other
+toolchains will emerge over time, exploring other aspects and proposing
+other solutions.  By their ``meta'' nature, these multiple approachs
+should be easier to bridge together than, say, multiple OO VMs with
+different object and runtime models.  We believe that further research
+in this area might open the door to better solutions for
+interoperability in general -- e.g. high-level bridges instead of
+(virtual-)machine-level ones, enabled by cross-translation.
+
+We believe this to be ultimately a better investment of efforts than the
+development of more advanced general-purpose OO VMs.
+
 % ---- Bibliography ----

### 2007-05-27 talk/dyla2007/dyla.tex
Fix an XXX.

@@
 \item
-\emph{A single unified implementation base:} The higher level of abstraction
-mentioned in the last point also helps to reduce the number of implementations
-of the language. Since the implementation on top of an OO VM is more abstract
-it is also easier to modify and to extend. XXX somewhat vague point
+\emph{A single unified implementation base:} The .NET and Java VMs are trying
+to position themselves as all-encompassing platforms; if one succeeds, then
+implementations of the dynamic language for other platforms might no longer
+be required.
 \end{itemize}

### 2007-05-27 talk/dyla2007/dyla.tex
Remove one XXX, add a (less purely aesthetic) one. Progress!
just remove one XXX. we have space problems anyway.
Fix refs.

@@
 approaches, using various Python implementations as the main example. In
-\ref{section sect:metaprogramming} we will describe our proposed
+section \ref{sect:metaprogramming} we will describe our proposed
 meta-programming approach.
@@
 \section{Approaches to Dynamic Language Implementation}
-\label{sect:appraches}
+\label{sect:approaches}
 
@@
 VMs}
-
-XXX mention that OO languages don't have good metaprogramming support
 
@@
 execution model of the underlying VM, which may be easy or not, depending very
-much on the language in question. XXX give example where it works well.
+much on the language in question.
 
@@
 that is quite unnatural both for the OO VM and for Prolog.
+
+Another important point that make implementations of languages on top of OO VMs
+harder is that typically they don't support meta-programming very well, or only
+at the bytecode level.
 \end{itemize}
@@
 
+%XXX doesn't look entirely nice
 \begin{itemize}

### 2007-05-27 talk/dyla2007/dyla.tex
the proper magic to be able to use utf-8

@@
-\documentclass{llncs}
+\documentclass[utf8x]{llncs}
 
@@
 \usepackage{graphicx}
+\usepackage[utf8x]{inputenc}
 
@@
 project, with which we can automatically insert many features into the
-generated virtual machines – including good just-in-time compilers tuned
+generated virtual machines -- including good just-in-time compilers tuned
 to the dynamic language at hand.
@@
 
-For these reasons writing a virtual machine in C has many problems because it
+For these reasons writing a virtual machine in C is problematic because it
 forces the language implementer to deal with many low-level details. Limitations
@@
 given programming language: as the language becomes popular, there will be a
-demand for having it run on various platforms – high-level VMs as well as
+demand for having it run on various platforms -- high-level VMs as well as
 C-level environments.
@@
 translation toolchain we can then generate whole virtual machines from
-this specification – either full custom VMs for C-level operating
+this specification -- either full custom VMs for C-level operating
 systems, or layers on top of various OO VMs.  In other words,

### 2007-05-27 talk/dyla2007/dyla.tex
okok. revert the last one since armin's latex doesn't seem to like it.
a quick try describing PyPy's approach without contrasting it to .NET/CLI too much - rather focusing on presenting PyPy as the next evolutional step to raise the level of language implementations, and still taking advantage of existing new approaches/VMs.

@@
-\documentclass[utf8x]{llncs}
+\documentclass{llncs}
 
@@
 \usepackage{graphicx}
-\usepackage[utf8x]{inputenc}
 
@@
 
-\title{How to Not write Virtual Machines for Dynamic Languages}
+\title{PyPy: raising the level for Dynamic Language implementations}
 
@@
 
-Dynamic languages are traditionally implemented by writing a virtual
-machine for them, centered around an interpreter and/or a built-in
-compiler and providing the object model and memory management. When
-a language becomes more successful, the limitations of such an implementation
-lead to the emergence of alternative implementations that try to solve some
-of the problems. Another reason for new implementations is the desire to have
-the language integrate well with existing, well-tuned object-oriented virtual
-machine like Java's. In this paper, we describe the mechanisms that lead to an
-abundance of implementations, and explore some of the limitations of standard
-VMs.  We propose a different complementary alternative: flexibly generating
-virtual machines instead
-of writing them by hand.  This approach has been validated by the PyPy
-project, with which we can automatically insert many features into the
-generated virtual machines -- including good just-in-time compilers tuned
-to the dynamic language at hand.
+Dynamic languages are usually implemented in lower level environments, 
+traditionally in C/Posix.  Recently .NET and the JVM are emerging as 
+new higher-level environments, offering more built-in support for 
+memory management, threading models, dynamic optimization techniques
+and also offering an abundance of libraries for application development.   
+However, manually implementing languages for a standardized runtime environment
+may not be the last word on bringing dynamic language implementations to a higher level. 
+The PyPy project successfully experimented with an approach that 
+uses extensive meta-programming techniques to analyse and transform the 
+single source of an language "specification" and automatically adds
+lower level and runtime-dependent aspects, successfully producing 
+efficent virtual machines for an runtime environment of choice. 
+The meta-programming facilities are largely re-useable for implementing 
+different computer languages. We thus argue in this paper that 
+one should not write interpreters manually and dependent on lower 
+level and runtime details - but rather explore and exploit 
+meta-programming facilities which themselves can be written 
+and tested in a very high language, overall raising the level 
+of implementing dynamic languages.  While there is considerable 
+complexity liying in such a meta-programming tool chain, it is 
+re-usable for different language implementations and generally
+helps to avoid having to implement language semantics and 
+optimizations redundantly.
+
 \footnote{This research is partially supported by the EU funded %research

### 2007-05-27 talk/dyla2007/dyla.tex
Shrink shrink shrink the start of section 2.  It says less, but I think that the point that we want implementations on top of OO VMs was made in the introduction already.

@@
 problem because it divides available resources. All these points also apply to
-the implementation of domain-specific-languages where it is important to keep
-implementation effort small.
+the implementation of domain-specific languages where it is important to keep
+the implementation effort small.
 
@@
 
-XXX the next paragraph is also the start of the introduction right now!
+\def\implname#1{\emph{#1}}
 
-Dynamic languages are traditionally implemented by writing a virtual
-machine for them in a low-level language like C, or in a language that
-can relatively easily be turned into C.  The machine implements an
-object model supporting the high level dynamic language's objects.  It
-typically provides features like automatic garbage collection.  Recent
-languages like Python, Ruby, Perl and JavaScript have complicated
-semantics which are most easily mapped to a simple interpreter operating
-on syntax trees or bytecode. Simpler languages like Lisp and Self
-typically have more efficient implementations based on just-in-time code
-generation.
+The fact that limitations of a C-based implementation of a dynamic
+language leads to the emergence of additional implementations is clear
+in the case of Python.  The reference implementation, \implname{CPython}
+\cite{cpy251}, is a simple recursive interpreter.  \implname{Stackless
+Python} \cite{stackless} is a fork that adds micro-threading
+capabilities to Python. One of the reasons for not incorporating it back
+into CPython was that it was felt that they would make the
+implementation too complex. Another implementation of the Python
+language is \implname{Psyco} \cite{psyco-software}, which adds a
+JIT-compiler to CPython.  Finally, \implname{Jython} is a
+re-implementation for the Java VM and \implname{IronPython} one for
+.NET.  All of these need to be kept in sync with the relatively fast
+evolution of the language.
 
-There are many good reasons for choosing C as an implementation language. It
-gives a decent base performance, is highly portable across different hardware
-architectures and operating system if some care is taken and it allows the reuse
-of many existing libraries.
-
-On the other hand the use of C comes with a set of problems. Since C has no ways
-to do automatic memory management and since the implemented languages are
-garbage collected, the VM needs to contain a garbage collector. Implementing a
-high-performance GC is a tedious task and so many implementations resort to
-relatively simple algorithms like mark-and-sweep or reference counting. Also the
-choice of garbage collector is fixed relatively early in the life of the project
-and then manifests itself throughout the complete VM source code. Thus it
-becomes extremely tedious to change it later.
-
-%A similar problem is the threading model the VM uses
-%  - simpler threading models
-%    - don't make single-threaded apps slower due to locks
-%    - global locks
-%    - green threading
-
-% - wrapping with C libraries simple but tedious
-
-These limitations of a C-based implementation often lead to more implementations
-of the language. For example in addition to the main C implementation of Python
-(``CPython'' \cite{cpy251}) there are a number of additional implementations
-that want to fix some of the limitations of CPython. One of them is ``Stackless
-Python'' \cite{stackless} which is a fork of CPython that adds micro-threading
-capabilities to Python. One of the reasons for not incorporating it back into
-CPython was that it was felt that they would make the implementation too
-complex. Another implementation of the Python language is the Psyco
-\cite{psyco-software}
-project which adds a JIT-compiler to CPython. As all other re-implementations,
-Psyco contains the Python semantics and needs to be kept synchronized with
-CPython manually.
-
-% - free threading attempts for Python
-
-As a language becomes more popular, yet another reason for different
-implementations emerges: People want to use it together with Java and C\#
-libraries and tools. Although it is in principle possible to do that by
-bridging the C-VM with the respective VMs of these languages (for example the
-``PythonNet'' project \cite{pythonnet} that brigdes CPython with a .NET VM)
-this is an incomplete solution since it prevents many use-cases.
-Therefore in the end implementations of the dynamic language on top of the
-object-oriented VMs are created to properly integrate with the complete
-environment.
+With the emergence of .NET and the JVM as interesting language
+implementation platforms, an argument that is sometimes made is that
+communities should only develop an implementation of their language for
+one of these platforms.\footnote{Preferably the argument author's favourite
+one.}
 
@@
 
-Implementing a language on top of an OO VM is in many ways easier than
+Implementing a language on top of an existing OO VM is in many ways easier than
 implementing it in C. Let's take a look at the advantages that are usually
@@
 languages is not actually significantly improved on OO VMs. 
-Jython\footnote{Python on the Java VM} is around 5
+Jython is around 5
 times slower than CPython, for IronPython\footnote{Python on .NET, which

### 2007-05-27 talk/dyla2007/dyla.tex
try to steal the good ideas of holgers abstract

@@
 
-Dynamic languages are usually implemented in lower level environments, 
-traditionally in C/Posix.  Recently .NET and the JVM are emerging as 
-new higher-level environments, offering more built-in support for 
-memory management, threading models, dynamic optimization techniques
-and also offering an abundance of libraries for application development.   
-However, manually implementing languages for a standardized runtime environment
-may not be the last word on bringing dynamic language implementations to a higher level. 
-The PyPy project successfully experimented with an approach that 
-uses extensive meta-programming techniques to analyse and transform the 
-single source of an language "specification" and automatically adds
-lower level and runtime-dependent aspects, successfully producing 
-efficent virtual machines for an runtime environment of choice. 
-The meta-programming facilities are largely re-useable for implementing 
-different computer languages. We thus argue in this paper that 
-one should not write interpreters manually and dependent on lower 
-level and runtime details - but rather explore and exploit 
-meta-programming facilities which themselves can be written 
-and tested in a very high language, overall raising the level 
-of implementing dynamic languages.  While there is considerable 
-complexity liying in such a meta-programming tool chain, it is 
-re-usable for different language implementations and generally
-helps to avoid having to implement language semantics and 
-optimizations redundantly.
+Dynamic languages are traditionally implemented by writing a virtual machine for
+them, centered around an interpreter and/or a built-in compiler and providing
+the object model and memory management. When a language becomes more successful,
+the limitations of such an implementation lead to the emergence of alternative
+implementations that try to solve some of the problems. Another reason for new
+implementations is the desire to have the language integrate well with existing,
+well-tuned object-oriented virtual machine like Java's. In this paper, we
+describe the mechanisms that lead to an abundance of implementations, and
+explore some of the limitations of standard VMs.  We propose a different
+complementary alternative to writing VMs by hand and dealing with low-level
+details: flexibly generating virtual machines from a single abstract language
+``specification''.
 
-\footnote{This research is partially supported by the EU funded %research
- project:
-IST 004779 PyPy (PyPy: Implementing Python in Python).} \\
+This approach has been validated by the PyPy project that tries to explore
+meta-programming facilities with which we can automatically insert many features
+and low-level details into an automatically generated virtual machines --
+including good just-in-time compilers tuned to the dynamic language at hand.
+We thus argue in this paper that one should not write interpreters manually but
+rather use meta-programming techniques and raise the overall level at which
+dynamic languages are implemented. We believe this to be ultimately a better
+investment of efforts than the development of more advanced general-purpose OO
+VMs. \footnote{This research is partially supported by the EU funded
+ project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\
 

### 2007-05-27 talk/dyla2007/dyla.tex
Swap the two paras of the abstract.  Experimental.

@@
 
+We argue in this paper that one should not write interpreters manually but
+rather use meta-programming techniques and raise the overall level at which
+dynamic languages are implemented. We believe this to be ultimately a better
+investment of efforts than the development of more advanced general-purpose OO
+VMs.
+
 Dynamic languages are traditionally implemented by writing a virtual machine for
@@
 complementary alternative to writing VMs by hand and dealing with low-level
-details: flexibly generating virtual machines from a single abstract language
-``specification''.
-
-This approach has been validated by the PyPy project that tries to explore
-meta-programming facilities with which we can automatically insert many features
-and low-level details into an automatically generated virtual machines --
-including good just-in-time compilers tuned to the dynamic language at hand.
-We thus argue in this paper that one should not write interpreters manually but
-rather use meta-programming techniques and raise the overall level at which
-dynamic languages are implemented. We believe this to be ultimately a better
-investment of efforts than the development of more advanced general-purpose OO
-VMs. \footnote{This research is partially supported by the EU funded
+details, validated by the PyPy project: flexibly generating virtual machines
+from a single abstract language ``specification'', inserting features and
+low-level details automatically -- including good just-in-time compilers tuned
+to the dynamic language at hand.
+\footnote{This research is partially supported by the EU funded
  project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

### 2007-05-27 talk/dyla2007/dyla.tex
slight rewrite of the first paragraph of the abstract. adding a parenthetical comment about GCs.

@@
 
-We argue in this paper that one should not write interpreters manually but
-rather use meta-programming techniques and raise the overall level at which
-dynamic languages are implemented. We believe this to be ultimately a better
-investment of efforts than the development of more advanced general-purpose OO
-VMs.
+We argue in this paper that one should not write interpreters for dynamic
+languages manually but rather use meta-programming techniques and raise the
+overall level at which they are are implemented. We believe this to be
+ultimately a better investment of efforts than the development of more advanced
+general-purpose object oriented VMs.
 
@@
 For these reasons writing a virtual machine in C is problematic because it
-forces the language implementer to deal with many low-level details. Limitations
+forces the language implementer to deal with many low-level details (like
+garbage collection and threading-issues). Limitations
 of the C implementation lead to alternative implementations which draw

### 2007-05-27 talk/dyla2007/dyla.tex
I find "fact" a bit strong here.

@@
 
-The fact that limitations of a C-based implementation of a dynamic
+The observation that limitations of a C-based implementation of a dynamic
 language leads to the emergence of additional implementations is clear

### 2007-05-27 talk/dyla2007/dyla.tex
add two noindents so make the single line between to itemizes stand out more

@@
 
+\noindent
 At a closer look, some of these advantages are only partially true in practice:
@@
 
+\noindent
 As a better alternative, we advocate a more general usage of

### 2007-05-27 talk/dyla2007/dyla.tex
Typo; spacing.

@@
 general-purpose object oriented VMs.
+
+\smallskip
 
@@
 forces the language implementer to deal with many low-level details (like
-garbage collection and threading-issues). Limitations
+garbage collection and threading issues). Limitations
 of the C implementation lead to alternative implementations which draw

### 2007-05-27 talk/dyla2007/dyla.tex
change to the old title but italicize the "not". is -> was in the footnote.

@@
 
-\title{PyPy: raising the level for Dynamic Language implementations}
+\title{How to \emph{not} write Virtual Machines for Dynamic Languages}
 
@@
 to the dynamic language at hand.
-\footnote{This research is partially supported by the EU funded
+\footnote{This research was partially supported by the EU funded
  project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

### 2007-05-27 talk/dyla2007/dyla.tex
Increase the skip.  Seems to have no visible effect when Carl Friedrich compiles the pdf...
53% of a review

@@
 overall level at which they are are implemented. We believe this to be
-ultimately a better investment of efforts than the development of more advanced
+ultimately a better investment of efforts than the development of more and more advanced
 general-purpose object oriented VMs.
 
-\smallskip
+\medskip
 
@@
 them, centered around an interpreter and/or a built-in compiler and providing
-the object model and memory management. When a language becomes more successful,
+the object model and memory management. When a language becomes more popular,
 the limitations of such an implementation lead to the emergence of alternative
@@
 implementations is the desire to have the language integrate well with existing,
-well-tuned object-oriented virtual machine like Java's. In this paper, we
+well-tuned object-oriented virtual machine like the Java Virtual Machine. In this paper, we
 describe the mechanisms that lead to an abundance of implementations, and
@@
 
-The efforts required to build a new virtual machine are relatively
-important.  This is particularly true for languages which are complex
+The effort required to build a new virtual machine are relatively
+large.  This is particularly true for languages which are complex
 and in constant evolution. Language implementation communities from an
@@
 virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
-the community anyway, since it leads to better re-usability of libraries of
+the community anyway, since it leads to the ability to re-use the libraries of
 these platforms. However, if a C implementation existed before the
-implementation of such a VM is started, this enters in conflict with the goal of
+implementation of such a VM is started, this enters into conflict with the goal of
 having to maintain essentially a single, simple enough implementation for a
@@
 implementation platforms, an argument that is sometimes made is that
-communities should only develop an implementation of their language for
-one of these platforms.\footnote{Preferably the argument author's favourite
-one.}
+communities should only develop an implementation of their language
+for one of these platforms (preferably the argument author's favourite
+one).
 
@@
 language implementation of a dynamic language on a standard object oriented
-virtual machine, for example the JVM or the CLR. The central theme of the
-benefits  of OO VMs is the ability to implement certain hard things only once
-and share the benefits between all language implementations on top of the OO VM.
+virtual machine, for example the JVM or the CLR.
 
@@
 high-level language like Java or C\# is used for the language implementation,
-which both offer the language implementer a much higher level of abstraction
+both of which offer the language implementer a much higher level of abstraction
 than when implementing in C.
@@
 \noindent
-At a closer look, some of these advantages are only partially true in practice:
+The central theme of the benefits of OO VMs is the ability to
+implement certain hard things only once and share the benefits between
+all language implementations on top of the OO VM.  At a closer look,
+some of these advantages are only partially true in practice:
 
@@
 reference implementations lead to a lot of duplicated work and division of
-efforts. This is especially true for open source languages which tend to evolve
+effort. This is especially true for open source languages which tend to evolve
 quickly. At any one point in time some of the implementations will lag behind
@@
 
-In the next section we are exploring a different approach for implementing
+In the next section we explore a different approach for implementing
 dynamic languages that we hope is able to solve many of the problems of

### 2007-05-27 talk/dyla2007/dyla.tex
Review by exarkun.  A couple of open issues, intermediate check-in.

@@
 languages manually but rather use meta-programming techniques and raise the
-overall level at which they are are implemented. We believe this to be
+overall level at which they are implemented. We believe this to be
 ultimately a better investment of efforts than the development of more and more advanced
@@
 
-Dynamic languages are traditionally implemented by writing a virtual machine for
-them, centered around an interpreter and/or a built-in compiler and providing
+Dynamic languages are traditionally implemented by writing a virtual machine
+centered around an interpreter and/or a built-in compiler and providing
 the object model and memory management. When a language becomes more popular,
@@
 well-tuned object-oriented virtual machine like the Java Virtual Machine. In this paper, we
-describe the mechanisms that lead to an abundance of implementations, and
-explore some of the limitations of standard VMs.  We propose a different
+describe the mechanisms that lead to an abundance of implementations and
+explore some of the limitations of standard VMs.  We propose a
 complementary alternative to writing VMs by hand and dealing with low-level
@@
 Dynamic languages are traditionally implemented by writing a virtual
-machine for them in a low-level language like C, or in a language that
+machine for them in a low-level language like C or in a language that
 can relatively easily be turned into C.  The machine implements an
@@
 on syntax trees or bytecode; simpler languages like Lisp and Self
-typically have more efficient implementations based on just-in-time code
+typically have more efficient implementations based on code
 generation.
 
-The effort required to build a new virtual machine are relatively
+The effort required to build a new virtual machine is relatively
 large.  This is particularly true for languages which are complex
@@
 open-source or academic context have only limited resources. Therefore they
-cannot afford to have a highly complex implementation and often chose simpler
-techniques even if that entails lower execution speed. Similarly fragmentation
+cannot afford to have a highly complex implementation and often choose simpler
+techniques even if that entails lower execution speed. Similarly, fragmentation
 (for example because of other implementations of the same language) is a
@@
 of the C implementation lead to alternative implementations which draw
-work-power from the reference implementation. An alternative to writing
+resources from the reference implementation. An alternative to writing
 implementations in C is to build them on top of one of the newer object oriented
@@
 given programming language: as the language becomes popular, there will be a
-demand for having it run on various platforms -- high-level VMs as well as
+demand to have it run on various platforms -- high-level VMs as well as
 C-level environments.
 
-The argument we will make in the present paper is that it is possible to
+In this paper, we will argue that it is possible to
 benefit from and integrate with OO VMs while keeping the dynamic
-language implemented by a single, simple source code base.  The idea is
+language implemented with a single, simple source code base.  The idea is
 to write an interpreter for that language in another sufficiently
 high-level but less dynamic language.  This interpreter plays the role
-of a specification for the dynamic language.  With a good enough
+of a specification for the dynamic language.  With a sufficiently capable
 translation toolchain we can then generate whole virtual machines from
-this specification -- either full custom VMs for C-level operating
-systems, or layers on top of various OO VMs.  In other words,
+this specification -- either wholely custom VMs for C-level operating
+systems or VMs layered on top of various OO VMs.  In other words,
 meta-programming techniques can be used to successfully replace a
@@
 
-The argument boils down to: VMs for dynamic languages should not be
-written by hand!  The justification is based on the
-PyPy project, which proves that the approach is
+The crux of the argument is that VMs for dynamic languages should not be
+written by hand!  The PyPy project is the justification,
+proving that the approach is
 feasible in practice.  Just as importantly, it also brings new insights
@@
 capabilities to Python. One of the reasons for not incorporating it back
-into CPython was that it was felt that they would make the
+into CPython was that it was felt that this would make the
 implementation too complex. Another implementation of the Python
@@
 JIT-compiler to CPython.  Finally, \implname{Jython} is a
-re-implementation for the Java VM and \implname{IronPython} one for
-.NET.  All of these need to be kept in sync with the relatively fast
+re-implementation for the Java VM and \implname{IronPython} for
+the CLR.  All of these need to be kept in sync with the relatively fast
 evolution of the language.
 
-With the emergence of .NET and the JVM as interesting language
-implementation platforms, an argument that is sometimes made is that
+With the emergence of the CLR and the JVM as interesting language
+implementation platforms, it is sometimes argued that
 communities should only develop an implementation of their language
@@
 allows reuse of libraries between all the implemented languages. This is
-typically the most important reason for wanting an implementation on the VM in
+typically the most important reason to want an implementation on the VM in
 the first place.
@@
 various hardware architectures and operating systems. The languages implemented
-on top can then be run without change in various environments.
+on top of it can then be run without change in various environments.
 
@@
 highly tuned just-in-time compiler to make them perform well without requiring
-ahead-of-time compilation to machine language. This in addition with the
-previous point leads to much better performance of the languages running on top
+ahead-of-time compilation to machine language. This, in addition to the
+previous point, leads to much better performance of the languages running on top
 of the VM.
@@
 \item
-\emph{A single unified implementation base:} The .NET and Java VMs are trying
-to position themselves as all-encompassing platforms; if one succeeds, then
+\emph{A single unified implementation base:} The CLR and JVM are trying
+to position themselves as all-encompassing platforms; if one succeeds,
 implementations of the dynamic language for other platforms might no longer
@@
 Jython is around 5
-times slower than CPython, for IronPython\footnote{Python on .NET, which
-gives up on some features to improve performance}
+times slower than CPython, for IronPython (which
+gives up on at least one feature -- frame objects -- to improve performance)
 the figures vary but it is mostly
@@
 this is that the VM's JIT compilers are optimized for specific usage patterns
-that are common in the main language of the OO VM. To get good speeds the
+that are common in the primary language of the OO VM. To achieve good speeds, the
 language implementers would have to carefully produce code that matches these
@@
 \item
-\emph{Cross-platform portability:} While this is true to some extend, the
+\emph{Cross-platform portability:} While this is true to some extent, the
 situation with regard to portability is not significantly improved compared to
-e.g.  C/Posix, which is relatively portable too. Also portability sometimes
+e.g.  C/POSIX, which is relatively portable as well. Also, portability sometimes
 comes at the price of performance, because even if the OO VM is running on a
 particular hardware architecture it is not clear that the JIT is tuned for this
-architecture too or working at all, which leads to significantly less
+architecture (or working at all), leading to significantly reduced
 speed.
@@
 \emph{Ease of implementation:} This point is disputable. On the one hand, OO
-VMs typically allow the language implementor to start at a higher level. On the
-other hand they also enforce a specific object and execution model. This means
+VMs typically allow the language implementer to start at a higher level. On the
+other hand, they also enforce a specific object and execution model. This means
 that the concepts of the implemented language need to be mapped to the
-execution model of the underlying VM, which may be easy or not, depending very
+execution model of the underlying VM, which may or may not be easy, depending very
 much on the language in question.
 
-An example where this mapping does not work too well is Prolog. While there
+An example where this mapping does not work very well is Prolog. While there
 exist several implementations of Prolog on top of the JVM \cite{prologcafe}
-\cite{InterProlog} and also one on .NET \cite{psharp},
+\cite{InterProlog} and one on .NET \cite{psharp},
 they are not particular efficient, especially when compared to good Prolog VMs
-in written in C. This is mostly because the Prolog execution model, which
-involves backtracking and deep recursion does not fit the JVM and .NET very
+written in C. This is mostly because the Prolog execution model, which
+involves backtracking and deep recursion, does not fit the JVM and .NET very
 well. Therefore the Prolog implementations on top of OO VMs resort to models
-that is quite unnatural both for the OO VM and for Prolog.
+that are quite unnatural both for the OO VM and for Prolog.
 
-Another important point that make implementations of languages on top of OO VMs
-harder is that typically they don't support meta-programming very well, or only
+Another important point that makes implementation of languages on top of OO VMs
+harder is that typically OO VMs don't support meta-programming very well, or do so only
 at the bytecode level.
@@
 
-On the other hand some of the benefits are real and very useful, the most
-prominent being the easy interaction with the rest of the VM. Furthermore there
-is better tool support and better GCs. Also for languages where the execution
+Nevertheless, some of the benefits are real and very useful, the most
+prominent of which being easy interaction with the rest of the VM. Furthermore, there
+is better tool support and better GCs. Also, for languages where the execution
 model fits the OO VM well, many of the disadvantages disappear.
@@
 
-The described proliferation of language implementations is a big problem for
+The described proliferation of language implementations is a large problem for
 language communities. Although most individual implementations exist for good
 reasons, the sum of all of them and the need to keep them synchronized with the
-reference implementations lead to a lot of duplicated work and division of
-effort. This is especially true for open source languages which tend to evolve
+reference implementations leads to a significant amount of duplicated work and division of
+effort; this is especially true for open source languages which tend to evolve
 quickly. At any one point in time some of the implementations will lag behind
@@
 people propose the solution of standardizing on one particular OO VM to not have
-to maintain implementations for several of them. While this would in theory
-alleviate the problem it is unlikely to happen. On the one hand many political
-issues are involved in such a decision. On the other hand deciding on one single
+to maintain implementations for several of them. While this would, in theory,
+alleviate the problem it is unlikely to happen. On the one hand, many political
+issues are involved in such a decision. On the other hand, deciding on a single
 object and execution model would not be an equally good fit for all languages.
 
-In the next section we explore a different approach for implementing
+In the next section, we explore a different approach for implementing
 dynamic languages that we hope is able to solve many of the problems of
@@
 virtual machine.  It also gives other benefits that we will describe --
-mostly in term of flexibility.  But most importantly, it lets a
+mostly in term of flexibility.  Most importantly, it lets a
 community write a single source implementation of the language, avoiding
-the time-consuming task of keeping multiple ones in sync.  The single
+the time-consuming task of keeping several of them in sync.  The single
 source can be used to generate either custom VMs for C-like
-environments, or interpreters running on top of OO VMs.  It makes it
+environments or interpreters running on top of OO VMs.  This makes it
 practical to experiment with large changes to the language and with
-entirely new languages, like domain-specific languages, while at any
+entirely new languages, such as domain-specific languages, while at any
 time being able to run the implemented language in a variety of
@@
 because of its size and rather intricate semantics, is a good target for
-our approach, in the following sense: its previous reimplementations
+our approach in the following sense: its previous reimplementations
 (Jython for the JVM and IronPython for .NET) have each proved to be very
-time-consuming to maintain.  Our implementation is by construction
-easier to maintain, and extremely portable (including to C/Posix, to the
+time-consuming to maintain.  Our implementation is, by construction,
+easier to maintain and extremely portable (including to C/Posix, to the
 JVM and to .NET).
@@
 \item
-we use a very expressive \emph{object language} (RPython -- an analyzable
+We use a very expressive \emph{object language} (RPython -- an analyzable
 subset of Python) as the language in which the complete Python
@@
 \item
-we use a very expressive metalanguage (namely regular Python) to
+We use a very expressive metalanguage (namely regular Python) to
 perform the analysis of RPython code (control flow and data flow
@@
 \item
-this meta-programming component of PyPy is called the \emph{translation
+This meta-programming component of PyPy is called the \emph{translation
 framework}, as it translates RPython source code (i.e. the full Python
@@
 runtime environment.  This either turns the interpreter into a
-standalone virtual machine, or integrates it into an existing OO VM.
+standalone virtual machine or integrates it into an existing OO VM.
 The necessary support code -- e.g. the garbage collector when
@@
 
-In the sequel, we will focus on the relative advantages and
-inconvenients of the PyPy approach compared to the approach of
+In the sequel (??? - what is the sequel), we will focus on the relative advantages and
+inconveniences of the PyPy approach compared to the approach of
 hand-writing a language implementation on top of an OO VM.
@@
 leads to language implementations that have various advantages over the
-``hand-written'' implementations.  First of all, it is a single-source
+``hand-written'' implementations.  Firstly, it is a single-source
 approach -- we explicitly seek to solve the problem of proliferation of
-implementations.  In the sequel, we will show that this goal can be
-achieved without giving up on the advantages of hand-written
-implementations for OO VMs.  Moreover, there are additional advantages
--- in our opinion significant enough to hint that meta-programming,
+implementations.  In the sequel (???), we will show that this goal can be
+achieved without giving up the advantages of hand-written
+implementations for OO VMs.  Moreover, there are additional advantages which,
+in our opinion, are significant enough to hint that meta-programming,
 though not widely used in general-purpose programming, is an essential
@@
 Python interpreter is written in RPython as a relatively simple
-interpreter, in some respects easier to understand than CPython.  We are
+interpreter and is, in some respects, easier to understand than CPython.  We are
 using its high level and flexibility to quickly experiment with features
@@
 
-Although it is able to generate, among other things, a complete custom
+Although it is able to generate, among other things, a complete, custom
 VM for C-like environments, we found that the required effort that must
@@
 them into RPython (or directly into our RPython-level intermediate
-representation) and integrate it with the rest of the program being
+representation) and integrates it with the rest of the program being
 translated.
@@
 work, but it can be done incrementally, it can reuse existing code, and
-it gives a toolchain that is itself highly reusable and flexible in
+it results in a toolchain that is itself highly reusable and flexible in
 nature.
@@
 languages.  It might be possible to tune a general-purpose JIT compiler
-enough, and write the dynamic language implementation accordingly, so
+enough and write the dynamic language implementation accordingly so
 that most of the bookkeeping work involved in running the dynamic
@@
 
-By far the fastest Python implementation, Psyco \cite{psyco-software} contains a
+By far the fastest Python implementation, Psyco \cite{psyco-software}, contains a
 hand-written language-specific dynamic compiler.  PyPy's translation
@@
 \item \emph{Do not write dynamic language implementations ``by hand''.}
-Writing them more abstractly, at a higher level, has mostly only
-advantages, among them avoiding a proliferation of implementations
+Writing them more abstractly, at a higher level, has primarily only
+advantages, among them the avoidance of a proliferation of implementations
 growing out of sync.  Writing interpreters both flexibly and efficiently
-is difficult, and meta-programming is a good way to achieve it.
+is difficult and meta-programming is a good way to achieve it.
 Moreover, this is not incompatible with targetting and benefiting from
-existing good object-oriented virtual machines like the Java and .NET
-ones.
+existing high-quality object-oriented virtual machines like those of the Java and .NET.
 
@@
 \ref{sect:metaprogramming}, a translation toolchain need not be
-standardized for inter-operability, but can be tailored to the needs of
+standardized for inter-operability but can be tailored to the needs of
 each project.  Diversity is good; there is no need to attempt to

### 2007-05-27 talk/dyla2007/dyla.tex
spell check and small remaining fixes

@@
 translation toolchain we can then generate whole virtual machines from
-this specification -- either wholely custom VMs for C-level operating
+this specification -- either wholly custom VMs for C-level operating
 systems or VMs layered on top of various OO VMs.  In other words,
@@
 time being able to run the implemented language in a variety of
-environments, from C/Posix to the JVM to .NET.
+environments, from C/POSIX to the JVM to .NET.
 
@@
 time-consuming to maintain.  Our implementation is, by construction,
-easier to maintain and extremely portable (including to C/Posix, to the
+easier to maintain and extremely portable (including to C/POSIX, to the
 JVM and to .NET).
@@
 interpreter of PyPy and the results we achieved by translating it to C, LLVM
-\cite{LLVM} and .NET are described in \cite{architecture} \cite{translationdoc}
+\cite{LLVM} and .NET are described in \cite{architecture} \cite{translationdoc}.
 These results show that the
@@
 is difficult and meta-programming is a good way to achieve it.
-Moreover, this is not incompatible with targetting and benefiting from
+Moreover, this is not incompatible with targeting and benefiting from
 existing high-quality object-oriented virtual machines like those of the Java and .NET.
@@
 toolchains will emerge over time, exploring other aspects and proposing
-other solutions.  By their ``meta'' nature, these multiple approachs
+other solutions.  By their ``meta'' nature, these multiple approaches
 should be easier to bridge together than, say, multiple OO VMs with

### 2007-05-27 talk/dyla2007/dyla.tex
Small fixes.

@@
 this specification -- either wholly custom VMs for C-level operating
-systems or VMs layered on top of various OO VMs.  In other words,
+systems or as a layer on top of various OO VMs.  In other words,
 meta-programming techniques can be used to successfully replace a
@@
 
-In the sequel (??? - what is the sequel), we will focus on the relative advantages and
+In the sequel, we will focus on the relative advantages and
 inconveniences of the PyPy approach compared to the approach of
@@
 approach -- we explicitly seek to solve the problem of proliferation of
-implementations.  In the sequel (???), we will show that this goal can be
+implementations.  In the sequel, we will show that this goal can be
 achieved without giving up the advantages of hand-written

### 2007-05-27 talk/dyla2007/dyla.tex
Define the abbreviation "VM".

@@
 ultimately a better investment of efforts than the development of more and more advanced
-general-purpose object oriented VMs.
+general-purpose object oriented virtual machines (VMs).
 

### 2007-05-27 talk/dyla2007/dyla.tex
Avoid "in sync" and "out of sync".

@@
 re-implementation for the Java VM and \implname{IronPython} for
-the CLR.  All of these need to be kept in sync with the relatively fast
+the CLR.  All of these need to be kept synchronized with the relatively fast
 evolution of the language.
@@
 community write a single source implementation of the language, avoiding
-the time-consuming task of keeping several of them in sync.  The single
+the time-consuming task of keeping several of them synchronized.  The single
 source can be used to generate either custom VMs for C-like
@@
 This result was one of the major goals and motivations for the whole
-approach.  By construction, the JIT does not get out of sync when the
-language evolves, and any code written in the dynamic language runs
+approach.  By construction, the JIT stays synchronized with its VM
+and with the language when it evolves,
+and any code written in the dynamic language runs
 correctly under the JIT.  Some very simple Python examples run more than
@@
 Writing them more abstractly, at a higher level, has primarily only
-advantages, among them the avoidance of a proliferation of implementations
-growing out of sync.  Writing interpreters both flexibly and efficiently
+advantages, among them the avoidance of a proliferation of diverging
+implementations.  Writing interpreters both flexibly and efficiently
 is difficult and meta-programming is a good way to achieve it.

### 2007-07-12 talk/dyla2007/dyla.tex
Start to answer to the reviewer's comments.

@@
 The crux of the argument is that VMs for dynamic languages should not be
-written by hand!  The PyPy project is the justification,
+written by hand!  The PyPy project \cite{pypy} is the justification,
 proving that the approach is
@@
 beyond the state of the art.
+
+PyPy contains a Python interpreter implemented in Python, from which
+Python VMs can be generated.  The reader is referred to
+\cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
+again that the argument we make in the present paper is \emph{not} that
+VMs for dynamic languages should be written in their own host language
+(as many projects like Squeak \cite{Squeak} and others have done) but
+instead that VMs should not be \emph{written} in the first place -- they
+should be generated from simple interpreters written in any suitable
+high-level language.
 
@@
 Jython is around 5
-times slower than CPython, for IronPython (which
+times slower than CPython; for IronPython (which
 gives up on at least one feature -- frame objects -- to improve performance)
@@
 \item
-\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a much
-higher memory overhead to start with (XXX ref)
+\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a
+higher memory overhead to start with.  For example, an instance of Sun's
+Java VM which just loaded Jython consumes XXX MB of non-shared memory
+(XXX for the JVM and an additional XXX after loading Jython), while a
+CPython process fits in XXX MB.
 

### 2007-07-12 talk/dyla2007/dyla.tex
Attempt...

@@
 should be generated from simple interpreters written in any suitable
-high-level language.
+high-level\footnote{``High-level'' is taken by opposition to languages
+like PreScheme \cite{kelsey-prescheme} which use the syntax and
+metaprogramming facilities of a high-level language but encode
+low-level details like memory management.} language.
 

### 2007-07-12 talk/dyla2007/dyla.tex
add slang too

--- 2007-07-12 talk/dyla2007/dyla.tex
@@
 should be generated from simple interpreters written in any suitable
-high-level language.
+high-level\footnote{``High-level'' is taken by opposition to languages
+like PreScheme \cite{kelsey-prescheme} or the subset of Smalltalk that the
+Squeak VM is written in \cite{Squeak} which use the syntax and
+metaprogramming facilities of a high-level language but encode
+low-level details like memory management.} language.
 

### 2007-07-15 talk/dyla2007/dyla.tex
More clarifications based on reviewers' feedback.

--- 2007-07-12 talk/dyla2007/dyla.tex
@@
 \usepackage{graphicx}
+\sloppy
 
@@
 languages like Python, Ruby, Perl and JavaScript have complicated
-semantics which are most easily mapped to a simple interpreter operating
-on syntax trees or bytecode; simpler languages like Lisp and Self
-typically have more efficient implementations based on code
-generation.
+semantics which are most easily mapped to a naive interpreter operating
+on syntax trees or bytecode; simpler languages\footnote
+{
+In the sense of the primitive semantics.  ``Simple'' here is
+as opposed to ``complicated'', not as opposed to ``complex'': Common
+Lisp for example is not a small language, but it can at least in theory
+be expressed from a smaller core of primitives.  In Python, all
+primitive operations have complicated semantics.  The argument developed
+in the present paper is more relevant to ``complicated'' dynamic languages.
+}
+like Lisp, Smalltalk and Self typically have more
+efficient implementations based on code generation.
 
@@
 should be generated from simple interpreters written in any suitable
-high-level language.
+high-level\footnote{``High-level'' is taken by opposition to languages
+like Scheme48's PreScheme \cite{kelsey-prescheme} or Squeak's \cite{Squeak}
+SLang which use the syntax and
+metaprogramming facilities of a high-level language but encode
+low-level details like object layout and memory management.} language.
 
@@
 higher memory overhead to start with.  For example, an instance of Sun's
-Java VM which just loaded Jython consumes XXX MB of non-shared memory
-(XXX for the JVM and an additional XXX after loading Jython), while a
-CPython process fits in XXX MB.
+Java VM which just loaded Jython consumes 34-42 MB of memory, while a
+CPython process fits in 3-4 MB.
 
@@
 Another important point that makes implementation of languages on top of OO VMs
-harder is that typically OO VMs don't support meta-programming very well, or do so only
-at the bytecode level.
+harder is that typically general-purpose OO VMs don't support meta-programming
+very well, or do so only at the bytecode level.
 \end{itemize}
@@
 language can be removed -- dispatching, boxing, unboxing...  However
-this has not been demonstrated yet.
+this has not been demonstrated yet.\footnote
+{Still in the draft stage, a proposed
+extension to the Java bytecode \cite{invokedynamic} might help achieve
+better integration between the Java JITs and dynamic language
+implementations running on top of JVMs.}
 
-By far the fastest Python implementation, Psyco \cite{psyco-software}, contains a
-hand-written language-specific dynamic compiler.  PyPy's translation
-tool-chain is able to extend the generated VMs with an automatically
-generated dynamic compiler that uses techniques similar to those of Psyco
-\cite{Psyco-paper}, derived from the
-interpreter.  This is achieved by a pragmatic application of partial
+By far the fastest Python implementation, Psyco \cite{psyco-software}
+contains a hand-written language-specific dynamic compiler.  It works by
+specializing (parts of) Python functions by feeding runtime information
+back into the compiler (typically, but not exclusively, object types).
+The reader is referred to \cite{Psyco-paper} for more details.
+
+PyPy abstracts on this approach: its translation tool-chain is able to
+extend the generated VMs with an \emph{automatically generated} dynamic
+compiler that uses techniques similar to those of Psyco, derived from
+the interpreter.  This is achieved by a pragmatic application of partial
 evaluation techniques guided by a few hints added to the source of the
@@
 \begin{itemize}
-\item \emph{Do not write dynamic language implementations ``by hand''.}
-Writing them more abstractly, at a higher level, has primarily only
-advantages, among them the avoidance of a proliferation of diverging
-implementations.  Writing interpreters both flexibly and efficiently
-is difficult and meta-programming is a good way to achieve it.
+\item \emph{High-level languages are suitable to implement dynamic languages.}
+They allow an interpreter to be written more abstractly, which has many
+advantages -- among them the avoidance of a proliferation of diverging
+implementations, and better ways to combine flexibility with efficiency.
 Moreover, this is not incompatible with targeting and benefiting from
-existing high-quality object-oriented virtual machines like those of the Java and .NET.
+existing high-quality object-oriented virtual machines like those of the
+Java and .NET.
 
 \item \emph{Do not write VMs ``by hand''.}
+In other words, write an \emph{interpreter} but not a
+\emph{virtual machine} for the language.
 Writing language-specific virtual machines is a time-consuming task for
@@
 standardized for inter-operability but can be tailored to the needs of
-each project.  Diversity is good; there is no need to attempt to
-standardize on a single OO VM.
+each project.  Diversity is good; translation toolchains offset the need
+to attempt to standardize on a single OO VM.
 \end{itemize}

### 2007-07-15 talk/dyla2007/dyla.tex
Shorten the abstract, trying to match more closely the content of the paper.

@@
 
-We argue in this paper that one should not write interpreters for dynamic
-languages manually but rather use meta-programming techniques and raise the
-overall level at which they are implemented. We believe this to be
-ultimately a better investment of efforts than the development of more and more advanced
-general-purpose object oriented virtual machines (VMs).
-
-\medskip
-
-Dynamic languages are traditionally implemented by writing a virtual machine
-centered around an interpreter and/or a built-in compiler and providing
-the object model and memory management. When a language becomes more popular,
-the limitations of such an implementation lead to the emergence of alternative
-implementations that try to solve some of the problems. Another reason for new
-implementations is the desire to have the language integrate well with existing,
-well-tuned object-oriented virtual machine like the Java Virtual Machine. In this paper, we
-describe the mechanisms that lead to an abundance of implementations and
-explore some of the limitations of standard VMs.  We propose a
-complementary alternative to writing VMs by hand and dealing with low-level
-details, validated by the PyPy project: flexibly generating virtual machines
-from a single abstract language ``specification'', inserting features and
-low-level details automatically -- including good just-in-time compilers tuned
-to the dynamic language at hand.
+Typical modern dynamic languages have a growing number of
+implementations.  We explore the reasons for this situation, and the
+limitations it imposes on open source or academic communities that lack
+the resources to fine-tune and maintain them all.  It is sometimes
+proposed that implementing dynamic languages on top of a standardized
+general-purpose object-oriented virtual machine (like Java or .NET)
+would help reduce this burden.  We propose a complementary alternative
+to writing custom virtual machine (VMs) by hand, validated by the PyPy
+project: flexibly generating VMs from a high-level ``specification'',
+inserting features and low-level details automatically -- including good
+just-in-time compilers tuned to the dynamic language at hand.  We
+believe this to be ultimately a better investment of efforts than the
+development of more and more advanced general-purpose object oriented
+VMs.  In this paper we compare these two approaches in detail.
 \footnote{This research was partially supported by the EU funded
@@
 implementations in C is to build them on top of one of the newer object oriented
-virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
+virtual machines (``OO VM'') such as the JVM (Java Virtual Machine) or the
+CLR (Common Language Runtime of the .NET framework). This is often wanted by
 the community anyway, since it leads to the ability to re-use the libraries of

### 2007-07-15 talk/dyla2007/dyla.tex
More reviewer comments.  Check in new pdf.

@@
 Python} \cite{stackless} is a fork that adds micro-threading
-capabilities to Python. One of the reasons for not incorporating it back
+capabilities to CPython. One of the reasons for not incorporating it back
 into CPython was that it was felt that this would make the
 implementation too complex. Another implementation of the Python
-language is \implname{Psyco} \cite{psyco-software}, which adds a
-JIT-compiler to CPython.  Finally, \implname{Jython} is a
+language is \implname{Psyco} \cite{psyco-software}, an extension of
+CPython which adds a JIT-compiler.  Finally, \implname{Jython} is a
 re-implementation for the Java VM and \implname{IronPython} for
-the CLR.  All of these need to be kept synchronized with the relatively fast
-evolution of the language.
+the CLR.  All of these ultimately need to be kept synchronized with the
+relatively fast evolution of the language.
 
@@
 \item
-\emph{Better tools:} Better IDEs, debuggers and profilers.
+\emph{Better tools:} Better and cross-language IDEs, debuggers and profilers.
 
@@
 all language implementations on top of the OO VM.  At a closer look,
-some of these advantages are only partially true in practice:
+some of these advantages are not quite true in practice:
 
@@
 execution model of the underlying VM, which may or may not be easy, depending very
-much on the language in question.
+much on the language in question.  In many cases, the mismatch between
+the abstractions provided by the underlying VM and the semantics of the
+dynamic language is too deep to allow a natural and/or efficient
+implementation.
 
@@
 targeting C -- is itself written in RPython in much the same spirit
-that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
+that the Jikes RVM's GCs are written in Java \cite{JikesGC}; if needed, it is
 translated together with the interpreter to form the final custom VM.
@@
 existing GCs can also be reused.  Good candidates are the GCs written in
-the Jikes RVM \cite{JikesGC}.  As they are in Java, it should be
-relatively straightforward to add a translation step that turns one of
-them into RPython (or directly into our RPython-level intermediate
-representation) and integrates it with the rest of the program being
-translated.
+the Jikes RVM \cite{JikesGC}.  As they are written in Java, it should be
+relatively straightforward to design a translation step that turns them
+into RPython (or directly into our RPython-level intermediate
+representation).  Our translation toolchain could then integrate any one
+of these GCs into the VMs it generates.
 

### 2007-07-15 talk/dyla2007/dyla.tex
Some more progress.

@@
 Dynamic languages are traditionally implemented by writing a virtual
-machine for them in a low-level language like C or in a language that
-can relatively easily be turned into C.  The machine implements an
+machine (VM) for them in a low-level language like C or in a language that
+can relatively easily be turned into C.  The VM implements an
 object model supporting the high level dynamic language's objects.  It
@@
 {
-In the sense of the primitive semantics.  ``Simple'' here is
-as opposed to ``complicated'', not as opposed to ``complex'': Common
-Lisp for example is not a small language, but it can at least in theory
-be expressed from a smaller core of primitives.  In Python, all
-primitive operations have complicated semantics.  The argument developed
-in the present paper is more relevant to ``complicated'' dynamic languages.
+In the sense of the primitive semantics.  For example, in Python most
+primitive operations have complicated semantics; by contrast, in Common
+Lisp complex features like the reader and printer can in theory be
+implemented in terms of simpler primitives as library code.
 }
@@
 The effort required to build a new virtual machine is relatively
-large.  This is particularly true for languages which are complex
+large.  This is particularly true for languages which are complicated
 and in constant evolution. Language implementation communities from an
@@
 forces the language implementer to deal with many low-level details (like
-garbage collection and threading issues). Limitations
-of the C implementation lead to alternative implementations which draw
+garbage collection and threading issues). If a language becomes popular,
+limitations of the C implementation eventually
+lead to alternative implementations which draw
 resources from the reference implementation. An alternative to writing
-implementations in C is to build them on top of one of the newer object oriented
+implementations in C is to build them on top of one of the newer
+general-purpose object-oriented
 virtual machines (``OO VM'') such as the JVM (Java Virtual Machine) or the
@@
 benefit from and integrate with OO VMs while keeping the dynamic
-language implemented with a single, simple source code base.  The idea is
+language implemented as a single, simple source code base.  The idea is
 to write an interpreter for that language in another sufficiently
@@
 meta-programming techniques can be used to successfully replace a
-foreseeable one-VM-fits-all standardization attempt.
+foreseeable one-OO-VM-fits-all standardization attempt.
 
@@
 \cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
-again that the argument we make in the present paper is \emph{not} that
+that the argument we make in the present paper is \emph{not} that
 VMs for dynamic languages should be written in their own host language
@@
 like Scheme48's PreScheme \cite{kelsey-prescheme} or Squeak's \cite{Squeak}
-SLang which use the syntax and
+SLang, which use the syntax and
 metaprogramming facilities of a high-level language but encode
@@
 section \ref{sect:metaprogramming} we will describe our proposed
-meta-programming approach.
+meta-programming approach and compare the two solutions.  We summarize
+our position and conclude in section \ref{sect:conclusion}.
 
@@
 
-The observation that limitations of a C-based implementation of a dynamic
-language leads to the emergence of additional implementations is clear
+Limitations of a C-based implementation of a dynamic
+language lead to the emergence of additional implementations --
+this observation is clear
 in the case of Python.  The reference implementation, \implname{CPython}
@@
 CPython which adds a JIT-compiler.  Finally, \implname{Jython} is a
-re-implementation for the Java VM and \implname{IronPython} for
+re-implementation for the Java VM and \implname{IronPython}
+a re-implementation for
 the CLR.  All of these ultimately need to be kept synchronized with the
@@
 language implementation of a dynamic language on a standard object oriented
-virtual machine, for example the JVM or the CLR.
+virtual machine.
 
@@
 \section{Conclusion}
+\label{sect:conclusion}
 

### 2007-07-15 talk/dyla2007/dyla.tex
Final re-read of the paper, part 1.  More small fixes.

@@
 higher memory overhead to start with.  For example, an instance of Sun's
-Java VM which just loaded Jython consumes 34-42 MB of memory, while a
-CPython process fits in 3-4 MB.
+JVM which just loaded Jython consumes between 34 and 42 MB of memory
+(Linux/IA32), while on the same machine a CPython process fits in 3 to 4
+MB.
 
@@
 \cite{InterProlog} and one on .NET \cite{psharp},
-they are not particular efficient, especially when compared to good Prolog VMs
+they are not particularly efficient, especially when compared to good Prolog VMs
 written in C. This is mostly because the Prolog execution model, which
@@
 reasons, the sum of all of them and the need to keep them synchronized with the
-reference implementations leads to a significant amount of duplicated work and division of
-effort; this is especially true for open source languages which tend to evolve
-quickly. At any one point in time some of the implementations will lag behind
-which makes writing code which can work on all of the implementations harder.
+reference implementation leads to a significant amount of duplicated work and division of
+effort; this is especially true for open source languages, which tend to evolve
+quickly. At any one point in time some of the implementations will lag behind;
+for developers, this makes writing code which can work on all of the
+implementations harder.
 
 Implementing a language on top of a OO VM has many advantages, so some
-people propose the solution of standardizing on one particular OO VM to not have
-to maintain implementations for several of them. While this would, in theory,
+people propose the solution of standardizing on one particular OO VM in order
+to not have
+to maintain implementations for several of them. While this would in theory
 alleviate the problem it is unlikely to happen. On the one hand, many political
@@
 In the next section, we explore a different approach for implementing
-dynamic languages that we hope is able to solve many of the problems of
-implementing a language, in particular the problem of an explosion of the number
+dynamic languages that we believe is able to solve many of the problems that
+implementers face, including the issue of the explosion of the number
 of implementations.
@@
 not be written by hand, but generated automatically ``around'' an
-interpreter playing the role of a high-level description of the language.  We
+interpreter\footnote
+{
+We do not explore the possibility of starting from even
+higher-level descriptions like declarative semantics; while interesting in
+theory, there are many open issues that make it impractical so far, from
+the question of whether it is feasible at all for large dynamic languages
+to the issue of how to compile such a description to efficient code.
+}
+playing the role of a high-level description of the language.  We
 argue that this approach gives many of the benefits usually expected by
@@
 
-In meta-programming terms, the PyPy architecture is as follows:
+In the terminology of meta-programming, the PyPy architecture is as follows:
 
@@
 interpreter is mostly free of low-level details -- no explicit memory
-management, no pieces of C or C-level code.
+management, no pieces of C or C-level code.  Any RPython code is also
+valid Python code, so for testing and bootstrapping it can be run
+directly on top of another Python implementation like CPython.
 
@@
 targeting C -- is itself written in RPython in much the same spirit
-that the Jikes RVM's GCs are written in Java \cite{JikesGC}; if needed, it is
+that the Jikes RVM's GCs are written in Java \cite{JikesGC}; this
+support code is
 translated together with the interpreter to form the final custom VM.
@@
 well-tuned CPython, the C reference implementation.  These figures do not
-include the spectacular speed-ups obtained in some cases by the JIT compiler
+include the spectacular speed-ups that PyPy obtains in some cases using the
+JIT compiler generator
 described in section \ref{subsect:dynamic_compilers}.

### 2007-07-15 talk/dyla2007/dyla.tex
My final review, 2nd half.

@@
 In the sequel, we will focus on the relative advantages and
-inconveniences of the PyPy approach compared to the approach of
+inconveniences of the PyPy approach compared to
 hand-writing a language implementation on top of an OO VM.
@@
 Of course, the translation toolchain, once written, can also be reused
-to implement other languages, and possibly tailored on a case-by-case
+to implement other languages, and tailored on a case-by-case
 basis to fit the specific needs of a language.  The process is
@@
 As mentioned above, the performance of the VMs generated by our
-translation framework are quite acceptable -- e.g. the Python VM
+translation framework is quite acceptable -- e.g. the Python VM
 generated via C code is much faster than Jython running on the best
-JVMs.  Of course, the JIT compilers in these JVMs are essential to
+JVMs.  Moreover, the JIT compilers in these JVMs are essential to
 achieve even this performance, which further proves the point that
-writing good OO VMs -- especially ones meant to support dynamic
+writing good general-purpose OO VMs -- especially ones meant to support dynamic
 languages -- is a lot of work.
@@
 The deeper problem with the otherwise highly-tuned JIT compilers of the
-OO VMs is that they are not a very good match for running dynamic
+OO VMs is that they are not a very good match for running arbitrary dynamic
 languages.  It might be possible to tune a general-purpose JIT compiler
@@
 extension to the Java bytecode \cite{invokedynamic} might help achieve
-better integration between the Java JITs and dynamic language
-implementations running on top of JVMs.}
+better integration between the Java JITs and some class of dynamic languages
+running on top of JVMs.}
 
@@
 contains a hand-written language-specific dynamic compiler.  It works by
-specializing (parts of) Python functions by feeding runtime information
-back into the compiler (typically, but not exclusively, object types).
+specializing (parts of) Python functions based on runtime information
+fed back into the compiler (typically, but not exclusively, object types).
 The reader is referred to \cite{Psyco-paper} for more details.
@@
 interpreter.  In other words, it is possible to produce a reasonably
-good language-specific JIT compiler and insert it into a VM, alongside
+good language-specific JIT compiler and insert it into a VM, along
 with the necessary support code and the rest of the regular interpreter.
@@
 approach.  By construction, the JIT stays synchronized with its VM
-and with the language when it evolves,
-and any code written in the dynamic language runs
-correctly under the JIT.  Some very simple Python examples run more than
+and with the language when it evolves.  Also by construction, the JIT
+immediately supports (and is correct for) arbitrary input code.
+Some very simple Python examples run more than
 100 times faster.  At the time of this writing this is still rather
@@
 They allow an interpreter to be written more abstractly, which has many
-advantages -- among them the avoidance of a proliferation of diverging
-implementations, and better ways to combine flexibility with efficiency.
+advantages.  Among these, it avoids the proliferation of diverging
+implementations, and gives implementers better ways to combine flexibility
+with efficiency.
 Moreover, this is not incompatible with targeting and benefiting from
@@
 the emergence of many implementations, a fact that is taxing precisely
-for a community with limited resources.  (This is of course even more
-true for general-purpose VMs.)
+for a community with limited resources.  This is of course even more
+true for VMs that are meant to be general-purpose.
 \end{itemize}
@@
 standardized for inter-operability but can be tailored to the needs of
-each project.  Diversity is good; translation toolchains offset the need
-to attempt to standardize on a single OO VM.
+each project.
+\item \emph{Diversity is good.}  Meta-programming translation
+toolchains offset the need for standardization of general-purpose OO VMs.
 \end{itemize}

### 2007-07-23 talk/dyla2007/talk.tex
empty beamer talk

--- 2007-05-01 talk/badhonnef2007/talk.tex
@@
-\documentclass[utf8]{beamer}
-
-% This file is a solution template for:
-
-% - Talk at a conference/colloquium.
-% - Talk length is about 20min.
-% - Style is ornate.
+\documentclass[utf8x]{beamer}
 
@@
   \usetheme{Warsaw}
-  % or ...
 
-  \setbeamercovered{transparent}
-  % or whatever (possibly just delete it)
+  %\setbeamercovered{transparent}
 }
@@
 
-\usepackage[utf8]{inputenc}
-% or whatever
+\usepackage[utf8x]{inputenc}
 
@@
 \usepackage[T1]{fontenc}
-% Or whatever. Note that the encoding and the font should match. If T1
-% does not look nice, try deleting the line with the fontenc.
 
-\title{A flexible Prolog interpreter in Python}
+\title{How to \emph{not} write Virtual Machines for Dynamic Languages}
 
-\author{Carl Friedrich Bolz}
-% - Give the names in the same order as the appear in the paper.
-% - Use the \inst{?} command only if the authors have different
-%   affiliation.
+\author{Carl Friedrich Bolz and Armin Rigo}
 
@@
 
-\date{24. Workshop der GI-Fachgruppe Programmiersprachen und Rechenkonzepte, 4. Mai 2007}
-% - Either use conference name or its abbreviation.
-% - Not really informative to the audience, more for people (including
-%   yourself) who are reading the slides online
-
-% If you have a file called "university-logo-filename.xxx", where xxx
-% is a graphic format that can be processed by latex or pdflatex,
-% resp., then you can add a logo as follows:
+\date{3rd Workshop on Dynamic Languages and Applications, July 31 2007}
 
@@
 
-\begin{frame}
-  \frametitle{Outline}
-  \tableofcontents
+%\begin{frame}
+%  \frametitle{Outline}
+%  \tableofcontents
   % You might wish to add the option [pausesections]
-\end{frame}
+%\end{frame}
 
-% Structuring a talk is a difficult task and the following structure
-% may not be suitable. Here are some rules that apply for this
-% solution:
-
-% - Exactly two or three sections (other than the summary).
-% - At *most* three subsections per section.
-% - Talk about 30s to 2min per frame. So there should be between about
-%   15 and 30 frames, all told.
-
-% - A conference audience is likely to know very little of what you
-%   are going to talk about. So *simplify*!
-% - In a 20min talk, getting the main ideas across is hard
-%   enough. Leave out details, even if it means being less precise than
-%   you think necessary.
-% - If you omit details that are vital to the proof/implementation,
-%   just say so once. Everybody will be happy with that.
-
-\section{What is Pyrolog?}
-
+% examples from my last beamer talk, shows the most important features
 \begin{frame}
@@
   \item
-    Pyrolog is a Prolog interpreter written in RPython
+    Pyrolog is a Prolog interpreter written in Python –
+  \item
+    want to compile it to get interesting performance
+  \item
+    translation tool-chain part of the PyPy project
+  \item
+    Python itself is too dynamic to be translatable to other languages, need a subset
   \item
@@
     RPython is designed to be significantly faster than regular Python
-  \item
-    translation tool-chain part of the PyPy project
   \end{itemize}
@@
 
-\section{The PyPy Approach to VM Construction}
-\subsection{Overview}
 \begin{frame}
-  \frametitle{What is PyPy?}
-  \begin{itemize}
-  \item
-    started as a Python VM implementation in Python
-  \item
-    Python itself is too dynamic to be translatable to other languages, need a subset
-  \item
-    includes a translation tool-chain for RPython
-  \item
-    is becoming a general environment for writing interpreters (JavaScript, Prolog started)
-  \item
-    Open Source project (MIT license)
-  \item
-    received EU funding for 2.5 years
-  \end{itemize}
-
-\end{frame}
-
-\subsection{Motivation}
-\begin{frame}
-  \frametitle{VMs are still hard}
-  Hard to reconcile:
-
-  \begin{itemize}
-  \item
-    flexibility
-  \item
-    maintainability
-  \item
-    performance (needs dynamic compilation techniques)
-  \end{itemize}
-  Especially with limited resources (like Open Source projects, research projects)
-\end{frame}
-
-\begin{frame}
-  \frametitle{The Python case (i)}
+  \frametitle{The Python case}
   CPython (the reference implementation) is a straightforward, portable VM.
@@
     No dynamic compilation
-%  \pause
   \end{itemize}
+  \pause
   \begin{block}{
@@
     \item
-      \alert{Stackless} (unlimited recursion, coroutines, serializable continuations)
+      \alert{Stackless} (unlimited recursion, coroutines, green threads)
     \item
-      \alert{Psyco} (run-time specializer)
+      \alert{Psyco} (run-time specializing compiler)
     \item
@@
   \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{The Python case (ii)}
-  \begin{block}{Problems of Extensions:}
-    \begin{itemize}
-    \item hard to maintain: need to keep track of CPython
-    \item Psyco very hard to port to other hardware architectures
-    \item tedious to write them, Python semantics need to be re-implemented
-    \end{itemize}
-  \end{block}
-  \begin{itemize}
-  \item
-    The community wants Python to run everywhere:
-    Jython (Java), IronPython (.NET).
-    Lots of effort and duplication.
-
-  \item
-    At various points various incompatibilities between
-    the implementations
-
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{The Prolog case (i)}
-  \begin{itemize}
-  \item
-    problem mitigated by the fact that Prolog the language does not change
-  \item
-    the core of Prolog is very simple (at least compared to Python)
-  \item
-    a lot of implementations out there
-  \item
-    well-tuned mature C implementations (Sicstus, XSB, SWI, GNU-Prolog)
-  \item
-    on CLR (P\#) and JVM (Prolog Café, tuProlog)
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{The Prolog case (ii)}
-  \begin{block}{mature C implementations}
-    \begin{itemize}
-    \item
-      interfacing with libraries is tedious
-    \item
-      changing the language to experiment is hard
-    \item
-      often extensions to core Prolog, incompatible between each other
-    \item
-      fixed implementation decisions (GC, how to generate code, etc.)
-    \end{itemize}
-  \end{block}
-  \begin{block}{implementations on CLR and JVM}
-    \begin{itemize}
-    \item
-      interfacing with libraries of the platform mostly easy
-    \item
-      no extensions to core Prolog (like tabling, coroutines)
-    \item
-      slow, compared to good C implementations
-    \end{itemize}
-  \end{block}
-\end{frame}
-
-\subsection{Approach}
-\begin{frame}
-  \frametitle{PyPy's Approach}
-    \alert{Goal:} generate VMs from a single high-level description of the
-     language, in a retargettable way.
-  \begin{itemize}
-  \item
-    Write an  interpreter for a dynamic language (Python, Prolog, JavaScript,
-    whatever) in a high-level language (Python)
-  \item
-    Leave out low-level details
-  \item
-    Favour simplicity and flexibility
-  \item
-    Define a mapping to low-level targets
-  \item
-    Generate VMs from the interpreter
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Mapping to low-level targets}
-  \begin{itemize}
-  \item
-    Mechanically translate the interpreter to multiple
-    lower-level targets
-    \begin{itemize}
-    \item C-like
-    \item Java
-    \item .NET
-    \end{itemize}
-  \item
-    Insert low-level aspects into the code as required by
-    the target (Object layout, memory management)
-    \begin{itemize}
-    \item object layout
-    \item memory management
-    \end{itemize}  \item
-    Optionally insert new pervasive features not expressed
-    in the source
-    \begin{itemize}
-    \item continuations, ``micro-threads''
-    \item dynamic compilation
-    \end{itemize}
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Translation Aspects (i)}
-  Features not present in the source can be added during translation.
-  \begin{itemize}
-  \item
-    \alert{memory management}: use different GC strategies (Boehm collector, custom mark-n-sweep)
-  \item
-    \alert{Stackless transformation}: allows program to control its stack (continuations, \dots)
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Translation Aspects (ii)}
-  A \alert{JIT compiler} as a translation aspect
-  \begin{itemize}
-  \item
-    Transform the interpreter into a JIT compiler, using partial evaluation and
-    specialization techniques
-  \item
-    Some hints in the interpreter source needed
-  \item
-    Current prototype applied to Python interpreter gives impressive speedups
-  \end{itemize}
-\end{frame}
-
-\section{The Prolog interpreter}
-\subsection{Interpreter}
-\begin{frame}
-  \frametitle{Prolog Interpreter Implementation}
-  \begin{itemize}
-  \item
-    naive, very simple interpreter
-    \begin{itemize}
-    \item uses "structure copying"
-    \item interprets Prolog terms directly, no bytecode
-    \end{itemize}
-  \item
-    uses continuation passing style inspired by BinProlog
-  \item
-    Prolog calls mapped to RPython calls
-  \item
-    implements large parts of the ISO standard (some builtins missing)
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Builtins}
-  \begin{itemize}
-  \item
-    builtins implemented in Python
-  \item
-    easy to add new ones to interface with libraries
-  \item
-    application-specific builtins
-  \item
-    examples:
-    \begin{itemize}
-    \item functions to download and analyze webpages
-    \item an imperative hashmap
-    \end{itemize}
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Interpreter Facts}
-  \begin{itemize}
-  \item
-    2500 lines of Python code in total
-  \item
-    700 of those are for builtins
-  \item
-    after translation to C: 14000 line of C code
-  \item
-    part of the PyPy distribution at:
-    \texttt{http://codespeak.net/pypy}
-  \end{itemize}
-\end{frame}
-
-\subsection{Performance}
-\begin{frame}
-  \frametitle{Performance (i)}
-  \includegraphics[scale=0.4]{image/bench.png}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Performance (ii)}
-  \begin{itemize}
-  \item
-    performance is quite bad compared to tuned C implementations
-  \item
-    performance is pretty good compared to Java and .NET implementations
-  \item
-    surprising, since those are often based on the WAM
-  \item
-    maybe the WAM model does not match these VMs very well?
-  \end{itemize}
-\end{frame}
-
-\section*{Summary}
-\subsection*{Summary}
-\begin{frame}
-  \frametitle{Summary}
-
-  % Keep the summary *very short*.
-  \begin{itemize}
-  \item
-    Very simple Prolog interpreter in RPython can compete with interpreters on the JVM, CLR
-  \item
-    Interpreter implementation eased by use of a high-level language
-  \item
-    Low-level details abstracted away but re-inserted later
-  \end{itemize}
-\end{frame}
-
-\subsection*{Outlook}
-\begin{frame}
-  \frametitle{Outlook}
-  \begin{itemize}
-  \item
-    complete the set of builtins
-  \item
-    tight language integration between Prolog and Python
-  \item
-    apply the dynamic compiler generator to the Prolog interpreter
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Backup slides}
-  \dots
-\end{frame}
-
-\begin{frame}
-  \frametitle{Translation Steps}
-  \begin{columns}[c]
-  \begin{column}{5cm}
-    \includegraphics[width=5cm]{image/arch.pdf}
-  \end{column}
-  \begin{column}{7cm}
-  \begin{itemize}
-    \item
-      Generate flow graphs from the RPython program
-    \item
-      Peform global type inference on the flow graphs
-    \item
-      Transform flow graphs through several steps until they match the level of
-      the target environment
-    \item
-      Weave in translation aspects in the process
-  \end{itemize}
-  \end{column}
-  \end{columns}
 \end{frame}

### 2007-07-24 talk/dyla2007/talk.tex
start the "problems with C" part

@@
 
-% examples from my last beamer talk, shows the most important features
 \begin{frame}
-  \frametitle{Pyrolog}
-
+  \frametitle{Introduction}
   \begin{itemize}
-  \item
-    Pyrolog is a Prolog interpreter written in Python –
-  \item
-    want to compile it to get interesting performance
-  \item
-    translation tool-chain part of the PyPy project
-  \item
-    Python itself is too dynamic to be translatable to other languages, need a subset
-  \item
-    RPython (``Restricted Python'') is a subset of Python translatable to other
-    languages
-  \item
-    RPython is designed to be significantly faster than regular Python
+  \item XXX intro, talk about scope
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{The Python case}
-  CPython (the reference implementation) is a straightforward, portable VM.
-
+  \frametitle{Common Approaches to VM construction}
+  \begin{block}{
+    Using C directly (or C disguised as another language)}
     \begin{itemize}
     \item
-    Pervasive decisions: reference counting, global lock \dots
+      CPython
     \item
-    No dynamic compilation
+      Ruby
+    \item
+      Spidermonkey (Mozilla's JavaScript VM)
+    \item
+      but also: Squeak, Scheme48
+    \end{itemize}
+  \end{block}
+  \begin{block}{
+    Building on top of an existing OO VM}
+    \begin{itemize}
+    \item
+      Jython, IronPython
+    \item
+      JRuby, IronRuby
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{The Problems of C}
+  When writing a VM in C it is hard to reconcile:
+  \begin{itemize}
+  \item
+    flexibility, maintainability
+  \item
+    simplicity of the VM
+  \item
+    performance (needs dynamic compilation techniques)
   \end{itemize}
@@
   \begin{block}{
-    Extensions:}
+    Python Case}
     \begin{itemize}
     \item
-      \alert{Stackless} (unlimited recursion, coroutines, green threads)
+      \alert{CPython} is a very simple bytecode VM, performance not great
     \item
-      \alert{Psyco} (run-time specializing compiler)
+      \alert{Psyco} is a just-in-time-specializer, very complex, hard to
+      maintain, but good performance
     \item
-      \alert{Jython}, \alert{IronPython}
+      \alert{Stackless} is a fork of CPython adding microthreads. It was never
+      incorporated into CPython for complexity reasons
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Compilers are a bad encoding of Semantics}
+  \begin{itemize}
+  \item to reach good performance levels, dynamic compilation is often needed
+  \item a dynamic compiler needs to encoding language semantics
+  \item this encoding is often relatively obscure and hard to change
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+      Psyco is a dynamic compiler for Python
+    \item
+      synchronizing with CPython's rapid development is a lot of effort
+    \item
+      many of CPython's new features not supported well
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Fixing of Early Design Decisions}
+  \begin{itemize}
+  \item when starting a VM in C, many decisions need to be made immediately
+  \item examples: memory management technique, threading model
+  \item the decision is manifested throughout the VM source
+  \item very hard to change later
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+      CPython uses reference counting, increfs and decrefs everywhere
+    \item
+      CPython uses OS threads with one global lock, hard to change to
+      lightweight threads or finer locking
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Implementation Proliferation}
+  \begin{itemize}
+  \item
+    restrictions of the original implementation lead to re-implementations,
+    forks
+  \item
+    all implementations need to be syncronized with language evolution
+  \item
+    lots of duplicate effort
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+      several serious implementations: CPython, Stackless, Psyco, Jython,
+      IronPython, PyPy
+    \item
+      the implementations have various grades of compliance
     \end{itemize}

### 2007-07-24 talk/dyla2007/talk.tex
Typoish changes.

@@
   \item to reach good performance levels, dynamic compilation is often needed
-  \item a dynamic compiler needs to encoding language semantics
-  \item this encoding is often relatively obscure and hard to change
+  \item a dynamic compiler needs to encode language semantics
+  \item this encoding is often obscure and hard to change
   \end{itemize}
@@
   \begin{itemize}
-  \item when starting a VM in C, many decisions need to be made immediately
+  \item when starting a VM in C, many design decisions need to be made upfront
   \item examples: memory management technique, threading model

### 2007-07-24 talk/dyla2007/talk.tex
something about OO VMs

@@
 \begin{frame}
-  \frametitle{The Problems of C}
+  \frametitle{Implementing VMs in C}
   When writing a VM in C it is hard to reconcile:
@@
 \begin{frame}
-  \frametitle{Title}
+  \frametitle{Implementing Languages on Top of OO VMs}
+  \begin{itemize}
+  \item
+    users wish to have easy interoperation with the OO VMs used by the industry
+    (JVM, CLR)
+  \item
+    bridges don't work that well (XXX maybe skip this)
+  \item
+    therefore re-implementations of the language on the OO VMs are started
+  \item
+    even more implementation proliferation
+  \item
+    implementing on top of an OO VM has its own set of problems
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+      \alert{Jython} is a Python-to-Java-bytecode compiler
+    \item
+      \alert{IronPython} is a Python-to-CLR-bytecode compiler
+    \item
+      both are slightly incompatible with the newest CPython version (especially
+      Jython)
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Benefits of implementing on top of OO VMs}
+  \begin{itemize}
+  \item
+    higher level of implementation
+  \item
+    the VM supplies a GC and mostly a JIT
+  \item
+    better interoperability than what the C level provides
+  \item
+    some proponents believe that eventually one single VM should be enough
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+      both Jython and IronPython allow easy access to the underlying VMs libraries
+    \item
+      XXX what else?
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{The problems of OO VMs}
+  \begin{itemize}
+  \item
+    some of the benefits of OO VMs don't work out in practice
+  \item
+    most immediate problem: it be hard to map concepts of the dynamic lang to
+    the host OO VM
+  \item
+    performance is often not improved, and can be very bad, because of the
+    semantic mismatch between the dynamic language and the host VM
+  \item
+    poor interoperability with everything outside the OO VM
+  \item
+    in practice, one OO VM is not enough
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+      Jython about 5 times slower than CPython
+    \item
+      IronPython is about as fast as CPython (but some introspection features
+      missing)
+    \item
+      XXX something about semantic mismatch?
+    \end{itemize}
+  \end{block}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{}
+  \begin{itemize}
+  \item
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{}
   \begin{itemize}

### 2007-07-24 talk/dyla2007/talk.tex
Typo - details.

@@
     \item
-      both Jython and IronPython allow easy access to the underlying VMs libraries
+      both Jython and IronPython integrate well with their host OO VM
     \item
@@
   \item
-    most immediate problem: it be hard to map concepts of the dynamic lang to
+    most immediate problem: it can be hard to map
+    concepts of the dynamic lang to
     the host OO VM

### 2007-07-24 talk/dyla2007/talk.tex
use "general-purpose OO VM" a few times

@@
 \begin{frame}
-  \frametitle{Implementing Languages on Top of OO VMs}
+  \frametitle{Implementing Languages on Top of General-Purpose OO VMs}
   \begin{itemize}
   \item
-    users wish to have easy interoperation with the OO VMs used by the industry
-    (JVM, CLR)
-  \item
-    bridges don't work that well (XXX maybe skip this)
+    users wish to have easy interoperation with the general-purpose OO VMs used
+    by the industry (JVM, CLR)
   \item

### 2007-07-24 talk/dyla2007/talk.tex
somewhat uninspired pypy section

@@
 \begin{frame}
-  \frametitle{}
+  \frametitle{PyPy's Approach to VM Construction}
   \begin{itemize}
   \item
+    Goal: achieve flexibility, simplicity and performance all at one
+  \item
+    Approach: auto-generate VMs from high-level descriptions of the language
+  \item
+    ... using meta-programming techniques and aspects (in the original sense)
+  \item
+    high-level description of the language done by writing an interpreter in
+    RPython
   \end{itemize}
@@
   \begin{block}{
-    Python Case}
+    What is RPython}
     \begin{itemize}
     \item
+      RPython is a subset of Python
+    \item
+      subset chosen in such a way that type-inference can be performed
+    \item
+      still a high-level language (unlike SLang or Prescheme)
     \end{itemize}
   \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Auto-generating VMs}
+  \begin{itemize}
+  \item
+    interpreters written in RPython can be translated to various targets:
+    C/Posix, CLR, JVM
+  \item
+    VMs generated by a flexible translation toolchain (really a compiler)
+  \end{itemize}
+  \begin{block}{
+    Translation Aspects}
+    \begin{itemize}
+    \item
+      the translation toolchain weaves in various aspect into the generated VM
+    \item
+      examples: GC strategy, threading model
+    \item
+      early design decisions not necessary
+    \item
+      non-trivial translation aspect: auto-generating a dynamic compiler from
+      the interpreter % XXX maybe we really want to write more about this
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Good Points of the Approach}
+  \begin{itemize}
+  \item
+    dynamic languages can implemented in a high level language
+  \item
+    doing this separates high-level from low-level concerns
+  \item
+    generating various VMs from a single source interpreter, which is easy to
+    write
+  \item
+    lots of flexibility at all levels:
+    \begin{itemize}
+    \item
+      when writing the interpreter
+    \item
+      also possible to adapt the translation toolchain if necessary
+    \item
+      this allows breaking of abstraction barriers, when necessary (e.g. tagged
+      integers)
+    \end{itemize}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Open Issues / Drawbacks / Further Work}
+  \begin{itemize}
+  \item
+    writing the translation toolchain in the first place takes lots of effort
+    (but it can be reused)
+  \item
+    writing a good GC is still necessary. but: can we maybe reuse existing good
+    GCs (e.g. from the Jikes RVM)?
+  \item
+    many abstraction layers
+  \item
+    dynamic compiler generation seems to work, but needs to be improved
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Conclusion / Meta-Points}
+  \begin{itemize}
+  \item
+    high-level languages are suitable to implement dynamic languages
+  \item
+    doing so has many benefits
+  \item
+    VM's shouldn't be written by hand
+  \item
+    PyPy's concrete approach is not so important
+  \item
+    diversity is good
+  \item
+    let's write more meta-programming tool-chains!
+  \end{itemize}
 \end{frame}

### 2007-07-25 talk/dyla2007/talk.tex
Some reshuffling of the first few slides of the PyPy part.

@@
   \frametitle{PyPy's Approach to VM Construction}
+  \emph{Goal: achieve flexibility, simplicity and performance together}
+
   \begin{itemize}
-  \item
-    Goal: achieve flexibility, simplicity and performance all at one
   \item
@@
   \item
-    ... using meta-programming techniques and aspects (in the original sense)
+    ... using meta-programming techniques and \emph{aspects}
   \item
-    high-level description of the language done by writing an interpreter in
-    RPython
+    high-level description: an interpreter written in a high-level language
+  \item
+    ... which we translate (i.e.\ compile) to VMs running on top of various
+    targets, like C/Posix, CLR, JVM
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{PyPy}
+  \begin{itemize}
+  \item
+    PyPy = Python interpreter written in RPython + translation toolchain
+    for RPython
   \end{itemize}
@@
       still a high-level language (unlike SLang or Prescheme)
+    \item
+      ...really a subset, can't give a small example of code that
+      doesn't just look like Python :-)
     \end{itemize}
@@
   \item
-    interpreters written in RPython can be translated to various targets:
-    C/Posix, CLR, JVM
+    high-level source: early design decisions not necessary
   \item
-    VMs generated by a flexible translation toolchain (really a compiler)
+    we need a custom \emph{translation toolchain} to compile the interpreter
+    to a full VM
+  \item
+    many aspects of the final VM are orthogonal from the interpreter source:
+    they are inserted during translation
+  \item
+    ``monads'', with more ad-hoc control
   \end{itemize}
+  \pause
   \begin{block}{
-    Translation Aspects}
+    Examples}
     \begin{itemize}
     \item
-      the translation toolchain weaves in various aspect into the generated VM
+      Garbage Collection strategy
     \item
-      examples: GC strategy, threading model
-    \item
-      early design decisions not necessary
+      Threading models (e.g.\ coroutines with CPS...)
     \item
@@
   \item
-    dynamic languages can implemented in a high level language
+    dynamic languages can be implemented in a high level language
   \item

### 2007-07-25 talk/dyla2007/talk.tex
More shuffling.

@@
   \item
-    ``monads'', with more ad-hoc control
+    translation aspect $\cong$ monads, with more ad-hoc control
   \end{itemize}
@@
       non-trivial translation aspect: auto-generating a dynamic compiler from
-      the interpreter % XXX maybe we really want to write more about this
+      the interpreter
     \end{itemize}
@@
   \frametitle{Good Points of the Approach}
+  {\bf Simplicity:}
+
   \begin{itemize}
@@
   \item
-    doing this separates high-level from low-level concerns
+    separation of concerns from low-level details
   \item
-    generating various VMs from a single source interpreter, which is easy to
-    write
+    a potential single-source-fits-all interpreter --
+    less duplication of efforts
   \item
-    lots of flexibility at all levels:
+    runs everywhere with the same semantics --
+    no outdated implementations, no ties to any standard platform
+  \end{itemize}
+  \pause
+  \begin{block}{PyPy}
+     arguably the most readable Python implementation so far
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Good Points of the Approach}
+  {\bf Flexibility} at all levels:
+
   \begin{itemize}
   \item
-      when writing the interpreter
+    when writing the interpreter (high-level languages rule!)
   \item
-      also possible to adapt the translation toolchain if necessary
+    when adapting the translation toolchain as necessary
   \item
-      this allows breaking of abstraction barriers, when necessary (e.g. tagged
-      integers)
+    to break abstraction barriers when necessary
   \end{itemize}
+  \pause
+  \begin{block}{
+    Example}
+    \begin{itemize}
+    \item
+      boxed integer objects, represented as tagged pointers
+    \item
+      manual system-level RPython code
     \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Good Points of the Approach}
+  {\bf Performance:}
+
+  \begin{itemize}
+  \item
+    ``reasonable'' performance
+  \item
+    can generate a dynamic compiler from the interpreter
+    (work in progress, 60x faster on very simple Python code)
+  \end{itemize}
+  \pause
+  \begin{block}{
+    JIT compiler generator}
+    \begin{itemize}
+    \item
+      almost orthogonal from the interpreter source - applicable to many
+      languages, follows language evolution ``for free''
+    \item
+      based on Partial Evaluation
+    \item
+      benefits from a high-level interpreter and a tweakable translation
+      toolchain
+    \item
+      generating a dynamic compiler is easier than generating a static one!
+    \end{itemize}
+  \end{block}
 \end{frame}
@@
   \item
-    writing a good GC is still necessary. but: can we maybe reuse existing good
-    GCs (e.g. from the Jikes RVM)?
+    writing a good GC is still necessary. but: maybe we can reuse existing good
+    GCs (e.g.\ from the Jikes RVM)?
   \item
-    many abstraction layers
+    conceptually simple approach but many abstraction layers
   \item
-    dynamic compiler generation seems to work, but needs to be improved
+    dynamic compiler generation seems to work, but needs more efforts.
+    also: can we layer it on top of the JIT of a general purpose OO VM?
   \end{itemize}
@@
   \item
-    VM's shouldn't be written by hand
+    VMs shouldn't be written by hand
   \item
@@
   \item
-    let's write more meta-programming tool-chains!
+    let's write more meta-programming toolchains!
   \end{itemize}

### 2007-07-26 talk/dyla2007/talk.tex
is this better?
this?
try to make "dynamic languages" be both in the second line in the title.
* birkenfeld sure knows LaTeX - fix title. * an introduction slide.

@@
 
-\title{How to \emph{not} write Virtual Machines for Dynamic Languages}
+\title{How to \underline{not} write Virtual Machines for \mbox{Dynamic Languages}}
 
@@
 \begin{frame}
-  \frametitle{Introduction}
+  \frametitle{Scope}
+  This talk is about:
+
   \begin{itemize}
-  \item XXX intro, talk about scope
+  \item
+    implementing dynamic languages \par(with a focus on complicated ones)
+  \item
+    in a context of limited resources \par(academic, open source, or
+    domain-specific)
   \end{itemize}
+  \pause
+  \begin{block}{
+    Our point}
+    \begin{itemize}
+    \item
+      Do not write virtual machines ``by hand''
+    \item
+      Instead, write interpreters in high-level languages
+    \item
+      Meta-programming is your friend
+    \end{itemize}
+  \end{block}
 \end{frame}

### 2007-07-28 talk/dyla2007/talk.tex
Insert "general-purpose" before the first "OO VM".
Write down a link to PyPy for interested people.
typos

@@
   \begin{block}{
-    Building on top of an existing OO VM}
+    Building on top of a general-purpose OO VM}
     \begin{itemize}
@@
   \item
-    all implementations need to be syncronized with language evolution
+    all implementations need to be synchronized with language evolution
   \item
@@
   \item
-    writing a good GC is still necessary. but: maybe we can reuse existing good
+    writing a good GC is still necessary. But: maybe we can reuse existing good
     GCs (e.g.\ from the Jikes RVM)?
@@
     dynamic compiler generation seems to work, but needs more efforts.
-    also: can we layer it on top of the JIT of a general purpose OO VM?
+    Also: can we layer it on top of the JIT of a general purpose OO VM?
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{}
-  \begin{itemize}
-  \item
-  \end{itemize}
+  \frametitle{Links}
+  \begin{block}{
+    PyPy}
+    \bigskip
+    \hskip 1cm \url{http://codespeak.net/pypy/}
+    \bigskip
+  \end{block}
 \end{frame}

### 2007-09-01 talk/esug2007/talk.tex
The ESUG2007 talk, based on the Dyla'07 one - sorry, I forgot to update to the latest version - but anyway the talk was mostly only the demo, not the slides.

--- 2007-07-28 talk/dyla2007/talk.tex
@@
 
-\title{How to \underline{not} write Virtual Machines for \mbox{Dynamic Languages}}
+\title{PyPy - How to \underline{not} write Virtual Machines for \mbox{Dynamic Languages}}
 
-\author{Carl Friedrich Bolz and Armin Rigo}
+\author{Armin Rigo}
 
@@
 
-\date{3rd Workshop on Dynamic Languages and Applications, July 31 2007}
+\date{ESUG 2007}
 
@@
   \begin{block}{
-    Our point}
+    Complicated = requiring a large VM}
+    \begin{itemize}
+    \item
+      Smalltalk (etc...): typically small core VM
+    \item
+      Python (etc...): the VM contains quite a lot
+    \end{itemize}
+  \end{block}
+  \pause
+  \begin{block}{
+    Limited resources}
+    \begin{itemize}
+    \item
+      Only near-complete implementations are really useful
+    \item
+      Minimize implementer's duplication of efforts
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Our point}
+  Our point:
+
     \begin{itemize}
@@
     \end{itemize}
-  \end{block}
 \end{frame}
@@
   \end{itemize}
-  \pause
-  \begin{block}{
-    Python Case}
-    \begin{itemize}
-    \item
-      both Jython and IronPython integrate well with their host OO VM
-    \item
-      XXX what else?
-    \end{itemize}
-  \end{block}
+%  \pause
+%  \begin{block}{
+%    Python Case}
+%    \begin{itemize}
+%    \item
+%      both Jython and IronPython integrate well with their host OO VM
+%    \item
+%      XXX what else?
+%    \end{itemize}
+%  \end{block}
 \end{frame}
@@
       missing)
-    \item
-      XXX something about semantic mismatch?
+    %\item
+    %  XXX something about semantic mismatch?
     \end{itemize}
@@
   \item
-    many aspects of the final VM are orthogonal from the interpreter source:
+    many aspects of the final VM are orthogonal to the interpreter source:
     they are inserted during translation
@@
 \begin{frame}
-  \frametitle{Links}
+  \frametitle{For more information}
   \begin{block}{
@@
   \end{block}
+  \begin{block}{
+    ``Sprints''}
+    \begin{itemize}
+    \item Main way we develop PyPy
+    \item They are programming camps, a few days to one week long
+    \item We may have one in Bern soon (PyPy+Squeak) and/or in
+          Germany (JIT and other topics)
+    \end{itemize}
+  \end{block}
+  \begin{block}{
+    See also}
+    Google for the full paper corresponding to these slides
+    that was submitted at Dyla'2007
+  \end{block}
 \end{frame}

### 2008-04-09 talk/rupy2008/talk.tex
first draft of the slides

\documentclass[utf8x]{beamer}

\mode<presentation>
{
  \usetheme{Warsaw}

  %\setbeamercovered{transparent}
}

\usepackage[english]{babel}

\usepackage[utf8x]{inputenc}

\usepackage{times}
\usepackage[T1]{fontenc}

\title{PyPy – status and next goals}

\author{Carl Friedrich Bolz}

\institute[Heinrich-Heine-Universität Düsseldorf]
{
  Institut für Informatik\\
  Heinrich-Heine-Universität Düsseldorf
}

\date{Ruby \& Python Conference, April 12th 2008, Poznań}

%\pgfdeclareimage[height=0.5cm]{pypy-logo}{image/py-web.png}
%\logo{\pgfuseimage{pypy-logo}}

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
%\AtBeginSubsection[]
%{
%  \begin{frame}<beamer>
%    \frametitle{Outline}
%    \tableofcontents[currentsection,currentsubsection]
%  \end{frame}
%}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

%\begin{frame}
%  \frametitle{Outline}
%  \tableofcontents
  % You might wish to add the option [pausesections]
%\end{frame}

\begin{frame}
  \frametitle{What is PyPy?}
  \begin{itemize}
  \item
    A flexible toolkit for writing interpreters
    \begin{itemize}
      \item
        giving you lots of ameninities
      \item
        translation to lots of environments
      \item
        a mostly good GC
      \item
        libraries
      \item
        a JIT
    \end{itemize}
  \pause
  \item
    the biggest implemented interpreter is a full Python interpreter
    \begin{itemize}
      \item
        implementing Python 2.4 semantics fully
      \item
        extra features
      \item
        lazy evaluation
      \item
        stackless
      \item
        various introspection enhancements
    \end{itemize}
  \item
    Open Source project (MIT licenced)
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{What is PyPy}
  ... pretty pictures and demos
\end{frame}

\begin{frame}
  \frametitle{Last year's developments}
  \begin{itemize}
  \item
    cleanups
  \item
    implementing ctypes
  \item
    .NET integration
  \item
    GC improvements
  \item
    JIT improvements
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Cleanups}
  \begin{itemize}
  \item PyPy got EU funding for 2.5 years
  \item in that time many features were added because the proposal required them
  \item not all of them were implemented well
  \item many of them were unmaintained
  \item some of them were preventing progress in other areas
  \item so we killed those
  \item improvements and cleanups everywhere, to get ready for the next round of features
  \item example: better implementation for unicode strings
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{CTypes}
  \begin{itemize}
  \item CTypes is a libffi wrapper for CPython
  \item implemented in C
  \item can be used to wrap C libraries rather effortlessly
  \item no need to write C code to interface with libraries

  \end{itemize}
  \pause
  \begin{block}{
    CTypes implementation for PyPy}
    \begin{itemize}
    \item two parts
    \item a very thin wrapper around libffi
    \item normal Python code for all the CTypes logic
    \item some things missing but already very useful
    \item ... demo
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{.NET Integration}
  \begin{itemize}
  \item giving PyPy.NET similar capabilities as CPython
  \item accessing .NET libraries seamlessly from PyPy
  \item ... demo
  \end{itemize}
\end

\begin{frame}
  \frametitle{Garbage Collection}
  \begin{itemize}
  \item RPython itself is garbage collected
  \item when translating to C, a GC is woven into the program
  \item GCs are completely written in Python
  \item similar to the MMTK toolkit of Jikes RVM
  \item two fast GCs: semispace, generational
  \item lots of care to get the semantics right:
    \begin{itemize}
    \item finalizers, weakrefs
    \item id function
    \item identity hashes
    \end{itemize}
  \item performance pretty good
  \item memory-bound benchmarks faster than CPython
  \item demo (id(x))
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Currently Working on: Just-in-Time Compiler}
  Problems:
  \begin{itemize}
  \item writing good JITs for dynamic languages is hard work
  \item keeping up with new language features is hard
  \item compilers are a bad encoding for languag semantics
  \pause
  \begin{block}{
    Existing hand-written JITs}
    \begin{itemize}
    \item Psyco (mostly unmaintained)
    \item LuaJit (slightly simpler language, genius at work)
    \item Tamarin (company support, who knows how good it is)
    \end{itemize}
  \end{block}
  \end{itemize}
  \pause
\end{frame}

\begin{frame}
  \frametitle{If you cannot write them, grow your own}
  \begin{itemize}
  \item If you cannot write a JIT, don't do that then
  \pause
  \item Idea: Automatically Generate the JIT from the interpreter
  \pause
  \end{itemize}
  \begin{block}{
    Partial Evaluation}
    \begin{itemize}
    \item old idea from CS research
    \item transforms an interpreter into a static compiler
    \item never worked well for real-life examples
    \item key insight: generating a JIT is \emph{easier}
    \end{itemize}
  \end{block}

\end{frame}

\begin{frame}
  \frametitle{Advantages of JIT Generation}
  Writing a JIT generator is hard, but:
  \begin{itemize}
  \item You only need to do it once
  \item the effort of writing a JIT generator is independent of the size of the language
  \item can keep up with language evolution
  \item XXX
  \item it's fun too
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{JIT Generation: Current Status}
  \begin{itemize}
  \item we had a working prototype at the end of the EU project
  \item speeds up simple arithmetic functions \emph{a lot}
  \item had limitations that were hard to fix with
  \pause
  \item refactoring happened in the last two months to improve the situation
  \item still in progress, but already allows experimentation
  \item we used the new flexibility to try new, advanced things
  \item goes into the direction of a tracing JIT
  \item right now we are starting to hit the hard bits
  \item we hope to get really good results for more general code in a few months
  \item ... demo
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
  \item PyPy approaching real world usefulness
  \item using CTypes to interface with the outside
  \item hopefully with a JIT soon
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Help Wanted!}
  we are looking for new contributors!
  \begin{itemize}
  \item try out your own apps that don't require extension modules
  \item use CTypes to implement missing modules
  \item experiment with extending Python (logic, distribution, what suits you)
  \item write a new JIT backend (x86-64 anyone?)
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Questions?}
  \begin{block}{
    PyPy}
    \bigskip
    \hskip 1cm \url{http://codespeak.net/pypy/}
    \bigskip
  \end{block}
\end{frame}

\end{document}

### 2008-04-09 talk/rupy2008/talk.tex
Typos.  Minor clarification.

@@
   \end{itemize}
-\end
+\end{frame}
 
@@
   \item keeping up with new language features is hard
-  \item compilers are a bad encoding for languag semantics
+  \item compilers are a bad encoding for language semantics
   \pause
@@
     \item never worked well for real-life examples
-    \item key insight: generating a JIT is \emph{easier}
+    \item key insight: generating a JIT is \emph{easier} than a static compiler
     \end{itemize}
@@
   \item speeds up simple arithmetic functions \emph{a lot}
-  \item had limitations that were hard to fix with
+  \item had limitations that were hard to fix
   \pause

### 2008-04-09 talk/rupy2008/talk.tex
mention anto, more advantages
Typo
drop few XXXs

@@
         giving you lots of ameninities
+        XXX don't use such words IMO, try to be more understandable
+        XXX audience is mostly students
       \item
@@
         lazy evaluation
+        XXX I don't like advertising features that don't work
       \item
@@
   \item no need to write C code to interface with libraries
+        XXX ctypes demo maybe...
 
@@
   \item RPython itself is garbage collected
+        XXX this is the first time you mention RPython, that's
+        XXX either too late, or not mention it at all
   \item when translating to C, a GC is woven into the program
-  \item GCs are completely written in Python
+        XXX another hard word
+  \item GCs are completely written in Python (XXX RPython)
   \item similar to the MMTK toolkit of Jikes RVM
@@
   \item can keep up with language evolution
-  \item XXX
+  \item retargetting the JIT is possible (e.g. to .NET, as Antonio Cuni is doing right now)
+  \item make the JIT behave radically differently is much easier with a generator
   \item it's fun too

### 2008-04-09 talk/rupy2008/talk.tex
try to fix the XXXs

--- 2008-04-09 talk/rupy2008/talk.tex
@@
       \item
-        giving you lots of ameninities
+        giving you lots of help
       \item
@@
       \item
-        lazy evaluation
-      \item
         stackless
@@
         various introspection enhancements
+      \item
+        experimental things like lazy evaluation
     \end{itemize}
@@
   \item RPython itself is garbage collected
-  \item when translating to C, a GC is woven into the program
-  \item GCs are completely written in Python
+%        XXX this is the first time you mention RPython, that's
+%        XXX either too late, or not mention it at all
+%        no, it is mentioned in the pictures part
+  \item when translating to C, a GC is integrated into the program
+  \item GCs are completely written in RPython
   \item similar to the MMTK toolkit of Jikes RVM

### 2008-05-08 talk/s3-2008/talk.tex
set up the latex files for the s3 talk

--- 2008-04-09 talk/rupy2008/talk.tex
@@
 
-\title{PyPy – status and next goals}
+\title{Back to the Future in one Week --- Implementing a Smalltalk VM in PyPy}
 
-\author{Carl Friedrich Bolz}
+\author{Carl Friedrich Bolz, Adrian Kuhn, Adrian Lienhard, Nicholas D. Matsakis, Oscar Nierstrasz, Lukas Renggli, Armin Rigo, Toon Verwaest}
 
-\institute[Heinrich-Heine-Universität Düsseldorf]
-{
-  Institut für Informatik\\
-  Heinrich-Heine-Universität Düsseldorf
-}
-
-\date{Ruby \& Python Conference, April 12th 2008, Poznań}
-
-%\pgfdeclareimage[height=0.5cm]{pypy-logo}{image/py-web.png}
-%\logo{\pgfuseimage{pypy-logo}}
+\date{Workshop on Self-sustaining Systems, May 16 2008}
 
@@
 \begin{frame}
-  \frametitle{What is PyPy?}
+  \frametitle{Scope}
+  This talk is about: XXX
   \begin{itemize}
   \item
-    A flexible toolkit for writing interpreters
-    \begin{itemize}
+    writing a Squeak implementation
   \item
-        giving you lots of ameninities
+    with eight people
   \item
-        translation to lots of environments
+    in five days
   \item
-        a mostly good GC
-      \item
-        libraries
-      \item
-        a JIT
-    \end{itemize}
-  \pause
-  \item
-    the biggest implemented interpreter is a full Python interpreter
-    \begin{itemize}
-      \item
-        implementing Python 2.4 semantics fully
-      \item
-        extra features
-      \item
-        lazy evaluation
-      \item
-        stackless
-      \item
-        various introspection enhancements
-    \end{itemize}
-  \item
-    Open Source project (MIT licenced)
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{What is PyPy}
-  ... pretty pictures and demos
-\end{frame}
-
-\begin{frame}
-  \frametitle{Last year's developments}
-  \begin{itemize}
-  \item
-    cleanups
-  \item
-    implementing ctypes
-  \item
-    .NET integration
-  \item
-    GC improvements
-  \item
-    JIT improvements
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Cleanups}
-  \begin{itemize}
-  \item PyPy got EU funding for 2.5 years
-  \item in that time many features were added because the proposal required them
-  \item not all of them were implemented well
-  \item many of them were unmaintained
-  \item some of them were preventing progress in other areas
-  \item so we killed those
-  \item improvements and cleanups everywhere, to get ready for the next round of features
-  \item example: better implementation for unicode strings
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{CTypes}
-  \begin{itemize}
-  \item CTypes is a libffi wrapper for CPython
-  \item implemented in C
-  \item can be used to wrap C libraries rather effortlessly
-  \item no need to write C code to interface with libraries
-
-  \end{itemize}
-  \pause
-  \begin{block}{
-    CTypes implementation for PyPy}
-    \begin{itemize}
-    \item two parts
-    \item a very thin wrapper around libffi
-    \item normal Python code for all the CTypes logic
-    \item some things missing but already very useful
-    \item ... demo
-    \end{itemize}
-  \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{.NET Integration}
-  \begin{itemize}
-  \item giving PyPy.NET similar capabilities as CPython
-  \item accessing .NET libraries seamlessly from PyPy
-  \item ... demo
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Garbage Collection}
-  \begin{itemize}
-  \item RPython itself is garbage collected
-  \item when translating to C, a GC is woven into the program
-  \item GCs are completely written in Python
-  \item similar to the MMTK toolkit of Jikes RVM
-  \item two fast GCs: semispace, generational
-  \item lots of care to get the semantics right:
-    \begin{itemize}
-    \item finalizers, weakrefs
-    \item id function
-    \item identity hashes
-    \end{itemize}
-  \item performance pretty good
-  \item memory-bound benchmarks faster than CPython
-  \item demo (id(x))
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Currently Working on: Just-in-Time Compiler}
-  Problems:
-  \begin{itemize}
-  \item writing good JITs for dynamic languages is hard work
-  \item keeping up with new language features is hard
-  \item compilers are a bad encoding for language semantics
-  \pause
-  \begin{block}{
-    Existing hand-written JITs}
-    \begin{itemize}
-    \item Psyco (mostly unmaintained)
-    \item LuaJit (slightly simpler language, genius at work)
-    \item Tamarin (company support, who knows how good it is)
-    \end{itemize}
-  \end{block}
-  \end{itemize}
-  \pause
-\end{frame}
-
-\begin{frame}
-  \frametitle{If you cannot write them, grow your own}
-  \begin{itemize}
-  \item If you cannot write a JIT, don't do that then
-  \pause
-  \item Idea: Automatically Generate the JIT from the interpreter
-  \pause
-  \end{itemize}
-  \begin{block}{
-    Partial Evaluation}
-    \begin{itemize}
-    \item old idea from CS research
-    \item transforms an interpreter into a static compiler
-    \item never worked well for real-life examples
-    \item key insight: generating a JIT is \emph{easier} than a static compiler
-    \end{itemize}
-  \end{block}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Advantages of JIT Generation}
-  Writing a JIT generator is hard, but:
-  \begin{itemize}
-  \item You only need to do it once
-  \item the effort of writing a JIT generator is independent of the size of the language
-  \item can keep up with language evolution
-  \item XXX
-  \item it's fun too
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{JIT Generation: Current Status}
-  \begin{itemize}
-  \item we had a working prototype at the end of the EU project
-  \item speeds up simple arithmetic functions \emph{a lot}
-  \item had limitations that were hard to fix
-  \pause
-  \item refactoring happened in the last two months to improve the situation
-  \item still in progress, but already allows experimentation
-  \item we used the new flexibility to try new, advanced things
-  \item goes into the direction of a tracing JIT
-  \item right now we are starting to hit the hard bits
-  \item we hope to get really good results for more general code in a few months
-  \item ... demo
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Conclusion}
-  \begin{itemize}
-  \item PyPy approaching real world usefulness
-  \item using CTypes to interface with the outside
-  \item hopefully with a JIT soon
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Help Wanted!}
-  we are looking for new contributors!
-  \begin{itemize}
-  \item try out your own apps that don't require extension modules
-  \item use CTypes to implement missing modules
-  \item experiment with extending Python (logic, distribution, what suits you)
-  \item write a new JIT backend (x86-64 anyone?)
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Questions?}
-  \begin{block}{
-    PyPy}
-    \bigskip
-    \hskip 1cm \url{http://codespeak.net/pypy/}
-    \bigskip
-  \end{block}
+    using PyPy
 \end{frame}

### 2008-05-09 talk/s3-2008/talk.tex
the slides about pypy in the s3 talk

@@
 
-\begin{frame}
+\frame{
     \frametitle{Scope}
     This talk is about: XXX
-  \begin{itemize}
-  \item
-    writing a Squeak implementation
-  \item
-    with eight people
-  \item
-    in five days
-  \item
-    using PyPy
-\end{frame}
+    \itemize{
+    \item writing a Squeak implementation
+    \item with eight people
+    \item in five days
+    \item using PyPy
+    }
+}
 
+\frame{
+    \frametitle{What is PyPy}
+    \itemize{
+    \item started as a Python implementation in Python
+    \item developed into a general environment for implementing dynamic languages
+    \item supports the language developer with a lot of infrastructure
+    \item Open Source project, MIT license
+    \item most important goal: abstracting over low-level details
+    \item don't fix decisions about low-level details 
+    }
+}
+
+\frame{
+    \frametitle{PyPy's Approach to VM Construction}
+    \itemize{
+    \item implement an interpreter for the dynamic language in RPython
+    \item translate this interpreter to a low-level language
+    \item translating inserts low-level details
+    \item a variety of target environment: C, LLVM, JVM, .NET
+    % XXX write about model-driven development?
+    }
+    \pause
+    \begin{block} {What is RPython?}
+        \itemize{
+        \item a more static subset of Python
+        \item static enough to enable type inference
+        \item still rather expressive: exceptions, inheritance, dynamic dispatch
+        \item analysis starts after importing of interpreter
+        \item enables compile-time metaprogramming
+        }
+    \end{block}
+}
+
+\frame{
+    \begin{block} {Translation Aspects}
+        \itemize{
+        \item many aspects of the final VM are orthogonal to language semantics
+        \item examples: GC strategy, threading model, many object details
+        \item non-trivial translation aspect: auto-generating a dynamic compiler
+        \item those shouldn't manifest in the interpreter source
+        \item they are inserted during translation
+        }
+    \end{block}
+    \pause
+    \begin{block} {Compile-Time Metaprogramming}
+        \itemize{
+        \item PyPy's translation toolchain starts analysis after importing
+        \item arbitrary (non-RPython) code can be executed during import
+        \item this allows metaprogramming of parts of the interpreter
+        }
+    \end{block}
+}
+
+\frame{
+    \frametitle{The SPy VM}
+    \itemize{
+        \item foo
+    }
+}
 \end{document}

### 2008-05-10 talk/s3-2008/talk.tex
some work on the slides

@@
 
-\title{Back to the Future in one Week --- Implementing a Smalltalk VM in PyPy}
+\title[Back to the Future in one Week]
+{%
+    Back to the Future in one Week — Implementing a Smalltalk VM in PyPy%
+}
 
-\author{Carl Friedrich Bolz, Adrian Kuhn, Adrian Lienhard, Nicholas D. Matsakis, Oscar Nierstrasz, Lukas Renggli, Armin Rigo, Toon Verwaest}
+\author[Bolz, Kuhn, Lienhard, Matsakis, Nierstrasz, Renggli, Rigo, Verwaest]
+{
+    \textcolor{green!50!black}{Carl~Friedrich~Bolz}\inst{1} \and
+    Adrian~Kuhn\inst{2} \and
+    Adrian~Lienhard\inst{2} \and
+    Nicholas~D.~Matsakis\inst{3} \and
+    Oscar~Nierstrasz\inst{2} \and
+    Lukas~Renggli\inst{2} \and
+    Armin~Rigo \and
+    Toon~Verwaest\inst{2}
+}
+
+\institute[Bern and others]
+{
+    \inst{2}%
+    Software Composition Group\\ University of Bern, Switzerland
+    \and%
+    \vskip-2mm
+    \inst{1}
+    Softwaretechnik und Programmiersprachen\\ Heinrich-Heine-Universit\"at D\"usseldorf
+    \and%
+    \vskip-2mm
+    \inst{3}%
+    ETH Zurich, Switzerland
+}
 
@@
 
-\frame{
+\begin{frame}
     \frametitle{Scope}
     This talk is about: XXX
-    \itemize{
+    \begin{itemize}
     \item writing a Squeak implementation
@@
     \item using PyPy
-    }
-}
+    \end{itemize}
+\end{frame}
 
-\frame{
+\begin{frame}
     \frametitle{What is PyPy}
-    \itemize{
+    \begin{itemize}
     \item started as a Python implementation in Python
@@
     \item don't fix decisions about low-level details 
-    }
-}
+    \end{itemize}
+\end{frame}
 
-\frame{
+\begin{frame}
     \frametitle{PyPy's Approach to VM Construction}
-    \itemize{
+    \begin{itemize}
     \item implement an interpreter for the dynamic language in RPython
@@
     % XXX write about model-driven development?
-    }
+    \end{itemize}
     \pause
     \begin{block} {What is RPython?}
-        \itemize{
+        \begin{itemize}
         \item a more static subset of Python
@@
         \item enables compile-time metaprogramming
-        }
+        \end{itemize}
     \end{block}
-}
+\end{frame}
 
-\frame{
+\begin{frame}
     \begin{block} {Translation Aspects}
-        \itemize{
+        \begin{itemize}
         \item many aspects of the final VM are orthogonal to language semantics
@@
         \item they are inserted during translation
-        }
+        \end{itemize}
     \end{block}
@@
     \begin{block} {Compile-Time Metaprogramming}
-        \itemize{
+        \begin{itemize}
         \item PyPy's translation toolchain starts analysis after importing
@@
         \item this allows metaprogramming of parts of the interpreter
-        }
+        \end{itemize}
     \end{block}
+\end{frame}
+
+\begin{frame}
+    \frametitle{The SPy VM}
+    \begin{itemize}
+        \item really simple, straight-forward Squeak interpreter in RPython
+        \item essentially free of low-level details, no GC
+        \item written in the course of five days
+        \item sprint-driven development
+    \end{itemize}
+    \pause
+    \begin{block} {Status}
+        \begin{itemize}
+        \item bytecodes fully implemented
+        \item many primitives implemented: arithmetic primitives, object primitives
+        \item can load Squeak images by decoding the bit patterns
+        \item runs simple benchmarks
+        \end{itemize}
+    \end{block}
+\end{frame}
+
+\begin{frame}
+    \frametitle{SPy Object Model}
+    \includegraphics[height=5cm]{objmodel}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Shadows}
+    Problem: 
+    \begin{itemize}
+    \item classes are just objects: can't really say which objects are used as classes
+    \item any object with the right instance fields can be used as a class
+    \item cryptic bitfields in the class which the VM needs to decode all the time
+    \end{itemize}
+    \pause
+    \begin{block}{Shadows}
+        \begin{itemize}
+        \item approach: potentially attach a \emph{shadow} to every object
+        \item the shadow caches VM-internal information
+        \item when the object is changed, the shadow is invalidated
+        \item so far only used for classes, later contexts, methods
+        \item conceptual cleanliness and nicer VM implementation
+        \end{itemize}
+    \end{block}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Tagged Pointers}
+    \begin{itemize}
+    \item Squeak implements integers as tagged pointers
+    \item doing that is orthogonal to language semantics
+    \item in PyPy implemented as a translation aspect
+    \item one class with exactly one int field can be implemented with a flag
+    \item when enabled, all method calls check for tag first
+    \item usually changing such a decision would be a major effort
+    \end{itemize}
+    \pause
+    \begin{block}{Results}
+        \begin{itemize}
+        \item small slowdown
+        \item memory advantage not measured
+        \item really easy to do
+        \end{itemize}
+    \end{block}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Primitives}
+    \begin{itemize}
+    \item try to make writing primitives easy
+    \item failure signaled by an exception
+    \item automatic popping from the stack and unwrapping of arguments
+    \item automatic pushing of the result
+    \item using a custom decorator \texttt{expose\_primitive}
+    \item compile-time metaprogramming
+    \end{itemize}
+\end{frame}
+
+\frame[containsverbatim, plain, shrink=10]{
+  \begin{verbatim}
+bool_ops = [
+    (LESSTHAN, operator.lt),
+    (GREATERTHAN, operator.gt),
+    (LESSOREQUAL, operator.le),
+    (GREATEROREQUAL,operator.ge),
+    (EQUAL, operator.eq),
+    (NOTEQUAL, operator.ne)
+    ]
+
+def make_func(code, op):
+
+    @expose_primitive(code, unwrap_spec=[int, int])
+    def func(interp, v1, v2):
+        res = op(v1, v2)
+        w_res = utility.wrap_bool(res)
+        return w_res
+
+for (code, op) in bool_ops:
+    make_func(code, op)
+  \end{verbatim}
 }
 
-\frame{
-    \frametitle{The SPy VM}
-    \itemize{
-        \item foo
+\begin{frame}
+    \frametitle{Image Loading}
+    \begin{itemize}
+    \item manual decoding of Squeak images by reading Squeak object format
+    \item image writing could be done in the same way later
+    \item not really a reason to directly use a memory dump
+    \item (hard anyway, on object-oriented platforms)
+    \end{itemize}
+\end{frame}
+
+\frame[plain]{
+    \frametitle{Performance (tiny Benchmark)}
+    \includegraphics[height=8cm]{speed}
 }
-}
+
+\begin{frame}
+    \frametitle{Results}
+    {\bf Good Points of the Approach:}
+    \begin{itemize}
+    \item simple, understandable, high-level Squeak implementation
+    \item mostly free of low-level details: no GC, no manual pointer tagging
+    \item written in a short amount of time
+    \item runs on various platforms
+    \end{itemize}
+    \pause
+    {\bf Bad Points of the Approach:}
+    \begin{itemize}
+    \item not really fast (yet)
+    \item XXX XXX
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Outlook}
+    \begin{itemize}
+    \item do graphical builtins, to actually start the full environment
+    \item Squeak-specific optimizations:
+    \item method-cache (should be easy with shadows)
+    \item JIT (see next slide)
+    \item lessons learned for a "SqueaSquea"?
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
+    \frametitle{JIT Outlook}
+    \begin{itemize}
+    \item PyPy currently explores automatic JIT generation
+    \item almost orthogonal from the interpreter source – applicable to many
+          languages, follows language evolution ``for free''
+    \item based on Partial Evaluation techniques
+    \item benefits from a high-level interpreter
+    \item generating a dynamic compiler is easier than generating a static one!
+    \item shares ideas with tracing JITs
+    \pause
+    \item current status: basic ideas seem to work, good speedups in prototypes
+    \item a lot of details still left
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Questions?}
+  \begin{block}{
+    PyPy}
+    \bigskip
+    \hskip 1cm \url{http://codespeak.net/pypy/}
+    \bigskip
+  \end{block}
+\end{frame}
+
 \end{document}

### 2008-05-16 talk/s3-2008/talk.tex
Put these two lines in a block (it's too easy to miss their apparition in the slide after the pause).
accumulated changes

@@
     \inst{3}%
-    ETH Zurich, Switzerland
+    ETH Zürich, Switzerland
 }
@@
     \frametitle{Scope}
-    This talk is about: XXX
+    This talk is about:
     \begin{itemize}
-    \item writing a Squeak implementation
+    \item writing a Squeak implementation (called "SPy") in Python
     \item with eight people
@@
 \begin{frame}
-    \frametitle{What is PyPy}
+    \frametitle{What is PyPy?}
     \begin{itemize}
     \item started as a Python implementation in Python
+    \item Open Source project, MIT license
     \item developed into a general environment for implementing dynamic languages
     \item supports the language developer with a lot of infrastructure
-    \item Open Source project, MIT license
     \item most important goal: abstracting over low-level details
-    \item don't fix decisions about low-level details 
+    \item don't fix decisions about low-level details early
     \end{itemize}
@@
     \item implement an interpreter for the dynamic language in RPython
-    \item translate this interpreter to a low-level language
-    \item translating inserts low-level details
+    \item translate this interpreter to a low-level language with PyPy toolchain
     \item a variety of target environment: C, LLVM, JVM, .NET
+    %\item translating inserts left-out low-level details
     % XXX write about model-driven development?
@@
         \item a more static subset of Python
-        \item static enough to enable type inference
-        \item still rather expressive: exceptions, inheritance, dynamic dispatch
-        \item analysis starts after importing of interpreter
-        \item enables compile-time metaprogramming
+        \item static enough to allow type inference
+        \item still rather expressive: exceptions, single inheritance, dynamic dispatch, interesting builtin types, garbage collection
+        %\item translations starts after importing of interpreter
         \end{itemize}
@@
         \begin{itemize}
-        \item many aspects of the final VM are orthogonal to language semantics
+        \item many low-level details of the final VM are orthogonal to language semantics
         \item examples: GC strategy, threading model, many object details
+        \item those shouldn't be visible in the interpreter source
+        \item they are inserted during translation
+        \pause
         \item non-trivial translation aspect: auto-generating a dynamic compiler
-        \item those shouldn't manifest in the interpreter source
-        \item they are inserted during translation
         \end{itemize}
@@
         \item really simple, straight-forward Squeak interpreter in RPython
-        \item essentially free of low-level details, no GC
-        \item written in the course of five days
-        \item sprint-driven development
+        \item goal is to fully support loading and running Squeak images
+        \item source code essentially free of low-level details, no GC
+        \item written during a five-day sprint in October in Bern
     \end{itemize}
@@
 
-\begin{frame}
+\begin{frame}[plain]
     \frametitle{SPy Object Model}
-    \includegraphics[height=5cm]{objmodel}
+    \includegraphics[height=7.5cm]{objmodel}
 \end{frame}
@@
     \item any object with the right instance fields can be used as a class
-    \item cryptic bitfields in the class which the VM needs to decode all the time
+    \item cryptic (bit)fields in the class which the VM needs to decode all the time
     \end{itemize}
     \pause
-    \begin{block}{Shadows}
+    \begin{block}{Approach}
         \begin{itemize}
-        \item approach: potentially attach a \emph{shadow} to every object
+        \item potentially attach a \emph{shadow} to every object
         \item the shadow caches VM-internal information
         \item when the object is changed, the shadow is invalidated
-        \item so far only used for classes, later contexts, methods
+        \item so far only used for classes, probably more later
         \item conceptual cleanliness and nicer VM implementation
+        \item still allows to expose low-level view to user code
         \end{itemize}
@@
     \begin{itemize}
-    \item Squeak implements integers as tagged pointers
+    \item Squeak implements small integers as tagged pointers
     \item doing that is orthogonal to language semantics
     \item in PyPy implemented as a translation aspect
-    \item one class with exactly one int field can be implemented with a flag
+    \item one class with exactly one int field can optionally be implemented
+    with tagged pointers
     \item when enabled, all method calls check for tag first
-    \item usually changing such a decision would be a major effort
+    \item usually changing such an implementation choice would be a major effort
     \end{itemize}
     \pause
-    \begin{block}{Results}
+    \begin{block}{Results of Tagging}
         \begin{itemize}
-        \item small slowdown
+        \item small slowdown due to need to check for flag
         \item memory advantage not measured
@@
     \item automatic pushing of the result
-    \item using a custom decorator \texttt{expose\_primitive}
+    \item using a custom function decorator \texttt{expose\_primitive}
     \item compile-time metaprogramming
@@
 
-def make_func(code, op):
+def make_primitive(prim_num, op):
 
-    @expose_primitive(code, unwrap_spec=[int, int])
-    def func(interp, v1, v2):
+    @expose_primitive(prim_num, unwrap_spec=[int, int])
+    def primitive(interp, v1, v2):
         res = op(v1, v2)
@@
 
-for (code, op) in bool_ops:
-    make_func(code, op)
+for (prim_num, op) in bool_ops:
+    make_primitive(prim_num, op)
   \end{verbatim}
@@
 
-\begin{frame}
-    \frametitle{Image Loading}
-    \begin{itemize}
-    \item manual decoding of Squeak images by reading Squeak object format
-    \item image writing could be done in the same way later
-    \item not really a reason to directly use a memory dump
-    \item (hard anyway, on object-oriented platforms)
-    \end{itemize}
-\end{frame}
+%\begin{frame}
+%    \frametitle{Image Loading}
+%    \begin{itemize}
+%    \item manual decoding of Squeak images by reading Squeak object format
+%    \item image writing could be done in the same way later
+%    \item not really possible to directly use a memory dump
+%    \item not really necessary either
+%    \item (hard anyway, on object-oriented platforms)
+%    \end{itemize}
+%\end{frame}
 
@@
     \frametitle{Performance (tiny Benchmark)}
+    \centering
     \includegraphics[height=8cm]{speed}
@@
     \item not really fast (yet)
-    \item XXX XXX
+    \item RPython isn't Python
+    \item toolchain needs quite some effort (but only once)
     \end{itemize}
 \end{frame}
+
+%<arigato> we could mention that we have many abstraction layers and levels in the translation toolchain,
+%<arigato> which is generally good and useful but makes things more complex
+%<arigato> that would be how I'd compare with cola
+%<arigato> of course the question "how do you compare with cola" will be asked, so we need to be ready for it
 
@@
     \end{itemize}
+  \begin{block}{
+    Join the Sprint!}
+    \bigskip
+    \hskip 1cm Saturday - Thursday, C-Base Berlin
+    \bigskip
+  \end{block}
 \end{frame}
@@
     \item benefits from a high-level interpreter
-    \item generating a dynamic compiler is easier than generating a static one!
     \item shares ideas with tracing JITs
+    \end{itemize}
     \pause
-    \item current status: basic ideas seem to work, good speedups in prototypes
+    \begin{block} {Current status}
+        \begin{itemize}
+        \item basic ideas seem to work, good speedups in prototypes
         \item a lot of details still left
         \end{itemize}
+    \end{block}
 \end{frame}
@@
   \frametitle{Questions?}
+  Outlook:
+  \begin{itemize}
+  \item do graphical builtins, to actually start the full environment
+  \item Squeak-specific optimizations:
+  \item method-cache (should be easy with shadows)
+  \item JIT (automatic generation of dynamic compilers)
+  \item lessons learned for a "SqueaSquea"?
+  \end{itemize}
+  \bigskip
   \begin{block}{
-    PyPy}
+    Join the Sprint!}
     \bigskip
-    \hskip 1cm \url{http://codespeak.net/pypy/}
+    \hskip 1cm Saturday - Thursday, C-Base Berlin
     \bigskip

### 2008-05-19 talk/s3-2008/talk-squeakev.tex
the talk we gave at the squeak e.V. meeting

--- 2008-05-09 talk/s3-2008/talk.tex
@@
 
-\title{Back to the Future in one Week --- Implementing a Smalltalk VM in PyPy}
+\title[Back to the Future in one Week]
+{%
+    Back to the Future in one Week — Implementing a Smalltalk VM in PyPy%
+}
 
-\author{Carl Friedrich Bolz, Adrian Kuhn, Adrian Lienhard, Nicholas D. Matsakis, Oscar Nierstrasz, Lukas Renggli, Armin Rigo, Toon Verwaest}
+\author[Bolz, Kuhn, Lienhard, Matsakis, Nierstrasz, Renggli, Rigo, Verwaest]
+{
+    \textcolor{green!50!black}{Carl~Friedrich~Bolz}\inst{1} \and
+    \textcolor{green!50!black}{Adrian~Kuhn\inst{2}} \and
+    Adrian~Lienhard\inst{2} \and
+    Nicholas~D.~Matsakis\inst{3} \and
+    Oscar~Nierstrasz\inst{2} \and
+    Lukas~Renggli\inst{2} \and
+    Armin~Rigo \and
+    \textcolor{green!50!black}{Toon~Verwaest}\inst{2}
+}
 
-\date{Workshop on Self-sustaining Systems, May 16 2008}
+\institute[Bern and others]
+{
+    \inst{2}%
+    Software Composition Group\\ University of Bern, Switzerland
+    \and%
+    \vskip-2mm
+    \inst{1}
+    Softwaretechnik und Programmiersprachen\\ Heinrich-Heine-Universit\"at D\"usseldorf
+    \and%
+    \vskip-2mm
+    \inst{3}%
+    ETH Zürich, Switzerland
+}
+
+\date{Squeak e.V. Meeting, May 17 2008 \\
+(presented at S3)
+}
 
@@
 
-\frame{
+\begin{frame}
     \frametitle{Scope}
-    This talk is about: XXX
-    \itemize{
-    \item writing a Squeak implementation
+    This demo is about:
+    \begin{itemize}
+    \item writing a Squeak implementation (called "SPy") in Python
     \item with eight people
@@
     \item using PyPy
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
+    \frametitle{What is PyPy?}
+    \begin{itemize}
+    \item started as a Python implementation in Python
+    \item Open Source project, MIT license
+    \item developed into a general environment for implementing dynamic languages
+    \item supports the language developer with a lot of infrastructure
+    \item most important goal: abstracting over low-level details
+    \item don't fix decisions about low-level details early
+    \end{itemize}
+\end{frame}
+
+\frame[plain]{
+    \frametitle{What is PyPy?}
+    \centering
+    \includegraphics[height=8cm]{dynlang}
 }
+\begin{frame}
+    \frametitle{The SPy VM}
+    \begin{itemize}
+        \item complete but simple, straight-forward Squeak interpreter in RPython
+        \item goal is to fully support loading and running Squeak images
+        \item source code essentially free of low-level details, no GC
+        \item written during a five-day sprint in October in Bern
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Current Status}
+    \begin{itemize}
+    \item bytecodes fully implemented
+    \item many primitives implemented: arithmetic primitives, object primitives
+    \item can load Squeak images (both 2.0 and 3.9)
+    \item runs tiny benchmarks
+    \item runs arbitrary code
+    \item resume image
+    \end{itemize}
+    \pause
+    Open Issues:
+    \begin{itemize}
+    \item any I/O
+    \item graphics
+    \item performance
+    \item \#someInstance, \#nextInstance
+    \item \#become
+    \item tests for threading and exceptions
+    \end{itemize}
+\end{frame}
+
+\frame[containsverbatim, plain, shrink=10]{
+  \begin{verbatim}
+  primitiveSquareRoot
+    | rcvr |
+    self var: #rcvr type: 'double '.
+    rcvr := self popFloat.
+    self success: rcvr >= 0.0.
+    successFlag
+        ifTrue: [self pushFloat:
+            (self cCode: 'sqrt(rcvr)'
+                  inSmalltalk: [rcvr sqrt])]
+        ifFalse: [self unPop: 1]
+
+  \end{verbatim}
 }
 
-\frame{
-    \frametitle{What is PyPy}
-    \itemize{
-    \item started as a Python implementation in Python
-    \item developed into a general environment for implementing dynamic languages
-    \item supports the language developer with a lot of infrastructure
-    \item Open Source project, MIT license
-    \item most important goal: abstracting over low-level details
-    \item don't fix decisions about low-level details 
-    }
+\frame[containsverbatim, plain, shrink=10]{
+  \begin{verbatim}
+  primitiveSquareRoot
+    | rcvr |
+    self var: #rcvr type: 'double '.
+    rcvr := self popFloat.
+    self success: rcvr >= 0.0.
+    successFlag
+        ifTrue: [self pushFloat:
+            (self cCode: 'sqrt(rcvr)'
+                  inSmalltalk: [rcvr sqrt])]
+        ifFalse: [self unPop: 1]
+
+  @expose_primitive(FLOAT_SQUARE_ROOT,
+                    unwrap_spec=[float])
+def func(interp, f): 
+    if f < 0.0:
+        raise PrimitiveFailedError
+    w_res = utility.wrap_float(math.sqrt(f))
+    return w_res
+  \end{verbatim}
 }
 
-\frame{
-    \frametitle{PyPy's Approach to VM Construction}
-    \itemize{
-    \item implement an interpreter for the dynamic language in RPython
-    \item translate this interpreter to a low-level language
-    \item translating inserts low-level details
-    \item a variety of target environment: C, LLVM, JVM, .NET
-    % XXX write about model-driven development?
-    }
-    \pause
-    \begin{block} {What is RPython?}
-        \itemize{
-        \item a more static subset of Python
-        \item static enough to enable type inference
-        \item still rather expressive: exceptions, inheritance, dynamic dispatch
-        \item analysis starts after importing of interpreter
-        \item enables compile-time metaprogramming
-        }
-    \end{block}
+\frame[plain]{
+    \frametitle{Performance (tiny Benchmark)}
+    \centering
+    \includegraphics[height=8cm]{speed}
 }
 
-\frame{
-    \begin{block} {Translation Aspects}
-        \itemize{
-        \item many aspects of the final VM are orthogonal to language semantics
-        \item examples: GC strategy, threading model, many object details
-        \item non-trivial translation aspect: auto-generating a dynamic compiler
-        \item those shouldn't manifest in the interpreter source
-        \item they are inserted during translation
-        }
+\begin{frame}
+    \frametitle{Results}
+    {\bf Good Points of the Approach:}
+    \begin{itemize}
+    \item simple, understandable, high-level Squeak implementation
+    \item mostly free of low-level details: no GC, no manual pointer tagging
+    \item written in a short amount of time
+    \item runs on various platforms
+    \item extensible tool chain
+    \end{itemize}
+    \pause
+    {\bf Bad Points of the Approach:}
+    \begin{itemize}
+    \item not really fast (yet)
+    \item RPython isn't Python
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Outlook}
+    \begin{itemize}
+    \item do graphical builtins, to actually start the full environment
+    \item Squeak-specific optimizations:
+    \item method-cache (should be easy with shadows)
+    \item JIT
+    \item lessons learned for a "SqueaSquea"?
+    \end{itemize}
+  \begin{block}{
+    Join the Sprint!}
+    \bigskip
+    \hskip 1cm Saturday - Thursday, C-Base Berlin
+    \bigskip
   \end{block}
-    \pause
-    \begin{block} {Compile-Time Metaprogramming}
-        \itemize{
-        \item PyPy's translation toolchain starts analysis after importing
-        \item arbitrary (non-RPython) code can be executed during import
-        \item this allows metaprogramming of parts of the interpreter
-        }
-    \end{block}
-}
+\end{frame}
 
-\frame{
-    \frametitle{The SPy VM}
-    \itemize{
-        \item foo
-    }
-}
 \end{document}

### 2008-12-02 talk/ecoop2009/abstract.tex
first draft of the ecoop paper

\begin{abstract}
Dynamic languages are increasingly popular etc. etc.
\end{abstract}

### 2008-12-02 talk/ecoop2009/clibackend.tex
first draft of the ecoop paper

\section{CLI backend for Rainbow interpreter}

\subsection{Promotion on CLI}

Implementing promotion on top of CLI is not straightforward, as it needs to
patch and modify the already generated code, and this is not possible on .NET.

To solve, we do xxx and yyy etc. etc.

### 2008-12-02 talk/ecoop2009/intro.tex
first draft of the ecoop paper

\section{Introduction}

Dynamic languages are increasingly popular etc. etc.

The easiest way to implement a dynamic language such as Python is to write an
interpreter; however, interpreters are slow.

The alternative is to write a compiler; writing a compiler that targets a high
level virtual machine like CLI or JVM is easier than targeting a real CPU, but
it still require a lot of work, as IronPython, Jython, JRuby demonstrate.

Moreover, writing a static compiler is often not enough to get high
performances; IronPython and JRuby are going in the direction of JIT compiling
specialized versions of the code depending on the actual values/types seen at
runtime; this approach seems to work, but write it manually requires an
enormous effort.

PyPy's idea is to automatize the generation of static/JIT compilers in order
to reduce at minimun the effort required to get a fast implementation of a
dynamic language; all you have to do is to write a high level specification of
the language (by writing an interpreter), and put it through PyPy's
translation toolchain.

\subsection{PyPy and RPython}

\begin{figure}[h]
\begin{center}
\includegraphics[width=.6\textwidth]{diagram0}
\caption{PyPy infrastracture for generating an interpreter of a
  language $L$ for several platforms}\label{pypy-fig}
\end{center}
\end{figure}

The \emph{PyPy} project\footnote{\texttt{http://codespeak.net/pypy/dist/pypy/doc/home.html}}
\cite{RigoPedroni06} was initially conceived to develop an implementation of Python which
could be easily portable and extensible without renouncing efficiency.
To achieve these aims, the PyPy implementation is based on a highly
modular design which allows high-level aspects
to be separated from lower-level implementation details.
The abstract semantics of Python is defined by an interpreter written
in a high-level language, called RPython \cite{AACM-DLS07}, which is in fact a subset of
Python where some dynamic features have been sacrificed to allow an
efficient translation of the interpreter to low-level code.

Compilation of the interpreter is implemented as a stepwise
refinement by means of a translation toolchain which performs type
analysis, code optimizations and several transformations aiming at 
incrementally providing implementation details as memory management or the threading model.
The different kinds of intermediate codes  which are refined 
during the translation process are all represented by a collection of control flow graphs,
at several levels of abstractions.

Finally, the low-level control flow-graphs produced by the toolchain
can be translated to executable code for a specific platform by a
corresponding backend.
Currently, three fully developed backends are available to produce
executable C/POSIX code, Java and CLI/.NET bytecode.

Despite the PyPy infrastructure was specifically developed 
for Python, in fact it can be used for implementing
other languages. Indeed, PyPy has been successfully experimented with
several languages as Smalltalk \cite{BolzEtAl08}, JavaScript, Scheme and Prolog.
As suggested by Figure~\ref{pypy-fig}, a portable interpreter for a
generic language $L$  can be
easily developed once an abstract interpreter for $L$ is implemented in
RPython.

Another interesting feature of PyPy
is that just-in-time compilers can be semi-automatically generated from the
interpreter source.

### 2008-12-02 talk/ecoop2009/jitgen.tex
first draft of the ecoop paper

\section{Automatic generation of JIT compilers}

Traditional JIT compilers are hard to write, time consuming, hard to evolve,
etc. etc.

\begin{figure}[h]
\begin{center}
\includegraphics[width=.6\textwidth]{diagram1}
\caption{PyPy infrastructure for generating an interpreter of a
  language $L$ with JIT compilation for the .NET platform}
\end{center}
\end{figure}

The JIT generation framework uses partial evaluation techniques to generate a
dynamic compiler from an interpreter; the idea is inspired by Psyco, which
uses the same techniques but it's manually written instead of being
automatically generated.

The original idea is by Futamura \cite{Futamura99}. He proposed to generate compilers
from interpreters with automatic specialization, but his work has had
relatively little practical impact so far.

\subsection{Partial evaluation}

Assume the Python bytecode to be constant, and constant-propagate it into the
Python interpreter.

Example::
\begin{lstlisting}[language=Python]
  def f(x, y):    
    x2 = x * x    
    y2 = y * y    
    return x2 + y2

**case x=3** ::

  def f_3(y):    
    y2 = y * y   
    return 9 + y2

**case x=10** ::

  def f_10(y):    
    y2 = y * y   
    return 100 + y2
\end{lstlisting}

A shortcoming of PE is that in many cases not much can be really assumed
constant at compile-time, and this leads to poor results.  Effective dynamic
compilation requires feedback of runtime information into compile-time; for a
dynamic language, types are a primary example.

\subsection{Execution steps}

* Translation time

  - pypy-c-jit is translated into an executable

  - the JIT compiler is automatically generated

* Compile-time: the JIT compiler runs

* Runtime: the JIT compiled code runs

* Compile-time and runtime are intermixed

### 2008-12-02 talk/ecoop2009/rainbow.tex
first draft of the ecoop paper

\section{The Rainbow Interpreter}

The JIT compiler is implemented by running the "Rainbow interpreter", whose
output is a compiled function.

Variables and opcodes are colored:

  - green values are known at compile time, and green operations are executed
    when interpreting the bytecode;

  - red values are not known until runtime; the result of a red operation is a
    piece of machine code that will compute the result at runtime.

The Rainbow bytecode is produced at translation time, when the JIT compiler is
generated.

Here are summarized the various phases of the JIT:

Translation time:

  * Low-level flowgraphs are produced

  * The *hint-annotator* colors the variables

  * The *rainbow codewriter* translates flowgraphs into rainbow bytecode

Compile-time:

  * The rainbow interpreter executes the bytecode

  * As a result, it produces executable code

Runtime:

  * The produced code is executed

\subsection{Example of Rainbow bytecode and execution}

TODO

\subsection{Promotion}

There are values that, if known at compile time, allow the JIT compiler to
produce very efficient code.  Unfortunately, these values are tipically red,
e.g. the exact type of a variable.

"Promotion" is a particular operation that convert a red value into a green
value; i.e., after the promotion of a variable, the JIT compiler knows its
value at compile time. Since the value of a red variable is not known until
runtime, we need to postpone the compilation phase after the runtime phase.

This is done by continuously intermixing compile time and runtime; a promotion
is implemented in this way:

  * (compile time): the rainbow interpreter produces machine code until it
    hits a promotion point; e.g.::

    \begin{lstlisting}[language=C]
        y = hint(x, promote=True)
        return y+10
    \end{lstlisting}

  * (compile time): at this point, it generates special machine code that when
    reached calls the JIT compiler again; the JIT compilation stops::

    \begin{lstlisting}[language=C]
        switch(y) {
            default: compile_more(y);
        }
    \end{lstlisting}

  * (runtime): the machine code is executed; when it reaches a promotion
    point, it executes the special machine code we described in the previous
    point; the JIT compiler is invoked again;

  * (compile time): now we finally know the exact value of our red variable,
    and we can promote it to green; suppose that the value of 'y' is 32::

    \begin{lstlisting}[language=C]
        switch(y) {
            32: return 42;
            default: compile_more(y);
        }
    \end{lstlisting}

    Note that the operation "y+10" has been constant-folded into "42", as it
    was a green operation.

  * (runtime) the execution restart from the point it stopped, until a new
    unhandled promotion point is reached.

\subsection{Virtuals and virtualizables}

TODO

### 2008-12-11 talk/ecoop2009/clibackend.tex
started integrating Antonio's post in the paper

@@
 
-\subsection{Promotion on CLI}
+%\subsection{Promotion on CLI}
+%
+%Implementing promotion on top of CLI is not straightforward, as it needs to
+%patch and modify the already generated code, and this is not possible on .NET.
+%
+%To solve, we do xxx and yyy etc. etc.
+%
 
-Implementing promotion on top of CLI is not straightforward, as it needs to
-patch and modify the already generated code, and this is not possible on .NET.
+\subsection{Flexswitches}
 
-To solve, we do xxx and yyy etc. etc.
+As already explained, \dacom{I guess flexswitches will be introduced
+  in the previous sect.} \emph{flexswitch} is one of the key
+concepts allowing the JIT compiler generator to produce code which can
+be incrementally specialized and compiled at run time.
+
+A flexswitch is a special kind of switch which can be dynamically
+extended with new cases; intuitively, its behavior can be described
+well in terms of flow graphs. Indeed, a flexswitch can be considered 
+as a special flow graph block where links to newly created blocks are
+dynamically added whenever new cases are needed. \dacom{in theory links could also be removed, but I
+  do not know whether this possibility has been considered in Rainbow}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[height=5cm]{flexswitch1}
+\includegraphics[height=5cm]{flexswitch2}
+\caption{An example of a flexswitch evolution: in the picture on the
+  right a new block has been dynamically added.}\label{flexswitch-fig}
+\end{center}
+\end{figure}
+
+In the pictures of Figure~\ref{flexswitch-fig}, the cyan block
+corresponds to a flexswitch; initially (picture on the left) 
+only the block containing the code to restart the JIT compilation
+is connected to the flexswitch; the picture on the right
+shows the graph after the first case has been dynamically added to the flexswitch,
+by linking the cyan block with a freshly created new block.
+
+\subsection{Implementing flexswitches in CLI}
+
+Implementing flexswitches for backends generating assembly code is
+quite straightforward: basically, a new jump has to be inserted in the
+existing code to point to the newly generated code fragment.
+
+Unfortunately, the CLI VM does not allow modification of code which
+has been already loaded and linked, therefore the simplest approach
+taken for low level architectures does not work for higher level 
+virtual machines as those for .NET and Java.
+
+Since in .NET methods are the basic units of compilation, a possible
+solution consists in creating a new method 
+any time a new case has to be added to a flexswitch.
+\dacom{comment for Antonio: I am not sure this is the best solution. This cannot work for Java where classes are the basic
+  units}
+
+\dacom{I still have to polish what comes next}
+
+\newcommand{\commentout}[1]{}
+\commentout{
+Because of all these constraints we cannot simply map each graph to its own method, since we saw that our graphs can grow after they have already been executed few times.
+
+Hence, we need to distinguish between the two concepts:
+
+        * a graph is the logical unit of code as seen by the JIT compiler: concretely, the CLI JIT backend renders it as one or more methods;
+        * a method is a collection of basic blocks; each method has the so called parent graph, i.e. the graph its blocks logically belongs to.
+
+The first method of a graph is called main method (which has nothing to do with the Main static methods found in .exe files); other methods are called children methods.
+
+When we want to add a new case to the flexswitch, we create a method containing all the new code; then we wrap the method inside a delegate (the .NET equivalent of a function pointer) and pass it to the flexswitch, so that it can later invoke it.
+The hard bit: non-local links
+
+Using this approach, after a while the blocks of our original graph are scattered over a lot of different methods; however, there are no constraints about how these blocks can be linked together, so it happens to have links between blocks which are not in the same method.
+In the following, we will refer to them as non-local links.
+
+If the non-local block we want to jump to happens to be at the beginning of its containing method, it is enough to invoke the method; but, what if we want to jump somewhere in the middle?
+What we really want is to produce a method which has multiple entry-points; again, doing it in assembly would be trivial, but the virtual machine does not provide any support for it, so we need a work around.
+
+Each method in a graph is assigned an unique 16 bit method id; each block in a method is assigned a progressive 16 bit block number.
+From this two numbers, we can compute the block id as an unsigned integer, by storing the method id in the first 16 bits and the block number in the second 16 bits.
+By construction, the block id is guaranteed to be unique in the graph.
+
+The following picture shows a graph composed of three methods; the id of each method is shown in red, while the block ids are shown in red (for the method id part) and black (for the block number part).
+The graph contains three non-local links; in particular, note the link between blocks 0x00020001 and 0x00010001 which connects two block that resides in different methods.
+
+Every method contains a special dispatch block, (not shown in the picture above) whose goal is to jump to the specified block number inside the method itself.
+The first argument of a child method is always a block id; when the method starts, it immediately jumps to the dispatch block, and thus to the desired block.
+
+For example, suppose to have a method which contains 3 blocks numbered 0, 1, 2; here is how its dispatch blocks looks like; for simplicity it is shown as C# code, but it is actually generated as IL bytecode:
+
+// dispatch block
+int methodid = (blockid & 0xFFFF0000) >> 16); // take the first 16 bits
+int blocknum = blockid && 0x0000FFFF;         // take the second 16 bits
+
+if (methodid != MY_METHOD_ID) {
+// jump_to_unknown block
+...
+}
+
+switch(blocknum) {
+case 0:
+goto block0;
+case 1:
+goto block1;
+case 2:
+goto block2;
+default:
+throw new Exception("Invalid block id");
+}
+
+Whenever we want to jump to a non-local block, it is enough to store the block id in the appropriate variable and jump to the dispatch block.
+If the block resides in a different method, the jump_to_unknown block is entered; this special block is implemented differently by the main method and the child methods, as we will see soon.
+
+Each time a new method is added to the graph, we build a delegate for it, and store it in a special array called method_map; since we assign the method id sequentially starting from 0, we are sure that to fetch the method whose id is n we can simply load the n-th element of the array.
+
+The jump_to_unknown block of the main method uses this array to select the right method, and calls it (FlexSwitchCase is the type of delegates for all children methods):
+
+// jump_to_unknown block of the main method
+FlexSwitchCase meth = method_map[methodid];
+blockid = meth(blockid, ...); // execute the method
+goto dispatch_block;
+
+Each child method returns a block id specifying the next block to jump to; after its execution, we assign the return value to the blockid variable, and jump again to the dispatch block, which will jump again to the appropriate block.
+
+Keeping this in mind, it is straightforward to implement the jump_to_unknown block of children methods: it is enough to return the target block id to the caller, and let its dispatch loop do the right thing.
+If the caller is also a child method, it will return it again, until we reach the dispatch loop of the main method, which will finally do the jump.
+In theory, we could implement things differently and jumping directly from a child method to another one, but in that case the call stack could grows indefinitely in case of a tight loop between two blocks residing in different methods.
+
+To implement the dispatch block we can exploit the switch opcode of the CLI; if the .NET JIT is smart enough, it can render it using an indirect jump; overall, jumping to a non-local block consists of an indirect function call (by invoking the delegate) plus an indirect jump (by executing the switch opcode); even if this is more costly than a simple direct jump, we will see in the next section that this not the main source of overhead when following a non-local link.
+
+Obviously, the slow dispatching logic is needed only when we want to jump to a non-local block; if the target block happens to reside in the same method as the current one, we can directly jump to it, completely removing the overhead.
+
+Moreover, the dispatch blocks are emitted only if needed, i.e.
+if the parent graph contains at least one flexswitch; graphs without flexswitches are rendered in the obvious way, by making one method per graph.
+The slow bit: passing arguments
+
+Jumping to the correct block is not enough to follow a link: as we said before, each link carries a set of arguments to be passed from the source to the target block.
+As usual, passing arguments across local links is easy, as we can just use local variables to hold their values; on the other hand, non-local links make things more complex.
+
+The only way to jump to a block is to invoke its containing method, so the first solution that comes to mind is to specify its input arguments as parameter of the method; however, each block has potentially a different number (and different types) of input arguments than every other block, so we need to think of something else.
+
+An alternative solution could be to compute the union of the sets of input arguments of all the blocks in the method, and use this set as a signature for the method; this way, there would be enough space to specify the input arguments for every block we might want to jump to, each block ignoring the exceeding unused parameters.
+
+Unfortunately, all the children methods must have the very same signature, as they are all called from the same calling site in the dispatch block of the main method.
+Since the union of the set of input arguments (and hence the computed signature) varies from method to method, this solution cannot work.
+
+We might think to determine the signature by computing the union of input arguments of all blocks in the graph; this way, all the children methods would have the same signature.
+But as we said above, the graph grows new blocks at runtime, so we cannot determine in advance which set of input arguments we will need.
+
+To solve the problem we need a way to pass a variable number of arguments without knowing in advance neither their number nor their types. Thus, we use an instance of this class:
+
+public class InputArgs {
+public int[] ints;
+public float[] floats;
+public object[] objs;
+...
+}
+
+Since the fields are arrays, they can grow as needed to contain any number of arguments; arguments whose type is primitive are stored in the ints or floats array, depending on their type; arguments whose type is a reference type are stored in the objs array: it's up to each block to cast each argument back to the needed type.
+
+This solution impose a huge overhead on both writing and reading arguments:
+
+        * when writing, we need to make sure that the arrays are big enough to contains all the arguments we need; if not, we need to allocate a bigger array.
+Moreover, for each argument we store into the array the virtual machine performs a bound-check, even if we know the index will never be out of bounds (because we checked the size of the array in advance);
+        * when reading, the same bound-check is performed for each argument read; moreover, for each value read from the objs array we need to insert a downcast.
+
+To mitigate the performance drop, we avoid to allocate a new InputArgs object each time we do a non-local jump; instead, we preallocate one at the beginning of the main method, and reuse it all the time.
+
+Our benchmarks show that passing arguments in arrays is about 10 times slower than passing them as real parameter of a method. Unfortunately, we couldn't come up with anything better.
+Implement flexswitches
+
+Now, we can exploit all this machinery to implement flexswitches, as this is our ultimate goal.
+As described above, the point is to be able to add new cases at runtime, each case represented as a delegate.
+Here is an excerpt of the C# class that implements a flexswitch that switches over an integer value:
+
+public class IntLowLevelFlexSwitch:
+{
+public uint default_blockid = 0xFFFFFFFF;
+public int numcases = 0;
+public int[] values = new int[4];
+public FlexSwitchCase[] cases = new FlexSwitchCase[4];
+
+public void add_case(int value, FlexSwitchCase c)
+{
+...
+}
+
+public uint execute(int value, InputArgs args)
+{
+for(int i=0; i<numcases; i++)
+if (values[i] == value) {
+ return cases[i](0, args);
+}
+return default_blockid;
+}
+}
+
+For each case, we store both the triggering value and the corresponding delegate; the add_case method takes care to append value and c to the values and cases arrays, respectively (and resize them if necessary).
+The interesting bit is the execute method: it takes a value and a set of input arguments to be passed across the link and jumps to the right block by performing a linear search in the values array.
+
+As shown by previous sections, the first argument of a FlexSwitchCase is the block id to jump to; since when we go through a flexswitch we always want to jump to the first block of the method, we pass the special value 0 as a block id, which precisely means jump to the first block.
+This little optimization let us not to have to explicitly store the block id for the first block of all the cases.
+
+The value returned by execute is the next block id to jump to; if the value is not found in the values array, we return the default_blockid, whose value has been set before by the JIT compiler; default_blockid usually points to a block containing code to restart the JIT compiler again; when the JIT compiler restarts, it emits more code for the missing case, then calls add_case on the flexswitch; from now on, the new blocks are wired into the existing graph, and we finally managed to implement growable graphs.
+Performances
+
+As we saw, implementing growable graphs for CLI is a pain, as the virtual machine offers very little support, so we need an incredible amount of workarounds.
+Moreover, the code generated is much worse than what an assembly backend could produce, and the cost of following a non-local link is very high compared to local links.
+
+However, our first blog post showed that we still get very good performances; how is it possible?
+
+As usual in computer science, most of the time of a running program in spent in a tiny fraction of the code; our benchmark is no exception, and the vast majority of the time is spent in the inner loop that multiplies numbers; the graph is built in such a way that all the blocks that are part of the inner loop reside in the same method, so that all links inside are local (and fast).
+
+Flexswitches and non-local links play a key role to select the right specialized implementation of the inner loop, but once it is selected they are not executed anymore until we have finished the computation.
+
+It is still unclear how things will look like when we will compile the full Python language instead of a toy one; depending on the code, it could be possible to have non-local links inside the inner loop, thus making performance much worse.
+Alternative implementations
+
+Before implementing the solution described here, we carefully studied a lot of possible alternatives, but all of them either didn't work because of a limitation of the virtual machine or they could work but with terrible performances.
+
+In particular, in theory it is possible to implement non-local links using tail calls, by putting each block in its own method and doing a tail call instead of a jump; this would also solve the problem of how to pass arguments, as each method could have its own signature matching the input args of the block.
+I would like to explain this solution in a more detailed way as I think it's really elegant and nice, but since this post is already too long, I'll stop here :-).
+
+In theory, if the .NET JIT were smart enough it could inline and optimize away the tail calls (or at least many of those) and give us very efficient code.
+However, one benchmark I wrote shows that tail calls are up to 10 times slower (!!!) than normal calls, thus making impractical to use them for our purposes.
+Conclusion
+
+Despite the complexity of the implementation, our result are extremely good; the speedup we got is impressive, and it proves that PyPy's approach to JIT compiler can work well also on top of object oriented virtual machines like .NET or the JVM.
+
+Generating bytecode for those machine at runtime is not a new idea; Jython, IronPython, JRuby and other languages have been doing this for years.
+However, Jython and IronPython do only a simple "static" translation, which doesn't take advantage of the informations gathered at runtime to generate better, faster and specialized code.
+Recently, JRuby grew a new strategy to JIT-compile only hotspots, taking advantage of some informations gathered while interpreting the code; this is still a "one-shot" compilation, where the compiled code does not change over time.
+
+To my knowledge, PyPy brings the first example of a language which implements a truly JIT compiler on top of the underlying JIT compiler of the virtual machine, emitting bytecode that changes and adapts over the time.
+If someone knows other languages doing that, I would really like to know more.
+
+Being so innovative, the problem of this approach is that the current virtual machines are not designed to support it in a native way, and this forces us to put a lot of workarounds that slow down the generated code.
+The hope is that in the future the virtual machines will grow features that help us to generate such kind of code.
+The experimental Da Vinci VM seems to go in the right direction, so it is possible that in the future I will try to write a JIT backend for it.
+
+At the moment, the CLI JIT backend is almost complete, and all the hardest problems seems to be solved; the next step is to fix all the remaining bugs and implement some minor feature that it's still missing, then try to apply it to the full Python language and see what is the outcome.
+}

### 2008-12-16 talk/ecoop2009/clibackend.tex

@@
 as a special flow graph block where links to newly created blocks are
-dynamically added whenever new cases are needed. \dacom{in theory links could also be removed, but I
-  do not know whether this possibility has been considered in Rainbow}
+dynamically added whenever new cases are needed.
 
@@
 
-Implementing flexswitches for backends generating assembly code is
-quite straightforward: basically, a new jump has to be inserted in the
+Implementing flexswitches for backends generating machine code is
+not too complex: basically, a new jump has to be inserted in the
 existing code to point to the newly generated code fragment.
@@
 \dacom{comment for Antonio: I am not sure this is the best solution. This cannot work for Java where classes are the basic
-  units}
+  units. Closures will be available only with Java Dolphin and I do
+  not know how much efficient will be}
+In this way, whereas flow graphs without flexswitches are translated
+to a single method, the translation of flow graphs which can dynamically grow because of
+flexswitches will be scattered over several methods.
+Summarizing, the backend behaves in the following way:
+\begin{itemize}
+\item Each flow graph is translated in a collection of methods which
+  can grow dynamically. \dacom{I propose primary/secondary instead of
+    the overloaded terms main/child} Each collection must contain at least one
+  method, called \emph{primary}, which is the first to be created.
+  All other methods, called \emph{secondary}, are added dynamically 
+  whenever a new case is added to a flexswitch.
 
-\dacom{I still have to polish what comes next}
+\item Each either primary or secondary method corresponds to the
+  translation of some of the blocks of a single flow graph. 
+\end{itemize}
 
-\newcommand{\commentout}[1]{}
+When  a new case is added to a flexswitch, new blocks are generated
+and translated by the backend in a new single method pointed
+by a delegate which is stored in the code implementing the flexswitch,
+so that the method can be invoked later.
+
+\subsubsection{Non local links}
+
 \commentout{
-Because of all these constraints we cannot simply map each graph to its own method, since we saw that our graphs can grow after they have already been executed few times.
 
-Hence, we need to distinguish between the two concepts:
-
-        * a graph is the logical unit of code as seen by the JIT compiler: concretely, the CLI JIT backend renders it as one or more methods;
-        * a method is a collection of basic blocks; each method has the so called parent graph, i.e. the graph its blocks logically belongs to.
-
-The first method of a graph is called main method (which has nothing to do with the Main static methods found in .exe files); other methods are called children methods.
-
-When we want to add a new case to the flexswitch, we create a method containing all the new code; then we wrap the method inside a delegate (the .NET equivalent of a function pointer) and pass it to the flexswitch, so that it can later invoke it.
 The hard bit: non-local links
@@
 if the parent graph contains at least one flexswitch; graphs without flexswitches are rendered in the obvious way, by making one method per graph.
+
 The slow bit: passing arguments
@@
 }
+
+% LocalWords:  flexswitches backend flexswitch

### 2008-12-16 talk/ecoop2009/benchmarks.tex
start the benchmark section

\section{The TLC language}

\anto{maybe we should move this section somewhere else, if we want to use TLC
  as a running example in other sections}

In this section, we will briefly describe \emph{TLC}, a simple dynamic
language that we developed to exercise our JIT compiler generator.  As most of
dynamic languages around, \emph{TLC} is implemented through a virtual machine
that interprets a custom bytecode. Since our main interest is in the runtime
performances of the VM, we did not implement the parser nor the bytecode
compiler, but only the VM itself.

TLC provides four different types:
\begin{enumerate}
\item Integers
\item \lstinline{nil}, whose only value is the null value
\item Objects
\item Lisp-like lists
\end{enumerate}

Objects represent a collection of named attributes (much like Javascript or
SELF) and named methods.  At creation time, it is necessary to specify the set
of attributes of the object, as well as its methods.  Once the object has been
created, it is not possible to add/remove attributes and methods.

The virtual machine is stack-based, and provides several operations:

\begin{itemize}
\item \textbf{Stack manipulation}: standard operations to manipulate the
  stack, such as \lstinline{PUSH}, \lstinline{POP}, \lstinline{SWAP}, etc.
\item \textbf{Flow control} to do conditional and unconditional jumps.
\item \textbf{Arithmetic}: numerical operations on integers, like
  \lstinline{ADD}, \lstinline{SUB}, etc.
\item \textbf{Comparisons} like \lstinline{EQ}, \lstinline{LT},
  \lstinline{GT}, etc.
\item \textbf{Object-oriented}: operations on objects: \lstinline{NEW},
  \lstinline{GETATTR}, \lstinline{SETATTR}, \lstinline{SEND}.
\item \textbf{List operations}: \lstinline{CONS}, \lstinline{CAR},
  \lstinline{CDR}.
\end{itemize}

Obviously, not all the operations are applicable to all objects. For example,
it is not possibile to \lstinline{ADD} an integer and an object, or reading an
attribute from an object which does not provide it.  Being a dynamic language,
the VM needs to do all these checks at runtime; in case one of the check
fails, the execution is simply aborted.

\anto{should we try to invent a syntax for TLC and provide some examples?}

\section{Benchmarks}

Despite being very simple and minimalistic, \lstinline{TLC} is a good
candidate as a language to run benchmarks, as it has some of the features that
makes most of current dynamic languages so slow:

\begin{itemize}

\item \textbf{Stack based VM}: this kind of VM requires all the operands to be
  on top of the evaluation stack.  As a consequence programs spend a lot of
  time pushing and popping values to/from the stack, or doing other stack
  related operations.  However, thanks ot its semplicity this is still the
  most common and preferred way to implement VMs.

\item \textbf{Boxed integers}: integer objects are internally represented as
  an instance of the \lstinline{IntObj} class, whose field \lstinline{value}
  contains the real value.  By having boxed integers, common ariithmetic
  operations are made very slow, because each time we want to load/store their
  value we need to go through an extra level of indirection.  Moreover, in
  case of a complex expression, it is necessary to create a lot of temporary
  objects to hold intermediate results.

\item \textbf{Dynamic lookup}: attributes and methods are looked up at
  runtime, because there is no way to know in advance if and where an object
  have that particular attribute or method.
\end{itemize}

### 2008-12-16 talk/ecoop2009/jitgen.tex
steal some sections from the EU report

@@
 Python interpreter.
+\cfbolz{note to self: steal bits from the master thesis?}
+
+\cfbolz{I would propose to use either TLC as an example here, or something that
+looks at least like an interpreter loop}
 
@@
 
+Partial evaluation (PE) comes in two flavors:
+
+\begin{itemize}
+\item \emph{On-line} partial evaluation: a compiler-like algorithm takes the
+source code of the function \texttt{f(x, y)} (or its intermediate representation,
+i.e. its control flow graph), and some partial
+information, e.g. \texttt{x=5}.  From this, it produces the residual function
+\texttt{g(y)} directly, by following in which operations the knowledge \texttt{x=5} can
+be used, which loops can be unrolled, etc.
+
+\item \emph{Off-line} partial evalution: in many cases, the goal of partial
+evaluation is to improve performance in a specific application.  Assume that we
+have a single known function \texttt{f(x, y)} in which we think that the value
+of \texttt{x} will change slowly during the execution of our program – much
+more slowly than the value of \texttt{y}.  An obvious example is a loop that
+calls \texttt{f(x, y)} many times with always the same value \texttt{x}.  We
+could then use an on-line partial evaluator to produce a \texttt{g(y)} for each
+new value of \texttt{x}.  In practice, the overhead of the partial evaluator
+might be too large for it to be executed at run-time.  However, if we know the
+function \texttt{f} in advance, and if we know \emph{which} arguments are the
+ones that we will want to partially evaluate \texttt{f} with, then we do not
+need a full compiler-like analysis of \texttt{f} every time the value of
+\texttt{x} changes.  We can precompute once and for all a specialized function
+\texttt{f1(x)}, which when called produces the residual function \texttt{g(y)}
+corresponding to \texttt{x}.  This is \emph{off-line partial evaluation}; the
+specialized function \texttt{f1(x)} is called a \emph{generating extension}.
+\end{itemize}
+
+Off-line partial evaluation is usually based on \emph{binding-time analysis}, which
+is the process of determining among the variables used in a function (or
+a set of functions) which ones are going to be known in advance and
+which ones are not.  In the example of \texttt{f(x, y)}, such an analysis
+would be able to infer that the constantness of the argument \texttt{x}
+implies the constantness of many intermediate values used in the
+function.  The \emph{binding time} of a variable determines how early the
+value of the variable will be known.
+
+\cfbolz{XXX: unclear how the next paragraph will fit into the text in the end.
+it's certainly wrong as it is}
+Once binding times have been determined, one possible approach to
+producing the generating extension itself is by self-applying on-line
+partial evaluators.  This is known as the second Futamura projection
+\cite{Futamura99}.  So far it is unclear if this approach can lead to optimal
+results, or even if it scales well.  In PyPy we selected a more direct
+approach: the generating extension is produced by transformation of the
+control flow graphs of the interpreter, guided by the binding times.  We
+call this process \emph{timeshifting}.
+
 \subsection{Execution steps}
 
-* Translation time
+PyPy contains a framework for generating just-in-time compilers using
+off-line partial evaluation.  As such, there are three distinct phases:
 
-  - pypy-c-jit is translated into an executable
+\begin{itemize}
+\item \emph{Translation time:} during the normal translation of an RPython
+program, say the TLC interpreter, we perform binding-time analysis on the
+interpreter.  This produces a generating extension, which is linked with the
+rest of the program. \cfbolz{XXX not quite right}
 
-  - the JIT compiler is automatically generated
+\item \emph{Compile time:} during the execution of the program, when a new
+bytecode is about to be interpreted, the generating extension is invoked
+instead.  As the generating extension is a compiler, all the computations it
+performs are called compile-time computations.  Its sole effect is to produce
+residual code.
 
-* Compile-time: the JIT compiler runs
+\item \emph{Run time:} the normal execution of the program (which includes the
+time spent running the residual code created by the generating extension).
+\end{itemize}
 
-* Runtime: the JIT compiled code runs
+Translation time is a purely off-line phase; compile time and run time are
+actually highly interleaved during the execution of the program.
 
-* Compile-time and runtime are intermixed
+\subsection{Binding Time Analysis}
+
+At translation time, PyPy performs binding-time analysis of the source
+RPython program after it has been turned to low-level graphs, i.e. at
+the level at which operations manipulate pointer-and-structure-like
+objects.
+
+The binding-time terminology that we are using in PyPy is based on the
+colors that we use when displaying the control flow graphs:
+
+\begin{itemize}
+\item \emph{Green} variables contain values that are known at compile-time;
+\item \emph{Red} variables contain values that are not known until run-time.
+\end{itemize}
+
+The binding-time analyzer of our translation tool-chain is based on the
+same type inference engine that is used on the source RPython program,
+the annotator.  In this mode, it is called the \emph{hint-annotator}; it
+operates over input graphs that are already low-level instead of
+RPython-level, and propagates annotations that do not track types but
+value dependencies and manually-provided binding time hints.
+
+The normal process of the hint-annotator is to propagate the binding
+time (i.e. color) of the variables using the following kind of rules:
+
+\begin{itemize}
+\item For a foldable operation (i.e. one without side effect and which depends
+only on its argument values), if all arguments are green, then the result can
+be green too.
+
+\item Non-foldable operations always produce a red result.
+
+\item At join points, where multiple possible values (depending on control
+flow) are meeting into a fresh variable, if any incoming value comes from a red
+variable, the result is red.  Otherwise, the color of the result might be
+green.  We do not make it eagerly green, because of the control flow
+dependency: the residual function is basically a constant-folded copy of the
+source function, so it might retain some of the same control flow.  The value
+that needs to be stored in the fresh join variable thus depends on which
+branches are taken in the residual graph.
+\end{itemize}
+
+\subsubsection{Hints}
+
+Our goal in designing our approach to binding-time analysis was to
+minimize the number of explicit hints that the user must provide in
+the source of the RPython program.  This minimalism was not pushed to
+extremes, though, to keep the hint-annotator reasonably simple.
+
+The driving idea was that hints should be need-oriented.  Indeed, in a
+program like an interpreter, there are a small number of places where it
+would be clearly beneficial for a given value to be known at
+compile-time, i.e. green: this is where we require the hints to be
+added.
+
+The hint-annotator assumes that all variables are red by default, and
+then propagates annotations that record dependency information.
+When encountering the user-provided hints, the dependency information
+is used to make some variables green.  All
+hints are in the form of an operation \texttt{hint(v1, someflag=True)}
+which semantically just returns its first argument unmodified.
+
+The crucial need-oriented hint is \texttt{v2 = hint(v1, concrete=True)}
+which should be used in places where the programmer considers the
+knowledge of the value to be essential.  This hint is interpreted by
+the hint-annotator as a request for both \texttt{v1} and \texttt{v2} to be green.  It
+has a \emph{global} effect on the binding times: it means that not only
+\texttt{v1} but all the values that \texttt{v1} depends on – recursively –
+are forced to be green.  The hint-annotator complains if the
+dependencies of \texttt{v1} include a value that cannot be green, like
+a value read out of a field of a non-immutable structure.
+
+Such a need-oriented backward propagation has advantages over the
+commonly used forward propagation, in which a variable is compile-time
+if and only if all the variables it depends on are also compile-time.  A
+known issue with forward propagation is that it may mark as compile-time
+either more variables than expected (which leads to over-specialization
+of the residual code), or less variables than expected (preventing
+specialization to occur where it would be the most useful).  Our
+need-oriented approach reduces the problem of over-specialization, and
+it prevents under-specialization: an unsatisfiable \texttt{hint(v1,
+concrete=True)} is reported as an error.
+
+In our context, though, such an error can be corrected.  This is done by
+promoting a well-chosen variable among the ones that \texttt{v1} depends on.
+
+Promotion is invoked with the use of a hint as well:
+\texttt{v2 = hint(v1, promote=True)}.
+This hint is a \emph{local} request for \texttt{v2} to be green, without
+requiring \texttt{v1} to be green.  Note that this amounts to copying
+a red value into a green one, which is not possible in classical
+approaches to partial evaluation.  See the Promotion section XXX ref for a
+complete discussion of promotion.

### 2008-12-16 talk/ecoop2009/clibackend.tex

@@
 \item Each either primary or secondary method corresponds to the
-  translation of some of the blocks of a single flow graph. 
+  translation of some of the blocks of a single flow graph. Each
+  method has an initial block whose input variables are the
+  parameters of the method; the input variables of all other blocks
+  are local variables of the method.
 \end{itemize}
@@
 
-\subsubsection{Non local links}
+\subsubsection{Internal and external links}
+
+A link is called \emph{internal} if it connects two blocks implemented in the same method,
+ \emph{external} otherwise.
+
+Following an internal link is not difficult in CLI bytecode: a jump to
+the corresponding code fragment in the same method is emitted 
+to execute the new block, whereas the appropriate local variables are
+used for passing arguments.
+Also following an external link whose target is the initial block of a
+method is not difficult: the corresponding method has to be invoked
+with the appropriate arguments.
 
@@
 
-The hard bit: non-local links
+Using this approach, after a while the blocks of our original graph are scattered over a lot of different methods; however, there are no constraints about how these blocks can be linked together, so it happens to have links between blocks which are not in the same method.
+In the following, we will refer to them as external links.
 
-Using this approach, after a while the blocks of our original graph are scattered over a lot of different methods; however, there are no constraints about how these blocks can be linked together, so it happens to have links between blocks which are not in the same method.
-In the following, we will refer to them as non-local links.
-
-If the non-local block we want to jump to happens to be at the beginning of its containing method, it is enough to invoke the method; but, what if we want to jump somewhere in the middle?
+If the external block we want to jump to happens to be at the beginning of its containing method, it is enough to invoke the method; but, what if we want to jump somewhere in the middle?
 What we really want is to produce a method which has multiple entry-points; again, doing it in assembly would be trivial, but the virtual machine does not provide any support for it, so we need a work around.
@@
 The following picture shows a graph composed of three methods; the id of each method is shown in red, while the block ids are shown in red (for the method id part) and black (for the block number part).
-The graph contains three non-local links; in particular, note the link between blocks 0x00020001 and 0x00010001 which connects two block that resides in different methods.
+The graph contains three external links; in particular, note the link between blocks 0x00020001 and 0x00010001 which connects two block that resides in different methods.
 
 Every method contains a special dispatch block, (not shown in the picture above) whose goal is to jump to the specified block number inside the method itself.
-The first argument of a child method is always a block id; when the method starts, it immediately jumps to the dispatch block, and thus to the desired block.
+The first argument of a secondary method is always a block id; when the method starts, it immediately jumps to the dispatch block, and thus to the desired block.
 
@@
 
-Whenever we want to jump to a non-local block, it is enough to store the block id in the appropriate variable and jump to the dispatch block.
-If the block resides in a different method, the jump_to_unknown block is entered; this special block is implemented differently by the main method and the child methods, as we will see soon.
+Whenever we want to jump to a external block, it is enough to store the block id in the appropriate variable and jump to the dispatch block.
+If the block resides in a different method, the jump_to_unknown block is entered; this special block is implemented differently by the main method and the secondary methods, as we will see soon.
 
@@
 
-The jump_to_unknown block of the main method uses this array to select the right method, and calls it (FlexSwitchCase is the type of delegates for all children methods):
+The jump_to_unknown block of the main method uses this array to select the right method, and calls it (FlexSwitchCase is the type of delegates for all secondary methods):
 
@@
 
-Each child method returns a block id specifying the next block to jump to; after its execution, we assign the return value to the blockid variable, and jump again to the dispatch block, which will jump again to the appropriate block.
+Each secondary method returns a block id specifying the next block to jump to; after its execution, we assign the return value to the blockid variable, and jump again to the dispatch block, which will jump again to the appropriate block.
 
-Keeping this in mind, it is straightforward to implement the jump_to_unknown block of children methods: it is enough to return the target block id to the caller, and let its dispatch loop do the right thing.
-If the caller is also a child method, it will return it again, until we reach the dispatch loop of the main method, which will finally do the jump.
-In theory, we could implement things differently and jumping directly from a child method to another one, but in that case the call stack could grows indefinitely in case of a tight loop between two blocks residing in different methods.
+Keeping this in mind, it is straightforward to implement the jump_to_unknown block of secondary methods: it is enough to return the target block id to the caller, and let its dispatch loop do the right thing.
+If the caller is also a secondary method, it will return it again, until we reach the dispatch loop of the main method, which will finally do the jump.
+In theory, we could implement things differently and jumping directly from a secondary method to another one, but in that case the call stack could grows indefinitely in case of a tight loop between two blocks residing in different methods.
 
-To implement the dispatch block we can exploit the switch opcode of the CLI; if the .NET JIT is smart enough, it can render it using an indirect jump; overall, jumping to a non-local block consists of an indirect function call (by invoking the delegate) plus an indirect jump (by executing the switch opcode); even if this is more costly than a simple direct jump, we will see in the next section that this not the main source of overhead when following a non-local link.
+To implement the dispatch block we can exploit the switch opcode of the CLI; if the .NET JIT is smart enough, it can render it using an indirect jump; overall, jumping to a external block consists of an indirect function call (by invoking the delegate) plus an indirect jump (by executing the switch opcode); even if this is more costly than a simple direct jump, we will see in the next section that this not the main source of overhead when following a external link.
 
-Obviously, the slow dispatching logic is needed only when we want to jump to a non-local block; if the target block happens to reside in the same method as the current one, we can directly jump to it, completely removing the overhead.
+Obviously, the slow dispatching logic is needed only when we want to jump to a external block; if the target block happens to reside in the same method as the current one, we can directly jump to it, completely removing the overhead.
 
@@
 Jumping to the correct block is not enough to follow a link: as we said before, each link carries a set of arguments to be passed from the source to the target block.
-As usual, passing arguments across local links is easy, as we can just use local variables to hold their values; on the other hand, non-local links make things more complex.
+As usual, passing arguments across internal links is easy, as we can just use local variables to hold their values; on the other hand, external links make things more complex.
 
@@
 
-Unfortunately, all the children methods must have the very same signature, as they are all called from the same calling site in the dispatch block of the main method.
+Unfortunately, all the secondary methods must have the very same signature, as they are all called from the same calling site in the dispatch block of the main method.
 Since the union of the set of input arguments (and hence the computed signature) varies from method to method, this solution cannot work.
 
-We might think to determine the signature by computing the union of input arguments of all blocks in the graph; this way, all the children methods would have the same signature.
+We might think to determine the signature by computing the union of input arguments of all blocks in the graph; this way, all the secondary methods would have the same signature.
 But as we said above, the graph grows new blocks at runtime, so we cannot determine in advance which set of input arguments we will need.
@@
 
-To mitigate the performance drop, we avoid to allocate a new InputArgs object each time we do a non-local jump; instead, we preallocate one at the beginning of the main method, and reuse it all the time.
+To mitigate the performance drop, we avoid to allocate a new InputArgs object each time we do a external jump; instead, we preallocate one at the beginning of the main method, and reuse it all the time.
 
@@
 As we saw, implementing growable graphs for CLI is a pain, as the virtual machine offers very little support, so we need an incredible amount of workarounds.
-Moreover, the code generated is much worse than what an assembly backend could produce, and the cost of following a non-local link is very high compared to local links.
+Moreover, the code generated is much worse than what an assembly backend could produce, and the cost of following a external link is very high compared to internal links.
 
@@
 
-As usual in computer science, most of the time of a running program in spent in a tiny fraction of the code; our benchmark is no exception, and the vast majority of the time is spent in the inner loop that multiplies numbers; the graph is built in such a way that all the blocks that are part of the inner loop reside in the same method, so that all links inside are local (and fast).
+As usual in computer science, most of the time of a running program in spent in a tiny fraction of the code; our benchmark is no exception, and the vast majority of the time is spent in the inner loop that multiplies numbers; the graph is built in such a way that all the blocks that are part of the inner loop reside in the same method, so that all links inside are internal (and fast).
 
-Flexswitches and non-local links play a key role to select the right specialized implementation of the inner loop, but once it is selected they are not executed anymore until we have finished the computation.
+Flexswitches and external links play a key role to select the right specialized implementation of the inner loop, but once it is selected they are not executed anymore until we have finished the computation.
 
-It is still unclear how things will look like when we will compile the full Python language instead of a toy one; depending on the code, it could be possible to have non-local links inside the inner loop, thus making performance much worse.
+It is still unclear how things will look like when we will compile the full Python language instead of a toy one; depending on the code, it could be possible to have external links inside the inner loop, thus making performance much worse.
 Alternative implementations
@@
 
-In particular, in theory it is possible to implement non-local links using tail calls, by putting each block in its own method and doing a tail call instead of a jump; this would also solve the problem of how to pass arguments, as each method could have its own signature matching the input args of the block.
+In particular, in theory it is possible to implement external links using tail calls, by putting each block in its own method and doing a tail call instead of a jump; this would also solve the problem of how to pass arguments, as each method could have its own signature matching the input args of the block.
 I would like to explain this solution in a more detailed way as I think it's really elegant and nice, but since this post is already too long, I'll stop here :-).

### 2008-12-16 talk/ecoop2009/clibackend.tex
added picture with method and block ids

@@
 
+What cannot be easily implemented in CLI is following an external link
+whose target is not an initial block; consider, for instance, the
+outgoing link of the block dynamically added in the right-hand side
+picture of Figure~\ref{flexswitch-fig}. How it is possible to pass the
+right arguments to the target block?
+
+To solve this problem a special block, called \emph{dispatcher}, is
+added to every method; whenever a method is invoked, its dispatcher is
+executed first\footnote{Recall that the dispatcher is a special block
+and must not be confused with the initial block of a method.} to
+determine which block has to be executed.
+This is done by passing to the method a 32 bits number, called 
+\emph{block id}, which uniquely identifies the next block of the graph to be executed.
+The high word of a block id is the id of the method to which the block
+belongs, whereas the low word is a progressive number identifying
+each block implemented by the method.
+
+The picture in Figure~\ref{block-id-fig} shows a graph composed of three methods (for
+simplicity, dispatchers are not shown); method ids are in red, whereas
+block numbers are in black. 
+The graph contains three external links; in particular, note the link
+between blocks \texttt{0x00020001} and \texttt{0x00010001} which
+connects two blocks implemented by different methods.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[height=6cm]{blockid}
+\caption{Method and block ids.}\label{block-id-fig}
+\end{center}
+\end{figure}
+
 \commentout{
-
-Using this approach, after a while the blocks of our original graph are scattered over a lot of different methods; however, there are no constraints about how these blocks can be linked together, so it happens to have links between blocks which are not in the same method.
-In the following, we will refer to them as external links.
-
-If the external block we want to jump to happens to be at the beginning of its containing method, it is enough to invoke the method; but, what if we want to jump somewhere in the middle?
-What we really want is to produce a method which has multiple entry-points; again, doing it in assembly would be trivial, but the virtual machine does not provide any support for it, so we need a work around.
-
 Each method in a graph is assigned an unique 16 bit method id; each block in a method is assigned a progressive 16 bit block number.

### 2008-12-17 talk/ecoop2009/benchmarks.tex
steal more text

@@
   as a running example in other sections}
+\cfbolz{yes please! how about making it the next section after the
+introduction?}
 
@@
 that interprets a custom bytecode. Since our main interest is in the runtime
-performances of the VM, we did not implement the parser nor the bytecode
+performance of the VM, we did not implement the parser nor the bytecode
 compiler, but only the VM itself.
@@
 
-Objects represent a collection of named attributes (much like Javascript or
-SELF) and named methods.  At creation time, it is necessary to specify the set
+Objects represent a collection of named attributes (much like JavaScript or
+Self) and named methods.  At creation time, it is necessary to specify the set
 of attributes of the object, as well as its methods.  Once the object has been
@@
 Obviously, not all the operations are applicable to all objects. For example,
-it is not possibile to \lstinline{ADD} an integer and an object, or reading an
+it is not possible to \lstinline{ADD} an integer and an object, or reading an
 attribute from an object which does not provide it.  Being a dynamic language,
@@
 \anto{should we try to invent a syntax for TLC and provide some examples?}
+\cfbolz{we should provide an example with the assembler syntax}
 
 \section{Benchmarks}
+
+\cfbolz{I think this should go to the beginning of the description of the TLC as
+it explains why it is written as it is written}
 
@@
   time pushing and popping values to/from the stack, or doing other stack
-  related operations.  However, thanks ot its semplicity this is still the
+  related operations.  However, thanks to its simplicity this is still the
   most common and preferred way to implement VMs.
@@
   an instance of the \lstinline{IntObj} class, whose field \lstinline{value}
-  contains the real value.  By having boxed integers, common ariithmetic
+  contains the real value.  By having boxed integers, common arithmetic
   operations are made very slow, because each time we want to load/store their
   value we need to go through an extra level of indirection.  Moreover, in
-  case of a complex expression, it is necessary to create a lot of temporary
+  case of a complex expression, it is necessary to create many temporary
   objects to hold intermediate results.

### 2008-12-17 talk/ecoop2009/jitgen.tex
steal more text

@@
 
-The original idea is by Futamura \cite{Futamura99}. He proposed to generate compilers
-from interpreters with automatic specialization, but his work has had
-relatively little practical impact so far.
-
 \subsection{Partial evaluation}
 
-Assume the Python bytecode to be constant, and constant-propagate it into the
-Python interpreter.
-\cfbolz{note to self: steal bits from the master thesis?}
+In 1971 Yoshihiko Futamura published a paper \cite{Futamura99} that proposed a
+technique to automatically transform an interpreter of a programming language
+into a compiler for the same language. This would solve the problem of having to
+write a compiler instead of a much simpler interpreter. He proposed to use
+partial evaluation to achieve this goal. He defined partial evaluation along the following lines:
+
+Given a program $P$ with $m + n$ input variables $s_1, ..., s_m$ and $d_1, ...,
+d_n$, the partial evaluation of $P$ with respect to concrete values $s'_1, ...,
+s'_m$ for the first $m$ variables is a program $P'$. The program $P'$ takes only
+the input variables $d_1, ..., d_n$ but behaves exactly like $P$ with the
+concrete values (but is hopefully more efficient). This transformation is done
+by a program $S$, the partial evaluator, which takes $P$ and $s_1, ..., s_m$ as
+input:
+
+    $$S(P, (s'_1, ..., s'_m)) = P'$$
+
+The variables $s_1, ..., s_m$ are called the \emph{static} variables, the
+variables $d_1, ..., d_n$ are called the \emph{dynamic} variables; $P'$ is the
+\emph{residual code}. Partial evaluation creates a version of $P$ that works
+only for a fixed set of inputs for the first $m$ arguments. This effect is
+called \emph{specialization}.
+
+When $P$ is an interpreter for a programming language, then the $s_1, ..., s_m$
+are chosen such that they represent the program that the interpreter is
+interpreting and the $d_1, ..., d_n$ represent the input of this program. Then
+$P'$ can be regarded as a compiled version of the program that the chosen $s'_1,
+..., s'_m$ represent, since it is a version of the interpreter that can only
+interpret this program. Now once the partial evaluator $S$ is implemented, it is
+actually enough to implement an interpreter for a new language and use $S$
+together with this interpreter to compile programs in that new language.
+
+A valid implementation for $S$ would be to just put the concrete values into $P$
+to get $P'$, which would not actually produce any performance benefits compared with
+directly using $P$. A good implementation for $S$ should instead make use of the
+information it has and evaluate all the parts of the program that actually
+depend only on the $s_1, ..., s_m$ and to remove parts of $P$ that cannot be
+reached given the concrete values.
 
@@
 
-\subsection{Execution steps}
+XXX write something about the problems of classical PE?
+
+\subsection{Partial Evaluation in PyPy}
 
@@
 \begin{itemize}
-\item \emph{Green} variables contain values that are known at compile-time;
-\item \emph{Red} variables contain values that are not known until run-time.
+\item \emph{Green} variables contain values that are known at compile-time.
+They correspond to static arguments.
+\item \emph{Red} variables contain values that are usually not known
+compile-time. They correspond to dynamic arguments.
 \end{itemize}
 
-The binding-time analyzer of our translation tool-chain is based on the
+The binding-time analyzer of our translation tool-chain is using a simple
+abstract-interpretation based analysis. It is based on the
 same type inference engine that is used on the source RPython program,
@@
 value dependencies and manually-provided binding time hints.
+XXX the above needs rewriting when the background section is there
 
@@
 \texttt{v1} but all the values that \texttt{v1} depends on – recursively –
-are forced to be green.  The hint-annotator complains if the
+are forced to be green.  The hint-annotator gives an error if the
 dependencies of \texttt{v1} include a value that cannot be green, like
-a value read out of a field of a non-immutable structure.
+a value read out of a field of a non-immutable instance.
 
@@
 
-In our context, though, such an error can be corrected.  This is done by
-promoting a well-chosen variable among the ones that \texttt{v1} depends on.
-
 Promotion is invoked with the use of a hint as well:

### 2008-12-17 talk/ecoop2009/rainbow.tex
make comments stand out more, visually
steal more text

@@
 
-Here are summarized the various phases of the JIT:
-
-Translation time:
-
-  * Low-level flowgraphs are produced
-
-  * The *hint-annotator* colors the variables
-
-  * The *rainbow codewriter* translates flowgraphs into rainbow bytecode
-
-Compile-time:
-
-  * The rainbow interpreter executes the bytecode
-
-  * As a result, it produces executable code
-
-Runtime:
-
-  * The produced code is executed
+\cfbolz{XXX I think we should be very careful with the rainbow interp. it is a
+total implementation-detail and we should only describe it as little as
+possible}
 
@@
 
-\subsection{Promotion}
+\section{Promotion}
+
+In the sequel, we describe in more details one of the main new
+techniques introduced in our approach, which we call \emph{promotion}.  In
+short, it allows an arbitrary run-time value to be turned into a
+compile-time value at any point in time.  Promotion is thus the central way by
+which we make use of the fact that the JIT is running interleaved with actual
+program execution. Each promotion point is explicitly defined with a hint that
+must be put in the source code of the interpreter.
+
+From a partial evaluation point of view, promotion is the converse of
+the operation generally known as "lift" \cite{XXX}.  Lifting a value means
+copying a variable whose binding time is compile-time into a variable
+whose binding time is run-time – it corresponds to the compiler
+"forgetting" a particular value that it knew about.  By contrast,
+promotion is a way for the compiler to gain \emph{more} information about
+the run-time execution of a program. Clearly, this requires
+fine-grained feedback from run-time to compile-time, thus a
+dynamic setting.
+
+Promotion requires interleaving compile-time and run-time phases,
+otherwise the compiler can only use information that is known ahead of
+time. It is impossible in the "classical" approaches to partial
+evaluation, in which the compiler always runs fully ahead of execution
+This is a problem in many large use cases.  For example, in an
+interpreter for a dynamic language, there is mostly no information
+that can be clearly and statically used by the compiler before any
+code has run.
+
+A very different point of view on promotion is as a generalization of
+techniques that already exist in dynamic compilers as found in modern
+object-oriented language virtual machines.  In this context feedback
+techniques are crucial for good results.  The main goal is to
+optimize and reduce the overhead of dynamic dispatching and indirect
+invocation.  This is achieved with variations on the technique of
+polymorphic inline caches \cite{XXX}: the dynamic lookups are cached and
+the corresponding generated machine code contains chains of
+compare-and-jump instructions which are modified at run-time.  These
+techniques also allow the gathering of information to direct inlining for even
+better optimization results.
+
+In the presence of promotion, dispatch optimization can usually be
+reframed as a partial evaluation task.  Indeed, if the type of the
+object being dispatched to is known at compile-time, the lookup can be
+folded, and only a (possibly inlined) direct call remains in the
+generated code.  In the case where the type of the object is not known
+at compile-time, it can first be read at run-time out of the object and
+promoted to compile-time.  As we will see in the sequel, this produces
+very similar machine code \footnote{This can also be seen as a generalization of
+a partial evaluation transformation called "The Trick" (see e.g. \cite{XXX}),
+which again produces similar code but which is only applicable for finite sets
+of values.}.
+
+The essential advantage is that it is no longer tied to the details of
+the dispatch semantics of the language being interpreted, but applies in
+more general situations.  Promotion is thus the central enabling
+primitive to make partial evaluation a practical approach to language
+independent dynamic compiler generation.
+
+\subsection{Promotion as Applied to the TLC}
+
+XXX
+
+\subsection{Promotion in Practise}
 
@@
 
-  * (compile time): the rainbow interpreter produces machine code until it
+\begin{itemize}
+  \item (compile time): the rainbow interpreter produces machine code until it
     hits a promotion point; e.g.::
@@
 
-  * (compile time): at this point, it generates special machine code that when
+  \item (compile time): at this point, it generates special machine code that when
     reached calls the JIT compiler again; the JIT compilation stops::
@@
 
-  * (runtime): the machine code is executed; when it reaches a promotion
+  \item (runtime): the machine code is executed; when it reaches a promotion
     point, it executes the special machine code we described in the previous
@@
 
-  * (compile time): now we finally know the exact value of our red variable,
+  \item (compile time): now we finally know the exact value of our red variable,
     and we can promote it to green; suppose that the value of 'y' is 32::
@@
 
-  * (runtime) the execution restart from the point it stopped, until a new
+  \item (runtime) the execution restart from the point it stopped, until a new
     unhandled promotion point is reached.
+\end{itemize}
 
-\subsection{Virtuals and virtualizables}
+\section{Automatic Unboxing of Intermediate Results}
 
-TODO
+XXX the following section needs a rewriting to be much more high-level and to
+compare more directly with classical escape analysis
+
+Interpreters for dynamic languages typically allocate a lot of small
+objects, for example due to boxing.  For this reason, we
+implemented a way for the compiler to generate residual memory
+allocations as lazily as possible.  The idea is to try to keep new
+run-time structures "exploded": instead of a single run-time pointer to
+a heap-allocated data structure, the structure is "virtualized" as a set
+of fresh variables, one per field.  In the compiler, the variable that
+would normally contain the pointer to the structure gets instead a
+content that is neither a run-time value nor a compile-time constant,
+but a special \emph{virtual structure} – a compile-time data structure that
+recursively contains new variables, each of which can again store a
+run-time, a compile-time, or a virtual structure value.
+
+This approach is based on the fact that the "run-time values" carried
+around by the compiler really represent run-time locations – the name of
+a CPU register or a position in the machine stack frame.  This is the
+case for both regular variables and the fields of virtual structures.
+It means that the compilation of a \texttt{getfield} or \texttt{setfield}
+operation performed on a virtual structure simply loads or stores such a
+location reference into the virtual structure; the actual value is not
+copied around at run-time.
+
+It is not always possible to keep structures virtual.  The main
+situation in which it needs to be "forced" (i.e. actually allocated at
+run-time) is when the pointer escapes to some non-virtual location like
+a field of a real heap structure.
+
+Virtual structures still avoid the run-time allocation of most
+short-lived objects, even in non-trivial situations.  The following
+example shows a typical case.  Consider the Python expression \texttt{a+b+c}.
+Assume that \texttt{a} contains an integer.  The PyPy Python interpreter
+implements application-level integers as boxes – instances of a
+\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
+addition of two integers:
+
+XXX needs to use TLC examples
+\begin{verbatim}
+    def add(w1, w2):            # w1, w2 are W_IntObject instances
+        value1 = w1.intval
+        value2 = w2.intval
+        result = value1 + value2
+        return W_IntObject(result)
+\end{verbatim}
+
+When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are
+issued; the intermediate \texttt{W\_IntObject} instance is built by the first
+call and thrown away after the second call.  By contrast, when the
+interpreter is turned into a compiler, the construction of the
+\texttt{W\_IntObject} object leads to a virtual structure whose \texttt{intval}
+field directly references the register in which the run-time addition
+put its result.  This location is read out of the virtual structure at
+the beginning of the second \texttt{add()}, and the second run-time addition
+directly operates on the same register.
+
+An interesting effect of virtual structures is that they play nicely with
+promotion.  Indeed, before the interpreter can call the proper \texttt{add()}
+function for integers, it must first determine that the two arguments
+are indeed integer objects.  In the corresponding dispatch logic, we
+have added two hints to promote the type of each of the two arguments.
+This produces a compiler that has the following behavior: in the general
+case, the expression \texttt{a+b} will generate two consecutive run-time
+switches followed by the residual code of the proper version of
+\texttt{add()}.  However, in \texttt{a+b+c}, the virtual structure representing
+the intermediate value will contain a compile-time constant as type.
+Promoting a compile-time constant is trivial – no run-time code is
+generated.  The whole expression \texttt{a+b+c} thus only requires three
+switches instead of four.  It is easy to see that even more switches can
+be skipped in larger examples; typically, in a tight loop manipulating
+only integers, all objects are virtual structures for the compiler and
+the residual code is theoretically optimal – all type propagation and
+boxing/unboxing occurs at compile-time.

### 2008-12-17 talk/ecoop2009/intro.tex
add an XXX
steal more text

@@
 Moreover, writing a static compiler is often not enough to get high
-performances; IronPython and JRuby are going in the direction of JIT compiling
+performance; IronPython and JRuby are going in the direction of JIT compiling
 specialized versions of the code depending on the actual values/types seen at
-runtime; this approach seems to work, but write it manually requires an
+runtime; this approach seems to work, but writing it manually requires an
 enormous effort.
 
-PyPy's idea is to automatize the generation of static/JIT compilers in order
-to reduce at minimun the effort required to get a fast implementation of a
+\cfbolz{we should cite the dyla paper somewhere here}
+
+PyPy's idea is to automatize the generation of JIT compilers in order
+to reduce to a minimum the effort required to get a fast implementation of a
 dynamic language; all you have to do is to write a high level specification of
-the language (by writing an interpreter), and put it through PyPy's
-translation toolchain.
+the language (by writing an interpreter), and putting it through PyPy's
+translation toolchain. The automatic generation of JIT compilers is done with
+the help of partial evaluation techniques.
 
 \subsection{PyPy and RPython}
+
+\anto{as Armin points out, the two CLI backends can be easily confused; what
+  about renaming the ``CLI Backend for flowgraphs'' into ``CLI bytecode
+  compiler''? Any better idea for the name?}
 
@@
 analysis, code optimizations and several transformations aiming at 
-incrementally providing implementation details as memory management or the threading model.
+incrementally providing implementation details such as memory management or the threading model.
 The different kinds of intermediate codes  which are refined 
@@
 
-Despite the PyPy infrastructure was specifically developed 
-for Python, in fact it can be used for implementing
-other languages. Indeed, PyPy has been successfully experimented with
-several languages as Smalltalk \cite{BolzEtAl08}, JavaScript, Scheme and Prolog.
+Despite having been specifically developed for Python, the PyPy infrastructure
+can in fact be used for implementing other languages. Indeed, there were
+successful experiments of using PyPy to implement several other languages such
+as Smalltalk \cite{BolzEtAl08}, JavaScript, Scheme and Prolog.
 As suggested by Figure~\ref{pypy-fig}, a portable interpreter for a
 generic language $L$  can be
-easily developed once an abstract interpreter for $L$ is implemented in
+easily developed once an interpreter for $L$ has been implemented in
 RPython.
+
+\subsection{PyPy and JIT-Generation}
+
+This section will give a high-level overview of how the JIT-generation process
+works. More details will be given in subsequent sections.
 
@@
 interpreter source.
+
+XXX list contributions clearly

### 2008-12-17 talk/ecoop2009/benchmarks.tex
more sections about benchmarks

@@
 \end{itemize}
+
+In the following sections, we will show some benchmarks that show how our
+generated JIT can handle all the features above very well.
+
+To measure the speedup we get with the JIT, we run each program three times:
+
+\begin{enumerate}
+\item By plain interpretation, without any jitting.
+\item With the jit enabled: this run includes the time spent by doing the
+  compilation itself, plus the time spent by running the produced code.
+\item Again with the jit enabled, but this time the compilation has already
+  been done, so we are actually measuring how good is the code we produced.
+\end{enumerate}
+
+The benchmarks have been run on machine XXX with hardware YYY etc. etc.
+
+\subsection{Arithmetic operations}
+
+To benchmark arithmetic operations between integers, we wrote a simple program
+that computes the factorial of a given number.  The algorithm is
+straightforward, and the loop contains only three operations: one
+multiplication, one subtraction, and one comparison to check if we have
+finished the job.
+
+When doing plain interpretation, we need to create and destroy three temporary
+objects at each iterations.  By contrast, the code generated by the JIT does
+much better.  At the first iteration, the classes of the two operands of the
+multiplication are promoted; then, the JIT compiler knows that both are
+integers, so it can inline the code to compute the result.  Moreover, it can
+\emph{virtualize} all the temporary objects, because they never escape from
+the inner loop.  The same remarks apply to the other two operations inside
+the loop.
+
+As a result, the code executed after the first iteration is close to optimal:
+the intermediate values are stored as \lstinline{int} local variables, and the
+multiplication, subtraction and \emph{less-than} comparison are mapped to a
+single CLI opcode (\lstinline{mul}, \lstinline{sub} and \lstinline{clt},
+respectively).
+
+Moreover, we also wrote a program to calculate the $n_{th}$ Fibonacci number,
+for which we can do the same reasoning as above.
+
+Table XXX and figure XXX show the time spent to calculate the factorial of
+various numbers, with and without the JIT.  Table XXX and figure XXX show the
+same informations for the Fibonacci program.
+
+\anto{Should we say that we get wrong results due to overflow but we don't care?}
+
+As we can see, the code generated by the JIT is almost 500 times faster than
+the non-jitted case, and it is only about 1.5 times slower than the same
+algorithm written in C\#, which can be considered the optimal goal.
+
+\subsection{Object-oriented features}
+
+To measure how the JIT handles object-oriented features, we wrote a very
+simple benchmark that involves attribute lookups and method calls.  We have an
+\emph{accumulator} object, which has a field \lstinline{value} and a method
+\lstinline{add}.  The method \lstinline{add} takes a parameter and adds it the
+field \lstinline{value}.
+
+XXX: update to the new version
+
+Our benchmark accepts a paramter \lstinline{n}, create an \emph{accumulator},
+and repeatedly calls \lstinline{add} on it, passing numbers from
+\lstinline{n-1} to \lstinline{0}.
+
+The JIT is able to completely remove the overhead of object creation,
+attribute lookups and method calls, and the generated code results in a simple
+loop doing additions in-place.
+
+Table XXX and figure XXX show the time spent to run the benchmark with various
+input arguments. Again, we can see that the jitted code is up to 500 times
+faster than the interpreted one.

### 2008-12-17 talk/ecoop2009/tlc.tex
move the section about tlc into its own file, and put it just after the introduction

--- 2008-12-17 talk/ecoop2009/benchmarks.tex
@@
 \section{The TLC language}
-
-\anto{maybe we should move this section somewhere else, if we want to use TLC
-  as a running example in other sections}
-\cfbolz{yes please! how about making it the next section after the
-introduction?}
 
@@
 \cfbolz{we should provide an example with the assembler syntax}
-
-\section{Benchmarks}
-
-\cfbolz{I think this should go to the beginning of the description of the TLC as
-it explains why it is written as it is written}
-
-Despite being very simple and minimalistic, \lstinline{TLC} is a good
-candidate as a language to run benchmarks, as it has some of the features that
-makes most of current dynamic languages so slow:
-
-\begin{itemize}
-
-\item \textbf{Stack based VM}: this kind of VM requires all the operands to be
-  on top of the evaluation stack.  As a consequence programs spend a lot of
-  time pushing and popping values to/from the stack, or doing other stack
-  related operations.  However, thanks to its simplicity this is still the
-  most common and preferred way to implement VMs.
-
-\item \textbf{Boxed integers}: integer objects are internally represented as
-  an instance of the \lstinline{IntObj} class, whose field \lstinline{value}
-  contains the real value.  By having boxed integers, common arithmetic
-  operations are made very slow, because each time we want to load/store their
-  value we need to go through an extra level of indirection.  Moreover, in
-  case of a complex expression, it is necessary to create many temporary
-  objects to hold intermediate results.
-
-\item \textbf{Dynamic lookup}: attributes and methods are looked up at
-  runtime, because there is no way to know in advance if and where an object
-  have that particular attribute or method.
-\end{itemize}

### 2008-12-17 talk/ecoop2009/benchmarks.tex
move the section about tlc into its own file, and put it just after the introduction

@@
-\section{The TLC language}
-
-\anto{maybe we should move this section somewhere else, if we want to use TLC
-  as a running example in other sections}
-\cfbolz{yes please! how about making it the next section after the
-introduction?}
-
-In this section, we will briefly describe \emph{TLC}, a simple dynamic
-language that we developed to exercise our JIT compiler generator.  As most of
-dynamic languages around, \emph{TLC} is implemented through a virtual machine
-that interprets a custom bytecode. Since our main interest is in the runtime
-performance of the VM, we did not implement the parser nor the bytecode
-compiler, but only the VM itself.
-
-TLC provides four different types:
-\begin{enumerate}
-\item Integers
-\item \lstinline{nil}, whose only value is the null value
-\item Objects
-\item Lisp-like lists
-\end{enumerate}
-
-Objects represent a collection of named attributes (much like JavaScript or
-Self) and named methods.  At creation time, it is necessary to specify the set
-of attributes of the object, as well as its methods.  Once the object has been
-created, it is not possible to add/remove attributes and methods.
-
-The virtual machine is stack-based, and provides several operations:
-
-\begin{itemize}
-\item \textbf{Stack manipulation}: standard operations to manipulate the
-  stack, such as \lstinline{PUSH}, \lstinline{POP}, \lstinline{SWAP}, etc.
-\item \textbf{Flow control} to do conditional and unconditional jumps.
-\item \textbf{Arithmetic}: numerical operations on integers, like
-  \lstinline{ADD}, \lstinline{SUB}, etc.
-\item \textbf{Comparisons} like \lstinline{EQ}, \lstinline{LT},
-  \lstinline{GT}, etc.
-\item \textbf{Object-oriented}: operations on objects: \lstinline{NEW},
-  \lstinline{GETATTR}, \lstinline{SETATTR}, \lstinline{SEND}.
-\item \textbf{List operations}: \lstinline{CONS}, \lstinline{CAR},
-  \lstinline{CDR}.
-\end{itemize}
-
-Obviously, not all the operations are applicable to all objects. For example,
-it is not possible to \lstinline{ADD} an integer and an object, or reading an
-attribute from an object which does not provide it.  Being a dynamic language,
-the VM needs to do all these checks at runtime; in case one of the check
-fails, the execution is simply aborted.
-
-\anto{should we try to invent a syntax for TLC and provide some examples?}
-\cfbolz{we should provide an example with the assembler syntax}
-
 \section{Benchmarks}

### 2008-12-17 talk/ecoop2009/tlc.tex
show an example of tlc assembler, and propose a higher level syntax to show more complex examples

@@
 
-\anto{should we try to invent a syntax for TLC and provide some examples?}
-\cfbolz{we should provide an example with the assembler syntax}
+\subsection{TLC examples}
+
+As we said above, TLC exists only at bytecode level; to ease the development
+of TLC programs, we wrote an assembler that generates TLC bytecode. The
+following example shows a simple program that computes the absolute value of
+the given integer:
+
+\begin{lstlisting}
+main:             # stack: []
+    PUSHARG       #        [n]
+    PUSH 0        #        [n, 0]
+    LT            #        [n<0]
+    BR_COND neg
+
+pos:              #        []
+    PUSHARG       #        [n]
+    RETURN
+
+neg:
+    PUSH 0        #        [0]
+    PUSHARG       #        [0,n]
+    SUB           #        [-n]
+    RETURN
+\end{lstlisting}
+
+Since reading TLC programs at bytecode level is hard, in this paper we will
+use an invented Python-like syntax to describe examples, even if we need to
+remind that the actual programs are written in the assembler language showed
+above. The following listing shows the same example as above written in the
+Python-like syntax:
+
+\begin{lstlisting}
+def main(n):
+    if n<0:
+        return -n
+    return n
+\end{lstlisting}

### 2008-12-17 talk/ecoop2009/tlc.tex
minor language tweaks

@@
 it is not possible to \lstinline{ADD} an integer and an object, or reading an
-attribute from an object which does not provide it.  Being a dynamic language,
+attribute from an object which does not provide it.  Being dynamically typed,
 the VM needs to do all these checks at runtime; in case one of the check
@@
 remind that the actual programs are written in the assembler language showed
-above. The following listing shows the same example as above written in the
-Python-like syntax:
+above. Thus, the example above can be written in the following way:
 

### 2008-12-17 talk/ecoop2009/benchmarks.tex
move the paragraph about tlc features to the TLC section, and add the benchmarks written in C#

@@
 
-\cfbolz{I think this should go to the beginning of the description of the TLC as
-it explains why it is written as it is written}
-
-Despite being very simple and minimalistic, \lstinline{TLC} is a good
-candidate as a language to run benchmarks, as it has some of the features that
-makes most of current dynamic languages so slow:
-
-\begin{itemize}
-
-\item \textbf{Stack based VM}: this kind of VM requires all the operands to be
-  on top of the evaluation stack.  As a consequence programs spend a lot of
-  time pushing and popping values to/from the stack, or doing other stack
-  related operations.  However, thanks to its simplicity this is still the
-  most common and preferred way to implement VMs.
-
-\item \textbf{Boxed integers}: integer objects are internally represented as
-  an instance of the \lstinline{IntObj} class, whose field \lstinline{value}
-  contains the real value.  By having boxed integers, common arithmetic
-  operations are made very slow, because each time we want to load/store their
-  value we need to go through an extra level of indirection.  Moreover, in
-  case of a complex expression, it is necessary to create many temporary
-  objects to hold intermediate results.
-
-\item \textbf{Dynamic lookup}: attributes and methods are looked up at
-  runtime, because there is no way to know in advance if and where an object
-  have that particular attribute or method.
-\end{itemize}
+In section \ref{sec:tlc-features}, we saw that TLC provides most of the
+features that usaully make dynamically typed language so slow, such as
+\emph{stack-based VM}, \emph{boxed arithmetic} and \emph{dynamic lookup} of
+methods and attributes.
 
 In the following sections, we will show some benchmarks that show how our
-generated JIT can handle all the features above very well.
+generated JIT can handle all these features very well.
 
@@
 \end{enumerate}
+
+Moreover, for each benchmark we also show the time taken by running the
+equivalent program written in C\#.  By comparing the results against C\#, we
+can see how far we are from the supposed optimal performances.  \anto{I
+  don't really like the last sentence, but right now I can't think of another
+  way to phrase it.  Rewording welcome :-)}
 
@@
 the non-jitted case, and it is only about 1.5 times slower than the same
-algorithm written in C\#, which can be considered the optimal goal.
+algorithm written in C\#: the difference in speed it is probably due to both
+the fact that the current CLI backend emits slightly non-optimal code and that
+the underyling .NET JIT compiler is highly optimized to handle bytecode
+generated by C\# compilers.
 

### 2008-12-17 talk/ecoop2009/tlc.tex
move the paragraph about tlc features to the TLC section, and add the benchmarks written in C#

@@
 
+\subsection{TLC features}
+\label{sec:tlc-features}
+
+Despite being very simple and minimalistic, \lstinline{TLC} is a good
+candidate as a language to test our JIT generator, as it has some of the
+features that makes most of current dynamic languages so slow:
+
+\begin{itemize}
+
+\item \textbf{Stack based VM}: this kind of VM requires all the operands to be
+  on top of the evaluation stack.  As a consequence programs spend a lot of
+  time pushing and popping values to/from the stack, or doing other stack
+  related operations.  However, thanks to its simplicity this is still the
+  most common and preferred way to implement VMs.
+
+\item \textbf{Boxed integers}: integer objects are internally represented as
+  an instance of the \lstinline{IntObj} class, whose field \lstinline{value}
+  contains the real value.  By having boxed integers, common arithmetic
+  operations are made very slow, because each time we want to load/store their
+  value we need to go through an extra level of indirection.  Moreover, in
+  case of a complex expression, it is necessary to create many temporary
+  objects to hold intermediate results.
+
+\item \textbf{Dynamic lookup}: attributes and methods are looked up at
+  runtime, because there is no way to know in advance if and where an object
+  have that particular attribute or method.
+\end{itemize}
+
 \subsection{TLC examples}

### 2008-12-17 talk/ecoop2009/benchmarks.tex
add high level syntax of two benchmarks.

@@
 
-The benchmarks have been run on machine XXX with hardware YYY etc. etc.
+The benchmarks have been run on machine with Intel Pentium 4 CPU running at
+3.20 GHz and 2 GB of RAM, running Microsoft Windows XP and Microsoft .NET
+Framework 2.0.
 
@@
 multiplication, one subtraction, and one comparison to check if we have
-finished the job.
+finished the job:
+
+\begin{lstlisting}
+def main(n):
+    result = 1
+    while n > 1:
+        result = result * n
+        n = n - 1
+    return n
+\end{lstlisting}
 
@@
 
-Moreover, we also wrote a program to calculate the $n_{th}$ Fibonacci number,
-for which we can do the same reasoning as above.
+Similarly, we wrote a program to calculate the $n_{th}$ Fibonacci number, for
+which we can do the same reasoning as above:
+
+\begin{lstlisting}
+def main(n):
+    a = 0
+    b = 1
+    while n > 1:
+        sum = a + b
+        a = b
+        b = sum
+        n = n - 1
+    return b
+\end{lstlisting}
 
@@
 
-\anto{Should we say that we get wrong results due to overflow but we don't care?}
+Note that do get meaningful timings, we had to calculate the factorial and
+Fibonacci of very high numbers.  This means that the results are incorrect due
+to overflow, but since all the runnings overflows in the very same way, the
+timings are still comparable. \anto{I think we should rephrase this sentence}.
 

### 2008-12-17 talk/ecoop2009/clibackend.tex
management of external links almost finished

@@
   can grow dynamically. \dacom{I propose primary/secondary instead of
-    the overloaded terms main/child} Each collection must contain at least one
+    the overloaded terms main/child} Each collection contains at least one
   method, called \emph{primary}, which is the first to be created.
@@
 
-\item Each either primary or secondary method corresponds to the
-  translation of some of the blocks of a single flow graph. Each
-  method has an initial block whose input variables are the
+\item Each either primary or secondary method implements a certain
+  number of blocks, all belonging to the same flow graph. Among these blocks
+  there always exists an initial block whose input variables are
   parameters of the method; the input variables of all other blocks
@@
 
-A link is called \emph{internal} if it connects two blocks implemented in the same method,
+A link is called \emph{internal} if it connects two blocks implemented
+by the same method,
  \emph{external} otherwise.
 
-Following an internal link is not difficult in CLI bytecode: a jump to
+Following an internal link is not difficult in IL bytecode: a jump to
 the corresponding code fragment in the same method is emitted 
@@
 
-To solve this problem a special block, called \emph{dispatcher}, is
-added to every method; whenever a method is invoked, its dispatcher is
-executed first\footnote{Recall that the dispatcher is a special block
-and must not be confused with the initial block of a method.} to
+To solve this problem every method contains a special code, called
+\emph{dispatcher}; whenever a method is invoked, its dispatcher is
+executed first\footnote{The dispatcher should not be
+confused with the initial block of a method.} to
 determine which block has to be executed.
@@
 The high word of a block id is the id of the method to which the block
-belongs, whereas the low word is a progressive number identifying
+belongs, whereas the low word is a progressive number univocally identifying
 each block implemented by the method.
@@
 
-\commentout{
-Each method in a graph is assigned an unique 16 bit method id; each block in a method is assigned a progressive 16 bit block number.
-From this two numbers, we can compute the block id as an unsigned integer, by storing the method id in the first 16 bits and the block number in the second 16 bits.
-By construction, the block id is guaranteed to be unique in the graph.
-
-The following picture shows a graph composed of three methods; the id of each method is shown in red, while the block ids are shown in red (for the method id part) and black (for the block number part).
-The graph contains three external links; in particular, note the link between blocks 0x00020001 and 0x00010001 which connects two block that resides in different methods.
-
-Every method contains a special dispatch block, (not shown in the picture above) whose goal is to jump to the specified block number inside the method itself.
-The first argument of a secondary method is always a block id; when the method starts, it immediately jumps to the dispatch block, and thus to the desired block.
-
-For example, suppose to have a method which contains 3 blocks numbered 0, 1, 2; here is how its dispatch blocks looks like; for simplicity it is shown as C# code, but it is actually generated as IL bytecode:
-
+For instance, the code\footnote{For simplicity we write C\# code instead of
+the actual IL bytecode.} generated for the dispatcher of method \texttt{0x0002}
+is similar to the following fragment: 
+\begin{small}
+\begin{lstlisting}[language={[Sharp]C}]
 // dispatch block
-int methodid = (blockid & 0xFFFF0000) >> 16); // take the first 16 bits
-int blocknum = blockid && 0x0000FFFF;         // take the second 16 bits
-
+int methodid = (blockid && 0xFFFF0000) >> 16; 
+int blocknum = blockid && 0x0000FFFF;         
 if (methodid != MY_METHOD_ID) {
-// jump_to_unknown block
+  // jump_to_ext 
   ...
 }
+switch(blocknum) {
+  case 0: goto block0;
+  case 1: goto block1;
+  default: throw new Exception("Invalid block id");
+}
+\end{lstlisting}
+\end{small}
+If the next block to be executed is implemented in the same method
+({\small\lstinline{methodid == MY_METHOD_ID}}), then the appropriate
+jump to the corresponding code is executed. Otherwise, the \lstinline{jump_to_ext}
+part of the dispatcher has to be executed.
+The code that actually jumps to an external block is contained in
+the dispatcher of the primary method, whereas the
+\lstinline{jump_to_ext} code of dispatchers of secondary methods
+simply delegates the dispatcher of the primary method of the same
+graph (see later).
 
-switch(blocknum) {
-case 0:
-goto block0;
-case 1:
-goto block1;
-case 2:
-goto block2;
-default:
-throw new Exception("Invalid block id");
-}
-
-Whenever we want to jump to a external block, it is enough to store the block id in the appropriate variable and jump to the dispatch block.
-If the block resides in a different method, the jump_to_unknown block is entered; this special block is implemented differently by the main method and the secondary methods, as we will see soon.
-
-Each time a new method is added to the graph, we build a delegate for it, and store it in a special array called method_map; since we assign the method id sequentially starting from 0, we are sure that to fetch the method whose id is n we can simply load the n-th element of the array.
-
-The jump_to_unknown block of the main method uses this array to select the right method, and calls it (FlexSwitchCase is the type of delegates for all secondary methods):
-
-// jump_to_unknown block of the main method
+The primary method is responsible for the bookkeeping of the secondary
+methods which are added to the same graph dynamically. This can be 
+simply implemented with an array mapping method id of secondary methods
+to the corresponding delegate. Therefore, the primary methods contain
+the following \lstinline{jump_to_ext} code (where
+\lstinline{FlexSwitchCase} is the type of delegates for secondary methods):
+\begin{small}
+\begin{lstlisting}[language={[Sharp]C}] 
+// jump_to_ext
 FlexSwitchCase meth = method_map[methodid];
@@
 goto dispatch_block;
+\end{lstlisting}
+\end{small}
+Each secondary method returns the block id of the next block to be
+executed; therefore, after the secondary method has returned, the
+dispatcher of the primary method will be executed again to jump
+to the correct next block.
 
-Each secondary method returns a block id specifying the next block to jump to; after its execution, we assign the return value to the blockid variable, and jump again to the dispatch block, which will jump again to the appropriate block.
+To avoid mutual recursion and an undesired growth of the stack,
+the \lstinline{jump_to_ext} code in dispatchers of secondary methods
+just returns the block id of the next block; since the primary method
+is always the first method of the graph which is called, the correct
+jump will be eventually executed by the dispatcher of the primary method.
 
-Keeping this in mind, it is straightforward to implement the jump_to_unknown block of secondary methods: it is enough to return the target block id to the caller, and let its dispatch loop do the right thing.
-If the caller is also a secondary method, it will return it again, until we reach the dispatch loop of the main method, which will finally do the jump.
-In theory, we could implement things differently and jumping directly from a secondary method to another one, but in that case the call stack could grows indefinitely in case of a tight loop between two blocks residing in different methods.
-
+\commentout{
 To implement the dispatch block we can exploit the switch opcode of the CLI; if the .NET JIT is smart enough, it can render it using an indirect jump; overall, jumping to a external block consists of an indirect function call (by invoking the delegate) plus an indirect jump (by executing the switch opcode); even if this is more costly than a simple direct jump, we will see in the next section that this not the main source of overhead when following a external link.
@@
 
-% LocalWords:  flexswitches backend flexswitch
+% LocalWords:  flexswitches backend flexswitch methodid blockid xFFFF blocknum
+% LocalWords:  FFFF goto FlexSwitchCase meth

### 2008-12-18 talk/ecoop2009/benchmarks.tex
finish benchmark section, only the tables are missing

@@
 To measure how the JIT handles object-oriented features, we wrote a very
-simple benchmark that involves attribute lookups and method calls.  We have an
-\emph{accumulator} object, which has a field \lstinline{value} and a method
-\lstinline{add}.  The method \lstinline{add} takes a parameter and adds it the
-field \lstinline{value}.
+simple benchmark that involves attribute lookups and polymorphic method calls:
 
-XXX: update to the new version
+\begin{lstlisting}
+def main(n):
+    if n < 0:
+        n = -n
+        obj = new(value, accumulate=count)
+    else:
+        obj = new(value, accumulate=add)
+    obj.value = 0
+    while n > 0:
+        n = n - 1
+        obj.accumulate(n)
+    return obj.value
 
-Our benchmark accepts a paramter \lstinline{n}, create an \emph{accumulator},
-and repeatedly calls \lstinline{add} on it, passing numbers from
-\lstinline{n-1} to \lstinline{0}.
+def count(x):
+    this.value = this.value + 1
 
-The JIT is able to completely remove the overhead of object creation,
-attribute lookups and method calls, and the generated code results in a simple
-loop doing additions in-place.
+def add(x):
+    this.value = this.value + x
+\end{lstlisting}
 
-Table XXX and figure XXX show the time spent to run the benchmark with various
-input arguments. Again, we can see that the jitted code is up to 500 times
-faster than the interpreted one.
+The two \lstinline{new} operations create an object with exactly one field
+\lstinline{value} and one method \lstinline{accumulate}, whose implementation
+is found in the functions \lstinline{count} and \lstinline{add}, respectively.
+When calling a method, the receiver is implicity passed and can be accessed
+through the special name \lstinline{this}.
+
+The computation \emph{per se} is trivial, as it calculates either $-n$ or
+$1+2...+n-1$, depending on the sign of $n$. The interesting part is the
+polymorphic call to \lstinline{accumulate} inside the loop, because the VM has
+no way to know in advance which method to call (unless it does flow analysis,
+which could be feasible in this case but not in general).  The equivalent C\#
+code we wrote uses two classes and a \lstinline{virtual} method call to
+implement this behaviour.
+
+However, our generated JIT does not compile the whole function at
+once. Instead, it compiles and executes code chunk by chunk, waiting until it
+knows enough informations to generate highly efficient code.  In particualr,
+at the time when it emits the code for the inner loop it exactly knows the
+type of \lstinline{obj}, thus it can remove the overhead of dynamic dispatch
+and inline the method call.  Moreover, since \lstinline{obj} never escapes the
+function, it is \emph{virtualized} and its field \lstinline{value} is stored
+as a local variable.  As a result, the generated code results in a simple loop
+doing additions in-place.
+
+Table XXX show the time spent to run the benchmark with various input
+arguments. Again, we can see that the jitted code is up to 500 times faster
+than the interpreted one.  Moreover, the code generated by the JIT is
+\textbf{faster} than the equivalent C\# code.
+
+Probably, the C\# code is slower because:
+
+\begin{itemize}
+\item The object is still allocated on the heap, and thus there is an extra
+  level of indirection to access the \lstinline{value} field.
+\item The method call is optimized through a \emph{polymorphic inline cache}
+  (XXX: citation needed), that requires a guard check at each iteration.
+\end{itemize}
+
+\anto{maybe we should move the following paragraph to
+  abstract/introduction/conclusion?}
+
+Despite being only a microbenchark, this result is very important as it proves
+that our strategy of intermixing compile time and runtime can yield to better
+performances than current techniques.  The result is even more impressive if
+we consider dynamically typed languages as TLC are usually considered much
+slower than the statically typed ones.

### 2008-12-18 talk/ecoop2009/tlc.tex
finish benchmark section, only the tables are missing

@@
 
-Obviously, not all the operations are applicable to all objects. For example,
+Obviously, not all the operations are applicable to all types. For example,
 it is not possible to \lstinline{ADD} an integer and an object, or reading an
@@
 Since reading TLC programs at bytecode level is hard, in this paper we will
-use an invented Python-like syntax to describe examples, even if we need to
+use an invented Python-like syntax to describe examples, although we need to
 remind that the actual programs are written in the assembler language showed

### 2008-12-18 talk/ecoop2009/clibackend.tex
almost finished. Need to discuss if we want to talk about alternative implementations

@@
 ({\small\lstinline{methodid == MY_METHOD_ID}}), then the appropriate
-jump to the corresponding code is executed. Otherwise, the \lstinline{jump_to_ext}
+jump to the corresponding code is executed, hence internal links
+can be managed efficiently.
+Otherwise, the \lstinline{jump_to_ext}
 part of the dispatcher has to be executed.
@@
 
-\commentout{
-To implement the dispatch block we can exploit the switch opcode of the CLI; if the .NET JIT is smart enough, it can render it using an indirect jump; overall, jumping to a external block consists of an indirect function call (by invoking the delegate) plus an indirect jump (by executing the switch opcode); even if this is more costly than a simple direct jump, we will see in the next section that this not the main source of overhead when following a external link.
+Clearly this complex translation is performed only for flow graphs
+having at least one flexswitch; flow graphs without flexswitches
+are implemented in a more efficient and direct way by a unique method
+with no dispatcher.
 
-Obviously, the slow dispatching logic is needed only when we want to jump to a external block; if the target block happens to reside in the same method as the current one, we can directly jump to it, completely removing the overhead.
+\subsubsection{Passing arguments to external links}
 
-Moreover, the dispatch blocks are emitted only if needed, i.e.
-if the parent graph contains at least one flexswitch; graphs without flexswitches are rendered in the obvious way, by making one method per graph.
+The main drawback of our solution is that passing arguments across
+external links cannot be done efficiently by using the parameters of
+methods for the following reasons:
+\begin{itemize}
+\item In general, the number and type of arguments is different for every block in a graph;
 
-The slow bit: passing arguments
+\item The number of blocks of a graph can grow dynamically, therefore
+  it is not possible to compute in advance the union of the arguments
+  of all blocks in a graph;
 
-Jumping to the correct block is not enough to follow a link: as we said before, each link carries a set of arguments to be passed from the source to the target block.
-As usual, passing arguments across internal links is easy, as we can just use local variables to hold their values; on the other hand, external links make things more complex.
+\item Since external jumps are implemented with a delegate, all the
+  secondary methods of a graph must have the same signature.
+\end{itemize}
 
-The only way to jump to a block is to invoke its containing method, so the first solution that comes to mind is to specify its input arguments as parameter of the method; however, each block has potentially a different number (and different types) of input arguments than every other block, so we need to think of something else.
-
-An alternative solution could be to compute the union of the sets of input arguments of all the blocks in the method, and use this set as a signature for the method; this way, there would be enough space to specify the input arguments for every block we might want to jump to, each block ignoring the exceeding unused parameters.
-
-Unfortunately, all the secondary methods must have the very same signature, as they are all called from the same calling site in the dispatch block of the main method.
-Since the union of the set of input arguments (and hence the computed signature) varies from method to method, this solution cannot work.
-
-We might think to determine the signature by computing the union of input arguments of all blocks in the graph; this way, all the secondary methods would have the same signature.
-But as we said above, the graph grows new blocks at runtime, so we cannot determine in advance which set of input arguments we will need.
-
-To solve the problem we need a way to pass a variable number of arguments without knowing in advance neither their number nor their types. Thus, we use an instance of this class:
-
+Therefore, the only solution we came up with is defining a class
+\lstinline{InputArgs} for passing sequences of arguments whose length
+and type is variable.
+\begin{small}
+\begin{lstlisting}[language={[Sharp]C}] 
 public class InputArgs {
@@
 }
+\end{lstlisting}
+\end{small}
+Unfortunately, with this solution passing arguments to external links
+becomes quite slow:
+\begin{itemize}
+\item When writing arguments, array re-allocation may be needed in
+  case the number of arguments exceeds the dimension of the
+  array. Furthermore the VM will always perform bound-checks, even
+  when the size is explicitly checked in advance;
 
-Since the fields are arrays, they can grow as needed to contain any number of arguments; arguments whose type is primitive are stored in the ints or floats array, depending on their type; arguments whose type is a reference type are stored in the objs array: it's up to each block to cast each argument back to the needed type.
+\item When reading arguments, a bound-check is performed by the VM for
+  accessing each argument; furthermore, an appropriate downcast must be
+  inserted anytime an argument of type object is read.
+\end{itemize}
+Of course, we do not need to create a new object of class
+\lstinline{InputArgs} any time we need to perform an external jump;
+instead, a unique object is created at the beginning of the execution
+of the primary method.
 
-This solution impose a huge overhead on both writing and reading arguments:
-
-        * when writing, we need to make sure that the arrays are big enough to contains all the arguments we need; if not, we need to allocate a bigger array.
-Moreover, for each argument we store into the array the virtual machine performs a bound-check, even if we know the index will never be out of bounds (because we checked the size of the array in advance);
-        * when reading, the same bound-check is performed for each argument read; moreover, for each value read from the objs array we need to insert a downcast.
-
-To mitigate the performance drop, we avoid to allocate a new InputArgs object each time we do a external jump; instead, we preallocate one at the beginning of the main method, and reuse it all the time.
-
-Our benchmarks show that passing arguments in arrays is about 10 times slower than passing them as real parameter of a method. Unfortunately, we couldn't come up with anything better.
-Implement flexswitches
-
-Now, we can exploit all this machinery to implement flexswitches, as this is our ultimate goal.
-As described above, the point is to be able to add new cases at runtime, each case represented as a delegate.
-Here is an excerpt of the C# class that implements a flexswitch that switches over an integer value:
-
-public class IntLowLevelFlexSwitch:
-{
+\subsubsection{Implementation of flexswitches}
+Finally, we can have a look at the implementation of flexswitches.
+The following snippet shows the special case of integer flexswitches.
+\begin{small}
+\begin{lstlisting}[language={[Sharp]C}] 
+public class IntLowLevelFlexSwitch:BaseLowLevelFlexSwitch {
   public uint default_blockid = 0xFFFFFFFF;
@@
 }
+\end{lstlisting}
+\end{small}
+The mapping from integers values to delegates (pointing to secondary
+methods) is just implemented by the two arrays \lstinline{values} and
+\lstinline{cases}. Method \lstinline{add_case} extends the mapping
+whenever a new case is added to the flexswitch.
 
-For each case, we store both the triggering value and the corresponding delegate; the add_case method takes care to append value and c to the values and cases arrays, respectively (and resize them if necessary).
-The interesting bit is the execute method: it takes a value and a set of input arguments to be passed across the link and jumps to the right block by performing a linear search in the values array.
+The most interesting part is the body of method \lstinline{execute},
+which takes a value and a set of input arguments to be passed across
+the link and jumps to the right block by performing a linear search in
+array \lstinline{values}.
 
-As shown by previous sections, the first argument of a FlexSwitchCase is the block id to jump to; since when we go through a flexswitch we always want to jump to the first block of the method, we pass the special value 0 as a block id, which precisely means jump to the first block.
-This little optimization let us not to have to explicitly store the block id for the first block of all the cases.
+Recall that the first argument of delegate \lstinline{FlexSwitchCase}
+is the block id to jump to; since the target of an external jump is
+always the initial block of the method, the first argument will be
+always 0.
 
-The value returned by execute is the next block id to jump to; if the value is not found in the values array, we return the default_blockid, whose value has been set before by the JIT compiler; default_blockid usually points to a block containing code to restart the JIT compiler again; when the JIT compiler restarts, it emits more code for the missing case, then calls add_case on the flexswitch; from now on, the new blocks are wired into the existing graph, and we finally managed to implement growable graphs.
-Performances
+The value returned by method \lstinline{execute} is the next block id
+to be executed; 
+in case no association is found for \lstinline{value},
+\lstinline{default_blockid} is returned. The value of
+\lstinline{default_blockid} is initially set by the JIT compiler and
+usually corresponds to a block containing code to restart the JIT
+compiler for creating a new secondary method with the new code for the
+missing case, and updating the flexswitch by calling method
+\lstinline{add_case}.
 
-As we saw, implementing growable graphs for CLI is a pain, as the virtual machine offers very little support, so we need an incredible amount of workarounds.
-Moreover, the code generated is much worse than what an assembly backend could produce, and the cost of following a external link is very high compared to internal links.
-
-However, our first blog post showed that we still get very good performances; how is it possible?
-
-As usual in computer science, most of the time of a running program in spent in a tiny fraction of the code; our benchmark is no exception, and the vast majority of the time is spent in the inner loop that multiplies numbers; the graph is built in such a way that all the blocks that are part of the inner loop reside in the same method, so that all links inside are internal (and fast).
-
-Flexswitches and external links play a key role to select the right specialized implementation of the inner loop, but once it is selected they are not executed anymore until we have finished the computation.
-
-It is still unclear how things will look like when we will compile the full Python language instead of a toy one; depending on the code, it could be possible to have external links inside the inner loop, thus making performance much worse.
-Alternative implementations
-
+\subsection{Alternative implementations}
+\dacom{need to be discussed with Antonio}
+\commentout{
 Before implementing the solution described here, we carefully studied a lot of possible alternatives, but all of them either didn't work because of a limitation of the virtual machine or they could work but with terrible performances.
@@
 In particular, in theory it is possible to implement external links using tail calls, by putting each block in its own method and doing a tail call instead of a jump; this would also solve the problem of how to pass arguments, as each method could have its own signature matching the input args of the block.
-I would like to explain this solution in a more detailed way as I think it's really elegant and nice, but since this post is already too long, I'll stop here :-).
+I would like to explain this solution in a more detailed way as I think it's really elegant and nice.
 
@@
 However, one benchmark I wrote shows that tail calls are up to 10 times slower (!!!) than normal calls, thus making impractical to use them for our purposes.
-Conclusion
-
-Despite the complexity of the implementation, our result are extremely good; the speedup we got is impressive, and it proves that PyPy's approach to JIT compiler can work well also on top of object oriented virtual machines like .NET or the JVM.
-
-Generating bytecode for those machine at runtime is not a new idea; Jython, IronPython, JRuby and other languages have been doing this for years.
-However, Jython and IronPython do only a simple "static" translation, which doesn't take advantage of the informations gathered at runtime to generate better, faster and specialized code.
-Recently, JRuby grew a new strategy to JIT-compile only hotspots, taking advantage of some informations gathered while interpreting the code; this is still a "one-shot" compilation, where the compiled code does not change over time.
-
-To my knowledge, PyPy brings the first example of a language which implements a truly JIT compiler on top of the underlying JIT compiler of the virtual machine, emitting bytecode that changes and adapts over the time.
-If someone knows other languages doing that, I would really like to know more.
-
-Being so innovative, the problem of this approach is that the current virtual machines are not designed to support it in a native way, and this forces us to put a lot of workarounds that slow down the generated code.
-The hope is that in the future the virtual machines will grow features that help us to generate such kind of code.
-The experimental Da Vinci VM seems to go in the right direction, so it is possible that in the future I will try to write a JIT backend for it.
-
-At the moment, the CLI JIT backend is almost complete, and all the hardest problems seems to be solved; the next step is to fix all the remaining bugs and implement some minor feature that it's still missing, then try to apply it to the full Python language and see what is the outcome.
 }
@@
 % LocalWords:  flexswitches backend flexswitch methodid blockid xFFFF blocknum
-% LocalWords:  FFFF goto FlexSwitchCase meth
+% LocalWords:  FFFF goto FlexSwitchCase meth InputArgs ints objs VM uint args
+% LocalWords:  IntLowLevelFlexSwitch BaseLowLevelFlexSwitch xFFFFFFFF numcases
+% LocalWords:  JIT

### 2008-12-18 talk/ecoop2009/benchmarks.tex
add tables showing the results. They need some restyling as they are very ugly.

@@
 
-Table XXX and figure XXX show the time spent to calculate the factorial of
-various numbers, with and without the JIT.  Table XXX and figure XXX show the
-same informations for the Fibonacci program.
+\anto{these tables are ugly}
 
-Note that do get meaningful timings, we had to calculate the factorial and
-Fibonacci of very high numbers.  This means that the results are incorrect due
-to overflow, but since all the runnings overflows in the very same way, the
-timings are still comparable. \anto{I think we should rephrase this sentence}.
+\begin{table}[ht]
+  \begin{tabular}{|l|r|r|r|r||r|r|}
+    \hline
+    \textbf{n} & 
+    \textbf{Interp} &
+    \textbf{JIT} &
+    \textbf{JIT 2} &
+    \textbf{C\#} &
+    \textbf{Interp/JIT 2} &
+    \textbf{JIT 2/C\#} \\
+    \hline
 
-As we can see, the code generated by the JIT is almost 500 times faster than
-the non-jitted case, and it is only about 1.5 times slower than the same
-algorithm written in C\#: the difference in speed it is probably due to both
-the fact that the current CLI backend emits slightly non-optimal code and that
-the underyling .NET JIT compiler is highly optimized to handle bytecode
-generated by C\# compilers.
+    $10$    &   0.031  &  0.422  &  0.000  &  0.000  &      N/A  &    N/A \\
+    $10^7$  &  30.984  &  0.453  &  0.047  &  0.031  &  661.000  &  1.500 \\
+    $10^8$  &     N/A  &  0.859  &  0.453  &  0.359  &      N/A  &  1.261 \\
+    $10^9$  &     N/A  &  4.844  &  4.641  &  3.438  &      N/A  &  1.350 \\
+
+    \hline
+
+  \end{tabular}
+  \caption{Factorial benchmark}
+  \label{tab:factorial}
+\end{table}
+
+\begin{table}[ht]
+  \begin{tabular}{|l|r|r|r|r||r|r|}
+    \hline
+    \textbf{n} & 
+    \textbf{Interp} &
+    \textbf{JIT} &
+    \textbf{JIT 2} &
+    \textbf{C\#} &
+    \textbf{Interp/JIT 2} &
+    \textbf{JIT 2/C\#} \\
+    \hline
+
+    $10$    &   0.031  &  0.453  &  0.000  &  0.000  &       N/A  &  N/A   \\
+    $10^7$  &  29.359  &  0.469  &  0.016  &  0.016  &  1879.962  &  0.999 \\
+    $10^8$  &     N/A  &  0.688  &  0.250  &  0.234  &       N/A  &  1.067 \\
+    $10^9$  &     N/A  &  2.953  &  2.500  &  2.453  &       N/A  &  1.019 \\
+
+    \hline
+
+  \end{tabular}
+  \caption{Fibonacci benchmark}
+  \label{tab:fibo}
+\end{table}
+
+Tables \ref{tab:factorial} and \ref{tab:fibo} show the time spent to calculate
+the factorial and Fibonacci for various $n$.  As we can see, for small values
+of $n$ the time spent running the JIT compiler is much higher than the time
+spent to simply interpret the program.  This is an expected result, as till
+now we only focused on optimizing the compiled code, not the compilation
+process itself.
+
+On the other hand, to get meaningful timings we had to use very high values of
+$n$.  This means that the results are incorrect due to overflow, but since all
+the runnings overflow in the very same way, the timings are still
+comparable. \anto{I think we should rephrase this sentence}.  For $n$ greater
+than $10^7$, we did not run the interpreted program as it would have took too
+much time, without adding anything to the discussion.
+
+As we can see, the code generated by the JIT can be up to ~1800 times faster
+than the non-jitted case.  Moreover, it often runs at the same speed as the
+equivalent program written in C\#, being only 1.5 slower in the worst case.
+
+The difference in speed it is probably due to both the fact that the current
+CLI backend emits slightly non-optimal code and that the underyling .NET JIT
+compiler is highly optimized to handle bytecode generated by C\# compilers.
 
@@
 
-Table XXX show the time spent to run the benchmark with various input
-arguments. Again, we can see that the jitted code is up to 500 times faster
-than the interpreted one.  Moreover, the code generated by the JIT is
-\textbf{faster} than the equivalent C\# code.
+\begin{table}[ht]
+  \begin{tabular}{|l|r|r|r|r||r|r|}
+    \hline
+    \textbf{n} & 
+    \textbf{Interp} &
+    \textbf{JIT} &
+    \textbf{JIT 2} &
+    \textbf{C\#} &
+    \textbf{Interp/JIT 2} &
+    \textbf{JIT 2/C\#} \\
+    \hline
+
+    $10$    &   0.031  &  0.453  &  0.000  &  0.000  &      N/A  &  N/A   \\
+    $10^7$  &  43.063  &  0.516  &  0.047  &  0.063  &  918.765  &  0.750 \\
+    $10^8$  &     N/A  &  0.875  &  0.453  &  0.563  &      N/A  &  0.806 \\
+    $10^9$  &     N/A  &  4.188  &  3.672  &  5.953  &      N/A  &  0.617 \\
+
+    \hline
+
+  \end{tabular}
+  \caption{Accumulator benchmark}
+  \label{tab:accumulator}
+\end{table}
+
+Table \ref{tab:accumulator} show the results for the benchmark.  Again, we can
+see that the speedup of the JIT over the interpreter is comparable to the
+other two benchmarks.  However, the really interesting part is the comparison
+with the equivalent C\# code, as the code generated by the JIT is
+\textbf{faster}.
 

### 2008-12-18 talk/ecoop2009/intro.tex
some notes about what the high-level overview should contain

@@
 
-\cfbolz{we should cite the dyla paper somewhere here}
+\cfbolz{we should cite the dyla paper somewhere here. also, we generally need
+more references}
 
-PyPy's idea is to automatize the generation of JIT compilers in order
+PyPy's approach is to automatize the generation of JIT compilers in order
 to reduce to a minimum the effort required to get a fast implementation of a
@@
 the help of partial evaluation techniques.
+
+The contributions of this paper are \emph{promotion}, a generalization of
+polymorphic inline caches that make partial evaluation practical for dynamic
+languages and trying out the idea of JIT-layering XXX.
 
@@
 
-The \emph{PyPy} project\footnote{\texttt{http://codespeak.net/pypy/dist/pypy/doc/home.html}}
+The \emph{PyPy} project\footnote{\texttt{http://codespeak.net/pypy/}}
 \cite{RigoPedroni06} was initially conceived to develop an implementation of Python which
@@
 
-Another interesting feature of PyPy
-is that just-in-time compilers can be semi-automatically generated from the
-interpreter source.
+- write an interpreter
+- add some hints to the interpreter
 
-XXX list contributions clearly
+translation time:
+- start the translation process, flow graphs
+- run a binding-time analysis using the hints, to figure out ...
+- turn the flow graphs into rainbow bytecode
+- turn the flow graphs of the rainbow interpreter and the original flow graphs
+into .NET code
+
+runtime:
+when the interpreter is started and a function should be interpreted:
+ - the generated JIT starts compiling the function 
+ - it produces .NET bytecode until it does not have enough information to
+ continue. then it stops the compilation and puts a callback into the compiler
+ into the generated bytecode.
+ - at this point the .NET bytecode that is produced so far is executed.
+ - when the callback into the compiler is hit, execution stops and compilation
+ starts again
+ - afterwards execution continues after the callback (running the code that was
+ newly produced)

### 2008-12-18 talk/ecoop2009/tlc.tex
fix some things in the tlc section

@@
 In this section, we will briefly describe \emph{TLC}, a simple dynamic
-language that we developed to exercise our JIT compiler generator.  As most of
-dynamic languages around, \emph{TLC} is implemented through a virtual machine
-that interprets a custom bytecode. Since our main interest is in the runtime
-performance of the VM, we did not implement the parser nor the bytecode
-compiler, but only the VM itself.
+language that we developed to exercise our JIT compiler generator and that will
+be used as a running example in this paper. The design goal of the language is
+to be very simple (the interpreter of the full language consists of about 600
+lines of RPython code) but to still have the typical properties of dynamic
+languages that make them hard to compile. \emph{TLC} is implemented with a small
+interpreter that interprets a custom bytecode instruction set. Since our main
+interest is in the runtime performance of the interpreter, we did not implement
+the parser nor the bytecode compiler, but only the interpreter itself.
 
@@
 
-The virtual machine is stack-based, and provides several operations:
+The interpreter for the language is stack-based and uses bytecode to represent
+the program. It provides the following bytecode instructions:
 
@@
 \item \textbf{Stack manipulation}: standard operations to manipulate the
-  stack, such as \lstinline{PUSH}, \lstinline{POP}, \lstinline{SWAP}, etc.
+  stack, such as \lstinline{POP}, \lstinline{PUSHARG}, \lstinline{SWAP}, etc.
 \item \textbf{Flow control} to do conditional and unconditional jumps.
+\cfbolz{what is the bytecode for conditional jumps?}
 \item \textbf{Arithmetic}: numerical operations on integers, like
@@
 \label{sec:tlc-features}
+\cfbolz{calling this sections "features" is a bit obscure, since it is more
+properties of the implementation}
 
@@
 candidate as a language to test our JIT generator, as it has some of the
-features that makes most of current dynamic languages so slow:
+properties that makes most of current dynamic languages so slow:
 
@@
 
-\item \textbf{Stack based VM}: this kind of VM requires all the operands to be
+\item \textbf{Stack based interpreter}: this kind of interpreter requires all the operands to be
   on top of the evaluation stack.  As a consequence programs spend a lot of
@@
   related operations.  However, thanks to its simplicity this is still the
-  most common and preferred way to implement VMs.
+  most common and preferred way to implement interpreters.
 
@@
     PUSH 0        #        [n, 0]
-    LT            #        [n<0]
+    LT            #        [0 or 1]
     BR_COND neg

### 2008-12-18 talk/ecoop2009/rainbow.tex
lots of fixes

@@
 techniques introduced in our approach, which we call \emph{promotion}.  In
-short, it allows an arbitrary run-time value to be turned into a
-compile-time value at any point in time.  Promotion is thus the central way by
+short, it allows an arbitrary run-time (i.e. red) value to be turned into a
+compile-time (i.e. green) value at any point in time.  Promotion is thus the central way by
 which we make use of the fact that the JIT is running interleaved with actual
@@
 evaluation, in which the compiler always runs fully ahead of execution
-This is a problem in many large use cases.  For example, in an
+This is a problem in many realistic use cases.  For example, in an
 interpreter for a dynamic language, there is mostly no information
@@
 
-A very different point of view on promotion is as a generalization of
-techniques that already exist in dynamic compilers as found in modern
-object-oriented language virtual machines.  In this context feedback
+A very different point of view on promotion is as a generalization of techniques
+that already exist in dynamic compilers as found in modern virtual machines for
+object-oriented language.  In this context feedback
 techniques are crucial for good results.  The main goal is to
@@
 object being dispatched to is known at compile-time, the lookup can be
-folded, and only a (possibly inlined) direct call remains in the
+folded, and only a (possibly even inlined) direct call remains in the
 generated code.  In the case where the type of the object is not known
@@
 promoted to compile-time.  As we will see in the sequel, this produces
-very similar machine code \footnote{This can also be seen as a generalization of
+machine code very similar to that of polymorphic inline
+caches\footnote{Promotion can also be seen as a generalization of
 a partial evaluation transformation called "The Trick" (see e.g. \cite{XXX}),
@@
 
-The essential advantage is that it is no longer tied to the details of
+The essential advantage of promotion is that it is no longer tied to the details of
 the dispatch semantics of the language being interpreted, but applies in
@@
 
-\subsection{Promotion as Applied to the TLC}
+\subsection{Implementing Promotion}
 
-XXX
+The implementation of promotion requires a tight coupling between
+compile-time and run-time: a \emph{callback}, put in the generated code,
+which can invoke the compiler again.  When the callback is actually
+reached at run-time, and only then, the compiler resumes and uses the
+knowledge of the actual run-time value to generate more code.
 
-\subsection{Promotion in Practise}
+The new generated code is potentially different for each run-time value
+seen.  This implies that the generated code needs to contain some sort
+of updatable switch, which can pick the right code path based on the
+run-time value.
 
-There are values that, if known at compile time, allow the JIT compiler to
-produce very efficient code.  Unfortunately, these values are tipically red,
-e.g. the exact type of a variable.
-
-"Promotion" is a particular operation that convert a red value into a green
-value; i.e., after the promotion of a variable, the JIT compiler knows its
-value at compile time. Since the value of a red variable is not known until
-runtime, we need to postpone the compilation phase after the runtime phase.
-
-This is done by continuously intermixing compile time and runtime; a promotion
-is implemented in this way:
+\cfbolz{I think this example is confusing, it is unclear in which order things
+happen how. I will try to come up with something different}.
 
@@
 
+\subsection{Promotion as Applied to the TLC}
+
+XXX maybe a tlc example can be used for the example above?
+
 \section{Automatic Unboxing of Intermediate Results}
@@
 
-Interpreters for dynamic languages typically allocate a lot of small
-objects, for example due to boxing.  For this reason, we
-implemented a way for the compiler to generate residual memory
-allocations as lazily as possible.  The idea is to try to keep new
-run-time structures "exploded": instead of a single run-time pointer to
-a heap-allocated data structure, the structure is "virtualized" as a set
-of fresh variables, one per field.  In the compiler, the variable that
-would normally contain the pointer to the structure gets instead a
-content that is neither a run-time value nor a compile-time constant,
-but a special \emph{virtual structure} – a compile-time data structure that
-recursively contains new variables, each of which can again store a
-run-time, a compile-time, or a virtual structure value.
-
-This approach is based on the fact that the "run-time values" carried
-around by the compiler really represent run-time locations – the name of
-a CPU register or a position in the machine stack frame.  This is the
-case for both regular variables and the fields of virtual structures.
-It means that the compilation of a \texttt{getfield} or \texttt{setfield}
-operation performed on a virtual structure simply loads or stores such a
-location reference into the virtual structure; the actual value is not
-copied around at run-time.
+Interpreters for dynamic languages typically continuously allocate a lot of small
+objects, for example due to boxing. This makes arithmetic operations extremely
+inefficient. For this reason, we
+implemented a way for the compiler to try to avoid memory allocations in the
+residual code as long as possible. The idea is to try to keep new
+run-time structures "exploded": instead of a single run-time object allocated on
+the heap, the object is "virtualized" as a set
+of fresh variables, one per field. Only when the object can be accessed by from
+somewhere else is it actually allocated on the heap. The effect of this is similar to that of
+escape analysis \cite{XXX}, which also prevents allocations of objects that can
+be proven to not escape a method or set of methods.
 
@@
 short-lived objects, even in non-trivial situations.  The following
-example shows a typical case.  Consider the Python expression \texttt{a+b+c}.
-Assume that \texttt{a} contains an integer.  The PyPy Python interpreter
+example shows a typical case.  XXX use TLC example
+
+The PyPy Python interpreter
 implements application-level integers as boxes – instances of a
@@
 
+XXX kill the rest?!‽
 When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are

### 2008-12-18 talk/ecoop2009/jitgen.tex
add a tlc-based example for partial evaluation.

@@
 
-\cfbolz{I would propose to use either TLC as an example here, or something that
-looks at least like an interpreter loop}
+\begin{figure}[h]
+\label{fig:tlc-main}
+\begin{center}
+\input{tlc-simplified.py}
+\caption{The main loop of the TLC interpreter}
+\end{center}
+\end{figure}
 
-Example::
-\begin{lstlisting}[language=Python]
-  def f(x, y):    
-    x2 = x * x    
-    y2 = y * y    
-    return x2 + y2
+Let us look at an example. Figure \ref{fig:tlc-main} shows parts of the
+main-loop of the TLC interpreter (in slightly simplified form). The static
+arguments of this functions would typically be the \lstinline{code} argument
+(containing the bytecode as a string), the \lstinline{pc} argument (containing
+the initial program counter) and the \lstinline{pool} argument (containing the
+constant pool). The \lstinline{args} argument is dynamic since it contains the
+user-input of the interpreted function. Since the \lstinline{while} loop and the
+contained conditions depend only on static arguments it will typically be
+unrolled by a partial evaluator.
 
-**case x=3** ::
+When the function is partially evaluated with respect to the TLC example
+function shown in Figure XXX (which computes the absolute value of a number),
+the residual code would look like in Figure \ref{fig:tlc-folded}. This version
+is already a great improvement over pure interpretation, all the bytecode
+dispatch overhead has been removed. However, the function as shown is still
+rather inefficient, due to lots of superfluous stack handling and also due to
+some indirect calls for implementing comparison (\lstinline{lt}) and
+subtraction (\lstinline{sub}).
 
-  def f_3(y):    
-    y2 = y * y   
-    return 9 + y2
+\begin{figure}[h]
+\label{fig:tlc-folded}
+\begin{center}
+\input{tlc-folded.py}
+\caption{The residual code of the TLC main loop with respect to the
+\lstinline{abs} function }
+\end{center}
+\end{figure}
 
-**case x=10** ::
+This shows a common shortcoming of partial evaluation when applied to a dynamic
+language: The partial evaluator (just like an ahead-of-time compiler) cannot
+make assumptions about the types of objects, which leads to poor results.
+Effective dynamic compilation requires feedback of runtime information into
+compile-time. This is no different for partial evaluation.
 
-  def f_10(y):    
-    y2 = y * y   
-    return 100 + y2
-\end{lstlisting}
-
-A shortcoming of PE is that in many cases not much can be really assumed
-constant at compile-time, and this leads to poor results.  Effective dynamic
-compilation requires feedback of runtime information into compile-time; for a
-dynamic language, types are a primary example.
-
-Partial evaluation (PE) comes in two flavors:
+Partial evaluation (PE) comes in two flavors: \cfbolz{if we have space problems
+we should kill the bits about online and offline PE}
 

### 2008-12-18 talk/ecoop2009/benchmarks.tex
kill a footnote, add a reference

@@
 \item The method call is optimized through a \emph{polymorphic inline cache}
-  (XXX: citation needed), that requires a guard check at each iteration.
+  \cite{hoelzle_optimizing_1991}, that requires a guard check at each iteration.
 \end{itemize}
@@
 
-Despite being only a microbenchark, this result is very important as it proves
+Despite being only a microbenchmarks, this result is very important as it proves
 that our strategy of intermixing compile time and runtime can yield to better

### 2008-12-18 talk/ecoop2009/rainbow.tex
kill a footnote, add a reference

@@
 invocation.  This is achieved with variations on the technique of
-polymorphic inline caches \cite{XXX}: the dynamic lookups are cached and
+polymorphic inline caches \cite{hoelzle_optimizing_1991}: the dynamic lookups are cached and
 the corresponding generated machine code contains chains of
@@
 machine code very similar to that of polymorphic inline
-caches\footnote{Promotion can also be seen as a generalization of
-a partial evaluation transformation called "The Trick" (see e.g. \cite{XXX}),
-which again produces similar code but which is only applicable for finite sets
-of values.}.
+caches.
 

### 2008-12-18 talk/ecoop2009/benchmarks.tex
answer to a comment, insert another one, fix a typo

@@
 
-Despite being only a microbenchmarks, this result is very important as it proves
+Despite being only a microbenchmark, this result is very important as it proves
 that our strategy of intermixing compile time and runtime can yield to better

### 2008-12-18 talk/ecoop2009/jitgen.tex
kill a footnote, add a reference
answer to a comment, insert another one, fix a typo

@@
 arguments of this functions would typically be the \lstinline{code} argument
-(containing the bytecode as a string), the \lstinline{pc} argument (containing
+(containing the bytecode as a string \anto{what about using array of bytes instead of string?
+For .NET/JVM people, there is a strong difference between the twos}), the \lstinline{pc} argument (containing
 the initial program counter) and the \lstinline{pool} argument (containing the
@@
 
-\subsubsection{Hints}
+\subsection{Hints}
 

### 2008-12-19 talk/ecoop2009/clibackend.tex
add a comment
minor fixes, some comments

@@
 any time a new case has to be added to a flexswitch.
-\dacom{comment for Antonio: I am not sure this is the best solution. This cannot work for Java where classes are the basic
-  units. Closures will be available only with Java Dolphin and I do
-  not know how much efficient will be}
 In this way, whereas flow graphs without flexswitches are translated
@@
 \item Each flow graph is translated in a collection of methods which
-  can grow dynamically. \dacom{I propose primary/secondary instead of
-    the overloaded terms main/child} Each collection contains at least one
+  can grow dynamically. Each collection contains at least one
   method, called \emph{primary}, which is the first to be created.
@@
   parameters of the method; the input variables of all other blocks
-  are local variables of the method.
+  are local variables of the method. \anto{This is wrong: the signature of the secondary methods is fixed, and input args are passed inside the InputArgs class, not as methodo parameters}
 \end{itemize}
@@
 and translated by the backend in a new single method pointed
-by a delegate which is stored in the code implementing the flexswitch,
+by a delegate \footnote{\emph{Delegates} are the .NET equivalent of function pointers}
+ of  which is stored in the code implementing the flexswitch,
 so that the method can be invoked later.
@@
 to execute the new block, whereas the appropriate local variables are
-used for passing arguments.
+used for passing arguments. \anto{this is wrong for the same reason as above}
 Also following an external link whose target is the initial block of a
@@
 \emph{block id}, which uniquely identifies the next block of the graph to be executed.
-The high word of a block id is the id of the method to which the block
+The high word \anto{a word is 32 bit, block num and method id are 16 bit each} of a block id is the id of the method to which the block
 belongs, whereas the low word is a progressive number univocally identifying
@@
 jump to the corresponding code is executed, hence internal links
-can be managed efficiently.
+can be managed efficiently. \anto{wrong: internal links don't go through the dispatcher}
 Otherwise, the \lstinline{jump_to_ext}
@@
 
-Recall that the first argument of delegate \lstinline{FlexSwitchCase}
-is the block id to jump to; since the target of an external jump is
-always the initial block of the method, the first argument will be
-always 0.
+Recall that the first argument of delegate \lstinline{FlexSwitchCase} is the
+block id to jump to. By construction, the target block of a flexswitch is
+always the first in a secondary method, and we use the special value
+\lstinline{0} to signal this.
 
@@
 \dacom{need to be discussed with Antonio}
+
+\anto{I propose to wait until the rest of paper is more or less in a final
+  state, and see how much space is left}
+
 \commentout{

### 2008-12-19 talk/ecoop2009/jitgen.tex
minor fixes, some comments

@@
 \input{tlc-simplified.py}
-\caption{The main loop of the TLC interpreter}
+\caption{The main loop of the TLC interpreter, written in RPython}
 \end{center}
@@
 same type inference engine that is used on the source RPython program,
-the annotator.  In this mode, it is called the \emph{hint-annotator}; it
+the annotator \anto{I'm not sure we should mention the annotator, as it is not referred anywhere else}.  In this mode, it is called the \emph{hint-annotator}; it
 operates over input graphs that are already low-level instead of

### 2008-12-19 talk/ecoop2009/rainbow.tex
minor fixes, some comments

@@
 \section{Automatic Unboxing of Intermediate Results}
+\label{sec:virtuals}
 
@@
 the heap, the object is "virtualized" as a set
-of fresh variables, one per field. Only when the object can be accessed by from
+of fresh local variables, one per field. Only when the object can be accessed by from
 somewhere else is it actually allocated on the heap. The effect of this is similar to that of

### 2008-12-19 talk/ecoop2009/benchmarks.tex
minor fixes, some comments

@@
 
-In section \ref{sec:tlc-features}, we saw that TLC provides most of the
+\anto{We should say somewhere that flexswitches are slow but benchmarks are so
+  good because they are not involved in the inner loops}
+
+In section \ref{sec:tlc-properties}, we saw that TLC provides most of the
 features that usaully make dynamically typed language so slow, such as
-\emph{stack-based VM}, \emph{boxed arithmetic} and \emph{dynamic lookup} of
+\emph{stack-based interpreter}, \emph{boxed arithmetic} and \emph{dynamic lookup} of
 methods and attributes.
@@
 \item By plain interpretation, without any jitting.
-\item With the jit enabled: this run includes the time spent by doing the
+\item With the JIT enabled: this run includes the time spent by doing the
   compilation itself, plus the time spent by running the produced code.
-\item Again with the jit enabled, but this time the compilation has already
+\item Again with the JIT enabled, but this time the compilation has already
   been done, so we are actually measuring how good is the code we produced.
@@
 integers, so it can inline the code to compute the result.  Moreover, it can
-\emph{virtualize} all the temporary objects, because they never escape from
+\emph{virtualize} (see section \ref{sec:virtuals} all the temporary objects, because they never escape from
 the inner loop.  The same remarks apply to the other two operations inside
@@
 $1+2...+n-1$, depending on the sign of $n$. The interesting part is the
-polymorphic call to \lstinline{accumulate} inside the loop, because the VM has
+polymorphic call to \lstinline{accumulate} inside the loop, because the interpreter has
 no way to know in advance which method to call (unless it does flow analysis,
@@
 knows enough informations to generate highly efficient code.  In particualr,
-at the time when it emits the code for the inner loop it exactly knows the
+at the time it emits the code for the inner loop it exactly knows the
 type of \lstinline{obj}, thus it can remove the overhead of dynamic dispatch

### 2008-12-19 talk/ecoop2009/tlc.tex
answer to a comment, insert another one, fix a typo
minor fixes, some comments

@@
 of attributes of the object, as well as its methods.  Once the object has been
-created, it is not possible to add/remove attributes and methods.
+created, it is possible to call methods and read or write attributes, but not
+to add or remove them.
 
@@
   stack, such as \lstinline{POP}, \lstinline{PUSHARG}, \lstinline{SWAP}, etc.
-\item \textbf{Flow control} to do conditional and unconditional jumps.
-\cfbolz{what is the bytecode for conditional jumps?}
+\item \textbf{Flow control} to do conditional (\lstinline{BR_COND}) and
+  unconditional (\lstinline{BR}) jumps.
 \item \textbf{Arithmetic}: numerical operations on integers, like
@@
 
-\subsection{TLC features}
-\label{sec:tlc-features}
-\cfbolz{calling this sections "features" is a bit obscure, since it is more
-properties of the implementation}
+\subsection{TLC properties}
+\label{sec:tlc-properties}
 

### 2008-12-19 talk/ecoop2009/jitstrategy.tex
start of the "how the generated jit works" section

\section{Effective JIT compilation of dynamic languages}

Writing efficient compilers for dynamic languages is hard.  Since these
languages are dynamically typed, usually the compiler does not have enough
information to produce efficient code, but instead it has to insert a lot of
runtime checks to select the appropriate implementation for each operation.

By generating code at runtime, JIT compilers can exploit some extra knowledge
compared to traditional static compilers.  However, we need to take special
care to choose a strategy for JIT compilation that lets the compiler to take
the best of this advantage.

Most JIT compilers for dynamic languages around (such as IronPython, Jython,
JRuby \anto{XXX insert some reference}) compile code at the method
granularity.  If on one hand they can exploit some of the knowledge gathered
at runtime (e.g. the types of method parameters), on the other hand they can
do little to optimize most of the operations inside, because their behaviour
depends on informations that are not available at compile time, because
e.g. the global state of the program can change at runtime. \anto{e.g., we can
  add/remove methods to classes, etc. Should we insert some example here?}

JIT compilers generated by PyPy solve this problem by delaying the compilation
until they know all the informations needed to generate efficient code.  If at
some point the JIT compiler does not know about something it needs, it
generates a callback into itself.

Later, when the generated code is executed, the callback is hit and the JIT
compiler is restarted again.  At this point, the JIT knows exactly the state
of the program and can exploit all this extra knowledge to generate highly
efficient code.  Finally, the old code is patched and linked to the newly
generated code, so that the next time the JIT compiler will not be invoked
again.  As a result, \textbf{runtime and JIT-compile time are continuously
  intermixed}. \anto{maybe we should put the translation/compile/run time
  distinction here}

XXX: insert an example of flexswitch

### 2008-12-19 talk/ecoop2009/clibackend.tex
fig.6 on method and block ids should be changed
changes in reaction of Antonio's comments

@@
   number of blocks, all belonging to the same flow graph. Among these blocks
-  there always exists an initial block whose input variables are
-  parameters of the method; the input variables of all other blocks
-  are local variables of the method. \anto{This is wrong: the signature of the secondary methods is fixed, and input args are passed inside the InputArgs class, not as methodo parameters}
+  there always exists an initial block whose input arguments 
+  might be passed as arguments of the method; however, for
+  implementation  reasons (see the details below) the input variables
+  of all blocks (including the initial one)
+  are implemented as local variables of the method. 
 \end{itemize}
@@
 
-Following an internal link is not difficult in IL bytecode: a jump to
-the corresponding code fragment in the same method is emitted 
-to execute the new block, whereas the appropriate local variables are
-used for passing arguments. \anto{this is wrong for the same reason as above}
-Also following an external link whose target is the initial block of a
-method is not difficult: the corresponding method has to be invoked
-with the appropriate arguments.
+Following an internal link would  not be difficult in IL bytecode: a jump to
+the corresponding code fragment in the same method can be emitted 
+to execute the new block, whereas the appropriate local variables can be
+used for passing arguments. 
+Also following an external link whose target is an initial block could
+be easily implemented, by just invoking the corresponding method.
 
@@
 \emph{block id}, which uniquely identifies the next block of the graph to be executed.
-The high word \anto{a word is 32 bit, block num and method id are 16 bit each} of a block id is the id of the method to which the block
-belongs, whereas the low word is a progressive number univocally identifying
+The high 2 bytes \dacom{word was meant as a fixed-sized group of bits} of a block id constitute the id of the method to which the block
+belongs, whereas the low 2 bytes constitute a progressive number univocally identifying
 each block implemented by the method.
@@
 \includegraphics[height=6cm]{blockid}
-\caption{Method and block ids.}\label{block-id-fig}
+\caption{Method and block ids.\dacom{in the picture ``Main method'' must be
+    replaced with ``Primary method'' and in the primary method the block with number 0 should be added}}\label{block-id-fig}
 \end{center}
@@
 ({\small\lstinline{methodid == MY_METHOD_ID}}), then the appropriate
-jump to the corresponding code is executed, hence internal links
-can be managed efficiently. \anto{wrong: internal links don't go through the dispatcher}
+jump to the corresponding code is executed.
 Otherwise, the \lstinline{jump_to_ext}

### 2008-12-19 talk/ecoop2009/conclusion.tex
add (empty) conclusion file

\section{Related Word}
XXX add me

\section{Conclusion}

XXX add me

### 2008-12-19 talk/ecoop2009/intro.tex
remove the jitstrategy section and integrate its contents into the introduction.

@@
 
-This section will give a high-level overview of how the JIT-generation process
-works. More details will be given in subsequent sections.
+One of the most important aspects that PyPy's translation toolchain can weave
+in is the \emph{automatic generation of a JIT compiler}.  This section will
+give a high-level overview of how the JIT-generation process works. More
+details will be given in subsequent sections.
 
-- write an interpreter
-- add some hints to the interpreter
+The first step is to write an interpreter for the chosen language.  Since it
+must be fed to the translation toolchain, the interpreter has to be written in
+RPython.  Then, to get best performances, we need to add few manual
+annotations to the interpreter, in order to teach the JIT generator which
+informations are important to know at compile-time.  Annotations are inserted
+as \emph{hints}, as described in section \ref{sec:hints}.
 
-translation time:
-- start the translation process, flow graphs
-- run a binding-time analysis using the hints, to figure out ...
-- turn the flow graphs into rainbow bytecode
-- turn the flow graphs of the rainbow interpreter and the original flow graphs
-into .NET code
+It is important to distinguish between three distinct phases of execution:
 
-runtime:
-when the interpreter is started and a function should be interpreted:
- - the generated JIT starts compiling the function 
- - it produces .NET bytecode until it does not have enough information to
- continue. then it stops the compilation and puts a callback into the compiler
- into the generated bytecode.
- - at this point the .NET bytecode that is produced so far is executed.
- - when the callback into the compiler is hit, execution stops and compilation
- starts again
- - afterwards execution continues after the callback (running the code that was
- newly produced)
+\begin{enumerate}
+\item \textbf{Translation-time}: when the translation toolchain runs and the
+  JIT compiler is automatically generated from the interpreter.  The result is
+  an executable that can be used to run programs written in the chosen
+  language.
+\item \textbf{Compile-time}: when the JIT compiler runs, generating executable
+  code on the fly.
+\item \textbf{Runtime}: when the code generated at compile-time is executed.
+\end{enumerate}
+
+Note that in this schema translation-time happens only once, on the
+developer's machine.  By contrast, compile-time and runtime happens every time
+the user wants to run some program.
+
+\subsection{Effective JIT compilation of dynamic languages}
+
+Generating efficient compilers for dynamic languages is hard.  Since these
+languages are dynamically typed, usually the compiler does not have enough
+information to produce efficient code, but instead it has to insert a lot of
+runtime checks to select the appropriate implementation for each operation.
+
+By emitting code at runtime, JIT compilers can exploit some extra knowledge
+compared to traditional static compilers.  However, we need to take special
+care to choose a strategy for JIT compilation that lets the compiler to take
+the best of this advantage.
+
+Most JIT compilers for dynamic languages around (such as IronPython, Jython,
+JRuby \anto{XXX insert some reference}) compile code at the method
+granularity.  If on one hand they can exploit some of the knowledge gathered
+at runtime (e.g. the types of method parameters), on the other hand they can
+do little to optimize most of the operations inside, because their behaviour
+depends on informations that are not available at compile-time, because
+e.g. the global state of the program can change at runtime. \anto{e.g., we can
+  add/remove methods to classes, etc. Should we insert some example here?}
+
+JIT compilers generated by PyPy solve this problem by delaying the compilation
+until they know all the informations needed to generate efficient code.  If at
+some point the JIT compiler does not know about something it needs, it
+generates a callback into itself.
+
+Later, when the generated code is executed, the callback is hit and the JIT
+compiler is restarted again.  At this point, the JIT knows exactly the state
+of the program and can exploit all this extra knowledge to generate highly
+efficient code.  Finally, the old code is patched and linked to the newly
+generated code, so that the next time the JIT compiler will not be invoked
+again.  As a result, \textbf{runtime and compile-time are continuously
+  intermixed}.
+
+Modifying the old code to link the newly generated one is very challenging on
+.NET, as the framework does not offer any primitive to do this.  Section
+\ref{sec:clibackend} explains how it is possible to simulate this behaviour.

### 2008-12-19 talk/ecoop2009/intro.tex
some fixes

@@
 must be fed to the translation toolchain, the interpreter has to be written in
-RPython.  Then, to get best performances, we need to add few manual
+RPython.  Then, to guide the process, we need to add few manual
 annotations to the interpreter, in order to teach the JIT generator which
@@
 
-\subsection{Effective JIT compilation of dynamic languages}
-
 Generating efficient compilers for dynamic languages is hard.  Since these
@@
 JRuby \anto{XXX insert some reference}) compile code at the method
-granularity.  If on one hand they can exploit some of the knowledge gathered
+granularity.  If on the one hand they can exploit some of the knowledge gathered
 at runtime (e.g. the types of method parameters), on the other hand they can
@@
   add/remove methods to classes, etc. Should we insert some example here?}
+\cfbolz{the last sentence is waaay too complicated}
 
@@
 some point the JIT compiler does not know about something it needs, it
-generates a callback into itself.
+generates a callback into itself and stops execution.
 
-Later, when the generated code is executed, the callback is hit and the JIT
+Later, when the generated code is executed, the callback might be hit and the JIT
 compiler is restarted again.  At this point, the JIT knows exactly the state
@@
 again.  As a result, \textbf{runtime and compile-time are continuously
-  intermixed}.
+interleaved}. The primitive to do this sort of interleaving is called promotion,
+it is described in Section \ref{sec:promotion}.
 
-Modifying the old code to link the newly generated one is very challenging on
+One of the most important optimizations that the generated JIT does is removing
+unnecessary allocations. This is described in Section \ref{sec:virtuals}
+
+Modifying the old code to link to the newly generated one is very challenging on
 .NET, as the framework does not offer any primitive to do this.  Section

### 2008-12-19 talk/ecoop2009/jitgen.tex
remove the jitstrategy section and integrate its contents into the introduction.
some fixes
Move the \labels to their proper location. Otherwise the references to them are bogus.
kill a now-unnecessary section

@@
 \begin{figure}[h]
-\label{fig:tlc-main}
 \begin{center}
@@
 \caption{The main loop of the TLC interpreter, written in RPython}
+\label{fig:tlc-main}
 \end{center}
@@
 \begin{figure}[h]
-\label{fig:tlc-folded}
 \begin{center}
@@
 \lstinline{abs} function }
+\label{fig:tlc-folded}
 \end{center}
@@
 
-\subsection{Partial Evaluation in PyPy}
+\subsection{Binding Time Analysis in PyPy}
 
-PyPy contains a framework for generating just-in-time compilers using
-off-line partial evaluation.  As such, there are three distinct phases:
-
-\begin{itemize}
-\item \emph{Translation time:} during the normal translation of an RPython
-program, say the TLC interpreter, we perform binding-time analysis on the
-interpreter.  This produces a generating extension, which is linked with the
-rest of the program. \cfbolz{XXX not quite right}
-
-\item \emph{Compile time:} during the execution of the program, when a new
-bytecode is about to be interpreted, the generating extension is invoked
-instead.  As the generating extension is a compiler, all the computations it
-performs are called compile-time computations.  Its sole effect is to produce
-residual code.
-
-\item \emph{Run time:} the normal execution of the program (which includes the
-time spent running the residual code created by the generating extension).
-\end{itemize}
-
-Translation time is a purely off-line phase; compile time and run time are
-actually highly interleaved during the execution of the program.
-
-\subsection{Binding Time Analysis}
-
+XXX need better transition text from last section
 At translation time, PyPy performs binding-time analysis of the source
@@
 \subsection{Hints}
+\label{sec:hints}
 
@@
 a red value into a green one, which is not possible in classical
-approaches to partial evaluation.  See the Promotion section XXX ref for a
+approaches to partial evaluation.  See Section \ref{sec:promotion} for a
 complete discussion of promotion.

### 2008-12-19 talk/ecoop2009/rainbow.tex
kill a now-unnecessary section

@@
 \section{Promotion}
+\label{sec:promotion}
 
@@
 short-lived objects, even in non-trivial situations.  The following
-example shows a typical case.  XXX use TLC example
-
-The PyPy Python interpreter
-implements application-level integers as boxes – instances of a
-\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
-addition of two integers:
+example shows a typical case.  The TLC interpreter implements application-level
+integers as boxes – instances of an \texttt{IntObject} class with a single
+field.
 

### 2008-12-19 talk/ecoop2009/rainbow.tex
swap promotion and virtuals sections

--- 2008-12-19 talk/ecoop2009/rainbow.tex
@@
-\section{The Rainbow Interpreter}
+\section{Automatic Unboxing of Intermediate Results}
+\label{sec:virtuals}
 
-The JIT compiler is implemented by running the "Rainbow interpreter", whose
-output is a compiled function.
+XXX the following section needs a rewriting to be much more high-level and to
+compare more directly with classical escape analysis
 
-Variables and opcodes are colored:
+Interpreters for dynamic languages typically continuously allocate a lot of small
+objects, for example due to boxing. This makes arithmetic operations extremely
+inefficient. For this reason, we
+implemented a way for the compiler to try to avoid memory allocations in the
+residual code as long as possible. The idea is to try to keep new
+run-time structures "exploded": instead of a single run-time object allocated on
+the heap, the object is "virtualized" as a set
+of fresh local variables, one per field. Only when the object can be accessed by from
+somewhere else is it actually allocated on the heap. The effect of this is similar to that of
+escape analysis \cite{XXX}, which also prevents allocations of objects that can
+be proven to not escape a method or set of methods.
 
-  - green values are known at compile time, and green operations are executed
-    when interpreting the bytecode;
+It is not always possible to keep structures virtual.  The main
+situation in which it needs to be "forced" (i.e. actually allocated at
+run-time) is when the pointer escapes to some non-virtual location like
+a field of a real heap structure.
 
-  - red values are not known until runtime; the result of a red operation is a
-    piece of machine code that will compute the result at runtime.
+Virtual structures still avoid the run-time allocation of most
+short-lived objects, even in non-trivial situations.  The following
+example shows a typical case.  The TLC interpreter implements application-level
+integers as boxes – instances of an \texttt{IntObject} class with a single
+field.
 
-The Rainbow bytecode is produced at translation time, when the JIT compiler is
-generated.
+The PyPy Python interpreter
+implements application-level integers as boxes – instances of a
+\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
+addition of two integers:
 
-\cfbolz{XXX I think we should be very careful with the rainbow interp. it is a
-total implementation-detail and we should only describe it as little as
-possible}
+XXX needs to use TLC examples
+\begin{verbatim}
+    def add(w1, w2):            # w1, w2 are W_IntObject instances
+        value1 = w1.intval
+        value2 = w2.intval
+        result = value1 + value2
+        return W_IntObject(result)
+\end{verbatim}
 
-\subsection{Example of Rainbow bytecode and execution}
+XXX kill the rest?!‽
+When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are
+issued; the intermediate \texttt{W\_IntObject} instance is built by the first
+call and thrown away after the second call.  By contrast, when the
+interpreter is turned into a compiler, the construction of the
+\texttt{W\_IntObject} object leads to a virtual structure whose \texttt{intval}
+field directly references the register in which the run-time addition
+put its result.  This location is read out of the virtual structure at
+the beginning of the second \texttt{add()}, and the second run-time addition
+directly operates on the same register.
 
-TODO
+An interesting effect of virtual structures is that they play nicely with
+promotion.  Indeed, before the interpreter can call the proper \texttt{add()}
+function for integers, it must first determine that the two arguments
+are indeed integer objects.  In the corresponding dispatch logic, we
+have added two hints to promote the type of each of the two arguments.
+This produces a compiler that has the following behavior: in the general
+case, the expression \texttt{a+b} will generate two consecutive run-time
+switches followed by the residual code of the proper version of
+\texttt{add()}.  However, in \texttt{a+b+c}, the virtual structure representing
+the intermediate value will contain a compile-time constant as type.
+Promoting a compile-time constant is trivial – no run-time code is
+generated.  The whole expression \texttt{a+b+c} thus only requires three
+switches instead of four.  It is easy to see that even more switches can
+be skipped in larger examples; typically, in a tight loop manipulating
+only integers, all objects are virtual structures for the compiler and
+the residual code is theoretically optimal – all type propagation and
+boxing/unboxing occurs at compile-time.
 
 \section{Promotion}
+\label{sec:promotion}
 
@@
 XXX maybe a tlc example can be used for the example above?
-
-\section{Automatic Unboxing of Intermediate Results}
-\label{sec:virtuals}
-
-XXX the following section needs a rewriting to be much more high-level and to
-compare more directly with classical escape analysis
-
-Interpreters for dynamic languages typically continuously allocate a lot of small
-objects, for example due to boxing. This makes arithmetic operations extremely
-inefficient. For this reason, we
-implemented a way for the compiler to try to avoid memory allocations in the
-residual code as long as possible. The idea is to try to keep new
-run-time structures "exploded": instead of a single run-time object allocated on
-the heap, the object is "virtualized" as a set
-of fresh local variables, one per field. Only when the object can be accessed by from
-somewhere else is it actually allocated on the heap. The effect of this is similar to that of
-escape analysis \cite{XXX}, which also prevents allocations of objects that can
-be proven to not escape a method or set of methods.
-
-It is not always possible to keep structures virtual.  The main
-situation in which it needs to be "forced" (i.e. actually allocated at
-run-time) is when the pointer escapes to some non-virtual location like
-a field of a real heap structure.
-
-Virtual structures still avoid the run-time allocation of most
-short-lived objects, even in non-trivial situations.  The following
-example shows a typical case.  XXX use TLC example
-
-The PyPy Python interpreter
-implements application-level integers as boxes – instances of a
-\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
-addition of two integers:
-
-XXX needs to use TLC examples
-\begin{verbatim}
-    def add(w1, w2):            # w1, w2 are W_IntObject instances
-        value1 = w1.intval
-        value2 = w2.intval
-        result = value1 + value2
-        return W_IntObject(result)
-\end{verbatim}
-
-XXX kill the rest?!‽
-When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are
-issued; the intermediate \texttt{W\_IntObject} instance is built by the first
-call and thrown away after the second call.  By contrast, when the
-interpreter is turned into a compiler, the construction of the
-\texttt{W\_IntObject} object leads to a virtual structure whose \texttt{intval}
-field directly references the register in which the run-time addition
-put its result.  This location is read out of the virtual structure at
-the beginning of the second \texttt{add()}, and the second run-time addition
-directly operates on the same register.
-
-An interesting effect of virtual structures is that they play nicely with
-promotion.  Indeed, before the interpreter can call the proper \texttt{add()}
-function for integers, it must first determine that the two arguments
-are indeed integer objects.  In the corresponding dispatch logic, we
-have added two hints to promote the type of each of the two arguments.
-This produces a compiler that has the following behavior: in the general
-case, the expression \texttt{a+b} will generate two consecutive run-time
-switches followed by the residual code of the proper version of
-\texttt{add()}.  However, in \texttt{a+b+c}, the virtual structure representing
-the intermediate value will contain a compile-time constant as type.
-Promoting a compile-time constant is trivial – no run-time code is
-generated.  The whole expression \texttt{a+b+c} thus only requires three
-switches instead of four.  It is easy to see that even more switches can
-be skipped in larger examples; typically, in a tight loop manipulating
-only integers, all objects are virtual structures for the compiler and
-the residual code is theoretically optimal – all type propagation and
-boxing/unboxing occurs at compile-time.

### 2008-12-19 talk/ecoop2009/rainbow.tex
A missing dot.

@@
 time. It is impossible in the "classical" approaches to partial
-evaluation, in which the compiler always runs fully ahead of execution
+evaluation, in which the compiler always runs fully ahead of execution.
 This is a problem in many realistic use cases.  For example, in an

### 2008-12-19 talk/ecoop2009/jitgen.tex
add note to not forget
remove note that we agreed on IRC to ignore

@@
 arguments of this functions would typically be the \lstinline{code} argument
-(containing the bytecode as a string \anto{what about using array of bytes instead of string?
-For .NET/JVM people, there is a strong difference between the twos}), the \lstinline{pc} argument (containing
+(containing the bytecode as a string), the \lstinline{pc} argument (containing
 the initial program counter) and the \lstinline{pool} argument (containing the
@@
 
+XXX move this to the promotion section
 Promotion is invoked with the use of a hint as well:

### 2008-12-19 talk/ecoop2009/jitgen.tex
make a figure

@@
 When the function is partially evaluated with respect to the TLC example
-function shown in Figure XXX (which computes the absolute value of a number),
+function shown in Figure \ref{fig:tlc-abs} (which computes the absolute value of a number),
 the residual code would look like in Figure \ref{fig:tlc-folded}. This version

### 2008-12-19 talk/ecoop2009/rainbow.tex
eh. not that many references were left to the rainbow interp anyway. die die
make a figure
add tlc example about virtuals

@@
 run-time) is when the pointer escapes to some non-virtual location like
-a field of a real heap structure.
+a field of a real heap structure.  Virtual structures still avoid the run-time
+ allocation of most short-lived objects, even in non-trivial situations.
 
-Virtual structures still avoid the run-time allocation of most
-short-lived objects, even in non-trivial situations.  The following
-example shows a typical case.  The TLC interpreter implements application-level
-integers as boxes – instances of an \texttt{IntObject} class with a single
-field.
+In addition to virtual structures, the compiler can also handle virtual
+containers, namely lists and dictionaries \footnote{(R)Python's dictionaries
+  are equivalent to .NET \lstinline{Hashtable}s}.  If the indexing operations
+can be evaluated at compile-time (i.e., if the variables holding the indexes
+are green), the compiler internally keeps track of the state of the container
+and store the items as local variables.
 
-The PyPy Python interpreter
-implements application-level integers as boxes – instances of a
-\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
-addition of two integers:
+Look again at figure \ref{fig:tlc-folded}: the list in the \lstinline{stack}
+variable never escapes from the function.  Moreover, all the indexing
+operations (either done explicitly or implicitly by \lstinline{append} and
+\lstinline{pop}) are evaluable at compile-time.  Thus, the list is kept
+\emph{virtual} and its elements are stored in variables $v_n$, where $n$
+represents the index in the list.  Figure \ref{fig:tlc-folded-virtualized}
+show how the resulting code looks like; to ease the reading, the state of the
+\lstinline{stack} as kept by the compiler is shown in the comments.
 
-XXX needs to use TLC examples
-\begin{verbatim}
-    def add(w1, w2):            # w1, w2 are W_IntObject instances
-        value1 = w1.intval
-        value2 = w2.intval
-        result = value1 + value2
-        return W_IntObject(result)
-\end{verbatim}
+\begin{figure}[h]
+\begin{center}
+\input{tlc-folded-virtualized.py}
+\caption{The result of virtualizing the \lstinline{stack} list}
+\label{fig:tlc-main}
+\end{center}
+\end{figure}
 
 XXX kill the rest?!‽
-When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are
-issued; the intermediate \texttt{W\_IntObject} instance is built by the first
-call and thrown away after the second call.  By contrast, when the
-interpreter is turned into a compiler, the construction of the
-\texttt{W\_IntObject} object leads to a virtual structure whose \texttt{intval}
-field directly references the register in which the run-time addition
-put its result.  This location is read out of the virtual structure at
-the beginning of the second \texttt{add()}, and the second run-time addition
-directly operates on the same register.
-
 An interesting effect of virtual structures is that they play nicely with
@@
 \begin{itemize}
+  %XXX remove mention of rainbow interp. but this needs to be rewritten anyway
   \item (compile time): the rainbow interpreter produces machine code until it

### 2008-12-19 talk/ecoop2009/rainbow.tex
add one more para about virtuals, and start the example about promotion

@@
 compare more directly with classical escape analysis
+
+\anto{Maybe we should talk about ``virtual instances'' and not structures,
+  considering the context}
 
@@
 \end{figure}
+
+Even if not shown in the example, \lstinline{stack} is not the only
+virtualized object.  In particular the two objects created by
+\lstinline{IntObj(0)} are also virtualized, and their fields are stored as
+local variables as well.  Virtualizion of instances is important not only
+because it avoids the allocation of unneeded temporary objects, but also
+because it makes possible to optimize method calls on them, as the JIT
+compiler knows their exact type in advance.
 
@@
 better optimization results. 
+\anto{What about this: While traditional PICs are only applied to indirect
+  calls, promotion is a more general operation that can be applied to any kind
+  of value, including instances of user-defined classes or integer numbers}
 
@@
 seen.  This implies that the generated code needs to contain some sort
-of updatable switch, which can pick the right code path based on the
+of updatable switch, or \emph{flexswitch}, which can pick the right code path based on the
 run-time value.
 
-\cfbolz{I think this example is confusing, it is unclear in which order things
-happen how. I will try to come up with something different}.
+Let us look again at the TLC example.  To ease the reading, figure
+\ref{fig:tlc-main} showed a simplified version of TLC's main loop, which did
+not include the hints.  The real implementation of the \lstinline{LT} opcode
+is shown in figure \ref{fig:tlc-main-hints}.
 
-\begin{itemize}
-  %XXX remove mention of rainbow interp. but this needs to be rewritten anyway
-  \item (compile time): the rainbow interpreter produces machine code until it
-    hits a promotion point; e.g.::
+\begin{figure}[h]
+\begin{center}
+\begin{lstlisting}[language=Python]
+        elif opcode == LT:
+            a, b = stack.pop(), stack.pop()
+            hint(a, promote_class=True)
+            hint(b, promote_class=True)
+            stack.append(IntObj(b.lt(a)))
+\end{lstlisting}
+\caption{Usage of hints in TLC's main loop}
+\label{fig:tlc-main-hints}
+\end{center}
+\end{figure}
 
-    \begin{lstlisting}[language=C]
-        y = hint(x, promote=True)
-        return y+10
-    \end{lstlisting}
-
-  \item (compile time): at this point, it generates special machine code that when
-    reached calls the JIT compiler again; the JIT compilation stops::
-
-    \begin{lstlisting}[language=C]
-        switch(y) {
-            default: compile_more(y);
-        }
-    \end{lstlisting}
-
-  \item (runtime): the machine code is executed; when it reaches a promotion
-    point, it executes the special machine code we described in the previous
-    point; the JIT compiler is invoked again;
-
-  \item (compile time): now we finally know the exact value of our red variable,
-    and we can promote it to green; suppose that the value of 'y' is 32::
-
-    \begin{lstlisting}[language=C]
-        switch(y) {
-            32: return 42;
-            default: compile_more(y);
-        }
-    \end{lstlisting}
-
-    Note that the operation "y+10" has been constant-folded into "42", as it
-    was a green operation.
-
-  \item (runtime) the execution restart from the point it stopped, until a new
-    unhandled promotion point is reached.
-\end{itemize}
-
-\subsection{Promotion as Applied to the TLC}
-
-XXX maybe a tlc example can be used for the example above?
+By promoting the class of \lstlisting{a} and \lstlisting{b}, we tell the JIT
+compiler not to generate code until it knows the exact RPython class of both.

### 2008-12-19 talk/ecoop2009/rainbow.tex
work in progress, but they are waiting for me for dinner :-)

@@
 
+\begin{figure}[h]
+\begin{center}
+\begin{lstlisting}[language=Python]
+class IntObj(Obj):
+    ...
+    def lt(self, other): 
+        return self.value < other.int_o()
+\end{lstlisting}
+\caption{Excerpt of the \lstlisting{IntObj} class}
+\label{fig:tlc-intobj}
+\end{center}
+\end{figure}
+
 By promoting the class of \lstlisting{a} and \lstlisting{b}, we tell the JIT
 compiler not to generate code until it knows the exact RPython class of both.
+Figure \ref{fig:tlc-abs-promotion-1} shows the
+code \footnote{\lstinline{switch} is not a legal (R)Python statement, it is
+  used here only as a pseudocode example} generated while compiling the usual
+\lstlisting{abs} function: note that, compared to figure
+\ref{fig:tlc-folded-virtualized}, the code stops just before the calls
+\lstlisting{b.lt(a)}.
+
+\begin{figure}[h]
+\begin{center}
+\begin{lstlisting}[language=Python]
+def interp_eval_abs(args):
+    v0 = args[0]
+    v1 = IntObj(0)
+    a, b = v0, v1
+    hint(a, promote_class=True) # no-op
+    cls_a = a.__class__
+    switch cls_a:
+        default: 
+            continue_compilation(jitstate, cls_a)
+\end{lstlisting}
+\caption{Promotion example 1}
+\label{fig:tlc-abs-promotion-1}
+\end{center}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\begin{lstlisting}[language=Python]
+def interp_eval_abs(args):
+    v0 = args[0]
+    v1 = IntObj(0)
+    a, b = v0, v1
+    hint(a, promote_class=True) # no-op
+    cls_a = a.__class__
+    switch cls_a:
+        IntObj:
+            hint(b, promote_class=True)
+            v0 = IntObj(a.value < b.int_o())
+            ...
+        default: 
+            continue_compilation(jitstate, cls_a)
+\end{lstlisting}
+\caption{Promotion example 2}
+\label{fig:tlc-abs-promotion-2}
+\end{center}
+\end{figure}
+
+The first time the flexswitch is executed, the \lstlisting{default} branch is
+taken, and the special function \lstlisting{continue_compilation} restarts the
+JIT compiler, passing it the just-seen value of \lstlisting{cls_a}.  The JIT
+compiler generates new specialized code, and \emph{patches} the flexswitch to
+add the new case, which is then executed.
+
+If later an instance of \lstlisting{IntObj} hits the flexswitch again, the
+code is executed without needing of more calls to the JIT compiler.  On the
+other hand, if the flexswitch is hit by an instance of some other class, the
+\lstlisting{default} branch will be selected again and the whole process will
+restart.
+
+Now, let us examine the content of the \lstlisting{IntObj} case: first, there
+is a hint to promote the class of \lstlisting{b}.  Although in general
+promotion is implemented through a flexswitch, in this case it is not needed
+as \lstlisting{b} holds a \emph{virtual instance}, whose class is already
+known (as described in previous section).
+
+%since at this
+%point we know the exact class of \lstinline{a} as we can see, the body of the
+%\lstlisting{IntObj.lt} method has been inlined,

### 2008-12-19 talk/ecoop2009/jitgen.tex
add one more para about virtuals, and start the example about promotion
fix an XXX

@@
 
-\cfbolz{XXX: unclear how the next paragraph will fit into the text in the end.
-it's certainly wrong as it is}
-Once binding times have been determined, one possible approach to
-producing the generating extension itself is by self-applying on-line
-partial evaluators.  This is known as the second Futamura projection
-\cite{Futamura99}.  So far it is unclear if this approach can lead to optimal
-results, or even if it scales well.  In PyPy we selected a more direct
-approach: the generating extension is produced by transformation of the
-control flow graphs of the interpreter, guided by the binding times.  We
-call this process \emph{timeshifting}.
-
 XXX write something about the problems of classical PE?
@@
 RPython program after it has been turned to low-level graphs, i.e. at
-the level at which operations manipulate pointer-and-structure-like
-objects.
+the level at which operations manipulate objects that have a fixed type
+attached.
 
@@
 abstract-interpretation based analysis. It is based on the
-same type inference engine that is used on the source RPython program,
-the annotator \anto{I'm not sure we should mention the annotator, as it is not referred anywhere else}.  In this mode, it is called the \emph{hint-annotator}; it
+same type inference engine that is used on the source RPython program.
+This is called the \emph{hint-annotator}; it
 operates over input graphs that are already low-level instead of
@@
 value dependencies and manually-provided binding time hints.
-XXX the above needs rewriting when the background section is there
 
@@
 concrete=True)} is reported as an error.
-
-XXX move this to the promotion section
-Promotion is invoked with the use of a hint as well:
-\texttt{v2 = hint(v1, promote=True)}.
-This hint is a \emph{local} request for \texttt{v2} to be green, without
-requiring \texttt{v1} to be green.  Note that this amounts to copying
-a red value into a green one, which is not possible in classical
-approaches to partial evaluation.  See Section \ref{sec:promotion} for a
-complete discussion of promotion.

### 2008-12-19 talk/ecoop2009/rainbow.tex
fix an XXX

@@
 
+Promotion is invoked with the use of a hint as well:
+\texttt{v2 = hint(v1, promote=True)}.
+This hint is a \emph{local} request for \texttt{v2} to be green, without
+requiring \texttt{v1} to be green.  Note that this amounts to copying
+a red value into a green one, which is not possible in classical
+approaches to partial evaluation. A slightly different hint can be used to
+promote the \emph{class} of an instance. This is done with \lstinline{hint(v1,
+promote_class=True)}. It does not have an effect on the bindings of any
+variable.
+
 \subsection{Implementing Promotion}

### 2008-12-19 talk/ecoop2009/rainbow.tex
finish the promotion example

@@
         return self.value < other.int_o()
+    def int_o(self):
+        return self.value
 \end{lstlisting}
@@
             hint(b, promote_class=True)
-            v0 = IntObj(a.value < b.int_o())
+            v0 = IntObj(b.value < a.value)
             ...
@@
 
-The first time the flexswitch is executed, the \lstlisting{default} branch is
-taken, and the special function \lstlisting{continue_compilation} restarts the
-JIT compiler, passing it the just-seen value of \lstlisting{cls_a}.  The JIT
+The first time the flexswitch is executed, the \lstinline{default} branch is
+taken, and the special function \lstinline{continue_compilation} restarts the
+JIT compiler, passing it the just-seen value of \lstinline{cls_a}.  The JIT
 compiler generates new specialized code, and \emph{patches} the flexswitch to
@@
 
-If later an instance of \lstlisting{IntObj} hits the flexswitch again, the
+If later an instance of \lstinline{IntObj} hits the flexswitch again, the
 code is executed without needing of more calls to the JIT compiler.  On the
 other hand, if the flexswitch is hit by an instance of some other class, the
-\lstlisting{default} branch will be selected again and the whole process will
+\lstinline{default} branch will be selected again and the whole process will
 restart.
 
-Now, let us examine the content of the \lstlisting{IntObj} case: first, there
-is a hint to promote the class of \lstlisting{b}.  Although in general
+Now, let us examine the content of the \lstinline{IntObj} case: first, there
+is a hint to promote the class of \lstinline{b}.  Although in general
 promotion is implemented through a flexswitch, in this case it is not needed
-as \lstlisting{b} holds a \emph{virtual instance}, whose class is already
+as \lstinline{b} holds a \emph{virtual instance}, whose class is already
 known (as described in previous section).
 
-%since at this
-%point we know the exact class of \lstinline{a} as we can see, the body of the
-%\lstlisting{IntObj.lt} method has been inlined,
+Then, the compiler knows the exact class of \lstinline{b}, thus it can inline
+the calls to \lstinline{lt}.  Moreover, inside \lstinline{lt} there is a
+call to \lstinline{a.int_o()}, which is inlined as well for the very same
+reason.
+
+Moreover, as we saw in section \ref{sec:virtuals}, the \lstinline{IntObj}
+instance can be virtualized, so that the subsequent \lstinline{BR_COND} opcode
+can be compiled efficiently without needing any more flexswitch.

### 2008-12-19 talk/ecoop2009/rainbow.tex
some improvements, notes

@@
 compiler knows their exact type in advance.
-
-XXX kill the rest?!‽
-An interesting effect of virtual structures is that they play nicely with
-promotion.  Indeed, before the interpreter can call the proper \texttt{add()}
-function for integers, it must first determine that the two arguments
-are indeed integer objects.  In the corresponding dispatch logic, we
-have added two hints to promote the type of each of the two arguments.
-This produces a compiler that has the following behavior: in the general
-case, the expression \texttt{a+b} will generate two consecutive run-time
-switches followed by the residual code of the proper version of
-\texttt{add()}.  However, in \texttt{a+b+c}, the virtual structure representing
-the intermediate value will contain a compile-time constant as type.
-Promoting a compile-time constant is trivial – no run-time code is
-generated.  The whole expression \texttt{a+b+c} thus only requires three
-switches instead of four.  It is easy to see that even more switches can
-be skipped in larger examples; typically, in a tight loop manipulating
-only integers, all objects are virtual structures for the compiler and
-the residual code is theoretically optimal – all type propagation and
-boxing/unboxing occurs at compile-time.
 
@@
 approaches to partial evaluation. A slightly different hint can be used to
-promote the \emph{class} of an instance. This is done with \lstinline{hint(v1,
-promote_class=True)}. It does not have an effect on the bindings of any
-variable.
+promote the \emph{class} of an instance. This is done with
+\lstinline{hint(v1, promote_class=True)}. It does not have an effect on the
+bindings of any variable.
 
@@
 \ref{fig:tlc-main} showed a simplified version of TLC's main loop, which did
-not include the hints.  The real implementation of the \lstinline{LT} opcode
-is shown in figure \ref{fig:tlc-main-hints}.
+not include the hints.  The implementation of the \lstinline{LT} opcode with
+hints added is shown in figure \ref{fig:tlc-main-hints}.
 
@@
 \begin{lstlisting}[language=Python]
+def interp_eval(code, pc, args, pool):
+    code_len = len(code)
+    stack = []
+    while pc < code_len:
+        opcode = ord(code[pc])
+        opcode = hint(opcode, promote_class=True)
+        pc += 1
+
+        if opcode == PUSH:
+            ...
         elif opcode == LT:
@@
         return self.value < other.int_o()
+    def sub(self, other):
+        return IntObj(self.value - other.int_o())
     def int_o(self):
@@
 \end{lstlisting}
-\caption{Excerpt of the \lstlisting{IntObj} class}
+\caption{Excerpt of the \lstinline{IntObj} class}
 \label{fig:tlc-intobj}
@@
 
-By promoting the class of \lstlisting{a} and \lstlisting{b}, we tell the JIT
+By promoting the class of \lstinline{a} and \lstinline{b}, we tell the JIT
 compiler not to generate code until it knows the exact RPython class of both.
@@
   used here only as a pseudocode example} generated while compiling the usual
-\lstlisting{abs} function: note that, compared to figure
+\lstinline{abs} function: note that, compared to figure
 \ref{fig:tlc-folded-virtualized}, the code stops just before the calls
-\lstlisting{b.lt(a)}.
+\lstinline{b.lt(a)}.
 
@@
     a, b = v0, v1
-    hint(a, promote_class=True) # no-op
+    # hint(a, promote_class=True) implemented as follows:
     cls_a = a.__class__
@@
         IntObj:
-            hint(b, promote_class=True)
+            # hint(b, promote_class=True) needs no code
             v0 = IntObj(b.value < a.value)
-            ...
+            if v0.value:
+                return a
+            else:
+                v0 = IntObj(0)
+                v1 = args[0]
+                a, b = v0, v1
+                v0 = IntObj(b.value - a.value)
+                return v0
         default: 

### 2008-12-19 talk/ecoop2009/conclusion.tex
add material to the related work section, needs an overhaul

@@
 \section{Related Word}
-XXX add me
+Promotion is a concept that we have already explored in other contexts. Psyco is
+a run-time specialiser for Python that uses promotion (called ``unlift'' in
+\cite{DBLP:conf/pepm/Rigo04}). However, Psyco is a manually written JIT, is
+not applicable to other languages and cannot be retargetted.
+
+ - reference standard jit stuff, PICs \cite{hoelzle_optimizing_1991}, this one: http://www.cs.ucsb.edu/~urs/oocsb/self/papers/type-feedback.html
+
+ - compare against tracing JITs \cite{gal_hotpathvm_2006}. tracing jits make
+ implementing jits much easier, maybe less need for JIT generation. also, they
+ concentrate on loops, which makes them produce a lot less code (which will need
+ to be addressed in our work as well).
+
+ so far tracing JITs have less general allocation removal techniques, which
+ makes them get less speedup in a dynamic language with boxing
+
+in my master I had this snippet:
+
+"""
+The generated code typically contains guards; in recent research
+\cite{gal_incremental_2006} on Java, these guards' behaviour is extended to be
+similar to our promotion.  This has been used twice to implement a dynamic
+language (JavaScript), by Tamarin\footnote{{\tt
+http://www.mozilla.org/projects/tamarin/}} and in \cite{chang_efficient_2007}.
+"""
+
+There has been an enormous amount of work on partial evaluation for compiler
+generation. A good introduction is given in \cite{Jones:peval}. However, most of
+it is for generating ahead-of-time compilers, which cannot produce very good
+performance results for dynamic languages.
+
+However, there is also some research on runtime partial evaluation. One of the
+earliest examples is Tempo for C
+\cite{DBLP:conf/popl/ConselN96,DBLP:conf/dagstuhl/ConselHNNV96}. However, it is
+essentially an offline specializer ``packaged as a library''; decisions about
+what can be specialized and how are pre-determined.
+
+Another work in this direction is DyC \cite{grant_dyc_2000}, another runtime
+specializer for C.  Specialization decisions are also pre-determined, but
+``polyvariant program-point specialization'' gives a coarse-grained equivalent
+of our promotion.  Targeting the C language makes higher-level specialization
+difficult, though (e.g.\ \texttt{mallocs} are not removed).
+
+Greg Sullivan introduced "Dynamic Partial Evaluation", which is a special
+form of partial evaluation at runtime \cite{sullivan_dynamic_2001} and describes
+an implementation for a small dynamic language based on lambda calculus. This
+work is conceptually very close to our own.
+% XXX there are no performance figures, we have no clue how much of this is
+% implemented. not sure how to write this
+
+ - apart from that we maybe should talk about escape analysis
+ \cite{Blanchet99escapeanalysis}, \cite{Choi99escapeanalysis}
+ our algorithm is totally simple-minded in comparison, but is useful in practise
 

### 2008-12-19 talk/ecoop2009/rainbow.tex
small fixes

@@
 \label{sec:virtuals}
-
-XXX the following section needs a rewriting to be much more high-level and to
-compare more directly with classical escape analysis
-
-\anto{Maybe we should talk about ``virtual instances'' and not structures,
-  considering the context}
 
@@
 residual code as long as possible. The idea is to try to keep new
-run-time structures "exploded": instead of a single run-time object allocated on
+run-time instances "exploded": instead of a single run-time object allocated on
 the heap, the object is "virtualized" as a set
@@
 somewhere else is it actually allocated on the heap. The effect of this is similar to that of
-escape analysis \cite{XXX}, which also prevents allocations of objects that can
-be proven to not escape a method or set of methods.
+escape analysis \cite{Blanchet99escapeanalysis}, \cite{Choi99escapeanalysis},
+which also prevents allocations of objects that can be proven to not escape a
+method or set of methods (the algorithms however are a lot more advanced than
+our very simple analysis).
 
-It is not always possible to keep structures virtual.  The main
+It is not always possible to keep instances virtual.  The main
 situation in which it needs to be "forced" (i.e. actually allocated at
 run-time) is when the pointer escapes to some non-virtual location like
-a field of a real heap structure.  Virtual structures still avoid the run-time
+a field of a real heap structure.  Virtual instances still avoid the run-time
  allocation of most short-lived objects, even in non-trivial situations.
 
-In addition to virtual structures, the compiler can also handle virtual
+In addition to virtual instances, the compiler can also handle virtual
 containers, namely lists and dictionaries \footnote{(R)Python's dictionaries
@@
 \end{lstlisting}
-\caption{Promotion example 1}
+\caption{Promotion step 1}
 \label{fig:tlc-abs-promotion-1}
@@
 \end{lstlisting}
-\caption{Promotion example 2}
+\caption{Promotion step 2}
 \label{fig:tlc-abs-promotion-2}

### 2008-12-20 talk/ecoop2009/rainbow.tex
I guess this was a typo

@@
         opcode = ord(code[pc])
-        opcode = hint(opcode, promote_class=True)
+        opcode = hint(opcode, concrete=True)
         pc += 1

### 2008-12-20 talk/ecoop2009/clibackend.tex
remove the jitstrategy section and integrate its contents into the introduction.
eh. not that many references were left to the rainbow interp anyway. die die
(arigato around) add an XXX
expand on the XXX
several small fixes.  Some rewording in the cli backend section

@@
-\section{CLI backend for Rainbow interpreter}
+\section{The CLI backend for the Generated JIT}
+\label{sec:clibackend}
 
@@
 
+\anto{XXX: add an overview section of the backend, else we only talk about
+  flexswitches}
+\cfbolz{it should maybe also have a nice introductory paragraph how the
+interface between the JIT works?}
+
 \subsection{Flexswitches}
 
-As already explained, \dacom{I guess flexswitches will be introduced
-  in the previous sect.} \emph{flexswitch} is one of the key
+As already explained, \emph{flexswitch} is one of the key
 concepts allowing the JIT compiler generator to produce code which can
@@
 A flexswitch is a special kind of switch which can be dynamically
-extended with new cases; intuitively, its behavior can be described
-well in terms of flow graphs. Indeed, a flexswitch can be considered 
+extended with new cases.  Intuitively, its behavior can be described
+well in terms of flow graphs: a flexswitch can be considered 
 as a special flow graph block where links to newly created blocks are
@@
 any time a new case has to be added to a flexswitch.
-In this way, whereas flow graphs without flexswitches are translated
-to a single method, the translation of flow graphs which can dynamically grow because of
-flexswitches will be scattered over several methods.
-Summarizing, the backend behaves in the following way:
+
+It is important to underline the difference between flow graphs and a methods:
+the first are the logical unit of code as seen by the JIT compiler, each of
+them being concretely implemented by \emph{one or more} methods.
+
+In this way, whereas flow graphs without flexswitches are translated to a
+single method, the translation of \emph{growable} flow graphs will be
+scattered over several methods.  Summarizing, the backend behaves in the
+following way:
 \begin{itemize}
@@
 \item Each either primary or secondary method implements a certain
-  number of blocks, all belonging to the same flow graph. Among these blocks
-  there always exists an initial block whose input arguments 
-  might be passed as arguments of the method; however, for
-  implementation  reasons (see the details below) the input variables
-  of all blocks (including the initial one)
-  are implemented as local variables of the method. 
+  number of blocks, all belonging to the same flow graph.
 \end{itemize}
 
-When  a new case is added to a flexswitch, new blocks are generated
-and translated by the backend in a new single method pointed
-by a delegate \footnote{\emph{Delegates} are the .NET equivalent of function pointers}
- of  which is stored in the code implementing the flexswitch,
-so that the method can be invoked later.
+When a new case is added to a flexswitch, the backend generates the new blocks
+into a new single method.  The newly created method is pointed by a
+delegate \footnote{\emph{Delegates} are the .NET equivalent of function
+  pointers} stored in the flexswitch, so that it can be invoked later when
+needed.
 
@@
 
-A link is called \emph{internal} if it connects two blocks implemented
-by the same method,
+A link is called \emph{internal} if it connects two blocks contained
+in the same method,
  \emph{external} otherwise.
 
-Following an internal link would  not be difficult in IL bytecode: a jump to
+Following an internal link is easy in IL bytecode: a jump to
 the corresponding code fragment in the same method can be emitted 
@@
 used for passing arguments.
-Also following an external link whose target is an initial block could
+
+Following an external link whose target is an initial block could also
 be easily implemented, by just invoking the corresponding method.
-
 What cannot be easily implemented in CLI is following an external link
@@
 outgoing link of the block dynamically added in the right-hand side
-picture of Figure~\ref{flexswitch-fig}. How it is possible to pass the
-right arguments to the target block?
+picture of Figure~\ref{flexswitch-fig}. How it is possible to jump in 
+the middle of a method?
 
 To solve this problem every method contains a special code, called
-\emph{dispatcher}; whenever a method is invoked, its dispatcher is
+\emph{dispatcher}: whenever a method is invoked, its dispatcher is
 executed first\footnote{The dispatcher should not be
@@
 \emph{block id}, which uniquely identifies the next block of the graph to be executed.
-The high 2 bytes \dacom{word was meant as a fixed-sized group of bits} of a block id constitute the id of the method to which the block
-belongs, whereas the low 2 bytes constitute a progressive number univocally identifying
-each block implemented by the method.
+The high 2 bytes of a block id constitute the \emph{method id}, which 
+univocally identifies a method in a graph, whereas the low 2 bytes constitute
+a progressive number univocally identifying a block inside each method.
 
@@
 
-Therefore, the only solution we came up with is defining a class
+Therefore, the solution we came up with is defining a class
 \lstinline{InputArgs} for passing sequences of arguments whose length

### 2008-12-20 talk/ecoop2009/tlc.tex
make a figure
several small fixes.  Some rewording in the cli backend section

@@
   \lstinline{GT}, etc.
-\item \textbf{Object-oriented}: operations on objects: \lstinline{NEW},
+\item \textbf{Object-oriented} operations: \lstinline{NEW},
   \lstinline{GETATTR}, \lstinline{SETATTR}, \lstinline{SEND}.
@@
 As we said above, TLC exists only at bytecode level; to ease the development
-of TLC programs, we wrote an assembler that generates TLC bytecode. The
-following example shows a simple program that computes the absolute value of
-the given integer:
+of TLC programs, we wrote an assembler that generates TLC bytecode. Figure \ref{fig:tlc-abs}
+shows a simple program that computes the absolute value of
+the given integer.
 
+\begin{figure}[h]
+\begin{center}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{The TLC bytecode for computing the absolute value of a function}
+\label{fig:tlc-abs}
+\end{center}
+\end{figure}
 

### 2008-12-20 talk/ecoop2009/jitgen.tex
several small fixes.  Some rewording in the cli backend section

@@
 The JIT generation framework uses partial evaluation techniques to generate a
-dynamic compiler from an interpreter; the idea is inspired by Psyco, which
+dynamic compiler from an interpreter; the idea is inspired by Psyco \cite{DBLP:conf/pepm/Rigo04}, which
 uses the same techniques but it's manually written instead of being
@@
 operates over input graphs that are already low-level instead of
-RPython-level, and propagates annotations that do not track types but
+RPython-level, \anto{we never make distinction between low-level and rpython-level 
+flowgraphs, do we? I propose to talk about ``intermediate flowgraphs''} 
+and propagates annotations that do not track types but
 value dependencies and manually-provided binding time hints.
@@
 concrete=True)} is reported as an error.
+
+\anto{maybe we should at least mention promotion, and refer to the proper
+  section for details?}

### 2008-12-20 talk/ecoop2009/rainbow.tex
several small fixes.  Some rewording in the cli backend section

@@
 residual code as long as possible. The idea is to try to keep new
-run-time instances "exploded": instead of a single run-time object allocated on
-the heap, the object is "virtualized" as a set
+run-time instances \emph{exploded}: instead of a single run-time object allocated on
+the heap, the object is \emph{virtualized} as a set
 of fresh local variables, one per field. Only when the object can be accessed by from
@@
 It is not always possible to keep instances virtual.  The main
-situation in which it needs to be "forced" (i.e. actually allocated at
+situation in which it needs to be \emph{forced} (i.e. actually allocated at
 run-time) is when the pointer escapes to some non-virtual location like
@@
 From a partial evaluation point of view, promotion is the converse of
-the operation generally known as "lift" \cite{XXX}.  Lifting a value means
+the operation generally known as \emph{lift} \cite{XXX}.  Lifting a value means
 copying a variable whose binding time is compile-time into a variable
 whose binding time is run-time – it corresponds to the compiler
-"forgetting" a particular value that it knew about.  By contrast,
+``forgetting'' a particular value that it knew about.  By contrast,
 promotion is a way for the compiler to gain \emph{more} information about
@@
 otherwise the compiler can only use information that is known ahead of
-time. It is impossible in the "classical" approaches to partial
+time. It is impossible in the ``classical'' approaches to partial
 evaluation, in which the compiler always runs fully ahead of execution.
@@
 
-A very different point of view on promotion is as a generalization of techniques
-that already exist in dynamic compilers as found in modern virtual machines for
-object-oriented language.  In this context feedback
-techniques are crucial for good results.  The main goal is to
-optimize and reduce the overhead of dynamic dispatching and indirect
-invocation.  This is achieved with variations on the technique of
-polymorphic inline caches \cite{hoelzle_optimizing_1991}: the dynamic lookups are cached and
-the corresponding generated machine code contains chains of
-compare-and-jump instructions which are modified at run-time.  These
-techniques also allow the gathering of information to direct inlining for even
-better optimization results. 
-\anto{What about this: While traditional PICs are only applied to indirect
-  calls, promotion is a more general operation that can be applied to any kind
-  of value, including instances of user-defined classes or integer numbers}
+A very different point of view on promotion is as a generalization of
+techniques that already exist in dynamic compilers as found in modern virtual
+machines for object-oriented language, like \emph{Polymorphic Inline Cache}
+(PIC, \cite{hoelzle_optimizing_1991}) and its variations, whose main goal is
+to optimize and reduce the overhead of dynamic dispatching and indirect
+invocation: the dynamic lookups are cached and the corresponding generated
+machine code contains chains of compare-and-jump instructions which are
+modified at run-time.  These techniques also allow the gathering of
+information to direct inlining for even better optimization results. Compared
+to PICs, promotion is more general because it can be applied not only to
+indirect calls but to any kind of value, including instances of user-defined
+classes or integer numbers.
 
@@
 as \lstinline{b} holds a \emph{virtual instance}, whose class is already
-known (as described in previous section).
+known (as described in the previous section).
 
@@
 call to \lstinline{a.int_o()}, which is inlined as well for the very same
-reason.
-
-Moreover, as we saw in section \ref{sec:virtuals}, the \lstinline{IntObj}
+reason.  Moreover, as we saw in section \ref{sec:virtuals}, the \lstinline{IntObj}
 instance can be virtualized, so that the subsequent \lstinline{BR_COND} opcode
 can be compiled efficiently without needing any more flexswitch.
+
+\anto{We should show the very final version of the code, with all instances virtualized}

### 2008-12-20 talk/ecoop2009/rainbow.tex
put two code snippets side by side, and fix a typo

@@
 \caption{The result of virtualizing the \lstinline{stack} list}
-\label{fig:tlc-main}
+\label{fig:tlc-folded-virtualized}
 \end{center}
@@
 Promotion is invoked with the use of a hint as well:
-\texttt{v2 = hint(v1, promote=True)}.
+\lstinline{v2 = hint(v1, promote=True)}.
 This hint is a \emph{local} request for \texttt{v2} to be green, without
@@
 \begin{center}
-\begin{lstlisting}[language=Python]
+\begin{tabular}{l|l}
+\begin{lstlisting}
 def interp_eval(code, pc, args, pool):
@@
 \end{lstlisting}
-\caption{Usage of hints in TLC's main loop}
-\label{fig:tlc-main-hints}
-\end{center}
-\end{figure}
+&
+\hspace{2pt}
+\begin{lstlisting}
+class IntObj(Obj):
 
-\begin{figure}[h]
-\begin{center}
-\begin{lstlisting}[language=Python]
-class IntObj(Obj):
-    ...
   def lt(self, other): 
-        return self.value < other.int_o()
+    return (self.value < 
+            other.int_o())
+
   def sub(self, other):
-        return IntObj(self.value - other.int_o())
+    return IntObj(self.value -
+                  other.int_o())
+
   def int_o(self):
     return self.value
+
+  ...
 \end{lstlisting}
-\caption{Excerpt of the \lstinline{IntObj} class}
-\label{fig:tlc-intobj}
+\end{tabular}
 \end{center}
+\caption{Usage of hints in TLC's main loop and excerpt of the \lstinline{IntObj} class}
+\label{fig:tlc-main-hints}
 \end{figure}

### 2008-12-20 talk/ecoop2009/rainbow.tex
add the final version of the code

@@
 \subsection{Implementing Promotion}
+\label{sec:implementing-promotion}
 
@@
             v0 = IntObj(b.value < a.value)
-            if v0.value:
+            cond = v0
+            if cond.value:
                 return a
@@
 
-\anto{We should show the very final version of the code, with all instances virtualized}
+Figure~\ref{fig:tlc-abs-final} shows the final, fully optimized version of the
+code, with all the instances virtualized and the unneeded temporary variables
+removed.
+
+\begin{figure}[h]
+\begin{center}
+\begin{lstlisting}[language=Python]
+def interp_eval_abs(args):
+    a = args[0]
+    cls_a = a.__class__
+    switch cls_a:
+        IntObj:
+            # 0 is the constant "value" field of the virtualized IntObj(0)
+            if 0 < a.value:
+                return a
+            else:
+                return IntObj(0 - a.value)
+        default: 
+            continue_compilation(jitstate, cls_a)
+\end{lstlisting}
+\caption{Final optimized version of the \lstinline{abs} example}
+\label{fig:tlc-abs-final}
+\end{center}
+\end{figure}

### 2008-12-20 talk/ecoop2009/clibackend.tex
Standardize the footnote format.

@@
 into a new single method.  The newly created method is pointed by a
-delegate \footnote{\emph{Delegates} are the .NET equivalent of function
-  pointers} stored in the flexswitch, so that it can be invoked later when
+delegate\footnote{\emph{Delegates} are the .NET equivalent of function
+  pointers.} stored in the flexswitch, so that it can be invoked later when
 needed.

### 2008-12-20 talk/ecoop2009/rainbow.tex
Standardize the footnote format.

@@
 In addition to virtual instances, the compiler can also handle virtual
-containers, namely lists and dictionaries \footnote{(R)Python's dictionaries
+containers, namely lists and dictionaries\footnote{(R)Python's dictionaries
   are equivalent to .NET \lstinline{Hashtable}s}.  If the indexing operations
@@
 Figure \ref{fig:tlc-abs-promotion-1} shows the
-code \footnote{\lstinline{switch} is not a legal (R)Python statement, it is
-  used here only as a pseudocode example} generated while compiling the usual
+code\footnote{\lstinline{switch} is not a legal (R)Python statement, it is
+  used here only as a pseudocode example.} generated while compiling the usual
 \lstinline{abs} function: note that, compared to figure
-\ref{fig:tlc-folded-virtualized}, the code stops just before the calls
+\ref{fig:tlc-folded-virtualized}, the code stops just before the call
 \lstinline{b.lt(a)}.

### 2008-12-20 talk/ecoop2009/clibackend.tex
a introduction about jit backends and jit layering, fix an XXX

@@
 
-%\subsection{Promotion on CLI}
-%
-%Implementing promotion on top of CLI is not straightforward, as it needs to
-%patch and modify the already generated code, and this is not possible on .NET.
-%
-%To solve, we do xxx and yyy etc. etc.
-%
+\subsection{JIT layering}
 
-\anto{XXX: add an overview section of the backend, else we only talk about
-  flexswitches}
-\cfbolz{it should maybe also have a nice introductory paragraph how the
-interface between the JIT works?}
+From the implementation point of view, the JIT generator is divided into a
+frontend and several backends.  The goal of the frontend is to generate a JIT
+compiler which works as described in the previous sections.  Internally, the
+frontend represents the compiled code as \emph{flow graphs}, and the role of
+the backends is to translate flowgraphs into machine code.
+
+At the moment of writing, three backends have been implemented: one for Intel
+\emph{x86} processors, one for \emph{PowerPC} processors, and one for the
+\emph{CLI Virtual Machine}.  The latter is special because instead of emitting
+code for a real CPU it emits code for a virtual machine: before being
+executed, the generated code will be compiled again by .NET's own JIT
+compiler.
+
+Thus, when using the CLI backend, we actually have two layers of
+JIT-compilation, each layer removing away different kinds of overhead.  By
+operating at a higher level, our JIT can potentially do a better job than the
+.NET one in some contexts, as our benchmarks demonstrate (see
+section~\ref{sec:benchmarks}).  On the other hand, the lower-level .NET JIT is
+very good at producing machine code, much more than PyPy's own \emph{x86}
+backend for example.  By combining the strengths of both we can get highly
+efficient machine code.
+
+As usual, the drawback is that programs that runs for a very short period of
+time could run slower with JIT than without, due to the time spent doing the
+initial (double) compilation.  It is important to underline that so far we
+mostly focused making the JIT able to produce very fast code, without trying
+to optimzed the compilation phase itself.
 
@@
 
-As already explained, \emph{flexswitch} is one of the key
-concepts allowing the JIT compiler generator to produce code which can
-be incrementally specialized and compiled at run time.
+For a large part, implementing the CLI backend is easy and straightforward, as
+there is a close corrispondence between most of the operations used by
+frontend's flowgraphs and the CLI instructions.  Thus, we will not dig into
+details for this part.
+
+However the concept of \emph{flexswitch}, as described in
+Section~\ref{sec:implementing-promotion}, does not have any direct equivalent
+in the CLI model, and it is hard to implement efficiently.
 
@@
 \includegraphics[height=6cm]{blockid}
-\caption{Method and block ids.\dacom{in the picture ``Main method'' must be
-    replaced with ``Primary method'' and in the primary method the block with number 0 should be added}}\label{block-id-fig}
+\caption{Method and block ids.}\label{block-id-fig}
 \end{center}

### 2008-12-20 talk/ecoop2009/benchmarks.tex
some improvements, notes
comment out a possible confusing example, and add an XXX
a introduction about jit backends and jit layering, fix an XXX
hide the python-like syntax for factorial and fibonacci, add a link to the sources

@@
 \section{Benchmarks}
+\label{sec:benchmarks}
 
@@
 Moreover, for each benchmark we also show the time taken by running the
-equivalent program written in C\#.  By comparing the results against C\#, we
-can see how far we are from the supposed optimal performances.  \anto{I
-  don't really like the last sentence, but right now I can't think of another
-  way to phrase it.  Rewording welcome :-)}
+equivalent program written in C\# \footnote{The sources for both TLC and C\#
+  programs are available at:
+  http://codespeak.net/svn/pypy/extradoc/talk/ecoop2009/benchmarks/}
 
@@
 that computes the factorial of a given number.  The algorithm is
-straightforward, and the loop contains only three operations: one
+straightforward, thus we are not showing the source code.  The loop contains only three operations: one
 multiplication, one subtraction, and one comparison to check if we have
-finished the job:
+finished the job.
+
+\commentout{
+\cfbolz{I think we can kill this for space reasons}
+\anto{If we decide to keep them, we must remember to explain the python-like
+  syntax, as it is no longer in tlc.tex}
 
@@
 \end{lstlisting}
+}
 
 When doing plain interpretation, we need to create and destroy three temporary
-objects at each iterations.  By contrast, the code generated by the JIT does
+objects at each iteration.  By contrast, the code generated by the JIT does
 much better.  At the first iteration, the classes of the two operands of the
@@
 Similarly, we wrote a program to calculate the $n_{th}$ Fibonacci number, for
-which we can do the same reasoning as above:
+which we can do the same reasoning as above.
 
+\commentout{
+\cfbolz{I think we can kill this for space reasons}
 \begin{lstlisting}
@@
 \end{lstlisting}
+}
 
@@
 
+\cfbolz{can we make one figure of all the tables?}
 \begin{table}[ht]
@@
 
-Tables \ref{tab:factorial} and \ref{tab:fibo} show the time spent to calculate
+Tables \ref{tab:factorial} and \ref{tab:fibo} show the seconds spent to calculate
 the factorial and Fibonacci for various $n$.  As we can see, for small values

### 2008-12-20 talk/ecoop2009/benchmarks.tex
explain that the syntax is invented

@@
 
-The benchmarks have been run on machine with Intel Pentium 4 CPU running at
+The benchmarks have been run on machine with an Intel Pentium 4 CPU running at
 3.20 GHz and 2 GB of RAM, running Microsoft Windows XP and Microsoft .NET
@@
 To measure how the JIT handles object-oriented features, we wrote a very
-simple benchmark that involves attribute lookups and polymorphic method calls:
+simple benchmark that involves attribute lookups and polymorphic method calls.
+Since the TLC assembler source is long and hard to read,
+Figure~\ref{fig:accumulator} shows the equivalent program written in an
+invented Python-like syntax.
 
+\begin{figure}[h]
+\begin{center}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{The \emph{accumulator} example, written in a invented Python-like syntax}
+\label{fig:accumulator}
+\end{center}
+\end{figure}
 

### 2008-12-20 talk/ecoop2009/clibackend.tex
small tweaks in the flexswitch section, kill a section that we decided not to write

@@
 
-For instance, the code\footnote{For simplicity we write C\# code instead of
-the actual IL bytecode.} generated for the dispatcher of method \texttt{0x0002}
+The code\footnote{For simplicity we write C\# code instead of
+the actual IL bytecode.} generated for the dispatcher of methods
 is similar to the following fragment: 
@@
 Otherwise, the \lstinline{jump_to_ext}
-part of the dispatcher has to be executed.
-The code that actually jumps to an external block is contained in
-the dispatcher of the primary method, whereas the
-\lstinline{jump_to_ext} code of dispatchers of secondary methods
-simply delegates the dispatcher of the primary method of the same
-graph (see later).
+part of the dispatcher has to be executed, which is implemented differently 
+by primary and secondary methods.
 
@@
 
-  public void add_case(int value, FlexSwitchCase c)
-  {
+  public void add_case(int value, FlexSwitchCase c) {
     ...
@@
 
-  public uint execute(int value, InputArgs args)
-  {
+  public uint execute(int value, InputArgs args) {
     for(int i=0; i<numcases; i++)
@@
 
-\subsection{Alternative implementations}
-\dacom{need to be discussed with Antonio}
-
-\anto{I propose to wait until the rest of paper is more or less in a final
-  state, and see how much space is left}
-
-\commentout{
-Before implementing the solution described here, we carefully studied a lot of possible alternatives, but all of them either didn't work because of a limitation of the virtual machine or they could work but with terrible performances.
-
-In particular, in theory it is possible to implement external links using tail calls, by putting each block in its own method and doing a tail call instead of a jump; this would also solve the problem of how to pass arguments, as each method could have its own signature matching the input args of the block.
-I would like to explain this solution in a more detailed way as I think it's really elegant and nice.
-
-In theory, if the .NET JIT were smart enough it could inline and optimize away the tail calls (or at least many of those) and give us very efficient code.
-However, one benchmark I wrote shows that tail calls are up to 10 times slower (!!!) than normal calls, thus making impractical to use them for our purposes.
-}
-
 % LocalWords:  flexswitches backend flexswitch methodid blockid xFFFF blocknum

### 2008-12-20 talk/ecoop2009/benchmarks.tex
refactor tables

@@
 
-\anto{these tables are ugly}
+\begin{table}[ht]
+  \begin{center}
 
-\cfbolz{can we make one figure of all the tables?}
-\begin{table}[ht]
-  \begin{tabular}{|l|r|r|r|r||r|r|}
+  \begin{tabular}{l|rrrrrr}
+    \multicolumn{5}{c}{\textbf{Factorial}} \\ [0.5ex]
+
+    \textbf{$n$}          & $10$  & $10^7$           & $10^8$         & $10^9$         \\
     \hline
-    \textbf{n} & 
-    \textbf{Interp} &
-    \textbf{JIT} &
-    \textbf{JIT 2} &
-    \textbf{C\#} &
-    \textbf{Interp/JIT 2} &
-    \textbf{JIT 2/C\#} \\
+    \textbf{Interp}       & 0.031 & 30.984           & N/A            & N/A            \\
+    \textbf{JIT}          & 0.422 &  0.453           & 0.859          & 4.844          \\
+    \textbf{JIT 2}        & 0.000 &  0.047           & 0.453          & 4.641          \\
+    \textbf{C\#}          & 0.000 &  0.031           & 0.359          & 3.438          \\
+    \textbf{Interp/JIT 2} & N/A   & \textbf{661.000} & N/A            & N/A            \\
+    \textbf{JIT 2/C\#}    & N/A   & \textbf{1.500}   & \textbf{1.261} & \textbf{1.350} \\ [3ex]
+
+    \multicolumn{5}{c}{\textbf{Fibonacci}} \\ [0.5ex]
+
+    \textbf{$n$}          & $10$  & $10^7$           & $10^8$         & $10^9$         \\
     \hline
+    \textbf{Interp}       & 0.031 & 29.359           & 0.000          & 0.000          \\
+    \textbf{JIT}          & 0.453 &  0.469           & 0.688          & 2.953          \\
+    \textbf{JIT 2}        & 0.000 &  0.016           & 0.250          & 2.500          \\ 
+    \textbf{C\#}          & 0.000 &  0.016           & 0.234          & 2.453          \\
+    \textbf{Interp/JIT 2} & N/A   & \textbf{1879.962}& N/A            & N/A            \\
+    \textbf{JIT 2/C\#}    & N/A   & \textbf{0.999}   & \textbf{1.067} & \textbf{1.019} \\
+  \end{tabular}
 
-    $10$    &   0.031  &  0.422  &  0.000  &  0.000  &      N/A  &    N/A \\
-    $10^7$  &  30.984  &  0.453  &  0.047  &  0.031  &  661.000  &  1.500 \\
-    $10^8$  &     N/A  &  0.859  &  0.453  &  0.359  &      N/A  &  1.261 \\
-    $10^9$  &     N/A  &  4.844  &  4.641  &  3.438  &      N/A  &  1.350 \\
-
-    \hline
-
-  \end{tabular}
-  \caption{Factorial benchmark}
-  \label{tab:factorial}
+  \end{center}
+  \caption{Factorial and Fibonacci benchmarks}
+  \label{tab:factorial-fibo}
 \end{table}
 
-\begin{table}[ht]
-  \begin{tabular}{|l|r|r|r|r||r|r|}
-    \hline
-    \textbf{n} & 
-    \textbf{Interp} &
-    \textbf{JIT} &
-    \textbf{JIT 2} &
-    \textbf{C\#} &
-    \textbf{Interp/JIT 2} &
-    \textbf{JIT 2/C\#} \\
-    \hline
-
-    $10$    &   0.031  &  0.453  &  0.000  &  0.000  &       N/A  &  N/A   \\
-    $10^7$  &  29.359  &  0.469  &  0.016  &  0.016  &  1879.962  &  0.999 \\
-    $10^8$  &     N/A  &  0.688  &  0.250  &  0.234  &       N/A  &  1.067 \\
-    $10^9$  &     N/A  &  2.953  &  2.500  &  2.453  &       N/A  &  1.019 \\
-
-    \hline
-
-  \end{tabular}
-  \caption{Fibonacci benchmark}
-  \label{tab:fibo}
-\end{table}
-
-Tables \ref{tab:factorial} and \ref{tab:fibo} show the seconds spent to calculate
+Table \ref{tab:factorial-fibo} shows the seconds spent to calculate
 the factorial and Fibonacci for various $n$.  As we can see, for small values
@@
 \begin{table}[ht]
-  \begin{tabular}{|l|r|r|r|r||r|r|}
+  \begin{center}
+
+  \begin{tabular}{l|rrrrrr}
+    \multicolumn{5}{c}{\textbf{Accumulator}} \\ [0.5ex]
+
+    \textbf{$n$}          & $10$  & $10^7$           & $10^8$         & $10^9$         \\
     \hline
-    \textbf{n} & 
-    \textbf{Interp} &
-    \textbf{JIT} &
-    \textbf{JIT 2} &
-    \textbf{C\#} &
-    \textbf{Interp/JIT 2} &
-    \textbf{JIT 2/C\#} \\
-    \hline
-
-    $10$    &   0.031  &  0.453  &  0.000  &  0.000  &      N/A  &  N/A   \\
-    $10^7$  &  43.063  &  0.516  &  0.047  &  0.063  &  918.765  &  0.750 \\
-    $10^8$  &     N/A  &  0.875  &  0.453  &  0.563  &      N/A  &  0.806 \\
-    $10^9$  &     N/A  &  4.188  &  3.672  &  5.953  &      N/A  &  0.617 \\
-
-    \hline
+    \textbf{Interp}       & 0.031 & 43.063           & N/A            & N/A            \\
+    \textbf{JIT}          & 0.453 &  0.516           & 0.875          & 4.188          \\
+    \textbf{JIT 2}        & 0.000 &  0.047           & 0.453          & 3.672          \\
+    \textbf{C\#}          & 0.000 &  0.063           & 0.563          & 5.953          \\
+    \textbf{Interp/JIT 2} & N/A   & \textbf{918.765} & N/A            & N/A            \\
+    \textbf{JIT 2/C\#}    & N/A   & \textbf{0.750}   & \textbf{0.806} & \textbf{0.617} \\
 
   \end{tabular}
+  \end{center}
   \caption{Accumulator benchmark}
@@
 other two benchmarks.  However, the really interesting part is the comparison
-with the equivalent C\# code, as the code generated by the JIT is
+with the equivalent C\# code, as the code generated by the JIT is up to 1.62 times
 \textbf{faster}.

### 2008-12-20 talk/ecoop2009/intro.tex
several small fixes.  Some rewording in the cli backend section
add reference to dyla paper, add institute for armin

@@
 
-\cfbolz{we should cite the dyla paper somewhere here. also, we generally need
-more references}
-
-PyPy's approach is to automatize the generation of JIT compilers in order
+PyPy's approach \cite{RiBo07_223} is to automatize the generation of JIT compilers in order
 to reduce to a minimum the effort required to get a fast implementation of a
@@
 \subsection{PyPy and RPython}
-
-\anto{as Armin points out, the two CLI backends can be easily confused; what
-  about renaming the ``CLI Backend for flowgraphs'' into ``CLI bytecode
-  compiler''? Any better idea for the name?}
 

### 2008-12-20 talk/ecoop2009/intro.tex
kill a picture, as it was only loosely referenced by the text

@@
 
+\commentout{
 \begin{figure}[h]
@@
 \end{figure}
+}
 
@@
 as Smalltalk \cite{BolzEtAl08}, JavaScript, Scheme and Prolog.
+
+\commentout{
 As suggested by Figure~\ref{pypy-fig}, a portable interpreter for a
@@
 RPython.
+}
 

### 2008-12-20 talk/ecoop2009/conclusion.tex
put two code snippets side by side, and fix a typo
try to write the related work section in proper english

@@
-\section{Related Word}
+\section{Related Work}
+
 Promotion is a concept that we have already explored in other contexts. Psyco is
@@
 
- - reference standard jit stuff, PICs \cite{hoelzle_optimizing_1991}, this one: http://www.cs.ucsb.edu/~urs/oocsb/self/papers/type-feedback.html
+Moreover, the idea of promotion of is a generalization of \emph{Polymorphic
+  Inline Caches} \cite{hoelzle_optimizing_1991}, as well as the idea of using
+runtime feedback to produce more efficient code
+\cite{hoelzle_type_feedback_1994}.
 
- - compare against tracing JITs \cite{gal_hotpathvm_2006}. tracing jits make
- implementing jits much easier, maybe less need for JIT generation. also, they
- concentrate on loops, which makes them produce a lot less code (which will need
- to be addressed in our work as well).
+PyPy-style JIT compilers are hard to write manually, thus we chose to write a
+JIT generator.  Tracing JIT compilers \cite{gal_hotpathvm_2006} also gives
+good results but are much easier to write, making the need for an automatic
+generator less urgent.  However so far tracing JITs have less general
+allocation removal techniques, which makes them get less speedup in a dynamic
+language with boxing.  Another difference is that tracing JITs concentrate on
+loops, which makes them produce a lot less code.  This issue will be addressed
+by future research in PyPy.
 
- so far tracing JITs have less general allocation removal techniques, which
- makes them get less speedup in a dynamic language with boxing
-
-in my master I had this snippet:
-
-"""
-The generated code typically contains guards; in recent research
+The code generated by tracing JITs code typically contains guards; in recent research
 \cite{gal_incremental_2006} on Java, these guards' behaviour is extended to be
@@
 http://www.mozilla.org/projects/tamarin/}} and in \cite{chang_efficient_2007}.
-"""
 
@@
 
- - apart from that we maybe should talk about escape analysis
- \cite{Blanchet99escapeanalysis}, \cite{Choi99escapeanalysis}
- our algorithm is totally simple-minded in comparison, but is useful in practise
+Our algorithm to avoid allocation of unneeded intermediate objects fits into
+the research area of escape analysis: in comparison to advanced techniques
+\cite{Blanchet99escapeanalysis}, \cite{Choi99escapeanalysis} our algorithm is
+totally simple-minded, but it is still useful in practise.
 

### 2008-12-20 talk/ecoop2009/conclusion.tex
add high-level structure of conclusions

@@
 
-XXX add me
+high level structure:
+
+1. We presented PyPy JIT generator
+
+2. interpreters can exploit the JIT generator by placing few hints here and
+there to guide the generation of code
+
+3. generated JITs produce fast code thanks to \textbf{promotion} (our first
+contribution) and automatic unboxing
+
+4. we presented a way to implement promotion/flexswitches on top of .NET
+
+5. our second contribution is to show that the idea of JIT layering can work
+well, as for some cases we can get performances even better than C\#

### 2008-12-20 talk/ecoop2009/intro.tex
expand a bit the idea of jit layering in the introduction

@@
 
-The contributions of this paper are \emph{promotion}, a generalization of
-polymorphic inline caches that make partial evaluation practical for dynamic
-languages and trying out the idea of JIT-layering XXX.
+Instead of emitting code for a real CPU, the generated JIT compilers described
+in this paper emit code for a virtual machine: before being executed, the
+emitted code will be compiled again by .NET's own JIT compiler, thus having
+two \emph{layers} of JIT compilation.
+
+The contributions of this paper are twofold: (1) \emph{promotion} is a
+generalization of polymorphic inline caches that make partial evaluation
+practical for dynamic languages; (2) we demonstrated that the idea of
+\emph{JIT layering} can give good results, as dynamic languages can be even
+faster than their static counterparts.
 

### 2008-12-20 talk/ecoop2009/intro.tex
fix few XXXs

@@
 
-Most JIT compilers for dynamic languages around (such as IronPython, Jython,
-JRuby \anto{XXX insert some reference}) compile code at the method
-granularity.  If on the one hand they can exploit some of the knowledge gathered
-at runtime (e.g. the types of method parameters), on the other hand they can
-do little to optimize most of the operations inside, because their behaviour
-depends on informations that are not available at compile-time, because
-e.g. the global state of the program can change at runtime. \anto{e.g., we can
-  add/remove methods to classes, etc. Should we insert some example here?}
-\cfbolz{the last sentence is waaay too complicated}
+Most JIT compilers for dynamic languages around (such as
+IronPython \footnote{http://www.codeplex.com/IronPython},
+Jython \footnote{http://www.jython.org/} or
+JRuby \footnote{http://jruby.codehaus.org/}) compile code at the method
+granularity.  If on the one hand they can exploit some of the knowledge
+gathered at runtime (e.g. the types of method parameters), on the other hand
+they can do little to optimize most of the operations inside, because few
+assumptions can be made about the global state of the program.
 

### 2008-12-20 talk/ecoop2009/jitgen.tex
make sources a bit smaller, and comment out a big image which is not referenced by the text
fix few XXXs

@@
 
-XXX write something about the problems of classical PE?
-
 \subsection{Binding Time Analysis in PyPy}
 
-XXX need better transition text from last section
 At translation time, PyPy performs binding-time analysis of the source
-RPython program after it has been turned to low-level graphs, i.e. at
-the level at which operations manipulate objects that have a fixed type
-attached.
-
-The binding-time terminology that we are using in PyPy is based on the
+RPython program.  The binding-time terminology that we are using in PyPy is based on the
 colors that we use when displaying the control flow graphs:
@@
 This is called the \emph{hint-annotator}; it
-operates over input graphs that are already low-level instead of
-RPython-level, \anto{we never make distinction between low-level and rpython-level 
-flowgraphs, do we? I propose to talk about ``intermediate flowgraphs''} 
+operates over input flowgraphs
 and propagates annotations that do not track types but

### 2008-12-20 talk/ecoop2009/benchmarks.tex
fix another XXX

@@
 \label{sec:benchmarks}
-
-\anto{We should say somewhere that flexswitches are slow but benchmarks are so
-  good because they are not involved in the inner loops}
 
@@
 
-\commentout{
-\cfbolz{I think we can kill this for space reasons}
-\anto{If we decide to keep them, we must remember to explain the python-like
-  syntax, as it is no longer in tlc.tex}
-
-\begin{lstlisting}
-def main(n):
-    result = 1
-    while n > 1:
-        result = result * n
-        n = n - 1
-    return n
-\end{lstlisting}
-}
-
 When doing plain interpretation, we need to create and destroy three temporary
@@
 which we can do the same reasoning as above.
-
-\commentout{
-\cfbolz{I think we can kill this for space reasons}
-\begin{lstlisting}
-def main(n):
-    a = 0
-    b = 1
-    while n > 1:
-        sum = a + b
-        a = b
-        b = sum
-        n = n - 1
-    return b
-\end{lstlisting}
-}
 
@@
 compiler is highly optimized to handle bytecode generated by C\# compilers.
+
+As we saw in Section~\ref{sec:flexswitches-cli}, the implementation of
+flexswitches on top of CLI is hard and inefficient.  However, our benchmarks
+show that this inefficiency does not affect the overall performances of the
+generated code.  This is because in most programs the vast majority of the
+time is spent in the inner loop: the graphs are built in such a way that all
+the blocks that are part of the inner loop reside in the same method, so that
+all links inside are internal (and fast).
 

### 2008-12-20 talk/ecoop2009/jitgen.tex
fix one of the last XXXs

@@
 
-\anto{maybe we should at least mention promotion, and refer to the proper
-  section for details?}
+There are cases in which having a green variable is essential for generating
+good code, but it is not possible to use the \texttt{concrete} due to an
+unsatisfiable dependency: Section~\ref{sec:promotion} introduces
+\emph{promotion}, the novel technique that makes possible to solve this
+problem.

### 2008-12-20 talk/ecoop2009/benchmarks.tex
some minore changes

@@
 integers, so it can inline the code to compute the result.  Moreover, it can
-\emph{virtualize} (see section \ref{sec:virtuals} all the temporary objects, because they never escape from
+\emph{virtualize} (see Section \ref{sec:virtuals}) all the temporary objects, because they never escape from
 the inner loop.  The same remarks apply to the other two operations inside
@@
 
-However, our generated JIT does not compile the whole function at
+As already discussed, our generated JIT does not compile the whole function at
 once. Instead, it compiles and executes code chunk by chunk, waiting until it

### 2008-12-20 talk/ecoop2009/clibackend.tex
fix another XXX
some minore changes

@@
 compiler which works as described in the previous sections.  Internally, the
-frontend represents the compiled code as \emph{flow graphs}, and the role of
+JIT represents the compiled code as \emph{flow graphs}, and the role of
 the backends is to translate flowgraphs into machine code.
@@
 .NET one in some contexts, as our benchmarks demonstrate (see
-section~\ref{sec:benchmarks}).  On the other hand, the lower-level .NET JIT is
+Section~\ref{sec:benchmarks}).  On the other hand, the lower-level .NET JIT is
 very good at producing machine code, much more than PyPy's own \emph{x86}
@@
 \subsection{Implementing flexswitches in CLI}
+\label{sec:flexswitches-cli}
 

### 2008-12-20 talk/ecoop2009/tlc.tex
comment out a possible confusing example, and add an XXX
some minore changes

@@
 attribute from an object which does not provide it.  Being dynamically typed,
-the VM needs to do all these checks at runtime; in case one of the check
+the interpreter needs to do all these checks at runtime; in case one of the check
 fails, the execution is simply aborted.
@@
 
-\subsection{TLC examples}
+\subsection{TLC example}
 
@@
 shows a simple program that computes the absolute value of
-the given integer.
+the given integer.  In the subsequent sections, we will examine step-by-step 
+how the generated JIT compiler manages to produce a fully optimized version of it.
 
@@
 
+\commentout{
+
+\anto{I think this example could cause confusion, because the synatx is TOO
+  similar to (R)Python, hence in the next section it could be unclear which is
+  the leves of the examples}
+
 Since reading TLC programs at bytecode level is hard, in this paper we will
@@
 \end{lstlisting}
+}

### 2008-12-21 talk/ecoop2009/conclusion.tex
some minore changes
add conclusion and future work

--- 2008-12-20 talk/ecoop2009/conclusion.tex
@@
 
-\section{Conclusion}
+\section{Conclusion and Future Work}
 
-XXX add me
+In this paper we presented PyPy's JIT compiler generator, based on partial
+evaluation techniques, which can automatically turn an interpreter into a JIT
+compiler, requiring the language developers to only add few \texttt{hint}s to
+guide the generation process.
+
+We showed that classical partial evaluation cannot remove all the overhead
+proper of dynamically typed languages, and how the new operation called
+\emph{promotion} solves the problem, by delaying compile-time until the JIT
+knows enough to produce efficient code, and by continuously intermixing
+compile-time and runtime.  Moreover, we showed that our simple but still
+practically useful technique to avoid allocation of intermediate unnecessary
+objects plays well with promotion and helps to produce even better code.
+
+Finally, we presented the CLI backend for PyPy's JIT compiler generator, whose
+goal is to produce .NET bytecode at runtime.  We showed how it is possible to
+circumvent intrinsic limitations of the virtual machine to implement
+promotion.  As a result, we proved that the idea of \emph{JIT layering} is
+worth of further exploration, as it makes possible for dynamically typed
+languages to be even faster than their statically typed counterpart in some
+circumstances.
+
+As a future work, we want to explore different strategies to make the frontend
+producing less code, maintaining comparable or better performances.  In
+particular, we are working on a way to automatically detect loops in the user
+code, as tracing JITs do \cite{gal_hotpathvm_2006}.  By compilining whole
+loops at once, the backends should be able to produce better code than today.
+
+At the moment, some bugs and minor missing features prevent the CLI JIT
+backend to handle more complex languages such as Python and Smalltalk.  We are
+confident that once these problems will be fixed, we will get performance
+results comparable to TLC, as the other backends already demonstrate
+\cite{PyPyJIT}.  Moreover, if the current implementation of flexswitches will
+prove to be too slow for some purposes, we want to explore alternative
+implementation strategies, also considering the new features that might be
+integrated into virtual machines.

### 2008-12-21 talk/ecoop2009/clibackend.tex
add conclusion and future work
add a footnote

@@
 \emph{CLI Virtual Machine}.  The latter is special because instead of emitting
-code for a real CPU it emits code for a virtual machine: before being
+code for a real CPU it emits code for a virtual machine \footnote{By using the 
+\lstinline{Reflection.Emit} namespace and creating \lstinline{DynamicMethod}s.}: 
+before being
 executed, the generated code will be compiled again by .NET's own JIT

### 2008-12-21 talk/ecoop2009/intro.tex
fix xxx

@@
 Most JIT compilers for dynamic languages around (such as
-IronPython \footnote{http://www.codeplex.com/IronPython},
-Jython \footnote{http://www.jython.org/} or
-JRuby \footnote{http://jruby.codehaus.org/}) compile code at the method
+IronPython\footnote{http://www.codeplex.com/IronPython},
+Jython\footnote{http://www.jython.org/} or
+JRuby\footnote{http://jruby.codehaus.org/}) compile code at the method
 granularity.  If on the one hand they can exploit some of the knowledge

### 2008-12-21 talk/ecoop2009/jitgen.tex
some minore changes
fix xxx

@@
 
-Traditional JIT compilers are hard to write, time consuming, hard to evolve,
-etc. etc.
+Traditional JIT compilers are hard to write, time consuming and hard to
+evolve.  On the other hand, interpreters are easy both to write and maintain,
+but they are usually slow.  By automatically generating a JIT compiler out of
+an interpreter, we take the best of both worlds.
 

### 2008-12-21 talk/ecoop2009/clibackend.tex
remove space before of footnote
typo

@@
 \emph{CLI Virtual Machine}.  The latter is special because instead of emitting
-code for a real CPU it emits code for a virtual machine \footnote{By using the 
+code for a real CPU it emits code for a virtual machine\footnote{By using the 
 \lstinline{Reflection.Emit} namespace and creating \lstinline{DynamicMethod}s.}: 
@@
 mostly focused making the JIT able to produce very fast code, without trying
-to optimzed the compilation phase itself.
+to optimze the compilation phase itself.
 

### 2008-12-21 talk/ecoop2009/benchmarks.tex
typo
minor modifications

@@
 In section \ref{sec:tlc-properties}, we saw that TLC provides most of the
-features that usaully make dynamically typed language so slow, such as
+features that usually make dynamically typed language so slow, such as
 \emph{stack-based interpreter}, \emph{boxed arithmetic} and \emph{dynamic lookup} of
@@
 
-In the following sections, we will show some benchmarks that show how our
+In the following sections, we present some benchmarks that show how our
 generated JIT can handle all these features very well.
@@
 Moreover, for each benchmark we also show the time taken by running the
-equivalent program written in C\# \footnote{The sources for both TLC and C\#
+equivalent program written in C\#\footnote{The sources for both TLC and C\#
   programs are available at:
@@
 
-As we can see, the code generated by the JIT can be up to ~1800 times faster
+As we can see, the code generated by the JIT can be up to about 1800 times faster
 than the non-jitted case.  Moreover, it often runs at the same speed as the
@@
 function, it is \emph{virtualized} and its field \lstinline{value} is stored
-as a local variable.  As a result, the generated code results in a simple loop
+as a local variable.  As a result, the generated code turns out to be a simple loop
 doing additions in-place.
@@
 
-\anto{maybe we should move the following paragraph to
-  abstract/introduction/conclusion?}
-
 Despite being only a microbenchmark, this result is very important as it proves
@@
 performances than current techniques.  The result is even more impressive if
-we consider dynamically typed languages as TLC are usually considered much
+we take in account that dynamically typed languages as TLC are usually considered much
 slower than the statically typed ones.

### 2008-12-21 talk/ecoop2009/abstract.tex
change a bit the abstract

@@
 \begin{abstract}
-Dynamic languages are increasingly popular etc. etc.
+Implementing dynamically typed languages in an efficient way is a hard task.
+Recent developments show that JIT compilers 
+can exploit runtime type information to generate quite efficient code.
+Unfortunately, writing a JIT compiler is far from being simple.
+
+In this paper we report our positive experience with automatic generation
+of JIT compilers as supported by the PyPy infrastructure, by
+focusing  on JIT compilation for .NET.
+The paper presents two main and novel contributions: we show that
+partial evaluation can be used in practice for generating a JIT compiler,
+and we experiment with the potentiality offered by the ability to
+add a further level of JIT compilation on top of .NET.
+
+The practicality of the approach is demonstrated by showing some
+promising experiments done with benchmarks written in a simple dynamic language.
 \end{abstract}
+
+% LocalWords:  JIT PyPy

### 2008-12-21 talk/ecoop2009/jitgen.tex
kill bits about online/offline PE

@@
 Effective dynamic compilation requires feedback of runtime information into
-compile-time. This is no different for partial evaluation.
-
-Partial evaluation (PE) comes in two flavors: \cfbolz{if we have space problems
-we should kill the bits about online and offline PE}
-
-\begin{itemize}
-\item \emph{On-line} partial evaluation: a compiler-like algorithm takes the
-source code of the function \texttt{f(x, y)} (or its intermediate representation,
-i.e. its control flow graph), and some partial
-information, e.g. \texttt{x=5}.  From this, it produces the residual function
-\texttt{g(y)} directly, by following in which operations the knowledge \texttt{x=5} can
-be used, which loops can be unrolled, etc.
-
-\item \emph{Off-line} partial evalution: in many cases, the goal of partial
-evaluation is to improve performance in a specific application.  Assume that we
-have a single known function \texttt{f(x, y)} in which we think that the value
-of \texttt{x} will change slowly during the execution of our program – much
-more slowly than the value of \texttt{y}.  An obvious example is a loop that
-calls \texttt{f(x, y)} many times with always the same value \texttt{x}.  We
-could then use an on-line partial evaluator to produce a \texttt{g(y)} for each
-new value of \texttt{x}.  In practice, the overhead of the partial evaluator
-might be too large for it to be executed at run-time.  However, if we know the
-function \texttt{f} in advance, and if we know \emph{which} arguments are the
-ones that we will want to partially evaluate \texttt{f} with, then we do not
-need a full compiler-like analysis of \texttt{f} every time the value of
-\texttt{x} changes.  We can precompute once and for all a specialized function
-\texttt{f1(x)}, which when called produces the residual function \texttt{g(y)}
-corresponding to \texttt{x}.  This is \emph{off-line partial evaluation}; the
-specialized function \texttt{f1(x)} is called a \emph{generating extension}.
-\end{itemize}
-
-Off-line partial evaluation is usually based on \emph{binding-time analysis}, which
-is the process of determining among the variables used in a function (or
-a set of functions) which ones are going to be known in advance and
-which ones are not.  In the example of \texttt{f(x, y)}, such an analysis
-would be able to infer that the constantness of the argument \texttt{x}
-implies the constantness of many intermediate values used in the
-function.  The \emph{binding time} of a variable determines how early the
-value of the variable will be known.
+compile-time. This is no different for partial evaluation, and subsequent sections shows how we managed to solve this problem.
 
@@
 
-At translation time, PyPy performs binding-time analysis of the source
-RPython program.  The binding-time terminology that we are using in PyPy is based on the
-colors that we use when displaying the control flow graphs:
+At translation time, PyPy performs binding-time analysis of the source RPython
+program, to determine which variables are static and which dynamic.  The
+binding-time terminology that we are using in PyPy is based on the colors that
+we use when displaying the control flow graphs:
 

### 2008-12-21 talk/ecoop2009/jitgen.tex
remove reference to year, as the cited paper is more recent

@@
 
-In 1971 Yoshihiko Futamura published a paper \cite{Futamura99} that proposed a
+Yoshihiko Futamura published a paper \cite{Futamura99} that proposed a
 technique to automatically transform an interpreter of a programming language

### 2008-12-21 talk/ecoop2009/intro.tex
first submission

@@
 \section{Introduction}
-
-Dynamic languages are increasingly popular etc. etc.
-
 The easiest way to implement a dynamic language such as Python is to write an
@@
 PyPy's approach \cite{RiBo07_223} is to automatize the generation of JIT compilers in order
-to reduce to a minimum the effort required to get a fast implementation of a
-dynamic language; all you have to do is to write a high level specification of
-the language (by writing an interpreter), and putting it through PyPy's
-translation toolchain. The automatic generation of JIT compilers is done with
-the help of partial evaluation techniques.
+to minimize the effort required to get a fast implementation of a
+dynamic language; automatic generation of JIT compilers is done with
+the help of partial evaluation techniques and requires the user only
+to provide an interpreter with some manual annotations which hint
+the generator how interpretation and JIT compilation has to be interleaved.
 
-Instead of emitting code for a real CPU, the generated JIT compilers described
-in this paper emit code for a virtual machine: before being executed, the
-emitted code will be compiled again by .NET's own JIT compiler, thus having
-two \emph{layers} of JIT compilation.
+More precisely, in this paper we focus on the ability of generating a JIT compiler able to emit code
+for the .NET virtual machine. To our knowledge, this is the first experiment with an interpreter with
+two \emph{layers} of JIT compilation, since, before being executed, the
+emitted code is eventually compiled again by .NET's own JIT compiler.
 

### 2008-12-21 talk/ecoop2009/benchmarks.tex
changed sentence which Antonio does not like

@@
 of $n$ the time spent running the JIT compiler is much higher than the time
-spent to simply interpret the program.  This is an expected result, as till
-now we only focused on optimizing the compiled code, not the compilation
-process itself.
+spent to simply interpret the program.  This is an expected result
+which, however, can be improved once we will have time
+to optimize compilation and not only the generated code.
 
-On the other hand, to get meaningful timings we had to use very high values of
-$n$.  This means that the results are incorrect due to overflow, but since all
-the runnings overflow in the very same way, the timings are still
-comparable. \anto{I think we should rephrase this sentence}.  For $n$ greater
-than $10^7$, we did not run the interpreted program as it would have took too
+On the other, for reasonably high values of $n$ we obtain very good
+results, which are valid despite the obvious overflow, since the 
+same operations are performed for all experiments.
+For $n$ greater than $10^7$, we did not run the interpreted program as it would have took too
 much time, without adding anything to the discussion.

### 2009-01-19 talk/ecoop2009/tutorial/proposal.tex
(cfbolz, arigo, antocuni): start tutorial propsal

\documentclass{llncs}

 %%%% Compression Light+: LNCS margin reduced by +/-7mm along all edges (RG).
%\textwidth=130mm   % LNCS: 122mm
%\textheight=203mm  % LNCS: 193mm

\renewcommand{\baselinestretch}{0.97}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[sans]{dsfont}
\usepackage{color}
\usepackage{ifthen}
\usepackage{xspace}
\usepackage{listings}
\usepackage{fancyvrb}
\usepackage{multirow}
\usepackage[pdftex]{graphicx}

%\input{macros}

\pagestyle{plain}

%\lstset{mathescape=true,language=Java,basicstyle=\tt,keywordstyle=\bf}
\lstset{language=Python,
        basicstyle=\scriptsize\ttfamily,
        keywordstyle=\color{blue}, % I couldn't find a way to make chars both bold and tt
        frame=none,
        stringstyle=\color{blue},
        fancyvrb=true,
        xleftmargin=20pt,xrightmargin=20pt,
        showstringspaces=false}

\setlength{\tabcolsep}{1ex}

%\renewcommand{\baselinestretch}{.98}
\newboolean{showcomments}
\setboolean{showcomments}{false}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\dacom[1]{\nb{DA}{#1}}
\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand{\commentout}[1]{}

\let\oldcite=\cite

\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}

\begin{document}
\title{Tutorial Proposal: Writing Interpreters for Dynamic Languages Using PyPy
(and Getting Them Fast)}

\author{Carl Friedrich Bolz\inst{2} \email{cfbolz@gmx.de} \and Antonio Cuni\inst{1} \email{anto.cuni@gmail.com} \and Armin Rigo\inst{2} \email{arigo@tunes.org}}

\institute{DISI, University of Genova, Italy 
\and 
Softwaretechnik und Programmiersprachen
 Heinrich-Heine-Universit\"at D\"usseldorf}

\maketitle

The easiest way to implement a dynamic language such as JavaScript or Python is
to write an interpreter for it; however, interpreters are slow. An alternative
is to write a compiler; writing a compiler that targets a high level virtual
machine like CLI or JVM is easier than targeting a real CPU, but it still
require a lot of work, as IronPython, Jython, JRuby demonstrate. Moreover,
the various platforms have to be targeted independently.

The PyPy project \cite{RiBo07_223} aims to make the implementation of dynamic
languages easier by providing a toolchain that allows to translate an
interpreter to various target platforms, including the JVM, .NET and C/Posix. 
In addition, we are working on being able to automatically transform the
interpreter into a specializing JIT-compiler to vastly increase performance.
The goal is to minimize the effort required to get a fast implementation of a
dynamic language.

Outline of the tutorial:

 - introduction to PyPy
 - present the small language and its interpreter
 - show how to compile the interpreter to various platforms
 - show how to apply JIT-generator

\bibliographystyle{plain}
\bibliography{main}

\end{document}

### 2009-01-19 talk/ecoop2009/tutorial/proposal.tex
(cfbolz, antocuni, arigo): expanded the proposal

@@
 
-The PyPy project \cite{RiBo07_223} aims to make the implementation of dynamic
+The PyPy project\footnote{\url{http://codespeak.net/pypy}} \cite{RiBo07_223}
+aims to make the implementation of dynamic
 languages easier by providing a toolchain that allows to translate an
@@
 
+This tutorial will give an insight into the usage of the PyPy translation
+toolchain by giving a step-by-step introduction. The topics covered are how to
+write an interpreter for a dynamic language for PyPy and make it possible to
+automatically generate a JIT for it.  We will showcase a typical interpreter for a
+small dynamic language as a running example.
+
 Outline of the tutorial:
+\begin{itemize}
+ \item \textbf{Introduction to PyPy:} Introduce the PyPy project and give an
+ overview of its goals.
+ \item \textbf{Overview of the example language:} Present the small language and
+ its interpreter that will be used as a running example throughout the tutorial.
+ \item \textbf{Translation:} Show how to compile the interpreter to various
+ platforms, including .NET and C/Posix.
+ \item \textbf{JIT-Generation:} Show how to apply the JIT-generator to the
+ interpreter to automatically get a JIT for the language.
+\end{itemize}
 
- - introduction to PyPy
- - present the small language and its interpreter
- - show how to compile the interpreter to various platforms
- - show how to apply JIT-generator
+There is a current resurgence in research on JIT-compilers for dynamic
+languages, mostly focusing on JavaScript implementations. PyPy generates such
+JITs so that they don't have to be hand-written. This requires much less effort
+for the language implementor.
+
+The presenters are part of the PyPy core development team and are all involved
+in the ongoing research on the JIT generator.
 
 \bibliographystyle{plain}
-\bibliography{main}
+\bibliography{proposal}
 

### 2009-01-19 talk/ecoop2009/tutorial/proposal.tex
(antocuni, arigo, cfbolz): write an exciting abstract.

@@
  Heinrich-Heine-Universit\"at D\"usseldorf}
+\maketitle
 
-\maketitle
+\begin{abstract}
+We propose to give a tutorial on how to use the PyPy project to write
+interpreters for dynamic languages. PyPy supports compiling such interpreters
+to various target environments (like .NET or C/Posix) and automatically
+generating a Just-In-Time compiler for them. PyPy is thus an emerging tool to
+help implement fast, portable multi-platform interpreters with less effort.
+\end{abstract}
 
@@
 machine like CLI or JVM is easier than targeting a real CPU, but it still
-require a lot of work, as IronPython, Jython, JRuby demonstrate. Moreover,
-the various platforms have to be targeted independently.
+require a lot of work, as demonstrated by the IronPython, Jython, JRuby
+projects. Moreover, the various platforms (either VM or real hardware) have to
+be targeted independently.
 
-The PyPy project\footnote{\url{http://codespeak.net/pypy}} \cite{RiBo07_223}
+The PyPy project\footnote{\url{http://codespeak.net/pypy}} \cite{RigoPedroni06}
+\cite{RiBo07_223}
 aims to make the implementation of dynamic
-languages easier by providing a toolchain that allows to translate an
+languages easier by providing a toolchain that allows to compile an
 interpreter to various target platforms, including the JVM, .NET and C/Posix. 
-In addition, we are working on being able to automatically transform the
+In addition, we are working automatically transforming the
 interpreter into a specializing JIT-compiler to vastly increase performance.
-The goal is to minimize the effort required to get a fast implementation of a
+The goal is to minimize the effort required to get a fast implementation for a
 dynamic language.
 
-This tutorial will give an insight into the usage of the PyPy translation
+This tutorial will provide insights into the usage of the PyPy translation
 toolchain by giving a step-by-step introduction. The topics covered are how to
-write an interpreter for a dynamic language for PyPy and make it possible to
+write an interpreter for a dynamic language with PyPy and make it possible to
 automatically generate a JIT for it.  We will showcase a typical interpreter for a

### 2009-01-19 talk/ecoop2009/intro.tex
(antocuni, cfbolz) minor fixes

@@
 level virtual machine like CLI or JVM is easier than targeting a real CPU, but
-it still require a lot of work, as IronPython, Jython, JRuby demonstrate.
+it still requires a lot of work, as IronPython, Jython, JRuby demonstrate.
 

### 2009-01-19 talk/ecoop2009/tutorial/proposal.tex
Typo.
(antocuni, cfbolz) minor fixes

@@
 machine like CLI or JVM is easier than targeting a real CPU, but it still
-require a lot of work, as demonstrated by the IronPython, Jython, JRuby
+requires a lot of work, as demonstrated by the IronPython, Jython, JRuby
 projects. Moreover, the various platforms (either VM or real hardware) have to
@@
 interpreter to various target platforms, including the JVM, .NET and C/Posix. 
-In addition, we are working automatically transforming the
+In addition, we are working on being able to automatically transform the
 interpreter into a specializing JIT-compiler to vastly increase performance.
@@
 automatically generate a JIT for it.  We will showcase a typical interpreter for a
-small dynamic language as a running example.
+small object-oriented dynamic language as a running example.
 
@@
  \item \textbf{JIT-Generation:} Show how to apply the JIT-generator to the
- interpreter to automatically get a JIT for the language.
+ interpreter by placing a few hints to automatically get a JIT for the language.
 \end{itemize}

### 2009-03-31 talk/icooolps2009/paper.tex
A draft for a paper about the tracing JIT that I will maybe submit to ICOOOLPS.

\documentclass{acm_proc_article-sp}

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}

\let\oldcite=\cite

\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}

\begin{document}

\title{Tracing the Meta-Level: PyPy's JIT Compiler}

\numberofauthors{2}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{Softwaretechnik und Programmiersprachen}\\
       \affaddr{Institut für Informatik}\\ 
       \affaddr{Universitätsstra{\ss}e 1}\\
       \affaddr{D-40225 Düsseldorf}\\
       \affaddr{Deutschland}\\
       \email{cfbolz@gmx.de}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\maketitle

%\category{D.3.4}{Programming Languages}{Processors}[code generation,
%interpreters, run-time environments]
%\category{F.3.2}{Logics and Meanings of Programs}{Semantics of Programming
%Languages}[program analysis]

\begin{abstract}
In this paper we describe the ongoing research in the PyPy project to write a
JIT compiler that is automatically adapted to various languages, given an
interpreter for that language. This is achieved with the help of a slightly
adapted tracing JIT compiler in combination with some hints by the author of the
interpreter.  XXX

\end{abstract}

XXX write somewhere that we will be predominantly occupied with bytecode
interpreters
XXX write somewhere that one problem of using tracing JITs for dynamic languages
is that dynamic languages have very complex bytecodes

\section{Introduction}

Dynamic languages, rise in popularity, bla bla XXX

One of the often-cited drawbacks of dynamic languages is the performance
penalties they impose. Typically they are slower than statically typed languages
\cite{XXX}. Even though there has been a lot of research into improving the
performance of dynamic languages \cite{XXX}, those techniques are not as widely
used as one would expect. Many dynamic language implementations use completely
straightforward bytecode-interpreters without any advanced implementation
techniques like just-in-time compilation. There are a number of reasons for
this. Most of them boil down to the inherent complexities of using compilation.
Interpreters are simple to understand and to implement whereas writing a
just-in-time compiler is an error-prone task that is even made harder by the
dynamic features of a language.

writing an interpreter has many advantages... XXX

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers. XXX

The PyPy project is trying to find approaches to generally ease the
implementation of dynamic languages. It started as a Python implementation in
Python, but has now extended its goals to be generally useful for implementing
other dynamic languages as well. The general approach is to implement an
interpreter for the language in a subset of Python. This subset is chosen in
such a way that programs in it can be compiled into various target environments,
such as C/Posix, the CLR or the JVM. The PyPy project is described in more
details in Section \ref{sect:pypy}.

In this paper we discuss ongoing work in the PyPy project to improve the
performance of interpreters written with the help of the PyPy toolchain. The
approach is that of a tracing JIT compiler. Opposed to the tracing JITs for dynamic
languages that exist so far, PyPy's tracing JIT operates "one level down",
e.g. traces the execution of the interpreter, as opposed to the execution
of the user program. The fact that the program the tracing JIT compiles is
in our case always an interpreter brings its own set of problems. We describe
tracing JITs and their application to interpreters in Section
\ref{sect:tracing}.  By this approach we hope to get a JIT compiler that can be
applied to a variety of dynamic languages, given an interpreter for them. The
process is not completely automatic but needs a small number of hints from the
interpreter author, to help the tracing JIT. The details of how the process
integrates into the rest of PyPy will be explained in Section
\ref{sect:implementation}. This work is not finished, but already produces some
promising results, which we will discuss in Section \ref{sect:evaluation}.

%- dynamic languages important
%- notoriously difficult to achieve good performance
%- even though the techniques exist since a while, not many implementations
%  actually use them
%    - hard to get all corner-cases right
%    - languages evolve
%    - modern dynamic languages are large
%    - open source/research communities don't have that many resources
%
%- PyPy project: trying find approaches to ease the implementation of dynamic
%languages
%- explore general ways to improve on the speed of dynamic languages with reduced
%effort
%- approach: write a tracing JIT that is applicable to many different languages,
%by tracing "one level done"
%- needs some hints by the interpreter-writer + slightly different optimizations
%- paper will describe the problems of applying a tracing jit to an interpreter
%- different integration needed than a typical tracing jit

\section{The PyPy Project}
\label{sect:pypy}

The PyPy project\footnote{http://codespeak.net/pypy} was started to implement a
new Python interpreter in Python but has now extended its goals to be an
environment where flexible implementation of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython. RPython ("Restricted Python") is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLR and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

The central idea of this way to implement VMs is that the interpreter
implementation in RPython should be as free as possible of low-level
implementation details, such as memory management strategy, threading model or
object layout. Instead, these details are inserted into the VM during the
translation process by the translation toolchain. This makes it possible to
change these details later, if that becomes necessary. This is something that is
hard to do with a traditional VM written in a low-level language such as C,
since the low-level details need to be fixed early in the development-process.
XXX is this paragraph really needed?

In the following we will describe some details of the translation process, since
they are needed in Section \ref{sect:implementation}. The first step is to produce
control flow graphs of all functions of the RPython program. Afterwards, type
inference is performed to gain type information about all the variables in the
flow graphs. Afterwards, the abstraction level of the operations in the graphs
is lowered in a stepwise fashion. At the end of this process, all operations in
the graphs correspond rather directly to a simple operation in C. The variables
are annotated with a C-like type system containing primitive types (like
\texttt{Signed}, \texttt{Bool}, etc.) or pointer types pointing to Structs,
Arrays or Functions.

XXX example. reuse the one of the tracing jit?

The translation process usually just turns these graphs into C code so that they
can be compiled into an executable. However, they can also be interpreted in
various ways. This is useful for testing and debugging the translation toolchain
because the produced error messages in case of a crash are a lot more helpful
than what would be produced after compilation to C. These low-level graphs are
also what the tracing JIT takes as input, as we will see later.

%- original goal: Python interpreter in Python
%- general way to write flexible VMs for dynamic languages
%- interpreter written in RPython, subset of Python to allow type inference
%- translation toolchain
%   - naive forward-propagation type inference
%   - lowering of abstractions
%   - lltype system, monomorphic C-level operations
%   - type system: primitives, pointers to structs and arrays
%   - still assumes presence of GC
%   - can be interpreted in various ways

\section{Tracing JIT Compilers}
\label{sect:tracing}

Tracing JITs are an idea explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{XXX}. The technique is now used by both and are now being used by both Mozilla's
TraceMonkey JavaScript VM \cite{XXX} and Adobe's Tamarin ActionScript VM
\cite{XXX}.

Tracing JITs are built on the following basic assumptions:

\begin{itemize}
 \item programs spend most of their runtime in loops
 \item several iterations of the same loop are likely to take similar code paths
\end{itemize}

The basic approach of a tracing JIT is to only generate machine code for the hot
code paths of commonly executed loops and to interpret the rest of the program.
The code for those common loops however should be highly optimized, including
aggressive inlining.

The generation of loops works as follows: At first, everything is interpreted.
The interpreter does a bit of lightweight profiling to figure out which loops
are run often. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
was executed. Since loops need a backward jump somewhere, this method finds
loops in the user program.

When a common loop is identified, the interpreter enters a
special mode (called tracing mode). When in tracing mode, the interpreter
records a history (the \emph{trace}) of all the operations it executes, in addition
to actually performing the operations. During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that is being interpreted at the moment anyway.

This process assumes that the path through the loop that was traced is a
"typical" example of possible paths (which is statistically likely). Of course
it is possible that later another path through the loop is taken, therefore the
machine code will contain \emph{guards}, which check that the path is still the same.
If a guard fails during execution of the machine code, the machine code is left
and execution falls back to using interpretation (there are more complex
mechanisms in place to still produce more code for the cases of guard failures,
but they are of no importance for this paper XXX is that true?).

It is important to understand when the tracer considers a loop in the trace to
be closed. This happens when the \emph{position key} is the same as at an earlier
point. The position key describes the position of the execution of the program,
e.g. usually contains things like the function currently being executed and the
program counter position of the tracing interpreter. The tracing interpreter
does not need to check all the time whether the position key already occurred
earlier, but only at instructions that are able to change the position key
to an earlier value, e.g. a backward branch instruction. Note that this is
already the second place where backward branches are treated specially: During
interpretation they are the place where the profiling is performed and where
tracing is started or already existing assembler code entered; during tracing
they are the place where the check for a closed loop is performed.

Let's look at a small example. Take the following (slightly contrived) RPython
code:
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}

At first those functions will be interpreted, but after a while, profiling shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often.  The
tracing JIT will then start trace the execution of that loop.  The trace would
look as follows:
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2) [result1]
jump(result1, n1)
\end{verbatim}

XXX add a note about the SSA-ness of the trace

This trace will then be turned into machine code. Note that the machine code
loop is by itself infinite and can only be left via a guard failure. Also note
\texttt{f} was inlined into the loop and how the common \texttt{else} case was
turned into machine code, while the other one is implemented via a guard
failure. The variables in square brackets after the guards are the state that
the interpreter will get when the guard fails.

%- general introduction to tracing
%- assumptions
%- mixed-mode execution environment: interpretation, tracing, compilation,
%  running native code
%- write why tracing jits are particularly well suited for dynamic languages

\subsection{Applying a Tracing JIT to an Interpreter}

XXX \cite{sullivan_dynamic_2003} somewhere

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and how to solve them (at least
partially). This means that there are two interpreters involved, and we need
terminology to distinguish them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that is running the
users programs, which we will call the \emph{language interpreter}. The program
that the language interpreter executes we will call the \emph{user program}
(from the point of view of a VM author, the "user" is a programmer using the
VM).

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is the bytecode dispatch loop. Usually it is also the only hot loop.
Tracing one iteration of this loop means that the execution of one bytecode was
seen. This means that the resulting machine code will correspond to a loop, that
assumes that this particular bytecode will be executed many times in a row,
which is clearly very unlikely.

\begin{figure}
\input{code/tlr-paper.py}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}

\begin{figure}
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

    # 4:
    MOV_R_A     0   # i--
    DECR_A
    MOV_A_R     0

    MOV_R_A     2   # res += a
    ADD_R_TO_A  1
    MOV_A_R     2

    MOV_R_A     0   # if i!=0: goto 4
    JUMP_IF_A   4

    MOV_R_A     2   # return res
    RETURN_A
\end{verbatim}
\caption{Example bytecode: Compute the square of the accumulator}
\label{fig:square}
\end{figure}

Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes and all register and the
accumulator are integers. A simple program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} bytecode, the
trace would look as follows:
\input{code/normal-tracing.txt}

To improve this situation, the tracing JIT could trace the execution of several
bytecodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode loop should be unrolled exactly so much, that the unrolled version
corresponds to a loop on the level of the user program. A loop in the user
program occurs when the program counter of the language interpreter has the
same value many times. This program counter is typically one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.

Since the tracing JIT cannot know which parts of the language interpreter are
the program counter, the author of the language interpreter needs to mark the
relevant variables of the language interpreter with the help of a \emph{hint}.
The tracing interpreter will then effectively add the values of these variables
to the position key. This means, that the loop will only be considered to be
closed, if these variables that are making up program counter at the language
interpreter level are the same a second time. Such a loop is a loop of the user
program. The program counter of the language interpreter can only be the same a
second time after an instruction in the user program sets it to an earlier
value. This happens only at backward jumps in the language interpreter. That
means that the tracing interpreter needs to check for a closed loop only when it
encounters a backward jump in the language interpreter. Again the tracing JIT
cannot known where the backward branch is located, so it needs to be told with
the help of a hint by the author of the language interpreter.

The condition for reusing already existing machine code needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. They correspond to different
unrollings and paths of that loop though. To figure out which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be entered. This
check again only needs to be performed at the backward branches of the language
interpreter.

There is a similar conceptual problem about the point where tracing is started.
Tracing starts when the tracing interpreter sees one particular loop often
enough. This loop is always going to be the bytecode dispatch loop of the
language interpreter, so the tracing interpreter will start tracing all the
time. This is not sensible. It makes more sense to start tracing only if a
particular loop in the user program would be seen often enough. Thus we
need to change the lightweight profiling to identify the loops of the user
program. Therefore profiling is also done at the backward branches of the
language interpreter, using one counter per seen program counter of the language
interpreter.

\begin{figure}
\input{code/tlr-paper-full.py}
\caption{Simple bytecode interpreter with hints applied}
\label{fig:tlr-full}
\end{figure}

Let's look at which hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. The basic thing needed to apply hints is a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red.

In addition to the classification of the variables, there are two methods of
\texttt{JitDriver} that need to be called. Both of them get as arguments the
current values of the variables listed in the definition of the driver. The
first one is \texttt{jit\_merge\_point} which needs to be put at the beginning
of the body of the bytecode dispatch loop. The other, more interesting one, is
\texttt{can\_enter\_jit}. This method needs to be called at the end of any
instruction that can set the program counter of the language interpreter to an
earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
instruction, and only if it is actually a backward jump. The place where this
method is called is where the language interpreter performs profiling to decide
when to start tracing. It is also the place where the tracing JIT checks
whether a loop is closed. This is considered to be the case when the values of
the "green" variables are the same as at an earlier call to the
\texttt{can\_enter\_jit} method.

For the small example the hints look like a lot of work. However, the amount of
hints is essentially constant no matter how large the interpreter is, which
makes it seem less significant for larger interpreters.

When executing the Square function of Figure \ref{fig:square}, the profiling
will identify the loop in the square function to be hot, and start tracing. It
traces the execution of the interpreter running the loop of the square function
for one iteration, thus unrolling the interpreter loop of the example
interpreter eight times. The resulting trace can be seen in Figure 
\ref{fig:trace-no-green-folding}.

\begin{figure}
\input{code/no-green-folding.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments.}
\label{fig:trace-no-green-folding}
\end{figure}

XXX summarize at which points the tracing interpreter needed changing
XXX all changes only to the position key and when to enter/leave the tracer!
XXX tracing remains essentially the same

\subsection{Improving the Result}

The critical problem of tracing the execution of just one bytecode has been
solved, the loop corresponds exactly to the loop in the square function.
However, the resulting trace is a bit too long. Most of its operations are not
actually doing any computation that is part of the square function. Instead,
they manipulate the data structures of the language interpreter. While this is
to be expected, given that the tracing interpreter looks at the execution of the
language interpreter, it would still be nicer if some of these operations could
be removed.

The simple insight how to greatly improve the situation is that most of the
operations in the trace are actually concerned with manipulating the
bytecode and the program counter. Those are stored in variables that are part of
the position key (they are "green"), that means that the tracer checks that they
are some fixed value at the beginning of the loop. In the example the check
would be that the \texttt{bytecode} variable is the bytecode string
corresponding to the square function and that the \texttt{pc} variable is
\texttt{4}. Therefore it is possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
Python, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} operations can be folded anyway.

With this optimization enabled, the trace looks as in Figure
\ref{fig:trace-full}. Now a lot of the language interpreter is actually gone
from the trace and what is left corresponds very closely to the loop of the
square function. The only vestige of the language interpreter is the fact that
the register list is still used to store the state of the computation. This
could be removed by some other optimization, but is maybe not really all that
bad anyway (in fact we have an experimental optimization that does exactly that,
but it is not finished).

\begin{figure}
\input{code/full.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments. The constant-folding of operations
on green variables is enabled.}
\label{fig:trace-full}
\end{figure}

%- problem: typical bytecode loops don't follow the general assumption of tracing
%- needs to unroll bytecode loop
%    - how often to unroll
%    - when to start tracing?
%    - unroll exactly so that unrolled loop corresponds to loop of the user
%      program
%- how to improve matters: introducing merge keys
%- constant-folding of operations on green things
%    - similarities to BTA of partial evaluation

\section{Implementation Issues}
\label{sect:implementation}

In this section we will describe some of the practical issues when implementing
the scheme described in the last section in PyPy. In particular we will describe
some of the problems of integrating the various parts with each other.

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It should be possible to choose when the interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
as if no hints were present in the interpreter at all.

If the JIT is enabled, things are more interesting. A classical tracing JIT will
interpret the program it is running until a common loop is identified, at which
point tracing and ultimately assembler generation starts. The tracing JIT in
PyPy is operating on the language interpreter, which is written in RPython. But
RPython programs are translatable to C. This means that interpreting the
language interpreter before a common loop is found is clearly not desirable,
since the overhead of this double-interpretation would be significantly too big
to be practical.

What is done instead is that the language interpreter keeps running as a C
program, until a common loop in the user program is found. To identify loops the
C version of the language interpreter is generated in such a way that at the
place that corresponds to the \texttt{can\_enter\_jit} hint profiling is
performed using the program counter of the language interpreter. Apart from this
bit of profiling, the language interpreter behaves in just the same way as
without a JIT.

When a hot loop in the user program is identified, tracing is started. The
tracing interpreter is invoked to start tracing the language interpreter that is
running the user program. Of course the tracing interpreter cannot actually
trace the execution of the C representation of the language interpreter. Instead
it takes the state of the execution of the language interpreter and starts
tracing using a bytecode representation of the language interpreter. That means
there are two "versions" of the language interpreter embedded in the final
executable of the VM: On the one hand it is there as executable machine code, on
the other hand as bytecode for the tracing interpreter. It also means that
tracing is costly as it incurs exactly a double interpretation overhead.

From then on things proceed like described in Section \ref{sect:tracing}. The
tracing interpreter tries to find a loop in the user program, if it found one it
will produce machine code for that loop and this machine code will be
immediately executed. The machine code is executed until a guard fails. Then the
execution should fall back to normal interpretation by the language interpreter.
This falling back is possibly a complex process, since the guard failure can
have occurred arbitrarily deep in a helper function of the language interpreter,
which would make it hard to rebuild the state of the language interpreter and
let it run from that point (e.g. this would involve building a potentially deep
C stack). Instead the falling back is achieved by a special \emph{fallback
interpreter} which runs the language interpreter and the user program from the
point of the guard failure. The fallback interpreter is essentially a variant of
the tracing interpreter that does not keep a trace. The fallback interpreter
runs until execution reaches a safe point where it is easy to let the C version
of the language interpreter resume its operation. Usually this means that the
fallback interpreter executes at most one bytecode operation of the language
interpreter. After the language interpreter takes over again, the whole process
starts again.

\subsection{Various Issues}

This section will hint at some other implementation issues and optimizations
that we have done that are beyond the scope of this paper (and will be subject
of a later publication).

\textbf{Assembler Backends:} The tracing interpreter uses a well-defined
interface to an assembler backend for code generation. This makes it possible to
easily port the tracing JIT to various architectures (including, we hope, to
virtual machines such as the JVM where backend could generate bytecode at
runtime).

\textbf{Trace Trees:} This paper ignored the problem of guards that fail in a
large percentage of cases because there are several equally likely paths through
a loop. This of course is not always practicable. Therefore we also start
tracing from guards that failed many times and produce assembler code for that
path, instead of always falling back to interpretation.

\textbf{Allocation Removal:} A key optimization for making the approach
produce good code for more complex dynamic language is to perform escape
analysis on the loop operation after tracing has been performed. In this way all
objects that are allocated during the loop and don't actually escape the loop do
not need to be allocated on the heap at all but can be exploded into their
respective fields.  This is very helpful for dynamic languages where primitive
types are often boxed, as the constant allocation of intermediate results is
very costly.

\textbf{Optimizing Frame Objects:} One problem with the removal of allocations
is that many dynamic languages are so reflective that they allow the
introspection of the frame object that the interpreter uses to store local
variables (e.g. SmallTalk, Python). This means that intermediate results always
escape because they are stored into the frame object, rendering the allocation
removal optimization ineffective. To remedy this problem we make it possible to
update the frame object lazily only when it is actually accessed from outside of
the code generated by the JIT.

\section{Evaluation}

\label{sect:evaluation}
%- benchmarks
%    - running example
%    - gameboy?

\section{Related Work}

% dynamorio stuff
% partial evaluation
% XXX

\section{Conclusion and Next Steps}

%\begin{verbatim}
%- next steps:
%  - Apply to other things, like smalltalk
%- conclusions
% - advantages + disadvantages in the meta-level approach
% - advantages are that the complex operations that occur in dynamic languages
%   are accessible to the tracer
\cite{bolz_back_2008}
\cite{Psyco}

\bigskip

\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

### 2009-03-31 talk/icooolps2009/code/tlr-paper-full.py

@@ -1,0 +1,32 @@
+\begin{verbatim}
+class TLRJitDriver(JitDriver):
+    greens = ['pc', 'bytecode']
+    reds   = ['a', 'regs']
+
+tlrjitdriver = TLRJitDriver()
+
+def interpret(bytecode, a):
+    regs = [0] * 256
+    pc = 0
+    while True:
+        tlrjitdriver.jit_merge_point(
+            bytecode=bytecode, pc=pc,
+            a=a, regs=regs)
+        opcode = ord(bytecode[pc])
+        pc += 1
+        if opcode == JUMP_IF_A:
+            target = ord(bytecode[pc])
+            pc += 1
+            if a:
+                if target < pc:
+                    tlrjitdriver.can_enter_jit(
+                        bytecode=bytecode, pc=target,
+                        a=a, regs=regs)
+                pc = target
+        elif opcode == MOV_A_R:
+            n = ord(bytecode[pc])
+            pc += 1
+            regs[n] = a
+        elif opcode == MOV_R_A:
+            ... # rest unmodified
+\end{verbatim}

### 2009-03-31 talk/icooolps2009/paper.tex
finished the review of the paper until section 3.1 (excluded)

@@
 \usepackage{color}
+\usepackage{ulem}
+
+\newboolean{showcomments}
+\setboolean{showcomments}{true}
+\ifthenelse{\boolean{showcomments}}
+  {\newcommand{\nb}[2]{
+    \fbox{\bfseries\sffamily\scriptsize#1}
+    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
+   }
+   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
+  }
+  {\newcommand{\nb}[2]{}
+   \newcommand{\version}{}
+  }
+
+\newcommand\cfbolz[1]{\nb{CFB}{#1}}
+\newcommand\anto[1]{\nb{ANTO}{#1}}
+\newcommand\arigo[1]{\nb{AR}{#1}}
+\newcommand{\commentout}[1]{}
 
@@
 
-\numberofauthors{2}
+\numberofauthors{3}
 \author{
@@
        \email{cfbolz@gmx.de}
+\alignauthor Antonio Cuni\\
+       \affaddr{University of Genova}\\
+       \affaddr{DISI}\\
+       \affaddr{Italy}\\
+       \email{cuni@disi.unige.it}
 \alignauthor Armin Rigo\\
@@
 such a way that programs in it can be compiled into various target environments,
-such as C/Posix, the CLR or the JVM. The PyPy project is described in more
+such as C/Posix, the CLI or the JVM. The PyPy project is described in more
 details in Section \ref{sect:pypy}.
@@
 To implement a dynamic language with PyPy, an interpreter for that language has
-to be written in RPython. RPython ("Restricted Python") is a subset of Python
+to be written in RPython \cite{AACM-DLS07}. RPython ("Restricted Python") is a subset of Python
 chosen in such a way that type inference can be performed on it. The language
 interpreter can then be translated with the help of PyPy into various target
-environments, such as C/Posix, the CLR and the JVM. This is done by a component
+environments, such as C/Posix, the CLI and the JVM. This is done by a component
 of PyPy called the \emph{translation toolchain}.
+
+\anto{XXX: are the following paragraphs really needed? I don't think this
+  details are needed in Section \ref{sect:implementation}.  I would say
+  something much shorter, see below.}
 
@@
 also what the tracing JIT takes as input, as we will see later.
+
+\anto{By writing VMs in a high-level language, we keep the implementation of
+  the language free of low-level details such as memory management strategy,
+  threading model or object layout.  These features are automatically added
+  during the translation process which consists in a series of steps, each
+  step transforming the representation of the program produced by the previous
+  one until we get the final executable.  As we will see later, this internal
+  low-level representation of the program is also used as an input for the
+  tracing JIT.}
 
@@
 
-The generation of loops works as follows: At first, everything is interpreted.
+\sout{The generation of loops works as follows: at first, everything is interpreted.}
+\anto{At first, when the program starts, everything is interpreted.}
 The interpreter does a bit of lightweight profiling to figure out which loops
@@
 
-When a common loop is identified, the interpreter enters a
-special mode (called tracing mode). When in tracing mode, the interpreter
+When a \sout{common}\anto{hot} loop is identified, the interpreter enters a
+special mode, called \emph{tracing mode}. When in tracing mode, the interpreter
 records a history (the \emph{trace}) of all the operations it executes, in addition
@@
 as it represents exactly the loop that is being interpreted at the moment anyway.
+
+\anto{XXX I think it's worth spending one more paragraph to explain what a
+  trace really is, i.e. that it's a list of \textbf{sequential} operations,
+  intermixed to guards which guarantee that this particular sequence is still
+  valid.  At the moment, the definition of trace is not given explicitly and
+  it's mixed with the details of how the JIT work}
 
@@
 to an earlier value, e.g. a backward branch instruction. Note that this is
-already the second place where backward branches are treated specially: During
+already the second place where backward branches are treated specially: during
 interpretation they are the place where the profiling is performed and where

### 2009-03-31 talk/icooolps2009/paper.tex
accepted most of anto's changes

@@
 
-\anto{XXX: are the following paragraphs really needed? I don't think this
-  details are needed in Section \ref{sect:implementation}.  I would say
-  something much shorter, see below.}
-
-The central idea of this way to implement VMs is that the interpreter
-implementation in RPython should be as free as possible of low-level
-implementation details, such as memory management strategy, threading model or
-object layout. Instead, these details are inserted into the VM during the
-translation process by the translation toolchain. This makes it possible to
-change these details later, if that becomes necessary. This is something that is
-hard to do with a traditional VM written in a low-level language such as C,
-since the low-level details need to be fixed early in the development-process.
-XXX is this paragraph really needed?
-
-In the following we will describe some details of the translation process, since
-they are needed in Section \ref{sect:implementation}. The first step is to produce
-control flow graphs of all functions of the RPython program. Afterwards, type
-inference is performed to gain type information about all the variables in the
-flow graphs. Afterwards, the abstraction level of the operations in the graphs
-is lowered in a stepwise fashion. At the end of this process, all operations in
-the graphs correspond rather directly to a simple operation in C. The variables
-are annotated with a C-like type system containing primitive types (like
-\texttt{Signed}, \texttt{Bool}, etc.) or pointer types pointing to Structs,
-Arrays or Functions.
-
-XXX example. reuse the one of the tracing jit?
-
-The translation process usually just turns these graphs into C code so that they
-can be compiled into an executable. However, they can also be interpreted in
-various ways. This is useful for testing and debugging the translation toolchain
-because the produced error messages in case of a crash are a lot more helpful
-than what would be produced after compilation to C. These low-level graphs are
-also what the tracing JIT takes as input, as we will see later.
-
-\anto{By writing VMs in a high-level language, we keep the implementation of
-  the language free of low-level details such as memory management strategy,
+By writing VMs in a high-level language, we keep the implementation of the
+language free of low-level details such as memory management strategy,
 threading model or object layout.  These features are automatically added
-  during the translation process which consists in a series of steps, each
-  step transforming the representation of the program produced by the previous
-  one until we get the final executable.  As we will see later, this internal
+during the translation process which consists in a series of steps, each step
+transforming the representation of the program produced by the previous one
+until we get the final executable.  As we will see later, this internal
 low-level representation of the program is also used as an input for the
-  tracing JIT.}
+tracing JIT.
 
@@
 
-\sout{The generation of loops works as follows: at first, everything is interpreted.}
-\anto{At first, when the program starts, everything is interpreted.}
+At first, when the program starts, everything is interpreted.
 The interpreter does a bit of lightweight profiling to figure out which loops
@@
 
-When a \sout{common}\anto{hot} loop is identified, the interpreter enters a
+When a hot loop is identified, the interpreter enters a
 special mode, called \emph{tracing mode}. When in tracing mode, the interpreter

### 2009-03-31 talk/icooolps2009/paper.tex
more comments

@@
 \newcommand{\commentout}[1]{}
+
+\normalem
 
@@
 
-A tracing JIT compiler finds the hot loops of the program it is compiling. In
+\anto{I find this para a bit confusing, but I don't really know how to improve
+  it. I try a rewriting, feel free to revert}
+
+\sout{A tracing JIT compiler finds the hot loops of the program it is compiling. In
 our case, this program is the language interpreter. The hot loop of the language
 interpreter is the bytecode dispatch loop. Usually it is also the only hot loop.
-Tracing one iteration of this loop means that the execution of one bytecode was
+Tracing one iteration of this loop means that the execution of one opcode was
 seen. This means that the resulting machine code will correspond to a loop, that
-assumes that this particular bytecode will be executed many times in a row,
-which is clearly very unlikely.
+assumes that this particular opcode will be executed many times in a row,
+which is clearly very unlikely.}
+
+\anto{Similarly, we need to distinguish two kind of loops: \emph{interpreter
+    loops} are loops \textbf{inside} the language interpreter (e.g., the
+  bytecode dispatch loop). \emph{User loops} are loops in the user program.}
+
+\anto{To improve the performances of the user programs, we need to find and
+  compile traces that correspond to user loops.  The strategy described in
+  section \ref{sect:tracing} does not work well, as it only detects
+  interpreter loops, and in particular the bytecode dispatch one.}
 
@@
 the square of the accumulator is shown in Figure \ref{fig:square}. If the
-tracing interpreter traces the execution of the \texttt{DECR\_A} bytecode, the
+tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose integer value is 7), the
 trace would look as follows:
@@
 
+\anto{ Because of the guard on \texttt{opcode0}, the code compiled from this
+  trace will be useful only when executing a \texttt{DECR\_A}.  For all the
+  other operations the guard will fail, making the whole system probably
+  slower than without JIT.}
+
+% YYY: (anto) I reviewd until here
+
 To improve this situation, the tracing JIT could trace the execution of several
-bytecodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
-bytecode loop should be unrolled exactly so much, that the unrolled version
+opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
+bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
 corresponds to a loop on the level of the user program. A loop in the user
@@
 
-The critical problem of tracing the execution of just one bytecode has been
+The critical problem of tracing the execution of just one opcode has been
 solved, the loop corresponds exactly to the loop in the square function.
@@
 \caption{Trace when executing the Square function of Figure \ref{fig:square},
-with the corresponding bytecodes as comments. The constant-folding of operations
+with the corresponding opcodes as comments. The constant-folding of operations
 on green variables is enabled.}

### 2009-03-31 talk/icooolps2009/paper.tex
Accepting one of Anto's changes, rejecting the rewrite: The problem with the rewrite was that it jumped to the conclusion of where the text gets to several paragraphs later. Instead, tried to rewrite it slightly to make it more clear.

@@
 
-\anto{I find this para a bit confusing, but I don't really know how to improve
-  it. I try a rewriting, feel free to revert}
+Similarly, we need to distinguish loops at two different levels:
+\emph{interpreter loops} are loops \emph{inside} the language interpreter. On
+the other hand, \emph{user loops} are loops in the user program.
 
-\sout{A tracing JIT compiler finds the hot loops of the program it is compiling. In
+A tracing JIT compiler finds the hot loops of the program it is compiling. In
 our case, this program is the language interpreter. The hot loop of the language
-interpreter is the bytecode dispatch loop. Usually it is also the only hot loop.
-Tracing one iteration of this loop means that the execution of one opcode was
-seen. This means that the resulting machine code will correspond to a loop, that
-assumes that this particular opcode will be executed many times in a row,
-which is clearly very unlikely.}
-
-\anto{Similarly, we need to distinguish two kind of loops: \emph{interpreter
-    loops} are loops \textbf{inside} the language interpreter (e.g., the
-  bytecode dispatch loop). \emph{User loops} are loops in the user program.}
-
-\anto{To improve the performances of the user programs, we need to find and
-  compile traces that correspond to user loops.  The strategy described in
-  section \ref{sect:tracing} does not work well, as it only detects
-  interpreter loops, and in particular the bytecode dispatch one.}
+interpreter is its bytecode dispatch loop. Usually that is is also the only hot
+loop of the language interpreter.  Tracing one iteration of this loop means that
+the recorded trace corresponds to execution of one opcode. This means that the
+assumption that the tracing JIT makes -- that several iterations of a hot loop
+take the same or similar code paths -- is just wrong in this case. It is very
+unlikely that the same particular opcode is executed many times in a row.
 
@@
 
-\anto{ Because of the guard on \texttt{opcode0}, the code compiled from this
-  trace will be useful only when executing a \texttt{DECR\_A}.  For all the
-  other operations the guard will fail, making the whole system probably
-  slower than without JIT.}
+Because of the guard on \texttt{opcode0}, the code compiled from this trace will
+be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
+the other operations the guard will fail, which will mean that performance is
+probably not improved at all.
 

### 2009-03-31 talk/icooolps2009/paper.tex
made this into a figure too

@@
 the square of the accumulator is shown in Figure \ref{fig:square}. If the
-tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose integer value is 7), the
-trace would look as follows:
-\input{code/normal-tracing.txt}
-
+tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
+integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
 Because of the guard on \texttt{opcode0}, the code compiled from this trace will
@@
 probably not improved at all.
+
+\begin{figure}
+\input{code/normal-tracing.txt}
+\caption{Trace when executing the \texttt{DECR\_A} opcode}
+\label{fig:trace-normal}
+\end{figure}
 

### 2009-03-31 talk/icooolps2009/paper.tex
fix an XXX

@@
 
-XXX write somewhere that we will be predominantly occupied with bytecode
-interpreters
 XXX write somewhere that one problem of using tracing JITs for dynamic languages
@@
 interpreter}. On the other hand, there is the interpreter that is running the
-users programs, which we will call the \emph{language interpreter}. The program
-that the language interpreter executes we will call the \emph{user program}
-(from the point of view of a VM author, the "user" is a programmer using the
-VM).
+users programs, which we will call the \emph{language interpreter}. In the
+following, we will assume that the language interpreter is bytecode-based. The
+program that the language interpreter executes we will call the \emph{user
+program} (from the point of view of a VM author, the "user" is a programmer
+using the VM).
 

### 2009-03-31 talk/icooolps2009/paper.tex
kill XXX

@@
 mechanisms in place to still produce more code for the cases of guard failures,
-but they are of no importance for this paper XXX is that true?).
+but they are of no importance for this paper).
 

### 2009-04-01 talk/icooolps2009/paper.tex
Some tweaks here and there, start with the related work.

@@
 
-The PyPy project\footnote{http://codespeak.net/pypy} was started to implement a
+The PyPy project\footnote{http://codespeak.net/pypy}
+\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} was started to implement a
 new Python interpreter in Python but has now extended its goals to be an
@@
 relatively simple way to implement a JIT compiler for a dynamic language
-\cite{XXX}. The technique is now used by both and are now being used by both Mozilla's
-TraceMonkey JavaScript VM \cite{XXX} and Adobe's Tamarin ActionScript VM
-\cite{XXX}.
+\cite{mason_chang_efficient_2007}. The technique is now used by both and are now
+being used by both Mozilla's TraceMonkey JavaScript VM \cite{XXX} and Adobe's
+Tamarin ActionScript VM \cite{XXX}.
 
@@
 
+XXX write somewhere here on which level the RPython tracer operates
+
 Let's look at a small example. Take the following (slightly contrived) RPython
@@
 \subsection{Applying a Tracing JIT to an Interpreter}
-
-XXX \cite{sullivan_dynamic_2003} somewhere
 
@@
 
-If the JIT is enabled, things are more interesting. A classical tracing JIT will
+If the JIT is enabled, things are more interesting. At the moment the JIT can
+only be enabled when translating the interpreter to C, but we hope to lift that
+restriction in the future. A classical tracing JIT will
 interpret the program it is running until a common loop is identified, at which
@@
 PyPy is operating on the language interpreter, which is written in RPython. But
-RPython programs are translatable to C. This means that interpreting the
+RPython programs are statically translatable to C anyway. This means that interpreting the
 language interpreter before a common loop is found is clearly not desirable,
@@
 virtual machines such as the JVM where backend could generate bytecode at
-runtime).
+runtime). At the moment the only implemented backend is a simple 32-bit
+Intel-x86 backend.
 
@@
 large percentage of cases because there are several equally likely paths through
-a loop. This of course is not always practicable. Therefore we also start
-tracing from guards that failed many times and produce assembler code for that
-path, instead of always falling back to interpretation.
+a loop. Just falling back to interpretation in this case is not practicable.
+Therefore we also start tracing from guards that failed many times and produce
+machine code for that path, instead of always falling back to interpretation.
 
@@
 \section{Evaluation}
+\label{sect:evaluation}
 
-\label{sect:evaluation}
+In this section we try to evaluate the work done so far by looking at some
+benchmark numbers. Since the work is not finished, these benchmarks can only be
+preliminary. All benchmarking was done on a machine with a 1.4 GHz Pentium M
+processor and 1GiB RAM, using Linux 2.6.27.
+
 %- benchmarks
@@
 
-% dynamorio stuff
-% partial evaluation
-% XXX
+Applying a trace-based optimizer to an interpreter, adding hints to help the
+tracer produce better results been tried before in the context of the DynamoRIO
+project \cite{sullivan_dynamic_2003}. This work is conceptually very close to
+ours. They achieve the same unrolling of the interpreter loop so that the
+unrolled version corresponds to the loops in the user program. However the
+approach is greatly hindered by the fact that they trace on the machine code
+level and thus have no high-level information available about the interpreter.
+This makes it necessary to add quite a large number of hints, because at the
+assembler level it is not really visible anymore that e.g. a bytecode string is
+really immutable. Also more advanced optimizations like allocation removal would
+not be possible with that approach.
+
+The standard approach for automatically producing a compiler for a programming
+language given an interpreter for it is that of partial evaluation \cite{XXX},
+\cite{XXX}. Conceptually there are some similarities to our work. In partial
+evaluation some arguments of the interpreter function are known (static) while
+the rest are unknown (dynamic). This separation of arguments is related to our
+separation of variables into those that should be part of the position key and
+the rest. In partial evaluation all parts of the interpreter that rely only on
+static arguments can be constant-folded so that only operations on the dynamic
+arguments remain.
+
+Classical partial evaluation has failed to be useful for dynamic language for
+much the same reasons why ahead-of-time compilers cannot compile them to
+efficient code. If the partial evaluator knows only the program it simply does
+not have enough information to produce good code. Therefore some work has been
+done to do partial evaluation at runtime. One of the earliest works on runtime
+specialisation is Tempo for C \cite{XXX}. However, it is essentially a normal
+partial evaluator ``packaged as a library''; decisions about what can be
+specialised and how are pre-determined. Another work in this direction is DyC
+\cite{grant_dyc_2000}, another runtime specialiser for C. Both of these projects
+have a similar problem as DynamoRIO.  Targeting the C language makes
+higher-level specialisation difficult (e.g.\ \texttt{malloc} can not be
+optimized).
+
+XXX what else?
 
@@
 \cite{bolz_back_2008}
-\cite{Psyco}
 

### 2009-04-01 talk/icooolps2009/paper.tex
more comments

@@
 
-% YYY: (anto) I reviewd until here
-
 To improve this situation, the tracing JIT could trace the execution of several
@@
 bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
-corresponds to a loop on the level of the user program. A loop in the user
-program occurs when the program counter of the language interpreter has the
-same value many times. This program counter is typically one or several
+corresponds to \emph{user loop}. User loops
+occur when the program counter of the language interpreter has the
+same value many times. This program counter is typically stored in one or several
 variables in the language interpreter, for example the bytecode object of the
 currently executed function of the user program and the position of the current
-bytecode within that.
+bytecode within that.  In the example above, the program counter is represented by 
+the \texttt{bytecode} and \texttt{pair} variables.
+\anto{XXX: why ``many times''? Twice is enough to have a loop, though is not hot}
 
@@
 The tracing interpreter will then effectively add the values of these variables
-to the position key. This means, that the loop will only be considered to be
-closed, if these variables that are making up program counter at the language
-interpreter level are the same a second time. Such a loop is a loop of the user
-program. The program counter of the language interpreter can only be the same a
+to the position key. This means that the loop will only be considered to be
+closed if these variables that are making up program counter at the language
+interpreter level are the same a second time. \sout{Such a loop is a loop of the user
+program.} \anto{Loops found in this way are, by definition, user loops}.
+
+The program counter of the language interpreter can only be the same a
 second time after an instruction in the user program sets it to an earlier
@@
 the help of a hint by the author of the language interpreter.
+\anto{XXX: the backward jumps are in the user program, not in the language
+  interprer. Am I missing something?}
 
@@
 
+\sout{
 There is a similar conceptual problem about the point where tracing is started.
@@
 interpreter.
+}
+
+\anto{I find the previous para a bit confusing. What about something more
+  lightweight, like the following?}
+
+\anto{Similarly, the interpreter uses the same techniques to detect \emph{hot
+    user loops}: the profiling is done at the backward branches of the user
+  program, using one counter per seen program counter of the language
+  interpreter.}
 
@@
 string is currently being interpreted. All other variables are red.
+
+\anto{XXX: they driver does not list \emph{all} the variables; e.g. \texttt{n}
+  is not listed.  But maybe we can just ignore this issue}
 
@@
 
+\anto{I propose to show also the trace with the malloc removal enabled, as it
+  is much nicer to see. Maybe we can say that the experimental optimization we
+  are working on would generate this and that}
+
 \begin{figure}
@@
 \end{figure}
+
+\anto{Once we get the highly optimized trace, we can pass it to the \emph{JIT
+    backend}, which generates the correspondent machine code. XXX: do we want
+  to say something more about backends?}
 
@@
 
+% YYY (anto)
+
 \section{Implementation Issues}
@@
 
-In this section we will describe some of the practical issues when implementing
-the scheme described in the last section in PyPy. In particular we will describe
-some of the problems of integrating the various parts with each other.
+In this section we will describe some of the practical issues when
+implementing the scheme described in the last section in PyPy. In particular
+we will describe some of the problems of integrating the various parts with
+each other.
+
+\anto{XXX: We shoud clarify the distinction between translation/compilation
+  somewhere in the introduction}
 

### 2009-04-01 talk/icooolps2009/paper.tex
finish the review

@@
 
-When a hot loop in the user program is identified, tracing is started. The
+When a hot user loop is identified, tracing is started. The
 tracing interpreter is invoked to start tracing the language interpreter that is
@@
 there are two "versions" of the language interpreter embedded in the final
-executable of the VM: On the one hand it is there as executable machine code, on
+executable of the VM: on the one hand it is there as executable machine code, on
 the other hand as bytecode for the tracing interpreter. It also means that
@@
 From then on things proceed like described in Section \ref{sect:tracing}. The
-tracing interpreter tries to find a loop in the user program, if it found one it
+tracing interpreter tries to find a loop in the user program, if it finds one it
 will produce machine code for that loop and this machine code will be
@@
 a loop. Just falling back to interpretation in this case is not practicable.
+\sout{
 Therefore we also start tracing from guards that failed many times and produce
 machine code for that path, instead of always falling back to interpretation. 
+}
+\anto{
+Therefore, if we find a guard that fails often enough, we start tracing from
+there and produce efficient machine code for that case, instead of alwayas
+falling back to interpretation.
+}
 
@@
 the code generated by the JIT.
+
+\anto{XXX: should we say that virtualizables are very cool, that nobody else
+  does that and that they are vital to get good performaces with python
+  without sacrificing compatibility?}
 
@@
 
+\anto{I would cite ourselves (maybe the JIT technical report?) and maybe
+  psyco}
+
 \section{Conclusion and Next Steps}

### 2009-04-01 talk/icooolps2009/paper.tex
fix/answer some of anto's comments

@@
 occur when the program counter of the language interpreter has the
-same value many times. This program counter is typically stored in one or several
+same value several times. This program counter is typically stored in one or several
 variables in the language interpreter, for example the bytecode object of the
@@
 the \texttt{bytecode} and \texttt{pair} variables.
-\anto{XXX: why ``many times''? Twice is enough to have a loop, though is not hot}
 
@@
 closed if these variables that are making up program counter at the language
-interpreter level are the same a second time. \sout{Such a loop is a loop of the user
-program.} \anto{Loops found in this way are, by definition, user loops}.
+interpreter level are the same a second time.  Loops found in this way are, by
+definition, user loops.
 
@@
 encounters a backward jump in the language interpreter. Again the tracing JIT
-cannot known where the backward branch is located, so it needs to be told with
-the help of a hint by the author of the language interpreter.
-\anto{XXX: the backward jumps are in the user program, not in the language
-  interprer. Am I missing something?}
+cannot known which part of the language interpreter implements backward jumps,
+so it needs to be told with the help of a hint by the author of the language
+interpreter.
 
@@
 
-\sout{
-There is a similar conceptual problem about the point where tracing is started.
-Tracing starts when the tracing interpreter sees one particular loop often
-enough. This loop is always going to be the bytecode dispatch loop of the
-language interpreter, so the tracing interpreter will start tracing all the
-time. This is not sensible. It makes more sense to start tracing only if a
-particular loop in the user program would be seen often enough. Thus we
-need to change the lightweight profiling to identify the loops of the user
-program. Therefore profiling is also done at the backward branches of the
-language interpreter, using one counter per seen program counter of the language
-interpreter.
-}
-
-\anto{I find the previous para a bit confusing. What about something more
-  lightweight, like the following?}
-
-\anto{Similarly, the interpreter uses the same techniques to detect \emph{hot
-    user loops}: the profiling is done at the backward branches of the user
-  program, using one counter per seen program counter of the language
-  interpreter.}
+The language interpreter uses a similar techniques to detect \emph{hot user
+loops}: the profiling is done at the backward branches of the user program,
+using one counter per seen program counter of the language interpreter.
 
@@
 string is currently being interpreted. All other variables are red.
-
-\anto{XXX: they driver does not list \emph{all} the variables; e.g. \texttt{n}
-  is not listed.  But maybe we can just ignore this issue}
 
@@
 
-\anto{I propose to show also the trace with the malloc removal enabled, as it
+\anto{XXX I propose to show also the trace with the malloc removal enabled, as it
   is much nicer to see. Maybe we can say that the experimental optimization we
-  are working on would generate this and that}
+  are working on would generate this and that} \cfbolz{This example is not about
+  mallocs! There are no allocations in the loop. The fix would be to use
+  maciek's lazy list stuff (or whatever it's called) which is disabled at the
+  moment}
 
@@
 
-\anto{Once we get the highly optimized trace, we can pass it to the \emph{JIT
-    backend}, which generates the correspondent machine code. XXX: do we want
-  to say something more about backends?}
+Once we get this highly optimized trace, we can pass it to the \emph{JIT
+backend}, which generates the correspondent machine code.
 
@@
 %    - similarities to BTA of partial evaluation
-
-% YYY (anto)
 
@@
 a loop. Just falling back to interpretation in this case is not practicable.
-\sout{
-Therefore we also start tracing from guards that failed many times and produce
-machine code for that path, instead of always falling back to interpretation. 
-}
-\anto{
 Therefore, if we find a guard that fails often enough, we start tracing from
@@
 falling back to interpretation.
-}
 
@@
   does that and that they are vital to get good performaces with python
-  without sacrificing compatibility?}
+  without sacrificing compatibility?} \cfbolz{no: feels a bit dishonest to not
+  describe them properly and then say that they are very cool and vital}
 

### 2009-04-01 talk/icooolps2009/paper.tex
give a quicky summary of the various phases of VMs before going into details
move the list after the assumptions
explain what is a trace

@@
 
+Typically, programs executed by a tracing VMs goes through various phases:
+
+\begin{itemize}
+\item Interpretation/profiling
+\item Tracing
+\item Code generation
+\item Execution of the generated code
+\end{itemize}
+
+The \emph{code generation} phase takes as input the trace generated during
+\emph{tracing}.
+
 At first, when the program starts, everything is interpreted.
@@
 When a hot loop is identified, the interpreter enters a
-special mode, called \emph{tracing mode}. When in tracing mode, the interpreter
+special mode, called \emph{tracing mode}. \sout{When in tracing mode, the interpreter
 records a history (the \emph{trace}) of all the operations it executes, in addition
-to actually performing the operations. During tracing, the trace is repeatedly
+to actually performing the operations.
+} \anto{During tracing, the interpreter records a history of all the
+  operations it executes.}
+
+\anto{
+Such a history is called a \emph{trace}: it is a sequential list of
+operations, togheter with their actual operands and results.  By examining the
+trace, it is possible to produce highly efficient machine code by emitting
+only the operations needed.  Being sequential, the trace represents only one
+of the many possible paths through the code: to ensure correctness, the trace
+contains a \emph{guard} at every possible point where the path could have
+followed another direction, for example \texttt{if}s or indirect/virtual
+calls.  When emitting the machine code, we turn every guard into a quick check
+to guarantee that the path we are executing is still valid.  If a guard fails,
+we immediately quit from the machine code and continue the execution in other
+ways.  
+}
+
+During tracing, the trace is repeatedly
 (XXX make this more precise: when does the check happen?)

### 2009-04-01 talk/icooolps2009/paper.tex
Typos, and a note about the introduction.

@@
 
+\arigo{We should not start from scratch and insert as little details what
+differs in our approach when compared to others; instead we should give a
+higher-level overview and then focus on these details, and a couple of
+references for more info about the "common" part.
+%
+In general there are many things that are never said at all.
+I think the introduction should really be written from the point of view of
+someone that has read already some papers for JavaScript.}
+
 Tracing JITs are an idea explored by the Dynamo project
@@
 relatively simple way to implement a JIT compiler for a dynamic language
-\cite{mason_chang_efficient_2007}. The technique is now used by both and are now
+\cite{mason_chang_efficient_2007}. The technique is now
 being used by both Mozilla's TraceMonkey JavaScript VM \cite{XXX} and Adobe's
@@
 Such a history is called a \emph{trace}: it is a sequential list of
-operations, togheter with their actual operands and results.  By examining the
+operations, together with their actual operands and results.  By examining the
 trace, it is possible to produce highly efficient machine code by emitting
@@
 of all the operations in it. The machine code can then be immediately executed,
-as it represents exactly the loop that is being interpreted at the moment anyway.
+as it represents exactly the loop that was being interpreted so far.
 

### 2009-04-01 talk/icooolps2009/paper.tex
Small typos.

@@
 interpreter is its bytecode dispatch loop. Usually that is is also the only hot
-loop of the language interpreter.  Tracing one iteration of this loop means that
+loop of the language interpreter \arigo{Uh?}.  Tracing one iteration of this loop means that
 the recorded trace corresponds to execution of one opcode. This means that the
@@
 
-Applying a trace-based optimizer to an interpreter, adding hints to help the
-tracer produce better results been tried before in the context of the DynamoRIO
+Applying a trace-based optimizer to an interpreter and adding hints to help the
+tracer produce better results has been tried before in the context of the DynamoRIO
 project \cite{sullivan_dynamic_2003}. This work is conceptually very close to

### 2009-04-01 talk/icooolps2009/paper.tex
Going over some changes by Anto. Still not quite happy with this area.

@@
 \usepackage{ulem}
+\usepackage[utf8]{inputenc}
 
@@
 
-\title{Tracing the Meta-Level: PyPy's JIT Compiler}
+\title{Tracing the Meta-Level: PyPy's Tracing JIT Compiler}
 
@@
 
+XXX contributions of this paper include:
+
 %- dynamic languages important
@@
 
-Tracing JITs are an idea explored by the Dynamo project
+Tracing JITs are an idea initially explored by the Dynamo project
 \cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
@@
 When a hot loop is identified, the interpreter enters a
-special mode, called \emph{tracing mode}. \sout{When in tracing mode, the interpreter
-records a history (the \emph{trace}) of all the operations it executes, in addition
-to actually performing the operations.
-} \anto{During tracing, the interpreter records a history of all the
-  operations it executes.}
+special mode, called \emph{tracing mode}. During tracing, the interpreter
+records a history of all the operations it executes.
 
-\anto{
 Such a history is called a \emph{trace}: it is a sequential list of
@@
 only the operations needed.  Being sequential, the trace represents only one
-of the many possible paths through the code: to ensure correctness, the trace
+of the many possible paths through the code. To ensure correctness, the trace
 contains a \emph{guard} at every possible point where the path could have
-followed another direction, for example \texttt{if}s or indirect/virtual
-calls.  When emitting the machine code, we turn every guard into a quick check
+followed another direction, for example conditions or indirect/virtual
+calls.  When emitting the machine code, every guard is turned into a quick check
 to guarantee that the path we are executing is still valid.  If a guard fails,
-we immediately quit from the machine code and continue the execution in other
+we immediately quit from the machine code and continue the execution by falling
 ways.
-}
 
 During tracing, the trace is repeatedly
-(XXX make this more precise: when does the check happen?)
 checked whether the interpreter is at a position in the program that it had seen
@@
 as it represents exactly the loop that was being interpreted so far.
-
-\anto{XXX I think it's worth spending one more paragraph to explain what a
-  trace really is, i.e. that it's a list of \textbf{sequential} operations,
-  intermixed to guards which guarantee that this particular sequence is still
-  valid.  At the moment, the definition of trace is not given explicitly and
-  it's mixed with the details of how the JIT work}
 

### 2009-04-02 talk/icooolps2009/paper.tex
minor clarification and some notes

@@
 \newcommand\arigo[1]{\nb{AR}{#1}}
+\newcommand\fijal[1]{\nb{FIJAL}{#1}}
 \newcommand{\commentout}[1]{}
@@
 A recent approach to getting better performance for dynamic languages is that of
-tracing JIT compilers. XXX
+tracing JIT compilers. XXX [fijal] cite andreas gal paper?
 
@@
 earlier in the trace. If this happens, the trace recorded corresponds to a loop
-in the program that the tracing interpreter is running. At this point, this loop
+in the interpreted program that the tracing interpreter is running. At this point, this loop
 is turned into machine code by taking the trace and making machine code versions
 of all the operations in it. The machine code can then be immediately executed,
+starting from the second iteration of the loop,
 as it represents exactly the loop that was being interpreted so far.
@@
 
-XXX write somewhere here on which level the RPython tracer operates
+\fijal{RPython tracer operates on a level which is slightly higher level than
+of the underlaying C. To be more precise: it operates on exception-transformed,
+non-gc-transformed graphs with an extra bit of abstraction inversion that
+finds out operations on loops and dicts, not sure how to express this}
 
@@
 
+\fijal{Following paragraph is more confusing than helpful. The trace contains
+all operations performed, including full inlining of functions called.
+The loop will loop over the path in the interpreter, which corresponds to
+path taken when tracing (which we assume is a common case).
+Later, which is outside of the scope of this paper, less common cases
+might be turned into assembler
+code as well, producing bridges, if they're common enough. Note that this
+loop is infinite, which means the only way to exit it is via guard failure}
 This trace will then be turned into machine code. Note that the machine code
@@
 
+\fijal{I'm lost, will continue later}
 A tracing JIT compiler finds the hot loops of the program it is compiling. In

### 2009-04-02 talk/icooolps2009/paper.tex
another round of reading

@@
 
-\fijal{I'm lost, will continue later}
+\fijal{I find following paragraph out of scope and completely confusing, we
+should instead simply state that we unroll the loop, how we do that and
+why we do that. Completely ignore aspect of an interpreter loop I suppose,
+because everything previously keeps talking about can\_enter\_jit that closes
+loop being available at jump back bytecodes}
 A tracing JIT compiler finds the hot loops of the program it is compiling. In
@@
 
+\fijal{This paragraph should go away as well}
 Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
@@
 
+\fijal{This is wrong. Without virtuals there is also at most one assembler
+loop per user loop. If it has more branches, we enter the loop as usual and
+then we create a bridge for a new situation}
 The condition for reusing already existing machine code needs to be adapted to
@@
 
+\fijal{Stopped reading at that point}
 Let's look at which hints would need to be applied to the example interpreter

### 2009-04-02 talk/icooolps2009/paper.tex
Fix a couple of XXXs.

@@
 this. Most of them boil down to the inherent complexities of using compilation.
-Interpreters are simple to understand and to implement whereas writing a
-just-in-time compiler is an error-prone task that is even made harder by the
+Interpreters are simple to implement, understand, extend and port whereas writing a
+just-in-time compiler is an error-prone task that is made even harder by the
 dynamic features of a language.
 
-writing an interpreter has many advantages... XXX
-
 A recent approach to getting better performance for dynamic languages is that of
-tracing JIT compilers. XXX [fijal] cite andreas gal paper?
+tracing JIT compilers \cite{XXX}. Writing a tracing JIT compiler is relatively
+simple, because it can be added to an existing interpreter for a language,
+because the interpreter takes over some of the functionality of the compiler and
+the machine code generation part can be simplified.
 
@@
 
-XXX contributions of this paper include:
+The contributions of this paper are:
+\begin{itemize}
+\item Techniques for improving the generated code when applying a tracing JIT to
+an interpreter
+\item 
+\end{itemize}
 
@@
 threading model or object layout.  These features are automatically added
-during the translation process which consists in a series of steps, each step
-transforming the representation of the program produced by the previous one
-until we get the final executable.  As we will see later, this internal
-low-level representation of the program is also used as an input for the
-tracing JIT.
+during the translation process. The process starts by performing control flow
+graph construction and type inferences, then followed by a series of steps, each step
+transforming the intermediate representation of the program produced by the
+previous one until we get the final executable.  The first transformation step
+makes details of the Python object model explicit in the intermediate
+representation, later steps introducing garbage collection and other low-level
+details. As we will see later, this internal representation of the program is
+also used as an input for the tracing JIT.
 
@@
 
-At first those functions will be interpreted, but after a while, profiling shows
+To trace this, a bytecode form of these functions needs to be introduced that
+the tracer understands. The tracer interprets a bytecode that is an encoding of
+the intermediate representation of PyPy's translation toolchain after type
+inference has been performed and Python-specifics have been made explicit. At
+first those functions will be interpreted, but after a while, profiling shows
 that the \texttt{while} loop in \texttt{strange\_sum} is executed often.  The
-tracing JIT will then start trace the execution of that loop.  The trace would
+tracing JIT will then start to trace the execution of that loop.  The trace would
 look as follows:
@@
 i2 = int_ge(n1, Const(0))
-guard_true(i2) [result1]
+guard_true(i2)
 jump(result1, n1)
@@
 
-XXX add a note about the SSA-ness of the trace
-
-\fijal{Following paragraph is more confusing than helpful. The trace contains
-all operations performed, including full inlining of functions called.
-The loop will loop over the path in the interpreter, which corresponds to
-path taken when tracing (which we assume is a common case).
-Later, which is outside of the scope of this paper, less common cases
-might be turned into assembler
-code as well, producing bridges, if they're common enough. Note that this
-loop is infinite, which means the only way to exit it is via guard failure}
-This trace will then be turned into machine code. Note that the machine code
-loop is by itself infinite and can only be left via a guard failure. Also note
-\texttt{f} was inlined into the loop and how the common \texttt{else} case was
-turned into machine code, while the other one is implemented via a guard
-failure. The variables in square brackets after the guards are the state that
-the interpreter will get when the guard fails.
+The operations in this sequence are operations of the mentioned intermediate
+representation (e.g. note that the generic modulo and equality operations in the
+function above have been recognized to always work on integers and are thus
+rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
+operations that were executed, is in SSA-form \cite{XXX} and ends with a jump
+to its own beginning, forming an endless loop that can only be left via a guard
+failure. The call to \texttt{f} was inlined into the trace. Of the condition in
+\texttt{f} the much more common \texttt{else} case was traced. The other case is
+implemented via a guard failure. This trace can then be turned into machine code
+and executed.
 
@@
 A tracing JIT compiler finds the hot loops of the program it is compiling. In
-our case, this program is the language interpreter. The hot loop of the language
-interpreter is its bytecode dispatch loop. Usually that is is also the only hot
-loop of the language interpreter \arigo{Uh?}.  Tracing one iteration of this loop means that
+our case, this program is the language interpreter. The most important hot loop
+of the language interpreter is its bytecode dispatch loop (for many simple
+interpreters it is also the only hot loops).  Tracing one iteration of this
+loop means that
 the recorded trace corresponds to execution of one opcode. This means that the
@@
 The first integration problem is how to \emph{not} integrate the tracing JIT at
-all. It should be possible to choose when the interpreter is translated to C
+all. It should be possible to choose when the language interpreter is translated to C
 whether the JIT should be built in or not. If the JIT is not enabled, all the
@@
 specialised and how are pre-determined. Another work in this direction is DyC
-\cite{grant_dyc_2000}, another runtime specialiser for C. Both of these projects
+\cite{grant_dyc_2000}, another runtime specializer for C. Both of these projects
 have a similar problem as DynamoRIO.  Targeting the C language makes

### 2009-04-02 talk/icooolps2009/paper.tex
add a paragraph to the related work

@@
 
+There has been some attempts to do \emph{dynamic partial evaluation}, which is
+partial evaluation that defers partial evaluation completely to runtime
+to make partial evaluation more useful for dynamic languages. This concept was
+introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
+small dynamic language based on lambda-calculus. There is some work by one of
+the authors to implement a dynamic partial evaluator for Prolog \cite{XXX}.
+
 XXX what else?

### 2009-04-02 talk/icooolps2009-dotnet/benchmarks.tex
start of the icooolps .NET paper. For now, I just wrote the abstract and blindly copied the last sections from the old ecoop paper

--- 2008-12-21 talk/ecoop2009/benchmarks.tex
@@
 Moreover, for each benchmark we also show the time taken by running the
-equivalent program written in C\#\footnote{The sources for both TLC and C\#
+equivalent program written in C\#.\footnote{The sources for both TLC and C\#
   programs are available at:
+
   http://codespeak.net/svn/pypy/extradoc/talk/ecoop2009/benchmarks/}
 
-The benchmarks have been run on machine with an Intel Pentium 4 CPU running at
+The benchmarks have been run on a machine with an Intel Pentium 4 CPU running at
 3.20 GHz and 2 GB of RAM, running Microsoft Windows XP and Microsoft .NET
@@
 Since the TLC assembler source is long and hard to read,
-Figure~\ref{fig:accumulator} shows the equivalent program written in an
+figure~\ref{fig:accumulator} shows the equivalent program written in an
 invented Python-like syntax.
@@
 once. Instead, it compiles and executes code chunk by chunk, waiting until it
-knows enough informations to generate highly efficient code.  In particualr,
+knows enough informations to generate highly efficient code.  In particular,
 at the time it emits the code for the inner loop it exactly knows the

### 2009-04-02 talk/icooolps2009-dotnet/paper.tex
start of the icooolps .NET paper. For now, I just wrote the abstract and blindly copied the last sections from the old ecoop paper

\documentclass{acm_proc_article-sp}

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{listings}

\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\davide[1]{\nb{DAV}{#1}}
\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand{\commentout}[1]{}

\normalem

\let\oldcite=\cite

\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}

\begin{document}

%% \title{Automatic generation of JIT compilers for dynamic languages in
%%   .NET\thanks{This work has been partially supported by MIUR EOS DUE -
%%     Extensible Object Systems for Dynamic and Unpredictable Environments and
%%     by the EU-funded project: IST 004779 PyPy (PyPy: Implementing Python in
%%     Python).}}

\title{Faster than C\#: efficient implementation of dynamic languages on
  .NET\thanks{This work has been partially supported by MIUR EOS DUE -
    Extensible Object Systems for Dynamic and Unpredictable Environments and
    by the EU-funded project: IST 004779 PyPy (PyPy: Implementing Python in
    Python).}}

%% Alternative title: Faster than C\#: the future of dynamic languages on .NET

\numberofauthors{3}
\author{
\alignauthor Antonio Cuni\\
       \affaddr{DISI, University of Genova}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\alignauthor Davide Ancona\\
       \affaddr{DISI, University of Genova}\\
       \affaddr{Italy}\\
       \email{davide@disi.unige.it}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\maketitle

\begin{abstract}
The CLI, i.e. the virtual machine at the core of the .NET environment, is a
statically typed virtual machine.  Although it is being used as a target for
many different languages, it is hard to provide higly efficient implementation
of languages whose model is too different than the one of the VM.  In
particular, dynamically typed languages are typically order of magnitude
slower than, e.g., C\#.

\anto{XXX: should we cite the DLR? How to prove that DLR programs are much
  slower than C\#?}

Recent developments show that JIT compilers can exploit runtime type
information to generate quite efficient code.  Unfortunately, writing a JIT
compiler is far from being simple.  In this paper we report our positive
experience with automatic generation of JIT compilers as supported by the PyPy
infrastructure, by focusing on JIT compilation for .NET.

JIT compilation for .NET means to generate .NET bytecode at runtime, which
will further compiled down to machine code by .NET's own JIT compiler.  The
main and novel contribution of this paper is to show that this
\emph{two-layers JIT} technique is effective, and can give very high speedups,
up to make dynamic languages faster than C\# itself under certain
circumstances.

The practicality of the approach is demonstrated by showing some promising
experiments done with benchmarks written in a simple dynamic language.
\end{abstract}

\section{Introduction}
XXX
\cite{PyPyJIT09}

\section{PyPy and automatic JIT compiler generation}
XXX

\include{clibackend}
\input{benchmarks}
\input{conclusion}

\bigskip

\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

% LocalWords:  JIT PyPy

### 2009-04-02 talk/icooolps2009/paper.tex
kill these comments, as I am not going to act on them
mark an unclear sentence, and some rewording

@@
 the machine code generation part can be simplified.
+\anto{XXX small english issue: I don't like the two consecutive ``because'',
+  I'm not even sure what is the right way to interpret the sentence}
 
@@
 to its own beginning, forming an endless loop that can only be left via a guard
-failure. The call to \texttt{f} was inlined into the trace. Of the condition in
+failure. The call to \texttt{f} was inlined into the trace. \sout{Of the condition in
 \texttt{f} the much more common \texttt{else} case was traced. The other case is
 implemented via a guard failure. This trace can then be turned into machine code
-and executed.
+and executed.}
+\anto{
+Note that the trace contains only the hot \texttt{else} case of the
+\texttt{if} test in \texttt{f}, while the other branch is implemented via a
+guard failure. This trace can then be turned into machine code and executed.
+}
 
@@
 
-\anto{XXX: should we say that virtualizables are very cool, that nobody else
-  does that and that they are vital to get good performaces with python
-  without sacrificing compatibility?} \cfbolz{no: feels a bit dishonest to not
-  describe them properly and then say that they are very cool and vital}
-
 \section{Evaluation}

### 2009-04-02 talk/icooolps2009/paper.tex
Add more references (slightly too large diff, as the reference-file is auto-generated).

@@
 To implement a dynamic language with PyPy, an interpreter for that language has
-to be written in RPython \cite{AACM-DLS07}. RPython ("Restricted Python") is a subset of Python
+to be written in RPython \cite{ancona_rpython:step_2007}. RPython ("Restricted Python") is a subset of Python
 chosen in such a way that type inference can be performed on it. The language
@@
 specialised and how are pre-determined. Another work in this direction is DyC
-\cite{grant_dyc_2000}, another runtime specializer for C. Both of these projects
+\cite{grant_dyc:expressive_2000}, another runtime specializer for C. Both of these projects
 have a similar problem as DynamoRIO.  Targeting the C language makes

### 2009-04-02 talk/icooolps2009/paper.tex
Kill one comment by maciek that I addressed, and one that I think is wrong.

@@
 
-\fijal{RPython tracer operates on a level which is slightly higher level than
-of the underlaying C. To be more precise: it operates on exception-transformed,
-non-gc-transformed graphs with an extra bit of abstraction inversion that
-finds out operations on loops and dicts, not sure how to express this}
-
 Let's look at a small example. Take the following (slightly contrived) RPython
@@
 
-\fijal{This is wrong. Without virtuals there is also at most one assembler
-loop per user loop. If it has more branches, we enter the loop as usual and
-then we create a bridge for a new situation}
 The condition for reusing already existing machine code needs to be adapted to

### 2009-04-02 talk/icooolps2009-dotnet/clibackend.tex
copy the introduction from the ecoop paper, and adapt it to this one. Add few XXXs

--- 2008-12-21 talk/ecoop2009/clibackend.tex
@@
-\section{The CLI backend for the Generated JIT}
+\section{The CLI JIT backend}
 \label{sec:clibackend}
@@
 For a large part, implementing the CLI backend is easy and straightforward, as
-there is a close corrispondence between most of the operations used by
+there is a close correspondence between most of the operations used by
 frontend's flowgraphs and the CLI instructions.  Thus, we will not dig into
@@
 When a new case is added to a flexswitch, the backend generates the new blocks
-into a new single method.  The newly created method is pointed by a
+into a new single method.  The newly created method is pointed to by a
 delegate\footnote{\emph{Delegates} are the .NET equivalent of function
@@
 outgoing link of the block dynamically added in the right-hand side
-picture of Figure~\ref{flexswitch-fig}. How it is possible to jump in 
+picture of Figure~\ref{flexswitch-fig}. How is it possible to jump into
 the middle of a method?

### 2009-04-02 talk/icooolps2009-dotnet/intro.tex
copy the introduction from the ecoop paper, and adapt it to this one. Add few XXXs

--- 2009-01-19 talk/ecoop2009/intro.tex
@@
 
-PyPy's approach \cite{RiBo07_223} is to automatize the generation of JIT compilers in order
+PyPy's approach \cite{RiBo07_223} is to automatize the generation of 
+specializing JIT compilers in order
 to minimize the effort required to get a fast implementation of a
@@
 to provide an interpreter with some manual annotations which hint
-the generator how interpretation and JIT compilation has to be interleaved.
+the generator how interpretation and JIT compilation has to be interleaved \cite{PyPyJIT09}.
 
@@
 
-The contributions of this paper are twofold: (1) \emph{promotion} is a
-generalization of polymorphic inline caches that make partial evaluation
-practical for dynamic languages; (2) we demonstrated that the idea of
+The main contribution of this paper is to demonstrate that the idea of
 \emph{JIT layering} can give good results, as dynamic languages can be even
@@
 
-\subsection{PyPy and RPython}
-
-\commentout{
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=.6\textwidth]{diagram0}
-\caption{PyPy infrastracture for generating an interpreter of a
-  language $L$ for several platforms}\label{pypy-fig}
-\end{center}
-\end{figure}
-}
+\subsection{Overview of PyPy}
 
@@
 give a high-level overview of how the JIT-generation process works. More
-details will be given in subsequent sections.
+details can be found in \cite{PyPyJIT} and \cite{PyPyJIT09}.
 
@@
 
+\anto{maybe we can avoid to talk about hints?}
+
+\commentout{
 It is important to distinguish between three distinct phases of execution:
@@
 the user wants to run some program.
+}
 
@@
 again.  As a result, \textbf{runtime and compile-time are continuously
-interleaved}. The primitive to do this sort of interleaving is called promotion,
+interleaved}.
+
+\commentout{
+The primitive to do this sort of interleaving is called promotion,
 it is described in Section \ref{sec:promotion}.
@@
 unnecessary allocations. This is described in Section \ref{sec:virtuals}
+}
 
@@
 \ref{sec:clibackend} explains how it is possible to simulate this behaviour.
+
+\anto{XXX: we need to explain what is a flexswitch!}

### 2009-04-02 talk/icooolps2009-dotnet/paper.tex
copy the introduction from the ecoop paper, and adapt it to this one. Add few XXXs

@@
 \usepackage{listings}
+
+\lstset{language=Python,
+        basicstyle=\scriptsize\ttfamily,
+        keywordstyle=\color{blue}, % I couldn't find a way to make chars both bold and tt
+        frame=none,
+        stringstyle=\color{blue},
+        fancyvrb=true,
+        xleftmargin=20pt,xrightmargin=20pt,
+        showstringspaces=false}
 
@@
 
-\section{Introduction}
-XXX
-\cite{PyPyJIT09}
-
-\section{PyPy and automatic JIT compiler generation}
-XXX
-
-\include{clibackend}
+\input{intro}
+\input{clibackend}
 \input{benchmarks}

### 2009-04-02 talk/icooolps2009/paper.tex
Fix Anto's remarks, add two more references.

@@
 tracing JIT compilers \cite{XXX}. Writing a tracing JIT compiler is relatively
-simple, because it can be added to an existing interpreter for a language,
-because the interpreter takes over some of the functionality of the compiler and
+simple. It can be added to an existing interpreter for a language,
+the interpreter takes over some of the functionality of the compiler and
 the machine code generation part can be simplified.
-\anto{XXX small english issue: I don't like the two consecutive ``because'',
-  I'm not even sure what is the right way to interpret the sentence}
 
@@
 to its own beginning, forming an endless loop that can only be left via a guard
-failure. The call to \texttt{f} was inlined into the trace. \sout{Of the condition in
-\texttt{f} the much more common \texttt{else} case was traced. The other case is
-implemented via a guard failure. This trace can then be turned into machine code
-and executed.}
-\anto{
-Note that the trace contains only the hot \texttt{else} case of the
-\texttt{if} test in \texttt{f}, while the other branch is implemented via a
-guard failure. This trace can then be turned into machine code and executed.
-}
+failure. The call to \texttt{f} was inlined into the trace. Note that the trace
+contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
+while the other branch is implemented via a guard failure. This trace can then
+be turned into machine code and executed.
 
@@
 done to do partial evaluation at runtime. One of the earliest works on runtime
-specialisation is Tempo for C \cite{XXX}. However, it is essentially a normal
+specialisation is Tempo for C \cite{consel_general_1996, consel_uniform_1996}.
+However, it is essentially a normal
 partial evaluator ``packaged as a library''; decisions about what can be
@@
 small dynamic language based on lambda-calculus. There is some work by one of
-the authors to implement a dynamic partial evaluator for Prolog \cite{XXX}.
+the authors to implement a dynamic partial evaluator for Prolog
+\cite{carl_friedrich_bolz_automatic_2008}.
 

### 2009-04-02 talk/icooolps2009-dotnet/clibackend.tex
try to make images and listings fit into a column

@@
 \begin{center}
-\includegraphics[height=5cm]{flexswitch1}
-\includegraphics[height=5cm]{flexswitch2}
+\includegraphics[height=4.5cm]{flexswitch1}
+\includegraphics[height=4.5cm]{flexswitch2}
 \caption{An example of a flexswitch evolution: in the picture on the
@@
 \begin{lstlisting}[language={[Sharp]C}] 
-public class IntLowLevelFlexSwitch:BaseLowLevelFlexSwitch {
+public class IntLowLevelFlexSwitch: 
+                        BaseLowLevelFlexSwitch {
   public uint default_blockid = 0xFFFFFFFF;
@@
   public int[] values = new int[4];
-  public FlexSwitchCase[] cases = new FlexSwitchCase[4];
+  public FlexSwitchCase[] cases = 
+                          new FlexSwitchCase[4];
 

### 2009-04-02 talk/icooolps2009-dotnet/paper.tex
try to make images and listings fit into a column

--- 2009-04-02 talk/icooolps2009-dotnet/paper.tex
@@
 \usepackage{listings}
+
+\lstset{language=Python,
+        basicstyle=\scriptsize\ttfamily,
+        keywordstyle=\color{blue}, % I couldn't find a way to make chars both bold and tt
+        frame=none,
+        stringstyle=\color{blue},
+        fancyvrb=true,
+        xleftmargin=10pt,xrightmargin=10pt,
+        showstringspaces=false}
 

### 2009-04-03 talk/icooolps2009-dotnet/clibackend.tex
describe what is a flexswitch and why it's useful to produce good code

@@
 However the concept of \emph{flexswitch}, as described in
-Section~\ref{sec:implementing-promotion}, does not have any direct equivalent
+Section~\ref{sec:jitgen}, does not have any direct equivalent
 in the CLI model, and it is hard to implement efficiently.

### 2009-04-03 talk/icooolps2009-dotnet/intro.tex
describe what is a flexswitch and why it's useful to produce good code

@@
 \subsection{PyPy and JIT-Generation}
+\label{sec:jitgen}
 
@@
 
+Implementing such a behavior requires a tight coupling between compile-time
+and run-time: a \emph{callback} is put in the generated code, which can invoke
+the compiler again.  When the callback is actually reached at run-time, and
+only then, the compiler resumes and uses the knowledge of the actual run-time
+value to generate more code.
+
+The new generated code is potentially different for each run-time value seen.
+This implies that the generated code needs to contain some sort of updatable
+switch, or \emph{flexswitch}, which can pick the right code path based on the
+run-time value.  Typically, the value we switch on is the runtime dynamc type
+of a value, so that the JIT compiler has all information needed to produce
+very good code for that specific case.
+
 \commentout{
@@
 \ref{sec:clibackend} explains how it is possible to simulate this behaviour.
-
-\anto{XXX: we need to explain what is a flexswitch!}

### 2009-04-03 talk/icooolps2009-dotnet/clibackend.tex
number each block and refer to them in the explanation

@@
 \begin{center}
-\includegraphics[height=4.5cm]{flexswitch1}
-\includegraphics[height=4.5cm]{flexswitch2}
+\includegraphics[height=5cm]{flexswitch1}
+\includegraphics[height=5cm]{flexswitch2}
 \caption{An example of a flexswitch evolution: in the picture on the
@@
 
-In the pictures of Figure~\ref{flexswitch-fig}, the cyan block
+In the pictures of Figure~\ref{flexswitch-fig}, block 5 (highlighted in grey)
 corresponds to a flexswitch; initially (picture on the left) 
-only the block containing the code to restart the JIT compilation
+only block 5, containing the code to restart the JIT compilation,
 is connected to the flexswitch; the picture on the right
 shows the graph after the first case has been dynamically added to the flexswitch,
-by linking the cyan block with a freshly created new block.
+by linking block 5 with the freshly created block number 7..
 

### 2009-04-03 talk/icooolps2009-dotnet/benchmarks.tex
copy the introduction from the ecoop paper, and adapt it to this one. Add few XXXs
minor fixes

@@
 \label{sec:benchmarks}
+
+\anto{XXX: we need to give an overview of TLC}
 
@@
 
+\anto{XXX: explain which temp objects are created}
 When doing plain interpretation, we need to create and destroy three temporary
@@
 same operations are performed for all experiments.
-For $n$ greater than $10^7$, we did not run the interpreted program as it would have took too
+For $n$ greater than $10^7$, we did not run the interpreted program as it would have taken too
 much time, without adding anything to the discussion.

### 2009-04-03 talk/icooolps2009-dotnet/clibackend.tex
minor fixes

@@
 corresponds to a flexswitch; initially (picture on the left) 
-only block 5, containing the code to restart the JIT compilation,
+only block 6, containing the code to restart the JIT compilation,
 is connected to the flexswitch; the picture on the right
 shows the graph after the first case has been dynamically added to the flexswitch,
-by linking block 5 with the freshly created block number 7..
+by linking block 5 with the freshly created block number 7.
 
@@
 whose target is not an initial block; consider, for instance, the
-outgoing link of the block dynamically added in the right-hand side
-picture of Figure~\ref{flexswitch-fig}. How is it possible to jump into
+outgoing link from block 7 to block 3 in Figure~\ref{flexswitch-fig}. How is it possible to jump into
 the middle of a method?
@@
     for(int i=0; i<numcases; i++)
-    if (values[i] == value) {
+      if (values[i] == value)
         return cases[i](0, args);
-    }
     return default_blockid;

### 2009-04-03 talk/icooolps2009-dotnet/intro.tex
minor fixes
Typos.

@@
 
-Finally, the low-level control flow-graphs produced by the toolchain
+Finally, the low-level control flow graphs produced by the toolchain
 can be translated to executable code for a specific platform by a
@@
 switch, or \emph{flexswitch}, which can pick the right code path based on the
-run-time value.  Typically, the value we switch on is the runtime dynamc type
+run-time value.  Typically, the value we switch on is the runtime dynamic type
 of a value, so that the JIT compiler has all information needed to produce
@@
 \ref{sec:clibackend} explains how it is possible to simulate this behaviour.
+
+\anto{XXX: mention virtuals/escape analysis?}

### 2009-04-03 talk/icooolps2009-dotnet/paper.tex
Typos.

@@
 statically typed virtual machine.  Although it is being used as a target for
-many different languages, it is hard to provide higly efficient implementation
+many different languages, it is hard to provide highly efficient implementation
 of languages whose model is too different than the one of the VM.  In
@@
 JIT compilation for .NET means to generate .NET bytecode at runtime, which
-will further compiled down to machine code by .NET's own JIT compiler.  The
+will be further compiled down to machine code by .NET's own JIT compiler.  The
 main and novel contribution of this paper is to show that this
 \emph{two-layers JIT} technique is effective, and can give very high speedups,
-up to make dynamic languages faster than C\# itself under certain
+up to making dynamic languages faster than C\# itself under certain
 circumstances.

### 2009-04-03 talk/icooolps2009-dotnet/intro.tex
Try to adapt section 1.2 to avoid repeating from the Introduction.

@@
 level virtual machine like CLI or JVM is easier than targeting a real CPU, but
-it still requires a lot of work, as IronPython, Jython, JRuby demonstrate.
+it still requires a lot of work, as
+IronPython\footnote{http://www.codeplex.com/IronPython},
+Jython\footnote{http://www.jython.org/}
+and JRuby\footnote{http://jruby.codehaus.org/} demonstrate.
 
 Moreover, writing a static compiler is often not enough to get high
-performance; IronPython and JRuby are going in the direction of JIT compiling
-specialized versions of the code depending on the actual values/types seen at
-runtime; this approach seems to work, but writing it manually requires an
-enormous effort.
+performance.  IronPython and JRuby are going in the direction of
+compiling just-in-time (JIT) specialized versions of the code depending
+on the actual values/types seen at runtime; this approach seems to work,
+but writing it manually requires an enormous effort.
 
@@
 
-The first step is to write an interpreter for the chosen language.  Since it
+The main difference between the JIT compilers generated by PyPy and the
+ones found in other projects like IronPython is that the latter compile
+code at the method granularity -- if on the one hand they can exploit
+some of the knowledge gathered at runtime (e.g.\ the types of method
+parameters), on the other hand they can do little to optimize most of
+the operations inside, because few assumptions can be made about the
+global state of the program.  The PyPy JITs, on the other hand, work at
+a sub-method granularity, as described next.
+
+When using PyPy, the first step is to write an interpreter for the chosen language.  Since it
 must be fed to the translation toolchain, the interpreter has to be written in
@@
 annotations to the interpreter, in order to teach the JIT generator which
-informations are important to know at compile-time.  Annotations are inserted
+information is important to know at compile-time.  Annotations are inserted
 as \emph{hints}, as described in section \ref{sec:hints}.
+From these hints, PyPy will statically generate an interpreter and a JIT
+compiler in a single executable (here a .NET executable).
 
@@
 
-Generating efficient compilers for dynamic languages is hard.  Since these
-languages are dynamically typed, usually the compiler does not have enough
-information to produce efficient code, but instead it has to insert a lot of
-runtime checks to select the appropriate implementation for each operation.
+\commentout{
+By emitting code at runtime, JIT compilers can exploit extra
+knowledge compared to traditional static compilers -- more than just the
+type of the function arguments.  However, special care is needed to
+choose a strategy for JIT compilation that lets the compiler take the
+best of this advantage.
+}
 
-By emitting code at runtime, JIT compilers can exploit some extra knowledge
-compared to traditional static compilers.  However, we need to take special
-care to choose a strategy for JIT compilation that lets the compiler to take
-the best of this advantage.
-
-Most JIT compilers for dynamic languages around (such as
-IronPython\footnote{http://www.codeplex.com/IronPython},
-Jython\footnote{http://www.jython.org/} or
-JRuby\footnote{http://jruby.codehaus.org/}) compile code at the method
-granularity.  If on the one hand they can exploit some of the knowledge
-gathered at runtime (e.g. the types of method parameters), on the other hand
-they can do little to optimize most of the operations inside, because few
-assumptions can be made about the global state of the program.
-
-JIT compilers generated by PyPy solve this problem by delaying the compilation
-until they know all the informations needed to generate efficient code.  If at
-some point the JIT compiler does not know about something it needs, it
-generates a callback into itself and stops execution.
+The interesting property of the generated JIT compiler is to delay the
+compilation until it knows all the information needed to generate
+efficient code.  In other words, at runtime, when the interpreter notice
+that it is useful to compile a given piece of code, it sends it to the
+JIT compiler; however, if at some point the JIT compiler does not know
+about something it needs, it generates a callback into itself and stops
+execution.
 

### 2009-04-03 talk/icooolps2009-dotnet/benchmarks.tex
briefly describe TLC

@@
 
-\anto{XXX: we need to give an overview of TLC}
+To measure the performances of the CLI JIT backend, we wrote a simple virtual
+machine for a dynamic toy languaged, called \emph{TLC}.
 
-In section \ref{sec:tlc-properties}, we saw that TLC provides most of the
-features that usually make dynamically typed language so slow, such as
-\emph{stack-based interpreter}, \emph{boxed arithmetic} and \emph{dynamic lookup} of
-methods and attributes.
+The design goal of the language is to be very simple (the interpreter of the
+full language consists of about 600 lines of RPython code) but to still have
+the typical properties of dynamic languages that make them hard to
+compile. TLC is implemented with a small interpreter that interprets a custom
+bytecode instruction set. Since our main interest is in the runtime
+performance of the interpreter, we did not implement the parser nor the
+bytecode compiler, but only the interpreter itself.
+
+Despite being very simple and minimalistic, \lstinline{TLC} is a good
+candidate as a language to test our JIT generator, as it has some of the
+properties that makes most of current dynamic languages (e.g. Python) so slow:
+
+\begin{itemize}
+
+\item \textbf{Stack based interpreter}: this kind of interpreter requires all the operands to be
+  on top of the evaluation stack.  As a consequence programs spend a lot of
+  time pushing and popping values to/from the stack, or doing other stack
+  related operations.  However, thanks to its simplicity this is still the
+  most common and preferred way to implement interpreters.
+
+\item \textbf{Boxed integers}: integer objects are internally represented as
+  an instance of the \lstinline{IntObj} class, whose field \lstinline{value}
+  contains the real value.  By having boxed integers, common arithmetic
+  operations are made very slow, because each time we want to load/store their
+  value we need to go through an extra level of indirection.  Moreover, in
+  case of a complex expression, it is necessary to create many temporary
+  objects to hold intermediate results.
+
+\item \textbf{Dynamic lookup}: attributes and methods are looked up at
+  runtime, because there is no way to know in advance if and where an object
+  have that particular attribute or method.
+\end{itemize}
 
@@
 \begin{enumerate}
-\item By plain interpretation, without any jitting.
+\item By plain interpretation, without any jitting (\emph{Interp}).
 \item With the JIT enabled: this run includes the time spent by doing the
-  compilation itself, plus the time spent by running the produced code.
+  compilation itself, plus the time spent by running the produced code (\emph{JIT}).
 \item Again with the JIT enabled, but this time the compilation has already
-  been done, so we are actually measuring how good is the code we produced.
+  been done, so we are actually measuring how good is the code we produced (\emph{JIT 2}).
 \end{enumerate}

### 2009-04-03 talk/icooolps2009/paper.tex
A minor simplification and add myself as an author

@@
        \email{arigo@tunes.org}
+\alignauthor Maciej Fijalkowski\\
+       \email{fijal@merlinux.eu}
 }
@@
 
-\fijal{Stopped reading at that point}
 Let's look at which hints would need to be applied to the example interpreter
@@
 Python, it is possible to constant-fold the \texttt{strgetitem} operation. The
-\texttt{int\_add} operations can be folded anyway.
+\texttt{int\_add} are additions of constant \texttt{pc} and a true constant,
+so they can be folded away.
 

### 2009-04-03 talk/icooolps2009-dotnet/conclusion.tex
adapt the conclusions

--- 2008-12-21 talk/ecoop2009/conclusion.tex
@@
 
-Moreover, the idea of promotion of is a generalization of \emph{Polymorphic
+Moreover, the idea of promotion is a generalization of \emph{Polymorphic
   Inline Caches} \cite{hoelzle_optimizing_1991}, as well as the idea of using
@@
 PyPy-style JIT compilers are hard to write manually, thus we chose to write a
-JIT generator.  Tracing JIT compilers \cite{gal_hotpathvm_2006} also gives
+JIT generator.  Tracing JIT compilers \cite{gal_hotpathvm_2006} also give
 good results but are much easier to write, making the need for an automatic
@@
 
-In this paper we presented PyPy's JIT compiler generator, based on partial
-evaluation techniques, which can automatically turn an interpreter into a JIT
+In this paper we gave an overview of PyPy's JIT compiler generator,
+which can automatically turn an interpreter into a JIT
 compiler, requiring the language developers to only add few \texttt{hint}s to
@@
 
-We showed that classical partial evaluation cannot remove all the overhead
-proper of dynamically typed languages, and how the new operation called
-\emph{promotion} solves the problem, by delaying compile-time until the JIT
-knows enough to produce efficient code, and by continuously intermixing
-compile-time and runtime.  Moreover, we showed that our simple but still
-practically useful technique to avoid allocation of intermediate unnecessary
-objects plays well with promotion and helps to produce even better code.
-
-Finally, we presented the CLI backend for PyPy's JIT compiler generator, whose
+Then, we presented the CLI backend for PyPy's JIT compiler generator, whose
 goal is to produce .NET bytecode at runtime.  We showed how it is possible to
 circumvent intrinsic limitations of the virtual machine to implement
-promotion.  As a result, we proved that the idea of \emph{JIT layering} is
+flexswitches.  As a result, we proved that the idea of \emph{JIT layering} is
 worth of further exploration, as it makes possible for dynamically typed
@@
 
+\anto{XXX: cite the pypy tracing paper}
+
 At the moment, some bugs and minor missing features prevent the CLI JIT
@@
 integrated into virtual machines.
+
+In particular, the \emph{Da Vinci Machine
+  Project} \footnote{http://openjdk.java.net/projects/mlvm/} is exploring and
+implementing new features to ease the implementation of dynamic languages on
+top of the JVM: some of these features, such as the new
+\emph{invokedynamic}\footnote{XXX} instruction and the \emph{tail call
+  optimization} can probably be exploited by a potential JVM backend to
+generate even more efficient code.

### 2009-04-03 talk/icooolps2009/paper.tex
it's not simple any more
Some more references. Finally take a stab at the benchmark section.

@@
 A recent approach to getting better performance for dynamic languages is that of
-tracing JIT compilers \cite{XXX}. Writing a tracing JIT compiler is relatively
+tracing JIT compilers \cite{gal_hotpathvm:effective_2006,
+mason_chang_efficient_2007}. Writing a tracing JIT compiler is relatively
 simple. It can be added to an existing interpreter for a language,
@@
 \cite{mason_chang_efficient_2007}. The technique is now
-being used by both Mozilla's TraceMonkey JavaScript VM \cite{XXX} and Adobe's
-Tamarin ActionScript VM \cite{XXX}.
+being used by both Mozilla's TraceMonkey JavaScript VM
+\cite{andreas_gal_trace-based_2009} and has been tried for Adobe's Tamarin
+ActionScript VM \cite{chang_tracing_2009}.
 
@@
 virtual machines such as the JVM where backend could generate bytecode at
-runtime). At the moment the only implemented backend is a simple 32-bit
+runtime). At the moment the only implemented backend is a 32-bit
 Intel-x86 backend.
@@
 benchmark numbers. Since the work is not finished, these benchmarks can only be
-preliminary. All benchmarking was done on a machine with a 1.4 GHz Pentium M
-processor and 1GiB RAM, using Linux 2.6.27.
+preliminary. All benchmarking was done on an otherwise idle machine with a 1.4
+GHz Pentium M processor and 1GiB RAM, using Linux 2.6.27.
+
+The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
+example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
+the square of 46340 (the smallest number so that the square still fits into a 32
+bit word) using the bytecode of Figure \ref{fig:square}. The results for various
+constellations are as follows:
+
+\begin{enumerate}
+\item The interpreter translated to C without any JIT inserted at all.
+\item The tracing JIT is enabled, but no interpreter-specific
+hints are applied. This corresponds to the trace in Figure
+\ref{fig:trace-normal}. The time includes the time it takes to trace and the
+production of the machine code, as well as the fallback interpreter to leave the
+machine code. The threshold when to consider a loop to be hot is 40 iterations.
+\item The hints as in Figure \ref{fig:tlr-full} are applied, which means the loop of
+the square function is reflected in the trace. Constant folding of green
+variables is disabled though. This corresponds to the trace in Figure
+\ref{fig:trace-no-green-folding}. XXX
+\item Same as before, but with constant folding enabled. This corresponds to the
+trace in Figure \ref{fig:trace-full}. This speeds up the square function nicely,
+making it about six times faster than the pure interpreter.
+\item Same as before, but with the threshold set so high that the tracer is
+never invoked. This measures the overhead of the profiling. For this interpreter
+the overhead seems rather large, with 50\% slowdown du to profiling. This is
+because the example interpreter needs to do one hash table lookup per loop
+iteration. For larger interpreters (e.g. the Python one) it seems likely that
+the overhead is less significant, given that many operations in Python need
+hash-table lookups themselves.
+\item Runs the whole computation on the tracing interpreter for estimating the
+involved overheads of tracing. The trace is not actually recorded (which would be a
+memory problem), so in reality the number is even higher. Due to the double
+interpretation, the overhead is huge. It remains to be seen whether that will be
+a problem for practical interpreters.
+\item For comparison, the time of running the interpreter on top of CPython
+(version 2.5.2).
+\end{enumerate}
+
+\begin{figure}
+\noindent
+\begin{tabular}{|l|r|}
+\hline
+ &ratio\tabularnewline
+\hline
+Interpreter compiled to C, no JIT &1\tabularnewline \hline
+Normal Trace Compilation &1.20\tabularnewline \hline
+Unfolding of Language Interpreter Loop &XXX\tabularnewline \hline
+Full Optimizations &0.17\tabularnewline \hline
+Profile Overhead &1.51\tabularnewline \hline
+Interpreter run by Tracing Interpreter &860.20\tabularnewline \hline
+Interpreter run by CPython &256.17\tabularnewline \hline
+\end{tabular}
+\label{fig:bench1}
+\caption{Benchmark results of example interpreter computing the square of
+46340}
+\end{figure}
 

### 2009-04-04 talk/icooolps2009-dotnet/paper.tex
some adjustments in the abstract

--- 2009-04-02 talk/icooolps2009-dotnet/paper.tex
@@
 \begin{abstract}
-The CLI, i.e. the virtual machine at the core of the .NET environment, is a
-statically typed virtual machine.  Although it is being used as a target for
-many different languages, it is hard to provide higly efficient implementation
-of languages whose model is too different than the one of the VM.  In
-particular, dynamically typed languages are typically order of magnitude
-slower than, e.g., C\#.
+The Common Language Infrastructure (CLI) is a virtual machine expressly
+designed for implementing statically typed languages as C\#, therefore
+programs written in dynamically typed languages are typically much slower than C\# when executed on .NET.
 
@@
   slower than C\#?}
+\davide{We can try writing something in the conclusion}
 
@@
 information to generate quite efficient code.  Unfortunately, writing a JIT
-compiler is far from being simple.  In this paper we report our positive
+compiler is far from being simple.
+
+In this paper we report our positive
 experience with automatic generation of JIT compilers as supported by the PyPy
 infrastructure, by focusing on JIT compilation for .NET.
+Following this approach, we have in fact added a second layer of JIT compilation, by allowing dynamic generation of more efficient .NET bytecode, which
+in turn can be compiled to machine code by the .NET JIT compiler.
 
-JIT compilation for .NET means to generate .NET bytecode at runtime, which
-will further compiled down to machine code by .NET's own JIT compiler.  The
-main and novel contribution of this paper is to show that this
-\emph{two-layers JIT} technique is effective, and can give very high speedups,
-up to make dynamic languages faster than C\# itself under certain
-circumstances.
+The main and novel contribution of this paper is to show that this
+\emph{two-layers JIT} technique is effective, since programs written in dynamic languages 
+can run on .NET as fast as (and in some cases even faster than) the equivalent C\# programs.
 

### 2009-04-04 talk/icooolps2009/paper.tex
fix a few things, roughly start with a conclusion

@@
 
+% compressing itemize env, in case we need it
+\newenvironment{zitemize}% zero - line spacing itemize environment
+   {\begin{list}{--}{
+   \setlength{\itemsep}{0 pt}
+   \setlength{\parsep}{0 pt}
+   \setlength{\topsep} {0 pt} }}% the end stuff
+   {\end{list}}
+
 \begin{document}
@@
        \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
-       \affaddr{Softwaretechnik und Programmiersprachen}\\
-       \affaddr{Institut für Informatik}\\ 
-       \affaddr{Universitätsstra{\ss}e 1}\\
-       \affaddr{D-40225 Düsseldorf}\\
+       \affaddr{STUPS Group}\\
        \affaddr{Deutschland}\\
@@
 code:
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
-
+}
 To trace this, a bytecode form of these functions needs to be introduced that
@@
 look as follows:
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
-
+}
 The operations in this sequence are operations of the mentioned intermediate
@@
 unlikely that the same particular opcode is executed many times in a row.
-
 \begin{figure}
@@
 \end{figure}
-
 \begin{figure}
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 \caption{Example bytecode: Compute the square of the accumulator}
@@
 as long as the operations are side-effect free. Since strings are immutable in
-Python, it is possible to constant-fold the \texttt{strgetitem} operation. The
-\texttt{int\_add} are additions of constant \texttt{pc} and a true constant,
-so they can be folded away.
+RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
+\texttt{int\_add} are additions of the green variable \texttt{pc} and a true
+constant, so they can be folded away as well.
 
@@
 never invoked. This measures the overhead of the profiling. For this interpreter
-the overhead seems rather large, with 50\% slowdown du to profiling. This is
+the overhead seems rather large, with 50\% slowdown due to profiling. This is
 because the example interpreter needs to do one hash table lookup per loop
@@
 
+We have shown techniques for improving the results when applying a tracing
+JIT to an interpreter. Our first benchmarks indicate that these techniques work
+and first experiments with PyPy's Python interpreter make it seems likely that
+they can be scaled up to realistic examples.
+
+Of course there is a lot of work still left to do. Various optimizations are not
+quite finished. Both tracing and leaving machine code is very slow due to a
+double interpretation overhead and we might need techniques for improving those.
+Furthermore we need to apply the JIT to the various interpreters that are
+written with PyPy (like the SPy-VM, a Smalltalk implementation \cite{XXX} or
+PyGirl, a Gameboy emulator \cite{XXX}) to evaluate how widely applicable the
+described techniques are.
+
+XXX would like a nice last sentence
 %\begin{verbatim}

### 2009-04-04 talk/icooolps2009-dotnet/intro.tex
add an XXX that I can't solve now
briefly describe TLC
started working at the intro

@@
 \section{Introduction}
-The easiest way to implement a dynamic language such as Python is to write an
-interpreter; however, interpreters are slow.
+Implementing a dynamic language such as Python with a compiler rather than with an interpreter improves performances at the cost of
+an increasing complexity. Furthermore, generating code for high level virtual machines like CLI or JVM enhances portability and inter-operability.
 
-The alternative is to write a compiler; writing a compiler that targets a high
-level virtual machine like CLI or JVM is easier than targeting a real CPU, but
+Writing a compiler that targets a high CLI or JVM is easier than targeting a real CPU, but
 it still requires a lot of work, as
@@
 faster than their static counterparts.
+
+\anto{XXX: we first say that IronPython\&co. does JIT compilation, then we say
+  we are the first to do JIT layering.  This seems a bit strange, though at
+  the moment I can't think of any better way to word this concept}
 

### 2009-04-06 talk/icooolps2009/paper.tex
Lots of fixes here and there.

@@
 \newboolean{showcomments}
-\setboolean{showcomments}{true}
+\setboolean{showcomments}{false}
 \ifthenelse{\boolean{showcomments}}
@@
 
+\textfloatsep 12pt plus 2pt minus 4pt
+
 \begin{document}
@@
 \begin{abstract}
-In this paper we describe the ongoing research in the PyPy project to write a
-JIT compiler that is automatically adapted to various languages, given an
-interpreter for that language. This is achieved with the help of a slightly
-adapted tracing JIT compiler in combination with some hints by the author of the
-interpreter.  XXX
+We present techniques for improving the results when a tracing JIT compiler is
+applied to an interpreter. An unmodified tracing JIT performs not as well as one
+would hope when the compiled program is itself a bytecode interpreter. We
+examine why that is the case, and how matters can be improved by adding hints to
+the interpreter, that help the tracing JIT to improve the results. We evaluate
+the techniques by using them both on a small example as well as on a full Python
+interpreter. This work has been done in the context of the PyPy project.
 
@@
 
-XXX write somewhere that one problem of using tracing JITs for dynamic languages
-is that dynamic languages have very complex bytecodes
-
 \section{Introduction}
 
-Dynamic languages, rise in popularity, bla bla XXX
+Dynamic languages have seen a steady rise in popularity in recent years.
+JavaScript is increasingly being used to implement full-scale applications
+running in browser, whereas other dynamic languages (such as Ruby, Perl, Python,
+PHP) are used for the server side of many web sites, as well as in areas
+unrelated to the web.
 
 One of the often-cited drawbacks of dynamic languages is the performance
-penalties they impose. Typically they are slower than statically typed languages
-\cite{XXX}. Even though there has been a lot of research into improving the
-performance of dynamic languages \cite{XXX}, those techniques are not as widely
+penalties they impose. Typically they are slower than statically typed
+languages. Even though there has been a lot of research into improving the
+performance of dynamic languages (in the SELF project, to name just one example
+\cite{XXX}), those techniques are not as widely
 used as one would expect. Many dynamic language implementations use completely
@@
 The contributions of this paper are:
-\begin{itemize}
-\item Techniques for improving the generated code when applying a tracing JIT to
-an interpreter
-\item 
-\end{itemize}
+\vspace{-0.3cm}
+\begin{zitemize}
+\item Applying a tracing JIT compiler to an interpreter.
+\item Finding techniques for improving the generated code.
+\item Integrating
+\end{zitemize}
 
@@
 Tracing JITs are built on the following basic assumptions:
-
-\begin{itemize}
+\vspace{-0.3cm}
+\begin{zitemize}
  \item programs spend most of their runtime in loops
  \item several iterations of the same loop are likely to take similar code paths
-\end{itemize}
+\end{zitemize}
 
@@
 Typically, programs executed by a tracing VMs goes through various phases:
-
-\begin{itemize}
+\vspace{-0.3cm}
+\begin{zitemize}
 \item Interpretation/profiling
@@
 \item Execution of the generated code
-\end{itemize}
+\end{zitemize}
 
@@
 we immediately quit from the machine code and continue the execution by falling
-ways.
+back to interpretation.
 
@@
 earlier in the trace. If this happens, the trace recorded corresponds to a loop
-in the interpreted program that the tracing interpreter is running. At this point, this loop
+in the interpreted program. At this point, this loop
 is turned into machine code by taking the trace and making machine code versions
 of all the operations in it. The machine code can then be immediately executed,
-starting from the second iteration of the loop,
-as it represents exactly the loop that was being interpreted so far.
+starting from the next iteration of the loop, as the machine code represents
+exactly the loop that was being interpreted so far.
 
@@
 "typical" example of possible paths (which is statistically likely). Of course
-it is possible that later another path through the loop is taken, therefore the
-machine code will contain \emph{guards}, which check that the path is still the same.
-If a guard fails during execution of the machine code, the machine code is left
-and execution falls back to using interpretation (there are more complex
-mechanisms in place to still produce more code for the cases of guard failures,
-but they are of no importance for this paper).
+it is possible that later another path through the loop is taken, in which case
+one of the guards that were put into the machine code will fail. There are more
+complex mechanisms in place to still produce more code for the cases of guard
+failures \cite{XXX}, but they are orthogonal to the issues discussed in this
+paper.
 
-It is important to understand when the tracer considers a loop in the trace to
-be closed. This happens when the \emph{position key} is the same as at an earlier
+It is important to understand how the tracer recognizes that the trace it
+recorded so far corresponds to a loop.
+This happens when the \emph{position key} is the same as at an earlier
 point. The position key describes the position of the execution of the program,
@@
 }
+\vspace{-0.4cm}
 To trace this, a bytecode form of these functions needs to be introduced that
@@
 }
+\vspace{-0.4cm}
 The operations in this sequence are operations of the mentioned intermediate
 representation (e.g. note that the generic modulo and equality operations in the
-function above have been recognized to always work on integers and are thus
+function above have been recognized to always take integers as arguments and are thus
 rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
-operations that were executed, is in SSA-form \cite{XXX} and ends with a jump
+operations that were executed, is in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
 to its own beginning, forming an endless loop that can only be left via a guard
@@
 interpreter}. On the other hand, there is the interpreter that is running the
-users programs, which we will call the \emph{language interpreter}. In the
+user's programs, which we will call the \emph{language interpreter}. In the
 following, we will assume that the language interpreter is bytecode-based. The
@@
 of the language interpreter is its bytecode dispatch loop (for many simple
-interpreters it is also the only hot loops).  Tracing one iteration of this
+interpreters it is also the only hot loop).  Tracing one iteration of this
 loop means that
@@
 \input{code/tlr-paper.py}
+\vspace{-0.4cm}
 \caption{A very simple bytecode interpreter with registers and an accumulator.}
@@
 }
+\vspace{-0.4cm}
 \caption{Example bytecode: Compute the square of the accumulator}
@@
 simple bytecode interpreter with 256 registers and an accumulator. The
-\texttt{bytecode} argument is a string of bytes and all register and the
-accumulator are integers. A simple program for this interpreter that computes
+\texttt{bytecode} argument is a string of bytes, all register and the
+accumulator are integers. A program for this interpreter that computes
 the square of the accumulator is shown in Figure \ref{fig:square}. If the
@@
 \input{code/normal-tracing.txt}
+\vspace{-0.4cm}
 \caption{Trace when executing the \texttt{DECR\_A} opcode}
@@
 bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
-corresponds to \emph{user loop}. User loops
-occur when the program counter of the language interpreter has the
+corresponds to a \emph{user loop}. User loops
+occur when the program counter of the \emph{language interpreter} has the
 same value several times. This program counter is typically stored in one or several
@@
 dispatch loop of the language interpreter. They correspond to different
-unrollings and paths of that loop though. To figure out which of them to use
+unrollings and paths through that loop though. To figure out which of them to use
 when trying to enter assembler code again, the program counter of the language
@@
 \input{code/tlr-paper-full.py}
+\vspace{-0.4cm}
 \caption{Simple bytecode interpreter with hints applied}
@@
 
-Let's look at which hints would need to be applied to the example interpreter
+Let's look at how hints would need to be applied to the example interpreter
 from Figure \ref{fig:tlr-basic}. The basic thing needed to apply hints is a
@@
 \input{code/no-green-folding.txt}
+\vspace{-0.4cm}
 \caption{Trace when executing the Square function of Figure \ref{fig:square},
@@
 \end{figure}
-
-XXX summarize at which points the tracing interpreter needed changing
-XXX all changes only to the position key and when to enter/leave the tracer!
-XXX tracing remains essentially the same
 
@@
 RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
-\texttt{int\_add} are additions of the green variable \texttt{pc} and a true
-constant, so they can be folded away as well.
+\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
+number, so they can be folded away as well.
 
@@
 bad anyway (in fact we have an experimental optimization that does exactly that,
-but it is not finished).
-
-\anto{XXX I propose to show also the trace with the malloc removal enabled, as it
-  is much nicer to see. Maybe we can say that the experimental optimization we
-  are working on would generate this and that} \cfbolz{This example is not about
-  mallocs! There are no allocations in the loop. The fix would be to use
-  maciek's lazy list stuff (or whatever it's called) which is disabled at the
-  moment}
+but it is not finished).  Once we get this optimized trace, we can pass it to
+the \emph{JIT backend}, which generates the correspondent machine code.
 
@@
 \input{code/full.txt}
+\vspace{-0.4cm}
 \caption{Trace when executing the Square function of Figure \ref{fig:square},
@@
 \end{figure}
-
-Once we get this highly optimized trace, we can pass it to the \emph{JIT
-backend}, which generates the correspondent machine code.
 
@@
 restriction in the future. A classical tracing JIT will
-interpret the program it is running until a common loop is identified, at which
+interpret the program it is running until a hot loop is identified, at which
 point tracing and ultimately assembler generation starts. The tracing JIT in
@@
 RPython programs are statically translatable to C anyway. This means that interpreting the
-language interpreter before a common loop is found is clearly not desirable,
+language interpreter before a hot loop is found is clearly not desirable,
 since the overhead of this double-interpretation would be significantly too big
@@
 What is done instead is that the language interpreter keeps running as a C
-program, until a common loop in the user program is found. To identify loops the
+program, until a hot loop in the user program is found. To identify loops the
 C version of the language interpreter is generated in such a way that at the
@@
 the other hand as bytecode for the tracing interpreter. It also means that
-tracing is costly as it incurs exactly a double interpretation overhead.
+tracing is costly as it incurs a double interpretation overhead.
 
@@
 
-\textbf{Trace Trees:} This paper ignored the problem of guards that fail in a
-large percentage of cases because there are several equally likely paths through
-a loop. Just falling back to interpretation in this case is not practicable.
+\textbf{Trace Trees:} This paper ignored the problem of guards that fail often
+because there are several equally likely paths through
+a loop. Always falling back to interpretation in this case is not practicable.
 Therefore, if we find a guard that fails often enough, we start tracing from
-there and produce efficient machine code for that case, instead of alwayas
+there and produce efficient machine code for that case, instead of always
 falling back to interpretation.
@@
 benchmark numbers. Since the work is not finished, these benchmarks can only be
-preliminary. All benchmarking was done on an otherwise idle machine with a 1.4
-GHz Pentium M processor and 1GiB RAM, using Linux 2.6.27.
+preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
+GHz Pentium M processor and 1GiB RAM, using Linux 2.6.27. All benchmarks where
+run 50 times, each in a newly started process. The first run was ignored. The
+final numbers were reached by computing the average of all other runs, the
+confidence intervals were computed using a 95\% confidence level. All times
+include the running of the tracer and machine code production to measure how
+costly those are.
 
@@
 
-\begin{enumerate}
-\item The interpreter translated to C without any JIT inserted at all.
-\item The tracing JIT is enabled, but no interpreter-specific
+\textbf{Benchmark 1:} The interpreter translated to C without any JIT inserted at all.
+
+\textbf{Benchmark 2:} The tracing JIT is enabled, but no interpreter-specific
 hints are applied. This corresponds to the trace in Figure
-\ref{fig:trace-normal}. The time includes the time it takes to trace and the
-production of the machine code, as well as the fallback interpreter to leave the
-machine code. The threshold when to consider a loop to be hot is 40 iterations.
-\item The hints as in Figure \ref{fig:tlr-full} are applied, which means the loop of
+\ref{fig:trace-normal}.  The threshold when to consider a loop to be hot is 40
+iterations.  As expected, this is not faster than the previous number. It is
+even quite a bit slower, probably due to the overheads of the JIT, as well as
+non-optimal generated machine code.
+
+\textbf{Benchmark 3:} The hints as in Figure \ref{fig:tlr-full} are applied, which means the loop of
 the square function is reflected in the trace. Constant folding of green
 variables is disabled though. This corresponds to the trace in Figure
-\ref{fig:trace-no-green-folding}. XXX
-\item Same as before, but with constant folding enabled. This corresponds to the
+\ref{fig:trace-no-green-folding}. This by alone brings no improvement over the
+previous case.
+
+\textbf{Benchmark 4:} Same as before, but with constant folding enabled. This corresponds to the
 trace in Figure \ref{fig:trace-full}. This speeds up the square function nicely,
 making it about six times faster than the pure interpreter.
-\item Same as before, but with the threshold set so high that the tracer is
-never invoked. This measures the overhead of the profiling. For this interpreter
-the overhead seems rather large, with 50\% slowdown due to profiling. This is
+
+\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
+never invoked to measure the overhead of the profiling. For this interpreter
+it to be rather large, with 50\% slowdown due to profiling. This is
 because the example interpreter needs to do one hash table lookup per loop
@@
 hash-table lookups themselves.
-\item Runs the whole computation on the tracing interpreter for estimating the
+
+\textbf{Benchmark 6:} Runs the whole computation on the tracing interpreter for estimating the
 involved overheads of tracing. The trace is not actually recorded (which would be a
@@
 a problem for practical interpreters.
-\item For comparison, the time of running the interpreter on top of CPython
+
+\textbf{Benchmark 7:} For comparison, the time of running the interpreter on top of CPython
 (version 2.5.2).
-\end{enumerate}
 
@@
 \noindent
+{\small
 \begin{tabular}{|l|r|}
@@
 \end{tabular}
+}
 \label{fig:bench1}
@@
 
-%- benchmarks
-%    - running example
-%    - gameboy?
+XXX insert some Python benchmarks
 
@@
 tracer produce better results has been tried before in the context of the DynamoRIO
-project \cite{sullivan_dynamic_2003}. This work is conceptually very close to
-ours. They achieve the same unrolling of the interpreter loop so that the
+project \cite{sullivan_dynamic_2003}, which has been a great inspiration for our
+work. They achieve the same unrolling of the interpreter loop so that the
 unrolled version corresponds to the loops in the user program. However the
@@
 assembler level it is not really visible anymore that e.g. a bytecode string is
-really immutable. Also more advanced optimizations like allocation removal would
+immutable. Also more advanced optimizations like allocation removal would
 not be possible with that approach.
@@
 The standard approach for automatically producing a compiler for a programming
-language given an interpreter for it is that of partial evaluation \cite{XXX},
-\cite{XXX}. Conceptually there are some similarities to our work. In partial
+language given an interpreter for it is that of partial evaluation
+\cite{futamura_partial_1999, jones_partial_1993}. Conceptually there are some
+similarities to our work. In partial
 evaluation some arguments of the interpreter function are known (static) while
@@
 the authors to implement a dynamic partial evaluator for Prolog
-\cite{carl_friedrich_bolz_automatic_2008}.
-
-XXX what else?
-
-\anto{I would cite ourselves (maybe the JIT technical report?) and maybe
-  psyco}
+\cite{carl_friedrich_bolz_automatic_2008}. There are also experiments within the
+PyPy project to use dynamic partial evaluation for automatically generating JIT
+compilers out of interpreters \cite{armin_rigo_jit_2007}. So far those have not been as
+successful as we would like and it seems likely that they will be supplanted
+with the work on tracing JITs described here.
 
@@
 
-We have shown techniques for improving the results when applying a tracing
+We have shown techniques for making it practical to apply a tracing
 JIT to an interpreter. Our first benchmarks indicate that these techniques work
-and first experiments with PyPy's Python interpreter make it seems likely that
-they can be scaled up to realistic examples.
+really well on small interpreters and first experiments with PyPy's Python
+interpreter make it seems likely that they can be scaled up to realistic
+examples.
 
@@
 Furthermore we need to apply the JIT to the various interpreters that are
-written with PyPy (like the SPy-VM, a Smalltalk implementation \cite{XXX} or
-PyGirl, a Gameboy emulator \cite{XXX}) to evaluate how widely applicable the
-described techniques are.
+written with PyPy to evaluate how widely applicable the described techniques
+are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
+implementation \cite{bolz_back_2008}, a Prolog interpreter or PyGirl, a Gameboy
+emulator \cite{XXX}; but also less immediately obvious ones, like Python's
+regular expression engine.
 
-XXX would like a nice last sentence
+If these experiments are successful we hope that we can reach a point where it
+becomes unnecessary to write a language specific JIT compiler and just apply a
+couple of hints to the interpreter to get reasonably good performance with
+relatively little effort.
+
 %\begin{verbatim}
@@
 %   are accessible to the tracer
-\cite{bolz_back_2008}
-
-\bigskip
 

### 2009-04-06 talk/icooolps2009-dotnet/intro.tex
fix an XXX

@@
 
-Writing a compiler that targets a high CLI or JVM is easier than targeting a real CPU, but
+Writing a compiler that targets the CLI or JVM is easier than targeting a real CPU, but
 it still requires a lot of work, as
@@
 Moreover, writing a static compiler is often not enough to get high
-performance.  IronPython and JRuby are going in the direction of
-compiling just-in-time (JIT) specialized versions of the code depending
-on the actual values/types seen at runtime; this approach seems to work,
-but writing it manually requires an enormous effort.
+performance.  IronPython and Jython compile code lazily to save memory, but
+since they do not exploit the extra information potentially available at
+runtime, it is more a delayed static compilation than a true JIT one.
+\anto{XXX: someone review the english of the last paragraph, please :-)}
 
@@
 faster than their static counterparts.
-
-\anto{XXX: we first say that IronPython\&co. does JIT compilation, then we say
-  we are the first to do JIT layering.  This seems a bit strange, though at
-  the moment I can't think of any better way to word this concept}
 

### 2009-04-06 talk/icooolps2009-dotnet/intro.tex
remove commented out text, a redundant paragraph and an XXX

@@
 annotations to the interpreter, in order to teach the JIT generator which
-information is important to know at compile-time.  Annotations are inserted
-as \emph{hints}, as described in section \ref{sec:hints}.
-From these hints, PyPy will statically generate an interpreter and a JIT
+information is important to know at compile-time.  
+From these annotations, PyPy will statically generate an interpreter and a JIT
 compiler in a single executable (here a .NET executable).
-
-\anto{maybe we can avoid to talk about hints?}
-
-\commentout{
-It is important to distinguish between three distinct phases of execution:
-
-\begin{enumerate}
-\item \textbf{Translation-time}: when the translation toolchain runs and the
-  JIT compiler is automatically generated from the interpreter.  The result is
-  an executable that can be used to run programs written in the chosen
-  language.
-\item \textbf{Compile-time}: when the JIT compiler runs, generating executable
-  code on the fly.
-\item \textbf{Runtime}: when the code generated at compile-time is executed.
-\end{enumerate}
-
-Note that in this schema translation-time happens only once, on the
-developer's machine.  By contrast, compile-time and runtime happens every time
-the user wants to run some program.
-}
-
-\commentout{
-By emitting code at runtime, JIT compilers can exploit extra
-knowledge compared to traditional static compilers -- more than just the
-type of the function arguments.  However, special care is needed to
-choose a strategy for JIT compilation that lets the compiler take the
-best of this advantage.
-}
 
@@
 JIT compiler; however, if at some point the JIT compiler does not know
-about something it needs, it generates a callback into itself and stops
+about something it needs, it generates a \emph{callback} into itself and stops
 execution.
@@
 
-Implementing such a behavior requires a tight coupling between compile-time
-and run-time: a \emph{callback} is put in the generated code, which can invoke
-the compiler again.  When the callback is actually reached at run-time, and
-only then, the compiler resumes and uses the knowledge of the actual run-time
-value to generate more code.
-
-The new generated code is potentially different for each run-time value seen.
+Potentially, the JIT compiler generates new code for each different run-time
+value seen in variables it is interested in.
 This implies that the generated code needs to contain some sort of updatable
@@
 
-\commentout{
-The primitive to do this sort of interleaving is called promotion,
-it is described in Section \ref{sec:promotion}.
-
-One of the most important optimizations that the generated JIT does is removing
-unnecessary allocations. This is described in Section \ref{sec:virtuals}
-}
-
 Modifying the old code to link to the newly generated one is very challenging on
@@
 \ref{sec:clibackend} explains how it is possible to simulate this behaviour.
-
-\anto{XXX: mention virtuals/escape analysis?}

### 2009-04-06 talk/icooolps2009-dotnet/benchmarks.tex
fix another XXX, and some other small fixes

@@
 To measure the performances of the CLI JIT backend, we wrote a simple virtual
-machine for a dynamic toy languaged, called \emph{TLC}.
+machine for a dynamic toy language, called \emph{TLC}.
 
@@
 
-\anto{XXX: explain which temp objects are created}
 When doing plain interpretation, we need to create and destroy three temporary
-objects at each iteration.  By contrast, the code generated by the JIT does
+objects (the results of each operation) at each iteration.  By contrast, the code generated by the JIT does
 much better.  At the first iteration, the classes of the two operands of the
-multiplication are promoted; then, the JIT compiler knows that both are
-integers, so it can inline the code to compute the result.  Moreover, it can
-\emph{virtualize} (see Section \ref{sec:virtuals}) all the temporary objects, because they never escape from
+multiplication go through a flexswitch; then, the JIT compiler knows that both are
+integers, so it can inline the code to compute the result.  Moreover, thanks to escape analysis,
+it can remove the allocation of all the temporary objects, because they never escape from
 the inner loop.  The same remarks apply to the other two operations inside

### 2009-04-06 talk/icooolps2009-dotnet/clibackend.tex
fix another XXX, and some other small fixes

@@
 has been already loaded and linked, therefore the simplest approach
-taken for low level architectures does not work for higher level 
-virtual machines as those for .NET and Java.
+taken for low level architectures does not work.
 

### 2009-04-06 talk/icooolps2009-dotnet/conclusion.tex
adapt che conclusions to the new paper

@@
 
-Promotion is a concept that we have already explored in other contexts. Psyco is
+Flexswitches are closely related to the concept of \emph{promotion}, as
+described by \cite{PyPyJIT}, \cite{PyPyJIT09}.
+Psyco is
 a run-time specialiser for Python that uses promotion (called ``unlift'' in
@@
 language with boxing.  Another difference is that tracing JITs concentrate on
-loops, which makes them produce a lot less code.  This issue will be addressed
-by future research in PyPy.
+loops, which makes them produce a lot less code.  This issue is being addressed
+by current research in PyPy \cite{PyPyTracing}.
 
@@
 
-There has been an enormous amount of work on partial evaluation for compiler
-generation. A good introduction is given in \cite{Jones:peval}. However, most of
-it is for generating ahead-of-time compilers, which cannot produce very good
-performance results for dynamic languages.
+IronPython and Jython are two popular implementations of Python for,
+respectively, the CLI and the JVM, whose approach differs fundamentally from
+PyPy.  The source code of PyPy contains a Python interpreter, which the JIT
+compiler is automatically generated from: the resulting executable contains
+both the interpreter and the compiler, so that it is possible to compile only
+the desired parts of the program.  On the other hand, both IronPython and
+Jython implements only the compiler: both compile code lazily (when a Python
+module is loaded), but since they do not exploit the extra information
+potentially available at runtime, it is more a delayed static compilation than
+a true JIT one.  As a result, they run Python programs much slower than their
+equivalent written in
+C\#\footnote{http://shootout.alioth.debian.org/gp4/\\benchmark.php?test=all\&lang=iron\&lang2=csharp}
+or
+Java\footnote{http://blog.dhananjaynene.com/2008/07/performance-comparison-c-java-python-ruby-jython-jruby-groovy/}.
 
-However, there is also some research on runtime partial evaluation. One of the
-earliest examples is Tempo for C
-\cite{DBLP:conf/popl/ConselN96,DBLP:conf/dagstuhl/ConselHNNV96}. However, it is
-essentially an offline specializer ``packaged as a library''; decisions about
-what can be specialized and how are pre-determined.
+The \emph{Dynamic Language Runtime}\footnote{http://www.codeplex.com/dlr}
+(DLR) is a library designed to ease the implementation of dynamic languages
+for .NET: IronPython is based on the DLR, so the same remarks apply.
 
-Another work in this direction is DyC \cite{grant_dyc_2000}, another runtime
-specializer for C.  Specialization decisions are also pre-determined, but
-``polyvariant program-point specialization'' gives a coarse-grained equivalent
-of our promotion.  Targeting the C language makes higher-level specialization
-difficult, though (e.g.\ \texttt{mallocs} are not removed).
-
-Greg Sullivan introduced "Dynamic Partial Evaluation", which is a special
-form of partial evaluation at runtime \cite{sullivan_dynamic_2001} and describes
-an implementation for a small dynamic language based on lambda calculus. This
-work is conceptually very close to our own.
-% XXX there are no performance figures, we have no clue how much of this is
-% implemented. not sure how to write this
-
-Our algorithm to avoid allocation of unneeded intermediate objects fits into
-the research area of escape analysis: in comparison to advanced techniques
-\cite{Blanchet99escapeanalysis}, \cite{Choi99escapeanalysis} our algorithm is
-totally simple-minded, but it is still useful in practise.
+\anto{XXX: please review the last two paragraphs}
 
@@
 
-\anto{XXX: cite the pypy tracing paper}
-
 At the moment, some bugs and minor missing features prevent the CLI JIT

### 2009-04-06 talk/icooolps2009-dotnet/paper.tex
fix an XXX
remove the fixed XXX

@@
 
-\anto{XXX: should we cite the DLR? How to prove that DLR programs are much
-  slower than C\#?}
-\davide{We can try writing something in the conclusion}
-
-Recent developments show that JIT compilers can exploit runtime type
+Recent developments show that \emph{Just In Time} (JIT) compilers can exploit runtime type
 information to generate quite efficient code.  Unfortunately, writing a JIT

### 2009-04-06 talk/icooolps2009-dotnet/intro.tex
Add an \arigo that could become a \footnote.

@@
 it still requires a lot of work, as
-IronPython\footnote{http://www.codeplex.com/IronPython},
-Jython\footnote{http://www.jython.org/}
-and JRuby\footnote{http://jruby.codehaus.org/} demonstrate.
+IronPython\footnote{\texttt{http://www.codeplex.com/IronPython}},
+Jython\footnote{\texttt{http://www.jython.org/}}
+and JRuby\footnote{\texttt{http://jruby.codehaus.org/}} demonstrate.
 
@@
 efficient translation of the interpreter to low-level code.
+\arigo{But note that it's a full Python interpreter; RPython is only the
+language in which this interpreter is written.}
 

### 2009-04-06 talk/icooolps2009/paper.tex
A round of language review.

@@
 would hope when the compiled program is itself a bytecode interpreter. We
-examine why that is the case, and how matters can be improved by adding hints to
+examine why that is the case, and how matters can be improved by adding markers to
 the interpreter, that help the tracing JIT to improve the results. We evaluate
@@
 JavaScript is increasingly being used to implement full-scale applications
-running in browser, whereas other dynamic languages (such as Ruby, Perl, Python,
+which run within a browser, whereas other dynamic languages (such as Ruby, Perl, Python,
 PHP) are used for the server side of many web sites, as well as in areas
@@
 performance of interpreters written with the help of the PyPy toolchain. The
-approach is that of a tracing JIT compiler. Opposed to the tracing JITs for dynamic
-languages that exist so far, PyPy's tracing JIT operates "one level down",
-e.g. traces the execution of the interpreter, as opposed to the execution
+approach is that of a tracing JIT compiler. Contrary to the tracing JITs for dynamic
+languages that currently exist, PyPy's tracing JIT operates "one level down",
+e.g. it traces the execution of the interpreter, as opposed to the execution
 of the user program. The fact that the program the tracing JIT compiles is
@@
 tracing JITs and their application to interpreters in Section
-\ref{sect:tracing}.  By this approach we hope to get a JIT compiler that can be
-applied to a variety of dynamic languages, given an interpreter for them. The
+\ref{sect:tracing}.  By this approach we hope to arrive at a JIT compiler that can be
+applied to a variety of dynamic languages, given an appropriate interpreter for each of them. The
 process is not completely automatic but needs a small number of hints from the
@@
 integrates into the rest of PyPy will be explained in Section
-\ref{sect:implementation}. This work is not finished, but already produces some
+\ref{sect:implementation}. This work is not finished, but has already produced some
 promising results, which we will discuss in Section \ref{sect:evaluation}.
@@
 
-\arigo{We should not start from scratch and insert as little details what
-differs in our approach when compared to others; instead we should give a
-higher-level overview and then focus on these details, and a couple of
-references for more info about the "common" part.
-%
-In general there are many things that are never said at all.
-I think the introduction should really be written from the point of view of
-someone that has read already some papers for JavaScript.}
-
 Tracing JITs are an idea initially explored by the Dynamo project
@@
 At first, when the program starts, everything is interpreted.
-The interpreter does a bit of lightweight profiling to figure out which loops
-are run often. This lightweight profiling is usually done by having a counter on
+The interpreter does a small amount of lightweight profiling to establish which loops
+are run most frequently. This lightweight profiling is usually done by having a counter on
 each backward jump instruction that counts how often this particular backward jump
-was executed. Since loops need a backward jump somewhere, this method finds
+was executed. Since loops need a backward jump somewhere, this method looks for
 loops in the user program.
@@
 operations, together with their actual operands and results.  By examining the
-trace, it is possible to produce highly efficient machine code by emitting
+trace, it is possible to produce highly efficient machine code by generating
 only the operations needed.  Being sequential, the trace represents only one
@@
 followed another direction, for example conditions or indirect/virtual
-calls.  When emitting the machine code, every guard is turned into a quick check
+calls.  When generating the machine code, every guard is turned into a quick check
 to guarantee that the path we are executing is still valid.  If a guard fails,
-we immediately quit from the machine code and continue the execution by falling
+we immediately quit the machine code and continue the execution by falling
 back to interpretation.
@@
 During tracing, the trace is repeatedly
-checked whether the interpreter is at a position in the program that it had seen
-earlier in the trace. If this happens, the trace recorded corresponds to a loop
+checked as to whether the interpreter is at a position in the program where it had been
+earlier. If this happens, the trace recorded corresponds to a loop
 in the interpreted program. At this point, this loop
 is turned into machine code by taking the trace and making machine code versions
-of all the operations in it. The machine code can then be immediately executed,
+of all the operations in it. The machine code can then be executed immediately,
 starting from the next iteration of the loop, as the machine code represents
@@
 complex mechanisms in place to still produce more code for the cases of guard
-failures \cite{XXX}, but they are orthogonal to the issues discussed in this
+failures \cite{XXX}, but they are independent of the issues discussed in this
 paper.
@@
 first those functions will be interpreted, but after a while, profiling shows
-that the \texttt{while} loop in \texttt{strange\_sum} is executed often.  The
+that the \texttt{while} loop in \texttt{strange\_sum} is executed often  The
 tracing JIT will then start to trace the execution of that loop.  The trace would
@@
 \vspace{-0.4cm}
-The operations in this sequence are operations of the mentioned intermediate
+The operations in this sequence are operations of the above-mentioned intermediate
 representation (e.g. note that the generic modulo and equality operations in the
@@
 while the other branch is implemented via a guard failure. This trace can then
-be turned into machine code and executed.
+be converted into machine code and executed.
 
@@
 user program, but to the interpreter running the user program. In this section
-we will explore what problems this brings, and how to solve them (at least
-partially). This means that there are two interpreters involved, and we need
-terminology to distinguish them. On the one hand, there is the interpreter that
+we will explore what problems this brings, and suggest how to solve them (at least
+partially). This means that there are two interpreters involved, and we need appropriate
+terminology to distinguish beween them. On the one hand, there is the interpreter that
 the tracing JIT uses to perform tracing. This we will call the \emph{tracing
-interpreter}. On the other hand, there is the interpreter that is running the
+interpreter}. On the other hand, there is the interpreter that runs the
 user's programs, which we will call the \emph{language interpreter}. In the
@@
 
-\fijal{I find following paragraph out of scope and completely confusing, we
-should instead simply state that we unroll the loop, how we do that and
-why we do that. Completely ignore aspect of an interpreter loop I suppose,
-because everything previously keeps talking about can\_enter\_jit that closes
-loop being available at jump back bytecodes}
 A tracing JIT compiler finds the hot loops of the program it is compiling. In
@@
 
-\fijal{This paragraph should go away as well}
 Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
@@
 encounters a backward jump in the language interpreter. Again the tracing JIT
-cannot known which part of the language interpreter implements backward jumps,
-so it needs to be told with the help of a hint by the author of the language
-interpreter.
+cannot know which part of the language interpreter implements backward jumps,
+so the author of the language interpreter needs to indicate this with the help
+of a hint.
 
@@
 described so far, \emph{all} pieces of assembler code correspond to the bytecode
-dispatch loop of the language interpreter. They correspond to different
-unrollings and paths through that loop though. To figure out which of them to use
+dispatch loop of the language interpreter. However, they correspond to different
+paths through the loop and different ways to unroll it. To ascertain which of them to use
 when trying to enter assembler code again, the program counter of the language
@@
 In addition to the classification of the variables, there are two methods of
-\texttt{JitDriver} that need to be called. Both of them get as arguments the
+\texttt{JitDriver} that need to be called. Both of them receive as arguments the
 current values of the variables listed in the definition of the driver. The
@@
 
-For the small example the hints look like a lot of work. However, the amount of
+For the small example the hints look like a lot of work. However, the number of
 hints is essentially constant no matter how large the interpreter is, which
@@
 to be expected, given that the tracing interpreter looks at the execution of the
-language interpreter, it would still be nicer if some of these operations could
+language interpreter, it would still be an improvement if some of these operations could
 be removed.
@@
 With this optimization enabled, the trace looks as in Figure
-\ref{fig:trace-full}. Now a lot of the language interpreter is actually gone
+\ref{fig:trace-full}. Now much of the language interpreter is actually gone
 from the trace and what is left corresponds very closely to the loop of the
@@
 bad anyway (in fact we have an experimental optimization that does exactly that,
-but it is not finished).  Once we get this optimized trace, we can pass it to
+but it is not yet finished).  Once we get this optimized trace, we can pass it to
 the \emph{JIT backend}, which generates the correspondent machine code.
@@
 translation process. In this way, the result of the translation is identical to
-as if no hints were present in the interpreter at all.
+that when no hints were present in the interpreter at all.
 
@@
 What is done instead is that the language interpreter keeps running as a C
-program, until a hot loop in the user program is found. To identify loops the
+program, until a hot loop in the user program is found. To identify loops, the
 C version of the language interpreter is generated in such a way that at the
@@
 
-This section will hint at some other implementation issues and optimizations
-that we have done that are beyond the scope of this paper (and will be subject
+This section will look at some other implementation issues and optimizations
+that we have done that are beyond the scope of this paper (and will be the subject
 of a later publication).
@@
 analysis on the loop operation after tracing has been performed. In this way all
-objects that are allocated during the loop and don't actually escape the loop do
+objects that are allocated during the loop and do not actually escape the loop do
 not need to be allocated on the heap at all but can be exploded into their
@@
 preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
-GHz Pentium M processor and 1GiB RAM, using Linux 2.6.27. All benchmarks where
+GHz Pentium M processor and 1 GB RAM, using Linux 2.6.27. All benchmarks where
 run 50 times, each in a newly started process. The first run was ignored. The
@@
 example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
-the square of 46340 (the smallest number so that the square still fits into a 32
+the square of 46340 (the smallest number whose square still fits into a 32
 bit word) using the bytecode of Figure \ref{fig:square}. The results for various
@@
 \textbf{Benchmark 4:} Same as before, but with constant folding enabled. This corresponds to the
-trace in Figure \ref{fig:trace-full}. This speeds up the square function nicely,
+trace in Figure \ref{fig:trace-full}. This speeds up the square function considerably,
 making it about six times faster than the pure interpreter.
@@
 partial evaluator ``packaged as a library''; decisions about what can be
-specialised and how are pre-determined. Another work in this direction is DyC
+specialised and how, are pre-determined. Another work in this direction is DyC
 \cite{grant_dyc:expressive_2000}, another runtime specializer for C. Both of these projects
-have a similar problem as DynamoRIO.  Targeting the C language makes
-higher-level specialisation difficult (e.g.\ \texttt{malloc} can not be
+have a problem similar to that of DynamoRIO.  Targeting the C language makes
+higher-level specialisation difficult (e.g.\ \texttt{malloc} cannot be
 optimized).
 
-There has been some attempts to do \emph{dynamic partial evaluation}, which is
+There have been some attempts to do \emph{dynamic partial evaluation}, which is
 partial evaluation that defers partial evaluation completely to runtime
@@
 
-We have shown techniques for making it practical to apply a tracing
+We have shown techniques for improving the results when applying a tracing
 JIT to an interpreter. Our first benchmarks indicate that these techniques work
 really well on small interpreters and first experiments with PyPy's Python
-interpreter make it seems likely that they can be scaled up to realistic
+interpreter make it appear likely that they can be scaled up to realistic
 examples.
@@
 Furthermore we need to apply the JIT to the various interpreters that are
-written with PyPy to evaluate how widely applicable the described techniques
+written in RPython to evaluate how widely applicable the described techniques
 are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk

### 2009-04-06 talk/icooolps2009-dotnet/intro.tex
turn arigo into a footnote

@@
 Python where some dynamic features have been sacrificed to allow an
-efficient translation of the interpreter to low-level code.
-\arigo{But note that it's a full Python interpreter; RPython is only the
-language in which this interpreter is written.}
+efficient translation of the interpreter to low-level code\footnote{But note that it's a full Python interpreter; RPython is only the
+language in which this interpreter is written.}.
 

### 2009-04-06 talk/icooolps2009-dotnet/conclusion.tex
turn arigo into a footnote

@@
 equivalent written in
-C\#\footnote{http://shootout.alioth.debian.org/gp4/\\benchmark.php?test=all\&lang=iron\&lang2=csharp}
+C\#\footnote{\texttt{http://shootout.alioth.debian.org/gp4/\\benchmark.php?test=all\&lang=iron\&lang2=csharp}}
 or
-Java\footnote{http://blog.dhananjaynene.com/2008/07/performance-comparison-c-java-python-ruby-jython-jruby-groovy/}.
+Java\footnote{\texttt{http://blog.dhananjaynene.com/2008/07/performance-\\comparison-c-java-python-ruby-jython-jruby-groovy/}}.
 
-The \emph{Dynamic Language Runtime}\footnote{http://www.codeplex.com/dlr}
+The \emph{Dynamic Language Runtime}\footnote{\texttt{http://www.codeplex.com/dlr}}
 (DLR) is a library designed to ease the implementation of dynamic languages
@@
 In particular, the \emph{Da Vinci Machine
-  Project} \footnote{http://openjdk.java.net/projects/mlvm/} is exploring and
+  Project} \footnote{\texttt{http://openjdk.java.net/projects/mlvm/}} is exploring and
 implementing new features to ease the implementation of dynamic languages on

### 2009-04-06 talk/icooolps2009-dotnet/benchmarks.tex
adapt che conclusions to the new paper
move the benchmarks to my home dir on codespeak, and add a link to jsr-292

@@
 
-  http://codespeak.net/svn/pypy/extradoc/talk/ecoop2009/benchmarks/}
+  \texttt{http://codespeak.net/~antocuni/tlc-benchmarks/}}
 
@@
 and inline the method call.  Moreover, since \lstinline{obj} never escapes the
-function, it is \emph{virtualized} and its field \lstinline{value} is stored
+function, the allocation is avoided and its field \lstinline{value} is stored
 as a local variable.  As a result, the generated code turns out to be a simple loop

### 2009-04-06 talk/icooolps2009-dotnet/conclusion.tex
move the benchmarks to my home dir on codespeak, and add a link to jsr-292

@@
 top of the JVM: some of these features, such as the new
-\emph{invokedynamic}\footnote{XXX} instruction and the \emph{tail call
+\emph{invokedynamic}\footnote{\texttt{http://jcp.org/en/jsr/detail?id=292}} instruction and the \emph{tail call
   optimization} can probably be exploited by a potential JVM backend to

### 2009-04-06 talk/icooolps2009-dotnet/clibackend.tex
display block ids in bold instead of red, else color-blind reviewers complain :-)

@@
 The picture in Figure~\ref{block-id-fig} shows a graph composed of three methods (for
-simplicity, dispatchers are not shown); method ids are in red, whereas
+simplicity, dispatchers are not shown); method ids are in bold, whereas
 block numbers are in black. 

### 2009-04-06 talk/icooolps2009-dotnet/benchmarks.tex
Typo.

@@
   runtime, because there is no way to know in advance if and where an object
-  have that particular attribute or method.
+  has that particular attribute or method.
 \end{itemize}

### 2009-04-06 talk/icooolps2009-dotnet/conclusion.tex
clarify the dlr paragraph

@@
 (DLR) is a library designed to ease the implementation of dynamic languages
-for .NET: IronPython is based on the DLR, so the same remarks apply.
+for .NET: the DLR is closely related to IronPython\footnote{In fact, the DLR
+  started as a spin-off of IronPython, and nowadays the latter is based on the
+  former.} and employs the same techniques described above: thus, the remarks
+about the differences between the PyPy and IronPython apply to all DLR based
+languages.
 

### 2009-04-06 talk/icooolps2009-dotnet/conclusion.tex
Typos.

--- 2009-04-06 talk/icooolps2009-dotnet/conclusion.tex
@@
 (DLR) is a library designed to ease the implementation of dynamic languages
-for .NET: IronPython is based on the DLR, so the same remarks apply.
+for .NET: the DLR is closely related to IronPython\footnote{In fact, the DLR
+  started as a spin-off of IronPython, and nowadays the latter is based on the
+  former.} and employs the techniques described above; thus, the remarks
+about the differences between PyPy and IronPython apply to all DLR based
+languages.
 

### 2009-04-06 talk/icooolps2009-dotnet/intro.tex
some statements rephrased

@@
 Writing a compiler that targets the CLI or JVM is easier than targeting a real CPU, but
-it still requires a lot of work, as
-IronPython\footnote{\texttt{http://www.codeplex.com/IronPython}},
-Jython\footnote{\texttt{http://www.jython.org/}}
-and JRuby\footnote{\texttt{http://jruby.codehaus.org/}} demonstrate.
+it still requires a lot of work, as IronPython\footnote{\texttt{http://www.codeplex.com/IronPython}},
+Jython\footnote{\texttt{http://www.jython.org/}} and JRuby\footnote{\texttt{http://jruby.codehaus.org/}} demonstrate.
+Finally, if one really seeks for an efficent language implementation, \emph{Just In Time} (JIT) compilation needs
+to be considered; only in this way  the compiler can exploit runtime type
+information to generate quite efficient code. Note that JIT compilation
+has not to be confused with lazy compilation of IronPython and Jython which is exploited to save memory, 
+since in these cases no runtime type information is ever used to generate more efficient code.
 
-Moreover, writing a static compiler is often not enough to get high
-performance.  IronPython and Jython compile code lazily to save memory, but
-since they do not exploit the extra information potentially available at
-runtime, it is more a delayed static compilation than a true JIT one.
-\anto{XXX: someone review the english of the last paragraph, please :-)}
-
-PyPy's approach \cite{RiBo07_223} is to automatize the generation of 
-specializing JIT compilers in order
-to minimize the effort required to get a fast implementation of a
-dynamic language; automatic generation of JIT compilers is done with
-the help of partial evaluation techniques and requires the user only
-to provide an interpreter with some manual annotations which hint
-the generator how interpretation and JIT compilation has to be interleaved \cite{PyPyJIT09}.
+Unfortunately, writing a JIT compiler is a very complex task.  
+To make this task simpler, the solution proposed by PyPy \cite{RiBo07_223}  
+is automatic generation of JIT compilers with the help of partial evaluation techniques: 
+the user has only to provide an interpreter manually annotated with \emph{hints}
+specifying how interpretation and JIT compilation has to be interleaved \cite{PyPyJIT09}.
 

### 2009-04-06 talk/icooolps2009-dotnet/intro.tex
small corrections, a comment added

@@
 ones found in other projects like IronPython is that the latter compile
-code at the method granularity -- if on the one hand they can exploit
-some of the knowledge gathered at runtime (e.g.\ the types of method
+code at the method granularity: if on the one hand they can exploit
+some of the knowledge gathered at runtime \davide{this contradicts what is stated in the intro} (e.g.\ the types of method
 parameters), on the other hand they can do little to optimize most of
@@
 RPython.  Then, to guide the process, we need to add few manual
-annotations to the interpreter, in order to teach the JIT generator which
+annotations (called hints) to the interpreter, in order to teach the JIT generator which
 information is important to know at compile-time.  
@@
 This implies that the generated code needs to contain some sort of updatable
-switch, or \emph{flexswitch}, which can pick the right code path based on the
+switch, called \emph{flexswitch}, which can pick the right code path based on the
 run-time value.  Typically, the value we switch on is the runtime dynamic type
@@
 .NET, as the framework does not offer any primitive to do this.  Section
-\ref{sec:clibackend} explains how it is possible to simulate this behaviour.
+\ref{sec:clibackend} explains how it is possible to obtain this behaviour.

### 2009-04-07 talk/icooolps2009/paper.tex
simplify this paragraph

@@
 the intermediate representation of PyPy's translation toolchain after type
-inference has been performed and Python-specifics have been made explicit. At
-first those functions will be interpreted, but after a while, profiling shows
-that the \texttt{while} loop in \texttt{strange\_sum} is executed often  The
-tracing JIT will then start to trace the execution of that loop.  The trace would
+inference has been performed.
+When the profiler shows
+that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
+tracing JIT will start to trace the execution of that loop.  The trace would
 look as follows:

### 2009-04-07 talk/icooolps2009-dotnet/benchmarks.tex
Some statments rephrased, minor corrections

@@
 bytecode instruction set. Since our main interest is in the runtime
-performance of the interpreter, we did not implement the parser nor the
-bytecode compiler, but only the interpreter itself.
+performance of the interpreter, we did not implement the whole language, but just its virtual machine.
 
@@
   programs are available at:
-
   \texttt{http://codespeak.net/~antocuni/tlc-benchmarks/}}
@@
 spent to simply interpret the program.  This is an expected result
-which, however, can be improved once we will have time
-to optimize compilation and not only the generated code.
+which, however, can be improved, since so far no effort has been direct to
+enhance the performance of the compiler itself.
 
@@
 once. Instead, it compiles and executes code chunk by chunk, waiting until it
-knows enough informations to generate highly efficient code.  In particular,
+knows enough information to generate highly efficient code.  In particular,
 at the time it emits the code for the inner loop it exactly knows the

### 2009-04-07 talk/icooolps2009-dotnet/clibackend.tex
Some statments rephrased, minor corrections

@@
 \lstinline{Reflection.Emit} namespace and creating \lstinline{DynamicMethod}s.}: 
-before being
-executed, the generated code will be compiled again by .NET's own JIT
+before being executed, the generated code will be compiled again by the .NET JIT
 compiler.
 
-Thus, when using the CLI backend, we actually have two layers of
-JIT-compilation, each layer removing away different kinds of overhead.  By
-operating at a higher level, our JIT can potentially do a better job than the
-.NET one in some contexts, as our benchmarks demonstrate (see
+Thus, when using the CLI backend, we actually have two JIT compilers at two different levels
+layers, each one specialized in different kinds of optimization.
+By operating at a higher level, our JIT can potentially do a better job 
+in some contexts, as our benchmarks demonstrate (see
 Section~\ref{sec:benchmarks}).  On the other hand, the lower-level .NET JIT is
 very good at producing machine code, much more than PyPy's own \emph{x86}
-backend for example.  By combining the strengths of both we can get highly
+backend, for example.  By combining the strengths of both we can get highly
 efficient machine code.
@@
 time could run slower with JIT than without, due to the time spent doing the
-initial (double) compilation.  It is important to underline that so far we
-mostly focused making the JIT able to produce very fast code, without trying
-to optimze the compilation phase itself.
+initial (double) compilation.  
+Finally, it is important to underline that while we have directed our efforts 
+to generate a JIT compiler able to emit very efficient code,
+the performance of the compiler itself has been neglected so far, but  
+it is certainly an issue which will have to be considered in our future research.
 
@@
 there is a close correspondence between most of the operations used by
-frontend's flowgraphs and the CLI instructions.  Thus, we will not dig into
+frontend's flowgraphs and the CLI instructions.  Thus, we will not go into
 details for this part.
@@
 \caption{An example of a flexswitch evolution: in the picture on the
-  right a new block has been dynamically added.}\label{flexswitch-fig}
+  right block 7 has been dynamically added.}\label{flexswitch-fig}
 \end{center}
@@
 
-Unfortunately, the CLI VM does not allow modification of code which
+Unfortunately, CLI does not allow modification of code which
 has been already loaded and linked, therefore the simplest approach
@@
   number of blocks, all belonging to the same flow graph.
+
+\item The parameters of a method correspond to the arguments of what we call the \emph{initial block} of the method.
+
 \end{itemize}
@@
 
-A link is called \emph{internal} if it connects two blocks contained
-in the same method,
+A link is called \emph{internal} if it connects two blocks implemented
+by the same method,
 \emph{external} otherwise.
@@
 is similar to the following fragment: 
-\begin{small}
 \begin{lstlisting}[language={[Sharp]C}]
@@
 \end{lstlisting}
-\end{small}
 If the next block to be executed is implemented in the same method
-({\small\lstinline{methodid == MY_METHOD_ID}}), then the appropriate
+(\lstinline{methodid == MY_METHOD_ID}), then the appropriate
 jump to the corresponding code is executed.
@@
 \lstinline{FlexSwitchCase} is the type of delegates for secondary methods):
-\begin{small}
 \begin{lstlisting}[language={[Sharp]C}] 
@@
 \end{lstlisting}
-\end{small}
 Each secondary method returns the block id of the next block to be
@@
 and type is variable.
-\begin{small}
 \begin{lstlisting}[language={[Sharp]C}] 
@@
 \end{lstlisting}
-\end{small}
 Unfortunately, with this solution passing arguments to external links
@@
 The following snippet shows the special case of integer flexswitches.
-\begin{small}
 \begin{lstlisting}[language={[Sharp]C}] 
@@
 \end{lstlisting}
-\end{small}
 The mapping from integers values to delegates (pointing to secondary

### 2009-04-07 talk/icooolps2009-dotnet/conclusion.tex
Some statments rephrased, minor corrections

@@
 
-Moreover, the idea of promotion is a generalization of \emph{Polymorphic
+The idea of promotion is a generalization of \emph{Polymorphic
   Inline Caches} \cite{hoelzle_optimizing_1991}, as well as the idea of using
@@
 which can automatically turn an interpreter into a JIT
-compiler, requiring the language developers to only add few \texttt{hint}s to
+compiler, requiring the language developers to only add few hints to
 guide the generation process.
@@
 languages to be even faster than their statically typed counterpart in some
-circumstances.
+cases.
 
@@
 particular, we are working on a way to automatically detect loops in the user
-code, as tracing JITs do \cite{gal_hotpathvm_2006}.  By compilining whole
-loops at once, the backends should be able to produce better code than today.
+code, as tracing JITs do \cite{gal_hotpathvm_2006}.  By compiling whole
+loops at once, the backends should be able to produce better code.
 
@@
 results comparable to TLC, as the other backends already demonstrate
-\cite{PyPyJIT}.  Moreover, if the current implementation of flexswitches will
-prove to be too slow for some purposes, we want to explore alternative
-implementation strategies, also considering the new features that might be
-integrated into virtual machines.
+\cite{PyPyJIT}.  However, if the current implementation of flexswitches will
+turn out to be too slow for some purposes, alternative
+implementation strategies could be explored by considering the novel features 
+offered the new generation of virtual machines.
 

### 2009-04-07 talk/icooolps2009-dotnet/paper.tex
keyword style now is bf

@@
         basicstyle=\scriptsize\ttfamily,
-        keywordstyle=\color{blue}, % I couldn't find a way to make chars both bold and tt
+        keywordstyle=\bf, %\color{blue}, % I couldn't find a way to make chars both bold and tt
         frame=none,

### 2009-04-07 talk/icooolps2009/paper.tex
use proper way of doing "

@@
 approach is that of a tracing JIT compiler. Contrary to the tracing JITs for dynamic
-languages that currently exist, PyPy's tracing JIT operates "one level down",
+languages that currently exist, PyPy's tracing JIT operates ``one level down'',
 e.g. it traces the execution of the interpreter, as opposed to the execution
@@
 %- approach: write a tracing JIT that is applicable to many different languages,
-%by tracing "one level done"
+%by tracing ``one level done''
 %- needs some hints by the interpreter-writer + slightly different optimizations
@@
 To implement a dynamic language with PyPy, an interpreter for that language has
-to be written in RPython \cite{ancona_rpython:step_2007}. RPython ("Restricted Python") is a subset of Python
+to be written in RPython \cite{ancona_rpython:step_2007}. RPython (``Restricted
+Python'') is a subset of Python
 chosen in such a way that type inference can be performed on it. The language
@@
 This process assumes that the path through the loop that was traced is a
-"typical" example of possible paths (which is statistically likely). Of course
+``typical'' example of possible paths (which is statistically likely). Of course
 it is possible that later another path through the loop is taken, in which case
@@
 program that the language interpreter executes we will call the \emph{user
-program} (from the point of view of a VM author, the "user" is a programmer
+program} (from the point of view of a VM author, the ``user'' is a programmer
 using the VM).
@@
 whether a loop is closed. This is considered to be the case when the values of
-the "green" variables are the same as at an earlier call to the
+the ``green'' variables are the same as at an earlier call to the
 \texttt{can\_enter\_jit} method.
@@
 bytecode and the program counter. Those are stored in variables that are part of
-the position key (they are "green"), that means that the tracer checks that they
+the position key (they are ``green''), that means that the tracer checks that they
 are some fixed value at the beginning of the loop. In the example the check
@@
 tracing using a bytecode representation of the language interpreter. That means
-there are two "versions" of the language interpreter embedded in the final
+there are two ``versions'' of the language interpreter embedded in the final
 executable of the VM: on the one hand it is there as executable machine code, on

### 2009-04-07 talk/icooolps2009/paper.tex
small fixes, and rephrasing some sentences that I found to heavyweight

@@
 A tracing JIT compiler finds the hot loops of the program it is compiling. In
-our case, this program is the language interpreter. The most important hot loop
-of the language interpreter is its bytecode dispatch loop (for many simple
+our case, this program is the language interpreter. The most important hot interpreter loop
+is the bytecode dispatch loop (for many simple
 interpreters it is also the only hot loop).  Tracing one iteration of this
@@
 bytecode within that.  In the example above, the program counter is represented by 
-the \texttt{bytecode} and \texttt{pair} variables.
+the \texttt{bytecode} and \texttt{pc} variables.
 
@@
 earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
-instruction, and only if it is actually a backward jump. The place where this
-method is called is where the language interpreter performs profiling to decide
+instruction, and only if it is actually a backward jump. Here is
+where the language interpreter performs profiling to decide
 when to start tracing. It is also the place where the tracing JIT checks
@@
 The first integration problem is how to \emph{not} integrate the tracing JIT at
-all. It should be possible to choose when the language interpreter is translated to C
+all. It is possible to choose when the language interpreter is translated to C
 whether the JIT should be built in or not. If the JIT is not enabled, all the
@@
 example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
-the square of 46340 (the smallest number whose square still fits into a 32
+the square of 46340 (the largest number whose square still fits into a 32
 bit word) using the bytecode of Figure \ref{fig:square}. The results for various
@@
 never invoked to measure the overhead of the profiling. For this interpreter
-it to be rather large, with 50\% slowdown due to profiling. This is
-because the example interpreter needs to do one hash table lookup per loop
-iteration. For larger interpreters (e.g. the Python one) it seems likely that
-the overhead is less significant, given that many operations in Python need
-hash-table lookups themselves.
+it to be rather large, with 50\% slowdown due to profiling. This is because the interpreter 
+is small and the opcodes simple. For larger interpreters (e.g. the Python one) it seems 
+likely that the overhead is less significant.
 

### 2009-04-07 talk/icooolps2009/paper.tex
some fixes by michael

@@
 This process assumes that the path through the loop that was traced is a
-``typical'' example of possible paths (which is statistically likely). Of course
+``typical'' example of possible paths. Of course
 it is possible that later another path through the loop is taken, in which case
-one of the guards that were put into the machine code will fail. There are more
+one of the guards that were put into the machine code will fail.\footnote{There are more
 complex mechanisms in place to still produce more code for the cases of guard
 failures \cite{XXX}, but they are independent of the issues discussed in this
-paper.
+paper.}
 
@@
 interpretation they are the place where the profiling is performed and where
-tracing is started or already existing assembler code entered; during tracing
+tracing is started or already existing assembler code executed; during tracing
 they are the place where the check for a closed loop is performed.
@@
 rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
-operations that were executed, is in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
-to its own beginning, forming an endless loop that can only be left via a guard
-failure. The call to \texttt{f} was inlined into the trace. Note that the trace
+operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
+to its beginning, forming an endless loop that can only be left via a guard
+failure. The call to \texttt{f} is inlined into the trace. Note that the trace
 contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
@@
 assumption that the tracing JIT makes -- that several iterations of a hot loop
-take the same or similar code paths -- is just wrong in this case. It is very
+take the same or similar code paths -- is wrong in this case. It is very
 unlikely that the same particular opcode is executed many times in a row.
@@
 interpreter needs to be checked. If it corresponds to the position key of one of
-the pieces of assembler code, then this assembler code can be entered. This
+the pieces of assembler code, then this assembler code can be executed. This
 check again only needs to be performed at the backward branches of the language
@@
 Let's look at how hints would need to be applied to the example interpreter
-from Figure \ref{fig:tlr-basic}. The basic thing needed to apply hints is a
+from Figure \ref{fig:tlr-basic}. To apply hints one generally needs a
 subclass of \texttt{JitDriver} that lists all the variables of the bytecode
@@
 solved, the loop corresponds exactly to the loop in the square function.
-However, the resulting trace is a bit too long. Most of its operations are not
+However, the resulting trace is not optimized enough. Most of its operations are not
 actually doing any computation that is part of the square function. Instead,
@@
 
-The simple insight how to greatly improve the situation is that most of the
+The simple insight how to improve the situation is that most of the
 operations in the trace are actually concerned with manipulating the
@@
 the position key (they are ``green''), that means that the tracer checks that they
-are some fixed value at the beginning of the loop. In the example the check
+are some fixed value at the beginning of the loop (they may well change over the
+course of the loop, though). In the example the check
 would be that the \texttt{bytecode} variable is the bytecode string
@@
 never invoked to measure the overhead of the profiling. For this interpreter
-it to be rather large, with 50\% slowdown due to profiling. This is because the interpreter 
+it seems to be rather large, with 50\% slowdown due to profiling. This is because the interpreter 
 is small and the opcodes simple. For larger interpreters (e.g. the Python one) it seems 

### 2009-04-07 talk/icooolps2009/paper.tex
I got another round of review

@@
        \affaddr{STUPS Group}\\
-       \affaddr{Deutschland}\\
+       \affaddr{Germany}\\
        \email{cfbolz@gmx.de}
@@
 We present techniques for improving the results when a tracing JIT compiler is
-applied to an interpreter. An unmodified tracing JIT performs not as well as one
+applied to an interpreter.XXXAbrupt start - is the relevanz really immediately clear?XXX An unmodified tracing JIT performs not as well as one
 would hope when the compiled program is itself a bytecode interpreter. We
-examine why that is the case, and how matters can be improved by adding markers to
-the interpreter, that help the tracing JIT to improve the results. We evaluate
+examine the reasons for that, and how matters can be improved by adding markers to
+the interpreter that help the tracing JIT to improve the results. We evaluate
 the techniques by using them both on a small example as well as on a full Python
@@
 such as C/Posix, the CLI or the JVM. The PyPy project is described in more
-details in Section \ref{sect:pypy}.
+detail in Section \ref{sect:pypy}.
 
@@
 The PyPy project\footnote{http://codespeak.net/pypy}
-\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} was started to implement a
-new Python interpreter in Python but has now extended its goals to be an
+\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} is an
 environment where flexible implementation of dynamic languages can be written.
@@
 the recorded trace corresponds to execution of one opcode. This means that the
-assumption that the tracing JIT makes -- that several iterations of a hot loop
+assumption made by the tracing JIT -- that several iterations of a hot loop
 take the same or similar code paths -- is wrong in this case. It is very
@@
 
-Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
+An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
 simple bytecode interpreter with 256 registers and an accumulator. The
@@
 opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
-bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
+bytecode dispatch loop should be unrolled exactly so much that the unrolled version
 corresponds to a \emph{user loop}. User loops
@@
 
-The language interpreter uses a similar techniques to detect \emph{hot user
+The language interpreter uses a similar technique to detect \emph{hot user
 loops}: the profiling is done at the backward branches of the user program,
@@
 
-Of course there is a lot of work still left to do. Various optimizations are not
-quite finished. Both tracing and leaving machine code is very slow due to a
+Of course there is a lot of work still left to do XXX fix this phrase. Various optimizations are not
+quite finished. Both tracing and leaving machine code are very slow due to a
 double interpretation overhead and we might need techniques for improving those.
@@
 implementation \cite{bolz_back_2008}, a Prolog interpreter or PyGirl, a Gameboy
-emulator \cite{XXX}; but also less immediately obvious ones, like Python's
+emulator \cite{XXX}; but also not immediately obvious ones, like Python's
 regular expression engine.

### 2009-04-07 talk/icooolps2009-dotnet/clibackend.tex
use letters instead of numbers to label the blocks in the picture, to avoid confusion with the "official" numbering described below

@@
 
-Thus, when using the CLI backend, we actually have two JIT compilers at two different levels
+Thus, when using the CLI backend, we actually have two JIT compilers at two different
 layers, each one specialized in different kinds of optimization.
@@
 \caption{An example of a flexswitch evolution: in the picture on the
-  right block 7 has been dynamically added.}\label{flexswitch-fig}
+  right block D has been dynamically added.}\label{flexswitch-fig}
 \end{center}
@@
 
-In the pictures of Figure~\ref{flexswitch-fig}, block 5 (highlighted in grey)
+In the pictures of Figure~\ref{flexswitch-fig}, block B (highlighted in grey)
 corresponds to a flexswitch; initially (picture on the left) 
-only block 6, containing the code to restart the JIT compilation,
+only block C, containing the code to restart the JIT compilation,
 is connected to the flexswitch; the picture on the right
 shows the graph after the first case has been dynamically added to the flexswitch,
-by linking block 5 with the freshly created block number 7.
+by linking block B with the freshly created block D.
 
@@
   number of blocks, all belonging to the same flow graph.
-
-\item The parameters of a method correspond to the arguments of what we call the \emph{initial block} of the method.
-
 \end{itemize}
@@
 whose target is not an initial block; consider, for instance, the
-outgoing link from block 7 to block 3 in Figure~\ref{flexswitch-fig}. How is it possible to jump into
+outgoing link from block D to block A in Figure~\ref{flexswitch-fig}. How is it possible to jump into
 the middle of a method?
@@
 \end{lstlisting}
+
+XXX: add note to explain why we use a for loop instead of an hashtable
+
+XXX: add note to explain who calls execute()
+
 The mapping from integers values to delegates (pointing to secondary

### 2009-04-07 talk/icooolps2009-dotnet/clibackend.tex
fix two XXXs

@@
 \subsubsection{Implementation of flexswitches}
-Finally, we can have a look at the implementation of flexswitches.
+
+To implement each flexswitch, the CLI backend creates an instace of a subclass
+of \lstinline{BaseLowLevelFlexSwitch}: such an instance stores the mapping
+between each value and the corresponding method we want to invoke.  Then, the
+generated code contains a call to the method \lstinline{execute}, which
+selects and invoke the right method depending on the actual value we are
+switching on.
+
 The following snippet shows the special case of integer flexswitches.
@@
 
-XXX: add note to explain why we use a for loop instead of an hashtable
-
-XXX: add note to explain who calls execute()
-
 The mapping from integers values to delegates (pointing to secondary
@@
 the link and jumps to the right block by performing a linear search in
-array \lstinline{values}.
+array \lstinline{values}\footnote{Our microbenchmarks indicate that a linear
+search is the fastest way to find the right method to call, since typically
+each flexswitch contains only a very small number of cases.}.
 

### 2009-04-07 talk/icooolps2009/paper.tex
small tweaks. can remove most of the compression stuff again, as there is an official compressing style sheet.

@@
-\documentclass{acm_proc_article-sp}
+%\documentclass{acm_proc_article-sp}
+\documentclass{sig-alternate}
 
@@
        \email{cuni@disi.unige.it}
+\and
 \alignauthor Armin Rigo\\
@@
 The contributions of this paper are:
-\vspace{-0.3cm}
 \begin{zitemize}
@@
 \item Finding techniques for improving the generated code.
-\item Integrating
 \end{zitemize}
@@
 Tracing JITs are built on the following basic assumptions:
-\vspace{-0.3cm}
 \begin{zitemize}
@@
 Typically, programs executed by a tracing VMs goes through various phases:
-\vspace{-0.3cm}
 \begin{zitemize}
@@
 complex mechanisms in place to still produce more code for the cases of guard
-failures \cite{XXX}, but they are independent of the issues discussed in this
+failures \cite{andreas_gal_incremental_2006}, but they are independent of the issues discussed in this
 paper.}
@@
 }
-\vspace{-0.4cm}
 To trace this, a bytecode form of these functions needs to be introduced that
@@
 }
-\vspace{-0.4cm}
 The operations in this sequence are operations of the above-mentioned intermediate
@@
 \input{code/tlr-paper.py}
-\vspace{-0.4cm}
 \caption{A very simple bytecode interpreter with registers and an accumulator.}
@@
 }
-\vspace{-0.4cm}
 \caption{Example bytecode: Compute the square of the accumulator}
@@
 \input{code/normal-tracing.txt}
-\vspace{-0.4cm}
 \caption{Trace when executing the \texttt{DECR\_A} opcode}
@@
 \input{code/tlr-paper-full.py}
-\vspace{-0.4cm}
 \caption{Simple bytecode interpreter with hints applied}
@@
 hints is essentially constant no matter how large the interpreter is, which
-makes it seem less significant for larger interpreters.
+makes it less significant for larger interpreters.
 
@@
 \input{code/no-green-folding.txt}
-\vspace{-0.4cm}
 \caption{Trace when executing the Square function of Figure \ref{fig:square},
@@
 \input{code/full.txt}
-\vspace{-0.4cm}
 \caption{Trace when executing the Square function of Figure \ref{fig:square},

### 2009-04-08 talk/icooolps2009/paper.tex
- finally insert the Python benchmark  - kill parts of Section 4 and move other parts to the future work section, as    recommended by Michael  - some more fixes and a new reference

@@
    {\end{list}}
-
-\textfloatsep 12pt plus 2pt minus 4pt
 
@@
 performance of dynamic languages (in the SELF project, to name just one example
-\cite{XXX}), those techniques are not as widely
+\cite{hlzle_adaptive_1994}), those techniques are not as widely
 used as one would expect. Many dynamic language implementations use completely
@@
 The contributions of this paper are:
-\begin{zitemize}
+\begin{itemize}
 \item Applying a tracing JIT compiler to an interpreter.
 \item Finding techniques for improving the generated code.
-\end{zitemize}
+\end{itemize}
 
@@
 Tracing JITs are built on the following basic assumptions:
-\begin{zitemize}
+\begin{itemize}
  \item programs spend most of their runtime in loops
  \item several iterations of the same loop are likely to take similar code paths
-\end{zitemize}
+\end{itemize}
 
@@
 Typically, programs executed by a tracing VMs goes through various phases:
-\begin{zitemize}
+\begin{itemize}
 \item Interpretation/profiling
@@
 \item Execution of the generated code
-\end{zitemize}
+\end{itemize}
 
@@
 
-\subsection{Various Issues}
-
-This section will look at some other implementation issues and optimizations
-that we have done that are beyond the scope of this paper (and will be the subject
-of a later publication).
-
-\textbf{Assembler Backends:} The tracing interpreter uses a well-defined
-interface to an assembler backend for code generation. This makes it possible to
-easily port the tracing JIT to various architectures (including, we hope, to
-virtual machines such as the JVM where backend could generate bytecode at
-runtime). At the moment the only implemented backend is a 32-bit
-Intel-x86 backend.
-
-\textbf{Trace Trees:} This paper ignored the problem of guards that fail often
-because there are several equally likely paths through
-a loop. Always falling back to interpretation in this case is not practicable.
-Therefore, if we find a guard that fails often enough, we start tracing from
-there and produce efficient machine code for that case, instead of always
-falling back to interpretation.
-
-\textbf{Allocation Removal:} A key optimization for making the approach
-produce good code for more complex dynamic language is to perform escape
-analysis on the loop operation after tracing has been performed. In this way all
-objects that are allocated during the loop and do not actually escape the loop do
-not need to be allocated on the heap at all but can be exploded into their
-respective fields.  This is very helpful for dynamic languages where primitive
-types are often boxed, as the constant allocation of intermediate results is
-very costly.
-
-\textbf{Optimizing Frame Objects:} One problem with the removal of allocations
-is that many dynamic languages are so reflective that they allow the
-introspection of the frame object that the interpreter uses to store local
-variables (e.g. SmallTalk, Python). This means that intermediate results always
-escape because they are stored into the frame object, rendering the allocation
-removal optimization ineffective. To remedy this problem we make it possible to
-update the frame object lazily only when it is actually accessed from outside of
-the code generated by the JIT.
+Machine code production is done via a well-defined interface to an assembler
+backend. This makes it possible to easily port the tracing JIT to various
+architectures (including, we hope, to virtual machines such as the JVM where
+backend could generate bytecode at runtime). At the moment the only implemented
+backend is a 32-bit Intel-x86 backend.
 
@@
 confidence intervals were computed using a 95\% confidence level. All times
-include the running of the tracer and machine code production to measure how
-costly those are.
+include the running of the tracer and machine code production.
 
@@
 example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
-the square of 46340 (the largest number whose square still fits into a 32
-bit word) using the bytecode of Figure \ref{fig:square}. The results for various
+the square of 10000000 (the result will overflow, but for smaller numbers the
+running time is not long enough to sensibly measure it)
+using the bytecode of Figure \ref{fig:square}. The results for various
 constellations are as follows:
@@
 
-\textbf{Benchmark 2:} The tracing JIT is enabled, but no interpreter-specific
+\textbf{Benchmark 2:} The tracing JIT is enabled, but no inter\-preter-specific
 hints are applied. This corresponds to the trace in Figure
@@
 \textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
-never invoked to measure the overhead of the profiling. For this interpreter
-it seems to be rather large, with 50\% slowdown due to profiling. This is because the interpreter 
+never invoked. In this way the overhead of the profiling is measured. For this interpreter
+it seems to be rather large, with about 20\% slowdown due to profiling. This is because the interpreter 
 is small and the opcodes simple. For larger interpreters (e.g. the Python one) it seems 
@@
 
-\textbf{Benchmark 6:} Runs the whole computation on the tracing interpreter for estimating the
-involved overheads of tracing. The trace is not actually recorded (which would be a
-memory problem), so in reality the number is even higher. Due to the double
-interpretation, the overhead is huge. It remains to be seen whether that will be
-a problem for practical interpreters.
+\begin{figure}
+\noindent
+\begin{tabular}{|l|l|r|r|}
+\hline
+& &Time (ms) &speedup\\
+\hline
+1 &compiled to C, no JIT &442.7 $\pm$ 3.4 &1.00\\
+2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
+3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
+4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
+5 &Profile Overhead &515.0 $\pm$ 7.2 &0.86\\
+\hline
+\end{tabular}
+\label{fig:bench1}
+\caption{Benchmark results of example interpreter computing the square of
+10000000}
+\end{figure}
 
-\textbf{Benchmark 7:} For comparison, the time of running the interpreter on top of CPython
-(version 2.5.2).
+Furthermore, to test the technique on a more realistic example, we did some
+preliminary benchmarks with PyPy's Python interpreter. The function we
+benchmarked as well as the results can be seen in Figure
+\ref{fig:bench-example}. The function is a bit arbitrary but executing it is
+still non-trivial, as a normal Python interpreter needs to dynamically dispatch
+nearly all of the involved operations (like indexing into the tuple, addition
+and comparison of \texttt{i}). We benchmarked PyPy's Python interpreter with the
+JIT disabled, with the JIT enabled and
+CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
+Python).
+
+The results show that the tracing JIT speeds up the execution of this Python
+function significantly, even outperforming CPython by a bit. The tracer needs to
+trace through quite a bit of dispatching machinery of the Python interpreter to
+achieve this, XXX.
 
 \begin{figure}
+\label{fig:bench-example}
+{\small
+\begin{verbatim}
+def f(a):
+    t = (1, 2, 3)
+    i = 0
+    while i < a:
+        t = (t[1], t[2], t[0])
+        i += t[0]
+    return i
+\end{verbatim}
+}
 \noindent
-{\small
-\begin{tabular}{|l|r|}
+\begin{tabular}{|l|l|r|r|}
 \hline
- &ratio\tabularnewline
+& &Time (s) &speedup\\
 \hline
-Interpreter compiled to C, no JIT &1\tabularnewline \hline
-Normal Trace Compilation &1.20\tabularnewline \hline
-Unfolding of Language Interpreter Loop &XXX\tabularnewline \hline
-Full Optimizations &0.17\tabularnewline \hline
-Profile Overhead &1.51\tabularnewline \hline
-Interpreter run by Tracing Interpreter &860.20\tabularnewline \hline
-Interpreter run by CPython &256.17\tabularnewline \hline
+1 &Compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
+2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
+3 &CPython 2.5.4 &4.96 $\pm$ 0.05 &4.73\\
+\hline
 \end{tabular}
-}
-\label{fig:bench1}
-\caption{Benchmark results of example interpreter computing the square of
-46340}
+\caption{Benchmarked function and results for the Python interpreter running
+\texttt{f(10000000)}}
 \end{figure}
-
-XXX insert some Python benchmarks
 
@@
 have a problem similar to that of DynamoRIO.  Targeting the C language makes
-higher-level specialisation difficult (e.g.\ \texttt{malloc} cannot be
-optimized).
+higher-level specialisation difficult.
 
@@
 
-Of course there is a lot of work still left to do XXX fix this phrase. Various optimizations are not
-quite finished. Both tracing and leaving machine code are very slow due to a
+A lot of work remains. We are working on two main optimizations at the moment.
+Those are:
+
+\textbf{Allocation Removal:} A key optimization for making the approach
+produce good code for more complex dynamic language is to perform escape
+analysis on the loop operation after tracing has been performed. In this way all
+objects that are allocated during the loop and do not actually escape the loop do
+not need to be allocated on the heap at all but can be exploded into their
+respective fields.  This is very helpful for dynamic languages where primitive
+types are often boxed, as the repeated allocation of intermediate results is
+very costly.
+
+\textbf{Optimizing Frame Objects:} One problem with the removal of allocations
+is that many dynamic languages are so reflective that they allow the
+introspection of the frame object that the interpreter uses to store local
+variables (e.g. SmallTalk, Python). This means that intermediate results always
+escape because they are stored into the frame object, rendering the allocation
+removal optimization ineffective. To remedy this problem we make it possible to
+update the frame object lazily only when it is actually accessed from outside of
+the code generated by the JIT.
+
+Furthermore both tracing and leaving machine code are very slow due to a
 double interpretation overhead and we might need techniques for improving those.
-Furthermore we need to apply the JIT to the various interpreters that are
+Eventually we will need to apply the JIT to the various interpreters that are
 written in RPython to evaluate how widely applicable the described techniques
 are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
-implementation \cite{bolz_back_2008}, a Prolog interpreter or PyGirl, a Gameboy
-emulator \cite{XXX}; but also not immediately obvious ones, like Python's
-regular expression engine.
+implementation \cite{bolz_back_2008}; a Prolog interpreter or PyGirl, a Gameboy
+emulator \cite{camillo_bruni_pygirl:_2009}; but also not immediately obvious
+ones, like Python's regular expression engine.
 
 If these experiments are successful we hope that we can reach a point where it
-becomes unnecessary to write a language specific JIT compiler and just apply a
-couple of hints to the interpreter to get reasonably good performance with
-relatively little effort.
+becomes unnecessary to write a language specific JIT compiler and instead
+possible to just apply a couple of hints to the interpreter to get reasonably
+good performance with relatively little effort.
 

### 2009-04-08 talk/icooolps2009/paper.tex
try to fix the XXX

@@
 The results show that the tracing JIT speeds up the execution of this Python
-function significantly, even outperforming CPython by a bit. The tracer needs to
+function significantly, even outperforming CPython. \sout{by a bit. The tracer needs to
 trace through quite a bit of dispatching machinery of the Python interpreter to
-achieve this, XXX.
+achieve this, XXX.}
+\anto{
+To achieve this, the tracer traces through the whole Python dispatching
+machinery, automatically inlining only the relevant fast paths.
+}
 

### 2009-04-08 talk/icooolps2009/paper.tex
some comments/fixes from toon

@@
 \newcommand\cfbolz[1]{\nb{CFB}{#1}}
+\newcommand\toon[1]{\nb{TOON}{#1}}
 \newcommand\anto[1]{\nb{ANTO}{#1}}
@@
 
+\toon{that part seems a bit out of place, it jumps back to a point in the story
+before what you explained right before it. (as far as I can tell)}
 During tracing, the trace is repeatedly
@@
 }
-To trace this, a bytecode form of these functions needs to be introduced that
+\toon{next sentence is strange} To trace this, a bytecode form of these functions needs to be introduced that
 the tracer understands. The tracer interprets a bytecode that is an encoding of
@@
 inference has been performed.
-When the profiler shows
+When the profiler discovers
 that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
@@
 The operations in this sequence are operations of the above-mentioned intermediate
-representation (e.g. note that the generic modulo and equality operations in the
+representation (e.g. the generic modulo and equality operations in the
 function above have been recognized to always take integers as arguments and are thus
@@
 to its beginning, forming an endless loop that can only be left via a guard
-failure. The call to \texttt{f} is inlined into the trace. Note that the trace
+failure. The call to \texttt{f} is inlined into the trace. The trace
 contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
@@
 Let's look at how hints would need to be applied to the example interpreter
-from Figure \ref{fig:tlr-basic}. To apply hints one generally needs a
+from Figure \ref{fig:tlr-basic}. To apply hints one needs a
 subclass of \texttt{JitDriver} that lists all the variables of the bytecode
@@
 \texttt{pc} variable is meaningless without the knowledge of which bytecode
-string is currently being interpreted. All other variables are red.
+string is currently being interpreted. All other variables are red (the fact
+that red variables need to be listed explicitly too is an implementation
+detail).
 
@@
 
+\toon{very difficult to read (actually so is the whole paragraph; rephrase)}
 The simple insight how to improve the situation is that most of the
@@
 but it is not yet finished).  Once we get this optimized trace, we can pass it to
-the \emph{JIT backend}, which generates the correspondent machine code.
+the \emph{JIT backend}, which generates the corresponding machine code.
 
@@
 translation process. In this way, the result of the translation is identical to
-that when no hints were present in the interpreter at all.
+that when no hints were present in the interpreter at all. \toon{strange
+sentence}
 
@@
 
-\textbf{Benchmark 1:} The interpreter translated to C without any JIT inserted at all.
+\textbf{Benchmark 1:} The interpreter translated to C without the JIT.
 

### 2009-04-08 talk/icooolps2009/paper.tex
a number of fixes

@@
 \newboolean{showcomments}
-\setboolean{showcomments}{false}
+\setboolean{showcomments}{true}
 \ifthenelse{\boolean{showcomments}}
@@
 machine code at runtime. The techniques were then successfully applied to Java
-VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
+VMs \cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006}. It also turned out that they are a
 relatively simple way to implement a JIT compiler for a dynamic language
@@
 code paths of commonly executed loops and to interpret the rest of the program.
-The code for those common loops however should be highly optimized, including
+The code for those common loops however is highly optimized, including
 aggressive inlining.
 
-Typically, programs executed by a tracing VMs goes through various phases:
+Typically, programs executed by a tracing VM go through various phases:
 \begin{itemize}
@@
 }
-\toon{next sentence is strange} To trace this, a bytecode form of these functions needs to be introduced that
-the tracer understands. The tracer interprets a bytecode that is an encoding of
+
+The tracer interprets these functions in a bytecode that is an encoding of
 the intermediate representation of PyPy's translation toolchain after type
@@
 \texttt{bytecode} argument is a string of bytes, all register and the
-accumulator are integers. A program for this interpreter that computes
+accumulator are integers.\footnote{The
+chain of \texttt{if}, \texttt{elif}, ... instructions that check the various
+opcodes is transformed into a \texttt{switch} statement by one of PyPy's
+optimizations. Python does not have a \texttt{switch} statement}
+A program for this interpreter that computes
 the square of the accumulator is shown in Figure \ref{fig:square}. If the
@@
 to the position key. This means that the loop will only be considered to be
-closed if these variables that are making up program counter at the language
+closed if these variables that are making up the program counter at the language
 interpreter level are the same a second time.  Loops found in this way are, by
@@
 
-\toon{very difficult to read (actually so is the whole paragraph; rephrase)}
-The simple insight how to improve the situation is that most of the
-operations in the trace are actually concerned with manipulating the
-bytecode and the program counter. Those are stored in variables that are part of
-the position key (they are ``green''), that means that the tracer checks that they
-are some fixed value at the beginning of the loop (they may well change over the
-course of the loop, though). In the example the check
-would be that the \texttt{bytecode} variable is the bytecode string
-corresponding to the square function and that the \texttt{pc} variable is
-\texttt{4}. Therefore it is possible to constant-fold computations on them away,
+The simple insight how to improve the situation is that most of the operations
+in the trace are actually concerned with manipulating the bytecode string and
+the program counter. Those are stored in variables that are ``green'' (e.g. they
+are part of the position key).  This means that the tracer checks that those
+variables have some fixed value at the beginning of the loop (they may well
+change over the course of the loop, though). In the example of Figure
+\ref{fig:trace-no-green-folding} the check would be that at the beginning of the
+trace the \texttt{pc} variable is \texttt{4} and the \texttt{bytecode} variable
+is the bytecode string corresponding to the square function. Therefore it is
+possible to constant-fold computations on them away,
 as long as the operations are side-effect free. Since strings are immutable in
@@
 hints that are possibly in the interpreter source are just ignored by the
-translation process. In this way, the result of the translation is identical to
-that when no hints were present in the interpreter at all. \toon{strange
-sentence}
+translation process.
 
@@
 The results show that the tracing JIT speeds up the execution of this Python
-function significantly, even outperforming CPython. \sout{by a bit. The tracer needs to
-trace through quite a bit of dispatching machinery of the Python interpreter to
-achieve this, XXX.}
-\anto{
-To achieve this, the tracer traces through the whole Python dispatching
-machinery, automatically inlining only the relevant fast paths.
-}
+function significantly, even outperforming CPython. To achieve this, the tracer
+traces through the whole Python dispatching machinery, automatically inlining
+the relevant fast paths.
 

### 2009-04-08 talk/icooolps2009/paper.tex
I think swapping those two paragraphs improves matters
grr. I tend to confuse i.e. and e.g.

@@
 \usepackage{ulem}
+\usepackage{xspace}
 \usepackage[utf8]{inputenc}
@@
 \newcommand{\commentout}[1]{}
+
+\newcommand\ie{i.e.,\xspace}
+\newcommand\eg{e.g.,\xspace}
 
@@
 languages that currently exist, PyPy's tracing JIT operates ``one level down'',
-e.g. it traces the execution of the interpreter, as opposed to the execution
+\ie it traces the execution of the interpreter, as opposed to the execution
 of the user program. The fact that the program the tracing JIT compiles is
@@
 point. The position key describes the position of the execution of the program,
-e.g. usually contains things like the function currently being executed and the
+\ie usually contains things like the function currently being executed and the
 program counter position of the tracing interpreter. The tracing interpreter
@@
 earlier, but only at instructions that are able to change the position key
-to an earlier value, e.g. a backward branch instruction. Note that this is
+to an earlier value, \eg a backward branch instruction. Note that this is
 already the second place where backward branches are treated specially: during
@@
 The operations in this sequence are operations of the above-mentioned intermediate
-representation (e.g. the generic modulo and equality operations in the
+representation (\eg the generic modulo and equality operations in the
 function above have been recognized to always take integers as arguments and are thus
@@
 
-The condition for reusing already existing machine code needs to be adapted to
+The language interpreter uses a similar technique to detect \emph{hot user
+loops}: the profiling is done at the backward branches of the user program,
+using one counter per seen program counter of the language interpreter.
+
+The condition for reusing already existing machine code also needs to be adapted to
 this new situation. In a classical tracing JIT there is at most one piece of
@@
 interpreter.
-
-The language interpreter uses a similar technique to detect \emph{hot user
-loops}: the profiling is done at the backward branches of the user program,
-using one counter per seen program counter of the language interpreter.
 
@@
 in the trace are actually concerned with manipulating the bytecode string and
-the program counter. Those are stored in variables that are ``green'' (e.g. they
+the program counter. Those are stored in variables that are ``green'' (\ie they
 are part of the position key).  This means that the tracer checks that those
@@
 which would make it hard to rebuild the state of the language interpreter and
-let it run from that point (e.g. this would involve building a potentially deep
+let it run from that point (\eg this would involve building a potentially deep
 C stack). Instead the falling back is achieved by a special \emph{fallback
@@
 trace in Figure \ref{fig:trace-full}. This speeds up the square function considerably,
-making it about six times faster than the pure interpreter.
+making it nearly three times faster than the pure interpreter.
 
@@
 it seems to be rather large, with about 20\% slowdown due to profiling. This is because the interpreter 
-is small and the opcodes simple. For larger interpreters (e.g. the Python one) it seems 
+is small and the opcodes simple. For larger interpreters (\eg the Python one) it seems 
 likely that the overhead is less significant.
@@
 \end{tabular}
+\caption{Benchmark results of example interpreter computing the square of
 \label{fig:bench1}
-\caption{Benchmark results of example interpreter computing the square of
 10000000}
@@
 benchmarked as well as the results can be seen in Figure
-\ref{fig:bench-example}. The function is a bit arbitrary but executing it is
+\ref{fig:bench-python}. The function is a bit arbitrary but executing it is
 still non-trivial, as a normal Python interpreter needs to dynamically dispatch
@@
 \end{tabular}
+
 \caption{Benchmarked function and results for the Python interpreter running
-\texttt{f(10000000)}}
+\texttt{f(10000000)} \label{fig:bench-python}}
 \end{figure}
@@
 This makes it necessary to add quite a large number of hints, because at the
-assembler level it is not really visible anymore that e.g. a bytecode string is
+assembler level it is not really visible anymore that \eg a bytecode string is
 immutable. Also more advanced optimizations like allocation removal would
@@
 introspection of the frame object that the interpreter uses to store local
-variables (e.g. SmallTalk, Python). This means that intermediate results always
+variables (\eg SmallTalk, Python). This means that intermediate results always
 escape because they are stored into the frame object, rendering the allocation

### 2009-04-08 talk/icooolps2009/paper.tex
tweaks to the benchmark section, thanks to toon again.

@@
 
-In this section we try to evaluate the work done so far by looking at some
+In this section we evaluate the work done so far by looking at some
 benchmark numbers. Since the work is not finished, these benchmarks can only be
@@
 confidence intervals were computed using a 95\% confidence level. All times
-include the running of the tracer and machine code production.
+include running the tracer and producing machine code.
 
 The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
-example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
-the square of 10000000 (the result will overflow, but for smaller numbers the
-running time is not long enough to sensibly measure it)
+example interpreter given in Figure \ref{fig:tlr-basic} computing
+the square of 10000000\footnote{the result will overflow, but for smaller numbers the
+running time is not long enough to sensibly measure it}
 using the bytecode of Figure \ref{fig:square}. The results for various
-constellations are as follows:
+configurations are as follows:
 
-\textbf{Benchmark 1:} The interpreter translated to C without the JIT.
+\textbf{Benchmark 1:} The interpreter translated to C without including a JIT
+compiler.
 
@@
 iterations.  As expected, this is not faster than the previous number. It is
-even quite a bit slower, probably due to the overheads of the JIT, as well as
+even quite a bit slower, probably due to the overheads, as well as
 non-optimal generated machine code.
 
-\textbf{Benchmark 3:} The hints as in Figure \ref{fig:tlr-full} are applied, which means the loop of
-the square function is reflected in the trace. Constant folding of green
-variables is disabled though. This corresponds to the trace in Figure
-\ref{fig:trace-no-green-folding}. This by alone brings no improvement over the
-previous case.
+\textbf{Benchmark 3:} The tracing JIT is enabled and hints as in Figure
+\ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
+so that it corresponds to the loop in the square function. However, constant folding of green
+variables is disabled, therefore the resulting machine code corresponds to the
+trace in Figure \ref{fig:trace-no-green-folding}. This by alone brings no
+improvement over the previous case.
 
@@
 it seems to be rather large, with about 20\% slowdown due to profiling. This is because the interpreter 
-is small and the opcodes simple. For larger interpreters (\eg the Python one) it seems 
-likely that the overhead is less significant.
+is small and the opcodes simple. For larger interpreters (\eg PyPy's Python
+interpreter) the overhead will likely be less significant.
 
@@
 
-Furthermore, to test the technique on a more realistic example, we did some
+To test the technique on a more realistic example, we did some
 preliminary benchmarks with PyPy's Python interpreter. The function we

### 2009-04-08 talk/icooolps2009/paper.tex
fix supposed holgerism

@@
 contains a \emph{guard} at every possible point where the path could have
-followed another direction, for example conditions or indirect/virtual
+followed another direction, for example conditions and indirect or virtual
 calls.  When generating the machine code, every guard is turned into a quick check

### 2009-04-08 talk/icooolps2009/paper.tex
more tweaks to the benchmark section

@@
 
-To test the technique on a more realistic example, we did some
-preliminary benchmarks with PyPy's Python interpreter. The function we
-benchmarked as well as the results can be seen in Figure
-\ref{fig:bench-python}. The function is a bit arbitrary but executing it is
-still non-trivial, as a normal Python interpreter needs to dynamically dispatch
-nearly all of the involved operations (like indexing into the tuple, addition
-and comparison of \texttt{i}). We benchmarked PyPy's Python interpreter with the
-JIT disabled, with the JIT enabled and
-CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
-Python).
-
-The results show that the tracing JIT speeds up the execution of this Python
-function significantly, even outperforming CPython. To achieve this, the tracer
-traces through the whole Python dispatching machinery, automatically inlining
-the relevant fast paths.
-
-\begin{figure}
+\begin{figure}[h]
 \label{fig:bench-example}
@@
 \end{figure}
+
+To test the technique on a more realistic example, we did some
+preliminary benchmarks with PyPy's Python interpreter. The function we
+benchmarked as well as the results can be seen in Figure
+\ref{fig:bench-python}. While the function may seem a bit arbitrary, executing it is
+still non-trivial, as a normal Python interpreter needs to dynamically dispatch
+nearly all of the involved operations, like indexing into the tuple, addition
+and comparison of \texttt{i}. We benchmarked PyPy's Python interpreter with the
+JIT disabled, with the JIT enabled and
+CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
+Python).
+
+The results show that the tracing JIT speeds up the execution of this Python
+function significantly, even outperforming CPython. To achieve this, the tracer
+traces through the whole Python dispatching machinery, automatically inlining
+the relevant fast paths.
 

### 2009-04-08 talk/icooolps2009/paper.tex
Move a figure to save some space (thanks toon). Actually reference another figure (thanks armin).
Benchmark Psyco

@@
 
+\begin{figure}
+\input{code/normal-tracing.txt}
+\caption{Trace when executing the \texttt{DECR\_A} opcode}
+\label{fig:trace-normal}
+\end{figure}
+
 An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
@@
 probably not improved at all.
-
-\begin{figure}
-\input{code/normal-tracing.txt}
-\caption{Trace when executing the \texttt{DECR\_A} opcode}
-\label{fig:trace-normal}
-\end{figure}
 
@@
 Let's look at how hints would need to be applied to the example interpreter
-from Figure \ref{fig:tlr-basic}. To apply hints one needs a
+from Figure \ref{fig:tlr-basic}. Figure \ref{fig:tlr-full} shows the relevant
+parts of the interpreter with hints applied. To apply hints one needs a
 subclass of \texttt{JitDriver} that lists all the variables of the bytecode
@@
 2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
-3 &CPython 2.5.4 &4.96 $\pm$ 0.05 &4.73\\
-\hline
+3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
+4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
 \end{tabular}
@@
 JIT disabled, with the JIT enabled and
-CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
-Python).
+CPython\footnote{\texttt{http://python.org}} 2.5.2 (the reference implementation of
+Python). In addition we benchmarked CPython using Psyco 1.6
+\cite{rigo_representation-based_2004}, a specializing JIT compiler for Python.
 
@@
 traces through the whole Python dispatching machinery, automatically inlining
-the relevant fast paths.
+the relevant fast paths. However, the manually tuned Psyco still performs a lot
+better than our prototype.
 

### 2009-04-08 talk/icooolps2009/paper.tex
Check in my changes, mostly typos, and let Carl Friedrich deal with the extra page :-)

@@
 from Figure \ref{fig:tlr-basic}. Figure \ref{fig:tlr-full} shows the relevant
-parts of the interpreter with hints applied. To apply hints one needs a
-subclass of \texttt{JitDriver} that lists all the variables of the bytecode
-loop. The variables are classified into two groups, red variables and green
+parts of the interpreter with hints applied.  One needs to instantiate
+\texttt{JitDriver} by listing all the variables of the bytecode loop.
+The variables are classified into two groups, ``green'' variables and ``red''
 variables. The green variables are those that the tracing JIT should consider to
 be part of the program counter of the language interpreter. In the case of the
-example, the \texttt{pc} variable is obviously part of the program counter.
-However, the \texttt{bytecode} variable is also counted as green, since the
+example, the \texttt{pc} variable is obviously part of the program counter;
+however, the \texttt{bytecode} variable is also counted as green, since the
 \texttt{pc} variable is meaningless without the knowledge of which bytecode
@@
 hints is essentially constant no matter how large the interpreter is, which
-makes it less significant for larger interpreters.
+makes the extra work negligible for larger interpreters.
 
@@
 
-The simple insight how to improve the situation is that most of the operations
+The simple insight on how to improve the situation is that most of the operations
 in the trace are actually concerned with manipulating the bytecode string and
@@
 
-From then on things proceed like described in Section \ref{sect:tracing}. The
+From then on things proceed as described in Section \ref{sect:tracing}. The
 tracing interpreter tries to find a loop in the user program, if it finds one it
@@
 runs until execution reaches a safe point where it is easy to let the C version
-of the language interpreter resume its operation. Usually this means that the
+of the language interpreter resume its operation.\footnote{This is the only
+reason for the \texttt{jit\_merge\_point} hint.}  This means that the
 fallback interpreter executes at most one bytecode operation of the language
-interpreter. After the language interpreter takes over again, the whole process
-starts again.
+interpreter and then falls back to the C version of the language interpreter.
+After this, the whole process of profiling may start again.
 
@@
 architectures (including, we hope, to virtual machines such as the JVM where
-backend could generate bytecode at runtime). At the moment the only implemented
-backend is a 32-bit Intel-x86 backend.
+our backend could generate JVM bytecode at runtime). At the moment the only
+implemented backend is a 32-bit Intel-x86 backend.
 
@@
 example interpreter given in Figure \ref{fig:tlr-basic} computing
-the square of 10000000\footnote{the result will overflow, but for smaller numbers the
-running time is not long enough to sensibly measure it}
+the square of 10000000\footnote{The result will overflow, but for smaller numbers the
+running time is not long enough to sensibly measure it.}
 using the bytecode of Figure \ref{fig:square}. The results for various
@@
 variables is disabled, therefore the resulting machine code corresponds to the
-trace in Figure \ref{fig:trace-no-green-folding}. This by alone brings no
+trace in Figure \ref{fig:trace-no-green-folding}. This alone brings no
 improvement over the previous case.
@@
 \hline
-1 &compiled to C, no JIT &442.7 $\pm$ 3.4 &1.00\\
+1 &Compiled to C, no JIT &442.7 $\pm$ 3.4 &1.00\\
 2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
@@
 \hline
-1 &Compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
-2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
+1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
+2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
 3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
@@
 the relevant fast paths. However, the manually tuned Psyco still performs a lot
-better than our prototype.
+better than our prototype (although it is interesting to note that Psyco
+improves the speed of CPython by only a factor 3.29 in this example, while
+our tracing JIT improves PyPy by a factor 6.54).
 
@@
 introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
-small dynamic language based on lambda-calculus. There is some work by one of
+small dynamic language based on lambda-calculus. It is also related to
+Psyco \cite{rigo_representation-based_2004}, a specializing JIT compiler for
+Python.  There is some work by one of
 the authors to implement a dynamic partial evaluator for Prolog
@@
 PyPy project to use dynamic partial evaluation for automatically generating JIT
-compilers out of interpreters \cite{armin_rigo_jit_2007}. So far those have not been as
+compilers out of interpreters \cite{armin_rigo_jit_2007}\cite{antocuni_2009}.
+So far those have not been as
 successful as we would like and it seems likely that they will be supplanted
@@
 double interpretation overhead and we might need techniques for improving those.
+
 Eventually we will need to apply the JIT to the various interpreters that are
@@
 are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
-implementation \cite{bolz_back_2008}; a Prolog interpreter or PyGirl, a Gameboy
-emulator \cite{camillo_bruni_pygirl:_2009}; but also not immediately obvious
+implementation \cite{bolz_back_2008}; a Prolog interpreter; PyGirl, a Gameboy
+emulator \cite{camillo_bruni_pygirl:_2009}; and also not immediately obvious
 ones, like Python's regular expression engine.

### 2009-04-08 talk/icooolps2009/paper.tex
switch armin and maciek to have alphabetical order. Try a new abstract (mostly by toon, again :-) ). It's not done but better than the old.

@@
 \and
+\alignauthor Maciej Fijalkowski\\
+       \email{fijal@merlinux.eu}
 \alignauthor Armin Rigo\\
        \email{arigo@tunes.org}
-\alignauthor Maciej Fijalkowski\\
-       \email{fijal@merlinux.eu}
 }
@@
 \begin{abstract}
-We present techniques for improving the results when a tracing JIT compiler is
-applied to an interpreter.XXXAbrupt start - is the relevanz really immediately clear?XXX An unmodified tracing JIT performs not as well as one
-would hope when the compiled program is itself a bytecode interpreter. We
-examine the reasons for that, and how matters can be improved by adding markers to
-the interpreter that help the tracing JIT to improve the results. We evaluate
-the techniques by using them both on a small example as well as on a full Python
-interpreter. This work has been done in the context of the PyPy project.
+XXX the first paragraph is a bit too long, the second too strange, but I think
+it is a step in the right direction.
+
+Tracing JIT compilers can greatly speed up programs that spend most of their time
+in loops in which they take similar code paths.  Applying an unmodified tracing
+JIT to an application that is itself a bytecode interpreter results in very
+limited or no speedup. While bytecode interpreters spend most of their time in a
+loop, \ie the bytecode dispatch, the code paths taken depend fully on the input
+bytecode which the interpreter evaluates and thus changes all the time.
+
+In this paper we show how to guide tracing JIT compilers to greatly improve
+the speed of bytecode interpreters towards input bytecode. We evaluate our
+technique by applying it to both a small example as well as to a full Python 
+interpreter. This research has been conducted in the context of the PyPy 
+project.
 

### 2009-04-08 talk/icooolps2009/paper.tex
Attempt to reformulate the abstract...

@@
 \begin{abstract}
-XXX the first paragraph is a bit too long, the second too strange, but I think
-it is a step in the right direction.
 
-Tracing JIT compilers can greatly speed up programs that spend most of their time
-in loops in which they take similar code paths.  Applying an unmodified tracing
-JIT to an application that is itself a bytecode interpreter results in very
-limited or no speedup. While bytecode interpreters spend most of their time in a
-loop, \ie the bytecode dispatch, the code paths taken depend fully on the input
-bytecode which the interpreter evaluates and thus changes all the time.
+We attempt to use the technique of Tracing JIT Compilers
+(see e.g.\ \cite{andreas_gal_incremental_2006}) in the context
+of the PyPy project, \ie on programs that are interpreters for some
+dynamic language (including Python).  Tracing JIT compilers can greatly
+speed up programs that spend most of their time in loops in which they
+take similar code paths.  However, applying an unmodified tracing JIT to
+a program that is itself a bytecode interpreter results in very limited
+or no speedup.  One of the reasons is that their main loop, \ie the
+bytecode dispatch, always executes different bytecodes and so follows
+different paths.
 
-In this paper we show how to guide tracing JIT compilers to greatly improve
-the speed of bytecode interpreters towards input bytecode. We evaluate our
-technique by applying it to both a small example as well as to a full Python 
-interpreter. This research has been conducted in the context of the PyPy 
-project.
+In this paper we show how to guide tracing JIT compilers to greatly
+improve the speed of bytecode interpreters. We evaluate our technique by
+applying it to two PyPy interpreters: one is a small example, and the
+other one is the full Python interpreter.
 
@@
 PyPy project to use dynamic partial evaluation for automatically generating JIT
-compilers out of interpreters \cite{armin_rigo_jit_2007}\cite{antocuni_2009}.
+compilers out of interpreters \cite{armin_rigo_jit_2007, antocuni_2009}.
 So far those have not been as

### 2009-04-08 talk/icooolps2009/paper.tex
try to refactor section 3 to contain less repetition and to fix the "backjumping" that toon disliked (and that michael complained about too).

@@
 
-The \emph{code generation} phase takes as input the trace generated during
-\emph{tracing}.
-
 At first, when the program starts, everything is interpreted.
@@
 
-When a hot loop is identified, the interpreter enters a
-special mode, called \emph{tracing mode}. During tracing, the interpreter
-records a history of all the operations it executes.
+When a hot loop is identified, the interpreter enters a special mode, called
+\emph{tracing mode}. During tracing, the interpreter records a history of all
+the operations it executes. It traces until it has recorded the execution of one
+iteration of the hot loop. To decide when this is the case, the trace is
+repeatedly checked during tracing as to whether the interpreter is at a position
+in the program where it had been earlier.
 
-Such a history is called a \emph{trace}: it is a sequential list of
+The history recorded by the tracer is called a \emph{trace}: it is a sequential list of
 operations, together with their actual operands and results.  By examining the
-trace, it is possible to produce highly efficient machine code by generating
-only the operations needed.  Being sequential, the trace represents only one
+trace, it is possible to produce efficient machine code by generating
+code from the operations in it.  The machine code can then be executed immediately,
+starting from the next iteration of the loop, as the machine code represents
+exactly the loop that was being interpreted so far.
+
+Being sequential, the trace represents only one
 of the many possible paths through the code. To ensure correctness, the trace
@@
 we immediately quit the machine code and continue the execution by falling
-back to interpretation.
-
-\toon{that part seems a bit out of place, it jumps back to a point in the story
-before what you explained right before it. (as far as I can tell)}
-During tracing, the trace is repeatedly
-checked as to whether the interpreter is at a position in the program where it had been
-earlier. If this happens, the trace recorded corresponds to a loop
-in the interpreted program. At this point, this loop
-is turned into machine code by taking the trace and making machine code versions
-of all the operations in it. The machine code can then be executed immediately,
-starting from the next iteration of the loop, as the machine code represents
-exactly the loop that was being interpreted so far.
-
-This process assumes that the path through the loop that was traced is a
-``typical'' example of possible paths. Of course
-it is possible that later another path through the loop is taken, in which case
-one of the guards that were put into the machine code will fail.\footnote{There are more
-complex mechanisms in place to still produce more code for the cases of guard
-failures \cite{andreas_gal_incremental_2006}, but they are independent of the issues discussed in this
-paper.}
+back to interpretation.\footnote{There are more complex mechanisms in place to
+still produce more code for the cases of guard failures
+\cite{andreas_gal_incremental_2006}, but they are independent of the issues
+discussed in this paper.}
 

### 2009-04-08 talk/icooolps2009/paper.tex
Add a sentence.
Typo.
try to streamline the beginning of the abstract a bit and kill a sentence that goes into a bit too much detail about the problem

@@
 We attempt to use the technique of Tracing JIT Compilers
-(see e.g.\ \cite{andreas_gal_incremental_2006}) in the context
+\cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006,
+mason_chang_efficient_2007} in the context
 of the PyPy project, \ie on programs that are interpreters for some
-dynamic language (including Python).  Tracing JIT compilers can greatly
+dynamic languages, including Python.  Tracing JIT compilers can greatly
 speed up programs that spend most of their time in loops in which they
@@
 a program that is itself a bytecode interpreter results in very limited
-or no speedup.  One of the reasons is that their main loop, \ie the
-bytecode dispatch, always executes different bytecodes and so follows
-different paths.
-
+or no speedup.
 In this paper we show how to guide tracing JIT compilers to greatly
-improve the speed of bytecode interpreters. We evaluate our technique by
+improve the speed of bytecode interpreters.  One crucial point is to
+unroll the bytecode dispatch loop, based on two hints provided by the
+implementer of the bytecode interpreter.  We evaluate our technique by
 applying it to two PyPy interpreters: one is a small example, and the

### 2009-04-08 talk/icooolps2009/paper.tex
Typo typo.

@@
 better than our prototype (although it is interesting to note that Psyco
-improves the speed of CPython by only a factor 3.29 in this example, while
-our tracing JIT improves PyPy by a factor 6.54).
+improves the speed of CPython by only a factor of 3.29 in this example, while
+our tracing JIT improves PyPy by a factor of 6.54).
 

### 2009-04-08 talk/icooolps2009/paper.tex
Fix it.
add merlinux
fix some things pointed out by toon

@@
 
-\numberofauthors{3}
+\numberofauthors{4}
 \author{
 \alignauthor Carl Friedrich Bolz\\
-       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
+       \affaddr{University of Düsseldorf}\\
        \affaddr{STUPS Group}\\
@@
 \alignauthor Maciej Fijalkowski\\
+       \affaddr{merlinux GmbH}\\
        \email{fijal@merlinux.eu}
@@
 We attempt to use the technique of Tracing JIT Compilers
-\cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006,
-mason_chang_efficient_2007} in the context
+in the context
 of the PyPy project, \ie on programs that are interpreters for some
@@
 
-Tracing JITs are an idea initially explored by the Dynamo project
-\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
-machine code at runtime. The techniques were then successfully applied to Java
-VMs \cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006}. It also turned out that they are a
-relatively simple way to implement a JIT compiler for a dynamic language
+Tracing optimizations were initially explored by the Dynamo project
+\cite{bala_dynamo:transparent_2000} to dynamically optimize
+machine code at runtime. Its techniques were then successfully used to implement
+a JIT compiler for a Java
+VM \cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006}.
+Subsequently these tracing JITs were discovered to be a
+relatively simple way to implement JIT compilers for dynamic languages
 \cite{mason_chang_efficient_2007}. The technique is now
@@
 
-Typically, programs executed by a tracing VM go through various phases:
+Typically tracing VMs go through various phases when they execute a program:
 \begin{itemize}
@@
 each backward jump instruction that counts how often this particular backward jump
-was executed. Since loops need a backward jump somewhere, this method looks for
+is executed. Since loops need a backward jump somewhere, this method looks for
 loops in the user program.
@@
 The history recorded by the tracer is called a \emph{trace}: it is a sequential list of
-operations, together with their actual operands and results.  By examining the
-trace, it is possible to produce efficient machine code by generating
-code from the operations in it.  The machine code can then be executed immediately,
-starting from the next iteration of the loop, as the machine code represents
-exactly the loop that was being interpreted so far.
+operations, together with their actual operands and results. Such a trace can be
+used to generate efficient machine code. This generated machine code is
+immediately executable, and can be used in the next iteration of the loop.
 
@@
 contains a \emph{guard} at every possible point where the path could have
-followed another direction, for example conditions and indirect or virtual
+followed another direction, for example at conditions and indirect or virtual
 calls.  When generating the machine code, every guard is turned into a quick check
@@
 back to interpretation.\footnote{There are more complex mechanisms in place to
-still produce more code for the cases of guard failures
+still produce extra code for the cases of guard failures
 \cite{andreas_gal_incremental_2006}, but they are independent of the issues
@@
 
-Let's look at a small example. Take the following (slightly contrived) RPython
-code:
+\begin{figure}
 {\small
@@
     return result
-\end{verbatim}
-}
 
-The tracer interprets these functions in a bytecode that is an encoding of
-the intermediate representation of PyPy's translation toolchain after type
-inference has been performed.
-When the profiler discovers
-that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
-tracing JIT will start to trace the execution of that loop.  The trace would
-look as follows:
-{\small
-\begin{verbatim}
+# corresponding trace:
 loop_header(result0, n0)
@@
 }
+\caption{A simple Python function and the recorded trace.}
+\label{fig:simple-trace}
+\end{figure}
+
+Let's look at a small example. Take the (slightly contrived) RPython code in
+Figure \ref{fig:simple-trace}.
+The tracer interprets these functions in a bytecode format that is an encoding of
+the intermediate representation of PyPy's translation tool\-chain after type
+inference has been performed.
+When the profiler discovers
+that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
+tracing JIT will start to trace the execution of that loop.  The trace would
+look as in the lower half of Figure \ref{fig:simple-trace}.
+
 The operations in this sequence are operations of the above-mentioned intermediate
@@
 example interpreter given in Figure \ref{fig:tlr-basic} computing
-the square of 10000000\footnote{The result will overflow, but for smaller numbers the
+the square of 10000000 using the bytecode of Figure \ref{fig:square}.\footnote{The result will overflow, but for smaller numbers the
 running time is not long enough to sensibly measure it.}
-using the bytecode of Figure \ref{fig:square}. The results for various
+The results for various
 configurations are as follows:
@@
 variables is disabled, therefore the resulting machine code corresponds to the
-trace in Figure \ref{fig:trace-no-green-folding}. This alone brings no
-improvement over the previous case.
+trace in Figure \ref{fig:trace-no-green-folding}. This alone brings an
+improvement over the previous case, but is still slower than pure
+interpretation.
 

### 2009-04-08 talk/icooolps2009/paper.tex
whack a bit till stuff fits into 8 pages

@@
 accumulator are integers.\footnote{The
-chain of \texttt{if}, \texttt{elif}, ... instructions that check the various
-opcodes is transformed into a \texttt{switch} statement by one of PyPy's
-optimizations. Python does not have a \texttt{switch} statement}
+chain of \texttt{if}, \texttt{elif}, ... instructions checking the various
+opcodes is turned into a \texttt{switch} statement by one of PyPy's
+optimizations. Python does not have a \texttt{switch} statement.}
 A program for this interpreter that computes
@@
 Machine code production is done via a well-defined interface to an assembler
-backend. This makes it possible to easily port the tracing JIT to various
+backend. This allows easy porting of the tracing JIT to various
 architectures (including, we hope, to virtual machines such as the JVM where
@@
 In this section we evaluate the work done so far by looking at some
-benchmark numbers. Since the work is not finished, these benchmarks can only be
+benchmarks. Since the work is not finished, these can only be
 preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
@@
 \ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
-so that it corresponds to the loop in the square function. However, constant folding of green
+so that it corresponds to the loop in the square function. Constant folding of green
 variables is disabled, therefore the resulting machine code corresponds to the

### 2009-04-08 talk/icooolps2009/paper.tex
a final tweak

@@
 
-We attempt to use the technique of Tracing JIT Compilers
+We attempt to apply the technique of Tracing JIT Compilers
 in the context
-of the PyPy project, \ie on programs that are interpreters for some
+of the PyPy project, \ie to programs that are interpreters for some
 dynamic languages, including Python.  Tracing JIT compilers can greatly

### 2009-06-09 talk/icooolps2009/paper.tex
starting to incorporate reviewer comments

@@
 improve the speed of bytecode interpreters.  One crucial point is to
-unroll the bytecode dispatch loop, based on two hints provided by the
+unroll the bytecode dispatch loop, based on two kinds of hints provided by the
 implementer of the bytecode interpreter.  We evaluate our technique by
@@
 \cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} is an
-environment where flexible implementation of dynamic languages can be written.
+environment where flexible implementations of dynamic languages can be written.
 To implement a dynamic language with PyPy, an interpreter for that language has
@@
 the other operations the guard will fail, which will mean that performance is
-probably not improved at all.
+not improved at all.
 
@@
 For the small example the hints look like a lot of work. However, the number of
-hints is essentially constant no matter how large the interpreter is, which
+hints that need to be put into the interpreter source remains small, which
 makes the extra work negligible for larger interpreters.
@@
 benchmarked as well as the results can be seen in Figure
-\ref{fig:bench-python}. While the function may seem a bit arbitrary, executing it is
+\ref{fig:bench-python}. While the function may seem a bit constructed, executing it is
 still non-trivial, as a normal Python interpreter needs to dynamically dispatch

### 2009-06-09 talk/icooolps2009/paper.tex
add acknowledgements
change the acks slightly
Simplify hints a bit, and add a footnote that it is different in reality. They confused reviewers quite a bit.

@@
 In addition to the classification of the variables, there are two methods of
-\texttt{JitDriver} that need to be called. Both of them receive as arguments the
-current values of the variables listed in the definition of the driver. The
+\texttt{JitDriver} that need to be called. The
 first one is \texttt{jit\_merge\_point} which needs to be put at the beginning
@@
 instruction that can set the program counter of the language interpreter to an
-earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
+earlier value.\footnote{The hints need to be written a bit differently in the
+actual implementation for purely technical reasons.} For the example this is
+only the \texttt{JUMP\_IF\_A}
 instruction, and only if it is actually a backward jump. Here is
@@
 
+\section*{Acknowledgements}
+The authors would like to thank Michael Leuschel, Toon Verwaest and the
+referees of ICOOOLPS'09 for helpful comments on earlier versions of this paper.
+
 \bibliographystyle{abbrv}

### 2009-06-09 talk/icooolps2009/paper.tex
add a paragraph about typical techniques to get rid of interpreter dispatch overhead

@@
 
+There are quite an number of approaches that try to minimally enhance
+interpreters to generate code at runtime without actually writing a native
+compiler by hand. The goal of these is to get rid of dispatch overhead of
+typical interpreters while retaining ease of implementation. Piumarta and
+Riccardi \cite{piumarta_optimizing_1998} propose to copy fragments of the
+interpreter together for commonly occurring bytecode sequences to reduce
+dispatch overhead. However, dispatching is still needed to jump between such
+sequences and also when non-copyable bytecodes occur. Ertl and Gregg
+\cite{ertl_retargeting_2004} go further and stitch together the concatenated
+sequences by patching the copied machine code. Thus they get rid of all dispatch
+overhead. Both techniques can speed up interpreters which large dispatch
+overhead a lot. However they will help less if the bytecodes themselves do a
+lot of work (as is the case with Python), whereas our tracing technique can
+trace into the implementation of those bytecodes.
+
 The standard approach for automatically producing a compiler for a programming

### 2009-06-09 talk/icooolps2009/paper.tex
Changes (thanks Anto).

@@
 
-There are quite an number of approaches that try to minimally enhance
+There are quite a number of approaches that try to minimally enhance
 interpreters to generate code at runtime without actually writing a native
@@
 overhead a lot. However they will help less if the bytecodes themselves do a
-lot of work (as is the case with Python), whereas our tracing technique can
-trace into the implementation of those bytecodes.
+lot of work (as is the case with Python) and the dispatch overhead is lower. On
+the other hand, our technique can do a better job by tracing inside the
+implementation of those bytecode and inlining common paths.
 

### 2009-06-09 talk/icooolps2009-dotnet/paper.tex
don't show comments
try to incorporate some of the reviewers' comments

@@
 The Common Language Infrastructure (CLI) is a virtual machine expressly
-designed for implementing statically typed languages as C\#, therefore
+designed for implementing statically typed languages such as C\#, therefore
 programs written in dynamically typed languages are typically much slower than C\# when executed on .NET.

### 2009-06-09 talk/icooolps2009-dotnet/conclusion.tex
fix last davide's comment
try to incorporate some of the reviewers' comments
apply ACM categories, add copyright info and acknowledgements (mostly stolen from cfbolz :-))

@@
 \cite{DBLP:conf/pepm/Rigo04}). However, Psyco is a manually written JIT, is
-not applicable to other languages and cannot be retargetted.
+not applicable to other languages and cannot be retargetted.  Psyco is a 
+good example of how to implement flexswitches for targets that don't have the
+limitations of the CLI.
 
@@
 runtime feedback to produce more efficient code
-\cite{hoelzle_type_feedback_1994}.
+\cite{hoelzle_type_feedback_1994}.  The main difference between the two is 
+that PICs only works on types, whereas promotion can work on every kind of value.
 
@@
 
-\anto{XXX: please review the last two paragraphs}
-
 \section{Conclusion and Future Work}
@@
 generate even more efficient code.
+
+\section*{Acknowledgements}
+
+The authors would like to thank Carl Friedrich Bolz, Maciej Fijalkowski and
+the referees of ICOOOLPS'09 for helpful comments on earlier versions of this
+paper.

### 2009-06-10 talk/icooolps2009/code/tlr-paper-full.py
Now that we need 9 pages anyway, make the font non-small again. Add another reference.

@@ -1,4 +1,4 @@
-{\small
+{
 \begin{verbatim}
 tlrjitdriver = JitDriver(greens = ['pc', 'bytecode'],
                          reds   = ['a', 'regs'])

### 2009-06-10 talk/icooolps2009/paper.tex
Enabling ACM categories.
Trying to get the copyright info right. No clue whether it is correct.
Now that we need 9 pages anyway, make the font non-small again. Add another reference.

@@
 }
+\conferenceinfo{ICOOOLPS}{'09 Genova, Italy}
+\CopyrightYear{2009}
+\crdata{978-1-60558-541-3/09/07}
+
 \maketitle
 
-%\category{D.3.4}{Programming Languages}{Processors}[code generation,
-%interpreters, run-time environments]
-%\category{F.3.2}{Logics and Meanings of Programs}{Semantics of Programming
-%Languages}[program analysis]
+\category{D.3.4}{Programming Languages}{Processors}[code generation,
+incremental compilers, interpreters, run-time environments]
 
@@
 \begin{figure}
-{\small
+{
 \begin{verbatim}
@@
 \label{fig:bench-example}
-{\small
+{
 \begin{verbatim}
@@
 overhead a lot. However they will help less if the bytecodes themselves do a
-lot of work (as is the case with Python) and the dispatch overhead is lower. On
+lot of work (as is the case with Python
+\cite{stefan_brunthaler_virtual-machine_2009}) and the dispatch overhead is lower. On
 the other hand, our technique can do a better job by tracing inside the

### 2009-06-10 talk/icooolps2009/code/tlr-paper.py

@@ -1,4 +1,4 @@
-{\small
+{
 \begin{verbatim}
 def interpret(bytecode, a):
     regs = [0] * 256

### 2009-06-10 talk/icooolps2009/code/tlr-paper-full.py
undo the smallification of the last commit.

@@ -1,4 +1,4 @@
-{
+{\small
 \begin{verbatim}
 tlrjitdriver = JitDriver(greens = ['pc', 'bytecode'],
                          reds   = ['a', 'regs'])

### 2009-06-10 talk/icooolps2009/code/tlr-paper.py

@@ -1,4 +1,4 @@
-{
+{\small
 \begin{verbatim}
 def interpret(bytecode, a):
     regs = [0] * 256

### 2009-06-10 talk/icooolps2009/paper.tex
undo the smallification of the last commit.
Completely horrible hacks to squeeze everything on 8 pages again.

@@
 Typically tracing VMs go through various phases when they execute a program:
-\begin{itemize}
-\item Interpretation/profiling
-\item Tracing
-\item Code generation
-\item Execution of the generated code
-\end{itemize}
-
-At first, when the program starts, everything is interpreted.
-The interpreter does a small amount of lightweight profiling to establish which loops
+Interpretation/profiling, tracing, code generation and execution of the
+generated code. When the program starts, everything is interpreted.
+The interpreter does lightweight profiling to establish which loops
 are run most frequently. This lightweight profiling is usually done by having a counter on
@@
 iteration of the hot loop. To decide when this is the case, the trace is
-repeatedly checked during tracing as to whether the interpreter is at a position
+repeatedly checked as to whether the interpreter is at a position
 in the program where it had been earlier.
 
-The history recorded by the tracer is called a \emph{trace}: it is a sequential list of
+The history recorded by the tracer is called a \emph{trace}: it is a list of
 operations, together with their actual operands and results. Such a trace can be
@@
 \begin{figure}
-{
+{\small
 \begin{verbatim}
@@
 
-Let's look at a small example. Take the (slightly contrived) RPython code in
+As a small example, take the (slightly contrived) RPython code in
 Figure \ref{fig:simple-trace}.
@@
 program} (from the point of view of a VM author, the ``user'' is a programmer
-using the VM).
-
-Similarly, we need to distinguish loops at two different levels:
+using the VM). Similarly, we need to distinguish loops at two different levels:
 \emph{interpreter loops} are loops \emph{inside} the language interpreter. On
@@
 A tracing JIT compiler finds the hot loops of the program it is compiling. In
-our case, this program is the language interpreter. The most important hot interpreter loop
+our case, this is the language interpreter. The most important hot interpreter loop
 is the bytecode dispatch loop (for many simple
-interpreters it is also the only hot loop).  Tracing one iteration of this
-loop means that
-the recorded trace corresponds to execution of one opcode. This means that the
+interpreters it is also the only hot loop).  One iteration of this loop
+corresponds to the execution of one opcode. This means that the
 assumption made by the tracing JIT -- that several iterations of a hot loop
@@
 
-The condition for reusing already existing machine code also needs to be adapted to
-this new situation. In a classical tracing JIT there is at most one piece of
+The condition for reusing existing machine code also needs to be adapted to
+this new situation. In a classical tracing JIT there is no or one piece of
 assembler code per loop of the jitted program, which in our case is the language
@@
 \texttt{pc} variable is meaningless without the knowledge of which bytecode
-string is currently being interpreted. All other variables are red (the fact
-that red variables need to be listed explicitly too is an implementation
-detail).
+string is currently being interpreted. All other variables are red.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \input{code/no-green-folding.txt}
@@
 \label{fig:bench-example}
-{
+{\small
 \begin{verbatim}
@@
 sequences and also when non-copyable bytecodes occur. Ertl and Gregg
-\cite{ertl_retargeting_2004} go further and stitch together the concatenated
-sequences by patching the copied machine code. Thus they get rid of all dispatch
-overhead. Both techniques can speed up interpreters which large dispatch
+\cite{ertl_retargeting_2004} go further and get rid of all dispatch overhead by
+stitching together the concatenated
+sequences by patching the copied machine code. 
+Both techniques can speed up interpreters which large dispatch
 overhead a lot. However they will help less if the bytecodes themselves do a
@@
 
-Furthermore both tracing and leaving machine code are very slow due to a
-double interpretation overhead and we might need techniques for improving those.
-
 Eventually we will need to apply the JIT to the various interpreters that are

### 2009-07-02 talk/icooolps2009/talk/talk.tex
Start with the ICOOOLPS slides for the tracing talk. This is how I have given the talk in Bad Honnef, needs updating.

\documentclass[utf8x]{beamer}

\mode<presentation>
{
  \usetheme{Warsaw}

  %\setbeamercovered{transparent}
}

\usepackage[english]{babel}

\usepackage[utf8x]{inputenc}

\usepackage{times}
\usepackage[T1]{fontenc}

\title[PyPy's Tracing JIT Compiler]{
    Tracing the Meta-Level: PyPy's Tracing JIT Compiler
}
\author[Bolz, Cuni, Fijalkowski, Rigo]
{
    \textcolor{green!50!black}{Carl~Friedrich~Bolz}\inst{1} \and
    Antonio Cuni\inst{2} \and
    Maciej Fijalkowski\inst{3} \and
    Armin Rigo
}

\institute[Düsseldorf]
{
    \inst{1}
    Softwaretechnik und Programmiersprachen\\ Heinrich-Heine-Universit\"at D\"usseldorf
    \and%
    \vskip-2mm
    \inst{2}
    University of Genova, Italy
    \and%
    \vskip-2mm
    \inst{3}
    merlinux GmbH
}

\date{ICOOOLPS 2009 XXX}

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
%\AtBeginSubsection[]
%{
%  \begin{frame}<beamer>
%    \frametitle{Outline}
%    \tableofcontents[currentsection,currentsubsection]
%  \end{frame}
%}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

%\begin{frame}
%  \frametitle{Outline}
%  \tableofcontents
  % You might wish to add the option [pausesections]
%\end{frame}

\begin{frame}
    \frametitle{Motivation}
    \begin{itemize}
    \item writing good JIT compilers for dynamic programming languages is hard and error-prone
    \item tracing JIT compilers are a new approach to JITs that are supposed to be easier
    \item what happens when a tracing JIT is applied ``one level down'', i.e. to an interpreter
    \item how to solve the occurring problems
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Context: The PyPy Project}
    \begin{itemize}
    \item a general environment for implementing dynamic languages
    \item contains a compiler for a subset of Python (``RPython'')
    \item interpreters for dynamic languages written in that subset
    \item various interpreters written with PyPy: Python, Prolog, Smalltalk, Scheme, JavaScript, GameBoy emulator
    \item can be translated to a variety of target environment: C, JVM, .NET
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Tracing JIT Compilers}
    \begin{itemize}
    \item idea from Dynamo project: dynamic rewriting of machine code
    \item later used for a lightweight Java JIT
    \item seems to also work for dynamic languages
    \item incorporated into Mozilla's JavaScript VM ("TraceMonkey")
    \end{itemize}
    \pause
    \begin{block}{Basic Assumption of a Tracing JIT}
        \begin{itemize}
        \item programs spend most of their time executing loops
        \item several iterations of a loop are likely to take similar code paths
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Tracing JIT Compilers}
    \begin{itemize}
    \item mixed-mode execuction environment
    \item at first, everything is interpreted
    \item lightweight profiling to discover hot loops
    \item code generation only for common paths of hot loops
    \item when a hot loop is discovered, start to produce a trace
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Tracing}
    \begin{itemize}
    \item a \emph{trace} is a sequential list of operations
    \item a trace is produced by recording every operation the interpreter executes
    \item tracing ends when the tracer sees a position in the program it has seen before
    \item to identify these places, the \emph{position key} is used
    \item the position key encodes the current point of execution
    \item a trace thus corresponds to exactly one loop
    \item that means it ends with a jump to its beginning
    \end{itemize}
    \pause
    \begin{block}{Guards}
        \begin{itemize}
        \item the trace is only one of the possible code paths through the loop
        \item at places where the path \emph{could} diverge, a guard is placed
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Code Generation and Execution}
    \begin{itemize}
    \item being linear, the trace can easily be turned into machine code
    \item the machine code can be immediately executed
    \item execution stops when a guard fails
    \item after a guard failure, go back to interpreting program
    \end{itemize}
\end{frame}

\frame[containsverbatim, plain, shrink=10]{
  \frametitle{Example}
  \begin{verbatim}
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result

def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b

\end{verbatim}
}

\frame[containsverbatim, plain, shrink=10]{
  \frametitle{Example}
  \begin{verbatim}
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result

def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b

# loop_header(result0, n0)
# i0 = int_mod(n0, Const(46))
# i1 = int_eq(i0, Const(41))
# guard_false(i1)
# result1 = int_add(result0, n0)
# n1 = int_sub(n0, Const(1))
# i2 = int_ge(n1, Const(0))
# guard_true(i2)
# jump(result1, n1)
\end{verbatim}
}

\begin{frame}
    \frametitle{(Dis-)Advantages of Tracing JITs}
    \begin{block}{Good Points of the Approach}
        \begin{itemize}
        \item easy and fast machine code generation: needs so support only one path
        \item interpreter does a lot of the work
        \item can be added to an existing interpreter unobtrusively
        \item automatic inlining
        \item produces very little code
        \end{itemize}
    \end{block}
    \pause
    \begin{block}{Bad Points of the Approach}
        \begin{itemize}
        \item unclear whether assumptions are true often enough
        \item switching between interpretation and machine code execution takes time
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Applying a Tracing JIT to an Interpreter}
    \begin{itemize}
    \item Question: What happens if the program is itself a bytecode interpreter?
    \item the (usually only) hot loop of a bytecode interpreter is the bytecode dispatch loop
    \item Assumption violated: two iterations of the dispatch loop will usually take very different code paths
    \end{itemize}
    \pause
    \begin{block}{Terminology}
        \begin{itemize}
        \item \emph{tracing interpreter:} the interpreter that originally runs the program and produces traces
        \item \emph{language interpreter:} the bytecode interpreter runs on top
        \item \emph{user program:} the program run by the language interpreter
        \end{itemize}
    \end{block}
\end{frame}

\frame[containsverbatim, plain, shrink=10]{
  \begin{verbatim}
def interpret(bytecode, a):
    regs = [0] * 256
    pc = 0
    while True:
        opcode = ord(bytecode[pc])
        pc += 1
        if opcode == JUMP_IF_A:
            target = ord(bytecode[pc])
            pc += 1
            if a:
                pc = target
        elif opcode == MOV_A_R:
            n = ord(bytecode[pc])
            pc += 1
            regs[n] = a
        elif opcode == MOV_R_A:
            n = ord(bytecode[pc])
            pc += 1
            a = regs[n]
        elif opcode == ADD_R_TO_A:
            n = ord(bytecode[pc])
            pc += 1
            a += regs[n]
        elif opcode == DECR_A:
            a -= 1
        elif opcode == RETURN_A:
            return a
  \end{verbatim}
}

\frame[containsverbatim, plain, shrink=10]{
  \begin{verbatim}
def interpret(bytecode, a):             |
    regs = [0] * 256                    |  # Example bytecode
    pc = 0                              |  # Square the accumulator:
    while True:                         |
        opcode = ord(bytecode[pc])      |  MOV_A_R     0   # i = a
        pc += 1                         |  MOV_A_R     1   # copy of 'a'
        if opcode == JUMP_IF_A:         |
            target = ord(bytecode[pc])  |  # 4:
            pc += 1                     |  MOV_R_A     0   # i--
            if a:                       |  DECR_A
                pc = target             |  MOV_A_R     0
        elif opcode == MOV_A_R:         |
            n = ord(bytecode[pc])       |  MOV_R_A     2   # res += a
            pc += 1                     |  ADD_R_TO_A  1
            regs[n] = a                 |  MOV_A_R     2
        elif opcode == MOV_R_A:         |
            n = ord(bytecode[pc])       |  MOV_R_A     0   # if i!=0:
            pc += 1                     |  JUMP_IF_A   4   #    goto 4
            a = regs[n]                 |
        elif opcode == ADD_R_TO_A:      |  MOV_R_A     2   # return res
            n = ord(bytecode[pc])       |  RETURN_A
            pc += 1                     |
            a += regs[n]                |
        elif opcode == DECR_A:          |
            a -= 1                      |
        elif opcode == RETURN_A:        |
            return a                    |
  \end{verbatim}
}

\frame[containsverbatim, plain, shrink=10]{
    \frametitle{Trace}
    ~\\
    Resulting trace when tracing bytecode \texttt{DECR\_A}:
    \begin{verbatim}

loop_start(a0, regs0, bytecode0, pc0)
opcode0 = strgetitem(bytecode0, pc0)
pc1 = int_add(pc0, Const(1))
guard_value(opcode0, Const(7))
a1 = int_sub(a0, Const(1))
jump(a1, regs0, bytecode0, pc1)
\end{verbatim}
}

\begin{frame}
    \frametitle{Idea for a Solution}
    \begin{itemize}
    \item key idea: try to trace the loops in the user program
    \item approach: add things to the position key of the tracer
    \item tracing interpreter needs information about the language interpreter
    \item provided by adding \emph{hints} to the language interpreter
    \end{itemize}
    \pause
    \begin{block}{Hints Give Information About:}
    \begin{itemize}
    \item which variables make up the program counter of the language interpreter
    \item where the bytecode dispatch loop is
    \item which bytecodes can correspond to backward jumps
    \end{itemize}
    \end{block}
\end{frame}

\frame[containsverbatim, plain, shrink=10]{
    \frametitle{Interpreter with Hints}
\begin{verbatim}
tlrjitdriver = JitDriver(['pc', 'bytecode'])

def interpret(bytecode, a):
    regs = [0] * 256
    pc = 0
    while True:
        tlrjitdriver.start_dispatch_loop()
        opcode = ord(bytecode[pc])
        pc += 1
        if opcode == JUMP_IF_A:
            target = ord(bytecode[pc])
            pc += 1
            if a:
                pc = target
                if target < pc:
                    tlrjitdriver.backward_jump()
        elif opcode == MOV_A_R:
            ... # rest unmodified
\end{verbatim}
}

\frame[containsverbatim, plain, shrink=20]{
    \frametitle{Result When Tracing \texttt{SQUARE}}
\begin{verbatim}
loop_start(a0, regs0, bytecode0, pc0)
# MOV_R_A 0
opcode0 = strgetitem(bytecode0, pc0)
pc1 = int_add(pc0, Const(1))
guard_value(opcode0, Const(2))
n1 = strgetitem(bytecode0, pc1)
pc2 = int_add(pc1, Const(1))
a1 = call(Const(<* fn list_getitem>), regs0, n1)
# DECR_A
...
# MOV_A_R 0
...
# MOV_R_A 2
...
# ADD_R_TO_A 1
...
# MOV_A_R 2
...
# MOV_R_A 0
...
# JUMP_IF_A 4
opcode6 = strgetitem(bytecode0, pc13)
pc14 = int_add(pc13, Const(1))
guard_value(opcode6, Const(3))
target0 = strgetitem(bytecode0, pc14)
pc15 = int_add(pc14, Const(1))
i1 = int_is_true(a5)
guard_true(i1)
jump(a5, regs0, bytecode0, target0)
\end{verbatim}
}

\begin{frame}
    \frametitle{What Have We Won?}
    \begin{itemize}
    \item trace corresponds to one loop of the user program
    \item however, most operations concerned with manipulating bytecode and program counter
    \item bytecode and program counter are part of the position key
    \item thus they are constant at the beginning of the loop
    \item therefore they can and should be constant-folded
    \end{itemize}
\end{frame}

\frame[containsverbatim, plain, shrink=20]{
    \frametitle{Result When Tracing \texttt{SQUARE} With Constant-Folding}
\begin{verbatim}
loop_start(a0, regs0)
# MOV_R_A 0
a1 = call(Const(<* fn list_getitem>), regs0, Const(0))
# DECR_A
a2 = int_sub(a1, Const(1))
# MOV_A_R 0    
call(Const(<* fn list_setitem>), regs0, Const(0), a2)
# MOV_R_A 2
a3 = call(Const(<* fn list_getitem>), regs0, Const(2))
# ADD_R_TO_A  1
i0 = call(Const(<* fn list_getitem>), regs0, Const(1))
a4 = int_add(a3, i0)
# MOV_A_R 2
call(Const(<* fn list_setitem>), regs0, Const(2), a4)
# MOV_R_A 0
a5 = call(Const(<* fn list_getitem>), regs0, Const(0))
# JUMP_IF_A 4
i1 = int_is_true(a5)
guard_true(i1)
jump(a5, regs0)
\end{verbatim}
}

\begin{frame}
    \frametitle{Results}
    \begin{itemize}
    \item almost only computations related to the user program remain
    \item list of registers is only vestige of language interpreter
    \end{itemize}
    \pause
    \begin{block}{Timing Results Computing Square of 10'000'000}
    \begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &No JIT &442.7 $\pm$ 3.4 &1.00\\
2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
\hline
\end{tabular}
\end{block}
\end{frame}

\begin{frame}
    \frametitle{Scaling to Large Interpreters?}
    \begin{itemize}
    \item we can apply this approach to PyPy's Python interpreter (70 KLOC)
    \item speed-ups promising: factor of 6 faster for simple loops with arithmetic
    \item no Python-specific bugs!
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Conclusions}
    \begin{itemize}
    \item some small changes to a tracing JIT makes it possible to effectively apply it to bytecode interpreters
    \item result is similar to a tracing JIT for that language
    \item bears resemblance to partial evaluation, arrived at by different means
    \item maybe enough to write exactly one tracing JIT?
    \end{itemize}
    \pause
    \begin{block}{Outlook}
        \begin{itemize}
        \item better optimizations of the traces
        \item escape analysis
        \item optimize frame objects
        \item speed up tracing itself
        \item apply to other interpreters and larger programs
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Thank you! Questions?}
    \begin{itemize}
    \item some small changes to a tracing JIT makes it possible to effectively apply it to bytecode interpreters
    \item result is similar to a tracing JIT for that language
    \item bears resemblance to partial evaluation, arrived at by different means
    \item maybe enough to write exactly one tracing JIT?
    \end{itemize}
    \begin{block}{Outlook}
        \begin{itemize}
        \item better optimizations of the traces
        \item escape analysis
        \item optimize frame objects
        \item speed up tracing itself
        \item apply to other interpreters and larger programs
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Backup Slides}
\end{frame}

\frame[containsverbatim, plain, shrink=10]{
    \frametitle{Timings for Python Interpreter}
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
\begin{block}{Timings}
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}
\end{block}
}

\end{document}

### 2009-07-02 talk/icooolps2009/talk/talk.tex
first round of simplification

@@
     \item when a hot loop is discovered, start to produce a trace
-    \end{itemize}
-\end{frame}
-
-\begin{frame}
-    \frametitle{Tracing}
-    \begin{itemize}
-    \item a \emph{trace} is a sequential list of operations
-    \item a trace is produced by recording every operation the interpreter executes
-    \item tracing ends when the tracer sees a position in the program it has seen before
-    \item to identify these places, the \emph{position key} is used
-    \item the position key encodes the current point of execution
-    \item a trace thus corresponds to exactly one loop
-    \item that means it ends with a jump to its beginning
-    \end{itemize}
-    \pause
-    \begin{block}{Guards}
-        \begin{itemize}
-        \item the trace is only one of the possible code paths through the loop
-        \item at places where the path \emph{could} diverge, a guard is placed
-        \end{itemize}
-    \end{block}
-\end{frame}
-
-\begin{frame}
-    \frametitle{Code Generation and Execution}
-    \begin{itemize}
-    \item being linear, the trace can easily be turned into machine code
-    \item the machine code can be immediately executed
-    \item execution stops when a guard fails
-    \item after a guard failure, go back to interpreting program
+    \item when a full loop is traced, the trace is converted to machine code
     \end{itemize}
@@
         \item easy and fast machine code generation: needs so support only one path
+        \item (things are more complex, but let's ignore that for now)
         \item interpreter does a lot of the work
@@
         \item automatic inlining
-        \item produces very little code
+        \item produces comparatively little code
         \end{itemize}
@@
     \item Question: What happens if the program is itself a bytecode interpreter?
-    \item the (usually only) hot loop of a bytecode interpreter is the bytecode dispatch loop
+    \item the (most important) hot loop of a bytecode interpreter is the bytecode dispatch loop
     \item Assumption violated: two iterations of the dispatch loop will usually take very different code paths
@@
 pc2 = int_add(pc1, Const(1))
-a1 = call(Const(<* fn list_getitem>), regs0, n1)
+a1 = list_getitem(regs0, n1)
 # DECR_A
@@
 # MOV_R_A 0
-a1 = call(Const(<* fn list_getitem>), regs0, Const(0))
+a1 = list_getitem(regs0, Const(0))
 # DECR_A
@@
 # MOV_A_R 0    
-call(Const(<* fn list_setitem>), regs0, Const(0), a2)
+list_setitem(regs0, Const(0), a2)
 # MOV_R_A 2
-a3 = call(Const(<* fn list_getitem>), regs0, Const(2))
+list_getitem(regs0, Const(2))
 # ADD_R_TO_A  1
-i0 = call(Const(<* fn list_getitem>), regs0, Const(1))
+i0 = list_getitem(regs0, Const(1))
 a4 = int_add(a3, i0)
 # MOV_A_R 2
-call(Const(<* fn list_setitem>), regs0, Const(2), a4)
+list_setitem(regs0, Const(2), a4)
 # MOV_R_A 0
-a5 = call(Const(<* fn list_getitem>), regs0, Const(0))
+a5 = list_getitem(regs0, Const(0))
 # JUMP_IF_A 4

### 2009-07-04 talk/icooolps2009/talk/talk.tex
If we're using tex, we can go for a correct spelling
Small reorganizion of the slide "Idea for a Solution"

@@
     Antonio Cuni\inst{2} \and
-    Maciej Fijalkowski\inst{3} \and
+    Maciej Fija\l{}kowski\inst{3} \and
     Armin Rigo
@@
     \item the (most important) hot loop of a bytecode interpreter is the bytecode dispatch loop
-    \item Assumption violated: two iterations of the dispatch loop will usually take very different code paths
+    \item Assumption violated: consecutive iterations of the dispatch loop will usually take very different code paths
     \end{itemize}
@@
         \item \emph{tracing interpreter:} the interpreter that originally runs the program and produces traces
-        \item \emph{language interpreter:} the bytecode interpreter runs on top
+        \item \emph{language interpreter:} the bytecode interpreter run on top
         \item \emph{user program:} the program run by the language interpreter
@@
     \begin{itemize}
-    \item key idea: try to trace the loops in the user program
-    \item approach: add things to the position key of the tracer
+    \item goal: try to trace the loops \emph{in the user program,}
+          and not just one iteration of the bytecode dispatch loop
     \item tracing interpreter needs information about the language interpreter
-    \item provided by adding \emph{hints} to the language interpreter
+    \item provided by adding \emph{three hints} to the language interpreter
     \end{itemize}
@@
     \item trace corresponds to one loop of the user program
-    \item however, most operations concerned with manipulating bytecode and program counter
+    \item however, most operations are concerned with manipulating bytecode and program counter
     \item bytecode and program counter are part of the position key
@@
 1 &No JIT &442.7 $\pm$ 3.4 &1.00\\
-2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
-3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
+2 &JIT, Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
+3 &JIT, Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
 4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\

### 2009-11-05 talk/rupy2009/talk.tex
progress..

\documentclass[utf8x]{beamer}

\mode<presentation>
{
  \usetheme{Warsaw}

  %\setbeamercovered{transparent}
}

\usepackage[english]{babel}

\usepackage[utf8x]{inputenc}

\usepackage{times}
\usepackage[T1]{fontenc}

\title{The speed of PyPy}

\author{Maciej Fijałkowski}

\institute[merlinux GmbH]
{ merlinux GmbH }

\date{RuPy, November 7th 2009, Poznań}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}

  \frametitle{Story about Python's speed}
  \begin{figure}
    \includegraphics[width=.8\textwidth]{img1.jpg}
  \end{figure}

\end{frame}

\begin{frame}
  \frametitle{Speed of Python}
  \begin{itemize}
     \item Python is slow
  \end{itemize}
  \pause
  \begin{figure}
    \includegraphics[width=.6\textwidth]{peasant_and_birdnester-400.jpg}
  \end{figure}

\end{frame}

\begin{frame}
  \frametitle{Speed of Python}
  \begin{itemize}
    \item Is Python really slow?
      \pause
    \item Sometimes
      \pause
    \item Let's have a look at some examples
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Nomenclature}
  \begin{itemize}
    \item Python - a programming language
    \item CPython - main implementation of Python
      xxx
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Example run}
  \begin{itemize}
    \item Float example, stolen from factor blog
  \end{itemize}
  \vspace{.5cm}
  \begin{tabular}{| l | c | r |}
    \hline
    & CPython & JVM (client mode) \\
    Average of 10 runs: & 7.6s & 0.77s \\
    \hline
  \end{tabular}
  \vspace{.5cm}
  \pause
  \begin{itemize}
    \item Python is 10x slower than Java
      \pause
    \item Python is 10x slower than Java on this particular benchmark
      \pause
    \item CPython is 10x slower than Java on this particular benchmark
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{More about this example}
  \begin{tabular}{| l | c | c | c | c |}
    \hline
    & CPython & JVM & Psyco & PyPy \\
    Average of 10 runs & 7.6s & 0.77s & 4.4s & 1.3s \\
    \hline
  \end{tabular}
  \vspace{.5cm}
  \pause
  \begin{itemize}
    \item So, it's CPython that is slow on this particular benchmark
  \end{itemize}
\end{frame}

\end{document}

### 2009-11-05 talk/rupy2009/talk.tex
update

@@
     \item CPython - main implementation of Python
-      xxx
+    \item JVM - Java Virtual Machine - VM used to run Java, among others
+    \item JIT - Just in time compiler
+    \item Psyco - JIT for Python
   \end{itemize}
@@
     \item So, it's CPython that is slow on this particular benchmark
+      \pause
+    \item Same example, using numpy and vectorization about 3x faster than JVM
   \end{itemize}
@@
 
+\begin{frame}
+  \frametitle{Python's speed}
+  \begin{itemize}
+    \item Instead of: ``Why is Python slow?''
+      \pause
+    \item Better: ``Why is Python hard to optimize?''
+      \pause
+    \item Even better: ``How are we going to fix it?''
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Why is Python hard to optimize?}
+  \begin{itemize}
+    \item Duck typing (dynamic dispatch)
+    \item Frames
+    \item Object encapsulation
+    \item Dictionaries of instances
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+\end{frame}
+
 \end{document}

### 2009-11-06 talk/rupy2009/talk.tex
something else I thought of
update + some examples

@@
 \usepackage[T1]{fontenc}
+\usepackage{color}
 
@@
       \pause
-    \item Sometimes
+    \item Sometimes, for some usecases
       \pause
@@
     \hline
-    & CPython & JVM (client mode) \\
+    & CPython & Java (hotspot client mode) \\
     Average of 10 runs: & 7.6s & 0.77s \\
@@
 \begin{frame}
+  \frametitle{Some evidence}
+  \begin{figure}
+    \includegraphics[width=1.0\textwidth]{time.png}
+  \end{figure}
+\end{frame}
+
+\begin{frame}
   \frametitle{Why is Python hard to optimize?}
@@
     \item Dictionaries of instances
+    \item Changing globals
+    \item Ability to dynamically change builtins
   \end{itemize}
@@
 \begin{frame}
+  \frametitle{Duck typing}
+  \begin{itemize}
+    \item Dispatching over item type
+    \item {\tt z = x + y}
+    \item Needs to check what the type of {\tt x} and {\tt y} is
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Frames}
+  \begin{itemize}
+    \item Python interpreters use frames on heap (instead of stack)
+    \item Locals are stored on those frames
+    \item Intermediate results are store on valuestack of frames
+    \item In fact, you can access frames via {\tt sys.\_getframe() or
+      traceback}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Object encapsulation}
+  \begin{itemize}
+    \item Also called boxing
+    \item Each object, even {\tt int}, has to be boxed
+    \item Requires allocations and indirection
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Example - addition}
+  \begin{itemize}
+    \item {\tt z = x + y}
+    \item read value for {\tt x} from frame, store on valuestack
+    \item read value for {\tt x} from frame, store on valuestack
+    \item allocate new integer
+    \item read two values from valuestack, add and store the result in
+      freshly allocated integer
+    \item move the result from valuestack to locals
+      \pause
+    \item {\color{red} in fact, should be one assembler instruction}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Dictionaries of instances}
+  \begin{itemize}
+    \item Need to perform a dictionary lookup for {\tt x.y}
+    \item There are {\bf three} lookups per method call
+      (descriptor, object, type)
+    \item {\bf Two} for attribute access
+      (descriptor, object)
+    \item Looks like list lookup should be enough
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Changing globals}
+  \begin{itemize}
+    \item Global symbols can change at any moment in time
+    \item Makes for example inlining hard
+    \item Requires global dict lookups, even for constants
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Ability to dynamically change builtins}
+  \begin{itemize}
+    \item You can say {\tt int = my\_function}
+    \item But you can't {\tt int.\_\_add\_\_ = my\_method}
+    \item Still messes up optimizations
+    \item Global lookup is also a dictionary lookup, even if globals
+      don't change
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Interpreting vs compiling}
+  \begin{itemize}
+    \item Apparently, processors are good at branch prediction
+    \item We didn't measure much of a difference, less than 2x overall
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{CPython specific problems}
+  \begin{itemize}
+    \item In general, CPython is fairly well optimized
+    \item refcounting is an inefficient garbage collection scheme
+    \item GIL
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Dynamic compilation to the rescue}
+  \begin{itemize}
+    \item You don't pay for feature, until you actually use it
+    \item In static compilation, compiler has to prove that bad
+      things can't happen
+    \item Impossible in Python
+    \item With dynamic compilation, you just throw away compiled
+      code in case things go wrong or start over
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Dealing with frames}
+  \begin{itemize}
+    \item Allocate frame
+      \pause
+    \item Use C stack, but remember where frame fields are
+      living on the stack
+      \pause
+    \item Be able to reconstruct frame on demand (for example
+      {\tt sys.\_getframe() was called})
+      \pause
+    \item The effect is that you don't pay for frames, unless
+      you really use them
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Dealing with dynamic dispatch}
+  \begin{itemize}
+    \item The answer is to simply specialize over types
+    \item Provides possibly multiple versions of compiled code for
+      single Python code
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Dealing with object encapsulation}
+  \begin{itemize}
+    \item ``Virtual objects''
+    \item Also known as escape analysis
+    \item If object does not ``escape'', don't allocate it at all
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{How does it work in practice?}
+  \begin{itemize}
+    \pause
+    \item Pretty well
+      \pause
+    \item XXXX speedup over CPython
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Dealing with attribute access}
+  \begin{itemize}
+    \item Fairly complex task
+    \item Sharing dict, more or less the same effect as V8's hidden
+      classes
+      \pause
+    \item Python is a very complex language
+    \item Shadowing methods with attributes
+    \item Descriptors before attributes
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Other important optimizations}
+  \begin{itemize}
+    \item Caching globals
+    \item Caching builtins
+    \item A lot of smaller ones
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{CPython vs PyPy}
+  \begin{itemize}
+    \item I use CPython for everyday usage
+      \pause
+    \item But personally, I hope to change it in next months
+      \pause
+    \item ... in places where performance matters, but that don't
+      depend on third party C modules (like numpy)
+      \pause
+    \item ... like building and developing PyPy
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Status of PyPy}
+  \begin{itemize}
+    \item Very compliant Python interpreter
+    \item Most of important stdlib modules
+    \item Differencies are agreed to be implementation details
+      \pause
+    \item {\color{red} or bugs}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Examples of working programs}
+  \begin{itemize}
+    \item Django (sqlite only)
+    \item Twisted
+    \item PyPy's translation toolchain
+    \item ctypes
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Status of JIT}
+  \begin{itemize}
+    \item Because the way it's constructed, handles all Python language
+      features (unlike for example {\bf Psyco})
+    \item Changes very quickly these days
+    \item Ready for cautious tests
+    \item Not ready as a drop-in replacement of CPython
+      \pause
+    \item {\color{green} yet!}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Adapt today!}
+  \begin{itemize}
+    \item A fact: People rely on deep obscure features of language
+    \item Examples:
+      \pause
+      \begin{itemize}
+        \item {\tt except ImportError, e: \\ \quad if str(e) != ...:
+          raise}
+          \pause
+        \item Exact naming of list comprehension variable
+          \pause
+        \item Immediate finalization
+      \end{itemize}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Profit tomorrow!}
+  \begin{itemize}
+    \item We plan to release JIT-ready version somewhere early 2010
+      xxx
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{How you can help}
+  \begin{itemize}
+    \item It's all open source after all ...
+    \item Try running existing programs
+    \item Profile, report bugs
+      \pause
+    \item Talk to your boss (XXX)
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Thank you!}
+  \begin{itemize}
+    \item This talk is already online (with all examples):
+      {\tt http://codespeak.net/svn/pypy/dist/extradoc/talk/rupy2009/talk.pdf}
+    \item {\tt http://morepypy.blogspot.com}
+    \item \#pypy on freenode
+    \item If you want to know more about PyPy, feel free to bug me
+      around (like, how does the JIT work?)
+    \item Any questions?
+  \end{itemize}
 \end{frame}

### 2009-11-06 talk/rupy2009/examples/escaping.py

@@ -1,0 +1,12 @@
+
+[x, y, z]
+v0 = IntObject(x.value % y.value)
+return IntObject(z.value + v0.value)
+
+
+
+
+
+[x, y, z]
+i0 = x.value % y.value
+return IntObject(z.value + i0)

### 2009-11-06 talk/rupy2009/examples/interp.py

@@ -1,0 +1,30 @@
+
+
+[x, y]
+# LOAD_FAST
+frame.valuestack.push(frame.locals[0])
+# LOAD_FAST
+frame.valuestack.push(frame.locals[1])
+# BINARY_ADD
+v1 = frame.valuestack.pop()
+v0 = frame.valuestack.pop()
+v2 = v0.add(v1)
+  return IntObject(self.value + other.value) 
+frame.valustack.push(v2)
+# RETURN_VALUE
+return frame.valuestack[-1]
+
+
+
+[x, y]
+v0 = IntObject(x.value + y.value)
+return v0
+
+
+
+
+[x, y]
+return x + y
+
+
+

### 2009-11-06 talk/rupy2009/examples/loop.py

@@ -1,0 +1,8 @@
+def f():
+    i = 0
+    s = 0
+    while i < 1000000:
+        s += (i % 10)
+        i += 1
+
+f()

### 2010-02-07 talk/pycon2010/keynote/talk.tex
Stuff I did so far

\documentclass[utf8x, 14pt]{beamer}

\mode<presentation>
{
  \usetheme{Warsaw}
}

\usepackage[english]{babel}

\usepackage[utf8x]{inputenc}
\usepackage{tikz}

\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{color}

\setbeamerfont{big}{size*={64pt}{0pt}}

\title{The status of PyPy}

\author{Maciej Fijałkowski}

\institute[merlinux GmbH]
{ merlinux GmbH }

\date{Pycon 2010, February 20th 2010, Atlanta}

\begin{document}

\begin{frame}
  \titlepage
  \begin{figure}
    \begin{tabular}{c c c}
    \includegraphics[width=.30\textwidth]{pypy-logo.png}
    &
    \hspace{2cm}
    &
    \includegraphics[width=.25\textwidth]{merlinux-logo.png}
    \end{tabular}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Introduction}
  \begin{itemize}
    \item What we did in the last year?
    \item What we achieved?
    \item Plans for near future
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Last year highlits}
  \pause
  \usebeamerfont{big}
  JIT
\end{frame}

\begin{frame}
  \frametitle{February 2009}
  \begin{tikzpicture}[remember picture, overlay]
    \node [shift={(1cm, -4cm)}] at (current page.north west) {
      \begin{tikzpicture}[remember picture, overlay, font={\fontsize{8pt}{0pt} \ttfamily}]
        \node [shift={(-1cm, 1.4cm)}, above right] {
          \includegraphics[width=.15\textwidth]{pypy-logo.png}};
        \draw [thick, ->] (1.2cm, 2.5cm) -- (10cm, 2.5cm);
      \end{tikzpicture}
    };
  \end{tikzpicture}
  \begin{itemize}
    \item We merge experiments into mailine PyPy
    \item xxx
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{March 2009}
  \begin{tikzpicture}[remember picture, overlay]
    \node [shift={(1cm, -4cm)}] at (current page.north west) {
      \begin{tikzpicture}[remember picture, overlay, font={\fontsize{8pt}{0pt} \ttfamily}]
        \node [shift={(-1cm, 1.4cm)}, above right] {
          \includegraphics[width=.15\textwidth]{pypy-logo.png}};
        \draw [thick, ->] (1.2cm, 2.5cm) -- (10cm, 2.5cm);
        \draw [fill=blue!20,very thin] (1.7cm, 2.5cm) circle (.06cm);
        \draw (1.7cm, 2.44cm) -- (1.7cm, 2.0cm);
        \draw (1.5cm, 1.5cm) node [rotate=45] {JIT merge};  
      \end{tikzpicture}
    };
  \end{tikzpicture}
  \begin{itemize}
    \item first python example works!
      \begin{verbatim}
i = 0
while i < 1000000:
  i = i + 1
      \end{verbatim}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{June 2009}
  \begin{tikzpicture}[remember picture, overlay]
    \node [shift={(1cm, -4cm)}] at (current page.north west) {
      \begin{tikzpicture}[remember picture, overlay, font={\fontsize{8pt}{0pt} \ttfamily}]
        \node [shift={(-1cm, 1.4cm)}, above right] {
          \includegraphics[width=.15\textwidth]{pypy-logo.png}};
        \draw [thick, ->] (1.2cm, 2.5cm) -- (10cm, 2.5cm);
        \draw [fill=blue!20,very thin] (1.7cm, 2.5cm) circle (.06cm);
        \draw (1.7cm, 2.44cm) -- (1.7cm, 2.0cm);
        \draw (1.5cm, 1.5cm) node [rotate=45] {JIT merge};
        \draw [fill=blue!20,very thin] (2.3cm, 2.5cm) circle (.06cm);
        \draw (2.3cm, 2.44cm) -- (2.3cm, 2.0cm);
        \draw (1.95cm, 1.5cm) node [rotate=45] {first example};  
      \end{tikzpicture}
    };
  \end{tikzpicture}
  xxx
\end{frame}

\begin{frame}
  \frametitle{Nice graphs}
\end{frame}

\begin{frame}
  \frametitle{Status}
  \begin{block}{Status of today}
    \begin{itemize}
    \item JIT is fairly good
      \pause
    \item sometimes can consume a lot of memory
      \pause
    \item sometimes keep on compiling all the time
      \pause
    \item GCs are pretty good
      \pause
    \item VMs algorithms are not as polished as CPython
      ones [xxx reword]
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Near future}
  \begin{itemize}
    \item release March 2010
      \pause
    \item first PyPy release to contain JIT!
    \item mostly to try it out
  \end{itemize}
\end{frame}

\end{document}

### 2010-02-08 talk/pycon2010/keynote/talk.tex
I'm sort of happy with slides as they're now.

@@
   \begin{itemize}
-    \item We merge experiments into mailine PyPy
+    \item we merge experiments into mailine PyPy
     \item xxx
@@
 \begin{frame}
-  \frametitle{June 2009}
+  \frametitle{September 2009}
   \begin{tikzpicture}[remember picture, overlay]
@@
         \draw (1.5cm, 1.5cm) node [rotate=45] {JIT merge};
-        \draw [fill=blue!20,very thin] (2.3cm, 2.5cm) circle (.06cm);
-        \draw (2.3cm, 2.44cm) -- (2.3cm, 2.0cm);
-        \draw (1.95cm, 1.5cm) node [rotate=45] {first example};  
+        \draw [fill=blue!20,very thin] (4.3cm, 2.5cm) circle (.06cm);
+        \draw (4.3cm, 2.44cm) -- (4.3cm, 2.0cm);
+        \draw (3.95cm, 1.5cm) node [rotate=45] {first example};  
       \end{tikzpicture}
@@
   \end{tikzpicture}
-  xxx
+  \begin{itemize}
+    \item we manage to speed first non-trivial benchmark
+    \item richards - a simple, object oriented code
+    \item a variety of optimizations involved
+  \end{itemize}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{Nice graphs}
+  \frametitle{February 2010 - now}
+  \begin{tikzpicture}[remember picture, overlay]
+    \node [shift={(1cm, -4cm)}] at (current page.north west) {
+      \begin{tikzpicture}[remember picture, overlay, font={\fontsize{8pt}{0pt} \ttfamily}]
+        \node [shift={(-1cm, 1.4cm)}, above right] {
+          \includegraphics[width=.15\textwidth]{pypy-logo.png}};
+        \draw [thick, ->] (1.2cm, 2.5cm) -- (10cm, 2.5cm);
+        \draw [fill=blue!20,very thin] (1.7cm, 2.5cm) circle (.06cm);
+        \draw (1.7cm, 2.44cm) -- (1.7cm, 2.0cm);
+        \draw (1.5cm, 1.5cm) node [rotate=45] {JIT merge};
+        \draw [fill=blue!20,very thin] (4.3cm, 2.5cm) circle (.06cm);
+        \draw (4.3cm, 2.44cm) -- (4.3cm, 2.0cm);
+        \draw (3.95cm, 1.5cm) node [rotate=45] {first example};  
+        \draw [fill=blue!20,very thin] (7.3cm, 2.5cm) circle (.06cm);
+        \draw (7.3cm, 2.44cm) -- (7.3cm, 2.0cm);
+        \draw (6.95cm, 1.5cm) node [rotate=45] {richards};  
+      \end{tikzpicture}
+    };
+  \end{tikzpicture}
+  We run a variety of benchmarks nightly.
 \end{frame}
@@
       \pause
-    \item sometimes can consume a lot of memory
+    \item can consume a lot of memory
       \pause
-    \item sometimes keep on compiling all the time
+    \item can recompile code too many times
       \pause
@@
       \pause
-    \item VMs algorithms are not as polished as CPython
-      ones [xxx reword]
+    \item runtime algorithms are not as polished as CPython
+      ones (ie strings, regexes)
     \end{itemize}
@@
     \item mostly to try it out
+      \pause
+    \item we just got a bit more {\bf EU funding}
+    \item we're actively looking for {\bf sponsoring} to continue PyPy development
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Come to my next talk!}
+  \begin{block}{speed of PyPy}
+    ... or how to write programs that play well with the JIT
+  \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{That's all}
+  Thank you for your attention
+  \begin{itemize}
+    \item http://morepypy.blogspot.com
+    \item http://pypy.org
+    \item http://merlinux.de
+    \item these slides are already online
   \end{itemize}

### 2010-02-15 talk/pycon2010/keynote/talk.tex
(alexgaynor) fix
typos and minor fixes

@@
 \begin{frame}
-  \frametitle{Last year highlits}
+  \frametitle{Last year highlights}
   \pause
@@
   \begin{itemize}
-    \item we merge experiments into mailine PyPy
-    \item xxx
+    \item we merge experiments into mainline PyPy
+      \pause
+    \item compiles assembler, slows things down
   \end{itemize}
@@
     \item we manage to speed first non-trivial benchmark
-    \item richards - a simple, object oriented code
+    \item {\bf richards} - a simple, object oriented code
     \item a variety of optimizations involved
@@
   \end{tikzpicture}
-  We run a variety of benchmarks nightly.
+  We run a variety of benchmarks nightly. Pictures!
 \end{frame}
@@
     \item runtime algorithms are not as polished as CPython
-      ones (ie strings, regexes)
+      ones (eg strings, regexes)
     \end{itemize}
@@
       \pause
-    \item first PyPy release to contain JIT!
+    \item first PyPy release to contain a working JIT!
     \item mostly to try it out
@@
     \item http://pypy.org
-    \item http://merlinux.de
-    \item these slides are already online
+    \item http://merlinux.eu
+    \item these slides are already online:
+    \item http://tinyurl.com/pypyslides
   \end{itemize}

### 2010-02-15 talk/pycon2010/pypyspeed/talk.tex
start working on talk

--- 2010-02-07 talk/pycon2010/keynote/talk.tex
@@
 
-\setbeamerfont{big}{size*={64pt}{0pt}}
-
-\title{The status of PyPy}
+\title{The speed of PyPy}
 
@@
     \begin{tabular}{c c c}
-    \includegraphics[width=.30\textwidth]{pypy-logo.png}
+    \includegraphics[width=.30\textwidth]{../common/pypy-logo.png}
     &
@@
     &
-    \includegraphics[width=.25\textwidth]{merlinux-logo.png}
+    \includegraphics[width=.25\textwidth]{../common/merlinux-logo.png}
     \end{tabular}
@@
 \begin{frame}
-  \frametitle{Introduction}
+  \frametitle{JIT - what's that about?}
   \begin{itemize}
-    \item What we did in the last year?
-    \item What we achieved?
-    \item Plans for near future
+    \item {\bf JIT is not a magical device!}
+      \pause
+    \item removes bytecode overhead
+    \item removes frame overhead
+    \item can make runtime decisions
+    \item more classic optimization that can follow
+\end{frame}
+
+\begin{frame}
+  \frametitle{The main idea}
+  \begin{itemize}
+    \item python has advanced features (frame introspection,
+      arbitrary code execution, overloading globals)
+    \item with JIT, you don't pay for them if you don't use
+      them
+    \item however, you pay if you use them, but they work
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{Last year highlits}
-  \pause
-  \usebeamerfont{big}
-  JIT
-\end{frame}
-
-\begin{frame}
-  \frametitle{February 2009}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node [shift={(1cm, -4cm)}] at (current page.north west) {
-      \begin{tikzpicture}[remember picture, overlay, font={\fontsize{8pt}{0pt} \ttfamily}]
-        \node [shift={(-1cm, 1.4cm)}, above right] {
-          \includegraphics[width=.15\textwidth]{pypy-logo.png}};
-        \draw [thick, ->] (1.2cm, 2.5cm) -- (10cm, 2.5cm);
-      \end{tikzpicture}
-    };
-  \end{tikzpicture}
+  \frametitle{A piece of advice}
   \begin{itemize}
-    \item We merge experiments into mailine PyPy
-    \item xxx
-  \end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]
-  \frametitle{March 2009}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node [shift={(1cm, -4cm)}] at (current page.north west) {
-      \begin{tikzpicture}[remember picture, overlay, font={\fontsize{8pt}{0pt} \ttfamily}]
-        \node [shift={(-1cm, 1.4cm)}, above right] {
-          \includegraphics[width=.15\textwidth]{pypy-logo.png}};
-        \draw [thick, ->] (1.2cm, 2.5cm) -- (10cm, 2.5cm);
-        \draw [fill=blue!20,very thin] (1.7cm, 2.5cm) circle (.06cm);
-        \draw (1.7cm, 2.44cm) -- (1.7cm, 2.0cm);
-        \draw (1.5cm, 1.5cm) node [rotate=45] {JIT merge};  
-      \end{tikzpicture}
-    };
-  \end{tikzpicture}
-  \begin{itemize}
-    \item first python example works!
-      \begin{verbatim}
-i = 0
-while i < 1000000:
-  i = i + 1
-      \end{verbatim}
+    \item Don't use advanced features if you don't have to
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{June 2009}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node [shift={(1cm, -4cm)}] at (current page.north west) {
-      \begin{tikzpicture}[remember picture, overlay, font={\fontsize{8pt}{0pt} \ttfamily}]
-        \node [shift={(-1cm, 1.4cm)}, above right] {
-          \includegraphics[width=.15\textwidth]{pypy-logo.png}};
-        \draw [thick, ->] (1.2cm, 2.5cm) -- (10cm, 2.5cm);
-        \draw [fill=blue!20,very thin] (1.7cm, 2.5cm) circle (.06cm);
-        \draw (1.7cm, 2.44cm) -- (1.7cm, 2.0cm);
-        \draw (1.5cm, 1.5cm) node [rotate=45] {JIT merge};
-        \draw [fill=blue!20,very thin] (2.3cm, 2.5cm) circle (.06cm);
-        \draw (2.3cm, 2.44cm) -- (2.3cm, 2.0cm);
-        \draw (1.95cm, 1.5cm) node [rotate=45] {first example};  
-      \end{tikzpicture}
-    };
-  \end{tikzpicture}
-  xxx
-\end{frame}
-
-\begin{frame}
-  \frametitle{Nice graphs}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Status}
-  \begin{block}{Status of today}
-    \begin{itemize}
-    \item JIT is fairly good
-      \pause
-    \item sometimes can consume a lot of memory
-      \pause
-    \item sometimes keep on compiling all the time
-      \pause
-    \item GCs are pretty good
-      \pause
-    \item VMs algorithms are not as polished as CPython
-      ones [xxx reword]
-    \end{itemize}
-  \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Near future}
-  \begin{itemize}
-    \item release March 2010
-      \pause
-    \item first PyPy release to contain JIT!
-    \item mostly to try it out
-  \end{itemize}
+  \frametitle{xxx}
 \end{frame}

### 2010-02-15 talk/pycon2010/pypyspeed/talk.tex
Finish slides

@@
     \item more classic optimization that can follow
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{How well it works in practice?}
 \end{frame}
@@
   \begin{itemize}
-    \item Don't use advanced features if you don't have to
+    \item don't use advanced features if you don't have to
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{xxx}
+  \frametitle{Tracing JIT}
+  \begin{itemize}
+    \item compiler traces the actual execution of Python program
+    \item then compiles linear path to assembler
+    \item example
+    \item mostly for speeding up loops and to certain extent
+      recursion
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Removing frame overhead}
+  \begin{verbatim}
+x = y + z
+  \end{verbatim}
+  \begin{itemize}
+    \item above has 5 frame accesses
+    \item they can all be removed (faster!)
+      \pause
+    \item they prevent optimizations from happening
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Removing object boxing}
+  \begin{verbatim}
+i = 0
+while i < 100:
+  i += 1
+  \end{verbatim}
+  \begin{itemize}
+    \item for each iteration we do a comparison and addition
+    \item xxx integers on valuestack and xxx integers in locals
+    \item all of those can be removed
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Access costs}
+  \begin{itemize}
+    \item local access costs nothing
+    \item global access is cheap, if you don't change global {\ttfamily \_\_dict\_\_} too much XXX rephrase
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Frame escapes}
+  \begin{itemize}
+    \item calling {\ttfamily sys.\_getframe()}, {\ttfamily sys.exc\_info()}
+    \item exception escaping
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Shared dicts (aka hidden classes)}
+  \begin{itemize}
+    \item instance {\ttfamily \_\_dict\_\_ } lookup becomes a list lookup
+    \item if you're evil, it'll bail back to dict lookup
+      \pause
+    \item only for newstyle classes!
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Version tags}
+  \begin{itemize}
+    \item dicts on types are version-controlled
+    \item this means methods are usually in known places
+      \pause
+    \item ... if you don't modify them too often
+    \item counters on classes are bad
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Call costs}
+  \begin{itemize}
+    \item calls can be inlined
+    \item simple arguments are by far the best
+    \item avoid {\ttfamily *args} and {\ttfamily **kwds}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Allocation patterns}
+  \begin{itemize}
+    \item PyPy uses a moving GC (like JVM, .NET, etc.)
+    \item pretty efficient for usecases with a lot of
+      short-living objects
+    \item objects are smaller than on CPython
+    \item certain behaviors are different than on CPython
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Differencies}
+  \begin{itemize}
+    \item no refcounting semantics
+    \item {\ttfamily id(obj)} can be expensive as it's a complex
+      operation on a moving GC
+    \item a large list of new objects is a bad case behavior
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{General rules}
+  \begin{itemize}
+    \item don't try to outsmart your compiler
+    \item simple is better than complex
+    \item metaprogramming is your friend
+    \item measurment is the only meaningful way to check
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Problems}
+  \begin{itemize}
+    \item long traces - tracing is slow
+    \item megamorphic calls
+    \item metaclasses
+    \item class global state
+      \pause
+    \item years of optimizations against CPython
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Future}
+  \begin{itemize}
+    \item release end March
+    \item to try it out
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{That's all!}
+  \begin{itemize}
+    \item Q \& A
+    \item http://morepypy.blogspot.com
+    \item http://pypy.org
+    \item http://merlinux.eu
+  \end{itemize}
 \end{frame}

### 2010-02-17 talk/pycon2010/crossinterp/talk.tex
A rewrite of slides - needs more work

\documentclass[utf8x, 14pt]{beamer}

\mode<presentation>
{
  \usetheme{Warsaw}
}

\usepackage[english]{babel}

\usepackage[utf8x]{inputenc}
\usepackage{tikz}

\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{color}

\title{How to write cross-interpreter programs}

\author{Maciej Fijałkowski}

\institute[merlinux GmbH]
{ merlinux GmbH }

\date{Pycon 2010, February 19th 2010, Atlanta}

\begin{document}

\begin{frame}
  \titlepage
  \begin{figure}
    \begin{tabular}{c c c}
    \includegraphics[width=.20\textwidth]{../common/pypy-logo.png}
    &
    \hspace{2cm}
    &
    \includegraphics[width=.18\textwidth]{../common/merlinux-logo.png}
    \end{tabular}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{This talk}
  \begin{itemize}
    \item most people only target CPython (or Jython or IronPython)
    \item sometimes, you want your program to run on each of those
    \item libraries are more often cross-interpreter
      \pause
    \item hope to give you more reasons tomorrow
      \pause
    \item won't talk about py3k
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Have a way to verify compatibility}
  \begin{itemize}
    \item extensive test suite
    \item good coverage
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Exceptions}
  \begin{itemize}
    \item TypeError vs AttributeError change often between implementations,
      even CPython versions
    \item don't rely on exception string messages (they may differ)
    \item also means - don't use doctests
  \end{itemize}
  \begin{verbatim}
try:
  ...
except ImportError, ie:
  if str(ie) != '...':
    raise
  \end{verbatim}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Subclasses of builtin types and overriding}
  \begin{itemize}
    \item in general overridden methods on subclassed builtin types are not 
      invoked by preexisting other methods
      \begin{verbatim}
class d(dict):
  def __getitem__(self, e):
    ...
      \end{verbatim}
    \item would {\ttfamily keys()} go via this getitem?
    \item there can be corner cases, for example when
      there are more than one object involved
    \item tests are your friend
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Access to 3rdy party libraries}
  \begin{itemize}
    \item there is no good story here
    \item ctypes based access is going to be supported by all Pythons
    \item are there pure Python replacements/options?
    \item separate out dependencies/especially optional ones
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Don't rely on refcounting}
  \begin{itemize}
    \item example {\ttfamily open('x', 'w').write('stuff')}
    \item on refcounting, flushes file immediately
    \item on any other GC, it might be deferred for
      a while
    \item the single most-common problem when fixing twisted
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{\_\_del\_\_}
  \begin{itemize}
    \item in case of resurrection CPython will call {\ttfamily \_\_del\_\_} multiple times,
      other Pythons exactly once
    \item cycles with \_\_del\_\_s are not collected by CPython,
      PyPy breaks them randomly instead
    \item in PyPy (xxx Jython?) \_\_del\_\_ cannot be attached to classes after creation
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Use new-style classes}
  \begin{itemize}
    \item 3.x ready
    \item faster on PyPy, too
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{{\ttfamily sys.prefix}}
  \begin{itemize}
    \item implementations may have different installation layouts
    \item open issue, at least for PyPy, has compatibility consequences with
  setuptools
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{IO bytes vs unicode}
  \begin{itemize}
    \item convert/decode as soon as possible, keep text and bytes apart
    \item for 2.x Pythons use str for bytes and unicode for text
    \item the distinction is deeper in 3.x (str is unicode, bytes exist with
  slightly different interface than old str)
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Obscure corners}
  \begin{itemize}
    \item non-string keys in type dictionaries
    \item introspection results, implementation objects (e.g. builtin methods
      etc), may have different types
    \item exact naming of things (like list-comprehension variable)
  \end{itemize}
\end{frame}

\end{document}

### 2010-02-17 talk/pycon2010/crossinterp/talk.tex
Progress

@@
     \item don't rely on exception string messages (they may differ)
-    \item also means - don't use doctests
   \end{itemize}
@@
   \end{verbatim}
+  \pause
+  \begin{itemize}
+    \item also means - don't use doctests
+  \end{itemize}
 \end{frame}
@@
 \begin{frame}[fragile]
-  \frametitle{Subclasses of builtin types and overriding}
+  \frametitle{Subclasses of builtin types}
   \begin{itemize}
@@
     \item would {\ttfamily keys()} go via this getitem?
-    \item there can be corner cases, for example when
-      there are more than one object involved
     \item tests are your friend
@@
 
-\begin{frame}
+\begin{frame}[fragile]
   \frametitle{Don't rely on refcounting}
   \begin{itemize}
-    \item example {\ttfamily open('x', 'w').write('stuff')}
+    \item example
+      \begin{verbatim}
+open('x', 'w').write('stuff')
+      \end{verbatim}
     \item on refcounting, flushes file immediately
@@
   \begin{itemize}
-    \item in case of resurrection CPython will call {\ttfamily \_\_del\_\_} multiple times,
+    \item resurrection on CPython will call {\ttfamily \_\_del\_\_} multiple times,
       other Pythons exactly once
@@
       PyPy breaks them randomly instead
-    \item in PyPy (xxx Jython?) \_\_del\_\_ cannot be attached to classes after creation
+    \item in PyPy and Jython \_\_del\_\_ cannot be attached to classes after creation
   \end{itemize}
@@
 
+\begin{frame}
+  \frametitle{Questions?}
+  \begin{itemize}
+    \item http://morepypy.blogspot.com
+    \item http://pypy.org
+  \end{itemize}
+\end{frame}
+
 \end{document}

### 2010-02-17 talk/pycon2010/pypyspeed/talk.tex
work on slides. They need a bit more work, but sort-of-ok-now
(hpk, fijal) minor stuff
(hpk, fijal) progress

@@
 \begin{frame}
+  \frametitle{How fast is PyPy?}
+  \pause
+  \begin{itemize}
+    \item pretty fast, in places
+    \item slower than cpython in other places
+    \item overall, it depends
+    \item graphs
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
   \frametitle{JIT - what's that about?}
   \begin{itemize}
+      \pause
     \item {\bf JIT is not a magical device!}
@@
   \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{How well it works in practice?}
 \end{frame}
@@
       \pause
-    \item they prevent optimizations from happening
+    \item {\bf this enables further optimizations}
   \end{itemize}
@@
     \item xxx integers on valuestack and xxx integers in locals
-    \item all of those can be removed
+    \item all boxing can be removed
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{Access costs}
+  \frametitle{Variable access costs}
   \begin{itemize}
@@
   \begin{itemize}
+    \item {\bf JIT} normally removes frame overhead, but
     \item calling {\ttfamily sys.\_getframe()}, {\ttfamily sys.exc\_info()}
     \item exception escaping
+    \item prevents a lot of optimizations
   \end{itemize}
@@
   \begin{itemize}
-    \item instance {\ttfamily \_\_dict\_\_ } lookup becomes a list lookup
+    \item instance {\ttfamily \_\_dict\_\_ } lookup becomes an array lookup
     \item if you're evil, it'll bail back to dict lookup
       \pause
-    \item only for newstyle classes!
+    \item {\bf only for newstyle classes}
   \end{itemize}
@@
     \item dicts on types are version-controlled
-    \item this means methods are usually in known places
+    \item this means method lookup can be removed
       \pause
@@
     \item avoid {\ttfamily *args} and {\ttfamily **kwds}
+    \item however, {\ttfamily f(a=3, b=c)} is fine
   \end{itemize}
@@
   \frametitle{Future}
+  \begin{block}{release end March}
     \begin{itemize}
-    \item release end March
-    \item to try it out
+      \item will contain a working JIT
+      \item will not speed up all cases
+      \item might eat all your memory
     \end{itemize}
+  \end{block}
 \end{frame}

### 2010-09-17 talk/pepm2011/paper.tex
optimistically adding a directory for a paper for PEPM 2011

\documentclass{sigplanconf}

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}

\newboolean{showcomments}
\setboolean{showcomments}{false}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\david[1]{\nb{DAVID}{#1}}
\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
\newcommand{\commentout}[1]{}

\newcommand\ie{i.e.,\xspace}
\newcommand\eg{e.g.,\xspace}

\normalem

\let\oldcite=\cite

\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}

%
\def\sharedaffiliation{%
\end{tabular}
\begin{tabular}{c}}
%
\begin{document}
\conferenceinfo{PEPM'11,} {XXX}
\CopyrightYear{XXX}
\copyrightdata{XXX}

\title{Escape Analysis and Specialization in a Tracing JIT}

\authorinfo{Carl Friedrich Bolz \and Armin Rigo \and Antion Cuni \and Maciek Fijałkowski}
           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany}
           {cfbolz@gmx.de}

%\numberofauthors{3}
%\author{
%\alignauthor Carl Friedrich Bolz\\
%       \email{cfbolz@gmx.de}
%\alignauthor Michael Leuschel\\
%       \email{leuschel@cs.uni-duesseldorf.de}
%\alignauthor David Schneider\\
%      \email{david.schneider@uni-duesseldorf.de}
%      \sharedaffiliation
%      \affaddr{Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany}\\
%}

\maketitle
\begin{abstract}
\footnote{This research is partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
Eureka Eurostars).}
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\category{D.3.4}{Programming Languages}{Processors}[code generation,
interpreters, run-time environments]

\terms
Languages, Performance, Experimentation

\keywords{XXX}%

\section{Introduction}

\section{Background}
\label{sec:Background}

\subsection{PyPy}
\label{sub:PyPy}

\subsection{Tracing JIT Compilers}
\label{sub:JIT_background}

\section{Evaluation}
\label{sec:Evaluation}

\section{Related Work}
\label{sec:related}

\section{Conclusions}
\label{sec:conclusions}

\bibliographystyle{abbrv}
\bibliography{paper,michael}

\end{document}

### 2010-09-17 talk/pepm2011/paper.tex
fix
paperify the blog post

@@
 
+The goal of a just-in-time compiler for a dynamic language is obviously to
+improve the speed of the language over an implementation of the language that
+uses interpretation. The first goal of a JIT is thus to remove the
+interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
+overhead of the interpreter's data structures, such as operand stack etc. The
+second important problem that any JIT for a dynamic language needs to solve is
+how to deal with the overhead of boxing of primitive types and of type
+dispatching. Those are problems that are usually not present in statically typed
+languages.
+
+Boxing of primitive types means that dynamic languages need to be able to handle
+all objects, even integers, floats, etc. in the same way as user-defined
+instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
+heap-structure is allocated for them, that contains the actual value.
+
+Type dispatching is the process of finding the concrete implementation that is
+applicable to the objects at hand when doing a generic operation at hand. An
+example would be the addition of two objects: The addition needs to check what
+the concrete objects are that should be added are, and choose the implementation
+that is fitting for them.
+
+Last year, we wrote a paper \cite{XXX} about how PyPy's meta-JIT
+approach works. These explain how the meta-tracing JIT can remove the overhead
+of bytecode dispatch. In this paper we want to explain how the traces that are
+produced by our meta-tracing JIT are then optimized to also remove some of the
+overhead more closely associated to dynamic languages, such as boxing overhead
+and type dispatching. The most important technique to achieve this is a form of
+escape analysis \cite{XXX} that we call \emph{virtual objects}. This is best
+explained via an example.
+
 \section{Background}
@@
 \label{sub:JIT_background}
+
+\section{Escape Analysis in a Tracing JIT}
+\label{sec:Escape Analysis in a Tracing JIT}
+
+\subsection{Running Example}
+
+For the purpose of this paper, we are going to use a very simple object
+model, that just supports an integer and a float type. The objects support only
+two operations, \texttt{add}, which adds two objects (promoting ints to floats in a
+mixed addition) and \texttt{is\_positive}, which returns whether the number is greater
+than zero. The implementation of \texttt{add} uses classical Smalltalk-like
+double-dispatching. These classes could be part of the implementation of a very
+simple interpreter written in RPython.
+
+\begin{verbatim}
+class Base(object):
+    def add(self, other):
+        """ add self to other """
+        raise NotImplementedError("abstract base")
+    def add__int(self, intother):
+        """ add intother to self, where intother is a Python integer """
+        raise NotImplementedError("abstract base")
+    def add__float(self, floatother):
+        """ add floatother to self, where floatother is a Python float """
+        raise NotImplementedError("abstract base")
+    def is_positive(self):
+        """ returns whether self is positive """
+        raise NotImplementedError("abstract base")
+
+class BoxedInteger(Base):
+    def __init__(self, intval):
+        self.intval = intval
+    def add(self, other):
+        return other.add__int(self.intval)
+    def add__int(self, intother):
+        return BoxedInteger(intother + self.intval)
+    def add__float(self, floatother):
+        return BoxedFloat(floatother + float(self.intval))
+    def is_positive(self):
+        return self.intval > 0
+
+class BoxedFloat(Base):
+    def __init__(self, floatval):
+        self.floatval = floatval
+    def add(self, other):
+        return other.add__float(self.floatval)
+    def add__int(self, intother):
+        return BoxedFloat(float(intother) + self.floatval)
+    def add__float(self, floatother):
+        return BoxedFloat(floatother + self.floatval)
+    def is_positive(self):
+        return self.floatval > 0.0
+\end{verbatim}
+
+Using these classes to implement arithmetic shows the basic problem that a
+dynamic language implementation has. All the numbers are instances of either
+\texttt{BoxedInteger} or \texttt{BoxedFloat}, thus they consume space on the
+heap. Performing many arithmetic operations produces lots of garbage quickly,
+thus putting pressure on the garbage collector. Using double dispatching to
+implement the numeric tower needs two method calls per arithmetic operation,
+which is costly due to the method dispatch.
+
+To understand the problems more directly, let us consider a simple function
+that uses the object model:
+
+\begin{verbatim}
+def f(y):
+    res = BoxedInteger(0)
+    while y.is_positive():
+        res = res.add(y).add(BoxedInteger(-100))
+        y = y.add(BoxedInteger(-1))
+    return res
+\end{verbatim}
+
+The loop iterates \texttt{y} times, and computes something in the process. To
+understand the reason why executing this function is slow, here is the trace
+that is produced by the tracing JIT when executing the function with \texttt{y}
+being a \texttt{BoxedInteger}:
+
+\begin{verbatim}
+# arguments to the trace: p0, p1
+# inside f: res.add(y)
+guard_class(p1, BoxedInteger)
+    # inside BoxedInteger.add
+    i2 = getfield_gc(p1, intval)
+    guard_class(p0, BoxedInteger)
+        # inside BoxedInteger.add__int
+        i3 = getfield_gc(p0, intval)
+        i4 = int_add(i2, i3)
+        p5 = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            setfield_gc(p5, i4, intval)
+# inside f: BoxedInteger(-100) 
+p6 = new(BoxedInteger)
+    # inside BoxedInteger.__init__
+    setfield_gc(p6, -100, intval)
+
+# inside f: .add(BoxedInteger(-100))
+guard_class(p5, BoxedInteger)
+    # inside BoxedInteger.add
+    i7 = getfield_gc(p5, intval)
+    guard_class(p6, BoxedInteger)
+        # inside BoxedInteger.add__int
+        i8 = getfield_gc(p6, intval)
+        i9 = int_add(i7, i8)
+        p10 = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            setfield_gc(p10, i9, intval)
+
+# inside f: BoxedInteger(-1)
+p11 = new(BoxedInteger)
+    # inside BoxedInteger.__init__
+    setfield_gc(p11, -1, intval)
+
+# inside f: y.add(BoxedInteger(-1))
+guard_class(p0, BoxedInteger)
+    # inside BoxedInteger.add
+    i12 = getfield_gc(p0, intval)
+    guard_class(p11, BoxedInteger)
+        # inside BoxedInteger.add__int
+        i13 = getfield_gc(p11, intval)
+        i14 = int_add(i12, i13)
+        p15 = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            setfield_gc(p15, i14, intval)
+
+# inside f: y.is_positive()
+guard_class(p15, BoxedInteger)
+    # inside BoxedInteger.is_positive
+    i16 = getfield_gc(p15, intval)
+    i17 = int_gt(i16, 0)
+# inside f
+guard_true(i17)
+jump(p15, p10)
+\end{verbatim}
+
+(indentation corresponds to the stack level of the traced functions).
+
+The trace is inefficient for a couple of reasons. One problem is that it checks
+repeatedly and redundantly for the class of the objects around, using a
+\texttt{guard\_class} instruction. In addition, some new \texttt{BoxedInteger} instances are
+constructed using the \texttt{new} operation, only to be used once and then forgotten
+a bit later. In the next section, we will see how this can be improved upon,
+using escape analysis.
+
+\subsection{Virtual Objects}
+
+The main insight to improve the code shown in the last section is that some of
+the objects created in the trace using a \texttt{new} operation don't survive very
+long and are collected by the garbage collector soon after their allocation.
+Moreover, they are used only inside the loop, thus we can easily prove that
+nobody else in the program stores a reference to them. The
+idea for improving the code is thus to analyze which objects never escape the
+loop and may thus not be allocated at all.
+
+This process is called \emph{escape analysis}. The escape analysis of
+our tracing JIT works by using \emph{virtual objects}: The trace is walked from
+beginning to end and whenever a \texttt{new} operation is seen, the operation is
+removed and a virtual object is constructed. The virtual object summarizes the
+shape of the object that is allocated at this position in the original trace,
+and is used by the escape analysis to improve the trace. The shape describes
+where the values that would be stored in the fields of the allocated objects
+come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
+object, that shape summary is thus updated and the operation can be removed.
+When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
+from the virtual object, and the operation is also removed.
+
+In the example from last section, the following operations would produce two
+virtual objects, and be completely removed from the optimized trace:
+
+\begin{verbatim}
+p5 = new(BoxedInteger)
+setfield_gc(p5, i4, intval)
+p6 = new(BoxedInteger)
+setfield_gc(p6, -100, intval)
+\end{verbatim}
+
+The virtual object stored in \texttt{p5} would know that it is an \texttt{BoxedInteger}, and that
+the \texttt{intval} field contains \texttt{i4}, the one stored in \texttt{p6} would know that
+its \texttt{intval} field contains the constant -100.
+
+The following operations, that use \texttt{p5} and \texttt{p6} could then be
+optimized using that knowledge:
+
+\begin{verbatim}
+guard_class(p5, BoxedInteger)
+i7 = getfield_gc(p5, intval)
+# inside BoxedInteger.add
+guard_class(p6, BoxedInteger)
+# inside BoxedInteger.add__int
+i8 = getfield_gc(p6, intval)
+i9 = int_add(i7, i8)
+\end{verbatim}
+
+The \texttt{guard\_class} operations can be removed, because the classes of \texttt{p5} and
+\texttt{p6} are known to be \texttt{BoxedInteger}. The \texttt{getfield\_gc} operations can be removed
+and \texttt{i7} and \texttt{i8} are just replaced by \texttt{i4} and -100. Thus the only
+remaining operation in the optimized trace would be:
+
+\begin{verbatim}
+i9 = int_add(i4, -100)
+\end{verbatim}
+
+The rest of the trace is optimized similarly.
+
+So far we have only described what happens when virtual objects are used in
+operations that read and write their fields. When the virtual object is used in
+any other operation, it cannot stay virtual. For example, when a virtual object
+is stored in a globally accessible place, the object needs to actually be
+allocated, as it will live longer than one iteration of the loop.
+
+This is what happens at the end of the trace above, when the \texttt{jump} operation
+is hit. The arguments of the jump are at this point virtual objects. Before the
+jump is emitted, they are \emph{forced}. This means that the optimizers produces code
+that allocates a new object of the right type and sets its fields to the field
+values that the virtual object has. This means that instead of the jump, the
+following operations are emitted:
+
+\begin{verbatim}
+p15 = new(BoxedInteger)
+setfield_gc(p15, i14, intval)
+p10 = new(BoxedInteger)
+setfield_gc(p10, i9, intval)
+jump(p15, p10)
+\end{verbatim}
+
+Note how the operations for creating these two instances has been moved down the
+trace. It looks like for these operations we actually didn't win much, because
+the objects are still allocated at the end. However, the optimization was still
+worthwhile even in this case, because some operations that have been performed
+on the forced virtual objects have been removed (some \texttt{getfield\_gc} operations
+and \texttt{guard\_class} operations).
+
+The final optimized trace of the example looks like this:
+
+\begin{verbatim}
+# arguments to the trace: p0, p1
+guard_class(p1, BoxedInteger)
+i2 = getfield_gc(p1, intval)
+guard_class(p0, BoxedInteger)
+i3 = getfield_gc(p0, intval)
+i4 = int_add(i2, i3)
+i9 = int_add(i4, -100)
+
+guard_class(p0, BoxedInteger)
+i12 = getfield_gc(p0, intval)
+i14 = int_add(i12, -1)
+
+i17 = int_gt(i14, 0)
+guard_true(i17)
+p15 = new(BoxedInteger)
+setfield_gc(p15, i14, intval)
+p10 = new(BoxedInteger)
+setfield_gc(p10, i9, intval)
+jump(p15, p10)
+\end{verbatim}
+
+The optimized trace contains only two allocations, instead of the original five,
+and only three \texttt{guard\_class} operations, from the original seven.
+
+%___________________________________________________________________________
+
+\subsection{Summary}
+
+In this section we described how simple escape analysis within the scope of one
+loop works. This optimizations reduces the allocation of many intermediate data
+structures that become garbage quickly in an interpreter. It also removes a lot
+of the type dispatching overhead. In the next section, we will explain how this
+optimization can be improved further.
+
+% section Escape Analysis in a Tracing JIT (end)
 
@@
 \bibliographystyle{abbrv}
-\bibliography{paper,michael}
+\bibliography{paper}
 

### 2010-09-22 talk/pepm2011/paper.tex
Two typos.

@@
 Type dispatching is the process of finding the concrete implementation that is
-applicable to the objects at hand when doing a generic operation at hand. An
+applicable to the objects at hand when doing a generic operation on them. An
 example would be the addition of two objects: The addition needs to check what
-the concrete objects are that should be added are, and choose the implementation
+the concrete objects that should be added are, and choose the implementation
 that is fitting for them.

### 2010-09-23 talk/pepm2011/paper.tex
import blog post

@@
 
+XXX drop the word "allocation removal" somewhere
+
 \section{Introduction}
@@
 
+\subsection{Tracing JIT Compilers}
+\label{sub:JIT_background}
+
+XXX object model and its reflection in traces (e.g. guard\_class before each method call)
+
+traces and bridges
+
+arguments to traces
+
+getting from the interpreter to traces
+
 \subsection{PyPy}
 \label{sub:PyPy}
-
-\subsection{Tracing JIT Compilers}
-\label{sub:JIT_background}
 
@@
 
+\begin{figure}
 \begin{verbatim}
@@
     def add__int(self, intother):
-        """ add intother to self, where intother is a Python integer """
+        """ add intother to self,
+            where intother is an integer """
         raise NotImplementedError("abstract base")
     def add__float(self, floatother):
-        """ add floatother to self, where floatother is a Python float """
+        """ add floatother to self,
+            where floatother is a float """
         raise NotImplementedError("abstract base")
@@
     def add__float(self, floatother):
-        return BoxedFloat(floatother + float(self.intval))
+        floatvalue = floatother + float(self.intval)
+        return BoxedFloat(floatvalue)
     def is_positive(self):
@@
     def add__int(self, intother):
-        return BoxedFloat(float(intother) + self.floatval)
+        floatvalue = float(intother) + self.floatval
+        return BoxedFloat(floatvalue)
     def add__float(self, floatother):
@@
 \end{verbatim}
+\caption{A simple object model}
+\end{figure}
 
@@
 that is produced by the tracing JIT when executing the function with \texttt{y}
-being a \texttt{BoxedInteger}:
+being a \texttt{BoxedInteger}: XXX make it clear that this is really a trace specific for BoxedInteger
 
+\begin{figure}
 \begin{verbatim}
@@
 \end{verbatim}
+\caption{unoptimized trace for the simple object model}
+\end{figure}
 
@@
 
+\section{Escape Analysis Across Loop Boundaries}
+\label{sec:crossloop}
+
+This section is a bit
+science-fictiony. The algorithm that PyPy currently uses is significantly more
+complex and much harder than the one that is described here. The resulting
+behaviour is very similar, however, so we will use the simpler version (and we
+might switch to that at some point in the actual implementation).
+
+In the last section we described how escape analysis can be used to remove
+many of the allocations of short-lived objects and many of the type dispatches
+that are present in a non-optimized trace. In this section we will improve the
+optimization to also handle more cases.
+
+To understand some more what the optimization described in the last section
+can achieve, look at the following figure:
+
+\includegraphics{figures/obj-lifetime.pdf}
+
+The figure shows a trace before optimization, together with the lifetime of
+various kinds of objects created in the trace. It is executed from top to
+bottom. At the bottom, a jump is used to execute the same loop another time.
+For clarity, the figure shows two iterations of the loop.
+The loop is executed until one of the guards in the trace fails, and the
+execution is aborted.
+
+Some of the operations within this trace are \texttt{new} operations, which each create a
+new instance of some class. These instances are used for a while, e.g. by
+calling methods on them, reading and writing their fields. Some of these
+instances escape, which means that they are stored in some globally accessible
+place or are passed into a function.
+
+Together with the \texttt{new} operations, the figure shows the lifetimes of the
+created objects. Objects in category 1 live for a while, and are then just not
+used any more. The creation of these objects is removed by the
+optimization described in the last section.
+
+Objects in category 2 live for a while and then escape. The optimization of the
+last section deals with them too: the \texttt{new} that creates them and
+the field accesses are deferred, until the point where the object escapes.
+
+The objects in category 3 and 4 are in principle like the objects in category 1
+and 2. They are created, live for a while, but are then passed as an argument
+to the \texttt{jump} operation. In the next iteration they can either die (category
+3) or escape (category 4).
+
+The optimization of the last section considered the passing of an object along a
+jump to be equivalent to escaping. It was thus treating objects in category 3
+and 4 like those in category 2.
+
+The improved optimization described in this section will make it possible to deal
+better with objects in category 3 and 4. This will have two consequences: on
+the one hand, more allocations are removed from the trace (which is clearly
+good). As a side-effect of this, the traces will also be type-specialized.
+
+%___________________________________________________________________________
+
+\subsection{Optimizing Across the Jump}
+
+Let's look at the final trace obtained in the last section for the example loop.
+The final trace was much better than the original one, because many allocations
+were removed from it. However, it also still contained allocations:
+
+\begin{figure}
+\includegraphics{figures/step1.pdf}
+\end{figure}
+
+The two new \texttt{BoxedIntegers} stored in \texttt{p15} and \texttt{p10} are passed into
+the next iteration of the loop. The next iteration will check that they are
+indeed \texttt{BoxedIntegers}, read their \texttt{intval} fields and then not use them
+any more. Thus those instances are in category 3.
+
+In its current state the loop
+allocates two \texttt{BoxedIntegers} at the end of every iteration, that then die
+very quickly in the next iteration. In addition, the type checks at the start
+of the loop are superfluous, at least after the first iteration.
+
+The reason why we cannot optimize the remaining allocations away is because
+their lifetime crosses the jump. To improve the situation, a little trick is
+needed. The trace above represents a loop, i.e. the jump at the end jumps to
+the beginning. Where in the loop the jump occurs is arbitrary, since the loop
+can only be left via failing guards anyway. Therefore it does not change the
+semantics of the loop to put the jump at another point into the trace and we
+can move the \texttt{jump} operation just above the allocation of the objects that
+appear in the current \texttt{jump}. This needs some care, because the arguments to
+\texttt{jump} are all currently live variables, thus they need to be adapted.
+
+If we do that for our example trace above, the trace looks like this:
+\begin{figure}
+\includegraphics{figures/step2.pdf}
+\end{figure}
+
+Now the lifetime of the remaining allocations no longer crosses the jump, and
+we can run our escape analysis a second time, to get the following trace:
+\begin{figure}
+\includegraphics{figures/step3.pdf}
+\end{figure}
+
+This result is now really good. The code performs the same operations than
+the original code, but using direct CPU arithmetic and no boxing, as opposed to
+the original version which used dynamic dispatching and boxing.
+
+Looking at the final trace it is also completely clear that specialization has
+happened. The trace corresponds to the situation in which the trace was
+originally recorded, which happened to be a loop where \texttt{BoxedIntegers} were
+used. The now resulting loop does not refer to the \texttt{BoxedInteger} class at
+all any more, but it still has the same behaviour. If the original loop had
+used \texttt{BoxedFloats}, the final loop would use \texttt{float\_*} operations
+everywhere instead (or even be very different, if the object model had
+user-defined classes).
+
+%___________________________________________________________________________
+
+\subsection{Entering the Loop}
+
+The approach of placing the \texttt{jump} at some other point in the loop leads to
+one additional complication that we glossed over so far. The beginning of the
+original loop corresponds to a point in the original program, namely the
+\texttt{while} loop in the function \texttt{f} from the last section.
+
+Now recall that in a VM that uses a tracing JIT, all programs start by being
+interpreted. This means that when \texttt{f} is executed by the interpreter, it is
+easy to go from the interpreter to the first version of the compiled loop.
+After the \texttt{jump} is moved and the escape analysis optimization is applied a
+second time, this is no longer easily possible.  In particular, the new loop
+expects two integers as input arguments, while the old one expected two
+instances.
+
+To make it possible to enter the loop directly from the intepreter, there
+needs to be some additional code that enters the loop by taking as input
+arguments what is available to the interpreter, i.e. two instances. This
+additional code corresponds to one iteration of the loop, which is thus
+peeled off \cite{XXX}:
+
+\begin{figure}
+\includegraphics{figures/step4.pdf}
+\end{figure}
+
+%___________________________________________________________________________
+
+\subsection{Summary}
+
+The optimization described in this section can be used to optimize away
+allocations in category 3 and improve allocations in category 4, by deferring
+them until they are no longer avoidable. A side-effect of these optimizations
+is also that the optimized loops are specialized for the types of the variables
+that are used inside them.
+
+% section Escape Analysis Across Loop Boundaries (end)
+
 \section{Evaluation}

### 2010-09-23 talk/pepm2011/paper.tex
move stuff around a bit

@@
 \label{sub:PyPy}
+
+\section{Object Lifetimes in a Tracing JIT}
+\label{sec:lifetimes}
+
+% section Object Lifetimes in a Tracing JIT (end)
+
+To understand the problems that this paper is trying to solve some more, we
+first need to understand various cases of object lifetimes that can occur in a
+tracing JIT compiler.
+
+\begin{figure}
+\includegraphics{figures/obj-lifetime.pdf}
+
+\caption{Object Lifetimes in a Trace}
+\label{fig:lifetimes}
+\end{figure}
+
+The figure shows a trace before optimization, together with the lifetime of
+various kinds of objects created in the trace. It is executed from top to
+bottom. At the bottom, a jump is used to execute the same loop another time.
+For clarity, the figure shows two iterations of the loop.
+The loop is executed until one of the guards in the trace fails, and the
+execution is aborted.
+
+Some of the operations within this trace are \texttt{new} operations, which each create a
+new instance of some class. These instances are used for a while, e.g. by
+calling methods on them, reading and writing their fields. Some of these
+instances escape, which means that they are stored in some globally accessible
+place or are passed into a function.
+
+Together with the \texttt{new} operations, the figure shows the lifetimes of the
+created objects. Objects in category 1 live for a while, and are then just not
+used any more. The creation of these objects is removed by the
+optimization described in the last section.
+
+Objects in category 2 live for a while and then escape. The optimization of the
+last section deals with them too: the \texttt{new} that creates them and
+the field accesses are deferred, until the point where the object escapes.
+
+The objects in category 3 and 4 are in principle like the objects in category 1
+and 2. They are created, live for a while, but are then passed as an argument
+to the \texttt{jump} operation. In the next iteration they can either die (category
+3) or escape (category 4).
 
@@
 
-This section is a bit
-science-fictiony. The algorithm that PyPy currently uses is significantly more
-complex and much harder than the one that is described here. The resulting
-behaviour is very similar, however, so we will use the simpler version (and we
-might switch to that at some point in the actual implementation).
-
 In the last section we described how escape analysis can be used to remove
@@
 optimization to also handle more cases.
-
-To understand some more what the optimization described in the last section
-can achieve, look at the following figure:
-
-\includegraphics{figures/obj-lifetime.pdf}
-
-The figure shows a trace before optimization, together with the lifetime of
-various kinds of objects created in the trace. It is executed from top to
-bottom. At the bottom, a jump is used to execute the same loop another time.
-For clarity, the figure shows two iterations of the loop.
-The loop is executed until one of the guards in the trace fails, and the
-execution is aborted.
-
-Some of the operations within this trace are \texttt{new} operations, which each create a
-new instance of some class. These instances are used for a while, e.g. by
-calling methods on them, reading and writing their fields. Some of these
-instances escape, which means that they are stored in some globally accessible
-place or are passed into a function.
-
-Together with the \texttt{new} operations, the figure shows the lifetimes of the
-created objects. Objects in category 1 live for a while, and are then just not
-used any more. The creation of these objects is removed by the
-optimization described in the last section.
-
-Objects in category 2 live for a while and then escape. The optimization of the
-last section deals with them too: the \texttt{new} that creates them and
-the field accesses are deferred, until the point where the object escapes.
-
-The objects in category 3 and 4 are in principle like the objects in category 1
-and 2. They are created, live for a while, but are then passed as an argument
-to the \texttt{jump} operation. In the next iteration they can either die (category
-3) or escape (category 4).
 
@@
 \subsection{Optimizing Across the Jump}
+
+\footnote{This section is a bit
+science-fictiony. The algorithm that PyPy currently uses is significantly more
+complex and much harder than the one that is described here. The resulting
+behaviour is very similar, however, so we will use the simpler version (and we
+might switch to that at some point in the actual implementation).}
 

### 2010-09-26 talk/pepm2011/paper.tex
whack at the introduction

@@
 how to deal with the overhead of boxing of primitive types and of type
-dispatching. Those are problems that are usually not present in statically typed
-languages.
+dispatching. Those are problems that are usually not present or at least less
+severe in statically typed languages.
 
 Boxing of primitive types means that dynamic languages need to be able to handle
-all objects, even integers, floats, etc. in the same way as user-defined
+all objects, even integers, floats, bools etc. in the same way as user-defined
 instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
-heap-structure is allocated for them, that contains the actual value.
+heap-structure is allocated for them, that contains the actual value. Boxing
+primitive types can be very costly, because XXX
 
@@
 the concrete objects that should be added are, and choose the implementation
-that is fitting for them.
+that is fitting for them. Type dispatching is a very common operation in a
+dynamic language because no types are known at compile time, so all operations
+need it.
 
-Last year, we wrote a paper \cite{XXX} about how PyPy's meta-JIT
-approach works. These explain how the meta-tracing JIT can remove the overhead
+A recently popular approach to implementing just-in-time compilers for dynamic
+languages is that of a tracing JIT. A tracing JIT often takes the form of an
+extension to an existing interpreter, which can be sped up that way. The PyPy
+project is an environment for implementing dynamic programming languages. It's
+approach to doing so is to straightforwardly implement an interpreter for the
+to-be-implemented language, and then use powerful tools to turn the interpreter
+into an efficient VM that also contains a just-in-time compiler. This compiler
+is automatically generated from the interpreter using partial-evaluation-like
+techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
+tracing JIT compilers is described in Section~\ref{sec:Background}
+
+The tracing JIT approach that the PyPy project is taking removes the overhead
 of bytecode dispatch. In this paper we want to explain how the traces that are
-produced by our meta-tracing JIT are then optimized to also remove some of the
+produced by our meta-tracing JIT can be optimized to also remove some of the
 overhead more closely associated to dynamic languages, such as boxing overhead
-and type dispatching. The most important technique to achieve this is a form of
-escape analysis \cite{XXX} that we call \emph{virtual objects}. This is best
-explained via an example.
+and type dispatching. To understand the problem more closely, we analyze the
+occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
+technique to achieve this is a form of escape analysis \cite{XXX} that we call
+\emph{virtual objects}, which is described in Section~\ref{sec:virtuals}. The
+basic approach of virtual objects can then be extended to also be used for
+type-specializing the traces that are produced by the tracing JIT
+(Section~\ref{sec:crossloop}). In Section~\ref{sec:XXX} we describe some
+supporting techniques that are not central to the approach, but are needed to
+improve the results. The introduced techniques are evaluated in
+Section~\ref{sec:Evaluation}.
+
+The contributions of this paper are:
+
+\begin{enumerate}
+    \item An efficient and effective escape analysis for a tracing JIT
+    \item XXX
+\end{enumerate}
 
@@
 \section{Escape Analysis in a Tracing JIT}
-\label{sec:Escape Analysis in a Tracing JIT}
+\label{sec:virtuals}
 

### 2010-09-26 talk/pepm2011/paper.tex
fix an XXX, add a thought

@@
 heap-structure is allocated for them, that contains the actual value. Boxing
-primitive types can be very costly, because XXX
+primitive types can be very costly, because a lot of common operations,
+particularly all arithmetic operations, have to produce a new box, in addition
+to the actual computation they do. Because the boxes are allocated on the heap,
+producing a lot of them puts pressure on the garbage collector.
 
@@
 
-XXX object model and its reflection in traces (e.g. guard\_class before each method call)
-
 traces and bridges
@@
 getting from the interpreter to traces
+
+XXX object model and its reflection in traces (e.g. guard\_class before each method call)
 
@@
 
+XXX optimization particularly effective for chains of operations
+
 %___________________________________________________________________________

### 2010-09-26 talk/pepm2011/paper.tex
reorder some stuff, write two paragraphs about pypy

@@
 of bytecode dispatch. In this paper we want to explain how the traces that are
-produced by our meta-tracing JIT can be optimized to also remove some of the
+produced by PyPy's tracing JIT can be optimized to also remove some of the
 overhead more closely associated to dynamic languages, such as boxing overhead
@@
 \emph{virtual objects}, which is described in Section~\ref{sec:virtuals}. The
-basic approach of virtual objects can then be extended to also be used for
+goal of virtual objects is to remove allocations of temporary objects that have
+a predictable lifetime and to optimize type dispatching in the process.
+
+The basic approach of virtual objects can then be extended to also be used for
 type-specializing the traces that are produced by the tracing JIT
@@
 improve the results. The introduced techniques are evaluated in
-Section~\ref{sec:Evaluation}.
+Section~\ref{sec:Evaluation} using PyPy's Python interpreter as a case study.
 
@@
 \begin{enumerate}
-    \item An efficient and effective escape analysis for a tracing JIT
+    \item An efficient and effective algorithm for removing objects allocations in a tracing JIT.
     \item XXX
@@
 \label{sec:Background}
+
+\subsection{PyPy}
+\label{sub:PyPy}
+
+The work described in this paper was done in the context of the PyPy project
+\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
+be implemented in a simple yet efficient way. The approach taken when
+implementing a language with PyPy is to write an interpreter for the language in
+\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
+is a subset of Python chosen in such a way, that type inference becomes
+possible. The language interpreter can thus be translated with the help of
+PyPy's tools into a VM on the C level. Because the interpreter is written at a
+relatively high level, the language implementation is kept free of low-level
+details, such as object layout, garbage collection or memory model. Those
+aspects of the final VM are woven into the generated code during the translation
+to C.
+
+The feature that makes PyPy more than a compiler with a runtime system is it's
+support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
+the translation to C, PyPy's tools can generate a just-in-time compiler for the
+language that the interpreter is implementing. This process is not fully
+automatic, but needs to be guided by the language implementer by some
+source-code hints.
 
@@
 XXX object model and its reflection in traces (e.g. guard\_class before each method call)
-
-\subsection{PyPy}
-\label{sub:PyPy}
-
-\section{Object Lifetimes in a Tracing JIT}
-\label{sec:lifetimes}
-
-% section Object Lifetimes in a Tracing JIT (end)
-
-To understand the problems that this paper is trying to solve some more, we
-first need to understand various cases of object lifetimes that can occur in a
-tracing JIT compiler.
-
-\begin{figure}
-\includegraphics{figures/obj-lifetime.pdf}
-
-\caption{Object Lifetimes in a Trace}
-\label{fig:lifetimes}
-\end{figure}
-
-The figure shows a trace before optimization, together with the lifetime of
-various kinds of objects created in the trace. It is executed from top to
-bottom. At the bottom, a jump is used to execute the same loop another time.
-For clarity, the figure shows two iterations of the loop.
-The loop is executed until one of the guards in the trace fails, and the
-execution is aborted.
-
-Some of the operations within this trace are \texttt{new} operations, which each create a
-new instance of some class. These instances are used for a while, e.g. by
-calling methods on them, reading and writing their fields. Some of these
-instances escape, which means that they are stored in some globally accessible
-place or are passed into a function.
-
-Together with the \texttt{new} operations, the figure shows the lifetimes of the
-created objects. Objects in category 1 live for a while, and are then just not
-used any more. The creation of these objects is removed by the
-optimization described in the last section.
-
-Objects in category 2 live for a while and then escape. The optimization of the
-last section deals with them too: the \texttt{new} that creates them and
-the field accesses are deferred, until the point where the object escapes.
-
-The objects in category 3 and 4 are in principle like the objects in category 1
-and 2. They are created, live for a while, but are then passed as an argument
-to the \texttt{jump} operation. In the next iteration they can either die (category
-3) or escape (category 4).
-
-\section{Escape Analysis in a Tracing JIT}
-\label{sec:virtuals}
 
@@
 using escape analysis.
+
+\section{Object Lifetimes in a Tracing JIT}
+\label{sec:lifetimes}
+
+% section Object Lifetimes in a Tracing JIT (end)
+
+To understand the problems that this paper is trying to solve some more, we
+first need to understand various cases of object lifetimes that can occur in a
+tracing JIT compiler.
+
+\begin{figure}
+\includegraphics{figures/obj-lifetime.pdf}
+
+\caption{Object Lifetimes in a Trace}
+\label{fig:lifetimes}
+\end{figure}
+
+The figure shows a trace before optimization, together with the lifetime of
+various kinds of objects created in the trace. It is executed from top to
+bottom. At the bottom, a jump is used to execute the same loop another time.
+For clarity, the figure shows two iterations of the loop.
+The loop is executed until one of the guards in the trace fails, and the
+execution is aborted.
+
+Some of the operations within this trace are \texttt{new} operations, which each create a
+new instance of some class. These instances are used for a while, e.g. by
+calling methods on them, reading and writing their fields. Some of these
+instances escape, which means that they are stored in some globally accessible
+place or are passed into a function.
+
+Together with the \texttt{new} operations, the figure shows the lifetimes of the
+created objects. Objects in category 1 live for a while, and are then just not
+used any more. The creation of these objects is removed by the
+optimization described in the last section.
+
+Objects in category 2 live for a while and then escape. The optimization of the
+last section deals with them too: the \texttt{new} that creates them and
+the field accesses are deferred, until the point where the object escapes.
+
+The objects in category 3 and 4 are in principle like the objects in category 1
+and 2. They are created, live for a while, but are then passed as an argument
+to the \texttt{jump} operation. In the next iteration they can either die (category
+3) or escape (category 4).
+
+\section{Escape Analysis in a Tracing JIT}
+\label{sec:virtuals}
 

### 2010-09-26 talk/pepm2011/paper.tex
more in the brackground section about tracing JITs

@@
 
-\authorinfo{Carl Friedrich Bolz \and Armin Rigo \and Antion Cuni \and Maciek Fijałkowski}
-           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany}
-           {cfbolz@gmx.de}
+\authorinfo{Carl Friedrich Bolz \and Antion Cuni \and Maciek Fijałkowski \and Samuele Pedroni \and Armin Rigo}
+           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}
+           {cfbolz@gmx.de XXX}
 
@@
 techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
-tracing JIT compilers is described in Section~\ref{sec:Background}
+tracing JIT compilers is described in Section~\ref{sec:Background}.
 
@@
 aspects of the final VM are woven into the generated code during the translation
-to C.
+to C. XXX languages that are done using PyPy
 
@@
 automatic, but needs to be guided by the language implementer by some
-source-code hints.
+source-code hints. Semi-automatically generating a JIT compiler has many advantages
+over writing one manually, which is an error-prone and tedious process. The
+generated JIT has the same semantics as the interpreter by construction, and all
+languages implemented using PyPy benefit from improvements to the JIT generator.
+The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
+compiler}, a concept which will be explained in more detail in the next section.
 
@@
 
-traces and bridges
+Tracing JITs are a recently more popular approach to write just-in-time
+compilers for dynamic languages \cite{XXX}. Their origins lie in the Dynamo
+project, that used a tracing approach to optimize machine code using execution
+traces \cite{XXX}. They were then adapted to be used for a very light-weight
+Java VM \cite{XXX} and afterwards used in several implementations of dynamic
+languages, such as JavaScript \cite{XXX}, Lua \cite{XXX} and now Python via
+PyPy.
+
+The core idea of tracing JITs is to focus the optimization effort of the JIT
+compiler on the hot paths of the core loops of the program and to just use an
+interpreter for the less commonly executed parts. VMs that use a tracing JIT are
+thus mixed-mode execution environments, they contain both an interpreter and a
+JIT compiler. By default the interpreter is used to execute the program, doing
+some light-weight profiling at the same time. This profiling is used to identify
+the hot loops of the program. If a hot loop is found in that way, the
+interpreter enters a special \emph{tracing mode}. In this tracing mode, the
+interpreter records all operations that it is executing while running one
+iteration of the hot loop. This history of executed operations of one loop is
+called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
+it always ends with a jump to its own beginning.
+
+This trace of operations is then the basis of the generated code. The trace is
+optimized in some ways, and then turned into machine code. Generating machine
+code is simple, because the traces are linear and the operations are very close
+to machine level. The trace corresponds to one concrete execution of a loop,
+therefore the code generated from it is only one possible path through the loop.
+To make sure that the trace is maintaining the correct semantics, it contains a
+\emph{guard} at all places where the execution could have diverged from the
+path. Those guards check the assumptions under which execution can stay on the
+trace. As an example, if a loop contains an \texttt{if} statement, the trace
+will contain the execution of one of the paths only, which is the path that was
+taken during the production of the trace. The trace will also contain a guard
+that checks that the condition of the \texttt{if} statement is true, because if
+it isn't, the rest of the trace is not valid.
+
+When generating machine code, every guard is be turned into a quick check to
+see whether the assumption still holds. When such a guard is hit during the
+execution of the machine code and the assumption does not hold, the execution of
+the machine code is stopped, and interpreter continues to run from that point
+on. These guards are the only mechanism to stop the execution of a trace, the
+loop end condition also takes the form of a guard.
 

### 2010-09-26 talk/pepm2011/paper.tex
whoops

@@
 
-\authorinfo{Carl Friedrich Bolz \and Antion Cuni \and Maciek Fijałkowski \and Samuele Pedroni \and Armin Rigo}
+\authorinfo{Carl Friedrich Bolz \and Antonio Cuni \and Maciek Fijałkowski \and Samuele Pedroni \and Armin Rigo}
            {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}

### 2010-09-27 talk/pepm2011/paper.tex
use more official form

--- 2010-09-26 talk/pepm2011/paper.tex
@@
 
-\authorinfo{Carl Friedrich Bolz \and Antion Cuni \and Maciek Fijałkowski \and Samuele Pedroni \and Armin Rigo}
+\authorinfo{Carl Friedrich Bolz \and Antonio Cuni \and Maciej Fijałkowski \and Samuele Pedroni \and Armin Rigo}
            {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}

### 2010-09-27 talk/pepm2011/paper.tex
refactor a number of things, add many XXXs

@@
 
+XXX define "escape analysis"
+
 \section{Introduction}
+
+XXX need to re-target introduction a bit to fit PEPMs focus
 
@@
 languages is that of a tracing JIT. A tracing JIT often takes the form of an
-extension to an existing interpreter, which can be sped up that way. The PyPy
-project is an environment for implementing dynamic programming languages. It's
+extension to an existing interpreter, which can be sped up that way. This
+approach is also the one taken by the PyPy project, which is an environment for
+implementing dynamic programming languages. PyPy's
 approach to doing so is to straightforwardly implement an interpreter for the
@@
 \begin{enumerate}
-    \item An efficient and effective algorithm for removing objects allocations in a tracing JIT.
+    \item An efficient and effective algorithm for removing object allocations in a tracing JIT.
     \item XXX
@@
 loop end condition also takes the form of a guard.
+
+bridges?
 
@@
 
-The loop iterates \texttt{y} times, and computes something in the process. To
-understand the reason why executing this function is slow, here is the trace
-that is produced by the tracing JIT when executing the function with \texttt{y}
-being a \texttt{BoxedInteger}: XXX make it clear that this is really a trace specific for BoxedInteger
+The loop iterates \texttt{y} times, and computes something in the process.
+Simply running this function is slow, because there are lots of virtual method
+calls inside the loop, one for each \texttt{is\_positive} and even two for each
+call to \texttt{add}. These method calls need to check the type of the involved
+objects repeatedly and redundantly. In addition, a lot of objects are created
+when executing that loop, many of these objects do not survive for very long.
+The actual computation that is performed by \texttt{f} is simply a number of
+float or integer additions.
 
@@
 \end{verbatim}
-\caption{unoptimized trace for the simple object model}
+\label{fig:unopt-trace}
+\caption{Unoptimized Trace for the Simple Object Model}
 \end{figure}
 
-(indentation corresponds to the stack level of the traced functions).
+If the function is executed using the tracing JIT, with \texttt{y} being a
+\texttt{BoxedInteger}, the produced trace looks like
+Figure~\ref{fig:unopt-trace}. The operations in the trace are indented to
+correspond to the stack level of the function that contains the traced
+operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
+looks at the number of \texttt{new}, \texttt{set/getfield\_gc} and
+\texttt{guard\_class} operations.
 
-The trace is inefficient for a couple of reasons. One problem is that it checks
-repeatedly and redundantly for the class of the objects around, using a
-\texttt{guard\_class} instruction. In addition, some new \texttt{BoxedInteger} instances are
-constructed using the \texttt{new} operation, only to be used once and then forgotten
-a bit later. In the next section, we will see how this can be improved upon,
-using escape analysis.
+Note how the functions that are called by \texttt{f} are automatically inlined
+into the trace. The method calls are always preceded by a \texttt{guard\_class}
+operation, to check that the class of the receiver is the same as the one that
+was observed during tracing.\footnote{\texttt{guard\_class} performs a precise
+class check, not checking for subclasses} These guards make the trace specific
+to the situation where \texttt{y} is really a \texttt{BoxedInteger}, it can
+already be said to be specialized for \texttt{BoxedIntegers}. When the trace is
+turned into machine code and then executed with \texttt{BoxedFloats}, the
+first \texttt{guard\_class} instruction will fail and execution will continue
+using the interpreter.
+
+XXX simplify traces a bit more
+get rid of \_gc suffix in set/getfield\_gc
+
+In the next section, we will see how this can be improved upon, using escape
+analysis. XXX
 
@@
 \begin{figure}
-\includegraphics{figures/obj-lifetime.pdf}
+\includegraphics[scale=0.7]{figures/obj-lifetime.pdf}
 
@@
 
-The figure shows a trace before optimization, together with the lifetime of
-various kinds of objects created in the trace. It is executed from top to
-bottom. At the bottom, a jump is used to execute the same loop another time.
-For clarity, the figure shows two iterations of the loop.
-The loop is executed until one of the guards in the trace fails, and the
-execution is aborted.
+Figure~\ref{fig:lifetimes} shows a trace before optimization, together with the
+lifetime of various kinds of objects created in the trace. It is executed from
+top to bottom. At the bottom, a jump is used to execute the same loop another
+time (for clarity, the figure shows two iterations of the loop). The loop is
+executed until one of the guards in the trace fails, and the execution is
+aborted and interpretation resumes.
 
-Some of the operations within this trace are \texttt{new} operations, which each create a
-new instance of some class. These instances are used for a while, e.g. by
-calling methods on them, reading and writing their fields. Some of these
-instances escape, which means that they are stored in some globally accessible
-place or are passed into a function.
+Some of the operations within this trace are \texttt{new} operations, which each
+create a new instance of some class. These instances are used for a while, e.g.
+by calling methods on them (which are inlined into the trace), reading and
+writing their fields. Some of these instances \emph{escape}, which means that
+they are stored in some globally accessible place or are passed into a function.
 
 Together with the \texttt{new} operations, the figure shows the lifetimes of the
-created objects. Objects in category 1 live for a while, and are then just not
-used any more. The creation of these objects is removed by the
-optimization described in the last section.
+created objects. The objects that are created within a trace using \texttt{new}
+fall into one of several categories:
 
-Objects in category 2 live for a while and then escape. The optimization of the
-last section deals with them too: the \texttt{new} that creates them and
-the field accesses are deferred, until the point where the object escapes.
+\begin{itemize}
+    \item Category 1: Objects that live for a while, and are then just not
+    used any more.
 
-The objects in category 3 and 4 are in principle like the objects in category 1
-and 2. They are created, live for a while, but are then passed as an argument
-to the \texttt{jump} operation. In the next iteration they can either die (category
-3) or escape (category 4).
+    \item Category 2: Objects that live for a while and then escape.
+
+    \item Category 3: Objects that live for a while, survive across the jump to
+    the beginning of the loop, and are then not used any more.
+
+    \item Category 4: Objects that live for a while, survive across the jump,
+    and then escape. To these we also count the objects that live across several
+    jumps and then either escape or stop being used\footnote{In theory, the
+    approach of Section~\ref{sec:XXX} works also for objects that live for
+    exactly $n>1$ iterations and then don't escape, but we expect this to be a
+    very rare case, so we do not handle it.}
+\end{itemize}
+
+The objects that are allocated in the example trace in
+Figure~\ref{fig:unopt-trace} fall into categories 1 and 3. Objects stored in
+\texttt{p5, p6, p11 XXX} are in category 1, objects in \texttt{p10, p15} are in
+category 3.
+
+The creation of objects in category 1 is removed by the optimization described
+in Section~\ref{sec:virtuals}. XXX
 
@@
 
-The main insight to improve the code shown in the last section is that some of
-the objects created in the trace using a \texttt{new} operation don't survive very
-long and are collected by the garbage collector soon after their allocation.
-Moreover, they are used only inside the loop, thus we can easily prove that
-nobody else in the program stores a reference to them. The
-idea for improving the code is thus to analyze which objects never escape the
-loop and may thus not be allocated at all.
+The main insight to improve the code shown in the last section is that objects
+in category 1 don't survive very long and are collected by the garbage collector
+soon after their allocation. Moreover, they are used only inside the loop and
+nobody else in the program stores a reference to them. The idea for improving
+the code is thus to analyze which objects fall in category 1 and which may thus
+not be allocated at all.
+
+XXX is "symbolic execution" the right word to drop?
 
@@
 shape of the object that is allocated at this position in the original trace,
-and is used by the escape analysis to improve the trace. The shape describes
+and is used by the optimization to improve the trace. The shapes describe
 where the values that would be stored in the fields of the allocated objects
@@
 from the virtual object, and the operation is also removed.
+
+XXX what happens on a guard\_class?
 
@@
 
-\begin{verbatim}
-p15 = new(BoxedInteger)
-setfield_gc(p15, i14, intval)
-p10 = new(BoxedInteger)
-setfield_gc(p10, i9, intval)
-jump(p15, p10)
-\end{verbatim}
+\texttt{
+\begin{tabular}{l} 
+$p_{15}$ = new(BoxedInteger) \\
+setfield\_gc($p_{15}$, $i_{14}$, intval) \\
+$p_{10}$ = new(BoxedInteger) \\
+setfield\_gc($p_{10}$, $i_{9}$, intval) \\
+jump($p_{15}$, $p_{10}$) \\
+\end{tabular}
+}
 
-Note how the operations for creating these two instances has been moved down the
+Note how the operations for creating these two instances have been moved down the
 trace. It looks like for these operations we actually didn't win much, because
@@
 
-The final optimized trace of the example looks like this:
+\begin{figure}
+\includegraphics{figures/step1.pdf}
+\label{fig:step1}
+\caption{Resulting Trace After Allocation Removal}
+\end{figure}
 
-\begin{verbatim}
-# arguments to the trace: p0, p1
-guard_class(p1, BoxedInteger)
-i2 = getfield_gc(p1, intval)
-guard_class(p0, BoxedInteger)
-i3 = getfield_gc(p0, intval)
-i4 = int_add(i2, i3)
-i9 = int_add(i4, -100)
-
-guard_class(p0, BoxedInteger)
-i12 = getfield_gc(p0, intval)
-i14 = int_add(i12, -1)
-
-i17 = int_gt(i14, 0)
-guard_true(i17)
-p15 = new(BoxedInteger)
-setfield_gc(p15, i14, intval)
-p10 = new(BoxedInteger)
-setfield_gc(p10, i9, intval)
-jump(p15, p10)
-\end{verbatim}
+The final optimized trace of the example can be seen in
+Figure~\ref{fig:step1}.
 
@@
 and only three \texttt{guard\_class} operations, from the original seven.
+
+\subsection{Algorithm}
+\label{sub:Algorithm}
+
+XXX want some sort of pseudo-code
+
+% subsection Algorithm (end)
 
@@
 optimization can be improved further.
+
+XXX Category 2 The optimization of
+Section~\ref{sec:virtuals} deals with them too: the \texttt{new} that creates them and
+the field accesses are deferred, until the point where the object escapes.
 
@@
 were removed from it. However, it also still contained allocations:
-
-\begin{figure}
-\includegraphics{figures/step1.pdf}
-\end{figure}
 

### 2010-09-27 talk/pepm2011/paper.tex
Tweaks.

@@
 technique to achieve this is a form of escape analysis \cite{XXX} that we call
-\emph{virtual objects}, which is described in Section~\ref{sec:virtuals}. The
+\emph{virtual objects},\footnote{The terminology comes from \cite{Psyco}}
+which is described in Section~\ref{sec:virtuals}. The
 goal of virtual objects is to remove allocations of temporary objects that have
@@
 
+XXX this is not an RPython interpreter; put a reference to the previous
+paper to show how we deal with an interpreted piece of code and remove
+the interpretation overhead, turning it into basically something
+equivalent to the example here, which is the start of the present paper.
+
 \begin{verbatim}
@@
 \texttt{BoxedInteger}, the produced trace looks like
-Figure~\ref{fig:unopt-trace}. The operations in the trace are indented to
+Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
 correspond to the stack level of the function that contains the traced
@@
 writing their fields. Some of these instances \emph{escape}, which means that
-they are stored in some globally accessible place or are passed into a function.
+they are stored in some globally accessible place or are passed into a
+non-inlined function via a residual call.
 
@@
     and then escape. To these we also count the objects that live across several
-    jumps and then either escape or stop being used\footnote{In theory, the
+    jumps and then either escape or stop being used.\footnote{In theory, the
     approach of Section~\ref{sec:XXX} works also for objects that live for
@@
 The main insight to improve the code shown in the last section is that objects
-in category 1 don't survive very long and are collected by the garbage collector
-soon after their allocation. Moreover, they are used only inside the loop and
+in category 1 don't survive very long -- they are used only inside the loop and
 nobody else in the program stores a reference to them. The idea for improving
-the code is thus to analyze which objects fall in category 1 and which may thus
+the code is thus to analyze which objects fall in category 1 and may thus
 not be allocated at all.
@@
 beginning to end and whenever a \texttt{new} operation is seen, the operation is
-removed and a virtual object is constructed. The virtual object summarizes the
+removed and a virtual object\footnote{XXX what I have in mind when I talk
+of ``virtual object'' is the run-time behavior -- i.e. a real object that
+would exist at run-time, except that it has be virtual-ized.  Here you seem
+to mean rather ``virtual object description'' or something.}
+is constructed. The virtual object summarizes the
 shape of the object that is allocated at this position in the original trace,
@@
 
+XXX ``the trace above'' is dangerous; should mention all figures by numbers
+
 This is what happens at the end of the trace above, when the \texttt{jump} operation
@@
 following operations are emitted:
+
+XXX should the variables be written in $math-style$ everywhere?
 
@@
 
+XXX the figure is moved elsewhere by latex
+
 Now the lifetime of the remaining allocations no longer crosses the jump, and
@@
 \end{figure}
+
+XXX the figure is moved elsewhere by latex
 
@@
 everywhere instead (or even be very different, if the object model had
-user-defined classes).
+more different classes).
 
@@
 
+XXX the figure is moved elsewhere by latex
+
 XXX optimization particularly effective for chains of operations

### 2010-09-27 talk/pepm2011/paper.tex
More tweaks.

@@
 
-The goal of a just-in-time compiler for a dynamic language is obviously to
+The goal of a just-in-time (JIT) compiler for a dynamic language is obviously to
 improve the speed of the language over an implementation of the language that
@@
 to-be-implemented language, and then use powerful tools to turn the interpreter
-into an efficient VM that also contains a just-in-time compiler. This compiler
+into an efficient virtual machine (VM) that also contains a just-in-time compiler.
+This compiler
 is automatically generated from the interpreter using partial-evaluation-like
@@
 be implemented in a simple yet efficient way. The approach taken when
-implementing a language with PyPy is to write an interpreter for the language in
+implementing a language with PyPy is to write an \emph{interpreter}
+for the language in
 \emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
-is a subset of Python chosen in such a way, that type inference becomes
-possible. The language interpreter can thus be translated with the help of
+is a subset of Python chosen in such a way that type inference becomes
+possible. The language interpreter can then be compiled (``translated'') with
 PyPy's tools into a VM on the C level. Because the interpreter is written at a
@@
 the translation to C, PyPy's tools can generate a just-in-time compiler for the
-language that the interpreter is implementing. This process is not fully
-automatic, but needs to be guided by the language implementer by some
-source-code hints. Semi-automatically generating a JIT compiler has many advantages
-over writing one manually, which is an error-prone and tedious process. The
-generated JIT has the same semantics as the interpreter by construction, and all
-languages implemented using PyPy benefit from improvements to the JIT generator.
+language that the interpreter is implementing. This process is mostly
+automatic; it only needs to be guided by the language implementer by a small number of
+source-code hints. Mostly-automatically generating a JIT compiler has many advantages
+over writing one manually, which is an error-prone and tedious process.
+By construction, the generated JIT has the same semantics as the interpreter, and
+the process benefits all languages implemented as an interpreter in RPython.
 The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
-compiler}, a concept which will be explained in more detail in the next section.
+compiler}, a concept which we now explain in more details.
 
@@
 
-Tracing JITs are a recently more popular approach to write just-in-time
+Tracing JITs are a recently popular approach to write just-in-time
 compilers for dynamic languages \cite{XXX}. Their origins lie in the Dynamo
-project, that used a tracing approach to optimize machine code using execution
-traces \cite{XXX}. They were then adapted to be used for a very light-weight
+project, which used a tracing approach to optimize machine code using execution
+traces \cite{XXX}. Tracing JITs have then be adapted to be used for a very light-weight
 Java VM \cite{XXX} and afterwards used in several implementations of dynamic
@@
 \begin{figure}
+\begin{center}
 \includegraphics[scale=0.7]{figures/obj-lifetime.pdf}
+\end{center}
 

### 2010-09-28 talk/pepm2011/paper.tex
convert all traces to use math style.

@@
 \begin{figure}
-\begin{verbatim}
-    # arguments to the trace: p0, p1
-    # inside f: res.add(y)
-    guard_class(p1, BoxedInteger)
-        # inside BoxedInteger.add
-        i2 = getfield_gc(p1, intval)
-        guard_class(p0, BoxedInteger)
-            # inside BoxedInteger.add__int
-            i3 = getfield_gc(p0, intval)
-            i4 = int_add(i2, i3)
-            p5 = new(BoxedInteger)
-                # inside BoxedInteger.__init__
-                setfield_gc(p5, i4, intval)
-    # inside f: BoxedInteger(-100) 
-    p6 = new(BoxedInteger)
-        # inside BoxedInteger.__init__
-        setfield_gc(p6, -100, intval)
-
-    # inside f: .add(BoxedInteger(-100))
-    guard_class(p5, BoxedInteger)
-        # inside BoxedInteger.add
-        i7 = getfield_gc(p5, intval)
-        guard_class(p6, BoxedInteger)
-            # inside BoxedInteger.add__int
-            i8 = getfield_gc(p6, intval)
-            i9 = int_add(i7, i8)
-            p10 = new(BoxedInteger)
-                # inside BoxedInteger.__init__
-                setfield_gc(p10, i9, intval)
-
-    # inside f: BoxedInteger(-1)
-    p11 = new(BoxedInteger)
-        # inside BoxedInteger.__init__
-        setfield_gc(p11, -1, intval)
-
-    # inside f: y.add(BoxedInteger(-1))
-    guard_class(p0, BoxedInteger)
-        # inside BoxedInteger.add
-        i12 = getfield_gc(p0, intval)
-        guard_class(p11, BoxedInteger)
-            # inside BoxedInteger.add__int
-            i13 = getfield_gc(p11, intval)
-            i14 = int_add(i12, i13)
-            p15 = new(BoxedInteger)
-                # inside BoxedInteger.__init__
-                setfield_gc(p15, i14, intval)
-
-    # inside f: y.is_positive()
-    guard_class(p15, BoxedInteger)
-        # inside BoxedInteger.is_positive
-        i16 = getfield_gc(p15, intval)
-        i17 = int_gt(i16, 0)
-    # inside f
-    guard_true(i17)
-    jump(p15, p10)
-\end{verbatim}
+\texttt{
+\begin{tabular}{l} 
+\# arguments to the trace: $p_{0}$, $p_{1}$ \\
+\# inside f: res.add(y) \\
+guard\_class($p_{1}$, BoxedInteger) \\
+~~~~\# inside BoxedInteger.add \\
+~~~~$i_{2}$ = getfield($p_{1}$, intval) \\
+~~~~guard\_class($p_{0}$, BoxedInteger) \\
+~~~~~~~~\# inside BoxedInteger.add\_\_int \\
+~~~~~~~~$i_{3}$ = getfield($p_{0}$, intval) \\
+~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
+~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
+~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
+~~~~~~~~~~~~setfield($p_{5}$, $i_{4}$, intval) \\
+\# inside f: BoxedInteger(-100)  \\
+$p_{6}$ = new(BoxedInteger) \\
+~~~~\# inside BoxedInteger.\_\_init\_\_ \\
+~~~~setfield($p_{6}$, -100, intval) \\
+~\\
+\# inside f: .add(BoxedInteger(-100)) \\
+guard\_class($p_{5}$, BoxedInteger) \\
+~~~~\# inside BoxedInteger.add \\
+~~~~$i_{7}$ = getfield($p_{5}$, intval) \\
+~~~~guard\_class($p_{6}$, BoxedInteger) \\
+~~~~~~~~\# inside BoxedInteger.add\_\_int \\
+~~~~~~~~$i_{8}$ = getfield($p_{6}$, intval) \\
+~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
+~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
+~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
+~~~~~~~~~~~~setfield($p_{10}$, $i_{9}$, intval) \\
+~\\
+\# inside f: BoxedInteger(-1) \\
+$p_{11}$ = new(BoxedInteger) \\
+~~~~\# inside BoxedInteger.\_\_init\_\_ \\
+~~~~setfield($p_{11}$, -1, intval) \\
+~\\
+\# inside f: y.add(BoxedInteger(-1)) \\
+guard\_class($p_{0}$, BoxedInteger) \\
+~~~~\# inside BoxedInteger.add \\
+~~~~$i_{12}$ = getfield($p_{0}$, intval) \\
+~~~~guard\_class($p_{11}$, BoxedInteger) \\
+~~~~~~~~\# inside BoxedInteger.add\_\_int \\
+~~~~~~~~$i_{13}$ = getfield($p_{11}$, intval) \\
+~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
+~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
+~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
+~~~~~~~~~~~~setfield($p_{15}$, $i_{14}$, intval) \\
+~\\
+\# inside f: y.is\_positive() \\
+guard\_class($p_{15}$, BoxedInteger) \\
+~~~~\# inside BoxedInteger.is\_positive \\
+~~~~$i_{16}$ = getfield($p_{15}$, intval) \\
+~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
+\# inside f \\
+guard\_true($i_{17}$) \\
+jump($p_{15}$, $p_{10}$) \\
+\end{tabular}
+}
 \label{fig:unopt-trace}
@@
 operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
-looks at the number of \texttt{new}, \texttt{set/getfield\_gc} and
+looks at the number of \texttt{new}, \texttt{set/getfield} and
 \texttt{guard\_class} operations.
@@
 XXX simplify traces a bit more
-get rid of \_gc suffix in set/getfield\_gc
 
@@
 Figure~\ref{fig:unopt-trace} fall into categories 1 and 3. Objects stored in
-\texttt{p5, p6, p11 XXX} are in category 1, objects in \texttt{p10, p15} are in
+$p_{5}$, $p_{6}$, $p_{11}$ are in category 1, objects in $p_{10}$, $p_{15}$ are in
 category 3.
@@
 
-\begin{verbatim}
-p5 = new(BoxedInteger)
-setfield_gc(p5, i4, intval)
-p6 = new(BoxedInteger)
-setfield_gc(p6, -100, intval)
-\end{verbatim}
+\texttt{
+\begin{tabular}{l} 
+$p_{5}$ = new(BoxedInteger) \\
+setfield($p_{5}$, $i_{4}$, intval) \\
+$p_{6}$ = new(BoxedInteger) \\
+setfield($p_{6}$, -100, intval) \\
+\end{tabular}
+}
 
-The virtual object stored in \texttt{p5} would know that it is an \texttt{BoxedInteger}, and that
-the \texttt{intval} field contains \texttt{i4}, the one stored in \texttt{p6} would know that
+The virtual object stored in $p_{5}$ would know that it is an \texttt{BoxedInteger}, and that
+the \texttt{intval} field contains $i_{4}$, the one stored in $p_{6}$ would know that
 its \texttt{intval} field contains the constant -100.
 
-The following operations, that use \texttt{p5} and \texttt{p6} could then be
+The following operations, that use $p_{5}$ and $p_{6}$ could then be
 optimized using that knowledge:
 
-\begin{verbatim}
-guard_class(p5, BoxedInteger)
-i7 = getfield_gc(p5, intval)
-# inside BoxedInteger.add
-guard_class(p6, BoxedInteger)
-# inside BoxedInteger.add__int
-i8 = getfield_gc(p6, intval)
-i9 = int_add(i7, i8)
-\end{verbatim}
+\texttt{
+\begin{tabular}{l} 
+guard\_class($p_{5}$, BoxedInteger) \\
+$i_{7}$ = getfield($p_{5}$, intval) \\
+\# inside BoxedInteger.add \\
+guard\_class($p_{6}$, BoxedInteger) \\
+\# inside BoxedInteger.add\_\_int \\
+$i_{8}$ = getfield($p_{6}$, intval) \\
+$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
+\end{tabular}
+}
 
-The \texttt{guard\_class} operations can be removed, because the classes of \texttt{p5} and
-\texttt{p6} are known to be \texttt{BoxedInteger}. The \texttt{getfield\_gc} operations can be removed
-and \texttt{i7} and \texttt{i8} are just replaced by \texttt{i4} and -100. Thus the only
+The \texttt{guard\_class} operations can be removed, because the classes of $p_{5}$ and
+$p_{6}$ are known to be \texttt{BoxedInteger}. The \texttt{getfield} operations can be removed
+and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
 remaining operation in the optimized trace would be:
 
-\begin{verbatim}
-i9 = int_add(i4, -100)
-\end{verbatim}
+\texttt{
+\begin{tabular}{l} 
+$i_{9}$ = int\_add($i_{4}$, -100) \\
+\end{tabular}
+}
 
@@
 
-XXX should the variables be written in $math-style$ everywhere?
-
 \texttt{
@@
 $p_{15}$ = new(BoxedInteger) \\
-setfield\_gc($p_{15}$, $i_{14}$, intval) \\
+setfield($p_{15}$, $i_{14}$, intval) \\
 $p_{10}$ = new(BoxedInteger) \\
-setfield\_gc($p_{10}$, $i_{9}$, intval) \\
+setfield($p_{10}$, $i_{9}$, intval) \\
 jump($p_{15}$, $p_{10}$) \\
@@
 worthwhile even in this case, because some operations that have been performed
-on the forced virtual objects have been removed (some \texttt{getfield\_gc} operations
+on the forced virtual objects have been removed (some \texttt{getfield} operations
 and \texttt{guard\_class} operations).
@@
 
-The two new \texttt{BoxedIntegers} stored in \texttt{p15} and \texttt{p10} are passed into
+The two new \texttt{BoxedIntegers} stored in $p_{15}$ and $p_{10}$ are passed into
 the next iteration of the loop. The next iteration will check that they are

### 2010-09-28 talk/pepm2011/paper.tex
cite psyco
swap last two arguments in setfield

@@
 technique to achieve this is a form of escape analysis \cite{XXX} that we call
-\emph{virtual objects},\footnote{The terminology comes from \cite{Psyco}}
+\emph{virtual objects}\footnote{The terminology comes from \cite{rigo_representation-based_2004}},
 which is described in Section~\ref{sec:virtuals}. The
@@
 ~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~setfield($p_{5}$, $i_{4}$, intval) \\
+~~~~~~~~~~~~setfield($p_{5}$, intval, $i_{4}$) \\
 \# inside f: BoxedInteger(-100)  \\
@@
 ~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~setfield($p_{6}$, -100, intval) \\
+~~~~setfield($p_{6}$, intval, -100) \\
 ~\\
@@
 ~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~setfield($p_{10}$, $i_{9}$, intval) \\
+~~~~~~~~~~~~setfield($p_{10}$, intval, $i_{9}$) \\
 ~\\
@@
 ~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~setfield($p_{11}$, -1, intval) \\
+~~~~setfield($p_{11}$, intval, -1) \\
 ~\\
@@
 ~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~setfield($p_{15}$, $i_{14}$, intval) \\
+~~~~~~~~~~~~setfield($p_{15}$, intval, $i_{14}$) \\
 ~\\
@@
 $p_{5}$ = new(BoxedInteger) \\
-setfield($p_{5}$, $i_{4}$, intval) \\
+setfield($p_{5}$, intval, $i_{4}$) \\
 $p_{6}$ = new(BoxedInteger) \\
-setfield($p_{6}$, -100, intval) \\
+setfield($p_{6}$, intval, -100) \\
 \end{tabular}
@@
 $p_{15}$ = new(BoxedInteger) \\
-setfield($p_{15}$, $i_{14}$, intval) \\
+setfield($p_{15}$, intval, $i_{14}$) \\
 $p_{10}$ = new(BoxedInteger) \\
-setfield($p_{10}$, $i_{9}$, intval) \\
+setfield($p_{10}$, intval, $i_{9}$) \\
 jump($p_{15}$, $p_{10}$) \\

### 2010-09-28 talk/pepm2011/paper.tex
fix some XXXs

@@
 type-specializing the traces that are produced by the tracing JIT
-(Section~\ref{sec:crossloop}). In Section~\ref{sec:XXX} we describe some
+(Section~\ref{sec:crossloop}). In Section~\ref{sec:support} we describe some
 supporting techniques that are not central to the approach, but are needed to
@@
     jumps and then either escape or stop being used.\footnote{In theory, the
-    approach of Section~\ref{sec:XXX} works also for objects that live for
+    approach of Section~\ref{sec:crossloop} works also for objects that live for
     exactly $n>1$ iterations and then don't escape, but we expect this to be a
@@
 beginning to end and whenever a \texttt{new} operation is seen, the operation is
-removed and a virtual object\footnote{XXX what I have in mind when I talk
+removed and a virtual object\arigo{XXX what I have in mind when I talk
 of ``virtual object'' is the run-time behavior -- i.e. a real object that
@@
 
-XXX ``the trace above'' is dangerous; should mention all figures by numbers
-
-This is what happens at the end of the trace above, when the \texttt{jump} operation
+This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
 is hit. The arguments of the jump are at this point virtual objects. Before the
@@
 their lifetime crosses the jump. To improve the situation, a little trick is
-needed. The trace above represents a loop, i.e. the jump at the end jumps to
+needed. The trace in Figure~\ref{fig:step1} represents a loop, i.e. the jump at the end jumps to
 the beginning. Where in the loop the jump occurs is arbitrary, since the loop
@@
 
-If we do that for our example trace above, the trace looks like this:
+If we do that for our example trace, the trace looks like this:
 \begin{figure}
@@
 
+\section{Supporting Techniques}
+\label{sec:support}
+
+XXX virtualizables, store sinking
+
+do we need this?
+
+% section Supporting Techniques (end)
+
 \section{Evaluation}

### 2010-09-28 talk/pepm2011/paper.tex
that figures

@@
 
-If we do that for our example trace, the trace looks like this:
+If we do that for our example trace, the trace looks like in Figure~\ref{fig:step2}.
+
 \begin{figure}
 \includegraphics{figures/step2.pdf}
+\label{fig:step2}
+\caption{Shifting the Jump}
 \end{figure}
 
-XXX the figure is moved elsewhere by latex
+Now the lifetime of the remaining allocations no longer crosses the jump, and
+we can run our escape analysis a second time, to get the trace in
+Figure~\ref{fig:step3}.
 
-Now the lifetime of the remaining allocations no longer crosses the jump, and
-we can run our escape analysis a second time, to get the following trace:
 \begin{figure}
 \includegraphics{figures/step3.pdf}
+\label{fig:step3}
+\caption{Removing Allocations a Second Time}
 \end{figure}
-
-XXX the figure is moved elsewhere by latex
 
@@
 additional code corresponds to one iteration of the loop, which is thus
-peeled off \cite{XXX}:
+peeled off \cite{XXX}, see Figure~\ref{fig:step3}.
 
@@
 \includegraphics{figures/step4.pdf}
+\label{fig:step3}
+\caption{A Way to Enter the Loop From the Interpreter}
 \end{figure}
-
-XXX the figure is moved elsewhere by latex
 

### 2010-09-28 talk/pepm2011/paper.tex
add two paragraphs about virtualizables, I guess they are cuttable

@@
 
-XXX virtualizables, store sinking
+\subsection{Virtualizables}
+\label{sub:Virtualizables}
+
+\cfbolz{probably can be cut in case of space problems}
+
+One problem to the successful application of the allocation removal techniques
+described in the previous sections is the presence of frame-introspection
+features in many dynamic languages. Languages such as Python and Smalltalk allow
+the programmer to get access to the frames object that the interpreter uses to
+store local variables. This is a useful feature, as makes the implementation of
+a debugger possible in Python without needing much support from the VM level. On
+the other hand, it severely hinders the effectiveness of allocation removal,
+because every time an object is stored into a local variable, it is stored into
+the frame-object, which makes it escape.
+
+This problem is solved by making it possible to the interpreter author to add
+some hints into the source code to declare instances of one class as frame
+objects. The JIT will then fill these objects only lazily when they are actually
+accessed (\eg because a debugger is used). Therefore in the common case, nothing
+is stored into the frame objects, making the problem of too much escaping go
+away. This is a common approach in VM implementations \cite{XXX}, the only
+novelty in our approach lays in its generality, because most other JITs are just
+specifically written for one particular language.
+
+% subsection Virtualizables (end)
+XXX store sinking
 

### 2010-09-28 talk/pepm2011/paper.tex
fix an XXX

@@
 from the virtual object, and the operation is also removed.
-
-XXX what happens on a guard\_class?
+Equivalently, a \texttt{guard\_class} on a virtual object can be removed as
+well, because the virtual object has a fixed and known class.
 
@@
 So far we have only described what happens when virtual objects are used in
-operations that read and write their fields. When the virtual object is used in
+operations that read and write their fields and in guards. When the virtual
+object is used in
 any other operation, it cannot stay virtual. For example, when a virtual object
 is stored in a globally accessible place, the object needs to actually be
-allocated, as it will live longer than one iteration of the loop.
+allocated, as it might live longer than one iteration of the loop.
 

### 2010-09-28 talk/pepm2011/paper.tex
list of benchmarks I want to use

@@
 
+Benchmarks from the Computer Language Benchmark Game are: fannkuch, nbody,
+meteor-contest, spectral-norm.
+
+\textbf{crypto\_pyaes}: AES implementation.
+\textbf{django}: The templating engine of the Django web
+framework\footnote{\texttt{http://www.djangoproject.com/}}.
+\textbf{go}: A Monte-Carlo Go
+AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/disco-elegant-python-go-player.html}}.
+\textbf{html5lib}: HTML5 parser
+\textbf{pyflate-fast}: BZ2 decoder
+\textbf{raytrace-simple}: ray tracer
+\textbf{richards}: The Richards benchmark \cite{XXX}
+\textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
+\textbf{telco}: A Python version of the Telco decimal
+benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a pure
+Python decimal floating point implementation.
+\textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
+framework\footnote{\texttt{http://twistedmatrix.com/trac/}}.
+
 \section{Related Work}

### 2010-09-28 talk/pepm2011/paper.tex
add some references

@@
 
-Tracing JITs are a recently popular approach to write just-in-time
-compilers for dynamic languages \cite{XXX}. Their origins lie in the Dynamo
-project, which used a tracing approach to optimize machine code using execution
-traces \cite{XXX}. Tracing JITs have then be adapted to be used for a very light-weight
-Java VM \cite{XXX} and afterwards used in several implementations of dynamic
-languages, such as JavaScript \cite{XXX}, Lua \cite{XXX} and now Python via
-PyPy.
+Tracing JITs are a recently popular approach to write just-in-time compilers for
+dynamic languages. Their origins lie in the Dynamo project, which used a tracing
+approach to optimize machine code using execution traces
+\cite{bala_dynamo:_2000}. Tracing JITs have then be adapted to be used for a
+very light-weight Java VM \cite{gal_hotpathvm:_2006} and afterwards used in
+several implementations of dynamic languages, such as JavaScript
+\cite{andreas_gal_trace-based_2009}, Lua \cite{XXX} and now Python (and other
+languages) via PyPy.
 

### 2010-09-28 talk/pepm2011/paper.tex
minimally fix an XXX

@@
 
-XXX object model and its reflection in traces (e.g. guard\_class before each method call)
-
 \subsection{Running Example}
@@
 operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
-looks at the number of \texttt{new}, \texttt{set/getfield} and
-\texttt{guard\_class} operations.
+looks at the number of \texttt{new} (corresponding to object creation),
+\texttt{set/getfield} (corresponding to attribute reads/writes) and
+\texttt{guard\_class} operations (corresponding to method calls).
 

### 2010-09-28 talk/pepm2011/paper.tex
minimally describe bridges
email addresses, no clue how to fit affiliations

@@
            {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}
-           {cfbolz@gmx.de XXX}
+           {cfbolz@gmx.de, antocuni@gmail.com, fijal@merlinux.eu,
+           samuele.pedroni@gmail.com, arigo@tunes.org}
 
@@
 Boxing of primitive types means that dynamic languages need to be able to handle
-all objects, even integers, floats, bools etc. in the same way as user-defined
+all objects, even integers, floats, booleans etc. in the same way as user-defined
 instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
@@
 
-bridges?
-
-arguments to traces
-
-getting from the interpreter to traces
+If one specific guard fails often enough, the tracing JIT will generate a new
+trace that starts exactly at the position of the failing guard. The existing
+assembler is patched to jump to the new trace when the guard fails
+\cite{andreas_gal_incremental_2006}.
 

### 2010-09-28 talk/pepm2011/paper.tex
fix my email address

@@
            {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}
-           {cfbolz@gmx.de, antocuni@gmail.com, fijal@merlinux.eu,
+           {cfbolz@gmx.de, anto.cuni@gmail.com, fijal@merlinux.eu,
            samuele.pedroni@gmail.com, arigo@tunes.org}

### 2010-10-06 talk/pepm2011/paper.tex
change title, add michael as an author, refactor some small things

--- 2010-09-28 talk/pepm2011/paper.tex
@@
 
-\title{Escape Analysis and Specialization in a Tracing JIT}
+\title{Allocation Removal by Partial Evaluation in a Tracing JIT}
 
-\authorinfo{Carl Friedrich Bolz \and Antonio Cuni \and Maciej Fijałkowski \and Samuele Pedroni \and Armin Rigo}
+\authorinfo{Carl Friedrich Bolz \and Antonio Cuni \and Maciej Fijałkowski \and Michael Leuschel \and Samuele Pedroni \and Armin Rigo}
            {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}
-           {cfbolz@gmx.de, antocuni@gmail.com, fijal@merlinux.eu,
+           {cfbolz@gmx.de, anto.cuni@gmail.com, fijal@merlinux.eu,
            samuele.pedroni@gmail.com, arigo@tunes.org}
@@
 
-\section{Escape Analysis in a Tracing JIT}
+\section{Escape Analysis in a Tracing JIT} % XXX imprecise title
 \label{sec:virtuals}
@@
 beginning to end and whenever a \texttt{new} operation is seen, the operation is
-removed and a virtual object\arigo{XXX what I have in mind when I talk
-of ``virtual object'' is the run-time behavior -- i.e. a real object that
-would exist at run-time, except that it has be virtual-ized.  Here you seem
-to mean rather ``virtual object description'' or something.}
-is constructed. The virtual object summarizes the
-shape of the object that is allocated at this position in the original trace,
-and is used by the optimization to improve the trace. The shapes describe
+removed and a shape description is constructed and associated with the variable
+that would have stored the result of \texttt{new}. This result, a newly
+allocated object, is called \emph{virtual}, because after the optimization is
+finished, it does not need to be allocated at all. The shape description summarizes
+the shape of this original object
+and is used by the optimization to improve the trace. The shape describes
 where the values that would be stored in the fields of the allocated objects
-come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
-object, that shape summary is thus updated and the operation can be removed.
-When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
-from the virtual object, and the operation is also removed.
-Equivalently, a \texttt{guard\_class} on a virtual object can be removed as
-well, because the virtual object has a fixed and known class.
+come from. Whenever the optimizer sees a \texttt{setfield} that writes into such
+an object, that shape description is updated and the operation can be removed.
+When the optimizer encounters a \texttt{getfield} from such an object, the result is read
+from the shape description, and the operation is also removed.
+Equivalently, a \texttt{guard\_class} on a variable that has a shape description can be removed as
+well, because the shape description stores the class.
 
 In the example from last section, the following operations would produce two
-virtual objects, and be completely removed from the optimized trace:
+shape descriptions, and be completely removed from the optimized trace:
 
@@
 
-The virtual object stored in $p_{5}$ would know that it is an \texttt{BoxedInteger}, and that
-the \texttt{intval} field contains $i_{4}$, the one stored in $p_{6}$ would know that
-its \texttt{intval} field contains the constant -100.
+The shape description associated with $p_{5}$ would know that it is an
+\texttt{BoxedInteger}, and that the \texttt{intval} field contains $i_{4}$, the
+associated with $p_{6}$ would know that its \texttt{intval} field contains the
+constant -100.
 
@@
 is stored in a globally accessible place, the object needs to actually be
-allocated, as it might live longer than one iteration of the loop.
+allocated, as it might live longer than one iteration of the loop. The actual
+allocation is inserted just before the operation that requires the object to be
+non-virtual.
 
@@
 
-\subsection{Algorithm}
-\label{sub:Algorithm}
+\section{Formal Description of the Algorithm}
+\label{sec:Algorithm}
 
@@
 
-% subsection Algorithm (end)
+% section Formal Description of the Algorithm (end)
 

### 2010-10-06 talk/pepm2011/paper.tex
first go at integrating the operational semantics. not completely happy.

@@
 \begin{center}
-\includegraphics[scale=0.7]{figures/obj-lifetime.pdf}
+\includegraphics[scale=0.6]{figures/obj-lifetime.pdf}
 \end{center}
@@
 
-XXX want some sort of pseudo-code
+\begin{figure*}
+\begin{center}
+\begin{tabular}{lcccc}
+\emph{new} & ${\displaystyle \frac{l\,\mathrm{fresh}}{v=\mathrm{new}(T),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto l\right],H\left[l\mapsto\left(T,\mathrm{null},\mathrm{null}\right)\right]}}$ & ~~~ &
+\emph{guard} & ${\displaystyle \frac{\mathrm{type}(H(E(v))=T}{\mathrm{guard}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H}}$\tabularnewline[3em]
+\emph{get} & ${\displaystyle \frac{\,}{v=\mathrm{get}(u,F),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto H\left(E\left(u\right)\right)_{F}\right],H}}$ & ~~~ &
+& ${\displaystyle \frac{\mathrm{type}(H(E(v))\neq T}{\mathrm{guard}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}\bot,\bot}}$\tabularnewline[3em]
+\emph{set} & ${\displaystyle \frac{\,}{\mathrm{set}\left(v,F,u\right),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H\left[E\left(v\right)\mapsto\left(H\left(E\left(v\right)\right)!_{F}E(u)\right)\right]}}$ & ~~~ &
+& \tabularnewline[4em]
+\end{tabular}
+
+\begin{minipage}[b]{7 cm}
+\emph{Object Domains:}
+$$\setlength\arraycolsep{0.1em}
+ \begin{array}{rcll}
+    u,v,w & \in & V & \mathrm{\ variables\ in\ trace}\\
+    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
+    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
+    l & \in & L & \mathrm{\ locations\ on\ heap}
+ \end{array}
+$$
+\end{minipage}
+\begin{minipage}[b]{5 cm}
+\emph{Semantic Values:}
+$$\setlength\arraycolsep{0.1em}
+ \begin{array}{rcll}
+    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
+    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
+    \\
+    \\
+ \end{array}
+$$
+\end{minipage}
+\end{center}
+\label{fig:semantics}
+\caption{The Operational Semantics of Simplified Traces}
+\end{figure*}
+
+In this section we want to give a formal description of the semantics of the
+traces and of the optimizer and liken the optimization to partial evaluation.
+We concentrate on the operations for manipulating dynamically allocated objects,
+as those are the only ones that are actually optimized. Without loss of
+generality we also consider only objects with two fields in this section.
+
+Traces are lists of operations. The operations considered here are new (to make
+a new object), get (to read a field out of an object), set (to write a field
+into an object) and guard (to check the type of an object). The values of all
+variables are locations (i.e.~pointers). Locations are mapped to objects, which
+are represented by triples of a type $T$, and two locations that represent the
+fields of the object. When a new object is created, the fields are initialized
+to null, but we require that they are immediately initialized to a real
+location, otherwise the trace is malformed.
+
+We use some abbreviations when dealing with object triples. To read the type of
+an object, $\mathrm{type}((T,l_1,l_2))=T$ is used. Reading a field $F$ from an
+object is written $(T,l_1,l_2)_F$ which either returns $l_1$ if $F=L$ or $l_2$
+if $F=R$. To set field $F$ to a new location $l$, we use the notation
+$(T,l_1,l_2)!_Fl$, which yields a new triple $(T,l,l_2)$ if $F=L$ or a new
+triple $(T,l_1,l)$ if $F=R$.
+
+Figure~\ref{fig:semantics} shows the operational semantics for traces. The
+interpreter formalized there executes one operation at a time. Its state is
+represented by an environment and a heap, which are potentially changed by the
+execution of an operation. The environment is a partial function from variables
+to locations and the heap is a partial function from locations to objects. Note
+that a variable can never be null in the environment, otherwise the trace would
+be malformed. We use some syntax for changing partial functions: $E[v\mapsto l]$
+denotes the environment which is just like $E$, but maps $v$ to $l$.
+
+The new operation creates a new object $(T,\mathrm{null},\mathrm{null})$, on the
+heap under a fresh location $l$ and adds the result variable to the environment,
+mapping to the new location $l$.
+
+The get operation reads a field $F$ out of an object and adds the result
+variable to the environment, mapping to the read location. The heap is
+unchanged.
+
+The set operation changes field $F$ of an object stored at the location that
+variable $v$ maps to. The new value of the field is the location in variable
+$u$. The environment is unchanged.
+
+The guard operation is used to check whether the object stored at the location
+that variable $v$ maps to is of type $T$. If that is the case, then execution
+continues without changing heap and environment. Otherwise, execution is
+stopped.
+
+\subsection{Optimizing Traces}
+\label{sub:formalopt}
+
+% subsection Optimizing Traces (end)
+
+\begin{figure*}
+\begin{center}
+\fbox{
+\begin{tabular}{lc}
+\emph{new} & ${\displaystyle \frac{v^{*}\,\mathrm{fresh}}{v=\mathrm{new}(T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto v^{*}\right],S\left[v^{*}\mapsto\left(T,\mathrm{null,null}\right)\right]}}$\tabularnewline[3em]
+\emph{get} & ${\displaystyle \frac{E(u)\in\mathrm{dom}(S)}{v=\mathrm{get}(u,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto S(E(u))_{F}\right],S}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(u)\notin\mathrm{dom}(S)\, v^{*}\,\mathrm{fresh}}{v=\mathrm{get}(u,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle v^{*}=\mathrm{get}(E(u),F)\right\rangle ,E\left[v\mapsto v^{*}\right],S}}$\tabularnewline[3em]
+\emph{set} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{\mathrm{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S\left[E\left(v\right)\mapsto\left(S(E(v))!_{F}E(u)\right)\right]}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\notin\mathrm{dom}\left(S\right),\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathrm{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathrm{set}\left(E(v),F,E(u)\right)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
+\emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathrm{guard}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathrm{guard}(E\left(v\right),T)\right\rangle ,E,S}}$\tabularnewline[3em]
+\emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{liftfields}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathrm{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathrm{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathrm{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline
+\end{tabular}
+}
+\end{center}
+\label{fig:optimization}
+\caption{Optimization Rules}
+\end{figure*}
 

### 2010-10-06 talk/pepm2011/paper.tex
some beautiful and rectangular but not very yellow text about the formal description of the optimization

@@
 
+XXX think of a way to format the operations to make them look distinct from
+normal text
+
 In this section we want to give a formal description of the semantics of the
@@
 that a variable can never be null in the environment, otherwise the trace would
-be malformed. We use some syntax for changing partial functions: $E[v\mapsto l]$
-denotes the environment which is just like $E$, but maps $v$ to $l$.
+be malformed. The environment could not directly map variables to object,
+because several variables can contain a pointer to the \emph{same} object. Thus
+the "indirection" is needed to express sharing.
+
+We use the following notation for updating partial functions:
+$E[v\mapsto l]$ denotes the environment which is just like $E$, but maps $v$ to
+$l$.
 
@@
 
+To optimize the simple traces from the last section, we use online partial
+evaluation. The partial evaluator optimizes one operation of the trace at a
+time. Every operation in the unoptimized trace is replaced by a list of
+operations in the optimized trace. This list is empty if the operation
+can be optimized away (which hopefully happens often).
+
+The state of the optimizer is stored in an environment $E$ and a \emph{static
+heap} $S$. The environment is a partial function from variables in the
+unoptimized trace to variables in the optimized trace (which are written with a
+$\ ^*$ for clarity). The reason for introducing new variables in the optimized
+trace is that several variables that appear in the unoptimized trace can turn
+into the same variables in the optimized trace. Thus the environment of the
+optimizer serves a function similar to that of the environment in the semantics.
+
+The static heap is a partial function from $V^*$ into the
+set of static objects, which are triples of a type, and two elements of $V^*$.
+An variable $v^*$ is in the domain of the static heap $S$ as long as the
+optimizer can fully keep track of the object. The image of $v^*$ is what is
+statically known about the object stored in it, \ie its type and its fields. The
+fields of objects in the static heap are also elements of $V^*$ (or null, for
+short periods of time).
+
+When the optimizer sees a new operation, it optimistically removes it and
+assumes that the resulting object can stay static. The optimization for all
+further operations is split into two cases. One case is for when the
+involved variables are in the static heap, which means that the operation can be
+removed. The other case is that nothing is known about the variables, which
+means the operation has to be residualized.
+
+If the argument $u$ of a get operation is mapped to something in the static
+heap, the get can be performed at optimization time. Otherwise, the get
+operation needs to be emitted.
+
+If the first argument $v$ to a set operation is mapped to something in the
+static heap, then the set can performed at optimization time and the static heap
+is updated. Otherwise the set operation needs to be emitted. This needs to be
+done carefully, because the new value for the field stored in the variable $u$
+could itself be static, in which case it needs to be lifted first.
+
+I a guard is performed on a variable that is in the static heap, the type check
+can be performed at optimization time, which means the operation can be removed
+if the types match. If the type check fails statically or if the object is not
+in the static heap, the guard is put into the residual trace. This also needs to
+lift the variable on which the guard is performed.
+
+Lifting takes a variable that is potentially in the static heap and makes sure
+that it is turned into a dynamic variable. This means that operations are
+emitted that construct an object that looks like the shape described in the
+static heap, and the variable is removed from the static heap.
+
+Lifting a static object needs to recursively lift its fields. Some care needs to
+be taken when lifting a static object, because the structures described by the
+static heap can be cyclic. To make sure that the same static object is not lifted
+twice, the liftfield operation removes it from the static heap \emph{before}
+recursively lifting its fields.
+
 % subsection Optimizing Traces (end)
@@
 \begin{center}
-\fbox{
 \begin{tabular}{lc}
@@
  & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{liftfields}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathrm{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
- & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathrm{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathrm{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline
+ & ${\displaystyle
+ \frac{\left(S\left(v^{*}\right)_{L},S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle
+ \mathrm{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathrm{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline[3em]
 \end{tabular}
-}
+
+\begin{minipage}[b]{7 cm}
+\emph{Object Domains:}
+$$\setlength\arraycolsep{0.1em}
+ \begin{array}{rcll}
+    u,v,w & \in & V & \mathrm{\ variables\ in\ trace}\\
+    u^*,v^*,w^* & \in & V^* & \mathrm{\ variables\ in\ optimized\ trace}\\
+    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
+    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
+ \end{array}
+$$
+\end{minipage}
+\begin{minipage}[b]{5 cm}
+\emph{Semantic Values:}
+$$\setlength\arraycolsep{0.1em}
+ \begin{array}{rcll}
+    E & \in & V\rightharpoonup V^* & \mathrm{\ Environment}\\
+    S & \in & V^*\rightharpoonup\mathfrak{T}\times(V^*\cup \{ \mathrm{null} \})\times (V^*\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
+    \\
+    \\
+ \end{array}
+$$
+\end{minipage}
 \end{center}

### 2010-10-06 talk/pepm2011/paper.tex
re-copy formulas from lyx file

@@
 \emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathrm{guard}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathrm{guard}(E\left(v\right),T)\right\rangle ,E,S}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\vee\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathrm{guard}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
 \emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{liftfields}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathrm{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
- & ${\displaystyle
- \frac{\left(S\left(v^{*}\right)_{L},S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle
- \mathrm{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathrm{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathrm{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathrm{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathrm{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline[3em]
 \end{tabular}

### 2010-10-06 talk/pepm2011/paper.tex
shuffle stuff around
mention similarity to execution semantics

@@
 
-XXX is "symbolic execution" the right word to drop?
+XXX rewrite to acknowledge the existence of the formal stuff
 
@@
 
-To optimize the simple traces from the last section, we use online partial
-evaluation. The partial evaluator optimizes one operation of the trace at a
-time. Every operation in the unoptimized trace is replaced by a list of
-operations in the optimized trace. This list is empty if the operation
-can be optimized away (which hopefully happens often).
-
-The state of the optimizer is stored in an environment $E$ and a \emph{static
-heap} $S$. The environment is a partial function from variables in the
-unoptimized trace to variables in the optimized trace (which are written with a
-$\ ^*$ for clarity). The reason for introducing new variables in the optimized
-trace is that several variables that appear in the unoptimized trace can turn
-into the same variables in the optimized trace. Thus the environment of the
-optimizer serves a function similar to that of the environment in the semantics.
-
-The static heap is a partial function from $V^*$ into the
-set of static objects, which are triples of a type, and two elements of $V^*$.
-An variable $v^*$ is in the domain of the static heap $S$ as long as the
-optimizer can fully keep track of the object. The image of $v^*$ is what is
-statically known about the object stored in it, \ie its type and its fields. The
-fields of objects in the static heap are also elements of $V^*$ (or null, for
-short periods of time).
-
-When the optimizer sees a new operation, it optimistically removes it and
-assumes that the resulting object can stay static. The optimization for all
-further operations is split into two cases. One case is for when the
-involved variables are in the static heap, which means that the operation can be
-removed. The other case is that nothing is known about the variables, which
-means the operation has to be residualized.
-
-If the argument $u$ of a get operation is mapped to something in the static
-heap, the get can be performed at optimization time. Otherwise, the get
-operation needs to be emitted.
-
-If the first argument $v$ to a set operation is mapped to something in the
-static heap, then the set can performed at optimization time and the static heap
-is updated. Otherwise the set operation needs to be emitted. This needs to be
-done carefully, because the new value for the field stored in the variable $u$
-could itself be static, in which case it needs to be lifted first.
-
-I a guard is performed on a variable that is in the static heap, the type check
-can be performed at optimization time, which means the operation can be removed
-if the types match. If the type check fails statically or if the object is not
-in the static heap, the guard is put into the residual trace. This also needs to
-lift the variable on which the guard is performed.
-
-Lifting takes a variable that is potentially in the static heap and makes sure
-that it is turned into a dynamic variable. This means that operations are
-emitted that construct an object that looks like the shape described in the
-static heap, and the variable is removed from the static heap.
-
-Lifting a static object needs to recursively lift its fields. Some care needs to
-be taken when lifting a static object, because the structures described by the
-static heap can be cyclic. To make sure that the same static object is not lifted
-twice, the liftfield operation removes it from the static heap \emph{before}
-recursively lifting its fields.
-
-% subsection Optimizing Traces (end)
-
 \begin{figure*}
@@
 
+To optimize the simple traces from the last section, we use online partial
+evaluation. The partial evaluator optimizes one operation of the trace at a
+time. Every operation in the unoptimized trace is replaced by a list of
+operations in the optimized trace. This list is empty if the operation
+can be optimized away (which hopefully happens often).
+
+The state of the optimizer is stored in an environment $E$ and a \emph{static
+heap} $S$. The environment is a partial function from variables in the
+unoptimized trace to variables in the optimized trace (which are written with a
+$\ ^*$ for clarity). The reason for introducing new variables in the optimized
+trace is that several variables that appear in the unoptimized trace can turn
+into the same variables in the optimized trace. Thus the environment of the
+optimizer serves a function similar to that of the environment in the semantics.
+
+The static heap is a partial function from $V^*$ into the
+set of static objects, which are triples of a type, and two elements of $V^*$.
+An variable $v^*$ is in the domain of the static heap $S$ as long as the
+optimizer can fully keep track of the object. The image of $v^*$ is what is
+statically known about the object stored in it, \ie its type and its fields. The
+fields of objects in the static heap are also elements of $V^*$ (or null, for
+short periods of time).
+
+When the optimizer sees a new operation, it optimistically removes it and
+assumes that the resulting object can stay static. The optimization for all
+further operations is split into two cases. One case is for when the
+involved variables are in the static heap, which means that the operation can be
+performed at optimization time and removed from the trace. These rules mirror
+the execution semantics closely. The other case is that nothing is known about
+the variables, which means the operation has to be residualized.
+
+If the argument $u$ of a get operation is mapped to something in the static
+heap, the get can be performed at optimization time. Otherwise, the get
+operation needs to be emitted.
+
+If the first argument $v$ to a set operation is mapped to something in the
+static heap, then the set can performed at optimization time and the static heap
+is updated. Otherwise the set operation needs to be emitted. This needs to be
+done carefully, because the new value for the field stored in the variable $u$
+could itself be static, in which case it needs to be lifted first.
+
+I a guard is performed on a variable that is in the static heap, the type check
+can be performed at optimization time, which means the operation can be removed
+if the types match. If the type check fails statically or if the object is not
+in the static heap, the guard is put into the residual trace. This also needs to
+lift the variable on which the guard is performed.
+
+Lifting takes a variable that is potentially in the static heap and makes sure
+that it is turned into a dynamic variable. This means that operations are
+emitted that construct an object that looks like the shape described in the
+static heap, and the variable is removed from the static heap.
+
+Lifting a static object needs to recursively lift its fields. Some care needs to
+be taken when lifting a static object, because the structures described by the
+static heap can be cyclic. To make sure that the same static object is not lifted
+twice, the liftfield operation removes it from the static heap \emph{before}
+recursively lifting its fields.
+
+% subsection Optimizing Traces (end)
+
+\subsection{Analysis of the Algorithm}
+\label{sub:analysis}
+
+XXX algorithm is linear in the length of the trace
+
+% subsection Analysis of the Algorithm (end)
+
 % section Formal Description of the Algorithm (end)

### 2010-10-07 talk/pepm2011/paper.tex
rewrite the introduction to mention partial evaluation more

@@
 
-Boxing of primitive types means that dynamic languages need to be able to handle
+Boxing of primitive types is necessary because dynamic languages need to be able to handle
 all objects, even integers, floats, booleans etc. in the same way as user-defined
@@
 A recently popular approach to implementing just-in-time compilers for dynamic
-languages is that of a tracing JIT. A tracing JIT often takes the form of an
-extension to an existing interpreter, which can be sped up that way. This
-approach is also the one taken by the PyPy project, which is an environment for
-implementing dynamic programming languages. PyPy's
-approach to doing so is to straightforwardly implement an interpreter for the
-to-be-implemented language, and then use powerful tools to turn the interpreter
-into an efficient virtual machine (VM) that also contains a just-in-time compiler.
-This compiler
-is automatically generated from the interpreter using partial-evaluation-like
-techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
-tracing JIT compilers is described in Section~\ref{sec:Background}.
-
-The tracing JIT approach that the PyPy project is taking removes the overhead
-of bytecode dispatch. In this paper we want to explain how the traces that are
-produced by PyPy's tracing JIT can be optimized to also remove some of the
+languages is that of a tracing JIT. A tracing JIT works by observing the running
+program and recording linear execution traces, which are then turned into
+machine code. One reason for the popularity of tracing JITs is their relative
+simplicity. They can often be added to an interpreter and a lot of the
+infrastructure of an interpreter can be reused. They give some important
+optimizations like inlining and constant-folding for free. A tracing JIT always
+produces linear pieces of code, which makes many optimizations that are usually
+hard in a compiler simpler, such as register allocation.
+
+The usage of a tracing JIT can remove the overhead of bytecode dispatch and that
+of the interpreter data structures. In this paper we want to present an approach
+how an optimization can be added to a tracing JIT that also removes some of the
 overhead more closely associated to dynamic languages, such as boxing overhead
-and type dispatching. To understand the problem more closely, we analyze the
-occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
-technique to achieve this is a form of escape analysis \cite{XXX} that we call
-\emph{virtual objects}\footnote{The terminology comes from \cite{rigo_representation-based_2004}},
-which is described in Section~\ref{sec:virtuals}. The
-goal of virtual objects is to remove allocations of temporary objects that have
-a predictable lifetime and to optimize type dispatching in the process.
-
-The basic approach of virtual objects can then be extended to also be used for
+and type dispatching. Our experimental platform is the PyPy project, which is an
+environment for implementing dynamic programming languages. PyPy and tracing
+JITs are described in more detail in Section~\ref{sec:Background}.
+Section~\ref{sec:lifetimes} analyzes the problem to be solved more closely.
+
+The most important technique we use to achieve to optimize traces is a
+straightforward application of partial evaluation. The partial evaluation
+performs a form of escape analysis \cite{XXX} on the traces and make some
+objects that are allocated in the trace \emph{static}\footnote{These objects are
+called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
+means that they do not occur any more in the optimized trace. This technique is
+informally described in Section~\ref{sec:virtuals}, a more formal description is
+given in Section~\ref{sec:formal}.
+
+The basic approach of static objects can then be extended to also be used for
 type-specializing the traces that are produced by the tracing JIT
@@
     \item An efficient and effective algorithm for removing object allocations in a tracing JIT.
-    \item XXX
+    \item A characterization of this algorithm as partial evaluation.
+    \item A rigorous evaluation of this algorithm.
 \end{enumerate}
@@
 \section{Formal Description of the Algorithm}
-\label{sec:Algorithm}
+\label{sec:formal}
 

### 2010-10-07 talk/pepm2011/paper.tex
rewrite to speak of "static objects" everywhere

@@
 means that they do not occur any more in the optimized trace. This technique is
-informally described in Section~\ref{sec:virtuals}, a more formal description is
+informally described in Section~\ref{sec:statics}, a more formal description is
 given in Section~\ref{sec:formal}.
@@
 
-In the next section, we will see how this can be improved upon, using escape
-analysis. XXX
+In the next section, we will see how this can be improved upon, using partial
+evaluation. XXX
 
@@
 The creation of objects in category 1 is removed by the optimization described
-in Section~\ref{sec:virtuals}. XXX
-
-\section{Escape Analysis in a Tracing JIT} % XXX imprecise title
-\label{sec:virtuals}
-
-\subsection{Virtual Objects}
+in Section~\ref{sec:statics}. XXX
+
+\section{Allocation Removal in Traces}
+\label{sec:statics}
+
+\subsection{Static Objects}
 
@@
 
-XXX rewrite to acknowledge the existence of the formal stuff
-
-This process is called \emph{escape analysis}. The escape analysis of
-our tracing JIT works by using \emph{virtual objects}: The trace is walked from
-beginning to end and whenever a \texttt{new} operation is seen, the operation is
-removed and a shape description is constructed and associated with the variable
-that would have stored the result of \texttt{new}. This result, a newly
-allocated object, is called \emph{virtual}, because after the optimization is
-finished, it does not need to be allocated at all. The shape description summarizes
-the shape of this original object
-and is used by the optimization to improve the trace. The shape describes
-where the values that would be stored in the fields of the allocated objects
-come from. Whenever the optimizer sees a \texttt{setfield} that writes into such
-an object, that shape description is updated and the operation can be removed.
-When the optimizer encounters a \texttt{getfield} from such an object, the result is read
-from the shape description, and the operation is also removed.
-Equivalently, a \texttt{guard\_class} on a variable that has a shape description can be removed as
-well, because the shape description stores the class.
+This is a process that is usually called \emph{escape analysis}. In this paper we will
+perform escape analysis by using partial evaluation. The partial evalution is a
+bit peculiar in that there are not actually any constant arguments to the trace,
+but it is only used to optimized operations within a trace. XXX mention Prolog.
+
+The partial evaluation works by walking the trace from beginning to end.
+Whenever a \texttt{new} operation is seen, the operation is removed and a static
+object is constructed and associated with the variable that would have stored
+the result of \texttt{new}. The static object describes the shape of the
+original object, \eg where the values that would be stored in the fields of the
+allocated object come from, as well as the type of the object. Whenever the
+optimizer sees a \texttt{setfield} that writes into such an object, that shape
+description is updated and the operation can be removed, which means that the
+operation was done at partial evaluation time. When the optimizer encounters a
+\texttt{getfield} from such an object, the result is read from the shape
+description, and the operation is also removed. Equivalently, a
+\texttt{guard\_class} on a variable that has a shape description can be removed
+as well, because the shape description stores the type.
 
 In the example from last section, the following operations would produce two
-shape descriptions, and be completely removed from the optimized trace:
+static objects, and be completely removed from the optimized trace:
 
@@
 
-The shape description associated with $p_{5}$ would know that it is an
+The static object associated with $p_{5}$ would know that it is a
 \texttt{BoxedInteger}, and that the \texttt{intval} field contains $i_{4}$, the
-associated with $p_{6}$ would know that its \texttt{intval} field contains the
-constant -100.
+one associated with $p_{6}$ would know that its \texttt{intval} field contains
+the constant -100.
 
@@
 
-So far we have only described what happens when virtual objects are used in
-operations that read and write their fields and in guards. When the virtual
-object is used in
-any other operation, it cannot stay virtual. For example, when a virtual object
-is stored in a globally accessible place, the object needs to actually be
-allocated, as it might live longer than one iteration of the loop. The actual
-allocation is inserted just before the operation that requires the object to be
-non-virtual.
+So far we have only described what happens when static objects are used in
+operations that read and write their fields and in guards. When the static
+object is used in any other operation, it cannot stay static. For example, when
+a static object is stored in a globally accessible place, the object needs to
+actually be allocated, as it might live longer than one iteration of the loop
+and because the partial evaluator looses track of it. This means that the static
+objects needs to be turned into a dynamic one, \ie lifted. This makes it
+necessary to put operations into the residual code that actually allocate the
+static object at runtime.
 
 This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
-is hit. The arguments of the jump are at this point virtual objects. Before the
-jump is emitted, they are \emph{forced}. This means that the optimizers produces code
+is hit. The arguments of the jump are at this point static objects. Before the
+jump is emitted, they are \emph{lifted}. This means that the optimizers produces code
 that allocates a new object of the right type and sets its fields to the field
-values that the virtual object has. This means that instead of the jump, the
-following operations are emitted:
+values that the static object has (if the static object points to other static
+objects, those need to be lifted as well) This means that instead of the jump,
+the following operations are emitted:
 
@@
 worthwhile even in this case, because some operations that have been performed
-on the forced virtual objects have been removed (some \texttt{getfield} operations
+on the lifted static objects have been removed (some \texttt{getfield} operations
 and \texttt{guard\_class} operations).
@@
 XXX Category 2 The optimization of
-Section~\ref{sec:virtuals} deals with them too: the \texttt{new} that creates them and
+Section~\ref{sec:statics} deals with them too: the \texttt{new} that creates them and
 the field accesses are deferred, until the point where the object escapes.

### 2010-10-07 talk/pepm2011/paper.tex
add which languages pypy supports

@@
 aspects of the final VM are woven into the generated code during the translation
-to C. XXX languages that are done using PyPy
+to C.
+
+A number of languages have been implemented with PyPy. The project was started
+to get a better Python implementation, which inspired the name of the project
+and is still the main focus of development. In addition a number of other
+languages were implemented, among them a Prolog interpreter
+\cite{carl_friedrich_bolz_towards_2010}, a Smalltalk VM
+\cite{carl_friedrich_bolz_back_2008} and a GameBoy emulator
+\cite{bruni_pygirl:_2009}.
 
@@
 over writing one manually, which is an error-prone and tedious process.
-By construction, the generated JIT has the same semantics as the interpreter, and
-the process benefits all languages implemented as an interpreter in RPython.
+By construction, the generated JIT has the same semantics as the interpreter.
+Many optimizations can benefit all languages implemented as an interpreter in RPython.
 The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
@@
 called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
-it always ends with a jump to its own beginning.
+it always ends with a jump to its own beginning. The trace also contains all
+operations that are performed in functions that were called in the loop, thus a
+tracing JIT automatically performs inlining.
 
 This trace of operations is then the basis of the generated code. The trace is
-optimized in some ways, and then turned into machine code. Generating machine
-code is simple, because the traces are linear and the operations are very close
-to machine level. The trace corresponds to one concrete execution of a loop,
-therefore the code generated from it is only one possible path through the loop.
+optimized in some ways, and then turned into machine code. Both optimizations
+and machine code generation is simple, because the traces are linear. This
+linearity makes many optimizations a lot more tractable, and the inlining that
+happens gives the optimizations automatically more context to work with.
+
+Since the trace corresponds to one concrete execution of a loop,
+the code generated from it is only one possible path through it.
 To make sure that the trace is maintaining the correct semantics, it contains a
@@
 The creation of objects in category 1 is removed by the optimization described
-in Section~\ref{sec:statics}. XXX
+in Sections~\ref{sec:statics} and \ref{sec:formal}. We will look at objects in
+category 3 in Section~\ref{sec:crossloop}.
 

### 2010-10-07 talk/pepm2011/paper.tex
rename operation to "get" and "set".

@@
 ~~~~\# inside BoxedInteger.add \\
-~~~~$i_{2}$ = getfield($p_{1}$, intval) \\
+~~~~$i_{2}$ = get($p_{1}$, intval) \\
 ~~~~guard\_class($p_{0}$, BoxedInteger) \\
 ~~~~~~~~\# inside BoxedInteger.add\_\_int \\
-~~~~~~~~$i_{3}$ = getfield($p_{0}$, intval) \\
+~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
 ~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
@@
 ~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~setfield($p_{5}$, intval, $i_{4}$) \\
+~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
 \# inside f: BoxedInteger(-100)  \\
@@
 ~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~setfield($p_{6}$, intval, -100) \\
+~~~~set($p_{6}$, intval, -100) \\
 ~\\
@@
 ~~~~\# inside BoxedInteger.add \\
-~~~~$i_{7}$ = getfield($p_{5}$, intval) \\
+~~~~$i_{7}$ = get($p_{5}$, intval) \\
 ~~~~guard\_class($p_{6}$, BoxedInteger) \\
 ~~~~~~~~\# inside BoxedInteger.add\_\_int \\
-~~~~~~~~$i_{8}$ = getfield($p_{6}$, intval) \\
+~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
 ~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
@@
 ~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~setfield($p_{10}$, intval, $i_{9}$) \\
+~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
 ~\\
@@
 ~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~setfield($p_{11}$, intval, -1) \\
+~~~~set($p_{11}$, intval, -1) \\
 ~\\
@@
 ~~~~\# inside BoxedInteger.add \\
-~~~~$i_{12}$ = getfield($p_{0}$, intval) \\
+~~~~$i_{12}$ = get($p_{0}$, intval) \\
 ~~~~guard\_class($p_{11}$, BoxedInteger) \\
 ~~~~~~~~\# inside BoxedInteger.add\_\_int \\
-~~~~~~~~$i_{13}$ = getfield($p_{11}$, intval) \\
+~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
 ~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
@@
 ~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~setfield($p_{15}$, intval, $i_{14}$) \\
+~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
 ~\\
@@
 ~~~~\# inside BoxedInteger.is\_positive \\
-~~~~$i_{16}$ = getfield($p_{15}$, intval) \\
+~~~~$i_{16}$ = get($p_{15}$, intval) \\
 ~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
@@
 looks at the number of \texttt{new} (corresponding to object creation),
-\texttt{set/getfield} (corresponding to attribute reads/writes) and
+\texttt{set/get} (corresponding to attribute reads/writes) and
 \texttt{guard\_class} operations (corresponding to method calls).
@@
 allocated object come from, as well as the type of the object. Whenever the
-optimizer sees a \texttt{setfield} that writes into such an object, that shape
+optimizer sees a \texttt{set} that writes into such an object, that shape
 description is updated and the operation can be removed, which means that the
 operation was done at partial evaluation time. When the optimizer encounters a
-\texttt{getfield} from such an object, the result is read from the shape
+\texttt{get} from such an object, the result is read from the shape
 description, and the operation is also removed. Equivalently, a
@@
 $p_{5}$ = new(BoxedInteger) \\
-setfield($p_{5}$, intval, $i_{4}$) \\
+set($p_{5}$, intval, $i_{4}$) \\
 $p_{6}$ = new(BoxedInteger) \\
-setfield($p_{6}$, intval, -100) \\
+set($p_{6}$, intval, -100) \\
 \end{tabular}
@@
 guard\_class($p_{5}$, BoxedInteger) \\
-$i_{7}$ = getfield($p_{5}$, intval) \\
+$i_{7}$ = get($p_{5}$, intval) \\
 \# inside BoxedInteger.add \\
@@
 \# inside BoxedInteger.add\_\_int \\
-$i_{8}$ = getfield($p_{6}$, intval) \\
+$i_{8}$ = get($p_{6}$, intval) \\
 $i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
@@
 The \texttt{guard\_class} operations can be removed, because the classes of $p_{5}$ and
-$p_{6}$ are known to be \texttt{BoxedInteger}. The \texttt{getfield} operations can be removed
+$p_{6}$ are known to be \texttt{BoxedInteger}. The \texttt{get} operations can be removed
 and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
@@
 $p_{15}$ = new(BoxedInteger) \\
-setfield($p_{15}$, intval, $i_{14}$) \\
+set($p_{15}$, intval, $i_{14}$) \\
 $p_{10}$ = new(BoxedInteger) \\
-setfield($p_{10}$, intval, $i_{9}$) \\
+set($p_{10}$, intval, $i_{9}$) \\
 jump($p_{15}$, $p_{10}$) \\
@@
 worthwhile even in this case, because some operations that have been performed
-on the lifted static objects have been removed (some \texttt{getfield} operations
+on the lifted static objects have been removed (some \texttt{get} operations
 and \texttt{guard\_class} operations).
@@
 
-XXX think of a way to format the operations to make them look distinct from
-normal text
-
 In this section we want to give a formal description of the semantics of the
@@
 
-Traces are lists of operations. The operations considered here are new (to make
-a new object), get (to read a field out of an object), set (to write a field
-into an object) and guard (to check the type of an object). The values of all
+Traces are lists of operations. The operations considered here are \texttt{new} (to make
+a new object), \texttt{get} (to read a field out of an object), \texttt{set} (to write a field
+into an object) and \texttt{guard\_class} (to check the type of an object). The values of all
 variables are locations (i.e.~pointers). Locations are mapped to objects, which
@@
 
-The get operation reads a field $F$ out of an object and adds the result
+The \texttt{get} operation reads a field $F$ out of an object and adds the result
 variable to the environment, mapping to the read location. The heap is
@@
 
-The set operation changes field $F$ of an object stored at the location that
+The \texttt{set} operation changes field $F$ of an object stored at the location that
 variable $v$ maps to. The new value of the field is the location in variable
@@
 
-The guard operation is used to check whether the object stored at the location
+The \texttt{guard\_class} operation is used to check whether the object stored at the location
 that variable $v$ maps to is of type $T$. If that is the case, then execution
@@
 
-If the argument $u$ of a get operation is mapped to something in the static
-heap, the get can be performed at optimization time. Otherwise, the get
+If the argument $u$ of a \texttt{get} operation is mapped to something in the static
+heap, the get can be performed at optimization time. Otherwise, the \texttt{get}
 operation needs to be emitted.
 
-If the first argument $v$ to a set operation is mapped to something in the
-static heap, then the set can performed at optimization time and the static heap
-is updated. Otherwise the set operation needs to be emitted. This needs to be
+If the first argument $v$ to a \texttt{set} operation is mapped to something in the
+static heap, then the \texttt{set} can performed at optimization time and the static heap
+is updated. Otherwise the \texttt{set} operation needs to be emitted. This needs to be
 done carefully, because the new value for the field stored in the variable $u$
@@
 
-I a guard is performed on a variable that is in the static heap, the type check
+I a \texttt{guard\_class} is performed on a variable that is in the static heap, the type check
 can be performed at optimization time, which means the operation can be removed
 if the types match. If the type check fails statically or if the object is not
-in the static heap, the guard is put into the residual trace. This also needs to
-lift the variable on which the guard is performed.
+in the static heap, the \texttt{guard\_class} is put into the residual trace. This also needs to
+lift the variable on which the \texttt{guard\_class} is performed.
 

### 2010-10-07 talk/pepm2011/paper.tex
some improvements

@@
     E & \in & V\rightharpoonup V^* & \mathrm{\ Environment}\\
-    S & \in & V^*\rightharpoonup\mathfrak{T}\times(V^*\cup \{ \mathrm{null} \})\times (V^*\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
+    S & \in & V^*\rightharpoonup\mathfrak{T}\times(V^*\cup \{ \mathrm{null} \})\times (V^*\cup \{ \mathrm{null}\} ) & \mathrm{\ Static\ Heap}\\
     \\
@@
 
-% subsection Analysis of the Algorithm (end)
-
-% section Formal Description of the Algorithm (end)
-
-%___________________________________________________________________________
-
-\subsection{Summary}
-
-In this section we described how simple escape analysis within the scope of one
-loop works. This optimizations reduces the allocation of many intermediate data
-structures that become garbage quickly in an interpreter. It also removes a lot
-of the type dispatching overhead. In the next section, we will explain how this
-optimization can be improved further.
-
 XXX Category 2 The optimization of
@@
 
+% subsection Analysis of the Algorithm (end)
+
+% section Formal Description of the Algorithm (end)
+
+%___________________________________________________________________________
+
 % section Escape Analysis in a Tracing JIT (end)
 
-\section{Escape Analysis Across Loop Boundaries}
+\section{Allocation Removal Across Loop Boundaries}
 \label{sec:crossloop}
 
-In the last section we described how escape analysis can be used to remove
+In the last sections we described how partial evaluation can be used to remove
 many of the allocations of short-lived objects and many of the type dispatches
@@
 
-Let's look at the final trace obtained in the last section for the example loop.
-The final trace was much better than the original one, because many allocations
-were removed from it. However, it also still contained allocations:
+Let's look at the final trace obtained in Section~\ref{sec:statics} for the
+example loop. The final trace (Figure~\ref{fig:step1}) was much better than the
+original one, because many allocations were removed from it. However, it also
+still contained allocations.
 
@@
 Now the lifetime of the remaining allocations no longer crosses the jump, and
-we can run our escape analysis a second time, to get the trace in
+we can run our partial evaluation a second time, to get the trace in
 Figure~\ref{fig:step3}.
@@
 
-% section Escape Analysis Across Loop Boundaries (end)
+% section Allocation Removal Across Loop Boundaries (end)
 

### 2010-10-07 talk/pepm2011/paper.tex
this won't fit
these are all done
kill more XXXs

@@
            {cfbolz@gmx.de, anto.cuni@gmail.com, fijal@merlinux.eu,
-           samuele.pedroni@gmail.com, arigo@tunes.org}
+           leuschel@cs.uni-duesseldorf.de, samuele.pedroni@gmail.com, arigo@tunes.org}
 
@@
 
-XXX drop the word "allocation removal" somewhere
-
-XXX define "escape analysis"
-
 \section{Introduction}
-
-XXX need to re-target introduction a bit to fit PEPMs focus
 
@@
 \texttt{guard\_class} operations (corresponding to method calls).
+In the rest of the paper we will see how this trace can be optimized using
+partial evaluation.
 
@@
 using the interpreter.
-
-XXX simplify traces a bit more
-
-In the next section, we will see how this can be improved upon, using partial
-evaluation. XXX
 
@@
 
-\footnote{This section is a bit
-science-fictiony. The algorithm that PyPy currently uses is significantly more
-complex and much harder than the one that is described here. The resulting
-behaviour is very similar, however, so we will use the simpler version (and we
-might switch to that at some point in the actual implementation).}
-
 Let's look at the final trace obtained in Section~\ref{sec:statics} for the
@@
 their lifetime crosses the jump. To improve the situation, a little trick is
-needed. The trace in Figure~\ref{fig:step1} represents a loop, i.e. the jump at the end jumps to
+needed.\footnote{The algorithm that PyPy currently uses is significantly more complex
+than the one that is described here. The resulting behaviour is nearly
+identical, however, so we will use the simpler version (and plan to switch to
+that at some point in the actual implementation).}
+The trace in Figure~\ref{fig:step1} represents a loop, i.e. the jump at the end jumps to
 the beginning. Where in the loop the jump occurs is arbitrary, since the loop
@@
 
-If we do that for our example trace, the trace looks like in Figure~\ref{fig:step2}.
-
 \begin{figure}
@@
 
+If we do that for our example trace, the trace looks like in Figure~\ref{fig:step2}.
 Now the lifetime of the remaining allocations no longer crosses the jump, and
@@
 % subsection Virtualizables (end)
-XXX store sinking
-
-do we need this?
 

### 2010-10-08 talk/pepm2011/paper.tex
preliminary numbers, psyco and pypy-nojit are missing

@@
 
+\begin{figure*}
+\begin{tabular}{lrrrr}
+		&CPython	&JIT no optimizations	&JIT allocation removal	&JIT full\\
+crypto\_pyaes	&1958.40 $\pm$ 0.32	&782.14 $\pm$ 3.08	&150.86 $\pm$ 3.62	&125.68 $\pm$ 3.02\\
+django		&630.91 $\pm$ 0.13	&365.90 $\pm$ 1.57	&124.41 $\pm$ 0.88	&117.36 $\pm$ 0.91\\
+fannkuch	&1344.15 $\pm$ 0.36	&378.18 $\pm$ 0.36	&309.04 $\pm$ 1.56	&299.12 $\pm$ 0.25\\
+go		&595.31 $\pm$ 0.51	&1067.26 $\pm$ 12.20	&121.76 $\pm$ 4.15	&125.57 $\pm$ 3.90\\
+html5lib	&8678.67 $\pm$ 23.37	&17665.16 $\pm$ 2695.45	&6739.95 $\pm$ 1225.25	&6699.24 $\pm$ 1295.92\\
+meteor-contest	&241.53 $\pm$ 0.09	&291.11 $\pm$ 0.86	&274.09 $\pm$ 0.37	&272.77 $\pm$ 0.30\\
+nbody\_modified	&396.55 $\pm$ 0.24	&84.77 $\pm$ 0.14	&69.01 $\pm$ 0.08	&68.71 $\pm$ 0.09\\
+pyflate-fast	&1991.33 $\pm$ 1.90	&1818.70 $\pm$ 4.30	&1062.03 $\pm$ 4.56	&1052.02 $\pm$ 3.24\\
+raytrace-simple	&1598.84 $\pm$ 1.07	&1126.93 $\pm$ 11.42	&93.45 $\pm$ 4.75	&92.17 $\pm$ 4.15\\
+richards	&216.47 $\pm$ 0.20	&169.84 $\pm$ 1.21	&11.81 $\pm$ 0.09	&11.90 $\pm$ 0.29\\
+spambayes	&193.30 $\pm$ 0.03	&331.37 $\pm$ 14.93	&238.83 $\pm$ 6.10	&237.23 $\pm$ 5.97\\
+spectral-norm	&375.91 $\pm$ 0.08	&221.24 $\pm$ 0.07	&31.96 $\pm$ 0.06	&28.79 $\pm$ 0.09\\
+telco		&789.67 $\pm$ 1.14	&622.67 $\pm$ 1.86	&155.33 $\pm$ 1.82	&154.67 $\pm$ 1.82\\
+twisted\_names	&6.61 $\pm$ 0.00	&8.85 $\pm$ 0.09	&4.35 $\pm$ 0.05 &4.35 $\pm$ 0.04\\
+\end{tabular}
+\caption{Benchmark Times}
+\label{fig:times}
+\end{figure*}
+
 \section{Related Work}

### 2010-10-12 talk/pepm2011/paper.tex
flesh out the benchmarking and related work sections a bit more. add some acknowlegdements

--- 2010-10-07 talk/pepm2011/paper.tex
@@
 
-\keywords{XXX}%
+\keywords{Tracing JIT, Partial Evaluation, Optimization}
 
@@
 
+XXX optimization particularly effective for chains of operations
+
 % subsection Analysis of the Algorithm (end)
@@
 
-XXX optimization particularly effective for chains of operations
-
 %___________________________________________________________________________
@@
 
-Benchmarks from the Computer Language Benchmark Game are: fannkuch, nbody,
-meteor-contest, spectral-norm.
-
-\textbf{crypto\_pyaes}: AES implementation.
-\textbf{django}: The templating engine of the Django web
+To evaluate the effectiveness of our allocation removal algorithm, we look at
+the effectiveness when used in the tracing JIT of PyPy's Python interpreter. The
+benchmarks we used are small-to-medium Python programs, some synthetic
+benchmarks, some real applications.
+
+Some of them are from the Computer Language Benchmark
+Game\footnote{\texttt{http://shootout.alioth.debian.org/}}: \textbf{fannkuch},
+\textbf{nbody}, \textbf{meteor-contest}, \textbf{spectral-norm}.
+
+Furthermore there are the following benchmarks:
+\begin{itemize}
+    \item \textbf{crypto\_pyaes}: AES implementation.
+    \item \textbf{django}: The templating engine of the Django web
     framework\footnote{\texttt{http://www.djangoproject.com/}}.
-\textbf{go}: A Monte-Carlo Go
-AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/disco-elegant-python-go-player.html}}.
-\textbf{html5lib}: HTML5 parser
-\textbf{pyflate-fast}: BZ2 decoder
-\textbf{raytrace-simple}: ray tracer
-\textbf{richards}: The Richards benchmark \cite{XXX}
-\textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
-\textbf{telco}: A Python version of the Telco decimal
-benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a pure
-Python decimal floating point implementation.
-\textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
-framework\footnote{\texttt{http://twistedmatrix.com/trac/}}.
+    \item \textbf{go}: A Monte-Carlo Go
+    AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
+    \item \textbf{html5lib}: HTML5 parser
+    \item \textbf{pyflate-fast}: BZ2 decoder
+    \item \textbf{raytrace-simple}: ray tracer
+    \item \textbf{richards}: The Richards benchmark \cite{XXX}
+    \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
+    \item \textbf{telco}: A Python version of the Telco decimal
+    benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
+    using a pure Python decimal floating point implementation.
+    \item \textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
+    framework\footnote{\texttt{http://twistedmatrix.com/}}.
+\end{itemize}
+
+We evaluate the allocation removal algorithm along two lines: First we want to
+know how many allocations could be optimized away. On the other hand, we want
+to know how much the run times of the benchmarks is improved.
+
+For the former we counted the occurring operations in all generated traces
+before and after the optimization phase for all benchmarks. The results can be
+seen in Figure~\ref{fig:numops}. The optimization removes as many as XXX and as
+little as XXX percent of allocation operations in the benchmarks. All benchmarks
+taken together, the optimization removes XXX percent of allocation operations.
+
+\begin{figure*}
+\begin{tabular}{lrrrrrrrrrrrrrrrrrrrrrrr}
+	&\#loops	&new	&	&	&get	&	&	&set	&	&	&guard	&	&	&rest	&	&	&	&	&	&total	&	&\\
+crypto\_pyaes	&75	&1663	&446	&27\%	&18842	&2127	&11\%	&14552	&2020	&14\%	&8952	&235	&3\%	&53339	&7794	&15\%	&44009	&4828	&11\%	&97348	&12622	&13\%\\
+django	&53	&453	&155	&34\%	&11970	&1457	&12\%	&8738	&991	&11\%	&4093	&275	&7\%	&33988	&4076	&12\%	&25254	&2878	&11\%	&59242	&6954	&12\%\\
+fannkuch	&40	&161	&76	&47\%	&574	&306	&53\%	&278	&139	&50\%	&1121	&202	&18\%	&2581	&1685	&65\%	&2134	&723	&34\%	&4715	&2408	&51\%\\
+go	&510	&5604	&1651	&29\%	&112403	&12612	&11\%	&88439	&11915	&13\%	&51621	&3194	&6\%	&300831	&30834	&10\%	&258067	&29372	&11\%	&558898	&60206	&11\%\\
+html5lib	&416	&5939	&1793	&30\%	&241402	&15863	&7\%	&178723	&13960	&8\%	&58486	&3037	&5\%	&682654	&40341	&6\%	&484550	&34653	&7\%	&1167204	&74994	&6\%\\
+meteor-contest	&56	&221	&141	&64\%	&2326	&606	&26\%	&2047	&645	&32\%	&1040	&169	&16\%	&7018	&2250	&32\%	&5634	&1561	&28\%	&12652	&3811	&30\%\\
+nbody	&10	&84	&48	&57\%	&287	&159	&55\%	&121	&74	&61\%	&411	&90	&22\%	&986	&737	&75\%	&903	&371	&41\%	&1889	&1108	&59\%\\
+pyflate-fast	&161	&1878	&723	&38\%	&22815	&2592	&11\%	&18415	&3690	&20\%	&8291	&628	&8\%	&64649	&11327	&18\%	&51399	&7633	&15\%	&116048	&18960	&16\%\\
+raytrace-simple	&117	&1906	&359	&19\%	&52995	&2202	&4\%	&39518	&1823	&5\%	&13451	&447	&3\%	&140723	&6533	&5\%	&107870	&4831	&4\%	&248593	&11364	&5\%\\
+richards	&79	&375	&191	&51\%	&27507	&2095	&8\%	&21217	&2004	&9\%	&3972	&332	&8\%	&76751	&6171	&8\%	&53071	&4622	&9\%	&129822	&10793	&8\%\\
+spambayes	&267	&2413	&505	&21\%	&53327	&3033	&6\%	&40938	&2428	&6\%	&19343	&718	&4\%	&143575	&10322	&7\%	&116021	&6684	&6\%	&259596	&17006	&7\%\\
+spectral-norm	&35	&266	&88	&33\%	&3004	&398	&13\%	&2544	&414	&16\%	&1084	&78	&7\%	&4803	&1184	&25\%	&6898	&978	&14\%	&11701	&2162	&18\%\\
+telco	&42	&660	&70	&11\%	&20716	&424	&2\%	&14444	&608	&4\%	&6574	&56	&1\%	&53674	&1224	&2\%	&42394	&1158	&3\%	&96068	&2382	&2\%\\
+twisted-names	&201	&2706	&317	&12\%	&64438	&2795	&4\%	&47960	&1968	&4\%	&24091	&555	&2\%	&171582	&6828	&4\%	&139195	&5635	&4\%	&310777	&12463	&4\%\\
+total	&2062	&24329	&6563	&27\%	&632606	&46669	&7\%	&477934	&42679	&9\%	&202530	&10016	&5\%	&1737154	&131306	&8\%	&1337399	&105927	&8\%	&3074553	&237233	&8\%\\
+\end{tabular}
+\caption{Number of Operations Before and After Optimization}
+\label{fig:numops}
+\end{figure*}
+
+In addition to the count of operations we also performed time measurements. The
+machine the benchmarks were run on is XXX. We compared the performance of
+various Python implementations on the benchmarks. As a baseline, we used the
+standard Python implementation in C, called
+CPython\footnote{\texttt{http://python.org}}, which uses a bytecode-based
+interpreter. Furthermore we compared against Psyco \cite{XXX}, an extension to
+CPython which is a just-in-time compiler that produces machine code at run-time.
+It is not based on traces. Of PyPy's Python interpreter we used three versions,
+one without a JIT, one including the JIT but not using the allocation removal
+optimization, and one using the allocation removal optimizations.
+
+All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
+code. The arithmetic mean of the times of the last 30 runs were used as the
+result. The errors were computed using a confidence interval with a 95\%
+confidence level \cite{XXX}. The results are reported in Figure~\ref{fig:times}.
+With the optimization turned on, PyPy's Python interpreter outperforms CPython
+in all benchmarks except spambayes (which heavily relies on regular expression
+performance). All benchmarks are improved by the allocation removal
+optimization, some as much as XXX. XXX Psyco
+
+XXX runtimes of the algorithm somehow?
+
+\begin{figure*}
+\begin{tabular}{lrrrr}
+		&CPython	&JIT no optimizations	&JIT allocation removal	&JIT full\\
+crypto\_pyaes	&1958.40 $\pm$ 0.32	&782.14 $\pm$ 3.08	&150.86 $\pm$ 3.62	&125.68 $\pm$ 3.02\\
+django		&630.91 $\pm$ 0.13	&365.90 $\pm$ 1.57	&124.41 $\pm$ 0.88	&117.36 $\pm$ 0.91\\
+fannkuch	&1344.15 $\pm$ 0.36	&378.18 $\pm$ 0.36	&309.04 $\pm$ 1.56	&299.12 $\pm$ 0.25\\
+go		&595.31 $\pm$ 0.51	&1067.26 $\pm$ 12.20	&121.76 $\pm$ 4.15	&125.57 $\pm$ 3.90\\
+html5lib	&8678.67 $\pm$ 23.37	&17665.16 $\pm$ 2695.45	&6739.95 $\pm$ 1225.25	&6699.24 $\pm$ 1295.92\\
+meteor-contest	&241.53 $\pm$ 0.09	&291.11 $\pm$ 0.86	&274.09 $\pm$ 0.37	&272.77 $\pm$ 0.30\\
+nbody		&396.55 $\pm$ 0.24	&84.77 $\pm$ 0.14	&69.01 $\pm$ 0.08	&68.71 $\pm$ 0.09\\
+pyflate-fast	&1991.33 $\pm$ 1.90	&1818.70 $\pm$ 4.30	&1062.03 $\pm$ 4.56	&1052.02 $\pm$ 3.24\\
+raytrace-simple	&1598.84 $\pm$ 1.07	&1126.93 $\pm$ 11.42	&93.45 $\pm$ 4.75	&92.17 $\pm$ 4.15\\
+richards	&216.47 $\pm$ 0.20	&169.84 $\pm$ 1.21	&11.81 $\pm$ 0.09	&11.90 $\pm$ 0.29\\
+spambayes	&193.30 $\pm$ 0.03	&331.37 $\pm$ 14.93	&238.83 $\pm$ 6.10	&237.23 $\pm$ 5.97\\
+spectral-norm	&375.91 $\pm$ 0.08	&221.24 $\pm$ 0.07	&31.96 $\pm$ 0.06	&28.79 $\pm$ 0.09\\
+telco		&789.67 $\pm$ 1.14	&622.67 $\pm$ 1.86	&155.33 $\pm$ 1.82	&154.67 $\pm$ 1.82\\
+twisted\_names	&6.61 $\pm$ 0.00	&8.85 $\pm$ 0.09	&4.35 $\pm$ 0.05 &4.35 $\pm$ 0.04\\
+\end{tabular}
+\caption{Benchmark Times}
+\label{fig:times}
+\end{figure*}
 
@@
 
+Escape analysis: XXX
+
+partial evaluation:
+
+Prolog
+
+partially static data structures: kenichi asai's thesis?
+
+hint in first js paper by michael franz \cite{mason_chang_efficient_2007}
+
+SPUR, a tracing JIT for C# seems to be able to remove allocations in a similar
+way to the approach described here, as hinted at in the technical report
+\cite{XXX}. However, no details for the approach and its implementation are
+given.
+
 \section{Conclusions}
@@
 
+\section*{Acknowledgements}
+
+The authors would like to thank Stefan Hallerstede and Thomas Stiehl for
+fruitful discussions during the writing of the paper.
+
 \bibliographystyle{abbrv}

### 2010-10-12 talk/pepm2011/paper.tex
Typos.  Swap $u$ and $v$ in \texttt{get}, to match the \texttt{set}.

@@
 The static object associated with $p_{5}$ would know that it is a
-\texttt{BoxedInteger}, and that the \texttt{intval} field contains $i_{4}$, the
-one associated with $p_{6}$ would know that its \texttt{intval} field contains
-the constant -100.
-
-The following operations, that use $p_{5}$ and $p_{6}$ could then be
+\texttt{BoxedInteger} whose \texttt{intval} field contains $i_{4}$; the
+one associated with $p_{6}$ would know that it is a \texttt{BoxedInteger}
+whose \texttt{intval} field contains the constant -100.
+
+The following operations on $p_{5}$ and $p_{6}$ could then be
 optimized using that knowledge:
@@
 and because the partial evaluator looses track of it. This means that the static
-objects needs to be turned into a dynamic one, \ie lifted. This makes it
+object needs to be turned into a dynamic one, \ie lifted. This makes it
 necessary to put operations into the residual code that actually allocate the
@@
 is hit. The arguments of the jump are at this point static objects. Before the
-jump is emitted, they are \emph{lifted}. This means that the optimizers produces code
+jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
 that allocates a new object of the right type and sets its fields to the field
@@
 Figure~\ref{fig:step1}.
+XXX why does it says ``Figure 4.1'' here but ``Figure 4'' in the label?
 
@@
  \begin{array}{rcll}
-    u,v,w & \in & V & \mathrm{\ variables\ in\ trace}\\
+    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
     T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
@@
 fields of the object. When a new object is created, the fields are initialized
-to null, but we require that they are immediately initialized to a real
-location, otherwise the trace is malformed.
+to null, but we require that they are initialized to a real
+location before being read, otherwise the trace is malformed.
 
@@
 interpreter formalized there executes one operation at a time. Its state is
-represented by an environment and a heap, which are potentially changed by the
+represented by an environment $E$ and a heap $H$, which are potentially changed by the
 execution of an operation. The environment is a partial function from variables
@@
 
-The new operation creates a new object $(T,\mathrm{null},\mathrm{null})$, on the
+The \texttt{new} operation creates a new object $(T,\mathrm{null},\mathrm{null})$ on the
 heap under a fresh location $l$ and adds the result variable to the environment,
-mapping to the new location $l$.
-
-The \texttt{get} operation reads a field $F$ out of an object and adds the result
-variable to the environment, mapping to the read location. The heap is
+mapping it to the new location $l$.
+
+The \texttt{get} operation reads a field $F$ out of an object, and adds the result
+variable to the environment, mapping it to the read location. The heap is
 unchanged.
@@
 heap} $S$. The environment is a partial function from variables in the
-unoptimized trace to variables in the optimized trace (which are written with a
+unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
+themselves written with a
 $\ ^*$ for clarity). The reason for introducing new variables in the optimized
@@
 into the same variables in the optimized trace. Thus the environment of the
-optimizer serves a function similar to that of the environment in the semantics.
+optimizer serves a function similar to that of the environment in the semantics: sharing.
 
 The static heap is a partial function from $V^*$ into the
-set of static objects, which are triples of a type, and two elements of $V^*$.
-An variable $v^*$ is in the domain of the static heap $S$ as long as the
+set of static objects, which are triples of a type and two elements of $V^*$.
+A variable $v^*$ is in the domain of the static heap $S$ as long as the
 optimizer can fully keep track of the object. The image of $v^*$ is what is
@@
 
-When the optimizer sees a new operation, it optimistically removes it and
+When the optimizer sees a \texttt{new} operation, it optimistically removes it and
 assumes that the resulting object can stay static. The optimization for all
@@
 performed at optimization time and removed from the trace. These rules mirror
-the execution semantics closely. The other case is that nothing is known about
-the variables, which means the operation has to be residualized.
-
-If the argument $u$ of a \texttt{get} operation is mapped to something in the static
+the execution semantics closely. The other case is for when not enough is known about
+the variables, and the operation has to be residualized.
+
+If the argument $v$ of a \texttt{get} operation is mapped to something in the static
 heap, the get can be performed at optimization time. Otherwise, the \texttt{get}
-operation needs to be emitted.
+operation needs to be residualized.
 
@@
 static heap, then the \texttt{set} can performed at optimization time and the static heap
-is updated. Otherwise the \texttt{set} operation needs to be emitted. This needs to be
-done carefully, because the new value for the field stored in the variable $u$
+updated. Otherwise the \texttt{set} operation needs to be residualized. This needs to be
+done carefully, because the new value for the field, from the variable $u$,
 could itself be static, in which case it needs to be lifted first.
 
-I a \texttt{guard\_class} is performed on a variable that is in the static heap, the type check
+If a \texttt{guard\_class} is performed on a variable that is in the static heap, the type check
 can be performed at optimization time, which means the operation can be removed
 if the types match. If the type check fails statically or if the object is not
-in the static heap, the \texttt{guard\_class} is put into the residual trace. This also needs to
+in the static heap, the \texttt{guard\_class} is residualized. This also needs to
 lift the variable on which the \texttt{guard\_class} is performed.
@@
 that it is turned into a dynamic variable. This means that operations are
-emitted that construct an object that looks like the shape described in the
+emitted that construct an object with the shape described in the
 static heap, and the variable is removed from the static heap.
@@
 static heap can be cyclic. To make sure that the same static object is not lifted
-twice, the liftfield operation removes it from the static heap \emph{before}
+twice, the \texttt{liftfield} operation removes it from the static heap \emph{before}
 recursively lifting its fields.
@@
 
-SPUR, a tracing JIT for C# seems to be able to remove allocations in a similar
+SPUR, a tracing JIT for C\# seems to be able to remove allocations in a similar
 way to the approach described here, as hinted at in the technical report

### 2010-10-12 talk/pepm2011/paper.tex
related work

@@
 \newcommand\eg{e.g.,\xspace}
+\newcommand{\etal}{\emph{et al.}\xspace}
 
@@
 
-Escape analysis: XXX
-
-partial evaluation:
-
-Prolog
-
-partially static data structures: kenichi asai's thesis?
-
-hint in first js paper by michael franz \cite{mason_chang_efficient_2007}
+There exists a large number of works on escape analysis, which is an program
+analysis that tries to find an upper bound for the lifetime of objects allocated
+at specific program points
+\cite{goldberg_higher_1990,park_escape_1992,choi_escape_1999,bruno_blanchet_escape_2003}.
+This information can then be used to decide that certain objects can be
+allocated on the stack, because their lifetime does not exceed that of the stack
+frame it is allocated in. The difference to our work is that escape analysis is
+split into an analysis and an optimization phase. The analysis can be a lot more
+complex than our simple one-pass optimization. Also, stack-allocation reduces
+garbage-collection pressure but does not optimize away the actual accesses to
+the stack-allocated object. In our case, an object is not needed at all any
+more.
+XXX more papers?
+
+Chang \etal describe a tracing JIT for JavaScript running on top of a JVM
+\cite{mason_chang_efficient_2007}. They mention in passing an approach to
+allocation removal that moves the allocation of an object of type 1 out of the
+loop to only allocate it once, instead of every iteration. No details are given
+for this optimization. The fact that the object is still allocated and needs to
+be written to means that only the allocations are optimized away, but not the
+reads and writes out of/into the object.
 
@@
 
+partial evaluation:
+
+Prolog
+
+partially static data structures: kenichi asai's thesis?
+
 \section{Conclusions}

### 2010-10-12 talk/pepm2011/paper.tex
added refs
unify stuff
Write a small bit of text for "Analysis of the Algorithm".

@@
 
-XXX algorithm is linear in the length of the trace
-
-XXX Category 2 The optimization of
-Section~\ref{sec:statics} deals with them too: the \texttt{new} that creates them and
-the field accesses are deferred, until the point where the object escapes.
-
-XXX optimization particularly effective for chains of operations
+While we do not offer a formal proof of it, it should be relatively clear
+that the algorithm presented above is sound: it works by delaying (and
+often completely removing) some operations.  The algorithm runs in a
+single pass over the list of operations.  We can check that altough
+recursively lifting a static object is not a constant-time operation,
+the algorithm only takes a total time linear in the length of the trace.
+Moreover, it gives the ``best'' possible result within its constrains,
+e.g.\ in term of the number of residual operations.  The
+algorithm itself is not particularly complex or innovative; our focus is
+rather that \emph{in the context of tracing JITs} it is possible to find a
+simple enough algorithm that still gives the best results.
+
+Note in particular that objects in category 1 (i.e.\ the ones that do
+not escape) are completely removed; moreover, objects in category 2
+(i.e.\ escaping) are still partially dealt with: if such an object
+escapes later than its creation point, all the operations inbetween that
+involve the object are removed.
+
+The optimization is particularly effective for chains of operations.
+For example, it is typical for an interpreter to generate sequences of
+writes-followed-by-reads, where one interpreted opcode writes to some
+object's field and the next interpreted opcode reads it back, possibly
+dispatching on the type of the object created just before.  In the case
+of PyPy's Python interpreter, this optimization can even remove the
+allocation of all intermediate frames that occur in the interpreter,
+corresponding to all calls that have been inlined in the trace.
 
@@
 
-Prolog
+Prolog:
+partially static datastructures are already built-in to Prolog and was built-into partial
+evaluation from the beginning \cite{lloyd_partial_1991}.
+ FP
 
 partially static data structures: kenichi asai's thesis?
+
+xxx: relation to compile-time garbage collection \cite{mazur_practical_2001};  separation logic; John Hughes: type specialization
 

### 2010-10-12 talk/pepm2011/paper.tex
add missing citations

@@
 straightforward application of partial evaluation. The partial evaluation
-performs a form of escape analysis \cite{XXX} on the traces and make some
+performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
 objects that are allocated in the trace \emph{static}\footnote{These objects are
@@
 several implementations of dynamic languages, such as JavaScript
-\cite{andreas_gal_trace-based_2009}, Lua \cite{XXX} and now Python (and other
-languages) via PyPy.
+\cite{andreas_gal_trace-based_2009}, Lua\footnote{\texttt{http://luajit.org/}}
+and now Python (and other languages) via PyPy.
 
@@
 is stored into the frame objects, making the problem of too much escaping go
-away. This is a common approach in VM implementations \cite{XXX}, the only
+away. This is a common approach in VM implementations \cite{miranda_context_1999}, the only
 novelty in our approach lays in its generality, because most other JITs are just
@@
     AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
-    \item \textbf{html5lib}: HTML5 parser
-    \item \textbf{pyflate-fast}: BZ2 decoder
-    \item \textbf{raytrace-simple}: ray tracer
-    \item \textbf{richards}: The Richards benchmark \cite{XXX}
+    \item \textbf{html5lib}: An HTML5 parser.
+    \item \textbf{pyflate-fast}: A BZ2 decoder.
+    \item \textbf{raytrace-simple}: A ray tracer.
+    \item \textbf{richards}: The Richards benchmark.
     \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
-    \item \textbf{telco}: A Python version of the Telco decimal
+    \item \textbf{telco}: A Python version of the Telco decimal.
     benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
@@
 CPython\footnote{\texttt{http://python.org}}, which uses a bytecode-based
-interpreter. Furthermore we compared against Psyco \cite{XXX}, an extension to
+interpreter. Furthermore we compared against Psyco \cite{rigo_representation-based_2004}, an extension to
 CPython which is a just-in-time compiler that produces machine code at run-time.
@@
 result. The errors were computed using a confidence interval with a 95\%
-confidence level \cite{XXX}. The results are reported in Figure~\ref{fig:times}.
+confidence level \cite{georges_statistically_2007}. The results are reported in Figure~\ref{fig:times}.
 With the optimization turned on, PyPy's Python interpreter outperforms CPython
@@
 way to the approach described here, as hinted at in the technical report
-\cite{XXX}. However, no details for the approach and its implementation are
-given.
+\cite{michael_bebenita_spur:_2010}. However, no details for the approach and its
+implementation are given.
 

### 2010-10-12 talk/pepm2011/paper.tex
use nice macros

@@
 Moreover, it gives the ``best'' possible result within its constrains,
-e.g.\ in term of the number of residual operations.  The
+\eg in term of the number of residual operations.  The
 algorithm itself is not particularly complex or innovative; our focus is
@@
 
-Note in particular that objects in category 1 (i.e.\ the ones that do
+Note in particular that objects in category 1 (\ie the ones that do
 not escape) are completely removed; moreover, objects in category 2
-(i.e.\ escaping) are still partially dealt with: if such an object
+(\ie escaping) are still partially dealt with: if such an object
 escapes later than its creation point, all the operations inbetween that

### 2010-10-12 talk/pepm2011/paper.tex
kill the description of the cross-loop optimization to save space :`-(

@@
 
-The basic approach of static objects can then be extended to also be used for
-type-specializing the traces that are produced by the tracing JIT
-(Section~\ref{sec:crossloop}). In Section~\ref{sec:support} we describe some
-supporting techniques that are not central to the approach, but are needed to
-improve the results. The introduced techniques are evaluated in
-Section~\ref{sec:Evaluation} using PyPy's Python interpreter as a case study.
+In Section~\ref{sec:support} we describe some supporting techniques that are not
+central to the approach, but are needed to improve the results. The introduced
+techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
+interpreter as a case study.
 
@@
     and then escape. To these we also count the objects that live across several
-    jumps and then either escape or stop being used.\footnote{In theory, the
-    approach of Section~\ref{sec:crossloop} works also for objects that live for
-    exactly $n>1$ iterations and then don't escape, but we expect this to be a
-    very rare case, so we do not handle it.}
+    jumps and then either escape or stop being used.
 \end{itemize}
@@
 The creation of objects in category 1 is removed by the optimization described
-in Sections~\ref{sec:statics} and \ref{sec:formal}. We will look at objects in
-category 3 in Section~\ref{sec:crossloop}.
+in Sections~\ref{sec:statics} and \ref{sec:formal}. Objects in the other
+categories are partially optimized by this approach as well.
 
@@
 
-\section{Allocation Removal Across Loop Boundaries}
-\label{sec:crossloop}
-
-In the last sections we described how partial evaluation can be used to remove
-many of the allocations of short-lived objects and many of the type dispatches
-that are present in a non-optimized trace. In this section we will improve the
-optimization to also handle more cases.
-
-The optimization of the last section considered the passing of an object along a
-jump to be equivalent to escaping. It was thus treating objects in category 3
-and 4 like those in category 2.
-
-The improved optimization described in this section will make it possible to deal
-better with objects in category 3 and 4. This will have two consequences: on
-the one hand, more allocations are removed from the trace (which is clearly
-good). As a side-effect of this, the traces will also be type-specialized.
-
-%___________________________________________________________________________
-
-\subsection{Optimizing Across the Jump}
-
-Let's look at the final trace obtained in Section~\ref{sec:statics} for the
-example loop. The final trace (Figure~\ref{fig:step1}) was much better than the
-original one, because many allocations were removed from it. However, it also
-still contained allocations.
-
-The two new \texttt{BoxedIntegers} stored in $p_{15}$ and $p_{10}$ are passed into
-the next iteration of the loop. The next iteration will check that they are
-indeed \texttt{BoxedIntegers}, read their \texttt{intval} fields and then not use them
-any more. Thus those instances are in category 3.
-
-In its current state the loop
-allocates two \texttt{BoxedIntegers} at the end of every iteration, that then die
-very quickly in the next iteration. In addition, the type checks at the start
-of the loop are superfluous, at least after the first iteration.
-
-The reason why we cannot optimize the remaining allocations away is because
-their lifetime crosses the jump. To improve the situation, a little trick is
-needed.\footnote{The algorithm that PyPy currently uses is significantly more complex
-than the one that is described here. The resulting behaviour is nearly
-identical, however, so we will use the simpler version (and plan to switch to
-that at some point in the actual implementation).}
-The trace in Figure~\ref{fig:step1} represents a loop, i.e. the jump at the end jumps to
-the beginning. Where in the loop the jump occurs is arbitrary, since the loop
-can only be left via failing guards anyway. Therefore it does not change the
-semantics of the loop to put the jump at another point into the trace and we
-can move the \texttt{jump} operation just above the allocation of the objects that
-appear in the current \texttt{jump}. This needs some care, because the arguments to
-\texttt{jump} are all currently live variables, thus they need to be adapted.
-
-\begin{figure}
-\includegraphics{figures/step2.pdf}
-\label{fig:step2}
-\caption{Shifting the Jump}
-\end{figure}
-
-If we do that for our example trace, the trace looks like in Figure~\ref{fig:step2}.
-Now the lifetime of the remaining allocations no longer crosses the jump, and
-we can run our partial evaluation a second time, to get the trace in
-Figure~\ref{fig:step3}.
-
-\begin{figure}
-\includegraphics{figures/step3.pdf}
-\label{fig:step3}
-\caption{Removing Allocations a Second Time}
-\end{figure}
-
-This result is now really good. The code performs the same operations than
-the original code, but using direct CPU arithmetic and no boxing, as opposed to
-the original version which used dynamic dispatching and boxing.
-
-Looking at the final trace it is also completely clear that specialization has
-happened. The trace corresponds to the situation in which the trace was
-originally recorded, which happened to be a loop where \texttt{BoxedIntegers} were
-used. The now resulting loop does not refer to the \texttt{BoxedInteger} class at
-all any more, but it still has the same behaviour. If the original loop had
-used \texttt{BoxedFloats}, the final loop would use \texttt{float\_*} operations
-everywhere instead (or even be very different, if the object model had
-more different classes).
-
-%___________________________________________________________________________
-
-\subsection{Entering the Loop}
-
-The approach of placing the \texttt{jump} at some other point in the loop leads to
-one additional complication that we glossed over so far. The beginning of the
-original loop corresponds to a point in the original program, namely the
-\texttt{while} loop in the function \texttt{f} from the last section.
-
-Now recall that in a VM that uses a tracing JIT, all programs start by being
-interpreted. This means that when \texttt{f} is executed by the interpreter, it is
-easy to go from the interpreter to the first version of the compiled loop.
-After the \texttt{jump} is moved and the escape analysis optimization is applied a
-second time, this is no longer easily possible.  In particular, the new loop
-expects two integers as input arguments, while the old one expected two
-instances.
-
-To make it possible to enter the loop directly from the intepreter, there
-needs to be some additional code that enters the loop by taking as input
-arguments what is available to the interpreter, i.e. two instances. This
-additional code corresponds to one iteration of the loop, which is thus
-peeled off \cite{XXX}, see Figure~\ref{fig:step3}.
-
-\begin{figure}
-\includegraphics{figures/step4.pdf}
-\label{fig:step3}
-\caption{A Way to Enter the Loop From the Interpreter}
-\end{figure}
-
-%___________________________________________________________________________
-
-\subsection{Summary}
-
-The optimization described in this section can be used to optimize away
-allocations in category 3 and improve allocations in category 4, by deferring
-them until they are no longer avoidable. A side-effect of these optimizations
-is also that the optimized loops are specialized for the types of the variables
-that are used inside them.
-
-% section Allocation Removal Across Loop Boundaries (end)
-
 \section{Supporting Techniques}

### 2010-10-12 talk/pepm2011/paper.tex
We plan to have only one conclusion, not many.
clean up figures
actually reference this figure
(cfbolz, arigo) Conclusion.

@@
 double-dispatching. These classes could be part of the implementation of a very
-simple interpreter written in RPython.
+simple interpreter written in RPython. The classes can be seen in
+Figure~\ref{fig:objmodel}.
 
@@
 \caption{A simple object model}
+\label{fig:objmodel}
 \end{figure}
@@
 }
+\caption{Unoptimized Trace for the Simple Object Model}
 \label{fig:unopt-trace}
-\caption{Unoptimized Trace for the Simple Object Model}
 \end{figure}
@@
 \includegraphics{figures/step1.pdf}
+\caption{Resulting Trace After Allocation Removal}
 \label{fig:step1}
-\caption{Resulting Trace After Allocation Removal}
 \end{figure}
 
-The final optimized trace of the example can be seen in
-Figure~\ref{fig:step1}.
-XXX why does it says ``Figure 4.1'' here but ``Figure 4'' in the label?
-
+The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
 The optimized trace contains only two allocations, instead of the original five,
@@
 \end{center}
+\caption{The Operational Semantics of Simplified Traces}
 \label{fig:semantics}
-\caption{The Operational Semantics of Simplified Traces}
 \end{figure*}
@@
 \end{center}
+\caption{Optimization Rules}
 \label{fig:optimization}
-\caption{Optimization Rules}
 \end{figure*}
@@
 operations in the optimized trace. This list is empty if the operation
-can be optimized away (which hopefully happens often).
+can be optimized away (which hopefully happens often). The optimization rules
+can be seen in Figure~\ref{fig:optimization}.
 
@@
 
-\section{Conclusions}
-\label{sec:conclusions}
+\section{Conclusion}
+\label{sec:conclusion}
+
+In this paper, we used a approach based on partial evaluation to
+optimize allocations in the traces of a tracing JIT.  In this context a
+simple approach to partial evaluation gives good results.  This is due
+to the fact that the tracing JIT itself is responsible for all control
+issues, which are usually the hardest part of partial evaluation: the
+tracing JIT selects the parts of the program that are worthwhile to
+optimize, and extracts linear paths through them, inlining functions as
+necessary.  What is left to optimize is only those linear paths.
+
+We expect a similar result for other optimizations that usually require
+a complex analysis phase and are thus normally too slow to use at
+runtime.  A tracing JIT selects interesting linear paths by itself;
+therefore, a naive version of many optimizations on such paths should
+give mostly the same results.  For example, we experimented with (and
+plan to write about) store-load propagation with a very simple alias
+analysis.
 

### 2010-10-12 talk/pepm2011/paper.tex
(cfbolz, arigo) Abstract.

@@
 \begin{abstract}
+The performance of many dynamic language implementations suffers from
+high allocation rates and runtime type checks.  This makes dynamic
+languages less applicable to purely algorithmic problems, despite their
+growing popularity.  In this paper, we present a simple optimization
+based on online partial evaluation to remove object allocations and
+runtime type checks in the context of a tracing JIT.  We evaluate the
+optimization using a Python VM and find that it gives good results for
+all our (real-life) benchmarks.
 \footnote{This research is partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
@@
 
-In this paper, we used a approach based on partial evaluation to
+In this paper, we used a approach based on online partial evaluation to
 optimize allocations in the traces of a tracing JIT.  In this context a

### 2010-10-12 talk/pepm2011/paper.tex
re-import math from lyx file

@@
 \begin{tabular}{lcccc}
-\emph{new} & ${\displaystyle \frac{l\,\mathrm{fresh}}{v=\mathrm{new}(T),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto l\right],H\left[l\mapsto\left(T,\mathrm{null},\mathrm{null}\right)\right]}}$ & ~~~ &
-\emph{guard} & ${\displaystyle \frac{\mathrm{type}(H(E(v))=T}{\mathrm{guard}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H}}$\tabularnewline[3em]
-\emph{get} & ${\displaystyle \frac{\,}{v=\mathrm{get}(u,F),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto H\left(E\left(u\right)\right)_{F}\right],H}}$ & ~~~ &
-& ${\displaystyle \frac{\mathrm{type}(H(E(v))\neq T}{\mathrm{guard}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}\bot,\bot}}$\tabularnewline[3em]
-\emph{set} & ${\displaystyle \frac{\,}{\mathrm{set}\left(v,F,u\right),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H\left[E\left(v\right)\mapsto\left(H\left(E\left(v\right)\right)!_{F}E(u)\right)\right]}}$ & ~~~ &
-& \tabularnewline[4em]
+\emph{new} & ${\displaystyle \frac{l\,\mathrm{fresh}}{v=\mathtt{new}(T),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto l\right],H\left[l\mapsto\left(T,\mathrm{null},\mathrm{null}\right)\right]}}$ & ~~~ & \emph{guard} & ${\displaystyle \frac{\mathrm{type}(H(E(v))=T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H}}$\tabularnewline[3em]
+\emph{get} & ${\displaystyle \frac{\,}{u=\mathtt{get}(v,F),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[u\mapsto H\left(E\left(v\right)\right)_{F}\right],H}}$ & ~~~ &  & ${\displaystyle \frac{\mathrm{type}(H(E(v))\neq T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}\bot,\bot}}$\tabularnewline[3em]
+\emph{set} & ${\displaystyle \frac{\,}{\mathtt{set}\left(v,F,u\right),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H\left[E\left(v\right)\mapsto\left(H\left(E\left(v\right)\right)!_{F}E(u)\right)\right]}}$ & ~~~ &  & \tabularnewline[4em]
 \end{tabular}
@@
 \begin{tabular}{lc}
-\emph{new} & ${\displaystyle \frac{v^{*}\,\mathrm{fresh}}{v=\mathrm{new}(T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto v^{*}\right],S\left[v^{*}\mapsto\left(T,\mathrm{null,null}\right)\right]}}$\tabularnewline[3em]
-\emph{get} & ${\displaystyle \frac{E(u)\in\mathrm{dom}(S)}{v=\mathrm{get}(u,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto S(E(u))_{F}\right],S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(u)\notin\mathrm{dom}(S)\, v^{*}\,\mathrm{fresh}}{v=\mathrm{get}(u,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle v^{*}=\mathrm{get}(E(u),F)\right\rangle ,E\left[v\mapsto v^{*}\right],S}}$\tabularnewline[3em]
-\emph{set} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{\mathrm{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S\left[E\left(v\right)\mapsto\left(S(E(v))!_{F}E(u)\right)\right]}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\notin\mathrm{dom}\left(S\right),\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathrm{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathrm{set}\left(E(v),F,E(u)\right)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
-\emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\vee\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathrm{guard}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
+\emph{new} & ${\displaystyle \frac{v^{*}\,\mathrm{fresh}}{v=\mathtt{new}(T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto v^{*}\right],S\left[v^{*}\mapsto\left(T,\mathrm{null,null}\right)\right]}}$\tabularnewline[3em]
+\emph{get} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[u\mapsto S(E(v))_{F}\right],S}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\, u^{*}\,\mathrm{fresh}}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle u^{*}=\mathtt{get}(E(v),F)\right\rangle ,E\left[u\mapsto u^{*}\right],S}}$\tabularnewline[3em]
+\emph{set} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S\left[E\left(v\right)\mapsto\left(S(E(v))!_{F}E(u)\right)\right]}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\notin\mathrm{dom}\left(S\right),\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathtt{set}\left(E(v),F,E(u)\right)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
+\emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\vee\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathtt{guard\_class}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
 \emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathrm{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
- & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathrm{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathrm{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathtt{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathtt{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathtt{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline[3em]
 \end{tabular}
@@
  \begin{array}{rcll}
-    u,v,w & \in & V & \mathrm{\ variables\ in\ trace}\\
-    u^*,v^*,w^* & \in & V^* & \mathrm{\ variables\ in\ optimized\ trace}\\
+    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
+    u^*,v^* & \in & V^* & \mathrm{\ variables\ in\ optimized\ trace}\\
     T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\

### 2010-10-13 talk/pepm2011/paper.tex
fix some typos, add few XXXs, improve sentences here and there

@@
 \newcommand\david[1]{\nb{DAVID}{#1}}
+\newcommand\anto[1]{\nb{ANTO}{#1}}
 \newcommand\reva[1]{\nb{Reviewer 1}{#1}}
@@
 languages is that of a tracing JIT. A tracing JIT works by observing the running
-program and recording linear execution traces, which are then turned into
+program and recording its hot parts into linear execution traces, which are then turned into
 machine code. One reason for the popularity of tracing JITs is their relative
@@
 optimizations like inlining and constant-folding for free. A tracing JIT always
-produces linear pieces of code, which makes many optimizations that are usually
-hard in a compiler simpler, such as register allocation.
+produces linear pieces of code, which simplifies many optimizations that are usually
+hard in a compiler, such as register allocation.
 
@@
 
-The feature that makes PyPy more than a compiler with a runtime system is it's
+The feature that makes PyPy more than a compiler with a runtime system is its
 support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
@@
 Many optimizations can benefit all languages implemented as an interpreter in RPython.
+
+Moreover, thanks to the internal design of the JIT generator, it is very easy
+to add new \emph{backends} for producing the actual machine code, in addition
+to the original backend for the Intel \emph{x86} architecture.  Examples of
+additional JIT backends are the one for Intel \emph{x86-64} and an
+experimental one for the CLI .NET Virtual Machine \cite{cuni_python_cli_2010}.
 The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
@@
 called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
-it always ends with a jump to its own beginning. The trace also contains all
+it always ends with a jump to its own beginning \anto{this is not true: what if we trace the last iteration?}. The trace also contains all
 operations that are performed in functions that were called in the loop, thus a
@@
 assembler is patched to jump to the new trace when the guard fails
-\cite{andreas_gal_incremental_2006}.
+\cite{andreas_gal_incremental_2006}.  This approach guarantees that all the
+hot paths in the program will eventually be traced and compiled into efficient
+code.
 
@@
 when executing that loop, many of these objects do not survive for very long.
-The actual computation that is performed by \texttt{f} is simply a number of
+The actual computation that is performed by \texttt{f} is simply a sequence of
 float or integer additions.
@@
 \begin{tabular}{l} 
+\# XXX: maybe we should specify that $p_{0}$, $p_{1}$ corresponds to y and res
 \# arguments to the trace: $p_{0}$, $p_{1}$ \\
@@
 \texttt{guard\_class} on a variable that has a shape description can be removed
-as well, because the shape description stores the type.
+as well, because the shape description stores the type and thus the result of 
+the guard is statically known.
 
@@
 whose \texttt{intval} field contains the constant -100.
+\anto{this works only because we use SSI and thus the value of $i_{4}$ never
+changes. However, SSI is not explained anywhere in the paper}
 
@@
 values that the static object has (if the static object points to other static
-objects, those need to be lifted as well) This means that instead of the jump,
+objects, those need to be lifted as well, recursively) This means that instead of the jump,
 the following operations are emitted:
@@
 on the lifted static objects have been removed (some \texttt{get} operations
-and \texttt{guard\_class} operations).
+and \texttt{guard\_class} operations).  Moreover, in real life example usually
+the loops are more complex and contain more objects of type 1, thus this
+technique is more effective.
 
@@
 If the argument $v$ of a \texttt{get} operation is mapped to something in the static
-heap, the get can be performed at optimization time. Otherwise, the \texttt{get}
+heap, the \texttt{get} can be performed at optimization time. Otherwise, the \texttt{get}
 operation needs to be residualized.
@@
     \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
-    \item \textbf{telco}: A Python version of the Telco decimal.
+    \item \textbf{telco}: A Python version of the Telco decimal
     benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
@@
 
-We evaluate the allocation removal algorithm along two lines: First we want to
+We evaluate the allocation removal algorithm along two lines: first we want to
 know how many allocations could be optimized away. On the other hand, we want
@@
 taken together, the optimization removes XXX percent of allocation operations.
+\anto{Actually, we can only know how many operations we removed from the traces, not from the actual execution}
 
@@
 CPython which is a just-in-time compiler that produces machine code at run-time.
-It is not based on traces. Of PyPy's Python interpreter we used three versions,
+It is not based on traces. Finally, we used three versions of PyPy's Python interpreter:
 one without a JIT, one including the JIT but not using the allocation removal

### 2010-10-13 talk/pepm2011/paper.tex
fix a few of the fixes/remarks

@@
 called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
-it always ends with a jump to its own beginning \anto{this is not true: what if we trace the last iteration?}. The trace also contains all
+it always ends with a jump to its own beginning. The trace also contains all
 operations that are performed in functions that were called in the loop, thus a
@@
 \begin{tabular}{l}
-\# XXX: maybe we should specify that $p_{0}$, $p_{1}$ corresponds to y and res
+\# XXX: maybe we should specify that $p_{0}$, $p_{1}$ corresponds to y and res \\
 \# arguments to the trace: $p_{0}$, $p_{1}$ \\
@@
 correspond to the stack level of the function that contains the traced
-operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
+operation. The trace is in single-assignment form, meaning that each variable is
+assigned to exactly once. The trace also shows the inefficiencies of \texttt{f} clearly, if one
 looks at the number of \texttt{new} (corresponding to object creation),
@@
 was observed during tracing.\footnote{\texttt{guard\_class} performs a precise
-class check, not checking for subclasses} These guards make the trace specific
+class check, not checking for subclasses.} These guards make the trace specific
 to the situation where \texttt{y} is really a \texttt{BoxedInteger}, it can
@@
 \texttt{guard\_class} on a variable that has a shape description can be removed
-as well, because the shape description stores the type and thus the result of 
-the guard is statically known.
+as well, because the shape description stores the type and thus the outcome of
+the type check che guard does is statically known.
 
@@
 whose \texttt{intval} field contains the constant -100.
-\anto{this works only because we use SSI and thus the value of $i_{4}$ never
-changes. However, SSI is not explained anywhere in the paper}
 
@@
 on the lifted static objects have been removed (some \texttt{get} operations
-and \texttt{guard\_class} operations).  Moreover, in real life example usually
-the loops are more complex and contain more objects of type 1, thus this
-technique is more effective.
+and \texttt{guard\_class} operations).
 
@@
 seen in Figure~\ref{fig:numops}. The optimization removes as many as XXX and as
-little as XXX percent of allocation operations in the benchmarks. All benchmarks
-taken together, the optimization removes XXX percent of allocation operations.
-\anto{Actually, we can only know how many operations we removed from the traces, not from the actual execution}
+little as XXX percent of allocation operations in the traces of the benchmarks.
+All benchmarks taken together, the optimization removes XXX percent of
+allocation operations.
 

### 2010-10-13 talk/pepm2011/paper.tex
move code around, another fix

@@
         return self.floatval > 0.0
+
+def f(y):
+    res = BoxedInteger(0)
+    while y.is_positive():
+        res = res.add(y).add(BoxedInteger(-100))
+        y = y.add(BoxedInteger(-1))
+    return res
 \end{verbatim}
-\caption{A simple object model}
+\caption{A Simple Object Model and an Example Function Using it}
 \label{fig:objmodel}
@@
 
-To understand the problems more directly, let us consider a simple function
-that uses the object model:
+To understand the problems more directly, let us consider the simple function
+\texttt{f} that uses the object model (see the bottom of
+Figure~\ref{fig:objmodel}).
 
@@
 
-\begin{verbatim}
-def f(y):
-    res = BoxedInteger(0)
-    while y.is_positive():
-        res = res.add(y).add(BoxedInteger(-100))
-        y = y.add(BoxedInteger(-1))
-    return res
-\end{verbatim}
-
-The loop iterates \texttt{y} times, and computes something in the process.
+The loop in \texttt{f} iterates \texttt{y} times, and computes something in the process.
 Simply running this function is slow, because there are lots of virtual method
@@
 \begin{tabular}{l}
-\# XXX: maybe we should specify that $p_{0}$, $p_{1}$ corresponds to y and res \\
 \# arguments to the trace: $p_{0}$, $p_{1}$ \\
@@
 operation. The trace is in single-assignment form, meaning that each variable is
-assigned to exactly once. The trace also shows the inefficiencies of \texttt{f} clearly, if one
+assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
+to the live variables \texttt{y} and \texttt{res} in the original function.
+
+The trace shows the inefficiencies of \texttt{f} clearly, if one
 looks at the number of \texttt{new} (corresponding to object creation),

### 2010-10-13 talk/pepm2011/paper.tex
typo

@@
 as well, because the shape description stores the type and thus the outcome of
-the type check che guard does is statically known.
+the type check the guard does is statically known.
 

### 2010-10-13 talk/pepm2011/paper.tex
(antocuni, cfbolz) a minimal fix for one of antos remarks that I just killed
some changes to Sect 1

@@
 languages is that of a tracing JIT. A tracing JIT works by observing the running
-program and recording its hot parts into linear execution traces, which are then turned into
+program and recording its hot spots into linear execution traces, which are then turned into
 machine code. One reason for the popularity of tracing JITs is their relative
@@
 The usage of a tracing JIT can remove the overhead of bytecode dispatch and that
-of the interpreter data structures. In this paper we want to present an approach
-how an optimization can be added to a tracing JIT that also removes some of the
+of the interpreter data structures. In this paper we want to present a new
+optimization that can be added to a tracing JIT that further removes some of the
 overhead more closely associated to dynamic languages, such as boxing overhead
@@
 
-The most important technique we use to achieve to optimize traces is a
-straightforward application of partial evaluation. The partial evaluation
+The core of our trace optimization technique can be
+viewed as partial evaluation: the partial evaluation
 performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
@@
 interpreter enters a special \emph{tracing mode}. In this tracing mode, the
-interpreter records all operations that it is executing while running one
+interpreter tries to record all operations that it is executing while running one
 iteration of the hot loop. This history of executed operations of one loop is

### 2010-10-13 talk/pepm2011/paper.tex
some more changes

@@
 \cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
-be implemented in a simple yet efficient way. The approach taken when
-implementing a language with PyPy is to write an \emph{interpreter}
+be implemented in a simple yet efficient way. 
+When implementing a language with PyPy one writes an \emph{interpreter}
 for the language in
@@
 
-A number of languages have been implemented with PyPy. The project was started
+A number of languages have been implemented with PyPy. The project was initiated
 to get a better Python implementation, which inspired the name of the project
@@
 language that the interpreter is implementing. This process is mostly
-automatic; it only needs to be guided by the language implementer by a small number of
+automatic; it only needs to be guided by the language implementer using a small number of
 source-code hints. Mostly-automatically generating a JIT compiler has many advantages
@@
 experimental one for the CLI .NET Virtual Machine \cite{cuni_python_cli_2010}.
-The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
+PyPy's JIT generator generates a \emph{tracing JIT
 compiler}, a concept which we now explain in more details.
@@
 This trace of operations is then the basis of the generated code. The trace is
-optimized in some ways, and then turned into machine code. Both optimizations
+first optimized, and then turned into machine code. Both optimizations
 and machine code generation is simple, because the traces are linear. This
@@
 
-For the purpose of this paper, we are going to use a very simple object
+For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
+ a very simple object
 model, that just supports an integer and a float type. The objects support only
@@
 than zero. The implementation of \texttt{add} uses classical Smalltalk-like
-double-dispatching. These classes could be part of the implementation of a very
-simple interpreter written in RPython. The classes can be seen in
-Figure~\ref{fig:objmodel}.
-
-\begin{figure}
-\begin{verbatim}
+double-dispatching.
+%These classes could be part of the implementation of a very
+%simple interpreter written in RPython.
+The classes can be seen in
+Figure~\ref{fig:objmodel} (written in RPython).
+
+\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
+\ignore{
 class Base(object):
@@
         raise NotImplementedError("abstract base")
+        }
+\begin{figure}
+\begin{verbatim}
+class Base(object):
 
@@
 \end{verbatim}
-\caption{A Simple Object Model and an Example Function Using it}
+\caption{An ``interpreter'' for a tiny Dynamic Language written in RPython}
+%\caption{A Simple Object Model and an Example Function Using it}
 \label{fig:objmodel}
@@
 
-To understand the problems more directly, let us consider the simple function
+To understand the problems more directly, let us consider the simple
+interpreter function
 \texttt{f} that uses the object model (see the bottom of
@@
 \texttt{BoxedInteger}, the produced trace looks like
-Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
+Figure~\ref{fig:unopt-trace} (lines starting with the hash ``\#'' are comments).
+
+XXX in which language is the trace written in ? still RPython ?
+
+The operations in the trace are shown indented to
 correspond to the stack level of the function that contains the traced
@@
 to the live variables \texttt{y} and \texttt{res} in the original function.
+
+XXX explain set and get + int_add briefly
 

### 2010-10-13 talk/pepm2011/paper.tex
make the code nicer

@@
 \usepackage[utf8]{inputenc}
+\usepackage{setspace}
+
+\usepackage{listings}
+
+\usepackage[T1]{fontenc}
+\usepackage{beramono}
+
+\definecolor{gray}{rgb}{0.3,0.3,0.3}
+
+\lstset{
+  basicstyle=\setstretch{1.1}\ttfamily\footnotesize,
+  language=Python,
+  keywordstyle=\bfseries,
+  stringstyle=\color{blue},
+  commentstyle=\color{gray}\textit,
+  fancyvrb=true,
+  showstringspaces=false,
+  keywords={def,while,if,elif,return,class,get,set,new,guard_class}
+}
 
@@
 path. Those guards check the assumptions under which execution can stay on the
-trace. As an example, if a loop contains an \texttt{if} statement, the trace
+trace. As an example, if a loop contains an \lstinline{if} statement, the trace
 will contain the execution of one of the paths only, which is the path that was
 taken during the production of the trace. The trace will also contain a guard
-that checks that the condition of the \texttt{if} statement is true, because if
+that checks that the condition of the \lstinline{if} statement is true, because if
 it isn't, the rest of the trace is not valid.
@@
 model, that just supports an integer and a float type. The objects support only
-two operations, \texttt{add}, which adds two objects (promoting ints to floats in a
-mixed addition) and \texttt{is\_positive}, which returns whether the number is greater
-than zero. The implementation of \texttt{add} uses classical Smalltalk-like
+two operations, \lstinline{add}, which adds two objects (promoting ints to floats in a
+mixed addition) and \lstinline{is_positive}, which returns whether the number is greater
+than zero. The implementation of \lstinline{add} uses classical Smalltalk-like
 double-dispatching.
@@
 \begin{figure}
-\begin{verbatim}
+\begin{lstlisting}[mathescape]
 class Base(object):
+      ...
 
@@
       self.intval = intval
+
    def add(self, other):
       return other.add__int(self.intval)
+
    def add__int(self, intother):
       return BoxedInteger(intother + self.intval)
+
    def add__float(self, floatother):
@@
       return BoxedFloat(floatvalue)
+
    def is_positive(self):
@@
       self.floatval = floatval
+
    def add(self, other):
       return other.add__float(self.floatval)
+
    def add__int(self, intother):
@@
       return BoxedFloat(floatvalue)
+
    def add__float(self, floatother):
       return BoxedFloat(floatother + self.floatval)
+
    def is_positive(self):
@@
    return res
-\end{verbatim}
+\end{lstlisting}
 \caption{An ``interpreter'' for a tiny Dynamic Language written in RPython}
@@
 dynamic language implementation has. All the numbers are instances of either
-\texttt{BoxedInteger} or \texttt{BoxedFloat}, thus they consume space on the
+\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, thus they consume space on the
 heap. Performing many arithmetic operations produces lots of garbage quickly,
@@
 interpreter function
-\texttt{f} that uses the object model (see the bottom of
+\lstinline{f} that uses the object model (see the bottom of
 Figure~\ref{fig:objmodel}).
@@
 
-The loop in \texttt{f} iterates \texttt{y} times, and computes something in the process.
+The loop in \lstinline{f} iterates \lstinline{y} times, and computes something in the process.
 Simply running this function is slow, because there are lots of virtual method
-calls inside the loop, one for each \texttt{is\_positive} and even two for each
-call to \texttt{add}. These method calls need to check the type of the involved
+calls inside the loop, one for each \lstinline{is_positive} and even two for each
+call to \lstinline{add}. These method calls need to check the type of the involved
 objects repeatedly and redundantly. In addition, a lot of objects are created
 when executing that loop, many of these objects do not survive for very long.
-The actual computation that is performed by \texttt{f} is simply a sequence of
+The actual computation that is performed by \lstinline{f} is simply a sequence of
 float or integer additions.
@@
 \begin{figure}
-\texttt{
-\begin{tabular}{l}
-\# arguments to the trace: $p_{0}$, $p_{1}$ \\
-\# inside f: res.add(y) \\
-guard\_class($p_{1}$, BoxedInteger) \\
-~~~~\# inside BoxedInteger.add \\
-~~~~$i_{2}$ = get($p_{1}$, intval) \\
-~~~~guard\_class($p_{0}$, BoxedInteger) \\
-~~~~~~~~\# inside BoxedInteger.add\_\_int \\
-~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
-~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
-~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
-~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
-\# inside f: BoxedInteger(-100)  \\
-$p_{6}$ = new(BoxedInteger) \\
-~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~set($p_{6}$, intval, -100) \\
-~\\
-\# inside f: .add(BoxedInteger(-100)) \\
-guard\_class($p_{5}$, BoxedInteger) \\
-~~~~\# inside BoxedInteger.add \\
-~~~~$i_{7}$ = get($p_{5}$, intval) \\
-~~~~guard\_class($p_{6}$, BoxedInteger) \\
-~~~~~~~~\# inside BoxedInteger.add\_\_int \\
-~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
-~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
-~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
-~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
-~\\
-\# inside f: BoxedInteger(-1) \\
-$p_{11}$ = new(BoxedInteger) \\
-~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~set($p_{11}$, intval, -1) \\
-~\\
-\# inside f: y.add(BoxedInteger(-1)) \\
-guard\_class($p_{0}$, BoxedInteger) \\
-~~~~\# inside BoxedInteger.add \\
-~~~~$i_{12}$ = get($p_{0}$, intval) \\
-~~~~guard\_class($p_{11}$, BoxedInteger) \\
-~~~~~~~~\# inside BoxedInteger.add\_\_int \\
-~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
-~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
-~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
-~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
-~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
-~\\
-\# inside f: y.is\_positive() \\
-guard\_class($p_{15}$, BoxedInteger) \\
-~~~~\# inside BoxedInteger.is\_positive \\
-~~~~$i_{16}$ = get($p_{15}$, intval) \\
-~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
-\# inside f \\
-guard\_true($i_{17}$) \\
-jump($p_{15}$, $p_{10}$) \\
-\end{tabular}
-}
+\begin{lstlisting}[mathescape]
+# arguments to the trace: $p_{0}$, $p_{1}$
+# inside f: res.add(y)
+guard_class($p_{1}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{2}$ = get($p_{1}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{3}$ = get($p_{0}$, intval)
+        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
+        $p_{5}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{5}$, intval, $i_{4}$)
+
+# inside f: BoxedInteger(-100) 
+$p_{6}$ = new(BoxedInteger)
+    # inside BoxedInteger.__init__
+    set($p_{6}$, intval, -100)
+
+# inside f: .add(BoxedInteger(-100))
+guard_class($p_{5}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{7}$ = get($p_{5}$, intval)
+    guard_class($p_{6}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{8}$ = get($p_{6}$, intval)
+        $i_{9}$ = int_add($i_{7}$, $i_{8}$)
+        $p_{10}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{10}$, intval, $i_{9}$)
+
+# inside f: BoxedInteger(-1)
+$p_{11}$ = new(BoxedInteger)
+    # inside BoxedInteger.__init__
+    set($p_{11}$, intval, -1)
+
+# inside f: y.add(BoxedInteger(-1))
+guard_class($p_{0}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{12}$ = get($p_{0}$, intval)
+    guard_class($p_{11}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{13}$ = get($p_{11}$, intval)
+        $i_{14}$ = int_add($i_{12}$, $i_{13}$)
+        $p_{15}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{15}$, intval, $i_{14}$)
+
+# inside f: y.is_positive()
+guard_class($p_{15}$, BoxedInteger)
+    # inside BoxedInteger.is_positive
+    $i_{16}$ = get($p_{15}$, intval)
+    $i_{17}$ = int_gt($i_{16}$, 0)
+# inside f
+guard_true($i_{17}$)
+jump($p_{15}$, $p_{10}$)
+\end{lstlisting}
 \caption{Unoptimized Trace for the Simple Object Model}
@@
 
-If the function is executed using the tracing JIT, with \texttt{y} being a
-\texttt{BoxedInteger}, the produced trace looks like
+If the function is executed using the tracing JIT, with \lstinline{y} being a
+\lstinline{BoxedInteger}, the produced trace looks like
 Figure~\ref{fig:unopt-trace} (lines starting with the hash ``\#'' are comments).
@@
 assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
-to the live variables \texttt{y} and \texttt{res} in the original function.
-
-XXX explain set and get + int_add briefly
-
-The trace shows the inefficiencies of \texttt{f} clearly, if one
-looks at the number of \texttt{new} (corresponding to object creation),
-\texttt{set/get} (corresponding to attribute reads/writes) and
-\texttt{guard\_class} operations (corresponding to method calls).
+to the live variables \lstinline{y} and \lstinline{res} in the original function.
+
+XXX explain set and get + int\_add briefly
+
+The trace shows the inefficiencies of \lstinline{f} clearly, if one
+looks at the number of \lstinline{new} (corresponding to object creation),
+\lstinline{set/get} (corresponding to attribute reads/writes) and
+\lstinline{guard_class} operations (corresponding to method calls).
 In the rest of the paper we will see how this trace can be optimized using
@@
 
-Note how the functions that are called by \texttt{f} are automatically inlined
-into the trace. The method calls are always preceded by a \texttt{guard\_class}
+Note how the functions that are called by \lstinline{f} are automatically inlined
+into the trace. The method calls are always preceded by a \lstinline{guard_class}
 operation, to check that the class of the receiver is the same as the one that
-was observed during tracing.\footnote{\texttt{guard\_class} performs a precise
+was observed during tracing.\footnote{\lstinline{guard_class} performs a precise
 class check, not checking for subclasses.} These guards make the trace specific
-to the situation where \texttt{y} is really a \texttt{BoxedInteger}, it can
-already be said to be specialized for \texttt{BoxedIntegers}. When the trace is
-turned into machine code and then executed with \texttt{BoxedFloats}, the
-first \texttt{guard\_class} instruction will fail and execution will continue
+to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}, it can
+already be said to be specialized for \lstinline{BoxedIntegers}. When the trace is
+turned into machine code and then executed with \lstinline{BoxedFloats}, the
+first \lstinline{guard_class} instruction will fail and execution will continue
 using the interpreter.
@@
 
-Some of the operations within this trace are \texttt{new} operations, which each
+Some of the operations within this trace are \lstinline{new} operations, which each
 create a new instance of some class. These instances are used for a while, e.g.
@@
 
-Together with the \texttt{new} operations, the figure shows the lifetimes of the
-created objects. The objects that are created within a trace using \texttt{new}
+Together with the \lstinline{new} operations, the figure shows the lifetimes of the
+created objects. The objects that are created within a trace using \lstinline{new}
 fall into one of several categories:
@@
 The partial evaluation works by walking the trace from beginning to end.
-Whenever a \texttt{new} operation is seen, the operation is removed and a static
+Whenever a \lstinline{new} operation is seen, the operation is removed and a static
 object is constructed and associated with the variable that would have stored
-the result of \texttt{new}. The static object describes the shape of the
+the result of \lstinline{new}. The static object describes the shape of the
 original object, \eg where the values that would be stored in the fields of the
 allocated object come from, as well as the type of the object. Whenever the
-optimizer sees a \texttt{set} that writes into such an object, that shape
+optimizer sees a \lstinline{set} that writes into such an object, that shape
 description is updated and the operation can be removed, which means that the
 operation was done at partial evaluation time. When the optimizer encounters a
-\texttt{get} from such an object, the result is read from the shape
+\lstinline{get} from such an object, the result is read from the shape
 description, and the operation is also removed. Equivalently, a
-\texttt{guard\_class} on a variable that has a shape description can be removed
+\lstinline{guard_class} on a variable that has a shape description can be removed
 as well, because the shape description stores the type and thus the outcome of
@@
 
-\texttt{
-\begin{tabular}{l}
-$p_{5}$ = new(BoxedInteger) \\
-set($p_{5}$, intval, $i_{4}$) \\
-$p_{6}$ = new(BoxedInteger) \\
-set($p_{6}$, intval, -100) \\
-\end{tabular}
-}
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+$p_{5}$ = new(BoxedInteger)
+set($p_{5}$, intval, $i_{4}$)
+$p_{6}$ = new(BoxedInteger)
+set($p_{6}$, intval, -100)
+\end{lstlisting}
 
 The static object associated with $p_{5}$ would know that it is a
-\texttt{BoxedInteger} whose \texttt{intval} field contains $i_{4}$; the
-one associated with $p_{6}$ would know that it is a \texttt{BoxedInteger}
-whose \texttt{intval} field contains the constant -100.
+\lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
+one associated with $p_{6}$ would know that it is a \lstinline{BoxedInteger}
+whose \lstinline{intval} field contains the constant -100.
 
@@
 
-\texttt{
-\begin{tabular}{l}
-guard\_class($p_{5}$, BoxedInteger) \\
-$i_{7}$ = get($p_{5}$, intval) \\
-\# inside BoxedInteger.add \\
-guard\_class($p_{6}$, BoxedInteger) \\
-\# inside BoxedInteger.add\_\_int \\
-$i_{8}$ = get($p_{6}$, intval) \\
-$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
-\end{tabular}
-}
-
-The \texttt{guard\_class} operations can be removed, because the classes of $p_{5}$ and
-$p_{6}$ are known to be \texttt{BoxedInteger}. The \texttt{get} operations can be removed
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+guard_class($p_{5}$, BoxedInteger)
+$i_{7}$ = get($p_{5}$, intval)
+# inside BoxedInteger.add
+guard_class($p_{6}$, BoxedInteger)
+# inside BoxedInteger.add__int
+$i_{8}$ = get($p_{6}$, intval)
+$i_{9}$ = int_add($i_{7}$, $i_{8}$)
+\end{lstlisting}
+
+The \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
+$p_{6}$ are known to be \lstinline{BoxedInteger}. The \lstinline{get} operations can be removed
 and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
@@
 
-\texttt{
-\begin{tabular}{l}
-$i_{9}$ = int\_add($i_{4}$, -100) \\
-\end{tabular}
-}
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+$i_{9}$ = int_add($i_{4}$, -100)
+\end{lstlisting}
 
@@
 
-This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
+This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
 is hit. The arguments of the jump are at this point static objects. Before the
@@
 
-\texttt{
-\begin{tabular}{l}
-$p_{15}$ = new(BoxedInteger) \\
-set($p_{15}$, intval, $i_{14}$) \\
-$p_{10}$ = new(BoxedInteger) \\
-set($p_{10}$, intval, $i_{9}$) \\
-jump($p_{15}$, $p_{10}$) \\
-\end{tabular}
-}
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+$p_{15}$ = new(BoxedInteger)
+set($p_{15}$, intval, $i_{14}$)
+$p_{10}$ = new(BoxedInteger)
+set($p_{10}$, intval, $i_{9}$)
+jump($p_{15}$, $p_{10}$)
+\end{lstlisting}
 
@@
 worthwhile even in this case, because some operations that have been performed
-on the lifted static objects have been removed (some \texttt{get} operations
-and \texttt{guard\_class} operations).
+on the lifted static objects have been removed (some \lstinline{get} operations
+and \lstinline{guard_class} operations).
 
@@
 The optimized trace contains only two allocations, instead of the original five,
-and only three \texttt{guard\_class} operations, from the original seven.
+and only three \lstinline{guard_class} operations, from the original seven.
 
@@
 
-Traces are lists of operations. The operations considered here are \texttt{new} (to make
-a new object), \texttt{get} (to read a field out of an object), \texttt{set} (to write a field
-into an object) and \texttt{guard\_class} (to check the type of an object). The values of all
+Traces are lists of operations. The operations considered here are \lstinline{new} (to make
+a new object), \lstinline{get} (to read a field out of an object), \lstinline{set} (to write a field
+into an object) and \lstinline{guard_class} (to check the type of an object). The values of all
 variables are locations (i.e.~pointers). Locations are mapped to objects, which
@@
 
-The \texttt{new} operation creates a new object $(T,\mathrm{null},\mathrm{null})$ on the
+The \lstinline{new} operation creates a new object $(T,\mathrm{null},\mathrm{null})$ on the
 heap under a fresh location $l$ and adds the result variable to the environment,
@@
 
-The \texttt{get} operation reads a field $F$ out of an object, and adds the result
+The \lstinline{get} operation reads a field $F$ out of an object, and adds the result
 variable to the environment, mapping it to the read location. The heap is
@@
 
-The \texttt{set} operation changes field $F$ of an object stored at the location that
+The \lstinline{set} operation changes field $F$ of an object stored at the location that
 variable $v$ maps to. The new value of the field is the location in variable
@@
 
-The \texttt{guard\_class} operation is used to check whether the object stored at the location
+The \lstinline{guard_class} operation is used to check whether the object stored at the location
 that variable $v$ maps to is of type $T$. If that is the case, then execution
@@
 
-When the optimizer sees a \texttt{new} operation, it optimistically removes it and
+When the optimizer sees a \lstinline{new} operation, it optimistically removes it and
 assumes that the resulting object can stay static. The optimization for all
@@
 
-If the argument $v$ of a \texttt{get} operation is mapped to something in the static
-heap, the \texttt{get} can be performed at optimization time. Otherwise, the \texttt{get}
+If the argument $v$ of a \lstinline{get} operation is mapped to something in the static
+heap, the \lstinline{get} can be performed at optimization time. Otherwise, the \lstinline{get}
 operation needs to be residualized.
 
-If the first argument $v$ to a \texttt{set} operation is mapped to something in the
-static heap, then the \texttt{set} can performed at optimization time and the static heap
-updated. Otherwise the \texttt{set} operation needs to be residualized. This needs to be
+If the first argument $v$ to a \lstinline{set} operation is mapped to something in the
+static heap, then the \lstinline{set} can performed at optimization time and the static heap
+updated. Otherwise the \lstinline{set} operation needs to be residualized. This needs to be
 done carefully, because the new value for the field, from the variable $u$,
@@
 
-If a \texttt{guard\_class} is performed on a variable that is in the static heap, the type check
+If a \lstinline{guard_class} is performed on a variable that is in the static heap, the type check
 can be performed at optimization time, which means the operation can be removed
 if the types match. If the type check fails statically or if the object is not
-in the static heap, the \texttt{guard\_class} is residualized. This also needs to
-lift the variable on which the \texttt{guard\_class} is performed.
+in the static heap, the \lstinline{guard_class} is residualized. This also needs to
+lift the variable on which the \lstinline{guard_class} is performed.
 
@@
 static heap can be cyclic. To make sure that the same static object is not lifted
-twice, the \texttt{liftfield} operation removes it from the static heap \emph{before}
+twice, the \lstinline{liftfield} operation removes it from the static heap \emph{before}
 recursively lifting its fields.

### 2010-10-13 talk/pepm2011/paper.tex
don't highlight get/set/new/guard_class in bold

@@
   showstringspaces=false,
-  keywords={def,while,if,elif,return,class,get,set,new,guard_class}
+  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
 }

### 2010-10-13 talk/pepm2011/paper.tex
Try to solve the XXXs of Michael. Add a new figure with the operation numbers.

@@
 \newcommand{\commentout}[1]{}
+\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
 
@@
 additional JIT backends are the one for Intel \emph{x86-64} and an
-experimental one for the CLI .NET Virtual Machine \cite{cuni_python_cli_2010}.
+experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.
 PyPy's JIT generator generates a \emph{tracing JIT
@@
 This trace of operations is then the basis of the generated code. The trace is
-first optimized, and then turned into machine code. Both optimizations
+first optimized, and then turned into machine code. Both optimization
 and machine code generation is simple, because the traces are linear. This
@@
 
-\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
-\ignore{
-class Base(object):
-    def add(self, other):
-        """ add self to other """
-        raise NotImplementedError("abstract base")
-    def add__int(self, intother):
-        """ add intother to self,
-            where intother is an integer """
-        raise NotImplementedError("abstract base")
-    def add__float(self, floatother):
-        """ add floatother to self,
-            where floatother is a float """
-        raise NotImplementedError("abstract base")
-    def is_positive(self):
-        """ returns whether self is positive """
-        raise NotImplementedError("abstract base")
-        }
 \begin{figure}
@@
 class Base(object):
-      ...
+   pass
 
@@
 \end{lstlisting}
-\caption{An ``interpreter'' for a tiny Dynamic Language written in RPython}
-%\caption{A Simple Object Model and an Example Function Using it}
+\caption{An ``Interpreter'' for a Tiny Dynamic Language Written in RPython}
 \label{fig:objmodel}
@@
 To understand the problems more directly, let us consider the simple
-interpreter function
-\lstinline{f} that uses the object model (see the bottom of
+interpreter function \lstinline{f} that uses the object model (see the bottom of
 Figure~\ref{fig:objmodel}).
-
-XXX this is not an RPython interpreter; put a reference to the previous
-paper to show how we deal with an interpreted piece of code and remove
-the interpretation overhead, turning it into basically something
-equivalent to the example here, which is the start of the present paper.
 
@@
 
-XXX in which language is the trace written in ? still RPython ?
-
 The operations in the trace are shown indented to
@@
 
-XXX explain set and get + int\_add briefly
-
-The trace shows the inefficiencies of \lstinline{f} clearly, if one
-looks at the number of \lstinline{new} (corresponding to object creation),
-\lstinline{set/get} (corresponding to attribute reads/writes) and
-\lstinline{guard_class} operations (corresponding to method calls).
-In the rest of the paper we will see how this trace can be optimized using
-partial evaluation.
+The operations in the trace correspond to the operations in the RPython program
+in Figure~\ref{fig:objmodel}:
+
+\begin{itemize}
+    \item \lstinline{new} corresponds to object creation.
+    \item \lstinline{get} correspond to attribute reads.
+    \item \lstinline{set} correspond to attribute writes.
+    \item \lstinline{guard\_class} correspond to method calls and are followed by
+    the trace of the called method.
+    \item \lstinline{int\_add} and \lstinline{int\_get} are integer addition and
+    comparison (``greater than''), respectively.
+\end{itemize}
 
@@
 
+The trace shows the inefficiencies of \lstinline{f} clearly, if one looks at the
+number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard\_class} operations.
+In the rest of the paper we will see how this trace can be optimized using
+partial evaluation.
+
 \section{Object Lifetimes in a Tracing JIT}
@@
 bit peculiar in that there are not actually any constant arguments to the trace,
-but it is only used to optimized operations within a trace. XXX mention Prolog.
+but it is only used to optimized operations within a trace.
 
@@
 \begin{figure*}
-\begin{tabular}{lrrrrrrrrrrrrrrrrrrrrrrr}
-	&\#loops	&new	&	&	&get	&	&	&set	&	&	&guard	&	&	&rest	&	&	&	&	&	&total	&	&\\
-crypto\_pyaes	&75	&1663	&446	&27\%	&18842	&2127	&11\%	&14552	&2020	&14\%	&8952	&235	&3\%	&53339	&7794	&15\%	&44009	&4828	&11\%	&97348	&12622	&13\%\\
-django	&53	&453	&155	&34\%	&11970	&1457	&12\%	&8738	&991	&11\%	&4093	&275	&7\%	&33988	&4076	&12\%	&25254	&2878	&11\%	&59242	&6954	&12\%\\
-fannkuch	&40	&161	&76	&47\%	&574	&306	&53\%	&278	&139	&50\%	&1121	&202	&18\%	&2581	&1685	&65\%	&2134	&723	&34\%	&4715	&2408	&51\%\\
-go	&510	&5604	&1651	&29\%	&112403	&12612	&11\%	&88439	&11915	&13\%	&51621	&3194	&6\%	&300831	&30834	&10\%	&258067	&29372	&11\%	&558898	&60206	&11\%\\
-html5lib	&416	&5939	&1793	&30\%	&241402	&15863	&7\%	&178723	&13960	&8\%	&58486	&3037	&5\%	&682654	&40341	&6\%	&484550	&34653	&7\%	&1167204	&74994	&6\%\\
-meteor-contest	&56	&221	&141	&64\%	&2326	&606	&26\%	&2047	&645	&32\%	&1040	&169	&16\%	&7018	&2250	&32\%	&5634	&1561	&28\%	&12652	&3811	&30\%\\
-nbody	&10	&84	&48	&57\%	&287	&159	&55\%	&121	&74	&61\%	&411	&90	&22\%	&986	&737	&75\%	&903	&371	&41\%	&1889	&1108	&59\%\\
-pyflate-fast	&161	&1878	&723	&38\%	&22815	&2592	&11\%	&18415	&3690	&20\%	&8291	&628	&8\%	&64649	&11327	&18\%	&51399	&7633	&15\%	&116048	&18960	&16\%\\
-raytrace-simple	&117	&1906	&359	&19\%	&52995	&2202	&4\%	&39518	&1823	&5\%	&13451	&447	&3\%	&140723	&6533	&5\%	&107870	&4831	&4\%	&248593	&11364	&5\%\\
-richards	&79	&375	&191	&51\%	&27507	&2095	&8\%	&21217	&2004	&9\%	&3972	&332	&8\%	&76751	&6171	&8\%	&53071	&4622	&9\%	&129822	&10793	&8\%\\
-spambayes	&267	&2413	&505	&21\%	&53327	&3033	&6\%	&40938	&2428	&6\%	&19343	&718	&4\%	&143575	&10322	&7\%	&116021	&6684	&6\%	&259596	&17006	&7\%\\
-spectral-norm	&35	&266	&88	&33\%	&3004	&398	&13\%	&2544	&414	&16\%	&1084	&78	&7\%	&4803	&1184	&25\%	&6898	&978	&14\%	&11701	&2162	&18\%\\
-telco	&42	&660	&70	&11\%	&20716	&424	&2\%	&14444	&608	&4\%	&6574	&56	&1\%	&53674	&1224	&2\%	&42394	&1158	&3\%	&96068	&2382	&2\%\\
-twisted-names	&201	&2706	&317	&12\%	&64438	&2795	&4\%	&47960	&1968	&4\%	&24091	&555	&2\%	&171582	&6828	&4\%	&139195	&5635	&4\%	&310777	&12463	&4\%\\
-total	&2062	&24329	&6563	&27\%	&632606	&46669	&7\%	&477934	&42679	&9\%	&202530	&10016	&5\%	&1737154	&131306	&8\%	&1337399	&105927	&8\%	&3074553	&237233	&8\%\\
+\begin{tabular}{|l||r|rr|rr|rr|rr|}
+\hline
+	&num loops	&new	&removed	&get/set	&removed &guard &removed	&all ops	&removed\\
+\hline
+crypto\_pyaes	&78	&3088	&50\%	&57148	&25\%	&9055	&95\%	&137189	&80\%\\
+django	&51	&673	&54\%	&19318	&18\%	&3876	&93\%	&55682	&85\%\\
+fannkuch	&43	&171	&49\%	&886	&63\%	&1159	&81\%	&4935	&45\%\\
+go	&517	&12234	&76\%	&200842	&21\%	&53138	&90\%	&568542	&84\%\\
+html5lib	&498	&14432	&68\%	&503390	&11\%	&71592	&94\%	&1405780	&91\%\\
+meteor-contest	&59	&277	&36\%	&4402	&31\%	&1078	&83\%	&12862	&68\%\\
+nbody	&13	&96	&38\%	&443	&69\%	&449	&78\%	&2107	&38\%\\
+pyflate-fast	&162	&2278	&55\%	&39126	&20\%	&8194	&92\%	&112857	&80\%\\
+raytrace-simple	&120	&3118	&59\%	&91982	&15\%	&13572	&95\%	&247436	&89\%\\
+richards	&87	&844	&4\%	&49875	&22\%	&4130	&91\%	&133898	&83\%\\
+spambayes	&314	&5608	&79\%	&117002	&11\%	&25313	&94\%	&324125	&90\%\\
+spectral-norm	&38	&360	&64\%	&5553	&20\%	&1122	&92\%	&11878	&77\%\\
+telco	&46	&1257	&90\%	&37470	&3\%	&6644	&99\%	&98590	&97\%\\
+twisted-names	&214	&5273	&84\%	&100010	&10\%	&23247	&96\%	&279667	&92\%\\
+\hline
+total	&2240	&49709	&70\%	&1227447	&14\%	&222569	&93\%	&3395548	&89\%\\
+\hline
 \end{tabular}
-\caption{Number of Operations Before and After Optimization}
+\caption{Number of Operations and Percentage Removed By Optimization}
 \label{fig:numops}

### 2010-10-13 talk/pepm2011/paper.tex
improve benchmark section

@@
 
-For the former we counted the occurring operations in all generated traces
-before and after the optimization phase for all benchmarks. The results can be
-seen in Figure~\ref{fig:numops}. The optimization removes as many as XXX and as
-little as XXX percent of allocation operations in the traces of the benchmarks.
-All benchmarks taken together, the optimization removes XXX percent of
-allocation operations.
+The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
+with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
+2.6.35. We compared the performance of various Python implementations on the
+benchmarks. As a baseline, we used the standard Python implementation in C,
+CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
+interpreter. Furthermore we compared against Psyco 1.6
+\cite{rigo_representation-based_2004}, an extension to CPython which is a
+just-in-time compiler that produces machine code at run-time. It is not based
+on traces. Finally, we used two versions of PyPy's Python interpreter (revision
+77823 of SVN trunk\footnote{\texttt{https://codespeak.net/svn/pypy/trunk}}): one
+including the JIT but not optimizing the traces, and one using the allocation
+removal optimizations (as well as some minor other optimizations, such as
+constant folding).
+
+As the first step, we counted the occurring operations in all generated traces
+before and after the optimization phase for all benchmarks. The resulting
+numbers can be
+seen in Figure~\ref{fig:numops}. The optimization removes as many as 90\% and as
+little as 4\% percent of allocation operations in the traces of the benchmarks.
+All benchmarks taken together, the optimization removes 70\% percent of
+allocation operations. The numbers look similar for reading and writing of
+attributes. There are even more \lstinline{guard} operations that are removed,
+however there is an additional optimization that removes guards, so not all the
+removed guards are an effect of the optimization described here.
 
@@
 
-In addition to the count of operations we also performed time measurements. The
-machine the benchmarks were run on is XXX. We compared the performance of
-various Python implementations on the benchmarks. As a baseline, we used the
-standard Python implementation in C, called
-CPython\footnote{\texttt{http://python.org}}, which uses a bytecode-based
-interpreter. Furthermore we compared against Psyco \cite{rigo_representation-based_2004}, an extension to
-CPython which is a just-in-time compiler that produces machine code at run-time.
-It is not based on traces. Finally, we used three versions of PyPy's Python interpreter:
-one without a JIT, one including the JIT but not using the allocation removal
-optimization, and one using the allocation removal optimizations.
-
+In addition to the count of operations we also performed time measurements.
 All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
@@
 in all benchmarks except spambayes (which heavily relies on regular expression
-performance). All benchmarks are improved by the allocation removal
+performance XXX). All benchmarks are improved by the allocation removal
 optimization, some as much as XXX. XXX Psyco

### 2010-10-13 talk/pepm2011/paper.tex
re-import benchmark times, fix one XXX

@@
 \begin{figure*}
+\begin{center}
 \begin{tabular}{|l||r|rr|rr|rr|rr|}
@@
 \end{tabular}
+\end{center}
 \caption{Number of Operations and Percentage Removed By Optimization}
@@
 in all benchmarks except spambayes (which heavily relies on regular expression
-performance XXX). All benchmarks are improved by the allocation removal
-optimization, some as much as XXX. XXX Psyco
+performance and thus is not helped much by our Python JIT) and meteor-contest.
+All benchmarks are improved by the allocation removal optimization, some as much
+as XXX. XXX Psyco
 
@@
 \begin{figure*}
-\begin{tabular}{lrrrr}
-		&CPython	&JIT no optimizations	&JIT allocation removal	&JIT full\\
-crypto\_pyaes	&1958.40 $\pm$ 0.32	&782.14 $\pm$ 3.08	&150.86 $\pm$ 3.62	&125.68 $\pm$ 3.02\\
-django		&630.91 $\pm$ 0.13	&365.90 $\pm$ 1.57	&124.41 $\pm$ 0.88	&117.36 $\pm$ 0.91\\
-fannkuch	&1344.15 $\pm$ 0.36	&378.18 $\pm$ 0.36	&309.04 $\pm$ 1.56	&299.12 $\pm$ 0.25\\
-go		&595.31 $\pm$ 0.51	&1067.26 $\pm$ 12.20	&121.76 $\pm$ 4.15	&125.57 $\pm$ 3.90\\
-html5lib	&8678.67 $\pm$ 23.37	&17665.16 $\pm$ 2695.45	&6739.95 $\pm$ 1225.25	&6699.24 $\pm$ 1295.92\\
-meteor-contest	&241.53 $\pm$ 0.09	&291.11 $\pm$ 0.86	&274.09 $\pm$ 0.37	&272.77 $\pm$ 0.30\\
-nbody		&396.55 $\pm$ 0.24	&84.77 $\pm$ 0.14	&69.01 $\pm$ 0.08	&68.71 $\pm$ 0.09\\
-pyflate-fast	&1991.33 $\pm$ 1.90	&1818.70 $\pm$ 4.30	&1062.03 $\pm$ 4.56	&1052.02 $\pm$ 3.24\\
-raytrace-simple	&1598.84 $\pm$ 1.07	&1126.93 $\pm$ 11.42	&93.45 $\pm$ 4.75	&92.17 $\pm$ 4.15\\
-richards	&216.47 $\pm$ 0.20	&169.84 $\pm$ 1.21	&11.81 $\pm$ 0.09	&11.90 $\pm$ 0.29\\
-spambayes	&193.30 $\pm$ 0.03	&331.37 $\pm$ 14.93	&238.83 $\pm$ 6.10	&237.23 $\pm$ 5.97\\
-spectral-norm	&375.91 $\pm$ 0.08	&221.24 $\pm$ 0.07	&31.96 $\pm$ 0.06	&28.79 $\pm$ 0.09\\
-telco		&789.67 $\pm$ 1.14	&622.67 $\pm$ 1.86	&155.33 $\pm$ 1.82	&154.67 $\pm$ 1.82\\
-twisted\_names	&6.61 $\pm$ 0.00	&8.85 $\pm$ 0.09	&4.35 $\pm$ 0.05 &4.35 $\pm$ 0.04\\
+\begin{center}
+\begin{tabular}{|l||r|r|r|r|}
+\hline
+	&Cpython	&Psyco	&PyPy w/o optimizations	&PyPy w/ optimizations \\
+\hline
+crypto\_pyaes	&2757.80 $\pm$ 0.98	&67.90 $\pm$ 0.47	&1652.00 $\pm$ 4.00 &266.86 $\pm$ 5.94 \\
+django	&993.19 $\pm$ 0.50	&913.51 $\pm$ 4.22	&694.73 $\pm$ 2.86 &259.53 $\pm$ 1.79 \\
+fannkuch	&1987.22 $\pm$ 2.02	&944.44 $\pm$ 0.61	&566.99 $\pm$ 1.06 &466.87 $\pm$ 1.85 \\
+go	&947.21 $\pm$ 1.58	&445.96 $\pm$ 0.68	&2197.71 $\pm$ 25.21 &316.15 $\pm$ 9.33 \\
+html5lib	&13987.12 $\pm$ 19.51	&17398.25 $\pm$ 36.50	&27194.45 $\pm$ 46.62	&10092.19 $\pm$ 23.50 \\
+meteor-contest	&346.98 $\pm$ 0.35	&215.66 $\pm$ 0.23	&433.04 $\pm$ 1.45 &392.85 $\pm$ 0.87 \\
+nbody\_modified	&637.90 $\pm$ 1.82	&256.78 $\pm$ 0.18	&135.55 $\pm$ 0.33 &103.93 $\pm$ 0.25 \\
+pyflate-fast	&3169.35 $\pm$ 1.89	&1278.16 $\pm$ 3.13	&3285.89 $\pm$ 8.51	&1822.36 $\pm$ 11.52 \\
+raytrace-simple	&2744.60 $\pm$ 51.72	&1072.66 $\pm$ 1.08	&2778.27 $\pm$ 15.13	&647.24 $\pm$ 5.44 \\
+richards	&354.06 $\pm$ 1.00	&63.48 $\pm$ 0.15	&383.93 $\pm$ 3.28 &88.32 $\pm$ 0.91 \\
+spambayes	&299.16 $\pm$ 0.35	&338.68 $\pm$ 3.14	&580.90 $\pm$ 24.68 &397.37 $\pm$ 10.60 \\
+spectral-norm	&478.63 $\pm$ 0.80	&139.83 $\pm$ 1.54	&353.51 $\pm$ 1.39 &112.10 $\pm$ 1.17 \\
+telco	&1207.67 $\pm$ 2.03	&730.00 $\pm$ 2.66	&1296.08 $\pm$ 4.37 &495.23 $\pm$ 2.14 \\
+twisted\_names	&9.58 $\pm$ 0.01	&10.43 $\pm$ 0.01	&17.99 $\pm$ 0.27	&7.13 $\pm$ 0.09 \\
+\hline
 \end{tabular}
+\end{center}
 \caption{Benchmark Times}

### 2010-10-14 talk/pepm2011/paper.tex
give a unit
two things from samuele, regen pdf

@@
 
-Note how the functions that are called by \lstinline{f} are automatically inlined
-into the trace. The method calls are always preceded by a \lstinline{guard_class}
+The method calls in the trace are always preceded by a \lstinline{guard_class}
 operation, to check that the class of the receiver is the same as the one that
@@
 \subsection{Static Objects}
+
+XXX add a footnote saying that "static" is meant in the sense of PE, not static
+allocation
 
@@
 \end{center}
-\caption{Benchmark Times}
+\caption{Benchmark Times in Milliseconds}
 \label{fig:times}

### 2010-10-14 talk/pepm2011/paper.tex
another comment by samuele
Add a comment
the escaping is not needed

@@
 severe in statically typed languages.
+
+XXX [fijal] I would say that the other goal of the JIT in case of dynamic
+languages is to compile only a common scenario and leave a guard (instead
+of proving that something will never ever happen)
 
@@
     \item \lstinline{set} correspond to attribute writes.
-    \item \lstinline{guard\_class} correspond to method calls and are followed by
+    \item \lstinline{guard_class} correspond to method calls and are followed by
     the trace of the called method.
-    \item \lstinline{int\_add} and \lstinline{int\_get} are integer addition and
+    \item \lstinline{int_add} and \lstinline{int_get} are integer addition and
     comparison (``greater than''), respectively.
@@
 The trace shows the inefficiencies of \lstinline{f} clearly, if one looks at the
-number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard\_class} operations.
+number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard_class} operations.
 In the rest of the paper we will see how this trace can be optimized using
 partial evaluation.
+
+XXX add comment that guards are not only slow, but also use lots of memory
 

### 2010-10-14 talk/pepm2011/paper.tex
typo+tweak
some changes to sect. 4

@@
 nobody else in the program stores a reference to them. The idea for improving
-the code is thus to analyze which objects fall in category 1 and may thus
-not be allocated at all.
+the code is thus to analyze which objects fall in category 1 and thus do
+not have to be allocated at all.
 
@@
 perform escape analysis by using partial evaluation. The partial evalution is a
-bit peculiar in that there are not actually any constant arguments to the trace,
-but it is only used to optimized operations within a trace.
-
-The partial evaluation works by walking the trace from beginning to end.
+bit peculiar in that it receives no static input arguments for the trace,
+but it is only used to optimize operations within a trace.
+
+The partial evaluation works by traversing the trace from beginning to end.
 Whenever a \lstinline{new} operation is seen, the operation is removed and a static
@@
 
-In the example from last section, the following operations would produce two
-static objects, and be completely removed from the optimized trace:
+In the example from last section, the following operations in the upper half
+of Fig.~\ref{fig:unopt-trace} would produce two
+static objects, and would thus be completely removed from the optimized trace:
 
@@
 
-The following operations on $p_{5}$ and $p_{6}$ could then be
+The subsequent operations in Fig.~\ref{fig:unopt-trace},
+ which use $p_{5}$ and $p_{6}$, could then be
 optimized using that knowledge:
@@
 
-The \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
-$p_{6}$ are known to be \lstinline{BoxedInteger}. The \lstinline{get} operations can be removed
+First, the \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
+$p_{6}$ are known to be \lstinline{BoxedInteger}. Second, the \lstinline{get} operations can be removed
 and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
@@
 
-The rest of the trace is optimized similarly.
-
-So far we have only described what happens when static objects are used in
-operations that read and write their fields and in guards. When the static
-object is used in any other operation, it cannot stay static. For example, when
-a static object is stored in a globally accessible place, the object needs to
-actually be allocated, as it might live longer than one iteration of the loop
+The rest of the trace from Fig.~\ref{fig:unopt-trace} is optimized similarly.
+
+So far we have only described what happens when static objects are used in guards and in
+operations that read and write fields. When the static
+object is used in any other operation, it cannot remain static. For example, when
+a static object is stored in a globally accessible place, the object has to
+be allocated, as it might live longer than one iteration of the loop
 and because the partial evaluator looses track of it. This means that the static
 object needs to be turned into a dynamic one, \ie lifted. This makes it
-necessary to put operations into the residual code that actually allocate the
+necessary to put operations into the residual code that allocate the
 static object at runtime.
 
-This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
+This is what happens at the end of the trace in Fig.~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
 is hit. The arguments of the jump are at this point static objects. Before the
@@
 values that the static object has (if the static object points to other static
-objects, those need to be lifted as well, recursively) This means that instead of the jump,
+objects, those need to be lifted as well, recursively). This means that instead of a simple jump,
 the following operations are emitted:
@@
 
-Note how the operations for creating these two instances have been moved down the
-trace. It looks like for these operations we actually didn't win much, because
-the objects are still allocated at the end. However, the optimization was still
-worthwhile even in this case, because some operations that have been performed
+Observe how the operations for creating these two instances have been moved to later point in the
+trace. 
+At first sight, it may look like for these operations we didn't gain much, as
+the objects are still allocated in the end. However, our optimizations were still
+worthwhile, because some operations that have been performed
 on the lifted static objects have been removed (some \lstinline{get} operations
@@
 
-In this paper, we used a approach based on online partial evaluation to
-optimize allocations in the traces of a tracing JIT.  In this context a
+In this paper, we used an approach based on online partial evaluation to
+optimize away allocations in the traces of a tracing JIT.  In this context a
 simple approach to partial evaluation gives good results.  This is due

### 2010-10-14 talk/pepm2011/paper.tex
try to have a marginally acceptable author block

@@
-\documentclass{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 
@@
 \begin{document}
-\conferenceinfo{PEPM'11,} {XXX}
-\CopyrightYear{XXX}
-\copyrightdata{XXX}
+\conferenceinfo{PEPM}{'11, Austin, USA}
+\CopyrightYear{2011}
+\copyrightdata{[to be supplied]}
 
@@
 
-\authorinfo{Carl Friedrich Bolz \and Antonio Cuni \and Maciej Fijałkowski \and Michael Leuschel \and Samuele Pedroni \and Armin Rigo}
-           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}
+\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^a$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
+            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
+           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
+
+            $^b$merlinux GmbH, Hildesheim, Germany
+
+            $^c$Open End, Göteborg, Sweden
+           }
            {cfbolz@gmx.de, anto.cuni@gmail.com, fijal@merlinux.eu,
            leuschel@cs.uni-duesseldorf.de, samuele.pedroni@gmail.com, arigo@tunes.org}
+%           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany} \and {merlinux GmbH, Hildesheim, Germany} \and {Open End, Göteborg, Sweden}
 
@@
 
-The authors would like to thank Stefan Hallerstede and Thomas Stiehl for
-fruitful discussions during the writing of the paper.
+The authors would like to thank Stefan Hallerstede, David Schneider and Thomas
+Stiehl for fruitful discussions during the writing of the paper.
 

### 2010-10-14 talk/pepm2011/paper.tex
better email addresses

@@
            }
-           {cfbolz@gmx.de, anto.cuni@gmail.com, fijal@merlinux.eu,
-           leuschel@cs.uni-duesseldorf.de, samuele.pedroni@gmail.com, arigo@tunes.org}
-%           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany} \and {merlinux GmbH, Hildesheim, Germany} \and {Open End, Göteborg, Sweden}
+           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu,
+           leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}
 

### 2010-10-14 talk/pepm2011/paper.tex
fix an XXX, add a new one

@@
 performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
-objects that are allocated in the trace \emph{static}\footnote{These objects are
-called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
+objects that are allocated in the trace \emph{static} which
 means that they do not occur any more in the optimized trace. This technique is
@@
 
-XXX add a footnote saying that "static" is meant in the sense of PE, not static
-allocation
-
 The main insight to improve the code shown in the last section is that objects
@@
 The partial evaluation works by traversing the trace from beginning to end.
-Whenever a \lstinline{new} operation is seen, the operation is removed and a static
-object is constructed and associated with the variable that would have stored
+Whenever a \lstinline{new} operation is seen, the operation is removed and a \empth{static
+object}\footnote{Here ``static'' is meant in the sense of partial
+evaluation, \ie known at partial evaluation time, not in the sense of static
+allocation or static method.} is constructed and associated with the variable
+that would have stored
 the result of \lstinline{new}. The static object describes the shape of the
@@
 
+XXX Psyco \cite{rigo_representation-based_2004}
+
+XXX cite SELF paper for the type propagation effects
+
 \section{Conclusion}

### 2010-10-14 talk/pepm2011/paper.tex
Two more comments
fix another an XXX and a typo

@@
 
-The trace shows the inefficiencies of \lstinline{f} clearly, if one looks at the
-number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard_class} operations.
+The trace shows the inefficiencies of \lstinline{f} clearly, if one looks at
+the number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard_class}
+operations. The number of \lstinline{guard_class} operation is particularly
+problematic, not only because of the time it takes to run them. All guards also
+have additional information attached that makes it possible to return to the
+interpreter, should the guard fail. This means that many guard operations also
+lead to a memory problem.
+
 In the rest of the paper we will see how this trace can be optimized using
 partial evaluation.
-
-XXX add comment that guards are not only slow, but also use lots of memory
 
@@
 The partial evaluation works by traversing the trace from beginning to end.
-Whenever a \lstinline{new} operation is seen, the operation is removed and a \empth{static
+Whenever a \lstinline{new} operation is seen, the operation is removed and a \emph{static
 object}\footnote{Here ``static'' is meant in the sense of partial
@@
 
+XXX [fijal] Well, in fact that's *a lot* of support for debugger from the VM.
+The support is just generic instead of specific.
+
 This problem is solved by making it possible to the interpreter author to add
@@
 benchmarks, some real applications.
+
+XXX [fijal] note that some of those have cpython-specific hacks removed or
+old versions used (nbody mostly I think).
 

### 2010-10-14 talk/pepm2011/paper.tex
Add the word "modern" here, and a footnote.

@@
 the concrete objects that should be added are, and choose the implementation
-that is fitting for them. Type dispatching is a very common operation in a
-dynamic language because no types are known at compile time, so all operations
+that is fitting for them. Type dispatching is a very common operation in
+modern\footnote{For languages in the LISP family, basic arithmetic operations
+are typically not overloaded; even in Smalltalk, type dispatching is much
+simpler than in Python or JavaScript.}
+dynamic languages because no types are known at compile time, so all operations
 need it.

### 2010-10-14 talk/pepm2011/paper.tex
Typos.

@@
 simplicity. They can often be added to an interpreter and a lot of the
-infrastructure of an interpreter can be reused. They give some important
+infrastructure of the interpreter can be reused. They give some important
 optimizations like inlining and constant-folding for free. A tracing JIT always
@@
 performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
-objects that are allocated in the trace \emph{static} which
+objects that are allocated in the trace \emph{static,} which
 means that they do not occur any more in the optimized trace. This technique is
-informally described in Section~\ref{sec:statics}, a more formal description is
+informally described in Section~\ref{sec:statics}; a more formal description is
 given in Section~\ref{sec:formal}.
@@
 CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
-interpreter. Furthermore we compared against Psyco 1.6
-\cite{rigo_representation-based_2004}, an extension to CPython which is a
+interpreter. Furthermore we compared against
+Psyco\cite{rigo_representation-based_2004} 1.6,
+an extension to CPython which is a
 just-in-time compiler that produces machine code at run-time. It is not based
 on traces. Finally, we used two versions of PyPy's Python interpreter (revision
-77823 of SVN trunk\footnote{\texttt{https://codespeak.net/svn/pypy/trunk}}): one
+77823 of SVN trunk\footnote{\texttt{http://codespeak.net/svn/pypy/trunk}}): one
 including the JIT but not optimizing the traces, and one using the allocation
@@
 \hline
-	&Cpython	&Psyco	&PyPy w/o optimizations	&PyPy w/ optimizations \\
+	&CPython	&Psyco	&PyPy w/o optimizations	&PyPy w/ optimizations \\
 \hline

### 2010-10-14 talk/pepm2011/paper.tex
A short paragraph about Psyco in Related Work.
expand for clarification

@@
 
-One problem to the successful application of the allocation removal techniques
-described in the previous sections is the presence of frame-introspection
-features in many dynamic languages. Languages such as Python and Smalltalk allow
-the programmer to get access to the frames object that the interpreter uses to
-store local variables. This is a useful feature, as makes the implementation of
-a debugger possible in Python without needing much support from the VM level. On
-the other hand, it severely hinders the effectiveness of allocation removal,
-because every time an object is stored into a local variable, it is stored into
-the frame-object, which makes it escape.
-
-XXX [fijal] Well, in fact that's *a lot* of support for debugger from the VM.
-The support is just generic instead of specific.
-
-This problem is solved by making it possible to the interpreter author to add
-some hints into the source code to declare instances of one class as frame
-objects. The JIT will then fill these objects only lazily when they are actually
-accessed (\eg because a debugger is used). Therefore in the common case, nothing
-is stored into the frame objects, making the problem of too much escaping go
-away. This is a common approach in VM implementations \cite{miranda_context_1999}, the only
-novelty in our approach lays in its generality, because most other JITs are just
-specifically written for one particular language.
+One problem to the successful application of the allocation removal
+techniques described in the previous sections is the presence of
+frame-introspection features in many dynamic languages. Languages such
+as Python and Smalltalk allow the programmer to get access to the
+frames object that the interpreter uses to store local variables. This
+is a useful feature, as makes the implementation of a debugger
+possible in Python without needing much explicit support from the VM
+level. On the other hand, it severely hinders the effectiveness of
+allocation removal, because every time an object is stored into a
+local variable, it is stored into the frame-object, which makes it
+escape.
+
+This problem is solved by making it possible to the interpreter author
+to add some hints into the source code to declare instances of one
+class as frame objects needing special treatment. The JIT will then
+fill these objects only lazily when they are actually accessed (\eg
+because a debugger is used) using values from JIT-emitted code runtime
+locations (typically the CPU stack). Therefore in the common case,
+nothing is stored into the frame objects, making the problem of too
+much escaping go away. This is a common approach in VM implementations
+\cite{miranda_context_1999}, the only novelty in our approach lays in
+its generality, all the delicate support code for this is generated, because
+most other JITs are just specifically written for one particular
+language.
 
@@
 
+Psyco \cite{rigo_representation-based_2004} is a (non-tracing) JIT for Python
+that fully implements allocation removal (the \emph{static} objects are called
+\emph{virtual} objects).  It is a hand-written extension module for CPython.
+Historically, PyPy's JIT can be seen as some successor of Psyco for a
+general context.
+
 partial evaluation:
@@
 xxx: relation to compile-time garbage collection \cite{mazur_practical_2001};  separation logic; John Hughes: type specialization
-
-XXX Psyco \cite{rigo_representation-based_2004}
 

### 2010-10-14 talk/pepm2011/paper.tex
(cfbolz, arigo for some of it):
    - add URLs for PyPy
    - add factors to the timing table, to read it more easily
    - change psyco description
    - some fixes in the virtualizables section

@@
 
-The work described in this paper was done in the context of the PyPy project
+The work described in this paper was done in the context of the PyPy project\footnote{\texttt{http://pypy.org}}
 \cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
@@
 frames object that the interpreter uses to store local variables. This
-is a useful feature, as makes the implementation of a debugger
+is a useful feature, as it makes the implementation of a debugger
 possible in Python without needing much explicit support from the VM
@@
 nothing is stored into the frame objects, making the problem of too
-much escaping go away. This is a common approach in VM implementations
-\cite{miranda_context_1999}, the only novelty in our approach lays in
-its generality, all the delicate support code for this is generated, because
-most other JITs are just specifically written for one particular
+much escaping go away. This special handling of frame objects is a common approach in VM implementations
+\cite{miranda_context_1999}. The only novelty in our approach lays in
+its generality: all the delicate support code for this is generated, as opposed to
+most other JITs, which are just specifically written for one particular
 language.
+
+XXX one of the JS tracing JITs does this as well
 
@@
 benchmarks we used are small-to-medium Python programs, some synthetic
-benchmarks, some real applications.
+benchmarks, some real applications.\footnote{All the source code of the
+benchmarks can be found at \texttt{http://codespeak.net/svn/pypy/benchmarks/}.
+There is also a website that monitors PyPy's performance nightly at
+\texttt{http://speed.pypy.org/}.}
 
@@
 
-XXX runtimes of the algorithm somehow?
-
 \begin{figure*}
 \begin{center}
-\begin{tabular}{|l||r|r|r|r|}
+\begin{tabular}{|l||r|r||r|r||r|r||r|r|}
 \hline
-	&CPython	&Psyco	&PyPy w/o optimizations	&PyPy w/ optimizations \\
+	&CPython [ms]	& $\times$ &Psyco [ms]	& $\times$ &PyPy w/o optim. [ms] & $\times$ &PyPy w/ optim. [ms]& $\times$ \\
 \hline
-crypto\_pyaes	&2757.80 $\pm$ 0.98	&67.90 $\pm$ 0.47	&1652.00 $\pm$ 4.00 &266.86 $\pm$ 5.94 \\
-django	&993.19 $\pm$ 0.50	&913.51 $\pm$ 4.22	&694.73 $\pm$ 2.86 &259.53 $\pm$ 1.79 \\
-fannkuch	&1987.22 $\pm$ 2.02	&944.44 $\pm$ 0.61	&566.99 $\pm$ 1.06 &466.87 $\pm$ 1.85 \\
-go	&947.21 $\pm$ 1.58	&445.96 $\pm$ 0.68	&2197.71 $\pm$ 25.21 &316.15 $\pm$ 9.33 \\
-html5lib	&13987.12 $\pm$ 19.51	&17398.25 $\pm$ 36.50	&27194.45 $\pm$ 46.62	&10092.19 $\pm$ 23.50 \\
-meteor-contest	&346.98 $\pm$ 0.35	&215.66 $\pm$ 0.23	&433.04 $\pm$ 1.45 &392.85 $\pm$ 0.87 \\
-nbody\_modified	&637.90 $\pm$ 1.82	&256.78 $\pm$ 0.18	&135.55 $\pm$ 0.33 &103.93 $\pm$ 0.25 \\
-pyflate-fast	&3169.35 $\pm$ 1.89	&1278.16 $\pm$ 3.13	&3285.89 $\pm$ 8.51	&1822.36 $\pm$ 11.52 \\
-raytrace-simple	&2744.60 $\pm$ 51.72	&1072.66 $\pm$ 1.08	&2778.27 $\pm$ 15.13	&647.24 $\pm$ 5.44 \\
-richards	&354.06 $\pm$ 1.00	&63.48 $\pm$ 0.15	&383.93 $\pm$ 3.28 &88.32 $\pm$ 0.91 \\
-spambayes	&299.16 $\pm$ 0.35	&338.68 $\pm$ 3.14	&580.90 $\pm$ 24.68 &397.37 $\pm$ 10.60 \\
-spectral-norm	&478.63 $\pm$ 0.80	&139.83 $\pm$ 1.54	&353.51 $\pm$ 1.39 &112.10 $\pm$ 1.17 \\
-telco	&1207.67 $\pm$ 2.03	&730.00 $\pm$ 2.66	&1296.08 $\pm$ 4.37 &495.23 $\pm$ 2.14 \\
-twisted\_names	&9.58 $\pm$ 0.01	&10.43 $\pm$ 0.01	&17.99 $\pm$ 0.27	&7.13 $\pm$ 0.09 \\
+crypto\_pyaes	&2757.80 $\pm$ 0.98	&10.33	&67.90 $\pm$ 0.47	&0.25	&1652.00 $\pm$ 4.00	&6.19	&266.86 $\pm$ 5.94	&1.00\\
+django	&993.19 $\pm$ 0.50	&3.83	&913.51 $\pm$ 4.22	&3.52	&694.73 $\pm$ 2.86	&2.68	&259.53 $\pm$ 1.79	&1.00\\
+fannkuch	&1987.22 $\pm$ 2.02	&4.26	&944.44 $\pm$ 0.61	&2.02	&566.99 $\pm$ 1.06	&1.21	&466.87 $\pm$ 1.85	&1.00\\
+go	&947.21 $\pm$ 1.58	&3.00	&445.96 $\pm$ 0.68	&1.41	&2197.71 $\pm$ 25.21	&6.95	&316.15 $\pm$ 9.33	&1.00\\
+html5lib	&13987.12 $\pm$ 19.51	&1.39	&17398.25 $\pm$ 36.50	&1.72	&27194.45 $\pm$ 46.62	&2.69	&10092.19 $\pm$ 23.50	&1.00\\
+meteor-contest	&346.98 $\pm$ 0.35	&0.88	&215.66 $\pm$ 0.23	&0.55	&433.04 $\pm$ 1.45	&1.10	&392.85 $\pm$ 0.87	&1.00\\
+nbody\_modified	&637.90 $\pm$ 1.82	&6.14	&256.78 $\pm$ 0.18	&2.47	&135.55 $\pm$ 0.33	&1.30	&103.93 $\pm$ 0.25	&1.00\\
+pyflate-fast	&3169.35 $\pm$ 1.89	&1.74	&1278.16 $\pm$ 3.13	&0.70	&3285.89 $\pm$ 8.51	&1.80	&1822.36 $\pm$ 11.52	&1.00\\
+raytrace-simple	&2744.60 $\pm$ 51.72	&4.24	&1072.66 $\pm$ 1.08	&1.66	&2778.27 $\pm$ 15.13	&4.29	&647.24 $\pm$ 5.44	&1.00\\
+richards	&354.06 $\pm$ 1.00	&4.01	&63.48 $\pm$ 0.15	&0.72	&383.93 $\pm$ 3.28	&4.35	&88.32 $\pm$ 0.91	&1.00\\
+spambayes	&299.16 $\pm$ 0.35	&0.75	&338.68 $\pm$ 3.14	&0.85	&580.90 $\pm$ 24.68	&1.46	&397.37 $\pm$ 10.60	&1.00\\
+spectral-norm	&478.63 $\pm$ 0.80	&4.27	&139.83 $\pm$ 1.54	&1.25	&353.51 $\pm$ 1.39	&3.15	&112.10 $\pm$ 1.17	&1.00\\
+telco	&1207.67 $\pm$ 2.03	&2.44	&730.00 $\pm$ 2.66	&1.47	&1296.08 $\pm$ 4.37	&2.62	&495.23 $\pm$ 2.14	&1.00\\
+twisted\_names	&9.58 $\pm$ 0.01	&1.34	&10.43 $\pm$ 0.01	&1.46	&17.99 $\pm$ 0.27	&2.52	&7.13 $\pm$ 0.09	&1.00\\
 \hline
@@
 \end{center}
-\caption{Benchmark Times in Milliseconds}
+\caption{Benchmark Times in Milliseconds, Together With Factor Over PyPy With Optimizations}
 \label{fig:times}
@@
 Psyco \cite{rigo_representation-based_2004} is a (non-tracing) JIT for Python
-that fully implements allocation removal (the \emph{static} objects are called
-\emph{virtual} objects).  It is a hand-written extension module for CPython.
-Historically, PyPy's JIT can be seen as some successor of Psyco for a
-general context.
+that implements a more ad-hoc version of the allocation removal described here.
+Our static objects could be related to what are called \emph{virtual} objects
+in Psyco.  It is a hand-written extension module for CPython. Historically,
+PyPy's JIT can be seen as some successor of Psyco for a general context (one of
+the authors of this paper is the author of Psyco).
 

### 2010-10-14 talk/pepm2011/paper.tex
clarifications

@@
 techniques described in the previous sections is the presence of
-frame-introspection features in many dynamic languages. Languages such
-as Python and Smalltalk allow the programmer to get access to the
-frames object that the interpreter uses to store local variables. This
-is a useful feature, as it makes the implementation of a debugger
-possible in Python without needing much explicit support from the VM
-level. On the other hand, it severely hinders the effectiveness of
-allocation removal, because every time an object is stored into a
-local variable, it is stored into the frame-object, which makes it
-escape.
+frame-introspection features in many dynamic languages and the fact
+that frames are heap allocated. Languages such as Python and Smalltalk
+allow the programmer to get access to the frames object that the
+interpreter uses to store local variables. This is a useful feature,
+as it makes the implementation of a debugger possible in Python
+without needing much explicit support from the VM level. On the other
+hand, it severely hinders the effectiveness of allocation removal,
+because every time an object is stored into a local variable, it is
+stored into the frame-object, which makes it escape.
 
@@
 
-In this paper, we used an approach based on online partial evaluation to
-optimize away allocations in the traces of a tracing JIT.  In this context a
-simple approach to partial evaluation gives good results.  This is due
-to the fact that the tracing JIT itself is responsible for all control
-issues, which are usually the hardest part of partial evaluation: the
-tracing JIT selects the parts of the program that are worthwhile to
-optimize, and extracts linear paths through them, inlining functions as
-necessary.  What is left to optimize is only those linear paths.
+In this paper, we used an approach based on online partial evaluation
+to optimize away allocations and type guards in the traces of a
+tracing JIT.  In this context a simple approach to partial evaluation
+gives good results.  This is due to the fact that the tracing JIT
+itself is responsible for all control issues, which are usually the
+hardest part of partial evaluation: the tracing JIT selects the parts
+of the program that are worthwhile to optimize, and extracts linear
+paths through them, inlining functions as necessary.  What is left to
+optimize is only those linear paths.
 

### 2010-10-14 talk/pepm2011/paper.tex
fix XXX, rename section

@@
            }
-           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu,
+           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu \and
            leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}
@@
 
-In Section~\ref{sec:support} we describe some supporting techniques that are not
-central to the approach, but are needed to improve the results. The introduced
+In Section~\ref{sec:frames} we describe some supporting techniques that are not
+central to the approach, but are essential to improve the results. The introduced
 techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
@@
 
-\section{Supporting Techniques}
-\label{sec:support}
-
-\subsection{Virtualizables}
-\label{sub:Virtualizables}
+\section{Taming Frame Objects}
+\label{sec:frames}
 
@@
 
-% subsection Virtualizables (end)
-
-% section Supporting Techniques (end)
+% section Taming Frame Objects (end)
 
@@
 result. The errors were computed using a confidence interval with a 95\%
-confidence level \cite{georges_statistically_2007}. The results are reported in Figure~\ref{fig:times}.
+confidence level \cite{georges_statistically_2007}. The results are reported in
+Figure~\ref{fig:times}. In addition to the run times the table also reports the
+speedup that PyPy with optimization turned on achieves.
 With the optimization turned on, PyPy's Python interpreter outperforms CPython
@@
 performance and thus is not helped much by our Python JIT) and meteor-contest.
-All benchmarks are improved by the allocation removal optimization, some as much
-as XXX. XXX Psyco
+All benchmarks are improved by the allocation removal optimization, by at least
+20\% and by as much as a factor of 6.95.
+
+Psyco is able to outperform PyPy's JIT in five out of 14 benchmarks. We hope to
+overtake Psyco (which is no longer being actively developped) by adding some
+further optimizations.
 

### 2010-10-14 talk/pepm2011/paper.tex
I think now that there is the explicit link to the source code of the benchmarks, this XXX can go
fix another XXX
I decided to ignore this comment, because that is dealt with by tracing itself, not by the optimization that the paper is about

@@
 severe in statically typed languages.
-
-XXX [fijal] I would say that the other goal of the JIT in case of dynamic
-languages is to compile only a common scenario and leave a guard (instead
-of proving that something will never ever happen)
 
@@
 much escaping go away. This special handling of frame objects is a common approach in VM implementations
-\cite{miranda_context_1999}. The only novelty in our approach lays in
+\cite{miranda_context_1999,andreas_gal_trace-based_2009}. The only novelty in our approach lays in
 its generality: all the delicate support code for this is generated, as opposed to
@@
 language.
-
-XXX one of the JS tracing JITs does this as well
 
@@
 \texttt{http://speed.pypy.org/}.}
-
-XXX [fijal] note that some of those have cpython-specific hacks removed or
-old versions used (nbody mostly I think).
 

### 2010-10-14 talk/pepm2011/paper.tex
mention self

@@
 
+The original SELF JIT compiler \cite{chambers_efficient_1989} used an algorithm for
+forward-propagating the types of variables as part of its optimizations. This
+makes it possible to remove all type checks on a variable but the first one.
+The optimization does not deal with removing the full object, if it is
+short-lived, but the type check removals are similar to what our optimization
+achieves.
+
 partial evaluation:
@@
 xxx: relation to compile-time garbage collection \cite{mazur_practical_2001};  separation logic; John Hughes: type specialization
-
-XXX cite SELF paper for the type propagation effects
 

### 2010-10-14 talk/pepm2011/paper.tex
flesh out section about tracing

@@
 
-Prolog:
-partially static datastructures are already built-in to Prolog and was built-into partial
-evaluation from the beginning \cite{lloyd_partial_1991}.
+Partially known data structures are built directly into Prolog (via unbound
+logic variables) and thus the optimization of partially static data structures
+was part of partial evaluation of Prolog programs from the beginning
+\cite{lloyd_partial_1991}. One effect of unfolding in Prolog is that terms that
+are constructed and immediately matched again completely disappear in the
+residual program, which is equivalent to what our optimization does for an
+imperative language.
+
  FP

### 2010-10-14 talk/pepm2011/paper.tex
fixes by david

@@
 improve the speed of the language over an implementation of the language that
-uses interpretation. The first goal of a JIT is thus to remove the
+uses interpretation. The first goal of a JIT is therefore to remove the
 interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
@@
 second important problem that any JIT for a dynamic language needs to solve is
-how to deal with the overhead of boxing of primitive types and of type
+how to deal with the overhead of boxing primitive types and of type
 dispatching. Those are problems that are usually not present or at least less
@@
 optimizations like inlining and constant-folding for free. A tracing JIT always
-produces linear pieces of code, which simplifies many optimizations that are usually
+produces linear pieces of code, which simplifies many algorithms that are usually
 hard in a compiler, such as register allocation.
@@
 interpreter for the less commonly executed parts. VMs that use a tracing JIT are
-thus mixed-mode execution environments, they contain both an interpreter and a
+mostly mixed-mode execution environments, they contain both an interpreter and a
 JIT compiler. By default the interpreter is used to execute the program, doing
@@
 first optimized, and then turned into machine code. Both optimization
-and machine code generation is simple, because the traces are linear. This
+and machine code generation are simple, because the traces are linear. This
 linearity makes many optimizations a lot more tractable, and the inlining that
@@
 taken during the production of the trace. The trace will also contain a guard
-that checks that the condition of the \lstinline{if} statement is true, because if
-it isn't, the rest of the trace is not valid.
+that checks that the condition of the \lstinline{if} statement is the same as
+during tracing, because if
+it isn't, the rest of the trace is not valid. \cfbolz{The "if" shouldn't be bold}
 
@@
 nobody else in the program stores a reference to them. The idea for improving
-the code is thus to analyze which objects fall in category 1 and thus do
+the code is to analyze which objects fall in category 1 and therefore do
 not have to be allocated at all.
@@
 A variable $v^*$ is in the domain of the static heap $S$ as long as the
-optimizer can fully keep track of the object. The image of $v^*$ is what is
+optimizer can fully keep track of the object. The object $S(v^*)$ is what is
 statically known about the object stored in it, \ie its type and its fields. The
@@
 Figure~\ref{fig:times}. In addition to the run times the table also reports the
-speedup that PyPy with optimization turned on achieves.
+speedup that PyPy achieves when the optimization is turned on.
+
 With the optimization turned on, PyPy's Python interpreter outperforms CPython
@@
 
-There exists a large number of works on escape analysis, which is an program
+There exists a large number of works on escape analysis, which is a program
 analysis that tries to find an upper bound for the lifetime of objects allocated
@@
 paths through them, inlining functions as necessary.  What is left to
-optimize is only those linear paths.
+optimize are only those linear paths.
 

### 2010-10-14 talk/pepm2011/paper.tex
remove a number of "thus"s. simplify two paragraphs in the background section.

@@
 languages is that of a tracing JIT. A tracing JIT works by observing the running
-program and recording its hot spots into linear execution traces, which are then turned into
-machine code. One reason for the popularity of tracing JITs is their relative
+program and recording its hot spots into linear execution traces. Working on
+traces is the central idea of a tracing JIT. Those traces are optimized and
+turned into machine code.
+
+One reason for the popularity of tracing JITs is their relative
 simplicity. They can often be added to an interpreter and a lot of the
@@
 possible. The language interpreter can then be compiled (``translated'') with
-PyPy's tools into a VM on the C level. Because the interpreter is written at a
-relatively high level, the language implementation is kept free of low-level
-details, such as object layout, garbage collection or memory model. Those
-aspects of the final VM are woven into the generated code during the translation
-to C.
+PyPy's tools into a VM on the C level. During translation to C, many low-level
+aspects of the final VM, such as object layout, garbage collection and memory
+model, are woven into the generated code. Therefore the interpreter itself can
+remain at a relatively high level of abstraction.
 
@@
 support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
-the translation to C, PyPy's tools can generate a just-in-time compiler for the
+the translation to C, PyPy's tools can generate a tracing just-in-time compiler for the
 language that the interpreter is implementing. This process is mostly
@@
 Moreover, thanks to the internal design of the JIT generator, it is very easy
-to add new \emph{backends} for producing the actual machine code, in addition
-to the original backend for the Intel \emph{x86} architecture.  Examples of
-additional JIT backends are the one for Intel \emph{x86-64} and an
+to add new \emph{backends} for producing the actual machine code.  Examples of
+JIT backends that are implemented are the one for Intel x86 and x86-64 and an
 experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.
-PyPy's JIT generator generates a \emph{tracing JIT
-compiler}, a concept which we now explain in more details.
 
@@
 dynamic language implementation has. All the numbers are instances of either
-\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, thus they consume space on the
+\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
 heap. Performing many arithmetic operations produces lots of garbage quickly,
-thus putting pressure on the garbage collector. Using double dispatching to
+which puts pressure on the garbage collector. Using double dispatching to
 implement the numeric tower needs two method calls per arithmetic operation,
@@
 In the example from last section, the following operations in the upper half
-of Fig.~\ref{fig:unopt-trace} would produce two
-static objects, and would thus be completely removed from the optimized trace:
+of Fig.~\ref{fig:unopt-trace} produce two
+static objects, and can be completely removed from the optimized trace:
 
@@
 $p_{6}$ are known to be \lstinline{BoxedInteger}. Second, the \lstinline{get} operations can be removed
-and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
+and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. The only
 remaining operation in the optimized trace would be:
@@
 be malformed. The environment could not directly map variables to object,
-because several variables can contain a pointer to the \emph{same} object. Thus
-the "indirection" is needed to express sharing.
+because several variables can contain a pointer to the \emph{same} object. 
+The "indirection" is needed to express sharing.
 
@@
 trace is that several variables that appear in the unoptimized trace can turn
-into the same variables in the optimized trace. Thus the environment of the
+into the same variables in the optimized trace. The environment of the
 optimizer serves a function similar to that of the environment in the semantics: sharing.

### 2010-10-14 talk/pepm2011/paper.tex
Small grammar fix.

@@
 viewed as partial evaluation: the partial evaluation
-performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
+performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and makes some
 objects that are allocated in the trace \emph{static,} which

### 2010-10-15 talk/pepm2011/paper.tex
minor edit

@@
 \end{lstlisting}
-\caption{Unoptimized Trace for the Simple Object Model}
+\caption{An Unoptimized Trace of the Example Interpreter}
 \label{fig:unopt-trace}
@@
     the trace of the called method.
-    \item \lstinline{int_add} and \lstinline{int_get} are integer addition and
+    \item \lstinline{int_add} and \lstinline{int_gt} are integer addition and
     comparison (``greater than''), respectively.
@@
 Partially known data structures are built directly into Prolog (via unbound
-logic variables) and thus the optimization of partially static data structures
-was part of partial evaluation of Prolog programs from the beginning
+logic variables) and thus the treatment of partially static data structures
+was part of partial evaluation of Prolog programs from the early stages
 \cite{lloyd_partial_1991}. One effect of unfolding in Prolog is that terms that
-are constructed and immediately matched again completely disappear in the
-residual program, which is equivalent to what our optimization does for an
+are constructed and immediately matched again, completely disappear in the
+residual program. This is similar to what our optimization does for an
 imperative language.
 
- FP
-
+xxx:
 partially static data structures: kenichi asai's thesis?
 
-xxx: relation to compile-time garbage collection \cite{mazur_practical_2001};  separation logic; John Hughes: type specialization
+xxx: 
+Other related work is compile-time garbage collection \cite{mazur_practical_2001}, which
+tries to ....;
+  separation logic; John Hughes: type specialization
 

### 2010-10-15 talk/pepm2011/paper.tex
added reference to constructor specialisation
some minor changes + requeusts

@@
 
+XXX Apply the rules of the algorithm to parts of the sample trace
+
 % subsection Optimizing Traces (end)
@@
 the algorithm only takes a total time linear in the length of the trace.
-Moreover, it gives the ``best'' possible result within its constrains,
+Moreover, it gives the ``best'' possible result within its constraints,
+\marginpar{Why? What are the constraints ? Justify}
 \eg in term of the number of residual operations.  The
-algorithm itself is not particularly complex or innovative; our focus is
+algorithm itself is not particularly complex; our focus is
 rather that \emph{in the context of tracing JITs} it is possible to find a
@@
 
+XXX unclear what the relationship is to the technique; has it been used in the experiments??
+XXX Maybe move to discussion and related work section?
+
 One problem to the successful application of the allocation removal
@@
 as it makes the implementation of a debugger possible in Python
-without needing much explicit support from the VM level. On the other
+without extra support from the VM. On the other
 hand, it severely hinders the effectiveness of allocation removal,
@@
 imperative language.
-
-xxx:
-partially static data structures: kenichi asai's thesis?
+In functional programming this idea was introduced as
+ constructor specialisation by Mogensen \cite{Mogensen:PEPM93}.)
+
+%xxx:
+%partially static data structures: kenichi asai's thesis?
 
@@
 tries to ....;
-  separation logic; John Hughes: type specialization
+  separation logic
+  %; John Hughes: type specialization
 

### 2010-10-15 talk/pepm2011/paper.tex
tone down claims a bit

@@
 often completely removing) some operations.  The algorithm runs in a
-single pass over the list of operations.  We can check that altough
+single pass over the list of operations.  We can check that although
 recursively lifting a static object is not a constant-time operation,
 the algorithm only takes a total time linear in the length of the trace.
-Moreover, it gives the ``best'' possible result within its constraints,
-\marginpar{Why? What are the constraints ? Justify}
-\eg in term of the number of residual operations.  The
-algorithm itself is not particularly complex; our focus is
+The algorithm itself is not particularly complex; our focus is
 rather that \emph{in the context of tracing JITs} it is possible to find a
-simple enough algorithm that still gives the best results.
+simple enough algorithm that still gives very good results.
 
@@
 (\ie escaping) are still partially dealt with: if such an object
-escapes later than its creation point, all the operations inbetween that
+escapes later than its creation point, all the operations in between that
 involve the object are removed.

### 2010-10-15 talk/pepm2011/paper.tex
kill the section about frames, as michael thinks its too confusing. Instead mention frames minimally in the evaluation section.

@@
 informally described in Section~\ref{sec:statics}; a more formal description is
-given in Section~\ref{sec:formal}.
-
-In Section~\ref{sec:frames} we describe some supporting techniques that are not
-central to the approach, but are essential to improve the results. The introduced
+given in Section~\ref{sec:formal}. The introduced
 techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
@@
 object's field and the next interpreted opcode reads it back, possibly
-dispatching on the type of the object created just before.  In the case
-of PyPy's Python interpreter, this optimization can even remove the
-allocation of all intermediate frames that occur in the interpreter,
-corresponding to all calls that have been inlined in the trace.
+dispatching on the type of the object created just before.  A typical example
+would be chains of arithmetic operations.
 
@@
 
-\section{Taming Frame Objects}
-\label{sec:frames}
-
-\cfbolz{probably can be cut in case of space problems}
-
-XXX unclear what the relationship is to the technique; has it been used in the experiments??
-XXX Maybe move to discussion and related work section?
-
-One problem to the successful application of the allocation removal
-techniques described in the previous sections is the presence of
-frame-introspection features in many dynamic languages and the fact
-that frames are heap allocated. Languages such as Python and Smalltalk
-allow the programmer to get access to the frames object that the
-interpreter uses to store local variables. This is a useful feature,
-as it makes the implementation of a debugger possible in Python
-without extra support from the VM. On the other
-hand, it severely hinders the effectiveness of allocation removal,
-because every time an object is stored into a local variable, it is
-stored into the frame-object, which makes it escape.
-
-This problem is solved by making it possible to the interpreter author
-to add some hints into the source code to declare instances of one
-class as frame objects needing special treatment. The JIT will then
-fill these objects only lazily when they are actually accessed (\eg
-because a debugger is used) using values from JIT-emitted code runtime
-locations (typically the CPU stack). Therefore in the common case,
-nothing is stored into the frame objects, making the problem of too
-much escaping go away. This special handling of frame objects is a common approach in VM implementations
-\cite{miranda_context_1999,andreas_gal_trace-based_2009}. The only novelty in our approach lays in
-its generality: all the delicate support code for this is generated, as opposed to
-most other JITs, which are just specifically written for one particular
-language.
-
-% section Taming Frame Objects (end)
-
-\section{Evaluation}
+\section{Implementation and Evaluation}
 \label{sec:Evaluation}
+
+The allocation removal techniques described in this paper were implemented in
+the optimizer of PyPy's tracing JIT. The optimization is independent of which
+interpreter a JIT is generated for. There are some practical issues beyond the
+techniques described in this paper. The actual implementation needs to deal with
+more operations than described in Section~\ref{sec:formal}, for example to
+also support static arrays in addition to static objects. The implementation of
+this optimization is about 400 lines of RPython code.
+
+A further complication is that most interpreters written with PyPy use
+heap-allocated frame objects to store local variables. Those severely hinders
+the effectiveness of allocation removal, because every time an object is stored
+into a local variable, it is stored into the frame-object, which makes it
+escape. We implemented a general technique to treat such frames objects in a
+special way to solve this problem. This is a common approach in VM
+implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}. The
+only novelty in our approach lays in its generality: most other JITs, are just
+specifically written for one particular language instead of being generated from
+an interpreter.
 

### 2010-10-15 talk/pepm2011/paper.tex
- shorten a bit  - mention size of python interpreter

@@
 A further complication is that most interpreters written with PyPy use
-heap-allocated frame objects to store local variables. Those severely hinders
+heap-allocated frame objects to store local variables. Those severely hinder
 the effectiveness of allocation removal, because every time an object is stored
 into a local variable, it is stored into the frame-object, which makes it
-escape. We implemented a general technique to treat such frames objects in a
+escape. We implemented a technique to treat such frames objects in a
 special way to solve this problem. This is a common approach in VM
-implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}. The
-only novelty in our approach lays in its generality: most other JITs, are just
-specifically written for one particular language instead of being generated from
-an interpreter.
-
-To evaluate the effectiveness of our allocation removal algorithm, we look at
-the effectiveness when used in the tracing JIT of PyPy's Python interpreter. The
+implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}, the
+novelty of our approach is that we generalized it enough to be usable for
+different interpreter.
+
+To evaluate our allocation removal algorithm, we look at the effectiveness when
+used in the generated tracing JIT of PyPy's Python interpreter. This interpreter
+is a full implementation of Python 2.5 language semantics and is about 30'000
+lines of code.
+
+The
 benchmarks we used are small-to-medium Python programs, some synthetic
@@
 \begin{itemize}
-    \item \textbf{crypto\_pyaes}: AES implementation.
+    \item \textbf{crypto\_pyaes}: An AES implementation.
     \item \textbf{django}: The templating engine of the Django web

### 2010-10-15 talk/pepm2011/paper.tex
Semicolonization.

@@
 special way to solve this problem. This is a common approach in VM
-implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}, the
+implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}; the
 novelty of our approach is that we generalized it enough to be usable for

### 2010-10-15 talk/pepm2011/paper.tex
add lifting example

@@
 
-XXX Apply the rules of the algorithm to parts of the sample trace
+As an example for lifting, consider the static heap $$\{v^* \mapsto (T_1, w^*,
+v^*), w^* \mapsto (T_2, u^*, u^*)\}$$ which contains two static objects. If $v^*$
+now needs to be lifted, the following residual operations are produced:
+
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+$v^*$ = new($T_1$)
+$w^*$ = new($T_2$)
+set($w^*$, $L$, $u^*$)
+set($w^*$, $R$, $u^*$)
+set($v^*$, $L$, $w^*$)
+set($v^*$, $R$, $v^*$)
+\end{lstlisting}
+
+After the lifting the static heap is the empty set, because both static objects
+were lifted.
 

### 2010-10-15 talk/pepm2011/paper.tex
enhance example
Rewording.
incorporate first round of comments by stephan

@@
 languages less applicable to purely algorithmic problems, despite their
-growing popularity.  In this paper, we present a simple optimization
+growing popularity.  In this paper we present a simple compiler optimization
 based on online partial evaluation to remove object allocations and
@@
 
-The goal of a just-in-time (JIT) compiler for a dynamic language is obviously to
+The objective of a just-in-time (JIT) compiler for a dynamic language is to
 improve the speed of the language over an implementation of the language that
@@
 all objects, even integers, floats, booleans etc. in the same way as user-defined
-instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
-heap-structure is allocated for them, that contains the actual value. Boxing
+instances. Thus those primitive types are usually \emph{boxed}, \ie a small
+heap-structure is allocated for them that contains the actual value. Boxing
 primitive types can be very costly, because a lot of common operations,
-particularly all arithmetic operations, have to produce a new box, in addition
+particularly all arithmetic operations, have to produce new boxes, in addition
 to the actual computation they do. Because the boxes are allocated on the heap,
-producing a lot of them puts pressure on the garbage collector.
+producing many of them puts pressure on the garbage collector.
 
 Type dispatching is the process of finding the concrete implementation that is
-applicable to the objects at hand when doing a generic operation on them. An
-example would be the addition of two objects: The addition needs to check what
-the concrete objects that should be added are, and choose the implementation
-that is fitting for them. Type dispatching is a very common operation in
+applicable to the objects at hand when performing a generic operation on them. An
+example would be the addition of two objects: For addition the types of the
+concrete objects need to be checked and the suiting implementation chosen.
+Type dispatching is a very common operation in
 modern\footnote{For languages in the LISP family, basic arithmetic operations
@@
 simpler than in Python or JavaScript.}
-dynamic languages because no types are known at compile time, so all operations
-need it.
+dynamic languages because no types are known at compile time. Therefore all
+operations need it.
 
@@
 languages is that of a tracing JIT. A tracing JIT works by observing the running
-program and recording its hot spots into linear execution traces. Working on
-traces is the central idea of a tracing JIT. Those traces are optimized and
-turned into machine code.
+program and recording its hot spots into \emph{linear execution traces}. Those
+traces are optimized and turned into machine code.
 
 One reason for the popularity of tracing JITs is their relative
-simplicity. They can often be added to an interpreter and a lot of the
-infrastructure of the interpreter can be reused. They give some important
+simplicity. They can often be added to an existing interpreter, reusing a lot of
+the interpreter's infrastructure. They give some important
 optimizations like inlining and constant-folding for free. A tracing JIT always
-produces linear pieces of code, which simplifies many algorithms that are usually
-hard in a compiler, such as register allocation.
-
-The usage of a tracing JIT can remove the overhead of bytecode dispatch and that
+produces linear pieces of code, which simplifies many of the hard algorithms in
+a compiler, such as register allocation.
+
+The use of a tracing JIT can remove the overhead of bytecode dispatch and that
 of the interpreter data structures. In this paper we want to present a new
@@
 techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
-interpreter as a case study.
-
-The contributions of this paper are:
+interpreter.
+
+The contributions made by this paper are:
 
 \begin{enumerate}
-    \item An efficient and effective algorithm for removing object allocations in a tracing JIT.
+    \item A description of an efficient and effective algorithm for removing
+          object allocations in a tracing JIT.
     \item A characterization of this algorithm as partial evaluation.
-    \item A rigorous evaluation of this algorithm.
+    \item Performance benchmarks for this algorithm.
 \end{enumerate}
@@
 possible. The language interpreter can then be compiled (``translated'') with
-PyPy's tools into a VM on the C level. During translation to C, many low-level
+PyPy's tools into a VM on C level. During translation to C, many low-level
 aspects of the final VM, such as object layout, garbage collection and memory
@@
 source-code hints. Mostly-automatically generating a JIT compiler has many advantages
-over writing one manually, which is an error-prone and tedious process.
+over writing one manually, an error-prone and tedious process.
 By construction, the generated JIT has the same semantics as the interpreter.
-Many optimizations can benefit all languages implemented as an interpreter in RPython.
+Optimizations can be shared between different languages implemented with PyPy.
 
@@
 to add new \emph{backends} for producing the actual machine code.  Examples of
-JIT backends that are implemented are the one for Intel x86 and x86-64 and an
+JIT backends that are implemented are those for Intel x86 and x86-64 and an
 experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.
@@
 The core idea of tracing JITs is to focus the optimization effort of the JIT
-compiler on the hot paths of the core loops of the program and to just use an
+compiler on the commonly executed, \ie \emph{hot} paths of the core loops of the program and to just use an
 interpreter for the less commonly executed parts. VMs that use a tracing JIT are
@@
 tracing JIT automatically performs inlining.
-
-This trace of operations is then the basis of the generated code. The trace is
+This trace of operations subsequently forms the basis of the generated code. The trace is
 first optimized, and then turned into machine code. Both optimization
@@
 Since the trace corresponds to one concrete execution of a loop,
-the code generated from it is only one possible path through it.
-To make sure that the trace is maintaining the correct semantics, it contains a
+the code generated from it is only one possible path through the loop.
+To make sure that the trace maintains the correct semantics, it contains a
 \emph{guard} at all places where the execution could have diverged from the
 path. Those guards check the assumptions under which execution can stay on the
-trace. As an example, if a loop contains an \lstinline{if} statement, the trace
+trace. As an example, if a loop contains an if-statement, the trace
 will contain the execution of one of the paths only, which is the path that was
 taken during the production of the trace. The trace will also contain a guard
-that checks that the condition of the \lstinline{if} statement is the same as
+that checks that the condition of the if-statement is the same as
 during tracing, because if
-it isn't, the rest of the trace is not valid. \cfbolz{The "if" shouldn't be bold}
+it isn't, the rest of the trace would not be valid.
 
@@
 
-Using these classes to implement arithmetic shows the basic problem that a
-dynamic language implementation has. All the numbers are instances of either
+Using these classes to implement arithmetic shows the basic problem of a
+dynamic language implementation. All the numbers are instances of either
 \lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
 heap. Performing many arithmetic operations produces lots of garbage quickly,
-which puts pressure on the garbage collector. Using double dispatching to
+putthing pressure on the garbage collector. Using double dispatching to
 implement the numeric tower needs two method calls per arithmetic operation,
@@
 objects repeatedly and redundantly. In addition, a lot of objects are created
-when executing that loop, many of these objects do not survive for very long.
+when executing that loop, many of these objects are short-lived.
 The actual computation that is performed by \lstinline{f} is simply a sequence of
@@
 In the example from last section, the following operations in the upper half
-of Fig.~\ref{fig:unopt-trace} produce two
+of Figure~\ref{fig:unopt-trace} produce two
 static objects, and can be completely removed from the optimized trace:
@@
 
-The subsequent operations in Fig.~\ref{fig:unopt-trace},
+The subsequent operations in Figure~\ref{fig:unopt-trace},
  which use $p_{5}$ and $p_{6}$, could then be
@@
 
-The rest of the trace from Fig.~\ref{fig:unopt-trace} is optimized similarly.
+The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized similarly.
 
@@
 
-This is what happens at the end of the trace in Fig.~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
+This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
 is hit. The arguments of the jump are at this point static objects. Before the
@@
 were lifted.
+If we had lifted $w^*$ instead of $v^*$, then the following operations would
+have been produced:
+
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+$w^*$ = new($T_2$)
+set($w^*$, $L$, $u^*$)
+set($w^*$, $R$, $u^*$)
+\end{lstlisting}
+
+In this case, the static heap afterwards would be
+$$\{v^* \mapsto (T_1, w^*, v^*)\}$$.
 

### 2010-10-15 talk/pepm2011/paper.tex
more suggestions by stephan

@@
 
-To understand the problems more directly, let us consider the simple
-interpreter function \lstinline{f} that uses the object model (see the bottom of
-Figure~\ref{fig:objmodel}).
-
+Let us now consider a simple interpreter function \lstinline{f} that uses the
+object model (see the bottom of Figure~\ref{fig:objmodel}).
 The loop in \lstinline{f} iterates \lstinline{y} times, and computes something in the process.
@@
 If the function is executed using the tracing JIT, with \lstinline{y} being a
-\lstinline{BoxedInteger}, the produced trace looks like
-Figure~\ref{fig:unopt-trace} (lines starting with the hash ``\#'' are comments).
-
-The operations in the trace are shown indented to
-correspond to the stack level of the function that contains the traced
+\lstinline{BoxedInteger}, the produced trace looks like the one of
+Figure~\ref{fig:unopt-trace} (lines starting with a hash ``\#'' are comments).
+The trace corresponds to one iteration of the while-loop in \lstinline{f}.
+
+The operations in the trace are indented
+corresponding to the stack level of the function that contains the traced
 operation. The trace is in single-assignment form, meaning that each variable is
-assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
+assigned a value exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
 to the live variables \lstinline{y} and \lstinline{res} in the original function.
@@
 \begin{itemize}
-    \item \lstinline{new} corresponds to object creation.
-    \item \lstinline{get} correspond to attribute reads.
-    \item \lstinline{set} correspond to attribute writes.
-    \item \lstinline{guard_class} correspond to method calls and are followed by
-    the trace of the called method.
+    \item \lstinline{new} creates a new object.
+    \item \lstinline{get} reads an attribute of an object.
+    \item \lstinline{set} writes to an attribute of an object.
+    \item \lstinline{guard_class} precedes an (inlined) method call and is
+    followed by the trace of the called method.
     \item \lstinline{int_add} and \lstinline{int_gt} are integer addition and
@@
 class check, not checking for subclasses.} These guards make the trace specific
-to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}, it can
-already be said to be specialized for \lstinline{BoxedIntegers}. When the trace is
-turned into machine code and then executed with \lstinline{BoxedFloats}, the
+to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}. When
+the trace is turned into machine code and afterwards executed with
+\lstinline{BoxedFloat}, the
 first \lstinline{guard_class} instruction will fail and execution will continue
@@
 have additional information attached that makes it possible to return to the
-interpreter, should the guard fail. This means that many guard operations also
-lead to a memory problem.
+interpreter, should the guard fail. This means that too many guard operations also
+consume a lot of memory.
 
@@
 
-To understand the problems that this paper is trying to solve some more, we
+To understand the problems that this paper is trying to solve in more detail, we
 first need to understand various cases of object lifetimes that can occur in a
@@
 Some of the operations within this trace are \lstinline{new} operations, which each
-create a new instance of some class. These instances are used for a while, e.g.
+create a new instance of some class. These instances are used for some time, \eg
 by calling methods on them (which are inlined into the trace), reading and
@@
 
-\begin{itemize}
-    \item Category 1: Objects that live for a while, and are then just not
-    used any more.
-
-    \item Category 2: Objects that live for a while and then escape.
-
-    \item Category 3: Objects that live for a while, survive across the jump to
+\begin{enumerate}
+    \item Objects that live for some time, and are then just not
+    used any more afterwards.
+
+    \item Objects that live for some time and then escape.
+
+    \item Objects that live for some time, survive across the jump to
     the beginning of the loop, and are then not used any more.
 
-    \item Category 4: Objects that live for a while, survive across the jump,
+    \item Objects that live for some time, survive across the jump,
     and then escape. To these we also count the objects that live across several
     jumps and then either escape or stop being used.
-\end{itemize}
+\end{enumerate}
 
@@
 in Sections~\ref{sec:statics} and \ref{sec:formal}. Objects in the other
-categories are partially optimized by this approach as well.
+categories are partially optimized by this approach as well.\footnote{We also started to
+work on optimizing objects in category 3, which will be the subject of a later
+paper.}
 
@@
 
-In this case, the static heap afterwards would be
-$$\{v^* \mapsto (T_1, w^*, v^*)\}$$.
+In this case, the static heap afterwards would be:
+$$\{v^* \mapsto (T_1, w^*, v^*)\}$$
 

### 2010-10-15 talk/pepm2011/paper.tex
make the paper sort of presentable

@@
 more.
-XXX more papers?
 
@@
 achieves.
-
-partial evaluation:
 
@@
 In functional programming this idea was introduced as
- constructor specialisation by Mogensen \cite{Mogensen:PEPM93}.)
+ constructor specialisation by Mogensen \cite{Mogensen:PEPM93}.
 
@@
 
-xxx: 
-Other related work is compile-time garbage collection \cite{mazur_practical_2001}, which
-tries to ....;
-  separation logic
+%xxx: 
+%Other related work is compile-time garbage collection \cite{mazur_practical_2001}, which
+%tries to ....;
+%  separation logic
   %; John Hughes: type specialization

### 2010-10-18 talk/pepm2011/paper.tex
comments by stephan in section 5

@@
 \end{tabular}
+\end{center}
 
@@
 \end{minipage}
-\end{center}
 \caption{The Operational Semantics of Simplified Traces}
@@
 traces and of the optimizer and liken the optimization to partial evaluation.
-We concentrate on the operations for manipulating dynamically allocated objects,
+We focus on the operations for manipulating dynamically allocated objects,
 as those are the only ones that are actually optimized. Without loss of
@@
 
-Traces are lists of operations. The operations considered here are \lstinline{new} (to make
-a new object), \lstinline{get} (to read a field out of an object), \lstinline{set} (to write a field
-into an object) and \lstinline{guard_class} (to check the type of an object). The values of all
-variables are locations (i.e.~pointers). Locations are mapped to objects, which
+Traces are lists of operations. The operations considered here are
+\lstinline{new}, \lstinline{get}, \lstinline{set} and \lstinline{guard_class}.
+The values of all
+variables are locations (\ie pointers). Locations are mapped to objects, which
 are represented by triples of a type $T$, and two locations that represent the
@@
 to null, but we require that they are initialized to a real
-location before being read, otherwise the trace is malformed.
+location before being read, otherwise the trace is malformed (this condition is
+guaranteed by how the traces are generated in PyPy).
 
@@
 interpreter formalized there executes one operation at a time. Its state is
-represented by an environment $E$ and a heap $H$, which are potentially changed by the
+represented by an environment $E$ and a heap $H$, which may be changed by the
 execution of an operation. The environment is a partial function from variables
 to locations and the heap is a partial function from locations to objects. Note
-that a variable can never be null in the environment, otherwise the trace would
-be malformed. The environment could not directly map variables to object,
-because several variables can contain a pointer to the \emph{same} object. 
-The "indirection" is needed to express sharing.
+that a variable can never be null in the environment, otherwise the trace would have
+been malformed. The environment could not directly map variables to objects,
+because several variables can point to the \emph{same} object, because of aliasing.
 
@@
 \emph{get} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[u\mapsto S(E(v))_{F}\right],S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\, u^{*}\,\mathrm{fresh}}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle u^{*}=\mathtt{get}(E(v),F)\right\rangle ,E\left[u\mapsto u^{*}\right],S}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S),\,\,\, u^{*}\,\mathrm{fresh}}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle u^{*}=\mathtt{get}(E(v),F)\right\rangle ,E\left[u\mapsto u^{*}\right],S}}$\tabularnewline[3em]
 \emph{set} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S\left[E\left(v\right)\mapsto\left(S(E(v))!_{F}E(u)\right)\right]}}$\tabularnewline[3em]
@@
 \emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\vee\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathtt{guard\_class}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\vee\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathtt{guard\_class}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
 \emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
@@
 \end{minipage}
-\end{center}
 \caption{Optimization Rules}
@@
 
-To optimize the simple traces from the last section, we use online partial
-evaluation. The partial evaluator optimizes one operation of the trace at a
+To optimize the simple traces of the last section, we use online partial
+evaluation. The partial evaluator optimizes one operation of a trace at a
 time. Every operation in the unoptimized trace is replaced by a list of
 operations in the optimized trace. This list is empty if the operation
-can be optimized away (which hopefully happens often). The optimization rules
-can be seen in Figure~\ref{fig:optimization}.
+can be optimized away. The optimization rules can be seen in
+Figure~\ref{fig:optimization}.
 
@@
 into the same variables in the optimized trace. The environment of the
-optimizer serves a function similar to that of the environment in the semantics: sharing.
+optimizer serves a function similar to that of the environment in the
+semantics: to express sharing.
 
@@
 A variable $v^*$ is in the domain of the static heap $S$ as long as the
-optimizer can fully keep track of the object. The object $S(v^*)$ is what is
-statically known about the object stored in it, \ie its type and its fields. The
+optimizer does not need to become dynamic XXX. The object $S(v^*)$ describes
+what is statically known about the object, \ie its type and its fields. The
 fields of objects in the static heap are also elements of $V^*$ (or null, for
@@
 involved variables are in the static heap, which means that the operation can be
-performed at optimization time and removed from the trace. These rules mirror
+performed at optimization time and can be removed from the trace. These rules mirror
 the execution semantics closely. The other case is for when not enough is known about
@@
 If the first argument $v$ to a \lstinline{set} operation is mapped to something in the
-static heap, then the \lstinline{set} can performed at optimization time and the static heap
-updated. Otherwise the \lstinline{set} operation needs to be residualized. This needs to be
+static heap, then the \lstinline{set} can be performed at optimization time
+(which updates the static heap). Otherwise the \lstinline{set} operation needs
+to be residualized. This needs to be
 done carefully, because the new value for the field, from the variable $u$,
@@
 
-Lifting takes a variable that is potentially in the static heap and makes sure
-that it is turned into a dynamic variable. This means that operations are
-emitted that construct an object with the shape described in the
-static heap, and the variable is removed from the static heap.
+Lifting takes a variable and turns it into a dynamic variable. If the variable
+is already dynamic, nothing needs to be done. If it is in the static heap,
+operations are emitted that construct an object with the shape described
+there, and the variable is removed from the static heap.
 
@@
 v^*), w^* \mapsto (T_2, u^*, u^*)\}$$ which contains two static objects. If $v^*$
-now needs to be lifted, the following residual operations are produced:
+needs to be lifted, the following residual operations are produced:
 
@@
 rather that \emph{in the context of tracing JITs} it is possible to find a
-simple enough algorithm that still gives very good results.
-
-Note in particular that objects in category 1 (\ie the ones that do
+simple enough algorithm that performs well.
+
+Note in particular that objects in category 1 (\ie those that do
 not escape) are completely removed; moreover, objects in category 2
-(\ie escaping) are still partially dealt with: if such an object
-escapes later than its creation point, all the operations in between that
-involve the object are removed.
+(\ie escaping) are still partially optimized: all the operations in between the
+creation of the object and the point where it escapes that involve the object
+are removed.
 
@@
 dispatching on the type of the object created just before.  A typical example
-would be chains of arithmetic operations.
+would be a chain of arithmetic operations.
 
@@
 techniques described in this paper. The actual implementation needs to deal with
-more operations than described in Section~\ref{sec:formal}, for example to
+more operations than described in Section~\ref{sec:formal}, \eg to
 also support static arrays in addition to static objects. The implementation of
@@
 novelty of our approach is that we generalized it enough to be usable for
-different interpreter.
+different interpreters.
 
@@
 numbers can be
-seen in Figure~\ref{fig:numops}. The optimization removes as many as 90\% and as
-little as 4\% percent of allocation operations in the traces of the benchmarks.
+seen in Figure~\ref{fig:numops}. The optimization removes between 4\% and 90\%
+and of allocation operations in the traces of the benchmarks.
 All benchmarks taken together, the optimization removes 70\% percent of
@@
 however there is an additional optimization that removes guards, so not all the
-removed guards are an effect of the optimization described here.
+removed guards are an effect of the optimization described here (for technical
+reasons, it would be very hard to separate the two effects).
 
 \begin{figure*}
+{\small
 \begin{center}
@@
 \end{center}
+}
 \caption{Number of Operations and Percentage Removed By Optimization}
@@
 \begin{figure*}
+{\small
 \begin{center}
@@
 \end{center}
+}
 \caption{Benchmark Times in Milliseconds, Together With Factor Over PyPy With Optimizations}
@@
 
-\section{Conclusion}
+\section{Conclusion and Future Work}
 \label{sec:conclusion}
@@
 to optimize away allocations and type guards in the traces of a
-tracing JIT.  In this context a simple approach to partial evaluation
+tracing JIT.  In this context a simple approach based on partial evaluation
 gives good results.  This is due to the fact that the tracing JIT
@@
 The authors would like to thank Stefan Hallerstede, David Schneider and Thomas
-Stiehl for fruitful discussions during the writing of the paper.
+Stiehl for fruitful discussions and detailed feedback during the writing of the
+paper.
 

### 2010-10-18 talk/pepm2011/paper.tex
small things, more fundamental changes to come

@@
 
-\subsection{Static Objects}
-
 The main insight to improve the code shown in the last section is that objects
-in category 1 don't survive very long -- they are used only inside the loop and
-nobody else in the program stores a reference to them. The idea for improving
+in category 1 do not survive very long -- they are used only inside the loop and
+there is no other outside reference to them. The idea for improving
 the code is to analyze which objects fall in category 1 and therefore do
@@
 
-This is a process that is usually called \emph{escape analysis}. In this paper we will
-perform escape analysis by using partial evaluation. The partial evalution is a
+This is a process that is usually called \emph{escape analysis}
+\cite{goldberg_higher_1990}. In this paper we will
+perform escape analysis by using partial evaluation. The use of partial evaluation is a
 bit peculiar in that it receives no static input arguments for the trace,
-but it is only used to optimize operations within a trace.
-
-The partial evaluation works by traversing the trace from beginning to end.
+but it is only used to optimize operations within the trace.
+
+To optimize the trace, it is traversed from beginning to end.
 Whenever a \lstinline{new} operation is seen, the operation is removed and a \emph{static

### 2010-10-18 talk/pepm2011/paper.tex
stop using the graphic for the traces

@@
 \begin{figure}
-\includegraphics{figures/step1.pdf}
+\begin{lstlisting}[mathescape]
+# arguments to the trace: $p_{0}$, $p_{1}$
+guard_class($p_1$, BoxedInteger)
+$i_2$ = getfield($p_1$, intval)
+guard_class($p_0$, BoxedInteger)
+$i_3$ = getfield($p_0$, intval)
+$i_4$ = int_add($i_2$, $i_3$)
+$i_9$ = int_add($i_4$, -100)
+
+guard_class($p_0$, BoxedInteger)
+$i_{12}$ = getfield($p_0$, intval)
+$i_{14}$ = int_add($i_{12}$, -1)
+
+$i_{17}$ = int_gt($i_{14}$, 0)
+guard_true($i_{17}$)
+
+$p_{15}$ = new(BoxedInteger)
+setfield($p_{15}$, intval, $i_{14}$)
+$p_{10}$ = new(BoxedInteger)
+setfield($p_{10}$, intval, $i_9$)
+
+jump($p_{15}$, $p_{10}$)
+\end{lstlisting}
 \caption{Resulting Trace After Allocation Removal}

### 2010-10-18 talk/pepm2011/paper.tex
gaaaah

--- 2010-10-18 talk/pepm2011/paper.tex
@@
 \begin{figure}
-\includegraphics{figures/step1.pdf}
+\begin{lstlisting}[mathescape]
+# arguments to the trace: $p_{0}$, $p_{1}$
+guard_class($p_1$, BoxedInteger)
+$i_2$ = get($p_1$, intval)
+guard_class($p_0$, BoxedInteger)
+$i_3$ = get($p_0$, intval)
+$i_4$ = int_add($i_2$, $i_3$)
+$i_9$ = int_add($i_4$, -100)
+
+guard_class($p_0$, BoxedInteger)
+$i_{12}$ = get($p_0$, intval)
+$i_{14}$ = int_add($i_{12}$, -1)
+
+$i_{17}$ = int_gt($i_{14}$, 0)
+guard_true($i_{17}$)
+
+$p_{15}$ = new(BoxedInteger)
+set($p_{15}$, intval, $i_{14}$)
+$p_{10}$ = new(BoxedInteger)
+set($p_{10}$, intval, $i_9$)
+
+jump($p_{15}$, $p_{10}$)
+\end{lstlisting}
 \caption{Resulting Trace After Allocation Removal}

### 2010-10-18 talk/pepm2011/paper.tex
more nonsense

@@
 guard_class($p_{5}$, BoxedInteger)
+# inside BoxedInteger.add
 $i_{7}$ = get($p_{5}$, intval)
-# inside BoxedInteger.add
 guard_class($p_{6}$, BoxedInteger)

### 2010-10-18 talk/pepm2011/paper.tex
more small tweaks, still missing a more aggressive rewrite

@@
 \subsection{Running Example}
+\label{sub:example}
 
@@
 \begin{figure}
+\lstset{
+    numbers = right,
+    numberstyle = \tiny,
+    numbersep = 0pt
+}
 \begin{lstlisting}[mathescape]
@@
 The main insight to improve the code shown in the last section is that objects
-in category 1 do not survive very long -- they are used only inside the loop and
-there is no other outside reference to them. The idea for improving
-the code is to analyze which objects fall in category 1 and therefore do
-not have to be allocated at all.
+in category 1 do not survive very long -- they are used only inside the loop
+and there is no other outside reference to them. Therefore the optimizer
+identifies objects in category 1 and removes the allocation of these objects,
+and all operations manipulating them.
 
@@
 
-In the example from last section, the following operations in the upper half
+In the example from Section~\ref{sub:example}, the following operations in the upper half
 of Figure~\ref{fig:unopt-trace} produce two
@@
 We focus on the operations for manipulating dynamically allocated objects,
-as those are the only ones that are actually optimized. Without loss of
-generality we also consider only objects with two fields in this section.
+as those are the only ones that are actually optimized. We also consider only
+objects with two fields in this section, generalizing to arbitrary many fields
+is straightforward.
 
@@
 set of static objects, which are triples of a type and two elements of $V^*$.
-A variable $v^*$ is in the domain of the static heap $S$ as long as the
-optimizer does not need to become dynamic XXX. The object $S(v^*)$ describes
+The object referenced by a variable $v^*$ is static, if $v^*$ is in the domain
+of the static heap $S$. The object $S(v^*)$ describes
 what is statically known about the object, \ie its type and its fields. The

### 2010-10-18 talk/pepm2011/paper.tex
customize line numbers in the code listing

@@
 \begin{figure}
-\begin{lstlisting}[mathescape]
-# arguments to the trace: $p_{0}$, $p_{1}$
+\begin{lstlisting}[mathescape,numbers=left,escapechar=|]
+# arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
 guard_class($p_1$, BoxedInteger)
@@
 $i_9$ = int_add($i_4$, -100)
-
+|\setcounter{lstnumber}{50}|
 guard_class($p_0$, BoxedInteger) 

### 2010-10-18 talk/pepm2011/paper.tex
add line numbers, try to gently rewrite things a bit

--- 2010-10-18 talk/pepm2011/paper.tex
@@
 
-To optimize the trace, it is traversed from beginning to end.
-Whenever a \lstinline{new} operation is seen, the operation is removed and a \emph{static
-object}\footnote{Here ``static'' is meant in the sense of partial
-evaluation, \ie known at partial evaluation time, not in the sense of static
-allocation or static method.} is constructed and associated with the variable
-that would have stored
-the result of \lstinline{new}. The static object describes the shape of the
-original object, \eg where the values that would be stored in the fields of the
-allocated object come from, as well as the type of the object. Whenever the
-optimizer sees a \lstinline{set} that writes into such an object, that shape
-description is updated and the operation can be removed, which means that the
-operation was done at partial evaluation time. When the optimizer encounters a
-\lstinline{get} from such an object, the result is read from the shape
-description, and the operation is also removed. Equivalently, a
-\lstinline{guard_class} on a variable that has a shape description can be removed
-as well, because the shape description stores the type and thus the outcome of
-the type check the guard does is statically known.
-
-In the example from Section~\ref{sub:example}, the following operations in the upper half
-of Figure~\ref{fig:unopt-trace} produce two
+To optimize the trace, it is traversed from beginning to end. Every
+optimization in the input trace is either removed, or new operations are
+produced. Whenever a \lstinline{new} operation is seen, the operation it is
+removed optimistically and a \emph{static object}\footnote{Here ``static'' is
+meant in the sense of partial evaluation, \ie known at partial evaluation time,
+not in the sense of static allocation or static method.} is constructed and
+associated with the result variable. The static object describes the shape of
+the original object, \ie where the values that would be stored in the fields of
+the allocated object come from, as well as the type of the object.
+
+When a \lstinline{set} that writes into a static object is optimized, the
+corresponding shape description is updated and the operation is removed. This
+means that the operation was done at partial evaluation time. When the
+optimizer encounters a \lstinline{get} from a static object, the result is read
+from the shape description, and the operation is also removed. Equivalently, a
+\lstinline{guard_class} on a variable that has a shape description can be
+removed as well, because the shape description stores the type and thus the
+outcome of the type check the guard does is statically known. Operations that
+have dynamic (\ie all other) objects as arguments are just left untouched by
+the optimizer.
+
+In the example from Section~\ref{sub:example}, the following operations
+of Figure~\ref{fig:unopt-trace} (lines 10-17) produce two
 static objects, and can be completely removed from the optimized trace:
@@
 
-The static object associated with $p_{5}$ would know that it is a
+The static object associated with $p_{5}$ would store the knowledge that it is a
 \lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
-one associated with $p_{6}$ would know that it is a \lstinline{BoxedInteger}
+one associated with $p_{6}$ would store that it is a \lstinline{BoxedInteger}
 whose \lstinline{intval} field contains the constant -100.
 
-The subsequent operations in Figure~\ref{fig:unopt-trace},
- which use $p_{5}$ and $p_{6}$, could then be
-optimized using that knowledge:
+The subsequent operations (line 19-29) in Figure~\ref{fig:unopt-trace}, which
+use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:
 
@@
 a static object is stored in a globally accessible place, the object has to
-be allocated, as it might live longer than one iteration of the loop
-and because the partial evaluator looses track of it. This means that the static
+be allocated, as it might live longer than one iteration of the loop and as
+arbitrary \lstinline{set} operations could change it due to aliasing. This
+means that the static
 object needs to be turned into a dynamic one, \ie lifted. This makes it
@@
 This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
-is hit. The arguments of the jump are at this point static objects. Before the
+is optimized. The arguments of the jump are at this point static objects. Before the
 jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
@@
 \begin{figure}
-\begin{lstlisting}[mathescape]
-# arguments to the trace: $p_{0}$, $p_{1}$
-guard_class($p_1$, BoxedInteger)
+\begin{lstlisting}[mathescape,numbers=right,escapechar=|,numberstyle = \tiny,numbersep=0pt, numberblanklines=false]
+# arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
+guard_class($p_1$, BoxedInteger)           |\setcounter{lstnumber}{4}|
 $i_2$ = get($p_1$, intval)
-guard_class($p_0$, BoxedInteger)
+guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{7}|
 $i_3$ = get($p_0$, intval)
-$i_4$ = int_add($i_2$, $i_3$)
-$i_9$ = int_add($i_4$, -100)
-
-guard_class($p_0$, BoxedInteger)
-$i_{12}$ = get($p_0$, intval)
-$i_{14}$ = int_add($i_{12}$, -1)
-
-$i_{17}$ = int_gt($i_{14}$, 0)
-guard_true($i_{17}$)
-
-$p_{15}$ = new(BoxedInteger)
-set($p_{15}$, intval, $i_{14}$)
-$p_{10}$ = new(BoxedInteger)
-set($p_{10}$, intval, $i_9$)
+$i_4$ = int_add($i_2$, $i_3$)              |\setcounter{lstnumber}{25}|
+$i_9$ = int_add($i_4$, -100)               |\setcounter{lstnumber}{35}|
+
+guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{38}|
+$i_{12}$ = get($p_0$, intval)              |\setcounter{lstnumber}{42}|
+$i_{14}$ = int_add($i_{12}$, -1)           |\setcounter{lstnumber}{50}|
+
+$i_{17}$ = int_gt($i_{14}$, 0)             |\setcounter{lstnumber}{53}|
+guard_true($i_{17}$)                       |\setcounter{lstnumber}{42}|
+
+$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
+set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
+$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
+set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{53}|
 
@@
 The optimized trace contains only two allocations, instead of the original five,
-and only three \lstinline{guard_class} operations, from the original seven.
+and only three \lstinline{guard_class} operations, from the original seven. The
+line numbers are the lines where the operations occurred in the original trace
+in Figure~\ref{fig:unopt-trace}.
 

### 2010-10-19 talk/pepm2011/paper.tex
fixes

@@
 To optimize the trace, it is traversed from beginning to end. Every
-optimization in the input trace is either removed, or new operations are
+operation in the input trace is either removed, or new operations are
 produced. Whenever a \lstinline{new} operation is seen, the operation it is
@@
 outcome of the type check the guard does is statically known. Operations that
-have dynamic (\ie all other) objects as arguments are just left untouched by
+have dynamic (\ie non-static) objects as arguments are just left untouched by
 the optimizer.

### 2010-10-21 talk/pepm2011/paper.tex
an attempt to rewrite section 4

@@
   %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
+  numberstyle = \tiny,
+  numbersep = 0pt,
 }
@@
 \begin{figure}
-\lstset{
-    numbers = right,
-    numberstyle = \tiny,
-    numbersep = 0pt
-}
-\begin{lstlisting}[mathescape]
+\begin{lstlisting}[mathescape,numbers = right]
 # arguments to the trace: $p_{0}$, $p_{1}$
@@
     \item \lstinline{set} writes to an attribute of an object.
-    \item \lstinline{guard_class} precedes an (inlined) method call and is
-    followed by the trace of the called method.
+    \item \lstinline{guard_class} is a precise type check and precedes an
+    (inlined) method call and is followed by the trace of the called method.
     \item \lstinline{int_add} and \lstinline{int_gt} are integer addition and
@@
 The main insight to improve the code shown in the last section is that objects
-in category 1 do not survive very long -- they are used only inside the loop
-and there is no other outside reference to them. Therefore the optimizer
-identifies objects in category 1 and removes the allocation of these objects,
-and all operations manipulating them.
+in category 1 do not survive very long -- they are used only inside the loop and
+there is no other outside reference to them. The optimizer identifies objects in
+category 1 and removes the allocation of these objects, and all operations
+manipulating them.
 
 This is a process that is usually called \emph{escape analysis}
-\cite{goldberg_higher_1990}. In this paper we will
-perform escape analysis by using partial evaluation. The use of partial evaluation is a
-bit peculiar in that it receives no static input arguments for the trace,
-but it is only used to optimize operations within the trace.
-
-To optimize the trace, it is traversed from beginning to end. Every
-operation in the input trace is either removed, or new operations are
-produced. Whenever a \lstinline{new} operation is seen, the operation it is
-removed optimistically and a \emph{static object}\footnote{Here ``static'' is
-meant in the sense of partial evaluation, \ie known at partial evaluation time,
-not in the sense of static allocation or static method.} is constructed and
-associated with the result variable. The static object describes the shape of
-the original object, \ie where the values that would be stored in the fields of
-the allocated object come from, as well as the type of the object.
-
-When a \lstinline{set} that writes into a static object is optimized, the
-corresponding shape description is updated and the operation is removed. This
-means that the operation was done at partial evaluation time. When the
-optimizer encounters a \lstinline{get} from a static object, the result is read
-from the shape description, and the operation is also removed. Equivalently, a
-\lstinline{guard_class} on a variable that has a shape description can be
-removed as well, because the shape description stores the type and thus the
-outcome of the type check the guard does is statically known. Operations that
-have dynamic (\ie non-static) objects as arguments are just left untouched by
-the optimizer.
-
-In the example from Section~\ref{sub:example}, the following operations
-of Figure~\ref{fig:unopt-trace} (lines 10-17) produce two
-static objects, and can be completely removed from the optimized trace:
+\cite{goldberg_higher_1990}. In this paper we will perform escape analysis by
+using partial evaluation. The use of partial evaluation is a bit peculiar in
+that it receives no static input arguments for the trace, but it is only used to
+optimize operations within the trace. This section will give an informal account
+of this process by examining the example trace in Figure~\ref{fig:unopt-trace}.
+The final trace after optimization can be seen in Figure~\ref{fig:step1} (the
+line numbers are the lines of the unoptimized trace where the operation comes
+from).
+
+\begin{figure}
+\begin{lstlisting}[mathescape,numbers=right,escapechar=|,numberstyle = \tiny,numbersep=0pt, numberblanklines=false]
+# arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
+guard_class($p_1$, BoxedInteger)           |\setcounter{lstnumber}{4}|
+$i_2$ = get($p_1$, intval)
+guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{7}|
+$i_3$ = get($p_0$, intval)
+$i_4$ = int_add($i_2$, $i_3$)              |\setcounter{lstnumber}{25}|
+$i_9$ = int_add($i_4$, -100)               |\setcounter{lstnumber}{35}|
+
+guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{38}|
+$i_{12}$ = get($p_0$, intval)              |\setcounter{lstnumber}{42}|
+$i_{14}$ = int_add($i_{12}$, -1)           |\setcounter{lstnumber}{50}|
+
+$i_{17}$ = int_gt($i_{14}$, 0)             |\setcounter{lstnumber}{53}|
+guard_true($i_{17}$)                       |\setcounter{lstnumber}{42}|
+
+$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
+set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
+$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
+set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{53}|
+
+jump($p_{15}$, $p_{10}$)
+\end{lstlisting}
+
+\caption{Resulting Trace After Allocation Removal}
+\label{fig:step1}
+\end{figure}
+
+To optimize the trace, it is traversed from beginning to end and an output trace
+is produced at the same time. Every operation in the input trace is either
+removed, or put into the output trace. Sometimes new operations need to be
+produced as well. The optimizer can only remove operations that manipulate
+objects that have been allocated within the trace, all others are copied to the
+output trace untouched.
+
+Looking at the example trace of Figure~\ref{fig:unopt-trace}, this is what
+happens with the operations in lines 1-9. They are manipulating objects that
+existed before the trace because they are passed in as arguments. Therefore the
+optimizer just puts them into the output trace.
+
+The following operations (lines 10-17) are more interesting:
 
@@
 \end{lstlisting}
+
+When the optimizer encounters a \lstinline{new}, it removes it optimistically,
+and assumes that the object is in category 1. When the object escapes later, it
+will be allocated at that point. The optimizer needs to keep track
+of how the object that the operation creates looks like at various points in
+the trace. This is done with the help of a \emph{static object}\footnote{Here
+``static'' is meant in the sense of partial evaluation, \ie known at partial
+evaluation time, not in the sense of ``static allocation'' or ``static
+method''.}. The static object describes the shape of the object that would have
+been allocated, \ie the type of the object and where the values that would be
+stored in the fields of the allocated object come from.
+
+In the snippet above, the two \lstinline{new} operations are removed and two
+static objects are constructed. The \lstinline{set} operations manipulate a
+static object, therefore they can be removed as well and their effect is
+remembered in the static objects.
 
@@
 
-First, the \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
-$p_{6}$ are known to be \lstinline{BoxedInteger}. Second, the \lstinline{get} operations can be removed
-and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. The only
-remaining operation in the optimized trace would be:
+The \lstinline{guard_class} operations can be removed, since their argument is a
+static object with the matching type \lstinline{BoxedInteger}. The
+\lstinline{get} operations can be removed as well, because each of them reads a
+field out of a static object. The results of the get operation are replaced with
+what the static object stores in these fields: $i_{7}$ and $i_{8}$ are just
+replaced by $i_{4}$ and -100. The only operation put into the optimized trace
+is the addition:
 
@@
 
-The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized similarly.
-
-So far we have only described what happens when static objects are used in guards and in
-operations that read and write fields. When the static
-object is used in any other operation, it cannot remain static. For example, when
-a static object is stored in a globally accessible place, the object has to
-be allocated, as it might live longer than one iteration of the loop and as
-arbitrary \lstinline{set} operations could change it due to aliasing. This
-means that the static
-object needs to be turned into a dynamic one, \ie lifted. This makes it
-necessary to put operations into the residual code that allocate the
-static object at runtime.
-
-This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
-is optimized. The arguments of the jump are at this point static objects. Before the
-jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
-that allocates a new object of the right type and sets its fields to the field
-values that the static object has (if the static object points to other static
-objects, those need to be lifted as well, recursively). This means that instead of a simple jump,
-the following operations are emitted:
+The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
+similar vein. The operations in lines 27-35 produce two more static objects and
+are removed. Those in line 36-39 are just put into the output trace because they
+manipulate objects that are allocated before the trace. Lines 40-42 are removed
+because they operate on a static object. Line 43 is put into the output trace.
+Lines 44-46 produce a new static object and are removed, lines 48-51 manipulate
+that static object and are removed as well. Lines 52-54 are put into the output
+trace.
+
+The last operation (line 55) is an interesting case. It is the \lstinline{jump}
+operation that passes control back to the beginning of the trace. The two
+arguments to this operation are at this point static objects. However, because
+they passed into the next iteration of the loop they live longer than the trace
+and therefore cannot remain static. They need to be turned into a dynamic
+(runtime) object before the actual \lstinline{jump} operation. This process of
+turning a static object into a dynamic one is called \emph{lifting}.
+
+Lifting a static object puts \lstinline{new} and \lstinline{set} operations into
+the output trace. Those operations produce an object at runtime that has the
+same shape that the static object describes. This process is a bit delicate,
+because the static objects could form an arbitrary graph structure. In our
+example is is simple, though:
 
@@
 
-Observe how the operations for creating these two instances have been moved to later point in the
-trace. 
-At first sight, it may look like for these operations we didn't gain much, as
-the objects are still allocated in the end. However, our optimizations were still
-worthwhile, because some operations that have been performed
-on the lifted static objects have been removed (some \lstinline{get} operations
-and \lstinline{guard_class} operations).
-
-\begin{figure}
-\begin{lstlisting}[mathescape,numbers=right,escapechar=|,numberstyle = \tiny,numbersep=0pt, numberblanklines=false]
-# arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
-guard_class($p_1$, BoxedInteger)           |\setcounter{lstnumber}{4}|
-$i_2$ = get($p_1$, intval)
-guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{7}|
-$i_3$ = get($p_0$, intval)
-$i_4$ = int_add($i_2$, $i_3$)              |\setcounter{lstnumber}{25}|
-$i_9$ = int_add($i_4$, -100)               |\setcounter{lstnumber}{35}|
-
-guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{38}|
-$i_{12}$ = get($p_0$, intval)              |\setcounter{lstnumber}{42}|
-$i_{14}$ = int_add($i_{12}$, -1)           |\setcounter{lstnumber}{50}|
-
-$i_{17}$ = int_gt($i_{14}$, 0)             |\setcounter{lstnumber}{53}|
-guard_true($i_{17}$)                       |\setcounter{lstnumber}{42}|
-
-$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
-set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
-$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
-set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{53}|
-
-jump($p_{15}$, $p_{10}$)
-\end{lstlisting}
-\caption{Resulting Trace After Allocation Removal}
-\label{fig:step1}
-\end{figure}
+Observe how the operations for creating these two instances have been moved to a
+later point in the trace. This is worthwhile even though the objects have to be
+allocated in the end because some \lstinline{get} operations and
+\lstinline{guard_class} operations on the lifted static objects could be
+removed.
+
+A bit more generally, lifting needs to occur if a static object is used in any
+operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
+It also needs to occur if \lstinline{set} is used to store a static object into
+a non-static one.
 
@@
 The optimized trace contains only two allocations, instead of the original five,
-and only three \lstinline{guard_class} operations, from the original seven. The
-line numbers are the lines where the operations occurred in the original trace
-in Figure~\ref{fig:unopt-trace}.
+and only three \lstinline{guard_class} operations, from the original seven.
 
@@
 time. Every operation in the unoptimized trace is replaced by a list of
-operations in the optimized trace. This list is empty if the operation
-can be optimized away. The optimization rules can be seen in
-Figure~\ref{fig:optimization}.
+operations in the optimized trace. This list is empty if the operation can be
+optimized away. The optimization rules can be seen in
+Figure~\ref{fig:optimization}. Lists are written using angular brackets $<...>$,
+list concatenation is expressed using two colons: $l_1::l_2$.
+
+XXX input/output of optimizer
 
@@
 creation of the object and the point where it escapes that involve the object
-are removed.
+are removed. Objects in category 3 and 4 are also partially optimized, their
+allocation is delayed till the end of the trace.
 
@@
 Figure~\ref{fig:times}. In addition to the run times the table also reports the
-speedup that PyPy achieves when the optimization is turned on.
+speedup that PyPy achieves when the optimization is turned on. XXX sounds sucky
 

### 2010-10-21 talk/pepm2011/paper.tex
review the new section 4

@@
 
-The main insight to improve the code shown in the last section is that objects
+The main insight to improve the code shown in section \ref{sub:example} is that objects
 in category 1 do not survive very long -- they are used only inside the loop and
@@
 produced as well. The optimizer can only remove operations that manipulate
-objects that have been allocated within the trace, all others are copied to the
+objects that have been allocated within the trace, while all other operations are copied to the
 output trace untouched.
 
-Looking at the example trace of Figure~\ref{fig:unopt-trace}, this is what
-happens with the operations in lines 1-9. They are manipulating objects that
-existed before the trace because they are passed in as arguments. Therefore the
-optimizer just puts them into the output trace.
+Looking at the example trace of Figure~\ref{fig:unopt-trace}, the operations
+in lines 1-9 are manipulating objects which existed before the trace and that
+are passed in as arguments: therefore the optimizer just puts them into the
+output trace.
 
 The following operations (lines 10-17) are more interesting:
+\anto{what about putting the line numbers also here? Moreover, figure 4 is put just after this listing, it could be confused}
 
@@
 
+\anto{I don't like ``how the object looks like''. I don't know what else to say, though :-(} 
 When the optimizer encounters a \lstinline{new}, it removes it optimistically,
-and assumes that the object is in category 1. When the object escapes later, it
+and assumes that the object is in category 1. If later the optimizer finds that the object escapes, it
 will be allocated at that point. The optimizer needs to keep track
@@
 field out of a static object. The results of the get operation are replaced with
-what the static object stores in these fields: $i_{7}$ and $i_{8}$ are just
+what the static object stores in these fields: all the occurences of $i_{7}$ and $i_{8}$ in the trace are just
 replaced by $i_{4}$ and -100. The only operation put into the optimized trace

### 2010-10-21 talk/pepm2011/paper.tex
fix some of anto's issues, will try to add line numbers now

@@
 
-The main insight to improve the code shown in section \ref{sub:example} is that objects
+The main insight to improve the code shown in Section~\ref{sub:example} is that objects
 in category 1 do not survive very long -- they are used only inside the loop and
@@
 
+To optimize the trace, it is traversed from beginning to end and an output trace
+is produced at the same time. Every operation in the input trace is either
+removed, or put into the output trace. Sometimes new operations need to be
+produced as well. The optimizer can only remove operations that manipulate
+objects that have been allocated within the trace, while all other operations are copied to the
+output trace untouched.
+
+Looking at the example trace of Figure~\ref{fig:unopt-trace}, the operations
+in lines 1-9 are manipulating objects which existed before the trace and that
+are passed in as arguments: therefore the optimizer just puts them into the
+output trace.
+
+The following operations (lines 10-17) are more interesting:
+% XXX line numbers
+
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+$p_{5}$ = new(BoxedInteger)
+set($p_{5}$, intval, $i_{4}$)
+$p_{6}$ = new(BoxedInteger)
+set($p_{6}$, intval, -100)
+\end{lstlisting}
+
+When the optimizer encounters a \lstinline{new}, it removes it optimistically,
+and assumes that the object is in category 1. If later the optimizer finds that the object escapes, it
+will be allocated at that point. The optimizer needs to keep track
+of the state of the object that the operation would have created.
+This is done with the help of a \emph{static object}\footnote{Here
+``static'' is meant in the sense of partial evaluation, \ie known at partial
+evaluation time, not in the sense of ``static allocation'' or ``static
+method''.}. The static object describes the shape of the object that would have
+been allocated, \ie the type of the object and where the values that would be
+stored in the fields of the allocated object come from.
+
+In the snippet above, the two \lstinline{new} operations are removed and two
+static objects are constructed. The \lstinline{set} operations manipulate a
+static object, therefore they can be removed as well and their effect is
+remembered in the static objects.
+
+The static object associated with $p_{5}$ would store the knowledge that it is a
+\lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
+one associated with $p_{6}$ would store that it is a \lstinline{BoxedInteger}
+whose \lstinline{intval} field contains the constant -100.
+
+The subsequent operations (line 19-29) in Figure~\ref{fig:unopt-trace}, which
+use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:
+
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+guard_class($p_{5}$, BoxedInteger)
+# inside BoxedInteger.add
+$i_{7}$ = get($p_{5}$, intval)
+guard_class($p_{6}$, BoxedInteger)
+# inside BoxedInteger.add__int
+$i_{8}$ = get($p_{6}$, intval)
+$i_{9}$ = int_add($i_{7}$, $i_{8}$)
+\end{lstlisting}
+
+The \lstinline{guard_class} operations can be removed, since their argument is a
+static object with the matching type \lstinline{BoxedInteger}. The
+\lstinline{get} operations can be removed as well, because each of them reads a
+field out of a static object. The results of the get operation are replaced with
+what the static object stores in these fields: all the occurences of $i_{7}$ and $i_{8}$ in the trace are just
+replaced by $i_{4}$ and -100. The only operation put into the optimized trace
+is the addition:
+
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+$i_{9}$ = int_add($i_{4}$, -100)
+\end{lstlisting}
+
+The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
+similar vein. The operations in lines 27-35 produce two more static objects and
+are removed. Those in line 36-39 are just put into the output trace because they
+manipulate objects that are allocated before the trace. Lines 40-42 are removed
+because they operate on a static object. Line 43 is put into the output trace.
+Lines 44-46 produce a new static object and are removed, lines 48-51 manipulate
+that static object and are removed as well. Lines 52-54 are put into the output
+trace.
+
+The last operation (line 55) is an interesting case. It is the \lstinline{jump}
+operation that passes control back to the beginning of the trace. The two
+arguments to this operation are at this point static objects. However, because
+they passed into the next iteration of the loop they live longer than the trace
+and therefore cannot remain static. They need to be turned into a dynamic
+(runtime) object before the actual \lstinline{jump} operation. This process of
+turning a static object into a dynamic one is called \emph{lifting}.
+
+Lifting a static object puts \lstinline{new} and \lstinline{set} operations into
+the output trace. Those operations produce an object at runtime that has the
+same shape that the static object describes. This process is a bit delicate,
+because the static objects could form an arbitrary graph structure. In our
+example is is simple, though:
+
+\begin{lstlisting}[mathescape,xleftmargin=20pt]
+$p_{15}$ = new(BoxedInteger)
+set($p_{15}$, intval, $i_{14}$)
+$p_{10}$ = new(BoxedInteger)
+set($p_{10}$, intval, $i_{9}$)
+jump($p_{15}$, $p_{10}$)
+\end{lstlisting}
+
+Observe how the operations for creating these two instances have been moved to a
+later point in the trace. This is worthwhile even though the objects have to be
+allocated in the end because some \lstinline{get} operations and
+\lstinline{guard_class} operations on the lifted static objects could be
+removed.
+
+A bit more generally, lifting needs to occur if a static object is used in any
+operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
+It also needs to occur if \lstinline{set} is used to store a static object into
+a non-static one.
+
+The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
+The optimized trace contains only two allocations, instead of the original five,
+and only three \lstinline{guard_class} operations, from the original seven.
+
 \begin{figure}
@@
 \end{figure}
-
-To optimize the trace, it is traversed from beginning to end and an output trace
-is produced at the same time. Every operation in the input trace is either
-removed, or put into the output trace. Sometimes new operations need to be
-produced as well. The optimizer can only remove operations that manipulate
-objects that have been allocated within the trace, while all other operations are copied to the
-output trace untouched.
-
-Looking at the example trace of Figure~\ref{fig:unopt-trace}, the operations
-in lines 1-9 are manipulating objects which existed before the trace and that
-are passed in as arguments: therefore the optimizer just puts them into the
-output trace.
-
-The following operations (lines 10-17) are more interesting:
-\anto{what about putting the line numbers also here? Moreover, figure 4 is put just after this listing, it could be confused}
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
-$p_{5}$ = new(BoxedInteger)
-set($p_{5}$, intval, $i_{4}$)
-$p_{6}$ = new(BoxedInteger)
-set($p_{6}$, intval, -100)
-\end{lstlisting}
-
-\anto{I don't like ``how the object looks like''. I don't know what else to say, though :-(} 
-When the optimizer encounters a \lstinline{new}, it removes it optimistically,
-and assumes that the object is in category 1. If later the optimizer finds that the object escapes, it
-will be allocated at that point. The optimizer needs to keep track
-of how the object that the operation creates looks like at various points in
-the trace. This is done with the help of a \emph{static object}\footnote{Here
-``static'' is meant in the sense of partial evaluation, \ie known at partial
-evaluation time, not in the sense of ``static allocation'' or ``static
-method''.}. The static object describes the shape of the object that would have
-been allocated, \ie the type of the object and where the values that would be
-stored in the fields of the allocated object come from.
-
-In the snippet above, the two \lstinline{new} operations are removed and two
-static objects are constructed. The \lstinline{set} operations manipulate a
-static object, therefore they can be removed as well and their effect is
-remembered in the static objects.
-
-The static object associated with $p_{5}$ would store the knowledge that it is a
-\lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
-one associated with $p_{6}$ would store that it is a \lstinline{BoxedInteger}
-whose \lstinline{intval} field contains the constant -100.
-
-The subsequent operations (line 19-29) in Figure~\ref{fig:unopt-trace}, which
-use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
-guard_class($p_{5}$, BoxedInteger)
-# inside BoxedInteger.add
-$i_{7}$ = get($p_{5}$, intval)
-guard_class($p_{6}$, BoxedInteger)
-# inside BoxedInteger.add__int
-$i_{8}$ = get($p_{6}$, intval)
-$i_{9}$ = int_add($i_{7}$, $i_{8}$)
-\end{lstlisting}
-
-The \lstinline{guard_class} operations can be removed, since their argument is a
-static object with the matching type \lstinline{BoxedInteger}. The
-\lstinline{get} operations can be removed as well, because each of them reads a
-field out of a static object. The results of the get operation are replaced with
-what the static object stores in these fields: all the occurences of $i_{7}$ and $i_{8}$ in the trace are just
-replaced by $i_{4}$ and -100. The only operation put into the optimized trace
-is the addition:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
-$i_{9}$ = int_add($i_{4}$, -100)
-\end{lstlisting}
-
-The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
-similar vein. The operations in lines 27-35 produce two more static objects and
-are removed. Those in line 36-39 are just put into the output trace because they
-manipulate objects that are allocated before the trace. Lines 40-42 are removed
-because they operate on a static object. Line 43 is put into the output trace.
-Lines 44-46 produce a new static object and are removed, lines 48-51 manipulate
-that static object and are removed as well. Lines 52-54 are put into the output
-trace.
-
-The last operation (line 55) is an interesting case. It is the \lstinline{jump}
-operation that passes control back to the beginning of the trace. The two
-arguments to this operation are at this point static objects. However, because
-they passed into the next iteration of the loop they live longer than the trace
-and therefore cannot remain static. They need to be turned into a dynamic
-(runtime) object before the actual \lstinline{jump} operation. This process of
-turning a static object into a dynamic one is called \emph{lifting}.
-
-Lifting a static object puts \lstinline{new} and \lstinline{set} operations into
-the output trace. Those operations produce an object at runtime that has the
-same shape that the static object describes. This process is a bit delicate,
-because the static objects could form an arbitrary graph structure. In our
-example is is simple, though:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
-$p_{15}$ = new(BoxedInteger)
-set($p_{15}$, intval, $i_{14}$)
-$p_{10}$ = new(BoxedInteger)
-set($p_{10}$, intval, $i_{9}$)
-jump($p_{15}$, $p_{10}$)
-\end{lstlisting}
-
-Observe how the operations for creating these two instances have been moved to a
-later point in the trace. This is worthwhile even though the objects have to be
-allocated in the end because some \lstinline{get} operations and
-\lstinline{guard_class} operations on the lifted static objects could be
-removed.
-
-A bit more generally, lifting needs to occur if a static object is used in any
-operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
-It also needs to occur if \lstinline{set} is used to store a static object into
-a non-static one.
-
-The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
-The optimized trace contains only two allocations, instead of the original five,
-and only three \lstinline{guard_class} operations, from the original seven.
 

### 2010-10-21 talk/pepm2011/paper.tex
add line numbers

@@
   numberstyle = \tiny,
-  numbersep = 0pt,
+  numbersep = -20pt,
 }
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
-$p_{5}$ = new(BoxedInteger)
-set($p_{5}$, intval, $i_{4}$)
-$p_{6}$ = new(BoxedInteger)
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=10]
+$p_{5}$ = new(BoxedInteger)   |\setcounter{lstnumber}{11}|
+set($p_{5}$, intval, $i_{4}$) |\setcounter{lstnumber}{14}|
+$p_{6}$ = new(BoxedInteger)   |\setcounter{lstnumber}{16}|
 set($p_{6}$, intval, -100)
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
-guard_class($p_{5}$, BoxedInteger)
-# inside BoxedInteger.add
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=20]
+guard_class($p_{5}$, BoxedInteger)  |\setcounter{lstnumber}{21}|
 $i_{7}$ = get($p_{5}$, intval)
-guard_class($p_{6}$, BoxedInteger)
-# inside BoxedInteger.add__int
+guard_class($p_{6}$, BoxedInteger)  |\setcounter{lstnumber}{24}|
 $i_{8}$ = get($p_{6}$, intval)
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=26]
 $i_{9}$ = int_add($i_{4}$, -100)
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
-$p_{15}$ = new(BoxedInteger)
-set($p_{15}$, intval, $i_{14}$)
-$p_{10}$ = new(BoxedInteger)
-set($p_{10}$, intval, $i_{9}$)
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=44]
+$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
+set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
+$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
+set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{54}|
 jump($p_{15}$, $p_{10}$)
@@
 \begin{figure}
-\begin{lstlisting}[mathescape,numbers=right,escapechar=|,numberstyle = \tiny,numbersep=0pt, numberblanklines=false]
+\begin{lstlisting}[mathescape,numbers=right,escapechar=|, numberblanklines=false]
 # arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|

### 2010-10-21 talk/pepm2011/paper.tex
A typo, and a semicolon (sorry, couldn't resist).

@@
 static objects are constructed. The \lstinline{set} operations manipulate a
-static object, therefore they can be removed as well and their effect is
+static object, therefore they can be removed as well; their effect is
 remembered in the static objects.
@@
 because the static objects could form an arbitrary graph structure. In our
-example is is simple, though:
+example it is simple, though:
 

### 2010-10-21 talk/pepm2011/paper.tex
did those
some more tweaks by stephan

@@
 The final trace after optimization can be seen in Figure~\ref{fig:step1} (the
-line numbers are the lines of the unoptimized trace where the operation comes
-from).
-
-To optimize the trace, it is traversed from beginning to end and an output trace
-is produced at the same time. Every operation in the input trace is either
-removed, or put into the output trace. Sometimes new operations need to be
+line numbers are the lines of the unoptimized trace where the operation originates).
+
+To optimize the trace, it is traversed from beginning to end while an output
+trace is produced. Every operation in the input trace is either
+removed or put into the output trace. Sometimes new operations need to be
 produced as well. The optimizer can only remove operations that manipulate
 objects that have been allocated within the trace, while all other operations are copied to the
-output trace untouched.
+output trace unchanged.
 
 Looking at the example trace of Figure~\ref{fig:unopt-trace}, the operations
-in lines 1-9 are manipulating objects which existed before the trace and that
+in lines 1--9 are manipulating objects which existed before the trace and that
 are passed in as arguments: therefore the optimizer just puts them into the
@@
 
-The following operations (lines 10-17) are more interesting:
-% XXX line numbers
+The following operations (lines 10--17) are more interesting:
 
@@
 
-The subsequent operations (line 19-29) in Figure~\ref{fig:unopt-trace}, which
+The subsequent operations (line 20--26) in Figure~\ref{fig:unopt-trace}, which
 use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:
@@
 what the static object stores in these fields: all the occurences of $i_{7}$ and $i_{8}$ in the trace are just
-replaced by $i_{4}$ and -100. The only operation put into the optimized trace
+replaced by $i_{4}$ and -100. The only operation copied into the optimized trace
 is the addition:
@@
 The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
-similar vein. The operations in lines 27-35 produce two more static objects and
-are removed. Those in line 36-39 are just put into the output trace because they
-manipulate objects that are allocated before the trace. Lines 40-42 are removed
+similar vein. The operations in lines 27--35 produce two more static objects and
+are removed. Those in line 36--39 are just put into the output trace because they
+manipulate objects that are allocated before the trace. Lines 40--42 are removed
 because they operate on a static object. Line 43 is put into the output trace.
-Lines 44-46 produce a new static object and are removed, lines 48-51 manipulate
-that static object and are removed as well. Lines 52-54 are put into the output
+Lines 44--46 produce a new static object and are removed, lines 48--51 manipulate
+that static object and are removed as well. Lines 52--54 are put into the output
 trace.
@@
 operation that passes control back to the beginning of the trace. The two
-arguments to this operation are at this point static objects. However, because
-they passed into the next iteration of the loop they live longer than the trace
-and therefore cannot remain static. They need to be turned into a dynamic
-(runtime) object before the actual \lstinline{jump} operation. This process of
+arguments to this operation at this point are static objects. However, because
+they are passed into the next iteration of the loop they live longer than the trace
+and therefore cannot remain static. They need to be turned into dynamic
+(runtime) objects before the actual \lstinline{jump} operation. This process of
 turning a static object into a dynamic one is called \emph{lifting}.
@@
 the output trace. Those operations produce an object at runtime that has the
-same shape that the static object describes. This process is a bit delicate,
+shape described by the static object. This process is a bit delicate,
 because the static objects could form an arbitrary graph structure. In our
@@
 
-A bit more generally, lifting needs to occur if a static object is used in any
+More generally, lifting needs to occur if a static object is used in any
 operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
@@
 The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
-The optimized trace contains only two allocations, instead of the original five,
-and only three \lstinline{guard_class} operations, from the original seven.
+The optimized trace contains only two allocations, instead of the original
+five, and only three \lstinline{guard_class} operations, compared to the
+original seven.
 
@@
 
-XXX input/output of optimizer
-
 The state of the optimizer is stored in an environment $E$ and a \emph{static
-heap} $S$. The environment is a partial function from variables in the
+heap} $S$. Each step of the optimizer takes an operation, an environment and a
+static heap and produces a list of operations, a new environment and a new
+static heap.
+
+The environment is a partial function from variables in the
 unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
@@
 confidence level \cite{georges_statistically_2007}. The results are reported in
-Figure~\ref{fig:times}. In addition to the run times the table also reports the
-speedup that PyPy achieves when the optimization is turned on. XXX sounds sucky
+Figure~\ref{fig:times}. For each implementation the table also reports the
+speedup that PyPy with optimization achieves over it.
 

### 2010-10-22 talk/pepm2011/paper.tex
various fixes, as well as two bugs in the math

@@
 automatic; it only needs to be guided by the language implementer using a small number of
-source-code hints. Mostly-automatically generating a JIT compiler has many advantages
+source-code hints in the interpreter. Mostly-automatically generating a JIT compiler has many advantages
 over writing one manually, an error-prone and tedious process.
@@
 
-When generating machine code, every guard is be turned into a quick check to
+When generating machine code, every guard is turned into a quick check to
 see whether the assumption still holds. When such a guard is hit during the
@@
 If one specific guard fails often enough, the tracing JIT will generate a new
-trace that starts exactly at the position of the failing guard. The existing
-assembler is patched to jump to the new trace when the guard fails
-\cite{andreas_gal_incremental_2006}.  This approach guarantees that all the
+trace that starts exactly at the position of the failing guard
+\cite{andreas_gal_incremental_2006}. The existing assembler is patched to jump
+to the new trace when the guard fails.  This approach guarantees that all the
 hot paths in the program will eventually be traced and compiled into efficient
@@
 
-Let us now consider a simple interpreter function \lstinline{f} that uses the
+Let us now consider a simple ``interpreter'' function \lstinline{f} that uses the
 object model (see the bottom of Figure~\ref{fig:objmodel}).
@@
 assigned a value exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
-to the live variables \lstinline{y} and \lstinline{res} in the original function.
+to the live variables \lstinline{y} and \lstinline{res} in the while-loop of
+the original function.
 
@@
     comparison (``greater than''), respectively.
+    \item \lstinline{guard_true} checks that a boolean is true.
 \end{itemize}
@@
 operation, to check that the class of the receiver is the same as the one that
-was observed during tracing.\footnote{\lstinline{guard_class} performs a precise
+was observed during tracing.\footnote{\lstinline{guard_class} XXX lstinline too large in footnotes
+performs a precise
 class check, not checking for subclasses.} These guards make the trace specific
@@
 
-To optimize the trace, it is traversed from beginning to end while an output
+To optimize the trace, it is traversed from beginning to end and an output
 trace is produced. Every operation in the input trace is either
-removed or put into the output trace. Sometimes new operations need to be
+removed or copied into the output trace. Sometimes new operations need to be
 produced as well. The optimizer can only remove operations that manipulate
@@
 in lines 1--9 are manipulating objects which existed before the trace and that
-are passed in as arguments: therefore the optimizer just puts them into the
+are passed in as arguments: therefore the optimizer just copies them into the
 output trace.
@@
 In the snippet above, the two \lstinline{new} operations are removed and two
-static objects are constructed. The \lstinline{set} operations manipulate a
-static object, therefore they can be removed as well; their effect is
+static objects are constructed. The \lstinline{set} operations manipulate
+static objects, therefore they can be removed as well; their effect is
 remembered in the static objects.
 
-The static object associated with $p_{5}$ would store the knowledge that it is a
+After the operations the static object associated with $p_{5}$ would store the knowledge that it is a
 \lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
@@
 
-The \lstinline{guard_class} operations can be removed, since their argument is a
-static object with the matching type \lstinline{BoxedInteger}. The
+The \lstinline{guard_class} operations can be removed, since their arguments are
+static objects with the matching type \lstinline{BoxedInteger}. The
 \lstinline{get} operations can be removed as well, because each of them reads a
@@
 similar vein. The operations in lines 27--35 produce two more static objects and
-are removed. Those in line 36--39 are just put into the output trace because they
+are removed. Those in line 36--39 are just copied into the output trace because they
 manipulate objects that are allocated before the trace. Lines 40--42 are removed
-because they operate on a static object. Line 43 is put into the output trace.
+because they operate on a static object. Line 43 is copied into the output trace.
 Lines 44--46 produce a new static object and are removed, lines 48--51 manipulate
-that static object and are removed as well. Lines 52--54 are put into the output
+that static object and are removed as well. Lines 52--54 are copied into the output
 trace.
@@
 traces and of the optimizer and liken the optimization to partial evaluation.
-We focus on the operations for manipulating dynamically allocated objects,
+We focus on the operations for manipulating heap allocated objects,
 as those are the only ones that are actually optimized. We also consider only
-objects with two fields in this section, generalizing to arbitrary many fields
+objects with two fields $L$ and $R$ in this section, generalizing to arbitrary many fields
 is straightforward.
@@
 variables are locations (\ie pointers). Locations are mapped to objects, which
-are represented by triples of a type $T$, and two locations that represent the
+are represented by triples $(T,l_1,l_2)$ of a type $T$, and two locations that represent the
 fields of the object. When a new object is created, the fields are initialized
@@
 an object, $\mathrm{type}((T,l_1,l_2))=T$ is used. Reading a field $F$ from an
-object is written $(T,l_1,l_2)_F$ which either returns $l_1$ if $F=L$ or $l_2$
+object is written $(T,l_1,l_2)_F$ which either is $l_1$ if $F=L$ or $l_2$
 if $F=R$. To set field $F$ to a new location $l$, we use the notation
@@
 \emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathtt{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
- & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathtt{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathtt{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathtt{new}\left(\mathrm{type}\left(S\left(v^{*}\right)\right)\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathtt{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathtt{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime\prime}}}$\tabularnewline[3em]
 \end{tabular}
@@
 
-While we do not offer a formal proof of it, it should be relatively clear
+While we do not offer a formal proof of it, it can argue informally
 that the algorithm presented above is sound: it works by delaying (and
@@
 is a full implementation of Python 2.5 language semantics and is about 30'000
-lines of code.
+lines of RPython code.
 
@@
 Psyco\cite{rigo_representation-based_2004} 1.6,
-an extension to CPython which is a
+a (hand-written) extension module to CPython which is a
 just-in-time compiler that produces machine code at run-time. It is not based
@@
 be written to means that only the allocations are optimized away, but not the
-reads and writes out of/into the object.
+reads out of and writes into the object.
 
@@
 way to the approach described here, as hinted at in the technical report
-\cite{michael_bebenita_spur:_2010}. However, no details for the approach and its
+\cite{bebenita_spur:_2010}. However, no details for the approach and its
 implementation are given.
@@
 Our static objects could be related to what are called \emph{virtual} objects
-in Psyco.  It is a hand-written extension module for CPython. Historically,
+in Psyco.  Historically,
 PyPy's JIT can be seen as some successor of Psyco for a general context (one of

### 2010-10-22 talk/pepm2011/paper.tex
fix citation
kill XXX, remove line overflow

@@
 
-The method calls in the trace are always preceded by a \lstinline{guard_class}
+Method calls in the trace are preceded by a \lstinline{guard_class}
 operation, to check that the class of the receiver is the same as the one that
-was observed during tracing.\footnote{\lstinline{guard_class} XXX lstinline too large in footnotes
+was observed during tracing.\footnote{\lstinline{guard_class}
 performs a precise
@@
 objects with two fields $L$ and $R$ in this section, generalizing to arbitrary many fields
-is straightforward.
-
-Traces are lists of operations. The operations considered here are
+is straightforward. Traces are lists of operations. The operations considered here are
 \lstinline{new}, \lstinline{get}, \lstinline{set} and \lstinline{guard_class}.
+
 The values of all
@@
 stopped.
+
+% subsection Operational Semantics of Traces (end)
 
@@
 In functional programming this idea was introduced as
- constructor specialisation by Mogensen \cite{Mogensen:PEPM93}.
+ constructor specialisation by Mogensen \cite{mogensen_constructor_1993}.
 

### 2010-11-17 talk/pepm2011/paper.tex
comments from the review

@@
 some light-weight profiling at the same time. This profiling is used to identify
-the hot loops of the program. If a hot loop is found in that way, the
+the hot loops\reva{what is that? clarify} of the program. If a hot loop is found in that way, the
 interpreter enters a special \emph{tracing mode}. In this tracing mode, the
@@
 
-If one specific guard fails often enough, the tracing JIT will generate a new
+If one specific guard fails often enough\reva{when is that?}, the tracing JIT will generate a new
 trace that starts exactly at the position of the failing guard
@@
 heap. Performing many arithmetic operations produces lots of garbage quickly,
-putthing pressure on the garbage collector. Using double dispatching to
+putting pressure on the garbage collector. Using double dispatching to
 implement the numeric tower needs two method calls per arithmetic operation,
@@
 the stack-allocated object. In our case, an object is not needed at all any
-more.
+more. \reva{Has this relation pointed out before in some paper? Is it a
+novel contribution of your work?}
 
@@
  constructor specialisation by Mogensen \cite{mogensen_constructor_1993}.
+\reva{I'd refer here to
+  deforestation (by Philip Wadler) and "short cut
+  deforestation" by Andrew Gill, John Launchbury, and Simon L 
+  Peyton Jones}
 

### 2010-11-17 talk/pepm2011/paper.tex
more comments

@@
 used in the generated tracing JIT of PyPy's Python interpreter. This interpreter
-is a full implementation of Python 2.5 language semantics and is about 30'000
+is a full implementation of Python 2.5 language semantics and is about 30,000
 lines of RPython code.
@@
   Peyton Jones}
+\revb{cite and relate to: boxing analysis,
+soft typing, dynamic typing, etc (by various authors: Henglein, Cartright, Felleisen,
+Leroy)}
 

### 2010-11-17 talk/pepm2011/paper.tex
fix a tiny problem in the formulas: something should not be in italics
ignore things where you have to change the language
- add two deforestation references - squeeze code a bit to add some space

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass{sigplanconf}
 
@@
 \CopyrightYear{2011}
-\copyrightdata{[to be supplied]}
+\copyrightdata{[to be supplied]} % XXX
 
@@
 \begin{figure}
-\begin{lstlisting}[mathescape]
+\begin{lstlisting}[mathescape,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 class Base(object):
@@
 \begin{figure}
-\begin{lstlisting}[mathescape,numbers = right]
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 # arguments to the trace: $p_{0}$, $p_{1}$
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=10]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=10,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 $p_{5}$ = new(BoxedInteger)   |\setcounter{lstnumber}{11}|
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=20]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=20,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 guard_class($p_{5}$, BoxedInteger)  |\setcounter{lstnumber}{21}|
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=26]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=26,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 $i_{9}$ = int_add($i_{4}$, -100)
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=44]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=44,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 $p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
@@
 \begin{figure}
-\begin{lstlisting}[mathescape,numbers=right,escapechar=|, numberblanklines=false]
+\begin{lstlisting}[mathescape,numbers=right,escapechar=|, numberblanklines=false,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 # arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
@@
  & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathtt{new}\left(\mathrm{type}\left(S\left(v^{*}\right)\right)\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
- & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathtt{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathtt{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime\prime}}}$\tabularnewline[3em]
+ & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::\mathrm{ops}_{R}::\left\langle \mathtt{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathtt{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime\prime}}}$\tabularnewline[3em]
 \end{tabular}
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 $v^*$ = new($T_1$)
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
 $w^*$ = new($T_2$)
@@
 \begin{figure*}
-{\small
+{\footnotesize
 \begin{center}
@@
  constructor specialisation by Mogensen \cite{mogensen_constructor_1993}.
-\reva{I'd refer here to
-  deforestation (by Philip Wadler) and "short cut
-  deforestation" by Andrew Gill, John Launchbury, and Simon L 
-  Peyton Jones}
-\revb{cite and relate to: boxing analysis,
-soft typing, dynamic typing, etc (by various authors: Henglein, Cartright, Felleisen,
-Leroy)}
+\reva{I'd refer here to \cite{wadler_deforestation:_1988}
+  and \cite{gill_short_1993}}
+\revb{cite and relate to: boxing analysis}
 

### 2010-11-18 talk/pepm2011/paper.tex
copyright data
add related work. now we are at 11 pages again.

@@
 \begin{document}
-\conferenceinfo{PEPM}{'11, Austin, USA}
+\conferenceinfo{PEPM'11,} {January 24--25, 2011, Austin, Texas, USA.}
 \CopyrightYear{2011}
-\copyrightdata{[to be supplied]} % XXX
+\copyrightdata{978-1-4503-0485-6/11/01}
 
@@
 \section{Introduction}
+
+% XXX insert "practical" somewhere
 
@@
  constructor specialisation by Mogensen \cite{mogensen_constructor_1993}.
-\reva{I'd refer here to \cite{wadler_deforestation:_1988}
-  and \cite{gill_short_1993}}
-\revb{cite and relate to: boxing analysis}
+
+A related optimization is also that of deforestation
+\cite{wadler_deforestation:_1988,gill_short_1993} which tries to remove
+intermediate lists or trees in functional languages. More explicitly geared at
+optimizing boxing is boxing analysis \cite{jrgensen_calculus_1996} which tries
+to fully remove pairs of calls to box/unbox in a functional language.
 

### 2010-11-18 talk/pepm2011/paper.tex
try to use the word "hot" only after it is defined

@@
 
-% XXX insert "practical" somewhere
-
 The objective of a just-in-time (JIT) compiler for a dynamic language is to
@@
 languages is that of a tracing JIT. A tracing JIT works by observing the running
-program and recording its hot spots into \emph{linear execution traces}. Those
+program and recording its commonly executed parts into \emph{linear execution traces}. Those
 traces are optimized and turned into machine code.
@@
 \begin{enumerate}
-    \item A description of an efficient and effective algorithm for removing
-          object allocations in a tracing JIT.
+    \item A description of a practical, efficient and effective algorithm for
+    removing object allocations in a tracing JIT.
     \item A characterization of this algorithm as partial evaluation.
@@
 some light-weight profiling at the same time. This profiling is used to identify
-the hot loops\reva{what is that? clarify} of the program. If a hot loop is found in that way, the
+the hot loops of the program. If a hot loop is found in that way, the
 interpreter enters a special \emph{tracing mode}. In this tracing mode, the

### 2010-11-18 talk/pepm2011/paper.tex
next review comment

@@
 
-If one specific guard fails often enough\reva{when is that?}, the tracing JIT will generate a new
+If one specific guard fails a lot (\ie more than some threshold), the tracing JIT will generate a new
 trace that starts exactly at the position of the failing guard

### 2010-11-18 talk/pepm2011/paper.tex
I will ignore this comment
get space back. bit non-beautiful, but well.

@@
 \lstset{
-  basicstyle=\setstretch{1.1}\ttfamily\footnotesize,
+  basicstyle=\setstretch{1.05}\ttfamily\footnotesize,
   language=Python,
@@
 \begin{figure}
-\begin{lstlisting}[mathescape,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 class Base(object):
@@
 \begin{figure}
-\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 # arguments to the trace: $p_{0}$, $p_{1}$
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=10,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=10,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 $p_{5}$ = new(BoxedInteger)   |\setcounter{lstnumber}{11}|
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=20,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=20,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 guard_class($p_{5}$, BoxedInteger)  |\setcounter{lstnumber}{21}|
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=26,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=26,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 $i_{9}$ = int_add($i_{4}$, -100)
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=44,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=44,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 $p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
@@
 \begin{figure}
-\begin{lstlisting}[mathescape,numbers=right,escapechar=|, numberblanklines=false,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,numbers=right,escapechar=|, numberblanklines=false,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 # arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 $v^*$ = new($T_1$)
@@
 
-\begin{lstlisting}[mathescape,xleftmargin=20pt,basicstyle=\setstretch{1.1}\ttfamily\scriptsize]
+\begin{lstlisting}[mathescape,xleftmargin=20pt,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
 $w^*$ = new($T_2$)
@@
 the stack-allocated object. In our case, an object is not needed at all any
-more. \reva{Has this relation pointed out before in some paper? Is it a
-novel contribution of your work?}
+more.
 

### 2010-11-18 talk/pepm2011/paper.tex
add dynamic typing reference

@@
 A related optimization is also that of deforestation
-\cite{wadler_deforestation:_1988,gill_short_1993} which tries to remove
-intermediate lists or trees in functional languages. More explicitly geared at
-optimizing boxing is boxing analysis \cite{jrgensen_calculus_1996} which tries
-to fully remove pairs of calls to box/unbox in a functional language.
-
-%xxx:
-%partially static data structures: kenichi asai's thesis?
-
-%xxx: 
-%Other related work is compile-time garbage collection \cite{mazur_practical_2001}, which
-%tries to ....;
-%  separation logic
-  %; John Hughes: type specialization
+\cite{wadler_deforestation:_1988,gill_short_1993} which removes intermediate
+lists or trees in functional languages. A more general approach is boxing
+analysis \cite{jrgensen_calculus_1996} which optimizes pairs of calls to
+box/unbox in a functional language. Similarly, "dynamic typing"
+\cite{Henglein:1994:DTS:190865.190867} tries to remove dynamic type coercions
+in a dynamically typed lambda-calculus. All these optimizations work by
+analyzing the program before execution, which makes them unsuitable for dynamic
+languages like Python, where almost nothing can be inferred purely by looking
+at the source code.
 

### 2011-01-05 talk/pepm2011/presentation/talk.tex
start the talk. will become too long if I continue like this

\documentclass[utf8x]{beamer}

% This file is a solution template for:

% - Talk at a conference/colloquium.
% - Talk length is about 20min.
% - Style is ornate.

\mode<presentation>
{
  \usetheme{Warsaw}
  % or ...

  %\setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}

\usepackage[english]{babel}
\usepackage{listings}

\usepackage[utf8x]{inputenc}
% or whatever

% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.

\title{Allocation Removal by Partial Evaluation in a Tracing JIT}

\author[Bolz et. al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{1} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and \\
            Samuele Pedroni\inst{3} \and Armin Rigo\inst{1}}
\author{\emph{Carl Friedrich Bolz} \and Antonio Cuni \and Maciej Fijałkowski \and Michael Leuschel \and Samuele Pedroni \and Armin Rigo}
% - Give the names in the same order as the appear in the paper.
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\institute[Heinrich-Heine-Universität Düsseldorf]
{Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany \and

 merlinux GmbH, Hildesheim, Germany \and

 Open End, Göteborg, Sweden \and
}

\date{2011 Workshop on Partial Evaluation and Program Manipulation, January 24, 2011}
% - Either use conference name or its abbreviation.
% - Not really informative to the audience, more for people (including
%   yourself) who are reading the slides online

% If you have a file called "university-logo-filename.xxx", where xxx
% is a graphic format that can be processed by latex or pdflatex,
% resp., then you can add a logo as follows:

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
%\AtBeginSubsection[]
%{
%  \begin{frame}<beamer>
%    \frametitle{Outline}
%    \tableofcontents[currentsection,currentsubsection]
%  \end{frame}
%}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

%\begin{frame}
%  \frametitle{Outline}
%  \tableofcontents
  % You might wish to add the option [pausesections]
%\end{frame}

% Structuring a talk is a difficult task and the following structure
% may not be suitable. Here are some rules that apply for this
% solution:

% - Exactly two or three sections (other than the summary).
% - At *most* three subsections per section.
% - Talk about 30s to 2min per frame. So there should be between about
%   15 and 30 frames, all told.

% - A conference audience is likely to know very little of what you
%   are going to talk about. So *simplify*!
% - In a 20min talk, getting the main ideas across is hard
%   enough. Leave out details, even if it means being less precise than
%   you think necessary.
% - If you omit details that are vital to the proof/implementation,
%   just say so once. Everybody will be happy with that.

\begin{frame}
  \frametitle{Dynamic Languages are Slow}
  \begin{itemize}
      \item Interpretation overhead
      \item Constant type dispatching
      \item Boxing of primitive types
      \pause
      \begin{itemize}
          \item A lot of allocation of short-lived objects
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Dynamic Languages are Slow: Example}
  Evaluate \texttt{x = a + b; y = x + c} in an interpreter:
  \pause
  \begin{enumerate}
      \item What's the type of a? \texttt{Integer}
      \item What's the type of b? \texttt{Integer}
  \pause
      \item unbox a
      \item unbox b
      \item compute the sum
      \item box the result
      \item store into x
  \pause
      \item What's the type of x? \texttt{Integer}
      \item What's the type of c? \texttt{Integer}
  \pause
      \item unbox x
      \item unbox c
      \item compute the sum
      \item box the result
      \item store into y
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{What to do?}
  \begin{itemize}
      \item Hard to improve in an interpreter
      \item Use a JIT compiler
      \item \textbf{Add a optimization that can deal with heap operations}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Context: The PyPy Project}
  A general environment for implementing dynamic languages
  \pause
  \begin{block}{Approach}
      \begin{itemize}
          \item write an interpreter for the language in RPython
          \item compilable to an efficient C-based VM
          \pause
          \item (RPython is a restricted subset of Python)
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{PyPy's Tracing JIT}
  the feature that makes PyPy interesting:
  \begin{itemize}
      \item a meta-JIT, applicable to many languages
      \item needs a few source-code hints (or user annotations) in the interpreter
      \item JIT is a tracing JIT compiler
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Tracing JITs}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \item for interesting, commonly executed code paths
      \item produces a linear list of operations (trace)
      \item trace is optimized turned into machine code
  \end{itemize}
\end{frame}

\begin{frame}[containsverbatim]
  \frametitle{Example Trace}
  Trace of \texttt{x = a + b; y = x + c}:
\begin{verbatim}
guard_class(a, Integer)
guard_class(b, Integer)
i1 = get(a, intval)
i2 = get(b, intval)
i3 = int_add(i1, i2)
x = new(Integer)
set(x, intval, i3)
\end{verbatim}
\pause
\begin{verbatim}
guard_class(x, Integer)
guard_class(c, Integer)
i4 = get(x, intval)
i5 = get(c, intval)
i6 = int_add(i4, i5)
y = new(Integer)
set(y, intval, i6)
\end{verbatim}
\end{frame}

\begin{frame}
  \frametitle{Tracing JIT: Advantages}
  \begin{itemize}
      \item Traces are interesting linear pieces of code
      \item most of the time correspond to loops
      \item everything called in the trace is inlined
      \item can perform good optimizations on the loop
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Optimizing the Heap Operations in a Trace}
  \begin{itemize}
      \item Contribution of our paper
      \item A simple, efficient and effective optimization of heap operations in a trace
      \item using online partial evaluation
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Heap Model}
  \includegraphics[scale=0.9]{figures/heap01}
\end{frame}

\begin{frame}
  \frametitle{Heap Model}
  \includegraphics[scale=0.9]{figures/heap02}
\end{frame}

\begin{frame}
  \frametitle{Heap Model}
  \includegraphics[scale=0.9]{figures/heap03}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: New}
  \includegraphics[scale=0.8]{figures/new01}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: New}
  \includegraphics[scale=0.8]{figures/new02}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: Get}
  \includegraphics[scale=0.8]{figures/get01}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: Get}
  \includegraphics[scale=0.8]{figures/get02}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: Set}
  \includegraphics[scale=0.8]{figures/set01}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: Set}
  \includegraphics[scale=0.8]{figures/set02}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: Guard}
  \includegraphics[scale=0.8]{figures/guard01}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: Guard}
  \includegraphics[scale=0.8]{figures/guard02}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: Guard}
  \includegraphics[scale=0.8]{figures/guard03}
\end{frame}

\begin{frame}[plain]
  \frametitle{Operations: Guard}
  \includegraphics[scale=0.8]{figures/guard04}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
      \item straightforward interpreter can be efficient, given enough technology
      \item successful application of partial evaluation
      \item Prolog can benefit from dynamic compilation
  \end{itemize}
  \pause
  \begin{block}{Future}
      \begin{itemize}
          \item Scale up to larger programs
          \item need some optimization on the interpreter level – indexing
          \item investigate memory usage
      \end{itemize}
  \end{block}
\end{frame}

\end{document}

### 2011-01-11 talk/pepm2011/presentation/talk.tex
continue working on the talk. will have to do things differently, the diagrams are nice but take way too much space.

@@
       \item using online partial evaluation
+      \item fully implemented an in use in large-scale interpreters
   \end{itemize}
+  \pause
+  \begin{block}{Ingredients}
+      \begin{itemize}
+          \item a slightly simplified model for objects on the heap
+          \item operational semantics of trace operations that manipulate the heap
+          \item optimization rules for those operations
+      \end{itemize}
+  \end{block}
 \end{frame}
@@
   \includegraphics[scale=0.9]{figures/heap03}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Operations Manipulating Heap Objects}
+  \begin{itemize}
+      \item \texttt{v = new(T)} makes a new object
+      \item \texttt{get(w, F)} reads a field out of an object
+      \item \texttt{u = set(v, F, w)} writes a field of an object
+      \item \texttt{guard(v, T)} checks the type of an object
+  \end{itemize}
 \end{frame}
@@
 \begin{frame}
+  \frametitle{Optimization by Online Partial Evaluation}
+  \begin{itemize}
+      \item Trace is optimized using online partial evaluation
+      \item part of the runtime heap is modelled in the \emph{static heap}
+      \item static heap contains objects that are allocated within the trace
+      \item (as opposed to before the trace is executed)
+      \pause
+      \item operations acting on the static heap can be removed
+      \item all others need to be residualized
+      \item all fairly straightforward
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing New}
+  \includegraphics[scale=0.8]{figures/opt_new1}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing New}
+  \includegraphics[scale=0.8]{figures/opt_new2}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Get}
+  \includegraphics[scale=0.8]{figures/opt_get1}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Get}
+  \includegraphics[scale=0.8]{figures/opt_get2}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Get}
+  \includegraphics[scale=0.8]{figures/opt_get3}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Get}
+  \includegraphics[scale=0.8]{figures/opt_get4}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Guard}
+  \includegraphics[scale=0.8]{figures/opt_guard1}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Guard}
+  \includegraphics[scale=0.8]{figures/opt_guard2}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Guard}
+  \includegraphics[scale=0.8]{figures/opt_guard3}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Guard}
+  \includegraphics[scale=0.8]{figures/opt_guard4}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Set}
+  \includegraphics[scale=0.8]{figures/opt_set1}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Set}
+  \includegraphics[scale=0.8]{figures/opt_set2}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Lifting}
+  Problem: What happens if we write a static object into a dynamic one?
+  \pause
+  \begin{itemize}
+      \item lose track of the static object because of possibility of aliasing
+      \item need to \emph{lift} the static object
+      \item lifting produces operations that recreate the static object
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Set}
+  \includegraphics[scale=0.8]{figures/opt_set_dynamic1}
+\end{frame}
+
+\begin{frame}[plain]
+  \frametitle{Optimizing Set}
+  \includegraphics[scale=0.8]{figures/opt_set_dynamic2}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Benchmark Results}
+  \begin{itemize}
+      \item to evaluate the optimization we used PyPy's Python interpreter with real-world programs
+      \item optimization can remove
+      \begin{itemize}
+          \item 70\% of all \texttt{new} operations
+          \item 14\% of all \texttt{get/set} operations
+          \item 93\% of all \texttt{guard} operations
+      \pause
+      \item Timings improve by a factor between 1.2 and 6.95
+      \item outperforming standard Python on all benchmarks but one
+      \pause
+      \item more details in the paper
+      \end{itemize}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
   \frametitle{Conclusion}

### 2011-01-11 talk/pepm2011/presentation/talk.tex
tweaks

@@
 \usepackage{listings}
+\usepackage{ulem}
+\usepackage{alltt}
 
@@
   \frametitle{Dynamic Languages are Slow: Example}
-  Evaluate \texttt{x = a + b; y = x + c} in an interpreter:
+  \texttt{x = a + b; y = x + c}
+
+  evaluated in an interpreter:
   \pause
@@
       \item compute the sum
-      \item box the result
+      \item box the result as an \texttt{Integer}
       \item store into x
@@
       \item compute the sum
-      \item box the result
+      \item box the result as an \texttt{Integer}
       \item store into y
@@
       \item \textbf{Add a optimization that can deal with heap operations}
+      \pause
+      \item optimize short-lived objects
   \end{itemize}
@@
 \pause
+\begin{verbatim}
+guard_class(x, Integer)
+guard_class(c, Integer)
+i4 = get(x, intval)
+i5 = get(c, intval)
+i6 = int_add(i4, i5)
+y = new(Integer)
+set(y, intval, i6)
+\end{verbatim}
+\end{frame}
+
+\begin{frame}[containsverbatim]
+  \frametitle{Example Trace}
+  Trace of \texttt{x = a + b; y = x + c}:
+\begin{verbatim}
+guard_class(a, Integer)
+guard_class(b, Integer)
+i1 = get(a, intval)
+i2 = get(b, intval)
+i3 = int_add(i1, i2)
+x = new(Integer)
+set(x, intval, i3)
+\end{verbatim}
 \begin{verbatim}
@@
 
+\begin{frame}[containsverbatim]
+  \frametitle{Optimized Example Trace}
+  Trace of \texttt{x = a + b; y = x + c}:
+\begin{alltt}
+guard_class(a, Integer)
+guard_class(b, Integer)
+i1 = get(a, intval)
+i2 = get(b, intval)
+i3 = int_add(i1, i2)
+\sout{x = new(Integer)}
+\sout{set(x, intval, i3)}
+\sout{guard_class(x, Integer)}
+guard_class(c, Integer)
+\sout{i4 = get(x, intval)}
+i5 = get(c, intval)
+i6 = int_add(i4, \emph{i3})
+y = new(Integer)
+set(y, intval, i6)
+\end{alltt}
+\end{frame}
+
 \begin{frame}
@@
   \begin{itemize}
-      \item straightforward interpreter can be efficient, given enough technology
+      \item very simple partial-evaluation-based
       \item successful application of partial evaluation

### 2011-01-11 talk/pepm2011/presentation/talk.tex
fix conclusion

@@
   \begin{itemize}
-      \item very simple partial-evaluation-based
-      \item successful application of partial evaluation
-      \item Prolog can benefit from dynamic compilation
+      \item very simple partial-evaluation-based optimization
+      \item can remove a lot of allocations and type checks in practical programs
+      \item no control issues
+      \item all control decisions made by the tracer
+      \item based on observing executing program
   \end{itemize}
-  \pause
-  \begin{block}{Future}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Questions?}
   \begin{itemize}
-          \item Scale up to larger programs
-          \item need some optimization on the interpreter level – indexing
-          \item investigate memory usage
+      \item very simple partial-evaluation-based optimization for tracing JITs of dynamic languages
+      \item can remove a lot of allocations and type checks in practical programs
+      \item no control issues
+      \item all control decisions made by the tracing JIT
+      \item based on observing executing program
   \end{itemize}
-  \end{block}
 \end{frame}

### 2011-01-11 talk/pepm2011/presentation/talk.tex
fixes and a small addition
Fix typo.

@@
       \item produces a linear list of operations (trace)
-      \item trace is optimized turned into machine code
+      \item trace is optimized and then turned into machine code
   \end{itemize}
@@
       \item can perform good optimizations on the loop
+      \item rarer paths run by the interpreter
   \end{itemize}
@@
       \item using online partial evaluation
-      \item fully implemented an in use in large-scale interpreters
+      \item fully implemented and in use in large-scale interpreters
   \end{itemize}
@@
 i5 = get(c, intval)
-i6 = int_add(i4, \emph{i3})
+i6 = int_add(\emph{i3}, i5)
 y = new(Integer)

### 2011-01-11 talk/pepm2011/presentation/talk.tex
backup slides

@@
 
+\begin{frame}
+  \frametitle{Backup Slides}
+\end{frame}
+
+\begin{frame}
+  \frametitle{What About Correctness?}
+  \begin{itemize}
+      \item We haven't proven correctness yet
+      \item should not be too hard
+      \item lifting needs to be carefully handled
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Comparison to Escape Analysis}
+  \begin{itemize}
+      \item Effect very similar to escape analysis
+      \item Escape analysis needs a complex upfront analysis
+      \item our optimization automatically has a lot of context, due to the inlining tracing does
+      \item our optimization can optimize operations on objects even if they escape later
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[containsverbatim]
+  \frametitle{Comparison to "Dynamic Typing"/Boxing Analysis}
+  \begin{itemize}
+      \item those optimizations work ahead of time
+      \item don't work for many dynamic languages, where the source simply does not contain enough information
+  \end{itemize}
+
+  \begin{block}{Python Example:}
+  \begin{verbatim}
+def sum(container, initial):
+    result = initial
+    for element in container:
+        result = result + element
+    return result
+  \end{verbatim}
+  \end{block}
+\end{frame}
+
 \end{document}

### 2011-01-19 talk/pepm2011/presentation/talk.tex
tweaks to the slides after the first test run

@@
 
-\author[Bolz et. al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{1} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and \\
-            Samuele Pedroni\inst{3} \and Armin Rigo\inst{1}}
-\author{\emph{Carl Friedrich Bolz} \and Antonio Cuni \and Maciej Fijałkowski \and Michael Leuschel \and Samuele Pedroni \and Armin Rigo}
+\author[Carl Friedrich Bolz et.
+al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{1} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and Samuele Pedroni\inst{3} \and Armin Rigo\inst{1}}
 % - Give the names in the same order as the appear in the paper.
@@
 \institute[Heinrich-Heine-Universität Düsseldorf]
-{Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany \and
+{$^1$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany \and
 
- merlinux GmbH, Hildesheim, Germany \and
+ $^2$merlinux GmbH, Hildesheim, Germany \and
 
- Open End, Göteborg, Sweden \and
+ $^3$Open End, Göteborg, Sweden \and
 }
@@
 
-%\begin{frame}
-%  \frametitle{Outline}
-%  \tableofcontents
-  % You might wish to add the option [pausesections]
-%\end{frame}
-
 % Structuring a talk is a difficult task and the following structure
@@
       \item Interpretation overhead
-      \item Constant type dispatching
+      \item Type dispatching
       \item Boxing of primitive types
@@
       \item compute the sum
+  \pause
       \item box the result as an \texttt{Integer}
@@
   \begin{itemize}
-      \item Hard to improve in an interpreter
       \item Use a JIT compiler
@@
       \item optimize short-lived objects
+      \item remove some of the redundancy
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{Context: The PyPy Project}
+  \frametitle{Overview}
+  \tableofcontents
+  % You might wish to add the option [pausesections]
+\end{frame}
+
+\section{Experimental Context}
+
+\begin{frame}
+  \frametitle{Our Experimental Context is the PyPy Project}
   A general environment for implementing dynamic languages
@@
 \begin{frame}
-  \frametitle{PyPy's Tracing JIT}
+  \frametitle{PyPy Uses a Novel Tracing JIT}
   the feature that makes PyPy interesting:
@@
 \begin{frame}
-  \frametitle{Tracing JITs}
+  \frametitle{Tracing JITs Compile by Observing an Interpreter}
   \begin{itemize}
@@
 
+\begin{frame}[containsverbatim]
+  \frametitle{Optimized Example Trace}
+  Trace of \texttt{x = a + b; y = x + c}:
+\begin{alltt}
+guard_class(a, Integer)
+guard_class(b, Integer)
+i1 = get(a, intval)
+i2 = get(b, intval)
+i3 = int_add(i1, i2)
+\sout{x = new(Integer)}
+\sout{set(x, intval, i3)}
+\end{alltt}
+\begin{alltt}
+\sout{guard_class(x, Integer)}
+guard_class(c, Integer)
+\sout{i4 = get(x, intval)}
+i5 = get(c, intval)
+i6 = int_add(\emph{i3}, i5)
+y = new(Integer)
+set(y, intval, i6)
+\end{alltt}
+\end{frame}
+
 \begin{frame}
-  \frametitle{Tracing JIT: Advantages}
+  \frametitle{The Advantages of Tracing JITs}
   \begin{itemize}
@@
 \end{frame}
+
+\section{Proposed Optimization}
 
@@
       \item \texttt{v = new(T)} makes a new object
-      \item \texttt{get(w, F)} reads a field out of an object
-      \item \texttt{u = set(v, F, w)} writes a field of an object
+      \item \texttt{u = get(w, F)} reads a field out of an object
+      \item \texttt{set(v, F, w)} writes a field of an object
       \item \texttt{guard(v, T)} checks the type of an object
@@
 
+\section{Benchmarks}
+
 \begin{frame}
@@
       \pause
-      \item Timings improve by a factor between 1.2 and 6.95
+      \item Timings improve by a factor between 1.1 and 6.95
       \item outperforming standard Python on all benchmarks but one
@@
   \begin{itemize}
-      \item very simple partial-evaluation-based optimization
-      \item can remove a lot of allocations and type checks in practical programs
-      \item no control issues
-      \item all control decisions made by the tracer
-      \item based on observing executing program
+      \item We propose a very simple partial-evaluation-based optimization for tracing
+      JITs of dynamic languages that:
+      \begin{itemize}
+          \item can remove a lot of allocations and type checks in practical
+          programs. 
+          \item is efficient and effective.
+          \item has no control issues because all control decisions are made by
+          the tracing JIT.
       \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Questions?}
-  \begin{itemize}
-      \item very simple partial-evaluation-based optimization for tracing JITs of dynamic languages
-      \item can remove a lot of allocations and type checks in practical programs
-      \item no control issues
-      \item all control decisions made by the tracing JIT
-      \item based on observing executing program
+      \pause
+      \item We claim that this is a general strategy to get rid of control
+      problems by simply observing the runtime behaviour of the program.
   \end{itemize}

### 2011-01-19 talk/pepm2011/presentation/talk.tex
a few visual bugs.
prize question: why does inkscape insert more and more empty lines into its svg
files?
feedback from the second training presentation

@@
 \usepackage{ulem}
+\usepackage{color}
 \usepackage{alltt}
@@
 \usepackage[utf8x]{inputenc}
+
+\newcommand\redsout[1]{{\color{red}\sout{\hbox{\color{black}{#1}}}}}
+
 % or whatever
@@
       \item Use a JIT compiler
-      \item \textbf{Add a optimization that can deal with heap operations}
+      \item \textbf{Add an optimization that can deal with heap operations}
       \pause
@@
       \item JIT works by observing and logging what the interpreter does
+      \begin{itemize}
           \item for interesting, commonly executed code paths
           \item produces a linear list of operations (trace)
+      \end{itemize}
       \item trace is optimized and then turned into machine code
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{The Advantages of Tracing JITs}
+  \begin{itemize}
+      \item Traces are interesting linear pieces of code
+      \item most of the time correspond to loops
+      \item everything called in the trace is inlined
+      \item can perform good optimizations on the trace
+      \item rarer paths run by the interpreter
   \end{itemize}
@@
 i3 = int_add(i1, i2)
-\sout{x = new(Integer)}
-\sout{set(x, intval, i3)}
+\redsout{x = new(Integer)}
+\redsout{set(x, intval, i3)}
 \end{alltt}
 \begin{alltt}
-\sout{guard_class(x, Integer)}
+\redsout{guard_class(x, Integer)}
 guard_class(c, Integer)
-\sout{i4 = get(x, intval)}
+\redsout{i4 = get(x, intval)}
 i5 = get(c, intval)
@@
 
-\begin{frame}
-  \frametitle{The Advantages of Tracing JITs}
-  \begin{itemize}
-      \item Traces are interesting linear pieces of code
-      \item most of the time correspond to loops
-      \item everything called in the trace is inlined
-      \item can perform good optimizations on the loop
-      \item rarer paths run by the interpreter
-  \end{itemize}
-\end{frame}
-
 \section{Proposed Optimization}
@@
 \begin{frame}
-  \frametitle{Optimizing the Heap Operations in a Trace}
+  \frametitle{Contribution of our Paper}
   \begin{itemize}
-      \item Contribution of our paper
-      \item A simple, efficient and effective optimization of heap operations in a trace
+      \item a simple, efficient and effective optimization of heap operations in a trace
       \item using online partial evaluation
@@
       \item lifting produces operations that recreate the static object
+      \pause
+      \item needs to be careful due to recursive structures
   \end{itemize}
@@
   \end{itemize}
-
   \begin{block}{Python Example:}

### 2011-01-21 talk/pepm2011/presentation/talk.tex
tweaks
todos left
- add 17 new diagrams that work out the example in detail - add a diagram for the benchmarks

@@
 \end{frame}
+
+% XXX todos:
+% note that two fields is a simplification
+% have a diagram for the example trace
+% lifting can never produce new operations
+% have some sort of overview or position indicator for the many graphical slides, too easy to get lost
+% more details about benchmarks, diagram?
 
@@
           \item operational semantics of trace operations that manipulate the heap
-          \item optimization rules for those operations
+          \item optimization rules for those operations, following the operational semantics
       \end{itemize}
@@
   \frametitle{Operations: New}
+%  \begin{columns}[c]
+%  \begin{column}{10cm}
     \includegraphics[scale=0.8]{figures/new01}
+%  \end{column}
+%  \begin{column}{2cm}
+%  {\tiny 
+%  \begin{itemize}
+%    \item new
+%    \item get
+%    \item set
+%    \item guard
+%  \end{itemize}
+%  }
+%  \end{column}
+%  \end{columns}
 \end{frame}
@@
       \pause
-      \item operations acting on the static heap can be removed
+      \item operations acting on the static heap can be executed
+      \item follows operational semantics
       \item all others need to be residualized
@@
 
-\begin{frame}[containsverbatim]
-  \frametitle{Optimized Example Trace}
-  Trace of \texttt{x = a + b; y = x + c}:
-\begin{alltt}
-guard_class(a, Integer)
-guard_class(b, Integer)
-i1 = get(a, intval)
-i2 = get(b, intval)
-i3 = int_add(i1, i2)
-\sout{x = new(Integer)}
-\sout{set(x, intval, i3)}
-\sout{guard_class(x, Integer)}
-guard_class(c, Integer)
-\sout{i4 = get(x, intval)}
-i5 = get(c, intval)
-i6 = int_add(\emph{i3}, i5)
-y = new(Integer)
-set(y, intval, i6)
-\end{alltt}
+\begin{frame}[plain]
+    \frametitle{Optimizing the Example Trace}
+\only<1>
+{\includegraphics[scale=0.8]{figures/ex00}}
+\only<2>
+{\includegraphics[scale=0.8]{figures/ex01}}
+\only<3>
+{\includegraphics[scale=0.8]{figures/ex02}}
+\only<4>
+{\includegraphics[scale=0.8]{figures/ex03}}
+\only<5>
+{\includegraphics[scale=0.8]{figures/ex04}}
+\only<6>
+{\includegraphics[scale=0.8]{figures/ex05}}
+\only<7>
+{\includegraphics[scale=0.8]{figures/ex06}}
+\only<8>
+{\includegraphics[scale=0.8]{figures/ex07}}
+\only<9>
+{\includegraphics[scale=0.8]{figures/ex08}}
+\only<10>
+{\includegraphics[scale=0.8]{figures/ex09}}
+\only<11>
+{\includegraphics[scale=0.8]{figures/ex10}}
+\only<12>
+{\includegraphics[scale=0.8]{figures/ex11}}
+\only<13>
+{\includegraphics[scale=0.8]{figures/ex12}}
+\only<14>
+{\includegraphics[scale=0.8]{figures/ex13}}
+\only<15>
+{\includegraphics[scale=0.8]{figures/ex14}}
+\only<16>
+{\includegraphics[scale=0.8]{figures/ex15}}
+\only<17>
+{\includegraphics[scale=0.8]{figures/ex16}}
 \end{frame}
@@
       \item Timings improve by a factor between 1.1 and 6.95
-      \item outperforming standard Python on all benchmarks but one
-      \pause
-      \item more details in the paper
+      \item outperforming standard Python on all benchmarks but two
       \end{itemize}
   \end{itemize}
+\end{frame}
+
+\section{Benchmarks}
+
+\begin{frame}
+  \frametitle{Benchmark}
+  \includegraphics[scale=0.4]{figures/benchmarks}
 \end{frame}

### 2011-02-07 talk/ibm-feb-2011/talk.tex
replace requirements stuff with a table

\documentclass[utf8x]{beamer}

\mode<presentation>
{
  \usetheme{Warsaw}

  %\setbeamercovered{transparent}
}

\usepackage[english]{babel}

\usepackage[utf8x]{inputenc}

\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{listings}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{alltt}
\usepackage{tikz}
\usetikzlibrary{arrows,positioning,shadows,shapes,calc,backgrounds,fit}

%\input{pygmentheader.tex}

\title[PyPy Implements Dynamic Languages Using a Tracing JIT]{
PyPy's Approach to \\
Implementing Dynamic Languages \\
Using a Tracing JIT Compiler}

\author{Carl Friedrich Bolz}

\institute[Heinrich-Heine-Universität Düsseldorf]
{
  Institut für Informatik\\
  Heinrich-Heine-Universität Düsseldorf
}

\date{
%\includegraphics[scale=0.25]{figures/hpi_logo_kl.jpg}\\
IBM Watson Research Center, February 8th, 2011
}

%\pgfdeclareimage[height=0.5cm]{pypy-logo}{image/py-web.png}
%\logo{\pgfuseimage{pypy-logo}}

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
%\AtBeginSubsection[]
%{
%  \begin{frame}<beamer>
%    \frametitle{Outline}
%    \tableofcontents[currentsection,currentsubsection]
%  \end{frame}
%}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
  \frametitle{Scope}
  This talk is about:

  \begin{itemize}
  \item implementing dynamic languages \par(with a focus on complicated ones)
  \item in a context of limited resources \par(academic, open source, or
    domain-specific)
  \item imperative, object-oriented languages
  \item single-threaded implementations
  \end{itemize}
  \pause
  \begin{block}{Goals}
    Reconciling:
    \begin{itemize}
    \item flexibility, maintainability (because languages evolve)
    \item simplicity (because teams are small)
    \item performance
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Outline}
  \tableofcontents[pausesections]
  % You might wish to add the option [pausesections]
\end{frame}

\section{The Difficulties of Implementing Dynamic Languages}

\subsection{Technical Factors}

\begin{frame}
  \frametitle{What is Needed Anyway}
  A lot of things are not really different from other languages:
  \begin{itemize}
      \item lexer, parser
      \item (bytecode) compiler
      \item garbage collector
      \item object system
  \end{itemize}
\end{frame}

% say, but don't write:
% from pypy perspective, but trying to do a general analysis
% feedback/discussion welcome
%  XXX random thought: a lot of semantics implemented on VM level, operations
%  XXX other direction: PyPy has granularity issues in opposite direction

\begin{frame}
  \frametitle{Control Flow}
    \begin{itemize}
        \item every language implementation needs a way to implement the control flow of the language
        \item trivially and slowly done in interpreters with AST or bytecode
        \item technically very well understood
        \item sometimes small difficulties, like generators in Python
        \pause
        \item some languages have more complex demands\\
        but this is rare
        \item examples: Prolog
    \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Late Binding}
  \begin{itemize}
      \item lookups can be done only at runtime
      \item historically, dynamic languages have moved to ever later binding times
      \item a large variety of mechanisms exist in various languages
      \item mechanism are often very ad-hoc because \\
      "it was easy to do in an interpreter"
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Late Binding in Python}
  In Python, the following things are late bound
  \begin{itemize}
      \item global names
      \item modules
      \item instance variables
      \item methods
      \pause
      \item the class of objects
      \item class hierarchy
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Dispatching}
  \begin{itemize}
      \item dispatching is a very important special case of late binding
      \item how are the operations on objects implemented?
      \item this is usually very complex, and different between languages
      \pause
      \item operations are internally split up into one or several lookup and call steps
      \item a huge space of paths ensues
      \item most of the paths are uncommon
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Example: Attribute Reads in Python}
  What happens when an attribute \texttt{x.m} is read? (simplified)
  \begin{itemize}
      \item check for the presence of \texttt{x.\_\_getattribute\_\_}, if there, call it
      \pause
      \item look for the name of the attribute in the object's dictionary, if it's there, return it
      \pause
      \item walk up the MRO of the object and look in each class' dictionary for the attribute
      \pause
      \item if the attribute is found, call its \texttt{\_\_get\_\_} attribute and return the result
      \pause
      \item if the attribute is not found, look for \texttt{x.\_\_getattr\_\_}, if there, call it
      \pause
      \item raise an \texttt{AttributeError}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Example: Addition in Python}
  \includegraphics[scale=0.5]{figures/add.pdf}
\end{frame}

\begin{frame}
  \frametitle{Dependencies Between Subsequent Dispatches}
  \begin{itemize}
      \item one dispatch operation is complex
      \item many in a sequence are worse
      \item take \texttt{(a + b) + c}
      \item the dispatch decision of the first operation influences the second
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Boxing of Primitive Values}
  \begin{itemize}
      \item primitive values often need to be boxed,\\
      to ensure uniform access
      \item a lot of pressure is put on the GC by arithmetic
      \item need a good GC (clear anyway)
      \item in arithmetic, lifetime of boxes is known
  \end{itemize}
%  \pause
%  \begin{block}{Escape Analysis?}
%    \begin{itemize}
%        \item should help in theory
%        \item nearly always there are some unlikely paths,\\
%        that lead to escapes
%        \item this is often due to user-overriding of operations
%    \end{itemize}
%  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Escaping Paths}
  \begin{itemize}
      \item considering again \texttt{(a + b) + c}
      \item assume \texttt{a} and \texttt{b} are ints
      \item then the result should not be allocated
      \item escaping path: if \texttt{c} has a user-defined class
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{(Frames)}
  \begin{itemize}
      \item side problem:
      \item many languages have reified frame access
      \item e.g. Python, Smalltalk, Ruby, ...
      \item support for in-language debuggers
      \item in an interpreter these are trivial,\\
      because the interpreter needs them anyway
      \item how should reified frames work efficiently when a compiler is used?
  \end{itemize}
\end{frame}

%\begin{frame}
%  \frametitle{(Fast Access to Native Libraries)}
%  \begin{itemize}
%      \item another side problem
%      \item how can native libraries be accessed efficiently?
%      \item important in dynamic languages, because they are often used as glue
%      \item even more important in browser, to interact with the DOM
%      \item can be solved either by compiling extensions, or via an FFI
%      \item both not very efficient
%  \end{itemize}
%\end{frame}

\subsection{Requirements}

\begin{frame}
  \frametitle{Summarizing the Requirements}
  \begin{enumerate}
      \item control flow
      \item late binding
      \item \textbf{dispatching}
      \item \textbf{dependencies between subsequent dispatches}
      \item boxing
      \item (reified frames)
%      \item (access to native libraries)
  \end{enumerate}
\end{frame}

\section{Approaches For Dynamic Language Implementation}

\begin{frame}
  \frametitle{Common Approaches to Language Implementation}
  \begin{itemize}
    \item Using C/C++
    \begin{itemize}
      \item for an interpreter
      \item for a static compiler
      \item for a method-based JIT
      \item for a tracing JIT
    \end{itemize}
    \item Building on top of a general-purpose OO VM
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Common Approaches to Language Implementation}
  \begin{block}{
    Using C/C++}
    \begin{itemize}
    \item CPython (interpreter)
    \item Ruby (interpreter)
    \item V8 (method-based JIT)
    \item TraceMonkey (tracing JIT)
    \item ...
    %\item but also: Scheme48, Squeak (interpreters)
    \end{itemize}
  \end{block}
  \pause
  \begin{block}{Building on top of a general-purpose OO VM}
    \begin{itemize}
    \item Jython, IronPython
    \item JRuby, IronRuby
    \item various Prolog, Lisp, even Smalltalk implementations
    \end{itemize}
  \end{block}
\end{frame}

\subsection{Implementing VMs in C/C++}

\begin{frame}
  \frametitle{Implementing VMs in C}
  When writing a VM in C it is hard to reconcile our goals
  \begin{itemize}
  \item flexibility, maintainability
  \item simplicity
  \item performance
  \end{itemize}
  \pause
  \begin{block}{Python Case}
    \begin{itemize}
    \item \alert{CPython} is a very simple bytecode VM, performance not great
    \item \alert{Psyco} is a just-in-time-specializer, very complex, hard to
      maintain, but good performance
    \item \alert{Stackless} is a fork of CPython adding microthreads. It was
    never incorporated into CPython for complexity reasons
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Interpreters in C/C++}
  \begin{itemize}
      \item mostly very easy
      \item well understood problem
      \item portable, maintainable
      \item slow
  \end{itemize}

\end{frame}

%\begin{frame}
%  \frametitle{Fixing of Early Design Decisions}
%  \begin{itemize}
%  \item when starting a VM in C, many design decisions need to be made upfront
%  \item examples: memory management technique, threading model
%  \item such decisions are manifested throughout the VM source
%  \item very hard to change later
%  \item low level details mixed with language semantics
%  \end{itemize}
%  \pause
%  \begin{block}{Python Case}
%    \begin{itemize}
%    \item CPython uses reference counting, increfs and decrefs everywhere
%    \item CPython uses OS threads with one global lock, hard to change to
%      lightweight threads or finer locking
%    \end{itemize}
%  \end{block}
%\end{frame}

\begin{frame}
  \frametitle{How do Interpreters Meet the Requirements?}
  \includegraphics[scale=0.6]{figures/output1.pdf}
%  \begin{enumerate}
%      \item \alert{control flow slowed by bytecode or AST dispatch overhead}
%      \item \alert{late binding works but slow}
%      \item \alert{dispatching works but slow}
%      \item \alert{no special dependencies support}
%      \item \alert{everything is boxed}
%      \item \alert{reified frames are easy but slow}
%%      \item \alert{access to native libraries easy but slow}
%  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{Static Compilers to C/C++}
  \begin{itemize}
      \item first reflex of many people is to blame it all on bytecode dispatch overhead
      \item thus static compilers are implemented that reuse the object model of an interpreter
      \item gets rid of interpretation overhead only
      \item seems to give about 2x speedup
      \pause
      \item dispatch, late-binding and boxing only marginally improved
      \item static analysis mostly never works
%      \item often easy access to native libraries
  \end{itemize}
  \pause
  \begin{block}{Python Case}
    \begin{itemize}
    \item \alert{Cython}, \alert{Pyrex} are compilers\\
    from large subsets of Python to C
    \item lots of older experiments, most discontinued
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{How do Static Compilers Meet the Requirements?}
  \includegraphics[scale=0.6]{figures/output2.pdf}
%  \begin{enumerate}
%      \item control flow works well
%      \item \alert{late binding not improved}
%      \item \alert{dispatching not improved}
%      \item \alert{dependencies not improved}
%      \item \alert{everything is boxed}
%      \item \alert{reified frames often not supported}
%%      \item direct access to native libraries
%  \end{enumerate}
\end{frame}

\subsection{Method-Based JIT Compilers}

\begin{frame}
  \frametitle{Method-Based JIT Compilers}
  \begin{itemize}
      \item to fundamentally attack some of the problems, \\
      a dynamic compiler is needed
      \item a whole new can of worms
      \pause
      \begin{itemize}
          \item type profiling
          \item inlining based on that
          \item general optimizations
          \item complex backends
      \end{itemize}
      \item very hard to pull off for a volunteer team
  \end{itemize}
  \pause
  \begin{block}{Examples}
      \begin{itemize}
          \item Smalltalk and SELF JITs
          \item V8 and JägerMonkey
          \item Psyco, sort of
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Compilers are a bad encoding of Semantics}
  \begin{itemize}
  \item to improve all complex corner cases of the language, \\
  a huge effort is needed
  \item often needs a big "bag of tricks"
  \item the interactions between all tricks is hard to foresee
  \item the encoding of language semantics in the compiler is thus often obscure and hard to change
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item Psyco is a dynamic compiler for Python
    \item synchronizing with CPython's development is a lot of effort
    \item many of CPython's new features not supported well
    \item not ported to 64-bit machines, and probably never will
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}[containsverbatim]
  \frametitle{Method-Based JITs and Dispatching Dependencies}
\begin{verbatim}
    x = add(a, b)
    r = add(x, c)
\end{verbatim}
\end{frame}

\begin{frame}[containsverbatim, plain]
  \frametitle{Method-Based JITs and Dispatching Dependencies}
\begin{verbatim}
    if isinstance(a, Integer):
        if isinstance(b, Integer):
            x = Integer(<perform int addition>)
        else:
            x = Float(<perform float addition>)
    else:
        x = Float(<perform float addition>)
    # x can be Float or Integer here
    if isinstance(x, Integer):
        if isinstance(c, Integer):
            r = Integer(<perform int addition>)
        else:
            r = Float(<perform float addition>)
    else:
        r = Float(<perform float addition>)
\end{verbatim}
\end{frame}

\begin{frame}
  \frametitle{How do Method-Based JIT Compilers Meet the Requirements?}
  \includegraphics[scale=0.6]{figures/output3.pdf}
%  \begin{enumerate}
%      \item control flow works well
%      \item late binding can be handled
%      \item dispatching can be handled
%      \item \alert{dependencies not necessarily improved}
%      \item boxing hard to support
%      \item \alert{reified frames hard to support}
%%      \item \alert{access to native libraries not improved}
%  \end{enumerate}
\end{frame}

\subsection{Tracing JIT Compilers}

\begin{frame}
  \frametitle{Tracing JIT Compilers}
  \begin{itemize}
      \item relatively recent approach to JIT compilers
      \item pioneered by Michael Franz and Andreas Gal for Java
      \item turned out to be well-suited for dynamic languages
  \end{itemize}
  \pause
  \begin{block}{Examples}
      \begin{itemize}
          \item \alert{TraceMonkey}
          \item \alert{LuaJIT}
          \item \alert{SPUR}, sort of
          \item \alert{PyPy}, sort of
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
    \frametitle{Tracing JIT Compilers}
    \begin{itemize}
    \item idea from Dynamo project: \\
    dynamic rewriting of machine code
    \item conceptually simpler than type profiling
    \end{itemize}
    \pause
    \begin{block}{Basic Assumption of a Tracing JIT}
        \begin{itemize}
        \item programs spend most of their time executing loops
        \item several iterations of a loop are likely to take similar code paths
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Tracing VMs}
    \begin{itemize}
    \item mixed-mode execution environment
    \item at first, everything is interpreted
    \item lightweight profiling to discover hot loops
    \item code generation only for common paths of hot loops
    \item when a hot loop is discovered, start to produce a trace
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Tracing}
    \begin{itemize}
    \item a \emph{trace} is a sequential list of operations
    \item a trace is produced by recording every operation the interpreter executes
    \item tracing ends when the tracer sees a position in the program it has seen before
    \item a trace thus corresponds to exactly one loop
    \item that means it ends with a jump to its beginning
    \end{itemize}
    \pause
    \begin{block}{Guards}
        \begin{itemize}
        \item the trace is only one of the possible code paths through the loop
        \item at places where the path \emph{could} diverge, a guard is placed
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Code Generation and Execution}
    \begin{itemize}
    \item being linear, the trace can easily be turned into machine code
    \item execution stops when a guard fails
    \item after a guard failure, go back to interpreting program
    \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Dealing With Control Flow}
  \begin{itemize}
      \item an if statement in a loop is turned into a guard
      \item if that guard fails often, things are inefficient
      \item solution: attach a new trace to a guard, if it fails often enough
      \item new trace can lead back to same loop
      \item or to some other loop
  \end{itemize}
\end{frame}

%\begin{frame}
%  \frametitle{Stages of Execution}
%  \includegraphics[scale=0.5]{figures/tracing.pdf}
%\end{frame}

\begin{frame}
  \frametitle{Dispatching in a Tracing JIT}
  \begin{itemize}
      \item trace contains bytecode operations
      \item bytecodes often have complex semantics
      \item optimizer often type-specializes the bytecodes
      \item according to the concrete types seen during tracing
      \item need to duplicate language semantics in optimizer for that
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Dispatching Dependencies in a Tracing JIT}
  \begin{itemize}
      \item one consequence of the tracing approach:
      \item paths are split aggressively
      \item control flow merging happens at beginning of loop only
      \item after a type check, the rest of the trace can assume that type
      \item only deal with paths that are actually seen
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Boxing Optimizations in a Tracing JIT}
  \begin{itemize}
      \item possibility to do escape analysis within the trace
      \item only optimize common path
      \item i.e. the one where the object doesn't escape
  \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Advantages of Tracing JITs}
    \begin{itemize}
    \item can be added to an existing interpreter unobtrusively
    \item interpreter does most of the work
    \item automatic inlining
    \item deals well with finding the few common paths through the large space
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Bad Points of the Approach}
    \begin{itemize}
        \item switching between interpretation and machine code execution takes time
        \item problems with really complex control flow
        \item granularity issues: often interpreter bytecode is too coarse
        \item if this is the case, the optimizer needs to carefully re-add the decision tree
    \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{How do Tracing JITs Meet the Requirements?}
  \includegraphics[scale=0.6]{figures/output4.pdf}
%  \begin{enumerate}
%      \item control flow works rather well
%      \item late binding can be handled
%      \item dispatching can be handled
%      \item dependencies improved by path splitting
%      \item unboxing optimization much simpler
%      \item reified frames can be implemented \\
%      by falling back to the interpreter
%%      \item \alert{access to native libraries not improved}
%  \end{enumerate}
\end{frame}

\subsection{Building on Top of an OO VM}

\begin{frame}
  \frametitle{Implementing Languages on Top of OO VMs}
  \begin{itemize}
  \item approach: implement on top of the JVM or the CLR
  \item usually by compiling to the target bytecode
  \item plus an object model implementation
  \item brings its own set of benefits of problems
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item \alert{Jython} is a Python-to-Java-bytecode compiler
    \item \alert{IronPython} is a Python-to-CLR-bytecode compiler
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Benefits of Implementing on Top of OO VMs}
  \begin{itemize}
  \item higher level of implementation
  \item the VM supplies a GC and a JIT
  \item better interoperability than what the C level provides
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item both Jython and IronPython integrate well with their host OO VM
    \item both have proper threading
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The Problems of OO VMs}
  \begin{itemize}
      \item often hard to map concepts of the dynamic language
      \item performance not improved because of the semantic mismatch
      \item untypical code in most object models
      \item object model typically has many megamorphic call sites
      \pause
      \item escape analysis cannot help with boxing,\\
      due to escaping paths
      \item to improve, very careful manual tuning is needed
      \item VM does not provide enough customization/feedback
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Examples of Problems}
  \begin{itemize}
      \item both Jython and IronPython are quite a bit slower than CPython
      \item IronPython misses reified frames
      \pause
      \item for languages like Prolog it is even harder to map the concepts
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Future of OO VMs?}
  \begin{itemize}
  \item the problems described might improve in the future
  \item JVM will add extra support for more languages
  \item i.e. tail calls, \texttt{InvokeDynamic}, ...
  \item has not really landed yet
  \item good performance needs a huge amount of tweaking
  \item controlling the VM's behaviour is brittle:\\
  VMs not meant for people who care about exact shape of assembler
  \end{itemize}
  \pause
  \begin{block}{Ruby Case}
    \begin{itemize}
    \item JRuby tries really hard to be a very good implementations
    \item took an enormous amount of effort
    \item tweaking is essentially Hotspot-specific
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{How do OO VMs Meet the Requirements?}
  \includegraphics[scale=0.6]{figures/output5.pdf}
%  \begin{enumerate}
%      \item control flow works well
%      \item late binding can be handled with a lot of effort
%      \item dispatching can be handled with a lot of effort
%      \item \alert{dependencies not improved}
%      \item \alert{boxing not necessarily improved}
%      \item \alert{reified frames are inefficient}
%%      \item access to native libraries not improved, but VM libraries help
%  \end{enumerate}
\end{frame}

\section{PyPy's Approach to VM Construction}

\begin{frame}
  \frametitle{The PyPy Project}
  \begin{itemize}
      \item started in 2003, received funding from the EU, Google, Nokia and some smaller companies
      \item goal: "The PyPy project aims at producing a flexible and fast Python implementation."
      \item technology should be reusable for other dynamic languages
  \end{itemize}
  \pause
  \begin{block}{Language Status}
    \begin{itemize}
        \item the fastest Python implementation, very complete
        \item contains a reasonably good Prolog
        \item full Squeak, but no JIT for that yet
        \item various smaller experiments (JavaScript, Scheme, Haskell)
    \end{itemize}

  \end{block}
\end{frame}

\begin{frame}
  \frametitle{PyPy's Approach to VM Construction}
  \emph{Goal: achieve flexibility, simplicity and performance together}

  \begin{itemize}
  \item Approach: auto-generate VMs from high-level descriptions of the language
  \item ... using meta-programming techniques and \emph{aspects}
  \item high-level description: an interpreter written in a high-level language
  \item ... which we translate (i.e.\ compile) to a VM running in various target
    environments, like C/Posix\pause, CLR, JVM
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{PyPy}
  \begin{itemize}
  \item PyPy = Python interpreter written in RPython + translation toolchain
    for RPython
  \end{itemize}
  \pause
  \begin{block}{What is RPython}
    \begin{itemize}
    \item RPython is a (large) subset of Python
    \item subset chosen in such a way that type-inference can be performed
    \item still a high-level language (unlike SLang or PreScheme)
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item we need a custom \emph{translation toolchain} to compile the interpreter
    to a full VM
  \item many aspects of the final VM are orthogonal from the interpreter source:
    they are inserted during translation
  \end{itemize}
  \pause
  \begin{block}{
    Examples}
    \begin{itemize}
    \item Garbage Collection strategy
%    \item Threading models (e.g.\ coroutines with CPS...)
    \item non-trivial translation aspect: auto-generating a tracing JIT compiler from
      the interpreter
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Good Points of the Approach}
  {\bf Simplicity:} separation of language semantics from low-level details
  \pause

  {\bf Flexibility} high-level implementation language eases things (meta-programming)
  \pause

  {\bf Performance:} ``reasonable'' baseline performance, can be very good with JIT
\end{frame}

\begin{frame}[plain]
  \includegraphics[scale=0.3]{figures/all_numbers.png}
\end{frame}

\subsection{PyPy's Meta-Tracing JIT Compiler}

%\begin{frame}
%  \frametitle{Example}
%XXX the idea of this section is to have a running example of a small
%interpreter that is extended with language constructs to provide examples for
%all the technologies
%
%\end{frame}

\begin{frame}
  \frametitle{Meta-Tracing}
  Problems of Tracing JITs:
  \begin{itemize}
      \item specific to one language's bytecode
      \item bytecode has wrong granularity
      \item internals of an operation not visible in trace
  \end{itemize}
  \pause
  \begin{block}{PyPy's Idea:}
      \begin{itemize}
          \item write interpreters in RPython
          \item trace the execution of the RPython code
          \item using one generic RPython tracer
          \item the process is customized via hints in the interpreter
          \item no language-specific bugs
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Interpreter Overhead}
  \begin{itemize}
      \item most immediate problem with meta-tracing
      \item interpreter typically has a bytecode dispatch loop
      \item not a good idea to trace that
      \pause
      \item solved by a simple trick:
      \item \emph{unroll} the bytecode dispatch loop
      \item control flow then taken care of
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Optimizing Late Binding and Dispatching}
  \begin{itemize}
      \item late binding and dispatching code in the interpreter is traced
      \item as in a normal tracing JIT, the meta-tracer is good at picking common paths
      \item a number of hints to fine-tune the process
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Optimizing Boxing Overhead}
  \begin{itemize}
      \item boxing optimized by a powerful general optimization on traces
      \item tries to defer allocations for as long as possible
      \item allocations only happen in those (rare) paths where they are needed
  \end{itemize}
  \pause
  \begin{block}{Use Cases}
      \begin{itemize}
          \item arithmetic
          \item argument holder objects
          \item frames of inlined functions
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Dealing With Reified Frames}
  \begin{itemize}
      \item interpreter needs a frame object to store its data anyway
      \item those frame objects are specially marked
      \item JIT special-cases them
      \item their attributes can live in CPU registers/stack
      \item on reflective access, machine code is left, interpreter continues
      \pause
      \item nothing deep, but a lot of engineering
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Feedback from the VM}
  \begin{itemize}
      \item in the beginning the hints are often not optimal yet
      \item to understand how to improve them, the traces must be read
      \item traces are in a machine-level intermediate representation
      \item not machine code
      \item corresponds quite closely to RPython interpreter code
      \item visualization and profiling tools
  \end{itemize}
\end{frame}

%\subsection{Customizing the Tracing of the Interpreter}
%
%\begin{frame}
%  \frametitle{Customizing the Tracing of the Interpreter}
%  - general building blocks for implementing many different dynamic languages well
%
%  - more expedient than minimal
%
%  without hints, behaviour is correct but not very fast
%\end{frame}
%
%
%\begin{frame}
%  \frametitle{The Extent of Tracing}
%  \begin{itemize}
%      \item tracer always starts at bytecode dispatch loop
%      \item at any interpreter-level function call, the tracer can decide to trace into call, or not
%      \item tracing should not continue arbitrarily deep into implementation of operations
%      \pause
%      \item basic mechanism: stop at functions that contain loops
%      \item hint to stop tracing explicitly
%      \item hint to force tracing into function with loops, unrolling them
%  \end{itemize}
%\end{frame}
%
%\begin{frame}
%  \frametitle{Example}
%  XXX
%\end{frame}
%
%
%\begin{frame}
%  \frametitle{Mechanisms for Dispatching and Late-Binding}
%  - tracing time computation
%
%  - promotion, turn a variable into a constant to insert a guard
%
%  - immutability, to fold away reads out of constants
%
%  - pure function hints, remove function calls if the arguments are pure
%
%\end{frame}
%
%\begin{frame}
%  \frametitle{Example Promotion}
%  XXX maps
%\end{frame}
%
%\begin{frame}
%  \frametitle{Example Immutability}
%  XXX code objects
%\end{frame}
%
%
%\begin{frame}
%  \frametitle{Example Pure Function}
%  XXX version tags
%\end{frame}
%
%\begin{frame}
%  \frametitle{Out-of-Line Guards}
%  - soon: slow-changing fields into out-of-line guards
%\end{frame}
%
%\begin{frame}
%  \frametitle{Example Out-of-Line Guards}
%  XXX version tags
%\end{frame}

\begin{frame}
  \frametitle{Drawbacks / Open Issues / Further Work}
  \begin{itemize}
  \item writing the translation toolchain in the first place takes lots of effort
    (but it can be reused)
  \item writing a good GC was still necessary, not perfect yet
  \item dynamic compiler generation seems to work now, but took \emph{very} long to get right
  \item granularity of tracing is sometimes not optimal, very low level
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
      \item PyPy solves many of the problems of dynamic language implementations
      \item it uses a high-level language
      \begin{itemize}
          \item to ease implementation
          \item for better analyzability
      \end{itemize}
      \item it gives good feedback to the language implementor
      \item and provides various mechanisms to express deeply different language semantics
      \pause
      \item only one solution in this design space (SPUR is another)
      \item more experiments needed
  \end{itemize}
\end{frame}

\end{document}

### 2011-02-07 talk/ibm-feb-2011/talk.tex
replace pseudo-code by diagram

@@
   \frametitle{Method-Based JITs and Dispatching Dependencies}
-\begin{verbatim}
-    if isinstance(a, Integer):
-        if isinstance(b, Integer):
-            x = Integer(<perform int addition>)
-        else:
-            x = Float(<perform float addition>)
-    else:
-        x = Float(<perform float addition>)
-    # x can be Float or Integer here
-    if isinstance(x, Integer):
-        if isinstance(c, Integer):
-            r = Integer(<perform int addition>)
-        else:
-            r = Float(<perform float addition>)
-    else:
-        r = Float(<perform float addition>)
-\end{verbatim}
+  \includegraphics[scale=0.5]{figures/add1.pdf}
+%\begin{verbatim}
+%    if (isinstance(a, Integer) and
+%            isinstance(b, Integer)):
+%        x = Integer(<perform int addition>)
+%    else:
+%        x = Float(<perform float addition>)
+%
+%    # x can be Float or Integer here
+%
+%    if (isinstance(x, Integer) and
+%            isinstance(c, Integer)):
+%        r = Integer(<perform int addition>)
+%    else:
+%        r = Float(<perform float addition>)
+%\end{verbatim}
+\end{frame}
+
+\begin{frame}[containsverbatim, plain]
+  \frametitle{Method-Based JITs and Dispatching Dependencies}
+  \includegraphics[scale=0.39]{figures/add2.pdf}
+\end{frame}
+
+\begin{frame}[containsverbatim, plain]
+  \frametitle{Method-Based JITs and Dispatching Dependencies}
+  \includegraphics[scale=0.41]{figures/add_fast.pdf}
 \end{frame}

### 2011-02-07 talk/ibm-feb-2011/talk.tex
add some more concrete examples for traces

@@
 \usepackage{alltt}
-\usepackage{tikz}
-\usetikzlibrary{arrows,positioning,shadows,shapes,calc,backgrounds,fit}
+
+\newcommand\redsout[1]{{\color{red}\sout{\hbox{\color{black}{#1}}}}}
 
@@
 
+\begin{frame}[containsverbatim]
+  \frametitle{Example: Dispatching in a Tracing JIT}
+\begin{verbatim}
+x = ADD(a : Integer, b : Integer)
+\end{verbatim}
+\end{frame}
+
+\begin{frame}[containsverbatim]
+  \frametitle{Example: Dispatching in a Tracing JIT}
+\begin{verbatim}
+guard_class(a, Integer)
+guard_class(b, Integer)
+u_a = unbox(a)
+u_b = unbox(b)
+u_x = int_add(a, b)
+x = new(Integer, u_x)
+\end{verbatim}
+\end{frame}
+
 \begin{frame}
@@
 
+\begin{frame}[containsverbatim]
+  \frametitle{Example: Dependencies in a Tracing JIT}
+\begin{alltt}
+guard_class(a, Integer)
+guard_class(b, Integer)
+u_a = unbox(a)
+u_b = unbox(b)
+u_x = int_add(u_a, u_b)
+x = new(Integer, u_x)
+
+guard_class(x, Integer)
+guard_class(c, Integer)
+u_x2 = unbox(x)
+u_c = unbox(c)
+u_r = int_add(u_x2, u_c)
+r = new(Integer, u_r)
+\end{alltt}
+\end{frame}
+
+\begin{frame}[containsverbatim]
+  \frametitle{Example: Dependencies in a Tracing JIT}
+\begin{alltt}
+guard_class(a, Integer)
+guard_class(b, Integer)
+u_a = unbox(a)
+u_b = unbox(b)
+u_x = int_add(u_a, u_b)
+x = new(Integer, u_x)
+
+\redsout{guard_class(x, Integer)}
+guard_class(c, Integer)
+u_x2 = unbox(x)
+u_c = unbox(c)
+u_r = int_add(u_x2, u_c)
+r = new(Integer, u_r)
+\end{alltt}
+\end{frame}
+
 \begin{frame}
@@
   \end{itemize}
+\end{frame}
+
+\begin{frame}[containsverbatim]
+  \frametitle{Example: Boxing in a Tracing JIT}
+\begin{alltt}
+guard_class(a, Integer)
+guard_class(b, Integer)
+u_a = unbox(a)
+u_b = unbox(b)
+u_x = int_add(u_a, u_b)
+x = new(Integer, u_x)
+
+guard_class(c, Integer)
+u_x2 = unbox(x)
+u_c = unbox(c)
+u_r = int_add(u_x2, u_c)
+r = new(Integer, u_r)
+\end{alltt}
+\end{frame}
+
+\begin{frame}[containsverbatim]
+  \frametitle{Example: Boxing in a Tracing JIT}
+\begin{alltt}
+guard_class(a, Integer)
+guard_class(b, Integer)
+u_a = unbox(a)
+u_b = unbox(b)
+u_x = int_add(u_a, u_b)
+\redsout{x = new(Integer, u_x)}
+
+guard_class(c, Integer)
+\redsout{u_x2 = unbox(x)}
+u_c = unbox(c)
+u_r = int_add(u_x, u_c)
+r = new(Integer, u_r)
+\end{alltt}
 \end{frame}

### 2011-03-03 talk/dyla2007/talk.tex
arigo? This must be one of Armin's because I did not edit this file. But if I have instead gone and clobbered his changes with an earlier version a) I apologise and b) I have no clue how this could happen given that he checked in things later than I.

@@
 
-\date{3rd Workshop on Dynamic Languages and Applications, July 31 2007}
+\date{3rd Workshop on Dynamic Languages and Applications, July 31st 2007}
 
-\pgfdeclareimage[height=0.5cm]{pypy-logo}{image/py-web.png}
-\logo{\pgfuseimage{pypy-logo}}
+%\pgfdeclareimage[height=0.5cm]{pypy-logo}{image/py-web.png}
+%\logo{\pgfuseimage{pypy-logo}}
 
@@
   \begin{block}{
-    Our point}
+    Our points}
     \begin{itemize}
@@
 \begin{frame}
-  \frametitle{Common Approaches to VM construction}
+  \frametitle{Common Approaches to Language Implementation}
   \begin{block}{
-    Using C directly (or C disguised as another language)}
+    Using C/C++ (potentially disguised as another language)}
     \begin{itemize}
@@
     \item
-      but also: Squeak, Scheme48
+      but also: Scheme48, Squeak
     \end{itemize}
   \end{block}
+  \pause
   \begin{block}{
@@
       JRuby, IronRuby
+    \item
+      various Prolog, Lisp, even Smalltalk implementations
     \end{itemize}
@@
   \item
-    simplicity of the VM
+    simplicity
   \item
-    performance (needs dynamic compilation techniques)
+    performance
   \end{itemize}
@@
 \begin{frame}
+  \frametitle{Fixing of Early Design Decisions}
+  \begin{itemize}
+  \item when starting a VM in C, many design decisions need to be made upfront
+  \item examples: memory management technique, threading model
+  \item such decisions are manifested throughout the VM source
+  \item very hard to change later
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+      CPython uses reference counting, increfs and decrefs everywhere
+    \item
+      CPython uses OS threads with one global lock, hard to change to
+      lightweight threads or finer locking
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}
   \frametitle{Compilers are a bad encoding of Semantics}
@@
   \item to reach good performance levels, dynamic compilation is often needed
-  \item a dynamic compiler needs to encode language semantics
+  \item a compiler (obviously) needs to encode language semantics
   \item this encoding is often obscure and hard to change
@@
 \begin{frame}
-  \frametitle{Fixing of Early Design Decisions}
-  \begin{itemize}
-  \item when starting a VM in C, many design decisions need to be made upfront
-  \item examples: memory management technique, threading model
-  \item the decision is manifested throughout the VM source
-  \item very hard to change later
-  \end{itemize}
-  \pause
-  \begin{block}{
-    Python Case}
-    \begin{itemize}
-    \item
-      CPython uses reference counting, increfs and decrefs everywhere
-    \item
-      CPython uses OS threads with one global lock, hard to change to
-      lightweight threads or finer locking
-    \end{itemize}
-  \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Implementation Proliferation}
-  \begin{itemize}
-  \item
-    restrictions of the original implementation lead to re-implementations,
-    forks
-  \item
-    all implementations need to be synchronized with language evolution
-  \item
-    lots of duplicate effort
-  \end{itemize}
-  \pause
-  \begin{block}{
-    Python Case}
-    \begin{itemize}
-    \item
-      several serious implementations: CPython, Stackless, Psyco, Jython,
-      IronPython, PyPy
-    \item
-      the implementations have various grades of compliance
-    \end{itemize}
-  \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Implementing Languages on Top of General-Purpose OO VMs}
+  \frametitle{Implementing Languages on Top of OO VMs}
   \begin{itemize}
@@
   \item
-    even more implementation proliferation
+    more implementations!
   \item
-    implementing on top of an OO VM has its own set of problems
+    implementing on top of an OO VM has its own set of benefits of problems
   \end{itemize}
@@
     better interoperability than what the C level provides
-  \item
-    some proponents believe that eventually one single VM should be enough
   \end{itemize}
@@
     \item
-      XXX what else?
+      Jython has free threading
     \end{itemize}
@@
   \item
-    some of the benefits of OO VMs don't work out in practice
-  \item
     most immediate problem: it can be hard to map
-    concepts of the dynamic lang to
+    concepts of the dynamic language to
     the host OO VM
@@
     semantic mismatch between the dynamic language and the host VM
-  \item
-    poor interoperability with everything outside the OO VM
-  \item
-    in practice, one OO VM is not enough
   \end{itemize}
@@
     \item
-      IronPython is about as fast as CPython (but some introspection features
-      missing)
+      IronPython is about as fast as CPython (but some stack introspection
+      features missing)
     \item
-      XXX something about semantic mismatch?
+      Python has very different semantics for method calls than Java
     \end{itemize}
@@
 
+\end{frame}
+
+\begin{frame}
+  \frametitle{Implementation Proliferation}
+  \begin{itemize}
+  \item
+    restrictions of the original implementation lead to re-implementations,
+    forks
+  \item
+    all implementations need to be synchronized with language evolution
+  \item
+    lots of duplicate effort, compatibility problems
+  \end{itemize}
+  \pause
+  \begin{block}{
+    Python Case}
+    \begin{itemize}
+    \item
+      several serious implementations: CPython, Stackless, Psyco, Jython,
+      IronPython, PyPy
+    \item
+      various incompatibilities
+    \end{itemize}
+  \end{block}
 \end{frame}
@@
   \item
-    ... which we translate (i.e.\ compile) to VMs running on top of various
-    targets, like C/Posix, CLR, JVM
+    ... which we translate (i.e.\ compile) to a VM running in various target
+    environments, like C/Posix, CLR, JVM
   \end{itemize}
@@
     \item
-      still a high-level language (unlike SLang or Prescheme)
+      still a high-level language (unlike SLang or PreScheme)
     \item
@@
   \begin{itemize}
-  \item
-    high-level source: early design decisions not necessary
   \item
@@
   \item
-    separation of concerns from low-level details
+    separation of language semantics from low-level details
   \item
-    a potential single-source-fits-all interpreter --
-    less duplication of efforts
-  \item
-    runs everywhere with the same semantics --
-    no outdated implementations, no ties to any standard platform
+    a single-source-fits-all interpreter
+    \begin{itemize}
+      \item runs everywhere with the same semantics
+      \item no outdated implementations, no ties to any standard platform
+      \item less duplication of efforts
+    \end{itemize}
   \end{itemize}
@@
     \item
-      almost orthogonal from the interpreter source - applicable to many
+      almost orthogonal from the interpreter source – applicable to many
       languages, follows language evolution ``for free''
     \item
-      based on Partial Evaluation
+      based on Partial Evaluation techniques
     \item
-      benefits from a high-level interpreter and a tweakable translation
-      toolchain
+      benefits from a high-level interpreter
     \item
@@
 \begin{frame}
-  \frametitle{Open Issues / Drawbacks / Further Work}
+  \frametitle{Drawbacks / Open Issues / Further Work}
   \begin{itemize}
@@
   \item
-    conceptually simple approach but many abstraction layers
-  \item
     dynamic compiler generation seems to work, but needs more efforts.
-    Also: can we layer it on top of the JIT of a general purpose OO VM?
   \end{itemize}
@@
   \item
+    VMs shouldn't be written by hand
+  \item
     high-level languages are suitable to implement dynamic languages
@@
   \item
-    VMs shouldn't be written by hand
-  \item
     PyPy's concrete approach is not so important
-  \item
-    diversity is good
   \item
@@
 \begin{frame}
-  \frametitle{Links}
+  \frametitle{Questions?}
   \begin{block}{

### 2011-03-23 talk/icooolps2011/paper.tex

%\documentclass{acm_proc_article-sp}
\documentclass{sig-alternate}

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage[utf8]{inputenc}

\makeatletter
\def\PY@reset{\let\PY@it=\relax \let\PY@bf=\relax%
    \let\PY@ul=\relax \let\PY@tc=\relax%
    \let\PY@bc=\relax \let\PY@ff=\relax}
\def\PY@tok#1{\csname PY@tok@#1\endcsname}
\def\PY@toks#1+{\ifx\relax#1\empty\else%
    \PY@tok{#1}\expandafter\PY@toks\fi}
\def\PY@do#1{\PY@bc{\PY@tc{\PY@ul{%
    \PY@it{\PY@bf{\PY@ff{#1}}}}}}}
\def\PY#1#2{\PY@reset\PY@toks#1+\relax+\PY@do{#2}}

\def\PY@tok@gd{\def\PY@bc##1{\fcolorbox[rgb]{0.80,0.00,0.00}{1.00,0.80,0.80}{##1}}}
\def\PY@tok@gu{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.00}{##1}}}
\def\PY@tok@gt{\def\PY@tc##1{\textcolor[rgb]{0.60,0.80,0.40}{##1}}}
\def\PY@tok@gs{\let\PY@bf=\textbf}
\def\PY@tok@gr{\def\PY@tc##1{\textcolor[rgb]{1.00,0.00,0.00}{##1}}}
\def\PY@tok@cm{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
\def\PY@tok@vg{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
\def\PY@tok@m{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@mh{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@cs{\let\PY@bf=\textbf\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
\def\PY@tok@ge{\let\PY@it=\textit}
\def\PY@tok@vc{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
\def\PY@tok@il{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@go{\def\PY@tc##1{\textcolor[rgb]{0.67,0.67,0.67}{##1}}}
\def\PY@tok@cp{\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,0.60}{##1}}}
\def\PY@tok@gi{\def\PY@bc##1{\fcolorbox[rgb]{0.00,0.80,0.00}{0.80,1.00,0.80}{##1}}}
\def\PY@tok@gh{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.00}{##1}}}
\def\PY@tok@ni{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,0.60}{##1}}}
\def\PY@tok@nl{\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,1.00}{##1}}}
\def\PY@tok@nn{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.80,1.00}{##1}}}
\def\PY@tok@no{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.00}{##1}}}
\def\PY@tok@na{\def\PY@tc##1{\textcolor[rgb]{0.20,0.00,0.60}{##1}}}
\def\PY@tok@nb{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.40}{##1}}}
\def\PY@tok@nc{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.67,0.53}{##1}}}
\def\PY@tok@nd{\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,1.00}{##1}}}
\def\PY@tok@ne{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.80,0.00,0.00}{##1}}}
\def\PY@tok@nf{\def\PY@tc##1{\textcolor[rgb]{0.80,0.00,1.00}{##1}}}
\def\PY@tok@si{\def\PY@tc##1{\textcolor[rgb]{0.67,0.00,0.00}{##1}}}
\def\PY@tok@s2{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@vi{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
\def\PY@tok@nt{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.20,0.00,0.60}{##1}}}
\def\PY@tok@nv{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
\def\PY@tok@s1{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@gp{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,0.60}{##1}}}
\def\PY@tok@sh{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@ow{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,0.00}{##1}}}
\def\PY@tok@sx{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@bp{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.40}{##1}}}
\def\PY@tok@c1{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
\def\PY@tok@kc{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@c{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
\def\PY@tok@mf{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@err{\def\PY@tc##1{\textcolor[rgb]{0.67,0.00,0.00}{##1}}\def\PY@bc##1{\colorbox[rgb]{1.00,0.67,0.67}{##1}}}
\def\PY@tok@kd{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@ss{\def\PY@tc##1{\textcolor[rgb]{1.00,0.80,0.20}{##1}}}
\def\PY@tok@sr{\def\PY@tc##1{\textcolor[rgb]{0.20,0.67,0.67}{##1}}}
\def\PY@tok@mo{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@mi{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
\def\PY@tok@kn{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@o{\def\PY@tc##1{\textcolor[rgb]{0.33,0.33,0.33}{##1}}}
\def\PY@tok@kr{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@s{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@kp{\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@w{\def\PY@tc##1{\textcolor[rgb]{0.73,0.73,0.73}{##1}}}
\def\PY@tok@kt{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.47,0.53}{##1}}}
\def\PY@tok@sc{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@sb{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@k{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
\def\PY@tok@se{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
\def\PY@tok@sd{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}

\def\PYZbs{\char`\\}
\def\PYZus{\char`\_}
\def\PYZob{\char`\{}
\def\PYZcb{\char`\}}
\def\PYZca{\char`\^}
% for compatibility with earlier versions
\def\PYZat{@}
\def\PYZlb{[}
\def\PYZrb{]}
\makeatother

\ifthenelse{\isundefined{\hypersetup}}{
  \usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
}{}
\hypersetup{
  pdftitle={Controlling the Tracing of an Interpreter With Hints, Part 1: Controlling the Extent of Tracing},
}

\newboolean{showcomments}
\setboolean{showcomments}{false}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\toon[1]{\nb{TOON}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand{\commentout}[1]{}

\newcommand\ie{i.e.,\xspace}
\newcommand\eg{e.g.,\xspace}

\normalem

\let\oldcite=\cite

\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}

% compressing itemize env, in case we need it
\newenvironment{zitemize}% zero - line spacing itemize environment
   {\begin{list}{--}{
   \setlength{\itemsep}{0 pt}
   \setlength{\parsep}{0 pt}
   \setlength{\topsep} {0 pt} }}% the end stuff
   {\end{list}}

\begin{document}

\title{XXX in a Tracing JIT Compiler for Efficient Dynamic Languages}

\numberofauthors{4}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{University of Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Germany}\\
       \email{cfbolz@gmx.de}
\alignauthor XXX
       \affaddr{XXX}\\
       \email{XXX}
}
\conferenceinfo{ICOOOLPS}{'09 Genova, Italy}
\CopyrightYear{2009}
\crdata{978-1-60558-541-3/09/07}

\maketitle

\category{D.3.4}{Programming Languages}{Processors}[code generation,
incremental compilers, interpreters, run-time environments]

\begin{abstract}

\end{abstract}

\section{Introduction}

\section{The PyPy Project}
\label{sect:pypy}

XXX

\section{Tracing JIT Compilers}
\label{sect:tracing}

XXX

\section{Controlling The Extent of Tracing}

The question I was asked most often during my recent \href{http://morepypy.blogspot.com/2011/03/us-trip-report-popl-microsoft-ibm.html}{US trip} was how exactly
the hints work that interpreter authors can use to improve the execution speed
of the programs running on their interpreters. Since those hints are not really
documented all that well, I decided to write blog posts about them. This is the
first one.

%___________________________________________________________________________

\subsection{Background}

First, let's recap some basics: PyPy's approach to implementing dynamic
languages is to write an interpreter for
the language in RPython. This interpreter can be translated to C and then
further to machine code. The interpreter consists of code in the form of a
large number of generated C functions and some data. Similarly, the user
program consists of functions in the language the interpreter executes.

As was explained in a \href{http://morepypy.blogspot.com/2009/03/applying-tracing-jit-to-interpreter.html}{blog post} and a \href{http://codespeak.net/svn/pypy/extradoc/talk/icooolps2009/bolz-tracing-jit.pdf}{paper} two years ago, PyPy's JIT is a
meta-tracer. Since we want to re-use our tracer for a variety of languages, we
don't trace the execution of the user program, but instead trace the execution
of the \emph{interpreter} that is running the program. This means that the traces
don't contain the bytecodes of the language in question, but RPython-level
operations that the interpreter did to execute the program.

On the other hand, the loops that are traced by the tracer are the loops in the
user program. This means that the tracer stops tracing after one iteration of
the loop in the user function that is being considered. At this point, it can
have traced many iterations of the interpreter main loop.

Here's a diagram of this process:

\begin{figure*}
\includegraphics[scale=0.5]{figures/trace-levels}
\caption{The levels involved in tracing}
\label{fig:trace-levels}
\end{figure*}

On the left you see the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. Some of these functions contain
loops, others don't. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

%___________________________________________________________________________

\subsection{How Far Should Tracing Go}

When the tracer encounters a function call at the interpreter level, e.g. the
interpreter main loop calling a helper function, it can do one of two things:

\begin{enumerate}
\item it can trace into the helper function, effectively inlining it into the trace.

\item it can not trace into the function and instead record a call to that function
as an operation in the trace. Such a call operation in the trace is sometimes
called \emph{residual call}.
\end{enumerate}

As a default, the tracer will try to trace into the helper because that will
give more information to the optimizer, allowing it to do a better job. This is
particularly important for the allocation removal optimization, because if a
freshly allocated object is passed as an argument to a residual call, its
allocation cannot be optimized away.

There is a problem however if the helper function itself contains a loop. The
tracer records the linear sequence of operations that are being executed. Thus
when it encounters a loop on the interpreter level it records all the
operations of every iteration of the loop itself, with the net effect of
unrolling it. The only places where the tracer stops and tries to close the
trace is in the main loop of the interpreter. When the tracer encounters the
main loop, it also checks whether the original user loop has been closed, and
thus whether it can stop tracing.

For most helper functions in the interpreter that contain loops, fully
unrolling does not make sense. If a loop is unrolled, the trace is specific to
the number of iteration that was seen during tracing. If the trace is later
executed with a different number of iterations, the trace will be left via a
guard failure, which is inefficient. Therefore the default behaviour of the
tracer is to never trace into a function on the interpreter level that contains
a loop, but to trace into all non-looping helper functions.

This default behaviour is essentially a heuristic, but one that usually makes
sense. We want to produce just enough traces to make the resulting code
efficient, but not more. Therefore we trace as much as possible (everything by
default) except the functions which loops where tracing would produce code that
is less general than it could be.

As an example for a helper with a loop, take string concatenation. It loops over
the characters of both arguments and copies them over into the result string. It
does not make sense to unroll the loops in this function. If we do that,
the resulting trace can only be used for strings of the length that was seen
during tracing. In practise, the string lengths are usually different each run,
meaning that the trace with unrolling is not run to completion in most cases.

%___________________________________________________________________________

\subsection{Influencing the Default Behaviour}

Sometimes the default behaviour is not actually what is wanted. This is
something the interpreter author has to decide, usually by looking at the traces
that are produced and deciding that they should be improved. There are two ways
in which the default is wrong:
%
\begin{itemize}

\item \textbf{false negatives:} if a helper function that \textbf{does} contain a loop should
be traced into, unrolling the loop.

\item \textbf{false positives:} if a helper function that \textbf{does not} contain a loop is
inlined into the trace, but the interpreter author decides that this is not
helpful.

\end{itemize}

If the interpreter author finds false negatives or false positives, she can fix
that by applying a hint to the tracer. These hints take the form of function
decorators (which both live in the \texttt{pypy.rlib.jit} module). In the next two
subsections I will describe these two function decorators and their use.

%___________________________________________________________________________

\subsubsection{Unrolling Functions With Loops}

The first decorator, used to fix false negatives, is the \texttt{unroll\_safe}
decorator. It is used to tell the tracer to always trace into a function that
has a loop, effectively unrolling the loop. This decorator should be used only
if the loop in the helper function is expected to always run for the same number
of iterations. This sounds like a strong restriction, in practise this is less
severe: The number of iterations needs to only be the same \emph{in the context where
the helper functions is traced from}.

It is easiest to understand this condition via an example. Let's look at the
\texttt{BUILD\_TUPLE} bytecode in Python. It takes one argument, the length \texttt{n} of
the tuple being built. The bytecode pops \texttt{n} arguments from the stack, turns
them into a tuple and pushes that tuple on the stack. Thus the function that
implements \texttt{BUILD\_TUPLE} in PyPy's Python interpreter calls a helper
\texttt{popvalues} which pops \texttt{n} values from the stack and returns them in a list.
This helper is implemented with a loop and would thus not be traced into by
default.  The loop in the helper can run for very different numbers of
iterations, because it is used in a variety of places. However, for every
concrete \texttt{BUILD\_TUPLE} bytecode, the argument will be constant. Therefore it
is safe (and even necessary) to annotate \texttt{popvalues} with the \texttt{unroll\_safe}
decorator.

A different example is the implementation of the \texttt{isinstance} builtin. It is
used to check whether an object \texttt{a} is an instance of a class \texttt{B} like
this: \texttt{isinstance(a, B)}. The second argument of the function can also be a
tuple of classes to check whether an object is an instance of one of a number of
classes: \texttt{isinstance(a, (A, B, C, D))}. To implement this second case, the
implementation of \texttt{isinstance} contains a loop iterating over the elements of
the tuple. The number of loop iterations can vary, but is usually fixed for each
individual call site which typically just lists a few classes in the source
code. Therefore it is also safe to annotate the implementation of \texttt{isinstance}
with the \texttt{unroll\_safe} decorator.

%___________________________________________________________________________

\subsubsection{Preventing the Tracing of Functions}

The second decorator \texttt{dont\_look\_inside} is used to fix false positives. It
tells the JIT to never trace into the decorated function and just always produce
a residual call instead. This decorator is in many ways less important than the
unrolling one (except for a special situation that I will describe in a
follow-up post). It is used if tracing into a function is not expected to yield
any speed benefits, because the optimizer will not be able to improve it much.
This is often the case if the called helper function does not contain any
``dynamic'' behaviour. In such a situation it is better to just leave the function
call in the trace, because that produces less code.

An example would be the import mechanism in Python. It's very unlikely that any
performance improvement can be had by turning part of it into assembler.
Therefore we hide it from the tracer by annotating them with
\texttt{dont\_look\_inside}.

%___________________________________________________________________________

\subsection{Conclusion}

In this post we discussed two hints that can be used to control precisely which
parts of the interpreter should be meta-traced. If these hints are used
carefully, this can go a long way to making the interpreter produce traces that
contain exactly the interesting part of the execution, and will contain calls to
the functions that can not be optimized by tracing techniques.

In the next part of this series I will discuss a different set of hints that can
be used to strongly optimize traces.

% Document title
\section{Controlling the Tracing of an Interpreter With Hints, Part 2: Controlling Optimization}

This is part 2 of a series on how to speed up an interpreter written with PyPy
by adding JIT hints to the interpreter. Part 1 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with.html}{control the
extent of tracing}. In this post I will describe how to add hints that
influence the optimizer.  If applied correctly these techniques can give
really big speedups by pre-computing parts of what happens at runtime. On the other
hand, if applied incorrectly they might lead to code bloat, thus making the
resulting program actually slower.

%___________________________________________________________________________

\subsection{Background}

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or reduce
the number of operations: most of these are well known \href{http://en.wikipedia.org/wiki/Compiler_optimization\#Optimization_techniques}{compiler optimization
techniques}, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques:
%
\begin{itemize}

\item \href{http://en.wikipedia.org/wiki/Constant_folding}{constant folding}

\item \href{http://en.wikipedia.org/wiki/Common_subexpression_elimination}{common subexpression elimination}

\item allocation removal, as described in the paper that I recently \href{http://morepypy.blogspot.com/2011/03/us-trip-report-popl-microsoft-ibm.html}{presented at
PEPM}

\item store/load propagation

\item \href{http://morepypy.blogspot.com/2011/01/loop-invariant-code-motion.html}{loop invariant code motion}

\end{itemize}

In some places it turns out that if the interpreter author rewrites some parts
of the interpreter with these optimizations in mind the traces that are produced
by the optimizer can be vastly improved.

In this post I will describe two hints that allow the interpreter author to
increase the optimization opportunities for constant folding. For constant
folding to work, two conditions need
to be met:
%
\begin{itemize}

\item the arguments of an operation actually need to all be constant,
i.e. statically known by the optimizer

\item the operation needs to be \emph{pure}, i.e. always yield the same result given
the same arguments.

\end{itemize}

The PyPy JIT generator automatically detects the majority of these conditions.
However, for the cases in which the automatic detection does not work, the
interpreter author can apply \textbf{hints} to improve the optimization
opportunities. There is one kind of hint for both of the conditions above.

\textbf{Note}: These hints are written by an interpreter developer and applied to the
RPython source of the interpreter. Normal Python users will never see them.

%___________________________________________________________________________

\subsection{Where Do All the Constants Come From}

It is worth clarifying what is a ``constant'' in this context.  A variable of
the trace is said to be constant if its value is statically known by the
optimizer.

The simplest example of constants are literal values.  For example, if in the
RPython source code we have a line like \texttt{y = x + 1}, the second operand will
be a constant in the trace.

However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{if} \PY{n}{x} \PY{o}{==} \PY{l+m+mi}{4}\PY{p}{:}
    \PY{n}{y} \PY{o}{=} \PY{n}{y} \PY{o}{+} \PY{n}{x}
\end{Verbatim}

If the fragment is traced with \texttt{x} being \texttt{4}, the following trace is
produced:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(x~==~4)\\
y~=~y~+~x
}
\end{quote}

In the trace above, the value of \texttt{x} is statically known thanks to the
guard. Remember that a guard is a runtime check. The above trace will run to
completion when \texttt{x == 4}. If the check fails, execution of the trace is
stopped and the interpreter continues to run.

There are cases in which it is useful to turn an arbitrary variable
into a constant value. This process is called \emph{promotion} and it is an old idea
in partial evaluation (it's called ``the trick'' there). Promotion is also heavily
used by \href{http://psyco.sourceforge.net/}{Psyco} and by all older versions of PyPy's JIT. Promotion is a technique
that only works well in JIT compilers, in
static compilers it is significantly less applicable.

Promotion is essentially a tool for trace specialization. In some places in the
interpreter it would be very useful if a variable were constant, even though it
could have different values in practice. In such a place, promotion is used. The
typical reason to do that is if there is
a lot of computation depending on the value of that variable.

Let's make this more concrete. If we trace a call to the following function:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{def} \PY{n+nf}{f1}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
    \PY{n}{z} \PY{o}{=} \PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
    \PY{k}{return} \PY{n}{z} \PY{o}{+} \PY{n}{y}
\end{Verbatim}

We get a trace that looks like this:
%
\begin{quote}{\ttfamily \raggedright \noindent
v1~=~x~*~2\\
z~=~v1~+~1\\
v2~=~z~+~y\\
return(v2)
}
\end{quote}

Observe how the first two operations could be constant-folded if the value of
\texttt{x} were known. Let's assume that the value of \texttt{x} can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{def} \PY{n+nf}{f2}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
    \PY{n}{x} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
    \PY{n}{z} \PY{o}{=} \PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
    \PY{k}{return} \PY{n}{z} \PY{o}{+} \PY{n}{y}
\end{Verbatim}

The meaning of this hint is that the tracer should pretend that \texttt{x} is a
constant
in the code that follows. When just running the code, the function has no
effect, as it simply returns its first argument. When tracing, some extra work
is done. Let's assume that this changed function is traced with
the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
operation at the beginning:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(x~==~4)\\
v1~=~x~*~2\\
z~=~v1~+~1\\
v2~=~z~+~y\\
return(v2)
}
\end{quote}

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the value of \texttt{x} as it was at runtime. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the example above. After the guard, the rest of the trace can
assume that \texttt{x} is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(x~==~4)\\
v2~=~9~+~y\\
return(v2)
}
\end{quote}

Notice how the first two arithmetic operations were constant folded. The hope is
that the guard is executed quicker than the multiplication and the addition that
was now optimized away.

If this trace is executed with values of \texttt{x} other than \texttt{4}, the guard will
fail, and execution will continue in the interpreter. If the guard fails often
enough, a new trace will be started from the guard. This other trace will
capture a different value of \texttt{x}. If it is e.g. \texttt{2}, then the optimized
trace looks like this:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(x~==~2)\\
v2~=~5~+~y\\
return(v2)
}
\end{quote}

This new trace will be attached to the guard instruction of the first trace. If
\texttt{x} takes on even more values, a new trace will eventually be made for all of them,
linking them into a chain. This is clearly not desirable, so we should promote
only variables that don't vary much. However, adding a promotion hint will never produce wrong
results. It might just lead to too much assembler code.

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend to not vary often. Therefore it is possible to
promote the types. In the next blog post I will give a complete example of how
this works.

%___________________________________________________________________________

\subsection{Declaring New Pure Operations}

In the last section we saw a way to turn arbitrary variables into constants. All
pure operations on these constants can be constant-folded. This works great for
constant folding of simple types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not simple types. The
operations on objects are often not pure and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

As an example, take the following class:

\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{A}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{=} \PY{n}{x}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n}{y}

    \PY{k}{def} \PY{n+nf}{f}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{val}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{compute}\PY{p}{(}\PY{p}{)} \PY{o}{+} \PY{n}{val}

    \PY{k}{def} \PY{n+nf}{compute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
\end{Verbatim}

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
\begin{quote}{\ttfamily \raggedright \noindent
x~=~a.x\\
v1~=~x~*~2\\
v2~=~v1~+~1\\
v3~=~v2~+~val\\
a.y~=~v3
}
\end{quote}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if \texttt{a} is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is a pure function. To communicate this, there is a \texttt{purefunction} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{A}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{=} \PY{n}{x}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n}{y}

    \PY{k}{def} \PY{n+nf}{f}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{val}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{compute}\PY{p}{(}\PY{p}{)} \PY{o}{+} \PY{n}{val}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{compute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
\end{Verbatim}

Now the trace will look like this:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(a~==~0xb73984a8)\\
v1~=~compute(a)\\
v2~=~v1~+~val\\
a.y~=~v2
}
\end{quote}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{quote}{\ttfamily \raggedright \noindent
guard(a~==~0xb73984a8)\\
v2~=~9~+~val\\
a.y~=~v2
}
\end{quote}

(assuming that the \texttt{x} field's value is \texttt{4}).

On the one hand, the \texttt{purefunction} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives you ample opportunity to mess things up. If a
function is annotated to be pure, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation.

%___________________________________________________________________________

\subsubsection{Observably Pure Functions}

Why can't we simply write an analysis to find out that the \texttt{x} fields of the
\texttt{A} instances is immutable and deduce that \texttt{compute} is a pure function,
since it only reads the \texttt{x} field and does not have side effects? This might
be possible in this particular case, but in practice the functions that are
annotated with the \texttt{purefunction} decorator are usually more complex.
The easiest example for this is that of a function that uses memoization to
cache its results. If you analyze this function, it looks like the function has
side effects, because it changes the memoizing dictionary. However, because this side
effect is not externally visible, the function from the outside is pure. This is
a property that is not easily detectable by analysis. Therefore, the purity
of this function needs to be annotated.

%___________________________________________________________________________

\subsubsection{Immutable Fields}

One of the most common cases of pure functions is reading immutable
values out of objects. Since this is so common, we have special syntactic sugar
for it. A RPython class can have a class attribute \texttt{\_immutable\_fields\_} set to
a list of strings, listing the fields that cannot be changed. This is equivalent
to using getters and annotating them with \texttt{purefunction}.

%___________________________________________________________________________

\subsection{Conclusion}

In this blog post I explained two more hints that can be used in the source code
of the interpreter. They are used to influence what the optimizer does with the
trace. I realize the examples given here are a bit too small, in the next
installment I will give a worked-out example that puts all the pieces together.

\section{Controlling the Tracing of an Interpreter With Hints, Part 3: Putting Things Together}

This is part 3 of the series on how to speed up an interpreter written with
PyPy by adding JIT hints to the interpreter. Part 1 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with.html}{control
the extent of tracing}. Part 2 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with_15.html}{influence the optimizer with
promotion and pure functions}. In this post I describe a worked-out example of
a small object model for a dynamic language and how to make it efficient using
the hints described in the previous posts.

%___________________________________________________________________________

\subsection{A Simple Object Model}

To implement a dynamic language efficiently, the operations on its objects need
to be fast. Most dynamic languages have object models that are made by using
dictionaries everywhere. Let's look at an example of how the JIT can be made to
optimize such operations.

For the purpose of this blog post we will use a very simple and bare-bones
object model that just supports very simple classes and instances, without any
inheritance or any fancy features. The model has classes, which contain methods.
Instances have a class. Instances have their own attributes. When looking up an
attribute on an instance, the instances attributes are searched. If the
attribute is not found there, the class' attributes are searched.

To implement this object model, we could use the following RPython code as part
of the interpreter source code:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{k}{def} \PY{n+nf}{instantiate}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n}{Instance}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
            \PY{k}{return} \PY{n}{result}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}

\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
            \PY{k}{return} \PY{n}{result}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}

    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{try}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
\end{Verbatim}

In this straightforward implementation the methods and attributes are just
stored in dictionaries on the classes/instances. While this object model is very
simple it already contains all the hard parts of Python's object model. Both
instances and classes can have arbitrary fields, and they are changeable at
any time.  Moreover, instances can change their class after they have been
created.

When using this object model in
an interpreter, a huge amount of time will be spent doing lookups in these
dictionaries. To make the language efficient using a tracing JIT, we need to
find a way to get rid of these dictionary lookups somehow.

Let's assume we trace through code that sums three attributes, such as:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)} \PY{o}{+} \PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)} \PY{o}{+} \PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\end{Verbatim}

The trace could look like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{attributes1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
\PY{n}{result1} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result1} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{c}{# inst.getattr("b")}
\PY{n}{attributes2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
\PY{n}{v1} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{v1} \PY{o+ow}{is} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods1} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}

\PY{c}{# inst.getattr("c")}
\PY{n}{attributes3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
\PY{n}{v3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{v3} \PY{o+ow}{is} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods2} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

In this example, the attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
five calls to \texttt{dict.get}, which is slow.

%___________________________________________________________________________

\subsection{Making Instance Attributes Faster Using Maps}

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints we have looked at in the
two earlier blog posts don't seem to help with the current object model. There is
no pure function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

This is a common problem when trying to apply hints. Often, the interpreter
needs a small rewrite to expose the pure functions and nearly-constant objects
that are implicitly there. In the case of instance fields this rewrite is not
entirely obvious. The basic idea is as follows. In theory instances can have
arbitrary fields. In practice however many instances share their layout (i.e.
their set of keys) with many other instances.

Therefore it makes sense to factor the layout information out of the instance
implementation into a shared object. This shared layout object is called a
\emph{map}. Maps are an old idea that comes originally from the SELF language. They are
also used by many JavaScript implementations such as V8. I've \href{http://morepypy.blogspot.com/2010/11/efficiently-implementing-python-objects.html}{written about maps
before}, so I won't explain them fully again.

The rewritten \texttt{Instance} class using maps looks like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{new\PYZus{}map\PYZus{}with\PYZus{}additional\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{if} \PY{n}{name} \PY{o+ow}{not} \PY{o+ow}{in} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{:}
            \PY{n}{newmap} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
            \PY{n}{newmap}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{o}{.}\PY{n}{update}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{)}
            \PY{n}{newmap}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n+nb}{len}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{)}
            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{newmap}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]}

\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}

\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}

    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]} \PY{o}{=} \PY{n}{value}
            \PY{k}{return}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{new\PYZus{}map\PYZus{}with\PYZus{}additional\PYZus{}attribute}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{n}{value}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{try}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
\end{Verbatim}

Instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. The maps are shared between
objects with the same layout. Therefore they have to be immutable, which means
that their \texttt{getindex} method is a pure function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{new\_map\_with\_additional\_attribute} method on the previous map. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

With this changed instance implementation, the trace we had above changes to the
following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
has been promoted:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index1} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index1} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{n}{index1}\PY{p}{]}

\PY{c}{# inst.getattr("b")}
\PY{n}{map2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map2} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index2} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index2} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods1} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}

\PY{c}{# inst.getattr("c")}
\PY{n}{map3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map3} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index3} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index3} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods2} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a pure function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. The optimized trace looks like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}

\PY{c}{# inst.getattr("b")}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{methods}
\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}

\PY{c}{# inst.getattr("c")}
\PY{n}{cls2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{methods2} \PY{o}{=} \PY{n}{cls2}\PY{o}{.}\PY{n}{methods}
\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
of the constant-folded \texttt{getindex} call. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

%___________________________________________________________________________

\subsection{Versioning of Classes}

Instances were optimized making the assumption that the total number of
Instance layouts is small compared to the number of instances. For classes we
will make an even stronger assumption. We simply assume that it is rare for
classes to change at all. This is not totally reasonable (sometimes classes contain
counters or similar things) but for this simple example it is good enough.

What we would really like is if the \texttt{Class.find\_method} method were pure.
But it cannot be, because it is always possible to change the class itself.
Every time the class changes, \texttt{find\_method} can potentially return a
new value.

Therefore, we give every class a version number, which is increased every time a
class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} looks like
this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{VersionTag}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{pass}

\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{version} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}
        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
            \PY{k}{return} \PY{n}{result}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}
\end{Verbatim}

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
pure (because when the version number changes, the result of the call might be
different than the previous one).

The trace with this new class implementation looks like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index1} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index1} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{n}{index1}\PY{p}{]}

\PY{c}{# inst.getattr("b")}
\PY{n}{map2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map2} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index2} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index2} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{guard}\PY{p}{(}\PY{n}{cls1} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
\PY{n}{version1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{version}
\PY{n}{guard}\PY{p}{(}\PY{n}{version1} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
\PY{n}{result2} \PY{o}{=} \PY{n}{Class}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{cls}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{,} \PY{n}{version1}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}

\PY{c}{# inst.getattr("c")}
\PY{n}{map3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map3} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{index3} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{index3} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
\PY{n}{cls2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{guard}\PY{p}{(}\PY{n}{cls2} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
\PY{n}{version2} \PY{o}{=} \PY{n}{cls2}\PY{o}{.}\PY{n}{version}
\PY{n}{guard}\PY{p}{(}\PY{n}{version2} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
\PY{n}{result3} \PY{o}{=} \PY{n}{Class}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{cls}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{,} \PY{n}{version2}\PY{p}{)}
\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}

\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

The calls to \texttt{Class.\_find\_method} can now be optimized away, also the
promotion of the class and the version, except for the first one. The final
optimized trace looks like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{c}{# inst.getattr("a")}
\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}

\PY{c}{# inst.getattr("b")}
\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
\PY{n}{guard}\PY{p}{(}\PY{n}{cls1} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
\PY{n}{version1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{version}
\PY{n}{guard}\PY{p}{(}\PY{n}{version1} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{l+m+mi}{41}

\PY{c}{# inst.getattr("c")}
\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{l+m+mi}{17}
\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
\end{Verbatim}

The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method`} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{change\_method} method on the class.

%___________________________________________________________________________

\subsection{Real-World Considerations}

The techniques used above for the simple object model are used for the object
model of PyPy's Python interpreter too. Since Python's object model is
considerably more complex, some additional work needs to be done.

The first problem that needs to be solved is that Python supports (multiple)
inheritance. Therefore looking up a method in a class needs to consider the
whole method resolution order. This makes the versioning of classes more
complex. If a class is changed its version changes. At the same time, the
versions of all the classes inheriting from it need to be changed as well,
recursively. This makes class changes expensive, but they should be rare.  On the
other hand, a method lookup in a complex class hierarchy is as optimized in the
trace as in our object model here.

A downside of the versioning of classes that we haven't yet fixed in PyPy, is
that some classes \emph{do} change a lot. An example would be a class that keeps a
counter of how many instances have been created so far. This is very slow right
now, but we have ideas about how to fix it in the future.

Another optimization is that in practice the shape of an instance is correlated
with its class. In our code above, we allow both to vary independently.
In PyPy's Python interpreter we act somewhat more cleverly. The class of
an instance is not stored on the instance itself, but on the map. This means
that we get one fewer promotion (and thus one fewer guard) in the trace, because the class doesn't need to
be promoted after the map has been.

%___________________________________________________________________________

\subsection{More General Patterns}

The techniques we used above to make instance and class lookups faster are
applicable in more general cases than the one we developed them for. A more
abstract view of maps is that of splitting a data-structure into a part that
changes slowly, and a part that changes quickly. In the concrete example of maps
we split the original dictionary into the map (the slow-changing part) and the
storage array (the quick-changing part). All the computation on the
slow-changing part can be constant-folded during tracing so that only the
manipulation of the quick-changing part remains.

Similarly, versions can be used to constant-fold arbitrary functions of large data
structures. The version needs to be updated carefully every time the result of
this function can change. Therefore this is useful only if the data structure is
expected to change slowly.

%___________________________________________________________________________

\subsection{Conclusion}

In this post I showed how to use \texttt{purefunction} and \texttt{promote} to make a
small but still relevant dynamic object model no longer use any dictionary lookups
after tracing. Instead a number of guards are inserted into the
trace to check whether the assumptions about the objects are still true. This
makes operations on objects seriously faster. I plan to write another small post
that shows the speed benefits for PyPy's Python interpreter for exactly these
operations.

\section{Evaluation}
\label{sect:evaluation}

\section{Related Work}

\section{Conclusion and Next Steps}

\section*{Acknowledgements}

\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

### 2011-03-24 talk/icooolps2011/paper.tex
- improve fixed-width font - turn some of the code into figures

@@
 \usepackage{xspace}
+\usepackage[scaled=0.8]{beramono}
 \usepackage[utf8]{inputenc}
 
-\makeatletter
-\def\PY@reset{\let\PY@it=\relax \let\PY@bf=\relax%
-    \let\PY@ul=\relax \let\PY@tc=\relax%
-    \let\PY@bc=\relax \let\PY@ff=\relax}
-\def\PY@tok#1{\csname PY@tok@#1\endcsname}
-\def\PY@toks#1+{\ifx\relax#1\empty\else%
-    \PY@tok{#1}\expandafter\PY@toks\fi}
-\def\PY@do#1{\PY@bc{\PY@tc{\PY@ul{%
-    \PY@it{\PY@bf{\PY@ff{#1}}}}}}}
-\def\PY#1#2{\PY@reset\PY@toks#1+\relax+\PY@do{#2}}
-
-\def\PY@tok@gd{\def\PY@bc##1{\fcolorbox[rgb]{0.80,0.00,0.00}{1.00,0.80,0.80}{##1}}}
-\def\PY@tok@gu{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.00}{##1}}}
-\def\PY@tok@gt{\def\PY@tc##1{\textcolor[rgb]{0.60,0.80,0.40}{##1}}}
-\def\PY@tok@gs{\let\PY@bf=\textbf}
-\def\PY@tok@gr{\def\PY@tc##1{\textcolor[rgb]{1.00,0.00,0.00}{##1}}}
-\def\PY@tok@cm{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
-\def\PY@tok@vg{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
-\def\PY@tok@m{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
-\def\PY@tok@mh{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
-\def\PY@tok@cs{\let\PY@bf=\textbf\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
-\def\PY@tok@ge{\let\PY@it=\textit}
-\def\PY@tok@vc{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
-\def\PY@tok@il{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
-\def\PY@tok@go{\def\PY@tc##1{\textcolor[rgb]{0.67,0.67,0.67}{##1}}}
-\def\PY@tok@cp{\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,0.60}{##1}}}
-\def\PY@tok@gi{\def\PY@bc##1{\fcolorbox[rgb]{0.00,0.80,0.00}{0.80,1.00,0.80}{##1}}}
-\def\PY@tok@gh{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.00}{##1}}}
-\def\PY@tok@ni{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,0.60}{##1}}}
-\def\PY@tok@nl{\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,1.00}{##1}}}
-\def\PY@tok@nn{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.80,1.00}{##1}}}
-\def\PY@tok@no{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.00}{##1}}}
-\def\PY@tok@na{\def\PY@tc##1{\textcolor[rgb]{0.20,0.00,0.60}{##1}}}
-\def\PY@tok@nb{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.40}{##1}}}
-\def\PY@tok@nc{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.67,0.53}{##1}}}
-\def\PY@tok@nd{\def\PY@tc##1{\textcolor[rgb]{0.60,0.60,1.00}{##1}}}
-\def\PY@tok@ne{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.80,0.00,0.00}{##1}}}
-\def\PY@tok@nf{\def\PY@tc##1{\textcolor[rgb]{0.80,0.00,1.00}{##1}}}
-\def\PY@tok@si{\def\PY@tc##1{\textcolor[rgb]{0.67,0.00,0.00}{##1}}}
-\def\PY@tok@s2{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-\def\PY@tok@vi{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
-\def\PY@tok@nt{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.20,0.00,0.60}{##1}}}
-\def\PY@tok@nv{\def\PY@tc##1{\textcolor[rgb]{0.00,0.20,0.20}{##1}}}
-\def\PY@tok@s1{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-\def\PY@tok@gp{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,0.60}{##1}}}
-\def\PY@tok@sh{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-\def\PY@tok@ow{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.00,0.00}{##1}}}
-\def\PY@tok@sx{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-\def\PY@tok@bp{\def\PY@tc##1{\textcolor[rgb]{0.20,0.40,0.40}{##1}}}
-\def\PY@tok@c1{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
-\def\PY@tok@kc{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
-\def\PY@tok@c{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.00,0.60,1.00}{##1}}}
-\def\PY@tok@mf{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
-\def\PY@tok@err{\def\PY@tc##1{\textcolor[rgb]{0.67,0.00,0.00}{##1}}\def\PY@bc##1{\colorbox[rgb]{1.00,0.67,0.67}{##1}}}
-\def\PY@tok@kd{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
-\def\PY@tok@ss{\def\PY@tc##1{\textcolor[rgb]{1.00,0.80,0.20}{##1}}}
-\def\PY@tok@sr{\def\PY@tc##1{\textcolor[rgb]{0.20,0.67,0.67}{##1}}}
-\def\PY@tok@mo{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
-\def\PY@tok@mi{\def\PY@tc##1{\textcolor[rgb]{1.00,0.40,0.00}{##1}}}
-\def\PY@tok@kn{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
-\def\PY@tok@o{\def\PY@tc##1{\textcolor[rgb]{0.33,0.33,0.33}{##1}}}
-\def\PY@tok@kr{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
-\def\PY@tok@s{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-\def\PY@tok@kp{\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
-\def\PY@tok@w{\def\PY@tc##1{\textcolor[rgb]{0.73,0.73,0.73}{##1}}}
-\def\PY@tok@kt{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.47,0.53}{##1}}}
-\def\PY@tok@sc{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-\def\PY@tok@sb{\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-\def\PY@tok@k{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.00,0.40,0.60}{##1}}}
-\def\PY@tok@se{\let\PY@bf=\textbf\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-\def\PY@tok@sd{\let\PY@it=\textit\def\PY@tc##1{\textcolor[rgb]{0.80,0.20,0.00}{##1}}}
-
-\def\PYZbs{\char`\\}
-\def\PYZus{\char`\_}
-\def\PYZob{\char`\{}
-\def\PYZcb{\char`\}}
-\def\PYZca{\char`\^}
-% for compatibility with earlier versions
-\def\PYZat{@}
-\def\PYZlb{[}
-\def\PYZrb{]}
-\makeatother
+\input{code/style.tex}
 
@@
 
+%___________________________________________________________________________
 \section{Introduction}
 
+%___________________________________________________________________________
 \section{The PyPy Project}
@@
 XXX
-
+\cite{armin_rigo_pypys_2006}
+
+%___________________________________________________________________________
 \section{Tracing JIT Compilers}
@@
 
+%___________________________________________________________________________
 \section{Controlling The Extent of Tracing}
 
-The question I was asked most often during my recent \href{http://morepypy.blogspot.com/2011/03/us-trip-report-popl-microsoft-ibm.html}{US trip} was how exactly
+XXX how exactly
 the hints work that interpreter authors can use to improve the execution speed
-of the programs running on their interpreters. Since those hints are not really
-documented all that well, I decided to write blog posts about them. This is the
-first one.
-
-%___________________________________________________________________________
+of the programs running on their interpreters?
 
@@
 
-As was explained in a \href{http://morepypy.blogspot.com/2009/03/applying-tracing-jit-to-interpreter.html}{blog post} and a \href{http://codespeak.net/svn/pypy/extradoc/talk/icooolps2009/bolz-tracing-jit.pdf}{paper} two years ago, PyPy's JIT is a
+XXX As was explained in a \href{http://morepypy.blogspot.com/2009/03/applying-tracing-jit-to-interpreter.html}{blog post} and a \href{http://codespeak.net/svn/pypy/extradoc/talk/icooolps2009/bolz-tracing-jit.pdf}{paper} two years ago, PyPy's JIT is a
 meta-tracer. Since we want to re-use our tracer for a variety of languages, we
@@
 in the user program.
-
-%___________________________________________________________________________
 
@@
 
-%___________________________________________________________________________
-
 \subsection{Influencing the Default Behaviour}
@@
 that by applying a hint to the tracer. These hints take the form of function
-decorators (which both live in the \texttt{pypy.rlib.jit} module). In the next two
-subsections I will describe these two function decorators and their use.
-
-%___________________________________________________________________________
+decorators (which both live in the \Verb|pypy.rlib.jit| module). In the next two
+subsections we describe these two function decorators and their use.
 
@@
 
-%___________________________________________________________________________
-
 \subsubsection{Preventing the Tracing of Functions}
@@
 a residual call instead. This decorator is in many ways less important than the
-unrolling one (except for a special situation that I will describe in a
-follow-up post). It is used if tracing into a function is not expected to yield
+unrolling one (except for a special situation that is described in
+Section XXX). It is used if tracing into a function is not expected to yield
 any speed benefits, because the optimizer will not be able to improve it much.
@@
 
-%___________________________________________________________________________
-
 \subsection{Conclusion}
 
-In this post we discussed two hints that can be used to control precisely which
+In this section we discussed two hints that can be used to control precisely which
 parts of the interpreter should be meta-traced. If these hints are used
@@
 
-In the next part of this series I will discuss a different set of hints that can
+In the next section we discuss a different set of hints that can
 be used to strongly optimize traces.
 
-% Document title
-\section{Controlling the Tracing of an Interpreter With Hints, Part 2: Controlling Optimization}
-
-This is part 2 of a series on how to speed up an interpreter written with PyPy
-by adding JIT hints to the interpreter. Part 1 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with.html}{control the
-extent of tracing}. In this post I will describe how to add hints that
+%___________________________________________________________________________
+
+\section{Controlling Optimization}
+
+The last section described how to control the
+extent of tracing. In this section we will describe how to add hints that
 influence the optimizer.  If applied correctly these techniques can give
@@
 resulting program actually slower.
-
-%___________________________________________________________________________
 
@@
 \begin{itemize}
-
-\item \href{http://en.wikipedia.org/wiki/Constant_folding}{constant folding}
-
-\item \href{http://en.wikipedia.org/wiki/Common_subexpression_elimination}{common subexpression elimination}
-
-\item allocation removal, as described in the paper that I recently \href{http://morepypy.blogspot.com/2011/03/us-trip-report-popl-microsoft-ibm.html}{presented at
-PEPM}
-
+    \item constant folding
+    \item common subexpression elimination
+    \item allocation removal \cite{bolz_allocation_2011}
     \item store/load propagation
-
-\item \href{http://morepypy.blogspot.com/2011/01/loop-invariant-code-motion.html}{loop invariant code motion}
-
+    \item loop invariant code motion
 \end{itemize}
@@
 
-In this post I will describe two hints that allow the interpreter author to
+In this section we describe two hints that allow the interpreter author to
 increase the optimization opportunities for constant folding. For constant
@@
 \begin{itemize}
-
     \item the arguments of an operation actually need to all be constant,
     i.e. statically known by the optimizer
-
     \item the operation needs to be \emph{pure}, i.e. always yield the same result given
     the same arguments.
-
 \end{itemize}
@@
 RPython source of the interpreter. Normal Python users will never see them.
-
-%___________________________________________________________________________
 
@@
 in practice the argument types tend to not vary often. Therefore it is possible to
-promote the types. In the next blog post I will give a complete example of how
+promote the types. The next section will present a complete example of how
 this works.
-
-%___________________________________________________________________________
 
@@
 annotation.
-
-%___________________________________________________________________________
 
@@
 
-%___________________________________________________________________________
-
 \subsubsection{Immutable Fields}
@@
 
-%___________________________________________________________________________
-
 \subsection{Conclusion}
 
-In this blog post I explained two more hints that can be used in the source code
+In this section we presented two more hints that can be used in the source code
 of the interpreter. They are used to influence what the optimizer does with the
-trace. I realize the examples given here are a bit too small, in the next
-installment I will give a worked-out example that puts all the pieces together.
-
-\section{Controlling the Tracing of an Interpreter With Hints, Part 3: Putting Things Together}
-
-This is part 3 of the series on how to speed up an interpreter written with
-PyPy by adding JIT hints to the interpreter. Part 1 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with.html}{control
-the extent of tracing}. Part 2 described how to \href{http://morepypy.blogspot.com/2011/03/controlling-tracing-of-interpreter-with_15.html}{influence the optimizer with
-promotion and pure functions}. In this post I describe a worked-out example of
+trace. The examples given here are a bit too small, the next
+section gives a worked-out example that puts all the pieces together.
+
+%___________________________________________________________________________
+
+\section{Putting Things Together}
+
+In this section we describe a worked-out example of
 a small object model for a dynamic language and how to make it efficient using
-the hints described in the previous posts.
+the hints described in the previous sections.
 
@@
 
-For the purpose of this blog post we will use a very simple and bare-bones
+For the purpose of this section we will use a very simple and bare-bones
 object model that just supports very simple classes and instances, without any
@@
 of the interpreter source code:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-
-    \PY{k}{def} \PY{n+nf}{instantiate}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n}{Instance}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
-            \PY{k}{return} \PY{n}{result}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-
-\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-
-    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
-            \PY{k}{return} \PY{n}{result}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-
-    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{try}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-\end{Verbatim}
+
+\begin{figure}
+\input{code/interpreter-slow.tex}
+\caption{Original Version of a Simple Object Model}
+\label{fig:interpreter-slow}
+\end{figure}
 
@@
 The trace could look like this:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{c}{# inst.getattr("a")}
-\PY{n}{attributes1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
-\PY{n}{result1} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result1} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-
-\PY{c}{# inst.getattr("b")}
-\PY{n}{attributes2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
-\PY{n}{v1} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{v1} \PY{o+ow}{is} \PY{n+nb+bp}{None}\PY{p}{)}
-\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{methods1} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
-\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}
-
-\PY{c}{# inst.getattr("c")}
-\PY{n}{attributes3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{attributes}
-\PY{n}{v3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{attributes3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{v3} \PY{o+ow}{is} \PY{n+nb+bp}{None}\PY{p}{)}
-\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{methods2} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
-\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-
-\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
-\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
-\end{Verbatim}
+
+\begin{figure}
+\input{code/trace1.tex}
+\caption{Trace Through the Object Model}
+\label{fig:trace1}
+\end{figure}
 
@@
 away the dictionary lookups on the instances. The hints we have looked at in the
-two earlier blog posts don't seem to help with the current object model. There is
+two previous sections don't seem to help with the current object model. There is
 no pure function to be seen, and the instance is not a candidate for promotion,
@@
 implementation into a shared object. This shared layout object is called a
-\emph{map}. Maps are an old idea that comes originally from the SELF language. They are
-also used by many JavaScript implementations such as V8. I've \href{http://morepypy.blogspot.com/2010/11/efficiently-implementing-python-objects.html}{written about maps
-before}, so I won't explain them fully again.
+\emph{map}. Maps are an old idea that comes originally from the SELF language \cite{XXX}. They are
+also used by many JavaScript implementations such as V8.
 
 The rewritten \texttt{Instance} class using maps looks like this:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{new\PYZus{}map\PYZus{}with\PYZus{}additional\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{if} \PY{n}{name} \PY{o+ow}{not} \PY{o+ow}{in} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{:}
-            \PY{n}{newmap} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
-            \PY{n}{newmap}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{o}{.}\PY{n}{update}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{)}
-            \PY{n}{newmap}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n+nb}{len}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attribute\PYZus{}indexes}\PY{p}{)}
-            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{newmap}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]}
-
-\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
-
-\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}
-
-    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
-            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-            \PY{k}{return}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{new\PYZus{}map\PYZus{}with\PYZus{}additional\PYZus{}attribute}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{n}{value}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{try}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-\end{Verbatim}
+
+\begin{figure}
+\input{code/map.tex}
+\caption{Simple Object Model With Maps}
+\label{fig:maps}
+\end{figure}
 
@@
 
-In this post I showed how to use \texttt{purefunction} and \texttt{promote} to make a
+In this section we saw how to use \texttt{purefunction} and \texttt{promote} to make a
 small but still relevant dynamic object model no longer use any dictionary lookups
@@
 trace to check whether the assumptions about the objects are still true. This
-makes operations on objects seriously faster. I plan to write another small post
-that shows the speed benefits for PyPy's Python interpreter for exactly these
-operations.
+makes operations on objects seriously faster.
 

### 2011-03-24 talk/icooolps2011/code/map.tex
- improve fixed-width font - turn some of the code into figures

\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{if} \PY{n}{name} \PY{o+ow}{not} \PY{o+ow}{in} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{:}
            \PY{n}{newmap} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
            \PY{n}{newmap}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{update}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{p}{)}
            \PY{n}{newmap}\PY{o}{.}\PY{n}{indexes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n+nb}{len}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{p}{)}
            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{newmap}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]}

\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}

\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}

    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]} \PY{o}{=} \PY{n}{value}
            \PY{k}{return}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{add\PYZus{}attribute}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{n}{value}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{try}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
\end{Verbatim}

### 2011-03-24 talk/icooolps2011/code/map.py

@@ -1,0 +1,47 @@
+class Map(object):
+    def __init__(self):
+        self.indexes = {}
+        self.other_maps = {}
+
+    @purefunction
+    def getindex(self, name):
+        return self.indexes.get(name, -1)
+
+    @purefunction
+    def add_attribute(self, name):
+        if name not in self.other_maps:
+            newmap = Map()
+            newmap.indexes.update(self.indexes)
+            newmap.indexes[name] = len(self.indexes)
+            self.other_maps[name] = newmap
+        return self.other_maps[name]
+
+EMPTY_MAP = Map()
+
+class Instance(object):
+    def __init__(self, cls):
+        self.cls = cls
+        self.map = EMPTY_MAP
+        self.storage = []
+
+    def getfield(self, name):
+        map = hint(self.map, promote=True)
+        index = map.getindex(name)
+        if index != -1:
+            return self.storage[index]
+        raise AttributeError(name)
+
+    def write_attribute(self, name, value):
+        map = hint(self.map, promote=True)
+        index = map.getindex(name)
+        if index != -1:
+            self.storage[index] = value
+            return
+        self.map = map.add_attribute(name)
+        self.storage.append(value)
+
+    def getattr(self, name):
+        try:
+            return self.getfield(name)
+        except AttributeError:
+            return self.cls.find_method(name)

### 2011-03-24 talk/icooolps2011/paper.tex
more figureyfication

@@
 
+XXX how exactly
+the hints work that interpreter authors can use to improve the execution speed
+of the programs running on their interpreters?
+
 %___________________________________________________________________________
@@
 \section{Controlling The Extent of Tracing}
-
-XXX how exactly
-the hints work that interpreter authors can use to improve the execution speed
-of the programs running on their interpreters?
 
@@
 
-Here's a diagram of this process:
-
 \begin{figure*}
@@
 
-On the left you see the levels of execution. The CPU executes the binary of
+Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
+see the levels of execution. The CPU executes the binary of
 PyPy's Python interpreter, which consists of RPython functions that have been
@@
 Let's assume we trace through code that sums three attributes, such as:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)} \PY{o}{+} \PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)} \PY{o}{+} \PY{n}{inst}\PY{o}{.}\PY{n}{getattr}\PY{p}{(}\PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
+
+\begin{Verbatim}
+inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
 \end{Verbatim}
-
-The trace could look like this:
 
@@
 
-In this example, the attribute \texttt{a} is found on the instance, but the
+The trace would look like in Figure~\ref{fig:trace1}. In this example, the
+attribute \texttt{a} is found on the instance, but the
 attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
@@
 following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
-has been promoted:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{c}{# inst.getattr("a")}
-\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
-\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
-\PY{n}{index1} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{index1} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
-\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{n}{index1}\PY{p}{]}
+has been promoted, see Figure~\ref{fig:trace2}.
 
-\PY{c}{# inst.getattr("b")}
-\PY{n}{map2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
-\PY{n}{guard}\PY{p}{(}\PY{n}{map2} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
-\PY{n}{index2} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{index2} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{methods1} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
-\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}
-
-\PY{c}{# inst.getattr("c")}
-\PY{n}{map3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
-\PY{n}{guard}\PY{p}{(}\PY{n}{map3} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
-\PY{n}{index3} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{index3} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{methods2} \PY{o}{=} \PY{n}{cls}\PY{o}{.}\PY{n}{methods}
-\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-
-\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
-\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
-\end{Verbatim}
+\begin{figure}
+\input{code/trace2.tex}
+\caption{Unoptimized Trace After the Introduction of Maps}
+\label{fig:trace2}
+\end{figure}
 
@@
 the map will be optimized away too, because the map cannot have changed in
-between. The optimized trace looks like this:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{c}{# inst.getattr("a")}
-\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
-\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
-\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
-\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}
+between. The optimized trace looks can be seen in Figure~\ref{fig:trace3}
 
-\PY{c}{# inst.getattr("b")}
-\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{methods1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{methods}
-\PY{n}{result2} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}
-
-\PY{c}{# inst.getattr("c")}
-\PY{n}{cls2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{methods2} \PY{o}{=} \PY{n}{cls2}\PY{o}{.}\PY{n}{methods}
-\PY{n}{result3} \PY{o}{=} \PY{n+nb}{dict}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{methods2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-
-\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
-\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
-\end{Verbatim}
+\begin{figure}
+\input{code/trace3.tex}
+\caption{Optimized Trace After the Introduction of Maps}
+\label{fig:trace3}
+\end{figure}
 
@@
 explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} looks like
-this:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{VersionTag}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{pass}
+in Figure~\ref{fig:version}
 
-\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{version} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}
-        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
-            \PY{k}{return} \PY{n}{result}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}
-\end{Verbatim}
+\begin{figure}
+\input{code/version.tex}
+\caption{Versioning of Classes}
+\label{fig:version}
+\end{figure}
 
@@
 
-The trace with this new class implementation looks like this:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{c}{# inst.getattr("a")}
-\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
-\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
-\PY{n}{index1} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map1}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{a}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{index1} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
-\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{n}{index1}\PY{p}{]}
+\begin{figure}
+\input{code/trace4.tex}
+\caption{Unoptimized Trace After Introduction of Versioned Classes}
+\label{fig:trace4}
+\end{figure}
 
-\PY{c}{# inst.getattr("b")}
-\PY{n}{map2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
-\PY{n}{guard}\PY{p}{(}\PY{n}{map2} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
-\PY{n}{index2} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map2}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{index2} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{guard}\PY{p}{(}\PY{n}{cls1} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
-\PY{n}{version1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{version}
-\PY{n}{guard}\PY{p}{(}\PY{n}{version1} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
-\PY{n}{result2} \PY{o}{=} \PY{n}{Class}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{cls}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{b}\PY{l+s}{"}\PY{p}{,} \PY{n}{version1}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result2} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{n}{result2}
-
-\PY{c}{# inst.getattr("c")}
-\PY{n}{map3} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
-\PY{n}{guard}\PY{p}{(}\PY{n}{map3} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
-\PY{n}{index3} \PY{o}{=} \PY{n}{Map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{map3}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{index3} \PY{o}{==} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-\PY{n}{cls2} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{guard}\PY{p}{(}\PY{n}{cls2} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
-\PY{n}{version2} \PY{o}{=} \PY{n}{cls2}\PY{o}{.}\PY{n}{version}
-\PY{n}{guard}\PY{p}{(}\PY{n}{version2} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
-\PY{n}{result3} \PY{o}{=} \PY{n}{Class}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{cls}\PY{p}{,} \PY{l+s}{"}\PY{l+s}{c}\PY{l+s}{"}\PY{p}{,} \PY{n}{version2}\PY{p}{)}
-\PY{n}{guard}\PY{p}{(}\PY{n}{result3} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{)}
-
-\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{n}{result3}
-\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
-\end{Verbatim}
-
+The trace with this new class implementation can be seen in
+Figure~\ref{fig:trace4}.
 The calls to \texttt{Class.\_find\_method} can now be optimized away, also the
 promotion of the class and the version, except for the first one. The final
-optimized trace looks like this:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{c}{# inst.getattr("a")}
-\PY{n}{map1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{map}
-\PY{n}{guard}\PY{p}{(}\PY{n}{map1} \PY{o}{==} \PY{l+m+mh}{0xb74af4a8}\PY{p}{)}
-\PY{n}{storage1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{storage}
-\PY{n}{result1} \PY{o}{=} \PY{n}{storage1}\PY{p}{[}\PY{l+m+mi}{0}\PY{p}{]}
+optimized trace can be seen in Figure~\ref{fig:trace5}.
 
-\PY{c}{# inst.getattr("b")}
-\PY{n}{cls1} \PY{o}{=} \PY{n}{inst}\PY{o}{.}\PY{n}{cls}
-\PY{n}{guard}\PY{p}{(}\PY{n}{cls1} \PY{o}{==} \PY{l+m+mh}{0xb7aaaaf8}\PY{p}{)}
-\PY{n}{version1} \PY{o}{=} \PY{n}{cls1}\PY{o}{.}\PY{n}{version}
-\PY{n}{guard}\PY{p}{(}\PY{n}{version1} \PY{o}{==} \PY{l+m+mh}{0xb7bbbb18}\PY{p}{)}
-\PY{n}{v2} \PY{o}{=} \PY{n}{result1} \PY{o}{+} \PY{l+m+mi}{41}
-
-\PY{c}{# inst.getattr("c")}
-\PY{n}{v4} \PY{o}{=} \PY{n}{v2} \PY{o}{+} \PY{l+m+mi}{17}
-\PY{k}{return}\PY{p}{(}\PY{n}{v4}\PY{p}{)}
-\end{Verbatim}
+\begin{figure}
+\input{code/trace5.tex}
+\caption{Optimized Trace After Introduction of Versioned Classes}
+\label{fig:trace5}
+\end{figure}
 

### 2011-03-24 talk/icooolps2011/code/interpreter-slow.tex
more figureyfication

--- 2011-03-24 talk/icooolps2011/code/map.tex
@@
 \begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
+\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
+    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
+        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
+        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
 
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
+    \PY{k}{def} \PY{n+nf}{instantiate}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
+        \PY{k}{return} \PY{n}{Instance}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}
 
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{if} \PY{n}{name} \PY{o+ow}{not} \PY{o+ow}{in} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{:}
-            \PY{n}{newmap} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
-            \PY{n}{newmap}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{update}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{p}{)}
-            \PY{n}{newmap}\PY{o}{.}\PY{n}{indexes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n+nb}{len}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{p}{)}
-            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{newmap}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]}
+    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
+        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
+            \PY{k}{return} \PY{n}{result}
+        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
 
-\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
+    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
+        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
 
@@
         \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}
+        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
 
     \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
+        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
+            \PY{k}{return} \PY{n}{result}
         \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
@@
     \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
-            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-            \PY{k}{return}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{add\PYZus{}attribute}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{n}{value}\PY{p}{)}
+        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
 

### 2011-03-24 talk/icooolps2011/code/version.tex
more figureyfication

--- 2011-03-24 talk/icooolps2011/code/interpreter-slow.tex
@@
 \begin{Verbatim}[commandchars=\\\{\}]
+\PY{k}{class} \PY{n+nc}{VersionTag}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
+    \PY{k}{pass}
+
 \PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
@@
         \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-
-    \PY{k}{def} \PY{n+nf}{instantiate}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n}{Instance}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}
+        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}
 
     \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+        \PY{n+nb+bp}{self} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
+        \PY{n}{version} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
+        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}
         \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
@@
 
+    \PY{n+nd}{@purefunction}
+    \PY{k}{def} \PY{n+nf}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}\PY{p}{:}
+        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+
     \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
         \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-
-\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-
-    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
-            \PY{k}{return} \PY{n}{result}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-
-    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{try}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}
 \end{Verbatim}

### 2011-03-24 talk/icooolps2011/paper.tex
kill part about unroll_safe and dont_look_inside. We have a space-problem and this material is not that important.

@@
 in the user program.
-
-\subsection{How Far Should Tracing Go}
-
-When the tracer encounters a function call at the interpreter level, e.g. the
-interpreter main loop calling a helper function, it can do one of two things:
-
-\begin{enumerate}
-\item it can trace into the helper function, effectively inlining it into the trace.
-
-\item it can not trace into the function and instead record a call to that function
-as an operation in the trace. Such a call operation in the trace is sometimes
-called \emph{residual call}.
-\end{enumerate}
-
-As a default, the tracer will try to trace into the helper because that will
-give more information to the optimizer, allowing it to do a better job. This is
-particularly important for the allocation removal optimization, because if a
-freshly allocated object is passed as an argument to a residual call, its
-allocation cannot be optimized away.
-
-There is a problem however if the helper function itself contains a loop. The
-tracer records the linear sequence of operations that are being executed. Thus
-when it encounters a loop on the interpreter level it records all the
-operations of every iteration of the loop itself, with the net effect of
-unrolling it. The only places where the tracer stops and tries to close the
-trace is in the main loop of the interpreter. When the tracer encounters the
-main loop, it also checks whether the original user loop has been closed, and
-thus whether it can stop tracing.
-
-For most helper functions in the interpreter that contain loops, fully
-unrolling does not make sense. If a loop is unrolled, the trace is specific to
-the number of iteration that was seen during tracing. If the trace is later
-executed with a different number of iterations, the trace will be left via a
-guard failure, which is inefficient. Therefore the default behaviour of the
-tracer is to never trace into a function on the interpreter level that contains
-a loop, but to trace into all non-looping helper functions.
-
-This default behaviour is essentially a heuristic, but one that usually makes
-sense. We want to produce just enough traces to make the resulting code
-efficient, but not more. Therefore we trace as much as possible (everything by
-default) except the functions which loops where tracing would produce code that
-is less general than it could be.
-
-As an example for a helper with a loop, take string concatenation. It loops over
-the characters of both arguments and copies them over into the result string. It
-does not make sense to unroll the loops in this function. If we do that,
-the resulting trace can only be used for strings of the length that was seen
-during tracing. In practise, the string lengths are usually different each run,
-meaning that the trace with unrolling is not run to completion in most cases.
-
-\subsection{Influencing the Default Behaviour}
-
-Sometimes the default behaviour is not actually what is wanted. This is
-something the interpreter author has to decide, usually by looking at the traces
-that are produced and deciding that they should be improved. There are two ways
-in which the default is wrong:
-%
-\begin{itemize}
-
-\item \textbf{false negatives:} if a helper function that \textbf{does} contain a loop should
-be traced into, unrolling the loop.
-
-\item \textbf{false positives:} if a helper function that \textbf{does not} contain a loop is
-inlined into the trace, but the interpreter author decides that this is not
-helpful.
-
-\end{itemize}
-
-If the interpreter author finds false negatives or false positives, she can fix
-that by applying a hint to the tracer. These hints take the form of function
-decorators (which both live in the \Verb|pypy.rlib.jit| module). In the next two
-subsections we describe these two function decorators and their use.
-
-\subsubsection{Unrolling Functions With Loops}
-
-The first decorator, used to fix false negatives, is the \texttt{unroll\_safe}
-decorator. It is used to tell the tracer to always trace into a function that
-has a loop, effectively unrolling the loop. This decorator should be used only
-if the loop in the helper function is expected to always run for the same number
-of iterations. This sounds like a strong restriction, in practise this is less
-severe: The number of iterations needs to only be the same \emph{in the context where
-the helper functions is traced from}.
-
-It is easiest to understand this condition via an example. Let's look at the
-\texttt{BUILD\_TUPLE} bytecode in Python. It takes one argument, the length \texttt{n} of
-the tuple being built. The bytecode pops \texttt{n} arguments from the stack, turns
-them into a tuple and pushes that tuple on the stack. Thus the function that
-implements \texttt{BUILD\_TUPLE} in PyPy's Python interpreter calls a helper
-\texttt{popvalues} which pops \texttt{n} values from the stack and returns them in a list.
-This helper is implemented with a loop and would thus not be traced into by
-default.  The loop in the helper can run for very different numbers of
-iterations, because it is used in a variety of places. However, for every
-concrete \texttt{BUILD\_TUPLE} bytecode, the argument will be constant. Therefore it
-is safe (and even necessary) to annotate \texttt{popvalues} with the \texttt{unroll\_safe}
-decorator.
-
-A different example is the implementation of the \texttt{isinstance} builtin. It is
-used to check whether an object \texttt{a} is an instance of a class \texttt{B} like
-this: \texttt{isinstance(a, B)}. The second argument of the function can also be a
-tuple of classes to check whether an object is an instance of one of a number of
-classes: \texttt{isinstance(a, (A, B, C, D))}. To implement this second case, the
-implementation of \texttt{isinstance} contains a loop iterating over the elements of
-the tuple. The number of loop iterations can vary, but is usually fixed for each
-individual call site which typically just lists a few classes in the source
-code. Therefore it is also safe to annotate the implementation of \texttt{isinstance}
-with the \texttt{unroll\_safe} decorator.
-
-\subsubsection{Preventing the Tracing of Functions}
-
-The second decorator \texttt{dont\_look\_inside} is used to fix false positives. It
-tells the JIT to never trace into the decorated function and just always produce
-a residual call instead. This decorator is in many ways less important than the
-unrolling one (except for a special situation that is described in
-Section XXX). It is used if tracing into a function is not expected to yield
-any speed benefits, because the optimizer will not be able to improve it much.
-This is often the case if the called helper function does not contain any
-``dynamic'' behaviour. In such a situation it is better to just leave the function
-call in the trace, because that produces less code.
-
-An example would be the import mechanism in Python. It's very unlikely that any
-performance improvement can be had by turning part of it into assembler.
-Therefore we hide it from the tracer by annotating them with
-\texttt{dont\_look\_inside}.
-
-\subsection{Conclusion}
-
-In this section we discussed two hints that can be used to control precisely which
-parts of the interpreter should be meta-traced. If these hints are used
-carefully, this can go a long way to making the interpreter produce traces that
-contain exactly the interesting part of the execution, and will contain calls to
-the functions that can not be optimized by tracing techniques.
-
-In the next section we discuss a different set of hints that can
-be used to strongly optimize traces.
-
-%___________________________________________________________________________
 

### 2011-03-24 talk/icooolps2011/paper.tex
use Verbatim environments
reshuffle some stuff, kill one of the traces

@@
 
+\definecolor{gray}{rgb}{0.5,0.5,0.5}
+
 \begin{document}
@@
 
-%___________________________________________________________________________
-\section{The PyPy Project}
+\section{Background}
+\label{sec:Background}
+
+\subsection{The PyPy Project}
 \label{sect:pypy}
 
-XXX
-\cite{armin_rigo_pypys_2006}
-
-%___________________________________________________________________________
-\section{Tracing JIT Compilers}
-\label{sect:tracing}
-
-XXX
-
-%___________________________________________________________________________
-\section{Controlling The Extent of Tracing}
-
-\subsection{Background}
-
-First, let's recap some basics: PyPy's approach to implementing dynamic
+PyPy's approach to implementing dynamic
 languages is to write an interpreter for
@@
 
-XXX As was explained in a \href{http://morepypy.blogspot.com/2009/03/applying-tracing-jit-to-interpreter.html}{blog post} and a \href{http://codespeak.net/svn/pypy/extradoc/talk/icooolps2009/bolz-tracing-jit.pdf}{paper} two years ago, PyPy's JIT is a
-meta-tracer. Since we want to re-use our tracer for a variety of languages, we
+XXX \cite{armin_rigo_pypys_2006}
+
+%___________________________________________________________________________
+\subsection{PyPy's Meta-Tracing JIT Compilers}
+\label{sect:tracing}
+
+PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. Since we want to re-use
+our tracer for a variety of languages, we
 don't trace the execution of the user program, but instead trace the execution
@@
 
-\section{Controlling Optimization}
-
-The last section described how to control the
-extent of tracing. In this section we will describe how to add hints that
-influence the optimizer.  If applied correctly these techniques can give
-really big speedups by pre-computing parts of what happens at runtime. On the other
-hand, if applied incorrectly they might lead to code bloat, thus making the
-resulting program actually slower.
-
-\subsection{Background}
+\subsection{Optimizing Traces}
+\label{sub:optimizing}
 
@@
 optimized.  The optimizer applies a number of techniques to remove or reduce
-the number of operations: most of these are well known \href{http://en.wikipedia.org/wiki/Compiler_optimization\#Optimization_techniques}{compiler optimization
-techniques}, with the difference that it is easier to apply them in a tracing
+the number of operations: most of these are well known compiler optimization
+techniques, with the difference that it is easier to apply them in a tracing
 JIT because it only has to deal with linear traces.  Among the techniques:
@@
 
-In this section we describe two hints that allow the interpreter author to
-increase the optimization opportunities for constant folding. For constant
-folding to work, two conditions need
-to be met:
-%
+% section Background (end)
+%___________________________________________________________________________
+
+\section{Controlling Optimization}
+
+In this section we will describe how to add two hints that allow the
+interpreter author to increase the optimization opportunities for constant
+folding. If applied correctly these techniques can give really big speedups by
+pre-computing parts of what happens at runtime. On the other
+hand, if applied incorrectly they might lead to code bloat, thus making the
+resulting program actually slower.
+
+For constant folding to work, two conditions need to be met:
+
 \begin{itemize}
@@
 opportunities. There is one kind of hint for both of the conditions above.
-
-\textbf{Note}: These hints are written by an interpreter developer and applied to the
-RPython source of the interpreter. Normal Python users will never see them.
 
@@
 %
-\begin{quote}{\ttfamily \raggedright \noindent
-guard(x~==~4)\\
-y~=~y~+~x
-}
-\end{quote}
+\begin{Verbatim}
+guard(x == 4)
+y = y + x
+\end{Verbatim}
 
@@
 into a constant value. This process is called \emph{promotion} and it is an old idea
-in partial evaluation (it's called ``the trick'' there). Promotion is also heavily
-used by \href{http://psyco.sourceforge.net/}{Psyco} and by all older versions of PyPy's JIT. Promotion is a technique
-that only works well in JIT compilers, in
-static compilers it is significantly less applicable.
+in partial evaluation (it's called ``the trick'' \cite{XXX} there). Promotion is also heavily
+used by Psyco \cite{rigo_representation-based_2004} and by all older versions
+of PyPy's JIT. Promotion is a technique that only works well in JIT compilers,
+in static compilers it is significantly less applicable.
 
@@
 We get a trace that looks like this:
-%
-\begin{quote}{\ttfamily \raggedright \noindent
-v1~=~x~*~2\\
-z~=~v1~+~1\\
-v2~=~z~+~y\\
+
+\begin{Verbatim}
+v1 = x * 2
+z = v1 + 1
+v2 = z + y
 return(v2)
-}
-\end{quote}
+\end{Verbatim}
 
@@
 operation at the beginning:
-%
-\begin{quote}{\ttfamily \raggedright \noindent
-guard(x~==~4)\\
-v1~=~x~*~2\\
-z~=~v1~+~1\\
-v2~=~z~+~y\\
+
+\begin{Verbatim}
+guard(x == 4)
+v1 = x * 2
+z = v1 + 1
+v2 = z + y
 return(v2)
-}
-\end{quote}
+\end{Verbatim}
 
@@
 trace into:
-%
-\begin{quote}{\ttfamily \raggedright \noindent
-guard(x~==~4)\\
-v2~=~9~+~y\\
+
+\begin{Verbatim}
+guard(x == 4)
+v2 = 9 + y
 return(v2)
-}
-\end{quote}
+\end{Verbatim}
 
@@
 trace looks like this:
-%
-\begin{quote}{\ttfamily \raggedright \noindent
-guard(x~==~2)\\
-v2~=~5~+~y\\
+
+\begin{Verbatim}
+guard(x == 2)
+v2 = 5 + y
 return(v2)
-}
-\end{quote}
+\end{Verbatim}
 
@@
 %
-\begin{quote}{\ttfamily \raggedright \noindent
-x~=~a.x\\
-v1~=~x~*~2\\
-v2~=~v1~+~1\\
-v3~=~v2~+~val\\
-a.y~=~v3
-}
-\end{quote}
+\begin{Verbatim}
+x = a.x
+v1 = x * 2
+v2 = v1 + 1
+v3 = v2 + val
+a.y = v3
+\end{Verbatim}
 
@@
 %
-\begin{quote}{\ttfamily \raggedright \noindent
-guard(a~==~0xb73984a8)\\
-v1~=~compute(a)\\
-v2~=~v1~+~val\\
-a.y~=~v2
-}
-\end{quote}
+\begin{Verbatim}
+guard(a == 0xb73984a8)
+v1 = compute(a)
+v2 = v1 + val
+a.y = v2
+\end{Verbatim}
 
@@
 %
-\begin{quote}{\ttfamily \raggedright \noindent
-guard(a~==~0xb73984a8)\\
-v2~=~9~+~val\\
-a.y~=~v2
-}
-\end{quote}
+\begin{Verbatim}
+guard(a == 0xb73984a8)
+v2 = 9 + val
+a.y = v2
+\end{Verbatim}
 
@@
 following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
-has been promoted, see Figure~\ref{fig:trace2}.
+has been promoted, see Figure~\ref{fig:trace2}. Operations that can be
+optimized away are grayed out.
+
+The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
+a pure function and they have constant arguments. That means that \texttt{index1/2/3}
+are constant and the guards on them can be removed. All but the first guard on
+the map will be optimized away too, because the map cannot have changed in
+between. This trace is already much better than
+the original one. Now we are down from five dictionary lookups to just two.
 
@@
 \end{figure}
-
-The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
-a pure function and they have constant arguments. That means that \texttt{index1/2/3}
-are constant and the guards on them can be removed. All but the first guard on
-the map will be optimized away too, because the map cannot have changed in
-between. The optimized trace looks can be seen in Figure~\ref{fig:trace3}
-
-\begin{figure}
-\input{code/trace3.tex}
-\caption{Optimized Trace After the Introduction of Maps}
-\label{fig:trace3}
-\end{figure}
-
-The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
-of the constant-folded \texttt{getindex} call. This trace is already much better than
-the original one. Now we are down from five dictionary lookups to just two.
 
@@
 
+The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
+of the constant-folded \texttt{getindex} call.
 The constants \texttt{41} and \texttt{17} are the results of the folding of the

### 2011-03-24 talk/icooolps2011/paper.tex
reorder and generally improve things. regenerate the figure

@@
 \input{code/style.tex}
-
-\ifthenelse{\isundefined{\hypersetup}}{
-  \usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
-}{}
-\hypersetup{
-  pdftitle={Controlling the Tracing of an Interpreter With Hints, Part 1: Controlling the Extent of Tracing},
-}
 
@@
 PyPy's Python interpreter, which consists of RPython functions that have been
-compiled first to C, then to machine code. Some of these functions contain
-loops, others don't. The interpreter runs a Python program written by a
+compiled first to C, then to machine code. The interpreter runs a Python program written by a
 programmer (the user). If the tracer is used, it traces operations on the level
@@
 Before sending the trace to the backend to produce actual machine code, it is
-optimized.  The optimizer applies a number of techniques to remove or reduce
-the number of operations: most of these are well known compiler optimization
+optimized.  The optimizer applies a number of techniques to remove or simplify
+the operations in the trace. Most of these are well known compiler optimization
 techniques, with the difference that it is easier to apply them in a tracing
@@
 by the optimizer can be vastly improved.
+
+\subsection{Running Example}
+\label{sub:running}
+
+As the running example of this paper we will use a very simple and bare-bones
+object model that just supports classes and instances, without any
+inheritance or other fancy features. The model has classes, which contain methods.
+Instances have a class. Instances have their own attributes (or fields). When looking up an
+attribute on an instance, the instances attributes are searched. If the
+attribute is not found there, the class' methods are searched.
+
+\begin{figure}
+\input{code/interpreter-slow.tex}
+\caption{Original Version of a Simple Object Model}
+\label{fig:interpreter-slow}
+\end{figure}
+
+To implement this object model, we could use the RPython code in
+Figure~\ref{fig:interpreter-slow} as part of the interpreter source code.
+In this straightforward implementation the methods and attributes are just
+stored in dictionaries (hash maps) on the classes and instances, respectively.
+While this object model is very
+simple it already contains all the hard parts of Python's object model. Both
+instances and classes can have arbitrary fields, and they are changeable at
+any time.  Moreover, instances can change their class after they have been
+created.
+
+When using this object model in
+an interpreter, a huge amount of time will be spent doing lookups in these
+dictionaries.
+Let's assume we trace through code that sums three attributes, such as:
+
+\begin{Verbatim}
+inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
+\end{Verbatim}
+
+\begin{figure}
+\input{code/trace1.tex}
+\caption{Trace Through the Object Model}
+\label{fig:trace1}
+\end{figure}
+
+The trace would look like in Figure~\ref{fig:trace1}. In this example, the
+attribute \texttt{a} is found on the instance, but the
+attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
+five calls to \texttt{dict.get}, which is slow. To make the language efficient
+using a tracing JIT, we need to find a way to get rid of these dictionary
+lookups somehow. How to achieve this will be topic of
+Section~\ref{sec:putting}.
+
+% subsection Running Example (end)
 
@@
 
-\section{Putting Things Together}
+\section{Putting It All Together}
 
-In this section we describe a worked-out example of
-a small object model for a dynamic language and how to make it efficient using
-the hints described in the previous sections.
-
-%___________________________________________________________________________
-
-\subsection{A Simple Object Model}
-
-To implement a dynamic language efficiently, the operations on its objects need
-to be fast. Most dynamic languages have object models that are made by using
-dictionaries everywhere. Let's look at an example of how the JIT can be made to
-optimize such operations.
-
-For the purpose of this section we will use a very simple and bare-bones
-object model that just supports very simple classes and instances, without any
-inheritance or any fancy features. The model has classes, which contain methods.
-Instances have a class. Instances have their own attributes. When looking up an
-attribute on an instance, the instances attributes are searched. If the
-attribute is not found there, the class' attributes are searched.
-
-To implement this object model, we could use the following RPython code as part
-of the interpreter source code:
-
-\begin{figure}
-\input{code/interpreter-slow.tex}
-\caption{Original Version of a Simple Object Model}
-\label{fig:interpreter-slow}
-\end{figure}
-
-In this straightforward implementation the methods and attributes are just
-stored in dictionaries on the classes/instances. While this object model is very
-simple it already contains all the hard parts of Python's object model. Both
-instances and classes can have arbitrary fields, and they are changeable at
-any time.  Moreover, instances can change their class after they have been
-created.
-
-When using this object model in
-an interpreter, a huge amount of time will be spent doing lookups in these
-dictionaries. To make the language efficient using a tracing JIT, we need to
-find a way to get rid of these dictionary lookups somehow.
-
-Let's assume we trace through code that sums three attributes, such as:
-
-\begin{Verbatim}
-inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
-\end{Verbatim}
-
-\begin{figure}
-\input{code/trace1.tex}
-\caption{Trace Through the Object Model}
-\label{fig:trace1}
-\end{figure}
-
-The trace would look like in Figure~\ref{fig:trace1}. In this example, the
-attribute \texttt{a} is found on the instance, but the
-attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
-five calls to \texttt{dict.get}, which is slow.
+In this section we describe how the simple object model from
+Section~\ref{sub:running} can be made efficient using the hints described in the
+previous the section. The object model there is typical for many current
+dynamic languages (such as Python, Ruby and JavaScript) as it relies heavily on
+hash-maps to implement its objects.
 
@@
 Therefore it makes sense to factor the layout information out of the instance
-implementation into a shared object. This shared layout object is called a
-\emph{map}. Maps are an old idea that comes originally from the SELF language \cite{XXX}. They are
-also used by many JavaScript implementations such as V8.
-
-The rewritten \texttt{Instance} class using maps looks like this:
+implementation into a shared object, called the \emph{map}. Maps are a well-known
+technique to efficiently implement instances and come from the SELF project
+\cite{XXX}. They are also used by many JavaScript implementations such as V8.
+The rewritten \texttt{Instance} class using maps can be seen in
+Figure~\ref{fig:maps}.
 
@@
 
-Instances no longer use dictionaries to store their fields. Instead, they have a
+In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
 reference to a map, which maps field names to indexes into a storage list. The
@@
 to an instance, a new map needs to be chosen, which is done with the
-\texttt{new\_map\_with\_additional\_attribute} method on the previous map. Now that we have
+\texttt{add\_attribute} method on the previous map (which is also pure). Now that we have
 introduced maps, it is safe to promote the map everywhere, because we assume
@@
 With this changed instance implementation, the trace we had above changes to the
-following, where \texttt{0xb74af4a8} is the memory address of the Map instance that
-has been promoted, see Figure~\ref{fig:trace2}. Operations that can be
-optimized away are grayed out.
+following that of see Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
+memory address of the \texttt{Map} instance that has been promoted. Operations
+that can be optimized away are grayed out.
 
@@
 Instances were optimized making the assumption that the total number of
-Instance layouts is small compared to the number of instances. For classes we
+different instance layouts is small compared to the number of instances. For classes we
 will make an even stronger assumption. We simply assume that it is rare for
@@
 
-Therefore, we give every class a version number, which is increased every time a
+Therefore, we give every class a version number, which is changed every time a
 class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
@@
 the JIT to detect this case, we factor it out in a helper method which is
-explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} looks like
-in Figure~\ref{fig:version}
+explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} can
+be seen in Figure~\ref{fig:version}
 
@@
 argument but it does not use it at all. Its only purpose is to make the call
-pure (because when the version number changes, the result of the call might be
-different than the previous one).
+pure, because when the version number changes, the result of the call might be
+different than the previous one.
 
@@
 code is executed with an instance that has another layout. The second guard
-checks that the class of \texttt{inst} is still the same. It will fail if trace is
+checks that the class of \texttt{inst} is still the same. It will fail if the trace is
 executed with an instance of another class. The third guard checks that the
@@
 The first problem that needs to be solved is that Python supports (multiple)
-inheritance. Therefore looking up a method in a class needs to consider the
+inheritance. Therefore looking up a method in a class needs to consider all the
+classes in the
 whole method resolution order. This makes the versioning of classes more
@@
 an instance is not stored on the instance itself, but on the map. This means
-that we get one fewer promotion (and thus one fewer guard) in the trace, because the class doesn't need to
+that we get one fewer promotion (and thus one fewer guard) in the trace,
+because the class doesn't need to
 be promoted after the map has been.

### 2011-03-24 talk/icooolps2011/paper.tex
write background and start intro

@@
 
-XXX how exactly
-the hints work that interpreter authors can use to improve the execution speed
-of the programs running on their interpreters?
+One of the hardest parts of implementing a dynamic language efficiently is to
+optimize its object model. This is made harder by the fact that many recent
+languages such as Python, JavaScript or Ruby have rather complex core object
+semantics. For them, implementing just an interpreter is already an arduous
+task. Implementing them efficiently with a just-in-time compiler is
+nigh-impossible, because or their many corner-cases.
+
+long dream of PE to not have to implement a compiler
+new life by meta-tracers, spur, pypy, other things
+trace execution of object model and get semantics right, not have to compile
+object model
+
+in spur and pypy, this can be improved by the interpreter author by adding
+hints, which influence tracer and optimizer
+
+pypy's hints go further than spurs, in this paper we present two very important
+ones and show how classical implementation techniques of dynlangs can be
+expressed by them
 
@@
 
-PyPy's approach to implementing dynamic
-languages is to write an interpreter for
-the language in RPython. This interpreter can be translated to C and then
-further to machine code. The interpreter consists of code in the form of a
-large number of generated C functions and some data. Similarly, the user
-program consists of functions in the language the interpreter executes.
+The PyPy project \cite{armin_rigo_pypys_2006} strives to be an environment where
+complex dynamic languages can be efficiently implemented. The approach taken
+when implement a language with PyPy is to write an interpreter for the language
+in \emph{RPython}. RPython is a restricted subset of Python chosen in such a way
+that it is possible to perform type inference on it. The interpreters in RPython
+can therefore be translated to efficient C code.
 
-XXX \cite{armin_rigo_pypys_2006}
+A number of languages have been implemented with PyPy, most importantly a full
+Python implementation, but also a Prolog interpreter \cite{XXX} and a Smalltalk
+VM \cite{XXX}.
+
+This translation to C code adds a number of implementation details into the
+final executable that are not present in the interpreter implementation, such as
+a garbage collector. The interpreter can therefore be kept free from low-level
+implementation details. Another aspect of the final VM that is added
+semi-automatically to the generated VM is a tracing JIT compiler.
+
+We call the code that runs on top of an interpreter implemented with PyPy the
+\emph{user code} or \emph{user program}.
 
@@
 \label{sect:tracing}
+
+XXX citations
+A recently popular approach to JIT compilers is that of tracing JITs. Tracing
+JITs record traces of concrete execution paths through the program. Those traces
+are therefore linear list of operations, which are optimized and then get turned
+into machine code. To be able to do this recording, VMs with a tracing JIT
+typically also contain an interpreter. After a user program is started the
+interpreter is used until the most important paths through the user program are
+turned into machine code.
+
+Because the traces always correspond to a concrete execution they cannot contain
+any control flow splits. Therefore they encode the control flow decisions needed
+to stay on the trace with the help of \emph{guards}. Those are operations that
+check that the assumptions are still true when the trace is later executed with different values.
+
+One disadvantage of tracing JITs which makes them not directly applicable to Pypy,
 
@@
 in the user program.
+
+XXX trace makes the object model operations explicit and transparent to the
+optimizer
 

### 2011-03-24 talk/icooolps2011/paper.tex
expand intro

@@
 
-long dream of PE to not have to implement a compiler
-new life by meta-tracers, spur, pypy, other things
-trace execution of object model and get semantics right, not have to compile
-object model
+It has long been an objective of the partial evaluation community to
+automatically produce compilers from interpreters. There has been a recent
+renaissance of this idea using the different technique of tracing just-in-time
+compilers. A number of projects have attempted this approach. SPUR \cite{XXX}
+is a tracing JIT for .NET together with a JavaScript implementation in C\#. XXX
 
-in spur and pypy, this can be improved by the interpreter author by adding
-hints, which influence tracer and optimizer
+All these projects have in common that they trace an implementation language
+which is then used to implement an object model of a dynamic language. The
+tracer then traces through this object model, which makes the object model
+transparent to the tracer and its optimizations. Therefore the semantics of the
+dynamic language does not have to be replicated in the JIT. We call this
+approach \emph{meta-tracing}. We will give an introduction to the PyPy project
+and to meta-tracing in Section~\ref{sec:Background}.
 
-pypy's hints go further than spurs, in this paper we present two very important
-ones and show how classical implementation techniques of dynlangs can be
-expressed by them
+Another commonality of all these approaches is that they require some
+annotations (or hints) in the dynamic language implementation to guide the
+meta-tracer. SPUR and PyPy both provide the interpreter author with more
+elaborate hints to influence the meta-tracer and its optimizer.
+
+PyPy's hints go even further than SPUR's in that they provide the interpreter
+author with a flexible toolset to make her implementation extremely efficient.
+In this paper we present the two most prominent ones and show how classical
+implementation techniques of dynamic languages can be expressed with them.
+
+The contributions of this paper are:
+\begin{itemize}
+ \item A hint to introduce arbitrary constants into the trace.
+ \item A way to define new pure operations which the optimizer then recognizes.
+ \item A worked-out example of a simple object model of a dynamic language and how it can be improved using these hints.
+\end{itemize}
 
@@
 A recently popular approach to JIT compilers is that of tracing JITs. Tracing
-JITs record traces of concrete execution paths through the program. Those traces
-are therefore linear list of operations, which are optimized and then get turned
-into machine code. To be able to do this recording, VMs with a tracing JIT
-typically also contain an interpreter. After a user program is started the
-interpreter is used until the most important paths through the user program are
-turned into machine code.
+JITs record traces of concrete execution paths through the program. Those
+traces are therefore linear list of operations, which are optimized and then
+get turned into machine code. To be able to do this recording, VMs with a
+tracing JIT typically also contain an interpreter. After a user program is
+started the interpreter is used until the most important paths through the user
+program are turned into machine code.
 
-Because the traces always correspond to a concrete execution they cannot contain
-any control flow splits. Therefore they encode the control flow decisions needed
-to stay on the trace with the help of \emph{guards}. Those are operations that
-check that the assumptions are still true when the trace is later executed with different values.
+Because the traces always correspond to a concrete execution they cannot
+contain any control flow splits. Therefore they encode the control flow
+decisions needed to stay on the trace with the help of \emph{guards}. Those are
+operations that check that the assumptions are still true when the trace is
+later executed with different values.
 
-One disadvantage of tracing JITs which makes them not directly applicable to Pypy,
-
-PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. Since we want to re-use
-our tracer for a variety of languages, we
-don't trace the execution of the user program, but instead trace the execution
-of the \emph{interpreter} that is running the program. This means that the traces
-don't contain the bytecodes of the language in question, but RPython-level
-operations that the interpreter did to execute the program.
+One disadvantage of tracing JITs which makes them not directly applicable to
+PyPy is that they encode the language semantics. Since PyPy wants to be a
+general framework, we want to reuse our tracer for different languages.
+Therefore PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. It does not
+trace the execution of the user program, but instead traces the execution of
+the \emph{interpreter} that is running the program. This means that the traces
+it produces don't contain the bytecodes of the language in question, but
+RPython-level operations that the interpreter did to execute the program.
 

### 2011-03-25 talk/icooolps2011/paper.tex
flesh out the introduction

@@
 renaissance of this idea using the different technique of tracing just-in-time
-compilers. A number of projects have attempted this approach. SPUR \cite{XXX}
-is a tracing JIT for .NET together with a JavaScript implementation in C\#. XXX
+compilers. A number of projects have attempted this approach. SPUR \cite{XXX} is
+a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
+\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython (a restricted
+subset of Python). This JIT is then used to trace a number of languages
+implementations written in RPython. A number of other experiments in this
+directions were done, such as an interpreter for Lua in JavaScript, which is run
+on and optimized with a tracing JIT for JavaScript
+\cite{yermolovich_optimization_2009}.
 
-All these projects have in common that they trace an implementation language
-which is then used to implement an object model of a dynamic language. The
-tracer then traces through this object model, which makes the object model
-transparent to the tracer and its optimizations. Therefore the semantics of the
-dynamic language does not have to be replicated in the JIT. We call this
-approach \emph{meta-tracing}. We will give an introduction to the PyPy project
-and to meta-tracing in Section~\ref{sec:Background}.
+These projects have in common that they implement a dynamic language in some
+implementation language. In addition they build a tracing JIT for that implementation
+language. The tracing JIT then traces through the object model of the dynamic
+language implementation. This makes the object model transparent to the tracer
+and its optimizations. Therefore the semantics of the dynamic language does not
+have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
+Another commonality of these approaches is that they allow some annotations (or
+hints) in the dynamic language implementation to guide the meta-tracer. This
+makes the process not completely automatic but can give good speedups over
+bare meta-tracing.
 
-Another commonality of all these approaches is that they require some
-annotations (or hints) in the dynamic language implementation to guide the
-meta-tracer. SPUR and PyPy both provide the interpreter author with more
-elaborate hints to influence the meta-tracer and its optimizer.
+In this paper we present two of these hints that are extensively used in the
+PyPy project to improve the performance of its Python interpreter.
+ The
 
@@
 In this paper we present the two most prominent ones and show how classical
-implementation techniques of dynamic languages can be expressed with them.
+implementation techniques of dynamic languages can be expressed with them. These
+hints are used to control how the optimizer of the tracing JIT can improve the
+traces of the object model. More specifically, these hints influence the
+constant folding optimization. The first hint make it possible to turn arbitrary
+variables in the trace into constants. The second hint allows the definition of
+additional foldable operations.
+
+Together these two hints can be used to express many classic implementation
+techniques used for object models of dynamic languages, such as maps and
+polymorphic inline caches.
 
@@
 \begin{itemize}
- \item A hint to introduce arbitrary constants into the trace.
- \item A way to define new pure operations which the optimizer then recognizes.
- \item A worked-out example of a simple object model of a dynamic language and how it can be improved using these hints.
+ \item A hint to turn arbitrary variables into constants in the trace.
+ \item A way to define new pure operations which the constant folding
+ optimization then recognizes.
+ \item A worked-out example of a simple object model of a dynamic language and
+ how it can be improved using these hints.
 \end{itemize}
+
+The paper is structured as follows: Section~\ref{sec:background} gives an
+introduction to the PyPy project and meta-tracing and presents an example of a
+tiny dynamic language object model. Section~\ref{sec:hints} presents the hints,
+what they do and how they are applied. Section~\ref{sec:fastobjmodel} shows how
+the hints are applied to the tiny object model and Section~\ref{sec:evaluation}
+presents benchmarks.
 
@@
 
-\section{Controlling Optimization}
+\section{Hints for Controlling Optimization}
+\label{sec:hints}
 
@@
 
-In this section we presented two more hints that can be used in the source code
+In this section we presented two hints that can be used in the source code
 of the interpreter. They are used to influence what the optimizer does with the
@@
 \section{Putting It All Together}
+\label{sec:fastobjmodel}
 
@@
 \section{Evaluation}
-\label{sect:evaluation}
+\label{sec:evaluation}
 

### 2011-03-25 talk/icooolps2011/paper.tex
improve intro and background

@@
 The PyPy project \cite{armin_rigo_pypys_2006} strives to be an environment where
-complex dynamic languages can be efficiently implemented. The approach taken
-when implement a language with PyPy is to write an interpreter for the language
+complex dynamic languages can be implemented efficiently. The approach taken
+when implementing a language with PyPy is to write an interpreter for the language
 in \emph{RPython}. RPython is a restricted subset of Python chosen in such a way
@@
 A number of languages have been implemented with PyPy, most importantly a full
-Python implementation, but also a Prolog interpreter \cite{XXX} and a Smalltalk
-VM \cite{XXX}.
+Python implementation, but also a Prolog interpreter
+\cite{carl_friedrich_bolz_towards_2010} and a Smalltalk VM
+\cite{carl_friedrich_bolz_back_2008}.
 
-This translation to C code adds a number of implementation details into the
+The translation of the interpreter to C code adds a number of implementation details into the
 final executable that are not present in the interpreter implementation, such as
@@
 
-XXX citations
 A recently popular approach to JIT compilers is that of tracing JITs. Tracing
-JITs record traces of concrete execution paths through the program. Those
+JITs have their origin in the Dynamo project which used the for dynamic
+assembler optimization \cite{XXX}. Later they were used for to implement
+a lightweight JIT for Java \cite{XXX} and for dynamic languages such as
+JavaScript \cite{XXX}.
+
+A tracing JIT works by recording traces of concrete execution paths through the
+program. Those
 traces are therefore linear list of operations, which are optimized and then
 get turned into machine code. To be able to do this recording, VMs with a
-tracing JIT typically also contain an interpreter. After a user program is
+tracing JIT typically contain an interpreter. After a user program is
 started the interpreter is used until the most important paths through the user
-program are turned into machine code.
+program are turned into machine code. The tracing JIT tries to produce traces
+that correspond to loops in the traced program, but most tracing JITs now also
+have support for tracing non-loops \cite{XXX}.
 
@@
 One disadvantage of tracing JITs which makes them not directly applicable to
-PyPy is that they encode the language semantics. Since PyPy wants to be a
+PyPy is that they need to encode the language semantics of the language they are
+tracing. Since PyPy wants to be a
 general framework, we want to reuse our tracer for different languages.
@@
 
-On the other hand, the loops that are traced by the tracer are the loops in the
+Tracing through the execution of an interpreter has many advantages. It makes
+the tracer, its optimizers and backends reusable for a variety of languages. The
+language semantics do not need to be encoded into the JIT. Instead the tracer
+just picks them up from the interpreter. XXX mention disadvantage of long
+traces?
+
+While the operations in a trace are those of the interpreter, the loops that are
+traced by the tracer are the loops in the
 user program. This means that the tracer stops tracing after one iteration of
@@
 in the user program.
-
-XXX trace makes the object model operations explicit and transparent to the
-optimizer
 

### 2011-03-25 talk/icooolps2011/paper.tex
more reference, a sentence about inlining

@@
 renaissance of this idea using the different technique of tracing just-in-time
-compilers. A number of projects have attempted this approach. SPUR \cite{XXX} is
+compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
 a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
@@
 JITs have their origin in the Dynamo project which used the for dynamic
-assembler optimization \cite{XXX}. Later they were used for to implement
-a lightweight JIT for Java \cite{XXX} and for dynamic languages such as
-JavaScript \cite{XXX}.
+assembler optimization \cite{bala_dynamo:_2000}. Later they were used for to implement
+a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
+JavaScript \cite{gal_trace-based_2009}.
 
@@
 traces are therefore linear list of operations, which are optimized and then
-get turned into machine code. To be able to do this recording, VMs with a
+get turned into machine code. This recording automatically inlines functions,
+when a function call is encountered the operations of the called functions are
+simply put into the trace too.
+
+To be able to do this recording, VMs with a
 tracing JIT typically contain an interpreter. After a user program is

### 2011-03-25 talk/icooolps2011/code/map.tex
use listings instead of pygmentize

@@
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+class Map(object):
+    def __init__(self):
+        self.indexes = {}
+        self.other_maps = {}
 
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
+    @purefunction
+    def getindex(self, name):
+        return self.indexes.get(name, -1)
 
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{if} \PY{n}{name} \PY{o+ow}{not} \PY{o+ow}{in} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{:}
-            \PY{n}{newmap} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
-            \PY{n}{newmap}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{update}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{p}{)}
-            \PY{n}{newmap}\PY{o}{.}\PY{n}{indexes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n+nb}{len}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{p}{)}
-            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{newmap}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]}
+    @purefunction
+    def add_attribute(self, name):
+        if name not in self.other_maps:
+            newmap = Map()
+            newmap.indexes.update(self.indexes)
+            newmap.indexes[name] = len(self.indexes)
+            self.other_maps[name] = newmap
+        return self.other_maps[name]
 
-\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
+EMPTY_MAP = Map()
 
-\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}
+class Instance(object):
+    def __init__(self, cls):
+        self.cls = cls
+        self.map = EMPTY_MAP
+        self.storage = []
 
-    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+    def getfield(self, name):
+        map = hint(self.map, promote=True)
+        index = map.getindex(name)
+        if index != -1:
+            return self.storage[index]
+        raise AttributeError(name)
 
-    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
-            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-            \PY{k}{return}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{add\PYZus{}attribute}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{n}{value}\PY{p}{)}
+    def write_attribute(self, name, value):
+        map = hint(self.map, promote=True)
+        index = map.getindex(name)
+        if index != -1:
+            self.storage[index] = value
+            return
+        self.map = map.add_attribute(name)
+        self.storage.append(value)
 
-    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{try}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-\end{Verbatim}
+    def getattr(self, name):
+        try:
+            return self.getfield(name)
+        except AttributeError:
+            return self.cls.find_method(name)
+\end{lstlisting}

### 2011-03-25 talk/icooolps2011/paper.tex
use listings instead of pygmentize

@@
 \usepackage[utf8]{inputenc}
+\usepackage{setspace}
+\usepackage{listings}
 
 \input{code/style.tex}
+
+\lstset{
+  basicstyle=\ttfamily\footnotesize,
+  language=Python,
+  keywordstyle=\bfseries,
+  stringstyle=\color{blue},
+  commentstyle=\color{gray}\textit,
+  fancyvrb=true,
+  showstringspaces=false,
+  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
+  numberstyle = \tiny,
+  numbersep = -20pt,
+}
 
@@
 
-\begin{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
 inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
-\end{Verbatim}
+\end{lstlisting}
 
@@
 following fragment of RPython code:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{if} \PY{n}{x} \PY{o}{==} \PY{l+m+mi}{4}\PY{p}{:}
-    \PY{n}{y} \PY{o}{=} \PY{n}{y} \PY{o}{+} \PY{n}{x}
-\end{Verbatim}
+
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+if x == 4:
+    y = y + x
+\end{lstlisting}
 
@@
 %
-\begin{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
 guard(x == 4)
 y = y + x
-\end{Verbatim}
+\end{lstlisting}
 
@@
 Let's make this more concrete. If we trace a call to the following function:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{def} \PY{n+nf}{f1}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
-    \PY{n}{z} \PY{o}{=} \PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
-    \PY{k}{return} \PY{n}{z} \PY{o}{+} \PY{n}{y}
-\end{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+def f2(x, y):
+    x = hint(x, promote=True)
+    z = x * 2 + 1
+    return z + y
+\end{lstlisting}
 
@@
 
-\begin{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
 v1 = x * 2
@@
 return(v2)
-\end{Verbatim}
+\end{lstlisting}
 
@@
 case, we can add a hint to promote \texttt{x}, like this:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{def} \PY{n+nf}{f2}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
-    \PY{n}{x} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-    \PY{n}{z} \PY{o}{=} \PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
-    \PY{k}{return} \PY{n}{z} \PY{o}{+} \PY{n}{y}
-\end{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+def f2(x, y):
+    x = hint(x, promote=True)
+    z = x * 2 + 1
+    return z + y
+\end{lstlisting}
 
@@
 
-\begin{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
 guard(x == 4)
@@
 
-\begin{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
 guard(x == 2)
@@
 
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{A}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{=} \PY{n}{x}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n}{y}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+class A(object):
+    def __init__(self, x, y):
+        self.x = x
+        self.y = y
 
-    \PY{k}{def} \PY{n+nf}{f}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{val}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{compute}\PY{p}{(}\PY{p}{)} \PY{o}{+} \PY{n}{val}
+    def f(self, val):
+        self.y = self.compute() + val
 
-    \PY{k}{def} \PY{n+nf}{compute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
-\end{Verbatim}
+    def compute(self):
+        return self.x * 2 + 1
+\end{lstlisting}
 
@@
 %
-\begin{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
 x = a.x
@@
 a.y = v3
-\end{Verbatim}
+\end{lstlisting}
 
@@
 class as follows:
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{A}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{=} \PY{n}{x}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n}{y}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+class A(object):
+    def __init__(self, x, y):
+        self.x = x
+        self.y = y
 
-    \PY{k}{def} \PY{n+nf}{f}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{val}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{compute}\PY{p}{(}\PY{p}{)} \PY{o}{+} \PY{n}{val}
+    def f(self, val):
+        self = hint(self, promote=True)
+        self.y = self.compute() + val
 
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{compute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
-\end{Verbatim}
+    @purefunction
+    def compute(self):
+        return self.x * 2 + 1
+\end{lstlisting}
 
@@
 %
-\begin{Verbatim}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
 guard(a == 0xb73984a8)
@@
 a.y = v2
-\end{Verbatim}
+\end{lstlisting}
 

### 2011-03-25 talk/icooolps2011/code/interpreter-slow.tex
use listings instead of pygmentize

@@
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+class Class(object):
+    def __init__(self, name):
+        self.name = name
+        self.methods = {}
 
-    \PY{k}{def} \PY{n+nf}{instantiate}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n}{Instance}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}
+    def instantiate(self):
+        return Instance(self)
 
-    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
-            \PY{k}{return} \PY{n}{result}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+    def find_method(self, name):
+        result = self.methods.get(name)
+        if result is not None:
+            return result
+        raise AttributeError(name)
 
-    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
+    def change_method(self, name, value):
+        self.methods[name] = value
 
-\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
+class Instance(object):
+    def __init__(self, cls):
+        self.cls = cls
+        self.attributes = {}
 
-    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
-            \PY{k}{return} \PY{n}{result}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+    def getfield(self, name):
+        result = self.attributes.get(name)
+        if result is not None:
+            return result
+        raise AttributeError(name)
 
-    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{attributes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
+    def write_attribute(self, name, value):
+        self.attributes[name] = value
 
-    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{try}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-\end{Verbatim}
+    def getattr(self, name):
+        try:
+            return self.getfield(name)
+        except AttributeError:
+            return self.cls.find_method(name)
+\end{lstlisting}

### 2011-03-25 talk/icooolps2011/code/version.tex
use listings instead of pygmentize

@@
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{VersionTag}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{pass}
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+class VersionTag(object):
+    pass
 
-\PY{k}{class} \PY{n+nc}{Class}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{name} \PY{o}{=} \PY{n}{name}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}
+class Class(object):
+    def __init__(self, name):
+        self.name = name
+        self.methods = {}
+        self.version = VersionTag()
 
-    \PY{k}{def} \PY{n+nf}{find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{version} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
-        \PY{n}{result} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}
-        \PY{k}{if} \PY{n}{result} \PY{o+ow}{is} \PY{o+ow}{not} \PY{n+nb+bp}{None}\PY{p}{:}
-            \PY{k}{return} \PY{n}{result}
-        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+    def find_method(self, name):
+        self = hint(self, promote=True)
+        version = hint(self.version, promote=True)
+        result = self._find_method(name, version)
+        if result is not None:
+            return result
+        raise AttributeError(name)
 
-    \PY{n+nd}{@purefunction}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}find\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{version}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{)}
+    @purefunction
+    def _find_method(self, name, version):
+        return self.methods.get(name)
 
-    \PY{k}{def} \PY{n+nf}{change\PYZus{}method}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{methods}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{value}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{version} \PY{o}{=} \PY{n}{VersionTag}\PY{p}{(}\PY{p}{)}
-\end{Verbatim}
+    def change_method(self, name, value):
+        self.methods[name] = value
+        self.version = VersionTag()
+\end{lstlisting}

### 2011-03-25 talk/icooolps2011/paper.tex
migrate to newer sigplan cls

@@
-%\documentclass{acm_proc_article-sp}
-\documentclass{sig-alternate}
+\documentclass{sigplanconf}
 
@@
 \usepackage{xspace}
-\usepackage[scaled=0.8]{beramono}
+\usepackage{epsfig}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsfonts}
 \usepackage[utf8]{inputenc}
 \usepackage{setspace}
+
 \usepackage{listings}
 
-\input{code/style.tex}
+\usepackage[T1]{fontenc}
+\usepackage[scaled=0.8]{beramono}
+
+\definecolor{commentgray}{rgb}{0.3,0.3,0.3}
 
@@
   stringstyle=\color{blue},
-  commentstyle=\color{gray}\textit,
+  commentstyle=\color{commentgray}\textit,
   fancyvrb=true,
@@
 
-\numberofauthors{4}
-\author{
-\alignauthor Carl Friedrich Bolz\\
-       \affaddr{University of Düsseldorf}\\
-       \affaddr{STUPS Group}\\
-       \affaddr{Germany}\\
-       \email{cfbolz@gmx.de}
-\alignauthor XXX
-       \affaddr{XXX}\\
-       \email{XXX}
+\authorinfo{Carl Friedrich Bolz \and XXX}
+           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
            }
-\conferenceinfo{ICOOOLPS}{'09 Genova, Italy}
-\CopyrightYear{2009}
-\crdata{978-1-60558-541-3/09/07}
+           {cfbolz@gmx.de \and XXX}
+
+\conferenceinfo{ICOOOLPS}{'11 Lancaster, UK}
+\CopyrightYear{2011}
+\crdata{XXX}
 

### 2011-03-25 talk/icooolps2011/code/interpreter-slow.tex
- simplify the code a bit - delete the .py files now that the tex files are not generated any more

@@
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily,numbers = right]
 class Class(object):
@@
     def find_method(self, name):
-        result = self.methods.get(name)
-        if result is not None:
-            return result
-        raise AttributeError(name)
+        return self.methods.get(name, None)
 
@@
     def getfield(self, name):
-        result = self.attributes.get(name)
-        if result is not None:
-            return result
-        raise AttributeError(name)
+        return self.attributes.get(name, None)
 
@@
     def getattr(self, name):
-        try:
-            return self.getfield(name)
-        except AttributeError:
-            return self.cls.find_method(name)
+	result = self.getfield(name)
+	if result is None:
+            result = self.cls.find_method(name)
+	    if result is None:
+		raise AttributeError
 \end{lstlisting}

### 2011-03-25 talk/icooolps2011/code/map.tex
- simplify the code a bit - delete the .py files now that the tex files are not generated any more

@@
             return self.storage[index]
-        raise AttributeError(name)
+        return None
 
@@
     def getattr(self, name):
-        try:
-            return self.getfield(name)
-        except AttributeError:
-            return self.cls.find_method(name)
+	... # as before
 \end{lstlisting}

### 2011-03-25 talk/icooolps2011/code/version.tex
- simplify the code a bit - delete the .py files now that the tex files are not generated any more

@@
         version = hint(self.version, promote=True)
-        result = self._find_method(name, version)
-        if result is not None:
-            return result
-        raise AttributeError(name)
+        return self._find_method(name, version)
 

### 2011-03-25 talk/icooolps2011/code/interpreter-slow.py

@@ -1,37 +1,0 @@
-class Class(object):
-    def __init__(self, name):
-        self.name = name
-        self.methods = {}
-
-    def instantiate(self):
-        return Instance(self)
-
-    def find_method(self, name):
-        result = self.methods.get(name)
-        if result is not None:
-            return result
-        raise AttributeError(name)
-
-    def change_method(self, name, value):
-        self.methods[name] = value
-
-
-class Instance(object):
-    def __init__(self, cls):
-        self.cls = cls
-        self.attributes = {}
-
-    def getfield(self, name):
-        result = self.attributes.get(name)
-        if result is not None:
-            return result
-        raise AttributeError(name)
-
-    def write_attribute(self, name, value):
-        self.attributes[name] = value
-
-    def getattr(self, name):
-        try:
-            return self.getfield(name)
-        except AttributeError:
-            return self.cls.find_method(name)

### 2011-03-25 talk/icooolps2011/paper.tex
add line numbers to the trace about where the operations originate from
move conclusion bits around

@@
 attribute \texttt{a} is found on the instance, but the
-attributes \texttt{b} and \texttt{c} are found on the class. The trace indeed contains
+attributes \texttt{b} and \texttt{c} are found on the class. The numbers line
+numbers in the trace correspond to the line numbers in
+Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
+trace indeed contains
 five calls to \texttt{dict.get}, which is slow. To make the language efficient
@@
 
-\subsection{Conclusion}
-
-In this section we presented two hints that can be used in the source code
-of the interpreter. They are used to influence what the optimizer does with the
-trace. The examples given here are a bit too small, the next
-section gives a worked-out example that puts all the pieces together.
-
 %___________________________________________________________________________
@@
 
-\subsection{Conclusion}
-
-In this section we saw how to use \texttt{purefunction} and \texttt{promote} to make a
-small but still relevant dynamic object model no longer use any dictionary lookups
-after tracing. Instead a number of guards are inserted into the
-trace to check whether the assumptions about the objects are still true. This
-makes operations on objects seriously faster.
-
 \section{Evaluation}
@@
 
+In this paper we presented two hints that can be used in the source code of an
+interpreter written with PyPy. They are used to influence what the optimizer
+does with the trace. We also showed how a small but still relevant dynamic
+object model can use these hints to no longer use any dictionary lookups after
+tracing. Instead a number of guards are inserted into the trace to check whether
+the assumptions about the objects are still true. This makes operations on
+objects seriously faster.
+
+XXX
+
 \section*{Acknowledgements}

### 2011-03-25 talk/icooolps2011/paper.tex
fix references

@@
 
-The paper is structured as follows: Section~\ref{sec:background} gives an
+The paper is structured as follows: Section~\ref{sec:Background} gives an
 introduction to the PyPy project and meta-tracing and presents an example of a
@@
 \subsection{The PyPy Project}
-\label{sect:pypy}
+\label{sub:pypy}
 
@@
 \subsection{PyPy's Meta-Tracing JIT Compilers}
-\label{sect:tracing}
+\label{sub:tracing}
 
@@
 lookups somehow. How to achieve this will be topic of
-Section~\ref{sec:putting}.
+Section~\ref{sec:fastobjmodel}.
 

### 2011-03-25 talk/icooolps2011/paper.tex
change diagram to make it fit into one column to save space
add a missing return, change_method -> write_method
remove duplication

@@
 In this paper we present two of these hints that are extensively used in the
-PyPy project to improve the performance of its Python interpreter.
- The
-
-PyPy's hints go even further than SPUR's in that they provide the interpreter
-author with a flexible toolset to make her implementation extremely efficient.
-In this paper we present the two most prominent ones and show how classical
-implementation techniques of dynamic languages can be expressed with them. These
+PyPy project to improve the performance of its Python interpreter. These
 hints are used to control how the optimizer of the tracing JIT can improve the
@@
 
-\begin{figure*}
+\begin{figure}
 \includegraphics[scale=0.5]{figures/trace-levels}
@@
 \label{fig:trace-levels}
-\end{figure*}
+\end{figure}
 
@@
 class did not change since the trace was produced. It will fail if somebody
-calls the \texttt{change\_method} method on the class.
+calls the \texttt{write\_method} method on the class.
 

### 2011-03-26 talk/icooolps2011/paper.tex
two more citations, fix one of the tiny functions

@@
 polymorphic inline caches.
+
+XXX replace "turn variable into constant" into "main tool to feedback language
+and semantics specific runtime information into the compilation"
 
@@
 into a constant value. This process is called \emph{promotion} and it is an old idea
-in partial evaluation (it's called ``the trick'' \cite{XXX} there). Promotion is also heavily
+in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993} there). Promotion is also heavily
 used by Psyco \cite{rigo_representation-based_2004} and by all older versions
@@
 def f2(x, y):
-    x = hint(x, promote=True)
     z = x * 2 + 1
@@
 technique to efficiently implement instances and come from the SELF project
-\cite{XXX}. They are also used by many JavaScript implementations such as V8.
+\cite{chambers_efficient_1989}. They are also used by many JavaScript implementations such as V8.
 The rewritten \texttt{Instance} class using maps can be seen in
@@
 
+\cite{mario_wolczko_towards_1999}
+
 \section{Conclusion and Next Steps}

### 2011-03-26 talk/icooolps2011/paper.tex
some notes about related work out of a discussion with samuele

@@
 
+partial evaluation \cite{futamura_partial_1999} \cite{jones_partial_1993}
+
+earlier examples of promotion \cite{carl_friedrich_bolz_towards_????}, \cite{armin_rigo_jit_2007}
+
+meta-tracers: SPUR \cite{bebenita_spur:_2010}
+tamarin/lua stuff \cite{yermolovich_optimization_2009}
+Dynamo \cite{sullivan_dynamic_2003}
+
+\cite{chambers_efficient_1989} maps
+\cite{hoelzle_optimizing_1994} Type feedback
+\cite{hoelzle_optimizing_1991} PICs
+find the web page of V8 about maps
+
+is there anything about versions? smalltalks tend to clear their method caches
+when new methods are added. self and java use dependency tracking and
+deoptimization. this is better what we have above, because we need runtime
+checks. mention out of line guard?
+
+jruby used versions at some point, jvm-l mailing list discusses them
+
 \cite{mario_wolczko_towards_1999}
+
+invokedynamic tries to give control to the language implementor about PICs and
+deoptimization
 

### 2011-03-26 talk/icooolps2011/paper.tex
use index notation for SSA variables everywhere

@@
 
-If the fragment is traced with \texttt{x} being \texttt{4}, the following trace is
+If the fragment is traced with $x_1$ being \texttt{4}, the following trace is
 produced:
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard(x == 4)
-y = y + x
+guard($x_1$ == 4)
+$y_2$ = $y_1$ + $x_1$
 \end{lstlisting}
 
-In the trace above, the value of \texttt{x} is statically known thanks to the
+In the trace above, the value of $x_1$ is statically known thanks to the
 guard. Remember that a guard is a runtime check. The above trace will run to
-completion when \texttt{x == 4}. If the check fails, execution of the trace is
+completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
 stopped and the interpreter continues to run.
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-def f2(x, y):
+def f1(x, y):
     z = x * 2 + 1
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-v1 = x * 2
-z = v1 + 1
-v2 = z + y
-return(v2)
+$v_1$ = $x_1$ * 2
+$z_1$ = $v_1$ + 1
+$v_2$ = $z_1$ + $y_1$
+return($v_2$)
 \end{lstlisting}
@@
 Observe how the first two operations could be constant-folded if the value of
-\texttt{x} were known. Let's assume that the value of \texttt{x} can vary, but does so
+$x_1$ were known. Let's assume that the value of \texttt{x} in the Python code can vary, but does so
 rarely, i.e. only takes a few different values at runtime. If this is the
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-def f2(x, y):
+def f1(x, y):
     x = hint(x, promote=True)
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard(x == 4)
-v1 = x * 2
-z = v1 + 1
-v2 = z + y
+guard($x_1$ == 4)
+$v_1$ = $x_1$ * 2
+$z_1$ = $v_1$ + 1
+$v_2$ = $z_1$ + $y_1$
 return(v2)
@@
 The promotion is turned into a \texttt{guard} operation in the trace. The guard
-captures the value of \texttt{x} as it was at runtime. From the point of view of the
+captures the value of $x_1$ as it was at runtime. From the point of view of the
 optimizer, this guard is not any different than the one produced by the \texttt{if}
 statement in the example above. After the guard, the rest of the trace can
-assume that \texttt{x} is equal to \texttt{4}, meaning that the optimizer will turn this
+assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
 trace into:
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard(x == 4)
-v2 = 9 + y
-return(v2)
+guard($x_1$ == 4)
+$v_2$ = 9 + $y_1$
+return($v_2$)
 \end{lstlisting}
@@
 
-If this trace is executed with values of \texttt{x} other than \texttt{4}, the guard will
+If this trace is executed with values of $x_1$ other than \texttt{4}, the guard will
 fail, and execution will continue in the interpreter. If the guard fails often
 enough, a new trace will be started from the guard. This other trace will
-capture a different value of \texttt{x}. If it is e.g. \texttt{2}, then the optimized
+capture a different value of $x_1$. If it is e.g. \texttt{2}, then the optimized
 trace looks like this:
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard(x == 2)
-v2 = 5 + y
-return(v2)
+guard($x_1$ == 2)
+$v_2$ = 5 + $y_1$
+return($v_2$)
 \end{lstlisting}
@@
 This new trace will be attached to the guard instruction of the first trace. If
-\texttt{x} takes on even more values, a new trace will eventually be made for all of them,
+$x_1$ takes on even more values, a new trace will eventually be made for all of them,
 linking them into a chain. This is clearly not desirable, so we should promote
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-x = a.x
-v1 = x * 2
-v2 = v1 + 1
-v3 = v2 + val
-a.y = v3
+$x_1$ = $a_1$.x
+$v_1$ = $x_1$ * 2
+$v_2$ = $v_1$ + 1
+$v_3$ = $v_2$ + $val_1$
+$a_1$.y = $v_3$
 \end{lstlisting}
@@
 In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
-computation of the first few operations does not help. Even if \texttt{a} is a
+computation of the first few operations does not help. Even if $a_1$ is a
 constant reference to an object, reading the \texttt{x} field does not necessarily
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard(a == 0xb73984a8)
-v1 = compute(a)
-v2 = v1 + val
-a.y = v2
+guard($a_1$ == 0xb73984a8)
+$v_1$ = compute($a_1$)
+$v_2$ = $v_1$ + $val_1$
+$a_1$.y = $v_2$
 \end{lstlisting}
@@
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard(a == 0xb73984a8)
-v2 = 9 + val
-a.y = v2
+guard($a_1$ == 0xb73984a8)
+$v_2$ = 9 + $val_1$
+$a_1$.y = $v_2$
 \end{lstlisting}

### 2011-03-27 talk/icooolps2011/paper.tex
vague first draft of some of the related work parts

@@
 \usepackage{xspace}
+\usepackage{relsize}
 \usepackage{epsfig}
@@
 
-partial evaluation \cite{futamura_partial_1999} \cite{jones_partial_1993}
+The very first meta-tracer is described by Sullivan et. al.
+\cite{sullivan_dynamic_2003}. They used Dynamo RIO, the successor of Dynamo
+\cite{XXX} to trace through a small synthetic interpreter. As in Dynamo, tracing
+happens on the machine code level. The tracer is instructed by some hints in the
+tiny interpreter where the main interpreter loop is and for how long to trace to
+match loops in the user-level functions. These hints are comparable to the one
+PyPy uses for the same reasons \cite{bolz_tracing_2009}. Their approach suffers
+mostly from the low abstraction level that machine code provides.
 
-earlier examples of promotion \cite{carl_friedrich_bolz_towards_????}, \cite{armin_rigo_jit_2007}
+Yermolovich et. al. describe the use of the Tamarin JavaScript tracing JIT as a
+meta-tracer for a Lua interpreter. They compile the normal Lua interpreter in C
+to ActionScript bytecode. Again, the interpreter is annotated with some hints
+that indicate the main interpreter loop to the tracer.  No further hints are
+described in the paper. There is no comparison of their system to the original
+Lua VM in C, which makes it hard to judge the effectiveness of the approach.
 
-meta-tracers: SPUR \cite{bebenita_spur:_2010}
-tamarin/lua stuff \cite{yermolovich_optimization_2009}
-Dynamo \cite{sullivan_dynamic_2003}
+SPUR \cite{bebenita_spur:_2010} is a tracing JIT for CIL bytecode, which is then
+used to trace through an JavaScript implementation written in C\#. The
+JavaScript implementation compiles JavaScript to CIL bytecode together with an
+implementation of the JavaScript object model. The object model uses maps
+and inline caches to speed up operations on objects. The tracer tracers through
+the compiled JavaScript functions and the object model. SPUR contains two hints
+that can be used to influence the tracer, one to prevent tracing of a C\#
+function and one to force unrolling of a loop (PyPy has equivalent hints, but
+they were not described in this paper).
+
+Partial evaluation \cite{jones_partial_1993} tries to automatically transform
+interpreters into compilers using the second futamura projection
+\cite{futamura_partial_1999}. XXX
+
+An early attempt at building a general environment for implementing languages
+efficiently is described by Wolczko et. al. \cite{mario_wolczko_towards_1999}.
+They implement Java and Smalltalk on top of the SELF VM by compiling the
+languages to SELF. The SELF JIT is good enough to optimize the compiled code
+very well. We believe the approach to be restricted to languages that are
+similar enough to SELF. XXX
+
+Somewhat relatedly, the proposed ``invokedynamic'' bytecode
+\cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
+implementation of dynamic languages on top of JVMs easier. The bytecode gives
+the language implementor control over how the JIT optimizes the language's
+features. XXX
+
+%We already explored promotion in other context, such as earlier examples of
+%promotion \cite{carl_friedrich_bolz_towards_????}, \cite{armin_rigo_jit_2007}
 
@@
 jruby used versions at some point, jvm-l mailing list discusses them
-
-\cite{mario_wolczko_towards_1999}
-
-invokedynamic tries to give control to the language implementor about PICs and
-deoptimization
 

### 2011-03-27 talk/icooolps2011/paper.tex
Typos and typographic details.

@@
 traces of the object model. More specifically, these hints influence the
-constant folding optimization. The first hint make it possible to turn arbitrary
+constant folding optimization. The first hint makes it possible to turn arbitrary
 variables in the trace into constants. The second hint allows the definition of
@@
 A recently popular approach to JIT compilers is that of tracing JITs. Tracing
-JITs have their origin in the Dynamo project which used the for dynamic
-assembler optimization \cite{bala_dynamo:_2000}. Later they were used for to implement
+JITs have their origin in the Dynamo project, which used one of them for dynamic
+assembler optimization \cite{bala_dynamo:_2000}. Later they were used to implement
 a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
@@
 traces are therefore linear list of operations, which are optimized and then
-get turned into machine code. This recording automatically inlines functions,
+get turned into machine code. This recording automatically inlines functions:
 when a function call is encountered the operations of the called functions are
@@
 tracing JIT typically contain an interpreter. After a user program is
-started the interpreter is used until the most important paths through the user
+started the interpreter is used; only the most frequently executed paths through the user
 program are turned into machine code. The tracing JIT tries to produce traces
@@
 
-One disadvantage of tracing JITs which makes them not directly applicable to
+One disadvantage of (tracing) JITs which makes them not directly applicable to
 PyPy is that they need to encode the language semantics of the language they are
@@
 attribute \texttt{a} is found on the instance, but the
-attributes \texttt{b} and \texttt{c} are found on the class. The numbers line
+attributes \texttt{b} and \texttt{c} are found on the class. The line
 numbers in the trace correspond to the line numbers in
@@
 used by Psyco \cite{rigo_representation-based_2004} and by all older versions
-of PyPy's JIT. Promotion is a technique that only works well in JIT compilers,
+of PyPy's JIT. Promotion is a technique that only works well in JIT compilers;
 in static compilers it is significantly less applicable.
@@
 though in principle the argument to a Python function could be any Python type,
-in practice the argument types tend to not vary often. Therefore it is possible to
+in practice the argument types tend not to vary often. Therefore it is possible to
 promote the types. The next section will present a complete example of how
@@
 
-In the last section we saw a way to turn arbitrary variables into constants. All
+In the previous section we saw a way to turn arbitrary variables into constants. All
 pure operations on these constants can be constant-folded. This works great for
@@
 during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
-has a chance to see it. Since \texttt{compute} function is marked as pure, and its
+has a chance to see it. Since the \texttt{compute} function is marked as pure, and its
 argument
@@
 
+XXX should we say ``version number'' all around when it is really an object?
+
 Therefore, we give every class a version number, which is changed every time a
@@
 The constants \texttt{41} and \texttt{17} are the results of the folding of the
-\texttt{\_find\_method`} calls. This final trace is now very good. It no longer performs any
+\texttt{\_find\_method} calls. This final trace is now very good. It no longer performs any
 dictionary lookups. Instead it contains several guards. The first guard
@@
 implementation of the JavaScript object model. The object model uses maps
-and inline caches to speed up operations on objects. The tracer tracers through
+and inline caches to speed up operations on objects. The tracer traces through
 the compiled JavaScript functions and the object model. SPUR contains two hints
-that can be used to influence the tracer, one to prevent tracing of a C\#
+that can be used to influence the tracer: one to prevent tracing of a C\#
 function and one to force unrolling of a loop (PyPy has equivalent hints, but

### 2011-03-27 talk/icooolps2011/code/interpreter-slow.tex
Typos and typographic details.

@@
 
-    def change_method(self, name, value):
+    def write_method(self, name, value):
         self.methods[name] = value
@@
 		raise AttributeError
+	return result
 \end{lstlisting}

### 2011-03-27 talk/icooolps2011/paper.tex
some changes, some comments

--- 2011-03-27 talk/icooolps2011/paper.tex
@@
 \newcommand\fijal[1]{\nb{FIJAL}{#1}}
+\newcommand\pedronis[1]{\nb{PEDRONIS}{#1}}
 \newcommand{\commentout}[1]{}
@@
 semantics. For them, implementing just an interpreter is already an arduous
-task. Implementing them efficiently with a just-in-time compiler is
-nigh-impossible, because or their many corner-cases.
+\pedronis{XXX ardous seems a bit too strong} task. Implementing them efficiently with a just-in-time compiler (JIT) is
+extremely challenging, because or their many corner-cases.
 
@@
 automatically produce compilers from interpreters. There has been a recent
-renaissance of this idea using the different technique of tracing just-in-time
+renaissance of this idea around the approach of tracing just-in-time
 compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
@@
 
-These projects have in common that they implement a dynamic language in some
-implementation language. In addition they build a tracing JIT for that implementation
-language. The tracing JIT then traces through the object model of the dynamic
+These projects have in common that they work one meta-level down, providing a tracing JIT for the implementation
+language used to implement the dynamic language, and not for the dynamic language itself.
+The tracing JIT then will trace through the object model of the dynamic
 language implementation. This makes the object model transparent to the tracer
@@
 A recently popular approach to JIT compilers is that of tracing JITs. Tracing
-JITs have their origin in the Dynamo project which used the for dynamic
+JITs have their origin in the Dynamo project which used them for dynamic
 assembler optimization \cite{bala_dynamo:_2000}. Later they were used for to implement
@@
 decisions needed to stay on the trace with the help of \emph{guards}. Those are
-operations that check that the assumptions are still true when the trace is
-later executed with different values.
+operations that check that the assumptions are still true when the compiled trace is later executed with different values.
 
@@
 Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
-see the levels of execution. The CPU executes the binary of
+\pedronis{XXX style: do we want to use you, or we and one} see the levels of execution. The CPU executes the binary of
 PyPy's Python interpreter, which consists of RPython functions that have been
@@
 
-Therefore, we give every class a version number, which is changed every time a
+Therefore, we give every class a version object, which is changed every time a
 class gets changed (i.e., the content of the \texttt{methods} dictionary changes).

### 2011-03-27 talk/icooolps2011/paper.tex
reworded something and fixed a typo

@@
 languages such as Python, JavaScript or Ruby have rather complex core object
-semantics. For them, implementing just an interpreter is already an arduous
-\pedronis{XXX ardous seems a bit too strong} task. Implementing them efficiently with a just-in-time compiler (JIT) is
-extremely challenging, because or their many corner-cases.
+semantics. For them, implementing just an interpreter is already a complex
+task. Implementing them efficiently with a just-in-time compiler (JIT) is
+extremely challenging, because of their many corner-cases.
 

### 2011-03-27 talk/icooolps2011/paper.tex
give a conceptual level view on the hints

@@
 In this paper we present two of these hints that are extensively used in the
-PyPy project to improve the performance of its Python interpreter. These
-hints are used to control how the optimizer of the tracing JIT can improve the
-traces of the object model. More specifically, these hints influence the
-constant folding optimization. The first hint make it possible to turn arbitrary
-variables in the trace into constants. The second hint allows the definition of
-additional foldable operations.
+PyPy project to improve the performance of its Python interpreter.
+
+Conceptually the significant speed-ups that can be achieved with
+dynamic compilation depends on feeding into compilation and exploiting
+values observed at runtime that are in practice mostly constant (XXX or should we say slow varying), and
+in structuring implementation code and data structures such that more
+of such values are at hand. The hints that we present allow exactly to implement such feedback and exploitation in a meta-tracing context.
+
+Concretely these hints are used to control how the optimizer of the
+tracing JIT can improve the traces of the object model. More
+specifically, these hints influence the constant folding
+optimization. The first hint make it possible to turn arbitrary
+variables in the trace into constant. The
+second hint allows the definition of additional foldable operations.
 
@@
 
-XXX replace "turn variable into constant" into "main tool to feedback language
-and semantics specific runtime information into the compilation"
-
 The contributions of this paper are:
 \begin{itemize}
- \item A hint to turn arbitrary variables into constants in the trace.
+ \item A hint to turn arbitrary variables into constants in the trace, that means the feedback of runtime information into compilation.
  \item A way to define new pure operations which the constant folding

### 2011-03-27 talk/icooolps2011/paper.tex
- a title - shorten a sentence

@@
 
-\title{XXX in a Tracing JIT Compiler for Efficient Dynamic Languages}
+\title{Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages}
 
@@
 Conceptually the significant speed-ups that can be achieved with
-dynamic compilation depends on feeding into compilation and exploiting
-values observed at runtime that are in practice mostly constant (XXX or should we say slow varying), and
-in structuring implementation code and data structures such that more
-of such values are at hand. The hints that we present allow exactly to implement such feedback and exploitation in a meta-tracing context.
+dynamic compilation depend on feeding into compilation and exploiting
+values observed at runtime that are slow-varying in practice. To exploit the
+runtime feedback, the implementation code and data structures need to be
+structured so that many such values are at hand. The hints that we present allow
+exactly to implement such feedback and exploitation in a meta-tracing context.
 
@@
 specifically, these hints influence the constant folding
-optimization. The first hint make it possible to turn arbitrary
-variables in the trace into constant. The
+optimization. The first hint makes it possible to turn arbitrary
+variables in the trace into constant by feeding back runtime values. The
 second hint allows the definition of additional foldable operations.
@@
 \begin{itemize}
- \item A hint to turn arbitrary variables into constants in the trace, that means the feedback of runtime information into compilation.
+ \item A hint to turn arbitrary variables into constants in the trace, that
+ means the feedback of runtime information into compilation.
  \item A way to define new pure operations which the constant folding

### 2011-03-27 talk/icooolps2011/paper.tex

--- 2011-03-27 talk/icooolps2011/paper.tex
@@
 A recently popular approach to JIT compilers is that of tracing JITs. Tracing
-JITs have their origin in the Dynamo project which used them for dynamic
-assembler optimization \cite{bala_dynamo:_2000}. Later they were used for to implement
+JITs have their origin in the Dynamo project, which used one of them for dynamic
+assembler optimization \cite{bala_dynamo:_2000}. Later they were used to implement
 a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
@@
 
+XXX should we say ``version number'' all around when it is really an object?
+
 Therefore, we give every class a version object, which is changed every time a

### 2011-03-27 talk/icooolps2011/paper.tex
Finish renaming 'version number' into 'version object'.

@@
 
-XXX should we say ``version number'' all around when it is really an object?
-
 Therefore, we give every class a version object, which is changed every time a
@@
 argument but it does not use it at all. Its only purpose is to make the call
-pure, because when the version number changes, the result of the call might be
+pure, because when the version object changes, the result of the call might be
 different than the previous one.

### 2011-03-27 talk/icooolps2011/paper.tex
second dynamo ref
few fixes, comment out a section to save space

@@
 
-Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
-\pedronis{XXX style: do we want to use you, or we and one} see the levels of execution. The CPU executes the binary of
+Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left are
+the levels of execution. The CPU executes the binary of
 PyPy's Python interpreter, which consists of RPython functions that have been
@@
 used to constant-fold arbitrary parts of the computation in the interpreter.
-However, the annotation also gives you ample opportunity to mess things up. If a
+However, the annotation also gives the interpreter author ample opportunity to mess things up. If a
 function is annotated to be pure, but is not really, the optimizer can produce
@@
 The easiest example for this is that of a function that uses memoization to
-cache its results. If you analyze this function, it looks like the function has
+cache its results. If this function is analyzed, it looks like the function has
 side effects, because it changes the memoizing dictionary. However, because this side
@@
 
-\subsection{More General Patterns}
-
-The techniques we used above to make instance and class lookups faster are
-applicable in more general cases than the one we developed them for. A more
-abstract view of maps is that of splitting a data-structure into a part that
-changes slowly, and a part that changes quickly. In the concrete example of maps
-we split the original dictionary into the map (the slow-changing part) and the
-storage array (the quick-changing part). All the computation on the
-slow-changing part can be constant-folded during tracing so that only the
-manipulation of the quick-changing part remains.
-
-Similarly, versions can be used to constant-fold arbitrary functions of large data
-structures. The version needs to be updated carefully every time the result of
-this function can change. Therefore this is useful only if the data structure is
-expected to change slowly.
+%\subsection{More General Patterns}
+%
+%The techniques we used above to make instance and class lookups faster are
+%applicable in more general cases than the one we developed them for. A more
+%abstract view of maps is that of splitting a data-structure into a part that
+%changes slowly, and a part that changes quickly. In the concrete example of maps
+%we split the original dictionary into the map (the slow-changing part) and the
+%storage array (the quick-changing part). All the computation on the
+%slow-changing part can be constant-folded during tracing so that only the
+%manipulation of the quick-changing part remains.
+%
+%Similarly, versions can be used to constant-fold arbitrary functions of large data
+%structures. The version needs to be updated carefully every time the result of
+%this function can change. Therefore this is useful only if the data structure is
+%expected to change slowly.
 
@@
 \cite{sullivan_dynamic_2003}. They used Dynamo RIO, the successor of Dynamo
-\cite{XXX} to trace through a small synthetic interpreter. As in Dynamo, tracing
+\cite{bala_dynamo:_2000} to trace through a small synthetic interpreter. As in Dynamo, tracing
 happens on the machine code level. The tracer is instructed by some hints in the
@@
 very well. We believe the approach to be restricted to languages that are
-similar enough to SELF. XXX
+similar enough to SELF as there were no mechanisms to control the underlying
+compiler.
 
@@
 
-is there anything about versions? smalltalks tend to clear their method caches
-when new methods are added. self and java use dependency tracking and
-deoptimization. this is better what we have above, because we need runtime
-checks. mention out of line guard?
+%is there anything about versions? smalltalks tend to clear their method caches
+%when new methods are added. self and java use dependency tracking and
+%deoptimization. this is better what we have above, because we need runtime
+%checks. mention out of line guard?
 

### 2011-03-27 talk/icooolps2011/paper.tex
old author block

@@
 
-\authorinfo{Carl Friedrich Bolz \and XXX}
-           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
+\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^a$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
+            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
+           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
+
+            $^b$merlinux GmbH, Hildesheim, Germany
+
+            $^c$Open End, Göteborg, Sweden
            }
-           {cfbolz@gmx.de \and XXX}
+           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu \and
+           leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}
 

### 2011-03-28 talk/icooolps2011/paper.tex
expand related work some more

@@
 that correspond to loops in the traced program, but most tracing JITs now also
-have support for tracing non-loops \cite{XXX}.
+have support for tracing non-loops \cite{andreas_gal_incremental_2006}.
 
@@
 language semantics do not need to be encoded into the JIT. Instead the tracer
-just picks them up from the interpreter. XXX mention disadvantage of long
-traces?
+just picks them up from the interpreter.
 
@@
 interpreters into compilers using the second futamura projection
-\cite{futamura_partial_1999}. XXX
+\cite{futamura_partial_1999}. Given that classical partial evaluation works
+strictly ahead of time, it inherently cannot support runtime feedback.
 
@@
 the language implementor control over how the JIT optimizes the language's
-features. XXX
+features and when optimized code needs to be deoptimized. XXX
 
-%We already explored promotion in other context, such as earlier examples of
-%promotion \cite{carl_friedrich_bolz_towards_????}, \cite{armin_rigo_jit_2007}
-
-\cite{chambers_efficient_1989} maps
-\cite{hoelzle_optimizing_1994} Type feedback
-\cite{hoelzle_optimizing_1991} PICs
-find the web page of V8 about maps
+We already explored promotion in other context, such as earlier versions of
+PyPy's JIT \cite{armin_rigo_jit_2007} as well as a Prolog partial evaluator
+\cite{carl_friedrich_bolz_towards_????}. Promotion is quite similar to
+(polymorphic) inline caching and runtime type feedback techniques which were
+first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
+\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
+Promotion is more general because any information can be cached in line, not
+just classes of method receivers.
 
@@
 
-jruby used versions at some point, jvm-l mailing list discusses them
+%jruby used versions at some point, jvm-l mailing list discusses them
 

### 2011-03-28 talk/icooolps2011/paper.tex
improve conclusion

@@
 \begin{abstract}
+
+XXX
 
@@
 
-\section{Conclusion and Next Steps}
+\section{Conclusion}
 
 In this paper we presented two hints that can be used in the source code of an
-interpreter written with PyPy. They are used to influence what the optimizer
-does with the trace. We also showed how a small but still relevant dynamic
-object model can use these hints to no longer use any dictionary lookups after
-tracing. Instead a number of guards are inserted into the trace to check whether
-the assumptions about the objects are still true. This makes operations on
-objects seriously faster.
-
-XXX
+interpreter written with PyPy. They give control over runtime feedback and
+optimization to the language implementor. They are expressive enough for
+building well-known virtual machine optimization techniques, such as maps and
+inline caches. We believe that they are flexible enough to express a wide
+variety of language semantics efficiently.
 

### 2011-03-30 talk/icooolps2011/paper.tex
add some benchmarks results, of course now we are over the page limit

@@
 a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
-\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython (a restricted
+\cite{armin_rigo_pypys_2006} contains a tracing JIT for Python (a restricted
 subset of Python). This JIT is then used to trace a number of languages
@@
 
+For space reasons we cannot perform a full evaluation here, but still want to
+present some benchmark numbers. We chose to present two benchmarks, a port of
+the classical Richards benchmark in RPython and a Python version of the Telco
+decimal benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
+using a pure Python decimal floating point implementation.
+
+The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
+with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
+2.6.35. We compared the performance of various Python implementations on the
+benchmarks. As a baseline, we used the standard Python implementation in C,
+CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
+interpreter. We compare it against four versions of PyPy's Python interpreter,
+all of them with JIT enabled. The PyPy baseline does not enable maps or type
+versions. Then we have a version each where maps and versions are enabled alone
+and finally a version with both.
+
+All benchmarks were run 50 times in the same process, to give the JIT time to
+produce machine code. The arithmetic mean of the times of the last 30 runs were
+used as the result. The errors were computed using a confidence interval with a
+95\% confidence level \cite{georges_statistically_2007}. The results are
+reported in Figure~\ref{fig:times}.
+
+XXX analysis
+
+\begin{figure}
+{\footnotesize
+\begin{center}
+\begin{tabular}{|l|r|r|}
+\hline
+        &richards[ms]       &telco[ms] \\
+\hline
+CPython &357.79 $\pm$ 1.32     &1209.67 $\pm$ 2.20\\
+speedup &1.00 $\times$ &1.00 $\times$\\
+\hline
+JIT baseline    &421.87 $\pm$ 0.48     &738.18 $\pm$ 3.29\\
+speedup &0.85 $\times$ &1.64 $\times$\\
+\hline
+JIT map &382.88 $\pm$ 4.40     &834.19 $\pm$ 4.91\\
+speedup &0.93 $\times$ &1.45 $\times$\\
+\hline
+JIT version     &49.87 $\pm$ 0.29      &157.88 $\pm$ 1.79    \\
+speedup &7.17 $\times$ &7.66 $\times$\\
+\hline
+JIT full        &17.89 $\pm$ 1.15      &153.48 $\pm$ 1.86    \\
+speedup &20.00 $\times$        &7.88 $\times$\\
+\hline
+\end{tabular}
+\end{center}
+}
+\caption{Benchmark Results}
+\label{fig:times}
+\end{figure}
+
 \section{Related Work}

### 2011-03-30 talk/icooolps2011/paper.tex
nonsense

@@
 present some benchmark numbers. We chose to present two benchmarks, a port of
-the classical Richards benchmark in RPython and a Python version of the Telco
+the classical Richards benchmark in Python and a Python version of the Telco
 decimal benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},

### 2011-03-30 talk/icooolps2011/paper.tex
analyze a bit

@@
 decimal benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
-using a pure Python decimal floating point implementation.
+using a pure Python decimal floating point implementation. The results we see in
+these two benchmarks seem to repeat themselves in other benchmarks using
+object-oriented code, for purely numerical algorithms the speedups are a lot
+lower.
 
@@
 
-XXX analysis
+Versioned types speed up both benchmarks by a significant factor of around 7.
+The speed advantage of maps alones is a lot less clear. Maps also have a memory
+advantage which we did not measure here. By themselves, maps improved the
+Richards benchmark slightly, but made the Telco benchmark slower. Enabling both
+maps and versioned types together yields a significant improvement over just
+versioned types for Richards. XXX good explanation. For Telco, enabling both
+does not change much.
 

### 2011-03-30 talk/icooolps2011/paper.tex
a minimal abstract

@@
 
-XXX
+A meta-tracing JIT is a JIT that is applicable to a variety of different
+languages without explicitly encoding language semantics into the compiler. So
+far, meta-tracing JITs lacked a way to feed back runtime information into the
+compiler, which restricted their performance. In this paper we describe the
+mechanisms in PyPy's meta-tracing JIT that can be used to control runtime
+feedback in flexible and language-specific ways. These mechanisms are flexible
+enough to implement classical VM techniques such as maps and polymorphic inline
+caches.
 

### 2011-03-30 talk/icooolps2011/paper.tex
minor language rephrasing

@@
 these two benchmarks seem to repeat themselves in other benchmarks using
-object-oriented code, for purely numerical algorithms the speedups are a lot
-lower.
+object-oriented code; for purely numerical algorithms the speedups are 
+significantly smaller.
 
@@
 all of them with JIT enabled. The PyPy baseline does not enable maps or type
-versions. Then we have a version each where maps and versions are enabled alone
-and finally a version with both.
+versions. We then benchmarked PyPy, first using each technique separately,
+and finally using both together.
 
@@
 Versioned types speed up both benchmarks by a significant factor of around 7.
-The speed advantage of maps alones is a lot less clear. Maps also have a memory
+The speed advantage of maps alone is a lot less clear. Maps also have a memory
 advantage which we did not measure here. By themselves, maps improved the
@@
 versioned types for Richards. XXX good explanation. For Telco, enabling both
-does not change much.
+has little effect over the gains for versioned types alone.
 

### 2011-03-30 talk/ep2008/examples/del.py

@@ -1,0 +1,6 @@
+l = []
+class A(object):
+    def __del__(self):
+        l.append(3)
+A()
+print l

### 2011-03-30 talk/icooolps2011/paper.tex
tweak abstract

--- 2011-03-30 talk/icooolps2011/paper.tex
@@
 
-XXX
+Meta-tracing JITs can be applied to a variety of different
+languages without explicitly encoding language semantics into the compiler. So
+far, they lacked a way to feed back runtime information into the
+compiler, which restricted their performance. In this paper we describe the
+flexible mechanisms in PyPy's meta-tracing JIT that can be used to control runtime feedback in language-specific ways. These mechanisms are flexible
+enough to implement classical VM techniques such as maps and polymorphic inline
+caches.
 

### 2011-03-30 talk/icooolps2011/paper.tex
Idea

@@
 
+\pedronis{XXX radical idea, given that there may be no space to discuss the subtle points and that this is really about showing that we can enable such mechanisms and there is already literature that shows that/how they work, to just consider the benchmarks with no maps and no versions and with both?}
+
 \begin{figure}

### 2011-03-30 talk/icooolps2011/paper.tex
change the accents of the invokedynamic consideration

@@
 \cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
-implementation of dynamic languages on top of JVMs easier. The bytecode gives
-the language implementor control over how the JIT optimizes the language's
-features and when optimized code needs to be deoptimized. XXX
+implementation of dynamic languages on top of JVMs easier.
+The bytecode gives access to user accessible generalized inline cache.
+It requires of course compilation to JVM bytecode instead of simply writing an interpreter, predictability of performance across JVMs is also an open question.
 

### 2011-04-01 talk/icooolps2011/paper.tex
small tweaks, add a citation, reword a paragraph

--- 2011-03-30 talk/icooolps2011/paper.tex
@@
 
-\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^a$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
-            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
-           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
-
-            $^b$merlinux GmbH, Hildesheim, Germany
-
-            $^c$Open End, Göteborg, Sweden
+\authorinfo{Carl Friedrich Bolz \and XXX}
+           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
            }
-           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu \and
-           leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}
+           {cfbolz@gmx.de \and XXX}
 
@@
 
-Meta-tracing JITs can be applied to a variety of different
-languages without explicitly encoding language semantics into the compiler. So
-far, they lacked a way to feed back runtime information into the
-compiler, which restricted their performance. In this paper we describe the
-flexible mechanisms in PyPy's meta-tracing JIT that can be used to control runtime feedback in language-specific ways. These mechanisms are flexible
-enough to implement classical VM techniques such as maps and polymorphic inline
-caches.
-
 \end{abstract}
@@
 languages such as Python, JavaScript or Ruby have rather complex core object
-semantics. For them, implementing just an interpreter is already a complex
+semantics. For them, even implementing just an interpreter is already a complex
 task. Implementing them efficiently with a just-in-time compiler (JIT) is
@@
 a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
-\cite{armin_rigo_pypys_2006} contains a tracing JIT for Python (a restricted
+\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython \cite{davide_ancona_rpython:_2007} (a restricted
 subset of Python). This JIT is then used to trace a number of languages
@@
 
-These projects have in common that they work one meta-level down, providing a tracing JIT for the implementation
+These projects have in common that they work one meta-level down, providing a tracing JIT for the
 language used to implement the dynamic language, and not for the dynamic language itself.
@@
 
-Conceptually the significant speed-ups that can be achieved with
-dynamic compilation depend on feeding into compilation and exploiting
-values observed at runtime that are slow-varying in practice. To exploit the
-runtime feedback, the implementation code and data structures need to be
-structured so that many such values are at hand. The hints that we present allow
-exactly to implement such feedback and exploitation in a meta-tracing context.
+% XXX: paragraph rephrased by anto; feel free to pick the one you like best
+
+Conceptually, it is possible to achieve significant speed-ups by feeding into
+the compiler some information that is observed at runtime: in particular, if
+there are values which vary very slowly, it is possible to compile multiple
+specialized versions of the same code, one for each actual value.  To exploit
+the runtime feedback, the implementation code and data structures need to be
+structured so that many such slow-varying values are at hand. The hints that
+we present allow exactly to implement such feedback and exploitation in a
+meta-tracing context.
+
+% Conceptually the significant speed-ups that can be achieved with
+% dynamic compilation depend on feeding into compilation and exploiting
+% values observed at runtime that are slow-varying in practice. To exploit the
+% runtime feedback, the implementation code and data structures need to be
+% structured so that many such values are at hand. The hints that we present allow
+% exactly to implement such feedback and exploitation in a meta-tracing context.
 
@@
 
+The advantage of this approach is that writing an interpreter is much easier
+and less error prone than manually writing a JIT compiler.  Similarly, writing
+in a high level language such as RPython is easier than writing in C.
+
 We call the code that runs on top of an interpreter implemented with PyPy the
@@
 that correspond to loops in the traced program, but most tracing JITs now also
-have support for tracing non-loops \cite{andreas_gal_incremental_2006}.
+have support for tracing non-loops \cite{XXX}.
 
@@
 language semantics do not need to be encoded into the JIT. Instead the tracer
-just picks them up from the interpreter.
+just picks them up from the interpreter. XXX mention disadvantage of long
+traces?
 
@@
 
-For space reasons we cannot perform a full evaluation here, but still want to
-present some benchmark numbers. We chose to present two benchmarks, a port of
-the classical Richards benchmark in Python and a Python version of the Telco
-decimal benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
-using a pure Python decimal floating point implementation. The results we see in
-these two benchmarks seem to repeat themselves in other benchmarks using
-object-oriented code, for purely numerical algorithms the speedups are a lot
-lower.
-
-The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
-with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
-2.6.35. We compared the performance of various Python implementations on the
-benchmarks. As a baseline, we used the standard Python implementation in C,
-CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
-interpreter. We compare it against four versions of PyPy's Python interpreter,
-all of them with JIT enabled. The PyPy baseline does not enable maps or type
-versions. Then we have a version each where maps and versions are enabled alone
-and finally a version with both.
-
-All benchmarks were run 50 times in the same process, to give the JIT time to
-produce machine code. The arithmetic mean of the times of the last 30 runs were
-used as the result. The errors were computed using a confidence interval with a
-95\% confidence level \cite{georges_statistically_2007}. The results are
-reported in Figure~\ref{fig:times}.
-
-Versioned types speed up both benchmarks by a significant factor of around 7.
-The speed advantage of maps alones is a lot less clear. Maps also have a memory
-advantage which we did not measure here. By themselves, maps improved the
-Richards benchmark slightly, but made the Telco benchmark slower. Enabling both
-maps and versioned types together yields a significant improvement over just
-versioned types for Richards. XXX good explanation. For Telco, enabling both
-does not change much.
-
-\pedronis{XXX radical idea, given that there may be no space to discuss the subtle points and that this is really about showing that we can enable such mechanisms and there is already literature that shows that/how they work, to just consider the benchmarks with no maps and no versions and with both?}
-
-\begin{figure}
-{\footnotesize
-\begin{center}
-\begin{tabular}{|l|r|r|}
-\hline
-        &richards[ms]       &telco[ms] \\
-\hline
-CPython &357.79 $\pm$ 1.32     &1209.67 $\pm$ 2.20\\
-speedup &1.00 $\times$ &1.00 $\times$\\
-\hline
-JIT baseline    &421.87 $\pm$ 0.48     &738.18 $\pm$ 3.29\\
-speedup &0.85 $\times$ &1.64 $\times$\\
-\hline
-JIT map &382.88 $\pm$ 4.40     &834.19 $\pm$ 4.91\\
-speedup &0.93 $\times$ &1.45 $\times$\\
-\hline
-JIT version     &49.87 $\pm$ 0.29      &157.88 $\pm$ 1.79    \\
-speedup &7.17 $\times$ &7.66 $\times$\\
-\hline
-JIT full        &17.89 $\pm$ 1.15      &153.48 $\pm$ 1.86    \\
-speedup &20.00 $\times$        &7.88 $\times$\\
-\hline
-\end{tabular}
-\end{center}
-}
-\caption{Benchmark Results}
-\label{fig:times}
-\end{figure}
-
 \section{Related Work}
@@
 interpreters into compilers using the second futamura projection
-\cite{futamura_partial_1999}. Given that classical partial evaluation works
-strictly ahead of time, it inherently cannot support runtime feedback.
+\cite{futamura_partial_1999}. XXX
 
@@
 the language implementor control over how the JIT optimizes the language's
-features and when optimized code needs to be deoptimized. XXX
+features. XXX
 
-We already explored promotion in other context, such as earlier versions of
-PyPy's JIT \cite{armin_rigo_jit_2007} as well as a Prolog partial evaluator
-\cite{carl_friedrich_bolz_towards_????}. Promotion is quite similar to
-(polymorphic) inline caching and runtime type feedback techniques which were
-first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
-\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
-Promotion is more general because any information can be cached in line, not
-just classes of method receivers.
+%We already explored promotion in other context, such as earlier examples of
+%promotion \cite{carl_friedrich_bolz_towards_????}, \cite{armin_rigo_jit_2007}
+
+\cite{chambers_efficient_1989} maps
+\cite{hoelzle_optimizing_1994} Type feedback
+\cite{hoelzle_optimizing_1991} PICs
+find the web page of V8 about maps
 
@@
 
-%jruby used versions at some point, jvm-l mailing list discusses them
+jruby used versions at some point, jvm-l mailing list discusses them
 
-\section{Conclusion}
+\section{Conclusion and Next Steps}
 
 In this paper we presented two hints that can be used in the source code of an
-interpreter written with PyPy. They give control over runtime feedback and
-optimization to the language implementor. They are expressive enough for
-building well-known virtual machine optimization techniques, such as maps and
-inline caches. We believe that they are flexible enough to express a wide
-variety of language semantics efficiently.
+interpreter written with PyPy. They are used to influence what the optimizer
+does with the trace. We also showed how a small but still relevant dynamic
+object model can use these hints to no longer use any dictionary lookups after
+tracing. Instead a number of guards are inserted into the trace to check whether
+the assumptions about the objects are still true. This makes operations on
+objects seriously faster.
+
+XXX
 

### 2011-04-01 talk/icooolps2011/paper.tex

--- 2011-03-30 talk/icooolps2011/paper.tex
@@
 semi-automatically to the generated VM is a tracing JIT compiler.
+
+The advantage of this approach is that writing an interpreter is much easier
+and less error prone than manually writing a JIT compiler.  Similarly, writing
+in a high level language such as RPython is easier than writing in C.
 

### 2011-04-04 talk/icooolps2011/paper.tex
swap these two paragraphs; I think it is more readable this way
another small tweak
try to take the best of both versions

@@
 
-% XXX: paragraph rephrased by anto; feel free to pick the one you like best
-
-Conceptually, it is possible to achieve significant speed-ups by feeding into
-the compiler some information that is observed at runtime: in particular, if
+Conceptually, the significant speed-ups that can be achieved with
+dynamic compilation depend on feeding into compilation and exploiting
+values observed at runtime. In particular, if
 there are values which vary very slowly, it is possible to compile multiple
@@
 meta-tracing context.
-
-% Conceptually the significant speed-ups that can be achieved with
-% dynamic compilation depend on feeding into compilation and exploiting
-% values observed at runtime that are slow-varying in practice. To exploit the
-% runtime feedback, the implementation code and data structures need to be
-% structured so that many such values are at hand. The hints that we present allow
-% exactly to implement such feedback and exploitation in a meta-tracing context.
 
@@
 
+Because the traces always correspond to a concrete execution they cannot
+contain any control flow splits. Therefore they encode the control flow
+decisions needed to stay on the trace with the help of \emph{guards}. Those are
+operations that check that the assumptions are still true when the compiled trace is later executed with different values.
+
 To be able to do this recording, VMs with a
@@
 have support for tracing non-loops \cite{andreas_gal_incremental_2006}.
-
-Because the traces always correspond to a concrete execution they cannot
-contain any control flow splits. Therefore they encode the control flow
-decisions needed to stay on the trace with the help of \emph{guards}. Those are
-operations that check that the assumptions are still true when the compiled trace is later executed with different values.
 
@@
 user program. This means that the tracer stops tracing after one iteration of
-the loop in the user function that is being considered. At this point, it can
-have traced many iterations of the interpreter main loop.
+the loop in the user function that is being considered. At this point, it probably
+traced many iterations of the interpreter main loop.
 
@@
 \label{sub:running}
+\anto{this example is not referenced until section \ref{sec:fastobjmodel}: I
+  would put it just before it}
 
@@
 Let's assume we trace through code that sums three attributes, such as:
+\anto{I still think it's a bit weird to call them ``methods'' and then use
+  them as attributes in the example}
 
@@
 
-Promotion is essentially a tool for trace specialization. In some places in the
-interpreter it would be very useful if a variable were constant, even though it
+Promotion is essentially a tool for trace specialization. There are places in
+the interpreter where knowing that a value is constant opens a lot of
+optimization opportunities, even though it
 could have different values in practice. In such a place, promotion is used. The
@@
 a property that is not easily detectable by analysis. Therefore, the purity
-of this function needs to be annotated.
+of this function needs to be annotated manually.
 

### 2011-04-05 talk/icooolps2011/paper.tex
add a sentence that the trace is SSA and stress where guards come from
implement Samuele's radical idea: just show the compounded speedups

@@
 Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
-trace indeed contains
+trace is in SSA form. Note how all the guards in trace correspond to a
+condition in the original code. The trace contains
 five calls to \texttt{dict.get}, which is slow. To make the language efficient
@@
 For space reasons we cannot perform a full evaluation here, but still want to
-present some benchmark numbers. We chose to present two benchmarks, a port of
-the classical Richards benchmark in Python and a Python version of the Telco
-decimal benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
-using a pure Python decimal floating point implementation. The results we see in
-these two benchmarks seem to repeat themselves in other benchmarks using
-object-oriented code; for purely numerical algorithms the speedups are 
-significantly smaller.
+present some benchmark numbers. We chose to present just some benchmarks: The
+templating engine of the Django web
+framework\footnote{\texttt{http://www.djangoproject.com/}}; a Monte-Carlo Go
+AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/
+disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
+Richards benchmark in Python; a Python version of the Telco decimal
+benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
+pure Python decimal floating point implementation. The results we see in these
+benchmarks seem to repeat themselves in other benchmarks using object-oriented
+code; for purely numerical algorithms the speedups are significantly smaller.
 
@@
 95\% confidence level \cite{georges_statistically_2007}. The results are
-reported in Figure~\ref{fig:times}.
+reported in Figure~\ref{fig:times}, together with the same numbers normed to
+those of the full JIT.
 
-Versioned types speed up both benchmarks by a significant factor of around 7.
-The speed advantage of maps alone is a lot less clear. Maps also have a memory
-advantage which we did not measure here. By themselves, maps improved the
-Richards benchmark slightly, but made the Telco benchmark slower. Enabling both
-maps and versioned types together yields a significant improvement over just
-versioned types for Richards. XXX good explanation. For Telco, enabling both
-has little effect over the gains for versioned types alone.
-
-\pedronis{XXX radical idea, given that there may be no space to discuss the subtle points and that this is really about showing that we can enable such mechanisms and there is already literature that shows that/how they work, to just consider the benchmarks with no maps and no versions and with both?}
+The optimizations give a speedup between 80\% and almost 20 times. The Richards
+is a particularly good case for the optimizations as it makes heavy uses of
+object-oriented features. Pyflate uses mostly imperative code, so does not
+benefit as much. Together with the optimization, PyPy outperforms CPython in
+all benchmarks, which is not surprising because CPython is a simple
+bytecode-based interpreter.
 
 \begin{figure}
+\begin{center}
 {\footnotesize
-\begin{center}
-\begin{tabular}{|l|r|r|}
+\begin{tabular}{|l|r|r|r|}
 \hline
-        &richards[ms]       &telco[ms] \\
+	&CPython	&JIT baseline	&JIT full\\
 \hline
-CPython &357.79 $\pm$ 1.32     &1209.67 $\pm$ 2.20\\
-speedup &1.00 $\times$ &1.00 $\times$\\
+django[ms]	&988.67 $\pm$ 0.49	&405.62 $\pm$ 4.80	&149.31 $\pm$ 1.37\\
+	&6.62 $\times$	&2.72 $\times$	&1.00 $\times$\\
 \hline
-JIT baseline    &421.87 $\pm$ 0.48     &738.18 $\pm$ 3.29\\
-speedup &0.85 $\times$ &1.64 $\times$\\
+go[ms]	&947.43 $\pm$ 1.30	&525.53 $\pm$ 7.67	&174.32 $\pm$ 7.78\\
+	&5.44 $\times$	&3.01 $\times$	&1.00 $\times$\\
 \hline
-JIT map &382.88 $\pm$ 4.40     &834.19 $\pm$ 4.91\\
-speedup &0.93 $\times$ &1.45 $\times$\\
+pyflate[ms]	&3209.20 $\pm$ 3.65	&2884.26 $\pm$ 21.11	&1585.48 $\pm$ 5.22\\
+	&2.02 $\times$	&1.82 $\times$	&1.00 $\times$\\
 \hline
-JIT version     &49.87 $\pm$ 0.29      &157.88 $\pm$ 1.79    \\
-speedup &7.17 $\times$ &7.66 $\times$\\
+richards[ms]	&357.79 $\pm$ 1.32	&421.87 $\pm$ 0.48	&17.89 $\pm$ 1.15\\
+	&20.00 $\times$	&23.58 $\times$	&1.00 $\times$\\
 \hline
-JIT full        &17.89 $\pm$ 1.15      &153.48 $\pm$ 1.86    \\
-speedup &20.00 $\times$        &7.88 $\times$\\
+telco[ms]	&1209.67 $\pm$ 2.20	&738.18 $\pm$ 3.29	&153.48 $\pm$ 1.86\\
+	&7.88 $\times$	&4.81 $\times$	&1.00 $\times$\\
 \hline
 \end{tabular}
+}
 \end{center}
-}
 \caption{Benchmark Results}

### 2011-04-05 talk/icooolps2011/paper.tex
the 1st part is a prerequisite for the 2nd

--- 2011-04-01 talk/icooolps2011/paper.tex
@@
 
-Promotion is essentially a tool for trace specialization. In some places in the
-interpreter it would be very useful if a variable were constant, even though it
+Promotion is essentially a tool for trace specialization. There are places in
+the interpreter where knowing that a value if constant opens a lot of
+optimization opportunities, even though it
 could have different values in practice. In such a place, promotion is used. The
@@
 a property that is not easily detectable by analysis. Therefore, the purity
-of this function needs to be annotated.
+of this function needs to be manually annotated.
 

### 2011-04-06 talk/icooolps2011/paper.tex
fix some things

--- 2011-04-05 talk/icooolps2011/paper.tex
@@
 It has long been an objective of the partial evaluation community to
-automatically produce compilers from interpreters. There has been a recent
+automatically produce compilers from interpreters. There has been a
 renaissance of this idea around the approach of tracing just-in-time
@@
 language used to implement the dynamic language, and not for the dynamic language itself.
-The tracing JIT then will trace through the object model of the dynamic
+The tracing JIT will then trace through the object model of the dynamic
 language implementation. This makes the object model transparent to the tracer
@@
 Concretely these hints are used to control how the optimizer of the
-tracing JIT can improve the traces of the object model. More
-specifically, these hints influence the constant folding
+tracing JIT can improve the traces of the object model. In particular the hints
+influence the constant folding
 optimization. The first hint makes it possible to turn arbitrary
@@
 when a function call is encountered the operations of the called functions are
-simply put into the trace too.
+simply put into the trace of the caller too. The tracing JIT tries to produce traces
+that correspond to loops in the traced program, but most tracing JITs now also
+have support for tracing non-loops \cite{andreas_gal_incremental_2006}.
 
@@
 started the interpreter is used; only the most frequently executed paths through the user
-program are turned into machine code. The tracing JIT tries to produce traces
-that correspond to loops in the traced program, but most tracing JITs now also
-have support for tracing non-loops \cite{andreas_gal_incremental_2006}.
+program are turned into machine code. The interpreter is also used when a guard
+fails to continue the execution from the failing guard.
 
@@
 general framework, we want to reuse our tracer for different languages.
-Therefore PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. It does not
+Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
 trace the execution of the user program, but instead traces the execution of
@@
 traced by the tracer are the loops in the
-user program. This means that the tracer stops tracing after one iteration of
+user program. To achieve this the tracer stops tracing after one iteration of
 the loop in the user function that is being considered. At this point, it probably
@@
 techniques, with the difference that it is easier to apply them in a tracing
-JIT because it only has to deal with linear traces.  Among the techniques:
-%
-\begin{itemize}
-    \item constant folding
-    \item common subexpression elimination
-    \item allocation removal \cite{bolz_allocation_2011}
-    \item store/load propagation
-    \item loop invariant code motion
-\end{itemize}
+JIT because it only has to deal with linear traces.  Among the techniques are
+constant folding, common subexpression elimination, allocation removal
+\cite{bolz_allocation_2011}, store/load propagation, loop invariant code
+motion.
 
@@
 \end{figure}
+
+\cfbolz{should we show the code that would create the inst use in tracing?}
 
@@
 
-In the trace above, the value of $x_1$ is statically known thanks to the
-guard. Remember that a guard is a runtime check. The above trace will run to
+In the trace above, the value of $x_1$ is statically known after the guard.
+Remember that a guard is a runtime check. The above trace will run to
 completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
@@
 used by Psyco \cite{rigo_representation-based_2004} and by all older versions
-of PyPy's JIT. Promotion is a technique that only works well in JIT compilers;
+of PyPy's JIT. It is a technique that only works well in JIT compilers;
 in static compilers it is significantly less applicable.
@@
 optimization opportunities, even though it
-could have different values in practice. In such a place, promotion is used. The
+could have different values in practice. In such a place, promotion can be used. The
 typical reason to do that is if there is
@@
 $v_2$ = $z_1$ + $y_1$
-return(v2)
+return($v_2$)
 \end{lstlisting}
@@
 The promotion is turned into a \texttt{guard} operation in the trace. The guard
-captures the value of $x_1$ as it was at runtime. From the point of view of the
+captures the value of $x_1$ as it was during tracing. \cfbolz{drop the word runtime feedback here?}
+From the point of view of the
 optimizer, this guard is not any different than the one produced by the \texttt{if}
@@
 only variables that don't vary much. However, adding a promotion hint will never produce wrong
-results. It might just lead to too much assembler code.
+results. It might just lead to too much assembler code being generated.
 
@@
 in practice the argument types tend not to vary often. Therefore it is possible to
-promote the types. The next section will present a complete example of how
+promote the types. Section~\ref{sec:} will present a complete example of how
 this works.
@@
 
+\cfbolz{should we mention that pure functions are not actually called by the optimizer, but the values that are seen during tracing are used?}
+
 Now the trace will look like this:
@@
 subtly wrong code. Therefore, a lot of care has to be taken when using this
-annotation.
+annotation\footnote{The most common use case of the \texttt{purefunction}
+annotation is indeed to declare the immutability of fields. Because it is so
+common, we have special syntactic sugar for it.}.
 
@@
 of this function needs to be annotated manually.
-
-\subsubsection{Immutable Fields}
-
-One of the most common cases of pure functions is reading immutable
-values out of objects. Since this is so common, we have special syntactic sugar
-for it. A RPython class can have a class attribute \texttt{\_immutable\_fields\_} set to
-a list of strings, listing the fields that cannot be changed. This is equivalent
-to using getters and annotating them with \texttt{purefunction}.
 
@@
 reference to a map, which maps field names to indexes into a storage list. The
-storage list contains the actual field values. The maps are shared between
-objects with the same layout. Therefore they have to be immutable, which means
+storage list contains the actual field values. Therefore they have to be immutable, which means
 that their \texttt{getindex} method is a pure function. When a new attribute is added
 to an instance, a new map needs to be chosen, which is done with the
-\texttt{add\_attribute} method on the previous map (which is also pure). Now that we have
+\texttt{add\_attribute} method on the previous map. This function is also pure,
+because it caches all new instances of \texttt{Map} that it creates, to make
+sure that objects with the same layout have the same map. Now that we have
 introduced maps, it is safe to promote the map everywhere, because we assume
@@
 Therefore, we give every class a version object, which is changed every time a
-class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
+class gets changed (i.e., the \texttt{methods} dictionary changes).
 This means that the result of \texttt{methods.get()} for a given \texttt{(name,
@@
 benchmarks seem to repeat themselves in other benchmarks using object-oriented
-code; for purely numerical algorithms the speedups are significantly smaller.
+code; for purely numerical algorithms the speedups introduced by the techniques
+in this paper are much smaller because they are already fast.
 
@@
 with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
-2.6.35. We compared the performance of various Python implementations on the
+2.6.35. We compared the performance of two Python implementations on the
 benchmarks. As a baseline, we used the standard Python implementation in C,
 CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
-interpreter. We compare it against four versions of PyPy's Python interpreter,
-all of them with JIT enabled. The PyPy baseline does not enable maps or type
-versions. We then benchmarked PyPy, first using each technique separately,
-and finally using both together.
+interpreter. We compare it against two versions of PyPy's Python interpreter,
+both of them with JIT enabled. The PyPy baseline does not enable maps or type
+version, the full JIT enables both.
 
@@
 95\% confidence level \cite{georges_statistically_2007}. The results are
-reported in Figure~\ref{fig:times}, together with the same numbers normed to
+reported in Figure~\ref{fig:times}, together with the same numbers normalized to
 those of the full JIT.

### 2011-04-07 talk/icooolps2011/paper.tex
add a comment
s/fancy/advanced/

@@
 
+XXX [fijal] yet another advantage is that JIT is by design supporting the whole
+    language
+
 We call the code that runs on top of an interpreter implemented with PyPy the
@@
 object model that just supports classes and instances, without any
-inheritance or other fancy features. The model has classes, which contain methods.
+inheritance or other advanced features. The model has classes, which contain methods.
 Instances have a class. Instances have their own attributes (or fields). When looking up an

### 2011-04-07 talk/icooolps2011/paper.tex
I think Python object model contains more hard features, but those are the major ones
add a note

@@
 While this object model is very
-simple it already contains all the hard parts of Python's object model. Both
+simple it already contains most hard parts of Python's object model. Both
 instances and classes can have arbitrary fields, and they are changeable at
@@
 
+XXX [fijal] maybe it's worth noting that those guards are removed out of
+    the loop by loop-invariant-code motion, unless our is so special
+    we want to write a special paper about it
+
 %___________________________________________________________________________

### 2011-04-11 talk/icooolps2011/paper.tex
add another note
fix one of maciej's comment, ignore two

@@
 
-XXX [fijal] yet another advantage is that JIT is by design supporting the whole
-    language
-
 We call the code that runs on top of an interpreter implemented with PyPy the
@@
 language semantics do not need to be encoded into the JIT. Instead the tracer
-just picks them up from the interpreter.
+just picks them up from the interpreter. This also means that the JIT by
+construction supports the full language.
 
@@
 
-XXX [fijal] maybe it's worth noting that those guards are removed out of
-    the loop by loop-invariant-code motion, unless our is so special
-    we want to write a special paper about it
-
 %___________________________________________________________________________

### 2011-04-13 talk/icooolps2011/paper.tex
some comments from Peng Wu

@@
 
+\cfbolz{XXX stress more that "the crux of the techniques and a significant
+portion of new contributions in the paper are from how to refactoring codes to
+expose likely runtime constants and pure functions"}
+
 \section{Background}
@@
 
+\cfbolz{XXX more precise definition of what promote does}
+\cfbolz{change the hint(x, promote=True) syntax to something more readable}
+
 In this section we will describe how to add two hints that allow the
@@
 
-The meaning of this hint is that the tracer should pretend that \texttt{x} is a
-constant
+The hint indicates that \texttt{x} is likely a runtime constant and the JIT
+should try to perform runtime specialization on it
 in the code that follows. When just running the code, the function has no
@@
 this works.
+
+\cfbolz{XXX explain how value specialization on the interpreter level can lead
+to type specialization on the language level}
 
@@
 \end{lstlisting}
+
+\cfbolz{XXX define the meaning of purefunction more precisely, particularly because add\_attribute has side effects, which is confusing}
 

### 2011-04-13 talk/icooolps2011/paper.tex
comments by rhy0lite

@@
 structured so that many such slow-varying values are at hand. The hints that
-we present allow exactly to implement such feedback and exploitation in a
+we present precisely allow us to implement such feedback and exploitation in a
 meta-tracing context.
@@
 
+\cfbolz{XXX also explain that some forwarding of guarded values is happening,
+make clearer which figures show optimized code and which show non-optimized
+code}
+
 The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
@@
 
+XXX Peng Wu and David Edelsohn
+
 \bibliographystyle{abbrv}

### 2011-04-13 talk/icooolps2011/code/map.tex
change x = hint(x, promote=True) to just promote(x) to reduce confusion

@@
     def getfield(self, name):
-        map = hint(self.map, promote=True)
+        map = self.map
+        promote(map)
         index = map.getindex(name)
@@
     def write_attribute(self, name, value):
-        map = hint(self.map, promote=True)
+        map = self.map
+        promote(map)
         index = map.getindex(name)

### 2011-04-13 talk/icooolps2011/code/version.tex
Typos and typographic details.
change x = hint(x, promote=True) to just promote(x) to reduce confusion

@@
     def find_method(self, name):
-        self = hint(self, promote=True)
-        version = hint(self.version, promote=True)
+        promote(self)
+        version = self.version
+        promote(version)
         return self._find_method(name, version)

### 2011-04-13 talk/icooolps2011/paper.tex
change x = hint(x, promote=True) to just promote(x) to reduce confusion

@@
 \cfbolz{XXX more precise definition of what promote does}
-\cfbolz{change the hint(x, promote=True) syntax to something more readable}
 
@@
 def f1(x, y):
-    x = hint(x, promote=True)
+    promote(x)
     z = x * 2 + 1
@@
 should try to perform runtime specialization on it
-in the code that follows. When just running the code, the function has no
-effect, as it simply returns its first argument. When tracing, some extra work
+in the code that follows.\footnote{For technical reasons the promote hint needs
+to be written down slightly differently in the actual code.}  When just running
+the code, the \texttt{promote} function has no
+effect. When tracing, some extra work
 is done. Let's assume that this changed function is traced with
@@
     def f(self, val):
-        self = hint(self, promote=True)
+        promote(self)
         self.y = self.compute() + val

### 2011-04-13 talk/icooolps2011/paper.tex
michael's notes

@@
 
-Meta-tracing JITs can be applied to a variety of different
+Meta-tracing JIT compilers can be applied to a variety of different
 languages without explicitly encoding language semantics into the compiler. So
@@
 caches.
+
+\cfbolz{XXX tracing is runtime feed back too, clarify}
 
@@
 optimize its object model. This is made harder by the fact that many recent
-languages such as Python, JavaScript or Ruby have rather complex core object
+languages such as Python, JavaScript or Ruby have a rather complex core object
 semantics. For them, even implementing just an interpreter is already a complex
-task. Implementing them efficiently with a just-in-time compiler (JIT) is
+task. Implementing these languages efficiently with a just-in-time compiler (JIT) is
 extremely challenging, because of their many corner-cases.
@@
 
+\cfbolz{XXX kill the next paragraph? the info is repeated in the list below}
 Concretely these hints are used to control how the optimizer of the
@@
 
-Together these two hints can be used to express many classic implementation
+Together these hints can be used to express many classic implementation
 techniques used for object models of dynamic languages, such as maps and
@@
 \begin{itemize}
- \item A hint to turn arbitrary variables into constants in the trace, that
- means the feedback of runtime information into compilation.
- \item A way to define new pure operations which the constant folding
+ \item A hint to turn arbitrary variables into constants in the trace by
+ feeding back runtime information into compilation.
+ \item A way to annotate operations as pure which the constant folding
  optimization then recognizes.
@@
 program. Those
-traces are therefore linear list of operations, which are optimized and then
+traces are linear lists of operations, which are optimized and then
 get turned into machine code. This recording automatically inlines functions:
@@
 
-One disadvantage of (tracing) JITs which makes them not directly applicable to
-PyPy is that they need to encode the language semantics of the language they are
-tracing. Since PyPy wants to be a
-general framework, we want to reuse our tracer for different languages.
+Since PyPy wants to be a general framework, we want to reuse our tracer for
+different languages, which makes classical tracers inapplicable, because they
+encode language semantics.
 Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
@@
 user program. To achieve this the tracer stops tracing after one iteration of
-the loop in the user function that is being considered. At this point, it probably
+the loop in the user function that is being considered; at this point, it probably
 traced many iterations of the interpreter main loop.
@@
 object model that just supports classes and instances, without any
-inheritance or other advanced features. The model has classes, which contain methods.
+inheritance or other advanced features. In the model classes contain methods.
 Instances have a class. Instances have their own attributes (or fields). When looking up an
@@
 
-To implement this object model, we could use the RPython code in
+To implement this object model, we use the RPython code in
 Figure~\ref{fig:interpreter-slow} as part of the interpreter source code.
@@
 When using this object model in
-an interpreter, a huge amount of time will be spent doing lookups in these
+an interpreter, a large amount of time will be spent doing lookups in these
 dictionaries.
@@
 
-The simplest example of constants are literal values.  For example, if in the
-RPython source code we have a line like \texttt{y = x + 1}, the second operand will
-be a constant in the trace.
-
+The simplest example of constants are literal values, such as \texttt{1}.
 However, the optimizer can statically know the value of a variable even if it
@@
 in practice the argument types tend not to vary often. Therefore it is possible to
-promote the types. Section~\ref{sec:} will present a complete example of how
+promote the types. Section~\ref{sec:fastobjmodel} will present a complete example of how
 this works.
@@
 In the previous section we saw a way to turn arbitrary variables into constants. All
-pure operations on these constants can be constant-folded. This works great for
+pure operations on these constants can be constant-folded. This works well for
 constant folding of simple types, e.g. integers. Unfortunately, in the context of an
@@
 common, we have special syntactic sugar for it.}.
+
+\cfbolz{XXX mention a possible debug mode for findings bugs in this, or too many values in promotion. stress more that promote is safe}
 

### 2011-04-13 talk/icooolps2011/code/map.tex
replace "pure" with "invariant", add a precise definition of the term

@@
 
-    @purefunction
+    @invariant
     def getindex(self, name):
@@
 
-    @purefunction
+    @invariant
     def add_attribute(self, name):

### 2011-04-13 talk/icooolps2011/paper.tex
replace "pure" with "invariant", add a precise definition of the term

@@
 In this paper we present two of these hints that are extensively used in the
-PyPy project to improve the performance of its Python interpreter.
+PyPy project to improve the performance of its Python interpreter, particularly
+of the object model.
 
@@
 \cfbolz{XXX kill the next paragraph? the info is repeated in the list below}
-Concretely these hints are used to control how the optimizer of the
-tracing JIT can improve the traces of the object model. In particular the hints
+In particular the hints
 influence the constant folding
@@
  feeding back runtime information into compilation.
- \item A way to annotate operations as pure which the constant folding
- optimization then recognizes.
+ \item A way to annotate operations which the constant folding
+ optimization then recognizes and exploits.
  \item A worked-out example of a simple object model of a dynamic language and
@@
 portion of new contributions in the paper are from how to refactoring codes to
-expose likely runtime constants and pure functions"}
+expose likely runtime constants and invariant functions"}
 
@@
     i.e. statically known by the optimizer
-    \item the operation needs to be \emph{pure}, i.e. always yield the same result given
-    the same arguments.
+    \item the operation needs to be \emph{constant-foldable}, i.e. always yield
+    the same result given the same arguments.
 \end{itemize}
@@
 
-\subsection{Declaring New Pure Operations}
+\subsection{Declaring New Foldable Operations}
 
 In the previous section we saw a way to turn arbitrary variables into constants. All
-pure operations on these constants can be constant-folded. This works well for
+foldable operations on these constants can be constant-folded. This works well for
 constant folding of simple types, e.g. integers. Unfortunately, in the context of an
@@
 language, most operations actually manipulate objects, not simple types. The
-operations on objects are often not pure and might even have side-effects. If
+operations on objects are often not foldable and might even have side-effects. If
 one reads a field out of a constant reference to an object this cannot
@@
 hint is needed.
+
+This hint can be used to mark functions as \emph{invariant}. A function is
+termed invariant if, during the execution of the program, successive calls to
+the function with identical arguments always return the same result. From this
+definition follows that a call to an invariant function with constant arguments
+in a trace can be replaced with the result of the call.
 
@@
 immutable, and therefore \texttt{compute}
-is a pure function. To communicate this, there is a \texttt{purefunction} decorator.
+is an invariant function. To communicate this, there is a \texttt{invariant} decorator.
 If the code in \texttt{compute} should be constant-folded away, we would change the
@@
 
-    @purefunction
+    @invariant
     def compute(self):
@@
 \end{lstlisting}
-
-\cfbolz{XXX define the meaning of purefunction more precisely, particularly because add\_attribute has side effects, which is confusing}
-
-\cfbolz{should we mention that pure functions are not actually called by the optimizer, but the values that are seen during tracing are used?}
 
@@
 during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
-has a chance to see it. Since the \texttt{compute} function is marked as pure, and its
+has a chance to see it. Since the \texttt{compute} function is marked as invariant, and its
 argument
@@
 
-On the one hand, the \texttt{purefunction} annotation is very powerful. It can be
+On the one hand, the \texttt{invariant} annotation is very powerful. It can be
 used to constant-fold arbitrary parts of the computation in the interpreter.
-However, the annotation also gives the interpreter author ample opportunity to mess things up. If a
-function is annotated to be pure, but is not really, the optimizer can produce
+However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
+function is annotated to be invariant, but is not really, the optimizer can produce
 subtly wrong code. Therefore, a lot of care has to be taken when using this
-annotation\footnote{The most common use case of the \texttt{purefunction}
+annotation\footnote{The most common use case of the \texttt{invariant}
 annotation is indeed to declare the immutability of fields. Because it is so
-common, we have special syntactic sugar for it.}.
+common, we have special syntactic sugar for it.}. We hope to introduce a
+debugging mode which would (slowly) check whether the annotation is applied
+incorrectly to mitigate this problem.
 
-\cfbolz{XXX mention a possible debug mode for findings bugs in this, or too many values in promotion. stress more that promote is safe}
+\subsubsection{Observably invariant Functions}
 
-\subsubsection{Observably Pure Functions}
+\cfbolz{XXX do we kill this section?}
 
 Why can't we simply write an analysis to find out that the \texttt{x} fields of the
-\texttt{A} instances is immutable and deduce that \texttt{compute} is a pure function,
+\texttt{A} instances is immutable and deduce that \texttt{compute} is an invariant function,
 since it only reads the \texttt{x} field and does not have side effects? This might
 be possible in this particular case, but in practice the functions that are
-annotated with the \texttt{purefunction} decorator are usually more complex.
+annotated with the \texttt{invariant} decorator are usually more complex.
 The easiest example for this is that of a function that uses memoization to
@@
 side effects, because it changes the memoizing dictionary. However, because this side
-effect is not externally visible, the function from the outside is pure. This is
-a property that is not easily detectable by analysis. Therefore, the purity
-of this function needs to be annotated manually.
+effect is not externally visible, the function is still invariant. This is
+a property that is not easily detectable by analysis.
 
@@
 two previous sections don't seem to help with the current object model. There is
-no pure function to be seen, and the instance is not a candidate for promotion,
+no invariant function to be seen, and the instance is not a candidate for promotion,
 because there tend to be many instances.
@@
 This is a common problem when trying to apply hints. Often, the interpreter
-needs a small rewrite to expose the pure functions and nearly-constant objects
+needs a small rewrite to expose the invariant functions and nearly-constant objects
 that are implicitly there. In the case of instance fields this rewrite is not
@@
 storage list contains the actual field values. Therefore they have to be immutable, which means
-that their \texttt{getindex} method is a pure function. When a new attribute is added
+that their \texttt{getindex} method is an invariant function. When a new attribute is added
 to an instance, a new map needs to be chosen, which is done with the
-\texttt{add\_attribute} method on the previous map. This function is also pure,
+\texttt{add\_attribute} method on the previous map. This function is also invariant,
 because it caches all new instances of \texttt{Map} that it creates, to make
@@
 The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
-a pure function and they have constant arguments. That means that \texttt{index1/2/3}
+an invariant function and they have constant arguments. That means that \texttt{index1/2/3}
 are constant and the guards on them can be removed. All but the first guard on
@@
 
-What we would really like is if the \texttt{Class.find\_method} method were pure.
+What we would really like is if the \texttt{Class.find\_method} method were invariant.
 But it cannot be, because it is always possible to change the class itself.
@@
 class gets changed (i.e., the \texttt{methods} dictionary changes).
-This means that the result of \texttt{methods.get()} for a given \texttt{(name,
-version)} pair will always be the same, i.e. it is a pure operation.  To help
+This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
+version)} pair will always be the same, i.e. it is an invariant operation.  To help
 the JIT to detect this case, we factor it out in a helper method which is
-explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} can
+explicitly marked as \texttt{@invariant}. The refactored \texttt{Class} can
 be seen in Figure~\ref{fig:version}
@@
 argument but it does not use it at all. Its only purpose is to make the call
-pure, because when the version object changes, the result of the call might be
+invariant, because when the version object changes, the result of the call might be
 different than the previous one.
@@
 
-XXX Peng Wu and David Edelsohn
+XXX Peng Wu and David Edelsohn, Laura Creighton
 

### 2011-04-13 talk/icooolps2011/paper.tex
fix a typo and an XXX

@@
 
-It is worth clarifying what is a ``constant'' in this context.  A variable of
+It is worth clarifying what a ``constant'' is in this context.  A variable of
 the trace is said to be constant if its value is statically known by the
@@
 have values that are variable but vary little in the context of parts of a user
-program. An example would be the types of variables in a user function. Even
-though in principle the argument to a Python function could be any Python type,
-in practice the argument types tend not to vary often. Therefore it is possible to
-promote the types. Section~\ref{sec:fastobjmodel} will present a complete example of how
-this works.
-
-\cfbolz{XXX explain how value specialization on the interpreter level can lead
-to type specialization on the language level}
+program. An example would be the types of variables in a user function, which
+rarely change in a dynamic language in practice (even though they could). In the
+interpreter, these user-level types are values. Thus promoting them will lead
+to type-specialization on the level of the user program.
+Section~\ref{sec:fastobjmodel} will present a complete example of how this
+works.
 

### 2011-04-13 talk/icooolps2011/paper.tex
footnotize a paragraph

@@
 classes to change at all. This is not totally reasonable (sometimes classes contain
-counters or similar things) but for this simple example it is good enough.
+counters or similar things) but for this simple example it is good
+enough.\footnote{There is a more complex variant of class versions that can
+accommodate class fields that change a lot better.}
 
@@
 other hand, a method lookup in a complex class hierarchy is as optimized in the
-trace as in our object model here.
-
-A downside of the versioning of classes that we haven't yet fixed in PyPy, is
-that some classes \emph{do} change a lot. An example would be a class that keeps a
-counter of how many instances have been created so far. This is very slow right
-now, but we have ideas about how to fix it in the future.
+trace as in our simple object model above.
 
@@
 with its class. In our code above, we allow both to vary independently.
-In PyPy's Python interpreter we act somewhat more cleverly. The class of
-an instance is not stored on the instance itself, but on the map. This means
-that we get one fewer promotion (and thus one fewer guard) in the trace,
-because the class doesn't need to
-be promoted after the map has been.
+Therefore we store the class of an instance on the map in PyPy's Python
+interpreter. This means that we get one fewer promotion (and thus one fewer
+guard) in the trace, because the class doesn't need to be promoted after the
+map has been.
 

### 2011-04-13 talk/icooolps2011/paper.tex
update definition, not too great and a bit circular

@@
 
+XXX not too happy with the definition
+
 This hint can be used to mark functions as \emph{invariant}. A function is
-termed invariant if, during the execution of the program, successive calls to
-the function with identical arguments always return the same result. From this
+termed invariant if, during the execution of the program, the results of
+subsequent calls to the function with identical arguments may be be replaced
+with the result of the first call without changing the program's behaviour.
+From this
 definition follows that a call to an invariant function with constant arguments
-in a trace can be replaced with the result of the call.
+in a trace can be replaced with the result of the call.\footnote{This property
+is less strict than that of a "pure" function, because it is only about actual
+calls during execution.}
 

### 2011-04-14 talk/icooolps2011/code/map.tex
use the (admittedly obscure) name "trace-elidable", simply because nobody has preconceptions abous it

@@
 
-    @invariant
+    @elidable
     def getindex(self, name):
@@
 
-    @invariant
+    @elidable
     def add_attribute(self, name):

### 2011-04-14 talk/icooolps2011/paper.tex
use the (admittedly obscure) name "trace-elidable", simply because nobody has preconceptions abous it

--- 2011-04-13 talk/icooolps2011/paper.tex
@@
 portion of new contributions in the paper are from how to refactoring codes to
-expose likely runtime constants and invariant functions"}
+expose likely runtime constants and trace-elidable functions"}
 
@@
 
-This hint can be used to mark functions as \emph{invariant}. A function is
-termed invariant if, during the execution of the program, successive calls to
-the function with identical arguments always return the same result. From this
-definition follows that a call to an invariant function with constant arguments
-in a trace can be replaced with the result of the call.
+XXX not too happy with the definition
+
+This hint can be used to mark functions as \emph{trace-elidable}. A function is
+termed trace-elidable if, during the execution of the program, the results of
+subsequent calls to the function with identical arguments may be be replaced
+with the result of the first call without changing the program's behaviour.
+From this
+definition follows that a call to an trace-elidable function with constant arguments
+in a trace can be replaced with the result of the call.\footnote{This property
+is less strict than that of a "pure" function, because it is only about actual
+calls during execution. All pure functions are trace-elidable though.}
 
@@
 immutable, and therefore \texttt{compute}
-is an invariant function. To communicate this, there is a \texttt{invariant} decorator.
+is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
 If the code in \texttt{compute} should be constant-folded away, we would change the
@@
 
-    @invariant
+    @elidable
     def compute(self):
@@
 during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
-has a chance to see it. Since the \texttt{compute} function is marked as invariant, and its
+has a chance to see it. Since the \texttt{compute} function is marked as trace-elidable, and its
 argument
@@
 
-On the one hand, the \texttt{invariant} annotation is very powerful. It can be
+On the one hand, the \texttt{elidable} annotation is very powerful. It can be
 used to constant-fold arbitrary parts of the computation in the interpreter.
 However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
-function is annotated to be invariant, but is not really, the optimizer can produce
+function is annotated to be trace-elidable, but is not really, the optimizer can produce
 subtly wrong code. Therefore, a lot of care has to be taken when using this
-annotation\footnote{The most common use case of the \texttt{invariant}
+annotation\footnote{The most common use case of the \texttt{elidable}
 annotation is indeed to declare the immutability of fields. Because it is so
@@
 
-\subsubsection{Observably invariant Functions}
+\subsubsection{Observably trace-elidable Functions}
 
@@
 Why can't we simply write an analysis to find out that the \texttt{x} fields of the
-\texttt{A} instances is immutable and deduce that \texttt{compute} is an invariant function,
+\texttt{A} instances is immutable and deduce that \texttt{compute} is a trace-elidable function,
 since it only reads the \texttt{x} field and does not have side effects? This might
 be possible in this particular case, but in practice the functions that are
-annotated with the \texttt{invariant} decorator are usually more complex.
+annotated with the \texttt{elidable} decorator are usually more complex.
 The easiest example for this is that of a function that uses memoization to
@@
 side effects, because it changes the memoizing dictionary. However, because this side
-effect is not externally visible, the function is still invariant. This is
+effect is not externally visible, the function is still trace-elidable. This is
 a property that is not easily detectable by analysis.
@@
 two previous sections don't seem to help with the current object model. There is
-no invariant function to be seen, and the instance is not a candidate for promotion,
+no trace-elidable function to be seen, and the instance is not a candidate for promotion,
 because there tend to be many instances.
@@
 This is a common problem when trying to apply hints. Often, the interpreter
-needs a small rewrite to expose the invariant functions and nearly-constant objects
+needs a small rewrite to expose the trace-elidable functions and nearly-constant objects
 that are implicitly there. In the case of instance fields this rewrite is not
@@
 storage list contains the actual field values. Therefore they have to be immutable, which means
-that their \texttt{getindex} method is an invariant function. When a new attribute is added
+that their \texttt{getindex} method is an trace-elidable function. When a new attribute is added
 to an instance, a new map needs to be chosen, which is done with the
-\texttt{add\_attribute} method on the previous map. This function is also invariant,
+\texttt{add\_attribute} method on the previous map. This function is also trace-elidable,
 because it caches all new instances of \texttt{Map} that it creates, to make
@@
 The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
-an invariant function and they have constant arguments. That means that \texttt{index1/2/3}
+a trace-elidable function and they have constant arguments. That means that \texttt{index1/2/3}
 are constant and the guards on them can be removed. All but the first guard on
@@
 
-What we would really like is if the \texttt{Class.find\_method} method were invariant.
+What we would really like is if the \texttt{Class.find\_method} method were trace-elidable.
 But it cannot be, because it is always possible to change the class itself.
@@
 This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
-version)} pair will always be the same, i.e. it is an invariant operation.  To help
+version)} pair will always be the same, i.e. it is a trace-elidable operation.  To help
 the JIT to detect this case, we factor it out in a helper method which is
-explicitly marked as \texttt{@invariant}. The refactored \texttt{Class} can
+explicitly marked as \texttt{@elidable}. The refactored \texttt{Class} can
 be seen in Figure~\ref{fig:version}
@@
 argument but it does not use it at all. Its only purpose is to make the call
-invariant, because when the version object changes, the result of the call might be
+trace-elidable, because when the version object changes, the result of the call might be
 different than the previous one.
@@
 %applicable in more general cases than the one we developed them for. A more
-%abstract view of maps is that of splitting a data-structure into a part that
-%changes slowly, and a part that changes quickly. In the concrete example of maps
-%we split the original dictionary into the map (the slow-changing part) and the
-%storage array (the quick-changing part). All the computation on the
-%slow-changing part can be constant-folded during tracing so that only the
-%manipulation of the quick-changing part remains.
+%abstract view of maps is that of splitting a data-structure into an immutable part (\eg the map)
+%and a part that changes (\eg the storage array). All the computation on the
+%immutable part is trace-elidable so that only the manipulation of the quick-changing
+%part remains in the trace after optimization.
 %

### 2011-04-14 talk/icooolps2011/paper.tex
go back to old definition and say something about side effects. kill section.

@@
 
-XXX not too happy with the definition
-
 This hint can be used to mark functions as \emph{trace-elidable}. A function is
-termed trace-elidable if, during the execution of the program, the results of
-subsequent calls to the function with identical arguments may be be replaced
-with the result of the first call without changing the program's behaviour.
-From this
-definition follows that a call to an trace-elidable function with constant arguments
-in a trace can be replaced with the result of the call.\footnote{This property
-is less strict than that of a "pure" function, because it is only about actual
-calls during execution. All pure functions are trace-elidable though.}
+termed trace-elidable if, during the execution of the program,
+successive calls to the function with identical arguments always return the
+same result. In addition the function needs to have no side effects or
+idempotent side effects\footnote{This property
+is less strict than that of a pure function, because it is only about actual
+calls during execution. All pure functions are trace-elidable though.}.
+From this definition follows that a call to an trace-elidable function with
+constant arguments in a trace can be replaced with the result of the call.
 
@@
 incorrectly to mitigate this problem.
-
-\subsubsection{Observably trace-elidable Functions}
-
-\cfbolz{XXX do we kill this section?}
-
-Why can't we simply write an analysis to find out that the \texttt{x} fields of the
-\texttt{A} instances is immutable and deduce that \texttt{compute} is a trace-elidable function,
-since it only reads the \texttt{x} field and does not have side effects? This might
-be possible in this particular case, but in practice the functions that are
-annotated with the \texttt{elidable} decorator are usually more complex.
-The easiest example for this is that of a function that uses memoization to
-cache its results. If this function is analyzed, it looks like the function has
-side effects, because it changes the memoizing dictionary. However, because this side
-effect is not externally visible, the function is still trace-elidable. This is
-a property that is not easily detectable by analysis.
 

### 2011-04-14 talk/icooolps2011/paper.tex
fix more XXXs, I also found another reference

@@
 
-\cfbolz{XXX kill the next paragraph? the info is repeated in the list below}
-In particular the hints
-influence the constant folding
-optimization. The first hint makes it possible to turn arbitrary
-variables in the trace into constant by feeding back runtime values. The
-second hint allows the definition of additional foldable operations.
-
 Together these hints can be used to express many classic implementation
@@
  how it can be improved using these hints.
+ \item This example also exemplifies general techniques for refactoring code to
+ expose likely runtime constants constant folding opportunities.
 \end{itemize}
@@
 presents benchmarks.
-
-\cfbolz{XXX stress more that "the crux of the techniques and a significant
-portion of new contributions in the paper are from how to refactoring codes to
-expose likely runtime constants and trace-elidable functions"}
 
@@
 into a constant value. This process is called \emph{promotion} and it is an old idea
-in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993} there). Promotion is also heavily
-used by Psyco \cite{rigo_representation-based_2004} and by all older versions
-of PyPy's JIT. It is a technique that only works well in JIT compilers;
-in static compilers it is significantly less applicable.
+in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993}
+there). The technique is substantially more powerful in a JIT compiler than in
+the static setting of classic partial evaluation.
 
@@
 The promotion is turned into a \texttt{guard} operation in the trace. The guard
-captures the value of $x_1$ as it was during tracing. \cfbolz{drop the word runtime feedback here?}
-From the point of view of the
+captures the value of $x_1$ as it was during tracing. Thus the runtime value of
+\texttt{x} is being made available to the compiler to exploit. The introduced
+guard specializes the trace, because it only works if the value of $x_1$ is
+\texttt{4}. From the point of view of the
 optimizer, this guard is not any different than the one produced by the \texttt{if}
@@
 memory address of the \texttt{Map} instance that has been promoted. Operations
-that can be optimized away are grayed out.
-
-\cfbolz{XXX also explain that some forwarding of guarded values is happening,
-make clearer which figures show optimized code and which show non-optimized
-code}
+that can be optimized away are grayed out, their results will be replaced by
+fixed values by the constant folding.
 
@@
 PyPy's JIT \cite{armin_rigo_jit_2007} as well as a Prolog partial evaluator
-\cite{carl_friedrich_bolz_towards_????}. Promotion is quite similar to
+\cite{carl_friedrich_bolz_towards_????}. Promotion is also heavily
+used by Psyco \cite{rigo_representation-based_2004} (promotion is called
+"unlifting" in this paper) a method-based JIT compiler for Python written by
+one of the authors. Promotion was also used in DyC \cite{grant_dyc:_2000}, a
+runtime partial evaluator for C. Promotion is quite similar to
 (polymorphic) inline caching and runtime type feedback techniques which were

### 2011-04-14 talk/icooolps2011/paper.tex
write acknowledgements
ruthlessly cut space, mostly by making font smaller. not pretty, but I have no idea what to cut atm.

@@
 \usepackage[T1]{fontenc}
-\usepackage[scaled=0.8]{beramono}
+\usepackage[scaled=0.82]{beramono}
 
@@
 Python implementation, but also a Prolog interpreter
-\cite{carl_friedrich_bolz_towards_2010} and a Smalltalk VM
-\cite{carl_friedrich_bolz_back_2008}.
+\cite{carl_friedrich_bolz_towards_2010}.
 
@@
 \begin{figure}
-\includegraphics[scale=0.5]{figures/trace-levels}
+\includegraphics[scale=0.45]{figures/trace-levels}
 \caption{The levels involved in tracing}
@@
 
+{\smaller
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 %
+{\smaller
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 Let's make this more concrete. If we trace a call to the following function:
+{\smaller
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 case, we can add a hint to promote \texttt{x}, like this:
+{\smaller
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 class as follows:
+{\smaller
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 \begin{center}
-{\footnotesize
+{\smaller
 \begin{tabular}{|l|r|r|r|}
@@
 We already explored promotion in other context, such as earlier versions of
-PyPy's JIT \cite{armin_rigo_jit_2007} as well as a Prolog partial evaluator
-\cite{carl_friedrich_bolz_towards_????}. Promotion is also heavily
+PyPy's JIT as well as a Prolog partial evaluator
+\cite{bolz_towards_2009}. Promotion is also heavily
 used by Psyco \cite{rigo_representation-based_2004} (promotion is called
@@
 
-XXX Peng Wu and David Edelsohn, Laura Creighton
+The authors would like to thank Peng Wu, David Edelsohn and Laura Creighton for
+encouragement, fruitful discussions and feedback during the writing of this paper.
 

### 2011-04-14 talk/icooolps2011/paper.tex
change abstract a bit

@@
 languages without explicitly encoding language semantics into the compiler. So
-far, they lacked a way to feed back runtime information into the
-compiler, which restricted their performance. In this paper we describe the
-flexible mechanisms in PyPy's meta-tracing JIT that can be used to control runtime feedback in language-specific ways. These mechanisms are flexible
-enough to implement classical VM techniques such as maps and polymorphic inline
+far, they lacked a way to give the language implementor control over runtime
+feedback. This restricted their performance. In this paper we describe the
+mechanisms in PyPy's meta-tracing JIT that can be used to control
+runtime feedback in language-specific ways. These mechanisms are flexible
+enough to express classical VM techniques such as maps and polymorphic inline
 caches.
-
-\cfbolz{XXX tracing is runtime feed back too, clarify}
 

### 2011-04-14 talk/icooolps2011/paper.tex
enable preprint option
will ignore this XXX
this is repetition

@@
 
-\cfbolz{XXX more precise definition of what promote does}
-
 In this section we will describe how to add two hints that allow the
@@
 
-The PyPy JIT generator automatically detects the majority of these conditions.
-However, for the cases in which the automatic detection does not work, the
-interpreter author can apply \textbf{hints} to improve the optimization
-opportunities. There is one kind of hint for both of the conditions above.
+There is one kind of hint for both of these conditions.
 

### 2011-04-14 talk/icooolps2011/paper.tex
a slightly different approach to saving space: for the small examples, put code and trace next to each other. this makes it possible to use the larger font again.

@@
 \usepackage{color}
+\usepackage{wrapfig}
 \usepackage{ulem}
@@
 \usepackage[T1]{fontenc}
-\usepackage[scaled=0.82]{beramono}
+\usepackage[scaled=0.81]{beramono}
 
@@
 \newcommand{\commentout}[1]{}
+
+\newcommand{\noop}{}
 
@@
 
-{\smaller
+{\noop
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 is not a constant in the original source code. For example, consider the
-following fragment of RPython code:
+following fragment of RPython code on the left. If the fragment is traced with
+$x_1$ being \texttt{4}, the trace on the left is produced:
 
-{\smaller
+\begin{minipage}[b]{0.5\linewidth}
+\centering
+{\noop
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 }
-
-If the fragment is traced with $x_1$ being \texttt{4}, the following trace is
-produced:
-%
-{\smaller
+\end{minipage}
+\vline
+\hspace{0.5cm}
+\begin{minipage}[b]{0.5\linewidth}
+{\noop
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 }
+\end{minipage}
 
-In the trace above, the value of $x_1$ is statically known after the guard.
+In the trace, the value of $x_1$ is statically known after the guard.
 Remember that a guard is a runtime check. The above trace will run to
@@
 
-Let's make this more concrete. If we trace a call to the following function:
-{\smaller
+Let's make this more concrete. If we trace a call to the function on the left, we get the trace on the right:
+
+\begin{minipage}[b]{0.5\linewidth}
+\centering
+{\noop
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 }
-
-We get a trace that looks like this:
-
-{\smaller
+\end{minipage}
+\vline
+\hspace{0.5cm}
+\begin{minipage}[b]{0.5\linewidth}
+{\noop
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 }
+\end{minipage}
 
@@
 case, we can add a hint to promote \texttt{x}, like this:
-{\smaller
+
+\begin{minipage}[b]{0.5\linewidth}
+\centering
+{\noop
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 }
+\end{minipage}
+\vline
+\hspace{0.5cm}
+\begin{minipage}[b]{0.5\linewidth}
+{\noop
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+guard($x_1$ == 4)
+$v_1$ = $x_1$ * 2
+$z_1$ = $v_1$ + 1
+$v_2$ = $z_1$ + $y_1$
+return($v_2$)
+\end{lstlisting}
+}
+\end{minipage}
 
@@
 the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
-operation at the beginning:
-
-{\smaller
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($x_1$ == 4)
-$v_1$ = $x_1$ * 2
-$z_1$ = $v_1$ + 1
-$v_2$ = $z_1$ + $y_1$
-return($v_2$)
-\end{lstlisting}
-}
+operation at the beginning.
 
@@
 
-As an example, take the following class:
+As an example, take the class on the left. Tracing the call \texttt{a.f(10)} of
+some instance of \texttt{A} yields the trace on the right (note how the call to
+\texttt{c} is inlined):
 
-{\smaller
+\begin{minipage}[t]{0.6\linewidth}
+\centering
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
     def f(self, val):
-        self.y = self.compute() + val
+        self.y = self.c() + val
 
-    def compute(self):
+    def c(self):
         return self.x * 2 + 1
 \end{lstlisting}
-}
-
-Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
-trace (note how the call to \texttt{compute} is inlined):
-%
-{\smaller
+\end{minipage}
+\vline
+\hspace{0.5cm}
+\begin{minipage}[t]{0.4\linewidth}
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
 \end{lstlisting}
-}
+\end{minipage}
 
@@
 this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
-immutable, and therefore \texttt{compute}
+immutable, and therefore \texttt{c}
 is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
-If the code in \texttt{compute} should be constant-folded away, we would change the
+If the code in \texttt{c} should be constant-folded away, we would change the
 class as follows:
-{\smaller
+
+\begin{minipage}[t]{0.6\linewidth}
+\centering
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
@@
         promote(self)
-        self.y = self.compute() + val
+        self.y = self.c() + val
 
     @elidable
-    def compute(self):
+    def c(self):
         return self.x * 2 + 1
 \end{lstlisting}
-}
-
-Now the trace will look like this:
-%
-{\smaller
+\end{minipage}
+\vline
+\hspace{0.4cm}
+\begin{minipage}[t]{0.4\linewidth}
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($a_1$ == 0xb73984a8)
-$v_1$ = compute($a_1$)
+guard($a_1$ == 
+      0xb73984a8)
+$v_1$ = c($a_1$)
 $v_2$ = $v_1$ + $val_1$
@@
 \end{lstlisting}
-}
+\end{minipage}
 
 Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
-during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
-has a chance to see it. Since the \texttt{compute} function is marked as trace-elidable, and its
+during tracing. The call to \texttt{c} is not inlined, so that the optimizer
+has a chance to see it. Since the \texttt{c} function is marked as trace-elidable, and its
 argument
@@
 %
-{\smaller
+{\noop
 \begin{lstlisting}[mathescape,basicstyle=\ttfamily]

### 2011-04-15 talk/icooolps2011/paper.tex
some typos and problems I found while reading again

@@
 Conceptually, the significant speed-ups that can be achieved with
-dynamic compilation depend on feeding into compilation and exploiting
-values observed at runtime. In particular, if
+dynamic compilation depend on feeding into compilation values observed at
+runtime and exploiting them. In particular, if
 there are values which vary very slowly, it is possible to compile multiple
@@
  \item This example also exemplifies general techniques for refactoring code to
- expose likely runtime constants constant folding opportunities.
+ expose constant folding opportunities of likely runtime constants.
 \end{itemize}
@@
 Python implementation, but also a Prolog interpreter
-\cite{carl_friedrich_bolz_towards_2010}.
+\cite{carl_friedrich_bolz_towards_2010} and some less mature experiments.
 
@@
 A recently popular approach to JIT compilers is that of tracing JITs. Tracing
-JITs have their origin in the Dynamo project, which used one of them for dynamic
-assembler optimization \cite{bala_dynamo:_2000}. Later they were used to implement
+JITs have their origin in the Dynamo project, which used the technique for dynamic
+machine code optimization \cite{bala_dynamo:_2000}. Later they were used to implement
 a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
@@
 started the interpreter is used; only the most frequently executed paths through the user
-program are turned into machine code. The interpreter is also used when a guard
+program are traced and turned into machine code. The interpreter is also used when a guard
 fails to continue the execution from the failing guard.
@@
 just picks them up from the interpreter. This also means that the JIT by
-construction supports the full language.
+construction supports the full language as correctly as the interpreter.
 
@@
 Instances have a class. Instances have their own attributes (or fields). When looking up an
-attribute on an instance, the instances attributes are searched. If the
+attribute on an instance, the instance's attributes are searched. If the
 attribute is not found there, the class' methods are searched.
@@
 
-In this section we will describe how to add two hints that allow the
+In this section we will describe two hints that allow the
 interpreter author to increase the optimization opportunities for constant
@@
 following fragment of RPython code on the left. If the fragment is traced with
-$x_1$ being \texttt{4}, the trace on the left is produced:
+\texttt{x} being \texttt{4}, the trace on the right is produced:
 
@@
 
-In the trace, the value of $x_1$ is statically known after the guard.
-Remember that a guard is a runtime check. The above trace will run to
-completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
-stopped and the interpreter continues to run.
+A guard is a runtime check. The above trace will run to completion when $x_1$
+\texttt{== 4}. If the check fails, execution of the trace is stopped and the
+interpreter continues to run. Therefore, the value of $x_1$ is statically known
+to be \texttt{4} after the guard.
 
@@
 typical reason to do that is if there is
-a lot of computation depending on the value of that variable.
-
-Let's make this more concrete. If we trace a call to the function on the left, we get the trace on the right:
+a lot of computation depending on the value of one variable.
+
+Let's make this more concrete. If we trace a call to the function (written in
+RPython) on the left, we get the trace on the right:
 
@@
 Observe how the first two operations could be constant-folded if the value of
-$x_1$ were known. Let's assume that the value of \texttt{x} in the Python code can vary, but does so
+$x_1$ were known. Let's assume that the value of \texttt{x} in the RPython code can vary, but does so
 rarely, i.e. only takes a few different values at runtime. If this is the
@@
 The promotion is turned into a \texttt{guard} operation in the trace. The guard
-captures the value of $x_1$ as it was during tracing. Thus the runtime value of
-\texttt{x} is being made available to the compiler to exploit. The introduced
+captures the runtime value of \texttt{x} as it was during tracing, which can
+then be exploited by the compiler. The introduced
 guard specializes the trace, because it only works if the value of $x_1$ is
@@
 optimizer, this guard is not any different than the one produced by the \texttt{if}
-statement in the example above. After the guard, the rest of the trace can
+statement in the first example. After the guard, the rest of the trace can
 assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
@@
 foldable operations on these constants can be constant-folded. This works well for
-constant folding of simple types, e.g. integers. Unfortunately, in the context of an
+constant folding of primitive types, e.g. integers. Unfortunately, in the context of an
 interpreter for a dynamic
-language, most operations actually manipulate objects, not simple types. The
+language, most operations actually manipulate objects, not primitive types. The
 operations on objects are often not foldable and might even have side-effects. If
@@
 From this definition follows that a call to an trace-elidable function with
-constant arguments in a trace can be replaced with the result of the call.
+constant arguments in a trace can be replaced with the result of the call seen during tracing.
 
@@
 immutable, and therefore \texttt{c}
-is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
+is an trace-elidable function. To communicate this, there is an \texttt{@elidable} decorator.
 If the code in \texttt{c} should be constant-folded away, we would change the
@@
       0xb73984a8)
-$v_1$ = c($a_1$)
+$v_1$ = A.c($a_1$)
 $v_2$ = $v_1$ + $val_1$
@@
 Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
-during tracing. The call to \texttt{c} is not inlined, so that the optimizer
-has a chance to see it. Since the \texttt{c} function is marked as trace-elidable, and its
+during tracing. The call to \texttt{A.c} is not inlined, so that the optimizer
+has a chance to see it. Since the \texttt{A.c} method is marked as trace-elidable, and its
 argument
 is a constant reference, the call will be removed by the optimizer. The final
-trace looks like this:
+trace looks like this (assuming that the \texttt{x} field's value is \texttt{4}):
 %
@@
 
-(assuming that the \texttt{x} field's value is \texttt{4}).
-
-On the one hand, the \texttt{elidable} annotation is very powerful. It can be
+On the one hand, the \texttt{@elidable} annotation is very powerful. It can be
 used to constant-fold arbitrary parts of the computation in the interpreter.
@@
 subtly wrong code. Therefore, a lot of care has to be taken when using this
-annotation\footnote{The most common use case of the \texttt{elidable}
+annotation\footnote{The most common use case of the \texttt{@elidable}
 annotation is indeed to declare the immutability of fields. Because it is so
@@
 reference to a map, which maps field names to indexes into a storage list. The
-storage list contains the actual field values. Therefore they have to be immutable, which means
+storage list contains the actual field values. Maps are shared between
+different instances, therefore they have to be immutable, which means
 that their \texttt{getindex} method is an trace-elidable function. When a new attribute is added
@@
 because it caches all new instances of \texttt{Map} that it creates, to make
-sure that objects with the same layout have the same map. Now that we have
+sure that objects with the same layout have the same map, which makes its side
+effects idempotent. Now that we have
 introduced maps, it is safe to promote the map everywhere, because we assume
@@
 
-With this changed instance implementation, the trace we had above changes to the
-following that of see Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
+With this changed instance implementation, the trace we saw in Section~\ref{sub:running} changes to the
+that of Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
 memory address of the \texttt{Map} instance that has been promoted. Operations
-that can be optimized away are grayed out, their results will be replaced by
+that can be optimized away are grayed out, their results will be replaced with
 fixed values by the constant folding.
@@
 The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
-a trace-elidable function and they have constant arguments. That means that \texttt{index1/2/3}
+a trace-elidable function and they have constant arguments. That means that $index_{1/2/3}$
 are constant and the guards on them can be removed. All but the first guard on
@@
 the original one. Now we are down from five dictionary lookups to just two.
+
+XXX separation of fast and slow-changing parts
 
@@
 
-Instances were optimized making the assumption that the total number of
+Instances were optimized by making the assumption that the total number of
 different instance layouts is small compared to the number of instances. For classes we
@@
 counters or similar things) but for this simple example it is good
-enough.\footnote{There is a more complex variant of class versions that can
-accommodate class fields that change a lot better.}
+enough.\footnote{There is a more complex variant of the presented technique that can
+accommodate quick-changing class fields a lot better.}
 
@@
 version)} pair will always be the same, i.e. it is a trace-elidable operation.  To help
-the JIT to detect this case, we factor it out in a helper method which is
-explicitly marked as \texttt{@elidable}. The refactored \texttt{Class} can
+the JIT to detect this case, we factor it out in a helper method \texttt{\_find\_method} which is
+marked as \texttt{@elidable}. The refactored \texttt{Class} can
 be seen in Figure~\ref{fig:version}
@@
 
-The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
+The index \texttt{0} that is used to read out of the \texttt{storage} list is the result
 of the constant-folded \texttt{getindex} call.
@@
 with its class. In our code above, we allow both to vary independently.
-Therefore we store the class of an instance on the map in PyPy's Python
-interpreter. This means that we get one fewer promotion (and thus one fewer
-guard) in the trace, because the class doesn't need to be promoted after the
+In PyPy's Python interpreter we store the class of an instance on its map. This
+means that we get one fewer promotion and thus one fewer
+guard in the trace, because the class doesn't need to be promoted after the
 map has been.
@@
 %abstract view of maps is that of splitting a data-structure into an immutable part (\eg the map)
-%and a part that changes (\eg the storage array). All the computation on the
+%and a part that changes (\eg the storage list). All the computation on the
 %immutable part is trace-elidable so that only the manipulation of the quick-changing
@@
 disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
-Richards benchmark in Python; a Python version of the Telco decimal
+Richards benchmark to Python; a Python version of the Telco decimal
 benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
@@
 
-The optimizations give a speedup between 80\% and almost 20 times. The Richards
+The optimizations give a speedup between 80\% and almost 20 times. The Richards benchmark
 is a particularly good case for the optimizations as it makes heavy uses of
@@
 
-Yermolovich et. al. describe the use of the Tamarin JavaScript tracing JIT as a
+Yermolovich et. al. \cite{yermolovich_optimization_2009} describe the use of the Tamarin JavaScript tracing JIT as a
 meta-tracer for a Lua interpreter. They compile the normal Lua interpreter in C
@@
 strictly ahead of time, it inherently cannot support runtime feedback.
+Some partial evaluator work at runtime, such as DyC \cite{grant_dyc:_2000},
+which also supports a concept similar to promotion (called dynamic-to-static promotion).
 
@@
 "unlifting" in this paper) a method-based JIT compiler for Python written by
-one of the authors. Promotion was also used in DyC \cite{grant_dyc:_2000}, a
-runtime partial evaluator for C. Promotion is quite similar to
+one of the authors. Promotion is quite similar to
 (polymorphic) inline caching and runtime type feedback techniques which were
@@
 encouragement, fruitful discussions and feedback during the writing of this paper.
+This research was partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
+Eureka Eurostars).
 

### 2011-04-15 talk/icooolps2011/paper.tex
:c

@@
 The first step in making \texttt{getattr} faster in our object model is to optimize
-away the dictionary lookups on the instances. The hints we have looked at in the
+away the dictionary lookups on the instances. The hints we have introduced in the
 two previous sections don't seem to help with the current object model. There is
@@
 
-With this changed instance implementation, the trace we saw in Section~\ref{sub:running} changes to the
+With this adapted instance implementation, the trace we saw in Section~\ref{sub:running} changes to the
 that of Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
@@
 
-Instances were optimized by making the assumption that the total number of
+Above we assumed that the total number of
 different instance layouts is small compared to the number of instances. For classes we
 will make an even stronger assumption. We simply assume that it is rare for
-classes to change at all. This is not totally reasonable (sometimes classes contain
+classes to change at all. This is not always reasonable (sometimes classes contain
 counters or similar things) but for this simple example it is good
@@
 \cite{bala_dynamo:_2000} to trace through a small synthetic interpreter. As in Dynamo, tracing
-happens on the machine code level. The tracer is instructed by some hints in the
-tiny interpreter where the main interpreter loop is and for how long to trace to
-match loops in the user-level functions. These hints are comparable to the one
-PyPy uses for the same reasons \cite{bolz_tracing_2009}. Their approach suffers
+happens on the machine code level. The system needs some hints to mark the main
+interpreter loop and where the backward jumps in user programs are.
+PyPy uses similar hints to achieve this \cite{bolz_tracing_2009}. Their approach suffers
 mostly from the low abstraction level that machine code provides.
@@
 strictly ahead of time, it inherently cannot support runtime feedback.
-Some partial evaluator work at runtime, such as DyC \cite{grant_dyc:_2000},
+Some partial evaluators work at runtime, such as DyC \cite{grant_dyc:_2000},
 which also supports a concept similar to promotion (called dynamic-to-static promotion).
@@
 implementation of dynamic languages on top of JVMs easier.
-The bytecode gives access to user accessible generalized inline cache.
+The bytecode gives the user access to generalized inline caches.
 It requires of course compilation to JVM bytecode instead of simply writing an interpreter, predictability of performance across JVMs is also an open question.

### 2011-04-15 talk/icooolps2011/paper.tex
replace one itemize by a sentence, which gives me enough space to resurrect the meta-paragraph.

@@
 
-For constant folding to work, two conditions need to be met:
-
-\begin{itemize}
-    \item the arguments of an operation actually need to all be constant,
-    i.e. statically known by the optimizer
-    \item the operation needs to be \emph{constant-foldable}, i.e. always yield
-    the same result given the same arguments.
-\end{itemize}
-
+For constant folding to work, two conditions need to be met: the arguments of
+an operation actually need to all be constant, i.e. statically known by the
+optimizer and the operation needs to be \emph{constant-foldable}, i.e. always
+yield the same result given the same arguments.
 There is one kind of hint for both of these conditions.
@@
 optimizer.
-
 The simplest example of constants are literal values, such as \texttt{1}.
@@
 The first step in making \texttt{getattr} faster in our object model is to optimize
-away the dictionary lookups on the instances. The hints we have introduced in the
-two previous sections don't seem to help with the current object model. There is
+away the dictionary lookups on the instances. The hints of the previous section
+don't seem to help with the current object model. There is
 no trace-elidable function to be seen, and the instance is not a candidate for promotion,
@@
 
-XXX separation of fast and slow-changing parts
+The technique to make instance lookups faster is applicable in more general
+cases. A more abstract view of maps is that of splitting a data-structure into
+an immutable part (\eg the map) and a part that changes (\eg the storage list).
+All the computation on the immutable part is trace-elidable so that only the
+manipulation of the quick-changing part remains in the trace after
+optimization.
 
@@
 %
-%The techniques we used above to make instance and class lookups faster are
-%applicable in more general cases than the one we developed them for. A more
-%abstract view of maps is that of splitting a data-structure into an immutable part (\eg the map)
-%and a part that changes (\eg the storage list). All the computation on the
-%immutable part is trace-elidable so that only the manipulation of the quick-changing
-%part remains in the trace after optimization.
 %

### 2011-04-15 talk/icooolps2011/paper.tex
small tweaks

@@
 
-We call the code that runs on top of an interpreter implemented with PyPy the
+We call the code that runs on top of a VM implemented with PyPy the
 \emph{user code} or \emph{user program}.
@@
 
-\subsection{Where Do All the Constants Come From}
+\subsection{Where Do All the Constants Come From?}
 
@@
 Section~\ref{sub:running} can be made efficient using the hints described in the
-previous the section. The object model there is typical for many current
+previous section. The object model there is typical for many current
 dynamic languages (such as Python, Ruby and JavaScript) as it relies heavily on

### 2011-04-21 talk/icooolps2011/paper.tex
submitted version
Stephen Simmons cannot make it.  But iko can.

@@
 
-\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^a$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
+\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^{a, c}$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
             Samuele Pedroni$^c$ \and Armin Rigo$^a$}
@@
 incremental compilers, interpreters, run-time environments]
+
+\terms
+Languages, Performance, Experimentation
+
+\keywords{Tracing JIT, Runtime Feedback, Interpreter, Code Generation, Meta-Programming}
 
@@
 We already explored promotion in other context, such as earlier versions of
-PyPy's JIT as well as a Prolog partial evaluator
-\cite{bolz_towards_2009}. Promotion is also heavily
+PyPy's JIT.
+%as well as a Prolog partial evaluator \cite{bolz_towards_2009}
+Promotion is also heavily
 used by Psyco \cite{rigo_representation-based_2004} (promotion is called

### 2011-04-21 talk/pepm2011/presentation/talk.tex
Stephen Simmons cannot make it.  But iko can.

@@
       \item optimize short-lived objects
-      \item remove some of the redundancy
+      \item remove some of the redundant type checks
   \end{itemize}
@@
 set(y, intval, i6)
+return(y)
 \end{verbatim}
@@
 set(y, intval, i6)
+return(y)
 \end{alltt}
@@
 
+\begin{frame}
+  \frametitle{Properties of the Optimization}
+  \begin{itemize}
+      \item output trace is never longer than input trace
+      \item runtime linear in the length of the trace
+  \end{itemize}
+  \pause
+  \begin{block}{Implementation}
+      \begin{itemize}
+          \item about 400 lines of code
+          \item some added complexity over presentation
+          \item objects with arbitrary numbers of fields
+          \item array support
+      \end{itemize}
+  \end{block}
+\end{frame}
+
 \begin{frame}[plain]
@@
       \item to evaluate the optimization we used PyPy's Python interpreter with real-world programs
+      \begin{itemize}
+          \item interpreter is about 30'000 lines of code
+      \end{itemize}
+      \pause
       \item optimization can remove
@@
           \item 93\% of all \texttt{guard} operations
+      \end{itemize}
       \pause
@@
   \end{itemize}
-  \end{itemize}
 \end{frame}
-
-\section{Benchmarks}
 

### 2011-04-21 pypy.org/confrest.py

@@ -1,0 +1,38 @@
+from confrest_oldpy import *
+
+class PyPyPage(Page):
+    def fill(self):
+        super(PyPyPage, self).fill()
+        self.menubar[:] = html.div(
+            html.a("home", href="index.html", class_="menu"), " ",
+            html.a("news", href="news.html", class_="menu"), " ",
+            html.a("consortium", href="consortium.html", class_="menu"), " ",
+            html.a("links", href="links.html", class_="menu"), " ",
+            html.a("community/coding",
+                   href="http://codespeak.net/pypy/dist/pypy/doc/index.html",
+                   class_="menu"), " ",
+            " ", id="menubar")
+
+    def get_doclink(self, target):
+        return relpath(self.targetpath.strpath,
+                       self.project.get_docpath().join(target).strpath)
+
+class Project(Project):
+    mydir = py.magic.autopath().dirpath()
+    title = "PyPy EU Project"
+    stylesheet = 'http://codespeak.net/pypy/dist/pypy/doc/style.css'
+    encoding = 'latin1'
+    prefix_title = "EU/PyPy"
+    logo = html.div(
+        html.a(html.img(alt="PyPy", id="pyimg",
+                        src="http://codespeak.net/pypy/img/py-web1.png",
+                        height=110, width=149)),
+        html.img(alt="EU Logo", id="extraimg",
+                 src="ist.png",
+                 height=105, width=213),
+        )
+    Page = PyPyPage
+    def get_docpath(self):
+        return self.mydir
+
+

### 2011-06-02 talk/icooolps2009/talk/talk.tex
Initial draft of a blog post on the survey results.

@@
 
-\date{ICOOOLPS 2009 XXX}
+\date{6th of July 2009, ICOOOLPS '09, Genova}
 
@@
     \item later used for a lightweight Java JIT
-    \item seems to also work for dynamic languages
-    \item incorporated into Mozilla's JavaScript VM ("TraceMonkey")
+    \item seems to also work for dynamic languages (see TraceMonkey)
     \end{itemize}
@@
         \item automatic inlining
-        \item produces comparatively little code
+        \item produces comparatively little machine code
         \end{itemize}
@@
     \item Assumption violated: consecutive iterations of the dispatch loop will usually take very different code paths
+    \item what can we do?
     \end{itemize}
@@
                 pc = target
-        elif opcode == MOV_A_R:
-            n = ord(bytecode[pc])
-            pc += 1
-            regs[n] = a
         elif opcode == MOV_R_A:
@@
             a = regs[n]
+        elif opcode == MOV_A_R:
+            ...
         elif opcode == ADD_R_TO_A:
-            n = ord(bytecode[pc])
-            pc += 1
-            a += regs[n]
+            ...
         elif opcode == DECR_A:
@@
             return a
+
   \end{verbatim}
@@
   \begin{verbatim}
-def interpret(bytecode, a):             |
-    regs = [0] * 256                    |  # Example bytecode
-    pc = 0                              |  # Square the accumulator:
-    while True:                         |
-        opcode = ord(bytecode[pc])      |  MOV_A_R     0   # i = a
-        pc += 1                         |  MOV_A_R     1   # copy of 'a'
-        if opcode == JUMP_IF_A:         |
-            target = ord(bytecode[pc])  |  # 4:
-            pc += 1                     |  MOV_R_A     0   # i--
-            if a:                       |  DECR_A
-                pc = target             |  MOV_A_R     0
-        elif opcode == MOV_A_R:         |
-            n = ord(bytecode[pc])       |  MOV_R_A     2   # res += a
-            pc += 1                     |  ADD_R_TO_A  1
-            regs[n] = a                 |  MOV_A_R     2
-        elif opcode == MOV_R_A:         |
-            n = ord(bytecode[pc])       |  MOV_R_A     0   # if i!=0:
-            pc += 1                     |  JUMP_IF_A   4   #    goto 4
-            a = regs[n]                 |
-        elif opcode == ADD_R_TO_A:      |  MOV_R_A     2   # return res
-            n = ord(bytecode[pc])       |  RETURN_A
-            pc += 1                     |
-            a += regs[n]                |
+def interpret(bytecode, a):
+    regs = [0] * 256                    |
+    pc = 0                              |  # Example bytecode
+    while True:                         |  # Square the accumulator:
+        opcode = ord(bytecode[pc])      |
+        pc += 1                         |  MOV_A_R     0  # i = a
+        if opcode == JUMP_IF_A:         |  MOV_A_R     1  # copy of 'a'
+            target = ord(bytecode[pc])  |
+            pc += 1                     |  # 4:
+            if a:                       |  MOV_R_A     0  # i--
+                pc = target             |  DECR_A
+        elif opcode == MOV_R_A:         |  MOV_A_R     0
+            n = ord(bytecode[pc])       |
+            pc += 1                     |  MOV_R_A     2  # res += a
+            a = regs[n]                 |  ADD_R_TO_A  1
+        elif opcode == MOV_A_R:         |  MOV_A_R     2
+            ...                         |
+        elif opcode == ADD_R_TO_A:      |  MOV_R_A     0  # if i!=0:
+            ...                         |  JUMP_IF_A   4  #    goto 4
         elif opcode == DECR_A:          |
-            a -= 1                      |
-        elif opcode == RETURN_A:        |
+            a -= 1                      |  MOV_R_A     2  # return res
+        elif opcode == RETURN_A:        |  RETURN_A                     
             return a                    |
+
   \end{verbatim}
@@
     \begin{itemize}
-    \item which variables make up the program counter of the language interpreter
+    \item which variables make up the program counter of the language interpreter (together those are called \emph{position key})
     \item where the bytecode dispatch loop is
@@
 
+\begin{frame}
+    \frametitle{Modifying Tracing}
+    \begin{itemize}
+    \item goal: try to trace the loops \emph{in the user program,}
+          and not just one iteration of the bytecode dispatch loop
+    \item tracing interpreter stops tracing only when:
+        \begin{itemize}
+        \item it sees a backward jump in the language interpreter
+        \item the position key of the language interpreter matches an ealier value
+        \end{itemize}
+    \item in this way, full user loops are traced
+    \end{itemize}
+\end{frame}
 \frame[containsverbatim, plain, shrink=20]{
@@
 # DECR_A
-...
 # MOV_A_R 0
-...
 # MOV_R_A 2
-...
 # ADD_R_TO_A 1
-...
 # MOV_A_R 2
-...
 # MOV_R_A 0
@@
     \item we can apply this approach to PyPy's Python interpreter (70 KLOC)
-    \item speed-ups promising: factor of 6 faster for simple loops with arithmetic
+    \item speed-ups promising (see next slide)
     \item no Python-specific bugs!
@@
 \end{frame}
+
+\frame[containsverbatim, plain, shrink=10]{
+    \frametitle{Timings for Python Interpreter}
+\begin{verbatim}
+def f(a):
+    t = (1, 2, 3)
+    i = 0
+    while i < a:
+        t = (t[1], t[2], t[0])
+        i += t[0]
+    return i
+\end{verbatim}
+\begin{block}{Timings of \texttt{f(10000000)}}
+\begin{tabular}{|l|l|r|r|}
+\hline
+& &Time (ms) &speedup\\
+\hline
+1 &PyPy compiled to C, no JIT &1793 $\pm$ 11 &1.00\\
+2 &PyPy comp'd to C, with JIT &483 $\pm$ 6 &3.71\\
+3 &CPython 2.6 &1869 $\pm$ 11 & 0.96\\
+4 &CPython 2.6 + Psyco 1.6 & 511 $\pm$ 7 &3.51\\\hline
+\end{tabular}
+\end{block}
+}
 
@@
 
-\begin{frame}
-    \frametitle{Backup Slides}
-\end{frame}
-
-\frame[containsverbatim, plain, shrink=10]{
-    \frametitle{Timings for Python Interpreter}
-\begin{verbatim}
-def f(a):
-    t = (1, 2, 3)
-    i = 0
-    while i < a:
-        t = (t[1], t[2], t[0])
-        i += t[0]
-    return i
-\end{verbatim}
-\begin{block}{Timings}
-\begin{tabular}{|l|l|r|r|}
-\hline
-& &Time (s) &speedup\\
-\hline
-1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
-2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
-3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
-4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
-\end{tabular}
-\end{block}
-}
-
 \end{document}

### 2011-06-08 talk/icooolps2011/paper.tex
fix the typos the reviewers pointed out

@@
 
-One of the hardest parts of implementing a dynamic language efficiently is to
-optimize its object model. This is made harder by the fact that many recent
-languages such as Python, JavaScript or Ruby have a rather complex core object
-semantics. For them, even implementing just an interpreter is already a complex
+One of the hardest parts of implementing an object-oriented dynamic language well is to
+optimize its object model. This is made harder by the complexity of the core
+object semantics of many recent languages such as Python, JavaScript or Ruby.
+For them, even implementing just an interpreter is already a difficult
 task. Implementing these languages efficiently with a just-in-time compiler (JIT) is
@@
 A recently popular approach to JIT compilers is that of tracing JITs. Tracing
-JITs have their origin in the Dynamo project, which used the technique for dynamic
+JITs were popularized by the Dynamo project, which used the technique for dynamic
 machine code optimization \cite{bala_dynamo:_2000}. Later they were used to implement
@@
 the \emph{interpreter} that is running the program. This means that the traces
-it produces don't contain the bytecodes of the language in question, but
+it produces do not contain the bytecodes of the language in question, but
 RPython-level operations that the interpreter did to execute the program.
@@
 Instances have a class. Instances have their own attributes (or fields). When looking up an
-attribute on an instance, the instance's attributes are searched. If the
+attribute of an instance, the instance's attributes are searched. If the
 attribute is not found there, the class' methods are searched.
@@
 dictionaries.
-Let's assume we trace through code that sums three attributes, such as:
+Let us assume we trace through code that sums three attributes, such as:
 \anto{I still think it's a bit weird to call them ``methods'' and then use
@@
 using a tracing JIT, we need to find a way to get rid of these dictionary
-lookups somehow. How to achieve this will be topic of
+lookups. How to achieve this will be the topic of
 Section~\ref{sec:fastobjmodel}.
@@
 
-Let's make this more concrete. If we trace a call to the function (written in
+Let us make this more concrete. If we trace a call to the function (written in
 RPython) on the left, we get the trace on the right:
@@
 Observe how the first two operations could be constant-folded if the value of
-$x_1$ were known. Let's assume that the value of \texttt{x} in the RPython code can vary, but does so
+$x_1$ were known. Let us assume that the value of \texttt{x} in the RPython code can vary, but does so
 rarely, i.e. only takes a few different values at runtime. If this is the
@@
 effect. When tracing, some extra work
-is done. Let's assume that this changed function is traced with
+is done. Let us assume that this changed function is traced with
 the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
@@
 \texttt{4}. From the point of view of the
-optimizer, this guard is not any different than the one produced by the \texttt{if}
-statement in the first example. After the guard, the rest of the trace can
-assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
-trace into:
+optimizer, this guard is not different frome the one produced by the \texttt{if}
+statement in the first example. After the guard, it can be assumed that $x_1$
+is equal to \texttt{4}, meaning that the optimizer will turn this trace into:
 
@@
 linking them into a chain. This is clearly not desirable, so we should promote
-only variables that don't vary much. However, adding a promotion hint will never produce wrong
-results. It might just lead to too much assembler code being generated.
+only variables that do not vary much. However, adding a promotion hint will never produce wrong
+results. It might just lead to too much machine code being generated.
 
@@
 calls during execution. All pure functions are trace-elidable though.}.
-From this definition follows that a call to an trace-elidable function with
+From this definition follows that a call to a trace-elidable function with
 constant arguments in a trace can be replaced with the result of the call seen during tracing.
@@
 immutable, and therefore \texttt{c}
-is an trace-elidable function. To communicate this, there is an \texttt{@elidable} decorator.
+is a trace-elidable function. To communicate this, there is an \texttt{@elidable} decorator.
 If the code in \texttt{c} should be constant-folded away, we would change the
@@
 away the dictionary lookups on the instances. The hints of the previous section
-don't seem to help with the current object model. There is
+do not seem to help with the current object model. There is
 no trace-elidable function to be seen, and the instance is not a candidate for promotion,
@@
 different instances, therefore they have to be immutable, which means
-that their \texttt{getindex} method is an trace-elidable function. When a new attribute is added
+that their \texttt{getindex} method is a trace-elidable function. When a new attribute is added
 to an instance, a new map needs to be chosen, which is done with the
@@
 
-With this adapted instance implementation, the trace we saw in Section~\ref{sub:running} changes to the
+With this adapted instance implementation, the trace we saw in Section~\ref{sub:running} changes to
 that of Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
@@
 
-What we would really like is if the \texttt{Class.find\_method} method were trace-elidable.
+What we would really like that the \texttt{Class.find\_method} method is trace-elidable.
 But it cannot be, because it is always possible to change the class itself.
@@
 trace-elidable, because when the version object changes, the result of the call might be
-different than the previous one.
+different from the previous one.
 
@@
 SPUR \cite{bebenita_spur:_2010} is a tracing JIT for CIL bytecode, which is then
-used to trace through an JavaScript implementation written in C\#. The
+used to trace through a JavaScript implementation written in C\#. The
 JavaScript implementation compiles JavaScript to CIL bytecode together with an
@@
 
-We already explored promotion in other context, such as earlier versions of
+We already explored promotion in other contexts, such as earlier versions of
 PyPy's JIT.

### 2011-06-08 talk/icooolps2011/paper.tex
don't stress polymorphic inline caches anymore, focus on runtime feedback.

@@
 runtime feedback in language-specific ways. These mechanisms are flexible
-enough to express classical VM techniques such as maps and polymorphic inline
-caches.
+enough to express classical VM techniques such as maps and runtime type feedback.
 
@@
 Together these hints can be used to express many classic implementation
-techniques used for object models of dynamic languages, such as maps and
-polymorphic inline caches.
+techniques used for object models of dynamic languages, such runtime type
+feedback and maps.
 
@@
 Promotion is essentially a tool for trace specialization. There are places in
-the interpreter where knowing that a value is constant opens a lot of
-optimization opportunities, even though it
-could have different values in practice. In such a place, promotion can be used. The
-typical reason to do that is if there is
-a lot of computation depending on the value of one variable.
+the interpreter where it would open a lot of optimization opportunities if a
+variable were constant, even though it could have different values in
+practice. In such a place, promotion can be used. The typical reason to do that
+is if there is a lot of computation depending on the value of one variable.
 
@@
 \cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
-implementation of dynamic languages on top of JVMs easier.
-The bytecode gives the user access to generalized inline caches.
-It requires of course compilation to JVM bytecode instead of simply writing an interpreter, predictability of performance across JVMs is also an open question.
+implementation of dynamic languages on top of JVMs easier. The bytecode gives
+the user access to generalized inline caches. It requires of course compilation
+to JVM bytecode instead of writing an interpreter.
 
@@
 one of the authors. Promotion is quite similar to
-(polymorphic) inline caching and runtime type feedback techniques which were
+runtime type feedback (and also inline caching) techniques which were
 first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
-\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
-Promotion is more general because any information can be cached in line, not
-just classes of method receivers.
+\cite{hoelzle_optimizing_1994} implementations.
+Promotion is more general because any information can be fed back into
+compilation, not just types.
 

### 2011-06-08 talk/icooolps2011/paper.tex
put another explicit sentence saying that the hints go into the interpreter, not the user code.

@@
 interpreter author to increase the optimization opportunities for constant
-folding. If applied correctly these techniques can give really big speedups by
+folding.
+If applied correctly these techniques can give really big speedups by
 pre-computing parts of what happens at runtime. On the other
 hand, if applied incorrectly they might lead to code bloat, thus making the
-resulting program actually slower.
+resulting program actually slower. Note that these hints are \emph{never} put into the user program, only into the interpreter.
 
@@
 should try to perform runtime specialization on it
-in the code that follows.\footnote{For technical reasons the promote hint needs
-to be written down slightly differently in the actual code.}  When just running
+in the code that follows. When just running
 the code, the \texttt{promote} function has no

### 2011-06-09 talk/iwtc11/paper.tex
stolen example from previous paper and adapted it for our needs. the description of the example probably also needs to be adapted...

--- 2010-11-18 talk/pepm2011/paper.tex
@@
-\documentclass{sigplanconf}
+%-----------------------------------------------------------------------------
+%
+%               Template for sigplanconf LaTeX Class
+%
+% Name:         sigplanconf-template.tex
+%
+% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
+%               file for SIGPLAN conference proceedings.
+%
+% Author:       Paul C. Anagnostopoulos
+%               Windfall Software
+%               978 371-2316
+%               paul@windfall.com
+%
+% Created:      15 February 2005
+%
+%-----------------------------------------------------------------------------
 
-\usepackage{ifthen}
-\usepackage{fancyvrb}
-\usepackage{color}
-\usepackage{ulem}
-\usepackage{xspace}
-\usepackage{epsfig}
-\usepackage{amssymb}
+\documentclass[preprint]{sigplanconf}
+
+% The following \documentclass options may be useful:
+%
+% 10pt          To set in 10-point type instead of 9-point.
+% 11pt          To set in 11-point type instead of 9-point.
+% authoryear    To obtain author/year citation style instead of numeric.
+
 \usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage[utf8]{inputenc}
 \usepackage{setspace}
-
 \usepackage{listings}
 
-\usepackage[T1]{fontenc}
-\usepackage{beramono}
+\begin{document}
 
-\definecolor{gray}{rgb}{0.3,0.3,0.3}
+\conferenceinfo{WXYZ '05}{date, City.} 
+\copyrightyear{2005} 
+\copyrightdata{[to be supplied]}
 
-\lstset{
-  basicstyle=\setstretch{1.05}\ttfamily\footnotesize,
-  language=Python,
-  keywordstyle=\bfseries,
-  stringstyle=\color{blue},
-  commentstyle=\color{gray}\textit,
-  fancyvrb=true,
-  showstringspaces=false,
-  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
-  numberstyle = \tiny,
-  numbersep = -20pt,
-}
+\titlebanner{banner above paper title}        % These are ignored unless
+\preprintfooter{short description of paper}   % 'preprint' option specified.
 
-\newboolean{showcomments}
-\setboolean{showcomments}{true}
-\ifthenelse{\boolean{showcomments}}
-  {\newcommand{\nb}[2]{
-    \fbox{\bfseries\sffamily\scriptsize#1}
-    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
-   }
-   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
-  }
-  {\newcommand{\nb}[2]{}
-   \newcommand{\version}{}
-  }
+\title{Title Text}
+\subtitle{Subtitle Text, if any}
 
-\newcommand\cfbolz[1]{\nb{CFB}{#1}}
-\newcommand\arigo[1]{\nb{AR}{#1}}
-\newcommand\fijal[1]{\nb{FIJAL}{#1}}
-\newcommand\david[1]{\nb{DAVID}{#1}}
-\newcommand\anto[1]{\nb{ANTO}{#1}}
-\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
-\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
-\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
-\newcommand{\commentout}[1]{}
-\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
-
-\newcommand\ie{i.e.,\xspace}
-\newcommand\eg{e.g.,\xspace}
-\newcommand{\etal}{\emph{et al.}\xspace}
-
-\normalem
-
-\let\oldcite=\cite
-
-\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}
-
-%
-\def\sharedaffiliation{%
-\end{tabular}
-\begin{tabular}{c}}
-%
-\begin{document}
-\conferenceinfo{PEPM'11,} {January 24--25, 2011, Austin, Texas, USA.}
-\CopyrightYear{2011}
-\copyrightdata{978-1-4503-0485-6/11/01}
-
-\title{Allocation Removal by Partial Evaluation in a Tracing JIT}
-
-\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^a$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
-            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
-           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
-
-            $^b$merlinux GmbH, Hildesheim, Germany
-
-            $^c$Open End, Göteborg, Sweden
-           }
-           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu \and
-           leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}
-
-%\numberofauthors{3}
-%\author{
-%\alignauthor Carl Friedrich Bolz\\
-%       \email{cfbolz@gmx.de}
-%\alignauthor Michael Leuschel\\
-%       \email{leuschel@cs.uni-duesseldorf.de}
-%\alignauthor David Schneider\\
-%      \email{david.schneider@uni-duesseldorf.de}
-%      \sharedaffiliation
-%      \affaddr{Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany}\\
-%}
+\authorinfo{Name1}
+           {Affiliation1}
+           {Email1}
+\authorinfo{Name2\and Name3}
+           {Affiliation2/3}
+           {Email2/3}
 
 \maketitle
+
 \begin{abstract}
-The performance of many dynamic language implementations suffers from
-high allocation rates and runtime type checks.  This makes dynamic
-languages less applicable to purely algorithmic problems, despite their
-growing popularity.  In this paper we present a simple compiler optimization
-based on online partial evaluation to remove object allocations and
-runtime type checks in the context of a tracing JIT.  We evaluate the
-optimization using a Python VM and find that it gives good results for
-all our (real-life) benchmarks.
-\footnote{This research is partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
-Eureka Eurostars).}
+This is the text of the abstract.
 \end{abstract}
 
-% A category with the (minimum) three required fields
-%\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-\category{D.3.4}{Programming Languages}{Processors}[code generation,
-interpreters, run-time environments]
+\category{CR-number}{subcategory}{third-level}
 
 \terms
-Languages, Performance, Experimentation
+term1, term2
 
-\keywords{Tracing JIT, Partial Evaluation, Optimization}
+\keywords
+keyword1, keyword2
 
@@
 
-The objective of a just-in-time (JIT) compiler for a dynamic language is to
-improve the speed of the language over an implementation of the language that
-uses interpretation. The first goal of a JIT is therefore to remove the
-interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
-overhead of the interpreter's data structures, such as operand stack etc. The
-second important problem that any JIT for a dynamic language needs to solve is
-how to deal with the overhead of boxing primitive types and of type
-dispatching. Those are problems that are usually not present or at least less
-severe in statically typed languages.
-
-Boxing of primitive types is necessary because dynamic languages need to be able to handle
-all objects, even integers, floats, booleans etc. in the same way as user-defined
-instances. Thus those primitive types are usually \emph{boxed}, \ie a small
-heap-structure is allocated for them that contains the actual value. Boxing
-primitive types can be very costly, because a lot of common operations,
-particularly all arithmetic operations, have to produce new boxes, in addition
-to the actual computation they do. Because the boxes are allocated on the heap,
-producing many of them puts pressure on the garbage collector.
-
-Type dispatching is the process of finding the concrete implementation that is
-applicable to the objects at hand when performing a generic operation on them. An
-example would be the addition of two objects: For addition the types of the
-concrete objects need to be checked and the suiting implementation chosen.
-Type dispatching is a very common operation in
-modern\footnote{For languages in the LISP family, basic arithmetic operations
-are typically not overloaded; even in Smalltalk, type dispatching is much
-simpler than in Python or JavaScript.}
-dynamic languages because no types are known at compile time. Therefore all
-operations need it.
-
-A recently popular approach to implementing just-in-time compilers for dynamic
-languages is that of a tracing JIT. A tracing JIT works by observing the running
-program and recording its commonly executed parts into \emph{linear execution traces}. Those
-traces are optimized and turned into machine code.
-
-One reason for the popularity of tracing JITs is their relative
-simplicity. They can often be added to an existing interpreter, reusing a lot of
-the interpreter's infrastructure. They give some important
-optimizations like inlining and constant-folding for free. A tracing JIT always
-produces linear pieces of code, which simplifies many of the hard algorithms in
-a compiler, such as register allocation.
-
-The use of a tracing JIT can remove the overhead of bytecode dispatch and that
-of the interpreter data structures. In this paper we want to present a new
-optimization that can be added to a tracing JIT that further removes some of the
-overhead more closely associated to dynamic languages, such as boxing overhead
-and type dispatching. Our experimental platform is the PyPy project, which is an
-environment for implementing dynamic programming languages. PyPy and tracing
-JITs are described in more detail in Section~\ref{sec:Background}.
-Section~\ref{sec:lifetimes} analyzes the problem to be solved more closely.
-
-The core of our trace optimization technique can be
-viewed as partial evaluation: the partial evaluation
-performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and makes some
-objects that are allocated in the trace \emph{static,} which
-means that they do not occur any more in the optimized trace. This technique is
-informally described in Section~\ref{sec:statics}; a more formal description is
-given in Section~\ref{sec:formal}. The introduced
-techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
-interpreter.
-
-The contributions made by this paper are:
-
-\begin{enumerate}
-    \item A description of a practical, efficient and effective algorithm for
-    removing object allocations in a tracing JIT.
-    \item A characterization of this algorithm as partial evaluation.
-    \item Performance benchmarks for this algorithm.
-\end{enumerate}
-
-\section{Background}
-\label{sec:Background}
-
-\subsection{PyPy}
-\label{sub:PyPy}
-
-The work described in this paper was done in the context of the PyPy project\footnote{\texttt{http://pypy.org}}
-\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
-be implemented in a simple yet efficient way. 
-When implementing a language with PyPy one writes an \emph{interpreter}
-for the language in
-\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
-is a subset of Python chosen in such a way that type inference becomes
-possible. The language interpreter can then be compiled (``translated'') with
-PyPy's tools into a VM on C level. During translation to C, many low-level
-aspects of the final VM, such as object layout, garbage collection and memory
-model, are woven into the generated code. Therefore the interpreter itself can
-remain at a relatively high level of abstraction.
-
-A number of languages have been implemented with PyPy. The project was initiated
-to get a better Python implementation, which inspired the name of the project
-and is still the main focus of development. In addition a number of other
-languages were implemented, among them a Prolog interpreter
-\cite{carl_friedrich_bolz_towards_2010}, a Smalltalk VM
-\cite{carl_friedrich_bolz_back_2008} and a GameBoy emulator
-\cite{bruni_pygirl:_2009}.
-
-The feature that makes PyPy more than a compiler with a runtime system is its
-support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
-the translation to C, PyPy's tools can generate a tracing just-in-time compiler for the
-language that the interpreter is implementing. This process is mostly
-automatic; it only needs to be guided by the language implementer using a small number of
-source-code hints in the interpreter. Mostly-automatically generating a JIT compiler has many advantages
-over writing one manually, an error-prone and tedious process.
-By construction, the generated JIT has the same semantics as the interpreter.
-Optimizations can be shared between different languages implemented with PyPy.
-
-Moreover, thanks to the internal design of the JIT generator, it is very easy
-to add new \emph{backends} for producing the actual machine code.  Examples of
-JIT backends that are implemented are those for Intel x86 and x86-64 and an
-experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.
-
-\subsection{Tracing JIT Compilers}
-\label{sub:JIT_background}
-
-Tracing JITs are a recently popular approach to write just-in-time compilers for
-dynamic languages. Their origins lie in the Dynamo project, which used a tracing
-approach to optimize machine code using execution traces
-\cite{bala_dynamo:_2000}. Tracing JITs have then be adapted to be used for a
-very light-weight Java VM \cite{gal_hotpathvm:_2006} and afterwards used in
-several implementations of dynamic languages, such as JavaScript
-\cite{andreas_gal_trace-based_2009}, Lua\footnote{\texttt{http://luajit.org/}}
-and now Python (and other languages) via PyPy.
-
-The core idea of tracing JITs is to focus the optimization effort of the JIT
-compiler on the commonly executed, \ie \emph{hot} paths of the core loops of the program and to just use an
-interpreter for the less commonly executed parts. VMs that use a tracing JIT are
-mostly mixed-mode execution environments, they contain both an interpreter and a
-JIT compiler. By default the interpreter is used to execute the program, doing
-some light-weight profiling at the same time. This profiling is used to identify
-the hot loops of the program. If a hot loop is found in that way, the
-interpreter enters a special \emph{tracing mode}. In this tracing mode, the
-interpreter tries to record all operations that it is executing while running one
-iteration of the hot loop. This history of executed operations of one loop is
-called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
-it always ends with a jump to its own beginning. The trace also contains all
-operations that are performed in functions that were called in the loop, thus a
-tracing JIT automatically performs inlining.
-This trace of operations subsequently forms the basis of the generated code. The trace is
-first optimized, and then turned into machine code. Both optimization
-and machine code generation are simple, because the traces are linear. This
-linearity makes many optimizations a lot more tractable, and the inlining that
-happens gives the optimizations automatically more context to work with.
-
-Since the trace corresponds to one concrete execution of a loop,
-the code generated from it is only one possible path through the loop.
-To make sure that the trace maintains the correct semantics, it contains a
-\emph{guard} at all places where the execution could have diverged from the
-path. Those guards check the assumptions under which execution can stay on the
-trace. As an example, if a loop contains an if-statement, the trace
-will contain the execution of one of the paths only, which is the path that was
-taken during the production of the trace. The trace will also contain a guard
-that checks that the condition of the if-statement is the same as
-during tracing, because if
-it isn't, the rest of the trace would not be valid.
-
-When generating machine code, every guard is turned into a quick check to
-see whether the assumption still holds. When such a guard is hit during the
-execution of the machine code and the assumption does not hold, the execution of
-the machine code is stopped, and interpreter continues to run from that point
-on. These guards are the only mechanism to stop the execution of a trace, the
-loop end condition also takes the form of a guard.
-
-If one specific guard fails a lot (\ie more than some threshold), the tracing JIT will generate a new
-trace that starts exactly at the position of the failing guard
-\cite{andreas_gal_incremental_2006}. The existing assembler is patched to jump
-to the new trace when the guard fails.  This approach guarantees that all the
-hot paths in the program will eventually be traced and compiled into efficient
-code.
+The text of the paper begins here.
 
@@
 def f(y):
-   res = BoxedInteger(0)
+   step = BoxedInteger(-1)
    while y.is_positive():
-      res = res.add(y).add(BoxedInteger(-100))
-      y = y.add(BoxedInteger(-1))
+      y = y.add(step)
    return res
@@
 # arguments to the trace: $p_{0}$, $p_{1}$
-# inside f: res.add(y)
+# inside f: y = y.add(step)
 guard_class($p_{1}$, BoxedInteger)
@@
             set($p_{5}$, intval, $i_{4}$)
-
-# inside f: BoxedInteger(-100) 
-$p_{6}$ = new(BoxedInteger)
-    # inside BoxedInteger.__init__
-    set($p_{6}$, intval, -100)
-
-# inside f: .add(BoxedInteger(-100))
-guard_class($p_{5}$, BoxedInteger)
-    # inside BoxedInteger.add
-    $i_{7}$ = get($p_{5}$, intval)
-    guard_class($p_{6}$, BoxedInteger)
-        # inside BoxedInteger.add__int
-        $i_{8}$ = get($p_{6}$, intval)
-        $i_{9}$ = int_add($i_{7}$, $i_{8}$)
-        $p_{10}$ = new(BoxedInteger)
-            # inside BoxedInteger.__init__
-            set($p_{10}$, intval, $i_{9}$)
-
-# inside f: BoxedInteger(-1)
-$p_{11}$ = new(BoxedInteger)
-    # inside BoxedInteger.__init__
-    set($p_{11}$, intval, -1)
-
-# inside f: y.add(BoxedInteger(-1))
-guard_class($p_{0}$, BoxedInteger)
-    # inside BoxedInteger.add
-    $i_{12}$ = get($p_{0}$, intval)
-    guard_class($p_{11}$, BoxedInteger)
-        # inside BoxedInteger.add__int
-        $i_{13}$ = get($p_{11}$, intval)
-        $i_{14}$ = int_add($i_{12}$, $i_{13}$)
-        $p_{15}$ = new(BoxedInteger)
-            # inside BoxedInteger.__init__
-            set($p_{15}$, intval, $i_{14}$)
-
-# inside f: y.is_positive()
-guard_class($p_{15}$, BoxedInteger)
-    # inside BoxedInteger.is_positive
-    $i_{16}$ = get($p_{15}$, intval)
-    $i_{17}$ = int_gt($i_{16}$, 0)
-# inside f
-guard_true($i_{17}$)
-jump($p_{15}$, $p_{10}$)
+jump($p_{0}$, $p_{5}$)
 \end{lstlisting}
@@
 
-\section{Object Lifetimes in a Tracing JIT}
-\label{sec:lifetimes}
+\appendix
+\section{Appendix Title}
 
-% section Object Lifetimes in a Tracing JIT (end)
+This is the text of the appendix, if you need one.
 
-To understand the problems that this paper is trying to solve in more detail, we
-first need to understand various cases of object lifetimes that can occur in a
-tracing JIT compiler.
+\acks
 
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.6]{figures/obj-lifetime.pdf}
-\end{center}
+Acknowledgments, if needed.
 
-\caption{Object Lifetimes in a Trace}
-\label{fig:lifetimes}
-\end{figure}
+% We recommend abbrvnat bibliography style.
 
-Figure~\ref{fig:lifetimes} shows a trace before optimization, together with the
-lifetime of various kinds of objects created in the trace. It is executed from
-top to bottom. At the bottom, a jump is used to execute the same loop another
-time (for clarity, the figure shows two iterations of the loop). The loop is
-executed until one of the guards in the trace fails, and the execution is
-aborted and interpretation resumes.
+\bibliographystyle{abbrvnat}
 
-Some of the operations within this trace are \lstinline{new} operations, which each
-create a new instance of some class. These instances are used for some time, \eg
-by calling methods on them (which are inlined into the trace), reading and
-writing their fields. Some of these instances \emph{escape}, which means that
-they are stored in some globally accessible place or are passed into a
-non-inlined function via a residual call.
+% The bibliography should be embedded for final submission.
 
-Together with the \lstinline{new} operations, the figure shows the lifetimes of the
-created objects. The objects that are created within a trace using \lstinline{new}
-fall into one of several categories:
+\begin{thebibliography}{}
+\softraggedright
 
-\begin{enumerate}
-    \item Objects that live for some time, and are then just not
-    used any more afterwards.
+\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
+P. Q. Smith, and X. Y. Jones. ...reference text...
 
-    \item Objects that live for some time and then escape.
-
-    \item Objects that live for some time, survive across the jump to
-    the beginning of the loop, and are then not used any more.
-
-    \item Objects that live for some time, survive across the jump,
-    and then escape. To these we also count the objects that live across several
-    jumps and then either escape or stop being used.
-\end{enumerate}
-
-The objects that are allocated in the example trace in
-Figure~\ref{fig:unopt-trace} fall into categories 1 and 3. Objects stored in
-$p_{5}$, $p_{6}$, $p_{11}$ are in category 1, objects in $p_{10}$, $p_{15}$ are in
-category 3.
-
-The creation of objects in category 1 is removed by the optimization described
-in Sections~\ref{sec:statics} and \ref{sec:formal}. Objects in the other
-categories are partially optimized by this approach as well.\footnote{We also started to
-work on optimizing objects in category 3, which will be the subject of a later
-paper.}
-
-\section{Allocation Removal in Traces}
-\label{sec:statics}
-
-The main insight to improve the code shown in Section~\ref{sub:example} is that objects
-in category 1 do not survive very long -- they are used only inside the loop and
-there is no other outside reference to them. The optimizer identifies objects in
-category 1 and removes the allocation of these objects, and all operations
-manipulating them.
-
-This is a process that is usually called \emph{escape analysis}
-\cite{goldberg_higher_1990}. In this paper we will perform escape analysis by
-using partial evaluation. The use of partial evaluation is a bit peculiar in
-that it receives no static input arguments for the trace, but it is only used to
-optimize operations within the trace. This section will give an informal account
-of this process by examining the example trace in Figure~\ref{fig:unopt-trace}.
-The final trace after optimization can be seen in Figure~\ref{fig:step1} (the
-line numbers are the lines of the unoptimized trace where the operation originates).
-
-To optimize the trace, it is traversed from beginning to end and an output
-trace is produced. Every operation in the input trace is either
-removed or copied into the output trace. Sometimes new operations need to be
-produced as well. The optimizer can only remove operations that manipulate
-objects that have been allocated within the trace, while all other operations are copied to the
-output trace unchanged.
-
-Looking at the example trace of Figure~\ref{fig:unopt-trace}, the operations
-in lines 1--9 are manipulating objects which existed before the trace and that
-are passed in as arguments: therefore the optimizer just copies them into the
-output trace.
-
-The following operations (lines 10--17) are more interesting:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=10,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$p_{5}$ = new(BoxedInteger)   |\setcounter{lstnumber}{11}|
-set($p_{5}$, intval, $i_{4}$) |\setcounter{lstnumber}{14}|
-$p_{6}$ = new(BoxedInteger)   |\setcounter{lstnumber}{16}|
-set($p_{6}$, intval, -100)
-\end{lstlisting}
-
-When the optimizer encounters a \lstinline{new}, it removes it optimistically,
-and assumes that the object is in category 1. If later the optimizer finds that the object escapes, it
-will be allocated at that point. The optimizer needs to keep track
-of the state of the object that the operation would have created.
-This is done with the help of a \emph{static object}\footnote{Here
-``static'' is meant in the sense of partial evaluation, \ie known at partial
-evaluation time, not in the sense of ``static allocation'' or ``static
-method''.}. The static object describes the shape of the object that would have
-been allocated, \ie the type of the object and where the values that would be
-stored in the fields of the allocated object come from.
-
-In the snippet above, the two \lstinline{new} operations are removed and two
-static objects are constructed. The \lstinline{set} operations manipulate
-static objects, therefore they can be removed as well; their effect is
-remembered in the static objects.
-
-After the operations the static object associated with $p_{5}$ would store the knowledge that it is a
-\lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
-one associated with $p_{6}$ would store that it is a \lstinline{BoxedInteger}
-whose \lstinline{intval} field contains the constant -100.
-
-The subsequent operations (line 20--26) in Figure~\ref{fig:unopt-trace}, which
-use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=20,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-guard_class($p_{5}$, BoxedInteger)  |\setcounter{lstnumber}{21}|
-$i_{7}$ = get($p_{5}$, intval)
-guard_class($p_{6}$, BoxedInteger)  |\setcounter{lstnumber}{24}|
-$i_{8}$ = get($p_{6}$, intval)
-$i_{9}$ = int_add($i_{7}$, $i_{8}$)
-\end{lstlisting}
-
-The \lstinline{guard_class} operations can be removed, since their arguments are
-static objects with the matching type \lstinline{BoxedInteger}. The
-\lstinline{get} operations can be removed as well, because each of them reads a
-field out of a static object. The results of the get operation are replaced with
-what the static object stores in these fields: all the occurences of $i_{7}$ and $i_{8}$ in the trace are just
-replaced by $i_{4}$ and -100. The only operation copied into the optimized trace
-is the addition:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=26,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$i_{9}$ = int_add($i_{4}$, -100)
-\end{lstlisting}
-
-The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
-similar vein. The operations in lines 27--35 produce two more static objects and
-are removed. Those in line 36--39 are just copied into the output trace because they
-manipulate objects that are allocated before the trace. Lines 40--42 are removed
-because they operate on a static object. Line 43 is copied into the output trace.
-Lines 44--46 produce a new static object and are removed, lines 48--51 manipulate
-that static object and are removed as well. Lines 52--54 are copied into the output
-trace.
-
-The last operation (line 55) is an interesting case. It is the \lstinline{jump}
-operation that passes control back to the beginning of the trace. The two
-arguments to this operation at this point are static objects. However, because
-they are passed into the next iteration of the loop they live longer than the trace
-and therefore cannot remain static. They need to be turned into dynamic
-(runtime) objects before the actual \lstinline{jump} operation. This process of
-turning a static object into a dynamic one is called \emph{lifting}.
-
-Lifting a static object puts \lstinline{new} and \lstinline{set} operations into
-the output trace. Those operations produce an object at runtime that has the
-shape described by the static object. This process is a bit delicate,
-because the static objects could form an arbitrary graph structure. In our
-example it is simple, though:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt,numbers=right,escapechar=|, firstnumber=44,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
-set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
-$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
-set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{54}|
-jump($p_{15}$, $p_{10}$)
-\end{lstlisting}
-
-Observe how the operations for creating these two instances have been moved to a
-later point in the trace. This is worthwhile even though the objects have to be
-allocated in the end because some \lstinline{get} operations and
-\lstinline{guard_class} operations on the lifted static objects could be
-removed.
-
-More generally, lifting needs to occur if a static object is used in any
-operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
-It also needs to occur if \lstinline{set} is used to store a static object into
-a non-static one.
-
-The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
-The optimized trace contains only two allocations, instead of the original
-five, and only three \lstinline{guard_class} operations, compared to the
-original seven.
-
-\begin{figure}
-\begin{lstlisting}[mathescape,numbers=right,escapechar=|, numberblanklines=false,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-# arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
-guard_class($p_1$, BoxedInteger)           |\setcounter{lstnumber}{4}|
-$i_2$ = get($p_1$, intval)
-guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{7}|
-$i_3$ = get($p_0$, intval)
-$i_4$ = int_add($i_2$, $i_3$)              |\setcounter{lstnumber}{25}|
-$i_9$ = int_add($i_4$, -100)               |\setcounter{lstnumber}{35}|
-
-guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{38}|
-$i_{12}$ = get($p_0$, intval)              |\setcounter{lstnumber}{42}|
-$i_{14}$ = int_add($i_{12}$, -1)           |\setcounter{lstnumber}{50}|
-
-$i_{17}$ = int_gt($i_{14}$, 0)             |\setcounter{lstnumber}{53}|
-guard_true($i_{17}$)                       |\setcounter{lstnumber}{42}|
-
-$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
-set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
-$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
-set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{53}|
-
-jump($p_{15}$, $p_{10}$)
-\end{lstlisting}
-
-\caption{Resulting Trace After Allocation Removal}
-\label{fig:step1}
-\end{figure}
-
-\section{Formal Description of the Algorithm}
-\label{sec:formal}
-
-\begin{figure*}
-\begin{center}
-\begin{tabular}{lcccc}
-\emph{new} & ${\displaystyle \frac{l\,\mathrm{fresh}}{v=\mathtt{new}(T),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto l\right],H\left[l\mapsto\left(T,\mathrm{null},\mathrm{null}\right)\right]}}$ & ~~~ & \emph{guard} & ${\displaystyle \frac{\mathrm{type}(H(E(v))=T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H}}$\tabularnewline[3em]
-\emph{get} & ${\displaystyle \frac{\,}{u=\mathtt{get}(v,F),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[u\mapsto H\left(E\left(v\right)\right)_{F}\right],H}}$ & ~~~ &  & ${\displaystyle \frac{\mathrm{type}(H(E(v))\neq T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}\bot,\bot}}$\tabularnewline[3em]
-\emph{set} & ${\displaystyle \frac{\,}{\mathtt{set}\left(v,F,u\right),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H\left[E\left(v\right)\mapsto\left(H\left(E\left(v\right)\right)!_{F}E(u)\right)\right]}}$ & ~~~ &  & \tabularnewline[4em]
-\end{tabular}
-\end{center}
-
-\begin{minipage}[b]{7 cm}
-\emph{Object Domains:}
-$$\setlength\arraycolsep{0.1em}
- \begin{array}{rcll}
-    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
-    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
-    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
-    l & \in & L & \mathrm{\ locations\ on\ heap}
- \end{array}
-$$
-\end{minipage}
-\begin{minipage}[b]{5 cm}
-\emph{Semantic Values:}
-$$\setlength\arraycolsep{0.1em}
- \begin{array}{rcll}
-    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
-    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
-    \\
-    \\
- \end{array}
-$$
-\end{minipage}
-\caption{The Operational Semantics of Simplified Traces}
-\label{fig:semantics}
-\end{figure*}
-
-In this section we want to give a formal description of the semantics of the
-traces and of the optimizer and liken the optimization to partial evaluation.
-We focus on the operations for manipulating heap allocated objects,
-as those are the only ones that are actually optimized. We also consider only
-objects with two fields $L$ and $R$ in this section, generalizing to arbitrary many fields
-is straightforward. Traces are lists of operations. The operations considered here are
-\lstinline{new}, \lstinline{get}, \lstinline{set} and \lstinline{guard_class}.
-
-The values of all
-variables are locations (\ie pointers). Locations are mapped to objects, which
-are represented by triples $(T,l_1,l_2)$ of a type $T$, and two locations that represent the
-fields of the object. When a new object is created, the fields are initialized
-to null, but we require that they are initialized to a real
-location before being read, otherwise the trace is malformed (this condition is
-guaranteed by how the traces are generated in PyPy).
-
-We use some abbreviations when dealing with object triples. To read the type of
-an object, $\mathrm{type}((T,l_1,l_2))=T$ is used. Reading a field $F$ from an
-object is written $(T,l_1,l_2)_F$ which either is $l_1$ if $F=L$ or $l_2$
-if $F=R$. To set field $F$ to a new location $l$, we use the notation
-$(T,l_1,l_2)!_Fl$, which yields a new triple $(T,l,l_2)$ if $F=L$ or a new
-triple $(T,l_1,l)$ if $F=R$.
-
-Figure~\ref{fig:semantics} shows the operational semantics for traces. The
-interpreter formalized there executes one operation at a time. Its state is
-represented by an environment $E$ and a heap $H$, which may be changed by the
-execution of an operation. The environment is a partial function from variables
-to locations and the heap is a partial function from locations to objects. Note
-that a variable can never be null in the environment, otherwise the trace would have
-been malformed. The environment could not directly map variables to objects,
-because several variables can point to the \emph{same} object, because of aliasing.
-
-We use the following notation for updating partial functions:
-$E[v\mapsto l]$ denotes the environment which is just like $E$, but maps $v$ to
-$l$.
-
-The \lstinline{new} operation creates a new object $(T,\mathrm{null},\mathrm{null})$ on the
-heap under a fresh location $l$ and adds the result variable to the environment,
-mapping it to the new location $l$.
-
-The \lstinline{get} operation reads a field $F$ out of an object, and adds the result
-variable to the environment, mapping it to the read location. The heap is
-unchanged.
-
-The \lstinline{set} operation changes field $F$ of an object stored at the location that
-variable $v$ maps to. The new value of the field is the location in variable
-$u$. The environment is unchanged.
-
-The \lstinline{guard_class} operation is used to check whether the object stored at the location
-that variable $v$ maps to is of type $T$. If that is the case, then execution
-continues without changing heap and environment. Otherwise, execution is
-stopped.
-
-% subsection Operational Semantics of Traces (end)
-
-\subsection{Optimizing Traces}
-\label{sub:formalopt}
-
-\begin{figure*}
-\begin{center}
-\begin{tabular}{lc}
-\emph{new} & ${\displaystyle \frac{v^{*}\,\mathrm{fresh}}{v=\mathtt{new}(T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto v^{*}\right],S\left[v^{*}\mapsto\left(T,\mathrm{null,null}\right)\right]}}$\tabularnewline[3em]
-\emph{get} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[u\mapsto S(E(v))_{F}\right],S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S),\,\,\, u^{*}\,\mathrm{fresh}}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle u^{*}=\mathtt{get}(E(v),F)\right\rangle ,E\left[u\mapsto u^{*}\right],S}}$\tabularnewline[3em]
-\emph{set} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S\left[E\left(v\right)\mapsto\left(S(E(v))!_{F}E(u)\right)\right]}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\notin\mathrm{dom}\left(S\right),\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathtt{set}\left(E(v),F,E(u)\right)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
-\emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\vee\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathtt{guard\_class}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
-\emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
- & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathtt{new}\left(\mathrm{type}\left(S\left(v^{*}\right)\right)\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
- & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::\mathrm{ops}_{R}::\left\langle \mathtt{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathtt{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime\prime}}}$\tabularnewline[3em]
-\end{tabular}
-\end{center}
-
-\begin{minipage}[b]{7 cm}
-\emph{Object Domains:}
-$$\setlength\arraycolsep{0.1em}
- \begin{array}{rcll}
-    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
-    u^*,v^* & \in & V^* & \mathrm{\ variables\ in\ optimized\ trace}\\
-    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
-    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
- \end{array}
-$$
-\end{minipage}
-\begin{minipage}[b]{5 cm}
-\emph{Semantic Values:}
-$$\setlength\arraycolsep{0.1em}
- \begin{array}{rcll}
-    E & \in & V\rightharpoonup V^* & \mathrm{\ Environment}\\
-    S & \in & V^*\rightharpoonup\mathfrak{T}\times(V^*\cup \{ \mathrm{null} \})\times (V^*\cup \{ \mathrm{null}\} ) & \mathrm{\ Static\ Heap}\\
-    \\
-    \\
- \end{array}
-$$
-\end{minipage}
-\caption{Optimization Rules}
-\label{fig:optimization}
-\end{figure*}
-
-To optimize the simple traces of the last section, we use online partial
-evaluation. The partial evaluator optimizes one operation of a trace at a
-time. Every operation in the unoptimized trace is replaced by a list of
-operations in the optimized trace. This list is empty if the operation can be
-optimized away. The optimization rules can be seen in
-Figure~\ref{fig:optimization}. Lists are written using angular brackets $<...>$,
-list concatenation is expressed using two colons: $l_1::l_2$.
-
-The state of the optimizer is stored in an environment $E$ and a \emph{static
-heap} $S$. Each step of the optimizer takes an operation, an environment and a
-static heap and produces a list of operations, a new environment and a new
-static heap.
-
-The environment is a partial function from variables in the
-unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
-themselves written with a
-$\ ^*$ for clarity). The reason for introducing new variables in the optimized
-trace is that several variables that appear in the unoptimized trace can turn
-into the same variables in the optimized trace. The environment of the
-optimizer serves a function similar to that of the environment in the
-semantics: to express sharing.
-
-The static heap is a partial function from $V^*$ into the
-set of static objects, which are triples of a type and two elements of $V^*$.
-The object referenced by a variable $v^*$ is static, if $v^*$ is in the domain
-of the static heap $S$. The object $S(v^*)$ describes
-what is statically known about the object, \ie its type and its fields. The
-fields of objects in the static heap are also elements of $V^*$ (or null, for
-short periods of time).
-
-When the optimizer sees a \lstinline{new} operation, it optimistically removes it and
-assumes that the resulting object can stay static. The optimization for all
-further operations is split into two cases. One case is for when the
-involved variables are in the static heap, which means that the operation can be
-performed at optimization time and can be removed from the trace. These rules mirror
-the execution semantics closely. The other case is for when not enough is known about
-the variables, and the operation has to be residualized.
-
-If the argument $v$ of a \lstinline{get} operation is mapped to something in the static
-heap, the \lstinline{get} can be performed at optimization time. Otherwise, the \lstinline{get}
-operation needs to be residualized.
-
-If the first argument $v$ to a \lstinline{set} operation is mapped to something in the
-static heap, then the \lstinline{set} can be performed at optimization time
-(which updates the static heap). Otherwise the \lstinline{set} operation needs
-to be residualized. This needs to be
-done carefully, because the new value for the field, from the variable $u$,
-could itself be static, in which case it needs to be lifted first.
-
-If a \lstinline{guard_class} is performed on a variable that is in the static heap, the type check
-can be performed at optimization time, which means the operation can be removed
-if the types match. If the type check fails statically or if the object is not
-in the static heap, the \lstinline{guard_class} is residualized. This also needs to
-lift the variable on which the \lstinline{guard_class} is performed.
-
-Lifting takes a variable and turns it into a dynamic variable. If the variable
-is already dynamic, nothing needs to be done. If it is in the static heap,
-operations are emitted that construct an object with the shape described
-there, and the variable is removed from the static heap.
-
-Lifting a static object needs to recursively lift its fields. Some care needs to
-be taken when lifting a static object, because the structures described by the
-static heap can be cyclic. To make sure that the same static object is not lifted
-twice, the \lstinline{liftfield} operation removes it from the static heap \emph{before}
-recursively lifting its fields.
-
-As an example for lifting, consider the static heap $$\{v^* \mapsto (T_1, w^*,
-v^*), w^* \mapsto (T_2, u^*, u^*)\}$$ which contains two static objects. If $v^*$
-needs to be lifted, the following residual operations are produced:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$v^*$ = new($T_1$)
-$w^*$ = new($T_2$)
-set($w^*$, $L$, $u^*$)
-set($w^*$, $R$, $u^*$)
-set($v^*$, $L$, $w^*$)
-set($v^*$, $R$, $v^*$)
-\end{lstlisting}
-
-After the lifting the static heap is the empty set, because both static objects
-were lifted.
-If we had lifted $w^*$ instead of $v^*$, then the following operations would
-have been produced:
-
-\begin{lstlisting}[mathescape,xleftmargin=20pt,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$w^*$ = new($T_2$)
-set($w^*$, $L$, $u^*$)
-set($w^*$, $R$, $u^*$)
-\end{lstlisting}
-
-In this case, the static heap afterwards would be:
-$$\{v^* \mapsto (T_1, w^*, v^*)\}$$
-
-% subsection Optimizing Traces (end)
-
-\subsection{Analysis of the Algorithm}
-\label{sub:analysis}
-
-While we do not offer a formal proof of it, it can argue informally
-that the algorithm presented above is sound: it works by delaying (and
-often completely removing) some operations.  The algorithm runs in a
-single pass over the list of operations.  We can check that although
-recursively lifting a static object is not a constant-time operation,
-the algorithm only takes a total time linear in the length of the trace.
-The algorithm itself is not particularly complex; our focus is
-rather that \emph{in the context of tracing JITs} it is possible to find a
-simple enough algorithm that performs well.
-
-Note in particular that objects in category 1 (\ie those that do
-not escape) are completely removed; moreover, objects in category 2
-(\ie escaping) are still partially optimized: all the operations in between the
-creation of the object and the point where it escapes that involve the object
-are removed. Objects in category 3 and 4 are also partially optimized, their
-allocation is delayed till the end of the trace.
-
-The optimization is particularly effective for chains of operations.
-For example, it is typical for an interpreter to generate sequences of
-writes-followed-by-reads, where one interpreted opcode writes to some
-object's field and the next interpreted opcode reads it back, possibly
-dispatching on the type of the object created just before.  A typical example
-would be a chain of arithmetic operations.
-
-% subsection Analysis of the Algorithm (end)
-
-% section Formal Description of the Algorithm (end)
-
-%___________________________________________________________________________
-
-% section Escape Analysis in a Tracing JIT (end)
-
-\section{Implementation and Evaluation}
-\label{sec:Evaluation}
-
-The allocation removal techniques described in this paper were implemented in
-the optimizer of PyPy's tracing JIT. The optimization is independent of which
-interpreter a JIT is generated for. There are some practical issues beyond the
-techniques described in this paper. The actual implementation needs to deal with
-more operations than described in Section~\ref{sec:formal}, \eg to
-also support static arrays in addition to static objects. The implementation of
-this optimization is about 400 lines of RPython code.
-
-A further complication is that most interpreters written with PyPy use
-heap-allocated frame objects to store local variables. Those severely hinder
-the effectiveness of allocation removal, because every time an object is stored
-into a local variable, it is stored into the frame-object, which makes it
-escape. We implemented a technique to treat such frames objects in a
-special way to solve this problem. This is a common approach in VM
-implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}; the
-novelty of our approach is that we generalized it enough to be usable for
-different interpreters.
-
-To evaluate our allocation removal algorithm, we look at the effectiveness when
-used in the generated tracing JIT of PyPy's Python interpreter. This interpreter
-is a full implementation of Python 2.5 language semantics and is about 30,000
-lines of RPython code.
-
-The
-benchmarks we used are small-to-medium Python programs, some synthetic
-benchmarks, some real applications.\footnote{All the source code of the
-benchmarks can be found at \texttt{http://codespeak.net/svn/pypy/benchmarks/}.
-There is also a website that monitors PyPy's performance nightly at
-\texttt{http://speed.pypy.org/}.}
-
-Some of them are from the Computer Language Benchmark
-Game\footnote{\texttt{http://shootout.alioth.debian.org/}}: \textbf{fannkuch},
-\textbf{nbody}, \textbf{meteor-contest}, \textbf{spectral-norm}.
-
-Furthermore there are the following benchmarks:
-\begin{itemize}
-    \item \textbf{crypto\_pyaes}: An AES implementation.
-    \item \textbf{django}: The templating engine of the Django web
-    framework\footnote{\texttt{http://www.djangoproject.com/}}.
-    \item \textbf{go}: A Monte-Carlo Go
-    AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
-    \item \textbf{html5lib}: An HTML5 parser.
-    \item \textbf{pyflate-fast}: A BZ2 decoder.
-    \item \textbf{raytrace-simple}: A ray tracer.
-    \item \textbf{richards}: The Richards benchmark.
-    \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
-    \item \textbf{telco}: A Python version of the Telco decimal
-    benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
-    using a pure Python decimal floating point implementation.
-    \item \textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
-    framework\footnote{\texttt{http://twistedmatrix.com/}}.
-\end{itemize}
-
-We evaluate the allocation removal algorithm along two lines: first we want to
-know how many allocations could be optimized away. On the other hand, we want
-to know how much the run times of the benchmarks is improved.
-
-The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
-with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
-2.6.35. We compared the performance of various Python implementations on the
-benchmarks. As a baseline, we used the standard Python implementation in C,
-CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
-interpreter. Furthermore we compared against
-Psyco\cite{rigo_representation-based_2004} 1.6,
-a (hand-written) extension module to CPython which is a
-just-in-time compiler that produces machine code at run-time. It is not based
-on traces. Finally, we used two versions of PyPy's Python interpreter (revision
-77823 of SVN trunk\footnote{\texttt{http://codespeak.net/svn/pypy/trunk}}): one
-including the JIT but not optimizing the traces, and one using the allocation
-removal optimizations (as well as some minor other optimizations, such as
-constant folding).
-
-As the first step, we counted the occurring operations in all generated traces
-before and after the optimization phase for all benchmarks. The resulting
-numbers can be
-seen in Figure~\ref{fig:numops}. The optimization removes between 4\% and 90\%
-and of allocation operations in the traces of the benchmarks.
-All benchmarks taken together, the optimization removes 70\% percent of
-allocation operations. The numbers look similar for reading and writing of
-attributes. There are even more \lstinline{guard} operations that are removed,
-however there is an additional optimization that removes guards, so not all the
-removed guards are an effect of the optimization described here (for technical
-reasons, it would be very hard to separate the two effects).
-
-\begin{figure*}
-{\footnotesize
-\begin{center}
-\begin{tabular}{|l||r|rr|rr|rr|rr|}
-\hline
-	&num loops      &new    &removed        &get/set        &removed &guard &removed        &all ops        &removed\\
-\hline
-crypto\_pyaes	&78	&3088	&50\%	&57148	&25\%	&9055	&95\%	&137189	&80\%\\
-django	&51	&673	&54\%	&19318	&18\%	&3876	&93\%	&55682	&85\%\\
-fannkuch	&43	&171	&49\%	&886	&63\%	&1159	&81\%	&4935	&45\%\\
-go	&517	&12234	&76\%	&200842	&21\%	&53138	&90\%	&568542	&84\%\\
-html5lib	&498	&14432	&68\%	&503390	&11\%	&71592	&94\%	&1405780	&91\%\\
-meteor-contest	&59	&277	&36\%	&4402	&31\%	&1078	&83\%	&12862	&68\%\\
-nbody	&13	&96	&38\%	&443	&69\%	&449	&78\%	&2107	&38\%\\
-pyflate-fast	&162	&2278	&55\%	&39126	&20\%	&8194	&92\%	&112857	&80\%\\
-raytrace-simple	&120	&3118	&59\%	&91982	&15\%	&13572	&95\%	&247436	&89\%\\
-richards	&87	&844	&4\%	&49875	&22\%	&4130	&91\%	&133898	&83\%\\
-spambayes	&314	&5608	&79\%	&117002	&11\%	&25313	&94\%	&324125	&90\%\\
-spectral-norm	&38	&360	&64\%	&5553	&20\%	&1122	&92\%	&11878	&77\%\\
-telco	&46	&1257	&90\%	&37470	&3\%	&6644	&99\%	&98590	&97\%\\
-twisted-names	&214	&5273	&84\%	&100010	&10\%	&23247	&96\%	&279667	&92\%\\
-\hline
-total	&2240	&49709	&70\%	&1227447	&14\%	&222569	&93\%	&3395548	&89\%\\
-\hline
-\end{tabular}
-\end{center}
-}
-\caption{Number of Operations and Percentage Removed By Optimization}
-\label{fig:numops}
-\end{figure*}
-
-In addition to the count of operations we also performed time measurements.
-All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
-code. The arithmetic mean of the times of the last 30 runs were used as the
-result. The errors were computed using a confidence interval with a 95\%
-confidence level \cite{georges_statistically_2007}. The results are reported in
-Figure~\ref{fig:times}. For each implementation the table also reports the
-speedup that PyPy with optimization achieves over it.
-
-With the optimization turned on, PyPy's Python interpreter outperforms CPython
-in all benchmarks except spambayes (which heavily relies on regular expression
-performance and thus is not helped much by our Python JIT) and meteor-contest.
-All benchmarks are improved by the allocation removal optimization, by at least
-20\% and by as much as a factor of 6.95.
-
-Psyco is able to outperform PyPy's JIT in five out of 14 benchmarks. We hope to
-overtake Psyco (which is no longer being actively developped) by adding some
-further optimizations.
-
-\begin{figure*}
-{\footnotesize
-\begin{center}
-\begin{tabular}{|l||r|r||r|r||r|r||r|r|}
-\hline
-	&CPython [ms]	& $\times$ &Psyco [ms]	& $\times$ &PyPy w/o optim. [ms] & $\times$ &PyPy w/ optim. [ms]& $\times$ \\
-\hline
-crypto\_pyaes	&2757.80 $\pm$ 0.98	&10.33	&67.90 $\pm$ 0.47	&0.25	&1652.00 $\pm$ 4.00	&6.19	&266.86 $\pm$ 5.94	&1.00\\
-django	&993.19 $\pm$ 0.50	&3.83	&913.51 $\pm$ 4.22	&3.52	&694.73 $\pm$ 2.86	&2.68	&259.53 $\pm$ 1.79	&1.00\\
-fannkuch	&1987.22 $\pm$ 2.02	&4.26	&944.44 $\pm$ 0.61	&2.02	&566.99 $\pm$ 1.06	&1.21	&466.87 $\pm$ 1.85	&1.00\\
-go	&947.21 $\pm$ 1.58	&3.00	&445.96 $\pm$ 0.68	&1.41	&2197.71 $\pm$ 25.21	&6.95	&316.15 $\pm$ 9.33	&1.00\\
-html5lib	&13987.12 $\pm$ 19.51	&1.39	&17398.25 $\pm$ 36.50	&1.72	&27194.45 $\pm$ 46.62	&2.69	&10092.19 $\pm$ 23.50	&1.00\\
-meteor-contest	&346.98 $\pm$ 0.35	&0.88	&215.66 $\pm$ 0.23	&0.55	&433.04 $\pm$ 1.45	&1.10	&392.85 $\pm$ 0.87	&1.00\\
-nbody\_modified	&637.90 $\pm$ 1.82	&6.14	&256.78 $\pm$ 0.18	&2.47	&135.55 $\pm$ 0.33	&1.30	&103.93 $\pm$ 0.25	&1.00\\
-pyflate-fast	&3169.35 $\pm$ 1.89	&1.74	&1278.16 $\pm$ 3.13	&0.70	&3285.89 $\pm$ 8.51	&1.80	&1822.36 $\pm$ 11.52	&1.00\\
-raytrace-simple	&2744.60 $\pm$ 51.72	&4.24	&1072.66 $\pm$ 1.08	&1.66	&2778.27 $\pm$ 15.13	&4.29	&647.24 $\pm$ 5.44	&1.00\\
-richards	&354.06 $\pm$ 1.00	&4.01	&63.48 $\pm$ 0.15	&0.72	&383.93 $\pm$ 3.28	&4.35	&88.32 $\pm$ 0.91	&1.00\\
-spambayes	&299.16 $\pm$ 0.35	&0.75	&338.68 $\pm$ 3.14	&0.85	&580.90 $\pm$ 24.68	&1.46	&397.37 $\pm$ 10.60	&1.00\\
-spectral-norm	&478.63 $\pm$ 0.80	&4.27	&139.83 $\pm$ 1.54	&1.25	&353.51 $\pm$ 1.39	&3.15	&112.10 $\pm$ 1.17	&1.00\\
-telco	&1207.67 $\pm$ 2.03	&2.44	&730.00 $\pm$ 2.66	&1.47	&1296.08 $\pm$ 4.37	&2.62	&495.23 $\pm$ 2.14	&1.00\\
-twisted\_names	&9.58 $\pm$ 0.01	&1.34	&10.43 $\pm$ 0.01	&1.46	&17.99 $\pm$ 0.27	&2.52	&7.13 $\pm$ 0.09	&1.00\\
-\hline
-\end{tabular}
-\end{center}
-}
-\caption{Benchmark Times in Milliseconds, Together With Factor Over PyPy With Optimizations}
-\label{fig:times}
-\end{figure*}
-
-\section{Related Work}
-\label{sec:related}
-
-There exists a large number of works on escape analysis, which is a program
-analysis that tries to find an upper bound for the lifetime of objects allocated
-at specific program points
-\cite{goldberg_higher_1990,park_escape_1992,choi_escape_1999,bruno_blanchet_escape_2003}.
-This information can then be used to decide that certain objects can be
-allocated on the stack, because their lifetime does not exceed that of the stack
-frame it is allocated in. The difference to our work is that escape analysis is
-split into an analysis and an optimization phase. The analysis can be a lot more
-complex than our simple one-pass optimization. Also, stack-allocation reduces
-garbage-collection pressure but does not optimize away the actual accesses to
-the stack-allocated object. In our case, an object is not needed at all any
-more.
-
-Chang \etal describe a tracing JIT for JavaScript running on top of a JVM
-\cite{mason_chang_efficient_2007}. They mention in passing an approach to
-allocation removal that moves the allocation of an object of type 1 out of the
-loop to only allocate it once, instead of every iteration. No details are given
-for this optimization. The fact that the object is still allocated and needs to
-be written to means that only the allocations are optimized away, but not the
-reads out of and writes into the object.
-
-SPUR, a tracing JIT for C\# seems to be able to remove allocations in a similar
-way to the approach described here, as hinted at in the technical report
-\cite{bebenita_spur:_2010}. However, no details for the approach and its
-implementation are given.
-
-Psyco \cite{rigo_representation-based_2004} is a (non-tracing) JIT for Python
-that implements a more ad-hoc version of the allocation removal described here.
-Our static objects could be related to what are called \emph{virtual} objects
-in Psyco.  Historically,
-PyPy's JIT can be seen as some successor of Psyco for a general context (one of
-the authors of this paper is the author of Psyco).
-
-The original SELF JIT compiler \cite{chambers_efficient_1989} used an algorithm for
-forward-propagating the types of variables as part of its optimizations. This
-makes it possible to remove all type checks on a variable but the first one.
-The optimization does not deal with removing the full object, if it is
-short-lived, but the type check removals are similar to what our optimization
-achieves.
-
-Partially known data structures are built directly into Prolog (via unbound
-logic variables) and thus the treatment of partially static data structures
-was part of partial evaluation of Prolog programs from the early stages
-\cite{lloyd_partial_1991}. One effect of unfolding in Prolog is that terms that
-are constructed and immediately matched again, completely disappear in the
-residual program. This is similar to what our optimization does for an
-imperative language.
-In functional programming this idea was introduced as
- constructor specialisation by Mogensen \cite{mogensen_constructor_1993}.
-
-A related optimization is also that of deforestation
-\cite{wadler_deforestation:_1988,gill_short_1993} which tries to remove
-intermediate lists or trees in functional languages. More explicitly geared at
-optimizing boxing is boxing analysis \cite{jrgensen_calculus_1996} which tries
-to fully remove pairs of calls to box/unbox in a functional language.
-
-%xxx:
-%partially static data structures: kenichi asai's thesis?
-
-%xxx: 
-%Other related work is compile-time garbage collection \cite{mazur_practical_2001}, which
-%tries to ....;
-%  separation logic
-  %; John Hughes: type specialization
-
-\section{Conclusion and Future Work}
-\label{sec:conclusion}
-
-In this paper, we used an approach based on online partial evaluation
-to optimize away allocations and type guards in the traces of a
-tracing JIT.  In this context a simple approach based on partial evaluation
-gives good results.  This is due to the fact that the tracing JIT
-itself is responsible for all control issues, which are usually the
-hardest part of partial evaluation: the tracing JIT selects the parts
-of the program that are worthwhile to optimize, and extracts linear
-paths through them, inlining functions as necessary.  What is left to
-optimize are only those linear paths.
-
-We expect a similar result for other optimizations that usually require
-a complex analysis phase and are thus normally too slow to use at
-runtime.  A tracing JIT selects interesting linear paths by itself;
-therefore, a naive version of many optimizations on such paths should
-give mostly the same results.  For example, we experimented with (and
-plan to write about) store-load propagation with a very simple alias
-analysis.
-
-\section*{Acknowledgements}
-
-The authors would like to thank Stefan Hallerstede, David Schneider and Thomas
-Stiehl for fruitful discussions and detailed feedback during the writing of the
-paper.
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
+\end{thebibliography}
 

### 2011-06-09 talk/iwtc11/paper.tex

@@
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-# arguments to the trace: $p_{0}$, $p_{1}$
+$l_0$($p_{0}$, $p_{1}$):
 # inside f: y = y.add(step)
@@
             set($p_{5}$, intval, $i_{4}$)
-jump($p_{0}$, $p_{5}$)
+jump($l_0$, $p_{0}$, $p_{5}$)
 \end{lstlisting}
@@
 the original function.
+
+The label of the loop is $l_0$ and is used by the jump instruction to
+identify it's jump target.
 
@@
 
+\section{Optimizations}
+Before the trace is passed to a backend compiling it into machine code
+it needs to be optimized to achieve better performance.
+The focus of this paper
+is loop invariant code motion. The goal of that is to move as many
+operations as possible out of the loop making them executed only once
+and not every iteration. This we propose to achieve by loop peeling. It
+leaves the loop body intact, but prefixes it with one iteration of the
+loop. This operation by itself will not achieve anything. But if it is
+combined with other optimizations it can increase the effectiveness of
+those optimizations. For many optimization of interest some care has
+to be taken when they are combined with loop peeling. This is
+described below by first explaining the loop peeling optimization
+followed by a set of other optimizations and how they interact with
+loop peeling.
+
+\subsection{Loop peeling}
+Loop peeling is achieved by inlining the trace at the end of
+itself. The input arguments of the second iteration are replaced with
+the jump arguments of the first iterations and then the arguments of all
+the operations are updated to operate on the new input arguments. To
+keep the single-assignment form new variables has to be introduced as
+the results of all the operations. The first iteration of the loop
+will end with a jump to the second iteration of the loop while the
+second iteration will end with a jump to itself. This way the first
+copy of the trace only be executed once while the second copy will be
+used for every other iteration. The rationality here is that the
+optimizations below typically will be able to optimize the second copy
+more efficiently than the first. The trace from Figure~\ref{fig:unopt-trace} would
+after this operation become the trace in Figure~\ref{fig:peeled-trace}.
+
+\begin{figure}
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_0$($p_{0}$, $p_{1}$):
+# inside f: y = y.add(step)
+guard_class($p_{1}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{2}$ = get($p_{1}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{3}$ = get($p_{0}$, intval)
+        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
+        $p_{5}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{5}$, intval, $i_{4}$)
+jump($l_1$, $p_{0}$, $p_{5}$)
+
+$l_1$($p_{0}$, $p_{5}$):
+# inside f: y = y.add(step)
+guard_class($p_{5}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{6}$ = get($p_{5}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{7}$ = get($p_{0}$, intval)
+        $i_{8}$ = int_add($i_{6}$, $i_{7}$)
+        $p_{9}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{9}$, intval, $i_{8}$)
+jump($l_1$, $p_{0}$, $p_{9}$)
+\end{lstlisting}
+\caption{An Unoptimized Trace of the Example Interpreter}
+\label{fig:peeled-trace}
+\end{figure}
+
+When applying the following optimizations to this two iteration trace
+some care has to taken as to how the jump arguments of both
+iterations and the input arguments of the second iteration are
+treated. It has to be ensured that the second iteration stays a proper
+trace in the sens that the operations within it only operations on
+variables that are either among the input arguments of the second iterations
+or are produced within the second iterations. To ensure this we need
+to introduce a bit of formalism.
+
+The original trace (prior too peeling) consists of three parts. 
+A vector of input
+variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non
+jump operations and a single
+jump operation. The jump operation contains a vector of jump variables,
+$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
+loop peeling there will be a second copy of this trace with input
+variables equal to the jump arguments of the first copy, $J$, and jump
+arguments $K$. Looking back at our example we have
+\begin{equation}
+  %\left\{
+    \begin{array}{lcl}
+      I &=& \left( p_0, p_1 \right) \\
+      J &=& \left( p_0, p_5 \right) \\
+      K &=& \left( p_0, p_9 \right) \\
+    \end{array}
+  %\right.
+  .
+\end{equation}
+To construct the second iteration from the first we also need a
+function, $m$, mapping the variables of the first iteration onto the
+variables of the second. This function is constructed during the
+inlining. It is initialized by mapping the input arguments, $I$, to
+the jump arguments $J$,
+\begin{equation}
+  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
+\end{equation}
+In the example that means (XXX which notation do we prefer?)
+\begin{equation}
+  m(v) = 
+  \left\{
+    \begin{array}{lcl}
+      p_0 &\text{if}& v=p_0 \\
+      p_5 &\text{if}& v=p_1 \\
+    \end{array}
+  \right.
+  .
+\end{equation}
+\begin{equation}
+  %\left\{
+    \begin{array}{lcl}
+      m\left(p_0\right) &=& p_0 \\
+      m\left(p_1\right) &=& p_5
+    \end{array}
+  %\right.
+  .
+\end{equation}
+Each operation in the trace is inlined in the order they are
+executed. To inline an operation with argument vector 
+$A=\left(A_1, A_2, \cdots, A_{|A|}\right)$ producing the variable $v$
+a new variable, $\hat v$ is introduced. The inlined operation will
+produce $\hat v$ from the input arguments 
+\begin{equation}
+  \left(m\left(A_1\right), m\left(A_2\right), 
+    \cdots, m\left(A_{|A|}\right)\right) . 
+\end{equation}
+Before the
+next operation is inlined, $m$ is extend by making $m\left(v\right) = \hat
+v$. After all the operations in the example have been inlined we have
+\begin{equation}
+  %\left\{
+    \begin{array}{lcl}
+      m\left(p_0\right) &=& p_0 \\
+      m\left(p_1\right) &=& p_5 \\
+      m\left(i_2\right) &=& i_6 \\
+      m\left(i_3\right) &=& i_7 \\
+      m\left(i_4\right) &=& i_8 \\
+      m\left(p_5\right) &=& p_9 \\
+    \end{array}
+  %\right.
+  .
+\end{equation}
+
+\subsection{Redundant guard removal}
+No special concerns needs to be taken when implementing redundant
+guard removal together with loop peeling. However the the guards from
+the first iteration might make the guards of the second iterations
+redundant and thus removed. So the net effect of combining redundant
+guard removal with loop peeling is that guards are moved out of the
+loop. The second iteraton of the example reduces to
+
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_1$($p_{0}$, $p_{5}$):
+# inside f: y = y.add(step)
+    # inside BoxedInteger.add
+    $i_{6}$ = get($p_{5}$, intval)
+        # inside BoxedInteger.add__int
+        $i_{7}$ = get($p_{0}$, intval)
+        $i_{8}$ = int_add($i_{6}$, $i_{7}$)
+        $p_{9}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{9}$, intval, $i_{8}$)
+jump($l_1$, $p_{0}$, $p_{9}$)
+\end{lstlisting}
+
+\subsection{Heap caching}
+
+To implement heap caching variables has to be passed from the first
+iteration to the second by XXX
+\begin{equation}
+  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
+\end{equation}
+\begin{equation}
+  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
+  .
+\end{equation}
+In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat K$.
+
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_0$($p_{0}$, $p_{1}$):
+# inside f: y = y.add(step)
+guard_class($p_{1}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{2}$ = get($p_{1}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{3}$ = get($p_{0}$, intval)
+        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
+        $p_{5}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{5}$, intval, $i_{4}$)
+jump($l_1$, $p_{0}$, $p_{5}$, $i_3$, $i_4$)
+
+$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
+# inside f: y = y.add(step)
+    # inside BoxedInteger.add
+        # inside BoxedInteger.add__int
+        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
+        $p_{9}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{9}$, intval, $i_{8}$)
+jump($l_1$, $p_{0}$, $p_{9}$, $i_3$, $i_8$)
+\end{lstlisting}
+
+\subsection{Virtualization}
+Using escape analysis we can XXX
+
+Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
+same order). Extend it with all non virtual fields, $H_i$, of the
+removed virtuals,
+\begin{equation}
+  \hat J = \left(\tilde J_1, \tilde J_2, \cdots, \tilde J_{|\tilde J|}, 
+                 H_1, H_2, \cdots, H_{|H}\right)
+\end{equation}
+and let
+\begin{equation}
+  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
+                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
+  .
+\end{equation}
+
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_0$($p_{0}$, $p_{1}$):
+# inside f: y = y.add(step)
+guard_class($p_{1}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{2}$ = get($p_{1}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{3}$ = get($p_{0}$, intval)
+        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
+jump($l_1$, $p_{0}$, $i_3$, $i_4$)
+
+$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
+# inside f: y = y.add(step)
+    # inside BoxedInteger.add
+        # inside BoxedInteger.add__int
+        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
+jump($l_1$, $p_{0}$, $i_3$, $i_8$)
+\end{lstlisting}
+
+And we're down to a single integer addition!
+
+\section{Benchmarks}
+
 \appendix

### 2011-06-10 talk/iwtc11/paper.tex
description of heap caching

@@
 
+The guard on $p_5$ on line 17 of Figure~\ref{fig:unopt-trace} can be
+removed since $p_5$ is allocated on line 10 with a known class. The
+guard on $p_0$ on line 20 can be removed since it is identical to the
+guard on line 6.
+
 \subsection{Heap caching}
+The objective of heap caching is to remove \lstinline{get} and
+\lstinline{set} operations whose results can be deduced from previous
+\lstinline{get} and \lstinline{set} operations. Exact details of the
+process are outside the scope of this paper We will here assume that
+it works perfectly and only consider the interactions with loop
+peeling.
 
-To implement heap caching variables has to be passed from the first
-iteration to the second by XXX
+The issue at hand is to keep the second iteration a proper
+trace. Consider the \lstinline{get} operation on line 19 of
+Figure~\ref{fig:unopt-trace}. The result of this operation can be
+deduced to be $i_4$ from the \lstinline{set} operation on line
+12. Also, the result of the \lstinline{get} operation on line 22 can
+be deduced to be $i_3$ from the \lstinline{get} operation on line
+8. The optimization will thus remove line 19 and 22 from the trace and
+replace $i_6$ with $i_4$ and $i_7$ with $i_3$.
+
+After that, the second
+iteration will no longer be proper as it operates on $i_3$ and $i_4$
+which are not part of it. The solution is to extend the input
+arguments, $J$, with those two variables. This will also extend the
+jump arguments of the first iteration, which is also $J$. 
+Implicitly that also extends the jump arguments of the second iteration, $K$,
+since they are the inlined versions of $J$. That is the, $I$ has to
+be replaced by $\hat I$ which is formed as a concatenation of $I$ and
+$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
+$\hat K$ which is formed as a concatenation of $K$ and 
+$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
+The variable $i_7$ will then be replaced by $i_3$ by the heap caching
+algorithm as it has removed the variable $i_7$. XXX: Maybe we should
+replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?
+
+In general what is needed is for the heap optimizer is to keep track of
+which variables from the first iterations it reuses in the second
+iteration. It has to construct a vector of such variables $H$ which
+can be used to update the input and jump arguments,
 \begin{equation}
@@
 \end{equation}
-In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat K$.
+In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
+K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into
 

### 2011-06-12 talk/iwtc11/paper.tex

@@
 
-\subsection{Virtualization}
-Using escape analysis we can XXX
+\subsection{Allocation removal}
+By using escape analysis it is possible to identify objects that are
+allocated within the loop but never escapes it. That is the object are
+short lived and no references to them exists outside the loop. This
+is performed by processing the operation from top to bottom and
+optimistically removing every \lstinline{new} operation. Later on if
+it is discovered that a reference to the object escapes the loop, the
+\lstinline{new} operation is inserted at this point. All operations
+(\lstinline{get} and \lstinline{set}) on the removed objects are also
+removed and the optimizer needs to keep track of the value of all
+attributes of the object.
 
-Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
-same order). Extend it with all non virtual fields, $H_i$, of the
-removed virtuals,
+Consider again the original unoptimized trace of
+Figure~\label{fig:peeled-trace}. Line 10 contains the first
+allocation. It is removed and $p_5$ is marked as virtual. This means
+that it refers to an virtual object that was not yet
+(and might never be) allocated. Line 12 sets the \lstinline{intval}
+attribute of $p_5$. This operation is also removed and the optimizer
+registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.
+
+When the optimizer reaches line 13 it needs to construct the
+arguments for the \lstinline{jump} operation, which contains the virtual
+reference $p_5$. This can be achieved by exploding $p_5$ into it's
+attributes. In this case there is only one attribute and it's value is
+$i_4$, which means the $p_5$ is replaced with $i_4$ in the jump
+arguments.
+
+In the general case, each virtual in the jump arguments is exploded into a
+vector of variables containing the values of all it's attributes. If some
+of the attributes are themselves virtuals they are recursively exploded
+to make the vector contain only non virtual variables. Some care has
+to be taken to always place the attributes in the same order when
+performing this explosion. Notation becomes somewhat simpler if also every non
+virtual variable of the jump arguments is exploded into a vector. This will
+be a vector containing the original variable only. To summarize, for
+every variable, $J_k$, of the original jump arguments, $J$, let
 \begin{equation}
-  \hat J = \left(\tilde J_1, \tilde J_2, \cdots, \tilde J_{|\tilde J|}, 
-                 H_1, H_2, \cdots, H_{|H}\right)
+  \tilde J^{\left(k\right)} = \left\{
+      \begin{array}{ll}
+        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
+        H^{\left(k\right)} & \text{if $J_k$ is virtual}
+      \end{array}
+  \right.
+  ,
 \end{equation}
-and let
+where $H^{\left(k\right)}$ is a vector containing all non virtual
+attributes of $J_k$. The arguments of the optimized \lstinline{jump}
+operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
+\begin{equation}
+  \hat J = \left( 
+    \begin{array}{cccc}
+      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
+      \tilde J^{\left(|J|\right)} \\
+    \end{array}
+  \right)      
+  .
+\end{equation}
+and the arguments of the \lstinline{jump} operation of the second
+operation, $K$, are replaced by inlining $\hat J$, 
 \begin{equation}
@@
 \end{equation}
+In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
+K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into
 
@@
         $i_{4}$ = int_add($i_{2}$, $i_{3}$)
-jump($l_1$, $p_{0}$, $i_3$, $i_4$)
+            # inside BoxedInteger.__init__
+jump($l_1$, $p_{0}$, $i_{4}$)
 
-$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
+$l_1$($p_{0}$, $i_{4}$):
 # inside f: y = y.add(step)
     # inside BoxedInteger.add
+    guard_class($p_{0}$, BoxedInteger)
         # inside BoxedInteger.add__int
-        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
-jump($l_1$, $p_{0}$, $i_3$, $i_8$)
+        $i_{7}$ = get($p_{0}$, intval)
+        $i_{8}$ = int_add($i_{4}$, $i_{7}$)
+            # inside BoxedInteger.__init__
+jump($l_1$, $p_{0}$, $i_8$)
 \end{lstlisting}
 
-And we're down to a single integer addition!
+Note that virtuals are only exploded into their attributes when
+constructing the arguments of the jump of the first iteration. This
+explosion can't be repeated when constructing the arguments of the
+jump of the second iteration as it has to mach the first. This means
+the objects that was passed as pointers (non virtuals) from the first
+iteration to the second also has to be passed as pointers from the
+second iteration to the third. If one of these objects are virtual
+at the end of the second iteration they need to be allocated right
+before the jump.
 

### 2011-06-12 talk/iwtc11/paper.tex
implementation detail

@@
 at the end of the second iteration they need to be allocated right
-before the jump.
+before the jump. With the simple objects considered in this paper,
+that is not a problem. However in more complicated interpreters such
+an allocation might, in combination with other optimizations, lead
+to additional variables from the first iteration being imported into
+the second. This extends both $\hat J$ and $\hat K$, which means that
+some care has to be taken, when implementing this, to allow $\hat J$ to
+grow while inlining it into $\hat K$.
 

### 2011-06-12 talk/iwtc11/paper.tex
started to describe some benchmarks

@@
 
+The loop peeling optimization was implemented in the PyPy
+framework. That means that the jit compilers generated for all
+interpreters implemented within PyPy now can take advantage of
+it. Benchmarks have been executed for a few different interpreters and
+we see improvements in several cases. The ideal loop for this optimization
+would be short numerical calculations with no failing guards and no
+external calls.
+
+\subsection{Python}
+The python interpreter of the PyPy framework is a complete python
+version 2.7 compatible interpreter. A set of numerical
+calculations where implemented in both python and in C and their
+runtimes compared. The benchmarks are
+\begin{itemize}
+\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
+  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
+  2$. There are three different versions of this benchmark where $x_k$
+  is represented with different type of objects: int's, float's and
+  Fix16's. The later, Fix16, is a custom class that implements
+  fixpoint arithmetic with 16 bits precision. In python there is only
+  a single implementation of the benchmark that gets specialized
+  depending on the class of it's input argument, $y$, while in C,
+  there is three different implementations.
+\item {\bf conv3}: one dimensional convolution with a kernel of fixed
+  size $3$.
+\item {\bf conv5}: one dimensional convolution with a kernel of fixed
+  size $5$.
+\item {\bf conv3x3}: two dimensional convolution with kernel of fixed
+  size $3 \times 3$ using a custom class to represent two dimensional
+  arrays.
+\item {\bf dilate3x3}: two dimensional dilation with kernel of fixed
+  size $3 \times 3$. This is similar to convolution but instead of
+  summing over the elements, the maximum is taken. That places a
+  external call to a max function within the loop that prevents some
+  of the optimizations.
+\item {\bf sobel}: an low level video processing algorithm used to
+  locate edges in an image. It calculated the gradient magnitude
+  using sobel derivatives. The algorithm is in python implemented
+  on top of a custom image class that is specially designed for the
+  problem. It ensures that there will be no failing guards, and makes
+  a lot of the two dimension index calculations loop invariant. The
+  intention there is twofold. It shows that the performance impact of
+  having wrapper classes giving objects some application specific
+  properties is negligible. This is due to the inlining performed
+  during the tracing and the allocation removal of the index objects
+  introduced. It also shows that it is possible to do some low level
+  hand optimizations of the python code and hide those optimization
+  under a nice interface without loosing performance.
+\end{itemize}
+
+\subsection{Numpy}
+XXX: Fijal?
+
+\subsection{Prolog}
+XXX: Carl?
+
 \appendix

### 2011-06-12 talk/iwtc11/paper.tex
fix some formatting stuff, start to import some papers that we will need into the bib file

--- 2011-06-12 talk/iwtc11/paper.tex
@@
 
+\usepackage{ifthen}
+\usepackage{fancyvrb}
+\usepackage{color}
+\usepackage{ulem}
+\usepackage{xspace}
+\usepackage{epsfig}
+\usepackage{amssymb}
 \usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage[utf8]{inputenc}
+\usepackage{setspace}
+
+\usepackage{listings}
+
+\usepackage[T1]{fontenc}
 \usepackage{setspace}
 \usepackage{listings}
+\usepackage{beramono}
+
+\definecolor{gray}{rgb}{0.3,0.3,0.3}
+
+\lstset{
+  basicstyle=\setstretch{1.05}\ttfamily\footnotesize,
+  language=Python,
+  keywordstyle=\bfseries,
+  stringstyle=\color{blue},
+  commentstyle=\color{gray}\textit,
+  fancyvrb=true,
+  showstringspaces=false,
+  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
+  numberstyle = \tiny,
+  numbersep = -20pt,
+}
+
+\newboolean{showcomments}
+\setboolean{showcomments}{true}
+\ifthenelse{\boolean{showcomments}}
+  {\newcommand{\nb}[2]{
+    \fbox{\bfseries\sffamily\scriptsize#1}
+    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
+   }
+   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
+  }
+  {\newcommand{\nb}[2]{}
+   \newcommand{\version}{}
+  }
+
+\newcommand\cfbolz[1]{\nb{CFB}{#1}}
+\newcommand\arigo[1]{\nb{AR}{#1}}
+\newcommand\fijal[1]{\nb{FIJAL}{#1}}
+\newcommand\david[1]{\nb{DAVID}{#1}}
+\newcommand\anto[1]{\nb{ANTO}{#1}}
+\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
+\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
+\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
+\newcommand{\commentout}[1]{}
+\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
+
+\newcommand\ie{i.e.,\xspace}
+\newcommand\eg{e.g.,\xspace}
+\newcommand{\etal}{\emph{et al.}\xspace}
+
+\normalem
+
+\let\oldcite=\cite
+
+\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}
 
@@
 
-\conferenceinfo{WXYZ '05}{date, City.} 
-\copyrightyear{2005} 
+\conferenceinfo{IWTC '11}{XXX} 
+\copyrightyear{2011} 
 \copyrightdata{[to be supplied]}
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
+\titlebanner{draft}        % These are ignored unless
+%\preprintfooter{short description of paper}   % 'preprint' option specified.
 
-\title{Title Text}
-\subtitle{Subtitle Text, if any}
+\title{Loop Invariant Code Motion in PyPy's Tracing JIT}
+%\subtitle{Subtitle Text, if any}
 
-\authorinfo{Name1}
+\authorinfo{Hakan Ardo XXX}
            {Affiliation1}
-           {Email1}
-\authorinfo{Name2\and Name3}
-           {Affiliation2/3}
-           {Email2/3}
+           {hakan@debian.org}
+\authorinfo{Carl Friedrich Bolz}
+           {Heinrich-Heine-Universität Düsseldorf}
+           {cfbolz@gmx.de}
 
@@
 
-\category{CR-number}{subcategory}{third-level}
+\category{D.3.4}{Programming Languages}{Processors}[code generation,
+incremental compilers, interpreters, run-time environments]
 
 \terms
-term1, term2
+Languages, Performance, Experimentation
 
-\keywords
-keyword1, keyword2
+\keywords{Tracing JIT, Optimization, Loop-Invariant Code Motion}
 
@@
 
-Using these classes to implement arithmetic shows the basic problem of a
-dynamic language implementation. All the numbers are instances of either
+Using these classes to implement arithmetic shows the basic problem of many
+dynamic language implementations. All the numbers are instances of either
 \lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
@@
 
-\subsection{Allocation removal}
-By using escape analysis it is possible to identify objects that are
-allocated within the loop but never escapes it. That is the object are
-short lived and no references to them exists outside the loop. This
-is performed by processing the operation from top to bottom and
-optimistically removing every \lstinline{new} operation. Later on if
-it is discovered that a reference to the object escapes the loop, the
-\lstinline{new} operation is inserted at this point. All operations
-(\lstinline{get} and \lstinline{set}) on the removed objects are also
-removed and the optimizer needs to keep track of the value of all
-attributes of the object.
+\subsection{Virtualization}
+Using escape analysis we can XXX
 
-Consider again the original unoptimized trace of
-Figure~\label{fig:peeled-trace}. Line 10 contains the first
-allocation. It is removed and $p_5$ is marked as virtual. This means
-that it refers to an virtual object that was not yet
-(and might never be) allocated. Line 12 sets the \lstinline{intval}
-attribute of $p_5$. This operation is also removed and the optimizer
-registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.
-
-When the optimizer reaches line 13 it needs to construct the
-arguments for the \lstinline{jump} operation, which contains the virtual
-reference $p_5$. This can be achieved by exploding $p_5$ into it's
-attributes. In this case there is only one attribute and it's value is
-$i_4$, which means the $p_5$ is replaced with $i_4$ in the jump
-arguments.
-
-In the general case, each virtual in the jump arguments is exploded into a
-vector of variables containing the values of all it's attributes. If some
-of the attributes are themselves virtuals they are recursively exploded
-to make the vector contain only non virtual variables. Some care has
-to be taken to always place the attributes in the same order when
-performing this explosion. Notation becomes somewhat simpler if also every non
-virtual variable of the jump arguments is exploded into a vector. This will
-be a vector containing the original variable only. To summarize, for
-every variable, $J_k$, of the original jump arguments, $J$, let
+Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
+same order). Extend it with all non virtual fields, $H_i$, of the
+removed virtuals,
 \begin{equation}
-  \tilde J^{\left(k\right)} = \left\{
-      \begin{array}{ll}
-        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
-        H^{\left(k\right)} & \text{if $J_k$ is virtual}
-      \end{array}
-  \right.
-  ,
+  \hat J = \left(\tilde J_1, \tilde J_2, \cdots, \tilde J_{|\tilde J|}, 
+                 H_1, H_2, \cdots, H_{|H}\right)
 \end{equation}
-where $H^{\left(k\right)}$ is a vector containing all non virtual
-attributes of $J_k$. The arguments of the optimized \lstinline{jump}
-operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
-\begin{equation}
-  \hat J = \left( 
-    \begin{array}{cccc}
-      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
-      \tilde J^{\left(|J|\right)} \\
-    \end{array}
-  \right)      
-  .
-\end{equation}
-and the arguments of the \lstinline{jump} operation of the second
-operation, $K$, are replaced by inlining $\hat J$, 
+and let
 \begin{equation}
@@
 \end{equation}
-In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
-K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into
 
@@
         $i_{4}$ = int_add($i_{2}$, $i_{3}$)
-            # inside BoxedInteger.__init__
-jump($l_1$, $p_{0}$, $i_{4}$)
+jump($l_1$, $p_{0}$, $i_3$, $i_4$)
 
-$l_1$($p_{0}$, $i_{4}$):
+$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
 # inside f: y = y.add(step)
     # inside BoxedInteger.add
-    guard_class($p_{0}$, BoxedInteger)
         # inside BoxedInteger.add__int
-        $i_{7}$ = get($p_{0}$, intval)
-        $i_{8}$ = int_add($i_{4}$, $i_{7}$)
-            # inside BoxedInteger.__init__
-jump($l_1$, $p_{0}$, $i_8$)
+        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
+jump($l_1$, $p_{0}$, $i_3$, $i_8$)
 \end{lstlisting}
 
-Note that virtuals are only exploded into their attributes when
-constructing the arguments of the jump of the first iteration. This
-explosion can't be repeated when constructing the arguments of the
-jump of the second iteration as it has to mach the first. This means
-the objects that was passed as pointers (non virtuals) from the first
-iteration to the second also has to be passed as pointers from the
-second iteration to the third. If one of these objects are virtual
-at the end of the second iteration they need to be allocated right
-before the jump.
+And we're down to a single integer addition!
 
@@
 
-\appendix
-\section{Appendix Title}
+%\appendix
+%\section{Appendix Title}
 
-This is the text of the appendix, if you need one.
+%This is the text of the appendix, if you need one.
 
@@
 \bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\begin{thebibliography}{}
-\softraggedright
-
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{thebibliography}
+\bibliography{paper}
 

### 2011-06-12 talk/iwtc11/paper.tex
the example is from some other paper, cite that

@@
  a very simple object
-model, that just supports an integer and a float type. The objects support only
+model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
 two operations, \lstinline{add}, which adds two objects (promoting ints to floats in a

### 2011-06-12 talk/iwtc11/paper.tex
various fixes/rewrites

@@
 \begin{figure}
+XXX the code for is\_positive is missing everywhere
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
@@
 
-\subsection{Loop peeling}
-Loop peeling is achieved by inlining the trace at the end of
-itself. The input arguments of the second iteration are replaced with
-the jump arguments of the first iterations and then the arguments of all
-the operations are updated to operate on the new input arguments. To
-keep the single-assignment form new variables has to be introduced as
-the results of all the operations. The first iteration of the loop
-will end with a jump to the second iteration of the loop while the
-second iteration will end with a jump to itself. This way the first
-copy of the trace only be executed once while the second copy will be
-used for every other iteration. The rationality here is that the
-optimizations below typically will be able to optimize the second copy
-more efficiently than the first. The trace from Figure~\ref{fig:unopt-trace} would
-after this operation become the trace in Figure~\ref{fig:peeled-trace}.
+\subsection{Loop Peeling}
+
+XXX find reference
+
+Loop peeling is achieved prefixing the loop with one iteration of itself. The
+peeled of iteration of the loop will end with a jump to the full loop, which
+ends with a jump to itself. This way the peeled of iteration will only be
+executed once while the second copy will be used for every further iteration.
+The trace from Figure~\ref{fig:unopt-trace} would after this operation become
+the trace in Figure~\ref{fig:peeled-trace}.
 
@@
 
-When applying the following optimizations to this two iteration trace
+When applying the following optimizations to this two-iteration trace
 some care has to taken as to how the jump arguments of both
@@
 treated. It has to be ensured that the second iteration stays a proper
-trace in the sens that the operations within it only operations on
+trace in the sense that the operations within it only operations on
 variables that are either among the input arguments of the second iterations
@@
 A vector of input
-variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non
+variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
 jump operations and a single
@@
 loop peeling there will be a second copy of this trace with input
-variables equal to the jump arguments of the first copy, $J$, and jump
+variables equal to the jump arguments of the peeled copy, $J$, and jump
 arguments $K$. Looking back at our example we have
@@
 To construct the second iteration from the first we also need a
-function, $m$, mapping the variables of the first iteration onto the
+function $m$, mapping the variables of the first iteration onto the
 variables of the second. This function is constructed during the
@@
 \end{equation}
-In the example that means (XXX which notation do we prefer?)
-\begin{equation}
-  m(v) = 
-  \left\{
-    \begin{array}{lcl}
-      p_0 &\text{if}& v=p_0 \\
-      p_5 &\text{if}& v=p_1 \\
-    \end{array}
-  \right.
-  .
-\end{equation}
+In the example that means:
+
 \begin{equation}
@@
 \end{equation}
-Each operation in the trace is inlined in the order they are
-executed. To inline an operation with argument vector 
-$A=\left(A_1, A_2, \cdots, A_{|A|}\right)$ producing the variable $v$
+
+Each operation in the trace is inlined in order.
+To inline an operation $v=op\left(A_1, A_2, \cdots, A_{|A|}\right)$
 a new variable, $\hat v$ is introduced. The inlined operation will
@@
 \begin{equation}
-  \left(m\left(A_1\right), m\left(A_2\right), 
+  \hat v = op\left(m\left(A_1\right), m\left(A_2\right), 
     \cdots, m\left(A_{|A|}\right)\right) . 
@@
 Before the
-next operation is inlined, $m$ is extend by making $m\left(v\right) = \hat
-v$. After all the operations in the example have been inlined we have
+next operation is inlined, $m$ is extend by assigning $m\left(v\right) = \hat
+v$. For the example above, after all the operations have been inlined we have
 \begin{equation}
@@
 
-\subsection{Redundant guard removal}
+\subsection{Redundant Guard Removal}
+
 No special concerns needs to be taken when implementing redundant
-guard removal together with loop peeling. However the the guards from
+guard removal together with loop peeling. The guards from
 the first iteration might make the guards of the second iterations
-redundant and thus removed. So the net effect of combining redundant
-guard removal with loop peeling is that guards are moved out of the
-loop. The second iteraton of the example reduces to
+redundant and thus removed. Therefore the net effect of combining redundant
+guard removal with loop peeling is that loop-invariant guards are moved out of the
+loop. The second iteration of the example reduces to
 
@@
 
-\subsection{Heap caching}
+\subsection{Heap Caching}
 The objective of heap caching is to remove \lstinline{get} and
@@
 \lstinline{get} and \lstinline{set} operations. Exact details of the
-process are outside the scope of this paper We will here assume that
-it works perfectly and only consider the interactions with loop
-peeling.
+process are outside the scope of this paper. We only consider the interaction
+with loop peeling.
 
@@
 After that, the second
-iteration will no longer be proper as it operates on $i_3$ and $i_4$
+iteration will no longer be in SSA form as it operates on $i_3$ and $i_4$
 which are not part of it. The solution is to extend the input
@@
 Implicitly that also extends the jump arguments of the second iteration, $K$,
-since they are the inlined versions of $J$. That is the, $I$ has to
+since they are the inlined versions of $J$. For the example $I$ has to
 be replaced by $\hat I$ which is formed as a concatenation of $I$ and
@@
 The variable $i_7$ will then be replaced by $i_3$ by the heap caching
-algorithm as it has removed the variable $i_7$. XXX: Maybe we should
+optimization as it has removed the variable $i_7$. XXX: Maybe we should
 replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?
@@
 In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
-K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into
+K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to:
 
@@
 
-\subsection{Virtualization}
+\subsection{Allocation Removals}
 Using escape analysis we can XXX
@@
 Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
-same order). Extend it with all non virtual fields, $H_i$, of the
+same order). Extend it with all non-virtual fields, $H_i$, of the
 removed virtuals,

### 2011-06-12 talk/iwtc11/paper.tex

@@
 \subsection{Allocation Removals}
-Using escape analysis we can XXX
+By using escape analysis it is possible to identify objects that are
+allocated within the loop but never escapes it. That is the object are
+short lived and no references to them exists outside the loop. This
+is performed by processing the operation from top to bottom and
+optimistically removing every \lstinline{new} operation. Later on if
+it is discovered that a reference to the object escapes the loop, the
+\lstinline{new} operation is inserted at this point. All operations
+(\lstinline{get} and \lstinline{set}) on the removed objects are also
+removed and the optimizer needs to keep track of the value of all
+attributes of the object.
 
-Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
-same order). Extend it with all non-virtual fields, $H_i$, of the
-removed virtuals,
+Consider again the original unoptimized trace of
+Figure~\label{fig:peeled-trace}. Line 10 contains the first
+allocation. It is removed and $p_5$ is marked as virtual. This means
+that it refers to an virtual object that was not yet
+(and might never be) allocated. Line 12 sets the \lstinline{intval}
+attribute of $p_5$. This operation is also removed and the optimizer
+registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.
+
+When the optimizer reaches line 13 it needs to construct the
+arguments for the \lstinline{jump} operation, which contains the virtual
+reference $p_5$. This can be achieved by exploding $p_5$ into it's
+attributes. In this case there is only one attribute and it's value is
+$i_4$, which means the $p_5$ is replaced with $i_4$ in the jump
+arguments.
+
+In the general case, each virtual in the jump arguments is exploded into a
+vector of variables containing the values of all it's attributes. If some
+of the attributes are themselves virtuals they are recursively exploded
+to make the vector contain only non virtual variables. Some care has
+to be taken to always place the attributes in the same order when
+performing this explosion. Notation becomes somewhat simpler if also every non
+virtual variable of the jump arguments is exploded into a vector. This will
+be a vector containing the original variable only. To summarize, for
+every variable, $J_k$, of the original jump arguments, $J$, let
 \begin{equation}
-  \hat J = \left(\tilde J_1, \tilde J_2, \cdots, \tilde J_{|\tilde J|}, 
-                 H_1, H_2, \cdots, H_{|H}\right)
+  \tilde J^{\left(k\right)} = \left\{
+      \begin{array}{ll}
+        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
+        H^{\left(k\right)} & \text{if $J_k$ is virtual}
+      \end{array}
+  \right.
+  ,
 \end{equation}
-and let
+where $H^{\left(k\right)}$ is a vector containing all non virtual
+attributes of $J_k$. The arguments of the optimized \lstinline{jump}
+operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
+\begin{equation}
+  \hat J = \left( 
+    \begin{array}{cccc}
+      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
+      \tilde J^{\left(|J|\right)} \\
+    \end{array}
+  \right)      
+  .
+\end{equation}
+and the arguments of the \lstinline{jump} operation of the second
+operation, $K$, are replaced by inlining $\hat J$, 
 \begin{equation}
@@
         $i_{4}$ = int_add($i_{2}$, $i_{3}$)
-jump($l_1$, $p_{0}$, $i_3$, $i_4$)
+            # inside BoxedInteger.__init__
+jump($l_1$, $p_{0}$, $i_{4}$)
 
-$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
+$l_1$($p_{0}$, $i_{4}$):
 # inside f: y = y.add(step)
     # inside BoxedInteger.add
+    guard_class($p_{0}$, BoxedInteger)
         # inside BoxedInteger.add__int
-        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
-jump($l_1$, $p_{0}$, $i_3$, $i_8$)
+        $i_{7}$ = get($p_{0}$, intval)
+        $i_{8}$ = int_add($i_{4}$, $i_{7}$)
+            # inside BoxedInteger.__init__
+jump($l_1$, $p_{0}$, $i_8$)
 \end{lstlisting}
 
-And we're down to a single integer addition!
+Note that virtuals are only exploded into their attributes when
+constructing the arguments of the jump of the first iteration. This
+explosion can't be repeated when constructing the arguments of the
+jump of the second iteration as it has to mach the first. This means
+the objects that was passed as pointers (non virtuals) from the first
+iteration to the second also has to be passed as pointers from the
+second iteration to the third. If one of these objects are virtual
+at the end of the second iteration they need to be allocated right
+before the jump. With the simple objects considered in this paper,
+that is not a problem. However in more complicated interpreters such
+an allocation might, in combination with other optimizations, lead
+to additional variables from the first iteration being imported into
+the second. This extends both $\hat J$ and $\hat K$, which means that
+some care has to be taken, when implementing this, to allow $\hat J$ to
+grow while inlining it into $\hat K$.
 
 \section{Benchmarks}
+
+The loop peeling optimization was implemented in the PyPy
+framework. That means that the jit compilers generated for all
+interpreters implemented within PyPy now can take advantage of
+it. Benchmarks have been executed for a few different interpreters and
+we see improvements in several cases. The ideal loop for this optimization
+would be short numerical calculations with no failing guards and no
+external calls.
+
+\subsection{Python}
+The python interpreter of the PyPy framework is a complete python
+version 2.7 compatible interpreter. A set of numerical
+calculations where implemented in both python and in C and their
+runtimes compared. The benchmarks are
+\begin{itemize}
+\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
+  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
+  2$. There are three different versions of this benchmark where $x_k$
+  is represented with different type of objects: int's, float's and
+  Fix16's. The later, Fix16, is a custom class that implements
+  fixpoint arithmetic with 16 bits precision. In python there is only
+  a single implementation of the benchmark that gets specialized
+  depending on the class of it's input argument, $y$, while in C,
+  there is three different implementations.
+\item {\bf conv3}: one dimensional convolution with a kernel of fixed
+  size $3$.
+\item {\bf conv5}: one dimensional convolution with a kernel of fixed
+  size $5$.
+\item {\bf conv3x3}: two dimensional convolution with kernel of fixed
+  size $3 \times 3$ using a custom class to represent two dimensional
+  arrays.
+\item {\bf dilate3x3}: two dimensional dilation with kernel of fixed
+  size $3 \times 3$. This is similar to convolution but instead of
+  summing over the elements, the maximum is taken. That places a
+  external call to a max function within the loop that prevents some
+  of the optimizations.
+\item {\bf sobel}: an low level video processing algorithm used to
+  locate edges in an image. It calculated the gradient magnitude
+  using sobel derivatives. The algorithm is in python implemented
+  on top of a custom image class that is specially designed for the
+  problem. It ensures that there will be no failing guards, and makes
+  a lot of the two dimension index calculations loop invariant. The
+  intention there is twofold. It shows that the performance impact of
+  having wrapper classes giving objects some application specific
+  properties is negligible. This is due to the inlining performed
+  during the tracing and the allocation removal of the index objects
+  introduced. It also shows that it is possible to do some low level
+  hand optimizations of the python code and hide those optimization
+  under a nice interface without loosing performance.
+\end{itemize}
+
+\subsection{Numpy}
+XXX: Fijal?
+
+\subsection{Prolog}
+XXX: Carl?
 

### 2011-06-12 talk/iwtc11/paper.tex
various typos and some XXXs

@@
 
-The original trace (prior too peeling) consists of three parts. 
+The original trace (prior to peeling) consists of three parts.
 A vector of input
@@
 By using escape analysis it is possible to identify objects that are
-allocated within the loop but never escapes it. That is the object are
+allocated within the loop but never escape it. That is the object are
 short lived and no references to them exists outside the loop. This
@@
 of the attributes are themselves virtuals they are recursively exploded
-to make the vector contain only non virtual variables. Some care has
+to make the vector contain only non-virtual variables. Some care has
 to be taken to always place the attributes in the same order when
-performing this explosion. Notation becomes somewhat simpler if also every non
+performing this explosion. Notation becomes somewhat simpler if also every non-
 virtual variable of the jump arguments is exploded into a vector. This will
@@
 The loop peeling optimization was implemented in the PyPy
-framework. That means that the jit compilers generated for all
+framework. That means that the JIT-compilers generated for all
 interpreters implemented within PyPy now can take advantage of
@@
 
+XXX reason why we use small numerical kernels for benchmarks
+
+XXX we either need to explain that we use C++ or consistently use C
+
 \subsection{Python}
-The python interpreter of the PyPy framework is a complete python
+The python interpreter of the PyPy framework is a complete Python
 version 2.7 compatible interpreter. A set of numerical
-calculations where implemented in both python and in C and their
+calculations were implemented in both Python and in C and their
 runtimes compared. The benchmarks are
@@
   is represented with different type of objects: int's, float's and
-  Fix16's. The later, Fix16, is a custom class that implements
-  fixpoint arithmetic with 16 bits precision. In python there is only
+  Fix16's. The latter, Fix16, is a custom class that implements
+  fixpoint arithmetic with 16 bits precision. In Python there is only
   a single implementation of the benchmark that gets specialized
@@
   there is three different implementations.
-\item {\bf conv3}: one dimensional convolution with a kernel of fixed
+\item {\bf conv3}: one-dimensional convolution with a kernel of fixed
   size $3$.
-\item {\bf conv5}: one dimensional convolution with a kernel of fixed
+\item {\bf conv5}: one-dimensional convolution with a kernel of fixed
   size $5$.
-\item {\bf conv3x3}: two dimensional convolution with kernel of fixed
-  size $3 \times 3$ using a custom class to represent two dimensional
+\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
+  size $3 \times 3$ using a custom class to represent two-dimensional
   arrays.
-\item {\bf dilate3x3}: two dimensional dilation with kernel of fixed
+\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
   size $3 \times 3$. This is similar to convolution but instead of
@@
   of the optimizations.
-\item {\bf sobel}: an low level video processing algorithm used to
-  locate edges in an image. It calculated the gradient magnitude
-  using sobel derivatives. The algorithm is in python implemented
+\item {\bf sobel}: a low-level video processing algorithm used to
+  locate edges in an image. It calculates the gradient magnitude
+  using sobel derivatives. In Python the algorithm is implemented
   on top of a custom image class that is specially designed for the
@@
   a lot of the two dimension index calculations loop invariant. The
-  intention there is twofold. It shows that the performance impact of
-  having wrapper classes giving objects some application specific
+  intention there is twofold. It shows that the performance-impact of
+  having wrapper classes giving objects some application-specific
   properties is negligible. This is due to the inlining performed
   during the tracing and the allocation removal of the index objects
-  introduced. It also shows that it is possible to do some low level
-  hand optimizations of the python code and hide those optimization
+  introduced. It also shows that it is possible to do some low-level
+  hand optimizations of the Python code and hide those optimization
   under a nice interface without loosing performance.
 \end{itemize}
+
+XXX we need Psyco numbers
 

### 2011-06-13 talk/iwtc11/paper.tex
Remove int_add and int_gt as well as mention. Use globally acknowledged '+' symbol

--- 2011-06-12 talk/iwtc11/paper.tex
@@
         $i_{3}$ = get($p_{0}$, intval)
-        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
+        $i_{4}$ = $i_{2} + i_{3}$
         $p_{5}$ = new(BoxedInteger)
@@
     (inlined) method call and is followed by the trace of the called method.
-    \item \lstinline{int_add} and \lstinline{int_gt} are integer addition and
-    comparison (``greater than''), respectively.
     \item \lstinline{guard_true} checks that a boolean is true.
@@
         $i_{3}$ = get($p_{0}$, intval)
-        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
+        $i_{4}$ = $i_{2}+i_{3}$
         $p_{5}$ = new(BoxedInteger)
@@
         $i_{7}$ = get($p_{0}$, intval)
-        $i_{8}$ = int_add($i_{6}$, $i_{7}$)
+        $i_{8}$ = $i_{6}+i_{7}$
         $p_{9}$ = new(BoxedInteger)
@@
         # inside BoxedInteger.add__int
-        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
+        $i_{8}$ = $i_{4}+i_{3}$
         $p_{9}$ = new(BoxedInteger)
@@
         $i_{3}$ = get($p_{0}$, intval)
-        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
+        $i_{4}$ = $i_{2}+i_{3}$
             # inside BoxedInteger.__init__
@@
         $i_{7}$ = get($p_{0}$, intval)
-        $i_{8}$ = int_add($i_{4}$, $i_{7}$)
+        $i_{8}$ = $i_{4}+i_{7}$
             # inside BoxedInteger.__init__

### 2011-06-13 talk/iwtc11/paper.tex
Copy-paste of the figure subtitle confused me greatly
Try to clarify a confusing para

@@
 
-Loop peeling is achieved prefixing the loop with one iteration of itself. The
-peeled of iteration of the loop will end with a jump to the full loop, which
-ends with a jump to itself. This way the peeled of iteration will only be
-executed once while the second copy will be used for every further iteration.
+Loop peeling is achieved by copying the traced iteration of the loop.
+The first part (called preamble) finishes with the jump the the second part
+(peeled loop). The second part ends up with the jump to itself. This way
+the preamble will be executed only once while the peeled loop will
+be used for every other iteration.
 The trace from Figure~\ref{fig:unopt-trace} would after this operation become
@@
 \end{lstlisting}
-\caption{An Unoptimized Trace of the Example Interpreter}
+\caption{A peeled trace of the Example Interpreter}
 \label{fig:peeled-trace}

### 2011-06-13 talk/iwtc11/paper.tex
add myself as an author
add XXX
another xxx

@@
            {cfbolz@gmx.de}
+\authorinfo{Maciej Fijałkowski}
+           {Affiliation2}
+           {fijall@gmail.com}
 
@@
 
+XXX should we have a mention where in the previous papers those optimizations
+are discussed? Is the previous XXX precisely about this?
+
 No special concerns needs to be taken when implementing redundant
@@
 \subsection{Heap Caching}
+
+XXX gcc calles this store-sinking and I'm sure there are some
+references in the literature (none at hand though). This is a ``typical''
+compiler optimization.
+
 The objective of heap caching is to remove \lstinline{get} and

### 2011-06-13 talk/iwtc11/paper.tex
typo
another, quite crucial IMO typo

@@
 In the general case, each virtual in the jump arguments is exploded into a
-vector of variables containing the values of all it's attributes. If some
+vector of variables containing the values of all used attributes. If some
 of the attributes are themselves virtuals they are recursively exploded

### 2011-06-13 talk/iwtc11/paper.tex
Explain numpy benchmark, start writing down limitations paragraph. I'm not even sure how much limitations paragraph belongs there.

@@
 
+\section{Limitations}
+
+XXX as of now?
+
+Loop invariant code motion as described has certain amount of limitations
+that prevent it from speeding up larger loops. Those limitations are a target
+of future work and might be lifted. Most important ones:
+
+\begin{itemize}
+\item Bridges are not well supported - if the flow is more complex than a single
+      loop, the bridge might need to jump to the beginning of the preamble,
+      making the optimization ineffective
+\item XXX write about flushing caches at calls?
+\end{itemize}
+
 \section{Benchmarks}
@@
 \subsection{Numpy}
-XXX: Fijal?
 
-\subsection{Prolog}
+As a part of the PyPy project, we implemented small numerical kernel for
+performing matrix operations. The exact extend of this kernel is besides
+the scope of this paper, however the basic idea is to unroll a series of
+array operations into a loop compiled into assembler. LICM is a very good
+optimization for those kind of operations. The example benchmark performs
+addition of five arrays, compiling it in a way that's equivalent to C's:
+
+\begin{figure}
+\begin{lstlisting}[mathescape,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+for (int i = 0; i < SIZE; i++) {
+   res[i] = a[i] + b[i] + c[i] + d[i] + e[i];
+}
+\end{lstlisting}
+\end{figure}
+
+Where $res$, $a$, $b$, $c$, $d$ and $e$ are $double$ arrays.
+
+\Subsection{Prolog}
 XXX: Carl?

### 2011-06-13 talk/iwtc11/paper.tex
cleared out some less relevant and inconsistent parts of the copied example

@@
 object model (see the bottom of Figure~\ref{fig:objmodel}).
-The loop in \lstinline{f} iterates \lstinline{y} times, and computes something in the process.
 Simply running this function is slow, because there are lots of virtual method
@@
 call to \lstinline{add}. These method calls need to check the type of the involved
-objects repeatedly and redundantly. In addition, a lot of objects are created
+objects every iteration. In addition, a lot of objects are created
 when executing that loop, many of these objects are short-lived.
@@
 using the interpreter.
-
-The trace shows the inefficiencies of \lstinline{f} clearly, if one looks at
-the number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard_class}
-operations. The number of \lstinline{guard_class} operation is particularly
-problematic, not only because of the time it takes to run them. All guards also
-have additional information attached that makes it possible to return to the
-interpreter, should the guard fail. This means that too many guard operations also
-consume a lot of memory.
-
-In the rest of the paper we will see how this trace can be optimized using
-partial evaluation.
 

### 2011-06-13 talk/iwtc11/paper.tex
consistenly use "preamble" and "peeled loop" instead of "first interation" and "second iteration"

@@
 is loop invariant code motion. The goal of that is to move as many
-operations as possible out of the loop making them executed only once
+operations as possible out of the loop making them executed at most once
 and not every iteration. This we propose to achieve by loop peeling. It
@@
 
-Loop peeling is achieved by copying the traced iteration of the loop.
+Loop peeling is achieved by appending a copy of the traced iteration at
+the end of the loop. The copy  is inlined to make the two parts form a
+consitant two iteration trace. 
 The first part (called preamble) finishes with the jump the the second part
-(peeled loop). The second part ends up with the jump to itself. This way
+(called peeled loop). The second part ends up with the jump to itself. This way
 the preamble will be executed only once while the peeled loop will
@@
 The trace from Figure~\ref{fig:unopt-trace} would after this operation become
-the trace in Figure~\ref{fig:peeled-trace}.
+the trace in Figure~\ref{fig:peeled-trace}. Line 1-13 shows the
+preamble while line 15-27 shows the peeled loop.
 
@@
 When applying the following optimizations to this two-iteration trace
-some care has to taken as to how the jump arguments of both
-iterations and the input arguments of the second iteration are
-treated. It has to be ensured that the second iteration stays a proper
-trace in the sense that the operations within it only operations on
-variables that are either among the input arguments of the second iterations
-or are produced within the second iterations. To ensure this we need
+some care has to taken as to how the arguments of the two
+\lstinline{jump} operations and the input arguments of the peeled loop are
+treated. It has to be ensured that the peeled loop stays a proper
+trace in the sense that the operations within it only operates on
+variables that are either among its input arguments 
+or produced within the peeled loop. To ensure this we need
 to introduce a bit of formalism.
@@
 loop peeling there will be a second copy of this trace with input
-variables equal to the jump arguments of the peeled copy, $J$, and jump
+variables equal to the jump arguments of the pereamble, $J$, and jump
 arguments $K$. Looking back at our example we have
@@
 To construct the second iteration from the first we also need a
-function $m$, mapping the variables of the first iteration onto the
-variables of the second. This function is constructed during the
+function $m$, mapping the variables of the preamble onto the
+variables of the peeled loop. This function is constructed during the
 inlining. It is initialized by mapping the input arguments, $I$, to
@@
 Each operation in the trace is inlined in order.
-To inline an operation $v=op\left(A_1, A_2, \cdots, A_{|A|}\right)$
+To inline an operation $v=\text{op}\left(A_1, A_2, \cdots, A_{|A|}\right)$
 a new variable, $\hat v$ is introduced. The inlined operation will
-produce $\hat v$ from the input arguments 
+produce $\hat v$ using
 \begin{equation}
-  \hat v = op\left(m\left(A_1\right), m\left(A_2\right), 
+  \hat v = \text{op}\left(m\left(A_1\right), m\left(A_2\right), 
     \cdots, m\left(A_{|A|}\right)\right) . 
@@
 guard removal together with loop peeling. The guards from
-the first iteration might make the guards of the second iterations
+the preamble might make the guards of the peeled loop
 redundant and thus removed. Therefore the net effect of combining redundant
 guard removal with loop peeling is that loop-invariant guards are moved out of the
-loop. The second iteration of the example reduces to
+loop. The peeled loop of the example reduces to
 
@@
 
-The issue at hand is to keep the second iteration a proper
+The issue at hand is to keep the peeled loop a proper
 trace. Consider the \lstinline{get} operation on line 19 of
@@
 
-After that, the second
-iteration will no longer be in SSA form as it operates on $i_3$ and $i_4$
+After that, the peeled loop
+will no longer be in SSA form as it operates on $i_3$ and $i_4$
 which are not part of it. The solution is to extend the input
 arguments, $J$, with those two variables. This will also extend the
-jump arguments of the first iteration, which is also $J$. 
-Implicitly that also extends the jump arguments of the second iteration, $K$,
+jump arguments of the preamble, which is also $J$. 
+Implicitly that also extends the jump arguments of the peeled loop, $K$,
 since they are the inlined versions of $J$. For the example $I$ has to
@@
 In general what is needed is for the heap optimizer is to keep track of
-which variables from the first iterations it reuses in the second
-iteration. It has to construct a vector of such variables $H$ which
+which variables from the preamble it reuses in the peeled loop.
+It has to construct a vector of such variables $H$ which
 can be used to update the input and jump arguments,
@@
 
+\subsection{Pure operation reusage}
+XXX
+
 \subsection{Allocation Removals}
 By using escape analysis it is possible to identify objects that are
-allocated within the loop but never escape it. That is the object are
-short lived and no references to them exists outside the loop. This
-is performed by processing the operation from top to bottom and
+allocated within the loop but never escape it. That is 
+short lived objects with no references outside the loop. This
+is performed by processing the operation in order and
 optimistically removing every \lstinline{new} operation. Later on if
@@
 removed and the optimizer needs to keep track of the value of all
-attributes of the object.
+used attributes of the object.
 
 Consider again the original unoptimized trace of
-Figure~\label{fig:peeled-trace}. Line 10 contains the first
+Figure~\ref{fig:peeled-trace}. Line 10 contains the first
 allocation. It is removed and $p_5$ is marked as virtual. This means
-that it refers to an virtual object that was not yet
+that it refers to an virtual object that was not yet been
 (and might never be) allocated. Line 12 sets the \lstinline{intval}
@@
 When the optimizer reaches line 13 it needs to construct the
-arguments for the \lstinline{jump} operation, which contains the virtual
+arguments of the \lstinline{jump} operation, which contains the virtual
 reference $p_5$. This can be achieved by exploding $p_5$ into it's
@@
 In the general case, each virtual in the jump arguments is exploded into a
-vector of variables containing the values of all used attributes. If some
+vector of variables containing the values of all registered attributes. If some
 of the attributes are themselves virtuals they are recursively exploded
@@
 \end{equation}
-and the arguments of the \lstinline{jump} operation of the second
-operation, $K$, are replaced by inlining $\hat J$, 
+and the arguments of the \lstinline{jump} operation of the peeled loop,
+$K$, constructed by inlining $\hat J$,
 \begin{equation}
@@
 Note that virtuals are only exploded into their attributes when
-constructing the arguments of the jump of the first iteration. This
+constructing the arguments of the jump of the preamble. This
 explosion can't be repeated when constructing the arguments of the
-jump of the second iteration as it has to mach the first. This means
+jump of the peeled loop as it has to mach the first. This means
 the objects that was passed as pointers (non virtuals) from the first
-iteration to the second also has to be passed as pointers from the
-second iteration to the third. If one of these objects are virtual
-at the end of the second iteration they need to be allocated right
+iteration to the second (from preamble to peeled loop) also has to be
+passed as pointers from the second iteration to the third (from peeled
+loop to peeled loop). If one of these objects are virtual 
+at the end of the peeled loop they need to be allocated right
 before the jump. With the simple objects considered in this paper,
@@
 some care has to be taken, when implementing this, to allow $\hat J$ to
-grow while inlining it into $\hat K$.
+grow while inlining it into $\hat K$. XXX: Maybe we can skip this?
 
@@
   depending on the class of it's input argument, $y$, while in C,
-  there is three different implementations.
+  there are three different implementations.
 \item {\bf conv3}: one dimensional convolution with a kernel of fixed
@@
   a lot of the two dimension index calculations loop invariant. The
-  intention there is twofold. It shows that the performance impact of
+  intention here is twofold. It shows that the performance impact of
   having wrapper classes giving objects some application specific
-  properties is negligible. This is due to the inlining performed
+  properties can be negligible. This is due to the inlining performed
   during the tracing and the allocation removal of the index objects
@@
 
-\Subsection{Prolog}
+\subsection{Prolog}
 XXX: Carl?

### 2011-06-13 talk/iwtc11/paper.tex
pure operations

@@
 \authorinfo{Hakan Ardo XXX}
-           {Affiliation1}
+           {Centre for Mathematical Sciences, Lund University}
            {hakan@debian.org}
@@
 It has to construct a vector of such variables $H$ which
-can be used to update the input and jump arguments,
+can be used to update the input and jump arguments using
 \begin{equation}
   \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
+  \label{eq:heap-inputargs}
 \end{equation}
+and
 \begin{equation}
@@
   .
+  \label{eq:heap-jumpargs}
 \end{equation}
@@
 \subsection{Pure operation reusage}
-XXX
+If a pure operation appears more than once in the trace with same input
+arguments, it only needs be executed the first time and then the result
+can be reused for all other appearances. When that is combined with loop
+peeling, the single execution of the operation is placed in the
+preamble. That is, loop invariant pure operations are moved out of the
+loop. The interactions here are the same as in the previous
+section. That is, a vector, $H$, of variables produced in the preamble
+and used in the peeled loop needs to be constructed. Then the jump and
+input arguments are updated according to
+Equation~\ref{eq:heap-inputargs} and Equation~\ref{eq:heap-jumpargs}.
 

### 2011-06-14 talk/iwtc11/paper.tex

@@
   there are three different implementations.
-\item {\bf conv3}: one dimensional convolution with a kernel of fixed
+\item {\bf conv3}: one-dimensional convolution with a kernel of fixed
   size $3$.
-\item {\bf conv5}: one dimensional convolution with a kernel of fixed
+\item {\bf conv5}: one-dimensional convolution with a kernel of fixed
   size $5$.
-\item {\bf conv3x3}: two dimensional convolution with kernel of fixed
-  size $3 \times 3$ using a custom class to represent two dimensional
+\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
+  size $3 \times 3$ using a custom class to represent two-dimensional
   arrays.
-\item {\bf dilate3x3}: two dimensional dilation with kernel of fixed
+\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
   size $3 \times 3$. This is similar to convolution but instead of
@@
   a lot of the two dimension index calculations loop invariant. The
-  intention here is twofold. It shows that the performance impact of
-  having wrapper classes giving objects some application specific
+  intention here is twofold. It shows that the performance-impact of
+  having wrapper classes giving objects some application-specific
   properties can be negligible. This is due to the inlining performed
   during the tracing and the allocation removal of the index objects
-  introduced. It also shows that it is possible to do some low level
-  hand optimizations of the python code and hide those optimization
+  introduced. It also shows that it is possible to do some low-level
+  hand optimizations of the Python code and hide those optimization
   under a nice interface without loosing performance.
 \end{itemize}
+
+XXX we need Psyco numbers
 

### 2011-06-14 talk/iwtc11/paper.tex
drafted some conclutions

@@
 
+\subsection{Conclutions}
+In this paper we have studied loop invariant code motion during trace
+compilation. We claim that loop peeling is a very convenient solution
+here since it fits well with other trace optimizations. By peeling of
+the first iteration and optimizing the resulting two iteration trace
+as a single trace, several standard optimizations can be 
+used unchanged. The only interaction needed between the loop peeling
+and the other 
+optimizations is during the constructing of the jump arguments
+connecting the peeled of iteration (the preamble) with the loop body. This
+approach 
+turns standard optimizations such as redundant guard removal, heap
+caching, pure operation reusage and allocation removals into loop
+invariant code motion optimizations.
+
+XXX: is ``loop body'' or ``peeled loop'' the preferable term?
+
+By using several benchmarks we show that the proposed algorithm can
+significantly
+improve the run time of small loops containing numerical
+calculations. 
+At least in cases where there are not too many guard
+failures. The standard way of handling guards that fail often is to 
+trace a bridge from it back to the start of some previously compiled
+loop. This is applicable here too. However the bridge will have to end
+with a jump to the preamble, which lessens the impact of the 
+proposed approach.
+In many situations it is however possible to make the bridge
+jump to the peeled loop instead. When and how this is possible will be
+focus of future work.
+
 %\appendix

### 2011-06-14 talk/iwtc11/paper.tex
typo
spellchecker

@@
 the end of the loop. The copy  is inlined to make the two parts form a
-consitant two iteration trace. 
+consistent two iteration trace. 
 The first part (called preamble) finishes with the jump the the second part
@@
 loop peeling there will be a second copy of this trace with input
-variables equal to the jump arguments of the pereamble, $J$, and jump
+variables equal to the jump arguments of the preamble, $J$, and jump
 arguments $K$. Looking back at our example we have
@@
 
-XXX gcc calles this store-sinking and I'm sure there are some
+XXX gcc calls this store-sinking and I'm sure there are some
 references in the literature (none at hand though). This is a ``typical''
@@
 
-\subsection{Pure operation reusage}
+\subsection{Pure operation reuse}
 If a pure operation appears more than once in the trace with same input
@@
 
-\subsection{Conclutions}
+\subsection{Conclusions}
 In this paper we have studied loop invariant code motion during trace
@@
 turns standard optimizations such as redundant guard removal, heap
-caching, pure operation reusage and allocation removals into loop
+caching, pure operation reuse and allocation removals into loop
 invariant code motion optimizations.

### 2011-06-15 talk/iwtc11/paper.tex
start drafting an intro very slowly

@@
 
-The text of the paper begins here.
+One of the advantages that tracing JIT compilers have above traditional tracing
+JITs is that their optimizers are much easier to write. Because a tracing JIT
+produces only linear pieces of code without control flow joins, many
+optimization passes on traces can have a very simple structure. They often
+consist of one forward pass through the trace, discarding and replacing
+operations in the trace by simpler ones as they walk along it. This makes
+optimization of traces be very similar to symbolic execution. Many
+difficult problems in traditional optimizers become tractable if the optimizer
+does not need to deal with control flow merges.
+
+One disadvantage of this simplicity is that such simple forward-passing
+optimizers ignore the only bit of control flow they have available, which is
+the fact that most traces actually represent loops. Making use of this
+information is necessary to perform such optimizations as loop-invariant code
+motion or optimizations that improve across several iterations of the loop.
+Having to deal with this property of traces makes optimizers immediately more
+complex, as a more global view of a trace needs to be considered when
+optimizing.
+
+In this paper we want to address this problem by proposing a simple scheme that
+makes it possible to turn simple optimizations using one forward pass into
+optimizations that can do loop invariant code motion and similar loop-aware
+improvements. Using this scheme one does not need to change the underlying
+optimization much to get these advantages.
+
+XXX "bridge" is not a standard term
 
@@
     \item \lstinline{set} writes to an attribute of an object.
-    \item \lstinline{guard_class} is a precise type check and precedes an
-    (inlined) method call and is followed by the trace of the called method.
+    \item \lstinline{guard_class} is a precise type check. It typically precedes
+    an (inlined) method call and is followed by the trace of the called method.
     \item \lstinline{guard_true} checks that a boolean is true.

### 2011-06-15 talk/iwtc11/paper.tex
tweaks

@@
 
-One of the advantages that tracing JIT compilers have above traditional tracing
+One of the advantages that tracing JIT compilers have above traditional
+method-based
 JITs is that their optimizers are much easier to write. Because a tracing JIT
@@
 operations in the trace by simpler ones as they walk along it. This makes
-optimization of traces be very similar to symbolic execution. Many
+optimization of traces very similar to symbolic execution. Many
 difficult problems in traditional optimizers become tractable if the optimizer
@@
 the fact that most traces actually represent loops. Making use of this
-information is necessary to perform such optimizations as loop-invariant code
+information is necessary to perform optimizations that take the whole loop into
+account, such as loop-invariant code
 motion or optimizations that improve across several iterations of the loop.
-Having to deal with this property of traces makes optimizers immediately more
-complex, as a more global view of a trace needs to be considered when
-optimizing.
+Having to deal with this property of traces complicates the optimization passes,
+as a more global view of a trace needs to be considered when optimizing.
 

### 2011-06-15 talk/iwtc11/paper.tex
remove is_positive fully

@@
 two operations, \lstinline{add}, which adds two objects (promoting ints to floats in a
-mixed addition) and \lstinline{is_positive}, which returns whether the number is greater
-than zero. The implementation of \lstinline{add} uses classical Smalltalk-like
+mixed addition). The implementation of \lstinline{add} uses classical Smalltalk-like
 double-dispatching.
@@
 
-   def is_positive(self):
-      return self.intval > 0
-
 class BoxedFloat(Base):
@@
 
-   def is_positive(self):
-      return self.floatval > 0.0
-
 def f(y):
    step = BoxedInteger(-1)
-   while y.is_positive():
+   while True:
       y = y.add(step)
-   return res
 \end{lstlisting}
@@
 Simply running this function is slow, because there are lots of virtual method
-calls inside the loop, one for each \lstinline{is_positive} and even two for each
+calls inside the loop, two for each
 call to \lstinline{add}. These method calls need to check the type of the involved
@@
 The actual computation that is performed by \lstinline{f} is simply a sequence of
-float or integer additions.
+float or integer additions (note that \lstinline{f} does not actually terminate,
+but it is still instructive to look at the produced traces).
 
 \begin{figure}
-XXX the code for is\_positive is missing everywhere
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]

### 2011-06-15 talk/iwtc11/paper.tex
fix h?kan's name
an XXX

@@
 
-\authorinfo{Hakan Ardo XXX}
+\authorinfo{H\aa kan Ardö}
            {Centre for Mathematical Sciences, Lund University}
@@
 
-\section{Optimizations}
+\section{Trace Optimizations}
+
+XXX make clear that the preamble is not necessarily the \emph{first} iteration
+of a loop
+
 Before the trace is passed to a backend compiling it into machine code

### 2011-06-16 talk/iwtc11/paper.tex
add a diagram to show loop peeling. first step in trying to convey they intuition better, more to come tomorrow.
tweak

@@
 optimization passes on traces can have a very simple structure. They often
-consist of one forward pass through the trace, discarding and replacing
-operations in the trace by simpler ones as they walk along it. This makes
-optimization of traces very similar to symbolic execution. Many
+consist of one forward pass replacing operations by simpler ones or even
+discarding them as they walk along it. This makes
+optimization of traces very similar to symbolic execution. Also, many
 difficult problems in traditional optimizers become tractable if the optimizer
@@
 \subsection{Loop Peeling}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=1]{figures/overview}
+\end{center}
+\end{figure}
 

### 2011-06-16 talk/iwtc11/paper.tex
loop peeling does more than loop-invariant code motion, add an XXX
use official name of this optimization

@@
 XXX "bridge" is not a standard term
+
+XXX loop peeling does a lot more than loop-invariant code motion
+% take this loop as an example:
+% [i1, i2]
+% i3 = i1 + 1
+% i4 = i2 + 1
+% escape(i4)
+% jump(i2, i3)
+% none of the operations is loop-invariant, but loop peeling will still remove the second addition
 
@@
 
-\subsection{Pure operation reuse}
+\subsection{Common Subexpression Elimination}
 If a pure operation appears more than once in the trace with same input

### 2011-06-16 talk/iwtc11/paper.tex
oops

@@
 
-\section{Trace Optimizations}
+\section{Making Trace Optimizations Loop Aware}
 
@@
 \end{center}
+\caption{Overview of Loop Peeling}
+\label{fig:overview}
 \end{figure}
 
-XXX find reference
+XXX find reference of prior work on this
 
 Loop peeling is achieved by appending a copy of the traced iteration at
-the end of the loop. The copy  is inlined to make the two parts form a
-consistent two iteration trace. 
-The first part (called preamble) finishes with the jump the the second part
-(called peeled loop). The second part ends up with the jump to itself. This way
+the end of the loop. See Figure~\ref{fig:overview}
+The first part (called \emph{preamble}) finishes with the jump the the second part
+(called the \emph{peeled loop}). The second part end with the jump to itself. This way
 the preamble will be executed only once while the peeled loop will
-be used for every other iteration.
-The trace from Figure~\ref{fig:unopt-trace} would after this operation become
-the trace in Figure~\ref{fig:peeled-trace}. Line 1-13 shows the
-preamble while line 15-27 shows the peeled loop.
-
-\begin{figure}
-\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$l_0$($p_{0}$, $p_{1}$):
-# inside f: y = y.add(step)
-guard_class($p_{1}$, BoxedInteger)
-    # inside BoxedInteger.add
-    $i_{2}$ = get($p_{1}$, intval)
-    guard_class($p_{0}$, BoxedInteger)
-        # inside BoxedInteger.add__int
-        $i_{3}$ = get($p_{0}$, intval)
-        $i_{4}$ = $i_{2}+i_{3}$
-        $p_{5}$ = new(BoxedInteger)
-            # inside BoxedInteger.__init__
-            set($p_{5}$, intval, $i_{4}$)
-jump($l_1$, $p_{0}$, $p_{5}$)
-
-$l_1$($p_{0}$, $p_{5}$):
-# inside f: y = y.add(step)
-guard_class($p_{5}$, BoxedInteger)
-    # inside BoxedInteger.add
-    $i_{6}$ = get($p_{5}$, intval)
-    guard_class($p_{0}$, BoxedInteger)
-        # inside BoxedInteger.add__int
-        $i_{7}$ = get($p_{0}$, intval)
-        $i_{8}$ = $i_{6}+i_{7}$
-        $p_{9}$ = new(BoxedInteger)
-            # inside BoxedInteger.__init__
-            set($p_{9}$, intval, $i_{8}$)
-jump($l_1$, $p_{0}$, $p_{9}$)
-\end{lstlisting}
-\caption{A peeled trace of the Example Interpreter}
-\label{fig:peeled-trace}
-\end{figure}
+be used for every further iteration.
 
@@
 
-Each operation in the trace is inlined in order.
-To inline an operation $v=\text{op}\left(A_1, A_2, \cdots, A_{|A|}\right)$
-a new variable, $\hat v$ is introduced. The inlined operation will
-produce $\hat v$ using
+Each operation in the trace is copied in order.
+To copy an operation $v=\text{op}\left(A_1, A_2, \cdots, A_{|A|}\right)$
+a new variable, $\hat v$ is introduced. The copied operation will
+return $\hat v$ using
 \begin{equation}
@@
 Before the
-next operation is inlined, $m$ is extend by assigning $m\left(v\right) = \hat
-v$. For the example above, after all the operations have been inlined we have
+next operation is copied, $m$ is extend by assigning $m\left(v\right) = \hat
+v$. For the example above, after all the operations have been copied we have
 \begin{equation}
@@
 
+The trace from Figure~\ref{fig:unopt-trace} would after this operation become
+the trace in Figure~\ref{fig:peeled-trace}. Line 1-13 shows the
+preamble while line 15-27 shows the peeled loop.
+
+\begin{figure}
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_0$($p_{0}$, $p_{1}$):
+# inside f: y = y.add(step)
+guard_class($p_{1}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{2}$ = get($p_{1}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{3}$ = get($p_{0}$, intval)
+        $i_{4}$ = $i_{2}+i_{3}$
+        $p_{5}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{5}$, intval, $i_{4}$)
+jump($l_1$, $p_{0}$, $p_{5}$)
+
+$l_1$($p_{0}$, $p_{5}$):
+# inside f: y = y.add(step)
+guard_class($p_{5}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{6}$ = get($p_{5}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{7}$ = get($p_{0}$, intval)
+        $i_{8}$ = $i_{6}+i_{7}$
+        $p_{9}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{9}$, intval, $i_{8}$)
+jump($l_1$, $p_{0}$, $p_{9}$)
+\end{lstlisting}
+\caption{A peeled trace of the Example Interpreter}
+\label{fig:peeled-trace}
+\end{figure}
+
+\section{Interaction of Optimizations with Loop Peeling}
+
 \subsection{Redundant Guard Removal}
@@
 XXX should we have a mention where in the previous papers those optimizations
-are discussed? Is the previous XXX precisely about this?
+are discussed?
 

### 2011-06-16 talk/iwtc11/benchmarks/benchmark.sh

@@ -1,0 +1,41 @@
+#!/bin/bash
+
+echo
+echo $*
+if [ "$1" == "gcc" ]; then
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_double.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_long.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_fix16.c
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 100
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 100
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1000
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1000
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/conv3x3.cc 1000000 3
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/conv3x3.cc 1000 1000
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/dilate3x3.cc 1000 1000
+    ./runner.py -n 5 -c "$* -lstdc++" image/sobel.cc 1002 1002
+    rm a.out
+else
+    $* ./runner.py -n 10 sqrt/sqrt.py main int
+    $* ./runner.py -n 10 sqrt/sqrt.py main float
+    $* ./runner.py -n 10 sqrt/sqrt.py main Fix16
+    $* ./runner.py -n 10 convolution/convolution.py conv3 1
+    $* ./runner.py -n 10 convolution/convolution.py conv5 1
+    $* ./runner.py -n 10 convolution/convolution.py conv3 100
+    $* ./runner.py -n 10 convolution/convolution.py conv5 100
+    $* ./runner.py -n 10 convolution/convolution.py conv3 1000
+    $* ./runner.py -n 10 convolution/convolution.py conv5 1000
+    $* ./runner.py -n 10 convolution/convolution.py conv3x3 1000000 3
+    $* ./runner.py -n 10 convolution/convolution.py conv3x3 1000 1000
+    $* ./runner.py -n 10 convolution/convolution.py dilate3x3 1000 1000
+    $* ./runner.py -n 10 image/noborder.py main NoBorderImagePadded
[…]

### 2011-06-16 talk/iwtc11/paper.tex
fixes

@@
 model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
-two operations, \lstinline{add}, which adds two objects (promoting ints to floats in a
+one operation, \lstinline{add}, which adds two objects (promoting ints to floats in a
 mixed addition). The implementation of \lstinline{add} uses classical Smalltalk-like
@@
 assigned a value exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
-to the live variables \lstinline{y} and \lstinline{res} in the while-loop of
+to the live variables \lstinline{y} and \lstinline{step} in the while-loop of
 the original function.
@@
 
-Loop peeling is achieved by appending a copy of the traced iteration at
-the end of the loop. See Figure~\ref{fig:overview}
+Loop peeling is achieved by appending an inlined copy of the traced iteration at
+the end of itselfe. See Figure~\ref{fig:overview}.
 The first part (called \emph{preamble}) finishes with the jump the the second part
@@
 variables equal to the jump arguments of the preamble, $J$, and jump
-arguments $K$. Looking back at our example we have
+arguments $K$. Looking  at the peeled version of our example in Figure~\ref{fig:peeled-trace} we have
 \begin{equation}
@@
 the preamble might make the guards of the peeled loop
-redundant and thus removed. Therefore the net effect of combining redundant
+redundant and thus removed. Therefore one effect of combining redundant
 guard removal with loop peeling is that loop-invariant guards are moved out of the
@@
 
-The guard on $p_5$ on line 17 of Figure~\ref{fig:unopt-trace} can be
+The guard on $p_5$ on line 17 of Figure~\ref{fig:peeled-trace} can be
 removed since $p_5$ is allocated on line 10 with a known class. The
@@
 guard on line 6.
+
+Note that the guard on $p_5$ is removed even though $p_5$ is not loop
+invariant, which shows that loop invariant code motion is not the only
+effect of loop peeling.
 
@@
 trace. Consider the \lstinline{get} operation on line 19 of
-Figure~\ref{fig:unopt-trace}. The result of this operation can be
+Figure~\ref{fig:peeled-trace}. The result of this operation can be
 deduced to be $i_4$ from the \lstinline{set} operation on line
@@
 which variables from the preamble it reuses in the peeled loop.
-It has to construct a vector of such variables $H$ which
+It has to construct a vector, $H$,  of such variables which
 can be used to update the input and jump arguments using
@@
 In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
-K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to:
+K$. The trace from Figure~\ref{fig:peeled-trace} will be optimized to:
 
@@
 \end{lstlisting}
+
+Note how the loop invaraint \lstinline{get} on $p_0$ was moved out of
+the loop, and how the non loop invariant \lstinline{get} on $p_5$ was
+removed entierly.
 
@@
 allocation. It is removed and $p_5$ is marked as virtual. This means
-that it refers to an virtual object that was not yet been
+that it refers to an virtual object that has not yet been
 (and might never be) allocated. Line 12 sets the \lstinline{intval}
@@
 attributes. In this case there is only one attribute and it's value is
-$i_4$, which means the $p_5$ is replaced with $i_4$ in the jump
+$i_4$, which means that $p_5$ is replaced with $i_4$ in the jump
 arguments.
@@
 \end{equation}
-and the arguments of the \lstinline{jump} operation of the peeled loop,
-$K$, constructed by inlining $\hat J$,
+The arguments of the \lstinline{jump} operation of the peeled loop,
+$K$, is constructed by inlining $\hat J$,
 \begin{equation}
@@
 explosion can't be repeated when constructing the arguments of the
-jump of the peeled loop as it has to mach the first. This means
+jump of the peeled loop as it has to mach the first jump. This means that
 the objects that was passed as pointers (non virtuals) from the first
@@
 an allocation might, in combination with other optimizations, lead
-to additional variables from the first iteration being imported into
+to additional variables from the preamble being imported into
 the second. This extends both $\hat J$ and $\hat K$, which means that
@@
 approach 
-turns standard optimizations such as redundant guard removal, heap
-caching, pure operation reuse and allocation removals into loop
+improves the effect of standard optimizations such as redundant guard removal, heap
+caching, common subexpression elimination and allocation removals. The
+most prominent effect is that they all become loop
 invariant code motion optimizations.
@@
 At least in cases where there are not too many guard
-failures. The standard way of handling guards that fail often is to 
+failures. A common way of handling a guard that fails often is to 
 trace a bridge from it back to the start of some previously compiled

### 2011-06-16 talk/iwtc11/paper.tex
a motivation example

@@
 
-\subsection{Running Example}
+\section{Motivation}
+\label{sec:Motivation}
+
+To motivate the approach we propose here, let's look at a trivial (unrealistic)
+trace which corresponds to an infinite loop:
+
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_0$($i_{0}$):
+$i_1$ = $i_0$ + 1
+print($i_1$)
+jump($l_0$, $i_0$)
+\end{lstlisting}
+
+The first line is a label $l_0$ with argument $i_0$. Every label has a list of
+arguments. The \lstinline{print} operation just prints its argument (it is not
+an operation that PyPy's tracing JIT really supports, we just use it for this
+example). The \lstinline{jump} operation jumps back to the beginning of the
+trace, listing the new values of the arguments of the trace. In this case, the
+new value of $i_0$ is $i_0$, making it a loop-invariant.
+
+Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
+However, we want to get this effect using our existing optimization passes
+without changing them too much. To achieve this, we peel one iteration off the
+loop before running the optimizations. This peeling gives the following trace:
+
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_0$($i_{0}$):
+$i_1$ = $i_0$ + 1
+print($i_1$)
+jump($l_1$, $i_0$)
+
+$l_1$($i_{0}$):
+$i_2$ = $i_0$ + 1
+print($i_2$)
+jump($l_1$, $i_0$)
+\end{lstlisting}
+
+The iteration of the loop that was peeled off (lines 1-4) is called the
+\emph{preamble}, the loop afterwards the \emph{peeled loop}.
+
+Now the optimizer optimizes both of these two iterations of the loop together,
+disregarding the \lstinline{jump} and the label in lines 4-6. Doing this, common
+subexpression elimination will discover that the two additions are the same, and
+replace $i_2$ with $i_1$. This leads to the following trace:
+
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_0$($i_{0}$):
+$i_1$ = $i_0$ + 1
+print($i_1$)
+jump($l_1$, $i_0$)
+
+$l_1$($i_{0}$):
+print($i_1$)
+jump($l_1$, $i_0$)
+\end{lstlisting}
+
+This trace is malformed, because $i_1$ is used after the label $l_1$ without
+being passed there, so we need to add $i_1$ as an argument to the label and pass
+it along the \lstinline{jump}s:
+
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$l_0$($i_{0}$):
+$i_1$ = $i_0$ + 1
+print($i_1$)
+jump($l_1$, $i_0$, $i_1$)
+
+$l_1$($i_{0}$, $i_1$):
+print($i_1$)
+jump($l_1$, $i_0$, $i_1$)
+\end{lstlisting}
+
+The final result is that the loop-invariant code was moved out of the loop into
+the peeled-off iteration. Thus the addition is only executed in the first
+iteration, while the result is reused in all further iterations.
+
+This scheme is quite powerful. It allows simple linear optimization passes to
+perform loop-aware optimizations, such as loop-invariant code motion without
+changing them at all. All that is needed is to peel off one iteration, then
+apply simple one-pass optimizations and make sure that the necessary extra
+arguments are inserted into the label of the loop itself and the jumps
+afterwards. The peeling off of one iteration gives the optimization enough
+context to remove operations from the peeled loop, because the optimization
+detects that the operation was performed in the preamble already.
+
+% section Motivation (end)
+
+\section{Running Example}
 \label{sub:example}
@@
 explosion can't be repeated when constructing the arguments of the
-jump of the peeled loop as it has to mach the first jump. This means that
+jump of the peeled loop as it has to match the first jump. This means that
 the objects that was passed as pointers (non virtuals) from the first

### 2011-06-16 talk/iwtc11/benchmarks/convolution/conv3x3.cc
killed inline

@@ -1,0 +1,46 @@
+// A safe array example.
+#include <stdio.h>
+#include <stdlib.h>
+
+class Array2D {
+  double *data;
+public:
+  int width, height;
+  Array2D(int w, int h) {
+    width = w;
+    height = h;
+    data = (double *) malloc(w*h*sizeof(double));
+  }
+  double &operator()(int x, int y) {
+    if (x >= 0 && x < width && y >= 0 && y < height) {
+	return data[y*width + x];
+    }
+    printf("IndexError\n");
+    exit(1);
+  }
+};
+
+void conv3x3(Array2D &a, Array2D &k, Array2D &b) {
+  int x, y;
+  for (y=1; y<a.height-1; y++) {
+    for (x=1; x<a.width-1; x++) {
+      b(x, y) = k(2,2)*a(x-1, y-1) + k(1,2)*a(x, y-1) + k(0,2)*a(x+1, y-1) +
+	        k(2,1)*a(x-1, y)   + k(1,1)*a(x, y)   + k(0,1)*a(x+1, y)   + 
+	        k(2,0)*a(x-1, y+1) + k(1,0)*a(x, y+1) + k(0,0)*a(x+1, y+1);
+
+    }
+  }
+}
+
+int main(int ac, char **av) {
+  int w = atoi(av[1]), h = atoi(av[2]);
+  int i;
+
+  for (i=0; i<10; i++) {
+    Array2D a(w, h), b(w, h), k(3, 3);
+    conv3x3(a, k, b);
+    printf("%f\n", b(1,1));
+  }
+  fprintf(stderr, "conv3x3(%d):  ", h);
+  return 0;
+}

### 2011-06-16 talk/iwtc11/benchmarks/image/noborder.py

@@ -1,0 +1,183 @@
+from array import array
+
+class NoBorderImage(object):
+    "An image class for people who dont care about border effects"
+    
+    def __init__(self, w, h, typecode='d', fromfile=None):
+        self.width = w
+        self.height = h
+        if fromfile is not None:
+            self.data = array(typecode)
+            self.data.fromfile(fromfile, w*h)
+        else:
+            self.data = array(typecode, [0]) * (w*h)
+        self.typecode = typecode
+
+    def _idx(self, p):
+        if isinstance(p, Pixel):
+            assert p.image.__class__ is self.__class__
+            assert p.image.width == self.width
+            idx = p.idx
+        else:
+            idx = p[1] * self.width + p[0]
+        return min(max(idx, 0), len(self.data)-1)
+
+    def __getitem__(self, p):
+        return self.data[self._idx(p)]
+
+    def __setitem__(self, p, val):
+        self.data[self._idx(p)] = val
+
+    def pixels(self):
+        for i in self.pixelrange():
+            yield Pixel(i, self)
+
+    def pixeliter(self):
+        return PixelIter(self)
+
+    def pixelrange(self):
+        return xrange(self.width * self.height)
+
+    def setup(self, data):
+        for y in xrange(self.height):
+            for x in xrange(self.width):
+                self[x, y] = data[y][x]
+        return self
+
+    def clone(self, **kwargs):
+        return self.__class__(self.width, self.height, **kwargs)
+
+    def tofile(self, f):
+        self.data.tofile(f)
+
+class NoBorderImagePadded(NoBorderImage):
+    def __init__(self, w, h, typecode='d', fromfile=None):
+        self.width = w
+        self.height = h
+        self.typecode = typecode
+        if fromfile is None:
+            self.data = array(typecode, [0]) * (w*(h+2)+2)
+        else:
+            self.data = array(typecode, [0]) * (w + 1)
+            self.data.fromfile(fromfile, w*h)
+            self.data += array(typecode, [0]) * (w + 1)
+
+    def _idx(self, p):
+        if isinstance(p, Pixel):
+            assert p.image.__class__ is self.__class__
[…]

### 2011-06-16 talk/iwtc11/paper.tex
killed inline

--- 2011-06-16 talk/iwtc11/paper.tex
@@
 
-Loop peeling is achieved by appending an inlined copy of the traced iteration at
+Loop peeling is achieved by appending an copy of the traced iteration at
 the end of itselfe. See Figure~\ref{fig:overview}.
@@
 the preamble will be executed only once while the peeled loop will
-be used for every further iteration.
+be used for every further iteration. New variable names have to be
+introduced in the entire copied trace in order to maintian the SSA-property.
 
@@
 \end{equation}
-To construct the second iteration from the first we also need a
+To construct the second copy of the trace (the peeled loop) from the
+first (the preeamble) we need a
 function $m$, mapping the variables of the preamble onto the
 variables of the peeled loop. This function is constructed during the
-inlining. It is initialized by mapping the input arguments, $I$, to
+copying. It is initialized by mapping the input arguments, $I$, to
 the jump arguments $J$,
@@
 Implicitly that also extends the jump arguments of the peeled loop, $K$,
-since they are the inlined versions of $J$. For the example $I$ has to
+since they are the image of $J$ under $m$. For the example $I$ has to
 be replaced by $\hat I$ which is formed as a concatenation of $I$ and

### 2011-06-16 talk/iwtc11/paper.tex
we have no guard_true any more

--- 2011-06-16 talk/iwtc11/paper.tex
@@
 the preamble will be executed only once while the peeled loop will
-be used for every further iteration.
-
+be used for every further iteration. New variable names have to be
+introduced in the entire copied trace in order to maintian the SSA-property.
 When applying the following optimizations to this two-iteration trace

### 2011-06-16 talk/iwtc11/paper.tex
clarify and stress things

@@
 
-This scheme is quite powerful. It allows simple linear optimization passes to
+This scheme is quite powerful and generalizes to other optimizations than just
+common subexpression elimination. It allows simple linear optimization passes to
 perform loop-aware optimizations, such as loop-invariant code motion without
@@
 arguments are inserted into the label of the loop itself and the jumps
-afterwards. The peeling off of one iteration gives the optimization enough
-context to remove operations from the peeled loop, because the optimization
-detects that the operation was performed in the preamble already.
+afterwards. Giving the optimizations two iterations together
+gives the optimization enough context to remove operations from the peeled loop,
+because it detects that the operation was performed in the preamble already.
 
@@
     an (inlined) method call and is followed by the trace of the called method.
-    \item \lstinline{guard_true} checks that a boolean is true.
 \end{itemize}

### 2011-06-16 talk/iwtc11/paper.tex
Use uppercase L for loops to prevent mixing up l and I.

@@
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$l_0$($i_{0}$):
+$L_0$($i_{0}$):
 $i_1$ = $i_0$ + 1
 print($i_1$)
-jump($l_0$, $i_0$)
+jump($L_0$, $i_0$)
 \end{lstlisting}
 
-The first line is a label $l_0$ with argument $i_0$. Every label has a list of
+The first line is a label $L_0$ with argument $i_0$. Every label has a list of
 arguments. The \lstinline{print} operation just prints its argument (it is not
@@
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$l_0$($i_{0}$):
+$L_0$($i_{0}$):
 $i_1$ = $i_0$ + 1
 print($i_1$)
-jump($l_1$, $i_0$)
+jump($L_1$, $i_0$)
 
-$l_1$($i_{0}$):
+$L_1$($i_{0}$):
 $i_2$ = $i_0$ + 1
 print($i_2$)
-jump($l_1$, $i_0$)
+jump($L_1$, $i_0$)
 \end{lstlisting}
@@
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$l_0$($i_{0}$):
+$L_0$($i_{0}$):
 $i_1$ = $i_0$ + 1
 print($i_1$)
-jump($l_1$, $i_0$)
+jump($L_1$, $i_0$)
 
-$l_1$($i_{0}$):
+$L_1$($i_{0}$):
 print($i_1$)
-jump($l_1$, $i_0$)
+jump($L_1$, $i_0$)
 \end{lstlisting}
 
-This trace is malformed, because $i_1$ is used after the label $l_1$ without
+This trace is malformed, because $i_1$ is used after the label $L_1$ without
 being passed there, so we need to add $i_1$ as an argument to the label and pass
@@
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$l_0$($i_{0}$):
+$L_0$($i_{0}$):
 $i_1$ = $i_0$ + 1
 print($i_1$)
-jump($l_1$, $i_0$, $i_1$)
+jump($L_1$, $i_0$, $i_1$)
 
-$l_1$($i_{0}$, $i_1$):
+$L_1$($i_{0}$, $i_1$):
 print($i_1$)
-jump($l_1$, $i_0$, $i_1$)
+jump($L_1$, $i_0$, $i_1$)
 \end{lstlisting}
@@
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$l_0$($p_{0}$, $p_{1}$):
+$L_0$($p_{0}$, $p_{1}$):
 # inside f: y = y.add(step)
@@
             set($p_{5}$, intval, $i_{4}$)
-jump($l_0$, $p_{0}$, $p_{5}$)
+jump($L_0$, $p_{0}$, $p_{5}$)
 \end{lstlisting}
@@
 
-The label of the loop is $l_0$ and is used by the jump instruction to
+The label of the loop is $L_0$ and is used by the jump instruction to
 identify it's jump target.
@@
 Loop peeling is achieved by appending an copy of the traced iteration at
-the end of itselfe. See Figure~\ref{fig:overview}.
+the end of itselfe. See Figure~\ref{fig:overview} for an illustration.
 The first part (called \emph{preamble}) finishes with the jump the the second part
@@
             set($p_{5}$, intval, $i_{4}$)
-jump($l_1$, $p_{0}$, $p_{5}$)
+jump($L_1$, $p_{0}$, $p_{5}$)
 
-$l_1$($p_{0}$, $p_{5}$):
+$L_1$($p_{0}$, $p_{5}$):
 # inside f: y = y.add(step)
@@
             set($p_{9}$, intval, $i_{8}$)
-jump($l_1$, $p_{0}$, $p_{9}$)
+jump($L_1$, $p_{0}$, $p_{9}$)
 \end{lstlisting}
@@
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$l_1$($p_{0}$, $p_{5}$):
+$L_1$($p_{0}$, $p_{5}$):
 # inside f: y = y.add(step)
@@
             set($p_{5}$, intval, $i_{4}$)
-jump($l_1$, $p_{0}$, $p_{5}$, $i_3$, $i_4$)
+jump($L_1$, $p_{0}$, $p_{5}$, $i_3$, $i_4$)
 
-$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
+$L_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
 # inside f: y = y.add(step)
@@
             set($p_{9}$, intval, $i_{8}$)
-jump($l_1$, $p_{0}$, $p_{9}$, $i_3$, $i_8$)
+jump($L_1$, $p_{0}$, $p_{9}$, $i_3$, $i_8$)
 \end{lstlisting}
@@
             # inside BoxedInteger.__init__
-jump($l_1$, $p_{0}$, $i_{4}$)
+jump($L_1$, $p_{0}$, $i_{4}$)
 
-$l_1$($p_{0}$, $i_{4}$):
+$L_1$($p_{0}$, $i_{4}$):
 # inside f: y = y.add(step)
@@
             # inside BoxedInteger.__init__
-jump($l_1$, $p_{0}$, $i_8$)
+jump($L_1$, $p_{0}$, $i_8$)
 \end{lstlisting}

### 2011-06-16 talk/iwtc11/paper.tex
make clear that we don't claim novelty

@@
 In this paper we want to address this problem by proposing a simple scheme that
-makes it possible to turn simple optimizations using one forward pass into
+makes it possible to turn optimizations using one forward pass into
 optimizations that can do loop invariant code motion and similar loop-aware
@@
 optimization much to get these advantages.
+
+The resulting optimizations one gets using this scheme are in no way novel, most
+of them are well-known loop optimizations. However, the way to implement them is
+a lot simpler than directly implementing loop-aware optimizations.
 
@@
 Loop peeling is achieved by appending an copy of the traced iteration at
-the end of itselfe. See Figure~\ref{fig:overview} for an illustration.
+the end of itself. See Figure~\ref{fig:overview} for an illustration.
 The first part (called \emph{preamble}) finishes with the jump the the second part

### 2011-06-16 talk/iwtc11/paper.tex
a minimal introduction section on PyPy

@@
 
-\title{Loop Invariant Code Motion in PyPy's Tracing JIT}
+\title{Loop-Aware Optimizations in PyPy's Tracing JIT}
 %\subtitle{Subtitle Text, if any}
@@
 % none of the operations is loop-invariant, but loop peeling will still remove the second addition
+
+\section{Background: PyPy}
+\label{sec:PyPy}
+
+The work described in this paper was done in the context of the PyPy
+project\footnote{\url{http://pypy.org}}. PyPy is a framework for implementing
+dynamic languages efficiently \cite{armin_rigo_pypys_2006}. When implementing a
+language with PyPy, one writes an interpreter for the language in RPython
+\cite{davide_ancona_rpython:_2007}. RPython (``Restricted Python``) is a subset
+of Python chosen in such a way that it can be efficiently translated to a
+C-based VM by performing type inference.
+
+Many low-level aspects of the final VM are not contained within the interpreter
+implementation but are inserted during translation to C. Examples for this are a
+garbage collector and also a tracing JIT compiler \cite{bolz_tracing_2009}.
+
+PyPy's tracing JIT compiler traces on the level of RPython programs. Thus it
+actually traces the execution of an interpreter written in RPython, not of the
+program itself. This makes the details of the object model of the implemented
+language transparent and optimizable by the tracing JIT. In the context of this
+paper, this aspect of PyPy's tracing JIT can be ignored. Instead, it is
+sufficient to view PyPy's tracing JIT as a JIT for RPython.
+
+% section PyPy (end)
 

### 2011-06-16 talk/iwtc11/paper.tex
fix url, use different bibliography style
always optimize original trace

@@
 The work described in this paper was done in the context of the PyPy
-project\footnote{\url{http://pypy.org}}. PyPy is a framework for implementing
+project\footnote{\texttt{http://pypy.org}}. PyPy is a framework for implementing
 dynamic languages efficiently \cite{armin_rigo_pypys_2006}. When implementing a
@@
 # inside f: y = y.add(step)
+guard_class($p_{5}$, BoxedInteger)
     # inside BoxedInteger.add
+    guard_class($p_{0}$, BoxedInteger)
         # inside BoxedInteger.add__int
@@
 optimizations is during the constructing of the jump arguments
-connecting the peeled of iteration (the preamble) with the loop body. This
+connecting the peeled of iteration (the preamble) with the peeled loop. This
 approach 
@@
 invariant code motion optimizations.
-
-XXX: is ``loop body'' or ``peeled loop'' the preferable term?
 
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{abbrv}
 \bibliography{paper}

### 2011-06-16 talk/iwtc11/paper.tex
start describing benchmarks

@@
 
+Benchmarks were run on Intel i7 M620 @2.67GHz with 4M cache and 8G of RAM in
+32bit mode.
+The machine was otherwise unoccupied. We run the following software
+for benchmarks:
+
+\begin{itemize}
+\item PyPy 1.5
+\item CPython 2.7.2
+\item Psyco 1.6 with CPython 2.6.6
+\item GCC 4.4.5 shipped with Ubuntu 11.4
+\end{itemize}
+
+We run GCC both with -O2 optimization and -O3 -march=native, disabling the
+automatic loop vectorization. In all cases, SSE2 instructions were used for
+floating point operations, except Psyco which uses x87 FPU instructions.
+We also run PyPy with loop peeling optimization and without (but otherwise
+identical).
+
+XXX discuss some more, explain why gcc is faster
 XXX we need Psyco numbers

### 2011-06-16 talk/iwtc11/benchmarks/benchmark.sh
start describing benchmarks

@@ -5,7 +5,7 @@ echo $*
 if [ "$1" == "gcc" ]; then
     ./runner.py -n 5 -c "$*" sqrt/sqrt_double.c
     ./runner.py -n 5 -c "$*" sqrt/sqrt_long.c
-    #./runner.py -n 5 -c "$*" sqrt/sqrt_fix16.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_fix16.c
     #./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1
     #./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1
     ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 100
@@ -20,7 +20,7 @@ if [ "$1" == "gcc" ]; then
 else
     $* ./runner.py -n 10 sqrt/sqrt.py main int
     $* ./runner.py -n 10 sqrt/sqrt.py main float
-    #$* ./runner.py -n 10 sqrt/sqrt.py main Fix16
+    $* ./runner.py -n 10 sqrt/sqrt.py main Fix16
     #$* ./runner.py -n 10 convolution/convolution.py conv3 1
     #$* ./runner.py -n 10 convolution/convolution.py conv5 1
     $* ./runner.py -n 10 convolution/convolution.py conv3 100

### 2011-06-16 talk/iwtc11/benchmarks/runall.sh

@@ -1,10 +1,10 @@
-#!/bin/sh
+#!/bin/bash
 
-#./benchmark.sh pypy
-./benchmark.sh pypy --jit enable_opts=intbounds:rewrite:virtualize:heap:unroll
+./benchmark.sh pypy
+#./benchmark.sh pypy --jit enable_opts=intbounds:rewrite:virtualize:heap:unroll
 ./benchmark.sh pypy --jit enable_opts=intbounds:rewrite:virtualize:heap
 #./benchmark.sh gcc
 ./benchmark.sh gcc -O2
 ./benchmark.sh gcc -O3 -march=native -fno-tree-vectorize
-#./benchmark.sh python2.7
-
+./benchmark.sh python2.7
+./benchmark.sh python2.6 psyco-wrapper.py

### 2011-06-17 talk/iwtc11/paper.tex
a para about warming up (and a bench file in-progress)
started to draft abstract

@@
 \begin{abstract}
-This is the text of the abstract.
+By introducing loop peeling into the optimization step of a tracing
+jit, the effect of optimizations already in place will be increased
+greatly. Not only will they become able to move loop invariant code
+out of loop. They will also become able to reuse results from the
+previous iteration. Also, the implementation of excising optimizations
+can be left almost intact as they will not have to interact much with
+the loop peeling.
+
+Several benchmarks executed on the PyPy python JIT show over 2
+times increase in speed when loop peeling was introduced. This makes
+some of them almost match optimized C performance and become over XXX
+times faster than cpython.
 \end{abstract}
@@
 
+For all JITted run (PyPy and Psyco) 3 iterations were run to warm up the JIT,
+followed by 10 iterations for averaging. For GCC and CPython, 5 iterations
+were run. In all cases, the standard deviation is very low, making benchmarks
+very well reproducible.
+
 XXX discuss some more, explain why gcc is faster

### 2011-06-17 talk/iwtc11/paper.tex
clairify

@@
 By introducing loop peeling into the optimization step of a tracing
-jit, the effect of optimizations already in place will be increased
+jit the effect of optimizations already in place will be increased
 greatly. Not only will they become able to move loop invariant code
@@
 
-Several benchmarks executed on the PyPy python JIT show over 2
+Several benchmarks, with few guard failures, executed on the
+PyPy python JIT show over 2 
 times increase in speed when loop peeling was introduced. This makes

### 2011-06-17 talk/iwtc11/paper.tex
minor

@@
   there are three different implementations.
-\item {\bf conv3}: one-dimensional convolution with a kernel of fixed
-  size $3$.
-\item {\bf conv5}: one-dimensional convolution with a kernel of fixed
-  size $5$.
+\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
+\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
 \item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
@@
 32bit mode.
-The machine was otherwise unoccupied. We run the following software
+The machine was otherwise unoccupied. We use the following software
 for benchmarks:
@@
 
-For all JITted run (PyPy and Psyco) 3 iterations were run to warm up the JIT,
+For all JIT:ed runs (PyPy and Psyco) 3 iterations were run to warm up the JIT,
 followed by 10 iterations for averaging. For GCC and CPython, 5 iterations
@@
 
-\begin{figure}
+%\begin{figure}
 \begin{lstlisting}[mathescape,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
@@
 \end{lstlisting}
-\end{figure}
+%\end{figure}
 

### 2011-06-17 talk/iwtc11/paper.tex
small improvements, write para about benchmarks

@@
 
-For all JIT:ed runs (PyPy and Psyco) 3 iterations were run to warm up the JIT,
-followed by 10 iterations for averaging. For GCC and CPython, 5 iterations
+For PyPy 10 iterations were run, prefaced with 3 iterations for warming up.
+Due to benchmarks taking large amounts of time on CPython, only one run
+was performed, prefaced with one warmup run for Psyco.
+For GCC 5 iterations
 were run. In all cases, the standard deviation is very low, making benchmarks
@@
 
-XXX discuss some more, explain why gcc is faster
-XXX we need Psyco numbers
+We can observe that PyPy (even without loop peeling) is orders of magnitude
+faster than either CPython or Psyco. This is due to the JIT compilation
+advantages and optimizations we discussed in XXX [ref to other paper]. Loop
+peeling gives an additional XXX on average, which makes benchmark times
+comparable with native-compiled C code. Missing performance we attribute to
+the relative immaturity of PyPy's JIT assembler backend as well as missing
+optimizations, like instruction scheduling.
 

### 2011-06-20 talk/iwtc11/paper.tex
explain that allocation removal leads to type specialization
Merge the CSE and heap optimization sections to save space, since they say mostly the same thing. From the trace, only remove one of the gets. This makes it easier to explain, and the other one is removed by allocation removal anyway

@@
 Several benchmarks, with few guard failures, executed on the
-PyPy python JIT show over 2 
+PyPy Python JIT show over 2
 times increase in speed when loop peeling was introduced. This makes
 some of them almost match optimized C performance and become over XXX
-times faster than cpython.
+times faster than CPython.
 \end{abstract}
@@
 invariant, which shows that loop invariant code motion is not the only
-effect of loop peeling.
+effect of loop peeling. Loop peeling can also remove guards that are implied by
+the guards of the previous iteration.
 
-\subsection{Heap Caching}
+\subsection{Common Subexpression Elimination and Heap Optimizations}
 
-XXX gcc calls this store-sinking and I'm sure there are some
-references in the literature (none at hand though). This is a ``typical''
-compiler optimization.
+If a pure operation appears more than once in the trace with the same input
+arguments, it only needs be executed the first time and then the result
+can be reused for all other appearances. PyPy's optimizers can also remove
+repeated heap reads if the intermediate operations cannot have changed their
+value\footnote{We perform a simple type-based alias analysis to know which
+writes can affect which reads. In addition writes on newly allocated objects
+can never change the value of old existing ones.}.
 
-The objective of heap caching is to remove \lstinline{get} and
-\lstinline{set} operations whose results can be deduced from previous
-\lstinline{get} and \lstinline{set} operations. Exact details of the
-process are outside the scope of this paper. We only consider the interaction
-with loop peeling.
+When that is combined with loop peeling, the single execution of the operation
+is placed in the preamble. That is, loop invariant pure operations and heap
+reads are moved out of the loop.
 
-The issue at hand is to keep the peeled loop a proper
-trace. Consider the \lstinline{get} operation on line 19 of
+Consider the \lstinline{get} operation on line 22 of
 Figure~\ref{fig:peeled-trace}. The result of this operation can be
-deduced to be $i_4$ from the \lstinline{set} operation on line
-12. Also, the result of the \lstinline{get} operation on line 22 can
-be deduced to be $i_3$ from the \lstinline{get} operation on line
-8. The optimization will thus remove line 19 and 22 from the trace and
-replace $i_6$ with $i_4$ and $i_7$ with $i_3$.
+deduced to be $i_3$ from the \lstinline{get} operation on line
+8. The optimization will thus remove line 22 from the trace and
+replace $i_7$ with $i_3$. Afterwards the trace is no longer in the correct
+form, because the argument $i_3$ is not passed along the loop arguments. It
+thus needs to be added there.
 
-After that, the peeled loop
-will no longer be in SSA form as it operates on $i_3$ and $i_4$
-which are not part of it. The solution is to extend the input
-arguments, $J$, with those two variables. This will also extend the
+The trace from Figure~\ref{fig:peeled-trace} will therefore be optimized to:
+
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$L_0$($p_{0}$, $p_{1}$):
+# inside f: y = y.add(step)
+guard_class($p_{1}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{2}$ = get($p_{1}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{3}$ = get($p_{0}$, intval)
+        $i_{4}$ = $i_{2}+i_{3}$
+        $p_{5}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{5}$, intval, $i_{4}$)
+jump($L_1$, $p_{0}$, $p_{5}$, $i_3$)
+
+$L_1$($p_{0}$, $p_{5}$, $i_3$):
+# inside f: y = y.add(step)
+guard_class($p_{5}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{6}$ = get($p_{5}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{8}$ = $i_{4}+i_{3}$
+        $p_{9}$ = new(BoxedInteger)
+            # inside BoxedInteger.__init__
+            set($p_{9}$, intval, $i_{8}$)
+jump($L_1$, $p_{0}$, $p_{9}$, $i_3$)
+\end{lstlisting}
+
+In general, after loop peeling and redundant operation removal the peeled loop
+will no longer be in SSA form as it operates on variables that are the result
+of pure operations in the preamble. The solution is to extend the input
+arguments, $J$, with those variables. This will also extend the
 jump arguments of the preamble, which is also $J$. 
@@
 since they are the image of $J$ under $m$. For the example $I$ has to
-be replaced by $\hat I$ which is formed as a concatenation of $I$ and
-$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
-$\hat K$ which is formed as a concatenation of $K$ and 
-$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
+be replaced by $\hat I$ which is formed by appending $i_3$ to $I$.
+At the same time $K$ has to be replaced by
+$\hat K$ which is formed by appending $m\left(i_3\right)=i_7$ to $K$.
 The variable $i_7$ will then be replaced by $i_3$ by the heap caching
-optimization as it has removed the variable $i_7$. XXX: Maybe we should
-replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?
+optimization as it has removed the variable $i_7$.
 
-In general what is needed is for the heap optimizer is to keep track of
+In general what is needed is to keep track of
 which variables from the preamble it reuses in the peeled loop.
@@
 In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
-K$. The trace from Figure~\ref{fig:peeled-trace} will be optimized to:
-
-\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-$L_0$($p_{0}$, $p_{1}$):
-# inside f: y = y.add(step)
-guard_class($p_{1}$, BoxedInteger)
-    # inside BoxedInteger.add
-    $i_{2}$ = get($p_{1}$, intval)
-    guard_class($p_{0}$, BoxedInteger)
-        # inside BoxedInteger.add__int
-        $i_{3}$ = get($p_{0}$, intval)
-        $i_{4}$ = $i_{2}+i_{3}$
-        $p_{5}$ = new(BoxedInteger)
-            # inside BoxedInteger.__init__
-            set($p_{5}$, intval, $i_{4}$)
-jump($L_1$, $p_{0}$, $p_{5}$, $i_3$, $i_4$)
-
-$L_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
-# inside f: y = y.add(step)
-guard_class($p_{5}$, BoxedInteger)
-    # inside BoxedInteger.add
-    guard_class($p_{0}$, BoxedInteger)
-        # inside BoxedInteger.add__int
-        $i_{8}$ = $i_{4}+i_{3}$
-        $p_{9}$ = new(BoxedInteger)
-            # inside BoxedInteger.__init__
-            set($p_{9}$, intval, $i_{8}$)
-jump($L_1$, $p_{0}$, $p_{9}$, $i_3$, $i_8$)
-\end{lstlisting}
-
-Note how the loop invaraint \lstinline{get} on $p_0$ was moved out of
-the loop, and how the non loop invariant \lstinline{get} on $p_5$ was
-removed entierly.
-
-\subsection{Common Subexpression Elimination}
-If a pure operation appears more than once in the trace with same input
-arguments, it only needs be executed the first time and then the result
-can be reused for all other appearances. When that is combined with loop
-peeling, the single execution of the operation is placed in the
-preamble. That is, loop invariant pure operations are moved out of the
-loop. The interactions here are the same as in the previous
-section. That is, a vector, $H$, of variables produced in the preamble
-and used in the peeled loop needs to be constructed. Then the jump and
-input arguments are updated according to
-Equation~\ref{eq:heap-inputargs} and Equation~\ref{eq:heap-jumpargs}.
+K$.
 
@@
 
+XXX explain that this is effectively type-specializing a loop
+
 \section{Limitations}
@@
 \subsection{Python}
-The python interpreter of the PyPy framework is a complete Python
+The Python interpreter of the PyPy framework is a complete Python
 version 2.7 compatible interpreter. A set of numerical

### 2011-06-20 talk/iwtc11/paper.tex
go over the allocation-removal section and replace "virtual" by "allocation-removed". Also some small simplifications.

@@
 \subsection{Allocation Removals}
-By using escape analysis it is possible to identify objects that are
-allocated within the loop but never escape it. That is 
-short lived objects with no references outside the loop. This
-is performed by processing the operation in order and
+PyPy's allocation removal optimization \cite{bolz_allocation_2011} makes it
+possible to identify objects that are allocated within the loop but never
+escape it. Those objects have to be allocated in the loop, but no outside
+object ever gets a reference short lived objects with no references outside the
+loop. This
+is performed by processing the operations in order and
 optimistically removing every \lstinline{new} operation. Later on if
@@
 \lstinline{new} operation is inserted at this point. All operations
-(\lstinline{get} and \lstinline{set}) on the removed objects are also
-removed and the optimizer needs to keep track of the value of all
-used attributes of the object.
+(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
+are also removed and the optimizer needs to keep track of the value of all used
+attributes of the object.
 
@@
 Figure~\ref{fig:peeled-trace}. Line 10 contains the first
-allocation. It is removed and $p_5$ is marked as virtual. This means
-that it refers to an virtual object that has not yet been
+allocation. It is removed and $p_5$ is marked as allocation-removed. This means
+that it refers to an object that has not yet been
 (and might never be) allocated. Line 12 sets the \lstinline{intval}
@@
 When the optimizer reaches line 13 it needs to construct the
-arguments of the \lstinline{jump} operation, which contains the virtual
-reference $p_5$. This can be achieved by exploding $p_5$ into it's
-attributes. In this case there is only one attribute and it's value is
+arguments of the \lstinline{jump} operation, which contains the
+reference to the allocation-removed object in $p_5$. This can be achieved by
+exploding $p_5$ into the fields of the allocation-removed object.
+In this case there is only one such field and its value is
 $i_4$, which means that $p_5$ is replaced with $i_4$ in the jump
@@
 
-In the general case, each virtual in the jump arguments is exploded into a
-vector of variables containing the values of all registered attributes. If some
-of the attributes are themselves virtuals they are recursively exploded
-to make the vector contain only non-virtual variables. Some care has
-to be taken to always place the attributes in the same order when
-performing this explosion. Notation becomes somewhat simpler if also every non-
-virtual variable of the jump arguments is exploded into a vector. This will
-be a vector containing the original variable only. To summarize, for
+In the general case, each allocation-removed object in the jump arguments is exploded into a
+vector of variables containing the values of all registered
+fields\footnote{This is sometimes called \emph{scalar replacement}. XXX check
+whether that's true}. If some of the fields are themselves references to
+allocation-removed objects they are recursively exploded
+to make the vector contain only concrete variables. Some care has
+to be taken to always place the fields in the same order when
+performing this explosion. Notation becomes somewhat simpler if also every
+concrete variable of the jump arguments is exploded into a vector containing
+itself. For
 every variable, $J_k$, of the original jump arguments, $J$, let
@@
       \begin{array}{ll}
-        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
-        H^{\left(k\right)} & \text{if $J_k$ is virtual}
+        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
+        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
       \end{array}
@@
 \end{equation}
-where $H^{\left(k\right)}$ is a vector containing all non virtual
+where $H^{\left(k\right)}$ is a vector containing all concrete
 attributes of $J_k$. The arguments of the optimized \lstinline{jump}
@@
 passed as pointers from the second iteration to the third (from peeled
-loop to peeled loop). If one of these objects are virtual 
+loop to peeled loop). If one of these objects are allocation-removed
 at the end of the peeled loop they need to be allocated right

### 2011-06-20 talk/iwtc11/paper.tex
I think that we can indeed skip this.

@@
 
-Note that virtuals are only exploded into their attributes when
-constructing the arguments of the jump of the preamble. This
-explosion can't be repeated when constructing the arguments of the
-jump of the peeled loop as it has to match the first jump. This means that
-the objects that was passed as pointers (non virtuals) from the first
-iteration to the second (from preamble to peeled loop) also has to be
-passed as pointers from the second iteration to the third (from peeled
-loop to peeled loop). If one of these objects are allocation-removed
-at the end of the peeled loop they need to be allocated right
-before the jump. With the simple objects considered in this paper,
-that is not a problem. However in more complicated interpreters such
-an allocation might, in combination with other optimizations, lead
-to additional variables from the preamble being imported into
-the second. This extends both $\hat J$ and $\hat K$, which means that
-some care has to be taken, when implementing this, to allow $\hat J$ to
-grow while inlining it into $\hat K$. XXX: Maybe we can skip this?
-
 XXX explain that this is effectively type-specializing a loop

### 2011-06-20 talk/iwtc11/paper.tex
fix XXX

@@
 would be short numerical calculations with no failing guards and no
-external calls.
-
-XXX reason why we use small numerical kernels for benchmarks
+external calls. Larger loops involving many operations on complex objects
+typically benefit less from it. Loop peeling never makes performance worse, in
+the worst case the peeled loop is exactly the same as the preamble. Therefore we
+chose to present benchmarks of small numeric kernels where loop peeling can show
+its use.
 

### 2011-06-20 talk/iwtc11/paper.tex
remove sections about numpy and prolog for space reasons

@@
 
-\subsection{Numpy}
-
-As a part of the PyPy project, we implemented small numerical kernel for
-performing matrix operations. The exact extend of this kernel is besides
-the scope of this paper, however the basic idea is to unroll a series of
-array operations into a loop compiled into assembler. LICM is a very good
-optimization for those kind of operations. The example benchmark performs
-addition of five arrays, compiling it in a way that's equivalent to C's:
-
-%\begin{figure}
-\begin{lstlisting}[mathescape,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
-for (int i = 0; i < SIZE; i++) {
-   res[i] = a[i] + b[i] + c[i] + d[i] + e[i];
-}
-\end{lstlisting}
-%\end{figure}
-
-Where $res$, $a$, $b$, $c$, $d$ and $e$ are $double$ arrays.
-
-\subsection{Prolog}
-XXX: Carl?
+XXX add a small note somewhere that numpy and prolog are helped by this
+optimization
 

### 2011-06-20 talk/iwtc11/paper.tex
kill the limitations section and mention them the conclusion, rewrite the conclusion to be more compact.

@@
 
-\section{Limitations}
-
-XXX as of now?
-
-Loop invariant code motion as described has certain amount of limitations
-that prevent it from speeding up larger loops. Those limitations are a target
-of future work and might be lifted. Most important ones:
-
-\begin{itemize}
-\item Bridges are not well supported - if the flow is more complex than a single
-      loop, the bridge might need to jump to the beginning of the preamble,
-      making the optimization ineffective
-\item XXX write about flushing caches at calls?
-\end{itemize}
-
 \section{Benchmarks}
@@
 
-\subsection{Conclusions}
+\section{Related Work}
+\label{sec:related}
+
+XXX
+% section Related Work (end)
+
+\section{Conclusions}
+
 In this paper we have studied loop invariant code motion during trace
 compilation. We claim that loop peeling is a very convenient solution
-here since it fits well with other trace optimizations. By peeling of
-the first iteration and optimizing the resulting two iteration trace
-as a single trace, several standard optimizations can be 
-used unchanged. The only interaction needed between the loop peeling
-and the other 
-optimizations is during the constructing of the jump arguments
-connecting the peeled of iteration (the preamble) with the peeled loop. This
-approach 
-improves the effect of standard optimizations such as redundant guard removal, heap
-caching, common subexpression elimination and allocation removals. The
-most prominent effect is that they all become loop
+here since it fits well with other trace optimizations and does not require
+large changes to them. This approach improves the effect of standard
+optimizations such as redundant guard removal, common subexpression elimination
+and allocation removal. The most prominent effect is that they all become loop
 invariant code motion optimizations.
@@
 By using several benchmarks we show that the proposed algorithm can
-significantly
-improve the run time of small loops containing numerical
+significantly improve the run time of small loops containing numerical
 calculations.
-At least in cases where there are not too many guard
-failures. A common way of handling a guard that fails often is to 
-trace a bridge from it back to the start of some previously compiled
-loop. This is applicable here too. However the bridge will have to end
-with a jump to the preamble, which lessens the impact of the 
-proposed approach.
-In many situations it is however possible to make the bridge
-jump to the peeled loop instead. When and how this is possible will be
-focus of future work.
+
+The current approach still has some limitations which we plan to lift in the
+future. In particular loop peeling is working less well in combination with
+trace trees or trace stitching. The side exits attached guards that fail often
+currently have to jump to the preamble which makes loops with several equally
+common paths less efficient than they could be.
 

### 2011-06-20 talk/iwtc11/paper.tex
Add benchmark table - too big as of now
squeeze a bit
intending to make benchmarks more reasonable (will update the code as well)

@@
 \usepackage{setspace}
+\usepackage{relsize}
 
@@
 external calls. Larger loops involving many operations on complex objects
-typically benefit less from it. Loop peeling never makes performance worse, in
+typically benefit less from it. Loop peeling never makes runtime performance worse, in
 the worst case the peeled loop is exactly the same as the preamble. Therefore we
@@
   locate edges in an image. It calculates the gradient magnitude
-  using sobel derivatives. In Python the algorithm is implemented
-  on top of a custom image class that is specially designed for the
-  problem. It ensures that there will be no failing guards, and makes
+  using sobel derivatives. 
+\end{itemize}
+
+The sobel and conv3x3 benchmarks are implemented
+on top of two different two-dimensional array classes: Array2D
+and NoBorderPadded. Array2D is 
+a simple straight forward implementation providing 2 dimensionall
+indexing with out of bounds checks. NoBorderPadded is optimized for
+this specific application.
+It ensures that there will be no failing guards, and makes
 a lot of the two dimension index calculations loop invariant. The
@@
 under a nice interface without loosing performance.
-\end{itemize}
 
@@
 the relative immaturity of PyPy's JIT assembler backend as well as missing
-optimizations, like instruction scheduling.
+optimizations, like instruction scheduling. Results:
+
+\begin{figure}
+\begin{center}
+{\smaller
+\begin{tabular}{|l|r|r|r|r|r|r|}
+\hline
+ & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
+\hline
+NoBorder & 537.31 & 329.98 & 2.22 +- 0.03 & 2.17 +- 0.02 & - & -\\
+\hline
+NoBorder(iter) & 548.32 & 304.13 & 1.45 +- 0.03 & 1.47 +- 0.02 & - & -\\
+\hline
+NoBorder(range) & 534.64 & 317.34 & 1.34 +- 0.03 & 1.40 +- 0.04 & - & -\\
+\hline
+NoBorderPadded & 543.73 & 333.20 & 2.09 +- 0.12 & 1.93 +- 0.05 & - & -\\
+\hline
+NoBorderPadded(iter) & 546.70 & 309.32 & 1.21 +- 0.02 & 0.49 +- 0.02 & - & -\\
+\hline
+NoBorderPadded(range) & 550.92 & 318.33 & 1.12 +- 0.03 & 0.48 +- 0.01 & - & -\\
+\hline
+conv3(1e5) & 77.89 & 9.52 & 1.77 +- 0.06 & 0.68 +- 0.02 & 0.70 +- 0.05 & 0.59 +- 0.01\\
+\hline
+conv3(1e6) & 77.15 & 9.58 & 1.69 +- 0.01 & 0.77 +- 0.01 & 0.84 +- 0.05 & 0.74 +- 0.01\\
+\hline
+conv3x3(1000) & 23.72 & 12.77 & 0.07 +- 0.00 & 0.05 +- 0.03 & 0.24 +- 0.00 & 0.25 +- 0.01\\
+\hline
+conv3x3(3) & 23.85 & 12.77 & 0.10 +- 0.00 & 0.07 +- 0.00 & 0.27 +- 0.01 & 0.27 +- 0.01\\
+\hline
+conv5(1e5) & 122.54 & 16.67 & 1.86 +- 0.02 & 1.05 +- 0.03 & 1.03 +- 0.05 & 0.65 +- 0.01\\
+\hline
+conv5(1e6) & 125.77 & 16.80 & 1.92 +- 0.03 & 1.09 +- 0.02 & 1.07 +- 0.05 & 0.80 +- 0.01\\
+\hline
+dilate3x3(1000) & 23.29 & 12.99 & 0.41 +- 0.04 & 0.39 +- 0.01 & 0.26 +- 0.00 & 0.26 +- 0.01\\
+\hline
+sobel(NoBorderPadded) & 461.14 & 258.02 & 1.01 +- 0.03 & 0.48 +- 0.03 & - & -\\
+\hline
+sobel\_magnitude & - & - & - & - & 0.19 +- 0.01 & 0.20 +- 0.01\\
+\hline
+sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 +- 0.01 & 0.51 +- 0.00 & - & -\\
+\hline
+sqrt(Fix16) & 744.35 & 421.65 & 3.93 +- 0.11 & 2.14 +- 0.02 & 0.97 +- 0.02 & 0.96 +- 0.01\\
+\hline
+sqrt(float) & 24.21 & 5.52 & 1.36 +- 0.03 & 1.00 +- 0.00 & 0.98 +- 0.00 & 0.98 +- 0.00\\
+\hline
+sqrt(int) & 20.84 & 1.78 & 2.26 +- 0.01 & 1.82 +- 0.01 & 0.81 +- 0.01 & 0.80 +- 0.01\\
+\hline
+
+\end{tabular}
+}
+\end{center}
+\end{figure}
 

### 2011-06-20 talk/iwtc11/paper.tex
trying to save some space

--- 2011-06-20 talk/iwtc11/paper.tex
@@
 the relative immaturity of PyPy's JIT assembler backend as well as missing
-optimizations, like instruction scheduling.
+optimizations, like instruction scheduling. Results:
+
+\begin{figure}
+\begin{center}
+{\smaller
+\begin{tabular}{|l|r|r|r|r|r|r|}
+\hline
+ & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
+\hline
+NoBorder & 537.31 & 329.98 & 2.22 $\pm$ 0.03 & 2.17 $\pm$ 0.02 & - & -\\
+\hline
+NoBorder(iter) & 548.32 & 304.13 & 1.45 $\pm$ 0.03 & 1.47 $\pm$ 0.02 & - & -\\
+\hline
+NoBorder(range) & 534.64 & 317.34 & 1.34 $\pm$ 0.03 & 1.40 $\pm$ 0.04 & - & -\\
+\hline
+NoBorderPadded & 543.73 & 333.20 & 2.09 $\pm$ 0.12 & 1.93 $\pm$ 0.05 & - & -\\
+\hline
+NoBorderPadded(iter) & 546.70 & 309.32 & 1.21 $\pm$ 0.02 & 0.49 $\pm$ 0.02 & - & -\\
+\hline
+NoBorderPadded(range) & 550.92 & 318.33 & 1.12 $\pm$ 0.03 & 0.48 $\pm$ 0.01 & - & -\\
+\hline
+conv3(1e5) & 77.89 & 9.52 & 1.77 $\pm$ 0.06 & 0.68 $\pm$ 0.02 & 0.70 $\pm$ 0.05 & 0.59 $\pm$ 0.01\\
+\hline
+conv3(1e6) & 77.15 & 9.58 & 1.69 $\pm$ 0.01 & 0.77 $\pm$ 0.01 & 0.84 $\pm$ 0.05 & 0.74 $\pm$ 0.01\\
+\hline
+conv3x3(1000) & 23.72 & 12.77 & 0.07 $\pm$ 0.00 & 0.05 $\pm$ 0.03 & 0.24 $\pm$ 0.00 & 0.25 $\pm$ 0.01\\
+\hline
+conv3x3(3) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
+\hline
+conv5(1e5) & 122.54 & 16.67 & 1.86 $\pm$ 0.02 & 1.05 $\pm$ 0.03 & 1.03 $\pm$ 0.05 & 0.65 $\pm$ 0.01\\
+\hline
+conv5(1e6) & 125.77 & 16.80 & 1.92 $\pm$ 0.03 & 1.09 $\pm$ 0.02 & 1.07 $\pm$ 0.05 & 0.80 $\pm$ 0.01\\
+\hline
+dilate3x3(1000) & 23.29 & 12.99 & 0.41 $\pm$ 0.04 & 0.39 $\pm$ 0.01 & 0.26 $\pm$ 0.00 & 0.26 $\pm$ 0.01\\
+\hline
+sobel(NoBorderPadded) & 461.14 & 258.02 & 1.01 $\pm$ 0.03 & 0.48 $\pm$ 0.03 & - & -\\
+\hline
+sobel\_magnitude & - & - & - & - & 0.19 $\pm$ 0.01 & 0.20 $\pm$ 0.01\\
+\hline
+%sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 $\pm$ 0.01 & 0.51 $\pm$ 0.00 & - & -\\
+\hline
+sqrt(Fix16) & 744.35 & 421.65 & 3.93 $\pm$ 0.11 & 2.14 $\pm$ 0.02 & 0.97 $\pm$ 0.02 & 0.96 $\pm$ 0.01\\
+\hline
+sqrt(float) & 24.21 & 5.52 & 1.36 $\pm$ 0.03 & 1.00 $\pm$ 0.00 & 0.98 $\pm$ 0.00 & 0.98 $\pm$ 0.00\\
+\hline
+sqrt(int) & 20.84 & 1.78 & 2.26 $\pm$ 0.01 & 1.82 $\pm$ 0.01 & 0.81 $\pm$ 0.01 & 0.80 $\pm$ 0.01\\
+\hline
+
+\end{tabular}
+}
+\end{center}
+\end{figure}
 

### 2011-06-20 talk/iwtc11/paper.tex
more consistent and descriptive naming of the benachmarks

--- 2011-06-20 talk/iwtc11/paper.tex
@@
 XXX we either need to explain that we use C++ or consistently use C
+\begin{figure*}
+\begin{center}
+{\smaller
+\begin{tabular}{|l|r|r|r|r|r|r|}
+\hline
+ & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
+\hline
+%NoBorder & 537.31 & 329.98 & 2.22 $\pm$ 0.03 & 2.17 $\pm$ 0.02 & - & -\\
+%\hline
+%NoBorder(iter) & 548.32 & 304.13 & 1.45 $\pm$ 0.03 & 1.47 $\pm$ 0.02 & - & -\\
+%\hline
+%NoBorder(range) & 534.64 & 317.34 & 1.34 $\pm$ 0.03 & 1.40 $\pm$ 0.04 & - & -\\
+%\hline
+conv3x3(NoBorderPadded(1000x1000)) & 543.73 & 333.20 & 2.09 $\pm$ 0.12 & 1.93 $\pm$ 0.05 & - & -\\
+\hline
+%NoBorderPadded(iter) & 546.70 & 309.32 & 1.21 $\pm$ 0.02 & 0.49 $\pm$ 0.02 & - & -\\
+%\hline
+%NoBorderPadded(range) & 550.92 & 318.33 & 1.12 $\pm$ 0.03 & 0.48 $\pm$ 0.01 & - & -\\
+%\hline
+conv3(array(1e5)) & 77.89 & 9.52 & 1.77 $\pm$ 0.06 & 0.68 $\pm$ 0.02 & 0.70 $\pm$ 0.05 & 0.59 $\pm$ 0.01\\
+\hline
+conv3(array(1e6)) & 77.15 & 9.58 & 1.69 $\pm$ 0.01 & 0.77 $\pm$ 0.01 & 0.84 $\pm$ 0.05 & 0.74 $\pm$ 0.01\\
+\hline
+conv3x3(Array2D(1000x1000)) & 23.72 & 12.77 & 0.07 $\pm$ 0.00 & 0.05 $\pm$ 0.03 & 0.24 $\pm$ 0.00 & 0.25 $\pm$ 0.01\\
+\hline
+conv3x3(Array2D(1000x3)) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
+\hline
+conv5(array(1e5)) & 122.54 & 16.67 & 1.86 $\pm$ 0.02 & 1.05 $\pm$ 0.03 & 1.03 $\pm$ 0.05 & 0.65 $\pm$ 0.01\\
+\hline
+conv5(array(1e6)) & 125.77 & 16.80 & 1.92 $\pm$ 0.03 & 1.09 $\pm$ 0.02 & 1.07 $\pm$ 0.05 & 0.80 $\pm$ 0.01\\
+\hline
+dilate3x3(Array2D(1000x1000)) & 23.29 & 12.99 & 0.41 $\pm$ 0.04 & 0.39 $\pm$ 0.01 & 0.26 $\pm$ 0.00 & 0.26 $\pm$ 0.01\\
+\hline
+sobel(NoBorderPadded(1000x1000)) & 461.14 & 258.02 & 1.01 $\pm$ 0.03 & 0.48 $\pm$ 0.03 & - & -\\
+\hline
+sobel(Array2D(1000x1000)) & - & - & - & - & 0.19 $\pm$ 0.01 & 0.20 $\pm$ 0.01\\
+\hline
+%sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 $\pm$ 0.01 & 0.51 $\pm$ 0.00 & - & -\\
+%\hline
+sqrt(Fix16) & 744.35 & 421.65 & 3.93 $\pm$ 0.11 & 2.14 $\pm$ 0.02 & 0.97 $\pm$ 0.02 & 0.96 $\pm$ 0.01\\
+\hline
+sqrt(float) & 24.21 & 5.52 & 1.36 $\pm$ 0.03 & 1.00 $\pm$ 0.00 & 0.98 $\pm$ 0.00 & 0.98 $\pm$ 0.00\\
+\hline
+sqrt(int) & 20.84 & 1.78 & 2.26 $\pm$ 0.01 & 1.82 $\pm$ 0.01 & 0.81 $\pm$ 0.01 & 0.80 $\pm$ 0.01\\
+\hline
+
+\end{tabular}
+}
+\end{center}
+\label{fig:benchmarks}
+\caption{XXX}
+\end{figure*}
 
@@
 calculations were implemented in both Python and in C and their
-runtimes compared. The benchmarks are
+runtimes are compared in Figure~\ref{fig:benchmarks}. The benchmarks are
 \begin{itemize}
@@
 the relative immaturity of PyPy's JIT assembler backend as well as missing
-optimizations, like instruction scheduling.
+optimizations, like instruction scheduling. Results:
 

### 2011-06-20 talk/iwtc11/benchmarks/convolution/convolution.py
rename in benchmark implementations aswell

@@ -1,5 +1,5 @@
 from array import array
-from math import log10
+from math import log10, sqrt
 
 def _conv3(a, k, n=1):
     assert len(k)==3
@@ -14,7 +14,7 @@ def conv3(args):
     n = int(args[0])
     _conv3(array('d', [1]) * (100000000/n),
            array('d', [-1, 0, 1]), n)
-    return 'conv3(1e%d)' % log10(100000000/n)
+    return 'conv3(array(1e%d))' % log10(100000000/n)
 
 def _conv5(a, k, n=1):
     assert len(k)==5
@@ -29,7 +29,7 @@ def conv5(args):
     n = int(args[0])
     _conv5(array('d', [1]) * (100000000/n),
            array('d', [1, 4, 6, 4, 1]), n)
-    return 'conv5(1e%d)' % log10(100000000/n)
+    return 'conv5(array(1e%d))' % log10(100000000/n)
 
 class Array2D(object):
     def __init__(self, w, h):
@@ -84,9 +84,28 @@ def _erode3x3(a, k):
     return morphology3x3(a, k, min)
 
 def conv3x3(args):
-    _conv3x3(Array2D(int(args[0]), int(args[1])), Array2D(3,3))
-    return 'conv3x3(%s)' % args[1]
+    for i in range(10):
+        _conv3x3(Array2D(int(args[0]), int(args[1])), Array2D(3,3))
+    return 'conv3x3(Array2D(%sx%s))' % tuple(args)
 
 def dilate3x3(args):
-    _dilate3x3(Array2D(int(args[0]), int(args[1])), Array2D(3,3))
-    return 'dilate3x3(%s)' % args[1]
+    for i in range(10):
+        _dilate3x3(Array2D(int(args[0]), int(args[1])), Array2D(3,3))
+    return 'dilate3x3(Array2D(%sx%s))' % tuple(args)
+
+def _sobel_magnitude(a):
+    b = Array2D(a.width, a.height)    
+    for y in xrange(1, a.height-1):
+        for x in xrange(1, a.width-1):
+            dx = -1.0 * a[x-1, y-1] + 1.0 * a[x+1, y-1] + \
+                 -2.0 * a[x-1, y]   + 2.0 * a[x+1, y] + \
+                 -1.0 * a[x-1, y+1] + 1.0 * a[x+1, y+1]
[…]

### 2011-06-21 talk/iwtc11/paper.tex
rename in benchmark implementations aswell
Fix my affiliation
add a small note about numpy and prolog

@@
 \authorinfo{Maciej Fijałkowski}
-           {Affiliation2}
+           {Unaffiliated}
            {fijall@gmail.com}
@@
 \hline
-conv3x3(Array2D(1000x3)) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
+conv3x3(Array2D(1000000x3)) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
 \hline
@@
 the relative immaturity of PyPy's JIT assembler backend as well as missing
-optimizations, like instruction scheduling. Results:
+optimizations, like instruction scheduling.
 
-XXX add a small note somewhere that numpy and prolog are helped by this
-optimization
+Other interesting interpreters that are helped greatly by this
+optimization are for
+example our Prolog interpreter written in RPython, as well as numerical
+kernel used for array manipulation. The exact extent is out of scope for
+this paper.
 

### 2011-06-21 talk/iwtc11/paper.tex
fix two XXXs, start adding some minimal notes to the related work section

@@
 
-XXX "bridge" is not a standard term
-
-XXX loop peeling does a lot more than loop-invariant code motion
+% loop peeling does a lot more than loop-invariant code motion
 % take this loop as an example:
@@
 introduced in the entire copied trace in order to maintian the SSA-property.
-When applying the following optimizations to this two-iteration trace
+Note that the peeled loop is not necessary the \emph{first} iteration of the
+loop execution, it is general enough to correspond to any iteration of the loop.
+However, the peeled loop can then be optimized using the assumption that a
+previous iteration has happened.
+
+When applying optimizations to this two-iteration trace
 some care has to taken as to how the arguments of the two
@@
 \subsection{Redundant Guard Removal}
-
-XXX should we have a mention where in the previous papers those optimizations
-are discussed?
 
@@
 The loop peeling optimization was implemented in the PyPy
-framework. That means that the JIT-compilers generated for all
+framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
 interpreters implemented within PyPy now can take advantage of
@@
 
+All the optimizations presented here are completely standard \cite{XXX} [dragon
+book or similar].
+
+LuaJIT does the same
+
+SPUR does all these optimizations manually, moves allocations out of loop but
+does not explode objects
+
 XXX

### 2011-06-21 talk/iwtc11/paper.tex
expand related work somewhat

@@
 \label{fig:benchmarks}
-\caption{XXX}
+\caption{Benchmark Results in Seconds}
 \end{figure*}
@@
 
-All the optimizations presented here are completely standard \cite{XXX} [dragon
-book or similar].
-
-LuaJIT does the same
-
-SPUR does all these optimizations manually, moves allocations out of loop but
-does not explode objects
+All the optimizations presented here are completely standard
+\cite{muchnick_advanced_1997}. XXX
+
+Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
+have developped the described technique independently. There are no papers about
+LuaJIT but the author of it writes on a mailing list: "The LOOP pass does
+synthetic unrolling of the recorded IR, combining copy-substitution with
+redundancy elimination to achieve code hoisting. The unrolled and
+copy-substituted instructions are simply fed back into the compiler pipeline,
+which allows reuse of all optimizations for redundancy elimination. Loop
+recurrences are detected on-the-fly and a minimized set of PHIs is generated."
+\cite{pall_luajit_2009}
+
+SPUR \cite{bebenita_spur:_2010} implements loop-invariant code motion
+directly, by explicitly marking as loop-invariant all variables that stay the
+same along all looping paths and then moving all pure computation that depends
+only on these variables out of the loop. SPUR can also hoist loads out of the
+loop if nothing in the loop can ever write to the memory location. It can also
+move allocations out of the loop, but does not replace the object by its fields.
+This saves only the allocation, not the access to the object fields.
 

### 2011-06-21 talk/iwtc11/paper.tex
english tweeks.

@@
 \hline
-%NoBorder & 537.31 & 329.98 & 2.22 $\pm$ 0.03 & 2.17 $\pm$ 0.02 & - & -\\
-%\hline
-%NoBorder(iter) & 548.32 & 304.13 & 1.45 $\pm$ 0.03 & 1.47 $\pm$ 0.02 & - & -\\
-%\hline
-%NoBorder(range) & 534.64 & 317.34 & 1.34 $\pm$ 0.03 & 1.40 $\pm$ 0.04 & - & -\\
-%\hline
-conv3x3(NoBorderPadded(1000x1000)) & 543.73 & 333.20 & 2.09 $\pm$ 0.12 & 1.93 $\pm$ 0.05 & - & -\\
-\hline
-%NoBorderPadded(iter) & 546.70 & 309.32 & 1.21 $\pm$ 0.02 & 0.49 $\pm$ 0.02 & - & -\\
-%\hline
-%NoBorderPadded(range) & 550.92 & 318.33 & 1.12 $\pm$ 0.03 & 0.48 $\pm$ 0.01 & - & -\\
-%\hline
 conv3(array(1e5)) & 77.89 & 9.52 & 1.77 $\pm$ 0.06 & 0.68 $\pm$ 0.02 & 0.70 $\pm$ 0.05 & 0.59 $\pm$ 0.01\\
@@
 dilate3x3(Array2D(1000x1000)) & 23.29 & 12.99 & 0.41 $\pm$ 0.04 & 0.39 $\pm$ 0.01 & 0.26 $\pm$ 0.00 & 0.26 $\pm$ 0.01\\
-\hline
-sobel(NoBorderPadded(1000x1000)) & 461.14 & 258.02 & 1.01 $\pm$ 0.03 & 0.48 $\pm$ 0.03 & - & -\\
 \hline
@@
 The sobel and conv3x3 benchmarks are implemented
-on top of two different two-dimensional array classes: Array2D
-and NoBorderPadded. Array2D is 
+on top of a custom two-dimensional array class,  Array2D.
+It is
 a simple straight forward implementation providing 2 dimensionall
-indexing with out of bounds checks. NoBorderPadded is optimized for
-this specific application.
-It ensures that there will be no failing guards, and makes
-a lot of the two dimension index calculations loop invariant. The
-intention here is twofold. It shows that the performance-impact of
-having wrapper classes giving objects some application-specific
-properties can be negligible. This is due to the inlining performed
-during the tracing and the allocation removal of the index objects
-introduced. It also shows that it is possible to do some low-level
-hand optimizations of the Python code and hide those optimization
-under a nice interface without loosing performance.
+indexing with out of bounds checks.
 
@@
 
-The current approach still has some limitations which we plan to lift in the
-future. In particular loop peeling is working less well in combination with
-trace trees or trace stitching. The side exits attached guards that fail often
+The current approach still has some limitations which we plan to address in the
+future. In particular loop peeling works poorly in combination with trace
+trees or trace stitching. The side exits attached guards that fail often
 currently have to jump to the preamble which makes loops with several equally

### 2011-06-21 talk/iwtc11/paper.tex
let the image class choose interpolation method, making multiple magnify() implemenattions unnecessery

@@
 
-The current approach still has some limitations which we plan to address in the
-future. In particular loop peeling works poorly in combination with trace
-trees or trace stitching. The side exits attached guards that fail often
+The current approach still has some limitations which we plan to lift in the
+future. In particular loop peeling is working less well in combination with
+trace trees or trace stitching. The side exits attached guards that fail often
 currently have to jump to the preamble which makes loops with several equally

### 2011-06-21 talk/iwtc11/benchmarks/convolution/convolution.py

@@ -1,0 +1,111 @@
+from array import array
+from math import log10, sqrt
+
+def _conv3(a, k, n=1):
+    assert len(k)==3
+    b = array(a.typecode, [0]) * (len(a) - 2)
+    while n:
+        n -= 1
+        for i in xrange(len(b)):
+            b[i] = k[2]*a[i] + k[1]*a[i+1] + k[0]*a[i+2]
+    return b
+
+def conv3(args):
+    n = int(args[0])
+    _conv3(array('d', [1]) * (100000000/n),
+           array('d', [-1, 0, 1]), n)
+    return 'conv3(array(1e%d))' % log10(100000000/n)
+
+def _conv5(a, k, n=1):
+    assert len(k)==5
+    b = array(a.typecode, [0]) * (len(a) - 4)
+    while n:
+        n -= 1
+        for i in xrange(len(b)):
+            b[i] = k[4]*a[i] + k[3]*a[i+1] + k[2]*a[i+2] + k[1]*a[i+3] + k[0]*a[i+4]
+    return b
+
+def conv5(args):
+    n = int(args[0])
+    _conv5(array('d', [1]) * (100000000/n),
+           array('d', [1, 4, 6, 4, 1]), n)
+    return 'conv5(array(1e%d))' % log10(100000000/n)
+
+class Array2D(object):
+    def __init__(self, w, h):
+        self.width = w
+        self.height = h
+        self.data = array('d', [0]) * (w*h)
+
+    def _idx(self, x, y):
+        if 0 <= x < self.width and 0 <= y < self.height:
+            return y*self.width + x
+        raise IndexError
+
+    def __getitem__(self, (x, y)):
+        return self.data[self._idx(x, y)]
+
+    def __setitem__(self, (x, y), val):
+        self.data[self._idx(x, y)] = val
+
+    def __cmp__(self, other):
+        return cmp(self.data, other.data)
+
+    def setup(self, data):
+        for y in xrange(self.height):
+            for x in xrange(self.width):
+                self[x, y] = data[y][x]
+        return self
+
+def _conv3x3(a, k):
+    assert k.width == k.height == 3
+    b = Array2D(a.width, a.height)
+    for y in xrange(1, a.height-1):
[…]

### 2011-06-21 talk/iwtc11/benchmarks/runall.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+./benchmark.sh pypy
+#./benchmark.sh pypy --jit enable_opts=intbounds:rewrite:virtualize:heap:unroll
+./benchmark.sh pypy --jit enable_opts=intbounds:rewrite:virtualize:heap
+#./benchmark.sh gcc
+./benchmark.sh gcc -O2
+./benchmark.sh gcc -O3 -march=native -fno-tree-vectorize
+./benchmark.sh python2.7
+./benchmark.sh python2.6 psyco-wrapper.py

### 2011-06-23 talk/iwtc11/paper.tex
compactifying
use a single standard deviation per binary
explain the shorter names

--- 2011-06-21 talk/iwtc11/paper.tex
@@
 XXX we either need to explain that we use C++ or consistently use C
-\begin{figure*}
+\begin{figure}
 \begin{center}
@@
 \hline
- & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
+ & CPython & Psyco & PyPy  & PyPy & GCC \\
+ &         &       & no LP &      & -O3 \\
 \hline
-conv3(array(1e5)) & 77.89 & 9.52 & 1.77 $\pm$ 0.06 & 0.68 $\pm$ 0.02 & 0.70 $\pm$ 0.05 & 0.59 $\pm$ 0.01\\
+conv3(1e5) & 77.89 & 9.52 & 1.77  & 0.68 &  0.59 \\
 \hline
-conv3(array(1e6)) & 77.15 & 9.58 & 1.69 $\pm$ 0.01 & 0.77 $\pm$ 0.01 & 0.84 $\pm$ 0.05 & 0.74 $\pm$ 0.01\\
+conv3(1e6) & 77.15 & 9.58 & 1.69  & 0.77 &  0.74 \\
 \hline
-conv3x3(Array2D(1000x1000)) & 23.72 & 12.77 & 0.07 $\pm$ 0.00 & 0.05 $\pm$ 0.03 & 0.24 $\pm$ 0.00 & 0.25 $\pm$ 0.01\\
+conv3x3(1000) & 23.72 & 12.77 & 0.07  & 0.05 &  0.25 \\
 \hline
-conv3x3(Array2D(1000000x3)) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
+conv3x3(3) & 23.85 & 12.77 & 0.10  & 0.07  &  0.27 \\
 \hline
-conv5(array(1e5)) & 122.54 & 16.67 & 1.86 $\pm$ 0.02 & 1.05 $\pm$ 0.03 & 1.03 $\pm$ 0.05 & 0.65 $\pm$ 0.01\\
+conv5(1e5) & 122.54 & 16.67 & 1.86  & 1.05 &  0.65\\
 \hline
-conv5(array(1e6)) & 125.77 & 16.80 & 1.92 $\pm$ 0.03 & 1.09 $\pm$ 0.02 & 1.07 $\pm$ 0.05 & 0.80 $\pm$ 0.01\\
+conv5(1e6) & 125.77 & 16.80 & 1.92 & 1.09 &  0.80 \\
 \hline
-dilate3x3(Array2D(1000x1000)) & 23.29 & 12.99 & 0.41 $\pm$ 0.04 & 0.39 $\pm$ 0.01 & 0.26 $\pm$ 0.00 & 0.26 $\pm$ 0.01\\
+dilate3x3(1000) & 23.29 & 12.99 & 0.41  & 0.39 &  0.26 \\
 \hline
-sobel(Array2D(1000x1000)) & - & - & - & - & 0.19 $\pm$ 0.01 & 0.20 $\pm$ 0.01\\
+sobel(1000) & - & - & - & - & 0.20 \\
 \hline
-%sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 $\pm$ 0.01 & 0.51 $\pm$ 0.00 & - & -\\
-%\hline
-sqrt(Fix16) & 744.35 & 421.65 & 3.93 $\pm$ 0.11 & 2.14 $\pm$ 0.02 & 0.97 $\pm$ 0.02 & 0.96 $\pm$ 0.01\\
+sqrt(Fix16) & 744.35 & 421.65 & 3.93  & 2.14  & 0.96 \\
 \hline
-sqrt(float) & 24.21 & 5.52 & 1.36 $\pm$ 0.03 & 1.00 $\pm$ 0.00 & 0.98 $\pm$ 0.00 & 0.98 $\pm$ 0.00\\
+sqrt(float) & 24.21 & 5.52 & 1.36 & 1.00 & 0.98\\
 \hline
-sqrt(int) & 20.84 & 1.78 & 2.26 $\pm$ 0.01 & 1.82 $\pm$ 0.01 & 0.81 $\pm$ 0.01 & 0.80 $\pm$ 0.01\\
+sqrt(int) & 20.84 & 1.78 & 2.26 & 1.82  & 0.80 \\
 \hline
-
+\hline
+Variations & - & - & $\pm 0.03$ & $\pm 0.01$ & $\pm 0.01$ \\
+\hline
 \end{tabular}
@@
 \label{fig:benchmarks}
-\caption{Benchmark Results in Seconds}
-\end{figure*}
+\caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
+  $10^6$ and matrixes of size $1000\times 1000$ and $1000000 \times
+  3$ are used. This is indicated in the leftmost column. For the
+  matrixes, only the number of rows are listed.} 
+\end{figure}
 

### 2011-06-23 talk/iwtc11/paper.tex
sobel numbers
explain specialization

@@
 
-XXX explain that this is effectively type-specializing a loop
+If all the optimizations presented above are applied, the resulting
+optimized peeled loop will consist of a single integer addition
+only. That is it will become type-specialized to the types of the
+variables \lstinline{step} and \lstinline{y}, and the overhead of
+using boxed values is removed.
 
@@
 \hline
-sobel(1000) & - & - & - & - & 0.20 \\
+sobel(1000) & 181.49 & 95.05 & 0.71 & 0.42 & 0.20 \\
 \hline
@@
   $10^6$ and matrixes of size $1000\times 1000$ and $1000000 \times
-  3$ are used. This is indicated in the leftmost column. For the
-  matrixes, only the number of rows are listed.} 
+  3$ are used. The one used in each benchmark is indicated in
+  the leftmost column. For the matrixes, only the number of rows are
+  specified.} 
 \end{figure}

### 2011-06-24 talk/iwtc11/paper.tex
rerun of fixed benchmark

@@
 
-XXX we either need to explain that we use C++ or consistently use C
 \begin{figure}
@@
 \hline
-conv3x3(1000) & 23.72 & 12.77 & 0.07  & 0.05 &  0.25 \\
+conv3x3(1000) & 236.96 & 128.88 & 0.70  & 0.41 &  0.25 \\
 \hline
@@
 The sobel and conv3x3 benchmarks are implemented
-on top of a custom two-dimensional array class,  Array2D.
+on top of a custom two-dimensional array class.
 It is
 a simple straight forward implementation providing 2 dimensionall
-indexing with out of bounds checks.
+indexing with out of bounds checks. For the C implementations it is
+implemented as a C++ class. The other benchmarks are implemented in
+plain C.
 

### 2011-06-27 talk/iwtc11/paper.tex
get latest number into the table
some clarifications

@@
 \authorinfo{Maciej Fijałkowski}
-           {Unaffiliated}
+           {}
            {fijall@gmail.com}
@@
 times increase in speed when loop peeling was introduced. This makes
-some of them almost match optimized C performance and become over XXX
+some of them almost match optimized C performance and become over 900
 times faster than CPython.
@@
 However, we want to get this effect using our existing optimization passes
-without changing them too much. To achieve this, we peel one iteration off the
-loop before running the optimizations. This peeling gives the following trace:
+without changing them too much. Simple optimizations with one forward pass
+cannot directly get this effect: They just look at the trace without taking
+into account that the trace executes many times in a row. Therefore to achieve
+loop-invariant code motion, we peel one iteration off the loop before running
+the optimizations. This peeling gives the following trace:
 
@@
 arguments are inserted into the label of the loop itself and the jumps
-afterwards. Giving the optimizations two iterations together
-gives the optimization enough context to remove operations from the peeled loop,
-because it detects that the operation was performed in the preamble already.
+afterwards.
+
+This is the key insight of the proposed implementation scheme: Giving an
+optimization two iterations together at the same time gives the optimization
+enough context to remove operations from the peeled loop, because it detects
+that the operation was performed in the preamble already. Thus at runtime these
+moved operations are only executed once when entering the loop and the results
+are reused in further iterations.
 
@@
 \hline
-conv3x3(1000) & 236.96 & 128.88 & 0.70  & 0.41 &  0.25 \\
-\hline
-conv3x3(3) & 23.85 & 12.77 & 0.10  & 0.07  &  0.27 \\
+conv3x3(1000) & 233.54 & 125.40 & 0.57 & 0.27 & 0.25 \\
+\hline
+conv3x3(3) & 234.45 & 126.28 & 0.60 & 0.31 & 0.28 \\
 \hline
@@
 \hline
-dilate3x3(1000) & 23.29 & 12.99 & 0.41  & 0.39 &  0.26 \\
+dilate3x3(1000) & 232.51 & 125.85 & 3.89 & 3.69 & 0.25 \\
 \hline
@@
 have developped the described technique independently. There are no papers about
-LuaJIT but the author of it writes on a mailing list: "The LOOP pass does
+LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
 synthetic unrolling of the recorded IR, combining copy-substitution with
@@
 which allows reuse of all optimizations for redundancy elimination. Loop
-recurrences are detected on-the-fly and a minimized set of PHIs is generated."
+recurrences are detected on-the-fly and a minimized set of PHIs is generated.''
 \cite{pall_luajit_2009}

### 2011-06-27 talk/iwtc11/paper.tex
XXXs

@@
 
-XXX make clear that the preamble is not necessarily the \emph{first} iteration
-of a loop
-
 Before the trace is passed to a backend compiling it into machine code
@@
 previous iteration has happened.
+
+XXX (samuele): the point about the first iteration is hard to understand
 
@@
 vector of variables containing the values of all registered
-fields\footnote{This is sometimes called \emph{scalar replacement}. XXX check
-whether that's true}. If some of the fields are themselves references to
+fields\footnote{This is sometimes called \emph{scalar replacement}.}.
+If some of the fields are themselves references to
 allocation-removed objects they are recursively exploded
@@
 faster than either CPython or Psyco. This is due to the JIT compilation
-advantages and optimizations we discussed in XXX [ref to other paper]. Loop
+advantages and optimizations we discussed in previous work
+\cite{bolz_allocation_2011, bolz_runtime_2011}. Loop
 peeling gives an additional XXX on average, which makes benchmark times
@@
 
-All the optimizations presented here are completely standard
-\cite{muchnick_advanced_1997}. XXX
+The effect of combining a one ass optimization with loop peeling gives
+completely standard loop invariant code motion optimizations
+\cite{muchnick_advanced_1997}. We do not claim any novelty in the effect, but
+think that our implementation scheme is a very simple one.
 

### 2011-06-27 talk/iwtc11/paper.tex
fix XXX

@@
 advantages and optimizations we discussed in previous work
-\cite{bolz_allocation_2011, bolz_runtime_2011}. Loop
-peeling gives an additional XXX on average, which makes benchmark times
-comparable with native-compiled C code. Missing performance we attribute to
+\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
+speedup of loop peeling is 70\%, which makes benchmark times
+comparable with native-compiled C code. We attribute the performance gap to C code to
 the relative immaturity of PyPy's JIT assembler backend as well as missing

### 2011-06-27 talk/iwtc11/paper.tex
more related work

@@
 \subsection{Allocation Removals}
+\label{sub:allocation}
+
 PyPy's allocation removal optimization \cite{bolz_allocation_2011} makes it
@@
 
-SPUR \cite{bebenita_spur:_2010} implements loop-invariant code motion
+Both the Hotpath VM \cite{gal_hotpathvm:_2006} and SPUR
+\cite{bebenita_spur:_2010} implements loop-invariant code motion
 directly, by explicitly marking as loop-invariant all variables that stay the
@@
 
-XXX
+The type specialization described by Gal \etal \cite{gal_trace-based_2009} can
+be seen as doing a similar optimization (again by manually implementing it)
+than the one described in Section~\ref{sub:allocation}: The effect of both is
+that type checks are fully done before a loop is even entered.
+
 % section Related Work (end)
@@
 
-\acks
-
-Acknowledgments, if needed.
+%\acks
+%Acknowledgments, if needed.
 

### 2011-06-27 talk/iwtc11/paper.tex
new abstract

@@
 \begin{abstract}
-By introducing loop peeling into the optimization step of a tracing
-jit the effect of optimizations already in place will be increased
-greatly. Not only will they become able to move loop invariant code
-out of loop. They will also become able to reuse results from the
-previous iteration. Also, the implementation of excising optimizations
-can be left almost intact as they will not have to interact much with
-the loop peeling.
-
-Several benchmarks, with few guard failures, executed on the
-PyPy Python JIT show over 2
-times increase in speed when loop peeling was introduced. This makes
-some of them almost match optimized C performance and become over 900
-times faster than CPython.
+One of the nice properties of a tracing JIT is that many of its optimization
+are simple requiring one forward pass. This is not true for loop-invariant code
+motion which is a very important optimization for code with tight kernels.
+In this paper we present a scheme for making simple optimizations loop-aware by
+using a simple pre-processing step on the trace and not changing the
+optimizations themselves. The scheme can give performance improvements of a
+factor over two for PyPy's Python JIT executing simple numerical kernels
+bringing the performance close to that of compiled C code.
 \end{abstract}

### 2011-06-28 talk/iwtc11/paper.tex
comment out XXXs
sprint planning tuesday

--- 2011-06-27 talk/iwtc11/paper.tex
@@
 \authorinfo{Maciej Fijałkowski}
-           {}
+           {Unaffiliated}
            {fijall@gmail.com}
@@
 However, we want to get this effect using our existing optimization passes
-without changing them too much. Simple optimizations with one forward pass
-cannot directly get this effect: They just look at the trace without taking
-into account that the trace executes many times in a row. Therefore to achieve
-loop-invariant code motion, we peel one iteration off the loop before running
-the optimizations. This peeling gives the following trace:
+without changing them too much. To achieve this, we peel one iteration off the
+loop before running the optimizations. This peeling gives the following trace:
 
@@
 arguments are inserted into the label of the loop itself and the jumps
-afterwards.
-
-This is the key insight of the proposed implementation scheme: Giving an
-optimization two iterations together at the same time gives the optimization
-enough context to remove operations from the peeled loop, because it detects
-that the operation was performed in the preamble already. Thus at runtime these
-moved operations are only executed once when entering the loop and the results
-are reused in further iterations.
+afterwards. Giving the optimizations two iterations together
+gives the optimization enough context to remove operations from the peeled loop,
+because it detects that the operation was performed in the preamble already.
 
@@
 
+XXX make clear that the preamble is not necessarily the \emph{first} iteration
+of a loop
+
 Before the trace is passed to a backend compiling it into machine code
@@
 previous iteration has happened.
-
-XXX (samuele): the point about the first iteration is hard to understand
 
@@
 vector of variables containing the values of all registered
-fields\footnote{This is sometimes called \emph{scalar replacement}.}.
-If some of the fields are themselves references to
+fields\footnote{This is sometimes called \emph{scalar replacement}. XXX check
+whether that's true}. If some of the fields are themselves references to
 allocation-removed objects they are recursively exploded
@@
 faster than either CPython or Psyco. This is due to the JIT compilation
-advantages and optimizations we discussed in previous work
-\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
-speedup of loop peeling is 70\%, which makes benchmark times
-comparable with native-compiled C code. We attribute the performance gap to C code to
+advantages and optimizations we discussed in XXX [ref to other paper]. Loop
+peeling gives an additional XXX on average, which makes benchmark times
+comparable with native-compiled C code. Missing performance we attribute to
 the relative immaturity of PyPy's JIT assembler backend as well as missing
@@
 
-The effect of combining a one ass optimization with loop peeling gives
-completely standard loop invariant code motion optimizations
-\cite{muchnick_advanced_1997}. We do not claim any novelty in the effect, but
-think that our implementation scheme is a very simple one.
+All the optimizations presented here are completely standard
+\cite{muchnick_advanced_1997}. XXX
 
@@
 have developped the described technique independently. There are no papers about
-LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
+LuaJIT but the author of it writes on a mailing list: "The LOOP pass does
 synthetic unrolling of the recorded IR, combining copy-substitution with
@@
 which allows reuse of all optimizations for redundancy elimination. Loop
-recurrences are detected on-the-fly and a minimized set of PHIs is generated.''
+recurrences are detected on-the-fly and a minimized set of PHIs is generated."
 \cite{pall_luajit_2009}

### 2011-06-29 talk/iwtc11/paper.tex
cleanups

--- 2011-06-27 talk/iwtc11/paper.tex
@@
 jit the effect of optimizations already in place will be increased
-greatly. Not only will they become able to move loop invariant code
+greatly. Not only does it make them move loop invariant code
 out of loop. They will also become able to reuse results from the
@@
 Before the trace is passed to a backend compiling it into machine code
-it needs to be optimized to achieve better performance.
-The focus of this paper
-is loop invariant code motion. The goal of that is to move as many
-operations as possible out of the loop making them executed at most once
+it is optimized to achieve better performance.
+One goal of that is to move 
+operations out of the loop making them executed only once
 and not every iteration. This we propose to achieve by loop peeling. It
@@
 to be taken when they are combined with loop peeling. This is
-described below by first explaining the loop peeling optimization
+described below by explaining the loop peeling optimization
 followed by a set of other optimizations and how they interact with
@@
 the end of itself. See Figure~\ref{fig:overview} for an illustration.
-The first part (called \emph{preamble}) finishes with the jump the the second part
-(called the \emph{peeled loop}). The second part end with the jump to itself. This way
+The first part (called \emph{preamble}) finishes with a jump the the second part
+(called the \emph{peeled loop}). The second part finishes with a jump to itself. This way
 the preamble will be executed only once while the peeled loop will
@@
 variables equal to the jump arguments of the preamble, $J$, and jump
-arguments $K$. Looking  at the peeled version of our example in Figure~\ref{fig:peeled-trace} we have
-\begin{equation}
-  %\left\{
-    \begin{array}{lcl}
-      I &=& \left( p_0, p_1 \right) \\
-      J &=& \left( p_0, p_5 \right) \\
-      K &=& \left( p_0, p_9 \right) \\
-    \end{array}
-  %\right.
-  .
-\end{equation}
+arguments $K$. 
+Figure~\ref{fig:overview} illustrates the general case. The running
+example in Figure~\ref{fig:unopt-trace} has  $I = \left( p_0, p_1
+\right)$ and $J = \left( p_0, p_5 \right)$. The result of applying
+loop peeling to it is shown in Figure~\ref{fig:peeled-trace} with 
+$K = \left( p_0, p_9 \right)$.
+
 To construct the second copy of the trace (the peeled loop) from the
@@
 next operation is copied, $m$ is extend by assigning $m\left(v\right) = \hat
-v$. For the example above, after all the operations have been copied we have
+v$. For the example above, that will extend $m$ with
 \begin{equation}
@@
     \begin{array}{lcl}
-      m\left(p_0\right) &=& p_0 \\
-      m\left(p_1\right) &=& p_5 \\
       m\left(i_2\right) &=& i_6 \\
@@
 \end{equation}
-
-The trace from Figure~\ref{fig:unopt-trace} would after this operation become
-the trace in Figure~\ref{fig:peeled-trace}. Line 1-13 shows the
-preamble while line 15-27 shows the peeled loop.
 
@@
 
-In general, after loop peeling and redundant operation removal the peeled loop
-will no longer be in SSA form as it operates on variables that are the result
-of pure operations in the preamble. The solution is to extend the input
+After loop peeling and redundant operation removal the peeled loop
+will typically no longer be in SSA form but operate on variables that are the result
+of operations in the preamble. The solution is to extend the input
 arguments, $J$, with those variables. This will also extend the
@@
 In general what is needed is to keep track of
-which variables from the preamble it reuses in the peeled loop.
-It has to construct a vector, $H$,  of such variables which
-can be used to update the input and jump arguments using
+which variables from the preamble are reused in the peeled loop.
+By constructing a vector, $H$,  of such variables, the input and jump
+arguments can be updated using
 \begin{equation}
@@
 possible to identify objects that are allocated within the loop but never
-escape it. Those objects have to be allocated in the loop, but no outside
-object ever gets a reference short lived objects with no references outside the
-loop. This
+escape it. That is, no outside
+object ever gets a reference to them. This
 is performed by processing the operations in order and
@@
 reference to the allocation-removed object in $p_5$. This can be achieved by
-exploding $p_5$ into the fields of the allocation-removed object.
-In this case there is only one such field and its value is
+exploding $p_5$ into the attributes of the allocation-removed object.
+In this case there is only one such attribute and its value is
 $i_4$, which means that $p_5$ is replaced with $i_4$ in the jump
@@
 vector of variables containing the values of all registered
-fields\footnote{This is sometimes called \emph{scalar replacement}.}.
-If some of the fields are themselves references to
+attributes\footnote{This is sometimes called \emph{scalar replacement}.}.
+If some of the attributes are themselves references to
 allocation-removed objects they are recursively exploded
 to make the vector contain only concrete variables. Some care has
-to be taken to always place the fields in the same order when
+to be taken to always place the attributes in the same order when
 performing this explosion. Notation becomes somewhat simpler if also every
@@
 we see improvements in several cases. The ideal loop for this optimization
-would be short numerical calculations with no failing guards and no
+is short and contains numerical calculations with no failing guards and no
 external calls. Larger loops involving many operations on complex objects
@@
 
-The effect of combining a one ass optimization with loop peeling gives
+The effect of combining a one pass optimization with loop peeling gives
 completely standard loop invariant code motion optimizations
@@
 loop if nothing in the loop can ever write to the memory location. It can also
-move allocations out of the loop, but does not replace the object by its fields.
-This saves only the allocation, not the access to the object fields.
+move allocations out of the loop, but does not replace the object by its attributes.
+This saves only the allocation, not the access to the object attributes.
 

### 2011-06-29 talk/iwtc11/paper.tex
squeeze into 6 pages

@@
 \begin{abstract}
-By introducing loop peeling into the optimization step of a tracing
-jit the effect of optimizations already in place will be increased
-greatly. Not only does it make them move loop invariant code
-out of loop. They will also become able to reuse results from the
-previous iteration. Also, the implementation of excising optimizations
-can be left almost intact as they will not have to interact much with
-the loop peeling.
-
-Several benchmarks, with few guard failures, executed on the
-PyPy Python JIT show over 2
-times increase in speed when loop peeling was introduced. This makes
-some of them almost match optimized C performance and become over 900
-times faster than CPython.
+One of the nice properties of a tracing JIT is that many of its optimization
+are simple requiring one forward pass. This is not true for loop-invariant code
+motion which is a very important optimization for code with tight kernels.
+In this paper we present a scheme for making simple optimizations loop-aware by
+using a simple pre-processing step on the trace and not changing the
+optimizations themselves. The scheme can give performance improvements of a
+factor over two for PyPy's Python JIT executing simple numerical kernels
+bringing the performance close to that of compiled C code.
 \end{abstract}
@@
 \begin{center}
-\includegraphics[scale=1]{figures/overview}
+\includegraphics[width=\columnwidth]{figures/overview}
 \end{center}
@@
 
-XXX find reference of prior work on this
+%XXX find reference of prior work on this
 
@@
 
-XXX (samuele): the point about the first iteration is hard to understand
+%XXX (samuele): the point about the first iteration is hard to understand
 

### 2011-07-01 talk/iwtc11/paper.tex
finetuning

@@
 One of the nice properties of a tracing JIT is that many of its optimization
-are simple requiring one forward pass. This is not true for loop-invariant code
+are simple requiring one forward pass only. This is not true for loop-invariant code
 motion which is a very important optimization for code with tight kernels.
@@
 The iteration of the loop that was peeled off (lines 1-4) is called the
-\emph{preamble}, the loop afterwards the \emph{peeled loop}.
+\emph{preamble}, the loop afterwards (lines 6-9) the \emph{peeled loop}.
 
@@
 
-Before the trace is passed to a backend compiling it into machine code
+Before a trace is passed to the backend compiling it into machine code
 it is optimized to achieve better performance.
@@
 combined with other optimizations it can increase the effectiveness of
-those optimizations. For many optimization of interest some care has
-to be taken when they are combined with loop peeling. This is
+those optimizations. For many optimization of interest only a few
+additional details has to be considered when they are combined with loop peeling. These are
 described below by explaining the loop peeling optimization
@@
 \end{figure}
-
-%XXX find reference of prior work on this
 

### 2011-07-23 talk/icooolps2011/talk/talk.tex
first draft of the talk

\documentclass[utf8x]{beamer}

% This file is a solution template for:

% - Talk at a conference/colloquium.
% - Talk length is about 20min.
% - Style is ornate.

\mode<presentation>
{
  \usetheme{Warsaw}
  % or ...

  %\setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}

\usepackage[english]{babel}
\usepackage{listings}
\usepackage{fancyvrb}
\usepackage{ulem}
\usepackage{color}
\usepackage{alltt}

\usepackage[utf8x]{inputenc}

\input{pygments}

\newcommand\redsout[1]{{\color{red}\sout{\hbox{\color{black}{#1}}}}}
\newcommand{\noop}{}

% or whatever

% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.

\title[Runtime Feedback in a Meta-Tracing JIT]{Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages}

\author[Carl Friedrich Bolz et. al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{3} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and Samuele Pedroni\inst{3} \and Armin Rigo\inst{1}}
% - Give the names in the same order as the appear in the paper.
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\institute[Heinrich-Heine-Universität Düsseldorf]
{$^1$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany \and

 $^2$merlinux GmbH, Hildesheim, Germany \and

 $^3$Open End, Göteborg, Sweden \and
}

\date{ICOOOLPS 2011, July 26, 2011}
% - Either use conference name or its abbreviation.
% - Not really informative to the audience, more for people (including
%   yourself) who are reading the slides online

% If you have a file called "university-logo-filename.xxx", where xxx
% is a graphic format that can be processed by latex or pdflatex,
% resp., then you can add a logo as follows:

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
%\AtBeginSubsection[]
%{
%  \begin{frame}<beamer>
%    \frametitle{Outline}
%    \tableofcontents[currentsection,currentsubsection]
%  \end{frame}
%}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
  \frametitle{Good JIT Compilers for Dynamic Languages are Hard}
  \begin{itemize}
      \item implementing the object model well is crucial
      \item recent languages like Python, Ruby, JS have complex core semantics
      \item many corner cases, even hard to interpret correctly
      \pause
      \item to get efficiency, the correct fast path through the tree of implementations must be generated XXX
      \item feedback of runtime information to the compiler is necessary
      \item correct exploitation of this information in the compiler
  \end{itemize}
  \pause
  \begin{block}{Problems}
      \begin{enumerate}
          \item implement all corner-cases of semantics correctly
          \item ... and the common cases efficiently
          \item feed back and exploit runtime information
      \end{enumerate}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{A Tracing JIT}
  \includegraphics[scale=0.5]{figures/trace01.pdf}
\end{frame}

\begin{frame}
  \frametitle{A Tracing JIT}
  \includegraphics[scale=0.5]{figures/trace02.pdf}
\end{frame}

\begin{frame}
  \frametitle{A Tracing JIT}
  \includegraphics[scale=0.5]{figures/trace03.pdf}
\end{frame}

\begin{frame}
  \frametitle{A Tracing JIT}
  \includegraphics[scale=0.5]{figures/trace04.pdf}
\end{frame}

\begin{frame}
  \frametitle{Tracing JITs}
  Advantages:
  \begin{itemize}
      \item can be added to existing VM
      \item interpreter does a lot of work
      \item can fall back to interpreter for uncommon paths
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Granularity Problems}
  \begin{itemize}
      \item if the tracer records bytecode, not enough information is there
      \item many dynamic languages have bytecodes that contain complex logic
      \item need to expand the bytecode in the trace into something more explicit
      \item this duplicates the lanuage semantics in the tracer/optimizer
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Example: Attribute Reads in Python}
  What happens when an attribute \texttt{x.m} is read? (simplified)
  \begin{itemize}
      \item check for \texttt{x.\_\_getattribute\_\_}, if there, call it
      \pause
      \item look for the attribute in the object's dictionary, if it's there, return it
      \pause
      \item walk up the MRO and look in each class' dictionary for the attribute
      \pause
      \item if the attribute is found, call its \texttt{\_\_get\_\_} attribute and return the result
      \pause
      \item if the attribute is not found, look for \texttt{x.\_\_getattr\_\_}, if there, call it
      \pause
      \item raise an \texttt{AttributeError}
  \end{itemize}
  \pause
  all this is one bytecode
\end{frame}

\begin{frame}
  \frametitle{Idea of Meta-Tracing}
  \includegraphics[scale=0.5]{figures/trace05.pdf}
\end{frame}

\begin{frame}
  \frametitle{Meta-Tracing}
  \includegraphics[scale=0.5]{figures/metatrace01.pdf}
\end{frame}

\begin{frame}
  \frametitle{Meta-Tracing JITs}
  \begin{block}{Advantages:}
    \begin{itemize}
        \item semantics is always like that of the interpreter
        \item trace fully contains language semantics
        \item meta-tracers can be reused for various interpreters
    \end{itemize}
  \end{block}
  \pause
  a few meta-tracing systems have been built:
  \begin{itemize}
      \item Sullivan et.al. describe a meta-tracer using the Dynamo RIO system
      \item Yermolovich et.al. run a Lua implementation on top of a tracing JS implementation
      \item SPUR is a tracing JIT for CLR bytecodes, which is used to speed up a JS implementation in C\#
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{PyPy}
  A general environment for implementing dynamic languages
  \pause
  \begin{block}{Approach}
      \begin{itemize}
          \item write an interpreter for the language in RPython
          \item compilable to an efficient C-based VM
          \pause
          \item (RPython is a restricted subset of Python)
      \end{itemize}
  \end{block}
  \pause
  \begin{block}{PyPy's Meta-Tracing JIT}
      \begin{itemize}
          \item PyPy contains a meta-tracing JIT for interpreters in RPython
          \item needs a few source-code hints (or annotations) in the interpreter
          \item powerful general optimizations
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Runtime Feedback}
  Problems of Naive Meta-Tracing:
  \begin{itemize}
      \item no runtime feedback of user-level types
      \item tracer does not know about invariants in the interpreter
  \end{itemize}
  \pause
  \begin{block}{Proposed Solutions}
      \begin{itemize}
          \item introduce \textit{hints} that the interpreter-author can use
          \item hints are annotation in the interpreter
          \item they give information to the meta-tracer
          \pause
          \item two hints presented here
          \item one to induce runtime feedback of arbitrary information
          \item the second one to influence constant folding
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Example: Instances with Maps}
  \includegraphics[scale=0.5]{figures/map01.pdf}
\end{frame}

\begin{frame}
  \frametitle{Example: Instances with Maps}
  \includegraphics[scale=0.5]{figures/map02.pdf}
\end{frame}

\begin{frame}[containsverbatim]
\frametitle{Map Implementation}

\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{indexes}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{n}{indexes}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{o}{.}\PY{o}{.}\PY{o}{.}

\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}\PY{p}{)}
\end{Verbatim}
\end{frame}

\begin{frame}[plain,containsverbatim]
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}

    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n}{index} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
        \PY{k}{return} \PY{n+nb+bp}{None}

    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{o}{.}\PY{o}{.}\PY{o}{.}
\end{Verbatim}
\end{frame}

\begin{frame}[plain,containsverbatim]
\frametitle{Trace for code \texttt{inst.a + inst.b}}
\begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
# $inst_1$.getattr("a")
$map_1$ = $inst_1$.map
$index_1$ = Map.getindex($map_1$, "a")
guard($index_1$ != -1)
$storage_1$ = $inst_1$.storage
$result_1$ = $storage_1$[$index_1$]
|\pause|
# $inst_1$.getattr("b")
$map_2$ = $inst_1$.map
$index_2$ = Map.getindex($map_2$, "b")
guard($index_2$ != -1)
$storage_2$ = $inst_1$.storage
$result_2$ = $storage_2$[$index_2$]

$v_1$ = $result_1$ + $result_2$
return($v_1$)
\end{lstlisting}
\end{frame}

\begin{frame}[plain,containsverbatim]
\frametitle{Trace for code \texttt{inst.a + inst.b}}
\begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
# $inst_1$.getattr("a")
$map_1$ = $inst_1$.map
$index_1$ = Map.getindex($map_1$, "a")
guard($index_1$ != -1)
$storage_1$ = $inst_1$.storage
$result_1$ = $storage_1$[$index_1$]

# $inst_1$.getattr("b")
$map_2$ = $inst_1$.map
$index_2$ = Map.getindex($map_2$, "b")
guard($index_2$ != -1)
$storage_2$ = $inst_1$.storage
$result_2$ = $storage_2$[$index_2$]

$v_1$ = $result_1$ + $result_2$
return($v_1$)
\end{lstlisting}
\end{frame}

\begin{frame}[containsverbatim]
  \frametitle{Runtime Feedback Controlled by the Interpreter Author}
  \begin{itemize}
      \item give the interpreter author a way to feed back runtime values into the trace
      \item written as \texttt{promote(x)}
      \item captures the argument's runtime value during tracing
      \item should be used only for variables that take few values
  \end{itemize}
  \pause

  \begin{minipage}[b]{6cm}
      \centering
      {\noop
      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    promote(x)
    z = x * 2 + 1
    return z + y
      \end{lstlisting}
      }
  \end{minipage}
  \vline
  \hspace{0.5cm}
  \begin{minipage}[b]{4cm}
      {\noop
      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
      \end{lstlisting}
      }
  \end{minipage}
\end{frame}

\begin{frame}[containsverbatim]
  \frametitle{Runtime Feedback Controlled by the Interpreter Author}
  \begin{itemize}
      \item give the interpreter author a way to feed back runtime values into the trace
      \item written as \texttt{promote(x)}
      \item captures the argument's runtime value during tracing
      \item should be used only for variables that take few values
  \end{itemize}

  \begin{minipage}[b]{6cm}
      \centering
      {\noop
      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    promote(x)
    z = x * 2 + 1
    return z + y

      \end{lstlisting}
      }
  \end{minipage}
\end{frame}

\begin{frame}[containsverbatim]
  \frametitle{Runtime Feedback Controlled by the Interpreter Author}
  \begin{itemize}
      \item give the interpreter author a way to feed back runtime values into the trace
      \item written as \texttt{promote(x)}
      \item captures the argument's runtime value during tracing
      \item should be used only for variables that take few values
  \end{itemize}

  \begin{minipage}[b]{6cm}
      \centering
      {\noop
      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    promote(x)
    z = x * 2 + 1
    return z + y

      \end{lstlisting}
      }
  \end{minipage}
  \vline
  \hspace{0.5cm}
  \begin{minipage}[b]{4cm}
      {\noop
      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
      \end{lstlisting}
      }
  \end{minipage}
\end{frame}

\begin{frame}
  \frametitle{Foldable Operations Defined by the Interpreter Author}
  \begin{itemize}
      \item let the interpreter author define foldable functions
      \item those functions typically don't look foldable
      \item otherwise there is no need for an annotation
      \item done via a function decorator \texttt{@elidable}
      \pause
      \item decorated functions should be pure
      \item or have idempotent side effects (such as a function that memoizes)
      \item trace optimizer will remove calls to such functions with constant arguments
  \end{itemize}
\end{frame}

\begin{frame}[containsverbatim]
\frametitle{Adding Hints to Maps}

\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{indexes}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{n}{indexes}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{n+nd}{@elidable}
    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}

    \PY{n+nd}{@elidable}
    \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{o}{.}\PY{o}{.}\PY{o}{.}

\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
\end{Verbatim}
\end{frame}

\begin{frame}[containsverbatim]
\frametitle{Adding Hints to Maps}

\begin{Verbatim}[commandchars=\\\{\}]
\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}\PY{p}{)}

\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}

    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n}{promote}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{)}
        \PY{n}{index} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
        \PY{k}{return} \PY{n+nb+bp}{None}

    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{o}{.}\PY{o}{.}\PY{o}{.}
\end{Verbatim}
\end{frame}

\begin{frame}[containsverbatim,plain]
  \frametitle{Trace with Hints}
\begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
# $inst_1$.getattr("a")
$map_1$ = $inst_1$.map
guard($map_1$ == 0xb74af4a8)
|{\color{gray}$index_1$ = Map.getindex($map_1$, "a")|
|{\color{gray}guard($index_1$ != -1)|
$storage_1$ = $inst_1$.storage
$result_1$ = $storage_1$[$index_1$]

# $inst_1$.getattr("b")
|{\color{gray}$map_2$ = $inst_1$.map|
|{\color{gray}guard($map_2$ == 0xb74af4a8)|
|{\color{gray}$index_2$ = Map.getindex($map_2$, "b")|
|{\color{gray}guard($index_2$ != -1)|
|{\color{gray}$storage_2$ = $inst_1$.storage|
$result_2$ = $storage_2$[$index_2$]

$v_1$ = $result_1$ + $result_2$
return($v_1$)
\end{lstlisting}
\end{frame}

\begin{frame}[containsverbatim,plain]
  \frametitle{Final Trace}
\begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
# $inst_1$.getattr("a")
$map_1$ = $inst_1$.map
guard($map_1$ == 0xb74af4a8)
$storage_1$ = $inst_1$.storage
$result_1$ = $storage_1$[$0$]
$result_2$ = $storage_2$[$1$]
$v_1$ = $result_1$ + $result_2$
return($v_1$)
\end{lstlisting}
\end{frame}

\begin{frame}
  \frametitle{Uses of These Hints}
  \begin{block}{\texttt{promote} lets one specialize on various things:}
      \begin{itemize}
          \item user-level types
          \item shapes of instances
          \item the current state of a classes' methods
          \item ...
      \end{itemize}
  \end{block}
  \pause
  \begin{block}{uses of \texttt{@elidable}}
      \begin{itemize}
          \item define immutable fields by decorating a getter
          \item declare arbitrary invariants
      \end{itemize}

  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Some Benchmarks}
  \begin{itemize}
      \item benchmarks done using PyPy's Python interpreter
      \item about 30'000 lines of code
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Some Benchmarks}
  \includegraphics[scale=0.5]{figures/bench.pdf}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
      \item meta-tracing can make the efficient implementation of complex dynamic languaes easier
      \item two kinds of hints to be added by the interpreter author allows arbitrary runtime feedback
      \item the hints are expressive enough to re-implement classical optimizations such as maps
      \item usage of the hints leads to good speedups for object-oriented code in PyPy's Python interpreter
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Bonus: Comparison with Partial Evaluation}
\end{frame}

\end{document}

### 2011-07-25 talk/icooolps2011/talk/talk.tex
tweaks

@@
   \begin{itemize}
-      \item implementing the object model well is crucial
       \item recent languages like Python, Ruby, JS have complex core semantics
@@
       \pause
-      \item to get efficiency, the correct fast path through the tree of implementations must be generated XXX
       \item feedback of runtime information to the compiler is necessary
@@
 \begin{frame}
-  \frametitle{A Tracing JIT}
+  \frametitle{An Interpreter}
   \includegraphics[scale=0.5]{figures/trace02.pdf}
@@
           \item PyPy contains a meta-tracing JIT for interpreters in RPython
-          \item needs a few source-code hints (or annotations) in the interpreter
+          \item needs a few source-code hints (or annotations) \emph{in the interpreter}
           \item powerful general optimizations
@@
         \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{n}{indexes}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
+        \PY{o}{.}\PY{o}{.}\PY{o}{.}
 
@@
 \begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
-# $inst_1$.getattr("a")
+# $inst_1$.getfield("a")
 $map_1$ = $inst_1$.map
@@
 |\pause|
-# $inst_1$.getattr("b")
+# $inst_1$.getfield("b")
 $map_2$ = $inst_1$.map
@@
 $v_1$ = $result_1$ + $result_2$
-return($v_1$)
 \end{lstlisting}
@@
 \begin{frame}[plain,containsverbatim]
-\frametitle{Trace for code \texttt{inst.a + inst.b}}
+\frametitle{Trace for Code \texttt{inst.a + inst.b}}
 \begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
-# $inst_1$.getattr("a")
+# $inst_1$.getfield("a")
 $map_1$ = $inst_1$.map
@@
 
-# $inst_1$.getattr("b")
+# $inst_1$.getfield("b")
 $map_2$ = $inst_1$.map
@@
   \pause
+\end{frame}
 
+\begin{frame}[containsverbatim]
+  \frametitle{Tiny Example}
   \begin{minipage}[b]{6cm}
@@
       {\noop
-      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
+      \begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]
 guard($x_1$ == 4)
-$v_1$ = $x_1$ * 2
-$z_1$ = $v_1$ + 1
-$v_2$ = $z_1$ + $y_1$
-return($v_2$)
-      \end{lstlisting}
-      }
-  \end{minipage}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-  \frametitle{Runtime Feedback Controlled by the Interpreter Author}
-  \begin{itemize}
-      \item give the interpreter author a way to feed back runtime values into the trace
-      \item written as \texttt{promote(x)}
-      \item captures the argument's runtime value during tracing
-      \item should be used only for variables that take few values
-  \end{itemize}
-
-  \begin{minipage}[b]{6cm}
-      \centering
-      {\noop
-      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-def f1(x, y):
-    promote(x)
-    z = x * 2 + 1
-    return z + y
-
-      \end{lstlisting}
-      }
-  \end{minipage}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-  \frametitle{Runtime Feedback Controlled by the Interpreter Author}
-  \begin{itemize}
-      \item give the interpreter author a way to feed back runtime values into the trace
-      \item written as \texttt{promote(x)}
-      \item captures the argument's runtime value during tracing
-      \item should be used only for variables that take few values
-  \end{itemize}
-
-  \begin{minipage}[b]{6cm}
-      \centering
-      {\noop
-      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-def f1(x, y):
-    promote(x)
-    z = x * 2 + 1
-    return z + y
-
-      \end{lstlisting}
-      }
-  \end{minipage}
-  \vline
-  \hspace{0.5cm}
-  \begin{minipage}[b]{4cm}
-      {\noop
-      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($x_1$ == 4)
-$v_1$ = $x_1$ * 2
-$z_1$ = $v_1$ + 1
+|{\color{gray}$v_1$ = $x_1$ * 2}|
+|{\color{gray}$z_1$ = $v_1$ + 1}|
 $v_2$ = $z_1$ + $y_1$
@@
 
-\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
+\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}\PY{p}{)}
 \end{Verbatim}
@@
 \begin{Verbatim}[commandchars=\\\{\}]
-\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}\PY{p}{)}
-
 \PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
@@
 \begin{frame}[containsverbatim,plain]
-  \frametitle{Trace with Hints}
+  \frametitle{Trace with Hints for Code \texttt{inst.a + inst.b}}
 \begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
-# $inst_1$.getattr("a")
+# $inst_1$.getfield("a")
 $map_1$ = $inst_1$.map
 guard($map_1$ == 0xb74af4a8)
-|{\color{gray}$index_1$ = Map.getindex($map_1$, "a")|
-|{\color{gray}guard($index_1$ != -1)|
+|{\color{gray}$index_1$ = Map.getindex($map_1$, "a")}|
+|{\color{gray}guard($index_1$ != -1)}|
 $storage_1$ = $inst_1$.storage
-$result_1$ = $storage_1$[$index_1$]
+$result_1$ = $storage_1$[$index_1$}]
 
-# $inst_1$.getattr("b")
+# $inst_1$.getfield("b")
 |{\color{gray}$map_2$ = $inst_1$.map|
-|{\color{gray}guard($map_2$ == 0xb74af4a8)|
-|{\color{gray}$index_2$ = Map.getindex($map_2$, "b")|
-|{\color{gray}guard($index_2$ != -1)|
-|{\color{gray}$storage_2$ = $inst_1$.storage|
+|{\color{gray}guard($map_2$ == 0xb74af4a8)}|
+|{\color{gray}$index_2$ = Map.getindex($map_2$, "b")}|
+|{\color{gray}guard($index_2$ != -1)}|
+|{\color{gray}$storage_2$ = $inst_1$.storage}|
 $result_2$ = $storage_2$[$index_2$]
@@
 \begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
-# $inst_1$.getattr("a")
 $map_1$ = $inst_1$.map
@@
       \item about 30'000 lines of code
+      \item 20 calls to \texttt{promote}
+      \item 10 applications of \texttt{@elidable}
   \end{itemize}
@@
   \begin{itemize}
-      \item meta-tracing can make the efficient implementation of complex dynamic languaes easier
-      \item two kinds of hints to be added by the interpreter author allows arbitrary runtime feedback
+      \item meta-tracing can make the efficient implementation of complex dynamic languages easier
+      \item only requires to write a correct interpreter
+      \item two kinds of hints to be added by the interpreter author allow arbitrary runtime feedback and its exploitation
       \item the hints are expressive enough to re-implement classical optimizations such as maps
@@
   \frametitle{Bonus: Comparison with Partial Evaluation}
+  \begin{itemize}
+      \pause
+      \item the only difference between meta-tracing and partial evaluation is that meta-tracing works
+      \pause
+      \item ... mostly kidding
+      \pause
+      \item very similar from the motivation and ideas
+      \item PE was never scaled up to perform well on large interpreters
+      \item classical PE mostly ahead of time
+      \item PE tried very carefully to select the right paths to inline and optimize
+      \item quite often this fails and inlines too much or too little
+      \item tracing is much more pragmatic: simply look what happens
+  \end{itemize}
 \end{frame}

### 2011-12-19 talk/icooolps2011/talk/talk.tex
this was done by the list-strategies branch
typo in talk slides

@@
 \begin{frame}
-  \frametitle{A Tracing JIT}
+  \frametitle{An Interpreter}
   \includegraphics[scale=0.5]{figures/trace01.pdf}
@@
 \begin{frame}
-  \frametitle{An Interpreter}
+  \frametitle{A Tracing JIT}
   \includegraphics[scale=0.5]{figures/trace02.pdf}
@@
   What happens when an attribute \texttt{x.m} is read? (simplified)
+  \pause
   \begin{itemize}
@@
     \begin{itemize}
-        \item semantics is always like that of the interpreter
+        \item semantics are always like that of the interpreter
         \item trace fully contains language semantics
@@
   \frametitle{Example: Instances with Maps}
-  \includegraphics[scale=0.5]{figures/map01.pdf}
+  \includegraphics[scale=0.7]{figures/map01.pdf}
 \end{frame}
@@
   \frametitle{Example: Instances with Maps}
-  \includegraphics[scale=0.5]{figures/map02.pdf}
+  \includegraphics[scale=0.7]{figures/map02.pdf}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Example: Instances with Maps}
+  \includegraphics[scale=0.7]{figures/map03.pdf}
 \end{frame}
@@
 \begin{frame}[plain,containsverbatim]
-\frametitle{Trace for code \texttt{inst.a + inst.b}}
+\frametitle{Trace for Code \texttt{inst.a + inst.b}}
 \begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
@@
 
-    \PY{n+nd}{@elidable}
     \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
@@
 $storage_1$ = $inst_1$.storage
-$result_1$ = $storage_1$[$index_1$}]
+$result_1$ = $storage_1$[$index_1$]
 

### 2012-03-07 talk/pycon2012/jit/slides.tex
add my jit talk

\documentclass[14pt]{beamer}

\mode<presentation>
{
  \usetheme{default}
  % or ...

  \setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}

% \usepackage{times}
% \usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.

\title{How the PyPy JIT works}

\author{Benjamin Peterson \\ benjamin@python.org}
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\date{PyCon 2012}

% If you have a file called "university-logo-filename.xxx", where xxx
% is a graphic format that can be processed by latex or pdflatex,
% resp., then you can add a logo as follows:

% \pgfdeclareimage[height=0.5cm]{university-logo}{university-logo-filename}
% \logo{\pgfuseimage{university-logo}}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\lstset{basicstyle=\scriptsize\ttfamily}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

% Since this a solution template for a generic talk, very little can
% be said about how it should be structured. However, the talk length
% of between 15min and 45min and the theme suggest that you stick to
% the following rules:

% - Exactly two or three sections (other than the summary).
% - At *most* three subsections per section.
% - Talk about 30s to 2min per frame. So there should be between about
%   15 and 30 frames, all told.

\begin{frame}
  % - A title should summarize the slide in an understandable fashion
  %   for anyone how does not follow everything on the slide itself.

  ``If the implementation is hard to explain, it's a bad idea.''
  \vskip0pt plus.5fill
  \uncover<2->{Except PyPy!}
\end{frame}

\begin{frame}{Design Principles}
\begin{itemize}
\item interpreter agnostic
\item but allow language implementation to help JIT
\item compile hot code: loops or functions
\item use runtime information to remove dynamic overhead
\item optimize common and idiomatic code
\end{itemize}
\end{frame}

\begin{frame}{Overview}
\begin{itemize}
\item translation
\item runtime
\begin{itemize}
\item tracing
\item optimization
\item assembly generation
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Adding JIT to an interpreter}
\begin{itemize}
\item JIT args
\begin{itemize}
\item ``greens'' = constants
\item ``reds'' = variables
\end{itemize}
\item hints
\begin{itemize}
\item \verb+can_enter_jit+
\item \verb+jit_merge_point+
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{JIT on the Python interpreter}
\begin{itemize}
\item greens: code object and instruction pointer
\item reds: frame object and execution context
\end{itemize}
\begin{lstlisting}[language=Python]
while True:
    jit_merge_point()
    next_instr = self.handle_bytecode(code, next_instr)
\end{lstlisting}
\begin{lstlisting}[language=Python]
def JUMP_ABSOLUTE(jump_to):
    can_enter_jit()
    return jump_to
\end{lstlisting}
\end{frame}

\begin{frame}{Generating the JIT}
\begin{itemize}
\item low-level flowgraphs converted to jitcodes
\item interpreter decides what functions to ``look into''
\item jitcodes stored in binary
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Main example: Primality Test}
\begin{lstlisting}[language=Python]
def is_prime(n):
    limit = int(n**.5)
    i = 2
    while i <= limit:
        if n % i == 0:
            return False
        i += 1
    return True
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]{Primality Test - Bytecode}
\tiny{
\begin{verbatim}
# while i <= limit:
22 SETUP_LOOP         46 (to 71)
25 LOAD_FAST          2 (i)
28 LOAD_FAST          1 (limit)
31 COMPARE_OP         1 (<=)
34 POP_JUMP_IF_FALSE  70
# if i % 1 === 0:
37 LOAD_FAST          0 (n)
40 LOAD_FAST          2 (i)
43 BINARY_MODULO       
44 LOAD_CONST         3 (0)
47 COMPARE_OP         2 (==)
50 POP_JUMP_IF_FALSE  57
#     return False
53 LOAD_GLOBAL        1 (False)
56 RETURN_VALUE        
# i += 1
57 LOAD_FAST          2 (i)
60 LOAD_CONST         4 (1)
63 INPLACE_ADD         
64 STORE_FAST         2 (i)
67 JUMP_ABSOLUTE      25
\end{verbatim}
}
\end{frame}

\begin{frame}[fragile]{Activating the JIT}
\begin{itemize}
\item every loop has a counter incremented when \verb+can_enter_jit+ is reached
\item tracing entered when threshold reached
\item slight fudging to avoid compiling everything suddenly
\end{itemize}
\end{frame}

\begin{frame}{Tracing}
\begin{itemize}
\item metainterpreter executes saved jitcodes
\item records operations in JIT IR
\item traces through function calls, flattening loop
\item program values stored in boxes (pointers, integers, floats)
\item tracing ends when loop is closed
\end{itemize}
\end{frame}

\begin{frame}{Guards}
\begin{itemize}
\item as the jit traces, it makes assumptions based on runtime information
\item these must hold otherwise trace is invalid
\item guards protect trace from different runtime conditions
\item examples: conditional branches, overflow, exceptions, other code accessing JIT data
\end{itemize}
\end{frame}

\begin{frame}[fragile]{JIT IR example - Head of loop}
\footnotesize{
\begin{verbatim}
# arguments to loop
[p0, p1, p2, p3, i4, p5, i6, i7, p8, p9, p10, p11, p12, p13]
#25 LOAD_FAST
guard_value(i6, 3)
guard_nonnull(p10)
guard_value(i4, 0)
#28 LOAD_FAST
guard_nonnull(p9)
#31 COMPARE_OP
guard_class(p10, W_IntObject)
guard_class(p9, W_IntObject)
i18 = getfield_gc_pure(p10, <W_IntObject.inst_intval>)
i19 = getfield_gc_pure(p9, <W_IntObject.inst_intval>)
i20 = int_le(i18, i19)
guard_true(i20)
#34 POP_JUMP_IF_FALSE
\end{verbatim}
}
\end{frame}

\begin{frame}[fragile]{JIT IR example - computing modulo}
\footnotesize{
\begin{verbatim}
#37 LOAD_FAST
guard_nonnull(p8)
#40 LOAD_FAST
#43 BINARY_MODULO
guard_class(p8, W_IntObject)
i22 = getfield_gc_pure(p8, <W_IntObject.inst_intval>)
i23 = int_is_zero(i18)
guard_false(i23)
i25 = int_eq(i22, -9223372036854775808)
guard_false(i25)
i26 = int_mod(i22, i18)
i28 = int_lt(i18, 0)
guard_false(i28)
i30 = int_rshift(i26, 63)
i31 = int_and(i18, i30)
i32 = int_add(i26, i31)
p34 = new_with_vtable(W_IntObject)
setfield_gc(p34, i32, <W_IntObject.inst_intval>)
\end{verbatim}
}
\end{frame}

\begin{frame}[fragile]{JIT IR example - check modulo}
\footnotesize{
\begin{verbatim}
#44 LOAD_CONST
guard_value(p3, ConstPtr(ptr35))
#47 COMPARE_OP')
i37 = int_eq(i32, 0)
guard_false(i37)
#50 POP_JUMP_IF_FALSE
\end{verbatim}
}
\end{frame}

\begin{frame}[fragile]{JIT IR example - finish loop}
\footnotesize{
\begin{verbatim}
#57 LOAD_FAST
#60 LOAD_FAST
#63 INPLACE_ADD
i38 = getfield_gc_pure(p10, <W_IntObject.inst_intval>)
i40 = int_add_ovf(i38, 1)
guard_no_overflow()
p42 = new_with_vtable(W_IntObject)
setfield_gc(p42, i40, <W_IntObject.inst_intval>)
#64 STORE_FAST
#67 JUMP_ABSOLUTE
\end{verbatim}
}
\end{frame}

\begin{frame}{Optimizations}
\begin{itemize}
\item limited by speed
\item classical compiler optimizations
\begin{itemize}
\item constant folding
\item strength reduction
\item integer bounds
\end{itemize}
\item most important is removing indirection: virtuals and virtualizables
\end{itemize}
\end{frame}

\begin{frame}{Virtuals}
\begin{itemize}
\item if objects allocated in the loop don't escape it, we don't have to
  allocate them
\item removes boxing
\item all field access can be killed
\item in our example, removes integer allocations
\end{itemize}
\end{frame}

\begin{frame}{Virtualizables}
\begin{itemize}
\item like virtuals but allowed to escape trace
\item handled at metainterp level
\item Python interpreter: the frame object
\item possibility of escape creates bookkeeping problems
\item advantage of JIT generator approach
\end{itemize}
\end{frame}

\begin{frame}{Unrolling}
\begin{itemize}
\item need some way to compute loop invariants before reaching optimized loop
\item solution: two optimization passes
\item first iteration computes loop invariants
\item second iteration is tight loop
\end{itemize}
\end{frame}

\begin{frame}{JIT viewer}
Our fancy toy for visualizing JIT
\end{frame}

\begin{frame}{Creating assembly}
\begin{itemize}
\item surprisingly straightforward
\item linear register allocator
\item gc has to be informed of dynamic allocations
\item assembly now ready to be run
\end{itemize}
\end{frame}

\begin{frame}{Bailing from the JIT}
\begin{itemize}
\item guards fail while assembly is running
\item assembler has written a compact description of hower to recover state
\item state reconstructed
\item blackhole interpreter
\item if gaurd fails repeatedly, compile a bridge
\end{itemize}
\end{frame}

\begin{frame}[fragile]{How the Python interpreter helps the JIT}
\begin{itemize}
\item constant promotion
\item decorators: \verb+unroll_safe+, \verb+elidable+
\item data structures optimized for the JIT
\begin{itemize}
\item map dicts
\item cell dicts
\item list and dict strategies
\item specialized tuples
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Map dicts}
\begin{itemize}
\item shares object attributes
\item instance data stored in array
\item memory efficient as well as good for the JIT
\end{itemize}
\includegraphics[scale=.45]{instancemap.png}
\end{frame}

\begin{frame}[fragile]{Thank you}
Questions?

Resources
\begin{itemize}
\item \#pypy on Freenode
\item http://morepypy.blogspot.com
\item pypy-dev@python.org
\item The Architecture of Open Source Applications, Volume 2
\end{itemize}

My email: benjamin@python.org
\end{frame}

\end{document}

### 2012-03-07 talk/pycon2012/jit/slides.tex
also this
hopefully make this clearer

@@
 \begin{itemize}
+\item RPython interpreter is in low-level flowgraphs
 \item low-level flowgraphs converted to jitcodes
@@
 [p0, p1, p2, p3, i4, p5, i6, i7, p8, p9, p10, p11, p12, p13]
+# while i < limit
 #25 LOAD_FAST
@@
 \begin{verbatim}
+# n % i
 #37 LOAD_FAST
@@
 guard_class(p8, W_IntObject)
-i22 = getfield_gc_pure(p8, <W_IntObject.inst_intval>)
+i22 = ((W_IntObject)p8)->inst_intval)
 i23 = int_is_zero(i18)
@@
 p34 = new_with_vtable(W_IntObject)
-setfield_gc(p34, i32, <W_IntObject.inst_intval>)
+((W_IntObject)p34)->inst_intval = i32
 \end{verbatim}
@@
 \begin{verbatim}
+# if <modulo> == 0:
 #44 LOAD_CONST
@@
 \begin{verbatim}
+# i += 1
 #57 LOAD_FAST
@@
 #63 INPLACE_ADD
-i38 = getfield_gc_pure(p10, <W_IntObject.inst_intval>)
+i38 = ((W_IntObject)p10)->inst_intval
 i40 = int_add_ovf(i38, 1)
@@
 p42 = new_with_vtable(W_IntObject)
-setfield_gc(p42, i40, <W_IntObject.inst_intval>)
+((W_IntObject)p42)->inst_inval = i40
 #64 STORE_FAST
@@
 \end{itemize}
+\item remove extra gaurds
 \item most important is removing indirection: virtuals and virtualizables

### 2012-03-07 talk/pycon2012/jit/slides.tex
another slide

@@
 
+\begin{frame}{Conclusion}
+\begin{itemize}
+\item JIT generated from RPython description of interpreter
+\item metainterpreter traces hot loops and functions
+\item optimizations remove indirection
+\item adapt to new runtime information with bridges
+\item interpreter optimized for JIT with special hints and data structures
+\end{itemize}
+\end{frame}
+
 \begin{frame}[fragile]{Thank you}
@@
 \item \#pypy on Freenode
+\item PyPy sprint
 \item http://morepypy.blogspot.com
 \item pypy-dev@python.org
-\item The Architecture of Open Source Applications, Volume 2
+\item The Architecture of Open Source Applications, Volume 2 (coming out soon)
 \end{itemize}

### 2012-06-30 talk/dls2012/paper.tex
work on slides a bit

--- 2011-07-01 talk/iwtc11/paper.tex
@@
 \section{Introduction}
+
+A dynamically typed language needs to do a lot of type
+checking and unwrapping. For tight computationally intensive loops a
+significant amount of the execution time might be spend on such tasks
+instead of the actual calculations. Moreover, the type checking and
+unwrapping is often loop invariant and performance could be increased
+by moving those operations out of the loop. We propose to design a
+loop-aware tracing JIT to perform such optimization at run time.
 

### 2012-07-03 talk/ecoop2009/tlc.tex
first part of my talk, roughly

@@
 \end{figure}
-
-\commentout{
-
-\anto{I think this example could cause confusion, because the synatx is TOO
-  similar to (R)Python, hence in the next section it could be unclear which is
-  the leves of the examples}
-
-Since reading TLC programs at bytecode level is hard, in this paper we will
-use an invented Python-like syntax to describe examples, although we need to
-remind that the actual programs are written in the assembler language showed
-above. Thus, the example above can be written in the following way:
-
-\begin{lstlisting}
-def main(n):
-    if n<0:
-        return -n
-    return n
-\end{lstlisting}
-}

### 2012-07-03 talk/vmil2012/paper.tex
get started with some ideas for the RPython section

--- 2011-06-08 talk/icooolps2011/paper.tex
@@
 
-% compressing itemize env, in case we need it
-\newenvironment{zitemize}% zero - line spacing itemize environment
-   {\begin{list}{--}{
-   \setlength{\itemsep}{0 pt}
-   \setlength{\parsep}{0 pt}
-   \setlength{\topsep} {0 pt} }}% the end stuff
-   {\end{list}}
-
 \definecolor{gray}{rgb}{0.5,0.5,0.5}
@@
 
-\title{Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages}
+\title{Efficiently Handling Guards in the low level design of RPython's tracing JIT}
 
-\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^{a, c}$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
-            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
+\authorinfo{Carl Friedrich Bolz$^a$ \and David Schneider$^{a}$}
            {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
+           }
+           {XXX emails}
 
-            $^b$merlinux GmbH, Hildesheim, Germany
-
-            $^c$Open End, Göteborg, Sweden
-           }
-           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu \and
-           leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}
-
-\conferenceinfo{ICOOOLPS}{'11 Lancaster, UK}
-\CopyrightYear{2011}
-\crdata{XXX}
+\conferenceinfo{VMIL'11}{}
+\CopyrightYear{2012}
+\crdata{}
 
@@
 
-\keywords{Tracing JIT, Runtime Feedback, Interpreter, Code Generation, Meta-Programming}
+\keywords{XXX}
 
 \begin{abstract}
-
-Meta-tracing JIT compilers can be applied to a variety of different
-languages without explicitly encoding language semantics into the compiler. So
-far, they lacked a way to give the language implementor control over runtime
-feedback. This restricted their performance. In this paper we describe the
-mechanisms in PyPy's meta-tracing JIT that can be used to control
-runtime feedback in language-specific ways. These mechanisms are flexible
-enough to express classical VM techniques such as maps and runtime type feedback.
 
@@
 
-One of the hardest parts of implementing an object-oriented dynamic language well is to
-optimize its object model. This is made harder by the complexity of the core
-object semantics of many recent languages such as Python, JavaScript or Ruby.
-For them, even implementing just an interpreter is already a difficult
-task. Implementing these languages efficiently with a just-in-time compiler (JIT) is
-extremely challenging, because of their many corner-cases.
-
-It has long been an objective of the partial evaluation community to
-automatically produce compilers from interpreters. There has been a
-renaissance of this idea around the approach of tracing just-in-time
-compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
-a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
-\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython \cite{davide_ancona_rpython:_2007} (a restricted
-subset of Python). This JIT is then used to trace a number of languages
-implementations written in RPython. A number of other experiments in this
-directions were done, such as an interpreter for Lua in JavaScript, which is run
-on and optimized with a tracing JIT for JavaScript
-\cite{yermolovich_optimization_2009}.
-
-These projects have in common that they work one meta-level down, providing a tracing JIT for the
-language used to implement the dynamic language, and not for the dynamic language itself.
-The tracing JIT will then trace through the object model of the dynamic
-language implementation. This makes the object model transparent to the tracer
-and its optimizations. Therefore the semantics of the dynamic language does not
-have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
-Another commonality of these approaches is that they allow some annotations (or
-hints) in the dynamic language implementation to guide the meta-tracer. This
-makes the process not completely automatic but can give good speedups over
-bare meta-tracing.
-
-In this paper we present two of these hints that are extensively used in the
-PyPy project to improve the performance of its Python interpreter, particularly
-of the object model.
-
-Conceptually, the significant speed-ups that can be achieved with
-dynamic compilation depend on feeding into compilation values observed at
-runtime and exploiting them. In particular, if
-there are values which vary very slowly, it is possible to compile multiple
-specialized versions of the same code, one for each actual value.  To exploit
-the runtime feedback, the implementation code and data structures need to be
-structured so that many such slow-varying values are at hand. The hints that
-we present precisely allow us to implement such feedback and exploitation in a
-meta-tracing context.
-
-Together these hints can be used to express many classic implementation
-techniques used for object models of dynamic languages, such runtime type
-feedback and maps.
-
 The contributions of this paper are:
 \begin{itemize}
- \item A hint to turn arbitrary variables into constants in the trace by
- feeding back runtime information into compilation.
- \item A way to annotate operations which the constant folding
- optimization then recognizes and exploits.
- \item A worked-out example of a simple object model of a dynamic language and
- how it can be improved using these hints.
- \item This example also exemplifies general techniques for refactoring code to
- expose constant folding opportunities of likely runtime constants.
+ \item
 \end{itemize}
 
-The paper is structured as follows: Section~\ref{sec:Background} gives an
-introduction to the PyPy project and meta-tracing and presents an example of a
-tiny dynamic language object model. Section~\ref{sec:hints} presents the hints,
-what they do and how they are applied. Section~\ref{sec:fastobjmodel} shows how
-the hints are applied to the tiny object model and Section~\ref{sec:evaluation}
-presents benchmarks.
+The paper is structured as follows:
 
@@
 
-\subsection{The PyPy Project}
+\subsection{RPython and the PyPy Project}
 \label{sub:pypy}
 
-The PyPy project \cite{armin_rigo_pypys_2006} strives to be an environment where
-complex dynamic languages can be implemented efficiently. The approach taken
-when implementing a language with PyPy is to write an interpreter for the language
-in \emph{RPython}. RPython is a restricted subset of Python chosen in such a way
-that it is possible to perform type inference on it. The interpreters in RPython
-can therefore be translated to efficient C code.
+The RPython language and the PyPy Project were started in 2002 with the goal of
+creating a python interpreter written in a High level language, allowing easy
+language experimentation and extension. PyPy is now a fully compatible
+alternative implementation of the Python language, xxx mention speed. The
+Implementation takes advantage of the language features provided by RPython
+such as the provided tracing just-in-time compiler described below.
 
-A number of languages have been implemented with PyPy, most importantly a full
-Python implementation, but also a Prolog interpreter
-\cite{carl_friedrich_bolz_towards_2010} and some less mature experiments.
+RPython, the language and the toolset originally developed to implement the
+Python interpreter have developed into a general environment for experimenting
+and developing fast and maintainable dynamic language implementations. xxx Mention
+the different language impls.
 
-The translation of the interpreter to C code adds a number of implementation details into the
-final executable that are not present in the interpreter implementation, such as
-a garbage collector. The interpreter can therefore be kept free from low-level
-implementation details. Another aspect of the final VM that is added
-semi-automatically to the generated VM is a tracing JIT compiler.
+RPython is built of two components, the language and the translation toolchain
+used to transform RPython programs to executable units.  The RPython language
+is a statically typed object oriented high level language. The language provides
+several features such as automatic memory management (aka. Garbage Collection)
+and just-in-time compilation. When writing an interpreter using RPython the
+programmer only has to write the interpreter for the language she is
+implementing.  The second RPython component, the translation toolchain, is used
+to transform the program to a low level representations suited to be compiled
+and run on one of the different supported target platforms/architectures such
+as C, .NET and Java. During the transformation process
+different low level aspects suited for the target environment are automatically
+added to program such as (if needed) a garbage collector and with some hints
+provided by the author a just-in-time compiler.
 
-The advantage of this approach is that writing an interpreter is much easier
-and less error prone than manually writing a JIT compiler.  Similarly, writing
-in a high level language such as RPython is easier than writing in C.
-
-We call the code that runs on top of a VM implemented with PyPy the
-\emph{user code} or \emph{user program}.
-
-%___________________________________________________________________________
 \subsection{PyPy's Meta-Tracing JIT Compilers}
@@
 
-A recently popular approach to JIT compilers is that of tracing JITs. Tracing
-JITs were popularized by the Dynamo project, which used the technique for dynamic
-machine code optimization \cite{bala_dynamo:_2000}. Later they were used to implement
-a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
-JavaScript \cite{gal_trace-based_2009}.
-
-A tracing JIT works by recording traces of concrete execution paths through the
-program. Those
-traces are linear lists of operations, which are optimized and then
-get turned into machine code. This recording automatically inlines functions:
-when a function call is encountered the operations of the called functions are
-simply put into the trace of the caller too. The tracing JIT tries to produce traces
-that correspond to loops in the traced program, but most tracing JITs now also
-have support for tracing non-loops \cite{andreas_gal_incremental_2006}.
-
-Because the traces always correspond to a concrete execution they cannot
-contain any control flow splits. Therefore they encode the control flow
-decisions needed to stay on the trace with the help of \emph{guards}. Those are
-operations that check that the assumptions are still true when the compiled trace is later executed with different values.
-
-To be able to do this recording, VMs with a
-tracing JIT typically contain an interpreter. After a user program is
-started the interpreter is used; only the most frequently executed paths through the user
-program are traced and turned into machine code. The interpreter is also used when a guard
-fails to continue the execution from the failing guard.
-
-Since PyPy wants to be a general framework, we want to reuse our tracer for
-different languages, which makes classical tracers inapplicable, because they
-encode language semantics.
-Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
-trace the execution of the user program, but instead traces the execution of
-the \emph{interpreter} that is running the program. This means that the traces
-it produces do not contain the bytecodes of the language in question, but
-RPython-level operations that the interpreter did to execute the program.
-
-Tracing through the execution of an interpreter has many advantages. It makes
-the tracer, its optimizers and backends reusable for a variety of languages. The
-language semantics do not need to be encoded into the JIT. Instead the tracer
-just picks them up from the interpreter. This also means that the JIT by
-construction supports the full language as correctly as the interpreter.
-
-While the operations in a trace are those of the interpreter, the loops that are
-traced by the tracer are the loops in the
-user program. To achieve this the tracer stops tracing after one iteration of
-the loop in the user function that is being considered; at this point, it probably
-traced many iterations of the interpreter main loop.
-
-\begin{figure}
-\includegraphics[scale=0.45]{figures/trace-levels}
-\caption{The levels involved in tracing}
-\label{fig:trace-levels}
-\end{figure}
-
-Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left are
-the levels of execution. The CPU executes the binary of
-PyPy's Python interpreter, which consists of RPython functions that have been
-compiled first to C, then to machine code. The interpreter runs a Python program written by a
-programmer (the user). If the tracer is used, it traces operations on the level
-of the interpreter. However, the extent of the trace is determined by the loops
-in the user program.
-
-\subsection{Optimizing Traces}
-\label{sub:optimizing}
-
-Before sending the trace to the backend to produce actual machine code, it is
-optimized.  The optimizer applies a number of techniques to remove or simplify
-the operations in the trace. Most of these are well known compiler optimization
-techniques, with the difference that it is easier to apply them in a tracing
-JIT because it only has to deal with linear traces.  Among the techniques are
-constant folding, common subexpression elimination, allocation removal
-\cite{bolz_allocation_2011}, store/load propagation, loop invariant code
-motion.
-
-In some places it turns out that if the interpreter author rewrites some parts
-of the interpreter with these optimizations in mind the traces that are produced
-by the optimizer can be vastly improved.
-
-\subsection{Running Example}
-\label{sub:running}
-\anto{this example is not referenced until section \ref{sec:fastobjmodel}: I
-  would put it just before it}
-
-As the running example of this paper we will use a very simple and bare-bones
-object model that just supports classes and instances, without any
-inheritance or other advanced features. In the model classes contain methods.
-Instances have a class. Instances have their own attributes (or fields). When looking up an
-attribute of an instance, the instance's attributes are searched. If the
-attribute is not found there, the class' methods are searched.
-
-\begin{figure}
-\input{code/interpreter-slow.tex}
-\caption{Original Version of a Simple Object Model}
-\label{fig:interpreter-slow}
-\end{figure}
-
-To implement this object model, we use the RPython code in
-Figure~\ref{fig:interpreter-slow} as part of the interpreter source code.
-In this straightforward implementation the methods and attributes are just
-stored in dictionaries (hash maps) on the classes and instances, respectively.
-While this object model is very
-simple it already contains most hard parts of Python's object model. Both
-instances and classes can have arbitrary fields, and they are changeable at
-any time.  Moreover, instances can change their class after they have been
-created.
-
-When using this object model in
-an interpreter, a large amount of time will be spent doing lookups in these
-dictionaries.
-Let us assume we trace through code that sums three attributes, such as:
-\anto{I still think it's a bit weird to call them ``methods'' and then use
-  them as attributes in the example}
-
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
-\end{lstlisting}
-}
-
-\begin{figure}
-\input{code/trace1.tex}
-\caption{Trace Through the Object Model}
-\label{fig:trace1}
-\end{figure}
-
-\cfbolz{should we show the code that would create the inst use in tracing?}
-
-The trace would look like in Figure~\ref{fig:trace1}. In this example, the
-attribute \texttt{a} is found on the instance, but the
-attributes \texttt{b} and \texttt{c} are found on the class. The line
-numbers in the trace correspond to the line numbers in
-Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
-trace is in SSA form. Note how all the guards in trace correspond to a
-condition in the original code. The trace contains
-five calls to \texttt{dict.get}, which is slow. To make the language efficient
-using a tracing JIT, we need to find a way to get rid of these dictionary
-lookups. How to achieve this will be the topic of
-Section~\ref{sec:fastobjmodel}.
-
-% subsection Running Example (end)
-
-% section Background (end)
-%___________________________________________________________________________
-
-\section{Hints for Controlling Optimization}
-\label{sec:hints}
-
-In this section we will describe two hints that allow the
-interpreter author to increase the optimization opportunities for constant
-folding.
-If applied correctly these techniques can give really big speedups by
-pre-computing parts of what happens at runtime. On the other
-hand, if applied incorrectly they might lead to code bloat, thus making the
-resulting program actually slower. Note that these hints are \emph{never} put into the user program, only into the interpreter.
-
-For constant folding to work, two conditions need to be met: the arguments of
-an operation actually need to all be constant, i.e. statically known by the
-optimizer and the operation needs to be \emph{constant-foldable}, i.e. always
-yield the same result given the same arguments.
-There is one kind of hint for both of these conditions.
-
-\subsection{Where Do All the Constants Come From?}
-
-It is worth clarifying what a ``constant'' is in this context.  A variable of
-the trace is said to be constant if its value is statically known by the
-optimizer.
-The simplest example of constants are literal values, such as \texttt{1}.
-However, the optimizer can statically know the value of a variable even if it
-is not a constant in the original source code. For example, consider the
-following fragment of RPython code on the left. If the fragment is traced with
-\texttt{x} being \texttt{4}, the trace on the right is produced:
-
-\begin{minipage}[b]{0.5\linewidth}
-\centering
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-if x == 4:
-    y = y + x
-\end{lstlisting}
-}
-\end{minipage}
-\vline
-\hspace{0.5cm}
-\begin{minipage}[b]{0.5\linewidth}
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($x_1$ == 4)
-$y_2$ = $y_1$ + $x_1$
-\end{lstlisting}
-}
-\end{minipage}
-
-A guard is a runtime check. The above trace will run to completion when $x_1$
-\texttt{== 4}. If the check fails, execution of the trace is stopped and the
-interpreter continues to run. Therefore, the value of $x_1$ is statically known
-to be \texttt{4} after the guard.
-
-There are cases in which it is useful to turn an arbitrary variable
-into a constant value. This process is called \emph{promotion} and it is an old idea
-in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993}
-there). The technique is substantially more powerful in a JIT compiler than in
-the static setting of classic partial evaluation.
-
-Promotion is essentially a tool for trace specialization. There are places in
-the interpreter where it would open a lot of optimization opportunities if a
-variable were constant, even though it could have different values in
-practice. In such a place, promotion can be used. The typical reason to do that
-is if there is a lot of computation depending on the value of one variable.
-
-Let us make this more concrete. If we trace a call to the function (written in
-RPython) on the left, we get the trace on the right:
-
-\begin{minipage}[b]{0.5\linewidth}
-\centering
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-def f1(x, y):
-    z = x * 2 + 1
-    return z + y
-\end{lstlisting}
-}
-\end{minipage}
-\vline
-\hspace{0.5cm}
-\begin{minipage}[b]{0.5\linewidth}
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-$v_1$ = $x_1$ * 2
-$z_1$ = $v_1$ + 1
-$v_2$ = $z_1$ + $y_1$
-return($v_2$)
-\end{lstlisting}
-}
-\end{minipage}
-
-Observe how the first two operations could be constant-folded if the value of
-$x_1$ were known. Let us assume that the value of \texttt{x} in the RPython code can vary, but does so
-rarely, i.e. only takes a few different values at runtime. If this is the
-case, we can add a hint to promote \texttt{x}, like this:
-
-\begin{minipage}[b]{0.5\linewidth}
-\centering
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-def f1(x, y):
-    promote(x)
-    z = x * 2 + 1
-    return z + y
-\end{lstlisting}
-}
-\end{minipage}
-\vline
-\hspace{0.5cm}
-\begin{minipage}[b]{0.5\linewidth}
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($x_1$ == 4)
-$v_1$ = $x_1$ * 2
-$z_1$ = $v_1$ + 1
-$v_2$ = $z_1$ + $y_1$
-return($v_2$)
-\end{lstlisting}
-}
-\end{minipage}
-
-The hint indicates that \texttt{x} is likely a runtime constant and the JIT
-should try to perform runtime specialization on it
-in the code that follows. When just running
-the code, the \texttt{promote} function has no
-effect. When tracing, some extra work
-is done. Let us assume that this changed function is traced with
-the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
-operation at the beginning.
-
-The promotion is turned into a \texttt{guard} operation in the trace. The guard
-captures the runtime value of \texttt{x} as it was during tracing, which can
-then be exploited by the compiler. The introduced
-guard specializes the trace, because it only works if the value of $x_1$ is
-\texttt{4}. From the point of view of the
-optimizer, this guard is not different frome the one produced by the \texttt{if}
-statement in the first example. After the guard, it can be assumed that $x_1$
-is equal to \texttt{4}, meaning that the optimizer will turn this trace into:
-
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($x_1$ == 4)
-$v_2$ = 9 + $y_1$
-return($v_2$)
-\end{lstlisting}
-}
-
-Notice how the first two arithmetic operations were constant folded. The hope is
-that the guard is executed quicker than the multiplication and the addition that
-was now optimized away.
-
-If this trace is executed with values of $x_1$ other than \texttt{4}, the guard will
-fail, and execution will continue in the interpreter. If the guard fails often
-enough, a new trace will be started from the guard. This other trace will
-capture a different value of $x_1$. If it is e.g. \texttt{2}, then the optimized
-trace looks like this:
-
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($x_1$ == 2)
-$v_2$ = 5 + $y_1$
-return($v_2$)
-\end{lstlisting}
-}
-
-This new trace will be attached to the guard instruction of the first trace. If
-$x_1$ takes on even more values, a new trace will eventually be made for all of them,
-linking them into a chain. This is clearly not desirable, so we should promote
-only variables that do not vary much. However, adding a promotion hint will never produce wrong
-results. It might just lead to too much machine code being generated.
-
-Promoting integers, as in the examples above, is not used that often.
-However, the internals of dynamic language interpreters often
-have values that are variable but vary little in the context of parts of a user
-program. An example would be the types of variables in a user function, which
-rarely change in a dynamic language in practice (even though they could). In the
-interpreter, these user-level types are values. Thus promoting them will lead
-to type-specialization on the level of the user program.
-Section~\ref{sec:fastobjmodel} will present a complete example of how this
-works.
-
-\subsection{Declaring New Foldable Operations}
-
-In the previous section we saw a way to turn arbitrary variables into constants. All
-foldable operations on these constants can be constant-folded. This works well for
-constant folding of primitive types, e.g. integers. Unfortunately, in the context of an
-interpreter for a dynamic
-language, most operations actually manipulate objects, not primitive types. The
-operations on objects are often not foldable and might even have side-effects. If
-one reads a field out of a constant reference to an object this cannot
-necessarily be folded away because the object can be mutated. Therefore, another
-hint is needed.
-
-This hint can be used to mark functions as \emph{trace-elidable}. A function is
-termed trace-elidable if, during the execution of the program,
-successive calls to the function with identical arguments always return the
-same result. In addition the function needs to have no side effects or
-idempotent side effects\footnote{This property
-is less strict than that of a pure function, because it is only about actual
-calls during execution. All pure functions are trace-elidable though.}.
-From this definition follows that a call to a trace-elidable function with
-constant arguments in a trace can be replaced with the result of the call seen during tracing.
-
-As an example, take the class on the left. Tracing the call \texttt{a.f(10)} of
-some instance of \texttt{A} yields the trace on the right (note how the call to
-\texttt{c} is inlined):
-
-\begin{minipage}[t]{0.6\linewidth}
-\centering
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-class A(object):
-    def __init__(self, x, y):
-        self.x = x
-        self.y = y
-
-    def f(self, val):
-        self.y = self.c() + val
-
-    def c(self):
-        return self.x * 2 + 1
-\end{lstlisting}
-\end{minipage}
-\vline
-\hspace{0.5cm}
-\begin{minipage}[t]{0.4\linewidth}
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-$x_1$ = $a_1$.x
-$v_1$ = $x_1$ * 2
-$v_2$ = $v_1$ + 1
-$v_3$ = $v_2$ + $val_1$
-$a_1$.y = $v_3$
-\end{lstlisting}
-\end{minipage}
-
-In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
-computation of the first few operations does not help. Even if $a_1$ is a
-constant reference to an object, reading the \texttt{x} field does not necessarily
-always yield the same value. To solve this problem, there is another annotation,
-which lets the interpreter author communicate invariants to the optimizer. In
-this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
-immutable, and therefore \texttt{c}
-is a trace-elidable function. To communicate this, there is an \texttt{@elidable} decorator.
-If the code in \texttt{c} should be constant-folded away, we would change the
-class as follows:
-
-\begin{minipage}[t]{0.6\linewidth}
-\centering
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-class A(object):
-    def __init__(self, x, y):
-        self.x = x
-        self.y = y
-
-    def f(self, val):
-        promote(self)
-        self.y = self.c() + val
-
-    @elidable
-    def c(self):
-        return self.x * 2 + 1
-\end{lstlisting}
-\end{minipage}
-\vline
-\hspace{0.4cm}
-\begin{minipage}[t]{0.4\linewidth}
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($a_1$ == 
-      0xb73984a8)
-$v_1$ = A.c($a_1$)
-$v_2$ = $v_1$ + $val_1$
-$a_1$.y = $v_2$
-\end{lstlisting}
-\end{minipage}
-
-Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
-during tracing. The call to \texttt{A.c} is not inlined, so that the optimizer
-has a chance to see it. Since the \texttt{A.c} method is marked as trace-elidable, and its
-argument
-is a constant reference, the call will be removed by the optimizer. The final
-trace looks like this (assuming that the \texttt{x} field's value is \texttt{4}):
-%
-{\noop
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-guard($a_1$ == 0xb73984a8)
-$v_2$ = 9 + $val_1$
-$a_1$.y = $v_2$
-\end{lstlisting}
-}
-
-On the one hand, the \texttt{@elidable} annotation is very powerful. It can be
-used to constant-fold arbitrary parts of the computation in the interpreter.
-However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
-function is annotated to be trace-elidable, but is not really, the optimizer can produce
-subtly wrong code. Therefore, a lot of care has to be taken when using this
-annotation\footnote{The most common use case of the \texttt{@elidable}
-annotation is indeed to declare the immutability of fields. Because it is so
-common, we have special syntactic sugar for it.}. We hope to introduce a
-debugging mode which would (slowly) check whether the annotation is applied
-incorrectly to mitigate this problem.
+ * Tracing JITs
+ * JIT Compiler
+   * describe the tracing jit stuff in pypy
+   * reference tracing the meta level paper for a high level description of what the JIT does
+   * JIT Architecture
+   * Explain the aspects of tracing and optimization
 
@@
 
-\section{Putting It All Together}
-\label{sec:fastobjmodel}
+\section{Resume Data}
+\label{sec:Resume Data}
 
-In this section we describe how the simple object model from
-Section~\ref{sub:running} can be made efficient using the hints described in the
-previous section. The object model there is typical for many current
-dynamic languages (such as Python, Ruby and JavaScript) as it relies heavily on
-hash-maps to implement its objects.
+* High level handling of resumedata
+   * trade-off fast tracing v/s memory usage
+   * creation in the frontendÂ
+   * optimization
+   * compression
+   * interaction with optimization
+   * tracing and attaching bridges and throwing away resume data
+   * compiling bridges
 
-%___________________________________________________________________________
+% section Resume Data (end)
 
-\subsection{Making Instance Attributes Faster Using Maps}
+\section{Guards in the Backend}
+\label{sec:Guards in the Backend}
 
-The first step in making \texttt{getattr} faster in our object model is to optimize
-away the dictionary lookups on the instances. The hints of the previous section
-do not seem to help with the current object model. There is
-no trace-elidable function to be seen, and the instance is not a candidate for promotion,
-because there tend to be many instances.
+* Low level handling of guards
+   * Fast guard checks v/s memory usage
+   * memory efficient encoding of low level resume data
+   * fast checks for guard conditions
+   * slow bail out
 
-This is a common problem when trying to apply hints. Often, the interpreter
-needs a small rewrite to expose the trace-elidable functions and nearly-constant objects
-that are implicitly there. In the case of instance fields this rewrite is not
-entirely obvious. The basic idea is as follows. In theory instances can have
-arbitrary fields. In practice however many instances share their layout (i.e.
-their set of keys) with many other instances.
-
-Therefore it makes sense to factor the layout information out of the instance
-implementation into a shared object, called the \emph{map}. Maps are a well-known
-technique to efficiently implement instances and come from the SELF project
-\cite{chambers_efficient_1989}. They are also used by many JavaScript implementations such as V8.
-The rewritten \texttt{Instance} class using maps can be seen in
-Figure~\ref{fig:maps}.
-
-\begin{figure}
-\input{code/map.tex}
-\caption{Simple Object Model With Maps}
-\label{fig:maps}
-\end{figure}
-
-In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
-reference to a map, which maps field names to indexes into a storage list. The
-storage list contains the actual field values. Maps are shared between
-different instances, therefore they have to be immutable, which means
-that their \texttt{getindex} method is a trace-elidable function. When a new attribute is added
-to an instance, a new map needs to be chosen, which is done with the
-\texttt{add\_attribute} method on the previous map. This function is also trace-elidable,
-because it caches all new instances of \texttt{Map} that it creates, to make
-sure that objects with the same layout have the same map, which makes its side
-effects idempotent. Now that we have
-introduced maps, it is safe to promote the map everywhere, because we assume
-that the number of different instance layouts is small.
-
-With this adapted instance implementation, the trace we saw in Section~\ref{sub:running} changes to
-that of Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
-memory address of the \texttt{Map} instance that has been promoted. Operations
-that can be optimized away are grayed out, their results will be replaced with
-fixed values by the constant folding.
-
-The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
-a trace-elidable function and they have constant arguments. That means that $index_{1/2/3}$
-are constant and the guards on them can be removed. All but the first guard on
-the map will be optimized away too, because the map cannot have changed in
-between. This trace is already much better than
-the original one. Now we are down from five dictionary lookups to just two.
-
-The technique to make instance lookups faster is applicable in more general
-cases. A more abstract view of maps is that of splitting a data-structure into
-an immutable part (\eg the map) and a part that changes (\eg the storage list).
-All the computation on the immutable part is trace-elidable so that only the
-manipulation of the quick-changing part remains in the trace after
-optimization.
-
-\begin{figure}
-\input{code/trace2.tex}
-\caption{Unoptimized Trace After the Introduction of Maps}
-\label{fig:trace2}
-\end{figure}
-
-%___________________________________________________________________________
-
-\subsection{Versioning of Classes}
-
-Above we assumed that the total number of
-different instance layouts is small compared to the number of instances. For classes we
-will make an even stronger assumption. We simply assume that it is rare for
-classes to change at all. This is not always reasonable (sometimes classes contain
-counters or similar things) but for this simple example it is good
-enough.\footnote{There is a more complex variant of the presented technique that can
-accommodate quick-changing class fields a lot better.}
-
-What we would really like that the \texttt{Class.find\_method} method is trace-elidable.
-But it cannot be, because it is always possible to change the class itself.
-Every time the class changes, \texttt{find\_method} can potentially return a
-new value.
-
-Therefore, we give every class a version object, which is changed every time a
-class gets changed (i.e., the \texttt{methods} dictionary changes).
-This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
-version)} pair will always be the same, i.e. it is a trace-elidable operation.  To help
-the JIT to detect this case, we factor it out in a helper method \texttt{\_find\_method} which is
-marked as \texttt{@elidable}. The refactored \texttt{Class} can
-be seen in Figure~\ref{fig:version}
-
-\begin{figure}
-\input{code/version.tex}
-\caption{Versioning of Classes}
-\label{fig:version}
-\end{figure}
-
-What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
-argument but it does not use it at all. Its only purpose is to make the call
-trace-elidable, because when the version object changes, the result of the call might be
-different from the previous one.
-
-\begin{figure}
-\input{code/trace4.tex}
-\caption{Unoptimized Trace After Introduction of Versioned Classes}
-\label{fig:trace4}
-\end{figure}
-
-The trace with this new class implementation can be seen in
-Figure~\ref{fig:trace4}.
-The calls to \texttt{Class.\_find\_method} can now be optimized away, also the
-promotion of the class and the version, except for the first one. The final
-optimized trace can be seen in Figure~\ref{fig:trace5}.
-
-\begin{figure}
-\input{code/trace5.tex}
-\caption{Optimized Trace After Introduction of Versioned Classes}
-\label{fig:trace5}
-\end{figure}
-
-The index \texttt{0} that is used to read out of the \texttt{storage} list is the result
-of the constant-folded \texttt{getindex} call.
-The constants \texttt{41} and \texttt{17} are the results of the folding of the
-\texttt{\_find\_method} calls. This final trace is now very good. It no longer performs any
-dictionary lookups. Instead it contains several guards. The first guard
-checks that the map is still the same. This guard will fail if the same
-code is executed with an instance that has another layout. The second guard
-checks that the class of \texttt{inst} is still the same. It will fail if the trace is
-executed with an instance of another class. The third guard checks that the
-class did not change since the trace was produced. It will fail if somebody
-calls the \texttt{write\_method} method on the class.
-
-%___________________________________________________________________________
-
-\subsection{Real-World Considerations}
-
-The techniques used above for the simple object model are used for the object
-model of PyPy's Python interpreter too. Since Python's object model is
-considerably more complex, some additional work needs to be done.
-
-The first problem that needs to be solved is that Python supports (multiple)
-inheritance. Therefore looking up a method in a class needs to consider all the
-classes in the
-whole method resolution order. This makes the versioning of classes more
-complex. If a class is changed its version changes. At the same time, the
-versions of all the classes inheriting from it need to be changed as well,
-recursively. This makes class changes expensive, but they should be rare.  On the
-other hand, a method lookup in a complex class hierarchy is as optimized in the
-trace as in our simple object model above.
-
-Another optimization is that in practice the shape of an instance is correlated
-with its class. In our code above, we allow both to vary independently.
-In PyPy's Python interpreter we store the class of an instance on its map. This
-means that we get one fewer promotion and thus one fewer
-guard in the trace, because the class doesn't need to be promoted after the
-map has been.
-
-%___________________________________________________________________________
-
-%\subsection{More General Patterns}
-%
-%
-%Similarly, versions can be used to constant-fold arbitrary functions of large data
-%structures. The version needs to be updated carefully every time the result of
-%this function can change. Therefore this is useful only if the data structure is
-%expected to change slowly.
+% section Guards in the Backend (end)
 
@@
 
-For space reasons we cannot perform a full evaluation here, but still want to
-present some benchmark numbers. We chose to present just some benchmarks: The
-templating engine of the Django web
-framework\footnote{\texttt{http://www.djangoproject.com/}}; a Monte-Carlo Go
-AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/
-disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
-Richards benchmark to Python; a Python version of the Telco decimal
-benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
-pure Python decimal floating point implementation. The results we see in these
-benchmarks seem to repeat themselves in other benchmarks using object-oriented
-code; for purely numerical algorithms the speedups introduced by the techniques
-in this paper are much smaller because they are already fast.
-
-The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
-with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
-2.6.35. We compared the performance of two Python implementations on the
-benchmarks. As a baseline, we used the standard Python implementation in C,
-CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
-interpreter. We compare it against two versions of PyPy's Python interpreter,
-both of them with JIT enabled. The PyPy baseline does not enable maps or type
-version, the full JIT enables both.
-
-All benchmarks were run 50 times in the same process, to give the JIT time to
-produce machine code. The arithmetic mean of the times of the last 30 runs were
-used as the result. The errors were computed using a confidence interval with a
-95\% confidence level \cite{georges_statistically_2007}. The results are
-reported in Figure~\ref{fig:times}, together with the same numbers normalized to
-those of the full JIT.
-
-The optimizations give a speedup between 80\% and almost 20 times. The Richards benchmark
-is a particularly good case for the optimizations as it makes heavy uses of
-object-oriented features. Pyflate uses mostly imperative code, so does not
-benefit as much. Together with the optimization, PyPy outperforms CPython in
-all benchmarks, which is not surprising because CPython is a simple
-bytecode-based interpreter.
-
-\begin{figure}
-\begin{center}
-{\smaller
-\begin{tabular}{|l|r|r|r|}
-\hline
-	&CPython	&JIT baseline	&JIT full\\
-\hline
-django[ms]	&988.67 $\pm$ 0.49	&405.62 $\pm$ 4.80	&149.31 $\pm$ 1.37\\
-	&6.62 $\times$	&2.72 $\times$	&1.00 $\times$\\
-\hline
-go[ms]	&947.43 $\pm$ 1.30	&525.53 $\pm$ 7.67	&174.32 $\pm$ 7.78\\
-	&5.44 $\times$	&3.01 $\times$	&1.00 $\times$\\
-\hline
-pyflate[ms]	&3209.20 $\pm$ 3.65	&2884.26 $\pm$ 21.11	&1585.48 $\pm$ 5.22\\
-	&2.02 $\times$	&1.82 $\times$	&1.00 $\times$\\
-\hline
-richards[ms]	&357.79 $\pm$ 1.32	&421.87 $\pm$ 0.48	&17.89 $\pm$ 1.15\\
-	&20.00 $\times$	&23.58 $\times$	&1.00 $\times$\\
-\hline
-telco[ms]	&1209.67 $\pm$ 2.20	&738.18 $\pm$ 3.29	&153.48 $\pm$ 1.86\\
-	&7.88 $\times$	&4.81 $\times$	&1.00 $\times$\\
-\hline
-\end{tabular}
-}
-\end{center}
-\caption{Benchmark Results}
-\label{fig:times}
-\end{figure}
+* Evaluation
+   * Measure guard memory consumption and machine code size
+   * Extrapolate memory consumption for guard other guard encodings
+      * compare to naive variant
+   * Measure how many guards survive optimization
+   * Measure the of guards and how many of these ever fail
 
@@
 
-The very first meta-tracer is described by Sullivan et. al.
-\cite{sullivan_dynamic_2003}. They used Dynamo RIO, the successor of Dynamo
-\cite{bala_dynamo:_2000} to trace through a small synthetic interpreter. As in Dynamo, tracing
-happens on the machine code level. The system needs some hints to mark the main
-interpreter loop and where the backward jumps in user programs are.
-PyPy uses similar hints to achieve this \cite{bolz_tracing_2009}. Their approach suffers
-mostly from the low abstraction level that machine code provides.
-
-Yermolovich et. al. \cite{yermolovich_optimization_2009} describe the use of the Tamarin JavaScript tracing JIT as a
-meta-tracer for a Lua interpreter. They compile the normal Lua interpreter in C
-to ActionScript bytecode. Again, the interpreter is annotated with some hints
-that indicate the main interpreter loop to the tracer.  No further hints are
-described in the paper. There is no comparison of their system to the original
-Lua VM in C, which makes it hard to judge the effectiveness of the approach.
-
-SPUR \cite{bebenita_spur:_2010} is a tracing JIT for CIL bytecode, which is then
-used to trace through a JavaScript implementation written in C\#. The
-JavaScript implementation compiles JavaScript to CIL bytecode together with an
-implementation of the JavaScript object model. The object model uses maps
-and inline caches to speed up operations on objects. The tracer traces through
-the compiled JavaScript functions and the object model. SPUR contains two hints
-that can be used to influence the tracer: one to prevent tracing of a C\#
-function and one to force unrolling of a loop (PyPy has equivalent hints, but
-they were not described in this paper).
-
-Partial evaluation \cite{jones_partial_1993} tries to automatically transform
-interpreters into compilers using the second futamura projection
-\cite{futamura_partial_1999}. Given that classical partial evaluation works
-strictly ahead of time, it inherently cannot support runtime feedback.
-Some partial evaluators work at runtime, such as DyC \cite{grant_dyc:_2000},
-which also supports a concept similar to promotion (called dynamic-to-static promotion).
-
-An early attempt at building a general environment for implementing languages
-efficiently is described by Wolczko et. al. \cite{mario_wolczko_towards_1999}.
-They implement Java and Smalltalk on top of the SELF VM by compiling the
-languages to SELF. The SELF JIT is good enough to optimize the compiled code
-very well. We believe the approach to be restricted to languages that are
-similar enough to SELF as there were no mechanisms to control the underlying
-compiler.
-
-Somewhat relatedly, the proposed ``invokedynamic'' bytecode
-\cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
-implementation of dynamic languages on top of JVMs easier. The bytecode gives
-the user access to generalized inline caches. It requires of course compilation
-to JVM bytecode instead of writing an interpreter.
-
-We already explored promotion in other contexts, such as earlier versions of
-PyPy's JIT.
-%as well as a Prolog partial evaluator \cite{bolz_towards_2009}
-Promotion is also heavily
-used by Psyco \cite{rigo_representation-based_2004} (promotion is called
-"unlifting" in this paper) a method-based JIT compiler for Python written by
-one of the authors. Promotion is quite similar to
-runtime type feedback (and also inline caching) techniques which were
-first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
-\cite{hoelzle_optimizing_1994} implementations.
-Promotion is more general because any information can be fed back into
-compilation, not just types.
-
-%is there anything about versions? smalltalks tend to clear their method caches
-%when new methods are added. self and java use dependency tracking and
-%deoptimization. this is better what we have above, because we need runtime
-%checks. mention out of line guard?
-
-%jruby used versions at some point, jvm-l mailing list discusses them
-
 \section{Conclusion}
 
-In this paper we presented two hints that can be used in the source code of an
-interpreter written with PyPy. They give control over runtime feedback and
-optimization to the language implementor. They are expressive enough for
-building well-known virtual machine optimization techniques, such as maps and
-inline caches. We believe that they are flexible enough to express a wide
-variety of language semantics efficiently.
-
 \section*{Acknowledgements}
-
-The authors would like to thank Peng Wu, David Edelsohn and Laura Creighton for
-encouragement, fruitful discussions and feedback during the writing of this paper.
-This research was partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
-Eureka Eurostars).
 

### 2012-07-03 talk/vmil2012/paper.tex
fix the animation

@@
 
-The RPython language and the PyPy Project were started in 2002 with the goal of
-creating a python interpreter written in a High level language, allowing easy
-language experimentation and extension. PyPy is now a fully compatible
-alternative implementation of the Python language, xxx mention speed. The
-Implementation takes advantage of the language features provided by RPython
-such as the provided tracing just-in-time compiler described below.
-
-RPython, the language and the toolset originally developed to implement the
-Python interpreter have developed into a general environment for experimenting
-and developing fast and maintainable dynamic language implementations. xxx Mention
-the different language impls.
-
-RPython is built of two components, the language and the translation toolchain
-used to transform RPython programs to executable units.  The RPython language
-is a statically typed object oriented high level language. The language provides
-several features such as automatic memory management (aka. Garbage Collection)
-and just-in-time compilation. When writing an interpreter using RPython the
-programmer only has to write the interpreter for the language she is
-implementing.  The second RPython component, the translation toolchain, is used
-to transform the program to a low level representations suited to be compiled
-and run on one of the different supported target platforms/architectures such
-as C, .NET and Java. During the transformation process
-different low level aspects suited for the target environment are automatically
-added to program such as (if needed) a garbage collector and with some hints
-provided by the author a just-in-time compiler.
-
 \subsection{PyPy's Meta-Tracing JIT Compilers}

### 2012-07-09 talk/ecoop2009/rainbow.tex
fix few XXXs
update year

@@
 From a partial evaluation point of view, promotion is the converse of
-the operation generally known as \emph{lift} \cite{XXX}.  Lifting a value means
+the operation generally known as \emph{lift}.  Lifting a value means
 copying a variable whose binding time is compile-time into a variable
@@
 If later an instance of \lstinline{IntObj} hits the flexswitch again, the
-code is executed without needing of more calls to the JIT compiler.  On the
+code is executed without needing more calls to the JIT compiler.  On the
 other hand, if the flexswitch is hit by an instance of some other class, the

### 2012-07-16 talk/vmil2012/paper.tex
update year
show comments and add a command for my comments
set font size to 10pt
write more about the low-level aspects of guards

--- 2012-07-03 talk/vmil2012/paper.tex
@@
-\documentclass{sigplanconf}
+\documentclass[10pt]{sigplanconf}
 
@@
 \newcommand\pedronis[1]{\nb{PEDRONIS}{#1}}
+\newcommand\bivab[1]{\nb{DAVID}{#1}}
 \newcommand{\commentout}[1]{}
@@
 
-\conferenceinfo{VMIL'11}{}
+\conferenceinfo{VMIL'12}{}
 \CopyrightYear{2012}
@@
 
+In this paper we describe and analyze how deoptimization works in the context
+of tracing just-in-time compilers. What instructions are used in the
+intermediate and low-level representation of the JIT instructions and how these
+are implemented.
+
+Although there are several publications about tracing jut-in-time compilers, to
+our knowledge, there are none that describe the use and implementaiton of
+guards in this context. With the following contributions we aim to sched some
+light (to much?) on this topic.
 The contributions of this paper are:
@@
 language experimentation and extension. PyPy is now a fully compatible
-alternative implementation of the Python language, xxx mention speed. The
+alternative implementation of the Python language\bivab{mention speed}. The
 Implementation takes advantage of the language features provided by RPython
@@
 Python interpreter have developed into a general environment for experimenting
-and developing fast and maintainable dynamic language implementations. xxx Mention
-the different language impls.
+and developing fast and maintainable dynamic language implementations.
+\bivab{Mention the different language impls}
 
@@
 
-* Low level handling of guards
-   * Fast guard checks v/s memory usage
-   * memory efficient encoding of low level resume data
-   * fast checks for guard conditions
-   * slow bail out
+Code generation consists of two passes over the lists of instructions, a
+backwards pass to calculate live ranges of IR-level variables and a forward one
+to emit the instructions. During the forward pass IR-level variables are
+assigned to registers and stack locations by the register allocator according to
+the requirements of the to be emitted instructions. Eviction/spilling is
+performed based on the live range information collected in the first pass. Each
+IR instruction is transformed into one or more machine level instructions that
+implement the required semantics.  Guards instructions are transformed into
+fast checks at the machine code level that verify the corresponding condition.
+In cases the value being checked by the guard is not used anywhere else the
+guard and the operation producing the value can merged, reducing even more the
+overhead of the guard.
 
+Each guard in the IR has attached to it a list of the IR-variables required to
+rebuild the execution state in case the trace is left through the side-exit
+corresponding to the guard. When a guard is compiled, additionally to the
+condition check two things are generated/compiled. First a special
+data structure is created that encodes the information provided by the register
+allocator about where the values corresponding to each IR-variable required by
+the guard will be stored when execution reaches the code emitted for the
+corresponding guard. \bivab{go into more detail here?!} This encoding needs to
+be as compact as possible to maintain an acceptable memory profile.
+
+\bivab{example goes here}
+
+Second a trampoline method stub is generated. Guards are usually implemented as
+a conditional jump to this stub, jumping to in case the guard condition check
+does not hold and a side-exit should be taken. This stub loads the pointer to
+the encoding of the locations mentioned above, preserves the execution state
+(stack and registers) and then jumps to generic bail-out handler that is used
+to leave the compiled trace if case of a guard failure.
+
+Using the encoded location information the bail-out handler reads from the
+saved execution state the values that the IR-variables had  at the time of the
+guard failure and stores them in a location that can be read by the fronted.
+After saving the information the control is passed to the frontend signaling
+which guard failed so the frontend can read the information passed and restore
+the state corresponding to the point in the program.
+
+As in previous sections the underlying idea for the design of guards is to have
+a fast on trace profile and a potentially slow one in the bail-out case where
+the execution takes one of the side exits due to a guard failure. At the same
+time the data stored in the backend needed to rebuild the state should be be
+as compact as possible to reduce the memory overhead produced by the large
+number of guards\bivab{back this}.
+
+%* Low level handling of guards
+%   * Fast guard checks v/s memory usage
+%   * memory efficient encoding of low level resume data
+%   * fast checks for guard conditions
+%   * slow bail out
+%
 % section Guards in the Backend (end)

### 2012-07-17 talk/vmil2012/paper.tex
a note

@@
    * trade-off fast tracing v/s memory usage
-   * creation in the frontendÂ
+   * creation in the frontend
    * optimization
@@
 Code generation consists of two passes over the lists of instructions, a
-backwards pass to calculate live ranges of IR-level variables and a forward one
+backwards pass to calculate live ranges of IR-level variables \cfbolz{doesn't the backward pass also remove dead instructions?} and a forward one
 to emit the instructions. During the forward pass IR-level variables are

### 2012-07-17 talk/vmil2012/paper.tex
expand notes somewhat

@@
 The RPython language and the PyPy Project were started in 2002 with the goal of
-creating a python interpreter written in a High level language, allowing easy
+creating a Python interpreter written in a high level language, allowing easy
 language experimentation and extension. PyPy is now a fully compatible
@@
 * High level handling of resumedata
-   * trade-off fast tracing v/s memory usage
-   * creation in the frontend
-   * optimization
-   * compression
-   * interaction with optimization
+
+- traces follow the execution path during tracing, other path not compiled at first
+- points of possible divergence from that path are guards
+- since path can later diverge, at the guards it must be possible to re-build interpreter state in the form of interpreter stack frames
+- tracing does inlining, therefore a guard must contain information to build a whole stack of frames
+- optimization rewrites traces, including removal of guards
+
+- frames consist of a PC and local variables
+- rebuild frame by taking local SSA variables in the trace and mapping them to variables in the frame
+
+two forces:
+- there are lots of guards, therefore the information must be stored in a compact way in the end
+- tracing must be fast
+
+compression approaches:
+- use fact that outer frames don't change in the part of the trace that is in the inner frame
+- compact bit-representation for constants/ssa vars
+
+interaction with optimization
+- guard coalescing
+- virtuals
+
    * tracing and attaching bridges and throwing away resume data

### 2012-07-18 talk/vmil2012/paper.tex
another note
clarify this point

@@
 - virtuals
+  - most virtuals not changed between guards
 
@@
 Code generation consists of two passes over the lists of instructions, a
-backwards pass to calculate live ranges of IR-level variables \cfbolz{doesn't the backward pass also remove dead instructions?} and a forward one
+backwards pass to calculate live ranges of IR-level variables and a forward one
 to emit the instructions. During the forward pass IR-level variables are
-assigned to registers and stack locations by the register allocator according to
-the requirements of the to be emitted instructions. Eviction/spilling is
+assigned to registers and stack locations by the register allocator according
+to the requirements of the to be emitted instructions. Eviction/spilling is
 performed based on the live range information collected in the first pass. Each
 IR instruction is transformed into one or more machine level instructions that
-implement the required semantics.  Guards instructions are transformed into
-fast checks at the machine code level that verify the corresponding condition.
-In cases the value being checked by the guard is not used anywhere else the
-guard and the operation producing the value can merged, reducing even more the
+implement the required semantics, operations withouth side effects whose result
+is not used are not emitted. Guards instructions are transformed into fast
+checks at the machine code level that verify the corresponding condition.  In
+cases the value being checked by the guard is not used anywhere else the guard
+and the operation producing the value can merged, reducing even more the
 overhead of the guard.

### 2012-07-18 talk/vmil2012/paper.tex
start writing about guards <-> bridges in the backend

@@
    * compiling bridges
-
+\bivab{mention that the failargs also go into the bridge}
 % section Resume Data (end)
@@
 and the operation producing the value can merged, reducing even more the
-overhead of the guard.
+overhead of the guard. \bivab{example for this}
 
@@
 
+As explained in previous sections, when a specific guard has failed often enogh
+a new trace, refered to as bridge, starting from this guard is recorded and
+compile. The goal for the execution of bridges, that become a part of the
+common path is to favor the performance while staying on the compiled trace.
+This means we want wo avoid switching back and forth to the frontend when a
+guard corresponding to the bridge fails before we can execute the brigde.
+Instead we want to have as little as possible overhead when switching from the
+loop path to the bridge.
+
+The process of compiling a bridge is very similar to compiling a loop,
+instructions and guards are processed in the same way as described above. The
+main difference is the setup phase, when compiling a trace we start with a
+clean slate, whilst the compilation of a bridge starts with state as it was
+when compiling the guard. To restore the state needed compile the bridge we use
+the encoded representation created for the guard to rebuild the bindings from
+IR-variables to stack locations and registers used in the register allocator.
+
+Once the bridge has been compiled the trampoline method stub is redirected to
+the code of the bridge. In future if the guard fails again it jumps to the
+trampoline and then jumps to the code compiled for the bridge, having alsmos no
+overhead compared to the execution of the original trace, behaving mainly as
+two paths in a conditional block.
+
 %* Low level handling of guards

### 2012-07-19 talk/vmil2012/paper.tex
- write about high-level handling of resume data - add a start of a bibliography

--- 2012-07-18 talk/vmil2012/paper.tex
@@
 
+Since tracing linearizes control flow by following one concrete execution,
+not the full control flow of a program is observed.
+The possible points of deviation from the trace are guard operations
+that check whether the same assumptions as during tracing still hold.
+In later executions of the trace the guards can fail.
+If that happens, execution needs to continue in the interpreter.
+This means it is necessary to attach enough information to a guard
+to construct the interpreter state when that guard fails.
+This information is called the \emph{resume data}.
+
+To do this reconstruction, it is necessary to take the values
+of the SSA variables of the trace
+and build interpreter stack frames.
+Tracing aggressively inlines functions.
+Therefore the reconstructed state of the interpreter
+can consist of several interpreter frames.
+
+If a guard fails often enough, a trace is started from it
+to create a trace tree.
+When that happens another use case of resume data
+is to construct the tracer state.
+
+There are several forces guiding the design of resume data handling.
+Guards are a very common operations in the traces.
+However, a large percentage of all operations
+are optimized away before code generation.
+Since there are a lot of guards
+the resume data needs to be stored in a very compact way.
+On the other hand, tracing should be as fast as possible,
+so the construction of resume data must not take too much time.
+
+\subsection{Capturing of Resume Data During Tracing}
+\label{sub:capturing}
+
+Every time a guard is recorded during tracing
+the tracer attaches preliminary resume data to it.
+The data is preliminary in that it is not particularly compact yet.
+The preliminary resume data takes the form of a stack of symbolic frames.
+The stack contains only those interpreter frames seen by the tracer.
+The frames are symbolic in that the local variables in the frames
+do not contain values.
+Instead, every local variables contains the SSA variable of the trace
+where the value would later come from, or a constant.
+
+\subsection{Compression of Resume Data}
+\label{sub:compression}
+
+The core idea of storing resume data as compactly as possible
+is to share parts of the data structure between subsequent guards.
+This is often useful because the density of guards in traces is so high,
+that quite often not much changes between them.
+Since resume data is a linked list of symbolic frames
+often only the information in the top frame changes from one guard to the next.
+The other frames can often be just reused.
+The reason for this is that during tracing only the variables
+of the currently executing frames can change.
+Therefore if two guards are generated from code in the same function
+the resume data of the rest of the stack can be reused.
+
+\subsection{Interaction With Optimization}
+\label{sub:optimization}
+
+Guards interact with optimizations in various ways.
+Most importantly optimizations try to remove as many operations
+and therefore guards as possible.
+This is done with many classical compiler optimizations.
+In particular guards can be remove by subexpression elimination.
+If the same guard is encountered a second time in the trace,
+the second one can be removed.
+This also works if a later guard is weaker implied by a earlier guard.
+
+One of the techniques in the optimizer specific to tracing for removing guards
+is guard strengthening~\cite{bebenita_spur:_2010}.
+The idea of guard strengthening is that if a later guard is stronger
+than an earlier guard it makes sense to move the stronger guard
+to the point of the earlier, weaker guard and to remove the weaker guard.
+Moving a guard to an earlier point is always valid,
+it just means that the guard fails earlier during the trace execution
+(the other direction is clearly not valid).
+
+The other important point of interaction between resume data and the optimizer
+is RPython's allocation removal optimization~\cite{bolz_allocation_2011}.
+This optimization discovers allocations in the trace that create objects
+that do not survive long.
+An example is the instance of \lstinline{Even} in the example\cfbolz{reference figure}.
+Allocation removal makes resume data more complex.
+Since allocations are removed from the trace it becomes necessary
+to reconstruct the objects that were not allocated so far when a guard fails.
+Therefore the resume data needs to store enough information
+to make this reconstruction possible.
+
+Adding this additional information is done as follows.
+So far, every variable in the symbolic frames
+contains a constant or an SSA variable.
+After allocation removal the variables in the symbolic frames can also contain
+``virtual'' objects.
+These are objects that were not allocated so far,
+because the optimizer removed their allocation.
+The virtual objects in the symbolic frames describe exactly
+how the heap objects that have to be allocated on guard failure look like.
+To this end, the content of the every field of the virtual object is described
+in the same way that the local variables of symbolic frames are described.
+The fields of the virtual objects can therefore be SSA variables, constants
+or other virtual objects.
+
+During the storing of resume data virtual objects are also shared
+between subsequent guards as much as possible.
+The same observation as about frames applies:
+Quite often a virtual object does not change from one guard to the next.
+Then the data structure is shared.
+
+% subsection Interaction With Optimization (end)
+
 * High level handling of resumedata
 
-- traces follow the execution path during tracing, other path not compiled at first
-- points of possible divergence from that path are guards
-- since path can later diverge, at the guards it must be possible to re-build interpreter state in the form of interpreter stack frames
-- tracing does inlining, therefore a guard must contain information to build a whole stack of frames
-- optimization rewrites traces, including removal of guards
-
 - frames consist of a PC and local variables
-- rebuild frame by taking local SSA variables in the trace and mapping them to variables in the frame
-
-two forces:
-- there are lots of guards, therefore the information must be stored in a compact way in the end
-- tracing must be fast
 
@@
 
+As explained in previous sections, when a specific guard has failed often enogh
+a new trace, refered to as bridge, starting from this guard is recorded and
+compiled. When compiling bridges the goal is that future failures of the guards
+that led to the compilation of the bridge should execute the bridge without
+additional overhead, in particular it the failure of the guard should not lead
+to leaving the compiled code prior to execution the code of the bridge.
+
+The process of compiling a bridge is very similar to compiling a loop,
+instructions and guards are processed in the same way as described above. The
+main difference is the setup phase, when compiling a trace we start with a lean
+slate. The compilation of a bridge is starts from a state (register and stack
+bindings) that corresponds to the state during the compilation of the original
+guard. To restore the state needed compile the bridge we use the encoded
+representation created for the guard to rebuild the bindings from IR-variables
+to stack locations and registers used in the register allocator.  With this
+reconstruction all bindings are restored to the state as they were in the
+original loop up to the guard.
+
+Once the bridge has been compiled the trampoline method stub is redirected to
+the code of the bridge. In future if the guard fails again it jumps to the code
+compiled for the bridge instead of bailing out. Once the guard has been
+compiled and attached to the loop the guard becomes just a point where
+control-flow can split. The loop after the guard and the bridge are just
+condional paths.
+
 %* Low level handling of guards

### 2012-07-19 talk/vmil2012/paper.tex
explain the compact bit representation

@@
 
+In addition to sharing as much as possible between subsequent guards
+a compact representation of the local variables of symbolic frames is used.
+Every variable in the symbolic frame is encoded using two bytes.
+Two bits are used as a tag to denote where the value of the variable
+comes from.
+The remaining 14 bits are a payload that depends on the tag bits.
+
+The possible source of information are:
+
+\begin{itemize}
+    \item For small integer constants
+        the payload contains the value of the constant.
+    \item For other constants
+        the payload contains an index into a per-loop list of constants.
+    \item For SSA variables,
+        the payload is the number of the variable.
+    \item For virtuals,
+        the payload is an index into a list of virtuals, see next section.
+\end{itemize}
+
 \subsection{Interaction With Optimization}
@@
 or other virtual objects.
+They are encoded using the same compact two-byte representation
+as local variables.
 
@@
 % subsection Interaction With Optimization (end)
-
-* High level handling of resumedata
-
-- frames consist of a PC and local variables
-
-compression approaches:
-- use fact that outer frames don't change in the part of the trace that is in the inner frame
-- compact bit-representation for constants/ssa vars
-
-interaction with optimization
-- guard coalescing
-- virtuals
-  - most virtuals not changed between guards
 

### 2012-07-19 talk/vmil2012/paper.tex
write a bit more about guards and bridges

--- 2012-07-18 talk/vmil2012/paper.tex
@@
 a new trace, refered to as bridge, starting from this guard is recorded and
-compile. The goal for the execution of bridges, that become a part of the
-common path is to favor the performance while staying on the compiled trace.
-This means we want wo avoid switching back and forth to the frontend when a
-guard corresponding to the bridge fails before we can execute the brigde.
-Instead we want to have as little as possible overhead when switching from the
-loop path to the bridge.
+compiled. When compiling bridges the goal is that future failures of the guards
+that led to the compilation of the bridge should execute the bridge without
+additional overhead, in particular it the failure of the guard should not lead
+to leaving the compiled code prior to execution the code of the bridge.
 
@@
 instructions and guards are processed in the same way as described above. The
-main difference is the setup phase, when compiling a trace we start with a
-clean slate, whilst the compilation of a bridge starts with state as it was
-when compiling the guard. To restore the state needed compile the bridge we use
-the encoded representation created for the guard to rebuild the bindings from
-IR-variables to stack locations and registers used in the register allocator.
+main difference is the setup phase, when compiling a trace we start with a lean
+slate. The compilation of a bridge is starts from a state (register and stack
+bindings) that corresponds to the state during the compilation of the original
+guard. To restore the state needed compile the bridge we use the encoded
+representation created for the guard to rebuild the bindings from IR-variables
+to stack locations and registers used in the register allocator.  With this
+reconstruction all bindings are restored to the state as they were in the
+original loop up to the guard.
 
 Once the bridge has been compiled the trampoline method stub is redirected to
-the code of the bridge. In future if the guard fails again it jumps to the
-trampoline and then jumps to the code compiled for the bridge, having alsmos no
-overhead compared to the execution of the original trace, behaving mainly as
-two paths in a conditional block.
+the code of the bridge. In future if the guard fails again it jumps to the code
+compiled for the bridge instead of bailing out. Once the guard has been
+compiled and attached to the loop the guard becomes just a point where
+control-flow can split. The loop after the guard and the bridge are just
+condional paths.
 

### 2012-07-20 talk/vmil2012/paper.tex
typos

--- 2012-07-19 talk/vmil2012/paper.tex
@@
 This is done with many classical compiler optimizations.
-In particular guards can be remove by subexpression elimination.
+In particular guards can be removed by subexpression elimination.
 If the same guard is encountered a second time in the trace,
@@
 how the heap objects that have to be allocated on guard failure look like.
-To this end, the content of the every field of the virtual object is described
+To this end, the content of every field of the virtual object is described
 in the same way that the local variables of symbolic frames are described.

### 2012-07-20 talk/vmil2012/paper.tex
more low level

@@
 
-Code generation consists of two passes over the lists of instructions, a
-backwards pass to calculate live ranges of IR-level variables and a forward one
-to emit the instructions. During the forward pass IR-level variables are
-assigned to registers and stack locations by the register allocator according
-to the requirements of the to be emitted instructions. Eviction/spilling is
-performed based on the live range information collected in the first pass. Each
-IR instruction is transformed into one or more machine level instructions that
-implement the required semantics, operations withouth side effects whose result
-is not used are not emitted. Guards instructions are transformed into fast
-checks at the machine code level that verify the corresponding condition.  In
-cases the value being checked by the guard is not used anywhere else the guard
-and the operation producing the value can merged, reducing even more the
-overhead of the guard. \bivab{example for this}
+After optimization the resulting trace is handed to the backend to be compiled
+to machine code. The compilation phase consists of two passes over the lists of
+instructions, a backwards pass to calculate live ranges of IR-level variables
+and a forward one to emit the instructions. During the forward pass IR-level
+variables are assigned to registers and stack locations by the register
+allocator according to the requirements of the to be emitted instructions.
+Eviction/spilling is performed based on the live range information collected in
+the first pass. Each IR instruction is transformed into one or more machine
+level instructions that implement the required semantics, operations withouth
+side effects whose result is not used are not emitted. Guards instructions are
+transformed into fast checks at the machine code level that verify the
+corresponding condition.  In cases the value being checked by the guard is not
+used anywhere else the guard and the operation producing the value can merged,
+reducing even more the overhead of the guard. \bivab{example for this}
 
@@
 corresponding to the guard. When a guard is compiled, additionally to the
-condition check two things are generated/compiled. First a special
-data structure is created that encodes the information provided by the register
-allocator about where the values corresponding to each IR-variable required by
-the guard will be stored when execution reaches the code emitted for the
-corresponding guard. \bivab{go into more detail here?!} This encoding needs to
-be as compact as possible to maintain an acceptable memory profile.
+condition check two things are generated/compiled. First a special data
+structure called \emph{low-level resume data} is created that encodes the
+information provided by the register allocator about where the values
+corresponding to each IR-variable required by the guard will be stored when
+execution reaches the code emitted for the corresponding guard. \bivab{go into
+more detail here?!} This encoding needs to be as compact as possible to
+maintain an acceptable memory profile.
 
-\bivab{example goes here}
+\bivab{example for low-level resume data goes here}
 
-Second a trampoline method stub is generated. Guards are usually implemented as
-a conditional jump to this stub, jumping to in case the guard condition check
-does not hold and a side-exit should be taken. This stub loads the pointer to
-the encoding of the locations mentioned above, preserves the execution state
-(stack and registers) and then jumps to generic bail-out handler that is used
-to leave the compiled trace if case of a guard failure.
+Second a piece of code is generated for each guard that acts as a trampoline.
+Guards are implemented as a conditional jump to this trampoline. In case the
+condition checked in the guard fails execution and a side-exit should be taken
+execution jumps the the trampoline. In the trampoline the pointer to the
+\emph{low-level resume data} is loaded and jumps to generic bail-out handler
+that is used to leave the compiled trace if case of a guard failure.
 

### 2012-07-20 talk/vmil2012/paper.tex
add the trace of the example as a figure
import example program
add an example figure showing how an operation and a following guard are compiled with and without merging
mention the figure in the text

@@
 
+\begin{figure}
+    \input{figures/example.tex}
+    \caption{Example Program}
+    \label{fig:trace-log}
+\end{figure}
+
 \section{Resume Data}
@@
 
+\begin{figure}
+    \input{figures/log.tex}
+    \caption{Optimized trace}
+    \label{fig:trace-log}
+\end{figure}
 \section{Guards in the Backend}
@@
 used anywhere else the guard and the operation producing the value can merged,
-reducing even more the overhead of the guard. \bivab{example for this}
+reducing even more the overhead of the guard. Figure \ref{fig:trace-compiled}
+shows how an \texttt{int\_eq} operation followed by a guard that checks the
+result of the operation are compiled to pseudo-assembler if the operation and
+the guard are compiled separeated or if they are merged.
+
+\bivab{Figure needs better formatting}
+\begin{figure}[ht]
+  \noindent
+  \centering
+  \begin{minipage}{1\columnwidth}
+    \begin{lstlisting}
+    i8 = int_eq(i6, 1)
+    guard_false(i8) [i6, i1, i0]
+    \end{lstlisting}
+  \end{minipage}
+  \begin{minipage}{.40\columnwidth}
+    \begin{lstlisting}
+CMP r6, #1
+MOVEQ r8, #1
+MOVNE r8, #0
+CMP r8, #0
+BEQ <bailout>
+    \end{lstlisting}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{.40\columnwidth}
+    \begin{lstlisting}
+CMP r6, #1
+BNE <bailout>
+...
+...
+...
+    \end{lstlisting}
+  \end{minipage}
+  \caption{Separated and merged compilation of operations and guards}
+  \label{fig:trace-compiled}
+\end{figure}
 

### 2012-07-23 talk/iwtc11/benchmarks/iter/xsum1d.c
add a list of the benchmarks used

@@ -1,0 +1,22 @@
+#include <stdio.h>
+#include <math.h>
+#include <stdlib.h>
+
+double result;
+
+double sum(double *a, int n) {
+  int i;
+  double sa = 0;
+  for (i=0; i<n; i++) sa += a[i] + i;
+  return sa;
+}
+
+#define N 100000000
+
+int main(int ac, char **av) {
+  double *a = malloc(N*sizeof(double));
+  int i, n = atoi(av[1]);
+  for (i=0; i<n; i++) result = sum(a, N);
+  fprintf(stderr, "xsum1d:    ");
+  return 0;
+}

### 2012-07-23 talk/vmil2012/paper.tex
make david first author
add a list of the benchmarks used

@@
 
-\authorinfo{Carl Friedrich Bolz$^a$ \and David Schneider$^{a}$}
+\authorinfo{David Schneider$^{a}$ \and Carl Friedrich Bolz$^a$}
            {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
@@
 Although there are several publications about tracing jut-in-time compilers, to
-our knowledge, there are none that describe the use and implementaiton of
-guards in this context. With the following contributions we aim to sched some
+our knowledge, there are none that describe the use and implementation of
+guards in this context. With the following contributions we aim to shed some
 light (to much?) on this topic.
@@
 \end{figure}
+% section Resume Data (end)
 \section{Guards in the Backend}
@@
 result of the operation are compiled to pseudo-assembler if the operation and
-the guard are compiled separeated or if they are merged.
+the guard are compiled separated or if they are merged.
 
@@
 condition checked in the guard fails execution and a side-exit should be taken
-execution jumps the the trampoline. In the trampoline the pointer to the
+execution jumps to the trampoline. In the trampoline the pointer to the
 \emph{low-level resume data} is loaded and jumps to generic bail-out handler
-that is used to leave the compiled trace if case of a guard failure.
+that is used to leave the compiled trace in case of a guard failure.
 
@@
 As in previous sections the underlying idea for the design of guards is to have
-a fast on trace profile and a potentially slow one in the bail-out case where
+a fast on-trace profile and a potentially slow one in the bail-out case where
 the execution takes one of the side exits due to a guard failure. At the same
@@
 
-As explained in previous sections, when a specific guard has failed often enogh
-a new trace, refered to as bridge, starting from this guard is recorded and
+As explained in previous sections, when a specific guard has failed often enough
+a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
 compiled. When compiling bridges the goal is that future failures of the guards
 that led to the compilation of the bridge should execute the bridge without
-additional overhead, in particular it the failure of the guard should not lead
+additional overhead, in particular the failure of the guard should not lead
 to leaving the compiled code prior to execution the code of the bridge.
 
-The process of compiling a bridge is very similar to compiling a loop,
-instructions and guards are processed in the same way as described above. The
-main difference is the setup phase, when compiling a trace we start with a lean
-slate. The compilation of a bridge is starts from a state (register and stack
+The process of compiling a bridge is very similar to compiling a loop.
+Instructions and guards are processed in the same way as described above. The
+main difference is the setup phase. When compiling a trace we start with a clean
+slate. The compilation of a bridge is started from a state (register and stack
 bindings) that corresponds to the state during the compilation of the original
-guard. To restore the state needed compile the bridge we use the encoded
+guard. To restore the state needed to compile the bridge we use the encoded
 representation created for the guard to rebuild the bindings from IR-variables
@@
 control-flow can split. The loop after the guard and the bridge are just
-condional paths.
+conditional paths. \cfbolz{maybe add the unpatched and patched assembler of the trampoline as well?}
 
@@
 \label{sec:evaluation}
+
+\include{figures/benchmarks_table}
 

### 2012-07-24 talk/vmil2012/paper.tex
write something about SPUR

@@
 
-Although there are several publications about tracing jut-in-time compilers, to
+Although there are several publications about tracing just-in-time compilers, to
 our knowledge, there are none that describe the use and implementation of
@@
 
+SPUR~\cite{bebenita_spur:_2010} is a tracing JIT compiler
+for a C\# virtual machine.
+It handles guards by always generating code for every one of them
+that transfers control back to the unoptimized code.
+Since the transfer code needs to reconstruct the stack frames
+of the unoptimized code,
+the transfer code is quite large.
+
 \section{Conclusion}

### 2012-07-24 talk/icooolps2011/code/interpreter-slow.tex
ruthlessly cut space, mostly by making font smaller. not pretty, but I have no idea what to cut atm.
a slightly different approach to saving space: for the small examples, put code and trace next to each other. this makes it possible to use the larger font again.
write something about SPUR

@@
-\begin{lstlisting}[mathescape,basicstyle=\ttfamily,numbers = right]
+{\noop
+\begin{lstlisting}[mathescape,basicstyle=\ttfamily,numbers = right,numberblanklines=false]
 class Class(object):

### 2012-07-24 talk/ep2012/stackless/demo/pickledtasklet.py

@@ -1,0 +1,25 @@
+import pickle, sys
+import stackless
+
+ch = stackless.channel()
+
+def recurs(depth, level=1):
+    print 'enter level %s%d' % (level*'  ', level)
+    if level >= depth:
+        ch.send('hi')
+    if level < depth:
+        recurs(depth, level+1)
+    print 'leave level %s%d' % (level*'  ', level)
+
+def demo(depth):
+    t = stackless.tasklet(recurs)(depth)
+    print ch.receive()
+    pickle.dump(t, file('tasklet.pickle', 'wb'))
+
+if __name__ == '__main__':
+    if len(sys.argv) > 1:
+        t = pickle.load(file(sys.argv[1], 'rb'))
+        t.insert()
+    else:
+        t = stackless.tasklet(demo)(9)
+    stackless.run()

### 2012-07-24 talk/vmil2012/paper.tex
remember to do store sinking
explain store sinking

@@
 
+Similarly, stores into the heap are delayed as long as possible.
+This can make it necessary to perform these delayed stores
+when leaving the trace via a guard.
+Therefore the resume data needs to contain a description
+of the delayed stores to be able to perform them when the guard fails.
+So far no special compression is done with this information.
+
 % subsection Interaction With Optimization (end)

### 2012-07-24 talk/dls2012/paper.tex
add to references to the related work section

--- 2011-07-01 talk/iwtc11/paper.tex
@@
 motion which is a very important optimization for code with tight kernels.
+Especially for dynamic languages that typically performs quite a lot of loop invariant
+type checking, boxed value unwrapping and virtual method lookups.
 In this paper we present a scheme for making simple optimizations loop-aware by
@@
 \section{Introduction}
+
+A dynamic language typically needs to do quite a lot of type
+checking, wrapping/unwrapping of boxed values, and virtual method dispatching. 
+For tight computationally intensive loops a
+significant amount of the execution time might be spend on such tasks
+instead of the actual computations. Moreover, the type checking,
+unwrapping and method lookups are often loop invariant and performance could be increased
+by moving those operations out of the loop. We propose a simple scheme
+to make a tracing JIT loop-aware by allowing it's existing optimizations to
+perform loop invariant code motion.
 
@@
 To copy an operation $v=\text{op}\left(A_1, A_2, \cdots, A_{|A|}\right)$
-a new variable, $\hat v$ is introduced. The copied operation will
+a new variable, $\hat v$, is introduced. The copied operation will
 return $\hat v$ using
@@
 \begin{equation}
-  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
+  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H|}\right)
   \label{eq:heap-inputargs}
@@
 \begin{equation}
-  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
+  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H|})\right)
   .
@@
 The arguments of the \lstinline{jump} operation of the peeled loop,
-$K$, is constructed by inlining $\hat J$,
+$K$, is constructed from $\hat J$ using the map $m$,
 \begin{equation}

### 2012-07-24 talk/vmil2012/paper.tex
add to references to the related work section
a paragraph about self

@@
 
+Deutsch et. al.~\cite{XXX} describe the use of stack descriptions
+to make it possible to do source-level debugging of JIT-compiled code.
+Self uses deoptimization to reach the same goal~\cite{XXX}.
+When a function is to be debugged, the optimized code version is left
+and one compiled without inlining and other optimizations is entered.
+Self uses scope descriptors to describe the frames
+that need to be re-created when leaving the optimized code.
+The scope descriptors are between 0.45 and 0.76 times
+the size of the generated machine code.
+
 SPUR~\cite{bebenita_spur:_2010} is a tracing JIT compiler
@@
 the transfer code is quite large.
+
+\bivab{mention Gal et al.~\cite{Gal:2009ux} trace stitching}
+and also mention \bivab{Dynamo's fragment linking~\cite{Bala:2000wv}} in
+relation to the low-level guard handling.
 

### 2012-07-25 talk/vmil2012/paper.tex
separate related work into two subsections
write about hotspot and escape analysis

@@
 
-Deutsch et. al.~\cite{XXX} describe the use of stack descriptions
-to make it possible to do source-level debugging of JIT-compiled code.
-Self uses deoptimization to reach the same goal~\cite{XXX}.
-When a function is to be debugged, the optimized code version is left
-and one compiled without inlining and other optimizations is entered.
-Self uses scope descriptors to describe the frames
-that need to be re-created when leaving the optimized code.
-The scope descriptors are between 0.45 and 0.76 times
-the size of the generated machine code.
+\subsection{Guards in Other Tracing JITs}
+\label{sub:Guards in Other Tracing JITs}
 
@@
 
+% subsection Guards in Other Tracing JITs (end)
+
+\subsection{Deoptimization in Method-Based JITs}
+\label{sub:Deoptimization in Method-Based JITs}
+
+Deoptimization in method-based JITs is used if one of the assumptions
+of the code generated by a JIT-compiler changes.
+This is often the case when new code is added to the system,
+or when the programmer tries to debug the program.
+
+Deutsch et. al.~\cite{XXX} describe the use of stack descriptions
+to make it possible to do source-level debugging of JIT-compiled code.
+Self uses deoptimization to reach the same goal~\cite{XXX}.
+When a function is to be debugged, the optimized code version is left
+and one compiled without inlining and other optimizations is entered.
+Self uses scope descriptors to describe the frames
+that need to be re-created when leaving the optimized code.
+The scope descriptors are between 0.45 and 0.76 times
+the size of the generated machine code.
+
+Java Hotspot~\cite{XXX} contains a deoptimization framework that is used
+for debugging and when an uncommon trap is triggered.
+To be able to do this, Hotspot stores a mapping from optimized states
+back to the interpreter state at various deoptimization points.
+There is no discussion of the memory use of this information.
+
+The deoptimization information of Hotspot is extended
+to support correct behaviour
+when scalar replacement of fields is done for non-escaping objects~\cite{XXX}.
+The approach is extremely similar to how RPython's JIT handles virtual objects.
+For every object that is not allocated in the code,
+the deoptimization information contains a description
+of the content of the fields.
+When deoptimizing code, these objects are reallocated
+and their fields filled with the values
+described by the deoptimization information.
+The paper does not describe any attempts to store this information compactly.
+
+% subsection Deoptimization in Method-Based JITs (end)
+
 \section{Conclusion}

### 2012-07-25 talk/vmil2012/paper.tex
need to mention luajit
split bibliography into an auto-generated one and the rest
add citations

@@
 
+LuaJIT, ...
+
 % subsection Guards in Other Tracing JITs (end)
@@
 
-Deutsch et. al.~\cite{XXX} describe the use of stack descriptions
+Deutsch et. al.~\cite{deutsch_efficient_1984} describe the use of stack descriptions
 to make it possible to do source-level debugging of JIT-compiled code.
@@
 
-Java Hotspot~\cite{XXX} contains a deoptimization framework that is used
+Java Hotspot~\cite{paleczny_java_2001} contains a deoptimization framework that is used
 for debugging and when an uncommon trap is triggered.
@@
 to support correct behaviour
-when scalar replacement of fields is done for non-escaping objects~\cite{XXX}.
+when scalar replacement of fields is done for non-escaping objects~\cite{kotzmann_escape_2005}.
 The approach is extremely similar to how RPython's JIT handles virtual objects.
@@
 \bibliographystyle{abbrv}
-\bibliography{paper}
+\bibliography{zotero,paper}
 

### 2012-07-25 talk/vmil2012/tool/build_tables.py
add code to build backend data tables about machine code and guard data sizes and add the latest results

@@ -2,17 +2,21 @@ from __future__ import division
 import csv
 import django
 from django.template import Template, Context
-import optparse
-from os import path
+import os
 import sys
 
-#
+# This line is required for Django configuration
+django.conf.settings.configure()
 
 
-def main(csvfile, template, texfile):
+def getlines(csvfile):
     with open(csvfile, 'rb') as f:
         reader = csv.DictReader(f, delimiter=',')
-        lines = [l for l in reader]
+        return [l for l in reader]
+
+
+def build_ops_count_table(csvfile, texfile, template):
+    lines = getlines(csvfile)
 
     head = ['Benchmark',
             'ops b/o',
@@ -20,7 +24,7 @@ def main(csvfile, template, texfile):
             'ops a/o',
             '\\% guards a/o',
             'opt. rate',
-            'guard opt. rate',]
+            'guard opt. rate']
 
     table = []
     # collect data
@@ -33,22 +37,43 @@ def main(csvfile, template, texfile):
         res = [
                 bench['bench'].replace('_', '\\_'),
                 ops_bo,
-                "%.2f (%s)" % (guards_bo / ops_bo * 100, bench['guard before']),
+                "%.2f (%s)" % (guards_bo / ops_bo * 100,
+                                 bench['guard before']),
                 ops_ao,
-                "%.2f (%s)" % (guards_ao / ops_ao * 100, bench['guard after']),
-                "%.2f" % ((1 - ops_ao/ops_bo) * 100,),
-                "%.2f" % ((1 - guards_ao/guards_bo) * 100,),
+                "%.2f (%s)" % (guards_ao / ops_ao * 100,
+                                  bench['guard after']),
+                "%.2f" % ((1 - ops_ao / ops_bo) * 100,),
+                "%.2f" % ((1 - guards_ao / guards_bo) * 100,),
               ]
         table.append(res)
     output = render_table(template, head, sorted(table))
+    write_table(output, texfile)
+
+
+def build_backend_count_table(csvfile, texfile, template):
+    lines = getlines(csvfile)
+
+    head = ['Benchmark',
+            'Machine code size (kB)',
+            'll resume data (kB)']
+
+    table = []
+    # collect data
+    for bench in lines:
+        bench['bench'] = bench['bench'].replace('_', '\\_')
+        keys = ['bench', 'asm size', 'guard map size']
+        table.append([bench[k] for k in keys])
+    output = render_table(template, head, sorted(table))
[…]

### 2012-07-25 talk/vmil2012/paper.tex
fix capitalization
add code to build backend data tables about machine code and guard data sizes and add the latest results
move the figure and caption definition from the table template to the paper
start explaining the contents of the tables

--- 2012-07-24 talk/vmil2012/paper.tex
@@
 
+The following analysis is based on a selection of benchmarks taken from the set
+of benchmarks used to measure the performance of PyPy as can be seen
+on\footnote{http://speed.pypy.org/}. The selection is based on the following
+criteria \bivab{??}. The benchmarks were taken from the PyPy benchmarks
+repository using revision
+\texttt{ff7b35837d0f}\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}.
+The benchmarks were run on a version of PyPy based on the
+tag~\texttt{release-1.9} and patched to collect additional data about the
+guards in the machine code
+backends\footnote{https://bitbucket.org/pypy/pypy/src/release-1.9}. All
+benchmark data was collected on a MacBook Pro 64 bit running Max OS
+10.7.4 \bivab{do we need more data for this kind of benchmarks} with the loop
+unrolling optimization disabled\bivab{rationale?}.
+
+Figure~\ref{fig:ops_count} shows the total number of operations that are
+recorded during tracing for each of the benchmarks on what percentage of these
+are guards. Figure~\ref{fig:ops_count} also shows the number of operations left
+after performing the different trace optimizations done by the trace optimizer,
+such as xxx. The last columns show the overall optimization rate and the
+optimization rate specific for guard operations, showing what percentage of the
+operations was removed during the optimizations phase.
+
+\begin{figure*}
     \include{figures/benchmarks_table}
+    \caption{Benchmark Results}
+    \label{fig:ops_count}
+\end{figure*}
+
+\bivab{should we rather count the trampolines as part of the guard data instead
+of counting it as part of the instructions}
+
+Figure~\ref{fig:backend_data} shows
+the total memory consumption of the code and of the data generated by the machine code
+backend for the different benchmarks mentioned above. Meaning the operations
+left after optimization take the space shown in Figure~\ref{fig:backend_data}
+after being compiled. Also the additional data stored for the guards to be used
+in case of a bailout and attaching a bridge.
+\begin{figure*}
+    \include{figures/backend_table}
+    \caption{Total size of generated machine code and guard data}
+    \label{fig:backend_data}
+\end{figure*}
+
+Both figures do not take into account garbage collection. Pieces of machine
+code can be globally invalidated or just become cold again. In both cases the
+generated machine code and the related data is garbage collected. The figures
+show the total amount of operations that are evaluated by the JIT and the
+total amount of code and data that is generated from the optimized traces.
 

### 2012-07-26 talk/vmil2012/paper.tex
update figure

@@
 
-\title{Efficiently Handling Guards in the low level design of RPython's tracing JIT}
+\title{Efficiently Handling Guards in the Low Level Design of RPython's tracing JIT}
 
@@
   \begin{minipage}{1\columnwidth}
-    \begin{lstlisting}
-    i8 = int_eq(i6, 1)
-    guard_false(i8) [i6, i1, i0]
+    \begin{lstlisting}[mathescape]
+$b_1$ = int_eq($i_2$, 1)
+guard_false($b_1$)
     \end{lstlisting}
@@
 
+LuaJIT, ...
+
+% subsection Guards in Other Tracing JITs (end)
+
+\subsection{Deoptimization in Method-Based JITs}
+\label{sub:Deoptimization in Method-Based JITs}
+
+Deoptimization in method-based JITs is used if one of the assumptions
+of the code generated by a JIT-compiler changes.
+This is often the case when new code is added to the system,
+or when the programmer tries to debug the program.
+
+Deutsch et. al.~\cite{deutsch_efficient_1984} describe the use of stack descriptions
+to make it possible to do source-level debugging of JIT-compiled code.
+Self uses deoptimization to reach the same goal~\cite{XXX}.
+When a function is to be debugged, the optimized code version is left
+and one compiled without inlining and other optimizations is entered.
+Self uses scope descriptors to describe the frames
+that need to be re-created when leaving the optimized code.
+The scope descriptors are between 0.45 and 0.76 times
+the size of the generated machine code.
+
+Java Hotspot~\cite{paleczny_java_2001} contains a deoptimization framework that is used
+for debugging and when an uncommon trap is triggered.
+To be able to do this, Hotspot stores a mapping from optimized states
+back to the interpreter state at various deoptimization points.
+There is no discussion of the memory use of this information.
+
+The deoptimization information of Hotspot is extended
+to support correct behaviour
+when scalar replacement of fields is done for non-escaping objects~\cite{kotzmann_escape_2005}.
+The approach is extremely similar to how RPython's JIT handles virtual objects.
+For every object that is not allocated in the code,
+the deoptimization information contains a description
+of the content of the fields.
+When deoptimizing code, these objects are reallocated
+and their fields filled with the values
+described by the deoptimization information.
+The paper does not describe any attempts to store this information compactly.
+
+% subsection Deoptimization in Method-Based JITs (end)
+
 \section{Conclusion}

### 2012-07-26 talk/vmil2012/tool/difflogs.py
consider some more operations as get/set, consider uint operations as numeric and print an overview of the operations considered "rest" at the end

@@ -16,19 +16,27 @@ from pypy.jit.metainterp.history import ConstInt
 from pypy.rpython.lltypesystem import llmemory, lltype
 
 categories = {
-    'setfield_gc': 'set',
-    'setarrayitem_gc': 'set',
-    'strsetitem': 'set',
-    'getfield_gc': 'get',
-    'getfield_gc_pure': 'get',
     'getarrayitem_gc': 'get',
     'getarrayitem_gc_pure': 'get',
-    'strgetitem': 'get',
+    'getarrayitem_raw': 'get',
+    'getfield_gc': 'get',
+    'getfield_gc_pure': 'get',
+    'getfield_raw': 'get',
+    'getinteriorfield_gc': 'get',
     'new': 'new',
     'new_array': 'new',
-    'newstr': 'new',
     'new_with_vtable': 'new',
+    'newstr': 'new',
+    'newunicode': 'new',
+    'setarrayitem_gc': 'set',
+    'setarrayitem_raw': 'set',
+    'setfield_gc': 'set',
+    'setfield_raw': 'set',
+    'setinteriorfield_gc': 'set',
+    'strgetitem': 'get',
+    'strsetitem': 'set',
 }
+rest_op_bucket = set()
 
 all_categories = 'new get set guard numeric rest'.split()
 
@@ -58,12 +66,17 @@ def summarize(loop, adding_insns={}):    # for debugging
             else:
                 assert categories.get(opname, "rest") == "get"
                 continue
-        if opname.startswith("int_") or opname.startswith("float_"):
+        if(opname.startswith("int_")
+                or opname.startswith("float_")
+                or opname.startswith('uint_')):
             opname = "numeric"
         elif opname.startswith("guard_"):
             opname = "guard"
         else:
-            opname = categories.get(opname, 'rest')
+            _opname = categories.get(opname, 'rest')
+            if _opname == 'rest':
+                rest_op_bucket.add(opname)
+            opname = _opname
         insns[opname] = insns.get(opname, 0) + 1
     assert seen_label
     return insns
@@ -178,3 +191,8 @@ if __name__ == '__main__':
         sys.exit(2)
     else:
         main(args[0], options)
+    if len(rest_op_bucket):
+        print "=" * 80
+        print "Elements considered as rest"
+        for x in sorted(rest_op_bucket):
+            print x

### 2012-07-26 talk/vmil2012/paper.tex
consider some more operations as get/set, consider uint operations as numeric and print an overview of the operations considered "rest" at the end

@@
 \usepackage{setspace}
+\usepackage[colorinlistoftodos]{todonotes}
 
@@
 
+\todo{add page numbers (draft) for review}
 In this paper we describe and analyze how deoptimization works in the context
@@
 The contributions of this paper are:
+\todo{more motivation}
+\todo{extend}
+\todo{contributions, description of PyPy's guard architecture, analysis on benchmarks}
 \begin{itemize}
@@
 alternative implementation of the Python language\bivab{mention speed}. The
-Implementation takes advantage of the language features provided by RPython
+implementation takes advantage of the language features provided by RPython
 such as the provided tracing just-in-time compiler described below.
@@
  * Tracing JITs
+ * Mention SSA
  * JIT Compiler
@@
 The possible points of deviation from the trace are guard operations
-that check whether the same assumptions as during tracing still hold.
+that check whether the same assumptions observed during tracing still hold during execution.
 In later executions of the trace the guards can fail.
@@
 This means it is necessary to attach enough information to a guard
-to construct the interpreter state when that guard fails.
+to reconstruct the interpreter state when that guard fails.
 This information is called the \emph{resume data}.
 
-To do this reconstruction, it is necessary to take the values
-of the SSA variables of the trace
-and build interpreter stack frames.
-Tracing aggressively inlines functions.
-Therefore the reconstructed state of the interpreter
-can consist of several interpreter frames.
+To do this reconstruction, it is necessary to take the values of the SSA
+variables of the trace and build interpreter stack frames.  Tracing
+aggressively inlines functions, therefore the reconstructed state of the
+interpreter can consist of several interpreter frames.
 
@@
 \end{itemize}
+\todo{figure showing linked resume-data}
 
@@
 % subsection Interaction With Optimization (end)
+\subsection{Compiling Side-Exits and Trace Stitching} % (fold)
+\label{sub:Compiling side-exits and trace stitching}
+   * tracing and attaching bridges and throwing away resume data
+   * restoring the state of the tracer
+     * keeping virtuals
+   * compiling bridges
+\todo{maybe mention that the failargs also go into the bridge}
 
-   * tracing and attaching bridges and throwing away resume data
-   * compiling bridges
-\bivab{mention that the failargs also go into the bridge}
+% subsection Compiling side-exits and trace stitching (end)
 % section Resume Data (end)
 
+\todo{set line numbers to the line numbers of the rpython example}
 \begin{figure}
@@
 control-flow can split. The loop after the guard and the bridge are just
-conditional paths. \cfbolz{maybe add the unpatched and patched assembler of the trampoline as well?}
-
+conditional paths. \todo{add figure of trace with trampoline and patched guard to a bridge}
 %* Low level handling of guards
@@
 of benchmarks used to measure the performance of PyPy as can be seen
-on\footnote{http://speed.pypy.org/}. The selection is based on the following
-criteria \bivab{??}. The benchmarks were taken from the PyPy benchmarks
+on.\footnote{http://speed.pypy.org/} The benchmarks were taken from the PyPy benchmarks
 repository using revision
-\texttt{ff7b35837d0f}\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}.
+\texttt{ff7b35837d0f}.\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}
 The benchmarks were run on a version of PyPy based on the
@@
 guards in the machine code
-backends\footnote{https://bitbucket.org/pypy/pypy/src/release-1.9}. All
-benchmark data was collected on a MacBook Pro 64 bit running Max OS
-10.7.4 \bivab{do we need more data for this kind of benchmarks} with the loop
-unrolling optimization disabled\bivab{rationale?}.
+backends.\footnote{https://bitbucket.org/pypy/pypy/src/release-1.9} All
+benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.7.4 with
+the loop unrolling optimization disabled.\footnote{Since loop unrolling
+duplicates the body of loops it would no longer be possible to meaningfully
+compare the number of operations before and after optimization. Loop unrolling
+is most effective for numeric kernels, so the benchmarks presented here are not
+affected much by its absence.}
 
@@
 
-\bivab{should we rather count the trampolines as part of the guard data instead
-of counting it as part of the instructions}
+\todo{resume data size estimates on 64bit}
+\todo{figure about failure counts of guards (histogram?)}
+\todo{integrate high level resume data size into Figure \ref{fig:backend_data}}
+\todo{count number of guards with bridges for \ref{fig:ops_count}}
+\todo{add resume data sizes without sharing}
 
@@
 
-LuaJIT, ...
+\todo{look into tracing papers for information about guards and deoptimization}
+LuaJIT \todo{link to mailing list discussion}
 
@@
 
+\todo{conclusion}
+
 \section*{Acknowledgements}
-
 \bibliographystyle{abbrv}
 \bibliography{zotero,paper}
-
+\listoftodos
 \end{document}

### 2012-07-26 talk/ep2006/fibonacci2.py

@@ -1,0 +1,40 @@
+"""
+This is a typical Functional Programming Languages demo, computing the
+Fibonacci sequence by using an infinite lazy linked list.
+"""
+
+try:
+    newvar    # only available in 'py.py -o logic'
+except NameError:
+    print __doc__
+    raise SystemExit(2)
+
+# ____________________________________________________________
+
+class Node:
+    def __init__(self, head, tail):
+        self.head = head
+        self.tail = tail
+
+Fibonacci = Node(1, Node(1, newvar()))
+
+def generate():
+    lst = Fibonacci
+    while 1:
+        a = newvar()
+        b = newvar()
+        tail = newvar()
+        unify(lst, Node(a, Node(b, tail)))
+        wait_needed(tail)
+        newtail = newvar()
+        bind(tail, Node(a+b, newtail))
+        lst = lst.tail
+
+def display(limit):
+    cur = Fibonacci
+    for i in range(limit):
+        print cur.head
+        cur = cur.tail
+
+uthread(generate)
+uthread(display, 15)

### 2012-07-26 talk/vmil2012/tool/build_tables.py

@@ -1,0 +1,113 @@
+from __future__ import division
+import csv
+import django
+from django.template import Template, Context
+import os
+import sys
+
+# This line is required for Django configuration
+django.conf.settings.configure()
+
+
+def getlines(csvfile):
+    with open(csvfile, 'rb') as f:
+        reader = csv.DictReader(f, delimiter=',')
+        return [l for l in reader]
+
+
+def build_ops_count_table(csvfiles, texfile, template):
+    lines = getlines(csvfiles[0])
+    bridge_lines = getlines(csvfiles[1])
+    bridgedata = {}
+    for l in bridge_lines:
+        bridgedata[l['bench']] = l
+
+    head = ['Benchmark',
+            'ops b/o',
+            '\\% guards b/o',
+            'ops a/o',
+            '\\% guards a/o',
+            'opt. rate in \\%',
+            'guard opt. rate in \\%',
+            'bridges']
+
+    table = []
+    # collect data
+    for bench in lines:
+        keys = 'numeric guard set get rest new'.split()
+        ops_bo = sum(int(bench['%s before' % s]) for s in keys)
+        ops_ao = sum(int(bench['%s after' % s]) for s in keys)
+        guards_bo = int(bench['guard before'])
+        guards_ao = int(bench['guard after'])
+        res = [
+                bench['bench'].replace('_', '\\_'),
+                ops_bo,
+                "%.2f" % (guards_bo / ops_bo * 100,),
+                ops_ao,
+                "%.2f" % (guards_ao / ops_ao * 100,),
+                "%.2f" % ((1 - ops_ao / ops_bo) * 100,),
+                "%.2f" % ((1 - guards_ao / guards_bo) * 100,),
+                bridgedata[bench['bench']]['bridges'],
+              ]
+        table.append(res)
+    output = render_table(template, head, sorted(table))
+    write_table(output, texfile)
+
+
+def build_backend_count_table(csvfiles, texfile, template):
+    lines = getlines(csvfiles[0])
+
+    head = ['Benchmark',
+            'Machine code size (kB)',
+            'll resume data (kB)',
+            '\\% of machine code size']
+
+    table = []
+    # collect data
+    for bench in lines:
+        bench['bench'] = bench['bench'].replace('_', '\\_')
+        keys = ['bench', 'asm size', 'guard map size']
+        gmsize = int(bench['guard map size'])
+        asmsize = int(bench['asm size'])
+        rel = "%.2f" % (gmsize / asmsize * 100,)
[…]

### 2012-07-26 talk/vmil2012/paper.tex
Add a table showing the relative numbers of all operation types to explain the motivation

@@
 
+\begin{figure*}
+    \include{figures/ops_count_table}
+    \caption{Relative numbers of operations in the traces generated for
+    different benchmarks}
+    \label{fig:ops_count}
+\end{figure*}
 Although there are several publications about tracing just-in-time compilers, to
@@
 
-Figure~\ref{fig:ops_count} shows the total number of operations that are
+Figure~\ref{fig:benchmarks} shows the total number of operations that are
 recorded during tracing for each of the benchmarks on what percentage of these
-are guards. Figure~\ref{fig:ops_count} also shows the number of operations left
+are guards. Figure~\ref{fig:benchmarks} also shows the number of operations left
 after performing the different trace optimizations done by the trace optimizer,
@@
     \caption{Benchmark Results}
-    \label{fig:ops_count}
+    \label{fig:benchmarks}
 \end{figure*}
@@
 \todo{integrate high level resume data size into Figure \ref{fig:backend_data}}
-\todo{count number of guards with bridges for \ref{fig:ops_count}}
 \todo{add resume data sizes without sharing}
+\todo{add a footnote about why guards have a threshold of 100}
 
@@
 LuaJIT \todo{link to mailing list discussion}
+http://lua-users.org/lists/lua-l/2009-11/msg00089.html
 

### 2012-07-27 talk/vmil2012/paper.tex
write a bit about the motivation

@@
 \todo{more motivation}
+Based on the informal observation that guards are among the most common
+operations in the traces produced by PyPy's tracing JIT and that guards are
+operations that are associated with an overhead to maintain information about
+state to be able to rebuild it, our goal is to present concrete numbers for the
+frecuency and the overhead produced by guards, explain how they are implemented
+in the diferent levels of PyPy's tracing JIT and explain the rationale behind
+the desing decisions based on the numbers.
 \todo{extend}

### 2012-07-27 talk/vmil2012/paper.tex
update os version for the machine running the benchmarks

@@
 backends.\footnote{https://bitbucket.org/pypy/pypy/src/release-1.9} All
-benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.7.4 with
+benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
 the loop unrolling optimization disabled.\footnote{Since loop unrolling

### 2012-07-30 talk/vmil2012/paper.tex
add calculated resume data size to asm and guard size table

@@
 state to be able to rebuild it, our goal is to present concrete numbers for the
-frecuency and the overhead produced by guards, explain how they are implemented
-in the diferent levels of PyPy's tracing JIT and explain the rationale behind
-the desing decisions based on the numbers.
+frequency and the overhead produced by guards, explain how they are implemented
+in the different levels of PyPy's tracing JIT and explain the rationale behind
+the design decisions based on the numbers.
+As can be seen on Figure~\ref{fig:ops_count} guards account for 14.42\% to
+22.32\% of the operations before and for 15.2\% to 20.12\% of after the
+optimization pass over the traced and compiled paths of the benchmarks.
+Figure~\ref{fig:benchmarks} shows the absolute number of operations for each
+benchmark, for every guard that stays alive after optimization there are
+several kinds of metadata created and stored at different levels of the JIT to
+be able to rebuild the interpreter or tracer state from a guard failure making
+the optimization \bivab{some good word} of guards an important aspect of the
+low-level design of a tracing just-in-time compiler.
 \todo{extend}

### 2012-07-30 talk/vmil2012/tool/build_tables.py
add calculated resume data size to asm and guard size table

@@ -89,24 +89,31 @@ def build_benchmarks_table(csvfiles, texfile, template):
 
 def build_backend_count_table(csvfiles, texfile, template):
     lines = getlines(csvfiles[0])
+    resume_lines = getlines(csvfiles[1])
+    resumedata = {}
+    for l in resume_lines:
+        resumedata[l['bench']] = l
 
     head = ['Benchmark',
             'Machine code size (kB)',
+            'hl resume data (kB)',
             'll resume data (kB)',
-            '\\% of machine code size']
+            'machine code resume data relation in \\%']
 
     table = []
     # collect data
     for bench in lines:
+        name = bench['bench']
         bench['bench'] = bench['bench'].replace('_', '\\_')
-        keys = ['bench', 'asm size', 'guard map size']
         gmsize = float(bench['guard map size'])
         asmsize = float(bench['asm size'])
-        rel = "%.2f" % (gmsize / asmsize * 100,)
+        rdsize = float(resumedata[name]['total resume data size'])
+        rel = "%.2f" % (asmsize / (gmsize + rdsize) * 100,)
         table.append([
             bench['bench'],
-            "%.2f" % (gmsize,),
             "%.2f" % (asmsize,),
+            "%.2f" % (rdsize,),
+            "%.2f" % (gmsize,),
             rel])
     output = render_table(template, head, sorted(table))
     write_table(output, texfile)
@@ -130,7 +137,7 @@ tables = {
         'benchmarks_table.tex':
             (['summary.csv', 'bridge_summary.csv'], build_benchmarks_table),
         'backend_table.tex':
-            (['backend_summary.csv'], build_backend_count_table),
+            (['backend_summary.csv', 'resume_summary.csv'], build_backend_count_table),
         'ops_count_table.tex':
             (['summary.csv'], build_ops_count_table),
         }

### 2012-07-30 talk/vmil2012/paper.tex
more related work

@@
 
-\todo{resume data size estimates on 64bit}
 \todo{figure about failure counts of guards (histogram?)}
-\todo{integrate high level resume data size into Figure \ref{fig:backend_data}}
 \todo{add resume data sizes without sharing}
@@
 
+Guards as described are a concept associated with tracing just-in-time
+compilers to represent possible divergent control flow paths.
+
 SPUR~\cite{bebenita_spur:_2010} is a tracing JIT compiler
@@
 
-\bivab{mention Gal et al.~\cite{Gal:2009ux} trace stitching}
-and also mention \bivab{Dynamo's fragment linking~\cite{Bala:2000wv}} in
-relation to the low-level guard handling.
+Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
+list different technologies and techniques used in the implementation of
+LuaJIT~\cite{Pall:2009}.\todo{decide if LuaJIT is a footnote or a reference and
+fix website citation} Pall explains that guards in LuaJIT use a datastucture
+called snapshots, similar to PyPy's resume data, to store the information about
+how to rebuild the state from a side-exit using the information in the snapshot
+and the machine execution state. Pall also acknowledges that snapshot for
+guards are associated with a large memory footprint. The solution used in
+LuaJIT is to store sparse snapshots, avoiding the creation of snapshots for
+every guard to reduce memory pressure. Snapshots are only created for guards
+after updates to the global state, after control flow points from the original
+program and for guards that are likely to fail. As an outlook Pall mentions the
+plans to switch to compressed snapshots to further reduce redundancy.
 
-\todo{look into tracing papers for information about guards and deoptimization}
-LuaJIT \todo{link to mailing list discussion}
-http://lua-users.org/lists/lua-l/2009-11/msg00089.html
+Linking side exits to pieces of later compiled machine code was described first
+in the context of Dynamo~\cite{Bala:2000wv} under the name of Fragment Linking.
+Once a new hot trace is emitted into the fragment cache it is linked to side
+exit that led to the compilation. Fragment Linking avoids the performance
+penalty involved in leaving the compiled and it to remove the compensation
+code used when restoring the machine state on a side exit.
+
+In~\cite{Gal:2006} Gal et. al describe that in the HotpathVM they experimented
+with having one generic compensation code block, like the RPython JIT, that
+uses a register variable mapping to restore the interpreter state. Later this
+was replaced by generting compensation code for each guard which produced a
+lower overhead in their benchmarks. HotpathVM also records secondary traces
+starting from failing guards that are connected directly to the original trace.
+Secondary traces are compiled by first restoring the register allocator state to
+the state at the side exit. The information is retrieved from a mapping stored
+in the guard that maps machine level registers and stack to Java level stack
+and variables.
+
+Gal et. al~\cite{Gal:2009ux} write about how TraceMonkey uses trace stitching
+to avoid th overhead of returning to the trace monitor and calling another
+trace when taking a side exit. In their approach it is required to write live
+values to an activation record before entering the new trace.
 

### 2012-07-31 talk/vmil2012/paper.tex
typo
Add a figure about the control flow in case of patched and unpatched guard failures and refer to the figure in the text

@@
 
-\bivab{example for low-level resume data goes here}
+\todo{example for low-level resume data showing how the current encoding works?}
 
@@
 the execution takes one of the side exits due to a guard failure. At the same
-time the data stored in the backend needed to rebuild the state should be be
-as compact as possible to reduce the memory overhead produced by the large
-number of guards\bivab{back this}.
+time the data stored in the backend needed to rebuild the state needs to be as
+compact as possible to reduce the memory overhead produced by the large number
+of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
+compressed encoding currently has about 25\% of the size of of the generated
+instructions on x86.
 
@@
 
-Once the bridge has been compiled the trampoline method stub is redirected to
-the code of the bridge. In future if the guard fails again it jumps to the code
-compiled for the bridge instead of bailing out. Once the guard has been
-compiled and attached to the loop the guard becomes just a point where
-control-flow can split. The loop after the guard and the bridge are just
-conditional paths. \todo{add figure of trace with trampoline and patched guard to a bridge}
+Once the bridge has been compiled the guard that led to compiling the birdge is
+patched to redirect control flow to the bridge in case the check fails. In
+future if the guard fails again it jumps to the code compiled for the bridge
+instead of bailing out. Once the guard has been compiled and attached to the
+loop the guard becomes just a point where control-flow can split. The loop
+after the guard and the bridge are just conditional paths.
+Figure~\ref{fig:trampoline} shows a digram of a compiled loop with two guards,
+Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
+then calls the compensation code, whereas Guard \#2 has already been patched
+and directly jumps to the corresponding bridge. The bridge also contains two
+guards that work based on the same principles.
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
+\caption{Trace control flow in case of guard failures with and without bridges}
+\label{fig:trampoline}
+\end{figure}
 %* Low level handling of guards
@@
 uses a register variable mapping to restore the interpreter state. Later this
-was replaced by generting compensation code for each guard which produced a
+was replaced by generating compensation code for each guard which produced a
 lower overhead in their benchmarks. HotpathVM also records secondary traces

### 2012-07-31 talk/vmil2012/paper.tex
set preprint option
refine point

@@
-\documentclass[10pt]{sigplanconf}
+\documentclass[10pt,preprint]{sigplanconf}
 
@@
 
-\todo{add page numbers (draft) for review}
 In this paper we describe and analyze how deoptimization works in the context
@@
 of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
-compressed encoding currently has about 25\% of the size of of the generated
-instructions on x86.
+compressed encoding currently has about 15\% to 25\% of the size of of the
+generated instructions on x86.
 

### 2012-08-01 talk/vmil2012/paper.tex
extend contribution and structure paragraphs of the introduction

@@
 low-level design of a tracing just-in-time compiler.
-\todo{extend}
-\todo{contributions, description of PyPy's guard architecture, analysis on benchmarks}
+The contributions of this paper are
+In this paper we want to substantiate the aforementioned and describe based on
+them the reasoning behind and the implementation of guards in PyPy's tracing
+just-in-time compiler.
 \begin{itemize}
- \item
+  \item An analysis of guards in the context of PyPy's tracing JIT to
+  substantiate the aforementioned observation, based on a set of benchmarks.
+  \item We provide a detailed measurements about the frequency and the
+  overhead associated with guards.
+  \item We provide a description about how guards are implemented in the high\-
+  and low-level parts of the JIT and describe the rationale behind the design.
 \end{itemize}
 
-The paper is structured as follows:
+The set of central concepts upon which this work is based is described in
+Section~\ref{sec:Background}, such as the PyPy project, the RPython language
+and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
+Data} we proceed to describe for PyPy's tracing JIT the details of guards in
+the frontend\bivab{better term for this?} related to recording and storing the
+information required to restore the interpreter state in case of a guard
+failure, once the frontend has traced and optimized a loop it invokes the
+backend to compile the operations to machine code, Section \ref{sec:Guards in
+the Backend} describes the low-level aspects of how guards are implemented in
+the JIT-backend. The frequency of guards and the overhead associated with the
+implementation described in this paper is discussed in
+Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
+overview about how guards are treated in the context of other just-in-time
+compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
+gives an outlook on further research topics.
 
@@
 
-\section{Resume Data}
+\section{Guards in the Frontend} %{Resume Data}
 \label{sec:Resume Data}
@@
 \section{Related Work}
+\label{sec:Related Work}
 
@@
 % subsection Deoptimization in Method-Based JITs (end)
+% section Related Work (end)
 
 \section{Conclusion}
+\label{sec:Conclusion}
 

### 2012-08-01 talk/vmil2012/paper.tex
generate a table showing only the percentage of guards before and after optimization for the set of benchmarks

@@
 
-\begin{figure*}
-    \include{figures/ops_count_table}
-    \caption{Relative numbers of operations in the traces generated for
-    different benchmarks}
-    \label{fig:ops_count}
-\end{figure*}
+%\begin{figure*}
+%    \include{figures/ops_count_table}
+%    \caption{Relative numbers of operations in the traces generated for
+%    different benchmarks}
+%    \label{fig:ops_count}
+%\end{figure*}
 Although there are several publications about tracing just-in-time compilers, to
@@
 just-in-time compiler.
+\begin{figure}
+    \include{figures/guard_table}
+    \caption{Percentage of guards before and after optimization for different benchmarks}
+    \label{fig:guard_percent}
+\end{figure}
 \begin{itemize}

### 2012-08-01 talk/vmil2012/tool/build_tables.py

@@ -41,6 +41,25 @@ def build_ops_count_table(csvfiles, texfile, template):
     output = render_table(template, head, sorted(table))
     write_table(output, texfile)
 
+def build_guard_table(csvfiles, texfile, template):
+    assert len(csvfiles) == 1
+    lines = getlines(csvfiles[0])
+    table = []
+    head = ['Benchmark', 'guards b/o in \%', 'guards a/o in \%']
+
+    keys = 'numeric set get rest new guard '.split()
+    for bench in lines:
+        ops = {'before': sum(int(bench['%s before' % s]) for s in keys),
+                'after': sum(int(bench['%s after' % s]) for s in keys)}
+
+        res = [bench['bench'].replace('_', '\\_'),]
+        for t in ('before', 'after'):
+            o = int(bench['guard %s' % t])
+            res.append('%.2f ' % (o / ops[t] * 100))
+        table.append(res)
+    output = render_table(template, head, sorted(table))
+    write_table(output, texfile)
+
 
 
 def build_benchmarks_table(csvfiles, texfile, template):
@@ -140,6 +159,8 @@ tables = {
             (['backend_summary.csv', 'resume_summary.csv'], build_backend_count_table),
         'ops_count_table.tex':
             (['summary.csv'], build_ops_count_table),
+        'guard_table.tex':
+            (['summary.csv'], build_guard_table),
         }
 
 

### 2012-08-01 talk/vmil2012/paper.tex
add a dummy abstract for length estimation

@@
 \begin{abstract}
-
+In pellentesque faucibus vestibulum. Nulla at nulla justo, eget luctus tortor.
+Nulla facilisi. Duis aliquet egestas purus in blandit. Curabitur vulputate,
+ligula lacinia scelerisque tempor, lacus lacus ornare ante, ac egestas est urna
+sit amet arcu. Class aptent taciti sociosqu ad litora torquent per conubia
+nostra, per inceptos himenaeos. Sed molestie augue sit amet leo consequat
+posuere. Vestibulum ante ipsum primis in faucibus orci luctus et ultrices
+posuere cubilia Curae; Proin vel ante a orci tempus eleifend ut et magna. Lorem
+ipsum dolor sit amet, consectetur adipiscing elit. Vivamus luctus urna sed urna
+ultricies ac tempor dui sagittis. In.
 \end{abstract}

### 2012-08-01 talk/vmil2012/paper.tex
extend and refactor the introduction

@@
 %\end{figure*}
-Although there are several publications about tracing just-in-time compilers, to
-our knowledge, there are none that describe the use and implementation of
-guards in this context. With the following contributions we aim to shed some
-light (to much?) on this topic.
-The contributions of this paper are:
-\todo{more motivation}
+Although there are several publications about tracing just-in-time compilers,
+to our knowledge, there are none that describe deoptimization and the use and
+implementation of guards in this context.
+
 Based on the informal observation that guards are among the most common
@@
 operations that are associated with an overhead to maintain information about
-state to be able to rebuild it, our goal is to present concrete numbers for the
-frequency and the overhead produced by guards, explain how they are implemented
-in the different levels of PyPy's tracing JIT and explain the rationale behind
-the design decisions based on the numbers.
-As can be seen on Figure~\ref{fig:ops_count} guards account for 14.42\% to
-22.32\% of the operations before and for 15.2\% to 20.12\% of after the
-optimization pass over the traced and compiled paths of the benchmarks.
-Figure~\ref{fig:benchmarks} shows the absolute number of operations for each
-benchmark, for every guard that stays alive after optimization there are
-several kinds of metadata created and stored at different levels of the JIT to
-be able to rebuild the interpreter or tracer state from a guard failure making
-the optimization \bivab{some good word} of guards an important aspect of the
-low-level design of a tracing just-in-time compiler.
-The contributions of this paper are
-In this paper we want to substantiate the aforementioned and describe based on
-them the reasoning behind and the implementation of guards in PyPy's tracing
-just-in-time compiler.
-\begin{figure}
-    \include{figures/guard_table}
-    \caption{Percentage of guards before and after optimization for different benchmarks}
-    \label{fig:guard_percent}
-\end{figure}
+state to be able to rebuild the execution state in case of deoptimization, our
+goal is to present concrete numbers for the frequency and the overhead produced
+by guards, explain how they are implemented in the different levels of PyPy's
+tracing JIT and explain the rationale behind the design decisions based on the
+numbers.
+
+The operations executed by an interpreter are recorded by the tracing JIT in
+case they are frequently executed, this process is described in more detail in
+Section~\ref{sec:Resume Data}, during the recording phase special operations,
+\texttt{guards}, are inserted into the recorded trace at all points where
+control flow could diverge. As can be seen on Figure~\ref{fig:guard_percent}
+guards account for 14.42\% to 22.32\% of the operations before and for 15.2\%
+to 20.12\% of the operations after the optimization pass over the traced and
+compiled parts of the benchmarks, making guards one of the most common
+operations. Many of these guards fail rarely on not all during execution. Given
+that associated with each guard information is stored, that is required to
+rebuild the execution state in case control flow diverges from the recorded
+path at a guard it is important to store the information associated with the
+guards in a manner that tries to keep the overhead for storing the information
+low while avoiding to put a burden on the execution of the recorded trace,
+making the optimization of guards an important aspect of
+the low-level design of a tracing just-in-time compiler.
+
+%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
+%operations for each benchmark, and the overhead produced by the information
+%stored at the different levels for the guards
+In this paper we want to substantiate the aforementioned observations and
+describe based on them the reasoning behind and the implementation of guards in
+PyPy's tracing just-in-time compiler, the contributions of this paper are:
 \begin{itemize}
@@
 \end{itemize}
+\begin{figure}
+    \include{figures/guard_table}
+    \caption{Percentage of guards before and after optimization for different benchmarks}
+    \label{fig:guard_percent}
+\end{figure}
 

### 2012-08-02 talk/vmil2012/paper.tex
add emails

@@
            }
-           {XXX emails}
+           {david.schneider@uni-duesseldorf.de \and cfbolz@gmx.de}
 

### 2012-08-03 talk/vmil2012/paper.tex
move figure to appendix

@@
 
-%\begin{figure*}
-%    \include{figures/ops_count_table}
-%    \caption{Relative numbers of operations in the traces generated for
-%    different benchmarks}
-%    \label{fig:ops_count}
-%\end{figure*}
 Although there are several publications about tracing just-in-time compilers,
@@
 
-After optimization the resulting trace is handed to the backend to be compiled
-to machine code. The compilation phase consists of two passes over the lists of
-instructions, a backwards pass to calculate live ranges of IR-level variables
-and a forward one to emit the instructions. During the forward pass IR-level
-variables are assigned to registers and stack locations by the register
-allocator according to the requirements of the to be emitted instructions.
-Eviction/spilling is performed based on the live range information collected in
-the first pass. Each IR instruction is transformed into one or more machine
-level instructions that implement the required semantics, operations withouth
-side effects whose result is not used are not emitted. Guards instructions are
-transformed into fast checks at the machine code level that verify the
-corresponding condition.  In cases the value being checked by the guard is not
-used anywhere else the guard and the operation producing the value can merged,
-reducing even more the overhead of the guard. Figure \ref{fig:trace-compiled}
-shows how an \texttt{int\_eq} operation followed by a guard that checks the
-result of the operation are compiled to pseudo-assembler if the operation and
-the guard are compiled separated or if they are merged.
+After optimization the resulting trace is handed to the over platform specific
+backend to be compiled to machine code. The compilation phase consists of two
+passes over the lists of instructions, a backwards pass to calculate live
+ranges of IR-level variables and a forward one to emit the instructions. During
+the forward pass IR-level variables are assigned to registers and stack
+locations by the register allocator according to the requirements of the to be
+emitted instructions.  Eviction/spilling is performed based on the live range
+information collected in the first pass. Each IR instruction is transformed
+into one or more machine level instructions that implement the required
+semantics, operations withouth side effects whose result is not used are not
+emitted. Guards instructions are transformed into fast checks at the machine
+code level that verify the corresponding condition.  In cases the value being
+checked by the guard is not used anywhere else the guard and the operation
+producing the value can merged, reducing even more the overhead of the guard.
+Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
+followed by a guard that checks the result of the operation are compiled to
+pseudo-assembler if the operation and the guard are compiled separated or if
+they are merged.
 
@@
 
-The following analysis is based on a selection of benchmarks taken from the set
-of benchmarks used to measure the performance of PyPy as can be seen
-on.\footnote{http://speed.pypy.org/} The benchmarks were taken from the PyPy benchmarks
-repository using revision
+The results presented in this section are based on numbers gathered by running
+a subset of the standard PyPy benchmarks. The PyPy benchmarks are used to
+measure the performance of PyPy and are composed of a series of
+micro-benchmarks and larger programs.\footnote{http://speed.pypy.org/} The
+benchmarks were taken from the PyPy benchmarks repository using revision
 \texttt{ff7b35837d0f}.\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}
 The benchmarks were run on a version of PyPy based on the
-tag~\texttt{release-1.9} and patched to collect additional data about the
+tag~\texttt{0b77afaafdd0} and patched to collect additional data about the
 guards in the machine code
-backends.\footnote{https://bitbucket.org/pypy/pypy/src/release-1.9} All
+backends.\footnote{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0} All
 benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
@@
 Figure~\ref{fig:benchmarks} shows the total number of operations that are
-recorded during tracing for each of the benchmarks on what percentage of these
-are guards. Figure~\ref{fig:benchmarks} also shows the number of operations left
-after performing the different trace optimizations done by the trace optimizer,
-such as xxx. The last columns show the overall optimization rate and the
-optimization rate specific for guard operations, showing what percentage of the
-operations was removed during the optimizations phase.
+recorded during tracing for each of the benchmarks and what percentage of these
+are guards. Figure~\ref{fig:benchmarks} also shows the number of operations
+left after performing the different trace optimizations done by the trace
+optimizer, such as xxx. The last columns show the overall optimization rate and
+the optimization rate specific for guard operations, showing what percentage of
+the operations were removed during the optimizations phase.
+Figure~\ref{fig:benchmarks} shows that as can also be seen on
+Figure~\ref{fig:guard_percent} the optimization rate for guards is on par with
+the average optimization rate for all operations in a trace. After optimization
+the amount of guards left in the trace still represents about 15.18\% to
+20.22\% of the operation, a bit less than before the optimization where guards
+represented between 15.85\% and 22.48\% of the operations. After performing the
+optimizations the most common operations are those that are difficult or
+impossible to optimize, such as JIT internal operations and different types of
+calls. These account for 14.53\% to 18.84\% of the operations before and for
+28.69\% to 46.60\% of the operations after optimization. These numbers show
+that about one fifth of the operations, making guards one of the most common
+operations, that are compiled are guards and have associated with them the
+high- and low-level datastructes that are reconstruct the state.
 
@@
 
-Figure~\ref{fig:backend_data} shows
-the total memory consumption of the code and of the data generated by the machine code
-backend for the different benchmarks mentioned above. Meaning the operations
-left after optimization take the space shown in Figure~\ref{fig:backend_data}
-after being compiled. Also the additional data stored for the guards to be used
-in case of a bailout and attaching a bridge.
+The overhead that is incurred by the JIT to manage the \texttt{resume data},
+the \texttt{low-level resume data} and the generated machine code is shown in
+Figure~\ref{fig:backend_data}. It shows the total memory consumption of the
+code and of the data generated by the machine code backend for the different
+benchmarks mentioned above. The size of the machine code is composed of the
+size of the compiled operations, the trampolines generated for the guards and a
+set of support functions that are generated when the JIT starts and are shared
+by all compiled traces. The size of the \texttt{low-level resume data} is the
+size of the registers and stack to IR-level variable mappings and finally the
+size of the \texttt{resume data} is an approximation of the size of the
+compressed high-level resume data. While the \texttt{low-level resume data} has
+a size of about 15\% to 20\% of the generated instructions the \texttt{resume
+data} is even in the compressed form larger than the generated machine code.
+
+Tracing JITs compilers only compile a subset of the executed program so the
+amount of generated machine code will be smaller than for function based JITs.
+At the same time there is a several times larger overhead for keeping the
+resume information for the guards. The generated machine code accounts for
+20.21\% to 37.97\% of the size required for storing the different kinds of
+resume data.
+
 \begin{figure*}
@@
 \section*{Acknowledgements}
+\section*{Appendix}
+\begin{figure*}
+    \include{figures/ops_count_table}
+    \caption{Relative numbers of operations in the traces generated for
+    different benchmarks}
+    \label{fig:ops_count}
+\end{figure*}
 \bibliographystyle{abbrv}

### 2012-08-06 talk/iwtc11/paper.tex
add all actionable reviewer comments to the paper

@@
 \newcommand\revc[1]{\nb{Reviewer 3}{#1}}
+\newcommand\revd[1]{\nb{Reviewer 4}{#1}}
 \newcommand{\commentout}[1]{}
@@
 
+\reva{
+You often use the word simple. While it might make sense to use it,
+it exact meaning in that context remains unclear.
+}
+
 One of the advantages that tracing JIT compilers have above traditional
@@
 
+\revc{
+Don't break code listings across pages, as at the start of section 3.  It makes
+them very hard to follow.
+}
+
 To motivate the approach we propose here, let's look at a trivial (unrealistic)
@@
 \label{sub:example}
+
+\reva{
+I think the motivation section is great, in particular for readers
+who are less familiar with compiler/JIT optimizations. However,
+section 4 starts with "yet another example" - at least this was my
+impression when reading it. I understand the differences and
+everything, but still, you might consider to improve the transition
+between sections 3 and 4.
+}
 
@@
 \section{Making Trace Optimizations Loop Aware}
+
+\revc{
+In general, the paper is over-long on generalities and too short on details.
+For example, the description of the basic technique at the beginning of section
+5 is the third time the idea is explained at basically the same level of detail
+(the others are in section 2 and section 4).  In contrast, the optimizations
+applied rely on a simple type analysis, but this is only briefly alluded to.
+}
 
@@
 
+\revc{
+This paper presents an elegant, if simple, technique, and demonstrates that
+it's effective in small cases.  The worked example is particularly helpful, and
+would be better if it were worked more thoroughly.  Some of the omitted steps
+are not entirely obvious, and the paper would be improved by making the
+clearer.  In particular, the final program presented on the bottom of page 5,
+first column, still has memory access, boxing, and type checks, which the paper
+then claims can be removed.  There's enough space to show this.
+}
+
 \section{Benchmarks}
+
+\revb{
+A nit: Section 7 says that loop peeling never makes runtime
+performance worse, but generating more code can potentially slow
+performance. I assume that non-numeric benchmarks show no slowdown in
+practice, and that might be worth noting.
+}
+
+\revb{
+Section 7 also mentions performance improvements for a Prolog
+interpreter. Consider adding a brief explanation of the benefit, since
+that example stands out as a non-numeric (I assume) performance
+improvement.
+}
+
+\revc{
+Providing source code for the benchmarks measured are needed for others to
+reproduce and build on your results.  I believe this should be the minimum
+standard for publishing measurements such as these.
+}
+
+\revc{
+I would have liked to have benchmark results for some larger applications.
+When is this optimization effective on a large scale, if ever?
+}
+
+\revd{
+It isn't clear from the paper, but a reader might conclude that the bulk of the
+time savings are from removing boxing/unboxing operations.
+}
+
+\revd{
+The benchmark results appear quite impressive -- especially the comparison with
+GCC -- but without additional information, I have no idea what is being
+compared.  Are these results from the same sizes of integers and/or floating
+point results?
+}
+
+\revd{
+This paper is relatively short, and could be significantly improved with a
+couple of pages of additional information about the details of the benchmarks
+-- both on the Python and on the C side.
+}
 
@@
 
+\revc{
+Include the OS with your benchmark specifications.
+}
+
 Benchmarks were run on Intel i7 M620 @2.67GHz with 4M cache and 8G of RAM in
@@
 
+\reva{
+First sentence of the related work section is kind of
+unfortunate. It is unclear what the reference at the end of the
+sentence is good for. To support the meaning of the entire sentence?
+Or is it just a reference to the standard loop invariant code motion
+techniques? The contribution of your paper seems much smaller than
+in the former case compared to the latter one. While I have not
+checked the content of the book, I believe the latter is the correct
+interpretation. You should remove this opportunity for
+misinterpretation.}
+
 The effect of combining a one pass optimization with loop peeling gives
@@
 think that our implementation scheme is a very simple one.
+
+\revc{
+The discussion of LuaJIT is unsatisfying.  It's not clear to me from that one
+quote that Mike is doing the same thing.  It might be worth including LuaJIT in
+the benchmarks, and/or examining the actual implementation of LuaJIT.
+}
 

### 2012-08-06 talk/dls2012/paper.tex
oops, add the review comments to the right version of the paper

@@
 
-Benchmarks were run on Intel i7 M620 @2.67GHz with 4M cache and 8G of RAM in
-32bit mode.
+Benchmarks were run on Intel i7 M620 @2.67GHz with 4M cache and 8G of RAM
+using Ubuntu Linux 11.4 in 32bit mode.
 The machine was otherwise unoccupied. We use the following software

### 2012-08-06 talk/dls2012/paper.tex
address the prolog comment

@@
 
-\revb{
-Section 7 also mentions performance improvements for a Prolog
-interpreter. Consider adding a brief explanation of the benefit, since
-that example stands out as a non-numeric (I assume) performance
-improvement.
-}
-
 \revc{
@@
 
-Other interesting interpreters that are helped greatly by this
-optimization are for
-example our Prolog interpreter written in RPython, as well as numerical
-kernel used for array manipulation. The exact extent is out of scope for
-this paper.
+Other interesting interpreters that are helped greatly by this optimization are
+for example our Prolog interpreter written in RPython
+\cite{carl_friedrich_bolz_towards_2010}. Prolog programs often contain tight
+loops that perform list processing. Furthermore we experimented with a language
+for writing numerical kernel used for array manipulation. The exact extent is
+out of scope for this paper.
 

### 2012-08-06 talk/dls2012/paper.tex
clarify

@@
 
-\reva{
-First sentence of the related work section is kind of
-unfortunate. It is unclear what the reference at the end of the
-sentence is good for. To support the meaning of the entire sentence?
-Or is it just a reference to the standard loop invariant code motion
-techniques? The contribution of your paper seems much smaller than
-in the former case compared to the latter one. While I have not
-checked the content of the book, I believe the latter is the correct
-interpretation. You should remove this opportunity for
-misinterpretation.}
-
-The effect of combining a one pass optimization with loop peeling gives
-completely standard loop invariant code motion optimizations
-\cite{muchnick_advanced_1997}. We do not claim any novelty in the effect, but
-think that our implementation scheme is a very simple one.
+Loop invariant code motion optimizations are completely
+standard~\cite{muchnick_advanced_1997}. Therefore, the effects that our
+optimization achieves is not in any way new. However, we think that achieving
+it as described in this paper is simpler than explicit algorithms.
 

### 2012-08-06 talk/dls2012/paper.tex
reduce the use of the word "simple"

@@
 
-\reva{
-You often use the word simple. While it might make sense to use it,
-it exact meaning in that context remains unclear.
-}
-
 One of the advantages that tracing JIT compilers have above traditional
@@
 produces only linear pieces of code without control flow joins, many
-optimization passes on traces can have a very simple structure. They often
-consist of one forward pass replacing operations by simpler ones or even
+optimization passes on traces can have a very simple structure: They often
+consist of one forward pass replacing operations by faster ones or even
 discarding them as they walk along it. This makes
@@
 
-One disadvantage of this simplicity is that such simple forward-passing
+One disadvantage of this simplicity is that such forward-passing
 optimizers ignore the only bit of control flow they have available, which is
@@
 
-In this paper we want to address this problem by proposing a simple scheme that
+In this paper we want to address this problem by proposing a scheme that
 makes it possible to turn optimizations using one forward pass into
@@
 However, we want to get this effect using our existing optimization passes
-without changing them too much. Simple optimizations with one forward pass
+without changing them too much. Optimizations with one forward pass
 cannot directly get this effect: They just look at the trace without taking
@@
 This scheme is quite powerful and generalizes to other optimizations than just
-common subexpression elimination. It allows simple linear optimization passes to
+common subexpression elimination. It allows linear optimization passes to
 perform loop-aware optimizations, such as loop-invariant code motion without
 changing them at all. All that is needed is to peel off one iteration, then
-apply simple one-pass optimizations and make sure that the necessary extra
+apply one-pass optimizations and make sure that the necessary extra
 arguments are inserted into the label of the loop itself and the jumps
@@
 For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
- a very simple object
+ a very small object
 model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
@@
 
-Let us now consider a simple ``interpreter'' function \lstinline{f} that uses the
+Let us now consider an ``interpreter'' function \lstinline{f} that uses the
 object model (see the bottom of Figure~\ref{fig:objmodel}).
@@
 repeated heap reads if the intermediate operations cannot have changed their
-value\footnote{We perform a simple type-based alias analysis to know which
-writes can affect which reads. In addition writes on newly allocated objects
+value\footnote{We perform a type-based alias analysis to know which
+writes can affect which reads \cite{XXX}. In addition writes on newly allocated objects
 can never change the value of old existing ones.}.
@@
 It is
-a simple straight forward implementation providing 2 dimensionall
+a straightforward implementation providing 2 dimensional
 indexing with out of bounds checks. For the C implementations it is

### 2012-08-06 talk/dls2012/paper.tex
clarify
improve transition

@@
 
-\reva{
-I think the motivation section is great, in particular for readers
-who are less familiar with compiler/JIT optimizations. However,
-section 4 starts with "yet another example" - at least this was my
-impression when reading it. I understand the differences and
-everything, but still, you might consider to improve the transition
-between sections 3 and 4.
-}
-
-For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
+The last section gave a motivating but unrealistically small example.
+This section will define a slightly larger example that the rest of the paper
+uses to demonstrate the effect of optimizations.
+For this we are going to use a tiny interpreter for a dynamic language with
  a very small object
@@
 one operation, \lstinline{add}, which adds two objects (promoting ints to floats in a
-mixed addition). The implementation of \lstinline{add} uses classical Smalltalk-like
+mixed addition). The implementation of \lstinline{add} uses classical
 double-dispatching.
@@
 \cite{carl_friedrich_bolz_towards_2010}. Prolog programs often contain tight
-loops that perform list processing. Furthermore we experimented with a language
-for writing numerical kernel used for array manipulation. The exact extent is
+loops that perform list processing. Furthermore we experimented with a Python library
+for writing numerical kernels doing array manipulation. The exact extent is
 out of scope for this paper.

### 2012-08-06 talk/dls2012/paper.tex
some typos

@@
 the end of itself. See Figure~\ref{fig:overview} for an illustration.
-The first part (called \emph{preamble}) finishes with a jump the the second part
+The first part (called \emph{preamble}) finishes with a jump to the second part
 (called the \emph{peeled loop}). The second part finishes with a jump to itself. This way
@@
 be used for every further iteration. New variable names have to be
-introduced in the entire copied trace in order to maintian the SSA-property.
+introduced in the entire copied trace in order to maintain the SSA-property.
 Note that the peeled loop is not necessary the \emph{first} iteration of the
@@
 \caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
-  $10^6$ and matrixes of size $1000\times 1000$ and $1000000 \times
+  $10^6$ and matrices of size $1000\times 1000$ and $1000000 \times
   3$ are used. The one used in each benchmark is indicated in
-  the leftmost column. For the matrixes, only the number of rows are
+  the leftmost column. For the matrices, only the number of rows are
   specified.} 
@@
 Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
-have developped the described technique independently. There are no papers about
+have developed the described technique independently. There are no papers about
 LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does

### 2012-08-06 talk/dls2012/paper.tex
use microtype
use footnote after full stop, protect space before citation

@@
 
+\usepackage{microtype}
 \usepackage{ifthen}
@@
 The work described in this paper was done in the context of the PyPy
-project\footnote{\texttt{http://pypy.org}}. PyPy is a framework for implementing
-dynamic languages efficiently \cite{armin_rigo_pypys_2006}. When implementing a
-language with PyPy, one writes an interpreter for the language in RPython
-\cite{davide_ancona_rpython:_2007}. RPython (``Restricted Python``) is a subset
+project.\footnote{\texttt{http://pypy.org}} PyPy is a framework for implementing
+dynamic languages efficiently~\cite{armin_rigo_pypys_2006}. When implementing a
+language with PyPy, one writes an interpreter for the language in RPython~\cite{davide_ancona_rpython:_2007}.
+RPython (``Restricted Python``) is a subset
 of Python chosen in such a way that it can be efficiently translated to a
@@
 implementation but are inserted during translation to C. Examples for this are a
-garbage collector and also a tracing JIT compiler \cite{bolz_tracing_2009}.
+garbage collector and also a tracing JIT compiler~\cite{bolz_tracing_2009}.
 
@@
  a very small object
-model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
+model, that just supports an integer and a float type (this example has been taken from a previous paper~\cite{bolz_allocation_2011}). The objects support only
 one operation, \lstinline{add}, which adds two objects (promoting ints to floats in a
@@
 repeated heap reads if the intermediate operations cannot have changed their
-value\footnote{We perform a type-based alias analysis to know which
-writes can affect which reads \cite{XXX}. In addition writes on newly allocated objects
-can never change the value of old existing ones.}.
+value.\footnote{We perform a type-based alias analysis to know which
+writes can affect which reads~\cite{XXX}. In addition writes on newly allocated objects
+can never change the value of old existing ones.}
 
@@
 
-PyPy's allocation removal optimization \cite{bolz_allocation_2011} makes it
+PyPy's allocation removal optimization~\cite{bolz_allocation_2011} makes it
 possible to identify objects that are allocated within the loop but never
@@
 vector of variables containing the values of all registered
-attributes\footnote{This is sometimes called \emph{scalar replacement}.}.
+attributes.\footnote{This is sometimes called \emph{scalar replacement}.}
 If some of the attributes are themselves references to
@@
 faster than either CPython or Psyco. This is due to the JIT compilation
-advantages and optimizations we discussed in previous work
-\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
+advantages and optimizations we discussed in previous
+work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
 speedup of loop peeling is 70\%, which makes benchmark times
@@
 Other interesting interpreters that are helped greatly by this optimization are
-for example our Prolog interpreter written in RPython
-\cite{carl_friedrich_bolz_towards_2010}. Prolog programs often contain tight
+for example our Prolog interpreter written in
+RPython~\cite{carl_friedrich_bolz_towards_2010}. Prolog programs often contain
+tight
 loops that perform list processing. Furthermore we experimented with a Python library
@@
 which allows reuse of all optimizations for redundancy elimination. Loop
-recurrences are detected on-the-fly and a minimized set of PHIs is generated.''
-\cite{pall_luajit_2009}
-
-Both the Hotpath VM \cite{gal_hotpathvm:_2006} and SPUR
-\cite{bebenita_spur:_2010} implements loop-invariant code motion
+recurrences are detected on-the-fly and a minimized set of PHIs is
+generated.''~\cite{pall_luajit_2009}
+
+Both the Hotpath VM~\cite{gal_hotpathvm:_2006} and
+SPUR~\cite{bebenita_spur:_2010} implements loop-invariant code motion
 directly, by explicitly marking as loop-invariant all variables that stay the
@@
 
-The type specialization described by Gal \etal \cite{gal_trace-based_2009} can
+The type specialization described by Gal \etal~\cite{gal_trace-based_2009} can
 be seen as doing a similar optimization (again by manually implementing it)

### 2012-08-06 talk/dls2012/paper.tex
replace some instances of "PyPy" with "RPython"

@@
 
-\section{Background: PyPy}
+\section{Background: RPython and PyPy}
 \label{sec:PyPy}
@@
 
-PyPy's tracing JIT compiler traces on the level of RPython programs. Thus it
+RPython's tracing JIT compiler traces on the level of RPython programs. Thus it
 actually traces the execution of an interpreter written in RPython, not of the
@@
 language transparent and optimizable by the tracing JIT. In the context of this
-paper, this aspect of PyPy's tracing JIT can be ignored. Instead, it is
-sufficient to view PyPy's tracing JIT as a JIT for RPython.
+paper, this aspect of RPython's tracing JIT can be ignored. Instead, it is
+sufficient to view RPython's tracing JIT as a JIT for RPython.
 
@@
 arguments. The \lstinline{print} operation just prints its argument (it is not
-an operation that PyPy's tracing JIT really supports, we just use it for this
+an operation that RPython's tracing JIT really supports, we just use it for this
 example). The \lstinline{jump} operation jumps back to the beginning of the
@@
 arguments, it only needs be executed the first time and then the result
-can be reused for all other appearances. PyPy's optimizers can also remove
+can be reused for all other appearances. RPython's optimizers can also remove
 repeated heap reads if the intermediate operations cannot have changed their
@@
 
-PyPy's allocation removal optimization~\cite{bolz_allocation_2011} makes it
+RPython's allocation removal optimization~\cite{bolz_allocation_2011} makes it
 possible to identify objects that are allocated within the loop but never
@@
 framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
-interpreters implemented within PyPy now can take advantage of
+interpreters implemented with RPython now can take advantage of
 it. Benchmarks have been executed for a few different interpreters and
@@
 \subsection{Python}
-The Python interpreter of the PyPy framework is a complete Python
+The Python interpreter of the RPython framework is a complete Python
 version 2.7 compatible interpreter. A set of numerical
@@
 comparable with native-compiled C code. We attribute the performance gap to C code to
-the relative immaturity of PyPy's JIT assembler backend as well as missing
+the relative immaturity of RPython's JIT assembler backend as well as missing
 optimizations, like instruction scheduling.

### 2012-08-06 talk/dls2012/paper.tex
some comments from me

@@
 }
+\cfbolz{let's do that only at the very end, just before submitting}
 
@@
 }
+\cfbolz{
+we have space now, can someone add the final optimized version of the loop?
+}
 
@@
 }
+\cfbolz{
+let's link to the bitbucket source code view. how about we move the benchmarks
+to the dls directory as well? or their own repository, we've been using them as
+demos
+}
 
@@
 }
+\cfbolz{I don't actually know. Does anybody?}
 
@@
 }
+\cfbolz{maybe we can look in the new LuaJIT wiki.
+how annoying would it be to rerun the benchmarks, if I can find somebody to write them?}
 

### 2012-08-06 talk/vmil2012/paper.tex
Rewrite paragraph about guard frequency

@@
 
-Figure~\ref{fig:benchmarks} shows the total number of operations that are
+\todo{a description about what each selected benchmark does}
+
+From the mentioned benchmarks we collected different datasets to evaluate the
+Frequency, the overhead and overall behaviour of guards.
+Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
 recorded during tracing for each of the benchmarks and what percentage of these
-are guards. Figure~\ref{fig:benchmarks} also shows the number of operations
-left after performing the different trace optimizations done by the trace
-optimizer, such as xxx. The last columns show the overall optimization rate and
-the optimization rate specific for guard operations, showing what percentage of
-the operations were removed during the optimizations phase.
-Figure~\ref{fig:benchmarks} shows that as can also be seen on
-Figure~\ref{fig:guard_percent} the optimization rate for guards is on par with
-the average optimization rate for all operations in a trace. After optimization
-the amount of guards left in the trace still represents about 15.18\% to
-20.22\% of the operation, a bit less than before the optimization where guards
-represented between 15.85\% and 22.48\% of the operations. After performing the
-optimizations the most common operations are those that are difficult or
-impossible to optimize, such as JIT internal operations and different types of
-calls. These account for 14.53\% to 18.84\% of the operations before and for
-28.69\% to 46.60\% of the operations after optimization. These numbers show
-that about one fifth of the operations, making guards one of the most common
-operations, that are compiled are guards and have associated with them the
-high- and low-level datastructes that are reconstruct the state.
+operations are guards. The number of operations was counted on the unoptimized
+and optimized traces. Showing that the overall optimization rate is between
+65.80\% and 86.23\% of all operations and that the optimization rate for guards
+is similar to the general one, as could be assumed based on
+Figure~\ref{fig:guard_percent}. These numbers show that guards are a rather
+common operation in the traces, which is a reason the put effort into
+optimizing them.
+\todo{some pie charts about operation distribution}
 

### 2012-08-06 talk/vmil2012/paper.tex
~add a pie chart to the appendix showing the distribution of operations before and after optimization (hmmmm pie)
refactor another evaluation paragraph and mark pending tasks for the evaluation section

@@
 The overhead that is incurred by the JIT to manage the \texttt{resume data},
-the \texttt{low-level resume data} and the generated machine code is shown in
-Figure~\ref{fig:backend_data}. It shows the total memory consumption of the
-code and of the data generated by the machine code backend for the different
-benchmarks mentioned above. The size of the machine code is composed of the
-size of the compiled operations, the trampolines generated for the guards and a
-set of support functions that are generated when the JIT starts and are shared
-by all compiled traces. The size of the \texttt{low-level resume data} is the
-size of the registers and stack to IR-level variable mappings and finally the
-size of the \texttt{resume data} is an approximation of the size of the
-compressed high-level resume data. While the \texttt{low-level resume data} has
-a size of about 15\% to 20\% of the generated instructions the \texttt{resume
-data} is even in the compressed form larger than the generated machine code.
+the \texttt{low-level resume data} as well as the generated machine code is
+shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
+of the code and of the data generated by the machine code backend and an
+approximation of the size of the \texttt{resume data} structures for the
+different benchmarks mentioned above. The size of the machine code is composed
+of the size of the compiled operations, the trampolines generated for the
+guards and a set of support functions that are generated when the JIT starts
+and are shared by all compiled traces. The size of the \texttt{low-level resume
+data} is the size of the compressed mapping from registers and stack to
+IR-level variable and finally the size of the \texttt{resume data} is an
+approximation of the size of the compressed high-level resume data\todo{explain
+why it is an approximation}.
 
-Tracing JITs compilers only compile a subset of the executed program so the
-amount of generated machine code will be smaller than for function based JITs.
-At the same time there is a several times larger overhead for keeping the
-resume information for the guards. The generated machine code accounts for
-20.21\% to 37.97\% of the size required for storing the different kinds of
-resume data.
+Compared to the size of the generated machine code the compressed
+\texttt{low-level resume data} is about 15\% to 20\% of that size, depending on
+the benchmark. On the other hand the generated machine code has only a size
+ranging from 20.21\% to 37.98\% of the size of the high and low-level
+\texttt{resume data} being compressed as described before.
+
+Tracing JIT compilers only compile the subset of the code executed in a program
+that is traced in a hot loop, for this reason the amount of generated machine
+code will be smaller than in other juts-in-time compilation approaches. Still
+the overhead associated to guards to resume execution from a side exit appears
+to be high.\bivab{put into relation to other JITs, compilers in general}
 
@@
 
-* Evaluation
-   * Measure guard memory consumption and machine code size
-   * Extrapolate memory consumption for guard other guard encodings
-      * compare to naive variant
-   * Measure how many guards survive optimization
-   * Measure the of guards and how many of these ever fail
+\todo{compare to naive variant of resume data}
+\todo{Measure the of guards and how many of these ever fail}
 
@@
 \end{figure*}
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
+\caption{Relative frequency of operations before and after optimization}
+\label{fig:ops_pie}
+\end{figure*}
 \bibliographystyle{abbrv}

### 2012-08-06 talk/vmil2012/paper.tex
typo

@@
 into one or more machine level instructions that implement the required
-semantics, operations withouth side effects whose result is not used are not
+semantics, operations without side effects whose result is not used are not
 emitted. Guards instructions are transformed into fast checks at the machine
@@
 corresponding to each IR-variable required by the guard will be stored when
-execution reaches the code emitted for the corresponding guard. \bivab{go into
-more detail here?!} This encoding needs to be as compact as possible to
-maintain an acceptable memory profile.
-
-\todo{example for low-level resume data showing how the current encoding works?}
+execution reaches the code emitted for the corresponding guard. This data
+structure stores the data in a compressed manner using an encoding the uses
+8bits to store 7bits of information. This encoding is efficient to create and
+provides a compact representation of the needed information. This encoding
+needs to be as compact as possible to maintain an acceptable memory profile.
 

### 2012-08-06 talk/vmil2012/paper.tex
for reference, the version that we submitted

@@
 into one or more machine level instructions that implement the required
-semantics, operations without side effects whose result is not used are not
+semantics, operations withouth side effects whose result is not used are not
 emitted. Guards instructions are transformed into fast checks at the machine

### 2012-08-07 talk/dls2012/paper.tex
typos

@@
 \begin{abstract}
-One of the nice properties of a tracing JIT is that many of its optimization
+One of the nice properties of a tracing JIT is that many of its optimizations
 are simple requiring one forward pass only. This is not true for loop-invariant code
 motion which is a very important optimization for code with tight kernels.
-Especially for dynamic languages that typically performs quite a lot of loop invariant
+Especially for dynamic languages that typically perform quite a lot of loop invariant
 type checking, boxed value unwrapping and virtual method lookups.
@@
 For tight computationally intensive loops a
-significant amount of the execution time might be spend on such tasks
+significant amount of the execution time might be spent on such tasks
 instead of the actual computations. Moreover, the type checking,
@@
 
-This is the key insight of the proposed implementation scheme: Giving an
-optimization two iterations together at the same time gives the optimization
-enough context to remove operations from the peeled loop, because it detects
+This is the key insight of the proposed implementation scheme: If an
+optimization is given two iterations together at the same time, the
+optimization has enough context to remove operations from the peeled loop,
+because it detects
 that the operation was performed in the preamble already. Thus at runtime these
@@
 operations out of the loop making them executed only once
-and not every iteration. This we propose to achieve by loop peeling. It
+and not every iteration. We propose to achieve this by loop peeling. It
 leaves the loop body intact, but prefixes it with one iteration of the
@@
 combined with other optimizations it can increase the effectiveness of
-those optimizations. For many optimization of interest only a few
-additional details has to be considered when they are combined with loop peeling. These are
+those optimizations. For many optimizations of interest only a few
+additional details have to be considered when they are combined with loop peeling. These are
 described below by explaining the loop peeling optimization
@@
 
-No special concerns needs to be taken when implementing redundant
+No special concern needs to be taken when implementing redundant
 guard removal together with loop peeling. The guards from

### 2012-08-07 talk/dls2012/paper.tex
I'm *pretty* sure that this needs to be J, not I
some more typos, and three references

@@
 If a pure operation appears more than once in the trace with the same input
-arguments, it only needs be executed the first time and then the result
+arguments, it only needs to be executed the first time and then the result
 can be reused for all other appearances. RPython's optimizers can also remove
@@
 replace $i_7$ with $i_3$. Afterwards the trace is no longer in the correct
-form, because the argument $i_3$ is not passed along the loop arguments. It
-thus needs to be added there.
-
-The trace from Figure~\ref{fig:peeled-trace} will therefore be optimized to:
+form, because the argument $i_3$ is not passed along the loop arguments.
+Therefore $i_3$ needs to be added to the loop arguments.
+
+Doing this, the trace from Figure~\ref{fig:peeled-trace} will be optimized to:
 
@@
 \end{equation}
-In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
+In the optimized trace $J$ is replaced by $\hat J$ and $K$ by $\hat
 K$.
@@
 \lstinline{new} operation is inserted at this point. All operations
-(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
+(\lstinline{get}, \lstinline{set} and \lstinline{guard_class}) on the removed objects
 are also removed and the optimizer needs to keep track of the value of all used
@@
 vector of variables containing the values of all registered
-attributes.\footnote{This is sometimes called \emph{scalar replacement}.}
+attributes.\footnote{This is sometimes called \emph{scalar replacement}~\cite{kotzmann_escape_2005}.}
 If some of the attributes are themselves references to
@@
 future. In particular loop peeling works poorly in combination with trace
-trees or trace stitching. The side exits attached guards that fail often
+trees~\cite{andreas_gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
+The side exits attached guards that fail often
 currently have to jump to the preamble which makes loops with several equally

### 2012-08-07 talk/dls2012/paper.tex
typos

@@
 standard~\cite{muchnick_advanced_1997}. Therefore, the effects that our
-optimization achieves is not in any way new. However, we think that achieving
-it as described in this paper is simpler than explicit algorithms.
+optimization achieves are not in any way new. However, we think that achieving
+them in the way described in this paper is simpler than writing explicit algorithms.
 

### 2012-08-07 talk/dls2012/paper.tex
try to make this point a bit clearer

@@
 introduced in the entire copied trace in order to maintain the SSA-property.
-Note that the peeled loop is not necessary the \emph{first} iteration of the
-loop execution, it is general enough to correspond to any iteration of the loop.
+
+When peeling the loop, no assumptions are made that the preamble is
+the \emph{first} iteration when later executing the loop. The preamble stays
+general enough to correspond to any iteration of the loop.
 However, the peeled loop can then be optimized using the assumption that a
-previous iteration has happened.
+previous iteration (the preamble) has been executed already.
 

### 2012-08-07 talk/dls2012/paper.tex
add the final version of the code

@@
 
-If all the optimizations presented above are applied, the resulting
-optimized peeled loop will consist of a single integer addition
+If all the optimizations presented above are applied, the resulting loop looks
+as in Figure~\ref{fig:opt-trace}.
+The resulting optimized peeled loop consists of a single integer addition
 only. That is it will become type-specialized to the types of the
@@
 
-\revc{
-This paper presents an elegant, if simple, technique, and demonstrates that
-it's effective in small cases.  The worked example is particularly helpful, and
-would be better if it were worked more thoroughly.  Some of the omitted steps
-are not entirely obvious, and the paper would be improved by making the
-clearer.  In particular, the final program presented on the bottom of page 5,
-first column, still has memory access, boxing, and type checks, which the paper
-then claims can be removed.  There's enough space to show this.
-}
-\cfbolz{
-we have space now, can someone add the final optimized version of the loop?
-}
+\begin{figure}
+\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
+$L_0$($p_{0}$, $p_{1}$):
+# inside f: y = y.add(step)
+guard_class($p_{1}$, BoxedInteger)
+    # inside BoxedInteger.add
+    $i_{2}$ = get($p_{1}$, intval)
+    guard_class($p_{0}$, BoxedInteger)
+        # inside BoxedInteger.add__int
+        $i_{3}$ = get($p_{0}$, intval)
+        $i_{4}$ = $i_{2}+i_{3}$
+            # inside BoxedInteger.__init__
+jump($L_1$, $p_{0}$, $i_{4}$)
+
+$L_1$($p_{0}$, $i_{3}$, $i_{4}$):
+    $i_{8}$ = $i_{4}+i_{3}$
+jump($L_1$, $p_{0}$, $i_{3}$, $i_8$)
+\end{lstlisting}
+\caption{The fully optimized loop of the Example Interpreter}
+\label{fig:opt-trace}
+\end{figure}
 

### 2012-08-07 talk/dls2012/paper.tex
make all traces figures

@@
 guard removal with loop peeling is that loop-invariant guards are moved out of the
-loop. The peeled loop of the example reduces to
-
+loop. The peeled loop of the example reduces to the trace in Figure~\ref{fig:guard-trace}.
+
+\begin{figure}
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
@@
 \end{lstlisting}
+\caption{Peeled loop after redundant guard removal}
+\label{fig:guard-trace}
+\end{figure}
 
@@
 
-Doing this, the trace from Figure~\ref{fig:peeled-trace} will be optimized to:
-
+Doing this, the trace from Figure~\ref{fig:peeled-trace} will be optimized to
+the trace in Figure~\ref{fig:cse-trace}.
+
+\begin{figure}
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
@@
 \end{lstlisting}
+\caption{Trace after common subexpression elimination}
+\label{fig:cse-trace}
+\end{figure}
 
@@
 
-\subsection{Allocation Removals}
+\subsection{Allocation Removal}
 \label{sub:allocation}
@@
 In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
-K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into
-
+K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to
+the trace in Figure~\ref{fig:virtual-trace}.
+
+\begin{figure}
 \begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
@@
 \end{lstlisting}
+\caption{Trace after allocation removal}
+\label{fig:virtual-trace}
+\end{figure}
 

### 2012-08-07 talk/vmil2012/paper.tex
add line numbers (puh)

--- 2012-08-06 talk/vmil2012/paper.tex
@@
 operations that are associated with an overhead to maintain information about
-state to be able to rebuild the execution state in case of deoptimization, our
-goal is to present concrete numbers for the frequency and the overhead produced
-by guards, explain how they are implemented in the different levels of PyPy's
+the execution state to be able to rebuild it in case of deoptimization, our
+goal is to present concrete numbers for the frequency and the overhead related
+to guards, explain how they are implemented in the different levels of PyPy's
 tracing JIT and explain the rationale behind the design decisions based on the
-numbers.
+numbers provided here.
 
@@
 Section~\ref{sec:Resume Data}, during the recording phase special operations,
-\texttt{guards}, are inserted into the recorded trace at all points where
-control flow could diverge. As can be seen on Figure~\ref{fig:guard_percent}
-guards account for 14.42\% to 22.32\% of the operations before and for 15.2\%
-to 20.12\% of the operations after the optimization pass over the traced and
-compiled parts of the benchmarks, making guards one of the most common
-operations. Many of these guards fail rarely on not all during execution. Given
-that associated with each guard information is stored, that is required to
-rebuild the execution state in case control flow diverges from the recorded
-path at a guard it is important to store the information associated with the
-guards in a manner that tries to keep the overhead for storing the information
-low while avoiding to put a burden on the execution of the recorded trace,
-making the optimization of guards an important aspect of
-the low-level design of a tracing just-in-time compiler.
+referred to as \texttt{guards}, are inserted into the recorded trace at all
+points where the control flow could diverge. As can be seen in
+Figure~\ref{fig:guard_percent} guards account for 14.42\% to 22.32\% of the
+operations before and for 15.2\% to 20.12\% of the operations after the
+optimization pass over the traced and later compiled parts of the benchmarks,
+making guards one of the most common types of operations. Many of these guards
+fail rarely or not all during execution. There are several aspects to consider
+in the design and optimization of guards, the first aspect is that due to the
+large number of guards the memory overhead related to storing the information
+needed for deoptimization should be kept low. A second aspect is that
+successfully checking guards, i.e. not leaving the compiled trace,  - which is
+the common case - should be a cheap operation to execute favouring the on-trace
+execution speed in contrast to the deoptimization case where the state has to
+be rebuilt using the stored information. These constraints and trade-offs are
+what make the design and optimization of guards an important and non-trivial
+aspect of the low-level design of a tracing just-in-time compiler.
 
@@
 
-\todo{set line numbers to the line numbers of the rpython example}
 \begin{figure}

### 2012-08-07 talk/vmil2012/paper.tex
typos

@@
 
-\subsection{PyPy's Meta-Tracing JIT Compilers}
+\subsection{RPython's Meta-Tracing JIT Compilers}
 \label{sub:tracing}
@@
 
-Once the bridge has been compiled the guard that led to compiling the birdge is
+Once the bridge has been compiled the guard that led to compiling the bridge is
 patched to redirect control flow to the bridge in case the check fails. In

### 2012-08-07 talk/vmil2012/paper.tex
add description of benchmarks

@@
 \usepackage{setspace}
+\usepackage[pdfpagelabels=true]{hyperref}
 \usepackage[colorinlistoftodos]{todonotes}
@@
 
-\todo{a description about what each selected benchmark does}
+We used the following benchmarks:
+
+\begin{description}
+    \item[chaos:] A Chaosgame implementation creating a fractal.
+    \item[crypto\_pyaes:] An AES implementation.
+    \item[django:] The templating engine of the Django Web
+        framework\footnote{\url{http://www.djangoproject.com/}}.
+
+    \item[go:] A Monte-Carlo Go
+        AI\footnote{\url{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
+    \item[pyflate\_fast:] A BZ2 decoder.
+    \item[raytrace\_simple:] A ray tracer.
+    \item[richards:] The Richards benchmark.
+    \item[spambayes:] A Bayesian spam filter\footnote{\url{http://spambayes.sourceforge.net/}}.
+    \item[simpy\_expand:] A computer algebra system.
+    \item[telco:] A Python version of the Telco decimal
+        benchmark\footnote{\url{http://speleotrove.com/decimal/telco.html}},
+        using a pure Python decimal floating point implementation.
+    \item[twisted\_names:] A DNS server benchmark using the Twisted networking
+        framework\footnote{\url{http://twistedmatrix.com/}}.
+\end{description}
 

### 2012-08-07 talk/vmil2012/paper.tex
some tweaks to the related work

@@
 
-In~\cite{Gal:2006} Gal et. al describe that in the HotpathVM they experimented
+Gal et. al~\cite{Gal:2006} describe that in the HotpathVM they experimented
 with having one generic compensation code block, like the RPython JIT, that
@@
 
-Gal et. al~\cite{Gal:2009ux} write about how TraceMonkey uses trace stitching
-to avoid th overhead of returning to the trace monitor and calling another
-trace when taking a side exit. In their approach it is required to write live
+TraceMonkey, a tracing JIT for JavaScript, uses trace stitching
+to avoid the overhead of returning to the trace monitor and calling another
+trace when taking a side exit~\cite{Gal:2009ux}. In this approach it is required to write live
 values to an activation record before entering the new trace.

### 2012-08-07 talk/vmil2012/paper.tex

--- 2012-08-07 talk/vmil2012/paper.tex
@@
 
-\subsection{RPython's Meta-Tracing JIT Compilers}
+\subsection{PyPy's Meta-Tracing JIT Compilers}
 \label{sub:tracing}
@@
 
+\todo{set line numbers to the line numbers of the rpython example}
 \begin{figure}
@@
 
-Once the bridge has been compiled the guard that led to compiling the bridge is
+Once the bridge has been compiled the guard that led to compiling the birdge is
 patched to redirect control flow to the bridge in case the check fails. In

### 2012-08-07 talk/vmil2012/paper.tex
a note
s/PyPy/RPython in some places

--- 2012-08-07 talk/vmil2012/paper.tex
@@
 Based on the informal observation that guards are among the most common
-operations in the traces produced by PyPy's tracing JIT and that guards are
+operations in the traces produced by RPython's tracing JIT and that guards are
 operations that are associated with an overhead to maintain information about
@@
 goal is to present concrete numbers for the frequency and the overhead related
-to guards, explain how they are implemented in the different levels of PyPy's
+to guards, explain how they are implemented in the different levels of RPython's
 tracing JIT and explain the rationale behind the design decisions based on the
@@
 describe based on them the reasoning behind and the implementation of guards in
-PyPy's tracing just-in-time compiler, the contributions of this paper are:
+RPython's tracing just-in-time compiler, the contributions of this paper are:
 \begin{itemize}
-  \item An analysis of guards in the context of PyPy's tracing JIT to
+  \item An analysis of guards in the context of RPython's tracing JIT to
   substantiate the aforementioned observation, based on a set of benchmarks.
@@
 and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
-Data} we proceed to describe for PyPy's tracing JIT the details of guards in
+Data} we proceed to describe for RPython's tracing JIT the details of guards in
 the frontend\bivab{better term for this?} related to recording and storing the
@@
 fix website citation} Pall explains that guards in LuaJIT use a datastucture
-called snapshots, similar to PyPy's resume data, to store the information about
+called snapshots, similar to RPython's resume data, to store the information about
 how to rebuild the state from a side-exit using the information in the snapshot
@@
 The paper does not describe any attempts to store this information compactly.
+This may not be needed in their approach, because method-based JITs have a lot
+fewer deoptimization points than tracing JITs.
 

### 2012-08-07 talk/vmil2012/paper.tex
urlify links, don't color them

@@
   numbersep = -20pt,
+}
+
+\hypersetup{%
+  plainpages=false,%
+  %hyperfootnotes=false,%
+  colorlinks=true,%
+  urlcolor=black,%
+  citecolor=black,%
+  linkcolor=black,%
+  pdftitle={Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT},%
+  pdfauthor={David Schneider},
 }
@@
 
-\title{Efficiently Handling Guards in the Low Level Design of RPython's tracing JIT}
+\title{Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT}
 
@@
 measure the performance of PyPy and are composed of a series of
-micro-benchmarks and larger programs.\footnote{http://speed.pypy.org/} The
+micro-benchmarks and larger programs.\footnote{\url{http://speed.pypy.org/}} The
 benchmarks were taken from the PyPy benchmarks repository using revision
-\texttt{ff7b35837d0f}.\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}
+\texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
 The benchmarks were run on a version of PyPy based on the
@@
 guards in the machine code
-backends.\footnote{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0} All
+backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} All
 benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with

### 2012-08-07 talk/vmil2012/paper.tex
argh

@@
     \item[django:] The templating engine of the Django Web
-        framework\footnote{\url{http://www.djangoproject.com/}}.
+        framework.\footnote{\url{http://www.djangoproject.com/}}
 
     \item[go:] A Monte-Carlo Go
-        AI\footnote{\url{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
+        AI.\footnote{\url{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}
     \item[pyflate\_fast:] A BZ2 decoder.
@@
     \item[richards:] The Richards benchmark.
-    \item[spambayes:] A Bayesian spam filter\footnote{\url{http://spambayes.sourceforge.net/}}.
+    \item[spambayes:] A Bayesian spam filter.\footnote{\url{http://spambayes.sourceforge.net/}}
     \item[simpy\_expand:] A computer algebra system.
     \item[telco:] A Python version of the Telco decimal
-        benchmark\footnote{\url{http://speleotrove.com/decimal/telco.html}},
+        benchmark,\footnote{\url{http://speleotrove.com/decimal/telco.html}}
         using a pure Python decimal floating point implementation.
     \item[twisted\_names:] A DNS server benchmark using the Twisted networking
-        framework\footnote{\url{http://twistedmatrix.com/}}.
+        framework.\footnote{\url{http://twistedmatrix.com/}}
 \end{description}

### 2012-08-07 talk/vmil2012/paper.tex
remove a space from the URL
remove a todo and make the url of the Lua mailing list post show up in the references

@@
     \item[go:] A Monte-Carlo Go
-        AI.\footnote{\url{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}
+        AI.\footnote{\url{http://shed-skin.blogspot.com/2009/07/disco-elegant-python-go-player.html}}
     \item[pyflate\_fast:] A BZ2 decoder.
@@
 list different technologies and techniques used in the implementation of
-LuaJIT~\cite{Pall:2009}.\todo{decide if LuaJIT is a footnote or a reference and
-fix website citation} Pall explains that guards in LuaJIT use a datastucture
+LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
 called snapshots, similar to RPython's resume data, to store the information about

### 2012-08-07 talk/vmil2012/paper.tex
(cfbolz, bivab) write section about RPython's tracing JIT

@@
 
-\subsection{RPython's Meta-Tracing JIT Compilers}
+\subsection{RPython's Tracing JIT Compilers}
 \label{sub:tracing}
+Tracing JITs are a technique of just-in-time compilers that generate code by
+observing the execution of a program. VMs using tracing JITs are typically
+mixed mode execution environments containing also an interpreter. The
+interpreter profiles the executed program and selects frequently executed code
+paths to be compiled to machine code. After profiling identified an interesting
+path, tracing is started, recording all operations that are executed on this
+path. Like in most compilers tracing JITs use an intermediate representation
+to store the recorded operations, which is typically in SSA form\todo{some ssa
+reference}. Since tracing follows actual execution the code that is recorded
+represents only one possible path through the control flow graph. Points of
+divergence from the recorded path are marked with special operations called
+\emph{guards}, these operations ensure that assumptions valid during the
+tracing phase are still valid when the code has been compiled and is executed.
+After a trace has been recorded it is optimized and then compiled to platform
+specific machine code.
 
- * Tracing JITs
- * Mention SSA
- * JIT Compiler
-   * describe the tracing jit stuff in pypy
-   * reference tracing the meta level paper for a high level description of what the JIT does
-   * JIT Architecture
-   * Explain the aspects of tracing and optimization
+When the check of a guard fails, the execution of the machine code must be
+stopped and the control is returned to the interpreter, after the interpreter's
+state has been restored. If a particular guard fails often a new trace is
+recorded starting from the guard. We will refer to this kind of trace as a
+\emph{bridge}. Once a bridge has been traced it is attached to the
+corresponding guard by patching the machine code. The next time the guard fails
+the bridge will be executed instead of leaving the machine code.
+
+RPython provides a tracing JIT that can be reused for a number of language
+implementations. This is possible, because it traces the execution of the
+language interpreter instead of tracing the user program directly. This
+approach is called \emph{meta-tracing}. For the purpose of this paper the fact
+that RPython's tracing JIT is a meta-tracing JIT can be ignored.
 

### 2012-08-07 talk/vmil2012/paper.tex
keywords and another todo
(cfbolz, bivab) write about bridges and trace stitching

@@
 
-\keywords{XXX}
+\keywords{tracing JIT, guards, deoptimization}
 
 \begin{abstract}
+\todo{write}
 In pellentesque faucibus vestibulum. Nulla at nulla justo, eget luctus tortor.
@@
 is to construct the tracer state.
+After the bridge has been recorded and compiled it is attached to the guard.
+If the guard fails later, the bridge is executed. Therefore the resume data of
+that guard is no longer needed.
 
@@
 
-% subsection Interaction With Optimization (end)
-\subsection{Compiling Side-Exits and Trace Stitching} % (fold)
-\label{sub:Compiling side-exits and trace stitching}
-   * tracing and attaching bridges and throwing away resume data
-   * restoring the state of the tracer
-     * keeping virtuals
-   * compiling bridges
-\todo{maybe mention that the failargs also go into the bridge}
-
-% subsection Compiling side-exits and trace stitching (end)
 % section Resume Data (end)
@@
 \end{figure}
-%* Low level handling of guards
-%   * Fast guard checks v/s memory usage
-%   * memory efficient encoding of low level resume data
-%   * fast checks for guard conditions
-%   * slow bail out
-%
 % section Guards in the Backend (end)

### 2012-08-07 talk/vmil2012/paper.tex
todo
(cfbolz, bivab) abstract

@@
 \begin{abstract}
-\todo{write}
-In pellentesque faucibus vestibulum. Nulla at nulla justo, eget luctus tortor.
-Nulla facilisi. Duis aliquet egestas purus in blandit. Curabitur vulputate,
-ligula lacinia scelerisque tempor, lacus lacus ornare ante, ac egestas est urna
-sit amet arcu. Class aptent taciti sociosqu ad litora torquent per conubia
-nostra, per inceptos himenaeos. Sed molestie augue sit amet leo consequat
-posuere. Vestibulum ante ipsum primis in faucibus orci luctus et ultrices
-posuere cubilia Curae; Proin vel ante a orci tempus eleifend ut et magna. Lorem
-ipsum dolor sit amet, consectetur adipiscing elit. Vivamus luctus urna sed urna
-ultricies ac tempor dui sagittis. In.
+Guards operations occur frequently in traces generated by tracing just-in-time
+(JIT) compilers. Therefore it is important to design and implement them
+carefully to find the right trade-off between execution speed, deoptimization,
+and memory overhead.  In this paper we describe the design decisions about
+guards taken in the implementation of the RPython tracing JIT. Furthermore we
+measure various properties of guards.
+% \o/
 \end{abstract}
@@
 
+\todo{explain example}
 %___________________________________________________________________________

### 2012-08-08 talk/vmil2012/paper.tex
remove a redundant todo
some improvements

@@
 The reason for this is that during tracing only the variables
-of the currently executing frames can change.
+of the currently executing frame can change.
 Therefore if two guards are generated from code in the same function
-the resume data of the rest of the stack can be reused.
+the resume data of the rest of the frame stack can be reused.
 
@@
 the second one can be removed.
-This also works if a later guard is weaker implied by a earlier guard.
+This also works if a later guard is weaker
+and therefore implied by a earlier guard.
 
@@
 The virtual objects in the symbolic frames describe exactly
-how the heap objects that have to be allocated on guard failure look like.
+how the heap objects look like which have to be allocated on guard failure.
 To this end, the content of every field of the virtual object is described
@@
 
-Similarly, stores into the heap are delayed as long as possible.
+A related optimization is the handling of heap stores by the optimizer.
+The optimizer tries to delay stores into the heap as long as possible.
+This is done because often heap stores become unnecessary
+due to another store to the same memory location later in the trace.
 This can make it necessary to perform these delayed stores
@@
 of the delayed stores to be able to perform them when the guard fails.
-So far no special compression is done with this information.
+So far no special compression is done with this information,
+compared to the other source of information delayed heap stores are quite rare.
 
@@
 \todo{compare to naive variant of resume data}
-\todo{Measure the of guards and how many of these ever fail}
 

### 2012-08-08 talk/vmil2012/tool/build_tables.py
(cfbolz, bivab) generate and embed table about resume data sizes

@@ -15,6 +15,22 @@ def getlines(csvfile):
         return [l for l in reader]
 
 
+def build_resume_data_table(csvfiles, texfile, template):
+    assert len(csvfiles) == 1
+    lines = getlines(csvfiles[0])
+    table = []
+    head = ['Benchmark', 'compressed', 'naive', 'xz compressed']
+
+    for bench in lines:
+        res = [bench['bench'].replace('_', '\\_'),
+                "%.2f" % float(bench['total resume data size']),
+                "%.2f" % float(bench['naive resume data size']),
+                "%.2f" % float(bench['compressed resume data size']),
+        ]
+        table.append(res)
+    output = render_table(template, head, sorted(table))
+    write_table(output, texfile)
+
 def build_ops_count_table(csvfiles, texfile, template):
     assert len(csvfiles) == 1
     lines = getlines(csvfiles[0])
@@ -161,6 +177,8 @@ tables = {
             (['summary.csv'], build_ops_count_table),
         'guard_table.tex':
             (['summary.csv'], build_guard_table),
+        'resume_data_table.tex':
+            (['resume_summary.csv'], build_resume_data_table),
         }
 
 

### 2012-08-08 talk/dls2012/paper.tex
add acknowledgements
start expanding the explanation of the benchmarks

@@
 \begin{itemize}
-\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
-  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
-  2$. There are three different versions of this benchmark where $x_k$
+\item {\bf sqrt}: approximates the square root of $y$. The approximation is 
+initiated to $x_0=y/2$ and the benchmark consists of a single loop updating this
+approximation using $x_i = \left( x_{i-1} + y/x_{i-1} \right) / 2$ for $1\leq i < 10^8$. 
+Only the latest calculated value $x_i$ is kept alive as a local variable within the loop.
+There are three different versions of this benchmark where $x_i$
   is represented with different type of objects: int's, float's and
@@
   there are three different implementations.
-\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
-\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
+\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$. A single loop
+is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_n\right)$ from a vector
+${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
+$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n$. Both the output vector, $\bf b$, 
+and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
+with $n=10^5$ and $n=10^6$.
+\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
+${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
+hardcoded into the implementation making the benchmark consist of a single loop too.
 \item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
@@
 
-%\acks
-%Acknowledgments, if needed.
+\acks
+We would like to thank Samuele Pedroni, Sven Hager and the anonymous reviewers
+for helpful comments on drafts of this paper.
 

### 2012-08-08 talk/dls2012/paper.tex
oups
more benchmark explanations

--- 2012-08-07 talk/dls2012/paper.tex
@@
   there are three different implementations.
-\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
-\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
+\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$. A single loop
+is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_n\right)$ from a vector
+${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
+$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n$. Both the output vector, $\bf b$, 
+and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
+with $n=10^5$ and $n=10^6$.
+\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
+${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
+hardcoded into the implementation making the benchmark consist of a single loop too.
 \item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
   size $3 \times 3$ using a custom class to represent two-dimensional
-  arrays.
+  arrays. It is implemented as a two nested loops that iterates over the elements of the 
+$n\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
+${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
+\begin{equation}
+  \label{eq:convsum}
+  \begin{array}{lclclc}
+    k_{3,3} a_{i-1,j-1} &+& k_{3,2} a_{i-1,j} &+& k_{3,1} a_{i-1,j+1} & + \\
+    k_{2,3} a_{i,j-1}   &+& k_{2,2} a_{i,j}   &+& k_{2,1} a_{i,j+1}   & + \\
+    k_{1,3} a_{i+1,j-1} &+& k_{1,2} a_{i+1,j} &+& k_{1,1} a_{i+1,j+1}  \\
+  \end{array}
+\end{equation}
+for $1 \leq i \leq n$ and $1 \leq j \leq n$.
+The memory for storing the matrices are again allocated outside the benchmark and $n=1000$ was used.
 \item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
   size $3 \times 3$. This is similar to convolution but instead of
-  summing over the elements, the maximum is taken. That places a
+  summing over the terms in Equation~\ref{eq:convsum}, the maximum over those terms is taken. That places a
   external call to a max function within the loop that prevents some
@@
   locate edges in an image. It calculates the gradient magnitude
-  using sobel derivatives. 
+  using sobel derivatives. A Sobel x-derivative $D_x$ of a $n \times n$ image ${I}$ is formed
+by convolving ${I}$ with
+\begin{equation}
+  {K} = \left(
+  \begin{array}{ccc}
+    -1 & 0 & 1 \\
+    -2 & 0 & 1 \\
+    -1 & 0 & 1 \\
+  \end{array}
+  \right) ,
+\end{equation}
+and a Sobel y-derivative $D_y$ is formed convolving with $K^\top$. The gradient magnitude is 
+then formed for each pixel independently by $\sqrt{D_x^2 + D_y^2}$. The two convolutions and the pixelwise
+magnitude calculation are combined in the implementation of this benchmark and calculated in a single pass over
+the input image. This single pass consists of two nested loops with a somewhat larger amount of calculations 
+performed each iteration as compared to the other benchmarks.
 \end{itemize}

### 2012-08-08 talk/vmil2012/tool/build_tables.py
The number of lines has grown

@@ -1,0 +1,197 @@
+from __future__ import division
+import csv
+import django
+from django.template import Template, Context
+import os
+import sys
+
+# This line is required for Django configuration
+django.conf.settings.configure()
+
+
+def getlines(csvfile):
+    with open(csvfile, 'rb') as f:
+        reader = csv.DictReader(f, delimiter=',')
+        return [l for l in reader]
+
+
+def build_resume_data_table(csvfiles, texfile, template):
+    assert len(csvfiles) == 1
+    lines = getlines(csvfiles[0])
+    table = []
+    head = ['Benchmark', 'compressed', 'naive', 'xz compressed']
+
+    for bench in lines:
+        res = [bench['bench'].replace('_', '\\_'),
+                "%.2f" % float(bench['total resume data size']),
+                "%.2f" % float(bench['naive resume data size']),
+                "%.2f" % float(bench['compressed resume data size']),
+        ]
+        table.append(res)
+    output = render_table(template, head, sorted(table))
+    write_table(output, texfile)
+
+def build_ops_count_table(csvfiles, texfile, template):
+    assert len(csvfiles) == 1
+    lines = getlines(csvfiles[0])
+    keys = 'numeric set get rest new guard '.split()
+    table = []
+    head = ['Benchmark']
+    head += ['%s b' % k for k in keys]
+    head += ['%s a' % k for k in keys]
+
+    for bench in lines:
+        ops = {'before': sum(int(bench['%s before' % s]) for s in keys),
+                'after': sum(int(bench['%s after' % s]) for s in keys)}
+
+        res = [bench['bench'].replace('_', '\\_'),]
+        for t in ('before', 'after'):
+            values = []
+            for key in keys:
+                o = int(bench['%s %s' % (key, t)])
+                values.append(o / ops[t] * 100)
+
+            assert 100.0 - sum(values) < 0.0001
+            res.extend(['%.2f ' % v for v in values])
+        table.append(res)
+    output = render_table(template, head, sorted(table))
+    write_table(output, texfile)
+
+def build_guard_table(csvfiles, texfile, template):
+    assert len(csvfiles) == 1
+    lines = getlines(csvfiles[0])
+    table = []
+    head = ['Benchmark', 'guards b/o in \%', 'guards a/o in \%']
+
+    keys = 'numeric set get rest new guard '.split()
+    for bench in lines:
+        ops = {'before': sum(int(bench['%s before' % s]) for s in keys),
[…]

### 2012-08-08 talk/dls2012/paper.tex
comment
The number of lines has grown

@@
 \newcommand\anto[1]{\nb{ANTO}{#1}}
+\newcommand\hakan[1]{\nb{HAKAN}{#1}}
 \newcommand\reva[1]{\nb{Reviewer 1}{#1}}
@@
   size $3 \times 3$ using a custom class to represent two-dimensional
-  arrays. It is implemented as a two nested loops that iterates over the elements of the 
+  arrays. It is implemented as two nested loops that iterates over the elements of the 
 $n\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
@@
   locate edges in an image. It calculates the gradient magnitude
-  using sobel derivatives. A Sobel x-derivative $D_x$ of a $n \times n$ image ${I}$ is formed
+  using sobel derivatives. A Sobel x-derivative, $D_x$, of a $n \times n$ image, ${I}$, is formed
 by convolving ${I}$ with
@@
     -1 & 0 & 1 \\
-    -2 & 0 & 1 \\
+    -2 & 0 & 2 \\
     -1 & 0 & 1 \\
@@
 \end{equation}
-and a Sobel y-derivative $D_y$ is formed convolving with $K^\top$. The gradient magnitude is 
+and a Sobel y-derivative, $D_y$, is formed convolving $I$ with $K^\top$. The gradient magnitude is 
 then formed for each pixel independently by $\sqrt{D_x^2 + D_y^2}$. The two convolutions and the pixelwise
@@
 how annoying would it be to rerun the benchmarks, if I can find somebody to write them?}
+\hakan{there is iwtc11/benchmarks/runall.sh which is supposed to run them all}
 

### 2012-08-09 talk/vmil2012/tool/build_tables.py
Add a table showing the percentage of guards that ever fail and the percentage of guards that fail more than 200 times

@@ -1,0 +1,228 @@
+from __future__ import division
+import csv
+import django
+import json
+import os
+import sys
+from django.template import Template, Context
+
+# This line is required for Django configuration
+django.conf.settings.configure()
+
+
+def getlines(csvfile):
+    with open(csvfile, 'rb') as f:
+        reader = csv.DictReader(f, delimiter=',')
+        return [l for l in reader]
+
+
+def build_failing_guards_table(files, texfile, template):
+    BRIDGE_THRESHOLD = 200
+    assert len(files) == 2
+    with open(files[1]) as f:
+        failures = json.load(f)
+    for l in getlines(files[0]):
+        failures[l['bench']]['nguards'] = float(l['number of guards'])
+
+    table = []
+    head = ['Benchmark',
+            'failing guards',
+            'over %d failures' % BRIDGE_THRESHOLD]
+
+    for bench, info in failures.iteritems():
+        total = failures[bench]['nguards']
+        total_failures = len(info['results'])
+        bridges = len([k for k,v in info['results'].iteritems() \
+                                            if v > BRIDGE_THRESHOLD])
+        res = [bench.replace('_', '\\_'),
+                "%.2f \\%%" % (100 * total_failures/total),
+                "%.2f \\%%" % (100 * bridges/total),
+        ]
+        table.append(res)
+    output = render_table(template, head, sorted(table))
+    write_table(output, texfile)
+
+
+def build_resume_data_table(csvfiles, texfile, template):
+    assert len(csvfiles) == 1
+    lines = getlines(csvfiles[0])
+    table = []
+    head = ['Benchmark', 'compressed', 'naive', 'xz compressed']
+
+    for bench in lines:
+        res = [bench['bench'].replace('_', '\\_'),
+                "%.2f" % float(bench['total resume data size']),
+                "%.2f" % float(bench['naive resume data size']),
+                "%.2f" % float(bench['compressed resume data size']),
+        ]
+        table.append(res)
+    output = render_table(template, head, sorted(table))
+    write_table(output, texfile)
+
+def build_ops_count_table(csvfiles, texfile, template):
+    assert len(csvfiles) == 1
+    lines = getlines(csvfiles[0])
+    keys = 'numeric set get rest new guard '.split()
+    table = []
+    head = ['Benchmark']
+    head += ['%s b' % k for k in keys]
+    head += ['%s a' % k for k in keys]
+
+    for bench in lines:
[…]

### 2012-08-09 talk/vmil2012/paper.tex
a very preliminary attempt at a figure for vizualizing resume data
(cfbolz, bivab) generate and embed table about resume data sizes
Add a table showing the percentage of guards that ever fail and the percentage of guards that fail more than 200 times
remove some todos and update one
Use SSA reference

@@
 path, tracing is started, recording all operations that are executed on this
-path. Like in most compilers tracing JITs use an intermediate representation
-to store the recorded operations, which is typically in SSA form\todo{some ssa
-reference}. Since tracing follows actual execution the code that is recorded
+path. Like in most compilers tracing JITs use an intermediate representation to
+store the recorded operations, which is typically in SSA
+form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
+code that is recorded
 represents only one possible path through the control flow graph. Points of
@@
 \end{itemize}
-\todo{figure showing linked resume-data}
 
@@
 compared to the other source of information delayed heap stores are quite rare.
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{figures/resume_data.pdf}
+\caption{The resume data for Figure~\ref{fig:trace-log}}
+\label{fig:resume-data}
+\end{figure}
 
@@
 
-\todo{figure about failure counts of guards (histogram?)}
-\todo{add resume data sizes without sharing}
-\todo{add a footnote about why guards have a threshold of 100}
+\begin{figure}
+    \include{figures/resume_data_table}
+    \caption{Resume Data sizes in KiB}
+    \label{fig:resume_data_sizes}
+\end{figure}
+
+\begin{figure}
+    \include{figures/failing_guards_table}
+    \caption{Failing guards}
+    \label{fig:failing_guards}
+\end{figure}
+
+\todo{add a footnote about why guards have a threshold of 200}
 

### 2012-08-09 talk/vmil2012/paper.tex
Move some figures around and add sub sections to the evaluation section

@@
 From the mentioned benchmarks we collected different datasets to evaluate the
-Frequency, the overhead and overall behaviour of guards.
+Frequency, the overhead and overall behaviour of guards, the results are
+summarized in the remainder of this section.
+
+\subsection{Frequency of Guards}
+\label{sub:guard_frequency}
+\begin{figure*}
+    \include{figures/benchmarks_table}
+    \caption{Benchmark Results}
+    \label{fig:benchmarks}
+\end{figure*}
 Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
@@
 optimizing them.
-\todo{some pie charts about operation distribution}
-
-\begin{figure*}
-    \include{figures/benchmarks_table}
-    \caption{Benchmark Results}
-    \label{fig:benchmarks}
-\end{figure*}
-
+\subsection{Overhead of Guards}
+\label{sub:guard_overhead}
 \begin{figure}
@@
 \end{figure}
-
-\begin{figure}
-    \include{figures/failing_guards_table}
-    \caption{Failing guards}
-    \label{fig:failing_guards}
-\end{figure}
-
-\todo{add a footnote about why guards have a threshold of 200}
 
@@
 
-\begin{figure*}
-    \include{figures/backend_table}
-    \caption{Total size of generated machine code and guard data}
-    \label{fig:backend_data}
-\end{figure*}
-
 Both figures do not take into account garbage collection. Pieces of machine
@@
 \todo{compare to naive variant of resume data}
+
+\begin{figure}
+    \include{figures/backend_table}
+    \caption{Total size of generated machine code and guard data}
+    \label{fig:backend_data}
+\end{figure}
+
+\subsection{Guard Failures}
+\label{sub:guard_failure}
+\begin{figure}
+    \include{figures/failing_guards_table}
+    \caption{Failing guards}
+    \label{fig:failing_guards}
+\end{figure}
+
+\todo{add a footnote about why guards have a threshold of 200}
 

### 2012-08-10 talk/vmil2012/paper.tex
more rewriting of the evaluation section

@@
 
-\bivab{Figure needs better formatting}
+\todo{Figure needs better formatting}
 \begin{figure}[ht]
@@
 Frequency, the overhead and overall behaviour of guards, the results are
-summarized in the remainder of this section.
+summarized in the remainder of this section. We want to point out three
+aspects of guards in particular
+\begin{itemize}
+  \item Guards are very common operations in traces.
+  \item There is overhead associated with guards.
+  \item Guard failures are local and rare.
+\end{itemize}
+
+All figures in this section do not take into account garbage collection. Pieces
+of machine code can be globally invalidated or just become cold again. In both
+cases the generated machine code and the related data is garbage collected. The
+figures show the total amount of operations that are evaluated by the JIT and
+the total amount of code and data that is generated from the optimized traces.
 
@@
 \end{figure*}
+
 Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
@@
 operations are guards. The number of operations was counted on the unoptimized
-and optimized traces. Showing that the overall optimization rate is between
-65.80\% and 86.23\% of all operations and that the optimization rate for guards
-is similar to the general one, as could be assumed based on
-Figure~\ref{fig:guard_percent}. These numbers show that guards are a rather
-common operation in the traces, which is a reason the put effort into
-optimizing them.
+and optimized traces. The Figure shows that the overall optimization rate for
+operations which is between 69.4\% and 83.89\% of the traced operations and the
+optimization rate of guards, which is between 65.8\% and 86.2\% of the
+operations, are very similar, as could be assumed based on
+Figure~\ref{fig:guard_percent}. This indicates that the optimizer can remove
+most of the guards, but after the optimization pass guards still account for
+15.2\% to 20.2\% of the operations being compiled and later executed, the
+frequency of this operation makes it important to store the associated
+information efficiently and also to make sure that guard checks are executed
+fast.
+
 \subsection{Overhead of Guards}
@@
 approximation of the size of the \texttt{resume data} structures for the
-different benchmarks mentioned above. The size of the machine code is composed
-of the size of the compiled operations, the trampolines generated for the
-guards and a set of support functions that are generated when the JIT starts
-and are shared by all compiled traces. The size of the \texttt{low-level resume
+different benchmarks mentioned above. The machine code taken into account is
+composed of the compiled operations, the trampolines generated for the guards
+and a set of support functions that are generated when the JIT starts and which
+are shared by all compiled traces. The size of the \texttt{low-level resume
 data} is the size of the compressed mapping from registers and stack to
-IR-level variable and finally the size of the \texttt{resume data} is an
-approximation of the size of the compressed high-level resume data\todo{explain
-why it is an approximation}.
+IR-level variables and finally the size of the \texttt{resume data} is an
+approximation of the size of the compressed high-level resume data as described
+in Section~\ref{sec:Resume Data}\todo{explain why it is an approximation}.
 
-Compared to the size of the generated machine code the compressed
-\texttt{low-level resume data} is about 15\% to 20\% of that size, depending on
-the benchmark. On the other hand the generated machine code has only a size
-ranging from 20.21\% to 37.98\% of the size of the high and low-level
-\texttt{resume data} being compressed as described before.
+For the different benchmarks the \texttt{low-level resume data} has a size of
+about 15\% to 20\% of the amount of memory compared to the size of the
+generated machine code. On the other hand the generated machine code has only a
+size ranging from 20.5\% to 37.98\% of the size of the high and low-level
+\texttt{resume data} combined and being compressed as described before.
 
@@
 that is traced in a hot loop, for this reason the amount of generated machine
-code will be smaller than in other juts-in-time compilation approaches. Still
-the overhead associated to guards to resume execution from a side exit appears
-to be high.\bivab{put into relation to other JITs, compilers in general}
-
-Both figures do not take into account garbage collection. Pieces of machine
-code can be globally invalidated or just become cold again. In both cases the
-generated machine code and the related data is garbage collected. The figures
-show the total amount of operations that are evaluated by the JIT and the
-total amount of code and data that is generated from the optimized traces.
+code will be smaller than in other juts-in-time compilation approaches.  This
+creates a larger discrepancy between the size of the \texttt{resume data} when
+compared to the illustrates why it is important to compress this information.
 

### 2012-08-10 talk/vmil2012/paper.tex
write about the size of resume data

@@
 
-All figures in this section do not take into account garbage collection. Pieces
+All figures in this section do not take garbage collection into account. Pieces
 of machine code can be globally invalidated or just become cold again. In both
@@
 
-\todo{compare to naive variant of resume data}
-
 \begin{figure}
@@
 \end{figure}
+
+Why the efficient storing of the \texttt{resume data} is a central concern in the design
+of guards is illustrated by Figure~\ref{fig:backend_data}, this Figure shows
+the size of the compressed \texttt{resume data}, the approximated size of
+storing the \texttt{resume data} without compression and the size of
+compressing the data to calculate the size of the resume data using the
+\texttt{xz} compression tool, which is a ``general-purpose data compression
+software with high compression ratio'' used to approximate the best possible
+compression for the \texttt{resume data}.\footnote{\url{http://tukaani.org/xz/}}.
+
+The results show that the current approach of compression and data sharing only
+requires 18.3\% to 31.1\% of the space compared to the naive approach. This
+shows that large parts of the resume data are redundant and can be stored more
+efficiently through using the techniques described above. On the other hand
+comparing the results to the xz compression which only requires between 17.1\%
+and 21.1\% of the space required by our compression shows that the compression
+is not optimal but a trade-off between the required space and the time needed
+to build a good compressed representation of the compressed resume data for the
+large amount of guards present in the traces.
 
@@
 LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
-called snapshots, similar to RPython's resume data, to store the information about
-how to rebuild the state from a side-exit using the information in the snapshot
-and the machine execution state. Pall also acknowledges that snapshot for
-guards are associated with a large memory footprint. The solution used in
-LuaJIT is to store sparse snapshots, avoiding the creation of snapshots for
-every guard to reduce memory pressure. Snapshots are only created for guards
-after updates to the global state, after control flow points from the original
-program and for guards that are likely to fail. As an outlook Pall mentions the
-plans to switch to compressed snapshots to further reduce redundancy.
+called snapshots, similar to RPython's resume data, to store the information
+about how to rebuild the state from a side-exit using the information in the
+snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
+snapshots for guards in LuaJIT are associated with a large memory footprint.
+The solution used in there is to store sparse snapshots, avoiding the creation
+of snapshots for every guard to reduce memory pressure. Snapshots are only
+created for guards after updates to the global state, after control flow points
+from the original program and for guards that are likely to fail. As an outlook
+Pall mentions the plans to switch to compressed snapshots to further reduce
+redundancy.
 

### 2012-08-10 talk/vmil2012/paper.tex
add a todo
Write about the guard failure frequency

@@
 \label{sec:evaluation}
+\todo{improve the table formatting}
 
@@
 \label{sub:guard_failure}
+The last point in this discussion is the frequency of guard failures.
+Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
+relative amounts of guards that ever fail and of guards that fail more than 200
+times. For guards that fail more than 200 times, as described before, a trace
+is recorded that starts from the guard, patching the guard so that later
+failures execute the new trace instead of taking the side-exit. Hence the
+numbers presented for guards that fail more than 200 times represent the 200
+failures up to the compilation of the bridge and all executions of the then
+attached bridge.
+
 \begin{figure}
@@
 \end{figure}
+
+From Figure~\ref{fig:failing_guards} we can see that only a very small amount
+of all the guards in the optimized traces ever fail. This amount varies between
+2.4\% and 5.7\% of all guards. As can be expected, even less guards fail often
+enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
+fail more than 200 times. Also of all failing guards a few fail extremely often
+and most fail rarely. The results emphasizes that as most of the guards never
+fail it is important to make sure that the successful execution of a guard does
+not have unnecessary overhead.
 

### 2012-08-10 talk/vmil2012/paper.tex
some improvements to the evaluation section

@@
 From the mentioned benchmarks we collected different datasets to evaluate the
-Frequency, the overhead and overall behaviour of guards, the results are
+frequency, the overhead and overall behaviour of guards, the results are
 summarized in the remainder of this section. We want to point out three
@@
 
-All figures in this section do not take garbage collection into account. Pieces
+All figures in this section do not take garbage collection of machine code into account. Pieces
 of machine code can be globally invalidated or just become cold again. In both
@@
 most of the guards, but after the optimization pass guards still account for
-15.2\% to 20.2\% of the operations being compiled and later executed, the
-frequency of this operation makes it important to store the associated
+15.2\% to 20.2\% of the operations being compiled and later executed.
+The frequency of guard operations makes it important to store the associated
 information efficiently and also to make sure that guard checks are executed
-fast.
+quickly.
 
@@
 approximation of the size of the compressed high-level resume data as described
-in Section~\ref{sec:Resume Data}\todo{explain why it is an approximation}.
+in Section~\ref{sec:Resume Data}.\footnote{
+The size of the resume data is not measured at runtime, but reconstructed from
+log files.}
 
@@
 Why the efficient storing of the \texttt{resume data} is a central concern in the design
-of guards is illustrated by Figure~\ref{fig:backend_data}, this Figure shows
+of guards is illustrated by Figure~\ref{fig:backend_data}. This figure shows
 the size of the compressed \texttt{resume data}, the approximated size of
-storing the \texttt{resume data} without compression and the size of
-compressing the data to calculate the size of the resume data using the
+storing the \texttt{resume data} without compression and
+an approximation of the best possible compression of the resume data by
+compressing the data using the
 \texttt{xz} compression tool, which is a ``general-purpose data compression
-software with high compression ratio'' used to approximate the best possible
-compression for the \texttt{resume data}.\footnote{\url{http://tukaani.org/xz/}}.
+software with high compression ratio''.\footnote{\url{http://tukaani.org/xz/}}
 
 The results show that the current approach of compression and data sharing only
-requires 18.3\% to 31.1\% of the space compared to the naive approach. This
+requires 18.3\% to 31.1\% of the space compared to a naive approach. This
 shows that large parts of the resume data are redundant and can be stored more
@@
 relative amounts of guards that ever fail and of guards that fail more than 200
-times. For guards that fail more than 200 times, as described before, a trace
-is recorded that starts from the guard, patching the guard so that later
+times.\footnote{
+    The threshold of 200 is rather high. It was picked experimentally to give
+    good results for long-running programs.
+}
+As described before, for guards that fail more than 200 times, a trace
+is recorded that starts from the guard. Afterwards the guard is patched so that later
 failures execute the new trace instead of taking the side-exit. Hence the
@@
 not have unnecessary overhead.
-
-\todo{add a footnote about why guards have a threshold of 200}
 

### 2012-08-10 talk/vmil2012/paper.tex
a note about luajit

@@
 Pall mentions the plans to switch to compressed snapshots to further reduce
-redundancy.
+redundancy. The approach of not creating snapshots at all for every guard is
+orthogonal to the resume data compression presented in this paper and could be
+reused within RPython to improve the memory usage further.
 
@@
 \label{sec:Conclusion}
+In this paper we have concentrated on guards, an operation typically found in
+tracing just-in-time compilers and used to denote points of possible control
+flow bifurcation in recorded traces.
+We described how, based on the observation that guards are a frequent operation
+in traces and that they do not fail often, guards have been implemented in the
+high and low level components of RPython's tracing JIT compiler.
 
-\todo{conclusion}
+Finally we have presented experimental data collected using the standard PyPy
+benchmark set to evaluate previous observations and assumptions. Our
+experiments showed that, as previously assumed, guards are a very common
+operation in traces. At the same time guards are associated with a high
+overhead, because for all compiled guards high- and low information needs to be
+stored to restore the execution state in case of a bail-out. The measurements
+showed that the compression techniques used in PyPy effectively reduce the
+overhead of guards, while it still produces a significant overhead. The results
+also showed that guard failure is a local event, meaning that there are few
+guards that fail at all, and ever fewer that fail very often, validating the
+concept of reducing the overhead of successful guard checks while incurring an
+additional overhead, besides leaving the optimized code, in the case of a
+bailout having to decode the compressed state representation, which on the
+other hand reduces the memory footprint storing data that is seldom used in a
+compact manner.
+
+\subsection{Future Work}
+% subsection Future Work (end)
+An aspect that requires further investigation is to experimentally validate
+different of the trade-off decisions that went into the design of guard in
+RPython's tracing JIT. Based on the observation that most guards do not fail
+very often or at all while there are many guard and a few fail very often it
+would be worth exploring if a more aggressive compression scheme for guards
+would be worth the memory saving in contrast to the increased decoding
+overhead. Based on the same observation we would like to explore the concept of
+LuaJIT's sparse snapshots and its applicability to PyPy. Taking the idea of
+data-driven optimization to investigate possible areas of optimization as was
+done here in retrospective to validate a design seems a promising approach in
+the context of tracing JITs.
 

### 2012-08-10 talk/vmil2012/paper.tex
Write some conclusion-ish words
rewrite conclusion/related work a bit

--- 2012-08-10 talk/vmil2012/paper.tex
@@
 \label{sec:Conclusion}
+In this paper we have concentrated on guards, an operation typically found in
+tracing just-in-time compilers and used to denote points of possible control
+flow divergence in recorded traces.
+We described how, based on the observation that guards are a frequent operation
+in traces and that they do not fail often, guards have been implemented in the
+high and low level components of RPython's tracing JIT compiler.
 
-\todo{conclusion}
+Finally we have presented experimental data collected using the standard PyPy
+benchmark set to evaluate previous observations and assumptions. Our
+experiments showed that, as previously assumed, guards are a very common
+operation in traces. At the same time guards are associated with a high
+overhead, because for all compiled guards information needs to be
+stored to restore the execution state in case of a bail-out. The measurements
+showed that the compression techniques used in PyPy effectively reduce the
+overhead of guards, while it still produces a significant overhead. The results
+also showed that guard failure is a local event: there are few
+guards that fail at all, and even fewer that fail very often.
+These numbers validate the design decision of reducing the overhead of
+successful guard checks as much as possible while paying a higher price in the
+case of bailout due to having to decode compressed state representation.
+The compressed state representation is reduces the memory footprint of rarely
+used data.
+
+Based on the observation that most guards do not fail very often or at all it
+would be worth exploring if a more aggressive compression scheme for guards
+would be worth the memory saving in contrast to the increased decoding
+overhead. Based on the same observation we would like to explore the concept of
+LuaJIT's sparse snapshots and its applicability to PyPy.
 

### 2012-08-10 talk/vmil2012/tool/build_tables.py
tweak figures and tables

@@ -26,8 +26,8 @@ def build_failing_guards_table(files, texfile, template):
 
     table = []
     head = ['Benchmark',
-            'failing guards',
-            'over %d failures' % BRIDGE_THRESHOLD]
+            'Failing guards',
+            'Over %d failures' % BRIDGE_THRESHOLD]
 
     for bench, info in failures.iteritems():
         total = failures[bench]['nguards']
@@ -47,16 +47,16 @@ def build_resume_data_table(csvfiles, texfile, template):
     assert len(csvfiles) == 1
     lines = getlines(csvfiles[0])
     table = []
-    head = ['Benchmark', 'compressed', 'naive', 'xz compressed']
+    head = ['Benchmark', 'Compressed', 'Naive', 'xz compressed']
 
     for bench in lines:
         total = float(bench['total resume data size'])
         naive = float(bench['naive resume data size'])
         xz = float(bench['compressed resume data size'])
         res = [bench['bench'].replace('_', '\\_'),
-                "%.2f (%.1f\\%%)" %  (total, (100*total/naive)),
-                "%.2f (%.1f\\%%)" % (naive, 100*naive/total),
-                "%.2f (%.1f\\%%)" % (xz, 100*xz/total),
+                "%.2f {\scriptsize KiB}" %  (total,),# (100*total/naive)),
+                "%.2f {\scriptsize KiB}" % (naive),#, 100*naive/total),
+                "%.2f {\scriptsize KiB}" % (xz),#, 100*xz/total),
         ]
         table.append(res)
     output = render_table(template, head, sorted(table))
@@ -92,7 +92,7 @@ def build_guard_table(csvfiles, texfile, template):
     assert len(csvfiles) == 1
     lines = getlines(csvfiles[0])
     table = []
-    head = ['Benchmark', 'guards b/o', 'guards a/o']
+    head = ['Benchmark', 'Guards before', 'Guards after']
 
     keys = 'numeric set get rest new guard '.split()
     for bench in lines:
@@ -119,12 +119,12 @@ def build_benchmarks_table(csvfiles, texfile, template):
         bridgedata[l['bench']] = l
 
     head = ['Benchmark',
-            'ops b/o',
-            'guards b/o',
-            'ops a/o',
-            'guards a/o',
-            'opt. rate',
-            'guard opt. rate',
+            'Ops. before',
+            'Guards before',
+            'Ops. after',
+            'Guards after',
+            'Opt. rate',
+            'Guard opt. rate',
             ]
 
[…]

### 2012-08-10 talk/vmil2012/paper.tex
update caption
tweak figures and tables

@@
     \include{figures/benchmarks_table}
-    \caption{Benchmark Results}
+    \caption{Number of operations in the recorded traces and the relative amount of guards before and after optimizations}
     \label{fig:benchmarks}
@@
     \include{figures/resume_data_table}
-    \caption{Resume Data sizes in KiB}
+    \caption{Resume data sizes}
     \label{fig:resume_data_sizes}
@@
     \include{figures/backend_table}
-    \caption{Total size of generated machine code and guard data}
+    \caption{Total size of generated machine code and resume data}
     \label{fig:backend_data}
@@
     \include{figures/failing_guards_table}
-    \caption{Failing guards}
+    \caption{Failing guards relative to the total number of guards}
     \label{fig:failing_guards}

### 2012-08-10 talk/vmil2012/paper.tex
add another todo
start explaining the example

@@
 
-\todo{explain example}
-%___________________________________________________________________________
-
 \begin{figure}
@@
     \caption{Example Program}
-    \label{fig:trace-log}
+    \label{fig:example}
+\end{figure}
+
+Figure~\ref{fig:example} shows an example RPython function that checks
+whether a number reduces to 1 with less than 100 steps of the Collatz process.
+It uses an \lstinline{Even} and an \lstinline{Odd} class to box the numbers, to
+make the example more interesting. If the loop in \lstinline{check_reduces} is
+traced when \lstinline{a} is a multiple of four, the unoptimized
+trace looks like in Figure~\ref{fig:unopt-trace}. The line numbers in the trace
+correspond to the line numbers in Figure~\ref{fig:trace-log}.
+
+\begin{figure}
+    \input{figures/unopt-log.tex}
+    \caption{Unoptimized trace}
+    \label{fig:unopt-trace}
 \end{figure}
@@
 \section*{Appendix}
+\todo{remove this section and the figures}
 \begin{figure*}

### 2012-08-10 talk/iwtc11/benchmarks/scimark.py
start explaining the example

@@ -1,0 +1,21 @@
+from convolution.convolution import Array2D
+
+def SOR_execute(omega, G, num_iterations):
+    for p in xrange(num_iterations):
+        for i in xrange(1, G.height - 1):
+            for j in xrange(1, G.width - 1):
+                G[j, i] = omega * 0.25 * (G[j, i-1] + G[j, i+1] + G[j-1, i] +
+                                          G[j+1, i] + (1.0 - omega) * G[j, i])
+def SOR(args):
+    n, cycles = map(int, args)
+    a = Array2D(n, n)
+    SOR_execute(1.25, a, cycles)
+
+if __name__ == '__main__':
+    from time import time
+    for i in range(10):
+        t0 = time()
+        #SOR([100, 32768]) # gcc -O3: 2.51, pypy-1.8: 3.83
+        SOR([1000, 256]) # gcc -O3 2.07, pypy-1.8: 3.03
+        t1 = time()
+        print t1-t0

### 2012-08-11 talk/vmil2012/paper.tex

@@
 trace looks like in Figure~\ref{fig:unopt-trace}. The line numbers in the trace
-correspond to the line numbers in Figure~\ref{fig:trace-log}.
+correspond to the line numbers in Figure~\ref{fig:trace-log}. The resulting
+trace repeatedly halves the current value and checks whether it is equal to
+one, or odd. In either of these cases the trace is left via a guard failure.
 
@@
 
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{figures/resume_data.pdf}
-\caption{The resume data for Figure~\ref{fig:trace-log}}
-\label{fig:resume-data}
-\end{figure}
+Figure~\ref{fig:trace-log} shows the optimized version of the trace in
+Figure~\ref{fig:fig:unopt-trace}. Allocation removal has removed the
+\lstinline{new} operation and other operations handling the boxes. The
+operations handle unboxed numbers now.
+
+Figure~\ref{fig:resume-data} sketches the symbolic frames of the first two
+guards in the trace. The frames for \lstinline{check_reduces} and
+\lstinline{Even.step} as well as the description of the allocation-removed
+virtual instance of \lstinline{Even} are shared between the two guards.
+
+\todo{fix labels in diagram}
 
@@
 \label{sec:Guards in the Backend}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{figures/resume_data.pdf}
+\caption{The resume data for Figure~\ref{fig:trace-log}}
+\label{fig:resume-data}
+\end{figure}
 

### 2012-08-12 talk/iwtc11/benchmarks/test_scimark.py
test for SOR

@@ -1,0 +1,30 @@
+from scimark import SOR_execute, Array2D
+from cffi import FFI
+import os
+
+ffi = FFI()
+ffi.cdef("""
+    typedef struct {...;} Random_struct, *Random;
+    Random new_Random_seed(int seed);
+    double **RandomMatrix(int M, int N, Random R);
+    void SOR_execute(int M, int N,double omega, double **G, int num_iterations);
+    """)
+C = ffi.verify("""
+    #include <SOR.h>
+    #include <Random.h>
+    """, 
+    extra_compile_args=['-I' + os.path.join(os.getcwd(), 'scimark')],
+    extra_link_args=['-fPIC'],
+    extra_objects=[os.path.join(os.getcwd(), 'scimark', f) 
+                   for f in ['SOR.c', 'Random.c']])
+
+def test_SOR():
+    width, height = 5, 7
+    rnd = C.new_Random_seed(7)
+    a = C.RandomMatrix(height, width, rnd)
+    b = Array2D(width, height, data=a)
+    C.SOR_execute(height, width, 1.25, a, 42)
+    SOR_execute(1.25, b, 42)
+    for x, y in b.indexes():
+        assert a[y][x] == b[x, y]
+

### 2012-08-12 talk/dls2012/paper.tex
correct the wrong depiction of luajit

@@
 type checking, boxed value unwrapping and virtual method lookups.
-In this paper we present a scheme for making simple optimizations loop-aware by
+In this paper we explain a scheme invented within the context of the LuaJIT project
+for making simple optimizations loop-aware by
 using a simple pre-processing step on the trace and not changing the
-optimizations themselves. The scheme can give performance improvements of a
+optimizations themselves.
+We have implemented the scheme in PyPy's tracing JIT compiler,
+where it can give performance improvements of a
 factor over two for PyPy's Python JIT executing simple numerical kernels
@@
 unwrapping and method lookups are often loop invariant and performance could be increased
-by moving those operations out of the loop. We propose a simple scheme
+by moving those operations out of the loop. We explain a simple scheme
 to make a tracing JIT loop-aware by allowing it's existing optimizations to
@@
 
-In this paper we want to address this problem by proposing a scheme that
-makes it possible to turn optimizations using one forward pass into
-optimizations that can do loop invariant code motion and similar loop-aware
-improvements. Using this scheme one does not need to change the underlying
-optimization much to get these advantages.
+Mike Pall pioneered a solution to address this problem in the context of a
+dynamic language using a tracing JIT compiler. He published his algorithm and
+its rationale in 2009~\cite{pall_luajit_2009} and implemented it in LuaJIT
+2.0\footnote{\url{http://luajit.org/}}, an open source JIT compiler for the Lua
+language. His approach allows to reuse all forward pass
+optimizations to achieve loop invariant code motion and other loop-related
+optimizations, which greatly simplifies the implementation. Using this scheme
+one does not need to change the underlying optimization much to get these
+advantages. We have implemented the same approach in PyPy's tracing JIT
+compiler the results of which we present here.
 
@@
 Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
-However, we want to get this effect using our existing optimization passes
+However, it is desirable to get this effect using our existing optimization passes
 without changing them too much. Optimizations with one forward pass
-cannot directly get this effect: They just look at the trace without taking
+cannot directly achieve this effect: They just look at the trace without taking
 into account that the trace executes many times in a row. Therefore to achieve
@@
 
-This is the key insight of the proposed implementation scheme: If an
+This is the key insight of the implementation scheme: If an
 optimization is given two iterations together at the same time, the
@@
 operations out of the loop making them executed only once
-and not every iteration. We propose to achieve this by loop peeling. It
+and not every iteration. This can be achieved by loop peeling. It
 leaves the loop body intact, but prefixes it with one iteration of the
@@
 \end{lstlisting}
-\caption{A peeled trace of the Example Interpreter}
+\caption{A peeled trace of the example interpreter}
 \label{fig:peeled-trace}
@@
 time savings are from removing boxing/unboxing operations.
-}
-
-\revd{
-The benchmark results appear quite impressive -- especially the comparison with
-GCC -- but without additional information, I have no idea what is being
-compared.  Are these results from the same sizes of integers and/or floating
-point results?
 }
@@
 implemented as a C++ class. The other benchmarks are implemented in
-plain C.
+plain C. All the benchmarks except sqrt operate on C double-precision floating
+point numbers, both in the Python and the C code.
 
@@
 
-We run GCC both with -O2 optimization and -O3 -march=native, disabling the
+We run GCC with -O3 -march=native, disabling the
 automatic loop vectorization. In all cases, SSE2 instructions were used for
@@
 comparable with native-compiled C code. We attribute the performance gap to C code to
-the relative immaturity of RPython's JIT assembler backend as well as missing
+the relative immaturity of RPython's JIT machine code backend as well as missing
 optimizations, like instruction scheduling.
@@
 
-Loop invariant code motion optimizations are completely
-standard~\cite{muchnick_advanced_1997}. Therefore, the effects that our
-optimization achieves are not in any way new. However, we think that achieving
-them in the way described in this paper is simpler than writing explicit algorithms.
-
-\revc{
-The discussion of LuaJIT is unsatisfying.  It's not clear to me from that one
-quote that Mike is doing the same thing.  It might be worth including LuaJIT in
-the benchmarks, and/or examining the actual implementation of LuaJIT.
-}
-\cfbolz{maybe we can look in the new LuaJIT wiki.
-how annoying would it be to rerun the benchmarks, if I can find somebody to write them?}
-\hakan{there is iwtc11/benchmarks/runall.sh which is supposed to run them all}
-
-Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
-have developed the described technique independently. There are no papers about
-LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
-synthetic unrolling of the recorded IR, combining copy-substitution with
-redundancy elimination to achieve code hoisting. The unrolled and
-copy-substituted instructions are simply fed back into the compiler pipeline,
-which allows reuse of all optimizations for redundancy elimination. Loop
-recurrences are detected on-the-fly and a minimized set of PHIs is
-generated.''~\cite{pall_luajit_2009}
+Loop invariant code motion optimizations are a well-known approach to optimize
+loops~\cite{muchnick_advanced_1997}. Therefore, the effects that the
+optimizations described here achieve are not in any way new. However, we think
+that achieving them in the way described in this paper is simpler than writing
+explicit algorithms.
+\cfbolz{more explicit listing of prior work goes here}
+
+As described in the introduction,
+Mike Pall pioneered the approach described in this paper.
+He showed that, unlike traditional loop-invariant code motion
+(LICM), this approach is effective, even in the presence of many
+guards and global control dependencies, which are caused by the
+semantics of dynamic languages.
+
+He writes on the Lua-users mailing list:
+``The LOOP pass does synthetic unrolling of the recorded IR, combining
+copy-substitution with redundancy elimination to achieve code hoisting. The
+unrolled and copy-substituted instructions are simply fed back into the
+compiler pipeline, which allows reuse of all optimizations for redundancy
+elimination. Loop recurrences are detected on-the-fly and a minimized set of
+PHIs is generated.''~\cite{pall_luajit_2009}
 
@@
 In this paper we have studied loop invariant code motion during trace
-compilation. We claim that loop peeling is a very convenient solution
-here since it fits well with other trace optimizations and does not require
-large changes to them. This approach improves the effect of standard
+compilation. We claim that the loop peeling approach of LuaJIT is a very convenient solution
+since it fits well with other trace optimizations and does not require
+large changes to them. The approach improves the effect of standard
 optimizations such as redundant guard removal, common subexpression elimination
@@
 We would like to thank Samuele Pedroni, Sven Hager and the anonymous reviewers
-for helpful comments on drafts of this paper.
+for helpful comments on drafts of this paper. We owe deep gratitude to Mike Pall
+for making his impressive work on LuaJIT available and for detailed help on a
+draft of the paper.
 

### 2012-08-12 talk/dls2012/paper.tex
seems we don't use \url yet

@@
 its rationale in 2009~\cite{pall_luajit_2009} and implemented it in LuaJIT
-2.0\footnote{\url{http://luajit.org/}}, an open source JIT compiler for the Lua
+2.0\footnote{\texttt{http://luajit.org/}}, an open source JIT compiler for the Lua
 language. His approach allows to reuse all forward pass

### 2012-08-13 talk/vmil2012/paper.tex
start adding David Edelsohn's comments

@@
   linkcolor=black,%
-  pdftitle={Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT},%
+  pdftitle={The Efficient Handling of Guards in the Design of RPython's Tracing JIT},%
   pdfauthor={David Schneider},
@@
 
-\title{Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT}
+\title{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}
 
@@
 
+\todo{the introduction needs some work}
+\cfbolz{the first two two paragraphs talk about deoptimization, then it
+switches to guards. I would say we should only talk about guards in the
+beginning}
 In this paper we describe and analyze how deoptimization works in the context
@@
 
-Although there are several publications about tracing just-in-time compilers,
-to our knowledge, there are none that describe deoptimization and the use and
-implementation of guards in this context.
-
+The goal of this paper is to understand the design constraints when
+implementing guards. Guards have a runtime cost, they take time to execute. On
+the other hand, guards are possible deoptimization points. They need to store
+enough information to rebuild the interpreter state.
 Based on the informal observation that guards are among the most common
-operations in the traces produced by RPython's tracing JIT and that guards are
-operations that are associated with an overhead to maintain information about
-the execution state to be able to rebuild it in case of deoptimization, our
+operations in the traces produced by RPython's tracing JIT, our
 goal is to present concrete numbers for the frequency and the overhead related
@@
 
+\cfbolz{this paragraph now suddenly \emph{introduces} guards, despite having talked about them already}
 The operations executed by an interpreter are recorded by the tracing JIT in
-case they are frequently executed, this process is described in more detail in
-Section~\ref{sec:Resume Data}, during the recording phase special operations,
+case they are frequently executed (this process is described in more detail in
+Section \ref{sec:Resume Data}). During the recording phase special operations,
 referred to as \texttt{guards}, are inserted into the recorded trace at all
 points where the control flow could diverge. As can be seen in
-Figure~\ref{fig:guard_percent} guards account for 14.42\% to 22.32\% of the
-operations before and for 15.2\% to 20.12\% of the operations after the
+Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
+operations before and for about 15\% to 20\% of the operations after the
 optimization pass over the traced and later compiled parts of the benchmarks,
@@
 needed for deoptimization should be kept low. A second aspect is that
-successfully checking guards, i.e. not leaving the compiled trace,  - which is
-the common case - should be a cheap operation to execute favouring the on-trace
+successfully checking guards, i.e. not leaving the compiled trace, – which is
+the common case – should be a cheap operation to execute favouring the on-trace
 execution speed in contrast to the deoptimization case where the state has to
@@
 describe based on them the reasoning behind and the implementation of guards in
-RPython's tracing just-in-time compiler, the contributions of this paper are:
+RPython's tracing just-in-time compiler. the contributions of this paper are:
 \begin{itemize}
   \item An analysis of guards in the context of RPython's tracing JIT to
-  substantiate the aforementioned observation, based on a set of benchmarks.
-  \item We provide a detailed measurements about the frequency and the
-  overhead associated with guards.
-  \item We provide a description about how guards are implemented in the high\-
+  substantiate the aforementioned observation, based on a set of benchmarks,
+  \item detailed measurements about the frequency and the
+  overhead associated with guards, and
+  \item a description about how guards are implemented in the high\-
   and low-level parts of the JIT and describe the rationale behind the design.
 \end{itemize}
+
 \begin{figure}
@@
 information required to restore the interpreter state in case of a guard
-failure, once the frontend has traced and optimized a loop it invokes the
+failure. Once the frontend has traced and optimized a loop it invokes the
 backend to compile the operations to machine code, Section \ref{sec:Guards in
@@
 
-The RPython language and the PyPy Project were started in 2002 with the goal of
+The RPython language and the PyPy project were started in 2002 with the goal of
 creating a Python interpreter written in a high level language, allowing easy
@@
 
-RPython is built of two components, the language and the translation toolchain
-used to transform RPython programs to executable units.  The RPython language
+RPython is constructed from two components:
+\begin{itemize}
+ \item the language itself
+ \item the translation toolchain used to transform RPython programs to executable units
+\end{itemize}
+The RPython language
 is a statically typed object oriented high level language. The language provides
-several features such as automatic memory management (aka. Garbage Collection)
+several features such as automatic memory management
 and just-in-time compilation. When writing an interpreter using RPython the
@@
 Tracing JITs are a technique of just-in-time compilers that generate code by
-observing the execution of a program. VMs using tracing JITs are typically
-mixed mode execution environments containing also an interpreter. The
-interpreter profiles the executed program and selects frequently executed code
-paths to be compiled to machine code. After profiling identified an interesting
+observing the execution of a program. VMs using tracing JITs typically are
+mixed-mode execution environments that also contain an interpreter. The
+interpreter profiles the executing program and selects frequently executed code
+paths to be compiled to machine code. Many tracing JIT compilers focus on
+selecting hot loops.
+
+After profiling identifies an interesting
 path, tracing is started, recording all operations that are executed on this
-path. Like in most compilers tracing JITs use an intermediate representation to
+path. This includes inlining functional calls.
+As in most compilers, tracing JITs use an intermediate representation to
 store the recorded operations, which is typically in SSA
@@
 tracing phase are still valid when the code has been compiled and is executed.
+In the case of dynamic languages, guards are also used to encode type checks
+that come from optimistic type specialization by recording the types of
+variables seen during tracing.
 After a trace has been recorded it is optimized and then compiled to platform
@@
 Since tracing linearizes control flow by following one concrete execution,
-not the full control flow of a program is observed.
+the full control flow of a program is not observed.
 The possible points of deviation from the trace are guard operations
-that check whether the same assumptions observed during tracing still hold during execution.
+that check whether the same assumptions observed during tracing
+still hold during execution.
+Similarly, in the case of dynamic languages guards can also encode type
+assumptions.
 In later executions of the trace the guards can fail.
@@
 
+After tracing has been finished the trace is optimized.
+During optimization a large percentage of operations can be removed.
+In the process the resume data is transformed into its final, compressed form.
+The rationale for not compressing the resume data during tracing
+is that a lot of guards will be optimized away.
+For them, the compression effort would be lost.
+
 The core idea of storing resume data as compactly as possible
@@
 often only the information in the top frame changes from one guard to the next.
-The other frames can often be just reused.
+The other symbolic frames can often just be reused.
 The reason for this is that during tracing only the variables
@@
 that do not survive long.
-An example is the instance of \lstinline{Even} in the example\cfbolz{reference figure}.
+An example is the instance of \lstinline{Even} in Figure~\ref{fig:unopt-trace}.
 Allocation removal makes resume data more complex.
@@
 because the optimizer removed their allocation.
-The virtual objects in the symbolic frames describe exactly
-how the heap objects look like which have to be allocated on guard failure.
+The structure of the heap objects that have to be allocated on guard failure
+is described by the virtual objects stored in the symbolic frames.
 To this end, the content of every field of the virtual object is described
@@
 Figure~\ref{fig:fig:unopt-trace}. Allocation removal has removed the
-\lstinline{new} operation and other operations handling the boxes. The
+\lstinline{new} operation and other operations handling the instance. The
 operations handle unboxed numbers now.
@@
 passes over the lists of instructions, a backwards pass to calculate live
-ranges of IR-level variables and a forward one to emit the instructions. During
+ranges of IR-level variables and a forward pass to emit the instructions. During
 the forward pass IR-level variables are assigned to registers and stack
@@
 checked by the guard is not used anywhere else the guard and the operation
-producing the value can merged, reducing even more the overhead of the guard.
+producing the value can often be merged, further reducing even more the overhead of the guard.
 Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
@@
 Each guard in the IR has attached to it a list of the IR-variables required to
-rebuild the execution state in case the trace is left through the side-exit
-corresponding to the guard. When a guard is compiled, additionally to the
+rebuild the execution state in case the trace is left through
+the guard. When a guard is compiled, in addition to the
 condition check two things are generated/compiled. First a special data
@@
 execution reaches the code emitted for the corresponding guard. This data
-structure stores the data in a compressed manner using an encoding the uses
+structure stores the data in a compressed manner using an encoding that uses
 8bits to store 7bits of information. This encoding is efficient to create and
-provides a compact representation of the needed information. This encoding
-needs to be as compact as possible to maintain an acceptable memory profile.
+provides a compact representation of the needed information,
+to maintain an acceptable memory profile.
 
@@
 Guards are implemented as a conditional jump to this trampoline. In case the
-condition checked in the guard fails execution and a side-exit should be taken
-execution jumps to the trampoline. In the trampoline the pointer to the
-\emph{low-level resume data} is loaded and jumps to generic bail-out handler
+condition checked in the guard fails
+execution jumps to the corresponding trampoline. In the trampoline the pointer to the
+\emph{low-level resume data} is loaded and jumps to generic bailout handler, also known as compensation code,
 that is used to leave the compiled trace in case of a guard failure.
 
-Using the encoded location information the bail-out handler reads from the
+Using the encoded location information the bailout handler reads from the
 saved execution state the values that the IR-variables had  at the time of the
-guard failure and stores them in a location that can be read by the fronted.
+guard failure and stores them in a location that can be read by the frontend.
 After saving the information the control is passed to the frontend signaling
@@
 As in previous sections the underlying idea for the design of guards is to have
-a fast on-trace profile and a potentially slow one in the bail-out case where
+a fast on-trace profile and a potentially slow one in the bailout case where
 the execution takes one of the side exits due to a guard failure. At the same
@@
 that led to the compilation of the bridge should execute the bridge without
-additional overhead, in particular the failure of the guard should not lead
+additional overhead. In particular the failure of the guard should not lead
 to leaving the compiled code prior to execution the code of the bridge.
@@
 reconstruction all bindings are restored to the state as they were in the
-original loop up to the guard.
+original loop up to the guard. This means that no register/stack reshuffling is
+needed before executing a bridge.
 
@@
 patched to redirect control flow to the bridge in case the check fails. In
-future if the guard fails again it jumps to the code compiled for the bridge
+the future, if the guard fails again it jumps to the code compiled for the bridge
 instead of bailing out. Once the guard has been compiled and attached to the
@@
 after the guard and the bridge are just conditional paths.
-Figure~\ref{fig:trampoline} shows a digram of a compiled loop with two guards,
+Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
 Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
-then calls the compensation code, whereas Guard \#2 has already been patched
+then calls the bailout handler, whereas Guard \#2 has already been patched
 and directly jumps to the corresponding bridge. The bridge also contains two
@@
 \todo{improve the table formatting}
+\todo{give a reference to the benchmark scripts to make things repeatable}
 
@@
 size ranging from 20.5\% to 37.98\% of the size of the high and low-level
-\texttt{resume data} combined and being compressed as described before.
+resume data combined and being compressed as described before.
 
@@
 creates a larger discrepancy between the size of the \texttt{resume data} when
-compared to the illustrates why it is important to compress this information.
+compared to the size of the generated machine code and illustrates why it is important to compress the \texttt{resume data} information.
 
@@
 is recorded that starts from the guard. Afterwards the guard is patched so that later
-failures execute the new trace instead of taking the side-exit. Hence the
+failures execute the new trace instead of jumping to the trampoline and returning to the interpreter. Hence the
 numbers presented for guards that fail more than 200 times represent the 200
@@
 of all the guards in the optimized traces ever fail. This amount varies between
-2.4\% and 5.7\% of all guards. As can be expected, even less guards fail often
+2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
 enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
@@
 called snapshots, similar to RPython's resume data, to store the information
-about how to rebuild the state from a side-exit using the information in the
+about how to rebuild the state from a guard failure using the information in the
 snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
@@
 The paper does not describe any attempts to store this information compactly.
-This may not be needed in their approach, because method-based JITs have a lot
+This may not be needed in their approach, because method-based JITs have
 fewer deoptimization points than tracing JITs.
@@
 overhead, because for all compiled guards information needs to be
-stored to restore the execution state in case of a bail-out. The measurements
+stored to restore the execution state in case of a bailout. The measurements
 showed that the compression techniques used in PyPy effectively reduce the

### 2012-08-13 talk/iwtc11/benchmarks/test_scimark.py
start adding David Edelsohn's comments

@@ -1,0 +1,51 @@
+from scimark import SOR_execute, Array2D, ArrayList, Random, MonteCarlo_integrate
+from cffi import FFI
+import os
+
+ffi = FFI()
+ffi.cdef("""
+    typedef struct {...;} Random_struct, *Random;
+    Random new_Random_seed(int seed);
+    double Random_nextDouble(Random R);
+    double **RandomMatrix(int M, int N, Random R);
+
+    void SOR_execute(int M, int N,double omega, double **G, int num_iterations);
+    double MonteCarlo_integrate(int Num_samples);    
+    """)
+C = ffi.verify("""
+    #include <SOR.h>
+    #include <Random.h>
+    #include <MonteCarlo.h>
+    """, 
+    extra_compile_args=['-I' + os.path.join(os.getcwd(), 'scimark')],
+    extra_link_args=['-fPIC'],
+    extra_objects=[os.path.join(os.getcwd(), 'scimark', f) 
+                   for f in ['SOR.c', 'Random.c', 'MonteCarlo.c']])
+
+class TestWithArray2D(object):
+    Array = Array2D
+
+    def test_SOR(self):
+        width, height = 5, 7
+        rnd = C.new_Random_seed(7)
+        a = C.RandomMatrix(height, width, rnd)
+        b = self.Array(width, height, data=a)
+        C.SOR_execute(height, width, 1.25, a, 42)
+        SOR_execute(1.25, b, 42)
+        for x, y in b.indexes():
+            assert a[y][x] == b[x, y]
+
+class TestWithArrayList(TestWithArray2D):
+    Array = ArrayList
+
+def test_random():
+    rnd_C = C.new_Random_seed(7)
+    rnd_py = Random(7)
+    for i in range(100000):
+        assert C.Random_nextDouble(rnd_C) == rnd_py.nextDouble()
+
+def test_montecarlo():
+    for n in [100, 200, 500, 1000]:
+        assert C.MonteCarlo_integrate(n) == MonteCarlo_integrate(n)
+
+

### 2012-08-13 talk/vmil2012/paper.tex
C runner for LU

@@
 
+\cfbolz{I would kill this paragraph}
+Although there are several publications about tracing just-in-time compilers,
+to our knowledge, there are none that describe deoptimization and the use and
+implementation of guards in this context.
+
 The goal of this paper is to understand the design constraints when
@@
 The operations executed by an interpreter are recorded by the tracing JIT in
-case they are frequently executed (this process is described in more detail in
-Section \ref{sec:Resume Data}). During the recording phase special operations,
+case they are frequently executed, this process is described in more detail in
+Section~\ref{sec:Resume Data}, during the recording phase special operations,
 referred to as \texttt{guards}, are inserted into the recorded trace at all
 points where the control flow could diverge. As can be seen in
-Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
-operations before and for about 15\% to 20\% of the operations after the
+Figure~\ref{fig:guard_percent} guards account for 14.42\% to 22.32\% of the
+operations before and for 15.2\% to 20.12\% of the operations after the
 optimization pass over the traced and later compiled parts of the benchmarks,
@@
 information required to restore the interpreter state in case of a guard
-failure. Once the frontend has traced and optimized a loop it invokes the
+failure, once the frontend has traced and optimized a loop it invokes the
 backend to compile the operations to machine code, Section \ref{sec:Guards in
@@
 
-RPython is constructed from two components:
-\begin{itemize}
- \item the language itself
- \item the translation toolchain used to transform RPython programs to executable units
-\end{itemize}
-The RPython language
+RPython is built of two components, the language and the translation toolchain
+used to transform RPython programs to executable units.  The RPython language
 is a statically typed object oriented high level language. The language provides
@@
 Tracing JITs are a technique of just-in-time compilers that generate code by
-observing the execution of a program. VMs using tracing JITs typically are
-mixed-mode execution environments that also contain an interpreter. The
-interpreter profiles the executing program and selects frequently executed code
+observing the execution of a program. VMs using tracing JITs are typically
+mixed mode execution environments containing also an interpreter. The
+interpreter profiles the executed program and selects frequently executed code
 paths to be compiled to machine code. Many tracing JIT compilers focus on
@@
 
-After profiling identifies an interesting
+After profiling identified an interesting
 path, tracing is started, recording all operations that are executed on this
 path. This includes inlining functional calls.
-As in most compilers, tracing JITs use an intermediate representation to
+Like most compilers, tracing JITs use an intermediate representation to
 store the recorded operations, which is typically in SSA
@@
 Since tracing linearizes control flow by following one concrete execution,
-the full control flow of a program is not observed.
+not the full control flow of a program is observed.
 The possible points of deviation from the trace are guard operations
@@
 because the optimizer removed their allocation.
-The structure of the heap objects that have to be allocated on guard failure
-is described by the virtual objects stored in the symbolic frames.
+The virtual objects in the symbolic frames describe exactly
+how the heap objects look like which have to be allocated on guard failure.
 To this end, the content of every field of the virtual object is described
@@
 checked by the guard is not used anywhere else the guard and the operation
-producing the value can often be merged, further reducing even more the overhead of the guard.
+producing the value can often merged, reducing even more the overhead of the guard.
 Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
@@
 Each guard in the IR has attached to it a list of the IR-variables required to
-rebuild the execution state in case the trace is left through
-the guard. When a guard is compiled, in addition to the
+rebuild the execution state in case the trace is left through the side-exit
+corresponding to the guard. When a guard is compiled, additionally to the
 condition check two things are generated/compiled. First a special data
@@
 Guards are implemented as a conditional jump to this trampoline. In case the
-condition checked in the guard fails
-execution jumps to the corresponding trampoline. In the trampoline the pointer to the
-\emph{low-level resume data} is loaded and jumps to generic bailout handler, also known as compensation code,
+condition checked in the guard fails execution and a side-exit should be taken
+execution jumps to the trampoline. In the trampoline the pointer to the
+\emph{low-level resume data} is loaded and jumps to generic bail-out handler
 that is used to leave the compiled trace in case of a guard failure.
 
-Using the encoded location information the bailout handler reads from the
+Using the encoded location information the bail-out handler reads from the
 saved execution state the values that the IR-variables had  at the time of the
-guard failure and stores them in a location that can be read by the frontend.
+guard failure and stores them in a location that can be read by the fronted.
 After saving the information the control is passed to the frontend signaling
@@
 As in previous sections the underlying idea for the design of guards is to have
-a fast on-trace profile and a potentially slow one in the bailout case where
+a fast on-trace profile and a potentially slow one in the bail-out case where
 the execution takes one of the side exits due to a guard failure. At the same
@@
 patched to redirect control flow to the bridge in case the check fails. In
-the future, if the guard fails again it jumps to the code compiled for the bridge
+future if the guard fails again it jumps to the code compiled for the bridge
 instead of bailing out. Once the guard has been compiled and attached to the
@@
 after the guard and the bridge are just conditional paths.
-Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
+Figure~\ref{fig:trampoline} shows a digram of a compiled loop with two guards,
 Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
-then calls the bailout handler, whereas Guard \#2 has already been patched
+then calls the compensation code, whereas Guard \#2 has already been patched
 and directly jumps to the corresponding bridge. The bridge also contains two
@@
 creates a larger discrepancy between the size of the \texttt{resume data} when
-compared to the size of the generated machine code and illustrates why it is important to compress the \texttt{resume data} information.
+compared to the illustrates why it is important to compress this information.
 
@@
 is recorded that starts from the guard. Afterwards the guard is patched so that later
-failures execute the new trace instead of jumping to the trampoline and returning to the interpreter. Hence the
+failures execute the new trace instead of taking the side-exit. Hence the
 numbers presented for guards that fail more than 200 times represent the 200
@@
 of all the guards in the optimized traces ever fail. This amount varies between
-2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
+2.4\% and 5.7\% of all guards. As can be expected, even less guards fail often
 enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
@@
 called snapshots, similar to RPython's resume data, to store the information
-about how to rebuild the state from a guard failure using the information in the
+about how to rebuild the state from a side-exit using the information in the
 snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
@@
 overhead, because for all compiled guards information needs to be
-stored to restore the execution state in case of a bailout. The measurements
+stored to restore the execution state in case of a bail-out. The measurements
 showed that the compression techniques used in PyPy effectively reduce the

### 2012-08-13 talk/iwtc11/benchmarks/scimark/run_SparseMatMult.c

@@ -1,0 +1,50 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <assert.h>
+
+#include "SparseCompRow.c"
+
+int main(int ac, char **av) {
+    assert(ac==4);
+    int N = atoi(av[1]);
+    int nz = atoi(av[2]);
+    int cycles = atoi(av[3]);
+    
+        double *x = (double*) calloc(sizeof(double), N); //RandomVector(N, R);
+        double *y = (double*) calloc(sizeof(double), N);
+
+        double result = 0.0;
+
+        int nr = nz/N;      /* average number of nonzeros per row  */
+        int anz = nr *N;    /* _actual_ number of nonzeros         */
+
+            
+        double *val = (double *) calloc(sizeof(double), anz); //RandomVector(anz, R);
+        int *col = (int*) calloc(sizeof(int), nz);
+        int *row = (int*) calloc(sizeof(int), N+1);
+        int r=0;
+
+        row[0] = 0; 
+        for (r=0; r<N; r++)
+        {
+            /* initialize elements for row r */
+
+            int rowr = row[r];
+            int step = r/ nr;
+            int i=0;
+
+            row[r+1] = rowr + nr;
+            if (step < 1) step = 1;   /* take at least unit steps */
+
+
+            for (i=0; i<nr; i++)
+                col[rowr+i] = i*step;
+                
+        }
+
+    SparseCompRow_matmult(N, y, val, row, col, x, cycles);
+    fprintf(stderr, "SparseMatMult(%d, %d, %d):  ", N, nz, cycles);
+    return 0;
+}
+
+

### 2012-08-13 talk/iwtc11/benchmarks/benchmark.sh
C runner for LU

@@ -20,7 +20,9 @@ if [ "$1" == "gcc" ]; then
     ./runner.py -n 5 -c "$*" scimark/run_SOR.c 1000 256
     ./runner.py -n 5 -c "$*" scimark/run_SparseMatMult.c 1000 5000 262144
     ./runner.py -n 5 -c "$*" scimark/run_SparseMatMult.c 100000 1000000 1024
-    ./runner.py -n 5 -c "$*" scimark/run_MonteCarlo 268435456
+    ./runner.py -n 5 -c "$*" scimark/run_MonteCarlo.c 268435456
+    ./runner.py -n 5 -c "$*" scimark/run_LU.c 100 4096
+    ./runner.py -n 5 -c "$*" scimark/run_LU.c 1000 2
     rm a.out
 else
     if [ "$1" == "python2.7" ]; then
@@ -58,4 +60,5 @@ else
     $* ./runner.py $EXTRA_OPTS scimark.py SparseMatMult 100000 1000000 1024
     $* ./runner.py $EXTRA_OPTS scimark.py MonteCarlo 268435456
     $* ./runner.py $EXTRA_OPTS scimark.py LU 100 4096
+    $* ./runner.py $EXTRA_OPTS scimark.py LU 1000 2
 fi

### 2012-08-13 talk/vmil2012/paper.tex
Changes based on Stephan's remarks

--- 2012-08-13 talk/vmil2012/paper.tex
@@
 beginning}
-In this paper we describe and analyze how deoptimization works in the context
-of tracing just-in-time compilers. What instructions are used in the
+\bivab{Introduce in a sentence what guards are}
+In this paper we describe and analyze how guards, as a concept of tracing just-in-time compilers, work. Explicating what concepts are used in the
 intermediate and low-level representation of the JIT instructions and how these
@@
 to guards, explain how they are implemented in the different levels of RPython's
-tracing JIT and explain the rationale behind the design decisions based on the
+tracing JIT and clarify the rationale behind the design decisions based on the
 numbers provided here.
@@
 making guards one of the most common types of operations. Many of these guards
-fail rarely or not all during execution. There are several aspects to consider
+fail rarely or not all during execution.
+
+There are several aspects to consider
 in the design and optimization of guards, the first aspect is that due to the
@@
 \begin{itemize}
-  \item An analysis of guards in the context of RPython's tracing JIT to
-  substantiate the aforementioned observation, based on a set of benchmarks,
+  \item An analysis and benchmark of guards in the context of RPython's tracing Jit.
+  %An analysis of guards in the context of RPython's tracing JIT to
+  %substantiate the aforementioned observation, based on a set of benchmarks,
   \item detailed measurements about the frequency and the
@@
 different low level aspects suited for the target environment are automatically
-added to program such as (if needed) a garbage collector and with some hints
-provided by the author a just-in-time compiler.
+added to the program such as (if needed) a garbage collector
+and a just-in-time compiler based on hints provided by the author.
 
@@
 \label{sub:tracing}
-Tracing JITs are a technique of just-in-time compilers that generate code by
+Tracing is a technique of just-in-time compilers that generate code by
 observing the execution of a program. VMs using tracing JITs typically are
@@
 After profiling identifies an interesting
-path, tracing is started, recording all operations that are executed on this
+path, tracing is started thus recording all operations that are executed on this
 path. This includes inlining functional calls.
@@
     \input{figures/unopt-log.tex}
-    \caption{Unoptimized trace}
+    \caption{Unoptimized trace. Numbers on the right represent the line numbers in the original program. Lines marked with -1 represent lines added by JIT}
     \label{fig:unopt-trace}
@@
 The possible points of deviation from the trace are guard operations
-that check whether the same assumptions observed during tracing
+that check whether the same assumptions observed while tracing
 still hold during execution.
@@
 
-To do this reconstruction, it is necessary to take the values of the SSA
+To do this reconstruction it is necessary to take the values of the SSA
 variables of the trace and build interpreter stack frames.  Tracing
@@
 If a guard fails often enough, a trace is started from it
-to create a trace tree.
+forming a trace tree.
 When that happens another use case of resume data
@@
 After the bridge has been recorded and compiled it is attached to the guard.
-If the guard fails later, the bridge is executed. Therefore the resume data of
+If the guard fails later the bridge is executed. Therefore the resume data of
 that guard is no longer needed.
@@
 is to share parts of the data structure between subsequent guards.
-This is often useful because the density of guards in traces is so high,
+This is useful because the density of guards in traces is so high,
 that quite often not much changes between them.
 Since resume data is a linked list of symbolic frames
-often only the information in the top frame changes from one guard to the next.
+in many cases only the information in the top frame changes from one guard to the next.
 The other symbolic frames can often just be reused.
@@
 Guards interact with optimizations in various ways.
-Most importantly optimizations try to remove as many operations
-and therefore guards as possible.
-This is done with many classical compiler optimizations.
+Using many classical compiler optimizations the JIT tries to remove as many
+operations, and therefore guards, as possible.
 In particular guards can be removed by subexpression elimination.
@@
 This also works if a later guard is weaker
-and therefore implied by a earlier guard.
+and hence implied by a earlier guard.
 
@@
 to reconstruct the objects that were not allocated so far when a guard fails.
-Therefore the resume data needs to store enough information
+Consequently the resume data needs to store enough information
 to make this reconstruction possible.
 
-Adding this additional information is done as follows.
+Adding this additional information is done as follows:
 So far, every variable in the symbolic frames
@@
 
-After optimization the resulting trace is handed to the over platform specific
+After optimization the resulting trace is handed over to the platform specific
 backend to be compiled to machine code. The compilation phase consists of two
@@
 the forward pass IR-level variables are assigned to registers and stack
-locations by the register allocator according to the requirements of the to be
+locations by the register allocator according to the requirements of the
 emitted instructions.  Eviction/spilling is performed based on the live range
@@
 into one or more machine level instructions that implement the required
-semantics, operations without side effects whose result is not used are not
+semantics. Operations without side effects whose result is not used are not
 emitted. Guards instructions are transformed into fast checks at the machine
@@
   \end{minipage}
-  \caption{Separated and merged compilation of operations and guards}
+  \caption{Result of separated (left) and merged (right) compilation of operations and guards (top).}
   \label{fig:trace-compiled}
@@
 
-Each guard in the IR has attached to it a list of the IR-variables required to
+Attached to each guard in the IR is a list of the IR-variables required to
 rebuild the execution state in case the trace is left through
 the guard. When a guard is compiled, in addition to the
-condition check two things are generated/compiled. First a special data
-structure called \emph{low-level resume data} is created that encodes the
-information provided by the register allocator about where the values
-corresponding to each IR-variable required by the guard will be stored when
-execution reaches the code emitted for the corresponding guard. This data
-structure stores the data in a compressed manner using an encoding that uses
-8bits to store 7bits of information. This encoding is efficient to create and
+condition check two things are generated/compiled.
+
+First a special data
+structure called \emph{low-level resume data} is created. This data structure encodes the
+information about where, i.e. which register or stack location, the IR-variables required to rebuild the state will be stored when the guard is executed.
+This data
+structure stores the values in a succinct manner using an encoding that uses
+8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
 provides a compact representation of the needed information,
@@
 Second a piece of code is generated for each guard that acts as a trampoline.
-Guards are implemented as a conditional jump to this trampoline. In case the
-condition checked in the guard fails
-execution jumps to the corresponding trampoline. In the trampoline the pointer to the
-\emph{low-level resume data} is loaded and jumps to generic bailout handler, also known as compensation code,
+Guards are implemented as a conditional jump to this trampoline in case the
+guard checked fails.
+In the trampoline the pointer to the
+\emph{low-level resume data} is loaded and execution jumps to a generic bailout handler, also known as compensation code,
 that is used to leave the compiled trace in case of a guard failure.
@@
 the execution takes one of the side exits due to a guard failure. At the same
-time the data stored in the backend needed to rebuild the state needs to be as
+time the data stored in the backend, required to rebuild the state, should be as
 compact as possible to reduce the memory overhead produced by the large number
@@
 a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
-compiled. When compiling bridges the goal is that future failures of the guards
-that led to the compilation of the bridge should execute the bridge without
-additional overhead. In particular the failure of the guard should not lead
+compiled.
+Since the goal of compiling bridges is to improve execution speed on the
+diverged path (failing guard) they should not introduce additional overhead.
+In particular the failure of the guard should not lead
 to leaving the compiled code prior to execution the code of the bridge.
@@
 
-Once the bridge has been compiled the guard that led to compiling the bridge is
+Once the bridge has been compiled the corresponding guard is
 patched to redirect control flow to the bridge in case the check fails. In
@@
 
-All figures in this section do not take garbage collection of machine code into account. Pieces
+All measurements presented in this section do not take garbage collection of machine code into account. Pieces
 of machine code can be globally invalidated or just become cold again. In both
@@
 Tracing JIT compilers only compile the subset of the code executed in a program
-that is traced in a hot loop, for this reason the amount of generated machine
+that occurs in a hot loop, for this reason the amount of generated machine
 code will be smaller than in other juts-in-time compilation approaches.  This
@@
 efficiently through using the techniques described above. On the other hand
-comparing the results to the xz compression which only requires between 17.1\%
+comparing the results to the xz compression which only needs between 17.1\%
 and 21.1\% of the space required by our compression shows that the compression
@@
 Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
-relative amounts of guards that ever fail and of guards that fail more than 200
-times.\footnote{
-    The threshold of 200 is rather high. It was picked experimentally to give
+relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.
+\footnote{
+    The threshold used is 200 failures. This rather high threshold was picked experimentally to give
     good results for long-running programs.
 }
-As described before, for guards that fail more than 200 times, a trace
-is recorded that starts from the guard. Afterwards the guard is patched so that later
-failures execute the new trace instead of jumping to the trampoline and returning to the interpreter. Hence the
-numbers presented for guards that fail more than 200 times represent the 200
+
+After the guard is patched
+failures execute the new bridge instead of jumping to the trampoline and returning to the interpreter. Hence the
+numbers presented for guards that have a bridge represent the
 failures up to the compilation of the bridge and all executions of the then
@@
 enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
-fail more than 200 times. Also of all failing guards a few fail extremely often
+fail often enough that a bridge is compiled. Also of all failing guards a few fail extremely often
 and most fail rarely. The results emphasizes that as most of the guards never
@@
 flow divergence in recorded traces.
-We described how, based on the observation that guards are a frequent operation
-in traces and that they do not fail often, guards have been implemented in the
+Based on the observation that guards are a frequent operation in traces and
+that they do not fail often, we described how they have been implemented in the
 high and low level components of RPython's tracing JIT compiler.
@@
 benchmark set to evaluate previous observations and assumptions. Our
-experiments showed that, as previously assumed, guards are a very common
+experiments confirmed that guards are a very common
 operation in traces. At the same time guards are associated with a high
@@
 showed that the compression techniques used in PyPy effectively reduce the
-overhead of guards, while it still produces a significant overhead. The results
+overhead of guards, but it still produces a significant overhead. The results
 also showed that guard failure is a local event: there are few
@@
 case of bailout due to having to decode compressed state representation.
-The compressed state representation is reduces the memory footprint of rarely
+The compressed state representation reduces the memory footprint of rarely
 used data.
 
-Based on the observation that most guards do not fail very often or at all it
+Based on the observation that guard failure is rare it
 would be worth exploring if a more aggressive compression scheme for guards

### 2012-08-13 talk/vmil2012/paper.tex
add acknowledgements
missing word
Address some of the open tasks in the introduction and abstract sections

@@
 \begin{abstract}
-Guards operations occur frequently in traces generated by tracing just-in-time
-(JIT) compilers. Therefore it is important to design and implement them
-carefully to find the right trade-off between execution speed, deoptimization,
+Tracing just-in-time (JIT) compilers record linear control flow paths,
+inserting operations called guards at points of possible divergence. These
+operations occur frequently generated traces and therefore it is important to
+design and implement them carefully to find the right trade-off between
+execution speed, deoptimization,
 and memory overhead.  In this paper we describe the design decisions about
@@
 beginning}
-\bivab{Introduce in a sentence what guards are}
-In this paper we describe and analyze how guards, as a concept of tracing just-in-time compilers, work. Explicating what concepts are used in the
-intermediate and low-level representation of the JIT instructions and how these
-are implemented.
+Tracing just-in-time (JIT) compilers record and compile linear control flow paths of operations executed by an interpreter
+inserting operations called guards at points of possible divergence.
+In this paper we describe and analyze how guards work, explaining what concepts
+are used in the intermediate and low-level representation of the JIT
+instructions and how these are implemented.
 
@@
 implementing guards. Guards have a runtime cost, they take time to execute. On
-the other hand, guards are possible deoptimization points. They need to store
+the other hand, guards are possible deoptimization points, meaning the recorded
+and compiled path has to be left returning control to the interpreter. They need to
+store
 enough information to rebuild the interpreter state.
@@
 goal is to present concrete numbers for the frequency and the overhead related
-to guards, explain how they are implemented in the different levels of RPython's
+to guards, explain how they are implemented in the different levels of this particular
 tracing JIT and clarify the rationale behind the design decisions based on the
@@
 
-\cfbolz{this paragraph now suddenly \emph{introduces} guards, despite having talked about them already}
 The operations executed by an interpreter are recorded by the tracing JIT in
 case they are frequently executed (this process is described in more detail in
-Section \ref{sec:Resume Data}). During the recording phase special operations,
-referred to as \texttt{guards}, are inserted into the recorded trace at all
+Section \ref{sec:Resume Data}). During the recording phase \texttt{guards} are
+inserted into the recorded trace at all
 points where the control flow could diverge. As can be seen in
 Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
-operations before and for about 15\% to 20\% of the operations after the
-optimization pass over the traced and later compiled parts of the benchmarks,
-making guards one of the most common types of operations. Many of these guards
+operations before and for about 15\% to 20\% of the operations after
+optimizing the traces generated for the different benchmarks used in this paper.
+This makes guards one of the most common types of operations. As this paper will show, many of these guards
 fail rarely or not all during execution.
@@
 In this paper we want to substantiate the aforementioned observations and
-describe based on them the reasoning behind and the implementation of guards in
+describe based on them the reasoning behind the implementation of guards in
 RPython's tracing just-in-time compiler. the contributions of this paper are:
 \begin{itemize}
-  \item An analysis and benchmark of guards in the context of RPython's tracing Jit.
+  \item An analysis and benchmark of guards in the context of RPython's tracing JIT.
   %An analysis of guards in the context of RPython's tracing JIT to
@@
   \item a description about how guards are implemented in the high\-
-  and low-level parts of the JIT and describe the rationale behind the design.
+  and low-level components of the JIT and describe the rationale behind the design.
 \end{itemize}
@@
 the frontend\bivab{better term for this?} related to recording and storing the
-information required to restore the interpreter state in case of a guard
+information required to rebuild the interpreter state in case of a guard
 failure. Once the frontend has traced and optimized a loop it invokes the
@@
 \section*{Acknowledgements}
+We would like to thank David Edelsohn and Stephan Zalewski for their helpful
+feedback and valuable comments while writing this paper.
+
 \section*{Appendix}

### 2012-08-13 talk/vmil2012/paper.tex
improvements for background section

@@
 language experimentation and extension. PyPy is now a fully compatible
-alternative implementation of the Python language\bivab{mention speed}. The
+alternative implementation of the Python language, that is on average about 5 times faster than the reference implementation. The
 implementation takes advantage of the language features provided by RPython
@@
 Python interpreter have developed into a general environment for experimenting
-and developing fast and maintainable dynamic language implementations.
-\bivab{Mention the different language impls}
+and developing fast and maintainable dynamic language implementations. There
+are, besides the Python interpreter, implementations of Prolog, Javascript, R,
+Smalltalk among other that are written in RPython at different levels of
+completeness.
 
@@
 added to the program such as (if needed) a garbage collector
-and a just-in-time compiler based on hints provided by the author.
+and based on hints provided by the author a just-in-time compiler.
 
@@
 Tracing is a technique of just-in-time compilers that generate code by
-observing the execution of a program. VMs using tracing JITs typically are
+observing the execution of a program. VMs using tracing JITs are typically
 mixed-mode execution environments that also contain an interpreter. The
@@
 Figure~\ref{fig:example} shows an example RPython function that checks
-whether a number reduces to 1 with less than 100 steps of the Collatz process.
+whether a number reduces to 1 with less than 100 steps of the Collatz process.\footnote{\url{http://en.wikipedia.org/wiki/Collatz_conjecture}}
 It uses an \lstinline{Even} and an \lstinline{Odd} class to box the numbers, to
@@
     \input{figures/unopt-log.tex}
-    \caption{Unoptimized trace. Numbers on the right represent the line numbers in the original program. Lines marked with -1 represent lines added by JIT}
+    \caption{Unoptimized trace, the line numbers in the trace correspond to the line numbers in Figure~\ref{fig:trace-log}.}
     \label{fig:unopt-trace}

### 2012-08-13 talk/vmil2012/paper.tex
improvements to guards in the frontend section

@@
 the full control flow of a program is not observed.
-The possible points of deviation from the trace are guard operations
+The possible points of deviation from the trace are denoted by guard operations
 that check whether the same assumptions observed while tracing
@@
 This also works if a later guard is weaker
-and hence implied by a earlier guard.
+and hence implied by an earlier guard.
 
@@
 Figure~\ref{fig:trace-log} shows the optimized version of the trace in
-Figure~\ref{fig:fig:unopt-trace}. Allocation removal has removed the
+Figure~\ref{fig:unopt-trace}. Allocation removal has removed the
 \lstinline{new} operation and other operations handling the instance. The

### 2012-08-13 talk/vmil2012/paper.tex
update figure with the correct variable indices and add the line numbers to the trace operations
improve the guards in the backend section

@@
 %___________________________________________________________________________
+\todo{find a better name for \texttt{low-level resume data}}
+\todo{find better names for JIT front- and backend}
 \section{Introduction}
@@
 checked by the guard is not used anywhere else the guard and the operation
-producing the value can often be merged, further reducing even more the overhead of the guard.
-Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
-followed by a guard that checks the result of the operation are compiled to
+producing the value can often be merged, further reducing the overhead of the guard.
+Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
+followed by a \texttt{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
 pseudo-assembler if the operation and the guard are compiled separated or if
@@
   \begin{minipage}{1\columnwidth}
-    \begin{lstlisting}[mathescape]
-$b_1$ = int_eq($i_2$, 1)
-guard_false($b_1$)
+\begin{lstlisting}[mathescape, numbers=right, escapechar=|, firstnumber=18]
+$b_3$ = int_eq($i_5$, 1)                    |\setcounter{lstnumber}{17}|
+guard_false($b_3$)                          |\setcounter{lstnumber}{-1}|
 \end{lstlisting}
@@
 8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
-provides a compact representation of the needed information,
+provides a compact representation of the needed information in order
 to maintain an acceptable memory profile.
 
-Second a piece of code is generated for each guard that acts as a trampoline.
+Second for each guard a piece of code is generated that acts as a trampoline.
 Guards are implemented as a conditional jump to this trampoline in case the
-guard checked fails.
+guard check fails.
 In the trampoline the pointer to the
-\emph{low-level resume data} is loaded and execution jumps to a generic bailout handler, also known as compensation code,
+\emph{low-level resume data} is loaded and after storing the current execution state
+(registers and stack) execution jumps to a generic bailout handler, also known
+as \texttt{compensation code},
 that is used to leave the compiled trace in case of a guard failure.
@@
 a fast on-trace profile and a potentially slow one in the bailout case where
-the execution takes one of the side exits due to a guard failure. At the same
+the execution has to return to the interpreter due to a guard failure. At the same
 time the data stored in the backend, required to rebuild the state, should be as

### 2012-08-13 talk/vmil2012/paper.tex
hide Appendix

@@
 Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
-followed by a \texttt{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
+followed by a \texttt{guard\_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
 pseudo-assembler if the operation and the guard are compiled separated or if
@@
 
-\section*{Appendix}
-\todo{remove this section and the figures}
-\begin{figure*}
-    \include{figures/ops_count_table}
-    \caption{Relative numbers of operations in the traces generated for
-    different benchmarks}
-    \label{fig:ops_count}
-\end{figure*}
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
-\caption{Relative frequency of operations before and after optimization}
-\label{fig:ops_pie}
-\end{figure*}
+%\section*{Appendix}
+%\todo{remove this section and the figures}
+%\begin{figure*}
+%    \include{figures/ops_count_table}
+%    \caption{Relative numbers of operations in the traces generated for
+%    different benchmarks}
+%    \label{fig:ops_count}
+%\end{figure*}
+%\begin{figure*}
+%\centering
+%\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
+%\caption{Relative frequency of operations before and after optimization}
+%\label{fig:ops_pie}
+%\end{figure*}
 \bibliographystyle{abbrv}

### 2012-08-13 talk/vmil2012/paper.tex
typos

--- 2012-08-13 talk/vmil2012/paper.tex
@@
 2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
-enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
+enough that a bridge is compiled for them, only 1.2\% to 3.6\% of all guards
 fail often enough that a bridge is compiled. Also of all failing guards a few fail extremely often
-and most fail rarely. The results emphasizes that as most of the guards never
+and most fail rarely. The results emphasize that as most of the guards never
 fail it is important to make sure that the successful execution of a guard does

### 2012-08-14 talk/vmil2012/paper.tex
Improve related work, evaluation and conclusion sections

@@
 \todo{find better names for JIT front- and backend}
+\todo{mention somewhere that it is to be expected that most guards do not fail}
 \section{Introduction}
@@
 
-The goal of this paper is to understand the design constraints when
+Our aim is to help understand the design constraints when
 implementing guards. Guards have a runtime cost, they take time to execute. On
 the other hand, guards are possible deoptimization points, meaning the recorded
-and compiled path has to be left returning control to the interpreter. They need to
-store
-enough information to rebuild the interpreter state.
+and compiled path has to be left returning control to the interpreter. They need
+enough associated information to enable rebuilding the interpreter state.
 Based on the informal observation that guards are among the most common
@@
 \todo{improve the table formatting}
-\todo{give a reference to the benchmark scripts to make things repeatable}
 
@@
 guards in the machine code
-backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} All
+backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} The
+tools used to run and evaluate the benchmarks including the patches applied to
+the PyPy sourcecode can be found in the repository for this
+paper.\footnote{\url{https://bitbucket.org/pypy/extradoc/src/tip/talk/vmil2012}}
+All
 benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
@@
 summarized in the remainder of this section. We want to point out three
-aspects of guards in particular
+aspects of guards in particular:
 \begin{itemize}
@@
 
-Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
+Figure~\ref{fig:benchmarks} extends Figure~\ref{fig:guard_percent} and summarizes the total number of operations that were
 recorded during tracing for each of the benchmarks and what percentage of these
@@
 and optimized traces. The Figure shows that the overall optimization rate for
-operations which is between 69.4\% and 83.89\% of the traced operations and the
+operations, which is between 69.4\% and 83.89\%, of the traced operations and the
 optimization rate of guards, which is between 65.8\% and 86.2\% of the
-operations, are very similar, as could be assumed based on
-Figure~\ref{fig:guard_percent}. This indicates that the optimizer can remove
-most of the guards, but after the optimization pass guards still account for
+operations, are very similar. This indicates that the optimizer can remove
+most of the guards, but after the optimization pass these still account for
 15.2\% to 20.2\% of the operations being compiled and later executed.
@@
 Why the efficient storing of the \texttt{resume data} is a central concern in the design
-of guards is illustrated by Figure~\ref{fig:backend_data}. This figure shows
+of guards is illustrated by Figure~\ref{fig:resume_data_sizes}. This figure shows
 the size of the compressed \texttt{resume data}, the approximated size of
@@
 shows that large parts of the resume data are redundant and can be stored more
-efficiently through using the techniques described above. On the other hand
+efficiently using the techniques described earlier. On the other hand
 comparing the results to the xz compression which only needs between 17.1\%
@@
 is not optimal but a trade-off between the required space and the time needed
-to build a good compressed representation of the compressed resume data for the
+to build a good, compressed representation of the resume data for the
 large amount of guards present in the traces.
@@
 
-After the guard is patched
-failures execute the new bridge instead of jumping to the trampoline and returning to the interpreter. Hence the
-numbers presented for guards that have a bridge represent the
+The numbers presented for guards that have a bridge represent the
 failures up to the compilation of the bridge and all executions of the then
@@
 enough that a bridge is compiled for them, only 1.2\% to 3.6\% of all guards
-fail often enough that a bridge is compiled. Also of all failing guards a few fail extremely often
+fail often enough that a bridge is compiled. Also, of all failing guards a few fail extremely often
 and most fail rarely. The results emphasize that as most of the guards never
@@
 
-Guards as described are a concept associated with tracing just-in-time
+Guards, as described, are a concept associated with tracing just-in-time
 compilers to represent possible divergent control flow paths.
@@
 snapshots for guards in LuaJIT are associated with a large memory footprint.
-The solution used in there is to store sparse snapshots, avoiding the creation
+The solution used there is to store sparse snapshots, avoiding the creation
 of snapshots for every guard to reduce memory pressure. Snapshots are only
@@
 Once a new hot trace is emitted into the fragment cache it is linked to side
-exit that led to the compilation. Fragment Linking avoids the performance
-penalty involved in leaving the compiled and it to remove the compensation
-code used when restoring the machine state on a side exit.
+exit that led to the compilation of the fragment. Fragment Linking avoids the
+performance penalty involved in leaving the compiled code. Fragment linking
+also allows to remove compensation code associated to the linked fragments that
+would have been required to restored the execution state on the side exit.
 
-Gal et. al~\cite{Gal:2006} describe that in the HotpathVM they experimented
+Gal et. al~\cite{Gal:2006} describe how in the HotpathVM they experimented
 with having one generic compensation code block, like the RPython JIT, that
@@
 \label{sec:Conclusion}
-In this paper we have concentrated on guards, an operation typically found in
+In this paper we have concentrated on guards, an operation found in
 tracing just-in-time compilers and used to denote points of possible control
@@
 
-Finally we have presented experimental data collected using the standard PyPy
+Additionally we have presented experimental data collected using the standard PyPy
 benchmark set to evaluate previous observations and assumptions. Our
@@
 successful guard checks as much as possible while paying a higher price in the
-case of bailout due to having to decode compressed state representation.
+case of bailout due to having to decode a compressed state representation.
 The compressed state representation reduces the memory footprint of rarely

### 2012-08-14 talk/iwtc11/benchmarks/benchmark.sh
lua version of runner

@@ -24,6 +24,9 @@ if [ "$1" == "gcc" ]; then
     ./runner.py -n 5 -c "$*" scimark/run_LU.c 100 4096
     ./runner.py -n 5 -c "$*" scimark/run_LU.c 1000 2
     rm a.out
+elif [[ "$1" == luajit* ]]; then
+    $* runner.lua SOR 100 32768
+    $* runner.lua SOR 1000 256
 else
     if [ "$1" == "python2.7" ]; then
         EXTRA_OPTS='-w 0 -n 1'

### 2012-08-14 talk/iwtc11/benchmarks/benchmark.sh
lua version of runner

@@ -1,0 +1,67 @@
+#!/bin/bash
+
+echo
+echo $*
+if [ "$1" == "gcc" ]; then
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_double.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_long.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_fix16.c
+    #./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1
+    #./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 100
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 100
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1000
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1000
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/conv3x3.cc 1000000 3
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/conv3x3.cc 1000 1000
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/dilate3x3.cc 1000 1000
+    ./runner.py -n 5 -c "$* -lstdc++" image/sobel.cc 1000 1000
+    ./runner.py -n 5 -c "$*" scimark/run_SOR.c 100 32768
+    ./runner.py -n 5 -c "$*" scimark/run_SOR.c 1000 256
+    ./runner.py -n 5 -c "$*" scimark/run_SparseMatMult.c 1000 5000 262144
+    ./runner.py -n 5 -c "$*" scimark/run_SparseMatMult.c 100000 1000000 1024
+    ./runner.py -n 5 -c "$*" scimark/run_MonteCarlo.c 268435456
+    ./runner.py -n 5 -c "$*" scimark/run_LU.c 100 4096
+    ./runner.py -n 5 -c "$*" scimark/run_LU.c 1000 2
+    rm a.out
+elif [[ "$1" == luajit* ]]; then
+    $* runner.lua SOR 100 32768
+    $* runner.lua SOR 1000 256
+else
+    if [ "$1" == "python2.7" ]; then
+        EXTRA_OPTS='-w 0 -n 1'
+    fi
+    if [ "$1" == "python2.6" ]; then
[…]

### 2012-08-14 talk/vmil2012/paper.tex
cite Self debugging paper

--- 2012-08-14 talk/vmil2012/paper.tex
@@
 to make it possible to do source-level debugging of JIT-compiled code.
-Self uses deoptimization to reach the same goal~\cite{XXX}.
+Self uses deoptimization to reach the same goal~\cite{holzle_debugging_1992}.
 When a function is to be debugged, the optimized code version is left
@@
 that need to be re-created when leaving the optimized code.
-The scope descriptors are between 0.45 and 0.76 times
+The scope descriptors are between 0.42 and 1.09 times
 the size of the generated machine code.
+All the information needed for debugging together
+is between 1.22 and 2.33 times the size of generated machine code,
+according to the paper.
 

### 2012-08-14 talk/vmil2012/paper.tex
replace front- and backend with high- and low-level components

@@
 \todo{find a better name for \texttt{low-level resume data}}
-\todo{find better names for JIT front- and backend}
 \todo{mention somewhere that it is to be expected that most guards do not fail}
@@
 Data} we proceed to describe for RPython's tracing JIT the details of guards in
-the frontend\bivab{better term for this?} related to recording and storing the
+the high-level, target independant, component of the JIT related to recording and storing the
 information required to rebuild the interpreter state in case of a guard
-failure. Once the frontend has traced and optimized a loop it invokes the
-backend to compile the operations to machine code, Section \ref{sec:Guards in
-the Backend} describes the low-level aspects of how guards are implemented in
-the JIT-backend. The frequency of guards and the overhead associated with the
+failure. Once the high-level component has traced and optimized a loop it invokes the
+low-level component to compile the operations to machine code, Section \ref{sec:Guards in
+the Backend} describes the aspects of the implementaion of guards in
+the low-level JIT component. The frequency of guards and the overhead associated with the
 implementation described in this paper is discussed in
@@
 
-\section{Guards in the Frontend} %{Resume Data}
+\section{Guards in the High-Level JIT Components} %{Resume Data}
 \label{sec:Resume Data}
@@
 % section Resume Data (end)
-\section{Guards in the Backend}
+\section{Guards in the Low-Level JIT Component}
 \label{sec:Guards in the Backend}
@@
 After optimization the resulting trace is handed over to the platform specific
-backend to be compiled to machine code. The compilation phase consists of two
+low-level implementation to be compiled to machine code. The compilation phase consists of two
 passes over the lists of instructions, a backwards pass to calculate live
@@
 saved execution state the values that the IR-variables had at the time of the
-guard failure and stores them in a location that can be read by the frontend.
-After saving the information the control is passed to the frontend signaling
-which guard failed so the frontend can read the information passed and restore
-the state corresponding to the point in the program.
+guard failure and stores them in a location that can be read by the high-level component.
+After saving the information the control is returned to the high-level component signaling
+which guard failed. The stored low-level information can be read and used to restore
+the execution state corresponding to the guard failure in the interpreter.
 
@@
 the execution has to return to the interpreter due to a guard failure. At the same
-time the data stored in the backend, required to rebuild the state, should be as
+time the data stored in the low-level component, required to rebuild the state, should be as
 compact as possible to reduce the memory overhead produced by the large number
@@
 \texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
-The benchmarks were run on a version of PyPy based on the
-tag~\texttt{0b77afaafdd0} and patched to collect additional data about the
-guards in the machine code
-backends.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} The
+The benchmarks were run on a version of PyPy based on
+revision~\texttt{0b77afaafdd0} and patched to collect additional data about the
+guards in the generated machine code.\footnote{\url{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0}} The
 tools used to run and evaluate the benchmarks including the patches applied to
@@
 shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
-of the code and of the data generated by the machine code backend and an
+of the code and of the data generated by the low-level component and an
 approximation of the size of the \texttt{resume data} structures for the

### 2012-08-14 talk/vmil2012/paper.tex
Settle for front- and backend

--- 2012-08-14 talk/vmil2012/paper.tex
@@
 Data} we proceed to describe for RPython's tracing JIT the details of guards in
-the frontend\bivab{better term for this?} related to recording and storing the
+the frontend related to recording and storing the
 information required to rebuild the interpreter state in case of a guard
@@
 the Backend} describes the low-level aspects of how guards are implemented in
-the JIT-backend. The frequency of guards and the overhead associated with the
+the machine specific JIT-backend. The frequency of guards and the overhead associated with the
 implementation described in this paper is discussed in

### 2012-08-14 talk/vmil2012/paper.tex
We are using a specific revision and not a tag anymore

@@
 The benchmarks were run on a version of PyPy based on the
-tag~\texttt{0b77afaafdd0} and patched to collect additional data about the
+revision~\texttt{0b77afaafdd0} and patched to collect additional data about the
 guards in the machine code

### 2012-08-14 talk/vmil2012/tool/build_tables.py
rename low-level resume data into backend map

@@ -163,7 +163,7 @@ def build_backend_count_table(csvfiles, texfile, template):
     head = [r'Benchmark',
             r'Code',
             r'Resume data',
-            r'll data',
+            r'Backend map',
             r'Relation']
 
     table = []

### 2012-08-14 talk/vmil2012/paper.tex
rename low-level resume data into backend map

@@
 %___________________________________________________________________________
-\todo{find a better name for \texttt{low-level resume data}}
 \todo{mention somewhere that it is to be expected that most guards do not fail}
@@
 First a special data
-structure called \emph{low-level resume data} is created. This data structure encodes the
-information about where, i.e. which register or stack location, the IR-variables required to rebuild the state will be stored when the guard is executed.
+structure called \emph{backend map} is created. This data structure encodes the
+mapping from the IR-variables needed by the guard to rebuild the state to the
+low-level locations (registers and stack) where the corresponding values will
+be stored when the guard is executed.
 This data
@@
 In the trampoline the pointer to the
-\emph{low-level resume data} is loaded and after storing the current execution state
+backend map is loaded and after storing the current execution state
 (registers and stack) execution jumps to a generic bailout handler, also known
-as \texttt{compensation code},
+as \emph{compensation code},
 that is used to leave the compiled trace in case of a guard failure.
@@
 Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
-Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
+Guard \#1 jumps to the trampoline, loads the \texttt{backend map} and
 then calls the bailout handler, whereas Guard \#2 has already been patched
@@
 The overhead that is incurred by the JIT to manage the \texttt{resume data},
-the \texttt{low-level resume data} as well as the generated machine code is
+the \texttt{backend map} as well as the generated machine code is
 shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
@@
 and a set of support functions that are generated when the JIT starts and which
-are shared by all compiled traces. The size of the \texttt{low-level resume
-data} is the size of the compressed mapping from registers and stack to
+are shared by all compiled traces. The size of the \texttt{backend map}
+is the size of the compressed mapping from registers and stack to
 IR-level variables and finally the size of the \texttt{resume data} is an
@@
 
-For the different benchmarks the \texttt{low-level resume data} has a size of
+For the different benchmarks the \texttt{backend map} has a size of
 about 15\% to 20\% of the amount of memory compared to the size of the
 generated machine code. On the other hand the generated machine code has only a
-size ranging from 20.5\% to 37.98\% of the size of the high and low-level
-resume data combined and being compressed as described before.
+size ranging from 20.5\% to 37.98\% of the size of the resume data and the backend map
+combined and being compressed as described before.
 

### 2012-08-14 talk/vmil2012/paper.tex
Remove most of \texttt annotations

@@
 case they are frequently executed (this process is described in more detail in
-Section \ref{sec:Resume Data}). During the recording phase \texttt{guards} are
+Section \ref{sec:Resume Data}). During the recording phase guards are
 inserted into the recorded trace at all
@@
 As explained in previous sections, when a specific guard has failed often enough
-a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
-compiled.
+a bridge starting from this guard is recorded and compiled.
 Since the goal of compiling bridges is to improve execution speed on the
@@
 Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
-Guard \#1 jumps to the trampoline, loads the \texttt{backend map} and
+Guard \#1 jumps to the trampoline, loads the backend map and
 then calls the bailout handler, whereas Guard \#2 has already been patched
@@
 
-The overhead that is incurred by the JIT to manage the \texttt{resume data},
-the \texttt{backend map} as well as the generated machine code is
+The overhead that is incurred by the JIT to manage the resume data,
+the backend map as well as the generated machine code is
 shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
 of the code and of the data generated by the machine code backend and an
-approximation of the size of the \texttt{resume data} structures for the
+approximation of the size of the resume data structures for the
 different benchmarks mentioned above. The machine code taken into account is
@@
 and a set of support functions that are generated when the JIT starts and which
-are shared by all compiled traces. The size of the \texttt{backend map}
+are shared by all compiled traces. The size of the backend map
 is the size of the compressed mapping from registers and stack to
-IR-level variables and finally the size of the \texttt{resume data} is an
+IR-level variables and finally the size of the resume data is an
 approximation of the size of the compressed high-level resume data as described
@@
 
-For the different benchmarks the \texttt{backend map} has a size of
+For the different benchmarks the backend map has a size of
 about 15\% to 20\% of the amount of memory compared to the size of the
@@
 code will be smaller than in other juts-in-time compilation approaches.  This
-creates a larger discrepancy between the size of the \texttt{resume data} when
-compared to the size of the generated machine code and illustrates why it is important to compress the \texttt{resume data} information.
+creates a larger discrepancy between the size of the resume data when
+compared to the size of the generated machine code and illustrates why it is important to compress the resume data information.
 
@@
 
-Why the efficient storing of the \texttt{resume data} is a central concern in the design
+Why the efficient storing of the resume data is a central concern in the design
 of guards is illustrated by Figure~\ref{fig:resume_data_sizes}. This figure shows
-the size of the compressed \texttt{resume data}, the approximated size of
-storing the \texttt{resume data} without compression and
+the size of the compressed resume data, the approximated size of
+storing the resume data without compression and
 an approximation of the best possible compression of the resume data by
 compressing the data using the
-\texttt{xz} compression tool, which is a ``general-purpose data compression
+\emph{xz} compression tool, which is a ``general-purpose data compression
 software with high compression ratio''.\footnote{\url{http://tukaani.org/xz/}}

### 2012-08-14 talk/vmil2012/paper.tex
extend paragraph about TraceMonkey (thanks arigo)

@@
 
-TraceMonkey, a tracing JIT for JavaScript, uses trace stitching
-to avoid the overhead of returning to the trace monitor and calling another
-trace when taking a side exit~\cite{Gal:2009ux}. In this approach it is required to write live
-values to an activation record before entering the new trace.
+For TraceMonkey, a tracing JIT for JavaScript, Gal et. al~\cite{Gal:2009ux}
+illustrate how it uses a small off-trace set of instructions that is
+executed in case a guard failure to return a structure describing the reason
+for the exit along with the information needed to restored the interpreter
+state. TraceMonkey uses trace stitching to avoid the overhead of returning to
+the trace monitor and calling another trace when taking a side exit. In this
+approach it is required to write live values to an activation record before
+entering the new trace.
 

### 2012-08-14 talk/vmil2012/paper.tex
Rephrase
Mention Armin's hint about continulet-jit-2

@@
 
-For the different benchmarks the backend map has a size of
-about 15\% to 20\% of the amount of memory compared to the size of the
+For the different benchmarks the backend map has
+about 15\% to 20\% of the size compared to the size of the
 generated machine code. On the other hand the generated machine code has only a
@@
 overhead. Based on the same observation we would like to explore the concept of
-LuaJIT's sparse snapshots and its applicability to PyPy.
+LuaJIT's sparse snapshots and its applicability to RPython's JIT.
+There is an ongoing effort to replace the backend map in RPython's JIT with a
+simpler technique that does not require decoding the backend map on each guard
+failure.
 

### 2012-08-14 talk/vmil2012/paper.tex
Some formatting for guard compilation figure

@@
 
-\todo{Figure needs better formatting}
+%\todo{Figure needs better formatting}
 \begin{figure}[ht]
@@
   \begin{minipage}{1\columnwidth}
-\begin{lstlisting}[mathescape, numbers=right, escapechar=|, firstnumber=18]
+\begin{lstlisting}[xleftmargin=20pt,xrightmargin=20pt,framexleftmargin=5pt,framexrightmargin=-10pt,mathescape, numbers=right, escapechar=|, firstnumber=18,frame=b]
 $b_3$ = int_eq($i_5$, 1)                    |\setcounter{lstnumber}{17}|
@@
   \end{minipage}
-  \caption{Result of separated (left) and merged (right) compilation of operations and guards (top).}
+  \caption{Result of separated (left) and merged (right) compilation of one guard and the following operation (top).}
   \label{fig:trace-compiled}

### 2012-08-14 talk/vmil2012/paper.tex
get rid of the strangeish full stops at the end of non-sentences

@@
 describe based on them the reasoning behind the implementation of guards in
-RPython's tracing just-in-time compiler. the contributions of this paper are:
+RPython's tracing just-in-time compiler. The contributions of this paper are:
 \begin{itemize}
-  \item An analysis and benchmark of guards in the context of RPython's tracing JIT.
+  \item An analysis and benchmark of guards in the context of RPython's tracing JIT,
   %An analysis of guards in the context of RPython's tracing JIT to
@@
   \item a description about how guards are implemented in the high\-
-  and low-level components of the JIT and describe the rationale behind the design.
+  and low-level components of the JIT and a description of the rationale behind the design.
 \end{itemize}

### 2012-08-14 talk/dls2012/paper.tex
copyright data
add remaining scimark lua benchmarks to runner

@@
 
-\conferenceinfo{IWTC '11}{XXX} 
-\copyrightyear{2011} 
-\copyrightdata{[to be supplied]}
+\conferenceinfo{DLS'12,} {October 22, 2012, Tucson, Arizona, USA.}
+\CopyrightYear{2012}
+\copyrightdata{978-1-4503-1564-7/12/10}
 
@@
 project.\footnote{\texttt{http://pypy.org}} PyPy is a framework for implementing
-dynamic languages efficiently~\cite{armin_rigo_pypys_2006}. When implementing a
-language with PyPy, one writes an interpreter for the language in RPython~\cite{davide_ancona_rpython:_2007}.
+dynamic languages efficiently~\cite{rigo_pypys_2006}. When implementing a
+language with PyPy, one writes an interpreter for the language in RPython~\cite{ancona_rpython:_2007}.
 RPython (``Restricted Python``) is a subset
@@
 value.\footnote{We perform a type-based alias analysis to know which
-writes can affect which reads~\cite{XXX}. In addition writes on newly allocated objects
-can never change the value of old existing ones.}
+writes can affect which reads~\cite{diwan_type-based_1998}. In addition writes
+on newly allocated objects can never change the value of old existing ones.}
 
@@
 for example our Prolog interpreter written in
-RPython~\cite{carl_friedrich_bolz_towards_2010}. Prolog programs often contain
+RPython~\cite{bolz_towards_2010}. Prolog programs often contain
 tight
@@
 explicit algorithms.
-\cfbolz{more explicit listing of prior work goes here}
+
+Loop invariant code motion has been part of early compilers in the 1960s and
+1970s~\cite{allen_catalogue_1971}. The approach for achieving loop invariant
+code motion is typically to perform partial redundancy elimination. The
+approach was first proposed by Morel and Renvoise~\cite{morel_global_1979}. It
+involves solving data flow problems usually involding bidirection data flow
+equations. After improvements~\cite{chow_portable_1984,
+dhamdhere_practical_1991} this approach was followed by the work of Knoop
+et.al.~\cite{knoop_lazy_1992} who cleany separated the problem into a backward
+and forward data flow analysis. Implementing partial redundancy elimination in
+compilers that use SSA form \cite{chow_new_1997} simplified the algorithms
+because no iterative data flow analysis is needed any more.
 
@@
 future. In particular loop peeling works poorly in combination with trace
-trees~\cite{andreas_gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
+trees~\cite{gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
 The side exits attached guards that fail often

### 2012-08-15 talk/dls2012/paper.tex
A comment and a typo.

@@
 floating point operations, except Psyco which uses x87 FPU instructions.
+% Psyco does not use the x87 FPU: all floating-point arithmetic is done with
+% residual calls to C helpers.  These can probably be compiled with SSE2.
+% But compiling CPython (and maybe Psyco) for x87 or SSE2 has probably
+% no measurable effect.
 We also run PyPy with loop peeling optimization and without (but otherwise
@@
 dhamdhere_practical_1991} this approach was followed by the work of Knoop
-et.al.~\cite{knoop_lazy_1992} who cleany separated the problem into a backward
+et.al.~\cite{knoop_lazy_1992} who cleanly separated the problem into a backward
 and forward data flow analysis. Implementing partial redundancy elimination in

### 2012-08-15 talk/iwtc11/benchmarks/benchmark.sh

@@ -1,0 +1,85 @@
+#!/bin/bash
+
+echo
+echo $*
+if [ "$1" == "gcc" ]; then
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_double.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_long.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_fix16.c
+    #./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1
+    #./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 100
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 100
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1000
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1000
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/conv3x3.cc 1000000 3
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/conv3x3.cc 1000 1000
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/dilate3x3.cc 1000 1000
+    ./runner.py -n 5 -c "$* -lstdc++" image/sobel.cc 1000 1000
+    ./runner.py -n 5 -c "$*" scimark/run_SOR.c 100 32768
+    ./runner.py -n 5 -c "$*" scimark/run_SOR.c 1000 256
+    ./runner.py -n 5 -c "$*" scimark/run_SparseMatMult.c 1000 5000 262144
+    ./runner.py -n 5 -c "$*" scimark/run_SparseMatMult.c 100000 1000000 1024
+    ./runner.py -n 5 -c "$*" scimark/run_MonteCarlo.c 268435456
+    ./runner.py -n 5 -c "$*" scimark/run_LU.c 100 4096
+    ./runner.py -n 5 -c "$*" scimark/run_LU.c 1000 2
+    rm a.out
+elif [[ "$1" == luajit* ]]; then
+    $* runner.lua sqrt int
+    $* runner.lua sqrt float
+    $* runner.lua sqrt Fix16
+    $* runner.lua convolution conv3 100
+    $* runner.lua convolution conv5 100
+    $* runner.lua convolution conv3 1000
+    $* runner.lua convolution conv5 1000
+    $* runner.lua convolution conv3x3 1000000 3
[…]

### 2012-08-15 talk/vmil2012/paper.tex
Typos, small fixes, rephrasing

--- 2012-08-14 talk/vmil2012/paper.tex
@@
 inserting operations called guards at points of possible divergence. These
-operations occur frequently generated traces and therefore it is important to
+operations occur frequently in generated traces and therefore it is important to
 design and implement them carefully to find the right trade-off between
@@
 
-Our aim is to help understand the design constraints when
+Our aim is to help understand the constraints when
 implementing guards. Guards have a runtime cost, they take time to execute. On
@@
 
-There are several aspects to consider
-in the design and optimization of guards, the first aspect is that due to the
+There are several aspects to be taken into account w.r.t. guards, the first aspect is that due to the
 large number of guards the memory overhead related to storing the information
@@
 what make the design and optimization of guards an important and non-trivial
-aspect of the low-level design of a tracing just-in-time compiler.
+aspect of the construction of a tracing just-in-time compiler.
 
@@
 
-The set of central concepts upon which this work is based is described in
+The set of central concepts upon which this work is based are described in
 Section~\ref{sec:Background}, such as the PyPy project, the RPython language
@@
 Data} we proceed to describe for RPython's tracing JIT the details of guards in
-the frontend related to recording and storing the
+the frontend. In this context the frontend is concerned with recording and storing the
 information required to rebuild the interpreter state in case of a guard
@@
 creating a Python interpreter written in a high level language, allowing easy
-language experimentation and extension. PyPy is now a fully compatible
+language experimentation and extension.\footnote{\url{http://pypy.org}} PyPy is now a fully compatible
 alternative implementation of the Python language, that is on average about 5 times faster than the reference implementation. The
-implementation takes advantage of the language features provided by RPython
+implementation takes advantage of RPython's language features
 such as the provided tracing just-in-time compiler described below.
 
-RPython, the language and the toolset originally developed to implement the
+RPython, the language and the toolset originally created to implement the
 Python interpreter have developed into a general environment for experimenting
-and developing fast and maintainable dynamic language implementations. There
-are, besides the Python interpreter, implementations of Prolog, Javascript, R,
-Smalltalk among other that are written in RPython at different levels of
-completeness.
+and developing fast and maintainable dynamic language implementations. Besides
+the Python interpreter there are several language implementation at different
+levels of completeness, e.g. for Smalltalk, Prolog, JavaScript and R.
 

### 2012-08-15 talk/dls2012/paper.tex
tune formulations

@@
 type checking, boxed value unwrapping and virtual method lookups.
-In this paper we explain a scheme invented within the context of the LuaJIT project
+In this paper we explain a scheme pioneered within the context of the LuaJIT project
 for making simple optimizations loop-aware by
@@
 floating point operations, except Psyco which uses x87 FPU instructions.
-% Psyco does not use the x87 FPU: all floating-point arithmetic is done with
-% residual calls to C helpers.  These can probably be compiled with SSE2.
-% But compiling CPython (and maybe Psyco) for x87 or SSE2 has probably
-% no measurable effect.
 We also run PyPy with loop peeling optimization and without (but otherwise
@@
 Loop invariant code motion has been part of early compilers in the 1960s and
-1970s~\cite{allen_catalogue_1971}. The approach for achieving loop invariant
-code motion is typically to perform partial redundancy elimination. The
+1970s~\cite{allen_catalogue_1971}. A common approach for achieving loop invariant
+code motion is to perform partial redundancy elimination. The
 approach was first proposed by Morel and Renvoise~\cite{morel_global_1979}. It
-involves solving data flow problems usually involding bidirection data flow
+involves solving data flow problems of bidirectional data flow
 equations. After improvements~\cite{chow_portable_1984,

### 2012-08-15 talk/vmil2012/paper.tex
mention why we expect guards to fail often

--- 2012-08-14 talk/vmil2012/paper.tex
@@
 %___________________________________________________________________________
-\todo{mention somewhere that it is to be expected that most guards do not fail}
 \section{Introduction}
@@
 
-\subsection{Overhead of Guards}
+\subsection{Space Overhead of Guards}
 \label{sub:guard_overhead}
@@
     \label{fig:backend_data}
-\end{figure}
+\end{figure}e.
 
@@
 not have unnecessary overhead.
+
+This low guard failure rate is expected. Most guards do not come from actual
+control flow divergences in the user program, but from type checks needed for
+type specialization. Various prior work has
+shown~\cite{holkner_evaluating_2009, richards_analysis_2010, callau_how_2011}
+that most programs in dynamic languages only use a limited amount of runtime
+variability. Therefore many guards are needed for making the traces behave
+correctly in all cases but fail rarely.
 

### 2012-08-15 talk/vmil2012/paper.tex
some protected whitespace

@@
 case they are frequently executed (this process is described in more detail in
-Section \ref{sec:Resume Data}). During the recording phase guards are
+Section~\ref{sec:Resume Data}). During the recording phase guards are
 inserted into the recorded trace at all
@@
 failure. Once the frontend has traced and optimized a loop it invokes the
-backend to compile the operations to machine code, Section \ref{sec:Guards in
+backend to compile the operations to machine code, Section~\ref{sec:Guards in
 the Backend} describes the low-level aspects of how guards are implemented in
@@
 producing the value can often be merged, further reducing the overhead of the guard.
-Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
+Figure~\ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
 followed by a \texttt{guard\_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
@@
 Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
-Guard \#1 jumps to the trampoline, loads the backend map and
-then calls the bailout handler, whereas Guard \#2 has already been patched
+Guard~\#1 jumps to the trampoline, loads the backend map and
+then calls the bailout handler, whereas Guard~\#2 has already been patched
 and directly jumps to the corresponding bridge. The bridge also contains two

### 2012-08-15 talk/vmil2012/paper.tex
attempt at a different introduction

@@
 beginning}
-Tracing just-in-time (JIT) compilers record and compile linear control flow paths of operations executed by an interpreter
-inserting operations called guards at points of possible divergence.
-In this paper we describe and analyze how guards work, explaining what concepts
-are used in the intermediate and low-level representation of the JIT
-instructions and how these are implemented.
 
-Our aim is to help understand the design constraints when
-implementing guards. Guards have a runtime cost, they take time to execute. On
-the other hand, guards are possible deoptimization points, meaning the recorded
-and compiled path has to be left returning control to the interpreter. They need
-enough associated information to enable rebuilding the interpreter state.
-Based on the informal observation that guards are among the most common
-operations in the traces produced by RPython's tracing JIT, our
-goal is to present concrete numbers for the frequency and the overhead related
-to guards, explain how they are implemented in the different levels of this particular
-tracing JIT and clarify the rationale behind the design decisions based on the
-numbers provided here.
+Tracing just-in-time (JIT) compilers record and compile commonly executed
+linear control flow paths consisting of operations executed by an interpreter.
+At points of possible divergence from the traced path operations called guards
+are inserted. Furthermore, type guards are inserted to specialize the trace
+based on the types observed during tracing. In this paper we describe and
+analyze how guards work and explain the concepts used in the intermediate and
+low-level representation of the JIT instructions and how these are implemented.
+This is done in the context of the RPython language and the PyPy project, which
+provides a tracing JIT compiler geared at dynamic language optimization.
 
-The operations executed by an interpreter are recorded by the tracing JIT in
-case they are frequently executed (this process is described in more detail in
-Section~\ref{sec:Resume Data}). During the recording phase guards are
-inserted into the recorded trace at all
-points where the control flow could diverge. As can be seen in
+Our aim is to help understand the design constraints when implementing guards
+and to describe the concrete techniques used in the various layers of RPython's
+tracing JIT. All design decisions will be motivated by concrete numbers for the
+frequency and the overhead related to guards.
+
+It is important to handle guards well, because they are very common operations
+in the traces produced by tracing JITs. As can be seen in
 Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
-operations before and for about 15\% to 20\% of the operations after
-optimizing the traces generated for the different benchmarks used in this paper.
-This makes guards one of the most common types of operations. As this paper will show, many of these guards
-fail rarely or not all during execution.
+operations before and for about 15\% to 20\% of the operations after optimizing
+the traces generated for the different benchmarks used in this paper. An
+additional property is that guard failure rates are very uneven. The majority
+of guards never fail at all, whereas those that do usually fail extremely
+often.
 
-There are several aspects to consider
-in the design and optimization of guards, the first aspect is that due to the
-large number of guards the memory overhead related to storing the information
-needed for deoptimization should be kept low. A second aspect is that
-successfully checking guards, i.e. not leaving the compiled trace, – which is
-the common case – should be a cheap operation to execute favouring the on-trace
-execution speed in contrast to the deoptimization case where the state has to
-be rebuilt using the stored information. These constraints and trade-offs are
-what make the design and optimization of guards an important and non-trivial
-aspect of the low-level design of a tracing just-in-time compiler.
+Besides being common, guards have various costs attached to them.
+Guards have a runtime cost, they take time to execute. Therefore it is
+important to make the on-trace execution of guards as efficient as possible. On
+the other hand, guards are possible deoptimization points. The recorded and
+compiled path has to be left if a guard fails, returning control to the
+interpreter. Therefore guards need enough associated information to enable
+rebuilding the interpreter state. The memory overhead of this information
+should be kept low. These constraints and trade-offs are what make the design
+and optimization of guards an important and non-trivial aspect of the low-level
+design of a tracing just-in-time compiler.
 

### 2012-08-15 talk/vmil2012/paper.tex
More small improvements

--- 2012-08-15 talk/vmil2012/paper.tex
@@
 %___________________________________________________________________________
+\todo{mention somewhere that it is to be expected that most guards do not fail}
+\todo{better formatting for lstinline}
 \section{Introduction}
@@
 describe based on them the reasoning behind the implementation of guards in
-RPython's tracing just-in-time compiler. The contributions of this paper are:
+RPython's tracing just-in-time compiler. the contributions of this paper are:
 \begin{itemize}
-  \item An analysis and benchmark of guards in the context of RPython's tracing JIT,
+  \item An analysis and benchmark of guards in the context of RPython's tracing JIT.
   %An analysis of guards in the context of RPython's tracing JIT to
@@
   \item a description about how guards are implemented in the high\-
-  and low-level components of the JIT and a description of the rationale behind the design.
+  and low-level components of the JIT and describe the rationale behind the design.
 \end{itemize}
@@
 As in most compilers, tracing JITs use an intermediate representation to
-store the recorded operations, which is typically in SSA
+store the recorded operations, typically in SSA
 form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
@@
 stopped and the control is returned to the interpreter, after the interpreter's
-state has been restored. If a particular guard fails often a new trace is
-recorded starting from the guard. We will refer to this kind of trace as a
-\emph{bridge}. Once a bridge has been traced it is attached to the
+state has been restored. If a particular guard fails often a new trace
+starting from the guard is recorded. We will refer to this kind of trace as a
+\emph{bridge}. Once a bridge has been traced and compiled it is attached to the
 corresponding guard by patching the machine code. The next time the guard fails
@@
 To do this reconstruction it is necessary to take the values of the SSA
-variables of the trace and build interpreter stack frames.  Tracing
+variables in the trace to build interpreter stack frames.  Tracing
 aggressively inlines functions, therefore the reconstructed state of the
@@
 If a guard fails often enough, a trace is started from it
-forming a trace tree.
+to create a bridge, forming a trace tree.
 When that happens another use case of resume data
-is to construct the tracer state.
+is to reconstruct the tracer state.
 After the bridge has been recorded and compiled it is attached to the guard.
@@
 Guards are a very common operations in the traces.
-However, a large percentage of all operations
+However, as will be shown, a large percentage of all operations
 are optimized away before code generation.
@@
 do not contain values.
-Instead, every local variables contains the SSA variable of the trace
+Instead, every local variable contains the SSA variable of the trace
 where the value would later come from, or a constant.
@@
 After tracing has been finished the trace is optimized.
-During optimization a large percentage of operations can be removed.
+During optimization a large percentage of operations can be removed.\todo{add a reference to the figure showing the optimization rates?}
 In the process the resume data is transformed into its final, compressed form.
@@
 In particular guards can be removed by subexpression elimination.
-If the same guard is encountered a second time in the trace,
+If the same guard is encountered a second time in a trace,
 the second one can be removed.
@@
 
-Adding this additional information is done as follows:
+Storing this additional information is done as follows:
 So far, every variable in the symbolic frames
@@
 The same observation as about frames applies:
-Quite often a virtual object does not change from one guard to the next.
-Then the data structure is shared.
+Quite often a virtual object does not change from one guard to the next,
+allowing the data structure to be shared.
 
@@
 
-After optimization the resulting trace is handed over to the platform specific
+After the recorded trace has been optimized it is handed over to the platform specific
 backend to be compiled to machine code. The compilation phase consists of two
@@
 checked by the guard is not used anywhere else the guard and the operation
-producing the value can often be merged, further reducing the overhead of the guard.
-Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
-followed by a \texttt{guard\_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
+producing the value can merged, further reducing the overhead of the guard.
+Figure \ref{fig:trace-compiled} shows how the \lstinline{int_eq} operation
+followed by a \lstinline{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
 pseudo-assembler if the operation and the guard are compiled separated or if
@@
 structure called \emph{backend map} is created. This data structure encodes the
-mapping from the IR-variables needed by the guard to rebuild the state to the
+mapping from IR-variables needed by the guard to rebuild the state to the
 low-level locations (registers and stack) where the corresponding values will
@@
 This data
-structure stores the values in a succinct manner using an encoding that uses
+structure stores the values in a succinct manner using an encoding that requires
 8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
@@
 as \emph{compensation code},
-that is used to leave the compiled trace in case of a guard failure.
+that is used to leave the compiled trace.
 
 Using the encoded location information the bailout handler reads from the
-saved execution state the values that the IR-variables had  at the time of the
+stored execution state the values that the IR-variables had at the time of the
 guard failure and stores them in a location that can be read by the frontend.
-After saving the information the control is passed to the frontend signaling
-which guard failed so the frontend can read the information passed and restore
+After saving the information the control is returned to the frontend signaling
+which guard failed so the frontend can read the stored information and rebuild
 the state corresponding to the point in the program.
 
-As in previous sections the underlying idea for the design of guards is to have
-a fast on-trace profile and a potentially slow one in the bailout case where
-the execution has to return to the interpreter due to a guard failure. At the same
+As in previous sections the underlying idea for the low-level design of guards is to have
+a fast on-trace profile and a potentially slow one in case
+the execution has to return to the interpreter. At the same
 time the data stored in the backend, required to rebuild the state, should be as
@@
 bindings) that corresponds to the state during the compilation of the original
-guard. To restore the state needed to compile the bridge we use the encoded
-representation created for the guard to rebuild the bindings from IR-variables
-to stack locations and registers used in the register allocator.  With this
+guard. To restore the state needed to compile the bridge we use the backend map
+created for the guard to rebuild the bindings from IR-variables
+to stack locations and registers.  With this
 reconstruction all bindings are restored to the state as they were in the
@@
 \texttt{ff7b35837d0f}.\footnote{\url{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}}
-The benchmarks were run on a version of PyPy based on the
-revision~\texttt{0b77afaafdd0} and patched to collect additional data about the
+The benchmarks were run on a version of PyPy based on
+revision~\texttt{0b77afaafdd0} and patched to collect additional data about
 guards in the machine code
@@
 
-All measurements presented in this section do not take garbage collection of machine code into account. Pieces
+All measurements presented in this section do not take garbage collection of resume data and machine code into account. Pieces
 of machine code can be globally invalidated or just become cold again. In both
@@
 
-\subsection{Space Overhead of Guards}
+\subsection{Overhead of Guards}
 \label{sub:guard_overhead}
@@
     \label{fig:backend_data}
-\end{figure}e.
+\end{figure}
 
@@
 not have unnecessary overhead.
-
-This low guard failure rate is expected. Most guards do not come from actual
-control flow divergences in the user program, but from type checks needed for
-type specialization. Various prior work has
-shown~\cite{holkner_evaluating_2009, richards_analysis_2010, callau_how_2011}
-that most programs in dynamic languages only use a limited amount of runtime
-variability. Therefore many guards are needed for making the traces behave
-correctly in all cases but fail rarely.
 

### 2012-08-15 talk/vmil2012/paper.tex
cite some of our papers and don't mention .net/java too promoninently as targets

--- 2012-08-15 talk/vmil2012/paper.tex
@@
 language experimentation and extension. PyPy is now a fully compatible
-alternative implementation of the Python language, that is on average about 5 times faster than the reference implementation. The
-implementation takes advantage of the language features provided by RPython
+alternative interpreter for the Python language.
+Using RPython's tracing JIT compiler it is on average about 5 times faster than
+CPython, the reference implementation.
+PyPy is an interpreter written in RPython and takes advantage of the language
+features provided by RPython
 such as the provided tracing just-in-time compiler described below.
@@
 and developing fast and maintainable dynamic language implementations. There
-are, besides the Python interpreter, implementations of Prolog, Javascript, R,
-Smalltalk among other that are written in RPython at different levels of
-completeness.
+are, besides the Python interpreter, experimental implementations of
+Prolog~\cite{bolz_towards_2010}, Javascript, R,
+Smalltalk~\cite{bolz_towards_2010} among other that are written in RPython at
+different levels of completeness.
 
-RPython is constructed from two components:
+RPython can mean one of two things:
 \begin{itemize}
@@
 The RPython language
-is a statically typed object oriented high level language. The language provides
+is a statically typed object-oriented high level language. The language provides
 several features such as automatic memory management
@@
 implementing.  The second RPython component, the translation toolchain, is used
-to transform the program to a low level representations suited to be compiled
-and run on one of the different supported target platforms/architectures such
-as C, .NET and Java. During the transformation process
+to transform the interpreter into a C program.\footnote{
+    RPython can also be used to translate programs to CLR and Java
+    bytecode~\cite{ancona_rpython:_2007}, but this feature is somewhat
+    experimental.
+}
+During the transformation process
 different low level aspects suited for the target environment are automatically
-added to the program such as (if needed) a garbage collector
-and based on hints provided by the author a just-in-time compiler.
+added to the program such as a garbage collector and a tracing JIT compiler.
+The process of inserting a tracing JIT is not fully automatic but is guided by
+hints from the interpreter author.
 

### 2012-08-15 talk/vmil2012/paper.tex
this is actually done
more project-internal citations

@@
 
-\todo{the introduction needs some work}
-\cfbolz{the first two two paragraphs talk about deoptimization, then it
-switches to guards. I would say we should only talk about guards in the
-beginning}
-
 Tracing just-in-time (JIT) compilers record and compile commonly executed
@@
 
-The RPython language and the PyPy project were started in 2002 with the goal of
+The RPython language and the PyPy project~\cite{rigo_pypys_2006} were started
+in 2002 with the goal of
 creating a Python interpreter written in a high level language, allowing easy
@@
 RPython provides a tracing JIT that can be reused for a number of language
-implementations. This is possible, because it traces the execution of the
+implementations~\cite{bolz_tracing_2009}. This is possible, because it traces
+the execution of the
 language interpreter instead of tracing the user program directly. This
@@
 that RPython's tracing JIT is a meta-tracing JIT can be ignored.
+The only point of interaction is that some of the guards that are inserted into
+the trace stem from an annotation provided by the interpreter
+author~\cite{bolz_runtime_2011}.
 

### 2012-08-15 talk/vmil2012/paper.tex
more tweaking

--- 2012-08-15 talk/vmil2012/paper.tex
@@
 figures show the total amount of operations that are evaluated by the JIT and
-the total amount of code and data that is generated from the optimized traces.
+the total amount of code and resume data that is generated.
 
@@
 operations are guards. The number of operations was counted on the unoptimized
-and optimized traces. The Figure shows that the overall optimization rate for
+and optimized traces. The Figure also shows the overall optimization rate for
 operations, which is between 69.4\% and 83.89\%, of the traced operations and the
 optimization rate of guards, which is between 65.8\% and 86.2\% of the
-operations, are very similar. This indicates that the optimizer can remove
+operations. This indicates that the optimizer can remove
 most of the guards, but after the optimization pass these still account for
@@
 From Figure~\ref{fig:failing_guards} we can see that only a very small amount
-of all the guards in the optimized traces ever fail. This amount varies between
+of all the guards in the compiled traces ever fail. This amount varies between
 2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
@@
 SPUR~\cite{bebenita_spur:_2010} is a tracing JIT compiler
-for a C\# virtual machine.
+for a CIL virtual machine.
 It handles guards by always generating code for every one of them
@@
 from the original program and for guards that are likely to fail. As an outlook
-Pall mentions the plans to switch to compressed snapshots to further reduce
+Pall mentions plans to switch to compressed snapshots to further reduce
 redundancy. The approach of not creating snapshots at all for every guard is
@@
 Linking side exits to pieces of later compiled machine code was described first
-in the context of Dynamo~\cite{Bala:2000wv} under the name of Fragment Linking.
-Once a new hot trace is emitted into the fragment cache it is linked to side
+in the context of Dynamo~\cite{Bala:2000wv} under the name of fragment linking.
+Once a new hot trace is emitted into the fragment cache it is linked to the side
 exit that led to the compilation of the fragment. Fragment Linking avoids the
@@
 
-Gal et. al~\cite{Gal:2006} describe how in the HotpathVM they experimented
+Gal et. al~\cite{Gal:2006} describe how in the HotpathVM, a JIT for a Java VM, they experimented
 with having one generic compensation code block, like the RPython JIT, that
@@
 that they do not fail often, we described how they have been implemented in the
-high and low level components of RPython's tracing JIT compiler.
+high- and low-level components of RPython's tracing JIT compiler.
 
 Additionally we have presented experimental data collected using the standard PyPy
-benchmark set to evaluate previous observations and assumptions. Our
+benchmark set to evaluate previous observations and assumptions about guards. Our
 experiments confirmed that guards are a very common
@@
 showed that the compression techniques used in PyPy effectively reduce the
-overhead of guards, but it still produces a significant overhead. The results
+overhead of guards, but they still produce a significant overhead. The results
 also showed that guard failure is a local event: there are few
@@
 \section*{Acknowledgements}
-We would like to thank David Edelsohn and Stephan Zalewski for their helpful
+We would like to thank David Edelsohn, Samuele Pedroni and Stephan Zalewski for their helpful
 feedback and valuable comments while writing this paper.

### 2012-08-15 talk/dls2012/paper.tex
describe the scimark benchmakrs

--- 2012-08-15 talk/dls2012/paper.tex
@@
 external calls. Larger loops involving many operations on complex objects
-typically benefit less from it. Loop peeling never makes runtime performance worse, in
-the worst case the peeled loop is exactly the same as the preamble. Therefore we
+typically benefit less from it. Loop peeling never makes the generated code worse, in
+the worst case the peeled loop is exactly the same as the preamble. 
+Therefore we
 chose to present benchmarks of small numeric kernels where loop peeling can show
@@
 version 2.7 compatible interpreter. A set of numerical
-calculations were implemented in both Python and in C and their
+calculations were implemented in both Python, C and Lua and their
 runtimes are compared in Figure~\ref{fig:benchmarks}. The benchmarks are
 \begin{itemize}
-\item {\bf sqrt}: approximates the square root of $y$. The approximation is 
+\item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 
 initiated to $x_0=y/2$ and the benchmark consists of a single loop updating this
@@
 There are three different versions of this benchmark where $x_i$
-  is represented with different type of objects: int's, float's and
+  is represented with different type of objects, $T$,: int's, float's and
   Fix16's. The latter, Fix16, is a custom class that implements
@@
   there are three different implementations.
-\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$. A single loop
+\item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
 is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_n\right)$ from a vector
@@
 with $n=10^5$ and $n=10^6$.
-\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
+\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
 ${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
 hardcoded into the implementation making the benchmark consist of a single loop too.
-\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
+\item {\bf conv3x3}$\left(n\right)$: two-dimensional convolution with kernel of fixed
   size $3 \times 3$ using a custom class to represent two-dimensional
@@
 The memory for storing the matrices are again allocated outside the benchmark and $n=1000$ was used.
-\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
+\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed
   size $3 \times 3$. This is similar to convolution but instead of
@@
   of the optimizations.
-\item {\bf sobel}: a low-level video processing algorithm used to
+\item {\bf sobel}$\left(n\right)$: a low-level video processing algorithm used to
   locate edges in an image. It calculates the gradient magnitude
@@
 a straightforward implementation providing 2 dimensional
-indexing with out of bounds checks. For the C implementations it is
+indexing with out of bounds checks and 
+data stored in row-major order.
+For the C implementations it is
 implemented as a C++ class. The other benchmarks are implemented in
@@
 point numbers, both in the Python and the C code.
+
+In addition we also ported the 
+SciMark\footnote{\texttt{http://math.nist.gov/scimark2/}} benchmakts to python, and compared 
+their runtimes with the already existing Lua and C implementations. 
+This port was performed after the release of the pypy used to run the benchmarks which means that 
+these benchmarks have not influenced the pypy implementation.
+SciMark consists of
+
+\begin{itemize}
+\item {\bf SOR}$\left(n, c\right)$: Jacobi successive over-relaxation on a $n\times n$ grid repreated $c$ times.
+The same custom two-dimensional array class as described above is used to represent
+the gird.
+\item {\bf SparseMatMult}$\left(n, z, c\right)$: Matrix multiplication between a $n\times n$ sparse matrix,
+stored in compressed-row format, and a full storage vector, stored in a normal array. The matrix has $z$ non-zero elements and the calculation is repeated $c$ times.
+\item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
+\item {\bf LU}$\left(n, c\right)$: Computes the LU factorization of a $n \times n$ matrix.
+The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable.
+Instead a list of arrays is used to represent the matrix.
+The calculation is repeated $c$ times.
+\item {\bf FFT}$\left(n, c\right)$: Fast Fourier Transform of a vector with $n$ elements, represented as an array, repeated $c$ times.
+\end{itemize}
 

### 2012-08-15 talk/dls2012/paper.tex
cleanup
correct some details

@@
 hardcoded into the implementation making the benchmark consist of a single loop too.
-\item {\bf conv3x3}$\left(n\right)$: two-dimensional convolution with kernel of fixed
+\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with kernel of fixed
   size $3 \times 3$ using a custom class to represent two-dimensional
   arrays. It is implemented as two nested loops that iterates over the elements of the 
-$n\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
+$m\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
 ${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
@@
 \end{equation}
-for $1 \leq i \leq n$ and $1 \leq j \leq n$.
-The memory for storing the matrices are again allocated outside the benchmark and $n=1000$ was used.
+for $1 \leq i \leq m$ and $1 \leq j \leq n$.
+The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
+as well as $(n,m)=(1000000,3)$ was used.
 \item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed
@@
 \item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
-\item {\bf LU}$\left(n, c\right)$: Computes the LU factorization of a $n \times n$ matrix.
+\item {\bf LU}$\left(n, c\right)$: LU factorization of an $n \times n$ matrix.
 The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable.

### 2012-08-16 talk/vmil2012/paper.tex
More fixes

@@
 goal is to present concrete numbers for the frequency and the overhead related
-to guards, explain how they are implemented in the different levels of this particular
-tracing JIT and clarify the rationale behind the design decisions based on the
+to guards, to explain how they are implemented in the different levels of this particular
+tracing JIT and to clarify the rationale behind the design decisions based on the
 numbers provided here.
@@
 
-There are several aspects to be taken into account w.r.t. guards, the first aspect is that due to the
+There are several aspects to be taken into account w.r.t. guards, the first of them is that due to the
 large number of guards the memory overhead related to storing the information
@@
 \begin{itemize}
-  \item An analysis and benchmark of guards in the context of RPython's tracing JIT.
+  \item an analysis and benchmark of guards in the context of RPython's tracing JIT,
   %An analysis of guards in the context of RPython's tracing JIT to
@@
   \item a description about how guards are implemented in the high\-
-  and low-level components of the JIT and describe the rationale behind the design.
+  and low-level components of the JIT and describe the rationale behind the design
 \end{itemize}
@@
 After tracing has been finished the trace is optimized.
-During optimization a large percentage of operations can be removed.\todo{add a reference to the figure showing the optimization rates?}
+During optimization a large percentage of operations can be removed. \todo{add a reference to the figure showing the optimization rates?}
 In the process the resume data is transformed into its final, compressed form.
@@
 
-The possible source of information are:
+The possible sources of information are:
 
@@
 operations are guards. The number of operations was counted on the unoptimized
-and optimized traces. The Figure also shows the overall optimization rate for
+and optimized traces. The figure also shows the overall optimization rate for
 operations, which is between 69.4\% and 83.89\%, of the traced operations and the
@@
 Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
-relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.
-\footnote{
+relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.\footnote{
     The threshold used is 200 failures. This rather high threshold was picked experimentally to give
@@
 Once a new hot trace is emitted into the fragment cache it is linked to the side
-exit that led to the compilation of the fragment. Fragment Linking avoids the
+exit that led to the compilation of the fragment. Fragment linking avoids the
 performance penalty involved in leaving the compiled code. Fragment linking

### 2012-08-16 talk/dls2012/paper.tex
fix size of beramono font
link to paper repository

--- 2012-08-15 talk/dls2012/paper.tex
@@
 \usepackage{listings}
-\usepackage{beramono}
+\usepackage[scaled=0.81]{beramono}
 
@@
 \revc{
-Providing source code for the benchmarks measured are needed for others to
-reproduce and build on your results.  I believe this should be the minimum
-standard for publishing measurements such as these.
-}
-\cfbolz{
-let's link to the bitbucket source code view. how about we move the benchmarks
-to the dls directory as well? or their own repository, we've been using them as
-demos
-}
-
-\revc{
 I would have liked to have benchmark results for some larger applications.
@@
 calculations were implemented in both Python and in C and their
-runtimes are compared in Figure~\ref{fig:benchmarks}. The benchmarks are
+runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
+    The benchmarks and the scripts to run them can be found in the repository for this paper:
+    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
+}
+
+The benchmarks are
 \begin{itemize}

### 2012-08-16 talk/dls2012/paper.tex

@@
 version 2.7 compatible interpreter. A set of numerical
-calculations were implemented in both Python and in C and their
+calculations were implemented in both Python, C and Lua and their
 runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
@@
 \begin{itemize}
-\item {\bf sqrt}: approximates the square root of $y$. The approximation is 
+\item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 
 initiated to $x_0=y/2$ and the benchmark consists of a single loop updating this
@@
 with $n=10^5$ and $n=10^6$.
-\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
+\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
 ${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
 hardcoded into the implementation making the benchmark consist of a single loop too.
-\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
+\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with kernel of fixed
   size $3 \times 3$ using a custom class to represent two-dimensional
   arrays. It is implemented as two nested loops that iterates over the elements of the 
-$n\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
+$m\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
 ${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
@@
 \end{equation}
-for $1 \leq i \leq n$ and $1 \leq j \leq n$.
-The memory for storing the matrices are again allocated outside the benchmark and $n=1000$ was used.
-\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
+for $1 \leq i \leq m$ and $1 \leq j \leq n$.
+The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
+as well as $(n,m)=(1000000,3)$ was used.
+\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed
   size $3 \times 3$. This is similar to convolution but instead of
@@
 point numbers, both in the Python and the C code.
+
+In addition we also ported the 
+SciMark\footnote{\texttt{http://math.nist.gov/scimark2/}} benchmakts to python, and compared 
+their runtimes with the already existing Lua and C implementations. 
+This port was performed after the release of the pypy used to run the benchmarks which means that 
+these benchmarks have not influenced the pypy implementation.
+SciMark consists of
+
+\begin{itemize}
+\item {\bf SOR}$\left(n, c\right)$: Jacobi successive over-relaxation on a $n\times n$ grid repreated $c$ times.
+The same custom two-dimensional array class as described above is used to represent
+the gird.
+\item {\bf SparseMatMult}$\left(n, z, c\right)$: Matrix multiplication between a $n\times n$ sparse matrix,
+stored in compressed-row format, and a full storage vector, stored in a normal array. The matrix has $z$ non-zero elements and the calculation is repeated $c$ times.
+\item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
+\item {\bf LU}$\left(n, c\right)$: LU factorization of an $n \times n$ matrix.
+The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable.
+Instead a list of arrays is used to represent the matrix.
+The calculation is repeated $c$ times.
+\item {\bf FFT}$\left(n, c\right)$: Fast Fourier Transform of a vector with $n$ elements, represented as an array, repeated $c$ times.
+\end{itemize}
 

### 2012-08-16 talk/vmil2012/paper.tex
(cfbolz, bivab) lstinline settings

@@
 
+\lstdefinelanguage{none}{
+      keywords={},
+}
 \lstset{
   basicstyle=\ttfamily\footnotesize,
-  language=Python,
+  language=none,
   keywordstyle=\bfseries,
@@
 
+\let\oldlstinline=\lstinline
+\renewcommand\lstinline[1]{\oldlstinline[basicstyle=\ttfamily]{#1}}
+
 \definecolor{gray}{rgb}{0.5,0.5,0.5}
@@
 %___________________________________________________________________________
-\todo{mention somewhere that it is to be expected that most guards do not fail}
-\todo{better formatting for lstinline}
 \section{Introduction}
 
-\todo{the introduction needs some work}
-\cfbolz{the first two two paragraphs talk about deoptimization, then it
-switches to guards. I would say we should only talk about guards in the
-beginning}
-Tracing just-in-time (JIT) compilers record and compile linear control flow paths of operations executed by an interpreter
-inserting operations called guards at points of possible divergence.
-In this paper we describe and analyze how guards work, explaining what concepts
-are used in the intermediate and low-level representation of the JIT
-instructions and how these are implemented.
+Tracing just-in-time (JIT) compilers record and compile commonly executed
+linear control flow paths consisting of operations executed by an interpreter.
+At points of possible divergence from the traced path operations called guards
+are inserted. Furthermore, type guards are inserted to specialize the trace
+based on the types observed during tracing. In this paper we describe and
+analyze how guards work and explain the concepts used in the intermediate and
+low-level representation of the JIT instructions and how these are implemented.
+This is done in the context of the RPython language and the PyPy project, which
+provides a tracing JIT compiler geared at dynamic language optimization.
 
-Our aim is to help understand the constraints when
-implementing guards. Guards have a runtime cost, they take time to execute. On
-the other hand, guards are possible deoptimization points, meaning the recorded
-and compiled path has to be left returning control to the interpreter. They need
-enough associated information to enable rebuilding the interpreter state.
-Based on the informal observation that guards are among the most common
-operations in the traces produced by RPython's tracing JIT, our
-goal is to present concrete numbers for the frequency and the overhead related
-to guards, to explain how they are implemented in the different levels of this particular
-tracing JIT and to clarify the rationale behind the design decisions based on the
-numbers provided here.
+Our aim is to help understand the constraints when implementing guards
+and to describe the concrete techniques used in the various layers of RPython's
+tracing JIT. All design decisions will be motivated by concrete numbers for the
+frequency and the overhead related to guards.
 
-The operations executed by an interpreter are recorded by the tracing JIT in
-case they are frequently executed (this process is described in more detail in
-Section \ref{sec:Resume Data}). During the recording phase guards are
-inserted into the recorded trace at all
-points where the control flow could diverge. As can be seen in
+It is important to handle guards well, because they are very common operations
+in the traces produced by tracing JITs. As can be seen in
 Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
-operations before and for about 15\% to 20\% of the operations after
-optimizing the traces generated for the different benchmarks used in this paper.
-This makes guards one of the most common types of operations. As this paper will show, many of these guards
-fail rarely or not all during execution.
+operations before and for about 15\% to 20\% of the operations after optimizing
+the traces generated for the different benchmarks used in this paper. An
+additional property is that guard failure rates are very uneven. The majority
+of guards never fail at all, whereas those that do usually fail extremely
+often.
 
-There are several aspects to be taken into account w.r.t. guards, the first of them is that due to the
-large number of guards the memory overhead related to storing the information
-needed for deoptimization should be kept low. A second aspect is that
-successfully checking guards, i.e. not leaving the compiled trace, – which is
-the common case – should be a cheap operation to execute favouring the on-trace
-execution speed in contrast to the deoptimization case where the state has to
-be rebuilt using the stored information. These constraints and trade-offs are
-what make the design and optimization of guards an important and non-trivial
-aspect of the construction of a tracing just-in-time compiler.
+Besides being common, guards have various costs attached to them.
+Guards have a runtime cost, they take time to execute. Therefore it is
+important to make the on-trace execution of guards as efficient as possible. On
+the other hand, guards are possible deoptimization points. The recorded and
+compiled path has to be left if a guard fails, returning control to the
+interpreter. Therefore guards need enough associated information to enable
+rebuilding the interpreter state. The memory overhead of this information
+should be kept low. These constraints and trade-offs are what make the design
+and optimization of guards an important and non-trivial aspect of the construction
+of a tracing just-in-time compiler.
 
@@
 
-The RPython language and the PyPy project were started in 2002 with the goal of
+The RPython language and the PyPy project~\cite{rigo_pypys_2006} were started
+in 2002 with the goal of
 creating a Python interpreter written in a high level language, allowing easy
 language experimentation and extension.\footnote{\url{http://pypy.org}} PyPy is now a fully compatible
-alternative implementation of the Python language, that is on average about 5 times faster than the reference implementation. The
-implementation takes advantage of RPython's language features
+alternative interpreter for the Python language.
+Using RPython's tracing JIT compiler it is on average about 5 times faster than
+CPython, the reference implementation.
+PyPy is an interpreter written in RPython and takes advantage of the language
+features provided by RPython
 such as the provided tracing just-in-time compiler described below.
@@
 and developing fast and maintainable dynamic language implementations. Besides
-the Python interpreter there are several language implementation at different
-levels of completeness, e.g. for Smalltalk, Prolog, JavaScript and R.
+the Python interpreter there are several experimental language implementation at different
+levels of completeness, e.g. for Prolog~\cite{bolz_towards_2010}, Smalltalk~\cite{bolz_towards_2010}, JavaScript and R.
 
-RPython is constructed from two components:
+different levels of completeness.
+
+RPython can mean one of two things:
 \begin{itemize}
@@
 producing the value can merged, further reducing the overhead of the guard.
-Figure \ref{fig:trace-compiled} shows how the \lstinline{int_eq} operation
+Figure~\ref{fig:trace-compiled} shows how the \lstinline{int_eq} operation
 followed by a \lstinline{guard_false} from the trace in Figure~\ref{fig:trace-log} are compiled to

### 2012-08-16 talk/vmil2012/paper.tex
Add PyPy and RPython to acknowledgements
tweaks to the introduction

@@
 
-Besides being common, guards have various costs attached to them.
+Besides being common, guards have various costs associated to them.
 Guards have a runtime cost, they take time to execute. Therefore it is
@@
 rebuilding the interpreter state. The memory overhead of this information
-should be kept low. These constraints and trade-offs are what make the design
+should be kept low. These constraints and trade-offs are what makes the design
 and optimization of guards an important and non-trivial aspect of the construction
@@
 %stored at the different levels for the guards
-In this paper we want to substantiate the aforementioned observations and
-describe based on them the reasoning behind the implementation of guards in
+In this paper we want to substantiate the aforementioned observations about guards and
+describe based on them the reasoning behind their implementation in
 RPython's tracing just-in-time compiler. the contributions of this paper are:
@@
 feedback and valuable comments while writing this paper.
+We thank the PyPy and RPython community for their continuous support and work:
+Armin Rigo, Antonio Cuni, Maciej Fijałkowski, Samuele Pedroni, and countless
+others. Any remaining errors are our own.
 

### 2012-08-16 talk/vmil2012/paper.tex
typo
calculate the percentage of guards responsible for 50% of the guard failures for the different benchmarks

@@
 It is important to handle guards well, because they are very common operations
-in the traces produced by tracing JITs. As can be seen in
-Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
+in the traces produced by tracing JITs. As we will see later (Figure~\ref{fig:benchmarks})
+guards account for about 14\% to 22\% of the
 operations before and for about 15\% to 20\% of the operations after optimizing
@@
 
-Besides being common, guards have various costs associated to them.
+Besides being common, guards have various costs associated with them.
 Guards have a runtime cost, they take time to execute. Therefore it is
@@
 \end{itemize}
-
-\begin{figure}
-    \include{figures/guard_table}
-    \caption{Percentage of guards before and after optimization for different benchmarks}
-    \label{fig:guard_percent}
-\end{figure}
 
@@
     \include{figures/failing_guards_table}
-    \caption{Failing guards relative to the total number of guards}
+    \caption{Failing guards, guards with more than 200 failures and guards responsible for 50\% of the failures relative to the total number of guards}
     \label{fig:failing_guards}
@@
 fail often enough that a bridge is compiled. Also, of all failing guards a few fail extremely often
-and most fail rarely. The results emphasize that as most of the guards never
+and most fail rarely. Reinforcing this notion the figure shows that, depending on the
+benchmark, between 0.008\% and 0.225\% of the guards are responsible for 50\%
+of the total guards failures.
+These results emphasize that as most of the guards never
 fail it is important to make sure that the successful execution of a guard does

### 2012-08-16 talk/vmil2012/tool/build_tables.py

@@ -26,22 +26,40 @@ def build_failing_guards_table(files, texfile, template):
 
     table = []
     head = ['Benchmark',
-            'Failing guards',
-            'Over %d failures' % BRIDGE_THRESHOLD]
+            'Failing',
+            '> %d failures' % BRIDGE_THRESHOLD,
+            '50\% of failures']
 
     for bench, info in failures.iteritems():
-        total = failures[bench]['nguards']
+        total = info['nguards']
         total_failures = len(info['results'])
         bridges = len([k for k,v in info['results'].iteritems() \
                                             if v > BRIDGE_THRESHOLD])
         res = [bench.replace('_', '\\_'),
                 "%.1f\\%%" % (100 * total_failures/total),
                 "%.1f\\%%" % (100 * bridges/total),
+                "%.3f\\%%"  % (100 * we_are_99_percent(info)),
         ]
         table.append(res)
     output = render_table(template, head, sorted(table))
     write_table(output, texfile)
 
+def we_are_50_percent(info):
+    total_guards = info['nguards']
+    failure_counts = info['results'].values()
+    print failure_counts
+    failure_counts.sort()
+    print failure_counts
+    failure_counts.reverse()
+    print failure_counts
+
+    total_failures = sum(failure_counts)
+    current_sum = 0
+    for i, f in enumerate(failure_counts):
+        current_sum += f
+        if current_sum > total_failures * 0.50:
+            return (i + 1)/total_guards
+    return -1
 
 def build_resume_data_table(csvfiles, texfile, template):
     assert len(csvfiles) == 1

### 2012-08-16 talk/vmil2012/paper.tex
kill todo
fix labes in diagram
referenc figure

@@
 After tracing has been finished the trace is optimized.
-During optimization a large percentage of operations can be removed. \todo{add a reference to the figure showing the optimization rates?}
+During optimization a large percentage of operations can be removed (Figure~\ref{fig:benchmarks}).
 In the process the resume data is transformed into its final, compressed form.
@@
 
-\todo{fix labels in diagram}
-
 % section Resume Data (end)
@@
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{figures/resume_data.pdf}
+\includegraphics[width=0.4\textwidth]{figures/resume_data.pdf}
 \caption{The resume data for Figure~\ref{fig:trace-log}}
@@
 \label{sec:evaluation}
-\todo{improve the table formatting}
 
@@
 
-Figure~\ref{fig:benchmarks} extends Figure~\ref{fig:guard_percent} and summarizes the total number of operations that were
+Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
 recorded during tracing for each of the benchmarks and what percentage of these

### 2012-08-16 talk/vmil2012/paper.tex
remove orphaned reference to figure

@@
 After tracing has been finished the trace is optimized.
-During optimization a large percentage of operations can be removed (Figure~\ref{fig:benchmarks}).
+During optimization a large percentage of operations can be removed. \todo{add a reference to the figure showing the optimization rates?}
 In the process the resume data is transformed into its final, compressed form.

### 2012-08-16 talk/vmil2012/paper.tex
remove list of todos
(cfbolz, bivab) reformat title
results so far, missing psyco

--- 2012-08-16 talk/vmil2012/paper.tex
@@
 
-\lstdefinelanguage{none}{
-      keywords={},
-}
 \lstset{
   basicstyle=\ttfamily\footnotesize,
-  language=none,
+  language=Python,
   keywordstyle=\bfseries,
@@
 
-\let\oldlstinline=\lstinline
-\renewcommand\lstinline[1]{\oldlstinline[basicstyle=\ttfamily]{#1}}
-
 \definecolor{gray}{rgb}{0.5,0.5,0.5}
@@
 %___________________________________________________________________________
+\todo{better formatting for lstinline}
 \section{Introduction}
@@
 It is important to handle guards well, because they are very common operations
-in the traces produced by tracing JITs. As we will see later (Figure~\ref{fig:benchmarks})
-guards account for about 14\% to 22\% of the
+in the traces produced by tracing JITs. As can be seen in
+Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
 operations before and for about 15\% to 20\% of the operations after optimizing
@@
 
-Besides being common, guards have various costs associated with them.
+Besides being common, guards have various costs attached to them.
 Guards have a runtime cost, they take time to execute. Therefore it is
@@
 rebuilding the interpreter state. The memory overhead of this information
-should be kept low. These constraints and trade-offs are what makes the design
+should be kept low. These constraints and trade-offs are what make the design
 and optimization of guards an important and non-trivial aspect of the construction
@@
 %stored at the different levels for the guards
-In this paper we want to substantiate the aforementioned observations about guards and
-describe based on them the reasoning behind their implementation in
+In this paper we want to substantiate the aforementioned observations and
+describe based on them the reasoning behind the implementation of guards in
 RPython's tracing just-in-time compiler. the contributions of this paper are:
@@
 \end{itemize}
+
+\begin{figure}
+    \include{figures/guard_table}
+    \caption{Percentage of guards before and after optimization for different benchmarks}
+    \label{fig:guard_percent}
+\end{figure}
 
@@
     \include{figures/failing_guards_table}
-    \caption{Failing guards, guards with more than 200 failures and guards responsible for 50\% of the failures relative to the total number of guards}
+    \caption{Failing guards relative to the total number of guards}
     \label{fig:failing_guards}
@@
 fail often enough that a bridge is compiled. Also, of all failing guards a few fail extremely often
-and most fail rarely. Reinforcing this notion the figure shows that, depending on the
-benchmark, between 0.008\% and 0.225\% of the guards are responsible for 50\%
-of the total guards failures.
-These results emphasize that as most of the guards never
+and most fail rarely. The results emphasize that as most of the guards never
 fail it is important to make sure that the successful execution of a guard does
@@
 feedback and valuable comments while writing this paper.
-We thank the PyPy and RPython community for their continuous support and work:
-Armin Rigo, Antonio Cuni, Maciej Fijałkowski, Samuele Pedroni, and countless
-others. Any remaining errors are our own.
 

### 2012-08-16 talk/dls2012/paper.tex

--- 2012-08-15 talk/dls2012/paper.tex
@@
 One of the nice properties of a tracing JIT is that many of its optimizations
-are simple requiring one forward pass only. This is not true for loop-invariant code
+are simple, requiring one forward pass only. This is not true for loop-invariant code
 motion which is a very important optimization for code with tight kernels.
@@
 \end{equation}
-In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
+In the optimized trace $J$ is replaced by $\hat J$ and $K$ by $\hat
 K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to
@@
 calculations were implemented in both Python, C and Lua and their
-runtimes are compared in Figure~\ref{fig:benchmarks}. The benchmarks are
+runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
+    The benchmarks and the scripts to run them can be found in the repository for this paper:
+    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
+}
+
+The benchmarks are
 \begin{itemize}
@@
   depending on the class of it's input argument, $y$, while in C,
-  there are three different implementations.
+  there are three different implementations. In Lua there is no support for
+  integers so only two versions are provided: float and Fix16. Here Fix16 is a custom class
+  that implements scaled floating point arithmetic.
 \item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
-is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_n\right)$ from a vector
+is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_{n-2}\right)$ from a vector
 ${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
-$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n$. Both the output vector, $\bf b$, 
+$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n-2$. Both the output vector, $\bf b$, 
 and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
@@
 \end{equation}
-for $1 \leq i \leq m$ and $1 \leq j \leq n$.
+for $2 \leq i \leq m-1$ and $2 \leq j \leq n-1$.
 The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
@@
 plain C. All the benchmarks except sqrt operate on C double-precision floating
-point numbers, both in the Python and the C code.
+point numbers, both in the Python, C and Lua code.
 
@@
 
-For PyPy 10 iterations were run, prefaced with 3 iterations for warming up.
+For PyPy and Lua 10 iterations were run, prefaced with 3 iterations for warming up.
 Due to benchmarks taking large amounts of time on CPython, only one run
@@
 the relative immaturity of RPython's JIT machine code backend as well as missing
-optimizations, like instruction scheduling.
+optimizations, like instruction scheduling. Also, in case of nested loops, 
+operations are only moved out of the 
+innermost loop. That is an issue when the innermost loop is 
+short and a significant amount of time is spent in the outer loops. This is the case 
+with for example SparseMatMult.
 
@@
 be seen as doing a similar optimization (again by manually implementing it)
-than the one described in Section~\ref{sub:allocation}: The effect of both is
+as the one described in Section~\ref{sub:allocation}: The effect of both is
 that type checks are fully done before a loop is even entered.

### 2012-08-16 talk/dls2012/paper.tex
explain the "type analysis"

@@
     an (inlined) method call and is followed by the trace of the called method.
+    The type that the guard checks for is the one that the variable had during
+    tracing.
 \end{itemize}
@@
 5 is the third time the idea is explained at basically the same level of detail
-(the others are in section 2 and section 4).  In contrast, the optimizations
-applied rely on a simple type analysis, but this is only briefly alluded to.
+(the others are in section 2 and section 4).
 }
@@
 {\smaller
-\begin{tabular}{|l|r|r|r|r|r|r|}
-\hline
- & CPython & Psyco & PyPy  & PyPy & GCC \\
- &         &       & no LP &      & -O3 \\
-\hline
-conv3(1e5) & 77.89 & 9.52 & 1.77  & 0.68 &  0.59 \\
-\hline
-conv3(1e6) & 77.15 & 9.58 & 1.69  & 0.77 &  0.74 \\
-\hline
-conv3x3(1000) & 233.54 & 125.40 & 0.57 & 0.27 & 0.25 \\
-\hline
-conv3x3(3) & 234.45 & 126.28 & 0.60 & 0.31 & 0.28 \\
-\hline
-conv5(1e5) & 122.54 & 16.67 & 1.86  & 1.05 &  0.65\\
-\hline
-conv5(1e6) & 125.77 & 16.80 & 1.92 & 1.09 &  0.80 \\
-\hline
-dilate3x3(1000) & 232.51 & 125.85 & 3.89 & 3.69 & 0.25 \\
-\hline
-sobel(1000) & 181.49 & 95.05 & 0.71 & 0.42 & 0.20 \\
-\hline
-sqrt(Fix16) & 744.35 & 421.65 & 3.93  & 2.14  & 0.96 \\
-\hline
-sqrt(float) & 24.21 & 5.52 & 1.36 & 1.00 & 0.98\\
-\hline
-sqrt(int) & 20.84 & 1.78 & 2.26 & 1.82  & 0.80 \\
-\hline
-\hline
-Variations & - & - & $\pm 0.03$ & $\pm 0.01$ & $\pm 0.01$ \\
+\begin{tabular}{|l|r|r|r|r|r|r|r|}
+\hline
+ & CPython & PyPy  & PyPy & GCC & luajit & luajit \\
+ &        & no LP &       & -O3 &        & no LP \\
+FFT(1024, 32768) & 469.07 & 20.83 +- 0.02 & 12.73 +- 0.01 & - & 2.76 +- 0.01 & 4.42 +- 0.01\\
+\hline
+FFT(1024,32768) & - & - & - & 1.40 +- 0.04 & - & -\\
+\hline
+FFT(1048576, 2) & 58.93 & 4.12 +- 0.01 & 2.05 +- 0.00 & - & 1.08 +- 0.01 & 1.33 +- 0.01\\
+\hline
+FFT(1048576,2) & - & - & - & 0.83 +- 0.02 & - & -\\
+\hline
+LU(100, 4096) & 1974.14 & 32.22 +- 0.14 & 13.39 +- 0.03 & - & 1.52 +- 0.01 & 8.57 +- 0.01\\
+\hline
+LU(100,4096) & - & - & - & 1.33 +- 0.04 & - & -\\
+\hline
+LU(1000, 2) & 955.31 & 14.98 +- 0.22 & 5.99 +- 0.21 & - & 0.67 +- 0.01 & 3.99 +- 0.01\\
+\hline
+LU(1000,2) & - & - & - & 0.65 +- 0.04 & - & -\\
+\hline
+MonteCarlo(268435456) & 618.89 & 20.60 +- 0.05 & 15.33 +- 0.08 & 1.69 +- 0.05 & 2.82 +- 0.00 & 3.92 +- 0.01\\
+\hline
+SOR(100, 32768) & 1458.12 & 8.24 +- 0.00 & 2.66 +- 0.00 & - & 1.31 +- 0.01 & 2.02 +- 0.00\\
+\hline
+SOR(100,32768) & - & - & - & 1.76 +- 0.04 & - & -\\
+\hline
+SOR(1000, 256) & 1210.45 & 6.48 +- 0.00 & 2.10 +- 0.00 & - & 1.08 +- 0.01 & 1.63 +- 0.00\\
+\hline
+SOR(1000,256) & - & - & - & 1.49 +- 0.02 & - & -\\
+\hline
+SparseMatMult(1000, 5000, 262144) & 371.66 & 24.25 +- 0.04 & 16.52 +- 0.04 & - & - & -\\
+\hline
+SparseMatMult(1000,5000,262144) & - & - & - & 1.84 +- 0.03 & 4.53 +- 0.02 & 9.64 +- 0.02\\
+\hline
+SparseMatMult(100000, 1000000, 1024) & 236.93 & 17.01 +- 0.01 & 8.75 +- 0.08 & - & - & -\\
+\hline
+SparseMatMult(100000,1000000,1024) & - & - & - & 1.20 +- 0.03 & 2.42 +- 0.01 & 7.19 +- 0.01\\
+\hline
+conv3(100, nil) & - & - & - & - & 0.18 +- 0.00 & 0.70 +- 0.00\\
+\hline
+conv3(1000, nil) & - & - & - & - & 0.12 +- 0.01 & 0.67 +- 0.01\\
+\hline
+conv3(1e5) & - & - & - & 0.52 +- 0.04 & - & -\\
+\hline
+conv3(1e6) & - & - & - & 0.60 +- 0.03 & - & -\\
+\hline
+conv3(array(1e5)) & 50.14 & 1.09 +- 0.01 & 0.49 +- 0.01 & - & - & -\\
+\hline
+conv3(array(1e6)) & 49.20 & 1.13 +- 0.02 & 0.51 +- 0.00 & - & - & -\\
+\hline
+conv3x3(1000) & - & - & - & 0.17 +- 0.04 & - & -\\
+\hline
+conv3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.01 & 0.49 +- 0.01\\
+\hline
+conv3x3(1000000, 3) & - & - & - & - & 0.13 +- 0.00 & 0.53 +- 0.00\\
+\hline
+conv3x3(3) & - & - & - & 0.19 +- 0.03 & - & -\\
+\hline
+conv3x3(Array2D(1000000x3)) & 139.81 & 0.70 +- 0.00 & 0.21 +- 0.00 & - & - & -\\
+\hline
+conv3x3(Array2D(1000x1000)) & 138.95 & 0.70 +- 0.00 & 0.20 +- 0.00 & - & - & -\\
+\hline
+conv5(100, nil) & - & - & - & - & 0.21 +- 0.01 & 0.87 +- 0.01\\
+\hline
+conv5(1000, nil) & - & - & - & - & 0.17 +- 0.01 & 0.84 +- 0.00\\
+\hline
+conv5(1e5) & - & - & - & 0.55 +- 0.02 & - & -\\
+\hline
+conv5(1e6) & - & - & - & 0.58 +- 0.03 & - & -\\
+\hline
+conv5(array(1e5)) & 74.65 & 1.22 +- 0.00 & 0.64 +- 0.00 & - & - & -\\
+\hline
+conv5(array(1e6)) & 77.94 & 1.26 +- 0.00 & 0.68 +- 0.01 & - & - & -\\
+\hline
+dilate3x3(1000) & - & - & - & 0.17 +- 0.03 & - & -\\
+\hline
+dilate3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.00 & 0.48 +- 0.01\\
+\hline
+dilate3x3(Array2D(1000x1000)) & 137.52 & 4.35 +- 0.01 & 3.91 +- 0.02 & - & - & -\\
+\hline
+sobel(Array2D(1000x1000)) & 104.02 & 0.49 +- 0.00 & 0.21 +- 0.00 & 0.17 +- 0.03 & - & -\\
+\hline
+sobel\_magnitude(1000, 1000) & - & - & - & - & 0.24 +- 0.01 & 0.60 +- 0.01\\
+\hline
+sqrt(Fix16) & 463.46 & 5.12 +- 0.00 & 2.96 +- 0.00 & 1.34 +- 0.03 & 1.14 +- 0.00 & 12.80 +- 0.04\\
+\hline
+sqrt(float) & 14.99 & 1.37 +- 0.00 & 0.89 +- 0.00 & 0.85 +- 0.04 & 0.83 +- 0.01 & 1.06 +- 0.01\\
+\hline
+sqrt(int) & 13.91 & 3.22 +- 0.02 & 2.65 +- 0.00 & 1.25 +- 0.03 & 0.83 +- 0.01 & 1.06 +- 0.00\\
 \hline
@@
 
-Benchmarks were run on Intel i7 M620 @2.67GHz with 4M cache and 8G of RAM
-using Ubuntu Linux 11.4 in 32bit mode.
+Benchmarks were run on Intel Xeon X5680 @3.33GHz with 12M cache and 16G of RAM
+using Ubuntu Linux 11.4 in 64bit mode.
 The machine was otherwise unoccupied. We use the following software
@@
 \begin{itemize}
-\item PyPy 1.5
-\item CPython 2.7.2
-\item Psyco 1.6 with CPython 2.6.6
-\item GCC 4.4.5 shipped with Ubuntu 11.4
+\item PyPy 1.9
+\item CPython 2.7.1
+\item GCC 4.5.2 shipped with Ubuntu 11.4
+\item LuaJIT 2.0 beta, commit ID 0dd175d9e711f039c663d35e96c149b705bcf450
 \end{itemize}

### 2012-08-16 talk/vmil2012/paper.tex
include Sven's notes

--- 2012-08-16 talk/vmil2012/paper.tex
@@
 
-\title{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}
+\title{The Efficient Handling of Guards \\in the Design of RPython's Tracing JIT}
 
@@
 describe based on them the reasoning behind their implementation in
-RPython's tracing just-in-time compiler. the contributions of this paper are:
+RPython's tracing just-in-time compiler. The contributions of this paper are:
 \begin{itemize}
@@
 overview about how guards are treated in the context of other just-in-time
-compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
+compilers. Finally, Section~\ref{sec:Conclusion} summarizes our conclusions and
 gives an outlook on further research topics.
@@
 the Python interpreter there are several experimental language implementation at different
-levels of completeness, e.g. for Prolog~\cite{bolz_towards_2010}, Smalltalk~\cite{bolz_towards_2010}, JavaScript and R.
+levels of completeness, e.g. for Prolog~\cite{bolz_towards_2010}, Smalltalk~\cite{bolz_back_2008}, JavaScript and R.
 
@@
     \input{figures/example.tex}
-    \caption{Example Program}
+    \caption{Example program}
     \label{fig:example}
@@
 that quite often not much changes between them.
-Since resume data is a linked list of symbolic frames
+Since resume data is a linked list of symbolic frames,
 in many cases only the information in the top frame changes from one guard to the next.
-The other symbolic frames can often just be reused.
-The reason for this is that during tracing only the variables
+The other symbolic frames can often be reused.
+The reason for this is that, during tracing only the variables
 of the currently executing frame can change.
@@
 
-In addition to sharing as much as possible between subsequent guards
+In addition to sharing as much as possible between subsequent guards,
 a compact representation of the local variables of symbolic frames is used.
@@
 
-After the recorded trace has been optimized it is handed over to the platform specific
+After the recorded trace has been optimized, it is handed over to the platform specific
 backend to be compiled to machine code. The compilation phase consists of two
@@
 semantics. Operations without side effects whose result is not used are not
-emitted. Guards instructions are transformed into fast checks at the machine
+emitted. Guard instructions are transformed into fast checks at the machine
 code level that verify the corresponding condition.  In cases the value being
@@
 
-First a special data
+First, a special data
 structure called \emph{backend map} is created. This data structure encodes the
@@
 
-Second for each guard a piece of code is generated that acts as a trampoline.
+Second, for each guard a piece of code is generated that acts as a trampoline.
 Guards are implemented as a conditional jump to this trampoline in case the
 guard check fails.
-In the trampoline the pointer to the
+In the trampoline, the pointer to the
 backend map is loaded and after storing the current execution state
@@
 
-As in previous sections the underlying idea for the low-level design of guards is to have
+As in previous sections, the underlying idea for the low-level design of guards is to have
 a fast on-trace profile and a potentially slow one in case
 the execution has to return to the interpreter. At the same
-time the data stored in the backend, required to rebuild the state, should be as
+time, the data stored in the backend, required to rebuild the state, should be as
 compact as possible to reduce the memory overhead produced by the large number
-of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
+of guards. The numbers in Figure~\ref{fig:backend_data} illustrate that the
 compressed encoding currently has about 15\% to 25\% of the size of of the
@@
 that occurs in a hot loop, for this reason the amount of generated machine
-code will be smaller than in other juts-in-time compilation approaches.  This
+code will be smaller than in other just-in-time compilation approaches.  This
 creates a larger discrepancy between the size of the resume data when
@@
 \section*{Acknowledgements}
-We would like to thank David Edelsohn, Samuele Pedroni and Stephan Zalewski for their helpful
+We would like to thank David Edelsohn, Samuele Pedroni, Stephan Zalewski and Sven Hager for their helpful
 feedback and valuable comments while writing this paper.
@@
 \bibliography{zotero,paper}
-\listoftodos
+%\listoftodos
 \end{document}

### 2012-08-16 talk/iwtc11/paper.tex
new benchmark results and computer info. formatting is screwed

@@
 {\smaller
-\begin{tabular}{|l|r|r|r|r|r|r|}
-\hline
- & CPython & Psyco & PyPy  & PyPy & GCC \\
- &         &       & no LP &      & -O3 \\
-\hline
-conv3(1e5) & 77.89 & 9.52 & 1.77  & 0.68 &  0.59 \\
-\hline
-conv3(1e6) & 77.15 & 9.58 & 1.69  & 0.77 &  0.74 \\
-\hline
-conv3x3(1000) & 233.54 & 125.40 & 0.57 & 0.27 & 0.25 \\
-\hline
-conv3x3(3) & 234.45 & 126.28 & 0.60 & 0.31 & 0.28 \\
-\hline
-conv5(1e5) & 122.54 & 16.67 & 1.86  & 1.05 &  0.65\\
-\hline
-conv5(1e6) & 125.77 & 16.80 & 1.92 & 1.09 &  0.80 \\
-\hline
-dilate3x3(1000) & 232.51 & 125.85 & 3.89 & 3.69 & 0.25 \\
-\hline
-sobel(1000) & 181.49 & 95.05 & 0.71 & 0.42 & 0.20 \\
-\hline
-sqrt(Fix16) & 744.35 & 421.65 & 3.93  & 2.14  & 0.96 \\
-\hline
-sqrt(float) & 24.21 & 5.52 & 1.36 & 1.00 & 0.98\\
-\hline
-sqrt(int) & 20.84 & 1.78 & 2.26 & 1.82  & 0.80 \\
-\hline
-\hline
-Variations & - & - & $\pm 0.03$ & $\pm 0.01$ & $\pm 0.01$ \\
-\hline
+\begin{tabular}{|l|r|r|r|r|r|r|r|}
+\hline
+ & CPython & PyPy  & PyPy & GCC & luajit & luajit \\
+ &        & no LP &       & -O3 &        & no LP \\
+FFT(1024, 32768) & 469.07 & 20.83 +- 0.02 & 12.73 +- 0.01 & - & 2.76 +- 0.01 & 4.42 +- 0.01\\
+\hline
+FFT(1024,32768) & - & - & - & 1.40 +- 0.04 & - & -\\
+\hline
+FFT(1048576, 2) & 58.93 & 4.12 +- 0.01 & 2.05 +- 0.00 & - & 1.08 +- 0.01 & 1.33 +- 0.01\\
+\hline
+FFT(1048576,2) & - & - & - & 0.83 +- 0.02 & - & -\\
+\hline
+LU(100, 4096) & 1974.14 & 32.22 +- 0.14 & 13.39 +- 0.03 & - & 1.52 +- 0.01 & 8.57 +- 0.01\\
+\hline
+LU(100,4096) & - & - & - & 1.33 +- 0.04 & - & -\\
+\hline
+LU(1000, 2) & 955.31 & 14.98 +- 0.22 & 5.99 +- 0.21 & - & 0.67 +- 0.01 & 3.99 +- 0.01\\
+\hline
+LU(1000,2) & - & - & - & 0.65 +- 0.04 & - & -\\
+\hline
+MonteCarlo(268435456) & 618.89 & 20.60 +- 0.05 & 15.33 +- 0.08 & 1.69 +- 0.05 & 2.82 +- 0.00 & 3.92 +- 0.01\\
+\hline
+SOR(100, 32768) & 1458.12 & 8.24 +- 0.00 & 2.66 +- 0.00 & - & 1.31 +- 0.01 & 2.02 +- 0.00\\
+\hline
+SOR(100,32768) & - & - & - & 1.76 +- 0.04 & - & -\\
+\hline
+SOR(1000, 256) & 1210.45 & 6.48 +- 0.00 & 2.10 +- 0.00 & - & 1.08 +- 0.01 & 1.63 +- 0.00\\
+\hline
+SOR(1000,256) & - & - & - & 1.49 +- 0.02 & - & -\\
+\hline
+SparseMatMult(1000, 5000, 262144) & 371.66 & 24.25 +- 0.04 & 16.52 +- 0.04 & - & - & -\\
+\hline
+SparseMatMult(1000,5000,262144) & - & - & - & 1.84 +- 0.03 & 4.53 +- 0.02 & 9.64 +- 0.02\\
+\hline
+SparseMatMult(100000, 1000000, 1024) & 236.93 & 17.01 +- 0.01 & 8.75 +- 0.08 & - & - & -\\
+\hline
+SparseMatMult(100000,1000000,1024) & - & - & - & 1.20 +- 0.03 & 2.42 +- 0.01 & 7.19 +- 0.01\\
+\hline
+conv3(100, nil) & - & - & - & - & 0.18 +- 0.00 & 0.70 +- 0.00\\
+\hline
+conv3(1000, nil) & - & - & - & - & 0.12 +- 0.01 & 0.67 +- 0.01\\
+\hline
+conv3(1e5) & - & - & - & 0.52 +- 0.04 & - & -\\
+\hline
+conv3(1e6) & - & - & - & 0.60 +- 0.03 & - & -\\
+\hline
+conv3(array(1e5)) & 50.14 & 1.09 +- 0.01 & 0.49 +- 0.01 & - & - & -\\
+\hline
+conv3(array(1e6)) & 49.20 & 1.13 +- 0.02 & 0.51 +- 0.00 & - & - & -\\
+\hline
+conv3x3(1000) & - & - & - & 0.17 +- 0.04 & - & -\\
+\hline
+conv3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.01 & 0.49 +- 0.01\\
+\hline
+conv3x3(1000000, 3) & - & - & - & - & 0.13 +- 0.00 & 0.53 +- 0.00\\
+\hline
+conv3x3(3) & - & - & - & 0.19 +- 0.03 & - & -\\
+\hline
+conv3x3(Array2D(1000000x3)) & 139.81 & 0.70 +- 0.00 & 0.21 +- 0.00 & - & - & -\\
+\hline
+conv3x3(Array2D(1000x1000)) & 138.95 & 0.70 +- 0.00 & 0.20 +- 0.00 & - & - & -\\
+\hline
+conv5(100, nil) & - & - & - & - & 0.21 +- 0.01 & 0.87 +- 0.01\\
+\hline
+conv5(1000, nil) & - & - & - & - & 0.17 +- 0.01 & 0.84 +- 0.00\\
+\hline
+conv5(1e5) & - & - & - & 0.55 +- 0.02 & - & -\\
+\hline
+conv5(1e6) & - & - & - & 0.58 +- 0.03 & - & -\\
+\hline
+conv5(array(1e5)) & 74.65 & 1.22 +- 0.00 & 0.64 +- 0.00 & - & - & -\\
+\hline
+conv5(array(1e6)) & 77.94 & 1.26 +- 0.00 & 0.68 +- 0.01 & - & - & -\\
+\hline
+dilate3x3(1000) & - & - & - & 0.17 +- 0.03 & - & -\\
+\hline
+dilate3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.00 & 0.48 +- 0.01\\
+\hline
+dilate3x3(Array2D(1000x1000)) & 137.52 & 4.35 +- 0.01 & 3.91 +- 0.02 & - & - & -\\
+\hline
+sobel(Array2D(1000x1000)) & 104.02 & 0.49 +- 0.00 & 0.21 +- 0.00 & 0.17 +- 0.03 & - & -\\
+\hline
+sobel\_magnitude(1000, 1000) & - & - & - & - & 0.24 +- 0.01 & 0.60 +- 0.01\\
+\hline
+sqrt(Fix16) & 463.46 & 5.12 +- 0.00 & 2.96 +- 0.00 & 1.34 +- 0.03 & 1.14 +- 0.00 & 12.80 +- 0.04\\
+\hline
+sqrt(float) & 14.99 & 1.37 +- 0.00 & 0.89 +- 0.00 & 0.85 +- 0.04 & 0.83 +- 0.01 & 1.06 +- 0.01\\
+\hline
+sqrt(int) & 13.91 & 3.22 +- 0.02 & 2.65 +- 0.00 & 1.25 +- 0.03 & 0.83 +- 0.01 & 1.06 +- 0.00\\
+\hline
+
 \end{tabular}
@@
 
-\revc{
-Include the OS with your benchmark specifications.
-}
-
-Benchmarks were run on Intel i7 M620 @2.67GHz with 4M cache and 8G of RAM in
-32bit mode.
+Benchmarks were run on Intel Xeon X5680 @3.33GHz with 12M cache and 16G of RAM
+using Ubuntu Linux 11.4 in 64bit mode.
 The machine was otherwise unoccupied. We use the following software

### 2012-08-16 talk/dls2012/paper.tex
some details

@@
 
-\revc{
-I would have liked to have benchmark results for some larger applications.
-When is this optimization effective on a large scale, if ever?
-}
-\cfbolz{I don't actually know. Does anybody?}
-
 \revd{
@@
 time savings are from removing boxing/unboxing operations.
-}
-
-\revd{
-This paper is relatively short, and could be significantly improved with a
-couple of pages of additional information about the details of the benchmarks
--- both on the Python and on the C side.
 }
@@
 }
+For benchmarks using larger Python applications the times are unaffected or
+slightly improved by the loop optimization of this paper.
 
@@
 In addition we also ported the 
-SciMark\footnote{\texttt{http://math.nist.gov/scimark2/}} benchmakts to python, and compared 
-their runtimes with the already existing Lua and C implementations. 
-This port was performed after the release of the pypy used to run the benchmarks which means that 
-these benchmarks have not influenced the pypy implementation.
-SciMark consists of
+SciMark\footnote{\texttt{http://math.nist.gov/scimark2/}} benchmarts to Python, and compared
+their runtimes with the already existing
+Lua\footnote{\texttt{http://luajit.org/download/scimark.lua}} and C
+implementations.
+
+SciMark consists of:
 

### 2012-08-16 talk/dls2012/benchmarks/benchmark.sh
kill the mention of psyco and analyze luajit a tiny bit

@@ -1,0 +1,89 @@
+#!/bin/bash
+
+echo
+echo $*
+if [ "$1" == "gcc" ]; then
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_double.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_long.c
+    ./runner.py -n 5 -c "$*" sqrt/sqrt_fix16.c
+    #./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1
+    #./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 100
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 100
+    ./runner.py -n 5 -c "$* -lm" convolution/conv3.c 1000
+    ./runner.py -n 5 -c "$* -lm" convolution/conv5.c 1000
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/conv3x3.cc 1000000 3
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/conv3x3.cc 1000 1000
+    ./runner.py -n 5 -c "$* -lstdc++" convolution/dilate3x3.cc 1000 1000
+    ./runner.py -n 5 -c "$* -lstdc++" image/sobel.cc 1000 1000
+    ./runner.py -n 5 -c "$*" scimark/run_SOR.c 100 32768
+    ./runner.py -n 5 -c "$*" scimark/run_SOR.c 1000 256
+    ./runner.py -n 5 -c "$*" scimark/run_SparseMatMult.c 1000 5000 262144
+    ./runner.py -n 5 -c "$*" scimark/run_SparseMatMult.c 100000 1000000 1024
+    ./runner.py -n 5 -c "$*" scimark/run_MonteCarlo.c 268435456
+    ./runner.py -n 5 -c "$*" scimark/run_LU.c 100 4096
+    ./runner.py -n 5 -c "$*" scimark/run_LU.c 1000 2
+    ./runner.py -n 5 -c "$* -lm" scimark/run_FFT.c 1024 32768
+    ./runner.py -n 5 -c "$* -lm" scimark/run_FFT.c 1048576 2
+    rm a.out
+elif [[ "$1" == luajit* ]]; then
+    $* runner.lua sqrt int
+    $* runner.lua sqrt float
+    $* runner.lua sqrt Fix16
+    $* runner.lua convolution conv3 100
+    $* runner.lua convolution conv5 100
[…]

### 2012-08-16 talk/dls2012/paper.tex
update the correct paper
make figure wide
this is done
kill the mention of psyco and analyze luajit a tiny bit

@@
 
-\revb{
-A nit: Section 7 says that loop peeling never makes runtime
-performance worse, but generating more code can potentially slow
-performance. I assume that non-numeric benchmarks show no slowdown in
-practice, and that might be worth noting.
-}
-
 \revd{
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
   specified.} 
-\end{figure}
+\end{figure*}
 
@@
 automatic loop vectorization. In all cases, SSE2 instructions were used for
-floating point operations, except Psyco which uses x87 FPU instructions.
-% Psyco does not use the x87 FPU: all floating-point arithmetic is done with
-% residual calls to C helpers.  These can probably be compiled with SSE2.
-% But compiling CPython (and maybe Psyco) for x87 or SSE2 has probably
-% no measurable effect.
-We also run PyPy with loop peeling optimization and without (but otherwise
+floating point operations.
+We also run PyPy and LuaJIT with loop peeling optimization and without (but otherwise
 identical).
 
-For PyPy and Lua 10 iterations were run, prefaced with 3 iterations for warming up.
+For PyPy and LuaJIT 10 iterations were run, prefaced with 3 iterations for warming up.
 Due to benchmarks taking large amounts of time on CPython, only one run
-was performed, prefaced with one warmup run for Psyco.
+was performed.
 For GCC 5 iterations
@@
 We can observe that PyPy (even without loop peeling) is orders of magnitude
-faster than either CPython or Psyco. This is due to the JIT compilation
+faster than CPython. This is due to the JIT compilation
 advantages and optimizations we discussed in previous
@@
 with for example SparseMatMult.
+
+The speedups that LuaJIT gains from the loop optimization pass are similar to
+those PyPy gains. In general, LuaJIT is even closer to C performance, sometimes
+even surpassing it. LuaJIT is generating machine code of higher quality because
+it has a much better register allocator than PyPy, among other things.
 

### 2012-08-16 talk/dls2012/paper.tex
collapse repetitions

--- 2012-08-16 talk/dls2012/paper.tex
@@
  &        & no LP &       & -O3 &        & no LP \\
-FFT(1024, 32768) & 469.07 & 20.83 +- 0.02 & 12.73 +- 0.01 & - & 2.76 +- 0.01 & 4.42 +- 0.01\\
-\hline
-FFT(1024,32768) & - & - & - & 1.40 +- 0.04 & - & -\\
-\hline
-FFT(1048576, 2) & 58.93 & 4.12 +- 0.01 & 2.05 +- 0.00 & - & 1.08 +- 0.01 & 1.33 +- 0.01\\
-\hline
-FFT(1048576,2) & - & - & - & 0.83 +- 0.02 & - & -\\
-\hline
-LU(100, 4096) & 1974.14 & 32.22 +- 0.14 & 13.39 +- 0.03 & - & 1.52 +- 0.01 & 8.57 +- 0.01\\
-\hline
-LU(100,4096) & - & - & - & 1.33 +- 0.04 & - & -\\
-\hline
-LU(1000, 2) & 955.31 & 14.98 +- 0.22 & 5.99 +- 0.21 & - & 0.67 +- 0.01 & 3.99 +- 0.01\\
-\hline
-LU(1000,2) & - & - & - & 0.65 +- 0.04 & - & -\\
+FFT(1024,32768) & 469.07 & 20.83 +- 0.02 & 12.73 +- 0.01 & 1.40 +- 0.04 & 2.76 +- 0.01 & 4.42 +- 0.01\\
+\hline
+FFT(1048576,2) & 58.93 & 4.12 +- 0.01 & 2.05 +- 0.00 & 0.83 +- 0.02 & 1.08 +- 0.01 & 1.33 +- 0.01\\
+\hline
+LU(100,4096) & 1974.14 & 32.22 +- 0.14 & 13.39 +- 0.03 & 1.33 +- 0.04 & 1.52 +- 0.01 & 8.57 +- 0.01\\
+\hline
+LU(1000,2) & 955.31 & 14.98 +- 0.22 & 5.99 +- 0.21 & 0.65 +- 0.04 & 0.67 +- 0.01 & 3.99 +- 0.01\\
 \hline
@@
 \hline
-SOR(100, 32768) & 1458.12 & 8.24 +- 0.00 & 2.66 +- 0.00 & - & 1.31 +- 0.01 & 2.02 +- 0.00\\
-\hline
-SOR(100,32768) & - & - & - & 1.76 +- 0.04 & - & -\\
-\hline
-SOR(1000, 256) & 1210.45 & 6.48 +- 0.00 & 2.10 +- 0.00 & - & 1.08 +- 0.01 & 1.63 +- 0.00\\
-\hline
-SOR(1000,256) & - & - & - & 1.49 +- 0.02 & - & -\\
-\hline
-SparseMatMult(1000, 5000, 262144) & 371.66 & 24.25 +- 0.04 & 16.52 +- 0.04 & - & - & -\\
-\hline
-SparseMatMult(1000,5000,262144) & - & - & - & 1.84 +- 0.03 & 4.53 +- 0.02 & 9.64 +- 0.02\\
-\hline
-SparseMatMult(100000, 1000000, 1024) & 236.93 & 17.01 +- 0.01 & 8.75 +- 0.08 & - & - & -\\
-\hline
-SparseMatMult(100000,1000000,1024) & - & - & - & 1.20 +- 0.03 & 2.42 +- 0.01 & 7.19 +- 0.01\\
-\hline
-conv3(100, nil) & - & - & - & - & 0.18 +- 0.00 & 0.70 +- 0.00\\
-\hline
-conv3(1000, nil) & - & - & - & - & 0.12 +- 0.01 & 0.67 +- 0.01\\
+SOR(100,32768) & 1458.12 & 8.24 +- 0.00 & 2.66 +- 0.00 & 1.76 +- 0.04 & 1.31 +- 0.01 & 2.02 +- 0.00\\
+\hline
+SOR(1000,256) & 1210.45 & 6.48 +- 0.00 & 2.10 +- 0.00 & 1.49 +- 0.02 & 1.08 +- 0.01 & 1.63 +- 0.00\\
+\hline
+SparseMatMult(1000,5000,262144) & 371.66 & 24.25 +- 0.04 & 16.52 +- 0.04 & 1.84 +- 0.03 & 4.53 +- 0.02 & 9.64 +- 0.02\\
+\hline
+SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 +- 0.01 & 8.75 +- 0.08 & 1.20 +- 0.03 & 2.42 +- 0.01 & 7.19 +- 0.01\\
+\hline
+conv3(100,nil) & - & - & - & - & 0.18 +- 0.00 & 0.70 +- 0.00\\
+\hline
+conv3(1000,nil) & - & - & - & - & 0.12 +- 0.01 & 0.67 +- 0.01\\
 \hline
@@
 \hline
-conv3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.01 & 0.49 +- 0.01\\
-\hline
-conv3x3(1000000, 3) & - & - & - & - & 0.13 +- 0.00 & 0.53 +- 0.00\\
+conv3x3(1000,1000) & - & - & - & - & 0.09 +- 0.01 & 0.49 +- 0.01\\
+\hline
+conv3x3(1000000,3) & - & - & - & - & 0.13 +- 0.00 & 0.53 +- 0.00\\
 \hline
@@
 \hline
-conv5(100, nil) & - & - & - & - & 0.21 +- 0.01 & 0.87 +- 0.01\\
-\hline
-conv5(1000, nil) & - & - & - & - & 0.17 +- 0.01 & 0.84 +- 0.00\\
+conv5(100,nil) & - & - & - & - & 0.21 +- 0.01 & 0.87 +- 0.01\\
+\hline
+conv5(1000,nil) & - & - & - & - & 0.17 +- 0.01 & 0.84 +- 0.00\\
 \hline
@@
 \hline
-dilate3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.00 & 0.48 +- 0.01\\
+dilate3x3(1000,1000) & - & - & - & - & 0.09 +- 0.00 & 0.48 +- 0.01\\
 \hline
@@
 \hline
-sobel\_magnitude(1000, 1000) & - & - & - & - & 0.24 +- 0.01 & 0.60 +- 0.01\\
+sobel_magnitude(1000,1000) & - & - & - & - & 0.24 +- 0.01 & 0.60 +- 0.01\\
 \hline

### 2012-08-16 talk/dls2012/paper.tex
address another reviewer comment

@@
 \section{Benchmarks}
-
-\revd{
-It isn't clear from the paper, but a reader might conclude that the bulk of the
-time savings are from removing boxing/unboxing operations.
-}
 
@@
 advantages and optimizations we discussed in previous
-work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
+work~\cite{bolz_allocation_2011, bolz_runtime_2011}, the main improvement for
+these concrete benchmarks come from the allocation removal/unboxing
+optimization.
+
+The geometric mean of the
 speedup of loop peeling is 70\%, which makes benchmark times

### 2012-08-16 talk/dls2012/benchmarks/parse.py
another round of deduplication

@@ -1,6 +1,19 @@
 
 import pdb, sys
 
+NAME_REPL = {
+    'dilate3x3(Array2D(1000x1000))': 'dilate3x3(1000,1000)',
+    'sobel_magnitude(1000,1000)': 'sobel(1000,1000)',
+    'conv3(array(1e5))': 'conv3(1e5)',
+    'conv3(array(1e6))': 'conv3(1e6)',
+    'conv5(array(1e5))': 'conv5(1e5)',
+    'conv5(array(1e6))': 'conv5(1e6)',
+    'sobel(Array2D(1000x1000))': 'sobel(1000,1000)',
+    'sobel(Array(1000x1000))': 'sobel(1000,1000)',
+    'conv3x3(Array2D(1000000x3))': 'conv3x3(1000000,3)',
+    'conv3x3(Array2D(1000x1000))': 'conv3x3(1000,1000)',
+}
+
 def main(name):
     interp = None
     res = {}
@@ -15,6 +28,7 @@ def main(name):
             else:
                 bench, rest = line.split(':')
                 bench = bench.replace(" ", "")
+                bench = NAME_REPL.get(bench, bench)
                 if '+-' in rest:
                     a, d = rest.split('+-')
                     res.setdefault(bench, {})[interp] = float(a), float(d)

### 2012-08-16 talk/dls2012/paper.tex
another round of deduplication
regenerate table

@@
 \hline
- & CPython & PyPy  & PyPy & GCC & luajit & luajit \\
- &        & no LP &       & -O3 &        & no LP \\
-FFT(1024,32768) & 469.07 & 20.83 +- 0.02 & 12.73 +- 0.01 & 1.40 +- 0.04 & 2.76 +- 0.01 & 4.42 +- 0.01\\
-\hline
-FFT(1048576,2) & 58.93 & 4.12 +- 0.01 & 2.05 +- 0.00 & 0.83 +- 0.02 & 1.08 +- 0.01 & 1.33 +- 0.01\\
-\hline
-LU(100,4096) & 1974.14 & 32.22 +- 0.14 & 13.39 +- 0.03 & 1.33 +- 0.04 & 1.52 +- 0.01 & 8.57 +- 0.01\\
-\hline
-LU(1000,2) & 955.31 & 14.98 +- 0.22 & 5.99 +- 0.21 & 0.65 +- 0.04 & 0.67 +- 0.01 & 3.99 +- 0.01\\
-\hline
-MonteCarlo(268435456) & 618.89 & 20.60 +- 0.05 & 15.33 +- 0.08 & 1.69 +- 0.05 & 2.82 +- 0.00 & 3.92 +- 0.01\\
-\hline
-SOR(100,32768) & 1458.12 & 8.24 +- 0.00 & 2.66 +- 0.00 & 1.76 +- 0.04 & 1.31 +- 0.01 & 2.02 +- 0.00\\
-\hline
-SOR(1000,256) & 1210.45 & 6.48 +- 0.00 & 2.10 +- 0.00 & 1.49 +- 0.02 & 1.08 +- 0.01 & 1.63 +- 0.00\\
-\hline
-SparseMatMult(1000,5000,262144) & 371.66 & 24.25 +- 0.04 & 16.52 +- 0.04 & 1.84 +- 0.03 & 4.53 +- 0.02 & 9.64 +- 0.02\\
-\hline
-SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 +- 0.01 & 8.75 +- 0.08 & 1.20 +- 0.03 & 2.42 +- 0.01 & 7.19 +- 0.01\\
-\hline
-conv3(100,nil) & - & - & - & - & 0.18 +- 0.00 & 0.70 +- 0.00\\
-\hline
-conv3(1000,nil) & - & - & - & - & 0.12 +- 0.01 & 0.67 +- 0.01\\
-\hline
-conv3(1e5) & - & - & - & 0.52 +- 0.04 & - & -\\
-\hline
-conv3(1e6) & - & - & - & 0.60 +- 0.03 & - & -\\
-\hline
-conv3(array(1e5)) & 50.14 & 1.09 +- 0.01 & 0.49 +- 0.01 & - & - & -\\
-\hline
-conv3(array(1e6)) & 49.20 & 1.13 +- 0.02 & 0.51 +- 0.00 & - & - & -\\
-\hline
-conv3x3(1000) & - & - & - & 0.17 +- 0.04 & - & -\\
-\hline
-conv3x3(1000,1000) & - & - & - & - & 0.09 +- 0.01 & 0.49 +- 0.01\\
-\hline
-conv3x3(1000000,3) & - & - & - & - & 0.13 +- 0.00 & 0.53 +- 0.00\\
-\hline
-conv3x3(3) & - & - & - & 0.19 +- 0.03 & - & -\\
-\hline
-conv3x3(Array2D(1000000x3)) & 139.81 & 0.70 +- 0.00 & 0.21 +- 0.00 & - & - & -\\
-\hline
-conv3x3(Array2D(1000x1000)) & 138.95 & 0.70 +- 0.00 & 0.20 +- 0.00 & - & - & -\\
-\hline
-conv5(100,nil) & - & - & - & - & 0.21 +- 0.01 & 0.87 +- 0.01\\
-\hline
-conv5(1000,nil) & - & - & - & - & 0.17 +- 0.01 & 0.84 +- 0.00\\
-\hline
-conv5(1e5) & - & - & - & 0.55 +- 0.02 & - & -\\
-\hline
-conv5(1e6) & - & - & - & 0.58 +- 0.03 & - & -\\
-\hline
-conv5(array(1e5)) & 74.65 & 1.22 +- 0.00 & 0.64 +- 0.00 & - & - & -\\
-\hline
-conv5(array(1e6)) & 77.94 & 1.26 +- 0.00 & 0.68 +- 0.01 & - & - & -\\
-\hline
-dilate3x3(1000) & - & - & - & 0.17 +- 0.03 & - & -\\
-\hline
-dilate3x3(1000,1000) & - & - & - & - & 0.09 +- 0.00 & 0.48 +- 0.01\\
-\hline
-dilate3x3(Array2D(1000x1000)) & 137.52 & 4.35 +- 0.01 & 3.91 +- 0.02 & - & - & -\\
-\hline
-sobel(Array2D(1000x1000)) & 104.02 & 0.49 +- 0.00 & 0.21 +- 0.00 & 0.17 +- 0.03 & - & -\\
-\hline
-sobel_magnitude(1000,1000) & - & - & - & - & 0.24 +- 0.01 & 0.60 +- 0.01\\
-\hline
-sqrt(Fix16) & 463.46 & 5.12 +- 0.00 & 2.96 +- 0.00 & 1.34 +- 0.03 & 1.14 +- 0.00 & 12.80 +- 0.04\\
-\hline
-sqrt(float) & 14.99 & 1.37 +- 0.00 & 0.89 +- 0.00 & 0.85 +- 0.04 & 0.83 +- 0.01 & 1.06 +- 0.01\\
-\hline
-sqrt(int) & 13.91 & 3.22 +- 0.02 & 2.65 +- 0.00 & 1.25 +- 0.03 & 0.83 +- 0.01 & 1.06 +- 0.00\\
+ & CPython & PyPy  & PyPy & LuaJIT & LuaJIT & GCC \\
+ &         & no LP &      & no LP  &        & -O3 \\
+\hline
+FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.42 $\pm$ 0.017 & 2.76 $\pm$ 0.017 & 1.40 $\pm$ 0.082\\
+\hline
+FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.33 $\pm$ 0.028 & 1.08 $\pm$ 0.014 & 0.83 $\pm$ 0.044\\
+\hline
+LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.012 & 1.52 $\pm$ 0.014 & 1.33 $\pm$ 0.070\\
+\hline
+LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 3.99 $\pm$ 0.014 & 0.67 $\pm$ 0.010 & 0.65 $\pm$ 0.077\\
+\hline
+MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.016 & 2.82 $\pm$ 0.009 & 1.69 $\pm$ 0.096\\
+\hline
+SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.009 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
+\hline
+SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.009 & 1.08 $\pm$ 0.010 & 1.49 $\pm$ 0.042\\
+\hline
+SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.64 $\pm$ 0.032 & 4.53 $\pm$ 0.032 & 1.84 $\pm$ 0.061\\
+\hline
+SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.016 & 2.42 $\pm$ 0.010 & 1.20 $\pm$ 0.053\\
+\hline
+\hline
+conv3(1e5) & 50.14 & 1.09 $\pm$ 0.022 & 0.49 $\pm$ 0.028 & 0.67 $\pm$ 0.010 & 0.12 $\pm$ 0.010 & 0.52 $\pm$ 0.084\\
+\hline
+conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.008 & 0.18 $\pm$ 0.000 & 0.60 $\pm$ 0.064\\
+\hline
+conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.49 $\pm$ 0.010 & 0.09 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
+\hline
+conv3x3(1000000,3) & 139.81 & 0.70 $\pm$ 0.005 & 0.21 $\pm$ 0.006 & 0.53 $\pm$ 0.008 & 0.13 $\pm$ 0.009 & 0.19 $\pm$ 0.061\\
+\hline
+conv5(1e5) & 74.65 & 1.22 $\pm$ 0.009 & 0.64 $\pm$ 0.005 & 0.84 $\pm$ 0.006 & 0.17 $\pm$ 0.010 & 0.55 $\pm$ 0.047\\
+\hline
+conv5(1e6) & 77.94 & 1.26 $\pm$ 0.009 & 0.68 $\pm$ 0.014 & 0.87 $\pm$ 0.010 & 0.21 $\pm$ 0.013 & 0.58 $\pm$ 0.049\\
+\hline
+dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.48 $\pm$ 0.014 & 0.09 $\pm$ 0.006 & 0.17 $\pm$ 0.061\\
+\hline
+sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.60 $\pm$ 0.012 & 0.24 $\pm$ 0.018 & 0.17 $\pm$ 0.061\\
+\hline
+sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.013 & 0.83 $\pm$ 0.010 & 0.85 $\pm$ 0.088\\
+\hline
+sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & 1.06 $\pm$ 0.009 & 0.83 $\pm$ 0.014 & 1.25 $\pm$ 0.053\\
+\hline
+sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & 12.80 $\pm$ 0.080 & 1.14 $\pm$ 0.009 & 1.34 $\pm$ 0.061\\
 \hline
@@
 \label{fig:benchmarks}
-\caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
+\caption{Benchmark results in seconds with 95\% confidence intervals. Arrays of length $10^5$ and
   $10^6$ and matrices of size $1000\times 1000$ and $1000000 \times

### 2012-08-16 talk/dls2012/paper.tex
date of luajit head
move stuff around

@@
 
-The loop peeling optimization was implemented in the PyPy
-framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
+The loop peeling optimization was implemented in RPython's tracing JIT
+in about 450 lines of RPython code. That means that the JIT-compilers generated for all
 interpreters implemented with RPython now can take advantage of
 it. Benchmarks have been executed for a few different interpreters and
-we see improvements in several cases. The ideal loop for this optimization
+we see improvements in several cases.
+
+An example of an RPython interpreter that is helped greatly by this
+optimization is our Prolog interpreter~\cite{bolz_towards_2010}. Prolog
+programs often contain tight loops that perform for example list processing.
+Furthermore we experimented with a Python library for writing numerical kernels
+doing array manipulation.
+
+The ideal loop for this optimization
 is short and contains numerical calculations with no failing guards and no
@@
 
-\subsection{Python}
 The Python interpreter of the RPython framework is a complete Python
@@
 \item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 
-initiated to $x_0=y/2$ and the benchmark consists of a single loop updating this
+initialized to $x_0=y/2$ and the benchmark consists of a single loop updating this
 approximation using $x_i = \left( x_{i-1} + y/x_{i-1} \right) / 2$ for $1\leq i < 10^8$. 
@@
 The same custom two-dimensional array class as described above is used to represent
-the gird.
+the grid.
 \item {\bf SparseMatMult}$\left(n, z, c\right)$: Matrix multiplication between a $n\times n$ sparse matrix,
@@
 \item GCC 4.5.2 shipped with Ubuntu 11.4
-\item LuaJIT 2.0 beta, commit ID 0dd175d9e711f039c663d35e96c149b705bcf450
+\item LuaJIT 2.0 beta, git head of August 15, 2012, commit ID 0dd175d9e711f039c663d35e96c149b705bcf450
 \end{itemize}
@@
 it has a much better register allocator than PyPy, among other things.
-
-Other interesting interpreters that are helped greatly by this optimization are
-for example our Prolog interpreter written in
-RPython~\cite{bolz_towards_2010}. Prolog programs often contain
-tight
-loops that perform list processing. Furthermore we experimented with a Python library
-for writing numerical kernels doing array manipulation. The exact extent is
-out of scope for this paper.
 

### 2012-08-16 talk/dls2012/paper.tex
it really doesn't make sense to run Fix16 with LuaJIT
I think these are fine

@@
 \label{sec:Motivation}
-
-\revc{
-Don't break code listings across pages, as at the start of section 3.  It makes
-them very hard to follow.
-}
-\cfbolz{let's do that only at the very end, just before submitting}
 
@@
 \section{Making Trace Optimizations Loop Aware}
-
-\revc{
-In general, the paper is over-long on generalities and too short on details.
-For example, the description of the basic technique at the beginning of section
-5 is the third time the idea is explained at basically the same level of detail
-(the others are in section 2 and section 4).
-}
 
@@
 \hline
-sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & 1.06 $\pm$ 0.009 & 0.83 $\pm$ 0.014 & 1.25 $\pm$ 0.053\\
-\hline
-sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & 12.80 $\pm$ 0.080 & 1.14 $\pm$ 0.009 & 1.34 $\pm$ 0.061\\
+sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & - & - & 1.25 $\pm$ 0.053\\
+\hline
+sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & - & - & 1.34 $\pm$ 0.061\\
 \hline
@@
   there are three different implementations. In Lua there is no support for
-  integers so only two versions are provided: float and Fix16. Here Fix16 is a custom class
-  that implements scaled floating point arithmetic.
+  integers so only the floating point number is provided.
 \item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop

### 2012-08-17 talk/dls2012/paper.tex
address review comments by mike pall

@@
 type checking, boxed value unwrapping and virtual method lookups.
+
 In this paper we explain a scheme pioneered within the context of the LuaJIT project
-for making simple optimizations loop-aware by
-using a simple pre-processing step on the trace and not changing the
+for making basic optimizations loop-aware by
+using a simple pre-processing step on the trace without changing the
 optimizations themselves.
+
 We have implemented the scheme in PyPy's tracing JIT compiler,
@@
 factor over two for PyPy's Python JIT executing simple numerical kernels
-bringing the performance close to that of compiled C code.
+bringing the performance into the ballpark of static language compilers.
 \end{abstract}
@@
 optimizations to achieve loop invariant code motion and other loop-related
-optimizations, which greatly simplifies the implementation. Using this scheme
-one does not need to change the underlying optimization much to get these
-advantages. We have implemented the same approach in PyPy's tracing JIT
-compiler the results of which we present here.
+optimizations, which greatly simplifies the implementation. We have implemented
+the same approach in PyPy's tracing JIT compiler, the results of which we
+present here.
 
@@
 \label{fig:benchmarks}
-\caption{Benchmark results in seconds with 95\% confidence intervals. Arrays of length $10^5$ and
-  $10^6$ and matrices of size $1000\times 1000$ and $1000000 \times
-  3$ are used. The one used in each benchmark is indicated in
-  the leftmost column. For the matrices, only the number of rows are
-  specified.} 
+\caption{Benchmark results in seconds with 95\% confidence intervals. The leftmost column gives the
+name of each benchmark and the values of the benchmark parameters used. The different benchmarks and the meaning of their parameters are described in Section~\ref{sec:benchmarks}.}
 \end{figure*}
@@
 comparable with native-compiled C code. We attribute the performance gap to C code to
-the relative immaturity of RPython's JIT machine code backend as well as missing
-optimizations, like instruction scheduling. Also, in case of nested loops, 
+the relative immaturity of RPython's JIT machine code backend and the naive register allocator.
+Also, in case of nested loops,
 operations are only moved out of the 
@@
 even surpassing it. LuaJIT is generating machine code of higher quality because
-it has a much better register allocator than PyPy, among other things.
+it has more optimizations\footnote{See
+\texttt{http://wiki.luajit.org/Optimizations}} and is a produces much better
+machine code than PyPy.
 
@@
 We would like to thank Samuele Pedroni, Sven Hager and the anonymous reviewers
-for helpful comments on drafts of this paper. We owe deep gratitude to Mike Pall
-for making his impressive work on LuaJIT available and for detailed help on a
-draft of the paper.
+for helpful comments on drafts of this paper. We owe gratitude to Mike Pall
+for making his impressive work on LuaJIT publicly available and for detailed
+reviews on drafts of the paper.
 

### 2012-08-17 talk/ep2011/training/src/html_fibo.py
Dont try to explain the parameters in this caption, its not that simple anymore

@@ -1,0 +1,47 @@
+"""
+The most complicate ever way to produce an HTML list of fibonacci numbers
+"""
+
+def fibo():
+    a, b = 1, 1
+    while True:
+        yield a
+        a, b = b, a+b
+
+
+class HtmlTag(object):
+    def __init__(self, f, indent, tag):
+        self.f = f
+        self.tag = tag
+        self.f.write(' ' * indent)
+        self.f.write('<%s>' % tag)
+
+    def __del__(self):
+        self.f.write('</%s>\n' % self.tag)
+
+def html_fibo(f):
+    f.write('<ul>\n')
+    try:
+        for n in fibo():
+            tag = HtmlTag(f, 4, 'li')
+            yield n
+            tag = None
+    finally:
+        tag = None
+        f.write('</ul>\n')
+
+
+def write_file():
+    f = open('fibo.txt', 'w')
+    for n in html_fibo(f):
+        f.write('%d' % n)
+        if n > 100:
+            break
+
+def main():
+    write_file()
+    content = open('fibo.txt').read()
+    print content
+
+if __name__ == '__main__':
+    main()

### 2012-08-17 talk/vmil2012/paper.tex

@@
 \begin{itemize}
-  \item an analysis and benchmark of guards in the context of RPython's tracing JIT,
-  %An analysis of guards in the context of RPython's tracing JIT to
-  %substantiate the aforementioned observation, based on a set of benchmarks,
+  \item An analysis guards in the context of RPython's tracing JIT,
   \item detailed measurements about the frequency and the
@@
   \item a description about how guards are implemented in the high\-
-  and low-level components of the JIT and describe the rationale behind the design
+  and low-level components of RPython's JIT and a description of the rationale behind the design.
 \end{itemize}
@@
 and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
-Data} we proceed to describe for RPython's tracing JIT the details of guards in
-the frontend. In this context the frontend is concerned with recording and storing the
-information required to rebuild the interpreter state in case of a guard
-failure. Once the frontend has traced and optimized a loop it invokes the
+Data} we proceed to describe the details of guards in
+the frontend of RPython's tracing JIT.
+Once the frontend has traced and optimized a loop it invokes the
 backend to compile the operations to machine code, Section~\ref{sec:Guards in
@@
 
-The RPython language and the PyPy project~\cite{rigo_pypys_2006} were started
+The RPython language and the PyPy project\footnote{\url{http://pypy.org}}~\cite{rigo_pypys_2006} were started
 in 2002 with the goal of
 creating a Python interpreter written in a high level language, allowing easy
-language experimentation and extension.\footnote{\url{http://pypy.org}} PyPy is now a fully compatible
+language experimentation and extension. PyPy is now a fully compatible
 alternative interpreter for the Python language.
@@
 
-different levels of completeness.
-
 RPython can mean one of two things:
@@
 \end{itemize}
-The RPython language
-is a statically typed object-oriented high level language. The language provides
+The RPython language, is a subset of Python that provides a
+statically typed object-oriented high level language. The subset of Python available in RPython is chosen in a way type inference is possible\cite{ancona_rpython:_2007}.
+The language provides
 several features such as automatic memory management
@@
 store the recorded operations, typically in SSA
-form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
+form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution, the
 code that is recorded
@@
 divergence from the recorded path are marked with special operations called
-\emph{guards}, these operations ensure that assumptions valid during the
+\emph{guards}. These operations ensure that assumptions valid during the
 tracing phase are still valid when the code has been compiled and is executed.
@@
 that come from optimistic type specialization by recording the types of
-variables seen during tracing.
+variables seen during tracing\cite{Gal:2009ux}.
 After a trace has been recorded it is optimized and then compiled to platform
@@
 
+In this context we refer to frontend as the component of the JIT that is
+concerned with recording and optimizing the traces as well as storing the
+information required to rebuild the interpreter state in case of a guard
+failure.
 Since tracing linearizes control flow by following one concrete execution,
@@
 MOVNE r8, #0
+...
 CMP r8, #0
@@
 BNE <bailout>
+...
 ...
@@
 This data
-structure stores the values in a succinct manner using an encoding that requires
-8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
+structure stores the values in a succinct manner. The encoding is efficient to create and
 provides a compact representation of the needed information in order
@@
 instead of bailing out. Once the guard has been compiled and attached to the
-loop the guard becomes just a point where control-flow can split. The loop
-after the guard and the bridge are just conditional paths.
+loop the guard becomes just a point where control-flow can split.
+The guard becomes the branching point of two conditional paths with no
+additional overhead.
 Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
@@
 
+\subsection{Guard Failures}
+\label{sub:guard_failure}
+The last point in this discussion is the frequency of guard failures.
+Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
+relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.\footnote{
+    The threshold used is 200 failures. This rather high threshold was picked experimentally to give
+    good results for long-running programs.
+}
+
+The numbers presented for guards that have a bridge represent the
+failures up to the compilation of the bridge and all executions of the then
+attached bridge.
+
+\begin{figure}
+    \include{figures/failing_guards_table}
+    \caption{Failing guards, guards with more than 200 failures and guards responsible for 50\% of the failures relative to the total number of guards}
+    \label{fig:failing_guards}
+\end{figure}
+
+From Figure~\ref{fig:failing_guards} we can see that only a very small amount
+of all the guards in the compiled traces ever fail. This amount varies between
+2.4\% and 5.7\% of all guards. As can be expected, even fewer, only 1.2\% to 3.6\% of all guards fail often
+enough that a bridge is compiled for them.
+Also, of all failing guards a few fail extremely often
+and most fail rarely. Reinforcing this notion the figure shows that, depending on the
+benchmark, between 0.008\% and 0.225\% of the guards are responsible for 50\%
+of the total guards failures.
+These results emphasize that as most of the guards never
+fail it is important to make sure that the successful execution of a guard does
+not have unnecessary overhead.
+
+This low guard failure rate is expected. Most guards do not come from actual
+control flow divergences in the user program, but from type checks needed for
+type specialization. Various prior work has
+shown~\cite{holkner_evaluating_2009, richards_analysis_2010, callau_how_2011}
+that most programs in dynamic languages only use a limited amount of runtime
+variability. Therefore many guards are needed for making the traces behave
+correctly in all cases but fail rarely.
+
 \subsection{Space Overhead of Guards}
 \label{sub:guard_overhead}
+
 \begin{figure}
-    \include{figures/resume_data_table}
-    \caption{Resume data sizes}
-    \label{fig:resume_data_sizes}
+    \include{figures/backend_table}
+    \caption{Total size of generated machine code and resume data}
+    \label{fig:backend_data}
 \end{figure}
@@
 \begin{figure}
-    \include{figures/backend_table}
-    \caption{Total size of generated machine code and resume data}
-    \label{fig:backend_data}
+    \include{figures/resume_data_table}
+    \caption{Resume data sizes}
+    \label{fig:resume_data_sizes}
 \end{figure}
@@
 and 21.1\% of the space required by our compression shows that the compression
-is not optimal but a trade-off between the required space and the time needed
-to build a good, compressed representation of the resume data for the
-large amount of guards present in the traces.
-
-\subsection{Guard Failures}
-\label{sub:guard_failure}
-The last point in this discussion is the frequency of guard failures.
-Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
-relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.\footnote{
-    The threshold used is 200 failures. This rather high threshold was picked experimentally to give
-    good results for long-running programs.
-}
-
-The numbers presented for guards that have a bridge represent the
-failures up to the compilation of the bridge and all executions of the then
-attached bridge.
-
-\begin{figure}
-    \include{figures/failing_guards_table}
-    \caption{Failing guards, guards with more than 200 failures and guards responsible for 50\% of the failures relative to the total number of guards}
-    \label{fig:failing_guards}
-\end{figure}
-
-From Figure~\ref{fig:failing_guards} we can see that only a very small amount
-of all the guards in the compiled traces ever fail. This amount varies between
-2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
-enough that a bridge is compiled for them, only 1.2\% to 3.6\% of all guards
-fail often enough that a bridge is compiled. Also, of all failing guards a few fail extremely often
-and most fail rarely. Reinforcing this notion the figure shows that, depending on the
-benchmark, between 0.008\% and 0.225\% of the guards are responsible for 50\%
-of the total guards failures.
-These results emphasize that as most of the guards never
-fail it is important to make sure that the successful execution of a guard does
-not have unnecessary overhead.
-
-This low guard failure rate is expected. Most guards do not come from actual
-control flow divergences in the user program, but from type checks needed for
-type specialization. Various prior work has
-shown~\cite{holkner_evaluating_2009, richards_analysis_2010, callau_how_2011}
-that most programs in dynamic languages only use a limited amount of runtime
-variability. Therefore many guards are needed for making the traces behave
-correctly in all cases but fail rarely.
+is not optimal and could be improved taking into account the trade-off between
+the required space and the time needed to build a good, compressed
+representation of the resume data for the large amount of guards present in the
+traces.
 
@@
 Pall mentions plans to switch to compressed snapshots to further reduce
-redundancy. The approach of not creating snapshots at all for every guard is
+redundancy.\footnote{This optimization is now implemented in LuaJIT, at the time of writing it has not been fully documented in the LuaJIT Wiki (\url{http://wiki.luajit.org/Optimizations\#1-D-Snapshot-Compression}).}
+The approach of not creating snapshots at all for every guard is
 orthogonal to the resume data compression presented in this paper and could be

### 2012-08-17 talk/dls2012/paper.tex
too much capitalization

--- 2012-08-17 talk/dls2012/paper.tex
@@
 \end{lstlisting}
-\caption{An Unoptimized Trace of the Example Interpreter}
+\caption{An Unoptimized Trace of the example interpreter}
 \label{fig:unopt-trace}
@@
 \end{lstlisting}
-\caption{The fully optimized loop of the Example Interpreter}
+\caption{The fully optimized loop of the example interpreter}
 \label{fig:opt-trace}
@@
 
+The large input parameters of the SciMark benchmarks are chosen in such a way
+to make the problem not fit into the CPU cache. This explains why PyPy is doing
+relatively better on them. The cache miss penalties are large relative to the
+time needed to perform the actual computations, which hides problems of the
+less efficient code generated by PyPy.
+
 The speedups that LuaJIT gains from the loop optimization pass are similar to
@@
 it has more optimizations\footnote{See
-\texttt{http://wiki.luajit.org/Optimizations}} and is a produces much better
+\texttt{http://wiki.luajit.org/Optimizations}} and produces much better
 machine code than PyPy.

### 2012-08-17 talk/vmil2012/paper.tex
too much capitalization

@@
 \end{itemize}
-The RPython language, is a subset of Python that provides a
-statically typed object-oriented high level language. The subset of Python available in RPython is chosen in a way type inference is possible\cite{ancona_rpython:_2007}.
+The RPython language is a
+statically typed object-oriented high-level subset of Python. The subset is chosen in such a way to make type inference possible\cite{ancona_rpython:_2007}.
 The language provides
@@
 Pall mentions plans to switch to compressed snapshots to further reduce
-redundancy.\footnote{This optimization is now implemented in LuaJIT, at the time of writing it has not been fully documented in the LuaJIT Wiki (\url{http://wiki.luajit.org/Optimizations\#1-D-Snapshot-Compression}).}
+redundancy.\footnote{This optimization is now implemented in LuaJIT, at the time of writing it has not been fully documented in the LuaJIT Wiki: \url{http://wiki.luajit.org/Optimizations\#1-D-Snapshot-Compression}}
 The approach of not creating snapshots at all for every guard is

### 2012-08-17 talk/dls2012/benchmarks/parse.py

@@ -1,0 +1,85 @@
+
+import pdb, sys
+import numpy as np
+import matplotlib.pyplot as plt
+
+NAME_REPL = {
+    'dilate3x3(Array2D(1000x1000))': 'dilate3x3(1000,1000)',
+    'sobel_magnitude(1000,1000)': 'sobel(1000,1000)',
+    'conv3(array(1e5))': 'conv3(1e5)',
+    'conv3(array(1e6))': 'conv3(1e6)',
+    'conv5(array(1e5))': 'conv5(1e5)',
+    'conv5(array(1e6))': 'conv5(1e6)',
+    'sobel(Array2D(1000x1000))': 'sobel(1000,1000)',
+    'sobel(Array(1000x1000))': 'sobel(1000,1000)',
+    'conv3x3(Array2D(1000000x3))': 'conv3x3(1000000,3)',
+    'conv3x3(Array2D(1000x1000))': 'conv3x3(1000,1000)',
+    'dilate3x3(1000)': 'dilate3x3(1000,1000)',
+    'conv3x3(1000)': 'conv3x3(1000,1000)',
+    'conv3x3(3)': 'conv3x3(1000000,3)',
+}
+
+def main(name):
+    interp = None
+    res = {}
+    order = ['python2.7', 'pypy --jit enable_opts=intbounds:rewrite:virtualize:string:earlyforce:pure:heap:ffi', 'pypy', 'luajit -O-loop', 'luajit', 'gcc -O3 -march=native -fno-tree-vectorize']
+    labels = ['CPython', 'PyPy no LP', 'PyPy', 'LuaJIT no LP', 'LuaJIT', 'gcc -O3']
+    with open(name) as f:
+        for line in f:
+            line = line.strip("\n")
+            if not line:
+                interp = None
+            elif interp is None:
+                interp = line
+            else:
+                bench, rest = line.split(':')
+                bench = bench.replace(" ", "")
+                bench = NAME_REPL.get(bench, bench)
+                if '+-' in rest:
+                    a, d = rest.split('+-')
+                    res.setdefault(bench, {})[interp] = float(a), float(d)
+                else:
+                    res.setdefault(bench, {})[interp] = float(rest)
+    resmat = np.zeros((len(res), len(order)))
+    for i, key in enumerate(sorted(res.keys())):
+        sys.stdout.write(key)
+        for j, ord in enumerate(order):
+            try:
+                e = res[key][ord]
+            except KeyError:
+                sys.stdout.write(" & -")
+            else:
[…]

### 2012-08-17 talk/dls2012/benchmarks/sqrt/sqrt.py

@@ -1,0 +1,54 @@
+def sqrt(y, n=10000):
+    x = y / 2
+    while n > 0:
+        #assert y > 0 and x > 0
+        n -= 1
+        x = (x + y/x) / 2
+    return x
+
+class Fix16(object):
+    def __init__(self, val, scale=True):
+        if isinstance(val, Fix16):
+            self.val = val.val
+        else:
+            if scale:
+                self.val = int(val * 2**16)
+            else:
+                self.val = val
+        #assert self.val <= 2147483647>>8
+
+    def __add__(self, other):
+        return  Fix16(self.val + Fix16(other).val, False)
+
+    def __sub__(self, other):
+        return  Fix16(self.val - Fix16(other).val, False)
+
+    def __mul__(self, other):
+        return  Fix16((self.val >> 8) * (Fix16(other).val >> 8), False)
+
+    def __div__(self, other):
+        return  Fix16((self.val << 8) / (Fix16(other).val >> 8), False)
+
+
+    def __float__(self):
+        return float(self.val) / float(2**16)
+
+    def __int__(self):
+        return self.val >> 16
+
+    def __cmp__(self, other):
+        return cmp(self.val, Fix16(other).val)
+
+    def __str__(self):
+        return str(float(self))
+
+    __radd__ = __add__
+    __rmul__ = __mul__
+    def __rsub__(self, other):
+        return  Fix16(Fix16(other).val - self.val, False)
+    def __rdiv__(self, other):
+        return  Fix16((Fix16(other).val << 8) / (self.val >> 8), False)
+
+def main(argv):
+    sqrt(eval(argv[0])(123), 100000000)
+    return 'sqrt(%s)' % argv[0]

### 2012-08-17 talk/badhonnef2007/talk.tex
- repetition - typo
Otherwise the talk looks very good to me.
tune plot

@@
 
-  \setbeamercovered{transparent}
+  %\setbeamercovered{transparent}
   % or whatever (possibly just delete it)
@@
 
-\begin{frame}
-  \frametitle{Outline}
-  \tableofcontents
+%\begin{frame}
+%  \frametitle{Outline}
+%  \tableofcontents
   % You might wish to add the option [pausesections]
-\end{frame}
+%\end{frame}
 
@@
   \item
-    Pyrolog is a Prolog interpreter written in RPython
+    Pyrolog is a Prolog interpreter written in Python
+  \item
+    want to compile it to get interesting performance
+  \item
+    translation tool-chain part of the PyPy project
+  \item
+    Python itself is too dynamic to be translatable to other languages, need a subset
   \item
@@
   \item
-    Python itself is too dynamic to be translatable to other languages, need a subset
-  \item
-    includes a translation tool-chain for RPython
+    most important part: translation tool-chain for RPython
   \item
@@
 \begin{frame}
-  \frametitle{The Python case (i)}
+  \frametitle{The Python case}
   CPython (the reference implementation) is a straightforward, portable VM.
@@
   \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{The Python case (ii)}
-  \begin{block}{Problems of Extensions:}
-    \begin{itemize}
-    \item hard to maintain: need to keep track of CPython
-    \item Psyco very hard to port to other hardware architectures
-    \item tedious to write them, Python semantics need to be re-implemented
-    \end{itemize}
-  \end{block}
-  \begin{itemize}
-  \item
-    The community wants Python to run everywhere:
-    Jython (Java), IronPython (.NET).
-    Lots of effort and duplication.
-
-  \item
-    At various points various incompatibilities between
-    the implementations
-
-  \end{itemize}
 \end{frame}
@@
       often extensions to core Prolog, incompatible between each other
-    \item
-      fixed implementation decisions (GC, how to generate code, etc.)
     \end{itemize}
   \end{block}
+  \pause
   \begin{block}{implementations on CLR and JVM}
@@
     \item
-      no extensions to core Prolog (like tabling, coroutines)
+      no extensions to core Prolog (like tabling, coroutines, constraints)
     \item
@@
     \item C-like
+    \item .NET
     \item Java
-    \item .NET
     \end{itemize}
+  \pause
   \item
     Insert low-level aspects into the code as required by
-    the target (Object layout, memory management)
+    the target
     \begin{itemize}
@@
     \item memory management
-    \end{itemize}  \item
+    \end{itemize}
+  \pause
+  \item
     Optionally insert new pervasive features not expressed
@@
   \frametitle{Translation Aspects (i)}
-  Features not present in the source can be added during translation.
+  Features not present in the source can be added during translation:
   \begin{itemize}
@@
   \item
-    Prolog calls mapped to RPython calls
-  \item
     implements large parts of the ISO standard (some builtins missing)
@@
   \item
-    after translation to C: 14000 line of C code
+    after translation to C: 14000 lines of C code
   \item
@@
 
-  % Keep the summary *very short*.
   \begin{itemize}
@@
   \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Questions?}
+  \begin{center}
+  \includegraphics[scale=0.5]{image/py-web.png}
+
+  \texttt{http://codespeak.net/pypy}
+  \end{center}
 \end{frame}

### 2012-08-17 talk/dls2012/paper.tex
tune plot
Update.

@@
 \end{lstlisting}
-\caption{An Unoptimized Trace of the example interpreter}
+\caption{An Unoptimized Trace of the Example Interpreter}
 \label{fig:unopt-trace}
@@
 \end{lstlisting}
-\caption{The fully optimized loop of the example interpreter}
+\caption{The fully optimized loop of the Example Interpreter}
 \label{fig:opt-trace}
@@
 \hline
-\hline
-conv3(1e5) & 50.14 & 1.09 $\pm$ 0.022 & 0.49 $\pm$ 0.028 & 0.67 $\pm$ 0.010 & 0.12 $\pm$ 0.010 & 0.52 $\pm$ 0.084\\
+%\hline
+%conv3(1e5) & 50.14 & 1.09 $\pm$ 0.022 & 0.49 $\pm$ 0.028 & 0.67 $\pm$ 0.010 & 0.12 $\pm$ 0.010 & 0.52 $\pm$ 0.084\\
 \hline
@@
 \hline
-conv3x3(1000000,3) & 139.81 & 0.70 $\pm$ 0.005 & 0.21 $\pm$ 0.006 & 0.53 $\pm$ 0.008 & 0.13 $\pm$ 0.009 & 0.19 $\pm$ 0.061\\
-\hline
-conv5(1e5) & 74.65 & 1.22 $\pm$ 0.009 & 0.64 $\pm$ 0.005 & 0.84 $\pm$ 0.006 & 0.17 $\pm$ 0.010 & 0.55 $\pm$ 0.047\\
-\hline
-conv5(1e6) & 77.94 & 1.26 $\pm$ 0.009 & 0.68 $\pm$ 0.014 & 0.87 $\pm$ 0.010 & 0.21 $\pm$ 0.013 & 0.58 $\pm$ 0.049\\
-\hline
+%conv3x3(1000000,3) & 139.81 & 0.70 $\pm$ 0.005 & 0.21 $\pm$ 0.006 & 0.53 $\pm$ 0.008 & 0.13 $\pm$ 0.009 & 0.19 $\pm$ 0.061\\
+%\hline
+%conv5(1e5) & 74.65 & 1.22 $\pm$ 0.009 & 0.64 $\pm$ 0.005 & 0.84 $\pm$ 0.006 & 0.17 $\pm$ 0.010 & 0.55 $\pm$ 0.047\\
+%\hline
+%conv5(1e6) & 77.94 & 1.26 $\pm$ 0.009 & 0.68 $\pm$ 0.014 & 0.87 $\pm$ 0.010 & 0.21 $\pm$ 0.013 & 0.58 $\pm$ 0.049\\
+%\hline
 dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.48 $\pm$ 0.014 & 0.09 $\pm$ 0.006 & 0.17 $\pm$ 0.061\\
@@
 and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
-with $n=10^5$ and $n=10^6$.
-\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
-${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
+with $n=10^5$.
+%\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
+%${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
 hardcoded into the implementation making the benchmark consist of a single loop too.
@@
 The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
-as well as $(n,m)=(1000000,3)$ was used.
+ was used.
 \item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed

### 2012-08-17 talk/dls2012/paper.tex

@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.5\textwidth]{benchmarks/result.pdf}
+\label{fig:benchmarks_plot}
+\caption{Benchmark results normalized with the runtime of the C version. The CPython results have been omitted to make the plot readable.}
+\end{center}
+\end{figure}
+
 The Python interpreter of the RPython framework is a complete Python
@@
 calculations were implemented in both Python, C and Lua and their
-runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
+runtimes are compared in Figuare~\ref{fig:benchmarks_plot} and Figure~\ref{fig:benchmarks}.\footnote{
     The benchmarks and the scripts to run them can be found in the repository for this paper:

### 2012-08-17 talk/dls2012/paper.tex
describe control dependencies

--- 2012-08-17 talk/dls2012/paper.tex
@@
 K$.
+
+It is interesting to note that the described approach automatically deals with
+implicit control dependencies correctly, whereas in other approaches this needs
+to be carefully programmed in. A commonly used example for a control dependency
+is a division operation that needs to be preceded by a check for the second
+argument being 0. In a trace, such a check would be done with a guard. The
+division operation must not be moved before that guard, and indeed, this is
+never done. If the division is loop invariant, the result computed in copy of
+the division operation in the preamble is reused. This division operation is
+preceded by a copy of the non-null guard, which ensures that it can be executed
+correctly.
 

### 2012-08-17 talk/vmil2012/paper.tex
(cfbolz, leuschel): some tweaks to the introduction and abstract

@@
 execution speed, deoptimization,
-and memory overhead.  In this paper we describe the design decisions about
-guards taken in the implementation of the RPython tracing JIT. Furthermore we
-measure various properties of guards.
+and memory overhead. In this paper, we perform an empirical analysis of runtime
+properties of guards. This is used to guide the design of guards in the RPython
+tracing JIT.
 % \o/
@@
 and to describe the concrete techniques used in the various layers of RPython's
-tracing JIT. All design decisions will be motivated by concrete numbers for the
+tracing JIT. All design decisions are be motivated by an empirical analysis of the
 frequency and the overhead related to guards.

### 2012-08-17 talk/dls2012/benchmarks/parse.py
Add a paragraph about the violation of immutability.

@@ -1,0 +1,89 @@
+
+import pdb, sys
+import numpy as np
+import matplotlib.pyplot as plt
+
+NAME_REPL = {
+    'dilate3x3(Array2D(1000x1000))': 'dilate3x3(1000,1000)',
+    'sobel_magnitude(1000,1000)': 'sobel(1000,1000)',
+    'conv3(array(1e5))': 'conv3(1e5)',
+    'conv3(array(1e6))': 'conv3(1e6)',
+    'conv5(array(1e5))': 'conv5(1e5)',
+    'conv5(array(1e6))': 'conv5(1e6)',
+    'sobel(Array2D(1000x1000))': 'sobel(1000,1000)',
+    'sobel(Array(1000x1000))': 'sobel(1000,1000)',
+    'conv3x3(Array2D(1000000x3))': 'conv3x3(1000000,3)',
+    'conv3x3(Array2D(1000x1000))': 'conv3x3(1000,1000)',
+    'dilate3x3(1000)': 'dilate3x3(1000,1000)',
+    'conv3x3(1000)': 'conv3x3(1000,1000)',
+    'conv3x3(3)': 'conv3x3(1000000,3)',
+}
+
+def main(name):
+    interp = None
+    res = {}
+    order = ['python2.7', 'pypy --jit enable_opts=intbounds:rewrite:virtualize:string:earlyforce:pure:heap:ffi', 'pypy', 'luajit -O-loop', 'luajit', 'gcc -O3 -march=native -fno-tree-vectorize']
+    labels = [None, 'PyPy no LP', 'PyPy', 'LuaJIT no LP', 'LuaJIT', None]
+    with open(name) as f:
+        for line in f:
+            line = line.strip("\n")
+            if not line:
+                interp = None
+            elif interp is None:
+                interp = line
+            else:
+                bench, rest = line.split(':')
+                bench = bench.replace(" ", "")
+                bench = NAME_REPL.get(bench, bench)
+                if '+-' in rest:
+                    a, d = rest.split('+-')
+                    res.setdefault(bench, {})[interp] = float(a), float(d)
+                else:
+                    res.setdefault(bench, {})[interp] = float(rest)
+    resmat = np.zeros((len(res), len(order)))
+    for i, key in enumerate(sorted(res.keys())):
+        sys.stdout.write(key)
+        for j, ord in enumerate(order):
+            try:
+                e = res[key][ord]
+            except KeyError:
+                sys.stdout.write(" & -")
+            else:
[…]

### 2012-08-17 talk/dls2012/paper.tex
regen paper
Write down Mike's comment here
More strongly put this XXX forward
improve the description of the sqrt(Fix16) benchmark

@@
 \hline
-FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.42 $\pm$ 0.017 & 2.76 $\pm$ 0.017 & 1.40 $\pm$ 0.082\\
-\hline
-FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.33 $\pm$ 0.028 & 1.08 $\pm$ 0.014 & 0.83 $\pm$ 0.044\\
-\hline
-LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.012 & 1.52 $\pm$ 0.014 & 1.33 $\pm$ 0.070\\
-\hline
-LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 3.99 $\pm$ 0.014 & 0.67 $\pm$ 0.010 & 0.65 $\pm$ 0.077\\
-\hline
-MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.016 & 2.82 $\pm$ 0.009 & 1.69 $\pm$ 0.096\\
-\hline
-SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.009 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
-\hline
-SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.009 & 1.08 $\pm$ 0.010 & 1.49 $\pm$ 0.042\\
-\hline
-SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.64 $\pm$ 0.032 & 4.53 $\pm$ 0.032 & 1.84 $\pm$ 0.061\\
-\hline
-SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.016 & 2.42 $\pm$ 0.010 & 1.20 $\pm$ 0.053\\
-\hline
-%\hline
-%conv3(1e5) & 50.14 & 1.09 $\pm$ 0.022 & 0.49 $\pm$ 0.028 & 0.67 $\pm$ 0.010 & 0.12 $\pm$ 0.010 & 0.52 $\pm$ 0.084\\
-\hline
-conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.008 & 0.18 $\pm$ 0.000 & 0.60 $\pm$ 0.064\\
-\hline
-conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.49 $\pm$ 0.010 & 0.09 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
-\hline
-%conv3x3(1000000,3) & 139.81 & 0.70 $\pm$ 0.005 & 0.21 $\pm$ 0.006 & 0.53 $\pm$ 0.008 & 0.13 $\pm$ 0.009 & 0.19 $\pm$ 0.061\\
-%\hline
-%conv5(1e5) & 74.65 & 1.22 $\pm$ 0.009 & 0.64 $\pm$ 0.005 & 0.84 $\pm$ 0.006 & 0.17 $\pm$ 0.010 & 0.55 $\pm$ 0.047\\
-%\hline
-%conv5(1e6) & 77.94 & 1.26 $\pm$ 0.009 & 0.68 $\pm$ 0.014 & 0.87 $\pm$ 0.010 & 0.21 $\pm$ 0.013 & 0.58 $\pm$ 0.049\\
-%\hline
-dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.48 $\pm$ 0.014 & 0.09 $\pm$ 0.006 & 0.17 $\pm$ 0.061\\
-\hline
-sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.60 $\pm$ 0.012 & 0.24 $\pm$ 0.018 & 0.17 $\pm$ 0.061\\
-\hline
-sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.013 & 0.83 $\pm$ 0.010 & 0.85 $\pm$ 0.088\\
+FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.45 $\pm$ 0.019 & 2.74 $\pm$ 0.021 & 1.40 $\pm$ 0.082\\
+\hline
+FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.25 $\pm$ 0.019 & 1.07 $\pm$ 0.050 & 0.83 $\pm$ 0.044\\
+\hline
+LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.018 & 1.52 $\pm$ 0.010 & 1.33 $\pm$ 0.070\\
+\hline
+LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 4.00 $\pm$ 0.018 & 0.67 $\pm$ 0.014 & 0.65 $\pm$ 0.077\\
+\hline
+MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.013 & 2.82 $\pm$ 0.010 & 1.69 $\pm$ 0.096\\
+\hline
+SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.011 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
+\hline
+SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.006 & 1.08 $\pm$ 0.014 & 1.49 $\pm$ 0.042\\
+\hline
+SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.69 $\pm$ 0.033 & 4.49 $\pm$ 0.036 & 1.84 $\pm$ 0.061\\
+\hline
+SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.019 & 2.43 $\pm$ 0.031 & 1.20 $\pm$ 0.053\\
+\hline
+\hline
+conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.009 & 0.18 $\pm$ 0.009 & 0.60 $\pm$ 0.064\\
+\hline
+conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.22 $\pm$ 0.009 & 0.15 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
+\hline
+dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.22 $\pm$ 0.008 & 0.16 $\pm$ 0.010 & 0.17 $\pm$ 0.061\\
+\hline
+sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.37 $\pm$ 0.014 & 0.24 $\pm$ 0.017 & 0.17 $\pm$ 0.061\\
+\hline
+sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.010 & 0.83 $\pm$ 0.014 & 0.85 $\pm$ 0.088\\
 \hline
@@
 \hline
-sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & - & - & 1.34 $\pm$ 0.061\\
+sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & 4.00 $\pm$ 0.040 & 1.47 $\pm$ 0.014 & 1.34 $\pm$ 0.061\\
 \hline
@@
 There are three different versions of this benchmark where $x_i$
-  is represented with different type of objects, $T$,: int's, float's and
+  is represented with different type $T$ of objects: int's, float's and
   Fix16's. The latter, Fix16, is a custom class that implements
-  fixpoint arithmetic with 16 bits precision. In Python there is only
+  fixpoint arithmetic with 16 bits precision. In Python and Lua there is only
   a single implementation of the benchmark that gets specialized
-  depending on the class of it's input argument, $y$, while in C,
-  there are three different implementations. In Lua there is no support for
-  integers so only the floating point number is provided.
+  depending on the class of it's input argument, $y$. In C,
+  there are three different implementations.
+
+The Fix16 type is a custom class with operator overloading in Lua and Python.
+The C version uses a C++ class. The goal of this variant of the benchmark is to
+check how large the overhead of a custom arithmetic class is, compared to
+builtin data types.
+
+In Lua there is no direct support for
+integers so the int version is not provided.
+
 \item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
@@
 
+The slowdown of sqrt(Fix16) compared to sqrt(int) or sqrt(float) show the
+overhead of using a custom class with operator overloading for arithmetic. For
+C/C++, this overhead is very low, for CPython the code becomes 30 times slower.
+In LuaJIT, the overhead is a slowdown of 70\%. For PyPy, sqrt(Fix16) is only
+slightly slower than sqrt(int), which is itself three times slower than
+sqrt(float). This is probably due to the additional overflow checking necessary
+for integer arithmetic in Python. The fact that LuaJIT and PyPy do so well on
+sqrt(Fix16) shows that the allocation removal/sinking optimizations work well
+in both JITs.
+
 \section{Related Work}

### 2012-08-17 talk/dls2012/paper.tex
use only eight characters for the hash
some improvements proposed by David

@@
 
-We have implemented the scheme in PyPy's tracing JIT compiler,
-where it can give performance improvements of a
-factor over two for PyPy's Python JIT executing simple numerical kernels
+We have implemented the scheme in RPython's tracing JIT compiler. PyPy's Python
+JIT executing simple numerical kernels can become up to two times faster,
 bringing the performance into the ballpark of static language compilers.
@@
 account, such as loop-invariant code
-motion or optimizations that improve across several iterations of the loop.
+motion or optimizations that improve several iterations of the loop.
 Having to deal with this property of traces complicates the optimization passes,
@@
     \item \lstinline{set} writes to an attribute of an object.
-    \item \lstinline{guard_class} is a precise type check. It typically precedes
-    an (inlined) method call and is followed by the trace of the called method.
-    The type that the guard checks for is the one that the variable had during
-    tracing.
+    \item \lstinline{guard_class} is a precise type check, not checking for subclasses.
 \end{itemize}
 
-Method calls in the trace are preceded by a \lstinline{guard_class}
+Inlined method calls in the trace are preceded by a \lstinline{guard_class}
 operation, to check that the class of the receiver is the same as the one that
-was observed during tracing.\footnote{\lstinline{guard_class}
-performs a precise
-class check, not checking for subclasses.} These guards make the trace specific
+was observed during tracing. These guards make the trace specific
 to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}. When
@@
 
-Before a trace is passed to the backend compiling it into machine code
+Before a trace is compiled to machine code by the backend,
 it is optimized to achieve better performance.
 One goal of that is to move 
-operations out of the loop making them executed only once
+operations out of the loop to execute them only once
 and not every iteration. This can be achieved by loop peeling. It
@@
 
-Loop peeling is achieved by appending an copy of the traced iteration at
+Loop peeling is achieved by appending a copy of the traced iteration at
 the end of itself. See Figure~\ref{fig:overview} for an illustration.
@@
 When peeling the loop, no assumptions are made that the preamble is
-the \emph{first} iteration when later executing the loop. The preamble stays
+the \emph{first} iteration, when later executing the loop. The preamble stays
 general enough to correspond to any iteration of the loop.
@@
 treated. It has to be ensured that the peeled loop stays a proper
-trace in the sense that the operations within it only operates on
+trace in the sense that the operations within it only operate on
 variables that are either among its input arguments 
@@
 
+Redundant guard removal removes guards that are implied by other guards earlier
+in the trace. The most common case is the removal of a guard that has already
+appeared.
 No special concern needs to be taken when implementing redundant
@@
 arguments, it only needs to be executed the first time and then the result
-can be reused for all other appearances. RPython's optimizers can also remove
+can be reused for all other appearances. This is achieved by common
+subexpression elimination. RPython's optimizers can also remove
 repeated heap reads if the intermediate operations cannot have changed their
@@
 
-It is interesting to note that the described approach automatically deals with
-implicit control dependencies correctly, whereas in other approaches this needs
+It is interesting to note that the described approach deals correctly with
+implicit control dependencies, whereas in other approaches this needs
 to be carefully programmed in. A commonly used example for a control dependency
@@
 division operation must not be moved before that guard, and indeed, this is
-never done. If the division is loop invariant, the result computed in copy of
+never done. If the division is loop invariant, the result computed by the copy of
 the division operation in the preamble is reused. This division operation is
-preceded by a copy of the non-null guard, which ensures that it can be executed
-correctly.
+preceded by a copy of the guard that checks that the second argument is not 0,
+which ensures that the division can be executed correctly.
+Such control dependencies are common in traces produced by dynamic languages.
+Reading a field out of an object is often preceded by checking the type of the
+object.
 
@@
 to be taken to always place the attributes in the same order when
-performing this explosion. Notation becomes somewhat simpler if also every
-concrete variable of the jump arguments is exploded into a vector containing
+performing this explosion. Notation becomes somewhat simpler if every
+concrete variable of the jump arguments is also exploded into a vector containing
 itself. For
@@
 as in Figure~\ref{fig:opt-trace}.
-The resulting optimized peeled loop consists of a single integer addition
-only. That is it will become type-specialized to the types of the
+The resulting optimized peeled loop consists of a single integer addition. That
+is it will become type-specialized to the types of the
 variables \lstinline{step} and \lstinline{y}, and the overhead of
@@
 \end{center}
-\label{fig:benchmarks}
 \caption{Benchmark results in seconds with 95\% confidence intervals. The leftmost column gives the
 name of each benchmark and the values of the benchmark parameters used. The different benchmarks and the meaning of their parameters are described in Section~\ref{sec:benchmarks}.}
+\label{fig:benchmarks}
 \end{figure*}
@@
 \includegraphics[width=0.5\textwidth]{benchmarks/result.pdf}
+\end{center}
+\caption{Benchmark results normalized to the runtime of the C version. The CPython results have been omitted to make the plot readable.}
 \label{fig:benchmarks_plot}
-\caption{Benchmark results normalized with the runtime of the C version. The CPython results have been omitted to make the plot readable.}
-\end{center}
 \end{figure}
@@
 calculations were implemented in both Python, C and Lua and their
-runtimes are compared in Figuare~\ref{fig:benchmarks_plot} and Figure~\ref{fig:benchmarks}.\footnote{
+runtimes are compared in Figure~\ref{fig:benchmarks_plot} and Figure~\ref{fig:benchmarks}.\footnote{
     The benchmarks and the scripts to run them can be found in the repository for this paper:
@@
 For benchmarks using larger Python applications the times are unaffected or
-slightly improved by the loop optimization of this paper.
+only slightly improved by the loop optimization of this paper.
 
@@
 hardcoded into the implementation making the benchmark consist of a single loop too.
-\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with kernel of fixed
+\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with a kernel of fixed
   size $3 \times 3$ using a custom class to represent two-dimensional
@@
  was used.
-\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed
+\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with a kernel of fixed
   size $3 \times 3$. This is similar to convolution but instead of
@@
 using Ubuntu Linux 11.4 in 64bit mode.
-The machine was otherwise unoccupied. We use the following software
+The machine was otherwise unoccupied. We used the following software
 for benchmarks:
@@
 \item GCC 4.5.2 shipped with Ubuntu 11.4
-\item LuaJIT 2.0 beta, git head of August 15, 2012, commit ID 0dd175d9e711f039c663d35e96c149b705bcf450
+\item LuaJIT 2.0 beta, git head of August 15, 2012, commit ID 0dd175d9
 \end{itemize}
 
-We run GCC with -O3 -march=native, disabling the
+We ran GCC with -O3 -march=native, disabling the
 automatic loop vectorization. In all cases, SSE2 instructions were used for
 floating point operations.
-We also run PyPy and LuaJIT with loop peeling optimization and without (but otherwise
+We also ran PyPy and LuaJIT with loop peeling optimization and without (but otherwise
 identical).
 
-For PyPy and LuaJIT 10 iterations were run, prefaced with 3 iterations for warming up.
+For PyPy and LuaJIT, 10 iterations were run, prefaced with 3 iterations for warming up.
 Due to benchmarks taking large amounts of time on CPython, only one run
 was performed.
-For GCC 5 iterations
+For GCC, 5 iterations
 were run. In all cases, the standard deviation is very low, making benchmarks
@@
 work~\cite{bolz_allocation_2011, bolz_runtime_2011}, the main improvement for
-these concrete benchmarks come from the allocation removal/unboxing
+these concrete benchmarks comes from the allocation removal/unboxing
 optimization.
@@
 
-Loop invariant code motion has been part of early compilers in the 1960s and
-1970s~\cite{allen_catalogue_1971}. A common approach for achieving loop invariant
+Loop invariant code motion has been part of early compilers since the
+1960s~\cite{allen_catalogue_1971}. A common approach for achieving loop
+invariant
 code motion is to perform partial redundancy elimination. The
@@
 and forward data flow analysis. Implementing partial redundancy elimination in
-compilers that use SSA form \cite{chow_new_1997} simplified the algorithms
-because no iterative data flow analysis is needed any more.
+compilers that use SSA form~\cite{chow_new_1997} simplified the algorithms,
+because no iterative data flow analysis was needed any more.
 
@@
 Both the Hotpath VM~\cite{gal_hotpathvm:_2006} and
-SPUR~\cite{bebenita_spur:_2010} implements loop-invariant code motion
+SPUR~\cite{bebenita_spur:_2010} implement loop-invariant code motion
 directly, by explicitly marking as loop-invariant all variables that stay the
@@
 
-The current approach still has some limitations which we plan to address in the
+The described approach still has some limitations which we plan to address in the
 future. In particular loop peeling works poorly in combination with trace
 trees~\cite{gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
-The side exits attached guards that fail often
-currently have to jump to the preamble which makes loops with several equally
-common paths less efficient than they could be.
+The side exits attached to guards that fail often
+currently have to jump to the preamble.
 
@@
 \acks
-We would like to thank Samuele Pedroni, Sven Hager and the anonymous reviewers
+We would like to thank Samuele Pedroni, Sven Hager, David Schneider, and the anonymous reviewers
 for helpful comments on drafts of this paper. We owe gratitude to Mike Pall

### 2012-08-17 talk/dls2012/paper.tex
PyPy's tracing JIT -> RPython's tracing JIT
don't insist on the pdf as a source
more correct explanation for the sqrt(Fix16) behaviour. Thanks, Mike.

@@
 optimizations, which greatly simplifies the implementation. We have implemented
-the same approach in PyPy's tracing JIT compiler, the results of which we
+the same approach in RPython's tracing JIT compiler, the results of which we
 present here.
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{benchmarks/result.pdf}
+\includegraphics[width=0.5\textwidth]{benchmarks/result}
 \end{center}
@@
 
-The slowdown of sqrt(Fix16) compared to sqrt(int) or sqrt(float) show the
+The performance of sqrt(Fix16) compared to the C version gives an indication of the
 overhead of using a custom class with operator overloading for arithmetic. For
-C/C++, this overhead is very low, for CPython the code becomes 30 times slower.
-In LuaJIT, the overhead is a slowdown of 70\%. For PyPy, sqrt(Fix16) is only
-slightly slower than sqrt(int), which is itself three times slower than
-sqrt(float). This is probably due to the additional overflow checking necessary
-for integer arithmetic in Python. The fact that LuaJIT and PyPy do so well on
+CPython the overhead over C is a lot larger than that of sqrt(int).
+In LuaJIT, the overhead is very small. For PyPy, sqrt(Fix16) 2.2 times slower
+than the C version. However, that is not actually due to the overhead of
+operator overloading but due to the additional overflow checking necessary
+for integer arithmetic in Python. The JIT does not manage to prove that the
+integer operations in these benchmarks cannot overflow and therefore cannot
+optimize away the overflow checking. This is also the reason why sqrt(float) is
+so much faster than sqrt(int) for PyPy.
+The fact that LuaJIT and PyPy do so well on
 sqrt(Fix16) shows that the allocation removal/sinking optimizations work well

### 2012-08-17 talk/dls2012/benchmarks/parse.py
be consistent about the order of the benchmarks

@@ -54,6 +54,9 @@ def main(name):
     resmat = np.zeros((len(res), len(order)))
     benchmarks = res.keys()
     benchmarks.sort()
+    assert benchmarks[-3] == "sqrt(Fix16)"
+    del benchmarks[-3]
+    benchmarks.append("sqrt(Fix16)")
     for i, key in enumerate(benchmarks):
         sys.stdout.write(key)
         for j, ord in enumerate(order):

### 2012-09-05 talk/vmil2012/paper.tex
Add a paragraph about the violation of immutability.
reviewers' comments

@@
 
+\newcommand{\bigtodo}[1]{\todo[inline,color=red!30]{#1}}
+
 \newboolean{showcomments}
@@
 
+\bigtodo{reviewer c: The use of an interpreter is a typical choice, but it is
+not the only choice (e.g., Jikes RVM is compile only) so it might make
+sense to introduce this idea earlier.}
+
 Tracing just-in-time (JIT) compilers record and compile commonly executed
@@
 and to describe the concrete techniques used in the various layers of RPython's
-tracing JIT. All design decisions are be motivated by an empirical analysis of the
+tracing JIT. All design decisions are motivated by an empirical analysis of the
 frequency and the overhead related to guards.
@@
 \begin{itemize}
-  \item An analysis guards in the context of RPython's tracing JIT,
+  \item An analysis of guards in the context of RPython's tracing JIT,
   \item detailed measurements about the frequency and the
@@
 several features such as automatic memory management
-and just-in-time compilation. When writing an interpreter using RPython the
+and just-in-time compilation.\todo{reviewer d: I argue that jit compilation is
+not a feature of a language, but a feature of a language
+\emph{implementation}.}
+When writing an interpreter using RPython the
 programmer only has to write the interpreter for the language she is
@@
 
-\subsection{RPython's Tracing JIT Compilers}
+\subsection{RPython's Tracing JIT Compiler}
 \label{sub:tracing}
+
+\bigtodo{reviewer c: In 2.2 there was a comment that guards can be used to
+specialize based on type in dynamic languages. Is this only applicable to
+dynamic languages, or could the same thing occur due to simple inheritance?}
+
 Tracing is a technique of just-in-time compilers that generate code by
@@
 
+\bigtodo{reviewer b: sadly no graphs}
+
+\bigtodo{reviewer b: One missing data set in the empirical study
+is excution time performance info for the 
+various guard implementation choices.
+The empirical data focuses on space not
+time.}
+
+\bigtodo{reviewer c: I found the evaluation less detailed than I was hoping
+for, which was a shame as I was very interested in learning how well everything
+performed after reading the earlier sections!}
+
+\bigtodo{reviewer c:
+I would have liked to see an evaluation based on execution time rather than
+operations. I do not have a good sense of the relative cost of different
+operations, and from my experience know that it can be misleading to use such a
+measure.}
+
+\bigtodo{reviewer c:
+For all of the results some aggregate values would be useful (max/min, etc.)
+}
+
 The results presented in this section are based on numbers gathered by running
@@
 
+\bigtodo{reviewer c: Figure 7 was printed out of order. It was not clear if
+    Figure 7 was talking about static measures (count once per compilation),
+    vs. dynamic measures (count once per actual execution). When I have done
+    studies like this in the past the difference between these two measures can
+    be very interesting. One has more of an impact on memory and compiler
+    issues, while the other will more significantly affect execution
+    performance.
+}
+
 \subsection{Frequency of Guards}
@@
 attached bridge.
+
+\bigtodo{reviewer c: It might be possible to use a chart for
+Figure~\ref{fig:failing_guards} to give more information?}
+
+\bigtodo{reviewer d:
+I would also like to see the number of guard that account for 99\% (or even
+99.9\%) of all guard failures, in addition to the 50\% number.}
 
@@
 
+\bigtodo{ reviewer c: I found it difficult to understand how the size of the
+resume structures are approximated, or what you mean in the footnote where you
+say it is reconstructed from log files. }
+
 For the different benchmarks the backend map has
@@
 the transfer code is quite large.
+
+\bigtodo{reviewer c: In 6.1 you claim that the elimination of snapshots is
+orthogonal to resume data compression. While they are techniques that can
+be used individually or together, I think there may be non-trivial
+interactions, in particular eliminating snapshots may alter opportunities
+for compression.}
 
@@
 
+\bigtodo{reviewer d: You state that Kotzmann et al. do not explain how the
+debug information is store in a compact way. They have a second paper that
+describes exactly that: Thomas Kotzmann, Hanspeter Mössenböck: Run-Time
+Support for Optimizations Based on Escape Analysis Proceedings of the
+International Symposium on Code Generation and Optimization, San Jose,
+March 2007.}
+
 % subsection Deoptimization in Method-Based JITs (end)
@@
 \bibliography{zotero,paper}
-%\listoftodos
+\listoftodos
 \end{document}

### 2012-09-05 talk/vmil2012/paper.tex
(cfbolz, bivab) address some of the comments by the reviewers

@@
 
-\bigtodo{reviewer c: The use of an interpreter is a typical choice, but it is
-not the only choice (e.g., Jikes RVM is compile only) so it might make
-sense to introduce this idea earlier.}
-
 Tracing just-in-time (JIT) compilers record and compile commonly executed
-linear control flow paths consisting of operations executed by an interpreter.
+linear control flow paths consisting of operations executed by an
+interpreter.\footnote{There are also virtual machines that have a tracing JIT
+compiler and do not use an interpreter~\cite{xxx}. This paper assumes that the
+baseline is provided by an interpreter. Similar design constraints would apply
+to a purely compiler-based system.}
 At points of possible divergence from the traced path operations called guards
@@
 statically typed object-oriented high-level subset of Python. The subset is chosen in such a way to make type inference possible\cite{ancona_rpython:_2007}.
-The language provides
+The language tool-set provides
 several features such as automatic memory management
-and just-in-time compilation.\todo{reviewer d: I argue that jit compilation is
-not a feature of a language, but a feature of a language
-\emph{implementation}.}
+and just-in-time compilation.
 When writing an interpreter using RPython the
@@
 
-\bigtodo{reviewer c: In 2.2 there was a comment that guards can be used to
-specialize based on type in dynamic languages. Is this only applicable to
-dynamic languages, or could the same thing occur due to simple inheritance?}
-
 Tracing is a technique of just-in-time compilers that generate code by
@@
 tracing phase are still valid when the code has been compiled and is executed.
-In the case of dynamic languages, guards are also used to encode type checks
+Guards are also used to encode type checks
 that come from optimistic type specialization by recording the types of
-variables seen during tracing\cite{Gal:2009ux}.
+variables seen during tracing\cite{Gal:2006, Gal:2009ux}.
 After a trace has been recorded it is optimized and then compiled to platform
@@
 
-\bigtodo{reviewer c: I found the evaluation less detailed than I was hoping
-for, which was a shame as I was very interested in learning how well everything
-performed after reading the earlier sections!}
-
 \bigtodo{reviewer c:
@@
 the total amount of code and resume data that is generated.
+%The measurements and the evaluation focus on trace properties and memory
+%consumption, and do not discuss the execution time of the benchmarks. These
+%topics were covered in earlier work~\cite{bolz_allocation_2011} and furthermore
+%are not influenced that much by the techniques described in this paper.
 
@@
 recorded during tracing for each of the benchmarks and what percentage of these
-operations are guards. The number of operations was counted on the unoptimized
+operations are guards. The static number of operations was counted on the unoptimized
 and optimized traces. The figure also shows the overall optimization rate for
@@
 
-\bigtodo{reviewer d:
-I would also like to see the number of guard that account for 99\% (or even
-99.9\%) of all guard failures, in addition to the 50\% number.}
-
-\begin{figure}
+\begin{figure*}
     \include{figures/failing_guards_table}
-    \caption{Failing guards, guards with more than 200 failures and guards responsible for 50\% of the failures relative to the total number of guards}
+    \caption{Failing guards, guards with more than 200 failures and guards
+    responsible for 50\%, 99\% and 99.9\% of the failures relative to the total number of guards}
     \label{fig:failing_guards}
-\end{figure}
+\end{figure*}
 
@@
 is the size of the compressed mapping from registers and stack to
-IR-level variables and finally the size of the resume data is an
-approximation of the size of the compressed high-level resume data as described
+IR-level variables and finally the size of the resume data is
+the size of the compressed high-level resume data as described
 in Section~\ref{sec:Resume Data}.\footnote{
-The size of the resume data is not measured at runtime, but reconstructed from
-log files.}
-
-\bigtodo{ reviewer c: I found it difficult to understand how the size of the
-resume structures are approximated, or what you mean in the footnote where you
-say it is reconstructed from log files. }
+Due to technical reasons the size of the resume data is hard to measure
+directly at runtime. Therefore the size given in the table is reconstructed
+from debugging information stored in log files produced by the JIT.}
 
@@
 the transfer code is quite large.
-
-\bigtodo{reviewer c: In 6.1 you claim that the elimination of snapshots is
-orthogonal to resume data compression. While they are techniques that can
-be used individually or together, I think there may be non-trivial
-interactions, in particular eliminating snapshots may alter opportunities
-for compression.}
 
@@
 redundancy.\footnote{This optimization is now implemented in LuaJIT, at the time of writing it has not been fully documented in the LuaJIT Wiki: \url{http://wiki.luajit.org/Optimizations\#1-D-Snapshot-Compression}}
-The approach of not creating snapshots at all for every guard is
-orthogonal to the resume data compression presented in this paper and could be
-reused within RPython to improve the memory usage further.
+It should be possible to combine the approaches of not creating snapshots at all
+for every guard and the resume data compression presented in this paper.
 
@@
 entering the new trace.
-
 % subsection Guards in Other Tracing JITs (end)

### 2012-09-06 talk/vmil2012/tool/build_tables.py
add spark line

@@ -26,6 +26,7 @@ def build_failing_guards_table(files, texfile, template):
 
     table = []
     head = ['Benchmark',
+            'Sparkline' + "~" * 8,
             'Failing',
             '> %d failures' % BRIDGE_THRESHOLD,
             '50\% of failures',
@@ -42,6 +43,7 @@ def build_failing_guards_table(files, texfile, template):
         num_99 = we_are_n_percent(info, 99)
         num_99_dot_9 = we_are_n_percent(info, 99.9)
         res = [bench.replace('_', '\\_'),
+                make_sparkline(info['results'], num_50 - 1, num_99 - 1, num_99_dot_9 - 1),
                 "%.1f\\%%" % (100 * total_failures/total),
                 "%.1f\\%%" % (100 * bridges/total),
                 "%d~~\\textasciitilde{}~~%.3f\\%%"  % (num_50, num_50 / total * 100),
@@ -67,6 +69,23 @@ def we_are_n_percent(info, n):
         if current_sum > total_failures * n/100.0:
             return (i + 1)
     return -1
+
+def make_sparkline(results, index_50, index_99, index_99_9):
+    results = results.values()
+    results.sort()
+    results.reverse()
+    running = 0
+    lines = ["\\begin{sparkline}{20}"]
+    lines.append("\\sparkdot %04f %04f blue" % (float(index_50) / (len(results) - 1), float(results[index_50]) / max(results)))
+    lines.append("\\sparkdot %04f %04f red" % (float(index_99) / (len(results) - 1), float(results[index_99]) / max(results)))
+    lines.append("\\sparkdot %04f %04f green" % (float(index_99_9) / (len(results) - 1), float(results[index_99_9]) / max(results)))
+    lines.append("\\spark")
+    for i, result in enumerate(results):
+        lines.append("%04f %04f" % (float(i) / (len(results) - 1), float(result) / max(results)))
+    lines.append("/")
+    lines.append("\\end{sparkline}")
+    return " ".join(lines)
+
 
 def build_resume_data_table(csvfiles, texfile, template):
     assert len(csvfiles) == 1

### 2012-09-06 talk/vmil2012/paper.tex
add spark line
this is "done"
this part is done too

@@
 \usepackage{ifthen}
+\usepackage{sparklines}
+\usepackage{booktabs}
 \usepackage{fancyvrb}
@@
 
-\bigtodo{reviewer b: sadly no graphs}
-
 \bigtodo{reviewer b: One missing data set in the empirical study
@@
 
-\bigtodo{reviewer c: Figure 7 was printed out of order. It was not clear if
-    Figure 7 was talking about static measures (count once per compilation),
-    vs. dynamic measures (count once per actual execution). When I have done
-    studies like this in the past the difference between these two measures can
-    be very interesting. One has more of an impact on memory and compiler
-    issues, while the other will more significantly affect execution
-    performance.
-}
+\bigtodo{reviewer c: Figure 7 was printed out of order.}
 
@@
 %\section*{Appendix}
+
 %\todo{remove this section and the figures}

### 2012-09-06 talk/vmil2012/paper.tex
done
cleanup
tweak colors
describe second kotzmann paper

@@
 \usepackage{color}
+\usepackage{colortbl}
 \usepackage{wrapfig}
@@
 \definecolor{commentgray}{rgb}{0.3,0.3,0.3}
+\definecolor{darkgray}{rgb}{0.5,0.5,0.5}
+\definecolor{lightgray}{rgb}{0.75,0.75,0.75}
 
@@
 
-\bigtodo{reviewer c: It might be possible to use a chart for
-Figure~\ref{fig:failing_guards} to give more information?}
-
 \begin{figure*}
@@
 to support correct behaviour
-when scalar replacement of fields is done for non-escaping objects~\cite{kotzmann_escape_2005}.
+when scalar replacement of fields is done for non-escaping objects~\cite{kotzmann_escape_2005, kotzmann_run-time_2007}.
 The approach is extremely similar to how RPython's JIT handles virtual objects.
@@
 described by the deoptimization information.
-The paper does not describe any attempts to store this information compactly.
-This may not be needed in their approach, because method-based JITs have
-fewer deoptimization points than tracing JITs.
-
-\bigtodo{reviewer d: You state that Kotzmann et al. do not explain how the
-debug information is store in a compact way. They have a second paper that
-describes exactly that: Thomas Kotzmann, Hanspeter Mössenböck: Run-Time
-Support for Optimizations Based on Escape Analysis Proceedings of the
-International Symposium on Code Generation and Optimization, San Jose,
-March 2007.}
+The data structures for the deoptimization information are very similar to
+those used by RPython's tracing JIT. For every compiled Java method there is a
+\emph{scope entry} for the stack and one for the local variables. The objects
+that are replaced by scalars are described by \emph{object entries}, which are
+equivalent to RPython's virtual objects.
+
+The papers does not describe any attempts to share the object entries and scope
+entries between different deoptimization safe points. This seems to not be
+needed in a method-based JIT compiler, because method-based JITs have fewer
+deoptimization points than tracing JITs. Indeed, in the evaluation presented in
+the second paper~\cite{kotzmann_run-time_2007} the number of safe points is low
+for the benchmarks presented there, between 167 and 1512. The size of the
+debugging information in the presented benchmarks is at most about half the
+size of the machine code generated.
+
+\bigtodo{for comparison we should add the total number of guards per benchmark to one of the tables}
 

### 2012-09-07 talk/vmil2012/paper.tex
another todo
a footnote

@@
 the second paper~\cite{kotzmann_run-time_2007} the number of safe points is low
-for the benchmarks presented there, between 167 and 1512. The size of the
+for the benchmarks presented there, between 167 and 1512.\footnote{The fact
+that the density of safe points is low also means that the sharing approaches
+of this chapter likely would not work well.} The size of the
 debugging information in the presented benchmarks is at most about half the
@@
 
-\bigtodo{for comparison we should add the total number of guards per benchmark to one of the tables}
+\bigtodo{for comparison we should add the total number of guards per benchmark to one of the tables, and maybe also the number of traces}
 

### 2012-09-07 talk/vmil2012/paper.tex
de-stress execution speed issues

@@
 design and implement them carefully to find the right trade-off between
-execution speed, deoptimization,
-and memory overhead. In this paper, we perform an empirical analysis of runtime
+deoptimization, memory overhead, and (partly) execution speed.
+In this paper, we perform an empirical analysis of runtime
 properties of guards. This is used to guide the design of guards in the RPython
@@
 Besides being common, guards have various costs associated with them.
-Guards have a runtime cost, they take time to execute. Therefore it is
-important to make the on-trace execution of guards as efficient as possible. On
-the other hand, guards are possible deoptimization points. The recorded and
+Guards are possible deoptimization points. The recorded and
 compiled path has to be left if a guard fails, returning control to the
@@
 rebuilding the interpreter state. The memory overhead of this information
-should be kept low. These constraints and trade-offs are what makes the design
+should be kept low. On the other hand,
+Guards have a runtime cost, they take time to execute. Therefore it is
+important to make the on-trace execution of guards as efficient as possible.
+These constraints and trade-offs are what makes the design
 and optimization of guards an important and non-trivial aspect of the construction
@@
   \item detailed measurements about the frequency and the
-  overhead associated with guards, and
+  memory overhead associated with guards, and
   \item a description about how guards are implemented in the high\-

### 2012-09-07 talk/vmil2012/paper.tex
comment in again
mention reviewers

@@
 the total amount of code and resume data that is generated.
-%The measurements and the evaluation focus on trace properties and memory
-%consumption, and do not discuss the execution time of the benchmarks. These
-%topics were covered in earlier work~\cite{bolz_allocation_2011} and furthermore
-%are not influenced that much by the techniques described in this paper.
+The measurements and the evaluation focus on trace properties and memory
+consumption, and do not discuss the execution time of the benchmarks. These
+topics were covered in earlier work~\cite{bolz_allocation_2011} and furthermore
+are not influenced that much by the techniques described in this paper.
 
@@
 \section*{Acknowledgements}
-We would like to thank David Edelsohn, Samuele Pedroni, Stephan Zalewski and Sven Hager for their helpful
+We would like to thank David Edelsohn, Samuele Pedroni, Stephan Zalewski, Sven
+Hager, and the anonymous reviewers for their helpful
 feedback and valuable comments while writing this paper.

### 2012-09-07 talk/vmil2012/paper.tex
tweak colors

@@
 \definecolor{commentgray}{rgb}{0.3,0.3,0.3}
-\definecolor{darkgray}{rgb}{0.5,0.5,0.5}
+\definecolor{darkgray}{rgb}{0.6,0.6,0.6}
 \definecolor{lightgray}{rgb}{0.75,0.75,0.75}
@@
 
-\bigtodo{reviewer b: One missing data set in the empirical study
-is excution time performance info for the
-various guard implementation choices.
-The empirical data focuses on space not
-time.}
-
-\bigtodo{reviewer c:
-I would have liked to see an evaluation based on execution time rather than
-operations. I do not have a good sense of the relative cost of different
-operations, and from my experience know that it can be misleading to use such a
-measure.}
-
-\bigtodo{reviewer c:
-For all of the results some aggregate values would be useful (max/min, etc.)
-}
-
 The results presented in this section are based on numbers gathered by running
@@
 
-\bigtodo{for comparison we should add the total number of guards per benchmark to one of the tables, and maybe also the number of traces}
-
 % subsection Deoptimization in Method-Based JITs (end)

### 2012-09-07 talk/vmil2012/paper.tex
these are done
fix reference

@@
 \usepackage[colorinlistoftodos]{todonotes}
+\usepackage{flafter}
 
@@
 interpreter.\footnote{There are also virtual machines that have a tracing JIT
-compiler and do not use an interpreter~\cite{xxx}. This paper assumes that the
+compiler and do not use an interpreter~\cite{bebenita_trace-based_2010}. This paper assumes that the
 baseline is provided by an interpreter. Similar design constraints would apply
@@
 
-\bigtodo{reviewer c: Figure 7 was printed out of order.}
-
 \subsection{Frequency of Guards}
@@
     \include{figures/benchmarks_table}
-    \caption{Number of operations in the recorded traces and the relative amount of guards before and after optimizations}
+    \caption{Number of operations and guards in the recorded traces before and after optimizations}
     \label{fig:benchmarks}

### 2012-09-07 talk/vmil2012/paper.tex
Use package flafter for float positioning
update caption
save a line
remove todo
zitemize environment, in case we need it

@@
 \usepackage{flafter}
+\usepackage{enumitem}
+\setitemize{noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt}
 
@@
 \usepackage[scaled=0.81]{beramono}
+
+\newenvironment{zitemize}% zero - line spacing itemize environment
+   {\begin{list}{--}{
+   \setlength{\itemsep}{0 pt}
+   \setlength{\parsep}{0 pt}
+   \setlength{\topsep} {0 pt} }}% the end stuff
+   {\end{list}}
 
@@
 \begin{itemize}
-  \item An analysis of guards in the context of RPython's tracing JIT,
+  \item An analysis of guards in the context of RPython's JIT,
   \item detailed measurements about the frequency and the
@@
 
-RPython can mean one of two things:
-\begin{itemize}
- \item the language itself
- \item the translation toolchain used to transform RPython programs to executable units
-\end{itemize}
+RPython can mean one of two things, the language itself and the translation
+toolchain used to transform RPython programs to executable units.
 The RPython language is a
@@
 
-\begin{figure}
+\begin{figure}[ht]
     \input{figures/example.tex}
@@
 
-\begin{figure}
+\begin{figure}[ht]
     \input{figures/unopt-log.tex}
@@
 The remaining 14 bits are a payload that depends on the tag bits.
-
 The possible sources of information are:
@@
 
-\begin{figure}
+\begin{figure}[ht]
     \input{figures/log.tex}
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics[width=0.4\textwidth]{figures/resume_data.pdf}

### 2012-09-07 talk/vmil2012/paper.tex
Trying to save some space
safe some lines

@@
 \includegraphics[width=0.4\textwidth]{figures/resume_data.pdf}
+\vspace{-3mm}
 \caption{The resume data for Figure~\ref{fig:trace-log}}
@@
 of the unoptimized code,
-the transfer code is quite large.
+the transfer code is large.
 
@@
 
-Gal et. al~\cite{Gal:2006} describe how in the HotpathVM, a JIT for a Java VM, they experimented
+Gal et. al~\cite{Gal:2006} describe the HotpathVM, a JIT for a Java VM. They experimented
 with having one generic compensation code block, like the RPython JIT, that
@@
 Secondary traces are compiled by first restoring the register allocator state to
-the state at the side exit. The information is retrieved from a mapping stored
-in the guard that maps machine level registers and stack to Java level stack
+the state at the side exit. The information is a mapping stored
+in the guard between machine level registers and stack to Java level stack
 and variables.
@@
 Deoptimization in method-based JITs is used if one of the assumptions
-of the code generated by a JIT-compiler changes.
+of the code generated by a JIT changes.
 This is often the case when new code is added to the system,
@@
 
-Deutsch et. al.~\cite{deutsch_efficient_1984} describe the use of stack descriptions
+Deutsch et. al.~\cite{deutsch_efficient_1984} use stack descriptions
 to make it possible to do source-level debugging of JIT-compiled code.
@@
 the size of the generated machine code.
-All the information needed for debugging together
+The information needed for debugging together
 is between 1.22 and 2.33 times the size of generated machine code,
@@
 
-Additionally we have presented experimental data collected using the standard PyPy
+Additionally we presented experimental data collected using the standard PyPy
 benchmark set to evaluate previous observations and assumptions about guards. Our

### 2012-09-07 talk/vmil2012/paper.tex
chapter -> paper

@@
 that the density of safe points is low also means that the sharing approaches
-of this chapter likely would not work well.} The size of the
+of this paper likely would not work well.} The size of the
 debugging information in the presented benchmarks is at most about half the

### 2012-09-07 talk/vmil2012/paper.tex
Add some notes about the changes to the tables

@@
 
-Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
+Figure~\ref{fig:benchmarks} summarizes\footnote{In all tables the minimum and
+maximum values for each column are highlighted in dark/light gray.} the total number of operations that were
 recorded during tracing for each of the benchmarks and what percentage of these
@@
 }
-
+It also contains sparklines depicting the failure rates for the failing guards
+in decreasing order, each normalized to the most failing guard.
 The numbers presented for guards that have a bridge represent the
@@
 benchmark, between 0.008\% and 0.225\% of the guards are responsible for 50\%
-of the total guards failures.
+of the total guards failures. Even considering 99.9\% of guard failures the
+relative amount of guards does not rise above 3\%.
+The colored dots in the sparklines correspond to 50\%, 99\% and 99.9\%.
 These results emphasize that as most of the guards never

### 2012-09-09 talk/vmil2012/paper.tex
don't use the footnote, they confuse google scholar and we have only one affiliation anyway

@@
 
-\authorinfo{David Schneider$^{a}$ \and Carl Friedrich Bolz$^a$}
-           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
+\authorinfo{David Schneider \and Carl Friedrich Bolz}
+           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
            }

### 2012-10-09 talk/vmil2012/presentation/talk.tex
start the talk

--- 2011-01-19 talk/pepm2011/presentation/talk.tex
@@
 
-\title{Allocation Removal by Partial Evaluation in a Tracing JIT}
+\title{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}
 
-\author[Carl Friedrich Bolz et.
-al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{1} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and Samuele Pedroni\inst{3} \and Armin Rigo\inst{1}}
+\author[David Schneider, Carl Friedrich Bolz]{David Schneider \and \emph{Carl Friedrich Bolz}}
 % - Give the names in the same order as the appear in the paper.
@@
 
-\institute[Heinrich-Heine-Universität Düsseldorf]
-{$^1$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany \and
-
- $^2$merlinux GmbH, Hildesheim, Germany \and
-
- $^3$Open End, Göteborg, Sweden \and
+\institute[Heinrich-Heine-Universität Düsseldorf]{
+Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
 }
 
-\date{2011 Workshop on Partial Evaluation and Program Manipulation, January 24, 2011}
+\date{2012 VMIL, XXX}
 % - Either use conference name or its abbreviation.
@@
 
+% XXX todos:
+% note that two fields is a simplification
+% have a diagram for the example trace
+% lifting can never produce new operations
+% have some sort of overview or position indicator for the many graphical slides, too easy to get lost
+% more details about benchmarks, diagram?
+
 % Structuring a talk is a difficult task and the following structure
@@
 \begin{frame}
-  \frametitle{Dynamic Languages are Slow}
-  \begin{itemize}
-      \item Interpretation overhead
-      \item Type dispatching
-      \item Boxing of primitive types
-      \pause
-      \begin{itemize}
-          \item A lot of allocation of short-lived objects
-      \end{itemize}
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Dynamic Languages are Slow: Example}
-  \texttt{x = a + b; y = x + c}
-
-  evaluated in an interpreter:
-  \pause
-  \begin{enumerate}
-      \item What's the type of a? \texttt{Integer}
-      \item What's the type of b? \texttt{Integer}
-  \pause
-      \item unbox a
-      \item unbox b
-      \item compute the sum
-  \pause
-      \item box the result as an \texttt{Integer}
-      \item store into x
-  \pause
-      \item What's the type of x? \texttt{Integer}
-      \item What's the type of c? \texttt{Integer}
-  \pause
-      \item unbox x
-      \item unbox c
-      \item compute the sum
-  \pause
-      \item box the result as an \texttt{Integer}
-      \item store into y
-  \end{enumerate}
-\end{frame}
-
-\begin{frame}
-  \frametitle{What to do?}
-  \begin{itemize}
-      \item Use a JIT compiler
-      \item \textbf{Add an optimization that can deal with heap operations}
-      \pause
-      \item optimize short-lived objects
-      \item remove some of the redundancy
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Overview}
-  \tableofcontents
-  % You might wish to add the option [pausesections]
-\end{frame}
-
-\section{Experimental Context}
-
-\begin{frame}
-  \frametitle{Our Experimental Context is the PyPy Project}
-  A general environment for implementing dynamic languages
-  \pause
-  \begin{block}{Approach}
-      \begin{itemize}
-          \item write an interpreter for the language in RPython
-          \item compilable to an efficient C-based VM
-          \pause
-          \item (RPython is a restricted subset of Python)
-      \end{itemize}
-  \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{PyPy Uses a Novel Tracing JIT}
-  the feature that makes PyPy interesting:
-  \begin{itemize}
-      \item a meta-JIT, applicable to many languages
-      \item needs a few source-code hints (or user annotations) in the interpreter
-      \item JIT is a tracing JIT compiler
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
   \frametitle{Tracing JITs Compile by Observing an Interpreter}
@@
 
-\begin{frame}
-  \frametitle{The Advantages of Tracing JITs}
-  \begin{itemize}
-      \item Traces are interesting linear pieces of code
-      \item most of the time correspond to loops
-      \item everything called in the trace is inlined
-      \item can perform good optimizations on the trace
-      \item rarer paths run by the interpreter
-  \end{itemize}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-  \frametitle{Example Trace}
-  Trace of \texttt{x = a + b; y = x + c}:
-\begin{verbatim}
-guard_class(a, Integer)
-guard_class(b, Integer)
-i1 = get(a, intval)
-i2 = get(b, intval)
-i3 = int_add(i1, i2)
-x = new(Integer)
-set(x, intval, i3)
-\end{verbatim}
-\pause
-\begin{verbatim}
-guard_class(x, Integer)
-guard_class(c, Integer)
-i4 = get(x, intval)
-i5 = get(c, intval)
-i6 = int_add(i4, i5)
-y = new(Integer)
-set(y, intval, i6)
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-  \frametitle{Example Trace}
-  Trace of \texttt{x = a + b; y = x + c}:
-\begin{verbatim}
-guard_class(a, Integer)
-guard_class(b, Integer)
-i1 = get(a, intval)
-i2 = get(b, intval)
-i3 = int_add(i1, i2)
-x = new(Integer)
-set(x, intval, i3)
-\end{verbatim}
-\begin{verbatim}
-guard_class(x, Integer)
-guard_class(c, Integer)
-i4 = get(x, intval)
-i5 = get(c, intval)
-i6 = int_add(i4, i5)
-y = new(Integer)
-set(y, intval, i6)
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-  \frametitle{Optimized Example Trace}
-  Trace of \texttt{x = a + b; y = x + c}:
-\begin{alltt}
-guard_class(a, Integer)
-guard_class(b, Integer)
-i1 = get(a, intval)
-i2 = get(b, intval)
-i3 = int_add(i1, i2)
-\redsout{x = new(Integer)}
-\redsout{set(x, intval, i3)}
-\end{alltt}
-\begin{alltt}
-\redsout{guard_class(x, Integer)}
-guard_class(c, Integer)
-\redsout{i4 = get(x, intval)}
-i5 = get(c, intval)
-i6 = int_add(\emph{i3}, i5)
-y = new(Integer)
-set(y, intval, i6)
-\end{alltt}
-\end{frame}
-
-\section{Proposed Optimization}
-
-\begin{frame}
-  \frametitle{Contribution of our Paper}
-  \begin{itemize}
-      \item a simple, efficient and effective optimization of heap operations in a trace
-      \item using online partial evaluation
-      \item fully implemented and in use in large-scale interpreters
-  \end{itemize}
-  \pause
-  \begin{block}{Ingredients}
-      \begin{itemize}
-          \item a slightly simplified model for objects on the heap
-          \item operational semantics of trace operations that manipulate the heap
-          \item optimization rules for those operations
-      \end{itemize}
-  \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Heap Model}
-  \includegraphics[scale=0.9]{figures/heap01}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Heap Model}
-  \includegraphics[scale=0.9]{figures/heap02}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Heap Model}
-  \includegraphics[scale=0.9]{figures/heap03}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Operations Manipulating Heap Objects}
-  \begin{itemize}
-      \item \texttt{v = new(T)} makes a new object
-      \item \texttt{u = get(w, F)} reads a field out of an object
-      \item \texttt{set(v, F, w)} writes a field of an object
-      \item \texttt{guard(v, T)} checks the type of an object
-  \end{itemize}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: New}
-  \includegraphics[scale=0.8]{figures/new01}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: New}
-  \includegraphics[scale=0.8]{figures/new02}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: Get}
-  \includegraphics[scale=0.8]{figures/get01}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: Get}
-  \includegraphics[scale=0.8]{figures/get02}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: Set}
-  \includegraphics[scale=0.8]{figures/set01}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: Set}
-  \includegraphics[scale=0.8]{figures/set02}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: Guard}
-  \includegraphics[scale=0.8]{figures/guard01}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: Guard}
-  \includegraphics[scale=0.8]{figures/guard02}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: Guard}
-  \includegraphics[scale=0.8]{figures/guard03}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Operations: Guard}
-  \includegraphics[scale=0.8]{figures/guard04}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Optimization by Online Partial Evaluation}
-  \begin{itemize}
-      \item Trace is optimized using online partial evaluation
-      \item part of the runtime heap is modelled in the \emph{static heap}
-      \item static heap contains objects that are allocated within the trace
-      \item (as opposed to before the trace is executed)
-      \pause
-      \item operations acting on the static heap can be removed
-      \item all others need to be residualized
-      \item all fairly straightforward
-  \end{itemize}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing New}
-  \includegraphics[scale=0.8]{figures/opt_new1}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing New}
-  \includegraphics[scale=0.8]{figures/opt_new2}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Get}
-  \includegraphics[scale=0.8]{figures/opt_get1}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Get}
-  \includegraphics[scale=0.8]{figures/opt_get2}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Get}
-  \includegraphics[scale=0.8]{figures/opt_get3}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Get}
-  \includegraphics[scale=0.8]{figures/opt_get4}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Guard}
-  \includegraphics[scale=0.8]{figures/opt_guard1}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Guard}
-  \includegraphics[scale=0.8]{figures/opt_guard2}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Guard}
-  \includegraphics[scale=0.8]{figures/opt_guard3}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Guard}
-  \includegraphics[scale=0.8]{figures/opt_guard4}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Set}
-  \includegraphics[scale=0.8]{figures/opt_set1}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Set}
-  \includegraphics[scale=0.8]{figures/opt_set2}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Lifting}
-  Problem: What happens if we write a static object into a dynamic one?
-  \pause
-  \begin{itemize}
-      \item lose track of the static object because of possibility of aliasing
-      \item need to \emph{lift} the static object
-      \item lifting produces operations that recreate the static object
-      \pause
-      \item needs to be careful due to recursive structures
-  \end{itemize}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Set}
-  \includegraphics[scale=0.8]{figures/opt_set_dynamic1}
-\end{frame}
-
-\begin{frame}[plain]
-  \frametitle{Optimizing Set}
-  \includegraphics[scale=0.8]{figures/opt_set_dynamic2}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-  \frametitle{Optimized Example Trace}
-  Trace of \texttt{x = a + b; y = x + c}:
-\begin{alltt}
-guard_class(a, Integer)
-guard_class(b, Integer)
-i1 = get(a, intval)
-i2 = get(b, intval)
-i3 = int_add(i1, i2)
-\sout{x = new(Integer)}
-\sout{set(x, intval, i3)}
-\sout{guard_class(x, Integer)}
-guard_class(c, Integer)
-\sout{i4 = get(x, intval)}
-i5 = get(c, intval)
-i6 = int_add(\emph{i3}, i5)
-y = new(Integer)
-set(y, intval, i6)
-\end{alltt}
-\end{frame}
-
-\section{Benchmarks}
-
-\begin{frame}
-  \frametitle{Benchmark Results}
-  \begin{itemize}
-      \item to evaluate the optimization we used PyPy's Python interpreter with real-world programs
-      \item optimization can remove
-      \begin{itemize}
-          \item 70\% of all \texttt{new} operations
-          \item 14\% of all \texttt{get/set} operations
-          \item 93\% of all \texttt{guard} operations
-      \pause
-      \item Timings improve by a factor between 1.1 and 6.95
-      \item outperforming standard Python on all benchmarks but one
-      \pause
-      \item more details in the paper
-      \end{itemize}
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Conclusion}
-  \begin{itemize}
-      \item We propose a very simple partial-evaluation-based optimization for tracing
-      JITs of dynamic languages that:
-      \begin{itemize}
-          \item can remove a lot of allocations and type checks in practical
-          programs. 
-          \item is efficient and effective.
-          \item has no control issues because all control decisions are made by
-          the tracing JIT.
-      \end{itemize}
-      \pause
-      \item We claim that this is a general strategy to get rid of control
-      problems by simply observing the runtime behaviour of the program.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Backup Slides}
-\end{frame}
-
-\begin{frame}
-  \frametitle{What About Correctness?}
-  \begin{itemize}
-      \item We haven't proven correctness yet
-      \item should not be too hard
-      \item lifting needs to be carefully handled
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Comparison to Escape Analysis}
-  \begin{itemize}
-      \item Effect very similar to escape analysis
-      \item Escape analysis needs a complex upfront analysis
-      \item our optimization automatically has a lot of context, due to the inlining tracing does
-      \item our optimization can optimize operations on objects even if they escape later
-  \end{itemize}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-  \frametitle{Comparison to "Dynamic Typing"/Boxing Analysis}
-  \begin{itemize}
-      \item those optimizations work ahead of time
-      \item don't work for many dynamic languages, where the source simply does not contain enough information
-  \end{itemize}
-  \begin{block}{Python Example:}
-  \begin{verbatim}
-def sum(container, initial):
-    result = initial
-    for element in container:
-        result = result + element
-    return result
-  \end{verbatim}
-  \end{block}
-\end{frame}
-
 \end{document}

### 2012-10-11 talk/vmil2012/presentation/talk.tex
a rough structure of the talk
fill the slides with a first rough draft

@@
 
-\date{2012 VMIL, XXX}
+\date{2012 VMIL, 21st of October, 2012}
 % - Either use conference name or its abbreviation.
@@
 
-% XXX todos:
-% note that two fields is a simplification
-% have a diagram for the example trace
-% lifting can never produce new operations
-% have some sort of overview or position indicator for the many graphical slides, too easy to get lost
-% more details about benchmarks, diagram?
-
-% Structuring a talk is a difficult task and the following structure
-% may not be suitable. Here are some rules that apply for this
-% solution:
-
-% - Exactly two or three sections (other than the summary).
-% - At *most* three subsections per section.
-% - Talk about 30s to 2min per frame. So there should be between about
-%   15 and 30 frames, all told.
-
-% - A conference audience is likely to know very little of what you
-%   are going to talk about. So *simplify*!
-% - In a 20min talk, getting the main ideas across is hard
-%   enough. Leave out details, even if it means being less precise than
-%   you think necessary.
-% - If you omit details that are vital to the proof/implementation,
-%   just say so once. Everybody will be happy with that.
+\section{Introduction}
 
@@
           \item produces a linear list of operations (trace)
+          \item automatically does (potentially deep) inlining
       \end{itemize}
@@
 
+\begin{frame}
+  \frametitle{Guards}
+  \begin{itemize}
+      \item Points of control flow divergence are marked with guards
+      \item Operations that check whether conditions are still true
+      \item When a guard fails, execution of the trace stops and continues in the interpreter
+      \pause
+      \begin{block}{Guard Characteristics}
+          \begin{itemize}
+              \item lots of them, up to 20\% guards
+              \item most never fail
+              \item costly to implement
+          \end{itemize}
+      \end{block}
+  \end{itemize}
+\end{frame}
+
+% this talk wants to go over a lot of details that are usually glossed over as
+% "easy" when tracing JITs are introduced.
+
+\begin{frame}
+  \frametitle{Bridges}
+  \begin{itemize}
+      \item When a trace fails often, it becomes worth to attach a new trace to it
+          \item This is called a bridge
+          \item The bridge is attached by patching the guard machine code
+          \item when this guard fails in the future, the new trace is executed instead
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{RPython and PyPy}
+  \begin{itemize}
+      \item Context: RPython
+      \item a generic tracing JIT, applicable to many languages
+      \item main use: PyPy, an efficient Python interpreter
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Running Example}
+\end{frame}
+
+\section{High-Level}
+
+\begin{frame}
+  \frametitle{Symbolic Frame Capturing}
+  \begin{itemize}
+      \item Guard can fail deep inside inlined function
+      \item when going back to the interpreter, call stack needs to be re-created
+      \item done with the help of symbolic frame stacks
+      \item these show how trace trace variables fill the to-be-built stack frames
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Symbolic Frame Compression}
+  \begin{itemize}
+      \item There are \emph{a lot of} guards
+      \item Naively storing symbolic frames would be costly in terms of memory
+      \item need to store them compactly
+      \item observation: from one guard to the next, the non-top stack frames don't change
+      \item share these between subsequent guards
+      \pause
+      \item also need a byte-saving binary representation, but that's just boring work
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Interaction with Optimization}
+  \begin{itemize}
+      \item Some optimizations make it necessary to store extra information in symbolic frames
+      \item examples:
+          \begin{itemize}
+              \item delayed heap stores (need to do stores before resuming interpreter)
+              \item allocation remvoal (need to allocate objects before resuming)
+          \end{itemize}
+      \item can be compressed using similar techniques
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Emitting Guards}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Patching Guards for Bridges}
+\end{frame}
+
+\section{Evaluation}
+
+%as in paper
+%fancy graphs
+%something about execution speed
+% <demo, if there is time>
 \end{document}

### 2012-10-12 talk/vmil2012/presentation/talk.tex
typo
add figures for guard compilation
some low-level notes

@@
       \item done with the help of symbolic frame stacks
-      \item these show how trace trace variables fill the to-be-built stack frames
+      \item these show how trace variables fill the to-be-built stack frames
   \end{itemize}
@@
               \item delayed heap stores (need to do stores before resuming interpreter)
-              \item allocation remvoal (need to allocate objects before resuming)
+              \item allocation removal (need to allocate objects before resuming)
           \end{itemize}
@@
   \frametitle{Emitting Guards}
+  Guards are compiled as
+  \begin{itemize}
+    \item Quick Check if the condition holds
+    \item and a mapping of machine locations to JIT-variables % indirection using the fail-boxes
+  \end{itemize}
+  In case of failure
+  \begin{itemize}
+    \item execution jumps to compensation code, decodes and stores mapping
+    \item returns to interpreter that rebuilds state
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \begin{figure}
+  \centering
+  \includegraphics[width=1\textwidth]{figures/loop.pdf}
+  \label{fig:trampoline}
+  \end{figure}
+\end{frame}
+
+\begin{frame}
+  \begin{figure}
+  \centering
+  \includegraphics[width=1\textwidth]{figures/bridge_compiled.pdf}
+  \label{fig:trampoline}
+  \end{figure}
+\end{frame}
+\begin{frame}
+  \begin{figure}
+  \centering
+  \includegraphics[width=1\textwidth]{figures/bridge_patched.pdf}
+  \label{fig:trampoline}
+  \end{figure}
 \end{frame}

### 2012-10-16 talk/vmil2012/presentation/talk.tex
some tweaks to the text
add charts for guard failure rates combined for all benchmarks and for go

@@
 \usepackage{alltt}
+\usepackage{tikz}
+\usepackage{pgfplots}
 
@@
 
-\section{Introduction}
+%\section{Introduction}
 
@@
       \pause
+      \item \emph{This talk:} describe technology and design decisions around guards
+      \pause
       \begin{block}{Guard Characteristics}
@@
               \item most never fail
-              \item costly to implement
+              \item need big information attached to them
           \end{itemize}
@@
 
-\section{High-Level}
+%\section{High-Level}
 
@@
           \begin{itemize}
+              \item allocation removal (need to allocate objects before resuming)
               \item delayed heap stores (need to do stores before resuming interpreter)
-              \item allocation removal (need to allocate objects before resuming)
           \end{itemize}
@@
   \begin{itemize}
-    \item execution jumps to compensation code, decodes and stores mapping
+    \item execution jumps to shared compensation code, decodes and stores mapping
     \item returns to interpreter that rebuilds state
@@
   \includegraphics[width=1\textwidth]{figures/loop.pdf}
-  \label{fig:trampoline}
   \end{figure}
@@
   \includegraphics[width=1\textwidth]{figures/bridge_compiled.pdf}
-  \label{fig:trampoline}
   \end{figure}
@@
   \includegraphics[width=1\textwidth]{figures/bridge_patched.pdf}
-  \label{fig:trampoline}
   \end{figure}
@@
 \end{frame}
+\begin{frame}[t,fragile]
+\pgfplotsset{tick label style={font=\tiny\bfseries},
+label style={font=\small},
+legend style={font=\tiny}
+}
+    \frametitle{Guard Failure Rates}
+    \begin{figure}
+        \centering
+        \begin{tikzpicture}
+            \begin{axis}[
+            xlabel= Guards by failures,
+            ylabel=Relative \# of failures,
+            xtick=\empty,
+            ytick=\empty,
+            ]
+                \input{figures/data}
+            \end{axis}
+        \end{tikzpicture}
+    \end{figure}
+\end{frame}
+\begin{frame}[t,fragile]
+\pgfplotsset{tick label style={font=\tiny\bfseries},
+label style={font=\small},
+legend style={font=\tiny}
+}
+    \frametitle{Guard Failure Rates / Go Benchmark}
+    \begin{figure}
+        \centering
+        \begin{tikzpicture}
+            \begin{axis}[
+            xlabel= Guards by failures,
+            ylabel=Relative \# of failures,
+            xtick=\empty,
+            ytick=\empty,
+            ]
+                \input{figures/go_data}
+            \end{axis}
+        \end{tikzpicture}
+    \end{figure}
+\end{frame}
 
-\section{Evaluation}
+%\section{Evaluation}
 

### 2012-10-16 talk/vmil2012/presentation/tool/data.py
add marks for 50%, 99% and 99.9% of failures

@@ -20,11 +20,8 @@ def getlines(csvfile):
 
 def we_are_n_percent(info, n):
     failure_counts = info['results'].values()
-    print failure_counts
     failure_counts.sort()
-    print failure_counts
     failure_counts.reverse()
-    print failure_counts
 
     total_failures = sum(failure_counts)
     current_sum = 0
@@ -44,23 +41,42 @@ def build_plot_data(files):
 
     output = []
     plot = """
-\\addplot[%(color)s] coordinates {
+\\addplot[%(color)s,mark=none] coordinates {
 %(data)s
 };
+\\addplot[%(color)s,only marks,mark=*] coordinates {%(marks)s};
     """
     for j, (bench, info) in enumerate(failures.iteritems()):
         data = []
+        marks = []
         results = info['results'].values()
         results.sort()
         results.reverse()
+
+        num_50 = we_are_n_percent(info, 50)
+        num_99 = we_are_n_percent(info, 99)
+        num_99_dot_9 = we_are_n_percent(info, 99.9)
+
+        marks.append("(%04f,%04f)" % (float(num_50) / (len(results) - 1), 
+            float(results[num_50]) / max(results)))
+        marks.append("(%04f,%04f)" % (float(num_99) / (len(results) - 1), 
+            float(results[num_99]) / max(results)))
+        marks.append("(%04f,%04f)" % (float(num_99_dot_9) / (len(results) - 1), 
+            float(results[num_99_dot_9]) / max(results)))
         for i, result in enumerate(results):
             data.append("(%04f,%04f)" % (float(i) / (len(results) - 1),
                         float(result) / max(results)))
 
         if bench == 'go':
             with open('figures/go_data.tex', 'w') as f:
-                f.write(plot % {'color': 'black', 'name': bench, 'data': " ".join(data)})
-        output.append(plot % {'color':COLORS[j], 'name': bench, 'data': " ".join(data)})
+                f.write(plot % {'color': 'green',
+                                'name': bench,
+                                'marks': " ".join(marks),
+                                'data': " ".join(data)})
+        output.append(plot % {'color': COLORS[j],
+                              'name': bench,
+                              'marks': " ".join(marks),
+                              'data': " ".join(data)})
 
     with open('figures/data.tex', 'w') as f:
         for l in output:

### 2012-10-17 talk/vmil2012/presentation/tool/data.py
use same color for group and single plot

@@ -49,6 +49,7 @@ def build_plot_data(files):
     for j, (bench, info) in enumerate(failures.iteritems()):
         data = []
         marks = []
+        color_map = {}
         results = info['results'].values()
         results.sort()
         results.reverse()
@@ -66,14 +67,15 @@ def build_plot_data(files):
         for i, result in enumerate(results):
             data.append("(%04f,%04f)" % (float(i) / (len(results) - 1),
                         float(result) / max(results)))
-
+        #
+        color_map[bench] = COLORS[j]
         if bench == 'go':
             with open('figures/go_data.tex', 'w') as f:
-                f.write(plot % {'color': 'green',
+                f.write(plot % {'color': color_map[bench],
                                 'name': bench,
                                 'marks': " ".join(marks),
                                 'data': " ".join(data)})
-        output.append(plot % {'color': COLORS[j],
+        output.append(plot % {'color': color_map[bench],
                               'name': bench,
                               'marks': " ".join(marks),
                               'data': " ".join(data)})

### 2012-10-18 talk/vmil2012/presentation/talk.tex
use same color for group and single plot
interleave example with explanation slides

@@
       \item JIT works by observing and logging what the interpreter does
-      \begin{itemize}
       \item for interesting, commonly executed code paths
       \item produces a linear list of operations (trace)
-          \item automatically does (potentially deep) inlining
   \end{itemize}
-      \item trace is optimized and then turned into machine code
-  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop01}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop02}
 \end{frame}
@@
       \pause
-      \item \emph{This talk:} describe technology and design decisions around guards
+      \item \emph{This talk:} technology and design decisions of guards
       \pause
@@
   \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop03}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Inlining}
+  Tracing automatically does (potentially deep) inlining
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop04}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop05}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop06}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop07}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop08}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop09}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop10}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop11}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop12}
 \end{frame}
@@
 
-\begin{frame}
-  \frametitle{Running Example}
-\end{frame}
-
 %\section{High-Level}
@@
   \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop07}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/framechain1}
 \end{frame}
@@
 \begin{frame}
+  \includegraphics[scale=0.4]{figures/loop07}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/framechain1}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/loop08}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.4]{figures/framechain2}
+\end{frame}
+
+\begin{frame}
   \frametitle{Interaction with Optimization}
@@
       \item Some optimizations make it necessary to store extra information in symbolic frames
+      \pause
       \item examples:
@@
   \begin{itemize}
-    \item Quick Check if the condition holds
+    \item quick Check if the condition holds
     \item and a mapping of machine locations to JIT-variables % indirection using the fail-boxes
   \end{itemize}
+  \pause
   In case of failure
@@
 \begin{frame}
+  \frametitle{Compiling a Trace}
   \begin{figure}
@@
 \begin{frame}
+  \frametitle{Compiling a Bridge}
   \begin{figure}
@@
 \begin{frame}
+  \frametitle{Patching Guards for Bridges}
   \begin{figure}
@@
   \end{figure}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Patching Guards for Bridges}
 \end{frame}

### 2012-11-01 talk/vmil2012/presentation/talk.tex
swap slides
trying to make diagrams, needs lots of steps
extremely minimalistic conclusion
tweak&comment

@@
 
-\title{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}
+\title[Guards in RPython's Tracing JIT]{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}
 
@@
 \begin{frame}
+  \frametitle{RPython and PyPy}
+  \begin{itemize}
+      \item Context: RPython
+      \item a language for writing interpreters for dynamic languages
+      \item a generic tracing JIT, applicable to many languages
+      \item used to implement PyPy, an efficient Python interpreter
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{How fast is PyPy?}
+  \includegraphics[scale=0.3]{figures/all_numbers.png}
+\end{frame}
+
+\begin{frame}
   \frametitle{Tracing JITs Compile by Observing an Interpreter}
@@
       \pause
+      \item similar to deoptimization points, but more common, and patchable
       \item \emph{This talk:} technology and design decisions of guards
@@
   \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.6]{figures/op_percentage_after}
+\end{frame}
+
+\begin{frame}[t,fragile]
+\pgfplotsset{tick label style={font=\tiny\bfseries},
+label style={font=\small},
+legend style={font=\tiny}
+}
+    \frametitle{Guard Failure Rates / Go Benchmark (29989 Guards)}
+    \begin{figure}
+        \centering
+        \begin{tikzpicture}
+            \begin{axis}[
+            xlabel= Guards by failures,
+            ylabel=Relative \# of failures,
+            xtick=\empty,
+            ytick=\empty,
+            ]
+                \input{figures/go_data}
+            \end{axis}
+        \end{tikzpicture}
+    \end{figure}
+\end{frame}
+
+\begin{frame}[t,fragile]
+\pgfplotsset{tick label style={font=\tiny\bfseries},
+label style={font=\small},
+legend style={font=\tiny}
+}
+    \frametitle{Guard Failure Rates}
+    \begin{figure}
+        \centering
+        \begin{tikzpicture}
+            \begin{axis}[
+            xlabel= Guards by failures,
+            ylabel=Relative \# of failures,
+            xtick=\empty,
+            ytick=\empty,
+            ]
+                \input{figures/data}
+            \end{axis}
+        \end{tikzpicture}
+    \end{figure}
 \end{frame}
@@
 
-\begin{frame}
-  \frametitle{Bridges}
-  \begin{itemize}
-      \item When a trace fails often, it becomes worth to attach a new trace to it
-          \item This is called a bridge
-          \item The bridge is attached by patching the guard machine code
-          \item when this guard fails in the future, the new trace is executed instead
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{RPython and PyPy}
-  \begin{itemize}
-      \item Context: RPython
-      \item a generic tracing JIT, applicable to many languages
-      \item main use: PyPy, an efficient Python interpreter
-  \end{itemize}
-\end{frame}
-
 %\section{High-Level}
@@
       \item done with the help of symbolic frame stacks
-      \item these show how trace variables fill the to-be-built stack frames
+      \item these show how trace variables fill the to-be-built interpreter stack frames
   \end{itemize}
@@
       \item share these between subsequent guards
-      \pause
-      \item also need a byte-saving binary representation, but that's just boring work
   \end{itemize}
@@
 \begin{frame}
+  \frametitle{Compact Representation}
+    also need a byte-saving binary representation, but that's just boring work
+\end{frame}
+
+\begin{frame}
   \frametitle{Interaction with Optimization}
@@
   \begin{itemize}
-    \item quick Check if the condition holds
+    \item quick check if the condition holds
     \item and a mapping of machine locations to JIT-variables % indirection using the fail-boxes
@@
 \begin{frame}
+  \frametitle{Bridges}
+  \begin{itemize}
+      \item When a trace fails often, it becomes worth to attach a new trace to it
+          \item This is called a bridge
+          \item The bridge is attached by patching the guard machine code
+          \item when this guard fails in the future, the new trace is executed instead
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
   \frametitle{Compiling a Bridge}
@@
 \end{frame}
-\begin{frame}[t,fragile]
-\pgfplotsset{tick label style={font=\tiny\bfseries},
-label style={font=\small},
-legend style={font=\tiny}
-}
-    \frametitle{Guard Failure Rates}
-    \begin{figure}
-        \centering
-        \begin{tikzpicture}
-            \begin{axis}[
-            xlabel= Guards by failures,
-            ylabel=Relative \# of failures,
-            xtick=\empty,
-            ytick=\empty,
-            ]
-                \input{figures/data}
-            \end{axis}
-        \end{tikzpicture}
-    \end{figure}
-\end{frame}
-\begin{frame}[t,fragile]
-\pgfplotsset{tick label style={font=\tiny\bfseries},
-label style={font=\small},
-legend style={font=\tiny}
-}
-    \frametitle{Guard Failure Rates / Go Benchmark}
-    \begin{figure}
-        \centering
-        \begin{tikzpicture}
-            \begin{axis}[
-            xlabel= Guards by failures,
-            ylabel=Relative \# of failures,
-            xtick=\empty,
-            ytick=\empty,
-            ]
-                \input{figures/go_data}
-            \end{axis}
-        \end{tikzpicture}
-    \end{figure}
+
+\begin{frame}
+  \frametitle{JIT memory overhead}
+  \includegraphics[width=\textwidth]{figures/jit_memory}
 \end{frame}
 
+\begin{frame}
+  \includegraphics[scale=0.6]{figures/resume_data_size}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Conclusion}
+  \begin{itemize}
+      \item Things that sound simple still often need careful engineering
+      \pause
+      \item guards are fundamental part of tracing JITs, need to be implemented well
+      \item not even any direct performance gains
+      \item keep memory usage sane
+      \item allows good bridges
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Thank you! Questions?}
+  \begin{itemize}
+      \item Things that sound simple still often need careful engineering
+      \item guards are fundamental part of tracing JITs, need to be implemented well
+      \item not even any direct performance gains
+      \item keep memory usage sane
+      \item allows good bridges
+  \end{itemize}
+\end{frame}
 %\section{Evaluation}

### 2012-11-01 talk/vmil2012/tool/backenddata.py

@@ -1,0 +1,94 @@
+#!/usr/bin/env python
+from __future__ import division
+"""
+Parse and summarize the traces produced by pypy-c-jit when PYPYLOG is set.
+only works for logs when unrolling is disabled
+"""
+
+import csv
+import optparse
+import os
+import re
+import sys
+from pypy.jit.metainterp.history import ConstInt
+from pypy.jit.tool.oparser import parse
+from pypy.rpython.lltypesystem import llmemory, lltype
+from pypy.tool import logparser
+
+
+def collect_logfiles(path):
+    if not os.path.isdir(path):
+        logs = [os.path.basename(path)]
+    else:
+        logs = os.listdir(path)
+    all = []
+    for log in logs:
+        parts = log.split(".")
+        if len(parts) != 3:
+            continue
+        l, exe, bench = parts
+        if l != "logbench":
+            continue
+        all.append((exe, bench, log))
+    all.sort()
+    return all
+
+
+def collect_guard_data(log):
+    """Calculate the total size in bytes of the locations maps for all guards
+    in a logfile"""
+    guards = logparser.extract_category(log, 'jit-backend-guard-size')
+    return sum(int(x[6:]) for x in guards if x.startswith('chars'))
+
+
+def collect_asm_size(log, guard_size=0):
+    """Calculate the size of the machine code pieces of a logfile. If
+    guard_size is passed it is substracted from result under the assumption
+    that the guard location maps are encoded in the instruction stream"""
+    asm = logparser.extract_category(log, 'jit-backend-dump')
+    asmlen = 0
+    for block in asm:
+        expr = re.compile("CODE_DUMP @\w+ \+\d+\s+(.*$)")
+        match = expr.search(block)
+        assert match is not None  # no match found
+        code = match.group(1)
+        asmlen += len(code)
+    return asmlen - guard_size
+
+
+def collect_data(dirname, logs):
+    for exe, name, log in logs:
+        path = os.path.join(dirname, log)
+        logfile = logparser.parse_log_file(path)
+        guard_size = collect_guard_data(logfile)
+        asm_size = collect_asm_size(logfile, guard_size)
+        yield (exe, name, log, asm_size, guard_size)
+
+
+def main(path):
+    logs = collect_logfiles(path)
+    if os.path.isdir(path):
+        dirname = path
+    else:
+        dirname = os.path.dirname(path)
+    results = collect_data(dirname, logs)
+
[…]

### 2012-11-09 talk/vmil2012/paper.tex
typos
Start preparing submission for publication
do not require pdf for figures
force page size to US Letter
Conditionally use hyperref or url packages depending on the target

@@
-\documentclass[10pt,preprint]{sigplanconf}
+\documentclass[letter,10pt]{sigplanconf}
+\special{papersize=8.5in,11in}
 
@@
 \usepackage{setspace}
-\usepackage[pdfpagelabels=true]{hyperref}
 \usepackage[colorinlistoftodos]{todonotes}
@@
 \usepackage[scaled=0.81]{beramono}
+\usepackage{ifpdf}
 
@@
 
+\ifpdf
+\usepackage[pdfpagelabels=true,hidelinks]{hyperref}
 \hypersetup{%
   plainpages=false,%
-  %hyperfootnotes=false,%
-  colorlinks=true,%
-  urlcolor=black,%
-  citecolor=black,%
-  linkcolor=black,%
   pdftitle={The Efficient Handling of Guards in the Design of RPython's Tracing JIT},%
@@
 }
+\else
+\usepackage[hyphens]{url}
+\fi
 
@@
 
-\conferenceinfo{VMIL'12}{}
+\conferenceinfo{VMIL'12,} {October 21, 2012, Tucson, Arizona, USA.}
 \CopyrightYear{2012}
+\copyrightdata{978-1-4503-1633-0/12/10}
 \crdata{}
@@
 There are several forces guiding the design of resume data handling.
-Guards are a very common operations in the traces.
+Guards are a very common operation in the traces.
 However, as will be shown, a large percentage of all operations
@@
 \begin{figure}[ht]
-\includegraphics[width=0.4\textwidth]{figures/resume_data.pdf}
+\includegraphics[width=0.4\textwidth]{figures/resume_data}
 \vspace{-3mm}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
+\includegraphics[width=0.5\textwidth]{figures/loop_bridge}
 \caption{Trace control flow in case of guard failures with and without bridges}
@@
 \label{sub:guard_failure}
-The last point in this discussion is the frequency of guard failures.
+The next point in this discussion is the frequency of guard failures.
 Figure~\ref{fig:failing_guards} presents for each benchmark a list of the

### 2012-11-12 talk/vmil2012/paper.tex
cleanup
update authors in pdfinfo

@@
   pdftitle={The Efficient Handling of Guards in the Design of RPython's Tracing JIT},%
-  pdfauthor={David Schneider},
+  pdfauthor={David Schneider and Carl Friedrich Bolz},
 }
@@
 
-%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
-%operations for each benchmark, and the overhead produced by the information
-%stored at the different levels for the guards
 In this paper we want to substantiate the aforementioned observations about guards and
@@
 
-%\todo{Figure needs better formatting}
 \begin{figure}[ht]
@@
 
-%\section*{Appendix}
-
-%\todo{remove this section and the figures}
-%\begin{figure*}
-%    \include{figures/ops_count_table}
-%    \caption{Relative numbers of operations in the traces generated for
-%    different benchmarks}
-%    \label{fig:ops_count}
-%\end{figure*}
-%\begin{figure*}
-%\centering
-%\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
-%\caption{Relative frequency of operations before and after optimization}
-%\label{fig:ops_pie}
-%\end{figure*}
 \bibliographystyle{abbrv}
 \bibliography{zotero,paper}
-%\listoftodos
 \end{document}

### 2012-11-19 talk/dls2012/benchmarks/scimark.py
Fix displaying of copyright data

@@ -1,0 +1,295 @@
+from convolution.convolution import Array2D
+from array import array
+import math
+
+class Random(object):
+    MDIG = 32
+    ONE = 1
+    m1 = (ONE << (MDIG-2)) + ((ONE << (MDIG-2) )-ONE)
+    m2 = ONE << MDIG/2
+    dm1  = 1.0 / float(m1);
+
+    def __init__(self, seed):
+        self.initialize(seed)
+        self.left = 0.0
+        self.right = 1.0
+        self.width = 1.0
+        self.haveRange = False
+
+    def initialize(self, seed):
+    
+        self.seed = seed
+        seed = abs(seed)
+        jseed = min(seed, self.m1)
+        if (jseed % 2 == 0):
+            jseed -= 1
+        k0 = 9069 % self.m2;
+        k1 = 9069 / self.m2;
+        j0 = jseed % self.m2;
+        j1 = jseed / self.m2;
+        self.m = array('d', [0]) * 17 
+        for iloop in xrange(17):
+            jseed = j0 * k0;
+            j1 = (jseed / self.m2 + j0 * k1 + j1 * k0) % (self.m2 / 2);
+            j0 = jseed % self.m2;
+            self.m[iloop] = j0 + self.m2 * j1;
+        self.i = 4;
+        self.j = 16;
+
+    def nextDouble(self):
+        I, J, m = self.i, self.j, self.m
+        k = m[I] - m[J];
+        if (k < 0):
+            k += self.m1;
+        self.m[J] = k;
+
+        if (I == 0):
+            I = 16;
+        else:
+            I -= 1;
+        self.i = I;
+
+        if (J == 0):
+            J = 16;
+        else:
+            J -= 1;
+        self.j = J;
+
+        if (self.haveRange):
+            return  self.left +  self.dm1 * float(k) * self.width;
+        else:
+            return self.dm1 * float(k);
+
+    def RandomMatrix(self, a):
+        for x, y in a.indexes():
+            a[x, y] = self.nextDouble()
+        return a
+
+    def RandomVector(self, n):
+        return array('d', [self.nextDouble() for i in xrange(n)])
+    
+
+class ArrayList(Array2D):
+    def __init__(self, w, h, data=None):
+        self.width = w
+        self.height = h
+        self.data = [array('d', [0]) * w for y in xrange(h)]
+        if data is not None:
[…]

### 2012-11-19 talk/pyconza2012/examples/jit01.py

@@ -1,0 +1,10 @@
+
+def f():
+    i = 0
+    while i < 1000000:
+        i = i + 1
+    return i
+
+if __name__ == '__main__':
+    f()
+

### 2013-01-09 talk/dls2012/dls04-ardo.tex
py3k's December update

--- 2012-08-17 talk/dls2012/paper.tex
@@
 
-\documentclass[preprint]{sigplanconf}
+\documentclass{sigplanconf}
 
@@
 \begin{abstract}
-One of the nice properties of a tracing JIT is that many of its optimizations
+One of the nice properties of a tracing just-in-time compiler (JIT) is that many of its optimizations
 are simple, requiring one forward pass only. This is not true for loop-invariant code
@@
 
-One of the advantages that tracing JIT compilers have above traditional
+One of the advantages that tracing just-in-time compilers (JITs) have above traditional
 method-based
@@
 of Python chosen in such a way that it can be efficiently translated to a
-C-based VM by performing type inference.
+C-based virtual machine (VM) by performing type inference.
 
@@
 \hline
-SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.69 $\pm$ 0.033 & 4.49 $\pm$ 0.036 & 1.84 $\pm$ 0.061\\
-\hline
-SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.019 & 2.43 $\pm$ 0.031 & 1.20 $\pm$ 0.053\\
+SparseMatMult(1e4,5e3,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.69 $\pm$ 0.033 & 4.49 $\pm$ 0.036 & 1.84 $\pm$ 0.061\\
+\hline
+SparseMatMult(1e5,1e6,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.019 & 2.43 $\pm$ 0.031 & 1.20 $\pm$ 0.053\\
 \hline
@@
 \begin{itemize}
-\item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 
-initialized to $x_0=y/2$ and the benchmark consists of a single loop updating this
-approximation using $x_i = \left( x_{i-1} + y/x_{i-1} \right) / 2$ for $1\leq i < 10^8$. 
-Only the latest calculated value $x_i$ is kept alive as a local variable within the loop.
-There are three different versions of this benchmark where $x_i$
-  is represented with different type $T$ of objects: int's, float's and
-  Fix16's. The latter, Fix16, is a custom class that implements
-  fixpoint arithmetic with 16 bits precision. In Python and Lua there is only
-  a single implementation of the benchmark that gets specialized
-  depending on the class of it's input argument, $y$. In C,
-  there are three different implementations.
-
-The Fix16 type is a custom class with operator overloading in Lua and Python.
-The C version uses a C++ class. The goal of this variant of the benchmark is to
-check how large the overhead of a custom arithmetic class is, compared to
-builtin data types.
-
-In Lua there is no direct support for
-integers so the int version is not provided.
-
 \item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
@@
 %${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
-hardcoded into the implementation making the benchmark consist of a single loop too.
+%hardcoded into the implementation making the benchmark consist of a single loop too.
 \item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with a kernel of fixed
@@
   external call to a max function within the loop that prevents some
-  of the optimizations.
+  of the optimizations for PyPy.
 \item {\bf sobel}$\left(n\right)$: a low-level video processing algorithm used to
@@
 performed each iteration as compared to the other benchmarks.
+
+\item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 
+initialized to $x_0=y/2$ and the benchmark consists of a single loop updating this
+approximation using $x_i = \left( x_{i-1} + y/x_{i-1} \right) / 2$ for $1\leq i < 10^8$. 
+Only the latest calculated value $x_i$ is kept alive as a local variable within the loop.
+There are three different versions of this benchmark where $x_i$
+  is represented with different type $T$ of objects: int's, float's and
+  Fix16's. The latter, Fix16, is a custom class that implements
+  fixpoint arithmetic with 16 bits precision. In Python and Lua there is only
+  a single implementation of the benchmark that gets specialized
+  depending on the class of it's input argument, $y$. In C,
+  there are three different implementations.
+
+The Fix16 type is a custom class with operator overloading in Lua and Python.
+The C version uses a C++ class. The goal of this variant of the benchmark is to
+check how large the overhead of a custom arithmetic class is, compared to
+builtin data types.
+
+In Lua there is no direct support for
+integers so the int version is not provided.
+
 \end{itemize}
@@
 \begin{itemize}
+\item {\bf FFT}$\left(n, c\right)$: Fast Fourier Transform of a vector with $n$ elements, represented as an array, repeated $c$ times.
+\item {\bf LU}$\left(n, c\right)$: LU factorization of an $n \times n$ matrix.
+The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable.
+Instead a list of arrays is used to represent the matrix.
+The calculation is repeated $c$ times.
+\item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
 \item {\bf SOR}$\left(n, c\right)$: Jacobi successive over-relaxation on a $n\times n$ grid repreated $c$ times.
@@
 stored in compressed-row format, and a full storage vector, stored in a normal array. The matrix has $z$ non-zero elements and the calculation is repeated $c$ times.
-\item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
-\item {\bf LU}$\left(n, c\right)$: LU factorization of an $n \times n$ matrix.
-The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable.
-Instead a list of arrays is used to represent the matrix.
-The calculation is repeated $c$ times.
-\item {\bf FFT}$\left(n, c\right)$: Fast Fourier Transform of a vector with $n$ elements, represented as an array, repeated $c$ times.
 \end{itemize}

### 2013-07-24 talk/dls2006/conftest.py
Update sprint task.

@@ -1,0 +1,11 @@
+
+import py
+
+class Directory(py.test.collect.Directory):
+    def run(self):
+        l = []
+        for x in super(Directory, self).run(): 
+            if x in ['draft.txt']: 
+                continue
+            l.append(x)
+        return l

### 2013-11-12 talk/rupy2008/talk.tex
typo-level fixes
py3k update #12

@@
 
-\title{PyPy – status and next goals}
+\title{PyPy – Status and Next Goals}
 
@@
   \frametitle{What is PyPy?}
+  ... pretty pictures and demos
+\end{frame}
+
+\begin{frame}
+  \frametitle{What is PyPy?}
   \begin{itemize}
@@
       \item
+        geared at dynamic languages
+      \item
         giving you lots of help
       \item
-        translation to lots of environments
+        translation to various of environments (C, .NET, Java)
       \item
@@
   \item
-    the biggest implemented interpreter is a full Python interpreter
+    a Python interpreter
     \begin{itemize}
+      \item
+        using the toolkit
       \item
@@
     \end{itemize}
+  \item most important goal: flexibility at all levels
   \item
@@
   \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{What is PyPy}
-  ... pretty pictures and demos
 \end{frame}
@@
   \item
-    implementing ctypes
+    implementing CTypes
   \item
@@
   \item CTypes is a libffi wrapper for CPython
-  \item implemented in C
+  \item no need to write C code to interface with libraries
   \item can be used to wrap C libraries rather effortlessly
-  \item no need to write C code to interface with libraries
 
@@
   \item accessing .NET libraries seamlessly from PyPy
+  \item various improvements
   \item ... demo
@@
   \item GCs are completely written in RPython
-  \item similar to the MMTK toolkit of Jikes RVM
   \item two fast GCs: semispace, generational
@@
   \item memory-bound benchmarks faster than CPython
-  \item demo (id(x))
+  \item ... tiny demo
   \end{itemize}
@@
   \item writing good JITs for dynamic languages is hard work
+  \item good performance needs lots of care
   \item keeping up with new language features is hard
@@
     \item Tamarin (company support, who knows how good it is)
+    \item Self
+    \item Smalltalk VMs
     \end{itemize}
@@
   \end{itemize}
-  \pause
 \end{frame}
@@
   \pause
-  \item Idea: Automatically Generate the JIT from the interpreter
+  \item Idea: Automatically generate the JIT from the interpreter
   \pause
@@
   \item You only need to do it once
-  \item the effort of writing a JIT generator is independent of the size of the language
   \item can keep up with language evolution
+  \item the JIT is correct by construction
   \item retargetting the JIT is possible (e.g. to .NET, as Antonio Cuni is doing right now)
@@
   \item we hope to get really good results for more general code in a few months
-  \item ... demo
   \end{itemize}

### 2013-11-12 talk/dls2012/benchmarks/parse.py

@@ -1,0 +1,121 @@
+
+import pdb, sys
+import numpy as np
+import matplotlib.pyplot as plt
+
+# force type 1 fonts
+import matplotlib
+
+matplotlib.rcParams['ps.useafm'] = True
+matplotlib.rcParams['pdf.use14corefonts'] = True
+matplotlib.rcParams['text.usetex'] = True
+
+NAME_REPL = {
+    'dilate3x3(Array2D(1000x1000))': 'dilate3x3(1000,1000)',
+    'sobel_magnitude(1000,1000)': 'sobel(1000,1000)',
+    'conv3(array(1e5))': 'conv3(1e5)',
+    'conv3(array(1e6))': 'conv3(1e6)',
+    'conv5(array(1e5))': 'conv5(1e5)',
+    'conv5(array(1e6))': 'conv5(1e6)',
+    'sobel(Array2D(1000x1000))': 'sobel(1000,1000)',
+    'sobel(Array(1000x1000))': 'sobel(1000,1000)',
+    'conv3x3(Array2D(1000000x3))': 'conv3x3(1000000,3)',
+    'conv3x3(Array2D(1000x1000))': 'conv3x3(1000,1000)',
+    'dilate3x3(1000)': 'dilate3x3(1000,1000)',
+    'conv3x3(1000)': 'conv3x3(1000,1000)',
+    'conv3x3(3)': 'conv3x3(1000000,3)',
+    'SparseMatMult(1000,5000,262144)': 'SparseMatMult(1e4,5e3,262144)',
+    'SparseMatMult(100000,1000000,1024)': 'SparseMatMult(1e5,1e6,1024)',
+}
+
+IGNORE = {
+    "conv3(1e5)",
+    "conv5(1e5)",
+    "conv5(1e6)",
+    "conv3x3(1000000,3)",
+}
+
+def main(name):
+    interp = None
+    res = {}
+    order = ['python2.7', 'pypy --jit enable_opts=intbounds:rewrite:virtualize:string:earlyforce:pure:heap:ffi', 'pypy', 'luajit -O-loop', 'luajit', 'gcc -O3 -march=native -fno-tree-vectorize']
+    labels = [None, 'PyPy no LP', 'PyPy', 'LuaJIT no LP', 'LuaJIT', None]
+    with open(name) as f:
+        for line in f:
+            line = line.strip("\n")
+            if not line:
+                interp = None
+            elif interp is None:
+                interp = line
+            else:
+                bench, rest = line.split(':')
[…]

### 2014-01-14 talk/ecoop2009/abstract.tex
I'm no longer late today (at least not by more than 5 minutes)

@@
 \begin{abstract}
-Implementing dynamically typed languages in an efficient way is a hard task.
-Recent developments show that JIT compilers 
+Writing an optimizing static compiler for dynamic languages is not an
+easy task, since quite complex static analysis is required.
+On the other hand, recent developments show that JIT compilers 
 can exploit runtime type information to generate quite efficient code.

### 2014-01-14 talk/icooolps2009-dotnet/benchmarks.tex
I'm no longer late today (at least not by more than 5 minutes)

@@
 
+The columns \emph{Interp/JIT2} and \emph{JIT2/C\#} measure the speedup of the
+generated code compared to plain interpretation and of the equivalent C\#
+program compared to the generated code, respectively.
+
 Moreover, for each benchmark we also show the time taken by running the
@@
   programs are available at:
-  \texttt{http://codespeak.net/~antocuni/tlc-benchmarks/}}
+  \texttt{http://codespeak.net/$\sim$antocuni/tlc-benchmarks/}}
 
@@
 
-The difference in speed it is probably due to both the fact that the current
+The difference in speed is probably due to both the fact that the current
 CLI backend emits slightly non-optimal code and that the underyling .NET JIT

### 2014-01-14 talk/dls2012/benchmarks/scimark/run_MonteCarlo.c

@@ -1,0 +1,15 @@
+#include <stdio.h>
+#include <assert.h>
+
+#include "Random.c"
+#include "MonteCarlo.c"
+
+int main(int ac, char **av) {
+    assert(ac==2);
+    int N = atoi(av[1]);
+    MonteCarlo_integrate(N);
+    fprintf(stderr, "MonteCarlo(%d):    ", N);
+    return 0;
+}
+
+

### 2014-01-14 talk/pypy-talk-pycon2005/ep.py

@@ -1,0 +1,651 @@
+from __future__ import generators
+import pygame
+from pygame.locals import *
+import sys, os
+from bubbob import pixmap
+from bubbob.patmap import patmap
+from bubbob import mnstrmap
+
+SubtitlesLine = 0
+#SubtitlesLine = 28
+
+bwidth = 784
+bheight = 592
+
+pygame.init()
+winstyle = HWSURFACE  # |FULLSCREEN
+bestdepth = pygame.display.mode_ok((bwidth, bheight+SubtitlesLine), winstyle, 32)
+screen = pygame.display.set_mode((bwidth, bheight+SubtitlesLine), winstyle,
+                                 bestdepth)
+
+
+class Ico:
+    Cache = {}
+    
+    def __init__(self, filename, rect, filter=None, colorkey=0x010101):
+        try:
+            img = Ico.Cache[filename, filter]
+        except KeyError:
+            data = open(filename, 'rb').read()
+            w, h, data = pixmap.decodepixmap(data)
+            if filter:
+                w, h, data = filter(w, h, data)
+            img = pygame.image.fromstring(data, (w, h), "RGB")
+            if colorkey is not None:
+                img.set_colorkey(color(colorkey), RLEACCEL)
+            Ico.Cache[filename, filter] = img
+        if filter:
+            rect = doublerect(rect)
+        self.img = img.subsurface(rect)
+        self.w, self.h = self.img.get_size()
+    def draw(self, x, y):
+        screen.blit(self.img, (x, y))
+
+sprmap = {}
+
+def doublesize(w, h, data):
+    scanline = 3*w
+    result = []
+    for position in range(0, scanline*h, scanline):
+        line = []
+        for p in range(position, position+scanline, 3):
+            line.append(2 * (data[p  ] + data[p+1] + data[p+2]))
+        line = ''.join(line)
+        result.append(line)
+        result.append(line)
+    return w*2, h*2, ''.join(result)
+
+def doublerect((x, y, w, h)):
+    return x*2, y*2, w*2, h*2
+
+def sprget(n, filter=None, spriconcache={}):
+    try:
+        return spriconcache[n, filter]
+    except KeyError:
+        filename, rect = sprmap[n]
+        ico = Ico(filename, rect, filter)
+        spriconcache[n, filter] = ico
+        return ico
+
+def loadpattern(n):
+    filename, rect = patmap[n]
+    filename = os.path.join('pat', filename)
[…]

### 2014-01-14 talk/vmil2012/tool/difflogs.py

@@ -1,0 +1,198 @@
+#!/usr/bin/env python
+"""
+Parse and summarize the traces produced by pypy-c-jit when PYPYLOG is set.
+only works for logs when unrolling is disabled
+"""
+
+import py
+import os
+import sys
+import csv
+import optparse
+from pprint import pprint
+from pypy.tool import logparser
+from pypy.jit.tool.oparser import parse
+from pypy.jit.metainterp.history import ConstInt
+from pypy.rpython.lltypesystem import llmemory, lltype
+
+categories = {
+    'getarrayitem_gc': 'get',
+    'getarrayitem_gc_pure': 'get',
+    'getarrayitem_raw': 'get',
+    'getfield_gc': 'get',
+    'getfield_gc_pure': 'get',
+    'getfield_raw': 'get',
+    'getinteriorfield_gc': 'get',
+    'new': 'new',
+    'new_array': 'new',
+    'new_with_vtable': 'new',
+    'newstr': 'new',
+    'newunicode': 'new',
+    'setarrayitem_gc': 'set',
+    'setarrayitem_raw': 'set',
+    'setfield_gc': 'set',
+    'setfield_raw': 'set',
+    'setinteriorfield_gc': 'set',
+    'strgetitem': 'get',
+    'strsetitem': 'set',
+}
+rest_op_bucket = set()
+
+all_categories = 'new get set guard numeric rest'.split()
+
+def extract_opnames(loop):
+    loop = loop.splitlines()
+    for line in loop:
+        if line.startswith('#') or line.startswith("[") or "end of the loop" in line:
+            continue
+        frontpart, paren, _ = line.partition("(")
+        assert paren
+        if " = " in frontpart:
+            yield frontpart.split(" = ", 1)[1]
+        elif ": " in frontpart:
+            yield frontpart.split(": ", 1)[1]
+        else:
+            yield frontpart
+
+def summarize(loop, adding_insns={}):    # for debugging
+    insns = adding_insns.copy()
+    seen_label = True
+    if "label" in loop:
+        seen_label = False
+    for opname in extract_opnames(loop):
+        if not seen_label:
+            if opname == 'label':
+                seen_label = True
+            else:
+                assert categories.get(opname, "rest") == "get"
+                continue
+        if(opname.startswith("int_")
+                or opname.startswith("float_")
+                or opname.startswith('uint_')):
+            opname = "numeric"
+        elif opname.startswith("guard_"):
+            opname = "guard"
+        else:
[…]

### 2014-01-14 talk/vmil2012/tool/guarddata.py

@@ -1,0 +1,67 @@
+#!/usr/bin/env python
+"""
+Parse and summarize the jit-summary data """
+
+import optparse
+import os
+import re
+import sys
+import json
+from pypy.jit.metainterp.history import ConstInt
+from pypy.jit.tool.oparser import parse
+from pypy.rpython.lltypesystem import llmemory, lltype
+from pypy.tool import logparser
+from backenddata import collect_logfiles
+
+def collect_data(dirname, logs):
+    for exe, bench, log in logs:
+        path = os.path.join(dirname, log)
+        logfile = logparser.parse_log_file(path)
+        counts = {}
+        guard_failures = \
+            logparser.extract_category(logfile, 'jit-guard-failure')
+        backend_counts = \
+            logparser.extract_category(logfile, 'jit-backend-counts')
+
+        assert len(guard_failures) > 0
+        assert len(backend_counts) > 0
+        # collect guard failures first
+        for g in guard_failures:
+            name = g.split(' ')[1].strip()
+            counts.setdefault(name, 0)
+            counts[name] += 1
+        for i in backend_counts:
+            if i == '':
+                continue
+            for l in i.splitlines():
+                if not l.startswith('bridge'):
+                    continue
+                colon = l.index(':')
+                n = l[len('bridge '):colon]
+                count = int(l[colon+1:])
+                counts[n] += count
+        yield (exe, bench, counts)
+
+
+def main(path):
+    logs = collect_logfiles(path)
+    if os.path.isdir(path):
+        dirname = path
+    else:
+        dirname = os.path.dirname(path)
+    results = collect_data(dirname, logs)
+    data = {}
+    for exe, bench, guards in results:
+        data[bench] = {'exe': exe, 'results': guards}
+    with file("logs/guard_summary.json", "w") as f:
+        print >>f, json.dumps(data, f, sort_keys=True, indent=4)
+
+if __name__ == '__main__':
+    parser = optparse.OptionParser(usage="%prog logdir_or_file")
+
+    options, args = parser.parse_args()
+    if len(args) != 1:
+        parser.print_help()
+        sys.exit(2)
+    else:
+        main(args[0])

### 2014-04-28 talk/icooolps2014/position-paper.tex
add paper outline

\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage[utf8]{inputenc}

\usepackage{amsmath}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
\copyrightyear{2014}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish,
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers,
                                  % short abstracts)

%% \titlebanner{banner above paper title}        % These are ignored unless
%% \preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Title Text}
\subtitle{Position Paper, ICOOOLPS'14}

\authorinfo{Remi Meier}
           {Department of Computer Science\\ ETH Zürich}
           {remi.meier@inf.ethz.ch}
\authorinfo{Armin Rigo}
           {www.pypy.org}
           {arigo@tunes.org}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore,
% you may leave them out
%% \terms
%% term1, term2

\keywords
transactional memory, dynamic languages, parallelism, global interpreter lock

\section{Introduction}

\subsection*{Issue}
efficiently supporting multi-CPU usage on dynamic languages that were designed with GIL semantics in
mind

(supporting (large) atomic blocks for synchronization)

\subsection*{Our Position}
Current solutions for replacing the GIL include STM, HTM, and
fine-grained locking. STM is usually too slow, HTM very limited, and
locking suffers from complexity that makes it hard to implement
correctly. We argue that the best way forward is still STM and that
its performance problem can be solved.

%% Current solutions like STM, HTM, and fine-grained locking are slow, hard
%% to implement correctly, and don't fit the specific problems of dynamic
%% language.  STM is the best way forward but has bad performance, so we
%% fix that.

\section{Discussion}
\paragraph{dynamic language VM problems}

- high allocation rate (short lived objects)\\
- (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)

\paragraph{GIL}

- nice semantics\\
- easy support of atomic blocks\\
- no parallelism

\paragraph{fine-grained locking}

- support of atomic blocks?\\
- hard to get right (deadlocks, performance, lock-granularity)\\
- very hard to get right for a large language\\
- hard to retro-fit, as all existing code assumes GIL semantics\\
- (there are some semantic differences, right? not given perfect lock-placement, but well)
( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )

\paragraph{multiprocessing / no-sharing models}

- often needs major restructuring of programs (explicit data exchange)\\
- sometimes communication overhead is too large\\
- shared memory is a problem, copies of memory are too expensive

\paragraph{HTM}

- false-sharing on cache-line level\\
- limited capacity (caches, undocumented)\\
- random aborts (haswell)\\
- generally: transaction-length limited (no atomic blocks)

\paragraph{STM}

- overhead (100-1000\%) (barrier reference resolution, kills performance on low \#cpu)
(FastLane: low overhead, not much gain)\\
- unlimited transaction length (easy atomic blocks)

\section{Potential Approach}
possible solution:\\
- use virtual memory paging to somehow lower the STM overhead\\
- tight integration with GC and jit?

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2014-04-29 talk/icooolps2014/position-paper.tex
add some text to the introduction

@@
 
-\title{Title Text}
+\title{The Way Forward in Parallelizing Dynamic Languages}
 \subtitle{Position Paper, ICOOOLPS'14}
@@
 \section{Introduction}
+In a world where computers get more and more cores and single-thread
+performance increases less and less every year, many dynamic languages
+have a problem. While there is certainly a lot of popularity around
+languages like Python and Ruby, their ability to make use of multiple
+cores is somewhat limited. For ease of implementation they chose to
+use a single, global interpreter lock (GIL) to synchronize the
+execution of code in multiple threads. While this is a
+straight-forward way to eliminate synchronization issues in the
+interpreter, it prevents parallel execution. Code executed in multiple
+threads will be serialized over this GIL so only one thread executes
+at a time.
 
-\subsection*{Issue}
-efficiently supporting multi-CPU usage on dynamic languages that were designed with GIL semantics in
-mind
+There exist several solutions and work-arounds to remove or avoid the
+GIL in order to benefit from multiple cores. We are going to discuss
+several of them and try to find the best way forward. The first
+approach uses fine-grained locking to replace the single GIL. Then
+there are shared-nothing models that use for example multiple
+processes with multiple interpreters and explicit message
+passing. Finally, one can also directly replace the GIL with
+transactional memory (TM), either software-based (STM) or
+hardware-based (HTM).
 
-(supporting (large) atomic blocks for synchronization)
+The approach that wins in the end should perform similarly for
+single-threaded execution as compared to the GIL and be able to
+execute code in parallel on multiple cores. Furthermore, we will also
+take into account the compatibility to existing code that already uses
+threads for concurrency, as well as the changes that are required to
+the interpreter itself.
 
-\subsection*{Our Position}
-Current solutions for replacing the GIL include STM, HTM, and
-fine-grained locking. STM is usually too slow, HTM very limited, and
-locking suffers from complexity that makes it hard to implement
-correctly. We argue that the best way forward is still STM and that
-its performance problem can be solved.
+These requirements are not easy to meet. We argue that STM is the
+overall winner. While it has a big performance problem currently, it
+gets more points in all the other categories. We think that it is the
+only solution that also enables a better synchronization mechanism to
+the application in the form of atomic blocks.
+
+%% \subsection{Issue}
+%% The issue that we want to discuss is how to efficiently support
+%% multi-core parallel execution of code in dynamic languages that were
+%% designed with GIL semantics in mind.
+
+%% Furthermore, a solution to this problem should also bring better
+%% synchronization mechanism with it...
+
+%% (supporting (large) atomic blocks for synchronization)
+
+%% \subsection{Our Position}
+%% Current solutions for replacing the GIL include STM, HTM, and
+%% fine-grained locking. STM is usually too slow, HTM very limited, and
+%% locking suffers from complexity that makes it hard to implement
+%% correctly. We argue that the best way forward is still STM and that
+%% its performance problem can be solved.
 
@@
 \section{Discussion}
+(why do we have the GIL? it makes the interpreter thread-safe)
+(now there is code relying on the semantics (atomicity of bytecode instructions))
+(synchronization with locks for application-level synchronization is still needed. So GIL is useless for that)
+
 \paragraph{dynamic language VM problems}

### 2014-04-29 talk/icooolps2014/position-paper.tex
more text

@@
 \section{Discussion}
-(why do we have the GIL? it makes the interpreter thread-safe)
-(now there is code relying on the semantics (atomicity of bytecode instructions))
-(synchronization with locks for application-level synchronization is still needed. So GIL is useless for that)
+\subsection{Why is there a GIL?}
+The GIL is a very simple synchronization mechanism for supporting
+multi-threading in the interpreter. The basic guarantee is that the
+GIL may only be released in-between bytecode instructions. The
+interpreter can thus rely on complete isolation and atomicity of these
+instructions. As a consequence, applications can rely on certain
+operations to be atomic. While this is probably not a good idea,
+it is used in practice. A solution replacing the GIL should therefore
+uphold these guarantees, while preferably also be as easily
+implementable as a GIL for the interpreter.
+
+The GIL also allows for easy integration with external C libraries that
+do not need to be thread-safe. For the duration of the calls, we
+simply do not release the GIL. External libraries that are explicitly
+thread-safe can voluntarily release the GIL themselves in order to
+still provide some parallelism. This is done for example for
+potentially long I/O operations. Consequently, I/O-bound,
+multi-threaded applications can actually parallelize to some
+degree. Again, a potential solution should be able to integrate with
+external libraries with similar ease. We will however focus our
+argumentation more on running code in the interpreted language in
+parallel, not the external C calls.
+
+Since the GIL is mostly an implementation detail of the interpreter,
+it is not exposed to the application running on top of it. To
+synchronize memory accesses in applications using threads, the
+state-of-the-art still means explicit locking everywhere. It is well
+known that using locks for synchronization is not easy.  They are
+non-composable, have overhead, may deadlock, limit scalability, and
+overall add a lot of complexity. For a better parallel programming
+model for dynamic languages, we propose another, well-known
+synchronization mechanism called \emph{atomic blocks}.
+
+Atomic blocks are composable, deadlock-free, higher-level and expose
+useful atomicity and isolation guarantees to the application for a
+series of instructions.  Interpreters using using a GIL can simply
+guarantee that the GIL is not released during the execution of the
+atomic block. Of course, this still means that no two atomic blocks
+can execute in parallel or even concurrently. Potential solutions
+that provide a good way to implement atomic blocks are therefore
+preferable.
+
+\subsection{Potential Solutions}
 

### 2014-04-29 talk/icooolps2014/position-paper.tex
xxx: should mention that the interpreter is typically very large and maintained by open-source communities
more text

@@
 interpreter, it prevents parallel execution. Code executed in multiple
-threads will be serialized over this GIL so only one thread executes
-at a time.
+threads will be serialized over this GIL so that only one thread can
+execute at a time.
 
@@
 gets more points in all the other categories. We think that it is the
-only solution that also enables a better synchronization mechanism to
+only solution that also provides a better synchronization mechanism to
 the application in the form of atomic blocks.
@@
 \section{Discussion}
+
+\paragraph{dynamic language VM problems}
+
+- high allocation rate (short lived objects)\\
+- (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)
+
 \subsection{Why is there a GIL?}
@@
 
-\paragraph{dynamic language VM problems}
+The list of criterias for evaluating potential solutions is as follows:
+\begin{itemize}
+\item Performance
+\item Changes required to existing applications
+\item Better synchronization options for applications (e.g. atomic blocks)
+\item Ease of implementation (interpreter-level)
+\item Integration with external libraries
+\end{itemize}
 
-- high allocation rate (short lived objects)\\
-- (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)
-
-\paragraph{GIL}
-
-- nice semantics\\
-- easy support of atomic blocks\\
-- no parallelism
-
-\paragraph{fine-grained locking}
+\subsubsection*{fine-grained locking}
 
@@
 
-\paragraph{multiprocessing / no-sharing models}
+\subsubsection*{Shared-nothing / multiple processes}
 
@@
 
+\subsubsection*{Transactional Memory}
 \paragraph{HTM}
@@
 
-\section{Potential Approach}
+\section{The Way Forward}
 possible solution:\\

### 2014-04-29 talk/icooolps2014/position-paper.tex
some changes

@@
 implementable as a GIL for the interpreter.
+[xxx mention that the interpreter is typically very large and maintained
+by open-source communities]
 
@@
 
-The list of criterias for evaluating potential solutions is as follows:
-\begin{itemize}
-\item Performance
-\item Changes required to existing applications
-\item Better synchronization options for applications (e.g. atomic blocks)
-\item Ease of implementation (interpreter-level)
-\item Integration with external libraries
-\end{itemize}
+For the discussion we define a set of criteria to evaluate the
+multiple potential solutions for removing or avoiding the GIL and its
+limitations:
 
-\subsubsection*{fine-grained locking}
+\begin{description}
+\item[Performance:] How well does the approach perform compared to the
+  GIL on single and multiple threads?
+\item[Existing applications:] How big are the changes required to
+  integrate with and parallelize existing applications?
+\item[Better synchronization:] Does the approach enable better
+  synchronization mechanisms for applications (e.g. atomic blocks)?
+\item[Implementation:] How difficult is it to implement the approach
+  in the interpreter?
+\item[External libraries:] Does the approach allow for easy
+  integration of external libraries?
+\end{description}
+
+\subsubsection{Fine-Grained Locking}
+
+The first obvious candidate to replace the GIL is to use multiple
+locks instead of a single global lock. By refining the granularity
+of the locking approach, we gain the ability to run code that does
+not access the same objects in parallel.
 
@@
 
-\subsubsection*{Shared-nothing / multiple processes}
+\subsubsection{Shared-Nothing / multiple processes}
 
@@
 
-\subsubsection*{Transactional Memory}
+\subsubsection{Transactional Memory}
 \paragraph{HTM}
@@
 
-\appendix
-\section{Appendix Title}
+%% \appendix
+%% \section{Appendix Title}
 
-This is the text of the appendix, if you need one.
+%% This is the text of the appendix, if you need one.
 

### 2014-05-01 talk/icooolps2014/position-paper.tex
fine-grained locking draft

@@
 The first obvious candidate to replace the GIL is to use multiple
-locks instead of a single global lock. By refining the granularity
-of the locking approach, we gain the ability to run code that does
-not access the same objects in parallel.
+locks instead of a single global lock. By refining the granularity of
+the locking approach, we gain the ability to run code that does not
+access the same objects in parallel. What we lose instead is the
+simplicity of the GIL approach. With every additional lock, the
+likeliness of deadlocks grows, as well as the overhead that acquiring
+and releasing locks produces.
 
-- support of atomic blocks?\\
-- hard to get right (deadlocks, performance, lock-granularity)\\
-- very hard to get right for a large language\\
-- hard to retro-fit, as all existing code assumes GIL semantics\\
-- (there are some semantic differences, right? not given perfect lock-placement, but well)
-( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )
+Jython\footnote{www.jython.org} is one project that implements an
+interpreter for Python on the JVM\footnote{Java Virtual Machine} and
+that uses fine-grained locking to correctly synchronize the
+interpreter. For a language like Python, one needs quite a few,
+carefully placed locks. Since there is no central location, the
+complexity of the implementation is quite a bit greater compared to
+using a GIL. Integrating external, non-thread-safe libraries should
+however be very simple too.  One could simply use one lock per library
+to avoid this issue.
+
+In the end, fine-grained locking can transparently replace the GIL
+and therefore parallelize existing applications without any
+changes. It does however not provide a better synchronization
+mechanism to the application like e.g. atomic blocks.
+
+%% - support of atomic blocks?\\
+%% - hard to get right (deadlocks, performance, lock-granularity)\\
+%% - very hard to get right for a large language\\
+%% - hard to retro-fit, as all existing code assumes GIL semantics\\
+%% - (there are some semantic differences, right? not given perfect lock-placement, but well)
+%% ( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )
 

### 2014-05-01 talk/icooolps2014/position-paper.tex
shared-nothing draft

@@
 
-- often needs major restructuring of programs (explicit data exchange)\\
-- sometimes communication overhead is too large\\
-- shared memory is a problem, copies of memory are too expensive
+There are also approaches that work around the GIL instead of trying
+to replace it. If an application can be split into completely
+independent parts that only very rarely need to share anything, or
+only do so via an external program like a database, then it is
+sensible to have one GIL per independent part. As an example of such
+an approach we look at the
+\emph{multiprocessing}\footnote{https://docs.python.org/2/library/multiprocessing.html}
+module of Python. In essence, it uses process-forking to provide the
+application with multiple interpreters that can run in parallel.
+Communication is then done explicitly through pipes.
+
+Obviously not every application fits well into this model and its
+applicability is therefore quite limited. Performance is good as
+long as the application does not need to communicate a lot, because
+inter-process communication is relatively expensive. Also the
+implementation of this approach is very cheap since one can
+actually take an unmodfied GIL-supported interpreter and run
+multiple of them in parallel. That way, we also inherit the
+easy integration of external libraries without any changes.
+While the model of explicit communication is often seen as a
+superior way to synchronize concurrent applications because
+of its explicitness, it does not actually introduce a better
+synchronization mechanism for applications.
+
+%% - often needs major restructuring of programs (explicit data exchange)\\
+%% - sometimes communication overhead is too large\\
+%% - shared memory is a problem, copies of memory are too expensive
 

### 2014-05-01 talk/icooolps2014/position-paper.tex
TM draft

@@
 interpreter can thus rely on complete isolation and atomicity of these
-instructions. As a consequence, applications can rely on certain
-operations to be atomic. While this is probably not a good idea,
-it is used in practice. A solution replacing the GIL should therefore
-uphold these guarantees, while preferably also be as easily
-implementable as a GIL for the interpreter.
+instructions. Additionally, it provides the application with a
+sequential consistency model. As a consequence, applications can rely
+on certain operations to be atomic and that they will always be
+executed in the order in which they appear in the code. While
+depending on this may not always be a good idea, it is done in
+practice. A solution replacing the GIL should therefore uphold these
+guarantees, while preferably also be as easily implementable as a GIL
+for the interpreter.
 [xxx mention that the interpreter is typically very large and maintained
@@
 \subsubsection{Transactional Memory}
+Transactional memory (TM) can be used as a direct replacement for a
+single global lock. Transactions provide the same atomicity and
+isolation guarantees as the GIL provides for the execution of bytecode
+instructions. So instead of acquiring and releasing the GIL between
+these instructions, this approach runs the protected instructions
+inside transactions.
+
+TM can be implemented in software (STM) or in hardware (HTM. There are
+also some hybrid approaches that combine the two. We count these
+hybrid approaches as STM, since they usually provide the same
+capabilities as software-only approaches but with different
+performance characteristics. We will now first look at HTM that
+recently gained a lot of popularity by its introduction in common
+desktop CPUs from Intel (Haswell generation).
+
 \paragraph{HTM}
 
-- false-sharing on cache-line level\\
-- limited capacity (caches, undocumented)\\
-- random aborts (haswell)\\
-- generally: transaction-length limited (no atomic blocks)
+HTM provides us with transactions like any TM system does. It can
+be used as a direct replacement for the GIL. However, as is common
+with hardware-only solutions, there are quite a few limitations
+that can not be lifted easily. For this comparison, we look at
+the implementation of Intel in recent Haswell generation CPUs.
+
+HTM in these CPUs works on the level of caches. This has a few
+consequences like false-sharing on the cache-line level, and most
+importantly it limits the amount of memory that can be accessed within
+a transaction. This transaction-length limitation makes it necessary
+to have a fallback in place in case this limit is reached. In recent
+attempts, the usual fallback is the GIL (XXX: cite). The current
+generation of HTM hits this limit very often for our use case (XXX:
+cite ruby GIL paper) and therefore does not parallelize that well.
+
+The performance of HTM is pretty good (XXX: cite again...) as it does
+not introduce much overhead. And it can transparently parallelize
+existing applications to some degree. The implementation is very
+straight-forward because it directly replaces the GIL in a central
+place. HTM is also directly compatible with any external library that
+needs to be integrated and synchronized for use in multiple
+threads. The one thing that is missing is support for a better
+synchronization mechanism for the application. It is not possible
+in general to expose the hardware-transactions to the application
+in the form of atomic blocks because that would require much
+longer transactions.
+
+%% - false-sharing on cache-line level\\
+%% - limited capacity (caches, undocumented)\\
+%% - random aborts (haswell)\\
+%% - generally: transaction-length limited (no atomic blocks)
 
@@
 
-- overhead (100-1000\%) (barrier reference resolution, kills performance on low \#cpu)
-(FastLane: low overhead, not much gain)\\
-- unlimited transaction length (easy atomic blocks)
+STM provides all the same benefits as HTM except for its performance.
+It is not unusual for the overhead introduced by STM to be between
+100\% to even 1000\%. While STM systems often scale very well to a big
+number of threads and eventually overtake the single-threaded
+execution, they often provide no benefits at all for low numbers of
+threads (1-8). There are some attempts (XXX: cite fastlane) that can
+reduce the overhead a lot, but also scale very badly so that their
+benefit on more than one thread is little.
+
+However, STM compared to HTM does not suffer from the same restricting
+limitations. Transactions can be arbitrarily long.  This makes it
+possible to actually expose transactions to the application in the
+form of atomic blocks. This is the only approach that enables a better
+synchronization mechanism than locks for applications \emph{and} still
+parallelizes when using it. We think this is a very important point
+because it not only gives dynamic languages the ability to parallelize
+(already commonplace in most other languages), but also pushes
+parallel programming forward. Together with sequential consistency it
+provides a lot of simplification for parallel applications.
+
+%% - overhead (100-1000\%) (barrier reference resolution, kills performance on low \#cpu)
+%% (FastLane: low overhead, not much gain)\\
+%% - unlimited transaction length (easy atomic blocks)
 
@@
 \end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2014-05-01 talk/icooolps2014/position-paper.tex
add some handwavy comparison table

@@
 \usepackage[utf8]{inputenc}
-
+\usepackage{array}
 \usepackage{amsmath}
@@
 \section{The Way Forward}
+\begin{table}[]
+  \centering
+  \begin{tabular}{|p{2cm}|c|c|c|c|c|}
+    \hline
+    & \textbf{GIL} & \parbox[t]{1cm}{\textbf{Fine-grained locking}}
+    & \parbox[t]{1cm}{\textbf{Shared-nothing}} & \textbf{HTM} & \textbf{STM}\\
+    \hline
+    Performance            & 0  & +  & ++   & +  & -{-} \\
+    \hline
+    Existing applications  & ++ & ++ & -{-} & ++ & ++   \\
+    \hline
+    Better synchronization & 0  & 0  & -    & -  & ++   \\
+    \hline
+    Implementation         & ++ & -  & ++   & ++ & ++   \\
+    \hline
+    External libraries     & ++ & ++ & ++   & ++ & ++   \\
+    \hline
+  \end{tabular}
+  \caption{Comparison (--/-/0/+/++)}
+  \label{tab:comparison}
+\end{table}
+
+Comparison in Table \ref{tab:comparison}
+
 possible solution:\\

### 2014-05-01 talk/icooolps2014/position-paper.tex
more text; corrections

--- 2014-05-01 talk/icooolps2014/position-paper.tex
@@
 cores is somewhat limited. For ease of implementation they chose to
-use a single, global interpreter lock (GIL) to synchronize the
+use a single, global interpreter lock (GIL) to synchronise the
 execution of code in multiple threads. While this is a
-straight-forward way to eliminate synchronization issues in the
+straight-forward way to eliminate synchronisation issues in the
 interpreter, it prevents parallel execution. Code executed in multiple
-threads will be serialized over this GIL so that only one thread can
+threads will be serialised over this GIL so that only one thread can
 execute at a time.
 
-There exist several solutions and work-arounds to remove or avoid the
+There exist several solutions and workarounds to remove or avoid the
 GIL in order to benefit from multiple cores. We are going to discuss
@@
 gets more points in all the other categories. We think that it is the
-only solution that also provides a better synchronization mechanism to
+only solution that also provides a better synchronisation mechanism to
 the application in the form of atomic blocks.
@@
 \paragraph{dynamic language VM problems}
-
+XXX:
 - high allocation rate (short lived objects)\\
@@
 \subsection{Why is there a GIL?}
-The GIL is a very simple synchronization mechanism for supporting
+The GIL is a very simple synchronisation mechanism for supporting
 multi-threading in the interpreter. The basic guarantee is that the
@@
 potentially long I/O operations. Consequently, I/O-bound,
-multi-threaded applications can actually parallelize to some
+multi-threaded applications can actually parallelise to some
 degree. Again, a potential solution should be able to integrate with
@@
 it is not exposed to the application running on top of it. To
-synchronize memory accesses in applications using threads, the
+synchronise memory accesses in applications using threads, the
 state-of-the-art still means explicit locking everywhere. It is well
-known that using locks for synchronization is not easy.  They are
+known that using locks for synchronisation is not easy.  They are
 non-composable, have overhead, may deadlock, limit scalability, and
@@
 model for dynamic languages, we propose another, well-known
-synchronization mechanism called \emph{atomic blocks}.
+synchronisation mechanism called \emph{atomic blocks}.
 
@@
 useful atomicity and isolation guarantees to the application for a
-series of instructions.  Interpreters using using a GIL can simply
-guarantee that the GIL is not released during the execution of the
-atomic block. Of course, this still means that no two atomic blocks
-can execute in parallel or even concurrently. Potential solutions
-that provide a good way to implement atomic blocks are therefore
+series of instructions.  Interpreters using a GIL can simply guarantee
+that the GIL is not released during the execution of the atomic
+block. Of course, this still means that no two atomic blocks can
+execute in parallel or even concurrently. Potential solutions that
+provide a good way to implement atomic blocks are therefore
 preferable.
@@
 \item[Existing applications:] How big are the changes required to
-  integrate with and parallelize existing applications?
-\item[Better synchronization:] Does the approach enable better
-  synchronization mechanisms for applications (e.g. atomic blocks)?
+  integrate with and parallelise existing applications?
+\item[Better synchronisation:] Does the approach enable better
+  synchronisation mechanisms for applications (e.g. atomic blocks)?
 \item[Implementation:] How difficult is it to implement the approach
@@
 interpreter for Python on the JVM\footnote{Java Virtual Machine} and
-that uses fine-grained locking to correctly synchronize the
+that uses fine-grained locking to correctly synchronise the
 interpreter. For a language like Python, one needs quite a few,
@@
 In the end, fine-grained locking can transparently replace the GIL
-and therefore parallelize existing applications without any
-changes. It does however not provide a better synchronization
+and therefore parallelise existing applications without any
+changes. It does however not provide a better synchronisation
 mechanism to the application like e.g. atomic blocks.
@@
 implementation of this approach is very cheap since one can
-actually take an unmodfied GIL-supported interpreter and run
+actually take an unmodified GIL-supported interpreter and run
 multiple of them in parallel. That way, we also inherit the
@@
 While the model of explicit communication is often seen as a
-superior way to synchronize concurrent applications because
+superior way to synchronise concurrent applications because
 of its explicitness, it does not actually introduce a better
-synchronization mechanism for applications.
+synchronisation mechanism for applications.
 
@@
 generation of HTM hits this limit very often for our use case (XXX:
-cite ruby GIL paper) and therefore does not parallelize that well.
+cite ruby GIL paper) and therefore does not parallelise that well.
 
 The performance of HTM is pretty good (XXX: cite again...) as it does
-not introduce much overhead. And it can transparently parallelize
+not introduce much overhead. And it can transparently parallelise
 existing applications to some degree. The implementation is very
@@
 place. HTM is also directly compatible with any external library that
-needs to be integrated and synchronized for use in multiple
+needs to be integrated and synchronised for use in multiple
 threads. The one thing that is missing is support for a better
-synchronization mechanism for the application. It is not possible
+synchronisation mechanism for the application. It is not possible
 in general to expose the hardware-transactions to the application
@@
 form of atomic blocks. This is the only approach that enables a better
-synchronization mechanism than locks for applications \emph{and} still
-parallelizes when using it. We think this is a very important point
-because it not only gives dynamic languages the ability to parallelize
+synchronisation mechanism than locks for applications \emph{and} still
+parallelises when using it. We think this is a very important point
+because it not only gives dynamic languages the ability to parallelise
 (already commonplace in most other languages), but also pushes
@@
 \section{The Way Forward}
-possible solution:\\
-- use virtual memory paging to somehow lower the STM overhead\\
-- tight integration with GC and jit?
+
+\begin{table*}[!ht]
+  \centering
+  \begin{tabular}{|l|c|c|c|c|c|}
+    \hline
+    & \textbf{GIL} & \textbf{Fine-grained locking}
+    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
+    \hline
+    Performance (single-threaded) & ++   & +  & ++   & ++ & -{-} \\
+    \hline
+    Performance (multi-threaded)  & -{-} & +  & +    & +  & +    \\
+    \hline
+    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
+    \hline
+    Better synchronisation        & -    & -  & -    & -  & ++   \\
+    \hline
+    Implementation                & ++   & -  & ++   & ++ & ++   \\
+    \hline
+    External libra\-ries          & ++   & ++ & ++   & ++ & ++   \\
+    \hline
+  \end{tabular}
+  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
+  \label{tab:comparison}
+\end{table*}
+
+Following the above argumentation for each approach we assembled a
+general overview in Table \ref{tab:comparison}. The general picture is
+everything else than clear. It looks like HTM may be a good solution
+to replace the GIL in the future. Current implementations are however
+far too limiting and do not provide good scaling.
+
+Just allowing for parallel execution only means that dynamic languages
+catch up to all other languages that already provide real
+parallelism. This is why we think that only the STM approach is a
+viable solution in the long-term. It provides the application with a
+simple memory model (sequential consistency) and a composable way to
+synchronise memory accesses using atomic blocks.
+
+STM has a big performance problem. We believe that further work
+to reduce the overhead by closely working together with the
+hardware should be the focus of research. Hybrid approaches that
+combine STM and HTM for performance may be able to overcome this
+obstacle.
+
+%% possible solution:\\
+%% - use virtual memory paging to somehow lower the STM overhead\\
+%% - tight integration with GC and jit?
 
@@
 
-Acknowledgments, if needed.
+Acknowledgments...
 

### 2014-05-01 talk/icooolps2014/position-paper.tex
say what we are working on

@@
 The GIL also allows for easy integration with external C libraries that
-do not need to be thread-safe. For the duration of the calls, we
+may not be thread-safe. For the duration of the calls, we
 simply do not release the GIL. External libraries that are explicitly
@@
 
-TM can be implemented in software (STM) or in hardware (HTM. There are
-also some hybrid approaches that combine the two. We count these
+TM can be implemented in software (STM) or in hardware (HTM). There
+are also hybrid approaches, which combine the two. We count these
 hybrid approaches as STM, since they usually provide the same
 capabilities as software-only approaches but with different
-performance characteristics. We will now first look at HTM that
+performance characteristics. We will now first look at HTM, which
 recently gained a lot of popularity by its introduction in common
@@
 everything else than clear. It looks like HTM may be a good solution
-to replace the GIL in the future. Current implementations are however
-far too limiting and do not provide good scaling.
+to replace the GIL in the near future. Current implementations are
+however far too limiting and do not provide good scaling.
 
-Just allowing for parallel execution only means that dynamic languages
+Allowing for parallel execution just means that dynamic languages
 catch up to all other languages that already provide real
@@
 
-STM has a big performance problem. We believe that further work
-to reduce the overhead by closely working together with the
-hardware should be the focus of research. Hybrid approaches that
-combine STM and HTM for performance may be able to overcome this
-obstacle.
+Unfortunately, STM has a big performance problem. One way to approach
+this problem is to make STM systems that use the available hardware
+better. We are currently working on a STM system that makes use of
+several hardware features like virtual memory and memory segmentation.
+We further tailor the system to the discussed use case which gives us
+an advantage over other STM systems that are more general.  With this
+approach, initial results suggest that we can keep the overhead of STM
+already below 50\%. A hybrid TM system, which also uses HTM to
+accelerate certain tasks, looks like a very promising direction of
+research too. In general we believe that further work to reduce the
+overhead of STM is very worthwhile.
 

### 2014-05-02 talk/icooolps2014/position-paper.tex
minor changes

@@
 
-\paragraph{dynamic language VM problems}
-XXX:
-- high allocation rate (short lived objects)\\
-- (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)
+%% \paragraph{dynamic language VM problems}
+%% XXX:
+%% - high allocation rate (short lived objects)\\
+%% - (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)
 
@@
 The GIL is a very simple synchronisation mechanism for supporting
-multi-threading in the interpreter. The basic guarantee is that the
+multithreading in the interpreter. The basic guarantee is that the
 GIL may only be released in-between bytecode instructions. The
@@
 depending on this may not always be a good idea, it is done in
-practice. A solution replacing the GIL should therefore uphold these
+practice. A GIL-replacement should therefore uphold these
 guarantees, while preferably also be as easily implementable as a GIL
@@
 potentially long I/O operations. Consequently, I/O-bound,
-multi-threaded applications can actually parallelise to some
+multithreaded applications can actually parallelise to some
 degree. Again, a potential solution should be able to integrate with
@@
 
-\paragraph{HTM}
-
-HTM provides us with transactions like any TM system does. It can
+\paragraph{HTM} provides us with transactions like any TM system does. It can
 be used as a direct replacement for the GIL. However, as is common
@@
 
-\paragraph{STM}
-
-STM provides all the same benefits as HTM except for its performance.
+\paragraph{STM} provides all the same benefits as HTM except for its performance.
 It is not unusual for the overhead introduced by STM to be between
@@
     \hline
-    Performance (single-threaded) & ++   & +  & ++   & ++ & -{-} \\
+    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
     \hline
-    Performance (multi-threaded)  & -{-} & +  & +    & +  & +    \\
+    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
     \hline
@@
     \hline
-    External libra\-ries          & ++   & ++ & ++   & ++ & ++   \\
+    External libraries            & ++   & ++ & ++   & ++ & ++   \\
     \hline
@@
 
-Acknowledgments...
+Acknowledgements...
 

### 2014-05-02 talk/icooolps2014/position-paper.tex
additions

@@
 likeliness of deadlocks grows, as well as the overhead that acquiring
-and releasing locks produces.
+and releasing locks produces. The former means that sometimes it is
+necessary to fall back to less fine-grained locking, preventing some
+potential parallelism in order to keep the complexity manageable.
+The latter means that we lose a bit of performance in the
+single-threaded case compared to the GIL, which requires much less
+acquire-release operations.
 
@@
 
-\subsubsection{Shared-Nothing / multiple processes}
+\subsubsection{Shared-Nothing}
 

### 2014-05-02 talk/icooolps2014/position-paper.tex
start citing things

@@
 \usepackage{array}
+\usepackage{hyperref}
 \usepackage{amsmath}
@@
 
-Jython\footnote{www.jython.org} is one project that implements an
+Jython\cite{webjython} is one project that implements an
 interpreter for Python on the JVM\footnote{Java Virtual Machine} and
@@
 
-\paragraph{HTM} provides us with transactions like any TM system does. It can
-be used as a direct replacement for the GIL. However, as is common
-with hardware-only solutions, there are quite a few limitations
-that can not be lifted easily. For this comparison, we look at
-the implementation of Intel in recent Haswell generation CPUs.
+\paragraph{HTM} provides us with transactions like any TM system does.
+It can be used as a direct replacement for the GIL. However, as is
+common with hardware-only solutions, there are quite a few limitations
+that can not be lifted easily. For this comparison, we look at the
+implementation of Intel in recent Haswell generation CPUs.
 
@@
 to have a fallback in place in case this limit is reached. In recent
-attempts, the usual fallback is the GIL (XXX: cite). The current
-generation of HTM hits this limit very often for our use case (XXX:
-cite ruby GIL paper) and therefore does not parallelise that well.
+attempts, the usual fallback is the GIL\cite{odaira14}. In our
+experiments, the current generation of HTM proved to be very fragile
+and thus needing the fallback very often. Consequently, scalability
+suffered a lot from this.
 
-The performance of HTM is pretty good (XXX: cite again...) as it does
-not introduce much overhead. And it can transparently parallelise
-existing applications to some degree. The implementation is very
-straight-forward because it directly replaces the GIL in a central
-place. HTM is also directly compatible with any external library that
-needs to be integrated and synchronised for use in multiple
-threads. The one thing that is missing is support for a better
-synchronisation mechanism for the application. It is not possible
-in general to expose the hardware-transactions to the application
-in the form of atomic blocks because that would require much
-longer transactions.
+The performance of HTM is pretty good as it does not introduce much
+overhead ($<40\%$ overhead\cite{odaira14}). And it can transparently
+parallelise existing applications to some degree. The implementation
+is very straight-forward because it directly replaces the GIL in a
+central place. HTM is also directly compatible with any external
+library that needs to be integrated and synchronised for use in
+multiple threads. The one thing that is missing is support for a
+better synchronisation mechanism for the application. It is not
+possible in general to expose the hardware-transactions to the
+application in the form of atomic blocks because that would require
+much longer transactions.
 
@@
 
-\paragraph{STM} provides all the same benefits as HTM except for its performance.
-It is not unusual for the overhead introduced by STM to be between
-100\% to even 1000\%. While STM systems often scale very well to a big
-number of threads and eventually overtake the single-threaded
-execution, they often provide no benefits at all for low numbers of
-threads (1-8). There are some attempts (XXX: cite fastlane) that can
-reduce the overhead a lot, but also scale very badly so that their
-benefit on more than one thread is little.
+\paragraph{STM} provides all the same benefits as HTM except for its
+performance.  It is not unusual for the overhead introduced by STM to
+be between 100\% to even 1000\% \cite{cascaval08,drago11}. While STM
+systems often scale very well to a big number of threads and
+eventually overtake the single-threaded execution, they often provide
+no benefits at all for low numbers of threads (1-8). There are some
+attempts \cite{warmhoff13} that can reduce the overhead a lot, but
+also scale very badly so that their benefit on more than one thread is
+little.
 
@@
 
+\bibitem{webjython}
+  The Jython Project, \url{www.jython.org}
+
+\bibitem{odaira14}
+  Odaira, Rei, Jose G. Castanos, and Hisanobu Tomari.  "Eliminating
+  global interpreter locks in Ruby through hardware transactional
+  memory."  \emph{Proceedings of the 19th ACM SIGPLAN symposium on
+    Principles and practice of parallel programming.} ACM, 2014.
+
+\bibitem{warmhoff13}
+  Wamhoff, Jons-Tobias, et al. "FastLane: improving performance of
+  software transactional memory for low thread counts."
+  \emph{Proceedings of the 18th ACM SIGPLAN symposium on Principles
+    and practice of parallel programming.} ACM, 2013.
+
+\bibitem{drago11}
+  Dragojević, Aleksandar, et al. "Why STM can be more than a research
+  toy." \emph{Communications of the ACM} 54.4 (2011): 70-77.
+
+\bibitem{cascaval08}
+  Cascaval, Calin, et al. "Software transactional memory: Why is it
+  only a research toy?." \emph{Queue} 6.5 (2008): 40.
+
 \end{thebibliography}

### 2014-05-02 talk/icooolps2014/position-paper.tex
Avoid using the word 'hardware' to describe our purely-STM system.

@@
 Unfortunately, STM has a big performance problem. One way to approach
-this problem is to make STM systems that use the available hardware
-better. We are currently working on a STM system that makes use of
-several hardware features like virtual memory and memory segmentation.
+this problem is to make STM systems that make better use of low-level
+features in existing OS kernels.
+We are currently working on a STM system that makes use of
+several such features like virtual memory and memory segmentation.
 We further tailor the system to the discussed use case which gives us

### 2014-05-02 talk/icooolps2014/position-paper.tex
more random citescites...

--- 2014-05-02 talk/icooolps2014/position-paper.tex
@@
 \paragraph{HTM} provides us with transactions like any TM system does.
-It can be used as a direct replacement for the GIL. However, as is
+It can be used as a direct replacement for the GIL\cite{nicholas06,odaira14,fuad10}. However, as is
 common with hardware-only solutions, there are quite a few limitations
@@
 to have a fallback in place in case this limit is reached. In recent
-attempts, the usual fallback is the GIL\cite{odaira14}. In our
+attempts, the usual fallback is the GIL\cite{odaira14,fuad10}. In our
 experiments, the current generation of HTM proved to be very fragile
@@
 approach, initial results suggest that we can keep the overhead of STM
-already below 50\%. A hybrid TM system, which also uses HTM to
-accelerate certain tasks, looks like a very promising direction of
-research too. In general we believe that further work to reduce the
-overhead of STM is very worthwhile.
+below 50\%. A hybrid TM system, which also uses HTM to accelerate
+certain tasks, looks like a very promising direction of research
+too. In general we believe that further work to reduce the overhead of
+STM is very worthwhile.
 
@@
 
+\bibitem{nicholas06}
+  Nicholas Riley and Craig Zilles. 2006. Hardware tansactional memory
+  support for lightweight dynamic language evolution. \emph{In
+    Companion to the 21st ACM SIGPLAN symposium on Object-oriented
+    programming systems, languages, and applications} (OOPSLA
+  '06). ACM, New York, NY, USA
+
+\bibitem{fuad10}
+  Fuad Tabba. 2010. Adding concurrency in python using a commercial
+  processor's hardware transactional memory support. \emph{SIGARCH
+  Comput. Archit. News 38}, 5 (April 2010)
+
 \end{thebibliography}

### 2014-05-02 talk/pycon2012/tutorial/examples/07_itertools.py

@@ -1,0 +1,32 @@
+
+from itertools import imap, izip
+import operator
+import timeit
+
+vector1 = [i for i in range(10000)]
+vector2 = [i for i in range(10000)]
+
+def one(vector1, vector2):
+    return sum(imap(operator.add, vector1, vector2))
+
+def two(vector1, vector2):
+    return sum(x + y for x, y in izip(vector1, vector2))
+
+def three(vector1, vector2):
+    s = 0
+    for i in range(len(vector1)):
+        s += vector1[i] + vector2[i]
+    return s
+
+def four(vector1, vector2):
+    return sum(vector1[i] + vector2[i] for i in range(len(vector1)))
+
+def five(vector1, vector2):
+    return sum([vector1[i] + vector2[i] for i in range(len(vector1))])
+
+print "one", timeit.timeit(lambda : one(vector1, vector2), number=1000)
+print "two", timeit.timeit(lambda : two(vector1, vector2), number=1000)
+print "three", timeit.timeit(lambda : three(vector1, vector2), number=1000)
+print "four", timeit.timeit(lambda : four(vector1, vector2), number=1000)
+print "five", timeit.timeit(lambda : five(vector1, vector2), number=1000)
+

### 2014-05-02 talk/icooolps2014/position-paper.tex
remove example citation
typo
add something about automatic barrier placement

@@
 
+While one can argue that STM requires the insertion of read and write
+barriers in the whole program, this can be done automatically and
+locally by a program transformation\cite{felber07}. There are attempts
+to do the same for fine-grained locking\cite{bill06} but they require
+a whole program analysis since locks are inherently non-composable.
+The effectiveness of these approaches still has to be proven for our
+use case.
+
 %% - overhead (100-1000\%) (barrier reference resolution, kills performance on low \#cpu)
@@
 
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-  P. Q. Smith, and X. Y. Jones. ...reference text...
-
 \bibitem{webjython}
@@
 \bibitem{nicholas06}
-  Nicholas Riley and Craig Zilles. 2006. Hardware tansactional memory
+  Nicholas Riley and Craig Zilles. 2006. Hardware transactional memory
   support for lightweight dynamic language evolution. \emph{In
@@
 
+\bibitem{felber07}
+  Felber, Pascal, et al. "Transactifying applications using an open
+  compiler framework." \emph{TRANSACT}, August (2007): 4-6.
+
+\bibitem{bill06}
+  Bill McCloskey, Feng Zhou, David Gay, and Eric
+  Brewer. 2006. Autolocker: synchronization inference for atomic
+  sections. \emph{In Conference record of the 33rd ACM SIGPLAN-SIGACT
+  symposium on Principles of programming languages (POPL '06)}. ACM,
+  New York, NY, USA
+
 \end{thebibliography}

### 2014-05-02 talk/icooolps2014/position-paper.tex
add some text

@@
 
-Unfortunately, STM has a big performance problem. One way to approach
-this problem is to make STM systems that make better use of low-level
-features in existing OS kernels.
-We are currently working on a STM system that makes use of
-several such features like virtual memory and memory segmentation.
-We further tailor the system to the discussed use case which gives us
-an advantage over other STM systems that are more general. With this
+Unfortunately, STM has a big performance problem. Particularly, for
+our use case there is not much static information available since we
+are executing a program only known at runtime. Additionally, replacing
+the GIL means running everything in transactions, so there is not much
+code that can run outside and be optimized better.
+
+One way to get more performance is to make STM systems that make
+better use of low-level features in existing OS kernels.  We are
+currently working on a STM system that makes use of several such
+features like virtual memory and memory segmentation.  We further
+tailor the system to the discussed use case which gives us an
+advantage over other STM systems that are more general. With this
 approach, initial results suggest that we can keep the overhead of STM

### 2014-05-02 talk/icooolps2014/position-paper.tex
minor changes

--- 2014-05-02 talk/icooolps2014/position-paper.tex
@@
 
-Unfortunately, STM has a big performance problem. One way to approach
-this problem is to make STM systems that make better use of low-level
-features in existing OS kernels.
-We are currently working on a STM system that makes use of
-several such features like virtual memory and memory segmentation.
-We further tailor the system to the discussed use case which gives us
-an advantage over other STM systems that are more general. With this
+Unfortunately, STM has a big performance problem. Particularly, for
+our use case there is not much static information available since we
+are executing a program only known at runtime. Additionally, replacing
+the GIL means running every part of the application in transactions,
+so there is not much code that can run outside and that can be
+optimized better. The performance of the TM system is vital.
+
+One way to get more performance is to develop STM systems that make
+better use of low-level features in existing OS kernels.  We are
+currently working on a STM system that makes use of several such
+features like virtual memory and memory segmentation.  We further
+tailor the system to the discussed use case which gives us an
+advantage over other STM systems that are more general. With this
 approach, initial results suggest that we can keep the overhead of STM
@@
 certain tasks, looks like a very promising direction of research
-too. In general we believe that further work to reduce the overhead of
-STM is very worthwhile.
+too. We believe that further work to reduce the overhead of STM is
+very worthwhile.
 

### 2014-05-02 talk/icooolps2014/position-paper.tex
additional cite

@@
 no benefits at all for low numbers of threads (1-8). There are some
-attempts \cite{warmhoff13} that can reduce the overhead a lot, but
-also scale very badly so that their benefit on more than one thread is
-little.
+attempts \cite{warmhoff13,spear09} that can reduce the overhead a lot,
+but scale badly or only for certain workloads. Often the benefits
+on more than one thread are too little in real world applications.
 
@@
 
+\bibitem{spear09}
+  Spear, Michael F., et al. "Transactional mutex locks." \emph{SIGPLAN
+    Workshop on Transactional Computing.} 2009.
+
 \end{thebibliography}

### 2014-05-02 talk/icooolps2014/position-paper.tex
add citation

@@
 The GIL is a very simple synchronisation mechanism for supporting
-multithreading in the interpreter. The basic guarantee is that the
-GIL may only be released in-between bytecode instructions. The
-interpreter can thus rely on complete isolation and atomicity of these
+multithreading in the interpreter. The basic guarantee is that the GIL
+may only be released in-between bytecode instructions. The interpreter
+can thus rely on complete isolation and atomicity of these
 instructions. Additionally, it provides the application with a
-sequential consistency model. As a consequence, applications can rely
-on certain operations to be atomic and that they will always be
-executed in the order in which they appear in the code. While
-depending on this may not always be a good idea, it is done in
-practice. A GIL-replacement should therefore uphold these
+sequential consistency model\cite{lamport79}. As a consequence,
+applications can rely on certain operations to be atomic and that they
+will always be executed in the order in which they appear in the
+code. While depending on this may not always be a good idea, it is
+done in practice. A GIL-replacement should therefore uphold these
 guarantees, while preferably also be as easily implementable as a GIL
 for the interpreter.
-[xxx mention that the interpreter is typically very large and maintained
-by open-source communities]
+[xxx mention that the interpreter is typically
+  very large and maintained by open-source communities]
 
@@
 
-\begin{table*}[!ht]
+\begin{table*}[h]
   \centering
@@
 
-\acks
+%% \acks
 
-Acknowledgements...
+%% Acknowledgements...
 
@@
 
+\bibitem{lamport79}
+  Lamport, Leslie. "How to make a multiprocessor computer that
+  correctly executes multiprocess programs." \emph{Computers, IEEE
+    Transactions} on 100.9 (1979): 690-691.
+
 \end{thebibliography}

### 2014-05-03 talk/icooolps2014/position-paper.tex
Add a suggestion
Precision

@@
 In the end, fine-grained locking can transparently replace the GIL
-and therefore parallelise existing applications without any
-changes. It does however not provide a better synchronisation
+and therefore parallelise existing applications, generally without any
+changes\footnote{There are rare cases where not having atomic
+bytecodes actually changes the semantics.}
+It does however not provide a better synchronisation
 mechanism to the application like e.g. atomic blocks.
@@
 use case.
+[xxx or maybe: "The effectiveness of these approaches is doubtful in our
+use case --- for example, it makes it close to impossible to order the
+locks consistently or to know in advance which locks a transaction will
+need.]
 

### 2014-05-04 talk/icooolps2014/position-paper.tex
change my name, add acks

@@
 
-\authorinfo{Remi Meier}
+\authorinfo{Remigius Meier}
            {Department of Computer Science\\ ETH Zürich}
@@
 
-%% \acks
-
-%% Acknowledgements...
+\acks
+We would like to thank Maciej Fijalkowski and Carl Friedrich Bolz for
+their valuable inputs and the many fruitful discussions.
 

### 2014-05-05 talk/icooolps2014/position-paper.tex
some extensions, do the XXXs

@@
 \copyrightyear{2014}
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
+%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
 \doi{nnnnnnn.nnnnnnn}
@@
 
-\title{The Way Forward in Parallelizing Dynamic Languages}
+\title{The Way Forward in Parallelising Dynamic Languages}
 \subtitle{Position Paper, ICOOOLPS'14}
@@
 \begin{abstract}
-This is the text of the abstract.
+  Dynamic languages became very popular in recent years. At some
+  point, the need for concurrency arised, and many of them made the
+  choice to use a single global interpreter lock (GIL) to synchronize
+  the interpreter in a multithreading scenario. This choice however
+  makes it impossible to actually run code in parallel.
+
+  Here we want to compare different approaches to replacing the GIL
+  with a technology that allows parallel execution. We look at
+  fine-grained locking, shared-nothing, and transactional memory (TM)
+  approaches. We argue that software-based TM systems are the most
+  promising, especially since they also enable the introduction of
+  atomic blocks as a better synchronization mechanism in the language.
 \end{abstract}
 
-\category{CR-number}{subcategory}{third-level}
+%\category{CR-number}{subcategory}{third-level}
 
@@
 
+In this section we examine the approaches and highlight their
+advantages and disadvantages.
 %% \paragraph{dynamic language VM problems}
@@
 guarantees, while preferably also be as easily implementable as a GIL
-for the interpreter.
-[xxx mention that the interpreter is typically
-  very large and maintained by open-source communities]
+for the interpreter. The latter can be especially important since
+many of these languages are developed and maintained by very large
+open-source communities, which are not easy to coordinate.
 
@@
 a whole program analysis since locks are inherently non-composable.
-The effectiveness of these approaches still has to be proven for our
-use case.
-[xxx or maybe: "The effectiveness of these approaches is doubtful in our
-use case --- for example, it makes it close to impossible to order the
-locks consistently or to know in advance which locks a transaction will
-need.]
+The effectiveness of these approaches is doubtful in our use case,
+since we execute bytecode instructions in any order defined by a
+script only known at runtime. This makes it close to impossible to
+order locks consistently or to know in advance which locks a
+transaction will need.
 

### 2014-05-05 talk/icooolps2014/position-paper.tex
change title
a few typos

@@
 
-\title{The Way Forward in Parallelising Dynamic Languages}
+\title{A Way Forward in Parallelising Dynamic Languages}
 \subtitle{Position Paper, ICOOOLPS'14}
@@
   Dynamic languages became very popular in recent years. At some
-  point, the need for concurrency arised, and many of them made the
-  choice to use a single global interpreter lock (GIL) to synchronize
+  point, the need for concurrency arose, and many of them made the
+  choice to use a single global interpreter lock (GIL) to synchronise
   the interpreter in a multithreading scenario. This choice however
@@
   promising, especially since they also enable the introduction of
-  atomic blocks as a better synchronization mechanism in the language.
+  atomic blocks as a better synchronisation mechanism in the language.
 \end{abstract}
@@
 %% Furthermore, a solution to this problem should also bring better
-%% synchronization mechanism with it...
+%% synchronisation mechanism with it...
 
-%% (supporting (large) atomic blocks for synchronization)
+%% (supporting (large) atomic blocks for synchronisation)
 
@@
 The GIL is a very simple synchronisation mechanism for supporting
-multithreading in the interpreter. The basic guarantee is that the GIL
-may only be released in-between bytecode instructions. The interpreter
+multithreading in an interpreter. The basic guarantee is that the GIL
+may only be released in between bytecode instructions. The interpreter
 can thus rely on complete isolation and atomicity of these
@@
 so there is not much code that can run outside and that can be
-optimized better. The performance of the TM system is vital.
+optimised better. The performance of the TM system is vital.
 

### 2014-05-05 talk/icooolps2014/position-paper.tex
use a ~ (non-breaking space) before citations

@@
 instructions. Additionally, it provides the application with a
-sequential consistency model\cite{lamport79}. As a consequence,
+sequential consistency model~\cite{lamport79}. As a consequence,
 applications can rely on certain operations to be atomic and that they
@@
 
-Jython\cite{webjython} is one project that implements an
+Jython~\cite{webjython} is one project that implements an
 interpreter for Python on the JVM\footnote{Java Virtual Machine} and
@@
 \paragraph{HTM} provides us with transactions like any TM system does.
-It can be used as a direct replacement for the GIL\cite{nicholas06,odaira14,fuad10}. However, as is
+It can be used as a direct replacement for the GIL~\cite{nicholas06,odaira14,fuad10}. However, as is
 common with hardware-only solutions, there are quite a few limitations
@@
 to have a fallback in place in case this limit is reached. In recent
-attempts, the usual fallback is the GIL\cite{odaira14,fuad10}. In our
+attempts, the usual fallback is the GIL~\cite{odaira14,fuad10}. In our
 experiments, the current generation of HTM proved to be very fragile
@@
 The performance of HTM is pretty good as it does not introduce much
-overhead ($<40\%$ overhead\cite{odaira14}). And it can transparently
+overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
 parallelise existing applications to some degree. The implementation
@@
 performance.  It is not unusual for the overhead introduced by STM to
-be between 100\% to even 1000\% \cite{cascaval08,drago11}. While STM
+be between 100\% to even 1000\% ~\cite{cascaval08,drago11}. While STM
 systems often scale very well to a big number of threads and
@@
 no benefits at all for low numbers of threads (1-8). There are some
-attempts \cite{warmhoff13,spear09} that can reduce the overhead a lot,
+attempts ~\cite{warmhoff13,spear09} that can reduce the overhead a lot,
 but scale badly or only for certain workloads. Often the benefits
@@
 barriers in the whole program, this can be done automatically and
-locally by a program transformation\cite{felber07}. There are attempts
-to do the same for fine-grained locking\cite{bill06} but they require
+locally by a program transformation~\cite{felber07}. There are attempts
+to do the same for fine-grained locking~\cite{bill06} but they require
 a whole program analysis since locks are inherently non-composable.

### 2014-05-05 talk/icooolps2014/position-paper.tex
a few tweaks, a few places where more citations are needed

@@
 \usepackage{array}
+\usepackage{color}
 \usepackage{hyperref}
 \usepackage{amsmath}
+\usepackage{amssymb}
+
+\newcommand{\mynote}[2]{%
+  \textcolor{red}{%
+    \fbox{\bfseries\sffamily\scriptsize#1}%
+    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
+  }%
+}
+
+\newcommand\cfbolz[1]{\mynote{Carl Friedrich}{#1}}
 
@@
 state-of-the-art still means explicit locking everywhere. It is well
-known that using locks for synchronisation is not easy.  They are
+known that using locks for synchronisation is not easy\cfbolz{citation needed
+:-). would be cool if you could find something}.  They are
 non-composable, have overhead, may deadlock, limit scalability, and
-overall add a lot of complexity. For a better parallel programming
+overall add a lot of complexity\cfbolz{same here, really}. For a better parallel programming
 model for dynamic languages, we propose another, well-known
-synchronisation mechanism called \emph{atomic blocks}.
+synchronisation mechanism called \emph{atomic blocks}\cfbolz{and here}.
 
@@
 \item[Performance:] How well does the approach perform compared to the
-  GIL on single and multiple threads?
+  GIL on a single and on multiple threads?
 \item[Existing applications:] How big are the changes required to
@@
 
+\cfbolz{I think you should mention the commented out point below, that a lot of existing code contains latent races / deadlocks that are just not exposed in a GIL-full world}
 %% - support of atomic blocks?\\
@@
 recently gained a lot of popularity by its introduction in common
-desktop CPUs from Intel (Haswell generation).
+desktop CPUs from Intel (Haswell generation).\cfbolz{citation}
 
@@
 However, STM compared to HTM does not suffer from the same restricting
-limitations. Transactions can be arbitrarily long.  This makes it
+limitations. Transactions can in principle be arbitrarily long.  This makes it
 possible to actually expose transactions to the application in the

### 2014-05-05 talk/icooolps2014/position-paper.tex
move the table up, to force latex to put it into a sensible place
more citations

@@
 
+\begin{table*}[ht]
+  \centering
+  \begin{tabular}{|l|c|c|c|c|c|}
+    \hline
+    & \textbf{GIL} & \textbf{Fine-grained locking}
+    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
+    \hline
+    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
+    \hline
+    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
+    \hline
+    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
+    \hline
+    Better synchronisation        & o    & o  & o    & -  & ++   \\
+    \hline
+    Implementation                & ++   & -  & ++   & ++ & ++   \\
+    \hline
+    External libraries            & ++   & ++ & ++   & ++ & ++   \\
+    \hline
+  \end{tabular}
+  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
+  \label{tab:comparison}
+\end{table*}
+
 \section{Discussion}
@@
 state-of-the-art still means explicit locking everywhere. It is well
-known that using locks for synchronisation is not easy\cfbolz{citation needed
-:-). would be cool if you could find something}.  They are
-non-composable, have overhead, may deadlock, limit scalability, and
-overall add a lot of complexity\cfbolz{same here, really}. For a better parallel programming
+known that using locks for synchronisation is not
+easy~\cite{christopher10,victor11,shan08}. They are non-composable,
+have overhead, may deadlock, limit scalability, and add to the overall
+complexity of the program logic. For a better parallel programming
 model for dynamic languages, we propose another, well-known
-synchronisation mechanism called \emph{atomic blocks}\cfbolz{and here}.
+synchronisation mechanism called \emph{atomic
+  blocks}~\cite{tim03,tim05}.
 
@@
 
-\begin{table*}[h]
-  \centering
-  \begin{tabular}{|l|c|c|c|c|c|}
-    \hline
-    & \textbf{GIL} & \textbf{Fine-grained locking}
-    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
-    \hline
-    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
-    \hline
-    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
-    \hline
-    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
-    \hline
-    Better synchronisation        & -    & -  & -    & -  & ++   \\
-    \hline
-    Implementation                & ++   & -  & ++   & ++ & ++   \\
-    \hline
-    External libraries            & ++   & ++ & ++   & ++ & ++   \\
-    \hline
-  \end{tabular}
-  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
-  \label{tab:comparison}
-\end{table*}
-
 Following the above argumentation for each approach we assembled a
@@
 
+\bibitem{victor11}
+  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. A study of
+  transactional memory vs. locks in practice. In \emph{Proceedings of
+    the twenty-third annual ACM symposium on Parallelism in algorithms
+    and architectures} (SPAA '11). ACM, New York, NY, USA
+
+\bibitem{christopher10}
+  Christopher J. Rossbach, Owen S. Hofmann, and Emmett
+  Witchel. 2010. Is transactional programming actually
+  easier?. \emph{SIGPLAN} Not. 45, 5 (January 2010), 47-56.
+
+\bibitem{tim03}
+  Tim Harris and Keir Fraser. 2003. Language support for lightweight
+  transactions. \emph{In Proceedings of the 18th annual ACM SIGPLAN
+    conference on Object-oriented programing, systems, languages, and
+    applications} (OOPSLA '03).
+
+\bibitem{tim05}
+  Tim Harris, Simon Marlow, Simon Peyton-Jones, and Maurice
+  Herlihy. 2005. Composable memory transactions. \emph{In Proceedings
+    of the tenth ACM SIGPLAN symposium on Principles and practice of
+    parallel programming} (PPoPP '05).
+
+\bibitem{shan08}
+  Shan Lu, Soyeon Park, Eunsoo Seo, and Yuanyuan Zhou. 2008. Learning
+  from mistakes: a comprehensive study on real world concurrency bug
+  characteristics. \emph{SIGARCH Comput. Archit. News} 36, 1 (March 2008),
+  329-339.
+
 \end{thebibliography}

### 2014-05-05 talk/icooolps2014/position-paper.tex
another comment

@@
 
-In the end, fine-grained locking can transparently replace the GIL
-and therefore parallelise existing applications, generally without any
+In the end, fine-grained locking can transparently replace the GIL and
+therefore parallelise existing applications, generally without any
 changes\footnote{There are rare cases where not having atomic
-bytecodes actually changes the semantics.}
-It does however not provide a better synchronisation
-mechanism to the application like e.g. atomic blocks.
+  bytecodes actually changes the semantics.}. An implementation has to
+follow the GIL semantics very closely, otherwise it may expose some
+latent data races in existing applications which are just not exposed
+with a GIL. This approach does however not provide a better parallelising
+synchronisation mechanism to the application like e.g. atomic blocks.
 
-\cfbolz{I think you should mention the commented out point below, that a lot of existing code contains latent races / deadlocks that are just not exposed in a GIL-full world}
 %% - support of atomic blocks?\\

### 2014-05-05 talk/icooolps2014/position-paper.tex
another citation

@@
 recently gained a lot of popularity by its introduction in common
-desktop CPUs from Intel (Haswell generation).\cfbolz{citation}
+desktop CPUs from Intel (Haswell generation)~\cite{odaira14,leis14}.
 
@@
 
+\bibitem{leis14}
+  Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
+  Hardware Transactional Memory in Main-Memory Databases."
+  \emph{Proc. of ICDE}. 2014.
+
 \end{thebibliography}

### 2014-05-05 talk/icooolps2014/position-paper.tex
at least give HTM a 0 too
emphasize atomic block parallelisation

@@
 only solution that also provides a better synchronisation mechanism to
-the application in the form of atomic blocks.
+the application in the form of paralleliseable atomic blocks.
 
@@
     \hline
-    Better synchronisation        & o    & o  & o    & -  & ++   \\
+    Better synchronisation        & o    & o  & o    & o  & ++   \\
     \hline
@@
 synchronise memory accesses in applications using threads, the
-state-of-the-art still means explicit locking everywhere. It is well
-known that using locks for synchronisation is not
-easy~\cite{christopher10,victor11,shan08}. They are non-composable,
+state-of-the-art still means explicit locking everywhere. It is
+known that using locks for synchronisation can be hard at
+times~\cite{christopher10,victor11,shan08}. They are non-composable,
 have overhead, may deadlock, limit scalability, and add to the overall
@@
 block. Of course, this still means that no two atomic blocks can
-execute in parallel or even concurrently. Potential solutions that
-provide a good way to implement atomic blocks are therefore
-preferable.
+execute in parallel or even concurrently.  Potential solutions are
+preferable if they provide a good way to implement atomic blocks that
+are also able to be executed in parallel.
 
@@
   integrate with and parallelise existing applications?
-\item[Better synchronisation:] Does the approach enable better
+\item[Better synchronisation:] Does the approach enable better, paralleliseable
   synchronisation mechanisms for applications (e.g. atomic blocks)?
@@
 possible in general to expose the hardware-transactions to the
-application in the form of atomic blocks because that would require
+application in the form of atomic blocks, because that would require
 much longer transactions.

### 2014-05-05 talk/icooolps2014/position-paper.tex
Tweaks tweaks tweaks

@@
   promising, especially since they also enable the introduction of
-  atomic blocks as a better synchronisation mechanism in the language.
+  large, parallelisable atomic blocks as a better synchronisation
+  mechanism in the language.
 \end{abstract}
@@
 execute code in parallel on multiple cores. Furthermore, we will also
-take into account the compatibility to existing code that already uses
+take into account the compatibility with existing code that may already use
 threads for concurrency, as well as the changes that are required to
@@
 only solution that also provides a better synchronisation mechanism to
-the application in the form of paralleliseable atomic blocks.
+the application in the form of parallelisable atomic blocks.
 
@@
     \hline
-    Better synchronisation        & o    & o  & o    & o  & ++   \\
+    Better synchronisation        & o    & o  &      & o  & ++   \\
     \hline
@@
 Jython~\cite{webjython} is one project that implements an
-interpreter for Python on the JVM\footnote{Java Virtual Machine} and
-that uses fine-grained locking to correctly synchronise the
+interpreter for Python on the Java Virtual Machine (JVM) and
+that uses fine-grained locking\footnote{The performance impact of
+fine-grained locking is milder in Java than it would be in a typical piece
+of C code; see e.g.~\cite{biased}.} to correctly synchronise the
 interpreter. For a language like Python, one needs quite a few,
 carefully placed locks. Since there is no central location, the
-complexity of the implementation is quite a bit greater compared to
+complexity of the implementation is quite a bit larger compared to
 using a GIL. Integrating external, non-thread-safe libraries should
@@
 only do so via an external program like a database, then it is
-sensible to have one GIL per independent part. As an example of such
-an approach we look at the
+sensible to have one GIL per independent part. At the extreme, there
+are applications that parallelise perfectly simply by running
+independent processes; some web servers and some numeric computations
+do. We will consider here a slightly more general approach: the
 \emph{multiprocessing}\footnote{https://docs.python.org/2/library/multiprocessing.html}
@@
 
-Obviously not every application fits well into this model and its
-applicability is therefore quite limited. Performance is good as
+The model of explicit communication is sometimes seen as a superior
+way to synchronise concurrent applications because of its explicitness.
+However, not every application fits well into this model and its
+applicability is therefore limited. Performance is good as
 long as the application does not need to communicate a lot, because
@@
 actually take an unmodified GIL-supported interpreter and run
-multiple of them in parallel. That way, we also inherit the
+several of them in parallel. That way, we also inherit the
 easy integration of external libraries without any changes.
-While the model of explicit communication is often seen as a
-superior way to synchronise concurrent applications because
-of its explicitness, it does not actually introduce a better
-synchronisation mechanism for applications.
 
@@
 better synchronisation mechanism for the application. It is not
-possible in general to expose the hardware-transactions to the
-application in the form of atomic blocks, because that would require
-much longer transactions.
+reasonable in general to expose the hardware-transactions to the
+application in the form of atomic blocks, because doing so would
+require the system to support much longer transactions.
 
@@
 but scale badly or only for certain workloads. Often the benefits
-on more than one thread are too little in real world applications.
+on more than one thread are too small in real world applications.
 
@@
 
-While one can argue that STM requires the insertion of read and write
-barriers in the whole program, this can be done automatically and
+On the implementation level,
+while one can argue that STM requires the insertion of read and write
+barriers in the whole interpreter, this can be done automatically and
 locally by a program transformation~\cite{felber07}. There are attempts
 to do the same for fine-grained locking~\cite{bill06} but they require
-a whole program analysis since locks are inherently non-composable.
-The effectiveness of these approaches is doubtful in our use case,
+a whole program analysis since locks are inherently non-composable
+--- and their effectiveness is doubtful in our use case,
 since we execute bytecode instructions in any order defined by a
@@
 approach, initial results suggest that we can keep the overhead of STM
-below 50\%. A hybrid TM system, which also uses HTM to accelerate
+well below 50\%. A hybrid TM system, which also uses HTM to accelerate
 certain tasks, looks like a very promising direction of research
@@
 
+\bibitem{biased}
+  Kenneth Russell and David Detlefs. 2006. Eliminating
+  synchronization-related atomic operations with biased locking and
+  bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
+    conference on Object-oriented programing, systems, languages, and
+    applications} (OOPSLA '06).
+
 \end{thebibliography}

### 2014-05-05 talk/icooolps2014/position-paper.tex
Move the table to page 3.
add concluding sentence

@@
 
-\begin{table*}[ht]
-  \centering
-  \begin{tabular}{|l|c|c|c|c|c|}
-    \hline
-    & \textbf{GIL} & \textbf{Fine-grained locking}
-    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
-    \hline
-    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
-    \hline
-    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
-    \hline
-    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
-    \hline
-    Better synchronisation        & o    & o  &      & o  & ++   \\
-    \hline
-    Implementation                & ++   & -  & ++   & ++ & ++   \\
-    \hline
-    External libraries            & ++   & ++ & ++   & ++ & ++   \\
-    \hline
-  \end{tabular}
-  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
-  \label{tab:comparison}
-\end{table*}
-
 \section{Discussion}
@@
 are also able to be executed in parallel.
+
+\begin{table*}[ht]
+  \centering
+  \begin{tabular}{|l|c|c|c|c|c|}
+    \hline
+    & \textbf{GIL} & \textbf{Fine-grained locking}
+    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
+    \hline
+    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
+    \hline
+    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
+    \hline
+    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
+    \hline
+    Better synchronisation        & o    & o  &      & o  & ++   \\
+    \hline
+    Implementation                & ++   & -  & ++   & ++ & ++   \\
+    \hline
+    External libraries            & ++   & ++ & ++   & ++ & ++   \\
+    \hline
+  \end{tabular}
+  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
+  \label{tab:comparison}
+\end{table*}
 
@@
 
-Following the above argumentation for each approach we assembled a
+Following the above argumentation for each approach, we assembled a
 general overview in Table \ref{tab:comparison}. The general picture is
@@
 certain tasks, looks like a very promising direction of research
-too. We believe that further work to reduce the overhead of STM is
-very worthwhile.
+too.
+
+We believe that further work to reduce the overhead of STM is
+very worthwhile. In fact, considering some analogies that have been
+drawn between garbage collection and transactional memory \cite{dan07},
+we believe that it is worthwhile to focus the STM research more
+specifically on the context shown in this paper --- for use in
+implementations of high-level languages, rather than as a tool
+directly used by the programmer.
 
@@
 \softraggedright
+
+\bibitem{dan07}
+  Dan Grossman. 2007. The transactional memory / garbage collection
+  analogy. \emph{In Proceedings of the 22nd annual ACM SIGPLAN
+    conference on Object-oriented programming systems and
+    applications} (OOPSLA '07).
 

### 2014-05-05 talk/icooolps2014/position-paper.tex
~\cite

@@
 very worthwhile. In fact, considering some analogies that have been
-drawn between garbage collection and transactional memory \cite{dan07},
+drawn between garbage collection and transactional memory~\cite{dan07},
 we believe that it is worthwhile to focus the STM research more

### 2014-05-06 talk/dls2014/report.tex
add dls2014 (technical report for now)

%% LyX 2.1.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\synctex=-1
\usepackage{color}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=false,bookmarksopen=false,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}

\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\usepackage{enumitem}		% customizable list environments
\newlength{\lyxlabelwidth}      % auxiliary length

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
% IEEE standard conference template; to be used with:
%   spconf.sty  - LaTeX style file, and
%   IEEEbib.bst - IEEE bibliography style file.
% --------------------------------------------------------------------------

\usepackage{spconf}
\usepackage{multicol}

% bold paragraph titles
\newcommand{\mypar}[1]{{\bf #1.}}

% Title.
% ------
\title{C7: Fast software transactional memory for dynamic languages}
%
% Single address.
% ---------------
%\name{Markus P\"uschel\thanks{The author thanks Jelena Kovacevic. This paper
%is a modified version of the template she used in her class.}} 
%\address{Department of Computer Science\\ ETH Z\"urich\\Z\"urich, Switzerland}

% For example:
% ------------
%\address{School\\
%		 Department\\
%		 Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
\twoauthors
 		{Remigius Meier}
 		{Department of Computer Science\\
 ETH Zürich\\
 		Switzerland}
  		{Armin Rigo}
 		{www.pypy.org}

% nice listings
\usepackage{xcolor}
\usepackage{newverbs}

\usepackage{color}
\definecolor{verylightgray}{rgb}{0.93,0.93,0.93}
\definecolor{darkblue}{rgb}{0.2,0.2,0.6}
\definecolor{commentgreen}{rgb}{0.25,0.5,0.37}
\usepackage{letltxmacro}

\usepackage{listings}
\makeatletter
\LetLtxMacro{\oldlstinline}{\lstinline}

\renewcommand\lstinline[1][]{%
\Collectverb{\@@myverb}%
}

\def\@@myverb#1{%
    \begingroup
    \fboxsep=0.2em
    \colorbox{verylightgray}{\oldlstinline|#1|}%
    \endgroup
}
\makeatother

\makeatother

\usepackage{listings}
\lstset{backgroundcolor={\color{verylightgray}},
basicstyle={\scriptsize\ttfamily},
commentstyle={\ttfamily\color{commentgreen}},
keywordstyle={\bfseries\color{darkblue}},
morecomment={[l]{//}},
morekeywords={foreach,in,def,type,dynamic,Int,Boolean,infer,void,super,if,boolean,int,else,while,do,extends,class,assert,for,switch,case,private,protected,public,const,final,static,interface,new,true,false,null,return}}
\renewcommand{\lstlistingname}{Listing}

\begin{document}
%\ninept
\maketitle 
\begin{abstract}
... 
\end{abstract}

\section{Introduction}

Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
regarded as very expressive but also very slow. In recent years, the
introduction of just-in-time compilers (JIT) for these languages (e.g.
PyPy, V8, Tracemonkey) started to change this perception by delivering
good performance that enables new applications. However, a parallel
programming model was not part of the design of those languages. Thus,
the reference implementations of e.g. Python and Ruby use a single,
global interpreter lock (GIL) to serialize the execution of code in
threads.

While this GIL prevents any parallelism from occurring, it also provides
some useful guarantees. Since this lock is always acquired while executing
bytecode instructions and it may only be released in-between such
instructions, it provides perfect isolation and atomicity between
multiple threads for a series of instructions. Another technology
that can provide the same guarantees is transactional memory (TM).

There have been several attempts at replacing the GIL with TM. Using
transactions to enclose multiple bytecode instructions, we can get
the very same semantics as the GIL while possibly executing several
transactions in parallel. Furthermore, by exposing these interpreter-level
transactions to the application in the form of \emph{atomic blocks},
we give dynamic languages a new synchronization mechanism that avoids
several of the problems of locks as they are used now.

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system that
performs well even on low numbers of CPUs. It uses a novel combination
of hardware features and garbage collector (GC) integration in order
to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in Python and is then
evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a backwards
compatible, composable synchronization mechanism for threads.
\end{itemize}

\section{Background}

\subsection{Transactional Memory}

Transactional memory (TM) is a concurrency control mechanism that
comes from database systems. Using transactions, we can group a series
of instructions performing operations on memory and make them happen
atomically and in complete isolations from other transactions. \emph{Atomicity}
means that all these instructions in the transaction and their effects
seem to happen at one, undividable point in time. Other transactions
never see inconsistent state of a partially executed transaction which
is called \emph{isolation}.

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serializable}. Meaning, the outcome is equal to some sequential
execution of these transactions. Overall, this is exactly what a single
global lock guarantees while still allowing the TM system to run transactions
in parallel as an optimization.

\subsection{Python}

We implement and evaluate our system for the Python language. For
the actual implementation, we chose the PyPy interpreter because
replacing the GIL there with a TM system is just a matter of adding
a new transformation to the translation process of the interpreter.

Over the years, Python added multiple ways to provide concurrency
and parallelism to its applications. We want to highlight two of them,
namely \emph{threading }and \emph{multiprocessing}.

\emph{Threading} employs operating system (OS) threads to provide
concurrency. It is, however, limited by the GIL and thus does not
provide parallelism. At this point we should mention that it is indeed
possible to run external functions written in C instead of Python
in parallel. Our work focuses on Python itself and ignores this aspect
as it requires writing in a different language.

The second approach, \emph{multiprocessing}, uses multiple instances
of the interpreter itself and runs them in separate OS processes.
Here we actually get parallelism because we have one GIL per interpreter,
but of course we have the overhead of multiple processes / interpreters
and also need to exchange data between them explicitly and expensively.

We focus on the \emph{threading }approach. This requires us to remove
the GIL from our interpreter in order to run code in parallel on multiple
threads. One approach to this is fine-grained locking instead of a
single global lock. Jython and IronPython are implementations of
this. It requires great care in order to avoid deadlocks, which is
why we follow the TM approach that provides a \emph{direct }replacement
for the GIL. It does not require careful placing of locks in the right
spots. We will compare our work with Jython for evaluation.

\subsection{Synchronization}

It is well known that using locks to synchronize multiple threads
is hard. They are non-composable, have overhead, may deadlock, limit
scalability, and overall add a lot of complexity. For a better parallel
programming model for dynamic languages, we want to add another, well-known
synchronization mechanism: \emph{atomic blocks}.

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions. An implementation using a GIL would simply
guarantee that the GIL is not released during the execution of the
atomic block. Using TM, we have the same effect by guaranteeing that
all instructions in an atomic block are executed inside a single transaction.

STM, how atomicity \& isolation

reasons for overhead

\section{Method}

\subsection{Transactional Memory Model}

In this section, we describe the general model of our TM system. This
should clarify the general semantics using commonly used terms from
the literature.

\subsubsection{Conflict Handling}

Our conflict detection works with \emph{object granularity}. Conceptually,
it is based on \emph{read }and \emph{write sets }of transactions.
Two transactions conflict if they have accessed a common object that
is now in the write set of at least one of them.

The \emph{concurrency control }works partly \emph{optimistically}
for reading of objects, where conflicts caused by just reading an
object in transactions are detected only when the transaction that
writes the object actually commits. For write-write conflicts we are
currently \emph{pessimistic}: Only one transaction may have a certain
object in its write set at any point in time, others trying to write
to it will have to wait or abort.

We use \emph{lazy version management }to ensure that modifications
by a transaction are not visible to another transaction before the
former commits.

\subsubsection{Semantics}

As required for TM systems, we guarantee complete \emph{isolation
}and \emph{atomicity }for transactions at all times. Furthermore,
the isolation provides full \emph{opacity }to always guarantee a consistent
read set.

We support the notion of \emph{inevitable transactions }that are always
guaranteed to commit. There is always at most one such transaction
running in the system. We use this kind of transaction to provide
\emph{strong isolation} by running non-transactional code in the context
of inevitable transactions and to still provide the \emph{serializability}
of all transaction schedules.

\subsubsection{Contention Management}

When a conflict is detected, we perform some simple contention management.
First, inevitable transactions always win. Second, the older transaction
wins. Different schemes are possible.

\subsubsection{Software Transactional Memory}

Generally speaking, the system is fully implemented in software. However,
we exploit some more advanced features of current CPUs, especially
\emph{memory segmentation, virtual memory, }and the 64-bit address
space.

\subsection{Implementation}

In this section, we will present the general idea of how the TM model
is implemented. Especially the aspects of providing isolation and
atomicity, as well as conflict detection are explained. We try to
do this without going into too much detail about the implementation.
The later section \ref{sub:Low-level-Implementation} will discuss
it in more depth.

\subsubsection{Memory Segmentation}

A naive approach to providing complete isolation between threads is
to partition the virtual memory of a process into $N$ segments, one
per thread. Each segment then holds a copy of all the memory available
to the program. Thus, each thread automatically has a private copy
of every object that it can modify in complete isolation from other
threads.

To get references to objects that are valid in all threads, we will
use the object's offset inside the segment. Since all segments are
copies of each other, the \emph{Segment Offset (SO)} will point to
the private version of an object in all threads/segments. To then
translate this SO to a real virtual memory address when used inside
a thread, we need to add the thread's segment start address to the
SO. The result of this operation is called a \emph{Linear Address
(LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

To make this address translation efficient, we use the segment register
$\%gs$. When this register points to a thread's segment start address,
we can instruct the CPU to perform the above translation from a reference
of the form $\%gs{::}SO$ to the right LA on its own.

In summary, we can use a single SO to reference the same object in
all threads, and it will be translated by the CPU to a LA that always
points to the thread's private version of this object. Thereby, threads
are fully isolated from each other. However, $N$ segments require
$N$-times the memory and modifications on an object need to be propagated
to all segments.

\begin{figure*}[t]
\begin{centering}
\includegraphics[scale=0.8]{\string"segment addressing\string".pdf}
\par\end{centering}

\protect\caption{Segment Addressing\label{fig:Segment-Addressing}}

\end{figure*}

\subsubsection{Page Sharing}

In order to eliminate the prohibitive memory requirements of keeping
around $N$ segment copies, we share memory between them. The segments
are initially allocated in a single range of virtual memory by a call
to \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}.
As illustrated in Figure \ref{fig:mmap()-Page-Mapping}, \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}
creates a mapping between a range of virtual memory pages and virtual
file pages. The virtual file pages are then mapped lazily by the kernel
to real physical memory pages. The mapping generated by \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}
is initially linear but can be changed arbitrarily. Especially, we
can remap so that multiple virtual memory pages map to a single virtual
file page. This is what we use to share memory between the segments
since then we also only require one page of physical memory.

\begin{figure}[h]
\begin{centering}
\includegraphics[scale=0.8]{\string"mmap pages\string".pdf}
\par\end{centering}

\protect\caption{\texttt{mmap()} Page Mapping\label{fig:mmap()-Page-Mapping}}
\end{figure}

As illustrated in Figure \ref{fig:Page-Remapping}, in our initial
configuration (I) all segments are backed by their own range of virtual
file pages. This is the share-nothing configuration.

We then designate segment 0 to be the \emph{Sharing-Segment}. No thread
gets this segment assigned to it, it simply holds the pages shared
between all threads. So in (II), we remap all virtual pages of the
segments $>0$ to the file pages of our sharing-segment. This is the
fully-shared configuration.

During runtime, we can then privatize single pages in segments $>0$
again by remapping single pages as seen in (III).

Looking back at address translation for object references, we see
now that this is actually a two-step process. First, $\%gs{::}SO$
gets translated to different linear addresses in different threads
by the CPU. Then, depending on the current mapping of virtual pages
to file pages, these LAs can map to a single file page in the sharing-segment,
or to privatized file pages in the corresponding segments. This mapping
is also performed efficiently by the CPU and can easily be done on
every access to an object.

In summary, $\%gs{::}SO$ is translated efficiently by the CPU to
either a physical memory location which is shared between several
threads/segments, or to a location in memory private to the segment/thread.
This makes the memory segmentation model for isolation memory efficient
again.

\begin{figure}[h]
\begin{centering}
\includegraphics[width=1\columnwidth]{\string"page remapping\string".pdf}
\par\end{centering}

\protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
segment 0, fully shared memory configuration. (III) privatize single
pages.\label{fig:Page-Remapping}}
\end{figure}

\subsubsection{Isolation: Copy-On-Write}

We now use these mechanisms to provide isolation for transactions.
Using write barriers, we implement a \emph{Copy-On-Write (COW) }on
the level of pages. Starting from the initial fully-shared configuration
(Figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

To detect when to privatize pages, we use write barriers before every
write. When the barrier detects that the object is not in a private
page (or any pages that belong to the object), we remap and copy the
pages to the thread's segment. From now on, the translation of $\%gs{::}SO$
in this particular segment will resolve to the private version of
the object. Note, the SO used to reference the object does not change
during that process.

\subsubsection{Isolation: Barriers}

The job of barriers is to ensure complete isolation between transactions
and to register the objects in the read or write set. We insert read
and write barriers before reading or modifying an object except if
we statically know an object to be readable or writable already. 
\begin{description}
\item [{Read~Barrier:}] Adds the object to the read set of the current
transaction. Since our two-step address translation automatically
resolves the reference to the private version of the object on every
access anyway, this is not the job of the read barrier anymore.
\item [{Write~Barrier:}] Adds the object to the read and write set of
the current transaction and checks if all pages of the object are
private, doing COW otherwise.\\
Furthermore, we currently allow only one transaction modifying an
object at a time. To ensure this, we acquire a write lock on the object
and also eagerly check for a write-write conflict at this point. If
there is a conflict, we do some contention management to decide which
transaction has to wait or abort. Eagerly detecting this kind of conflict
is not inherent to our system, future experiments may show that we
want to lift this restriction.
\end{description}

\subsubsection{Atomicity: Commit \& Abort}

To provide atomicity for a transaction, we want to make changes visible
on commit. We also need to be able to completely abort a transaction
without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronize all threads
so that all of them are waiting in a safe point. In the committing
transaction, we go through all objects in the write set and check
if another transaction in a different segment read the same object.
Conflicts are resolved again by either the committing or the other
transaction waiting or aborting.\\
We then push all changes of modified objects in private pages to all
the pages in other segments, including the sharing-segment (segment
0).
\item [{Abort:}] On abort the transaction will forget about all the changes
it has done. All objects in the write set are reset by copying their
previous version from the sharing-segment into the private pages of
the aborting transaction.
\item [{}]~
\end{description}

\subsubsection{Summary}

We provide isolation between transactions by privatizing the pages
of the segments belonging to the threads the transactions run in.
To detect when and which pages need privatization, we use write barriers
that trigger a COW of one or several pages. Conflicts, however, are
detected on the level of objects; based on the concept of read and
write sets. Barriers before reading and writing add objects to the
corresponding set; particularly detecting write-write conflicts eagerly.
On commit, we resolve read-write conflicts and push modifications
to other segments. Aborting transactions simply undo their changes
by copying from the sharing-segment.

\subsection{Low-level Implementation\label{sub:Low-level-Implementation}}

In this section, we will provide details about the actual implementation
of the system and discuss some of the issues that we encountered.

\subsubsection{Architecture}

Our TM system is designed as a library that covers all aspects around
transactions and object management. The library consists of two parts:
(I) It provides a simple interface to starting and committing transactions,
as well as the required read and write barriers. (II) It also includes
a \emph{garbage collector (GC) }that is closely integrated with the
TM part (e.g. it shares the write barrier). The close integration
helps in order to know more about the lifetime of an object, as will
be explained in the following sections.

\subsubsection{Application Programming Interface\label{sub:Application-Programming-Interfac}}

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_start_transaction(tl, jmpbuf)
void stm_commit_transaction()
void stm_read(object_t *obj)
void stm_write(object_t *obj)
object_t *stm_allocate(ssize_t size_rounded)
STM_PUSH_ROOT(tl, obj)
STM_POP_ROOT(tl, obj)
\end{lstlisting}
\inputencoding{utf8}

\inputencoding{latin9}\lstinline!stm_start_transaction()!\inputencoding{utf8}
starts a transaction. It requires two arguments, the first being a
thread-local data structure and the second a buffer for use by \inputencoding{latin9}\lstinline!setjmp()!\inputencoding{utf8}.
\inputencoding{latin9}\lstinline!stm_commit_transaction()!\inputencoding{utf8}
tries to commit the current transaction. \inputencoding{latin9}\lstinline!stm_read()!\inputencoding{utf8},
\inputencoding{latin9}\lstinline!stm_write()!\inputencoding{utf8}
perform a read or a write barrier on an object and \inputencoding{latin9}\lstinline!stm_allocate()!\inputencoding{utf8}
allocates a new object with the specified size (must be a multiple
of 16). \inputencoding{latin9}\lstinline!STM_PUSH_ROOT()!\inputencoding{utf8}
and \inputencoding{latin9}\lstinline!STM_POP_ROOT()!\inputencoding{utf8}
push and pop objects on the shadow stack%
\footnote{A stack for pointers to GC objects that allows for precise garbage
collection. All objects on that stack are never seen as garbage and
are thus always kept alive.%
}. Objects have to be saved using this stack around calls that may
cause a GC cycle to happen, and also while there is no transaction
running. In this simplified API, only \inputencoding{latin9}\lstinline!stm_allocate()!\inputencoding{utf8}
and \inputencoding{latin9}\lstinline!stm_commit_transaction()!\inputencoding{utf8}
require saving object references.

The type \inputencoding{latin9}\lstinline!object_t!\inputencoding{utf8}
is special as it causes the compiler%
\footnote{Clang 3.5 with some patches to this address-space 256 feature%
} to make all accesses through it relative to the $\%gs$ register.
With exceptions, nearly all accesses to objects managed by the TM
system should use this type so that the CPU will translate the reference
to the right version of the object.

\subsubsection{Setup\label{sub:Setup}}

On startup, we reserve a big range of virtual memory with a call to
\inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8} and
partition this space into $N+1$ segments. We want to run $N$ threads
in parallel while segment 0 is designated as the \emph{sharing-segment
}that is never assigned to a thread.

The next step involves using \inputencoding{latin9}\lstinline!remap_file_pages()!\inputencoding{utf8},
a Linux system call, to establish the fully-shared configuration.
All pages of segments $>0$ map to the pages of the sharing-segment.

However, the layout of a segment is not uniform and we actually privatize
a few areas again right away. These areas are illustrated in Figure
\ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a segmentation
violation when accessed. We use this to detect erroneous dereferencing
of \inputencoding{latin9}\lstinline!NULL!\inputencoding{utf8} references.
All $\%gs{::}SO$ translated to linear addresses will point to NULL
pages if SO is set to \inputencoding{latin9}\lstinline!NULL!\inputencoding{utf8}.
\item [{Segment-local~data:}] Some area private to the segment that contains
segment-local information.
\item [{Read~markers:}] These are pages that store information about which
objects were read in the current transaction running in this segment.
\item [{Nursery:}] This area contains all the freshly allocated objects
(\emph{young objects}) of the current transaction. The GC uses pointer-bump
allocation in this area to allocate objects in the first generation.
\item [{Old~object~space:}] These pages are the ones that are really
shared between segments. They mostly contain old objects but also
some young ones that were too big to allocate in the nursery.
\end{description}
\begin{figure*}[t]
\begin{centering}
\includegraphics[scale=0.8]{\string"segment layout\string".pdf}
\par\end{centering}

\protect\caption{Segment Layout\label{fig:Segment-Layout}}
\end{figure*}

\subsubsection{Assigning Segments}

From the above setup it is clear that the number of segments is statically
set to some $N$. That means that at any point in time, a maximum
of $N$ threads and their transactions can be running in parallel.
To support an unlimited number of threads in applications that use
this TM system, we assign segments dynamically to threads.

At the start of a transaction, the thread it is running in acquires
a segment. It may have to wait until another thread finishes its transaction
and releases a segment. Fairness is not guaranteed yet, as we simply
assume a fair scheduling policy in the operating system when waiting
on a condition variable.

Therefore, a thread may be assigned to different segments each time
it starts a transaction. Although, we try to assign it the same segment
again if possible. And a maximum of $N$ transactions may run in parallel.

\subsubsection{Garbage Collection}

Garbage collection plays a big role in our TM system. The GC is generational
and has two generations.

The \textbf{first generation}, where objects are considered to be
\emph{young }and reside in the \emph{Nursery}, is collected by \emph{minor
collections}. These collections move the surviving objects out of
the nursery into the old object space, which can be done without stopping
other threads. This is done either if the nursery has no space left
anymore or if we are committing the current transaction. Consequently,
all objects are old and the nursery empty after a transaction commits.
Furthermore, all objects in the nursery were always created in the
current transaction. This fact is useful since we do not need to call
any barrier on this kind of objects.

To improve this situation even more, we introduce the concept of \emph{overflow
objects}. If a minor collection needs to occur during a transaction,
we empty the nursery and mark each surviving object in the old object
space with an \inputencoding{latin9}\lstinline!overflow_number!\inputencoding{utf8}
globally unique to the current transaction. That way we can still
detect in a medium-fast path inside barriers that the object still
belongs to the current transaction.

The \textbf{second generation}, where objects are considered to be
\emph{old }and never move again, is collected by \emph{major collections}.
These collections are implemented in a stop-the-world kind of way
and first force minor collections in all threads. The major goal is
to free objects in the old objects space. Furthermore, we optimistically
re-share pages that do not need to be private anymore.

As seen in the API (section \ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack }in order to provide precise garbage collection.
Any time we call a function that possibly triggers a collection, we
need to save the objects that we need afterwards on the shadow stack
using \inputencoding{latin9}\lstinline!STM_PUSH_ROOT()!\inputencoding{utf8}.
That way, they will not be freed. And in case they were young, we
get their new location in the old object space when getting them back
from the stack using \inputencoding{latin9}\lstinline!STM_POP_ROOT()!\inputencoding{utf8}.

\subsubsection{Read Barrier}

The point of the read barrier is to add the object to the read set
of the transaction. This information is needed to detect conflicts
between transactions. Usually, it also resolves an object reference
to a private copy, but since the CPU performs our address translation
on every object access efficiently, we do not need to do that in our
barrier.

To add the object to the read set, for us it is enough to mark it
as read. Since this information needs to be local to the segment,
we need to store it in private pages. The area is called \emph{read
markers }and already mentioned in section \ref{sub:Setup}. This area
can be seen as a continuous array of bytes that is indexed from the
start of the segment by an object's reference ($SO$) divided by 16
(this requires objects of at least 16 bytes in size). Instead of just
setting the byte to \inputencoding{latin9}\lstinline!true!\inputencoding{utf8}
if the corresponding object was read, we set it to a \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
belonging to the transaction, which will be incremented on each commit.
Thereby, we can avoid resetting the bytes to \inputencoding{latin9}\lstinline!false!\inputencoding{utf8}
on commit and only need to do this every 255 transactions. The whole
code for the barrier is easily optimizable for compilers as well as
perfectly predictable for CPUs:

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_read(SO):
    *(SO >> 4) = read_version
\end{lstlisting}
\inputencoding{utf8}

\subsubsection{Write Barrier}

The job of the write barrier is twofold: first, it serves as a write
barrier for the garbage collector and second, it supports copy-on-write
and adds objects to the write set of the transaction.

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \inputencoding{latin9}\lstinline!WRITE_BARRIER!\inputencoding{utf8}
in the object's header and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier transaction,
or if there was a minor collection which will add the flag again on
all objects. It is never set on freshly allocated objects that still
reside in the nursery.

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_write(SO):
	if SO->flags & WRITE_BARRIER:
		write_slowpath(SO)
\end{lstlisting}
\inputencoding{utf8}

The \textbf{slow path} is shown here:

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void write_slowpath(SO):
	// GC part:
	list_append(to_trace, SO)
	if is_overflow_obj(SO):
		SO->flags &= ~WRITE_BARRIER
		return
	// STM part
	stm_read(SO)
	lock_idx = SO >> 4
  retry:
	if write_locks[lock_idx] == our_num:
		// we already own it
	else if write_locks[lock_idx] == 0:
		if cmp_and_swap(&write_locks[lock_idx],
					    0, our_num):
			list_append(modified_old_objects, SO)
			privatize_pages(SO)
		else:
			goto retry
	else:
		w_w_contention_management()
		goto retry
	SO->flags &= ~WRITE_BARRIER
\end{lstlisting}
\inputencoding{utf8}

First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\inputencoding{latin9}\lstinline!to_trace!\inputencoding{utf8}).
This is necessary in case we write a reference to it that points to
a young object. We then need to trace it during the next minor collection
in order to mark the young object alive and to update its reference
to the new location it gets moved to. The check for \inputencoding{latin9}\lstinline!is_overflow_obj()!\inputencoding{utf8}
tells us if the object was actually created in this transaction. In
that case, we do not need to execute the following \emph{TM part}.
We especially do not need to privatize the page since no other transaction
knows about these ``old'' objects.

For TM, we first perform a read barrier on the object. We then try
to acquire its write lock. \inputencoding{latin9}\lstinline!write_locks!\inputencoding{utf8}
again is a simple global array of bytes that is indexed with the SO
of the object divided by 16. If we already own the lock, we are done.
If someone else owns the lock, we will do a write-write contention
management that will abort either us or the current owner of the object.
If we succeed in acquiring the lock using an atomic \inputencoding{latin9}\lstinline!cmp_and_swap!\inputencoding{utf8},
we need to add the object to the write set (a simple list called \inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
and privatize all pages belonging to it (copy-on-write).

In all cases, we remove the \inputencoding{latin9}\lstinline!WRITE_BARRIER!\inputencoding{utf8}
flag from the object before we return. Thus, we never trigger the
slow path again before we do the next minor collection (also part
of a commit) or we start the next transaction.

\subsubsection{Abort}

Aborting a transaction is rather easy. The first step is to reset
the nursery and all associated data structures. The second step is
to go over all objects in the write set (\inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
and reset any modifications in our private pages by copying from the
sharing-segment. What is left is to use \inputencoding{latin9}\lstinline!longjmp()!\inputencoding{utf8}
to jump back to the location initialized by a \inputencoding{latin9}\lstinline!setjmp()!\inputencoding{utf8}
in \inputencoding{latin9}\lstinline!stm_start_transaction()!\inputencoding{utf8}.
Increasing the \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
is also done there.

\subsubsection{Commit}

Committing a transaction needs a bit more work. First, we synchronize
all threads so that the committing one is the only one running and
all the others are waiting in a safe point. We then go through the
write set (\inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
and check the corresponding \inputencoding{latin9}\lstinline!read_markers!\inputencoding{utf8}
in other threads/segments. If we detect a read-write conflict, we
do contention management to either abort us or the other transaction,
or to simply wait a bit.

After verifying that there are no conflicts anymore, we copy all our
changes done to the objects in the write set to all other segments,
including the sharing-segment. This is safe since we synchronized
all threads. We also need to push overflow objects generated by minor
collections to other segments, since they may reside partially in
private pages. At that point we also get a new \inputencoding{latin9}\lstinline!overflow_number!\inputencoding{utf8}
by increasing a global one, so that it stays globally unique for each
transaction. Increasing the \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
is then done at the start of a new transaction.

\subsubsection{Thread Synchronization}

A requirement for performing a commit is to synchronize all threads
so that we can safely update objects in other segments. To make this
synchronization fast and cheap, we do not want to insert an additional
check regularly in order to see if synchronization is requested. We
use a trick relying on the fact that dynamic languages are usually
very high-level and thus allocate a lot of objects very regularly.
This is done through the function \inputencoding{latin9}\lstinline!stm_allocate!\inputencoding{utf8}
shown below:

\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
object_t *stm_allocate(ssize_t size_rounded):
    result = nursery_current
	nursery_current += size_rounded
	if nursery_current > nursery_end:
		return allocate_slowpath(size_rounded)
	return result
\end{lstlisting}
\inputencoding{utf8}

This code does simple pointer-bump allocation in the nursery. If there
is still space left in the nursery, we return \inputencoding{latin9}\lstinline!nursery_current!\inputencoding{utf8}
and bump it up by \inputencoding{latin9}\lstinline!size_rounded!\inputencoding{utf8}.
The interesting part is the check \inputencoding{latin9}\lstinline!nursery_current > nursery_end!\inputencoding{utf8}
which will trigger the slow path of the function to possibly perform
a minor collection in order to free up space in the nursery.

If we want to synchronize all threads, we can rely on this check being
performed regularly. So what we do is to set the \inputencoding{latin9}\lstinline!nursery_end!\inputencoding{utf8}
to $0$ in all segments that we want to synchronize. The mentioned
check will then fail in those segments and call the slow path. In
\inputencoding{latin9}\lstinline!allocate_slowpath!\inputencoding{utf8}
they can simply check for this condition and enter a safe point.

For other synchronization requirements, for example:
\begin{itemize}[noitemsep]
\item waiting for a segment to be released,
\item waiting for a transaction to abort or commit,
\item waiting for all threads to reach their safe points,
\end{itemize}
we use a set of condition variables to wait or signal other threads.

\subsubsection{Contention Management}

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
\item to support \emph{inevitable} transactions, we always prefer them to
others since they cannot abort
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and another
transaction already read the object. We can then signal the other
transaction to commit as soon as possible and wait. After waiting,
there is now no conflict between our write and the already committed
read anymore.

\textbf{}

\section{Experimental Results}

compare some programs between
\begin{itemize}[noitemsep]
\item pypy
\item pypy-jit
\item pypy-stm
\item pypy-stm-jit
\item cpython
\item jython
\end{itemize}

\section{Related Work}

\section{Conclusions}

\section{Further comments}

Some citation: \cite{Higham:98}

\bibliographystyle{IEEEbib}
\bibliography{bibl_conf}

\end{document}

### 2014-05-06 talk/dls2014/report.tex
fix some formatting

@@
 
+\newcommand{\mynote}[2]{%
+  \textcolor{red}{%
+    \fbox{\bfseries\sffamily\scriptsize#1}%
+    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
+  }%
+}
+\newcommand\remi[1]{\mynote{Remi}{#1}}
+
 % Title.
@@
     {Department of Computer Science\\
- ETH Zürich\\
- 		Switzerland}
+      ETH Zürich, Switzerland\\
+      \nolinkurl{remi.meier@inf.ethz.ch}}
     {Armin Rigo}
- 		{www.pypy.org}
+    {www.pypy.org\\
+      \nolinkurl{arigo@tunes.org}}
 
@@
 
-\makeatother
+\usepackage{listings}
 
-\usepackage{listings}
 \lstset{backgroundcolor={\color{verylightgray}},
@@
   morecomment={[l]{//}},
-morekeywords={foreach,in,def,type,dynamic,Int,Boolean,infer,void,super,if,boolean,int,else,while,do,extends,class,assert,for,switch,case,private,protected,public,const,final,static,interface,new,true,false,null,return}}
+  morekeywords={foreach,in,def,type,dynamic,Int,
+    Boolean,infer,void,super,if,boolean,int,else,
+    while,do,extends,class,assert,for,switch,case,
+    private,protected,public,const,final,static,
+    interface,new,true,false,null,return}}
 \renewcommand{\lstlistingname}{Listing}
@@
 \begin{document}
-%\ninept
+\ninept
 \maketitle
@@
 
-While this GIL prevents any parallelism from occurring, it also provides
-some useful guarantees. Since this lock is always acquired while executing
-bytecode instructions and it may only be released in-between such
-instructions, it provides perfect isolation and atomicity between
-multiple threads for a series of instructions. Another technology
-that can provide the same guarantees is transactional memory (TM).
+While this GIL prevents any parallelism from occurring, it also
+provides some useful guarantees. Since this lock is always acquired
+while executing bytecode instructions and it may only be released
+in-between such instructions, it provides perfect isolation and
+atomicity between multiple threads for a series of
+instructions. Another technology that can provide the same guarantees
+is transactional memory (TM).
 
 There have been several attempts at replacing the GIL with TM. Using
-transactions to enclose multiple bytecode instructions, we can get
-the very same semantics as the GIL while possibly executing several
-transactions in parallel. Furthermore, by exposing these interpreter-level
-transactions to the application in the form of \emph{atomic blocks},
-we give dynamic languages a new synchronization mechanism that avoids
-several of the problems of locks as they are used now.
+transactions to enclose multiple bytecode instructions, we can get the
+very same semantics as the GIL while possibly executing several
+transactions in parallel. Furthermore, by exposing these
+interpreter-level transactions to the application in the form of
+\emph{atomic blocks}, we give dynamic languages a new synchronization
+mechanism that avoids several of the problems of locks as they are
+used now.
 
@@
 \begin{itemize}[noitemsep]
-\item We introduce a new software transactional memory (STM) system that
-performs well even on low numbers of CPUs. It uses a novel combination
-of hardware features and garbage collector (GC) integration in order
-to keep the overhead of STM very low.
-\item This new STM system is used to replace the GIL in Python and is then
-evaluated extensively.
-\item We introduce atomic blocks to the Python language to provide a backwards
-compatible, composable synchronization mechanism for threads.
+\item We introduce a new software transactional memory (STM) system
+  that performs well even on low numbers of CPUs. It uses a novel
+  combination of hardware features and garbage collector (GC)
+  integration in order to keep the overhead of STM very low.
+\item This new STM system is used to replace the GIL in Python and is
+  then evaluated extensively.
+\item We introduce atomic blocks to the Python language to provide a
+  backwards compatible, composable synchronization mechanism for
+  threads.
 \end{itemize}
@@
 of instructions performing operations on memory and make them happen
-atomically and in complete isolations from other transactions. \emph{Atomicity}
-means that all these instructions in the transaction and their effects
-seem to happen at one, undividable point in time. Other transactions
-never see inconsistent state of a partially executed transaction which
-is called \emph{isolation}.
+atomically and in complete isolations from other
+transactions. \emph{Atomicity} means that all these instructions in
+the transaction and their effects seem to happen at one, undividable
+point in time. Other transactions never see inconsistent state of a
+partially executed transaction which is called \emph{isolation}.
 
@@
 \emph{serializable}. Meaning, the outcome is equal to some sequential
-execution of these transactions. Overall, this is exactly what a single
-global lock guarantees while still allowing the TM system to run transactions
-in parallel as an optimization.
+execution of these transactions. Overall, this is exactly what a
+single global lock guarantees while still allowing the TM system to
+run transactions in parallel as an optimization.
 
@@
 
-We implement and evaluate our system for the Python language. For
-the actual implementation, we chose the PyPy interpreter because
-replacing the GIL there with a TM system is just a matter of adding
-a new transformation to the translation process of the interpreter.
+We implement and evaluate our system for the Python language. For the
+actual implementation, we chose the PyPy interpreter because replacing
+the GIL there with a TM system is just a matter of adding a new
+transformation to the translation process of the interpreter.
 
-Over the years, Python added multiple ways to provide concurrency
-and parallelism to its applications. We want to highlight two of them,
+Over the years, Python added multiple ways to provide concurrency and
+parallelism to its applications. We want to highlight two of them,
 namely \emph{threading }and \emph{multiprocessing}.
@@
 provide parallelism. At this point we should mention that it is indeed
-possible to run external functions written in C instead of Python
-in parallel. Our work focuses on Python itself and ignores this aspect
-as it requires writing in a different language.
+possible to run external functions written in C instead of Python in
+parallel. Our work focuses on Python itself and ignores this aspect as
+it requires writing in a different language.
 
@@
 of the interpreter itself and runs them in separate OS processes.
-Here we actually get parallelism because we have one GIL per interpreter,
-but of course we have the overhead of multiple processes / interpreters
-and also need to exchange data between them explicitly and expensively.
+Here we actually get parallelism because we have one GIL per
+interpreter, but of course we have the overhead of multiple processes
+/ interpreters and also need to exchange data between them explicitly
+and expensively.
 
 We focus on the \emph{threading }approach. This requires us to remove
-the GIL from our interpreter in order to run code in parallel on multiple
-threads. One approach to this is fine-grained locking instead of a
-single global lock. Jython and IronPython are implementations of
-this. It requires great care in order to avoid deadlocks, which is
-why we follow the TM approach that provides a \emph{direct }replacement
+the GIL from our interpreter in order to run code in parallel on
+multiple threads. One approach to this is fine-grained locking instead
+of a single global lock. Jython and IronPython are implementations of
+this. It requires great care in order to avoid deadlocks, which is why
+we follow the TM approach that provides a \emph{direct }replacement
 for the GIL. It does not require careful placing of locks in the right
@@
 
-It is well known that using locks to synchronize multiple threads
-is hard. They are non-composable, have overhead, may deadlock, limit
-scalability, and overall add a lot of complexity. For a better parallel
-programming model for dynamic languages, we want to add another, well-known
-synchronization mechanism: \emph{atomic blocks}.
+It is well known that using locks to synchronize multiple threads is
+hard. They are non-composable, have overhead, may deadlock, limit
+scalability, and overall add a lot of complexity. For a better
+parallel programming model for dynamic languages, we want to add
+another, well-known synchronization mechanism: \emph{atomic blocks}.
 
@@
 atomic block. Using TM, we have the same effect by guaranteeing that
-all instructions in an atomic block are executed inside a single transaction.
+all instructions in an atomic block are executed inside a single
+transaction.
 
-STM, how atomicity \& isolation
-
-reasons for overhead
+\remi{STM, how atomicity \& isolation; reasons for overhead}
 
@@
 
-Our conflict detection works with \emph{object granularity}. Conceptually,
-it is based on \emph{read }and \emph{write sets }of transactions.
-Two transactions conflict if they have accessed a common object that
-is now in the write set of at least one of them.
+Our conflict detection works with \emph{object
+  granularity}. Conceptually, it is based on \emph{read }and
+\emph{write sets }of transactions.  Two transactions conflict if they
+have accessed a common object that is now in the write set of at least
+one of them.
 
-The \emph{concurrency control }works partly \emph{optimistically}
-for reading of objects, where conflicts caused by just reading an
-object in transactions are detected only when the transaction that
-writes the object actually commits. For write-write conflicts we are
-currently \emph{pessimistic}: Only one transaction may have a certain
-object in its write set at any point in time, others trying to write
-to it will have to wait or abort.
+The \emph{concurrency control }works partly \emph{optimistically} for
+reading of objects, where conflicts caused by just reading an object
+in transactions are detected only when the transaction that writes the
+object actually commits. For write-write conflicts we are currently
+\emph{pessimistic}: Only one transaction may have a certain object in
+its write set at any point in time, others trying to write to it will
+have to wait or abort.
 
-We use \emph{lazy version management }to ensure that modifications
-by a transaction are not visible to another transaction before the
-former commits.
+We use \emph{lazy version management }to ensure that modifications by
+a transaction are not visible to another transaction before the former
+commits.
 
@@
 
-When a conflict is detected, we perform some simple contention management.
-First, inevitable transactions always win. Second, the older transaction
-wins. Different schemes are possible.
+When a conflict is detected, we perform some simple contention
+management.  First, inevitable transactions always win. Second, the
+older transaction wins. Different schemes are possible.
 
@@
 
-Generally speaking, the system is fully implemented in software. However,
-we exploit some more advanced features of current CPUs, especially
-\emph{memory segmentation, virtual memory, }and the 64-bit address
-space.
+Generally speaking, the system is fully implemented in
+software. However, we exploit some more advanced features of current
+CPUs, especially \emph{memory segmentation, virtual memory, }and the
+64-bit address space.
 
@@
 is implemented. Especially the aspects of providing isolation and
-atomicity, as well as conflict detection are explained. We try to
-do this without going into too much detail about the implementation.
-The later section \ref{sub:Low-level-Implementation} will discuss
-it in more depth.
+atomicity, as well as conflict detection are explained. We try to do
+this without going into too much detail about the implementation.  The
+later section \ref{sub:Low-level-Implementation} will discuss it in
+more depth.
 
@@
 
-A naive approach to providing complete isolation between threads is
-to partition the virtual memory of a process into $N$ segments, one
-per thread. Each segment then holds a copy of all the memory available
-to the program. Thus, each thread automatically has a private copy
-of every object that it can modify in complete isolation from other
+A naive approach to providing complete isolation between threads is to
+partition the virtual memory of a process into $N$ segments, one per
+thread. Each segment then holds a copy of all the memory available to
+the program. Thus, each thread automatically has a private copy of
+every object that it can modify in complete isolation from other
 threads.
@@
 use the object's offset inside the segment. Since all segments are
-copies of each other, the \emph{Segment Offset (SO)} will point to
-the private version of an object in all threads/segments. To then
-translate this SO to a real virtual memory address when used inside
-a thread, we need to add the thread's segment start address to the
+copies of each other, the \emph{Segment Offset (SO)} will point to the
+private version of an object in all threads/segments. To then
+translate this SO to a real virtual memory address when used inside a
+thread, we need to add the thread's segment start address to the
 SO. The result of this operation is called a \emph{Linear Address
@@
 
-To make this address translation efficient, we use the segment register
-$\%gs$. When this register points to a thread's segment start address,
-we can instruct the CPU to perform the above translation from a reference
-of the form $\%gs{::}SO$ to the right LA on its own.
+To make this address translation efficient, we use the segment
+register $\%gs$. When this register points to a thread's segment start
+address, we can instruct the CPU to perform the above translation from
+a reference of the form $\%gs{::}SO$ to the right LA on its own.
 
-In summary, we can use a single SO to reference the same object in
-all threads, and it will be translated by the CPU to a LA that always
-points to the thread's private version of this object. Thereby, threads
-are fully isolated from each other. However, $N$ segments require
-$N$-times the memory and modifications on an object need to be propagated
-to all segments.
+In summary, we can use a single SO to reference the same object in all
+threads, and it will be translated by the CPU to a LA that always
+points to the thread's private version of this object. Thereby,
+threads are fully isolated from each other. However, $N$ segments
+require $N$-times the memory and modifications on an object need to be
+propagated to all segments.
 
@@
     \protect\caption{Segment Addressing\label{fig:Segment-Addressing}}
-
 \end{figure*}
@@
 are initially allocated in a single range of virtual memory by a call
-to \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}.
-As illustrated in Figure \ref{fig:mmap()-Page-Mapping}, \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}
-creates a mapping between a range of virtual memory pages and virtual
-file pages. The virtual file pages are then mapped lazily by the kernel
-to real physical memory pages. The mapping generated by \inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8}
-is initially linear but can be changed arbitrarily. Especially, we
-can remap so that multiple virtual memory pages map to a single virtual
+to \lstinline!mmap()!.  As illustrated in Figure
+\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
+between a range of virtual memory pages and virtual file pages. The
+virtual file pages are then mapped lazily by the kernel to real
+physical memory pages. The mapping generated by \lstinline!mmap()! is
+initially linear but can be changed arbitrarily. Especially, we can
+remap so that multiple virtual memory pages map to a single virtual
 file page. This is what we use to share memory between the segments
@@
 As illustrated in Figure \ref{fig:Page-Remapping}, in our initial
-configuration (I) all segments are backed by their own range of virtual
-file pages. This is the share-nothing configuration.
+configuration (I) all segments are backed by their own range of
+virtual file pages. This is the share-nothing configuration.
 
-We then designate segment 0 to be the \emph{Sharing-Segment}. No thread
-gets this segment assigned to it, it simply holds the pages shared
-between all threads. So in (II), we remap all virtual pages of the
-segments $>0$ to the file pages of our sharing-segment. This is the
-fully-shared configuration.
+We then designate segment 0 to be the \emph{Sharing-Segment}. No
+thread gets this segment assigned to it, it simply holds the pages
+shared between all threads. So in (II), we remap all virtual pages of
+the segments $>0$ to the file pages of our sharing-segment. This is
+the fully-shared configuration.
 
@@
 
-Looking back at address translation for object references, we see
-now that this is actually a two-step process. First, $\%gs{::}SO$
-gets translated to different linear addresses in different threads
-by the CPU. Then, depending on the current mapping of virtual pages
-to file pages, these LAs can map to a single file page in the sharing-segment,
-or to privatized file pages in the corresponding segments. This mapping
-is also performed efficiently by the CPU and can easily be done on
-every access to an object.
+Looking back at address translation for object references, we see now
+that this is actually a two-step process. First, $\%gs{::}SO$ gets
+translated to different linear addresses in different threads by the
+CPU. Then, depending on the current mapping of virtual pages to file
+pages, these LAs can map to a single file page in the sharing-segment,
+or to privatized file pages in the corresponding segments. This
+mapping is also performed efficiently by the CPU and can easily be
+done on every access to an object.
 
@@
 either a physical memory location which is shared between several
-threads/segments, or to a location in memory private to the segment/thread.
-This makes the memory segmentation model for isolation memory efficient
-again.
+threads/segments, or to a location in memory private to the
+segment/thread.  This makes the memory segmentation model for
+isolation memory efficient again.
 
@@
 We now use these mechanisms to provide isolation for transactions.
-Using write barriers, we implement a \emph{Copy-On-Write (COW) }on
-the level of pages. Starting from the initial fully-shared configuration
+Using write barriers, we implement a \emph{Copy-On-Write (COW) }on the
+level of pages. Starting from the initial fully-shared configuration
 (Figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
@@
 page (or any pages that belong to the object), we remap and copy the
-pages to the thread's segment. From now on, the translation of $\%gs{::}SO$
-in this particular segment will resolve to the private version of
-the object. Note, the SO used to reference the object does not change
-during that process.
+pages to the thread's segment. From now on, the translation of
+$\%gs{::}SO$ in this particular segment will resolve to the private
+version of the object. Note, the SO used to reference the object does
+not change during that process.
 
@@
 
-To provide atomicity for a transaction, we want to make changes visible
-on commit. We also need to be able to completely abort a transaction
-without a trace, like it never happened.
+To provide atomicity for a transaction, we want to make changes
+visible on commit. We also need to be able to completely abort a
+transaction without a trace, like it never happened.
 \begin{description}
@@
   so that all of them are waiting in a safe point. In the committing
-transaction, we go through all objects in the write set and check
-if another transaction in a different segment read the same object.
+  transaction, we go through all objects in the write set and check if
+  another transaction in a different segment read the same object.
   Conflicts are resolved again by either the committing or the other
@@
   0).
-\item [{Abort:}] On abort the transaction will forget about all the changes
-it has done. All objects in the write set are reset by copying their
-previous version from the sharing-segment into the private pages of
-the aborting transaction.
-\item [{}]~
+\item [{Abort:}] On abort the transaction will forget about all the
+  changes it has done. All objects in the write set are reset by
+  copying their previous version from the sharing-segment into the
+  private pages of the aborting transaction.
 \end{description}
@@
 
-We provide isolation between transactions by privatizing the pages
-of the segments belonging to the threads the transactions run in.
-To detect when and which pages need privatization, we use write barriers
+We provide isolation between transactions by privatizing the pages of
+the segments belonging to the threads the transactions run in.  To
+detect when and which pages need privatization, we use write barriers
 that trigger a COW of one or several pages. Conflicts, however, are
@@
 write sets. Barriers before reading and writing add objects to the
-corresponding set; particularly detecting write-write conflicts eagerly.
-On commit, we resolve read-write conflicts and push modifications
-to other segments. Aborting transactions simply undo their changes
-by copying from the sharing-segment.
+corresponding set; particularly detecting write-write conflicts
+eagerly.  On commit, we resolve read-write conflicts and push
+modifications to other segments. Aborting transactions simply undo
+their changes by copying from the sharing-segment.
 
@@
 
-In this section, we will provide details about the actual implementation
-of the system and discuss some of the issues that we encountered.
+In this section, we will provide details about the actual
+implementation of the system and discuss some of the issues that we
+encountered.
 
@@
 transactions and object management. The library consists of two parts:
-(I) It provides a simple interface to starting and committing transactions,
-as well as the required read and write barriers. (II) It also includes
-a \emph{garbage collector (GC) }that is closely integrated with the
-TM part (e.g. it shares the write barrier). The close integration
-helps in order to know more about the lifetime of an object, as will
-be explained in the following sections.
+(I) It provides a simple interface to starting and committing
+transactions, as well as the required read and write barriers. (II) It
+also includes a \emph{garbage collector (GC) }that is closely
+integrated with the TM part (e.g. it shares the write barrier). The
+close integration helps in order to know more about the lifetime of an
+object, as will be explained in the following sections.
 
@@
 
-\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
   void stm_start_transaction(tl, jmpbuf)
@@
 \end{lstlisting}
-\inputencoding{utf8}
 
-\inputencoding{latin9}\lstinline!stm_start_transaction()!\inputencoding{utf8}
-starts a transaction. It requires two arguments, the first being a
-thread-local data structure and the second a buffer for use by \inputencoding{latin9}\lstinline!setjmp()!\inputencoding{utf8}.
-\inputencoding{latin9}\lstinline!stm_commit_transaction()!\inputencoding{utf8}
-tries to commit the current transaction. \inputencoding{latin9}\lstinline!stm_read()!\inputencoding{utf8},
-\inputencoding{latin9}\lstinline!stm_write()!\inputencoding{utf8}
-perform a read or a write barrier on an object and \inputencoding{latin9}\lstinline!stm_allocate()!\inputencoding{utf8}
-allocates a new object with the specified size (must be a multiple
-of 16). \inputencoding{latin9}\lstinline!STM_PUSH_ROOT()!\inputencoding{utf8}
-and \inputencoding{latin9}\lstinline!STM_POP_ROOT()!\inputencoding{utf8}
-push and pop objects on the shadow stack%
-\footnote{A stack for pointers to GC objects that allows for precise garbage
-collection. All objects on that stack are never seen as garbage and
-are thus always kept alive.%
-}. Objects have to be saved using this stack around calls that may
-cause a GC cycle to happen, and also while there is no transaction
-running. In this simplified API, only \inputencoding{latin9}\lstinline!stm_allocate()!\inputencoding{utf8}
-and \inputencoding{latin9}\lstinline!stm_commit_transaction()!\inputencoding{utf8}
+\lstinline!stm_start_transaction()!  starts a transaction. It requires
+two arguments, the first being a thread-local data structure and the
+second a buffer for use by \lstinline!setjmp()!.
+\lstinline!stm_commit_transaction()!  tries to commit the current
+transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
+a read or a write barrier on an object and \lstinline!stm_allocate()!
+allocates a new object with the specified size (must be a multiple of
+16). \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
+and pop objects on the shadow stack~\footnote{A stack for pointers to
+  GC objects that allows for precise garbage collection. All objects
+  on that stack are never seen as garbage and are thus always kept
+  alive.}.  Objects have to be saved using this stack around calls
+that may cause a GC cycle to happen, and also while there is no
+transaction running. In this simplified API, only
+\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
 require saving object references.
 
-The type \inputencoding{latin9}\lstinline!object_t!\inputencoding{utf8}
-is special as it causes the compiler%
-\footnote{Clang 3.5 with some patches to this address-space 256 feature%
-} to make all accesses through it relative to the $\%gs$ register.
-With exceptions, nearly all accesses to objects managed by the TM
-system should use this type so that the CPU will translate the reference
-to the right version of the object.
+The type \lstinline!object_t!  is special as it causes the
+compiler~\footnote{Clang 3.5 with some patches to this address-space
+  256 feature} to make all accesses through it relative to the $\%gs$
+register.  With exceptions, nearly all accesses to objects managed by
+the TM system should use this type so that the CPU will translate the
+reference to the right version of the object.
 
@@
 On startup, we reserve a big range of virtual memory with a call to
-\inputencoding{latin9}\lstinline!mmap()!\inputencoding{utf8} and
-partition this space into $N+1$ segments. We want to run $N$ threads
-in parallel while segment 0 is designated as the \emph{sharing-segment
-}that is never assigned to a thread.
+\lstinline!mmap()! and partition this space into $N+1$ segments. We
+want to run $N$ threads in parallel while segment 0 is designated as
+the \emph{sharing-segment }that is never assigned to a thread.
 
-The next step involves using \inputencoding{latin9}\lstinline!remap_file_pages()!\inputencoding{utf8},
-a Linux system call, to establish the fully-shared configuration.
-All pages of segments $>0$ map to the pages of the sharing-segment.
+The next step involves using \lstinline!remap_file_pages()!, a Linux
+system call, to establish the fully-shared configuration.  All pages
+of segments $>0$ map to the pages of the sharing-segment.
 
-However, the layout of a segment is not uniform and we actually privatize
-a few areas again right away. These areas are illustrated in Figure
-\ref{fig:Segment-Layout} and explained here:
+However, the layout of a segment is not uniform and we actually
+privatize a few areas again right away. These areas are illustrated in
+Figure \ref{fig:Segment-Layout} and explained here:
 \begin{description}[noitemsep]
-\item [{NULL~page:}] This page is unmapped and will produce a segmentation
-violation when accessed. We use this to detect erroneous dereferencing
-of \inputencoding{latin9}\lstinline!NULL!\inputencoding{utf8} references.
-All $\%gs{::}SO$ translated to linear addresses will point to NULL
-pages if SO is set to \inputencoding{latin9}\lstinline!NULL!\inputencoding{utf8}.
-\item [{Segment-local~data:}] Some area private to the segment that contains
-segment-local information.
-\item [{Read~markers:}] These are pages that store information about which
-objects were read in the current transaction running in this segment.
-\item [{Nursery:}] This area contains all the freshly allocated objects
-(\emph{young objects}) of the current transaction. The GC uses pointer-bump
-allocation in this area to allocate objects in the first generation.
+\item [{NULL~page:}] This page is unmapped and will produce a
+  segmentation violation when accessed. We use this to detect
+  erroneous dereferencing of \lstinline!NULL! references.  All
+  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
+  if SO is set to \lstinline!NULL!.
+\item [{Segment-local~data:}] Some area private to the segment that
+  contains segment-local information.
+\item [{Read~markers:}] These are pages that store information about
+  which objects were read in the current transaction running in this
+  segment.
+\item [{Nursery:}] This area contains all the freshly allocated
+  objects (\emph{young objects}) of the current transaction. The GC
+  uses pointer-bump allocation in this area to allocate objects in the
+  first generation.
 \item [{Old~object~space:}] These pages are the ones that are really
@@
 \end{description}
+
 \begin{figure*}[t]
@@
 
-From the above setup it is clear that the number of segments is statically
-set to some $N$. That means that at any point in time, a maximum
-of $N$ threads and their transactions can be running in parallel.
-To support an unlimited number of threads in applications that use
-this TM system, we assign segments dynamically to threads.
+From the above setup it is clear that the number of segments is
+statically set to some $N$. That means that at any point in time, a
+maximum of $N$ threads and their transactions can be running in
+parallel.  To support an unlimited number of threads in applications
+that use this TM system, we assign segments dynamically to threads.
 
-At the start of a transaction, the thread it is running in acquires
-a segment. It may have to wait until another thread finishes its transaction
-and releases a segment. Fairness is not guaranteed yet, as we simply
-assume a fair scheduling policy in the operating system when waiting
-on a condition variable.
+At the start of a transaction, the thread it is running in acquires a
+segment. It may have to wait until another thread finishes its
+transaction and releases a segment. Fairness is not guaranteed yet, as
+we simply assume a fair scheduling policy in the operating system when
+waiting on a condition variable.
 
-Therefore, a thread may be assigned to different segments each time
-it starts a transaction. Although, we try to assign it the same segment
-again if possible. And a maximum of $N$ transactions may run in parallel.
+Therefore, a thread may be assigned to different segments each time it
+starts a transaction. Although, we try to assign it the same segment
+again if possible. And a maximum of $N$ transactions may run in
+parallel.
 
@@
 
-Garbage collection plays a big role in our TM system. The GC is generational
-and has two generations.
+Garbage collection plays a big role in our TM system. The GC is
+generational and has two generations.
 
 The \textbf{first generation}, where objects are considered to be
-\emph{young }and reside in the \emph{Nursery}, is collected by \emph{minor
-collections}. These collections move the surviving objects out of
-the nursery into the old object space, which can be done without stopping
-other threads. This is done either if the nursery has no space left
-anymore or if we are committing the current transaction. Consequently,
-all objects are old and the nursery empty after a transaction commits.
-Furthermore, all objects in the nursery were always created in the
-current transaction. This fact is useful since we do not need to call
-any barrier on this kind of objects.
+\emph{young }and reside in the \emph{Nursery}, is collected by
+\emph{minor collections}. These collections move the surviving objects
+out of the nursery into the old object space, which can be done
+without stopping other threads. This is done either if the nursery has
+no space left anymore or if we are committing the current
+transaction. Consequently, all objects are old and the nursery empty
+after a transaction commits.  Furthermore, all objects in the nursery
+were always created in the current transaction. This fact is useful
+since we do not need to call any barrier on this kind of objects.
 
-To improve this situation even more, we introduce the concept of \emph{overflow
-objects}. If a minor collection needs to occur during a transaction,
-we empty the nursery and mark each surviving object in the old object
-space with an \inputencoding{latin9}\lstinline!overflow_number!\inputencoding{utf8}
-globally unique to the current transaction. That way we can still
-detect in a medium-fast path inside barriers that the object still
-belongs to the current transaction.
+To improve this situation even more, we introduce the concept of
+\emph{overflow objects}. If a minor collection needs to occur during a
+transaction, we empty the nursery and mark each surviving object in
+the old object space with an \lstinline!overflow_number!  globally
+unique to the current transaction. That way we can still detect in a
+medium-fast path inside barriers that the object still belongs to the
+current transaction.
 
 The \textbf{second generation}, where objects are considered to be
-\emph{old }and never move again, is collected by \emph{major collections}.
-These collections are implemented in a stop-the-world kind of way
-and first force minor collections in all threads. The major goal is
-to free objects in the old objects space. Furthermore, we optimistically
-re-share pages that do not need to be private anymore.
+\emph{old }and never move again, is collected by \emph{major
+  collections}.  These collections are implemented in a stop-the-world
+kind of way and first force minor collections in all threads. The
+major goal is to free objects in the old objects space. Furthermore,
+we optimistically re-share pages that do not need to be private
+anymore.
 
-As seen in the API (section \ref{sub:Application-Programming-Interfac}),
-we use a \emph{shadow stack }in order to provide precise garbage collection.
-Any time we call a function that possibly triggers a collection, we
-need to save the objects that we need afterwards on the shadow stack
-using \inputencoding{latin9}\lstinline!STM_PUSH_ROOT()!\inputencoding{utf8}.
-That way, they will not be freed. And in case they were young, we
-get their new location in the old object space when getting them back
-from the stack using \inputencoding{latin9}\lstinline!STM_POP_ROOT()!\inputencoding{utf8}.
+As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
+we use a \emph{shadow stack} in order to provide precise garbage
+collection.  Any time we call a function that possibly triggers a
+collection, we need to save the objects that we need afterwards on the
+shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
+not be freed. And in case they were young, we get their new location
+in the old object space when getting them back from the stack using
+\lstinline!STM_POP_ROOT()!.
 
@@
 
-The point of the read barrier is to add the object to the read set
-of the transaction. This information is needed to detect conflicts
-between transactions. Usually, it also resolves an object reference
-to a private copy, but since the CPU performs our address translation
-on every object access efficiently, we do not need to do that in our
+The point of the read barrier is to add the object to the read set of
+the transaction. This information is needed to detect conflicts
+between transactions. Usually, it also resolves an object reference to
+a private copy, but since the CPU performs our address translation on
+every object access efficiently, we do not need to do that in our
 barrier.
 
-To add the object to the read set, for us it is enough to mark it
-as read. Since this information needs to be local to the segment,
-we need to store it in private pages. The area is called \emph{read
-markers }and already mentioned in section \ref{sub:Setup}. This area
-can be seen as a continuous array of bytes that is indexed from the
-start of the segment by an object's reference ($SO$) divided by 16
-(this requires objects of at least 16 bytes in size). Instead of just
-setting the byte to \inputencoding{latin9}\lstinline!true!\inputencoding{utf8}
-if the corresponding object was read, we set it to a \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
-belonging to the transaction, which will be incremented on each commit.
-Thereby, we can avoid resetting the bytes to \inputencoding{latin9}\lstinline!false!\inputencoding{utf8}
-on commit and only need to do this every 255 transactions. The whole
-code for the barrier is easily optimizable for compilers as well as
-perfectly predictable for CPUs:
+To add the object to the read set, for us it is enough to mark it as
+read. Since this information needs to be local to the segment, we need
+to store it in private pages. The area is called \emph{read markers
+}and already mentioned in section \ref{sub:Setup}. This area can be
+seen as a continuous array of bytes that is indexed from the start of
+the segment by an object's reference ($SO$) divided by 16 (this
+requires objects of at least 16 bytes in size). Instead of just
+setting the byte to \lstinline!true!  if the corresponding object was
+read, we set it to a \lstinline!read_version!  belonging to the
+transaction, which will be incremented on each commit.  Thereby, we
+can avoid resetting the bytes to \lstinline!false!  on commit and only
+need to do this every 255 transactions. The whole code for the barrier
+is easily optimizable for compilers as well as perfectly predictable
+for CPUs:
 
-\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
 void stm_read(SO):
@@
 \end{lstlisting}
-\inputencoding{utf8}
 
@@
 The job of the write barrier is twofold: first, it serves as a write
-barrier for the garbage collector and second, it supports copy-on-write
-and adds objects to the write set of the transaction.
+barrier for the garbage collector and second, it supports
+copy-on-write and adds objects to the write set of the transaction.
 
 The \textbf{fast path} of the write barrier is very simple. We only
-need to check for the flag \inputencoding{latin9}\lstinline!WRITE_BARRIER!\inputencoding{utf8}
-in the object's header and call the slow path if it is set. This flag
-is set either if the object is old and comes from an earlier transaction,
-or if there was a minor collection which will add the flag again on
-all objects. It is never set on freshly allocated objects that still
+need to check for the flag \lstinline!WRITE_BARRIER!  in the object's
+header and call the slow path if it is set. This flag is set either if
+the object is old and comes from an earlier transaction, or if there
+was a minor collection which will add the flag again on all
+objects. It is never set on freshly allocated objects that still
 reside in the nursery.
 
-\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
 void stm_write(SO):
@@
 
-\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
 void write_slowpath(SO):
@@
 to the list of objects that need tracing in the next minor collection
-(\inputencoding{latin9}\lstinline!to_trace!\inputencoding{utf8}).
-This is necessary in case we write a reference to it that points to
-a young object. We then need to trace it during the next minor collection
-in order to mark the young object alive and to update its reference
-to the new location it gets moved to. The check for \inputencoding{latin9}\lstinline!is_overflow_obj()!\inputencoding{utf8}
-tells us if the object was actually created in this transaction. In
-that case, we do not need to execute the following \emph{TM part}.
-We especially do not need to privatize the page since no other transaction
-knows about these ``old'' objects.
+(\lstinline!to_trace!).  This is necessary in case we write a
+reference to it that points to a young object. We then need to trace
+it during the next minor collection in order to mark the young object
+alive and to update its reference to the new location it gets moved
+to. The check for \lstinline!is_overflow_obj()!  tells us if the
+object was actually created in this transaction. In that case, we do
+not need to execute the following \emph{TM part}.  We especially do
+not need to privatize the page since no other transaction knows about
+these ``old'' objects.
 
-For TM, we first perform a read barrier on the object. We then try
-to acquire its write lock. \inputencoding{latin9}\lstinline!write_locks!\inputencoding{utf8}
-again is a simple global array of bytes that is indexed with the SO
-of the object divided by 16. If we already own the lock, we are done.
-If someone else owns the lock, we will do a write-write contention
-management that will abort either us or the current owner of the object.
-If we succeed in acquiring the lock using an atomic \inputencoding{latin9}\lstinline!cmp_and_swap!\inputencoding{utf8},
-we need to add the object to the write set (a simple list called \inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
-and privatize all pages belonging to it (copy-on-write).
+For TM, we first perform a read barrier on the object. We then try to
+acquire its write lock. \lstinline!write_locks!  again is a simple
+global array of bytes that is indexed with the SO of the object
+divided by 16. If we already own the lock, we are done.  If someone
+else owns the lock, we will do a write-write contention management
+that will abort either us or the current owner of the object.  If we
+succeed in acquiring the lock using an atomic
+\lstinline!cmp_and_swap!, we need to add the object to the write set
+(a simple list called \lstinline!modified_old_objects!)  and privatize
+all pages belonging to it (copy-on-write).
 
-In all cases, we remove the \inputencoding{latin9}\lstinline!WRITE_BARRIER!\inputencoding{utf8}
-flag from the object before we return. Thus, we never trigger the
-slow path again before we do the next minor collection (also part
-of a commit) or we start the next transaction.
+In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
+object before we return. Thus, we never trigger the slow path again
+before we do the next minor collection (also part of a commit) or we
+start the next transaction.
 
@@
 
-Aborting a transaction is rather easy. The first step is to reset
-the nursery and all associated data structures. The second step is
-to go over all objects in the write set (\inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
+Aborting a transaction is rather easy. The first step is to reset the
+nursery and all associated data structures. The second step is to go
+over all objects in the write set (\lstinline!modified_old_objects!)
 and reset any modifications in our private pages by copying from the
-sharing-segment. What is left is to use \inputencoding{latin9}\lstinline!longjmp()!\inputencoding{utf8}
-to jump back to the location initialized by a \inputencoding{latin9}\lstinline!setjmp()!\inputencoding{utf8}
-in \inputencoding{latin9}\lstinline!stm_start_transaction()!\inputencoding{utf8}.
-Increasing the \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
-is also done there.
+sharing-segment. What is left is to use \lstinline!longjmp()!  to jump
+back to the location initialized by a \lstinline!setjmp()!  in
+\lstinline!stm_start_transaction()!.  Increasing the
+\lstinline!read_version!  is also done there.
 
@@
 Committing a transaction needs a bit more work. First, we synchronize
-all threads so that the committing one is the only one running and
-all the others are waiting in a safe point. We then go through the
-write set (\inputencoding{latin9}\lstinline!modified_old_objects!\inputencoding{utf8})
-and check the corresponding \inputencoding{latin9}\lstinline!read_markers!\inputencoding{utf8}
-in other threads/segments. If we detect a read-write conflict, we
-do contention management to either abort us or the other transaction,
-or to simply wait a bit.
+all threads so that the committing one is the only one running and all
+the others are waiting in a safe point. We then go through the write
+set (\lstinline!modified_old_objects!)  and check the corresponding
+\lstinline!read_markers!  in other threads/segments. If we detect a
+read-write conflict, we do contention management to either abort us or
+the other transaction, or to simply wait a bit.
 
@@
 changes done to the objects in the write set to all other segments,
-including the sharing-segment. This is safe since we synchronized
-all threads. We also need to push overflow objects generated by minor
+including the sharing-segment. This is safe since we synchronized all
+threads. We also need to push overflow objects generated by minor
 collections to other segments, since they may reside partially in
-private pages. At that point we also get a new \inputencoding{latin9}\lstinline!overflow_number!\inputencoding{utf8}
-by increasing a global one, so that it stays globally unique for each
-transaction. Increasing the \inputencoding{latin9}\lstinline!read_version!\inputencoding{utf8}
-is then done at the start of a new transaction.
+private pages. At that point we also get a new
+\lstinline!overflow_number!  by increasing a global one, so that it
+stays globally unique for each transaction. Increasing the
+\lstinline!read_version!  is then done at the start of a new
+transaction.
 
@@
 
-A requirement for performing a commit is to synchronize all threads
-so that we can safely update objects in other segments. To make this
+A requirement for performing a commit is to synchronize all threads so
+that we can safely update objects in other segments. To make this
 synchronization fast and cheap, we do not want to insert an additional
@@
 very high-level and thus allocate a lot of objects very regularly.
-This is done through the function \inputencoding{latin9}\lstinline!stm_allocate!\inputencoding{utf8}
-shown below:
+This is done through the function \lstinline!stm_allocate!  shown
+below:
 
-\inputencoding{latin9}\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
 object_t *stm_allocate(ssize_t size_rounded):
@@
 \end{lstlisting}
-\inputencoding{utf8}
 
 This code does simple pointer-bump allocation in the nursery. If there
-is still space left in the nursery, we return \inputencoding{latin9}\lstinline!nursery_current!\inputencoding{utf8}
-and bump it up by \inputencoding{latin9}\lstinline!size_rounded!\inputencoding{utf8}.
-The interesting part is the check \inputencoding{latin9}\lstinline!nursery_current > nursery_end!\inputencoding{utf8}
-which will trigger the slow path of the function to possibly perform
-a minor collection in order to free up space in the nursery.
+is still space left in the nursery, we return
+\lstinline!nursery_current!  and bump it up by
+\lstinline!size_rounded!.  The interesting part is the check
+\lstinline!nursery_current > nursery_end!  which will trigger the slow
+path of the function to possibly perform a minor collection in order
+to free up space in the nursery.
 
 If we want to synchronize all threads, we can rely on this check being
-performed regularly. So what we do is to set the \inputencoding{latin9}\lstinline!nursery_end!\inputencoding{utf8}
-to $0$ in all segments that we want to synchronize. The mentioned
-check will then fail in those segments and call the slow path. In
-\inputencoding{latin9}\lstinline!allocate_slowpath!\inputencoding{utf8}
-they can simply check for this condition and enter a safe point.
+performed regularly. So what we do is to set the
+\lstinline!nursery_end!  to $0$ in all segments that we want to
+synchronize. The mentioned check will then fail in those segments and
+call the slow path. In \lstinline!allocate_slowpath!  they can simply
+check for this condition and enter a safe point.
 
@@
 \item prefer transactions that started earlier to younger transactions
-\item to support \emph{inevitable} transactions, we always prefer them to
-others since they cannot abort
+\item to support \emph{inevitable} transactions, we always prefer them
+  to others since they cannot abort
 \end{itemize}
@@
 or we can also wait until the other transaction committed. The latter
-is an interesting option if we are trying to commit a write and another
-transaction already read the object. We can then signal the other
-transaction to commit as soon as possible and wait. After waiting,
-there is now no conflict between our write and the already committed
-read anymore.
-
-\textbf{}
+is an interesting option if we are trying to commit a write and
+another transaction already read the object. We can then signal the
+other transaction to commit as soon as possible and wait. After
+waiting, there is now no conflict between our write and the already
+committed read anymore.
 

### 2014-05-15 talk/dls2014/report/report.tex
move report to its own directory

--- 2014-05-06 talk/dls2014/report.tex
@@
 parallelism to its applications. We want to highlight two of them,
-namely \emph{threading }and \emph{multiprocessing}.
+namely \emph{threading} and \emph{multiprocessing}.
 
@@
 
-We focus on the \emph{threading }approach. This requires us to remove
+We focus on the \emph{threading} approach. This requires us to remove
 the GIL from our interpreter in order to run code in parallel on
@@
 this. It requires great care in order to avoid deadlocks, which is why
-we follow the TM approach that provides a \emph{direct }replacement
+we follow the TM approach that provides a \emph{direct} replacement
 for the GIL. It does not require careful placing of locks in the right
@@
 Our conflict detection works with \emph{object
-  granularity}. Conceptually, it is based on \emph{read }and
-\emph{write sets }of transactions.  Two transactions conflict if they
+  granularity}. Conceptually, it is based on \emph{read} and
+\emph{write sets} of transactions.  Two transactions conflict if they
 have accessed a common object that is now in the write set of at least
@@
 
-The \emph{concurrency control }works partly \emph{optimistically} for
+The \emph{concurrency control} works partly \emph{optimistically} for
 reading of objects, where conflicts caused by just reading an object
@@
 
-We use \emph{lazy version management }to ensure that modifications by
+We use \emph{lazy version management} to ensure that modifications by
 a transaction are not visible to another transaction before the former
@@
 As required for TM systems, we guarantee complete \emph{isolation
-}and \emph{atomicity }for transactions at all times. Furthermore,
-the isolation provides full \emph{opacity }to always guarantee a consistent
+}and \emph{atomicity} for transactions at all times. Furthermore,
+the isolation provides full \emph{opacity} to always guarantee a consistent
 read set.
 
-We support the notion of \emph{inevitable transactions }that are always
+We support the notion of \emph{inevitable transactions} that are always
 guaranteed to commit. There is always at most one such transaction
@@
 software. However, we exploit some more advanced features of current
-CPUs, especially \emph{memory segmentation, virtual memory, }and the
+CPUs, especially \emph{memory segmentation, virtual memory,} and the
 64-bit address space.
@@
 We now use these mechanisms to provide isolation for transactions.
-Using write barriers, we implement a \emph{Copy-On-Write (COW) }on the
+Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
 level of pages. Starting from the initial fully-shared configuration
@@
 transactions, as well as the required read and write barriers. (II) It
-also includes a \emph{garbage collector (GC) }that is closely
+also includes a \emph{garbage collector (GC)} that is closely
 integrated with the TM part (e.g. it shares the write barrier). The
@@
 want to run $N$ threads in parallel while segment 0 is designated as
-the \emph{sharing-segment }that is never assigned to a thread.
+the \emph{sharing-segment} that is never assigned to a thread.
 
@@
 The \textbf{first generation}, where objects are considered to be
-\emph{young }and reside in the \emph{Nursery}, is collected by
+\emph{young} and reside in the \emph{Nursery}, is collected by
 \emph{minor collections}. These collections move the surviving objects
@@
 The \textbf{second generation}, where objects are considered to be
-\emph{old }and never move again, is collected by \emph{major
+\emph{old} and never move again, is collected by \emph{major
   collections}.  These collections are implemented in a stop-the-world

### 2014-05-15 talk/dls2014/paper/paper.tex
add paper template

--- 2014-05-05 talk/icooolps2014/position-paper.tex
@@
 \begin{abstract}
-  Dynamic languages became very popular in recent years. At some
-  point, the need for concurrency arose, and many of them made the
-  choice to use a single global interpreter lock (GIL) to synchronise
-  the interpreter in a multithreading scenario. This choice however
-  makes it impossible to actually run code in parallel.
-
-  Here we want to compare different approaches to replacing the GIL
-  with a technology that allows parallel execution. We look at
-  fine-grained locking, shared-nothing, and transactional memory (TM)
-  approaches. We argue that software-based TM systems are the most
-  promising, especially since they also enable the introduction of
-  large, parallelisable atomic blocks as a better synchronisation
-  mechanism in the language.
+....
 \end{abstract}
@@
 \keywords
-transactional memory, dynamic languages, parallelism, global interpreter lock
+...
 
 \section{Introduction}
-In a world where computers get more and more cores and single-thread
-performance increases less and less every year, many dynamic languages
-have a problem. While there is certainly a lot of popularity around
-languages like Python and Ruby, their ability to make use of multiple
-cores is somewhat limited. For ease of implementation they chose to
-use a single, global interpreter lock (GIL) to synchronise the
-execution of code in multiple threads. While this is a
-straight-forward way to eliminate synchronisation issues in the
-interpreter, it prevents parallel execution. Code executed in multiple
-threads will be serialised over this GIL so that only one thread can
-execute at a time.
+...
 
-There exist several solutions and workarounds to remove or avoid the
-GIL in order to benefit from multiple cores. We are going to discuss
-several of them and try to find the best way forward. The first
-approach uses fine-grained locking to replace the single GIL. Then
-there are shared-nothing models that use for example multiple
-processes with multiple interpreters and explicit message
-passing. Finally, one can also directly replace the GIL with
-transactional memory (TM), either software-based (STM) or
-hardware-based (HTM).
-
-The approach that wins in the end should perform similarly for
-single-threaded execution as compared to the GIL and be able to
-execute code in parallel on multiple cores. Furthermore, we will also
-take into account the compatibility with existing code that may already use
-threads for concurrency, as well as the changes that are required to
-the interpreter itself.
-
-These requirements are not easy to meet. We argue that STM is the
-overall winner. While it has a big performance problem currently, it
-gets more points in all the other categories. We think that it is the
-only solution that also provides a better synchronisation mechanism to
-the application in the form of parallelisable atomic blocks.
-
-%% \subsection{Issue}
-%% The issue that we want to discuss is how to efficiently support
-%% multi-core parallel execution of code in dynamic languages that were
-%% designed with GIL semantics in mind.
-
-%% Furthermore, a solution to this problem should also bring better
-%% synchronisation mechanism with it...
-
-%% (supporting (large) atomic blocks for synchronisation)
-
-%% \subsection{Our Position}
-%% Current solutions for replacing the GIL include STM, HTM, and
-%% fine-grained locking. STM is usually too slow, HTM very limited, and
-%% locking suffers from complexity that makes it hard to implement
-%% correctly. We argue that the best way forward is still STM and that
-%% its performance problem can be solved.
-
-%% Current solutions like STM, HTM, and fine-grained locking are slow, hard
-%% to implement correctly, and don't fit the specific problems of dynamic
-%% language.  STM is the best way forward but has bad performance, so we
-%% fix that.
-
-\begin{table*}[ht]
-  \centering
-  \begin{tabular}{|l|c|c|c|c|c|}
-    \hline
-    & \textbf{GIL} & \textbf{Fine-grained locking}
-    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
-    \hline
-    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
-    \hline
-    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
-    \hline
-    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
-    \hline
-    Better synchronisation        & o    & o  &      & o  & ++   \\
-    \hline
-    Implementation                & ++   & -  & ++   & ++ & ++   \\
-    \hline
-    External libraries            & ++   & ++ & ++   & ++ & ++   \\
-    \hline
-  \end{tabular}
-  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
-  \label{tab:comparison}
-\end{table*}
-
-\section{Discussion}
-
-In this section we examine the approaches and highlight their
-advantages and disadvantages.
-%% \paragraph{dynamic language VM problems}
-%% XXX:
-%% - high allocation rate (short lived objects)\\
-%% - (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)
-
-\subsection{Why is there a GIL?}
-The GIL is a very simple synchronisation mechanism for supporting
-multithreading in an interpreter. The basic guarantee is that the GIL
-may only be released in between bytecode instructions. The interpreter
-can thus rely on complete isolation and atomicity of these
-instructions. Additionally, it provides the application with a
-sequential consistency model~\cite{lamport79}. As a consequence,
-applications can rely on certain operations to be atomic and that they
-will always be executed in the order in which they appear in the
-code. While depending on this may not always be a good idea, it is
-done in practice. A GIL-replacement should therefore uphold these
-guarantees, while preferably also be as easily implementable as a GIL
-for the interpreter. The latter can be especially important since
-many of these languages are developed and maintained by very large
-open-source communities, which are not easy to coordinate.
-
-The GIL also allows for easy integration with external C libraries that
-may not be thread-safe. For the duration of the calls, we
-simply do not release the GIL. External libraries that are explicitly
-thread-safe can voluntarily release the GIL themselves in order to
-still provide some parallelism. This is done for example for
-potentially long I/O operations. Consequently, I/O-bound,
-multithreaded applications can actually parallelise to some
-degree. Again, a potential solution should be able to integrate with
-external libraries with similar ease. We will however focus our
-argumentation more on running code in the interpreted language in
-parallel, not the external C calls.
-
-Since the GIL is mostly an implementation detail of the interpreter,
-it is not exposed to the application running on top of it. To
-synchronise memory accesses in applications using threads, the
-state-of-the-art still means explicit locking everywhere. It is
-known that using locks for synchronisation can be hard at
-times~\cite{christopher10,victor11,shan08}. They are non-composable,
-have overhead, may deadlock, limit scalability, and add to the overall
-complexity of the program logic. For a better parallel programming
-model for dynamic languages, we propose another, well-known
-synchronisation mechanism called \emph{atomic
-  blocks}~\cite{tim03,tim05}.
-
-Atomic blocks are composable, deadlock-free, higher-level and expose
-useful atomicity and isolation guarantees to the application for a
-series of instructions. Interpreters using a GIL can simply guarantee
-that the GIL is not released during the execution of the atomic
-block. Of course, this still means that no two atomic blocks can
-execute in parallel or even concurrently.  Potential solutions are
-preferable if they provide a good way to implement atomic blocks that
-are also able to be executed in parallel.
-
-\subsection{Potential Solutions}
-
-For the discussion we define a set of criteria to evaluate the
-multiple potential solutions for removing or avoiding the GIL and its
-limitations:
-
-\begin{description}
-\item[Performance:] How well does the approach perform compared to the
-  GIL on a single and on multiple threads?
-\item[Existing applications:] How big are the changes required to
-  integrate with and parallelise existing applications?
-\item[Better synchronisation:] Does the approach enable better, paralleliseable
-  synchronisation mechanisms for applications (e.g. atomic blocks)?
-\item[Implementation:] How difficult is it to implement the approach
-  in the interpreter?
-\item[External libraries:] Does the approach allow for easy
-  integration of external libraries?
-\end{description}
-
-\subsubsection{Fine-Grained Locking}
-
-The first obvious candidate to replace the GIL is to use multiple
-locks instead of a single global lock. By refining the granularity of
-the locking approach, we gain the ability to run code that does not
-access the same objects in parallel. What we lose instead is the
-simplicity of the GIL approach. With every additional lock, the
-likeliness of deadlocks grows, as well as the overhead that acquiring
-and releasing locks produces. The former means that sometimes it is
-necessary to fall back to less fine-grained locking, preventing some
-potential parallelism in order to keep the complexity manageable.
-The latter means that we lose a bit of performance in the
-single-threaded case compared to the GIL, which requires much less
-acquire-release operations.
-
-Jython~\cite{webjython} is one project that implements an
-interpreter for Python on the Java Virtual Machine (JVM) and
-that uses fine-grained locking\footnote{The performance impact of
-fine-grained locking is milder in Java than it would be in a typical piece
-of C code; see e.g.~\cite{biased}.} to correctly synchronise the
-interpreter. For a language like Python, one needs quite a few,
-carefully placed locks. Since there is no central location, the
-complexity of the implementation is quite a bit larger compared to
-using a GIL. Integrating external, non-thread-safe libraries should
-however be very simple too. One could simply use one lock per library
-to avoid this issue.
-
-In the end, fine-grained locking can transparently replace the GIL and
-therefore parallelise existing applications, generally without any
-changes\footnote{There are rare cases where not having atomic
-  bytecodes actually changes the semantics.}. An implementation has to
-follow the GIL semantics very closely, otherwise it may expose some
-latent data races in existing applications which are just not exposed
-with a GIL. This approach does however not provide a better parallelising
-synchronisation mechanism to the application like e.g. atomic blocks.
-
-%% - support of atomic blocks?\\
-%% - hard to get right (deadlocks, performance, lock-granularity)\\
-%% - very hard to get right for a large language\\
-%% - hard to retro-fit, as all existing code assumes GIL semantics\\
-%% - (there are some semantic differences, right? not given perfect lock-placement, but well)
-%% ( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )
-
-\subsubsection{Shared-Nothing}
-
-There are also approaches that work around the GIL instead of trying
-to replace it. If an application can be split into completely
-independent parts that only very rarely need to share anything, or
-only do so via an external program like a database, then it is
-sensible to have one GIL per independent part. At the extreme, there
-are applications that parallelise perfectly simply by running
-independent processes; some web servers and some numeric computations
-do. We will consider here a slightly more general approach: the
-\emph{multiprocessing}\footnote{https://docs.python.org/2/library/multiprocessing.html}
-module of Python. In essence, it uses process-forking to provide the
-application with multiple interpreters that can run in parallel.
-Communication is then done explicitly through pipes.
-
-The model of explicit communication is sometimes seen as a superior
-way to synchronise concurrent applications because of its explicitness.
-However, not every application fits well into this model and its
-applicability is therefore limited. Performance is good as
-long as the application does not need to communicate a lot, because
-inter-process communication is relatively expensive. Also the
-implementation of this approach is very cheap since one can
-actually take an unmodified GIL-supported interpreter and run
-several of them in parallel. That way, we also inherit the
-easy integration of external libraries without any changes.
-
-%% - often needs major restructuring of programs (explicit data exchange)\\
-%% - sometimes communication overhead is too large\\
-%% - shared memory is a problem, copies of memory are too expensive
-
-\subsubsection{Transactional Memory}
-Transactional memory (TM) can be used as a direct replacement for a
-single global lock. Transactions provide the same atomicity and
-isolation guarantees as the GIL provides for the execution of bytecode
-instructions. So instead of acquiring and releasing the GIL between
-these instructions, this approach runs the protected instructions
-inside transactions.
-
-TM can be implemented in software (STM) or in hardware (HTM). There
-are also hybrid approaches, which combine the two. We count these
-hybrid approaches as STM, since they usually provide the same
-capabilities as software-only approaches but with different
-performance characteristics. We will now first look at HTM, which
-recently gained a lot of popularity by its introduction in common
-desktop CPUs from Intel (Haswell generation)~\cite{odaira14,leis14}.
-
-\paragraph{HTM} provides us with transactions like any TM system does.
-It can be used as a direct replacement for the GIL~\cite{nicholas06,odaira14,fuad10}. However, as is
-common with hardware-only solutions, there are quite a few limitations
-that can not be lifted easily. For this comparison, we look at the
-implementation of Intel in recent Haswell generation CPUs.
-
-HTM in these CPUs works on the level of caches. This has a few
-consequences like false-sharing on the cache-line level, and most
-importantly it limits the amount of memory that can be accessed within
-a transaction. This transaction-length limitation makes it necessary
-to have a fallback in place in case this limit is reached. In recent
-attempts, the usual fallback is the GIL~\cite{odaira14,fuad10}. In our
-experiments, the current generation of HTM proved to be very fragile
-and thus needing the fallback very often. Consequently, scalability
-suffered a lot from this.
-
-The performance of HTM is pretty good as it does not introduce much
-overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
-parallelise existing applications to some degree. The implementation
-is very straight-forward because it directly replaces the GIL in a
-central place. HTM is also directly compatible with any external
-library that needs to be integrated and synchronised for use in
-multiple threads. The one thing that is missing is support for a
-better synchronisation mechanism for the application. It is not
-reasonable in general to expose the hardware-transactions to the
-application in the form of atomic blocks, because doing so would
-require the system to support much longer transactions.
-
-%% - false-sharing on cache-line level\\
-%% - limited capacity (caches, undocumented)\\
-%% - random aborts (haswell)\\
-%% - generally: transaction-length limited (no atomic blocks)
-
-\paragraph{STM} provides all the same benefits as HTM except for its
-performance.  It is not unusual for the overhead introduced by STM to
-be between 100\% to even 1000\% ~\cite{cascaval08,drago11}. While STM
-systems often scale very well to a big number of threads and
-eventually overtake the single-threaded execution, they often provide
-no benefits at all for low numbers of threads (1-8). There are some
-attempts ~\cite{warmhoff13,spear09} that can reduce the overhead a lot,
-but scale badly or only for certain workloads. Often the benefits
-on more than one thread are too small in real world applications.
-
-However, STM compared to HTM does not suffer from the same restricting
-limitations. Transactions can in principle be arbitrarily long.  This makes it
-possible to actually expose transactions to the application in the
-form of atomic blocks. This is the only approach that enables a better
-synchronisation mechanism than locks for applications \emph{and} still
-parallelises when using it. We think this is a very important point
-because it not only gives dynamic languages the ability to parallelise
-(already commonplace in most other languages), but also pushes
-parallel programming forward. Together with sequential consistency it
-provides a lot of simplification for parallel applications.
-
-On the implementation level,
-while one can argue that STM requires the insertion of read and write
-barriers in the whole interpreter, this can be done automatically and
-locally by a program transformation~\cite{felber07}. There are attempts
-to do the same for fine-grained locking~\cite{bill06} but they require
-a whole program analysis since locks are inherently non-composable
---- and their effectiveness is doubtful in our use case,
-since we execute bytecode instructions in any order defined by a
-script only known at runtime. This makes it close to impossible to
-order locks consistently or to know in advance which locks a
-transaction will need.
-
-%% - overhead (100-1000\%) (barrier reference resolution, kills performance on low \#cpu)
-%% (FastLane: low overhead, not much gain)\\
-%% - unlimited transaction length (easy atomic blocks)
-
-\section{The Way Forward}
-
-Following the above argumentation for each approach we assembled a
-general overview in Table \ref{tab:comparison}. The general picture is
-everything else than clear. It looks like HTM may be a good solution
-to replace the GIL in the near future. Current implementations are
-however far too limiting and do not provide good scaling.
-
-Allowing for parallel execution just means that dynamic languages
-catch up to all other languages that already provide real
-parallelism. This is why we think that only the STM approach is a
-viable solution in the long-term. It provides the application with a
-simple memory model (sequential consistency) and a composable way to
-synchronise memory accesses using atomic blocks.
-
-Unfortunately, STM has a big performance problem. Particularly, for
-our use case there is not much static information available since we
-are executing a program only known at runtime. Additionally, replacing
-the GIL means running every part of the application in transactions,
-so there is not much code that can run outside and that can be
-optimised better. The performance of the TM system is vital.
-
-One way to get more performance is to develop STM systems that make
-better use of low-level features in existing OS kernels.  We are
-currently working on a STM system that makes use of several such
-features like virtual memory and memory segmentation.  We further
-tailor the system to the discussed use case which gives us an
-advantage over other STM systems that are more general. With this
-approach, initial results suggest that we can keep the overhead of STM
-well below 50\%. A hybrid TM system, which also uses HTM to accelerate
-certain tasks, looks like a very promising direction of research
-too. We believe that further work to reduce the overhead of STM is
-very worthwhile.
-
-%% possible solution:\\
-%% - use virtual memory paging to somehow lower the STM overhead\\
-%% - tight integration with GC and jit?
+\section{Conclusion}
 
@@
 \acks
-We would like to thank Maciej Fijalkowski and Carl Friedrich Bolz for
-their valuable inputs and the many fruitful discussions.
+...
 

### 2014-05-15 talk/dls2014/report/report.tex
some tweaks and several notes

@@
 \newcommand\remi[1]{\mynote{Remi}{#1}}
+\newcommand\cfbolz[1]{\mynote{cfbolz}{#1}}
 
@@
 
+\cfbolz{the above is good, here is something missing: problems with current STM approaches, outlining the intuition behind the new one}
+
 Our contributions include:
@@
 \emph{serializable}. Meaning, the outcome is equal to some sequential
-execution of these transactions. Overall, this is exactly what a
-single global lock guarantees while still allowing the TM system to
+execution of these transactions. This means that the approach provides the same
+semantics as using the GIL
+while still allowing the TM system to
 run transactions in parallel as an optimization.
@@
 \subsection{Python}
+
+\cfbolz{a pypy introduction needs to go somewhere, a paragraph or so. maybe in the evaluation section}
 
@@
 
+ cfbolz{citation again needed for the whole subsection}
+
 It is well known that using locks to synchronize multiple threads is
@@
 scalability, and overall add a lot of complexity. For a better
-parallel programming model for dynamic languages, we want to add
+parallel programming model for dynamic languages, we want to implement
 another, well-known synchronization mechanism: \emph{atomic blocks}.
@@
 the literature.
+
+\cfbolz{there is an overview paragraph of the idea missing, maybe in the introduction}
+
+\cfbolz{this all feels very much dumping details, needs more overview. why is this info important? the subsubsections don't have any connections}
 
@@
 read set.
+
+\cfbolz{this paragraph is hard to understand without giving an example (eg console printing) when it is useful}
 
@@
 To get references to objects that are valid in all threads, we will
-use the object's offset inside the segment. Since all segments are
+use \cfbolz{use for what?} the object's offset inside the segment. Since all segments are
 copies of each other, the \emph{Segment Offset (SO)} will point to the
@@
   (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.
+
+\cfbolz{here it needs to say that this is x86 specific}
 
@@
   resolves the reference to the private version of the object on every
-  access anyway, this is not the job of the read barrier anymore.
+  access anyway, the read barrier does not need to do address translation anymore.
 \item [{Write~Barrier:}] Adds the object to the read and write set of
@@
 
-To provide atomicity for a transaction, we want to make changes
+To provide atomicity for a transaction, we want to make changes globally
 visible on commit. We also need to be able to completely abort a
@@
   the pages in other segments, including the sharing-segment (segment
-  0).
+  0). \cfbolz{can it really happen that you push pages to other segments? I thought it's always just back to the sharing segment}
 \item [{Abort:}] On abort the transaction will forget about all the
@@
   private pages of the aborting transaction.
+  \cfbolz{why doing any copying? aren't the pages re-shared instead?}
 \end{description}
+
+\cfbolz{random question: did we investigate the extra memory requirements? we should characterize memory overhead somewhere, eg at least one byte per object for the read markers}
 
@@
   shared between segments. They mostly contain old objects but also
-  some young ones that were too big to allocate in the nursery.
+  some young ones that were too big to be allocated in the nursery.
 \end{description}
@@
 starts a transaction. Although, we try to assign it the same segment
-again if possible. And a maximum of $N$ transactions may run in
-parallel.
+again if possible.
 
@@
 generational and has two generations.
+
+\cfbolz{maybe use "young" and "old" generation, if there are only two}
 
@@
 the transaction. This information is needed to detect conflicts
-between transactions. Usually, it also resolves an object reference to
+between transactions. In other STM systems, it also resolves an object reference to
 a private copy, but since the CPU performs our address translation on
@@
 read-write conflict, we do contention management to either abort us or
-the other transaction, or to simply wait a bit.
+the other transaction, or to simply wait a bit. \cfbolz{why does waiting help?}
 

### 2014-05-15 talk/dls2014/paper/paper.tex
move report's content to the paper

@@
 % authoryear    To obtain author/year citation style instead of numeric.
+\synctex=-1
 
@@
 \usepackage{color}
-\usepackage{hyperref}
 \usepackage{amsmath}
 \usepackage{amssymb}
+\usepackage{fixltx2e}
+\usepackage{graphicx}
+\usepackage[unicode=true,pdfusetitle,
+ bookmarks=true,bookmarksnumbered=false,bookmarksopen=false,
+ breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]{hyperref}
+
+\makeatletter
+\usepackage{enumitem}
+\usepackage{multicol}
+
+% nice listings
+\usepackage{xcolor}
+\usepackage{newverbs}
+
+\usepackage{color}
+\definecolor{verylightgray}{rgb}{0.93,0.93,0.93}
+\definecolor{darkblue}{rgb}{0.2,0.2,0.6}
+\definecolor{commentgreen}{rgb}{0.25,0.5,0.37}
+\usepackage{letltxmacro}
+
+\usepackage{listings}
+
+\makeatletter
+\LetLtxMacro{\oldlstinline}{\lstinline}
+
+\renewcommand\lstinline[1][]{%
+  \Collectverb{\@@myverb}%
+}
+
+\def\@@myverb#1{%
+    \begingroup
+    \fboxsep=0.2em
+    \colorbox{verylightgray}{\oldlstinline|#1|}%
+    \endgroup
+}
+\makeatother
+
+\lstset{backgroundcolor={\color{verylightgray}},
+  basicstyle={\scriptsize\ttfamily},
+  commentstyle={\ttfamily\color{commentgreen}},
+  keywordstyle={\bfseries\color{darkblue}},
+  morecomment={[l]{//}},
+  tabsize=4,
+  morekeywords={foreach,in,def,type,dynamic,Int,
+    Boolean,infer,void,super,if,boolean,int,else,
+    while,do,extends,class,assert,for,switch,case,
+    private,protected,public,const,final,static,
+    interface,new,true,false,null,return}}
+\renewcommand{\lstlistingname}{Listing}
 
@@
 
-\newcommand\cfbolz[1]{\mynote{Carl Friedrich}{#1}}
+\newcommand\remi[1]{\mynote{Remi}{#1}}
+\newcommand\cfbolz[1]{\mynote{cfbolz}{#1}}
 
@@
 
-\title{A Way Forward in Parallelising Dynamic Languages}
-\subtitle{Position Paper, ICOOOLPS'14}
+\title{Virtual Memory Assisted Transactional Memory for Dynamic Languages}
+\subtitle{DLS'14}
 
@@
 \section{Introduction}
-...
 
-\section{Conclusion}
+Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
+regarded as very expressive but also very slow. In recent years, the
+introduction of just-in-time compilers (JIT) for these languages (e.g.
+PyPy, V8, Tracemonkey) started to change this perception by delivering
+good performance that enables new applications. However, a parallel
+programming model was not part of the design of those languages. Thus,
+the reference implementations of e.g. Python and Ruby use a single,
+global interpreter lock (GIL) to serialize the execution of code in
+threads.
+
+While this GIL prevents any parallelism from occurring, it also
+provides some useful guarantees. Since this lock is always acquired
+while executing bytecode instructions and it may only be released
+in-between such instructions, it provides perfect isolation and
+atomicity between multiple threads for a series of
+instructions. Another technology that can provide the same guarantees
+is transactional memory (TM).
+
+There have been several attempts at replacing the GIL with TM. Using
+transactions to enclose multiple bytecode instructions, we can get the
+very same semantics as the GIL while possibly executing several
+transactions in parallel. Furthermore, by exposing these
+interpreter-level transactions to the application in the form of
+\emph{atomic blocks}, we give dynamic languages a new synchronization
+mechanism that avoids several of the problems of locks as they are
+used now.
+
+\cfbolz{the above is good, here is something missing: problems with current STM approaches, outlining the intuition behind the new one}
+
+Our contributions include:
+\begin{itemize}[noitemsep]
+\item We introduce a new software transactional memory (STM) system
+  that performs well even on low numbers of CPUs. It uses a novel
+  combination of hardware features and garbage collector (GC)
+  integration in order to keep the overhead of STM very low.
+\item This new STM system is used to replace the GIL in Python and is
+  then evaluated extensively.
+\item We introduce atomic blocks to the Python language to provide a
+  backwards compatible, composable synchronization mechanism for
+  threads.
+\end{itemize}
+
+\section{Background}
+
+\subsection{Transactional Memory}
+
+Transactional memory (TM) is a concurrency control mechanism that
+comes from database systems. Using transactions, we can group a series
+of instructions performing operations on memory and make them happen
+atomically and in complete isolations from other
+transactions. \emph{Atomicity} means that all these instructions in
+the transaction and their effects seem to happen at one, undividable
+point in time. Other transactions never see inconsistent state of a
+partially executed transaction which is called \emph{isolation}.
+
+If we start multiple such transactions in multiple threads, the TM
+system guarantees that the outcome of running the transactions is
+\emph{serializable}. Meaning, the outcome is equal to some sequential
+execution of these transactions. This means that the approach provides the same
+semantics as using the GIL
+while still allowing the TM system to
+run transactions in parallel as an optimization.
+
+\subsection{Python}
+
+\cfbolz{a pypy introduction needs to go somewhere, a paragraph or so. maybe in the evaluation section}
+
+We implement and evaluate our system for the Python language. For the
+actual implementation, we chose the PyPy interpreter because replacing
+the GIL there with a TM system is just a matter of adding a new
+transformation to the translation process of the interpreter.
+
+Over the years, Python added multiple ways to provide concurrency and
+parallelism to its applications. We want to highlight two of them,
+namely \emph{threading} and \emph{multiprocessing}.
+
+\emph{Threading} employs operating system (OS) threads to provide
+concurrency. It is, however, limited by the GIL and thus does not
+provide parallelism. At this point we should mention that it is indeed
+possible to run external functions written in C instead of Python in
+parallel. Our work focuses on Python itself and ignores this aspect as
+it requires writing in a different language.
+
+The second approach, \emph{multiprocessing}, uses multiple instances
+of the interpreter itself and runs them in separate OS processes.
+Here we actually get parallelism because we have one GIL per
+interpreter, but of course we have the overhead of multiple processes
+/ interpreters and also need to exchange data between them explicitly
+and expensively.
+
+We focus on the \emph{threading} approach. This requires us to remove
+the GIL from our interpreter in order to run code in parallel on
+multiple threads. One approach to this is fine-grained locking instead
+of a single global lock. Jython and IronPython are implementations of
+this. It requires great care in order to avoid deadlocks, which is why
+we follow the TM approach that provides a \emph{direct} replacement
+for the GIL. It does not require careful placing of locks in the right
+spots. We will compare our work with Jython for evaluation.
+
+\subsection{Synchronization}
+
+\cfbolz{citation again needed for the whole subsection}
+
+It is well known that using locks to synchronize multiple threads is
+hard. They are non-composable, have overhead, may deadlock, limit
+scalability, and overall add a lot of complexity. For a better
+parallel programming model for dynamic languages, we want to implement
+another, well-known synchronization mechanism: \emph{atomic blocks}.
+
+Atomic blocks are composable, deadlock-free, higher-level and expose
+useful atomicity and isolation guarantees to the application for a
+series of instructions. An implementation using a GIL would simply
+guarantee that the GIL is not released during the execution of the
+atomic block. Using TM, we have the same effect by guaranteeing that
+all instructions in an atomic block are executed inside a single
+transaction.
+
+\remi{STM, how atomicity \& isolation; reasons for overhead}
+
+\section{Method}
+
+\subsection{Transactional Memory Model}
+
+In this section, we describe the general model of our TM system. This
+should clarify the general semantics in commonly used terms from
+the literature.
+
+\cfbolz{there is an overview paragraph of the idea missing, maybe in the introduction}
+
+\cfbolz{this all feels very much dumping details, needs more overview. why is this info important? the subsubsections don't have any connections}
+
+\subsubsection{Conflict Handling}
+
+Our conflict detection works with \emph{object
+  granularity}. Conceptually, it is based on \emph{read} and
+\emph{write sets} of transactions.  Two transactions conflict if they
+have accessed a common object that is now in the write set of at least
+one of them.
+
+The \emph{concurrency control} works partly \emph{optimistically} for
+reading of objects, where conflicts caused by just reading an object
+in transactions are detected only when the transaction that writes the
+object actually commits. For write-write conflicts we are currently
+\emph{pessimistic}: Only one transaction may have a certain object in
+its write set at any point in time, others trying to write to it will
+have to wait or abort.
+
+We use \emph{lazy version management} to ensure that modifications by
+a transaction are not visible to another transaction before the former
+commits.
+
+\subsubsection{Semantics}
+
+As required for TM systems, we guarantee complete \emph{isolation}
+and \emph{atomicity} for transactions at all times. Furthermore,
+the isolation provides full \emph{opacity} to always guarantee a consistent
+read set.
+
+To support irreversible operations that cannot be undone when we abort
+a transaction (e.g. I/O, syscalls, and non-transactional code in
+general), we employ \emph{irrevocable} or \emph{inevitable
+transactions}. These transactions are always guaranteed to
+commit. There is always at most one such transaction running in the
+system, thus their execution is serialised. With this guarantee,
+providing \emph{strong isolation} and \emph{serializability} between
+non-transactional code is possible by making the current transaction
+inevitable right before running irreversible operations.
+
+\subsubsection{Contention Management}
+
+When a conflict is detected, we perform some simple contention
+management.  First, inevitable transactions always win. Second, the
+older transaction wins. Different schemes are possible.
+
+\subsubsection{Software Transactional Memory}
+
+Generally speaking, the system is fully implemented in
+software. However, we exploit some more advanced features of current
+CPUs, especially \emph{memory segmentation, virtual memory,} and the
+64-bit address space.
+
+\subsection{Implementation}
+
+In this section, we will present the general idea of how the TM model
+is implemented. Especially the aspects of providing isolation and
+atomicity, as well as conflict detection are explained. We try to do
+this without going into too much detail about the implementation.  The
+later section \ref{sub:Low-level-Implementation} will discuss it in
+more depth.
+
+\subsubsection{Memory Segmentation}
+
+A naive approach to providing complete isolation between threads is to
+partition the virtual memory of a process into $N$ segments, one per
+thread. Each segment then holds a copy of all the memory available to
+the program. Thus, each thread automatically has a private copy of
+every object that it can modify in complete isolation from other
+threads.
+
+In order to reference these objects, we need references that are valid
+in all threads and automatically point to the private copies. Since
+an object's offset inside a segment is the same in all segments, we
+can use this offset to reference objects. Because all segments are
+copies of each other, this \emph{Segment Offset (SO)} points to the
+private version of an object in all threads\,/\,segments. To then
+translate this SO to a real virtual memory address when used inside a
+thread, we need to add the thread's segment start address to the
+SO. The result of this operation is called a \emph{Linear Address
+  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.
+
+x86-CPUs provide a feature called \emph{memory segmentation}. It
+performs this translation from a SO to a LA directly in hardware.  We
+can use the segment register $\%gs$, which is mostly unused in current
+applications. When this register points to a thread's segment start
+address, we can instruct the CPU to perform the above translation from
+a reference of the form $\%gs{::}SO$ to the right LA on its own. This
+process is efficient enough that we can do it on every access to an
+object.
+
+In summary, we can use a single SO to reference the same object in all
+threads, and it will be translated by the CPU to a LA that always
+points to the thread's private version of this object. Thereby,
+threads are fully isolated from each other. However, $N$ segments
+require $N$-times the memory and modifications on an object need to be
+propagated to all segments.
+
+\begin{figure*}[t]
+  \begin{centering}
+    \includegraphics[scale=0.8]{\string"segment addressing\string".pdf}
+    \par\end{centering}
+
+    \protect\caption{Segment Addressing\label{fig:Segment-Addressing}}
+\end{figure*}
+
+\subsubsection{Page Sharing}
+
+In order to eliminate the prohibitive memory requirements of keeping
+around $N$ segment copies, we share memory between them. The segments
+are initially allocated in a single range of virtual memory by a call
+to \lstinline!mmap()!.  As illustrated in Figure
+\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
+between a range of virtual memory pages and virtual file pages. The
+virtual file pages are then mapped lazily by the kernel to real
+physical memory pages. The mapping generated by \lstinline!mmap()! is
+initially linear but can be changed arbitrarily. Especially, we can
+remap so that multiple virtual memory pages map to a single virtual
+file page. This is what we use to share memory between the segments
+since then we also only require one page of physical memory.
+
+\begin{figure}[h]
+  \begin{centering}
+    \includegraphics[scale=0.8]{\string"mmap pages\string".pdf}
+    \par\end{centering}
+
+    \protect\caption{\texttt{mmap()} Page Mapping\label{fig:mmap()-Page-Mapping}}
+\end{figure}
+
+As illustrated in Figure \ref{fig:Page-Remapping}, in our initial
+configuration (I) all segments are backed by their own range of
+virtual file pages. This is the share-nothing configuration.
+
+We then designate segment 0 to be the \emph{Sharing-Segment}. No
+thread gets this segment assigned to it, it simply holds the pages
+shared between all threads. So in (II), we remap all virtual pages of
+the segments $>0$ to the file pages of our sharing-segment. This is
+the fully-shared configuration.
+
+During runtime, we can then privatize single pages in segments $>0$
+again by remapping single pages as seen in (III).
+
+Looking back at address translation for object references, we see now
+that this is actually a two-step process. First, $\%gs{::}SO$ gets
+translated to different linear addresses in different threads by the
+CPU. Then, depending on the current mapping of virtual pages to file
+pages, these LAs can map to a single file page in the sharing-segment,
+or to privatized file pages in the corresponding segments. This
+mapping is also performed efficiently by the CPU and can easily be
+done on every access to an object.
+
+In summary, $\%gs{::}SO$ is translated efficiently by the CPU to
+either a physical memory location which is shared between several
+threads/segments, or to a location in memory private to the
+segment/thread.  This makes the memory segmentation model for
+isolation memory efficient again.
+
+\begin{figure}[h]
+  \begin{centering}
+    \includegraphics[width=1\columnwidth]{\string"page remapping\string".pdf}
+    \par\end{centering}
+
+    \protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
+      segment 0, fully shared memory configuration. (III) privatize single
+      pages.\label{fig:Page-Remapping}}
+\end{figure}
+
+\subsubsection{Isolation: Copy-On-Write}
+
+We now use these mechanisms to provide isolation for transactions.
+Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
+level of pages. Starting from the initial fully-shared configuration
+(Figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
+object without other threads seeing the changes immediately, we ensure
+that all pages belonging to the object are private to our segment.
+
+To detect when to privatize pages, we use write barriers before every
+write. When the barrier detects that the object is not in a private
+page (or any pages that belong to the object), we remap and copy the
+pages to the thread's segment. From now on, the translation of
+$\%gs{::}SO$ in this particular segment will resolve to the private
+version of the object. Note, the SO used to reference the object does
+not change during that process.
+
+\subsubsection{Isolation: Barriers}
+
+The job of barriers is to ensure complete isolation between transactions
+and to register the objects in the read or write set. We insert read
+and write barriers before reading or modifying an object except if
+we statically know an object to be readable or writable already.
+\begin{description}
+\item [{Read~Barrier:}] Adds the object to the read set of the current
+  transaction. Since our two-step address translation automatically
+  resolves the reference to the private version of the object on every
+  access anyway, the read barrier does not need to do address translation anymore.
+\item [{Write~Barrier:}] Adds the object to the read and write set of
+  the current transaction and checks if all pages of the object are
+  private, doing COW otherwise.\\
+  Furthermore, we currently allow only one transaction modifying an
+  object at a time. To ensure this, we acquire a write lock on the object
+  and also eagerly check for a write-write conflict at this point. If
+  there is a conflict, we do some contention management to decide which
+  transaction has to wait or abort. Eagerly detecting this kind of conflict
+  is not inherent to our system, future experiments may show that we
+  want to lift this restriction.
+\end{description}
+
+\subsubsection{Atomicity: Commit \& Abort}
+
+To provide atomicity for a transaction, we want to make changes globally
+visible on commit. We also need to be able to completely abort a
+transaction without a trace, like it never happened.
+\begin{description}
+\item [{Commit:}] If a transaction commits, we synchronize all threads
+  so that all of them are waiting in a safe point. In the committing
+  transaction, we go through all objects in the write set and check if
+  another transaction in a different segment read the same object.
+  Conflicts are resolved again by either the committing or the other
+  transaction waiting or aborting.\\
+  We then push all changes of modified objects in private pages to all
+  the pages in other segments, including the sharing-segment (segment
+  0).
+\item [{Abort:}] On abort the transaction will forget about all the
+  changes it has done. All objects in the write set are reset by
+  copying their previous version from the sharing-segment into the
+  private pages of the aborting transaction.\\
+  Re-sharing these pages instead of resetting the changes would not
+  only be slower because it involves a system call, it is also very
+  likely the page would get privatised again immediately after re-trying
+  the transaction.  Since this again involves a full page copy,
+  resetting should be faster than re-sharing.
+\end{description}
+
+\cfbolz{random question: did we investigate the extra memory requirements? we should characterize memory overhead somewhere, eg at least one byte per object for the read markers}
+
+\subsubsection{Summary}
+
+We provide isolation between transactions by privatizing the pages of
+the segments belonging to the threads the transactions run in.  To
+detect when and which pages need privatization, we use write barriers
+that trigger a COW of one or several pages. Conflicts, however, are
+detected on the level of objects; based on the concept of read and
+write sets. Barriers before reading and writing add objects to the
+corresponding set; particularly detecting write-write conflicts
+eagerly.  On commit, we resolve read-write conflicts and push
+modifications to other segments. Aborting transactions simply undo
+their changes by copying from the sharing-segment.
+
+\subsection{Low-level Implementation\label{sub:Low-level-Implementation}}
+
+In this section, we will provide details about the actual
+implementation of the system and discuss some of the issues that we
+encountered.
+
+\subsubsection{Architecture}
+
+Our TM system is designed as a library that covers all aspects around
+transactions and object management. The library consists of two parts:
+(I) It provides a simple interface to starting and committing
+transactions, as well as the required read and write barriers. (II) It
+also includes a \emph{garbage collector (GC)} that is closely
+integrated with the TM part (e.g. it shares the write barrier). The
+close integration helps in order to know more about the lifetime of an
+object, as will be explained in the following sections.
+
+\subsubsection{Application Programming Interface\label{sub:Application-Programming-Interfac}}
+
+\begin{lstlisting}
+void stm_start_transaction(tl, jmpbuf)
+void stm_commit_transaction()
+void stm_read(object_t *obj)
+void stm_write(object_t *obj)
+object_t *stm_allocate(ssize_t size_rounded)
+STM_PUSH_ROOT(tl, obj)
+STM_POP_ROOT(tl, obj)
+\end{lstlisting}
+
+\lstinline!stm_start_transaction()!  starts a transaction. It requires
+two arguments, the first being a thread-local data structure and the
+second a buffer for use by \lstinline!setjmp()!.
+\lstinline!stm_commit_transaction()!  tries to commit the current
+transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
+a read or a write barrier on an object and \lstinline!stm_allocate()!
+allocates a new object with the specified size (must be a multiple of
+16). \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
+and pop objects on the shadow stack~\footnote{A stack for pointers to
+  GC objects that allows for precise garbage collection. All objects
+  on that stack are never seen as garbage and are thus always kept
+  alive.}.  Objects have to be saved using this stack around calls
+that may cause a GC cycle to happen, and also while there is no
+transaction running. In this simplified API, only
+\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
+require saving object references.
+
+The type \lstinline!object_t!  is special as it causes the
+compiler~\footnote{Clang 3.5 with some patches to this address-space
+  256 feature} to make all accesses through it relative to the $\%gs$
+register.  With exceptions, nearly all accesses to objects managed by
+the TM system should use this type so that the CPU will translate the
+reference to the right version of the object.
+
+\subsubsection{Setup\label{sub:Setup}}
+
+On startup, we reserve a big range of virtual memory with a call to
+\lstinline!mmap()! and partition this space into $N+1$ segments. We
+want to run $N$ threads in parallel while segment 0 is designated as
+the \emph{sharing-segment} that is never assigned to a thread.
+
+The next step involves using \lstinline!remap_file_pages()!, a Linux
+system call, to establish the fully-shared configuration.  All pages
+of segments $>0$ map to the pages of the sharing-segment.
+
+However, the layout of a segment is not uniform and we actually
+privatize a few areas again right away. These areas are illustrated in
+Figure \ref{fig:Segment-Layout} and explained here:
+\begin{description}[noitemsep]
+\item [{NULL~page:}] This page is unmapped and will produce a
+  segmentation violation when accessed. We use this to detect
+  erroneous dereferencing of \lstinline!NULL! references.  All
+  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
+  if SO is set to \lstinline!NULL!.
+\item [{Segment-local~data:}] Some area private to the segment that
+  contains segment-local information.
+\item [{Read~markers:}] These are pages that store information about
+  which objects were read in the current transaction running in this
+  segment.
+\item [{Nursery:}] This area contains all the freshly allocated
+  objects (\emph{young objects}) of the current transaction. The GC
+  uses pointer-bump allocation in this area to allocate objects in the
+  first generation.
+\item [{Old~object~space:}] These pages are the ones that are really
+  shared between segments. They mostly contain old objects but also
+  some young ones that were too big to be allocated in the nursery.
+\end{description}
+
+\begin{figure*}[t]
+  \begin{centering}
+    \includegraphics[scale=0.8]{\string"segment layout\string".pdf}
+    \par\end{centering}
+
+    \protect\caption{Segment Layout\label{fig:Segment-Layout}}
+\end{figure*}
+
+\subsubsection{Assigning Segments}
+
+From the above setup it is clear that the number of segments is
+statically set to some $N$. That means that at any point in time, a
+maximum of $N$ threads and their transactions can be running in
+parallel.  To support an unlimited number of threads in applications
+that use this TM system, we assign segments dynamically to threads.
+
+At the start of a transaction, the thread it is running in acquires a
+segment. It may have to wait until another thread finishes its
+transaction and releases a segment. Fairness is not guaranteed yet, as
+we simply assume a fair scheduling policy in the operating system when
+waiting on a condition variable.
+
+Therefore, a thread may be assigned to different segments each time it
+starts a transaction. Although, we try to assign it the same segment
+again if possible.
+
+\subsubsection{Garbage Collection}
+
+Garbage collection plays a big role in our TM system. The GC is
+generational and has two generations: the \emph{young} and the
+\emph{old} generation.
+
+The \textbf{young generation}, where objects are considered to be
+\emph{young} and reside in the \emph{Nursery}, is collected by
+\emph{minor collections}. These collections move the surviving objects
+out of the nursery into the old object space, which can be done
+without stopping other threads. This is done either if the nursery has
+no space left anymore or if we are committing the current
+transaction. Consequently, all objects are old and the nursery empty
+after a transaction commits.  Furthermore, all objects in the nursery
+were always created in the current transaction. This fact is useful
+since we do not need to call any barrier on this kind of objects.
+
+To improve this situation even more, we introduce the concept of
+\emph{overflow objects}. If a minor collection needs to occur during a
+transaction, we empty the nursery and mark each surviving object in
+the old object space with an \lstinline!overflow_number!  globally
+unique to the current transaction. That way we can still detect in a
+medium-fast path inside barriers that the object still belongs to the
+current transaction.
+
+The \textbf{old generation}, where objects are considered to be
+\emph{old} and never move again, is collected by \emph{major
+  collections}.  These collections are implemented in a stop-the-world
+kind of way and first force minor collections in all threads. The
+major goal is to free objects in the old objects space. Furthermore,
+we optimistically re-share pages that do not need to be private
+anymore.
+
+As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
+we use a \emph{shadow stack} in order to provide precise garbage
+collection.  Any time we call a function that possibly triggers a
+collection, we need to save the objects that we need afterwards on the
+shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
+not be freed. And in case they were young, we get their new location
+in the old object space when getting them back from the stack using
+\lstinline!STM_POP_ROOT()!.
+
+\subsubsection{Read Barrier}
+
+The point of the read barrier is to add the object to the read set of
+the transaction. This information is needed to detect conflicts
+between transactions. In other STM systems, it also resolves an object reference to
+a private copy, but since the CPU performs our address translation on
+every object access efficiently, we do not need to do that in our
+barrier.
+
+To add the object to the read set, for us it is enough to mark it as
+read. Since this information needs to be local to the segment, we need
+to store it in private pages. The area is called \emph{read markers
+}and already mentioned in section \ref{sub:Setup}. This area can be
+seen as a continuous array of bytes that is indexed from the start of
+the segment by an object's reference ($SO$) divided by 16 (this
+requires objects of at least 16 bytes in size). Instead of just
+setting the byte to \lstinline!true!  if the corresponding object was
+read, we set it to a \lstinline!read_version!  belonging to the
+transaction, which will be incremented on each commit.  Thereby, we
+can avoid resetting the bytes to \lstinline!false!  on commit and only
+need to do this every 255 transactions. The whole code for the barrier
+is easily optimizable for compilers as well as perfectly predictable
+for CPUs:
+
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+void stm_read(SO):
+    *(SO >> 4) = read_version
+\end{lstlisting}
+
+\subsubsection{Write Barrier}
+
+The job of the write barrier is twofold: first, it serves as a write
+barrier for the garbage collector and second, it supports
+copy-on-write and adds objects to the write set of the transaction.
+
+The \textbf{fast path} of the write barrier is very simple. We only
+need to check for the flag \lstinline!WRITE_BARRIER!  in the object's
+header and call the slow path if it is set. This flag is set either if
+the object is old and comes from an earlier transaction, or if there
+was a minor collection which will add the flag again on all
+objects. It is never set on freshly allocated objects that still
+reside in the nursery.
+
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+void stm_write(SO):
+	if SO->flags & WRITE_BARRIER:
+		write_slowpath(SO)
+\end{lstlisting}
+
+The \textbf{slow path} is shown here:
+
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+void write_slowpath(SO):
+	// GC part:
+	list_append(to_trace, SO)
+	if is_overflow_obj(SO):
+		SO->flags &= ~WRITE_BARRIER
+		return
+	// STM part
+	stm_read(SO)
+	lock_idx = SO >> 4
+  retry:
+	if write_locks[lock_idx] == our_num:
+		// we already own it
+	else if write_locks[lock_idx] == 0:
+		if cmp_and_swap(&write_locks[lock_idx],
+					    0, our_num):
+			list_append(modified_old_objects, SO)
+			privatize_pages(SO)
+		else:
+			goto retry
+	else:
+		w_w_contention_management()
+		goto retry
+	SO->flags &= ~WRITE_BARRIER
+\end{lstlisting}
+
+First comes the \emph{GC part}: In any case, the object will be added
+to the list of objects that need tracing in the next minor collection
+(\lstinline!to_trace!).  This is necessary in case we write a
+reference to it that points to a young object. We then need to trace
+it during the next minor collection in order to mark the young object
+alive and to update its reference to the new location it gets moved
+to. The check for \lstinline!is_overflow_obj()!  tells us if the
+object was actually created in this transaction. In that case, we do
+not need to execute the following \emph{TM part}.  We especially do
+not need to privatize the page since no other transaction knows about
+these ``old'' objects.
+
+For TM, we first perform a read barrier on the object. We then try to
+acquire its write lock. \lstinline!write_locks!  again is a simple
+global array of bytes that is indexed with the SO of the object
+divided by 16. If we already own the lock, we are done.  If someone
+else owns the lock, we will do a write-write contention management
+that will abort either us or the current owner of the object.  If we
+succeed in acquiring the lock using an atomic
+\lstinline!cmp_and_swap!, we need to add the object to the write set
+(a simple list called \lstinline!modified_old_objects!)  and privatize
+all pages belonging to it (copy-on-write).
+
+In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
+object before we return. Thus, we never trigger the slow path again
+before we do the next minor collection (also part of a commit) or we
+start the next transaction.
+
+\subsubsection{Abort}
+
+Aborting a transaction is rather easy. The first step is to reset the
+nursery and all associated data structures. The second step is to go
+over all objects in the write set (\lstinline!modified_old_objects!)
+and reset any modifications in our private pages by copying from the
+sharing-segment. What is left is to use \lstinline!longjmp()!  to jump
+back to the location initialized by a \lstinline!setjmp()!  in
+\lstinline!stm_start_transaction()!.  Increasing the
+\lstinline!read_version!  is also done there.
+
+\subsubsection{Commit}
+
+Committing a transaction needs a bit more work. First, we synchronize
+all threads so that the committing one is the only one running and all
+the others are waiting in a safe point. We then go through the write
+set (\lstinline!modified_old_objects!)  and check the corresponding
+\lstinline!read_markers!  in other threads/segments. If we detect a
+read-write conflict, we do contention management to either abort us or
+the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).
+
+After verifying that there are no conflicts anymore, we copy all our
+changes done to the objects in the write set to all other segments,
+including the sharing-segment. This is safe since we synchronised all
+threads. We also need to push overflow objects generated by minor
+collections to other segments, since they may reside partially in
+private pages. At that point we also get a new
+\lstinline!overflow_number!  by increasing a global one, so that it
+stays globally unique for each transaction. Increasing the
+\lstinline!read_version!  is then done at the start of a new
+transaction.
+
+\subsubsection{Thread Synchronization}
+
+A requirement for performing a commit is to synchronize all threads so
+that we can safely update objects in other segments. To make this
+synchronization fast and cheap, we do not want to insert an additional
+check regularly in order to see if synchronization is requested. We
+use a trick relying on the fact that dynamic languages are usually
+very high-level and thus allocate a lot of objects very regularly.
+This is done through the function \lstinline!stm_allocate!  shown
+below:
+
+\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+object_t *stm_allocate(ssize_t size_rounded):
+    result = nursery_current
+	nursery_current += size_rounded
+	if nursery_current > nursery_end:
+		return allocate_slowpath(size_rounded)
+	return result
+\end{lstlisting}
+
+This code does simple pointer-bump allocation in the nursery. If there
+is still space left in the nursery, we return
+\lstinline!nursery_current!  and bump it up by
+\lstinline!size_rounded!.  The interesting part is the check
+\lstinline!nursery_current > nursery_end!  which will trigger the slow
+path of the function to possibly perform a minor collection in order
+to free up space in the nursery.
+
+If we want to synchronize all threads, we can rely on this check being
+performed regularly. So what we do is to set the
+\lstinline!nursery_end!  to $0$ in all segments that we want to
+synchronize. The mentioned check will then fail in those segments and
+call the slow path. In \lstinline!allocate_slowpath!  they can simply
+check for this condition and enter a safe point.
+
+For other synchronization requirements, for example:
+\begin{itemize}[noitemsep]
+\item waiting for a segment to be released,
+\item waiting for a transaction to abort or commit,
+\item waiting for all threads to reach their safe points,
+\end{itemize}
+we use a set of condition variables to wait or signal other threads.
+
+\subsubsection{Contention Management\label{subsub:contentionmanagement}}
+
+On encountering conflicts, we employ contention management to solve
+the problem as well as we can. The general rules are:
+\begin{itemize}[noitemsep]
+\item prefer transactions that started earlier to younger transactions
+\item to support \emph{inevitable} transactions, we always prefer them
+  to others since they cannot abort
+\end{itemize}
+We can either simply abort a transaction to let the other one succeed,
+or we can also wait until the other transaction committed. The latter
+is an interesting option if we are trying to commit a write and
+another transaction already read the object. We can then signal the
+other transaction to commit as soon as possible and wait. After
+waiting, there is now no conflict between our write and the already
+committed read anymore.
+
+\section{Experimental Results}
+
+compare some programs between
+\begin{itemize}[noitemsep]
+\item pypy
+\item pypy-jit
+\item pypy-stm
+\item pypy-stm-jit
+\item cpython
+\item jython
+\end{itemize}
+
+\section{Related Work}
+
+\section{Conclusions}
 

### 2014-05-15 talk/dls2014/report/report.tex
move report's content to the paper

@@
 % ------
-\title{C7: Fast software transactional memory for dynamic languages}
+\title{Virtual Memory Assisted Transactional Memory for Dynamic Languages}
 %
@@
 
- cfbolz{citation again needed for the whole subsection}
+\cfbolz{citation again needed for the whole subsection}
 
@@
 In this section, we describe the general model of our TM system. This
-should clarify the general semantics using commonly used terms from
+should clarify the general semantics in commonly used terms from
 the literature.
@@
 
-As required for TM systems, we guarantee complete \emph{isolation
-}and \emph{atomicity} for transactions at all times. Furthermore,
+As required for TM systems, we guarantee complete \emph{isolation}
+and \emph{atomicity} for transactions at all times. Furthermore,
 the isolation provides full \emph{opacity} to always guarantee a consistent
@@
 
-\cfbolz{this paragraph is hard to understand without giving an example (eg console printing) when it is useful}
-
-We support the notion of \emph{inevitable transactions} that are always
-guaranteed to commit. There is always at most one such transaction
-running in the system. We use this kind of transaction to provide
-\emph{strong isolation} by running non-transactional code in the context
-of inevitable transactions and to still provide the \emph{serializability}
-of all transaction schedules.
+To support irreversible operations that cannot be undone when we abort
+a transaction (e.g. I/O, syscalls, and non-transactional code in
+general), we employ \emph{irrevocable} or \emph{inevitable
+transactions}. These transactions are always guaranteed to
+commit. There is always at most one such transaction running in the
+system, thus their execution is serialised. With this guarantee,
+providing \emph{strong isolation} and \emph{serializability} between
+non-transactional code is possible by making the current transaction
+inevitable right before running irreversible operations.
 
@@
 
-To get references to objects that are valid in all threads, we will
-use \cfbolz{use for what?} the object's offset inside the segment. Since all segments are
-copies of each other, the \emph{Segment Offset (SO)} will point to the
-private version of an object in all threads/segments. To then
+In order to reference these objects, we need references that are valid
+in all threads and automatically point to the private copies. Since
+an object's offset inside a segment is the same in all segments, we
+can use this offset to reference objects. Because all segments are
+copies of each other, this \emph{Segment Offset (SO)} points to the
+private version of an object in all threads\,/\,segments. To then
 translate this SO to a real virtual memory address when used inside a
@@
 
-\cfbolz{here it needs to say that this is x86 specific}
-
-To make this address translation efficient, we use the segment
-register $\%gs$. When this register points to a thread's segment start
+x86-CPUs provide a feature called \emph{memory segmentation}. It
+performs this translation from a SO to a LA directly in hardware.  We
+can use the segment register $\%gs$, which is mostly unused in current
+applications. When this register points to a thread's segment start
 address, we can instruct the CPU to perform the above translation from
-a reference of the form $\%gs{::}SO$ to the right LA on its own.
+a reference of the form $\%gs{::}SO$ to the right LA on its own. This
+process is efficient enough that we can do it on every access to an
+object.
 
@@
   the pages in other segments, including the sharing-segment (segment
-  0). \cfbolz{can it really happen that you push pages to other segments? I thought it's always just back to the sharing segment}
+  0).
 \item [{Abort:}] On abort the transaction will forget about all the
@@
   copying their previous version from the sharing-segment into the
-  private pages of the aborting transaction.
-  \cfbolz{why doing any copying? aren't the pages re-shared instead?}
+  private pages of the aborting transaction.\\
+  Re-sharing these pages instead of resetting the changes would not
+  only be slower because it involves a system call, it is also very
+  likely the page would get privatised again immediately after re-trying
+  the transaction.  Since this again involves a full page copy,
+  resetting should be faster than re-sharing.
 \end{description}
@@
 Garbage collection plays a big role in our TM system. The GC is
-generational and has two generations.
+generational and has two generations: the \emph{young} and the
+\emph{old} generation.
 
-\cfbolz{maybe use "young" and "old" generation, if there are only two}
-
-The \textbf{first generation}, where objects are considered to be
+The \textbf{young generation}, where objects are considered to be
 \emph{young} and reside in the \emph{Nursery}, is collected by
@@
 
-The \textbf{second generation}, where objects are considered to be
+The \textbf{old generation}, where objects are considered to be
 \emph{old} and never move again, is collected by \emph{major
@@
 read-write conflict, we do contention management to either abort us or
-the other transaction, or to simply wait a bit. \cfbolz{why does waiting help?}
+the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).
 
@@
 changes done to the objects in the write set to all other segments,
-including the sharing-segment. This is safe since we synchronized all
+including the sharing-segment. This is safe since we synchronised all
 threads. We also need to push overflow objects generated by minor
@@
 
-\subsubsection{Contention Management}
+\subsubsection{Contention Management\label{subsub:contentionmanagement}}
 

### 2014-05-15 talk/dls2014/paper/paper.tex
paper updates

@@
 the reference implementations of e.g. Python and Ruby use a single,
-global interpreter lock (GIL) to serialize the execution of code in
+global interpreter lock (GIL) to serialise the execution of code in
 threads.
@@
 instructions. Another technology that can provide the same guarantees
-is transactional memory (TM).
+is transactional memory (TM). \remi{cite our position paper}
 
@@
 interpreter-level transactions to the application in the form of
-\emph{atomic blocks}, we give dynamic languages a new synchronization
+\emph{atomic blocks}, we give dynamic languages a new synchronisation
 mechanism that avoids several of the problems of locks as they are
@@
 
-\cfbolz{the above is good, here is something missing: problems with current STM approaches, outlining the intuition behind the new one}
+\remi{cite and extract from (our pos. paper):}
+TM systems come in can be broadly categorised as hardware based (HTM),
+software based (STM), or hybrid systems (HyTM). HTM systems are limited
+by hardware constraints, while STM systems have a lot of overhead.
+In this paper, we describe how we manage to lower the overhead of our
+STM system so that it can be seen as a viable replacement for the GIL.
 
@@
 \item We introduce atomic blocks to the Python language to provide a
-  backwards compatible, composable synchronization mechanism for
+  backwards compatible, composable synchronisation mechanism for
   threads.
@@
 of instructions performing operations on memory and make them happen
-atomically and in complete isolations from other
+atomically and in complete isolation from other
 transactions. \emph{Atomicity} means that all these instructions in
-the transaction and their effects seem to happen at one, undividable
+the transaction and their effects seem to happen at one, indivisible
 point in time. Other transactions never see inconsistent state of a
@@
 system guarantees that the outcome of running the transactions is
-\emph{serializable}. Meaning, the outcome is equal to some sequential
+\emph{serialisable}. Meaning, the outcome is equal to some sequential
 execution of these transactions. This means that the approach provides the same
@@
 while still allowing the TM system to
-run transactions in parallel as an optimization.
+run transactions in parallel as an optimisation.
 
@@
 
-\subsection{Synchronization}
+\subsection{Synchronisation}
 
@@
 
-It is well known that using locks to synchronize multiple threads is
+It is well known that using locks to synchronise multiple threads is
 hard. They are non-composable, have overhead, may deadlock, limit
@@
 parallel programming model for dynamic languages, we want to implement
-another, well-known synchronization mechanism: \emph{atomic blocks}.
+another, well-known synchronisation mechanism: \emph{atomic blocks}.
 
@@
 
-In this section, we describe the general model of our TM system. This
-should clarify the general semantics in commonly used terms from
-the literature.
+In this section, we characterise the model of our TM system and its
+guarantees as well as some of the design choices we made. This should
+clarify the general semantics in commonly used terms from the
+literature.\remi{cite Transactional Memory 2nd edition}
 
-\cfbolz{there is an overview paragraph of the idea missing, maybe in the introduction}
-
-\cfbolz{this all feels very much dumping details, needs more overview. why is this info important? the subsubsections don't have any connections}
+Our TM system is fully implemented in software. However, we do exploit
+some more advanced features of current CPUs, particularly \emph{memory
+segmentation, virtual memory,} and the 64-bit address space. Still,
+it cannot be classified as a hybrid TM system since it currently
+makes no use of any HTM present in the CPU.
 
@@
 
-Our conflict detection works with \emph{object
-  granularity}. Conceptually, it is based on \emph{read} and
-\emph{write sets} of transactions.  Two transactions conflict if they
-have accessed a common object that is now in the write set of at least
-one of them.
+We implement an object-based TM system, thus it makes sense to detect
+conflicts with \emph{object granularity}. With this choice, if two
+transactions access the same object and at least one access is a
+write, we count it as a conflict. Conceptually, it is based on
+\emph{read} and \emph{write sets} of transactions. Reading from an
+object adds the object to the read set, writing to it adds it to both
+sets. Two transactions conflict if they have accessed a common object
+that is in the write set of at least one of them.
 
-The \emph{concurrency control} works partly \emph{optimistically} for
-reading of objects, where conflicts caused by just reading an object
-in transactions are detected only when the transaction that writes the
-object actually commits. For write-write conflicts we are currently
+The detection, or \emph{concurrency control}, works partly
+\emph{optimistically} for reading objects. Read-write conflicts
+between two transactions are detected in both exactly at the time when
+the writing one commits. For write-write conflicts we are currently
 \emph{pessimistic}: Only one transaction may have a certain object in
 its write set at any point in time, others trying to write to it will
-have to wait or abort.
+have to wait or abort. This decision needs to be evaluated further
+in the future.
 
-We use \emph{lazy version management} to ensure that modifications by
-a transaction are not visible to another transaction before the former
-commits.
+When a conflict is detected, we perform some simple contention
+management that generally prefers the older transaction to the younger.
 
@@
 
-As required for TM systems, we guarantee complete \emph{isolation}
-and \emph{atomicity} for transactions at all times. Furthermore,
-the isolation provides full \emph{opacity} to always guarantee a consistent
-read set.
+As required for TM systems, we guarantee complete \emph{isolation} and
+\emph{atomicity} for transactions at all times. Our method of choice
+is \emph{lazy version management}. Modifications by a transaction are
+not visible to another transaction before the former commits.
+Furthermore, the isolation provides full \emph{opacity} to always
+guarantee a consistent read set even for non-committed transactions.
+\remi{cite On the Correctness of Transactional Memory}
 
-To support irreversible operations that cannot be undone when we abort
-a transaction (e.g. I/O, syscalls, and non-transactional code in
-general), we employ \emph{irrevocable} or \emph{inevitable
-transactions}. These transactions are always guaranteed to
-commit. There is always at most one such transaction running in the
-system, thus their execution is serialised. With this guarantee,
-providing \emph{strong isolation} and \emph{serializability} between
-non-transactional code is possible by making the current transaction
-inevitable right before running irreversible operations.
-
-\subsubsection{Contention Management}
-
-When a conflict is detected, we perform some simple contention
-management.  First, inevitable transactions always win. Second, the
-older transaction wins. Different schemes are possible.
-
-\subsubsection{Software Transactional Memory}
-
-Generally speaking, the system is fully implemented in
-software. However, we exploit some more advanced features of current
-CPUs, especially \emph{memory segmentation, virtual memory,} and the
-64-bit address space.
+To also support these properties for irreversible operations that
+cannot be undone when we abort a transaction (e.g. I/O, syscalls, and
+non-transactional code in general), we use \emph{irrevocable} or
+\emph{inevitable transactions}. These transactions are always
+guaranteed to commit, which is why they always have to win in case
+there is a conflict with another, normal transaction. There is always
+at most one such transaction running in the system, thus their
+execution is serialised. With this guarantee, providing \emph{strong
+isolation} and \emph{serialisability} between non-transactional code
+is possible by making the current transaction inevitable right before
+running irreversible operations.
 
@@
 
-During runtime, we can then privatize single pages in segments $>0$
+During runtime, we can then privatise single pages in segments $>0$
 again by remapping single pages as seen in (III).
@@
 pages, these LAs can map to a single file page in the sharing-segment,
-or to privatized file pages in the corresponding segments. This
+or to privatised file pages in the corresponding segments. This
 mapping is also performed efficiently by the CPU and can easily be
@@
     \protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
-      segment 0, fully shared memory configuration. (III) privatize single
+      segment 0, fully shared memory configuration. (III) privatise single
       pages.\label{fig:Page-Remapping}}
@@
 
-To detect when to privatize pages, we use write barriers before every
+To detect when to privatise pages, we use write barriers before every
 write. When the barrier detects that the object is not in a private
@@
 \begin{description}
-\item [{Commit:}] If a transaction commits, we synchronize all threads
+\item [{Commit:}] If a transaction commits, we synchronise all threads
   so that all of them are waiting in a safe point. In the committing
@@
 
-\cfbolz{random question: did we investigate the extra memory requirements? we should characterize memory overhead somewhere, eg at least one byte per object for the read markers}
+\cfbolz{random question: did we investigate the extra memory requirements? we should characterise memory overhead somewhere, eg at least one byte per object for the read markers}
 
@@
 
-We provide isolation between transactions by privatizing the pages of
+We provide isolation between transactions by privatising the pages of
 the segments belonging to the threads the transactions run in.  To
-detect when and which pages need privatization, we use write barriers
+detect when and which pages need privatisation, we use write barriers
 that trigger a COW of one or several pages. Conflicts, however, are
@@
 However, the layout of a segment is not uniform and we actually
-privatize a few areas again right away. These areas are illustrated in
+privatise a few areas again right away. These areas are illustrated in
 Figure \ref{fig:Segment-Layout} and explained here:
@@
 need to do this every 255 transactions. The whole code for the barrier
-is easily optimizable for compilers as well as perfectly predictable
+is easily optimisable for compilers as well as perfectly predictable
 for CPUs:
 
-\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}
 void stm_read(SO):
@@
 
-\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}
 void stm_write(SO):
@@
 
-\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}
 void write_slowpath(SO):
@@
 not need to execute the following \emph{TM part}.  We especially do
-not need to privatize the page since no other transaction knows about
+not need to privatise the page since no other transaction knows about
 these ``old'' objects.
@@
 \lstinline!cmp_and_swap!, we need to add the object to the write set
-(a simple list called \lstinline!modified_old_objects!)  and privatize
+(a simple list called \lstinline!modified_old_objects!)  and privatise
 all pages belonging to it (copy-on-write).
@@
 sharing-segment. What is left is to use \lstinline!longjmp()!  to jump
-back to the location initialized by a \lstinline!setjmp()!  in
+back to the location initialised by a \lstinline!setjmp()!  in
 \lstinline!stm_start_transaction()!.  Increasing the
@@
 
-Committing a transaction needs a bit more work. First, we synchronize
+Committing a transaction needs a bit more work. First, we synchronise
 all threads so that the committing one is the only one running and all
@@
 
-\subsubsection{Thread Synchronization}
+\subsubsection{Thread Synchronisation}
 
-A requirement for performing a commit is to synchronize all threads so
+A requirement for performing a commit is to synchronise all threads so
 that we can safely update objects in other segments. To make this
-synchronization fast and cheap, we do not want to insert an additional
-check regularly in order to see if synchronization is requested. We
+synchronisation fast and cheap, we do not want to insert an additional
+check regularly in order to see if synchronisation is requested. We
 use a trick relying on the fact that dynamic languages are usually
@@
 
-\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
+\begin{lstlisting}
 object_t *stm_allocate(ssize_t size_rounded):
@@
 
-If we want to synchronize all threads, we can rely on this check being
+If we want to synchronise all threads, we can rely on this check being
 performed regularly. So what we do is to set the
 \lstinline!nursery_end!  to $0$ in all segments that we want to
-synchronize. The mentioned check will then fail in those segments and
+synchronise. The mentioned check will then fail in those segments and
 call the slow path. In \lstinline!allocate_slowpath!  they can simply
@@
 
-For other synchronization requirements, for example:
+For other synchronisation requirements, for example:
 \begin{itemize}[noitemsep]

### 2014-05-15 talk/dls2014/paper/paper.tex
add some missing points

@@
 can use this offset to reference objects. Because all segments are
-copies of each other, this \emph{Segment Offset (SO)} points to the
+copies of each other, this \emph{Segment Offset ($SO$)} points to the
 private version of an object in all threads\,/\,segments. To then
-translate this SO to a real virtual memory address when used inside a
+translate this $SO$ to a real virtual memory address when used inside a
 thread, we need to add the thread's segment start address to the
-SO. The result of this operation is called a \emph{Linear Address
+$SO$. The result of this operation is called a \emph{Linear Address
   (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.
@@
 x86-CPUs provide a feature called \emph{memory segmentation}. It
-performs this translation from a SO to a LA directly in hardware.  We
+performs this translation from a $SO$ to a LA directly in hardware.  We
 can use the segment register $\%gs$, which is mostly unused in current
@@
 
-In summary, we can use a single SO to reference the same object in all
+In summary, we can use a single $SO$ to reference the same object in all
 threads, and it will be translated by the CPU to a LA that always
@@
 $\%gs{::}SO$ in this particular segment will resolve to the private
-version of the object. Note, the SO used to reference the object does
+version of the object. Note, the $SO$ used to reference the object does
 not change during that process.
@@
   $\%gs{::}SO$ translated to linear addresses will point to NULL pages
-  if SO is set to \lstinline!NULL!.
+  if $SO$ is set to \lstinline!NULL!.
 \item [{Segment-local~data:}] Some area private to the segment that
@@
 read. Since this information needs to be local to the segment, we need
-to store it in private pages. The area is called \emph{read markers
-}and already mentioned in section \ref{sub:Setup}. This area can be
-seen as a continuous array of bytes that is indexed from the start of
-the segment by an object's reference ($SO$) divided by 16 (this
-requires objects of at least 16 bytes in size). Instead of just
-setting the byte to \lstinline!true!  if the corresponding object was
-read, we set it to a \lstinline!read_version!  belonging to the
-transaction, which will be incremented on each commit.  Thereby, we
-can avoid resetting the bytes to \lstinline!false!  on commit and only
-need to do this every 255 transactions. The whole code for the barrier
-is easily optimisable for compilers as well as perfectly predictable
-for CPUs:
+to store it in private pages. The area is called \emph{read markers}
+and already mentioned in section \ref{sub:Setup}.
+
+This area can be seen as a continuous array of bytes that is indexed
+from the start of the segment by an object's reference ($SO$) divided
+by 16 (this is where the requirement of objects to be of at least 16
+bytes in size comes from). Instead of just setting the byte to
+\lstinline!true!  if the corresponding object was read, we set it to a
+\lstinline!read_version!  belonging to the transaction, which will be
+incremented on each commit.  Thereby, we can avoid resetting the bytes
+to \lstinline!false!  on commit and only need to do this every 255
+transactions. The whole code for the barrier is easily optimisable for
+compilers as well as perfectly predictable for CPUs:
 
@@
 acquire its write lock. \lstinline!write_locks!  again is a simple
-global array of bytes that is indexed with the SO of the object
+global array of bytes that is indexed with the $SO$ of the object
 divided by 16. If we already own the lock, we are done.  If someone
@@
 
-\section{Experimental Results}
-
-compare some programs between
+\section{Evaluation}
+
+\subsection{Memory Requirements}
+
+\begin{itemize}
+\item stm\_flags per object
+\item read markers and other sections
+\item private pages
+\end{itemize}
+
+maybe some memory usage graph over time
+
+\subsection{Overhead Breakdown}
+
+\begin{itemize}
+\item time taken by read \& write barriers
+\item time spent committing \& aborting (maybe with different numbers
+  of threads)
+\item time in GC
+\end{itemize}
+
+\subsection{Scaling}
+
+maybe some simple micro benchmarks with adaptable conflict rate
+
+\subsection{Real-World Benchmarks}
+
+more real benchmarks comparing multiple implementations:
 \begin{itemize}[noitemsep]

### 2014-05-16 talk/dls2014/paper/paper.tex
adjust some details

@@
 run transactions in parallel as an optimisation.
+\remi{maybe some more explanation of how exactly TM replaces the GIL}
 
@@
 Here we actually get parallelism because we have one GIL per
-interpreter, but of course we have the overhead of multiple processes
-/ interpreters and also need to exchange data between them explicitly
+interpreter, but of course we have the overhead of multiple processes~/
+interpreters and also need to exchange data between them explicitly
 and expensively.
@@
 copies of each other, this \emph{Segment Offset ($SO$)} points to the
-private version of an object in all threads\,/\,segments. To then
+private version of an object in all threads~/ segments. To then
 translate this $SO$ to a real virtual memory address when used inside a
@@
 virtual file pages are then mapped lazily by the kernel to real
-physical memory pages. The mapping generated by \lstinline!mmap()! is
-initially linear but can be changed arbitrarily. Especially, we can
-remap so that multiple virtual memory pages map to a single virtual
-file page. This is what we use to share memory between the segments
-since then we also only require one page of physical memory.
+physical memory pages.
+
+The mapping generated by \lstinline!mmap()! is initially linear but
+can be changed arbitrarily. Especially, we can remap so that multiple
+virtual memory pages map to a single virtual file page. This is what
+we use to share memory between the segments, since we then only
+require one page of physical memory for all of them.
 
@@
 configuration (I) all segments are backed by their own range of
-virtual file pages. This is the share-nothing configuration.
-
-We then designate segment 0 to be the \emph{Sharing-Segment}. No
+virtual file pages. This is the share-nothing configuration where
+all threads have private versions of all objects.
+
+We then designate segment~0 to be the \emph{sharing-segment}. No
 thread gets this segment assigned to it, it simply holds the pages
-shared between all threads. So in (II), we remap all virtual pages of
-the segments $>0$ to the file pages of our sharing-segment. This is
-the fully-shared configuration.
-
-During runtime, we can then privatise single pages in segments $>0$
-again by remapping single pages as seen in (III).
+shared between threads. So in step (II), we remap all virtual pages of
+the segments~$>0$ to the file pages of our sharing-segment. This is
+the fully-shared configuration where no threads have private versions
+of any objects.
+
+During runtime, we can then privatise single pages in segments~$>0$
+again by remapping single pages as seen in (III). All objects in that
+page now have a private version in some thread.
 
@@
 or to privatised file pages in the corresponding segments. This
-mapping is also performed efficiently by the CPU and can easily be
-done on every access to an object.
+mapping is also performed efficiently by CPUs that have a Memory
+Management Unit (MMU) and can easily be done on every access to an
+object.
 
@@
 either a physical memory location which is shared between several
-threads/segments, or to a location in memory private to the
-segment/thread.  This makes the memory segmentation model for
-isolation memory efficient again.
+threads~/ segments or to a location in memory private to the segment~/
+thread. Page sharing makes the memory segmentation model for isolation
+memory efficient again.
 
@@
     \protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
-      segment 0, fully shared memory configuration. (III) privatise single
+      segment~0, fully shared memory configuration. (III) privatise single
       pages.\label{fig:Page-Remapping}}
@@
 To detect when to privatise pages, we use write barriers before every
-write. When the barrier detects that the object is not in a private
-page (or any pages that belong to the object), we remap and copy the
-pages to the thread's segment. From now on, the translation of
-$\%gs{::}SO$ in this particular segment will resolve to the private
-version of the object. Note, the $SO$ used to reference the object does
-not change during that process.
+write to an object. When the barrier detects that the object is not in
+a private page (or any pages that belong to the object), we remap and
+copy the pages to the thread's segment. From now on, the translation
+of $\%gs{::}SO$ in this particular thread will resolve to a private
+version of the object automatically. Note that the $SO$ used to reference
+the object does not change during that process.
 
@@
 and to register the objects in the read or write set. We insert read
-and write barriers before reading or modifying an object except if
+and write barriers before reading or modifying an object, except if
 we statically know an object to be readable or writable already.
@@
   resolves the reference to the private version of the object on every
-  access anyway, the read barrier does not need to do address translation anymore.
+  access anyway, the read barrier does not have the job to find the
+  private version. This job is fully performed by the CPU so that
+  read barriers have very little work to do.
 \item [{Write~Barrier:}] Adds the object to the read and write set of
@@
   there is a conflict, we do some contention management to decide which
-  transaction has to wait or abort. Eagerly detecting this kind of conflict
-  is not inherent to our system, future experiments may show that we
-  want to lift this restriction.
+  transaction has to wait or abort.\\
+  Eagerly detecting this kind of conflict is not inherent to our
+  system, future experiments may show that we want to lift this
+  restriction.
 \end{description}
@@
 Our TM system is designed as a library that covers all aspects around
-transactions and object management. The library consists of two parts:
-(I) It provides a simple interface to starting and committing
-transactions, as well as the required read and write barriers. (II) It
-also includes a \emph{garbage collector (GC)} that is closely
-integrated with the TM part (e.g. it shares the write barrier). The
-close integration helps in order to know more about the lifetime of an
-object, as will be explained in the following sections.
+transactions and object management. It is designed for object-oriented
+dynamic language VMs as a replacement for the GIL.
+
+The library consists of two parts: (I) It provides a simple interface
+to starting and committing transactions, as well as the required read
+and write barriers. (II) It also includes a \emph{garbage collector
+(GC)} that is closely integrated with the TM part (e.g. it shares the
+write barrier). The close integration helps in order to know more
+about the lifetime of an object, as will be explained in the following
+sections.
 
@@
 a read or a write barrier on an object and \lstinline!stm_allocate()!
-allocates a new object with the specified size (must be a multiple of
-16). \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
+allocates a new object with the specified size.
+ \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
 and pop objects on the shadow stack~\footnote{A stack for pointers to
@@
 The type \lstinline!object_t!  is special as it causes the
-compiler~\footnote{Clang 3.5 with some patches to this address-space
+compiler~\footnote{Clang 3.5 with some patches to its address-space
  256 feature} to make all accesses through it relative to the $\%gs$
@@
 \lstinline!mmap()! and partition this space into $N+1$ segments. We
-want to run $N$ threads in parallel while segment 0 is designated as
+want to run $N$ threads in parallel while segment~0 is designated as
 the \emph{sharing-segment} that is never assigned to a thread.
@@
 generational and has two generations: the \emph{young} and the
-\emph{old} generation.
+\emph{old} generation. It is optimised for dynamic languages with
+high allocation rates.
 
@@
 medium-fast path inside barriers that the object still belongs to the
-current transaction.
+current transaction. \remi{so this is where we mention how the GC-STM
+integration is useful. highlight more or move to own section?}
 
@@
 in the old object space when getting them back from the stack using
-\lstinline!STM_POP_ROOT()!.
+\lstinline!STM_POP_ROOT()!. \remi{cite something which explains
+shadowstacks in more detail}
 
@@
 from the start of the segment by an object's reference ($SO$) divided
-by 16 (this is where the requirement of objects to be of at least 16
-bytes in size comes from). Instead of just setting the byte to
-\lstinline!true!  if the corresponding object was read, we set it to a
-\lstinline!read_version!  belonging to the transaction, which will be
-incremented on each commit.  Thereby, we can avoid resetting the bytes
-to \lstinline!false!  on commit and only need to do this every 255
+by 16 (this requires objects to have a size which is dividable by 16).
+Instead of just setting the byte to \lstinline!true!  if the
+corresponding object was read, we set it to a \lstinline!read_version!
+belonging to the transaction, which will be incremented on each
+commit.  Thereby, we can avoid resetting the bytes to
+\lstinline!false!  on commit and only need to do this every 255
 transactions. The whole code for the barrier is easily optimisable for
@@
 The job of the write barrier is twofold: first, it serves as a write
-barrier for the garbage collector and second, it supports
-copy-on-write and adds objects to the write set of the transaction.
+barrier for the garbage collector and second, it supports COW and adds
+objects to the write set of the transaction.
 
@@
 alive and to update its reference to the new location it gets moved
-to. The check for \lstinline!is_overflow_obj()!  tells us if the
-object was actually created in this transaction. In that case, we do
-not need to execute the following \emph{TM part}.  We especially do
-not need to privatise the page since no other transaction knows about
-these ``old'' objects.
-
-For TM, we first perform a read barrier on the object. We then try to
-acquire its write lock. \lstinline!write_locks!  again is a simple
-global array of bytes that is indexed with the $SO$ of the object
-divided by 16. If we already own the lock, we are done.  If someone
-else owns the lock, we will do a write-write contention management
-that will abort either us or the current owner of the object.  If we
-succeed in acquiring the lock using an atomic
+to.
+
+The check for \lstinline!is_overflow_obj()! looks at the
+\lstinline!overflow_number!  and tells us if the object was actually
+created in this transaction. In that case, we do not need to execute
+the following \emph{TM part}.  We especially do not need to privatise
+its pages since no other transaction knows about these overflow
+objects. Even if they reside in non-private pages, it is guaranteed
+that no other transaction can have a reference to them.
+
+For the \emph{TM part}, we first perform a read barrier on the
+object. We then try to acquire its write lock. \lstinline!write_locks!
+again is a simple global array of bytes that is indexed with the $SO$
+of the object divided by 16. If we already own the lock, we are done.
+If someone else owns the lock, we will do a write-write contention
+management that will abort either us or the current owner of the
+object.  If we succeed in acquiring the lock using an atomic
 \lstinline!cmp_and_swap!, we need to add the object to the write set
@@
 object before we return. Thus, we never trigger the slow path again
-before we do the next minor collection (also part of a commit) or we
-start the next transaction.
+before we do the next minor collection or we start the next
+transaction (we always do a minor collection during a commit).
 
@@
 \lstinline!stm_start_transaction()!.  Increasing the
-\lstinline!read_version!  is also done there.
+\lstinline!read_version! for the next transaction is also done there.
 
@@
 all threads so that the committing one is the only one running and all
-the others are waiting in a safe point. We then go through the write
+the others are waiting in safe points. We then go through the write
 set (\lstinline!modified_old_objects!)  and check the corresponding
-\lstinline!read_markers!  in other threads/segments. If we detect a
+\lstinline!read_markers!  in other threads~/ segments. If we detect a
 read-write conflict, we do contention management to either abort us or
@@
 performed regularly. So what we do is to set the
-\lstinline!nursery_end!  to $0$ in all segments that we want to
-synchronise. The mentioned check will then fail in those segments and
-call the slow path. In \lstinline!allocate_slowpath!  they can simply
-check for this condition and enter a safe point.
+\lstinline!nursery_end!  to some small number in all segments that we
+want to synchronise. The mentioned check will then fail in those
+segments and call the slow path. In \lstinline!allocate_slowpath!
+they can simply check for this condition and enter a safe point.
 

### 2014-05-22 talk/dls2014/paper/paper.tex
Comment

@@
 \newcommand\cfbolz[1]{\mynote{cfbolz}{#1}}
+\newcommand\arigo[1]{\mynote{arigo}{#1}}
 
@@
   that performs well even on low numbers of CPUs. It uses a novel
-  combination of hardware features and garbage collector (GC)
+  combination of hardware features\arigo{"OS-level feature" maybe.
+  "Hardware feature" implies it only works on custom chips}
+  and garbage collector (GC)
   integration in order to keep the overhead of STM very low.

### 2014-05-27 talk/dls2014/paper/paper.tex
talk a bit about jit-integration

@@
 
+We evaluate our system in a Python interpreter called
+PyPy\footnote{www.pypy.org}. PyPy is an implementation of an
+interpreter for the Python language. It has a special focus on speed,
+as it provides a just-in-time (JIT) compiler to speed up applications
+running on top of it. For comparison, we also do evaluation on other
+Python interpreters:
+\begin{description}
+\item[CPython] is the reference implementation of the Python
+  language. It is the most widely used interpreter for this language.
+  The implementation uses a GIL for synchronisation in multi-threaded
+  execution and it does not feature a JIT compiler.
+\item[Jython] is an implementation of Python on top of the Java Virtual
+  Machine (JVM). Instead of a GIL, this interpreter uses fine-grained
+  locking for synchronisation. This enables true parallelism when
+  executing code on multiple threads. In addition, its integration
+  with the JVM provides it with a JIT compiler for faster execution.
+\end{description}
+
+Here, we will not go into detail about the integration of our STM
+system with PyPy's JIT. In fact, we will disable it for all benchmarks
+except those in section \ref{sec:real-world-bench}. We would like to
+regard it as a simple performance enhancement, but that is not what
+happens in reality. First, since the JIT is a tracing
+JIT\remi{explain?}  running in multiple threads, it may compile
+different things in each run because of the non-deterministic
+thread-scheduling of the operating system (OS). Second, it is able to
+remove some allocations in some cases. Because compilation is already
+non-deterministic, so is this allocation-removal. And third, we did
+not have enough time to optimise integration with STM so that the JIT
+exposes the overhead of STM more by speeding up all the rest.
+
+Overall, we believe that disabling it on all benchmarks except the
+real-world benchmarks in section \ref{sec:real-world-bench} is better
+because we can minimise non-determinism. We also do not want to depend
+on the capabilities of the JIT in these experiments.
+
+% benchmarks with: pypy-c--Ojit-d1454093dd48+-14-05-26-17:16
+% that's with stmgc 70c403598485
+
+% Sometimes with JIT, sometimes without.
+% For scaling & memory w/o jit, since the jit can optimize away
+% many allocations and exposes the overhead more.
+
 \subsection{Memory Requirements}
@@
 
-\subsection{Real-World Benchmarks}
+\subsection{Real-World Benchmarks\label{sec:real-world-bench}}
 

### 2014-05-27 talk/dls2014/paper/paper.tex
add plot to paper
add test

@@
 
+There are several sources of extra memory requirements in our
+STM system. First, we need to keep track of the state an object
+is in. We do this using flags and an overflow number. Both currently
+fit in an additional header of 4~bytes per object.
+
+Second, there are areas in memory private to each segment (see
+section \ref{sub:Setup}). The Nursery for example is 4~MiB in
+size. Additionally, the area used for the read markers is lazily
+mapped by the kernel and only needed for objects in the old object
+space. Thus, it can be at most the total size of memory used by
+old objects divided by 16.
+
+Third, since we do copy on write, if all objects were to be
+written to at the same time, all pages would need to be privatised
+for all objects in the old object space. In that case we would
+need the total amount of memory required by old objects multiplied
+by $N+1$ (incl. the sharing segment).
+\remi{maybe collect some statistics about pages privatised per segment}
+
 \begin{itemize}
@@
 maybe some memory usage graph over time
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\columnwidth]{plots/richards_mem.pdf}
+  \caption{Actual memory managed by the GC and resident set size
+    over time in Richards benchmark\label{fig:richards_mem}}
+\end{figure}
 

### 2014-05-27 talk/dls2014/paper/paper.tex
text

@@
 We evaluate our system in a Python interpreter called
-PyPy\footnote{www.pypy.org}. PyPy is an implementation of an
+PyPy\footnote{www.pypy.org} (version 2.3). PyPy is an implementation of an
 interpreter for the Python language. It has a special focus on speed,
 as it provides a just-in-time (JIT) compiler to speed up applications
-running on top of it. For comparison, we also do evaluation on other
-Python interpreters:
+running on top of it. For comparison, we compare between normal PyPy
+using a GIL and a PyPy with STM, as well as between other Python interpreters:
 \begin{description}
-\item[CPython] is the reference implementation of the Python
+\item[CPython] (version 2.7.6) is the reference implementation of the Python
   language. It is the most widely used interpreter for this language.
@@
   execution and it does not feature a JIT compiler.
-\item[Jython] is an implementation of Python on top of the Java Virtual
-  Machine (JVM). Instead of a GIL, this interpreter uses fine-grained
-  locking for synchronisation. This enables true parallelism when
-  executing code on multiple threads. In addition, its integration
-  with the JVM provides it with a JIT compiler for faster execution.
+\item[Jython] (version 2.7b1) is an implementation of Python on top of
+  the Java Virtual Machine (JVM). Instead of a GIL, this interpreter
+  uses fine-grained locking for synchronisation. This enables true
+  parallelism when executing code on multiple threads. In addition, its
+  integration with the JVM provides it with a JIT compiler for faster
+  execution.
 \end{description}
@@
 
+We performed all benchmarks on a machine with a Intel Core i7-4770
+CPU~@3.40GHz (4 cores, 8 threads).  There are 16~GiB of memory
+available and we ran them under Ubuntu 14.04 with a Linux 3.13.0
+kernel. The STM system was compiled with a number of segments $N=4$
+and a maximum amount of memory of 1.5~GiB (both are configurable at
+compile time).
+
 % benchmarks with: pypy-c--Ojit-d1454093dd48+-14-05-26-17:16
@@
 is in. We do this using flags and an overflow number. Both currently
-fit in an additional header of 4~bytes per object.
+fit in a single additional header of 4~bytes per object.
 
@@
 need the total amount of memory required by old objects multiplied
-by $N+1$ (incl. the sharing segment).
+by $N+1$ (incl. the sharing segment). Pages get re-shared during
+major collections if possible.
 \remi{maybe collect some statistics about pages privatised per segment}
 
-\begin{itemize}
-\item stm\_flags per object
-\item read markers and other sections
-\item private pages
-\end{itemize}
-
-maybe some memory usage graph over time
+\remi{The following discussion about richards mem usage does not
+say that much... Also, RSS is not a good measure but it's hard to
+get something better.}
+In figure \ref{fig:richards_mem} we look at the memory usage of
+one of our benchmarks called Richards\footnote{OS kernel simulation
+benchmark}. The \emph{Resident Set Size} (RSS) shows the physical memory
+assigned to the process. From it, we see that the process' memory
+usage does not explode during the benchmark but actually stays pretty
+much the same after start-up. Since it is the job of the OS to map
+physical memory, this RSS number should be seen as a maximum. It is
+possible that some of the memory is not required any more but still
+assigned to our process.
+
+The \emph{GC managed memory} counts all memory used in the old object
+space including the memory required for private pages. The sharp drops
+in memory usage come from major collections that free old objects and
+re-share pages. Again the overall memory usage stays the same and
+we see that in this benchmark we have around 1 major collection every
+second.
+
+For PyPy-STM the average memory requirement is 29~MiB and there are
+$\sim 11$ major collections during the runtime. Normal PyPy with a GIL
+grows its memory up to just 7~MiB and does not do a single major
+collection in that time.
+
+We are missing a memory optimisation to store small objects in a more
+compact way, which is done by a normal PyPy not using STM.
+Additionally, since normal PyPy uses a GIL, it does not need to
+duplicate any data structures like e.g. the Nursery for each
+thread. This, the missing optimisation, and the additional memory
+requirements for STM explained above account for this difference.
+\remi{I don't know how much sense it makes to go deeper. We will
+improve this in the future, but right now this is the overall picture.}
 
@@
 
+\remi{gs:segment prefix overhead is virtually none (maybe instruction cache)}
+\remi{update numbers in pypy/TODO}
+
 \begin{itemize}
@@
 \item time spent committing \& aborting (maybe with different numbers
-  of threads)
+  of threads; maybe split conflict detection and obj sync on commit)
 \item time in GC

### 2014-05-27 talk/dls2014/paper/paper.tex
remove rss comparison and try with some page privatisation statistics

@@
 major collections if possible.
-\remi{maybe collect some statistics about pages privatised per segment}
-
-\remi{The following discussion about richards mem usage does not
-say that much... Also, RSS is not a good measure but it's hard to
-get something better.}
-In figure \ref{fig:richards_mem} we look at the memory usage of
-one of our benchmarks called Richards\footnote{OS kernel simulation
-benchmark}. The \emph{Resident Set Size} (RSS) shows the physical memory
-assigned to the process. From it, we see that the process' memory
-usage does not explode during the benchmark but actually stays pretty
-much the same after start-up. Since it is the job of the OS to map
-physical memory, this RSS number should be seen as a maximum. It is
-possible that some of the memory is not required any more but still
-assigned to our process.
-
-The \emph{GC managed memory} counts all memory used in the old object
-space including the memory required for private pages. The sharp drops
-in memory usage come from major collections that free old objects and
-re-share pages. Again the overall memory usage stays the same and
-we see that in this benchmark we have around 1 major collection every
-second.
+
+In figure \ref{fig:richards_mem} we look at the memory usage of one of
+our benchmarks called Richards\footnote{OS kernel simulation
+benchmark}. The \emph{GC managed memory} counts all memory used in the
+old object space including the memory required for private pages. The
+sharp drops in memory usage come from major collections that free old
+objects and re-share pages. The average memory usage stays around
+29~MiB and we see that in this benchmark we have around 1 major
+collection every second. The \emph{page privatisation}, which
+represents the percentage of used pages with at least one private
+copy, shows the same spikes as the memory usage. These come directly from
+re-sharing the pages. The maximum page privatisation is around $20\%$
+between major collections. Thus we can say that $~20\%$ of the old
+objects get modified between collections in this benchmark.
 
@@
 grows its memory up to just 7~MiB and does not do a single major
-collection in that time.
-
-We are missing a memory optimisation to store small objects in a more
-compact way, which is done by a normal PyPy not using STM.
-Additionally, since normal PyPy uses a GIL, it does not need to
+collection in that time. Compared to normal PyPy, we are missing a
+memory optimisation to store small objects in a more compact
+way. Additionally, since normal PyPy uses a GIL, it does not need to
 duplicate any data structures like e.g. the Nursery for each
@@
   \includegraphics[width=1\columnwidth]{plots/richards_mem.pdf}
-  \caption{Actual memory managed by the GC and resident set size
+  \caption{Actual memory managed by the GC and the page privatisation
     over time in Richards benchmark\label{fig:richards_mem}}
@@
 
+\remi{do it on a non-jit build (see reason above)}
 \remi{gs:segment prefix overhead is virtually none (maybe instruction cache)}

### 2014-05-27 talk/dls2014/paper/paper.tex
again change the metric for explaining memory usage (not sure what's best)

@@
 \begin{figure*}[t]
-  \begin{centering}
+  \centering
   \includegraphics[scale=0.8]{\string"segment addressing\string".pdf}
-    \par\end{centering}
-
-    \protect\caption{Segment Addressing\label{fig:Segment-Addressing}}
+  \caption{Segment Addressing\label{fig:Segment-Addressing}}
 \end{figure*}
@@
 \begin{figure}[h]
-  \begin{centering}
+  \centering
   \includegraphics[scale=0.8]{\string"mmap pages\string".pdf}
-    \par\end{centering}
-
-    \protect\caption{\texttt{mmap()} Page Mapping\label{fig:mmap()-Page-Mapping}}
+  \caption{\texttt{mmap()} Page Mapping\label{fig:mmap()-Page-Mapping}}
 \end{figure}
@@
 \begin{figure}[h]
-  \begin{centering}
+  \centering
   \includegraphics[width=1\columnwidth]{\string"page remapping\string".pdf}
-    \par\end{centering}
-
-    \protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
+  \caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
     segment~0, fully shared memory configuration. (III) privatise single
@@
 \begin{figure*}[t]
-  \begin{centering}
+  \centering
   \includegraphics[scale=0.8]{\string"segment layout\string".pdf}
-    \par\end{centering}
-
-    \protect\caption{Segment Layout\label{fig:Segment-Layout}}
+  \caption{Segment Layout\label{fig:Segment-Layout}}
 \end{figure*}
@@
 
-Third, since we do copy on write, if all objects were to be
-written to at the same time, all pages would need to be privatised
-for all objects in the old object space. In that case we would
-need the total amount of memory required by old objects multiplied
-by $N+1$ (incl. the sharing segment). Pages get re-shared during
-major collections if possible.
+Third, since we do copy on write, if all objects were to be written to
+at the same time, all pages would need to be privatised for all
+objects in the old object space. In that case we would need the total
+amount of memory required by old objects multiplied by $N+1$
+(incl. the sharing segment). During major collections we re-share the
+pages if possible.
 
@@
 29~MiB and we see that in this benchmark we have around 1 major
-collection every second. The \emph{page privatisation}, which
-represents the percentage of used pages with at least one private
-copy, shows the same spikes as the memory usage. These come directly from
-re-sharing the pages. The maximum page privatisation is around $20\%$
-between major collections. Thus we can say that $~20\%$ of the old
-objects get modified between collections in this benchmark.
+collection every second.
+
+The \emph{page privatisation}, which represents the ratio between
+private pages and shared pages, gives us an idea about how many
+private copies exist for each shared page. If the ratio is $>1.0$, it
+means that on average we have more than one private copy for each
+shared page. In the worst case described above, this number would
+reach $N$ and means we need $N\times$ the memory for the private
+pages alone. In this benchmark, we see the same spikes as the memory
+usage. These come directly from re-sharing the pages. The maximum page
+privatisation is around $0.5$ between major collections, thus the
+private pages are responsible for a $50\%$ increase in the required
+memory. Since the spikes in the GC managed memory line actually
+show increases by $~80\%$, it means that the rest comes from actual
+garbage objects that were collected.
 

### 2014-05-27 talk/dls2014/paper/plots/plot_scaling.py
re-run with no-jit build and add scaling micro-bench

@@ -1,9 +1,7 @@
 #!/usr/bin/python
 
-# obtained log-file with
-#   pypy-c --jit off ~/pypy/benchmarks/multithread/multithread-richards.py 60 4 2>richards_mem.log
-# rss using measure_memusage.sh
-
+# obtained with time on
+#   pypy-c --jit off bench_scaling.py [1-4]
 
 
 import matplotlib
@@ -21,37 +19,30 @@ args = None
 import matplotlib.pyplot as plt
 # import pprint - slow as hell
 
-with open('richards_mem.log') as f:
-    xs = []
-    y1s = []
-    y2s = []
-    first_time = None
-    for line in f.readlines():
-        line = line.strip().strip("{").strip("}")
-        time, mems = line.split(":")
-        if not first_time:
-            first_time = float(time)
-        xs.append(float(time) - first_time)
-        real_mem, max_rss, page_util = mems.split("/")
-        y1s.append(int(real_mem) / 1024. / 1024)
-        y2s.append(float(page_util))
-
-# RSS:
-# x2s = range(12)
-# y2s = [152304, 180060, 180428,
-#        180448, 180460, 180696,
-#        180124, 180552, 180584,
-#        180588, 180544, 180252]
-# y2s = map(lambda x: x / 1024., y2s)
+xs = range(1,5)
+ys = [[1.73, 1.74, 1.73, 1.73, 1.74],
+      [1.75, 1.77, 1.78, 1.75, 1.75],
+      [1.8, 1.79, 1.76, 1.76, 1.79],
+      [1.82, 2.1, 1.84, 1.9, 2.13]]
 
 
-def plot_mems(ax, ax2):
-    print sum(y1s) / len(xs)
-    print sum(y2s) / len(xs)
-    a, = ax.plot(xs, y1s, 'b-o', ms=3)
-    b, = ax2.plot(xs, y2s, 'r-x', ms=3)
-    return ax.legend((a, b),
-                     ('GC managed memory', 'Page privatisation'))
+
+def plot_mems(ax):
+    import numpy as np
+    y = []
+    yerr = []
+    opt_y = [1.0] * len(xs)
+    first_time = np.mean(ys[0])
+    for x, d in zip(xs, ys):
+        normalized = map(lambda x:x/first_time, d)
+        y.append(np.mean(normalized))
+        yerr.append(np.std(normalized))
+
+    print y
[…]

### 2014-05-27 talk/dls2014/paper/paper.tex
re-run with no-jit build and add scaling micro-bench

@@
 usage. These come directly from re-sharing the pages. The maximum page
-privatisation is around $0.5$ between major collections, thus the
-private pages are responsible for a $50\%$ increase in the required
+privatisation is around $0.9$ between major collections, thus the
+private pages are responsible for a $90\%$ increase in the required
 memory. Since the spikes in the GC managed memory line actually
-show increases by $~80\%$, it means that the rest comes from actual
+show increases by $~100\%$, it means that the rest comes from actual
 garbage objects that were collected.
 
-For PyPy-STM the average memory requirement is 29~MiB and there are
-$\sim 11$ major collections during the runtime. Normal PyPy with a GIL
+For PyPy-STM the average memory requirement is 21~MiB and there are
+$\sim 10$ major collections during the runtime. Normal PyPy with a GIL
 grows its memory up to just 7~MiB and does not do a single major
@@
 requirements for STM explained above account for this difference.
+We expect to improve this aspect in the future, in this paper we
+want to focus first on performance.
 \remi{I don't know how much sense it makes to go deeper. We will
@@
 
-maybe some simple micro benchmarks with adaptable conflict rate
+To asses how well the STM system scales on its own (without any real
+workload), we execute the following loop on 1 to 4 threads:
+\begin{lstlisting}
+def workload():
+    i = 20000000
+    while i:
+        i -= 1
+\end{lstlisting}
+
+For the results in figure \ref{fig:scaling}, we averaged
+over 5 runs and normalised the average runtimes to the
+time it took on a single thread. From this we see that there
+is additional overhead introduced by each thread ($13\%$
+for all 4 threads together).
+
+\remi{what we don't show is by how much this overhead is influenced
+by allocations}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\columnwidth]{plots/scaling.pdf}
+  \caption{Scalability of the STM system\label{fig:scaling}}
+\end{figure}
 

### 2014-05-28 talk/dls2014/paper/paper.tex
add first (bad) results for JIT in our benchmarks

@@
 system with PyPy's JIT. In fact, we will disable it for all benchmarks
-except those in section \ref{sec:real-world-bench}. We would like to
+except those in section \ref{sec:performance-bench}. We would like to
 regard it as a simple performance enhancement, but that is not what
@@
 Overall, we believe that disabling it on all benchmarks except the
-real-world benchmarks in section \ref{sec:real-world-bench} is better
+performance benchmarks in section \ref{sec:performance-bench} is better
 because we can minimise non-determinism. We also do not want to depend
@@
 
-\subsection{Real-World Benchmarks\label{sec:real-world-bench}}
+\subsection{Performance Benchmarks\label{sec:performance-bench}}
+
+\remi{For performance we first look at no-JIT behaviour of STM. Since
+we cannot compete well even with CPython, we later show JIT benchmarks
+where we see the unstable performance but also that we can still scale.
+(with more work we can use our STM system to parallelise jitted code
+too)}
 
@@
 
+% TODO: Jython
+\remi{Some benchmarks (figure \ref{fig:performance-jit} with enabled
+JIT show that we can be competitive with the other solutions. It also
+shows that more work is needed in that area to make performance more
+stable.}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\columnwidth]{plots/performance.pdf}
+  \caption{Comparing runtime between interpreters with JIT\label{fig:performance-jit}}
+\end{figure}
+
 \section{Related Work}

### 2014-05-28 talk/dls2014/paper/plots/plot_performance.py

@@ -1,0 +1,227 @@
+#!/usr/bin/python
+
+# benchmarks-repo at 0d81c9b1ec8e
+
+# for now: avg & stddev of the best
+
+#    pypy-c-paper-jit bench.py -k5 raytrace/raytrace.py 1-4
+#    pypy-c-paper-jit bench.py -k5 btree/btree.py 1-4
+#    pypy-c-paper-jit bench.py -k5 skiplist/skiplist.py 1-4
+#    pypy-c-paper-jit bench.py -k5 threadworms/threadworms.py 1-4
+#    pypy-c-paper-jit bench.py -k5 mandelbrot/mandelbrot.py 1-4 64
+#    pypy-c-paper-jit multithread-richards.py 10000 1-4 # report runtime
+
+
+
+import matplotlib
+import os
+import sys
+matplotlib.use('gtkagg')
+
+from matplotlib import rc
+#rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
+## for Palatino and other serif fonts use:
+rc('font',**{'family':'serif','serif':['Palatino']})
+rc('text', usetex=True)
+
+args = None
+import matplotlib.pyplot as plt
+# import pprint - slow as hell
+
+# threads
+ts = range(1,5)
+
+interps_styles = {
+    "pypy-stm-jit": 'r-',
+    "pypy-jit": 'b--',
+    "best": "k:"
+}
+
+benchs = {
+    "raytrace":{
+        "pypy-stm-jit":[
+            [3.91, 3.87],
+            [2.53, 2.52],
+            [2.23],
+            [2.46, 2.6]
+        ],
+        "pypy-jit":[
+            [1.6],
+            [2.17],
+            [3.33],
+            [4.16]
+        ]},
+
+    "btree":{
+        "pypy-stm-jit":[
+            [1.68],
+            [1.3],
+            [1.39],
+            [1.66]
+        ],
+        "pypy-jit":[
+            [1.6],
+            [3.3],
+            [5.1],
+            [5.8]
+        ]},
+
+    "skiplist":{
+        "pypy-stm-jit":[
+            [2.9],
+            [3.0],
+            [3.4],
+            [3.8]
+        ],
+        "pypy-jit":[
+            [2.14],
+            [4.5],
+            [6.2],
+            [6.58]
+        ]},
+
+    "threadworms":{
+        "pypy-stm-jit":[
+            [4.23],
+            [3.4],
+            [3.16],
[…]

### 2014-05-28 talk/dls2014/paper/plots/plot_performance.py
add some no-jit performance graphs

@@ -29,7 +29,7 @@ import matplotlib.pyplot as plt
 # import pprint - slow as hell
 
 # threads
-ts = range(1,5)
+
 
 interps_styles = {
     "pypy-stm-jit": 'r-',
@@ -126,13 +126,13 @@ benchs = {
 
 
 
-def plot_speedups(plt):
+def plot_speedups(plt, w, h, benchs, interps_styles):
     import numpy as np
     from collections import OrderedDict
     fig = plt.figure()
 
+    ts = range(1,5) # threads
     legend = OrderedDict()
-    w, h = 2, 3
     axs = {}
     for i, (name, contestants) in enumerate(benchs.items()):
         if i >= w:
@@ -180,7 +180,7 @@ def main():
     global fig
 
     print "Draw..."
-    legend = plot_speedups(plt)
+    legend = plot_speedups(plt, 2, 3, benchs, interps_styles)
 
     #axs[0].set_ylim(0, len(x))
     #ax.set_yticks([r+0.5 for r in range(len(logs))])

### 2014-05-28 talk/dls2014/paper/paper.tex
add some no-jit performance graphs

@@
 (with more work we can use our STM system to parallelise jitted code
-too)}
-
-more real benchmarks comparing multiple implementations:
-\begin{itemize}[noitemsep]
-\item pypy
-\item pypy-jit
-\item pypy-stm
-\item pypy-stm-jit
-\item cpython
-\item jython
-\end{itemize}
-
-% TODO: Jython
+too)} See figure \ref{fig:performance-nojit}
+
+% TODO: pypy-nostm, Jython?
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
+  \caption{Comparing runtime between interpreters without a JIT\label{fig:performance-nojit}}
+\end{figure}
+
+% TODO: Jython, compare to cpython? or just jython as common baseline with no-jit?
 \remi{Some benchmarks (figure \ref{fig:performance-jit} with enabled

### 2014-05-28 talk/dls2014/paper/plots/plot_performance.py
add jython to the mix

@@ -32,9 +32,10 @@ import matplotlib.pyplot as plt
 
 
 interps_styles = {
-    "pypy-stm-jit": 'r-',
-    "pypy-jit": 'b--',
-    "best": "k:"
+    "pypy-stm-jit": {'fmt':'r-'},
+    "pypy-jit": {'fmt':'b', 'dashes':(1,1)},
+    "jython": {'fmt':'m', 'dashes':(2, 5)},
+    "best": {'fmt':"k:"}        # only fmt allowed
 }
 
 benchs = {
@@ -156,12 +157,12 @@ def plot_speedups(plt, w, h, benchs, interps_styles):
                     best_y = new_y
 
             artist = ax.errorbar(ts, y, yerr=yerr,
-                                 fmt=interps_styles[interp])
+                                 **interps_styles[interp])
             if interp not in legend:
                 legend[interp] = artist
 
         legend["best"], = ax.plot(ts, [best_y] * len(ts),
-                                  interps_styles["best"])
+                                  interps_styles["best"]['fmt'])
 
         if i // w == h-1:
             ax.set_xlim(0, 5)
@@ -172,7 +173,7 @@ def plot_speedups(plt, w, h, benchs, interps_styles):
         ax.set_title(name)
 
     return axs[w*(h-1)].legend(tuple(legend.values()), tuple(legend.keys()),
-                               ncol=3,
+                               ncol=4,
                                loc=(0,-0.4))
 
 

### 2014-05-29 talk/dls2014/paper/plots/plot_performance.py
more text in the performance section

@@ -32,7 +32,7 @@ import matplotlib.pyplot as plt
 
 
 interps_styles = {
-    "pypy-stm-jit": {'fmt':'r-'},
+    "pypy-stm-jit": {'fmt':'r-', 'linewidth':2},
     "pypy-jit": {'fmt':'b', 'dashes':(1,1)},
     "jython": {'fmt':'m', 'dashes':(2, 5)},
     "best": {'fmt':"k:"}        # only fmt allowed
@@ -161,8 +161,8 @@ def plot_speedups(plt, w, h, benchs, interps_styles):
             if interp not in legend:
                 legend[interp] = artist
 
-        legend["best"], = ax.plot(ts, [best_y] * len(ts),
-                                  interps_styles["best"]['fmt'])
+        # legend["best"], = ax.plot(ts, [best_y] * len(ts),
+        #                           interps_styles["best"]['fmt'])
 
         if i // w == h-1:
             ax.set_xlim(0, 5)
@@ -174,7 +174,7 @@ def plot_speedups(plt, w, h, benchs, interps_styles):
 
     return axs[w*(h-1)].legend(tuple(legend.values()), tuple(legend.keys()),
                                ncol=4,
-                               loc=(0,-0.4))
+                               loc=(-0.15,-0.5))
 
 
 def main():

### 2014-05-29 talk/dls2014/paper/paper.tex
add jython to the mix
more text in the performance section

@@
 
-\remi{For performance we first look at no-JIT behaviour of STM. Since
-we cannot compete well even with CPython, we later show JIT benchmarks
-where we see the unstable performance but also that we can still scale.
-(with more work we can use our STM system to parallelise jitted code
-too)} See figure \ref{fig:performance-nojit}
-
-% TODO: pypy-nostm, Jython?
+In this section we want to look at the general performance of our
+system. As explained above, the JIT does not simply speed up the
+performance by a certain factor. In order to better understand the
+behaviour of our system, we therefore split this section into two
+parts. We first look at how it behaves without the JIT, and then with
+the JIT. We look at six benchmarks:
+
+\begin{itemize}
+\item \emph{btree} and \emph{skiplist}, which are both inserting,
+  removing, and finding elements in a data structure
+\item \emph{threadworms}, which simulates worms walking on a grid in
+  parallel and checking for collisions with each other
+\item \emph{mandelbrot}, \emph{raytrace}, and \emph{richards}, which
+  all perform simple, independent computations in parallel
+\end{itemize}
+
+We use coarse-grained locking for the first three benchmarks and no
+locking for the last three. This is important because Jython, which
+uses fine-grained locking instead of a GIL, is only expected to scale
+with the number of threads for the latter group. It is not able to
+scale when using coarse-grained locking. STM, however, uses atomic
+blocks instead, so it may still be able to scale since they are
+implemented as simple transactions.
+
+% To isolate factors we look at performance w/o JIT and perf w JIT.
+% w/o JIT:
+%  - it scales
+%  - slower than reference CPython
+%  - gil scales negatively
+% w JIT:
+%  - it sometimes still scales (more work needed)
+%  - changed working set size because generally ~50x faster
+%  - competing w our own JIT on non-stm a challenge
+%  - gil scales negatively
+
+\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
+different interpreters: Jython (fine-grained locking), CPython (GIL),
+and PyPy with STM and with the GIL (both without the JIT). The results
+are shown in \ref{fig:performance-nojit}.
+
+As expected, all interpreters with a GIL do not scale with the number
+of threads. They even become slower because of the overhead of
+thread-switching and GIL handling. We also see Jython scale when we
+expect it to (mandelbrot, raytrace, richards), and behave similar to
+the GIL interpreters in the other cases.
+
+PyPy using our STM system (pypy-stm-nojit) scales in all benchmarks to
+a certain degree. We see that the average overhead from switching from
+GIL to STM is \remi{$35.5\%$}, the maximum in richards is
+\remi{$63\%$}. pypy-stm-nojit beats pypy-nojit already on two threads;
+however, it never even beats CPython, the reference implementation of
+Python. This means that without the JIT, our performance is not
+competitive. We now look at how well our system works when we enable
+the JIT.
+
 \begin{figure}[h]
@@
 
-% TODO: Jython, compare to cpython? or just jython as common baseline with no-jit?
-\remi{Some benchmarks (figure \ref{fig:performance-jit} with enabled
-JIT show that we can be competitive with the other solutions. It also
-shows that more work is needed in that area to make performance more
-stable.}
+\paragraph{JIT benchmarks:} The speedups from enabling the JIT in
+these benchmarks range from $10-50\times$. This is why we had to do
+without Jython and CPython here, since they would be much further up
+in the plots. Also, in order to get more stable results, we increased
+the input size of all benchmarks to get reasonable execution times.
+
+The results are shown in \ref{fig:performance-nojit}. We see that the
+performance is much less stable. There is certainly more work required
+in this area. In general, we see that the group of non-locked
+benchmarks certainly scales best. The other three scale barely or not
+at all with the number of threads. The slowdown factor from GIL to STM
+ranges around \remi{$1-2.4\times$} and we beat GIL performance in half
+of the benchmarks.
 
@@
 \end{figure}
+
+Overall PyPy needs the JIT in order for its performance to be
+competitive.  It would be interesting to see how using our STM system
+in CPython would perform, but it is a lot of work. On its own, our
+system scales well so we hope to also see that with the JIT in the
+future.
 

### 2014-05-29 talk/dls2014/paper/paper.tex
tweaks

@@
 
-We performed all benchmarks on a machine with a Intel Core i7-4770
+We performed all benchmarks on a machine with an Intel Core i7-4770
 CPU~@3.40GHz (4 cores, 8 threads).  There are 16~GiB of memory
@@
 pages if possible.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\columnwidth]{plots/richards_mem.pdf}
+  \caption{Actual memory managed by the GC and the page privatisation
+    over time in Richards benchmark\label{fig:richards_mem}}
+\end{figure}
 
@@
 requirements for STM explained above account for this difference.
-We expect to improve this aspect in the future, in this paper we
+We expect to improve this aspect in the future; in this paper we
 want to focus first on performance.
@@
 improve this in the future, but right now this is the overall picture.}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=1\columnwidth]{plots/richards_mem.pdf}
-  \caption{Actual memory managed by the GC and the page privatisation
-    over time in Richards benchmark\label{fig:richards_mem}}
-\end{figure}
 
@@
 To asses how well the STM system scales on its own (without any real
-workload), we execute the following loop on 1 to 4 threads:
+workload), we execute the following loop on 1 to 4 threads on our
+Python interpreter with STM:
 \begin{lstlisting}
@@
 
-For the results in figure \ref{fig:scaling}, we averaged
-over 5 runs and normalised the average runtimes to the
-time it took on a single thread. From this we see that there
-is additional overhead introduced by each thread ($13\%$
-for all 4 threads together).
+For the results in figure \ref{fig:scaling}, we averaged over 5 runs
+and normalised the average runtimes to the time it took on a single
+thread. From this we see that there is additional overhead introduced
+by each thread (\remi{$13\%$} for all 4 threads together). While this
+is not ideal, we think that \remi{$13\%$} are acceptable on four
+threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
+more iterations per second than a single thread.
 

### 2014-06-02 talk/dls2014/paper/paper.tex
Some work on the introduction, background

@@
 introduction of just-in-time compilers (JIT) for these languages (e.g.
-PyPy, V8, Tracemonkey) started to change this perception by delivering
+PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
+started to change this perception by delivering
 good performance that enables new applications. However, a parallel
@@
 atomicity between multiple threads for a series of
-instructions. Another technology that can provide the same guarantees
-is transactional memory (TM). \remi{cite our position paper}
-
-There have been several attempts at replacing the GIL with TM. Using
-transactions to enclose multiple bytecode instructions, we can get the
-very same semantics as the GIL while possibly executing several
-transactions in parallel. Furthermore, by exposing these
-interpreter-level transactions to the application in the form of
-\emph{atomic blocks}, we give dynamic languages a new synchronisation
-mechanism that avoids several of the problems of locks as they are
-used now.
-
-\remi{cite and extract from (our pos. paper):}
+instructions. Additionally, it provides the application with a
+sequential consistency model~\cite{lamport79}. Another technology that
+can provide the same guarantees is transactional memory
+(TM). \remi{cite our position paper}
+
+There have been several attempts at replacing the GIL with
+TM~\cite{nicholas06,odaira14,fuad10}. Using transactions to enclose
+multiple bytecode instructions, we can get the very same semantics as
+the GIL while possibly executing several transactions in
+parallel. Furthermore, by exposing these interpreter-level
+transactions to the application in the form of \emph{atomic
+blocks}~\cite{tim03,tim05}, we give dynamic languages a new
+synchronisation mechanism that avoids several of the problems of locks
+as they are used now.
+
 TM systems come in can be broadly categorised as hardware based (HTM),
 software based (STM), or hybrid systems (HyTM). HTM systems are limited
-by hardware constraints, while STM systems have a lot of overhead.
+by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
+a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
+we argue that STM is still the best way forward, especially since it
+supports large atomic blocks as a new way for synchronising multiple
+threads. There have been several attempts at lowering the overhead
+of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
 In this paper, we describe how we manage to lower the overhead of our
@@
   that performs well even on low numbers of CPUs. It uses a novel
-  combination of hardware features\arigo{"OS-level feature" maybe.
-  "Hardware feature" implies it only works on custom chips}
-  and garbage collector (GC)
+  combination of long-existing CPU features and garbage collector (GC)
   integration in order to keep the overhead of STM very low.
@@
 
+\subsection{Global Interpreter Lock}
+
+The GIL is a very simple synchronisation mechanism for supporting
+multithreading in an interpreter. The basic guarantee is that the GIL
+may only be released in between bytecode instructions. Thus, these
+instructions are always executed atomically and in complete isolation
+from others running in other threads. \emph{Atomicity} means that each
+instruction and its effects seem to happen at one, indivisible point
+in time. Other instructions never see inconsistent state of a
+partially executed instruction (\emph{isolation}).
+
+In addition to these guarantees, instructions are executed in a
+sequential consistency model~\cite{lamport79}. This means that
+the outcome of any execution of instructions in multiple threads is
+equal to \emph{some} sequential execution of them.
+
 \subsection{Transactional Memory}
@@
 of instructions performing operations on memory and make them happen
-atomically and in complete isolation from other
-transactions. \emph{Atomicity} means that all these instructions in
-the transaction and their effects seem to happen at one, indivisible
-point in time. Other transactions never see inconsistent state of a
-partially executed transaction which is called \emph{isolation}.
+atomically and in complete isolation from other transactions.
+Atomicity and isolation are basic properties of transactions.
 
@@
 \emph{serialisable}. Meaning, the outcome is equal to some sequential
-execution of these transactions. This means that the approach provides the same
-semantics as using the GIL
-while still allowing the TM system to
-run transactions in parallel as an optimisation.
-\remi{maybe some more explanation of how exactly TM replaces the GIL}
+execution of these transactions. By that, we can again provide a
+sequentially consistent model for programming in multiple threads. We
+can therefore use TM to directly replace the GIL. Instead of releasing
+and acquiring the GIL between bytecode instructions, we commit and
+start the transactions our instructions are running in.
 
@@
 
-\cfbolz{a pypy introduction needs to go somewhere, a paragraph or so. maybe in the evaluation section}
-
-We implement and evaluate our system for the Python language. For the
-actual implementation, we chose the PyPy interpreter because replacing
-the GIL there with a TM system is just a matter of adding a new
-transformation to the translation process of the interpreter.
-
+We implement and evaluate our system for the Python language. Python
+is a dynamic programming language that was designed with GIL semantics
+in mind. Its reference implementation, CPython~\cite{cpython}, uses a
+GIL to synchronise instructions in multiple threads.
 Over the years, Python added multiple ways to provide concurrency and
@@
 multiple threads. One approach to this is fine-grained locking instead
-of a single global lock. Jython and IronPython are implementations of
-this. It requires great care in order to avoid deadlocks, which is why
-we follow the TM approach that provides a \emph{direct} replacement
-for the GIL. It does not require careful placing of locks in the right
-spots. We will compare our work with Jython for evaluation.
+of a single global lock. Jython~\cite{webjython} and
+IronPython~\cite{ironpython} are implementations of this. It requires
+great care in order to avoid deadlocks, which is why we follow the TM
+approach that provides a \emph{direct} replacement for the GIL. It
+does not require careful placing of locks in the right spots. We will
+compare our work with Jython for evaluation.
 
@@
 
-\cfbolz{citation again needed for the whole subsection}
-
-It is well known that using locks to synchronise multiple threads is
-hard. They are non-composable, have overhead, may deadlock, limit
-scalability, and overall add a lot of complexity. For a better
-parallel programming model for dynamic languages, we want to implement
-another, well-known synchronisation mechanism: \emph{atomic blocks}.
+In Python, since the GIL is not directly exposed to the interpreter,
+applications still need to synchronise memory accesses from multiple
+threads using locks. Locks can be very hard to get
+right~\cite{christopher10,victor11,shan08}.  They are non-composable,
+have overhead, may deadlock, limit scalability, and add to the overall
+complexity of the program logic. We think that \emph{atomic
+blocks}~\cite{tim03,tim05} provide a better way for synchronisation.
 
@@
 useful atomicity and isolation guarantees to the application for a
-series of instructions. An implementation using a GIL would simply
-guarantee that the GIL is not released during the execution of the
-atomic block. Using TM, we have the same effect by guaranteeing that
-all instructions in an atomic block are executed inside a single
-transaction.
-
-\remi{STM, how atomicity \& isolation; reasons for overhead}
+series of instructions. This is why we think that the introduction
+of atomic blocks to Python is a valuable contribution. Since atomicity
+is a property of transactions, TM and atomic blocks are a natural fit.
 
 \section{Method}
+
+We now take a closer look at how our TM system that we use to replace
+the GIL works, what properties it has, and how it is implemented.
 
@@
 and PyPy with STM and with the GIL (both without the JIT). The results
-are shown in \ref{fig:performance-nojit}.
+are shown in figure \ref{fig:performance-nojit}.
 
@@
 
-The results are shown in \ref{fig:performance-nojit}. We see that the
-performance is much less stable. There is certainly more work required
-in this area. In general, we see that the group of non-locked
-benchmarks certainly scales best. The other three scale barely or not
-at all with the number of threads. The slowdown factor from GIL to STM
-ranges around \remi{$1-2.4\times$} and we beat GIL performance in half
-of the benchmarks.
+The results are presented in figure \ref{fig:performance-nojit}. We
+see that the performance is much less stable. There is certainly more
+work required in this area. In general, we see that the group of
+non-locked benchmarks certainly scales best. The other three scale
+barely or not at all with the number of threads. The slowdown factor
+from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
+performance in half of the benchmarks.
+
+\remi{Reason for bad scaling: acceleration of code that produces
+conflicts $-->$ more iterations $-->$ more conflicts. The overhead
+doesn't get accelerated by the JIT.}
 
@@
 
+\bibitem{cfbolz09} Carl Friedrich Bolz, Antonio Cuni, Maciej
+  Fijalkowski, and Armin Rigo. 2009. Tracing the meta-level: PyPy's
+  tracing JIT compiler.  \emph{In Proceedings of the 4th workshop on the
+    Implementation, Compilation, Optimization of Object-Oriented Languages
+    and Programming Systems} (ICOOOLPS '09).
+
+\bibitem{kevin10} Kevin Millikin, Florian Schneider. 2010.  A New
+  Crankshaft for V8.
+  \url{http://blog.chromium.org/2010/12/new-crankshaft-for-v8.html}
+
+\bibitem{ionmonkey} IonMonkey from Mozilla. 2014.
+  \url{https://wiki.mozilla.org/IonMonkey/Overview}
+
+\bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. A Way Forward
+  in Parallelising Dynamic Languages. Under review in ICOOOLPS'14.
+
+\bibitem{cpython} CPython. \url{www.python.org}
+\bibitem{webjython} The Jython Project, \url{www.jython.org}
+\bibitem{ironpython} IronPython. \url{www.ironpython.net}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \bibitem{dan07}
@@
     applications} (OOPSLA '07).
-
-\bibitem{webjython}
-  The Jython Project, \url{www.jython.org}
 

### 2014-06-02 talk/dls2014/paper/paper.tex
add some refs

@@
 clarify the general semantics in commonly used terms from the
-literature.\remi{cite Transactional Memory 2nd edition}
+literature~\cite{harris10}.
 
@@
 not visible to another transaction before the former commits.
-Furthermore, the isolation provides full \emph{opacity} to always
-guarantee a consistent read set even for non-committed transactions.
-\remi{cite On the Correctness of Transactional Memory}
+Furthermore, the isolation provides full
+\emph{opacity}~\cite{guerraoui08} to always guarantee a consistent
+read set even for non-committed transactions.
 
@@
 non-transactional code in general), we use \emph{irrevocable} or
-\emph{inevitable transactions}. These transactions are always
+\emph{inevitable transactions}~\cite{blundell06,spear08}. These transactions are always
 guaranteed to commit, which is why they always have to win in case
@@
 $SO$. The result of this operation is called a \emph{Linear Address
-  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.
+  (LA)}. This is illustrated in figure \ref{fig:Segment-Addressing}.
 
@@
 are initially allocated in a single range of virtual memory by a call
-to \lstinline!mmap()!.  As illustrated in Figure
+to \lstinline!mmap()!.  As illustrated in figure
 \ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
@@
 
-As illustrated in Figure \ref{fig:Page-Remapping}, in our initial
+As illustrated in figure \ref{fig:Page-Remapping}, in our initial
 configuration (I) all segments are backed by their own range of
@@
 level of pages. Starting from the initial fully-shared configuration
-(Figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
+(figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
 object without other threads seeing the changes immediately, we ensure
@@
 
-\cfbolz{random question: did we investigate the extra memory requirements? we should characterise memory overhead somewhere, eg at least one byte per object for the read markers}
-
 \subsubsection{Summary}
@@
   on that stack are never seen as garbage and are thus always kept
-  alive.}.  Objects have to be saved using this stack around calls
+  alive.~\cite{fergus02}}.  Objects have to be saved using this stack around calls
 that may cause a GC cycle to happen, and also while there is no
@@
 privatise a few areas again right away. These areas are illustrated in
-Figure \ref{fig:Segment-Layout} and explained here:
+figure \ref{fig:Segment-Layout} and explained here:
 \begin{description}[noitemsep]
@@
 As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
-we use a \emph{shadow stack} in order to provide precise garbage
+we use a \emph{shadow stack}~\cite{fergus02} in order to provide precise garbage
 collection.  Any time we call a function that possibly triggers a
@@
 in the old object space when getting them back from the stack using
-\lstinline!STM_POP_ROOT()!. \remi{cite something which explains
-shadowstacks in more detail}
+\lstinline!STM_POP_ROOT()!.
 
@@
 \item to support \emph{inevitable} transactions, we always prefer them
-  to others since they cannot abort
+  to others since they cannot abort (similar to \cite{blundell06})
 \end{itemize}
@@
 of threads. They even become slower because of the overhead of
-thread-switching and GIL handling. We also see Jython scale when we
-expect it to (mandelbrot, raytrace, richards), and behave similar to
-the GIL interpreters in the other cases.
+thread-switching and GIL handling (see \cite{beazley10} for a detailed
+analysis). We also see Jython scale when we expect it to (mandelbrot,
+raytrace, richards), and behave similar to the GIL interpreters in the
+other cases.
 
@@
 
+Eliminate GIL:
+\begin{itemize}
+\item Previous attempts with HTM: \cite{nicholas06,odaira14,fuad10}
+\item Previous attempts with STM: \cite{stmupdate13}
+\end{itemize}
+
+Similar STMs:
+\begin{itemize}
+\item FastLane: \cite{warmhoff13}
+\item TML: \cite{spear09}
+\item Virtualizing HTM: \cite{rajwar05}
+\item Page-based virtualizing HyTM: \cite{chung06} (XTM can be
+  implemented either in the OS as part of the virtual memory manager or
+  between underlying TM systems and the OS, like virtual machines;
+  Conflicts for overflowed transactions are tracked at page granularity;
+  XTM-e allows conflict detection at cache line granu-
+  larity, even for overflowed data in virtual memory)
+\item using mmap(): Memory-Mapped Transactions
+\item mem-protected conflict detection: \cite{martin09}
+\end{itemize}
+
 \section{Conclusions}
@@
 \bibitem{ironpython} IronPython. \url{www.ironpython.net}
+
+\bibitem{beazley10} Beazley, David. "Understanding the python gil."
+  \emph{PyCON Python Conference}. Atlanta, Georgia. 2010.
+
+\bibitem{harris10} Harris, Tim, James Larus, and Ravi
+  Rajwar. "Transactional memory." \emph{Synthesis Lectures on Computer
+  Architecture 5.1} (2010): 1-263.
+
+\bibitem{guerraoui08} Guerraoui, Rachid, and Michal Kapalka. "On the
+  correctness of transactional memory." \emph{Proceedings of the 13th
+    ACM SIGPLAN Symposium on Principles and practice of parallel
+    programming.} ACM, 2008.
+
+\bibitem{blundell06} Blundell, Colin, E. Christopher Lewis, and Milo
+  Martin. "Unrestricted transactional memory: Supporting I/O and system
+  calls within transactions." (2006).
+\bibitem{spear08} Spear, Michael F., et al. "Implementing and
+  exploiting inevitability in software transactional memory."
+  \emph{Parallel Processing, 2008}. ICPP'08. 37th International
+  Conference on. IEEE, 2008.
+
+\bibitem{fergus02} Fergus Henderson. 2002. Accurate garbage collection
+  in an uncooperative environment. \emph{In Proceedings of the 3rd
+    international symposium on Memory management} (ISMM '02).
+
+\bibitem{stmupdate13} Armin Rigo, Remigius Meier. Update on
+  STM. \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}
+
+\bibitem{rajwar05} Rajwar, Ravi, Maurice Herlihy, and Konrad
+  Lai. "Virtualizing transactional memory." \emph{Computer
+    Architecture}, 2005. ISCA'05. Proceedings. 32nd International
+  Symposium on. IEEE, 2005.
+
+\bibitem{chung06} Chung, JaeWoong, et al. "Tradeoffs in transactional
+  memory virtualization." \emph{ACM SIGARCH Computer Architecture
+  News}. Vol. 34. No. 5. ACM, 2006.
+
+\bibitem{martin09} Martín Abadi, Tim Harris, and Mojtaba
+  Mehrara. 2009. Transactional memory with strong atomicity using
+  off-the-shelf memory protection hardware. SIGPLAN Not. 44, 4 (February
+  2009), 185-196.
 

### 2014-06-02 talk/dls2014/paper/paper.tex
Tweaks

@@
 
-TM systems come in can be broadly categorised as hardware based (HTM),
+TM systems come in\arigo{typo?} can be broadly categorised as hardware based (HTM),
 software based (STM), or hybrid systems (HyTM). HTM systems are limited
@@
   integration in order to keep the overhead of STM very low.
-\item This new STM system is used to replace the GIL in Python and is
-  then evaluated extensively.
+\item This new STM system is used to replace the GIL in one implementation
+  of Python and is then evaluated extensively.
 \item We introduce atomic blocks to the Python language to provide a
@@
 or to privatised file pages in the corresponding segments. This
-mapping is also performed efficiently by CPUs that have a Memory
-Management Unit (MMU) and can easily be done on every access to an
-object.
+mapping is cost-free: it is part of the normal operations done by the
+Memory Management Unit (MMU) of the CPU on every memory access.
 
@@
 Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
-level of pages. Starting from the initial fully-shared configuration
+level of pages~\footnote{Conflict detection still occurs on the level
+of objects.}. Starting from the initial fully-shared configuration
 (figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
@@
 
-To detect when to privatise pages, we use write barriers before every
-write to an object. When the barrier detects that the object is not in
-a private page (or any pages that belong to the object), we remap and
-copy the pages to the thread's segment. From now on, the translation
+More precisely, this is done by a write barrier that detects that we are
+about to write to an old (pre-transaction) object that we did not record
+in the write-set yet.  When this occurs, the slow-path of the write barrier
+will also check if the page (or pages) containing the object is still
+shared, and if so, privatise it.  This is done by remapping and copying
+the page to the thread's own segment. From now on, the translation
 of $\%gs{::}SO$ in this particular thread will resolve to a private

### 2014-06-02 talk/dls2014/paper/paper.tex
tweaks and performance-section updates

@@
 sequential consistency model~\cite{lamport79}. Another technology that
-can provide the same guarantees is transactional memory
-(TM). \remi{cite our position paper}
+can provide the same guarantees is transactional memory (TM).
 
@@
 
-TM systems come in\arigo{typo?} can be broadly categorised as hardware based (HTM),
+TM systems can be broadly categorised as hardware based (HTM),
 software based (STM), or hybrid systems (HyTM). HTM systems are limited
@@
 
-We now use these mechanisms to provide isolation for transactions.
-Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
-level of pages~\footnote{Conflict detection still occurs on the level
-of objects.}. Starting from the initial fully-shared configuration
+We now use these mechanisms to provide isolation for transactions.  We
+implement a \emph{Copy-On-Write (COW)} on the level of
+pages~\footnote{Conflict detection still occurs on the level of
+objects.}. Starting from the initial fully-shared configuration
 (figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
@@
 
-More precisely, this is done by a write barrier that detects that we are
-about to write to an old (pre-transaction) object that we did not record
+More precisely, this is done with a write barrier that detects that we are
+about to write to an object that we did not record
 in the write-set yet.  When this occurs, the slow-path of the write barrier
@@
 scale when using coarse-grained locking. STM, however, uses atomic
-blocks instead, so it may still be able to scale since they are
-implemented as simple transactions.
+blocks instead of a single lock to synchronise accesses to the
+shared data structures. Since atomic blocks map to transactions,
+our STM system may still be able to get a speedup on more than one
+thread by running transactions in parallel.
 
@@
 thread-switching and GIL handling (see \cite{beazley10} for a detailed
-analysis). We also see Jython scale when we expect it to (mandelbrot,
-raytrace, richards), and behave similar to the GIL interpreters in the
-other cases.
-
-PyPy using our STM system (pypy-stm-nojit) scales in all benchmarks to
-a certain degree. We see that the average overhead from switching from
-GIL to STM is \remi{$35.5\%$}, the maximum in richards is
-\remi{$63\%$}. pypy-stm-nojit beats pypy-nojit already on two threads;
+analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
+raytrace, richards}), and behave similar to the GIL interpreters in the
+other cases. The reason again being the coarse grained locking.
+
+PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
+benchmarks to a certain degree. It scales best in the ones where
+Jython scales as well and a little less in the others. The reason
+for that is that in the former group there are no real, logical
+conflicts -- all threads do independent calculations. In the latter
+case, the threads work on a common data structure and therefore
+create much more conflicts, which limits the scalability.
+
+Looking at the average overhead from switching from GIL to STM, we see
+that it is \remi{$\approx 35.5\%$}. The maximum in richards is
+\remi{$63\%$}.
+
+\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
 however, it never even beats CPython, the reference implementation of
 Python. This means that without the JIT, our performance is not
-competitive. We now look at how well our system works when we enable
-the JIT.
+competitive. So we will now look at how well our system works when we
+enable the JIT.
 
@@
 see that the performance is much less stable. There is certainly more
-work required in this area. In general, we see that the group of
-non-locked benchmarks certainly scales best. The other three scale
-barely or not at all with the number of threads. The slowdown factor
-from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
-performance in half of the benchmarks.
-
-\remi{Reason for bad scaling: acceleration of code that produces
-conflicts $-->$ more iterations $-->$ more conflicts. The overhead
-doesn't get accelerated by the JIT.}
+work required in this area. The slowdown factor for switching from GIL
+to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
+in half of the benchmarks.
+
+We see that generally, the group of non-locked benchmarks scales
+best. The other three scale barely or not at all with the number of
+threads. The reason for this is likely again the conflicts in the
+latter group. Since the JIT accelerates all code but not the STM
+overhead, we do more work per transaction. And this increases the
+likelihood of conflicts between them and therefore limits scalability
+even more than in the no-JIT benchmarks.
+
+Overall PyPy needs the JIT in order for its performance to be
+competitive. It would be interesting to see how using our STM system
+in CPython would turn out, but it is a lot of work. On its own, our
+system scales well, so we hope to also see that with the JIT in the
+future.
 
@@
 
-Overall PyPy needs the JIT in order for its performance to be
-competitive.  It would be interesting to see how using our STM system
-in CPython would perform, but it is a lot of work. On its own, our
-system scales well so we hope to also see that with the JIT in the
-future.
-
 \section{Related Work}
@@
 \item Virtualizing HTM: \cite{rajwar05}
-\item Page-based virtualizing HyTM: \cite{chung06} (XTM can be
+\item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
+  detection, otherwise hardware extensions required; assumes most
+  transactions fit HTM capacities (not so true here); COW using page-faults;
+  they assume OS-level access to page-tables (maybe not inherent to their
+  design); eval on simulator; value-based confl detection;
+
+ (XTM can be
   implemented either in the OS as part of the virtual memory manager or

### 2014-06-02 talk/dls2014/paper/paper.tex
comment some currently unused references

@@
   \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
-  \caption{Comparing runtime between interpreters without a JIT\label{fig:performance-nojit}}
+  \caption{Comparing execution time between interpreters without a JIT\label{fig:performance-nojit}}
 \end{figure}
@@
   \includegraphics[width=1\columnwidth]{plots/performance.pdf}
-  \caption{Comparing runtime between interpreters with JIT\label{fig:performance-jit}}
+  \caption{Comparing execution time between interpreters with JIT\label{fig:performance-jit}}
 \end{figure}
@@
 
-\bibitem{dan07}
-  Dan Grossman. 2007. The transactional memory / garbage collection
-  analogy. \emph{In Proceedings of the 22nd annual ACM SIGPLAN
-    conference on Object-oriented programming systems and
-    applications} (OOPSLA '07).
+% \bibitem{dan07}
+%   Dan Grossman. 2007. The transactional memory / garbage collection
+%   analogy. \emph{In Proceedings of the 22nd annual ACM SIGPLAN
+%     conference on Object-oriented programming systems and
+%     applications} (OOPSLA '07).
 
@@
 
-\bibitem{felber07}
-  Felber, Pascal, et al. "Transactifying applications using an open
-  compiler framework." \emph{TRANSACT}, August (2007): 4-6.
-
-\bibitem{bill06}
-  Bill McCloskey, Feng Zhou, David Gay, and Eric
-  Brewer. 2006. Autolocker: synchronization inference for atomic
-  sections. \emph{In Conference record of the 33rd ACM SIGPLAN-SIGACT
-  symposium on Principles of programming languages (POPL '06)}. ACM,
-  New York, NY, USA
+% \bibitem{felber07}
+%   Felber, Pascal, et al. "Transactifying applications using an open
+%   compiler framework." \emph{TRANSACT}, August (2007): 4-6.
+
+% \bibitem{bill06}
+%   Bill McCloskey, Feng Zhou, David Gay, and Eric
+%   Brewer. 2006. Autolocker: synchronization inference for atomic
+%   sections. \emph{In Conference record of the 33rd ACM SIGPLAN-SIGACT
+%   symposium on Principles of programming languages (POPL '06)}. ACM,
+%   New York, NY, USA
 
@@
 
-\bibitem{leis14}
-  Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
-  Hardware Transactional Memory in Main-Memory Databases."
-  \emph{Proc. of ICDE}. 2014.
-
-\bibitem{biased}
-  Kenneth Russell and David Detlefs. 2006. Eliminating
-  synchronization-related atomic operations with biased locking and
-  bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
-    conference on Object-oriented programing, systems, languages, and
-    applications} (OOPSLA '06).
+% \bibitem{leis14}
+%   Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
+%   Hardware Transactional Memory in Main-Memory Databases."
+%   \emph{Proc. of ICDE}. 2014.
+
+% \bibitem{biased}
+%   Kenneth Russell and David Detlefs. 2006. Eliminating
+%   synchronization-related atomic operations with biased locking and
+%   bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
+%     conference on Object-oriented programing, systems, languages, and
+%     applications} (OOPSLA '06).
 

### 2014-06-03 talk/dls2014/paper/paper.tex
some work on related work and contributions

@@
 \item We introduce a new software transactional memory (STM) system
-  that performs well even on low numbers of CPUs. It uses a novel
-  combination of long-existing CPU features and garbage collector (GC)
-  integration in order to keep the overhead of STM very low.
-\item This new STM system is used to replace the GIL in one implementation
-  of Python and is then evaluated extensively.
+  that performs well on low numbers of CPUs.
+\item We make novel use of the memory management capabilities of
+  off-the-shelf CPUs, e.g. memory segmentation and 64bit address space.
+\item We integrate our STM system closely with a garbage collector
+  (GC) in order to lower the overhead of STM.
+\item This new STM system is used to replace the GIL in one
+  implementation of Python and is then evaluated extensively.
 \item We introduce atomic blocks to the Python language to provide a
@@
 
-Eliminate GIL:
-\begin{itemize}
-\item Previous attempts with HTM: \cite{nicholas06,odaira14,fuad10}
-\item Previous attempts with STM: \cite{stmupdate13}
-\end{itemize}
-
-Similar STMs:
-\begin{itemize}
-\item FastLane: \cite{warmhoff13}
-\item TML: \cite{spear09}
-\item Virtualizing HTM: \cite{rajwar05}
-\item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
-  detection, otherwise hardware extensions required; assumes most
-  transactions fit HTM capacities (not so true here); COW using page-faults;
-  they assume OS-level access to page-tables (maybe not inherent to their
-  design); eval on simulator; value-based confl detection;
-
- (XTM can be
-  implemented either in the OS as part of the virtual memory manager or
-  between underlying TM systems and the OS, like virtual machines;
-  Conflicts for overflowed transactions are tracked at page granularity;
-  XTM-e allows conflict detection at cache line granu-
-  larity, even for overflowed data in virtual memory)
-\item using mmap(): Memory-Mapped Transactions
-\item mem-protected conflict detection: \cite{martin09}
-\end{itemize}
+There have been several attempts at removing the GIL using TM. We
+ourselves did a few earlier attempts with more conventional STM
+systems~\cite{stmupdate13}. Because of the huge overhead we
+encountered, the result was mostly useless on $\le 4$ CPU cores.
+Since this could be considered the main operating conditions of
+dynamic languages, it was never as useful as we wanted it to be.
+
+Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
+promising results, especially the very recent \cite{odaira14} that
+examines the current generation of Intel CPUs. However, they did a
+simplification by essentially disabling the GC during the
+benchmarks. This may be valid for the hand-tuned allocations in their
+interpreter. For us the GC is essential, and because it is a moving
+GC, it unfortunately touches a lot more memory and thereby reaches the
+HTM capacity much faster.  This, and the fact that HTM's capacity
+limits do not allow the implementation of arbitrary atomic blocks, let
+us to believe that STM is the right choice at the moment.
+
+There are multiple attempts at removing the limits of HTM by means of
+\emph{virtualising} HTM~\cite{rajwar05,chung06}. Indeed this may also
+be the future direction of our research. So far, some of them depend
+on additional hardware features. Or in the case of \cite{chung06},
+they have the assumption that most transactions fit the HTM capacity
+and that virtualisation is only a fallback. Our use case, however,
+features arbitrarily long transactions, which is why this assumption
+first has to be verified in our case.
+
+From the other end, there also have been attempts at lowering the
+overhead of STM~\cite{warmhoff13,spear09} especially for a low number
+of CPU cores. They have completely different approaches and it would
+be interesting to see how well they work for our specific use
+case. \remi{enough?}
+
+Similarly to us, \cite{martin09} use memory management techniques to
+provide isolation between transactions. They focus on providing strong
+atomicity between transactional and non-transactional code.  We,
+however, use inevitable transactions to provide strong atomicity.
+Additionally, while they use page fault handlers to detect conflicts,
+we use traditional write barriers. Again, there are only some
+superficial similarities between our approach and theirs.
+
+% Similar STMs:
+% \begin{itemize}
+% \item FastLane: \cite{warmhoff13}
+% \item TML: \cite{spear09}
+% \item Virtualizing HTM: \cite{rajwar05}
+% \item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
+%   detection, otherwise hardware extensions required; assumes most
+%   transactions fit HTM capacities (not so true here); COW using page-faults;
+%   they assume OS-level access to page-tables (maybe not inherent to their
+%   design); eval on simulator; value-based confl detection;
+
+%  (XTM can be
+%   implemented either in the OS as part of the virtual memory manager or
+%   between underlying TM systems and the OS, like virtual machines;
+%   Conflicts for overflowed transactions are tracked at page granularity;
+%   XTM-e allows conflict detection at cache line granu-
+%   larity, even for overflowed data in virtual memory)
+% \item using mmap(): Memory-Mapped Transactions
+% \item mem-protected conflict detection: \cite{martin09}
+% \end{itemize}
 

### 2014-06-03 talk/dls2014/paper/paper.tex
more explanations

@@
 before we do the next minor collection or we start the next
-transaction (we always do a minor collection during a commit).
+transaction~\footnote{We always do a minor collection during a commit.}.
+
+Note that we have three kinds of objects: \emph{young, overflow} and
+\emph{old} objects. Young and overflow objects were created in the
+same transaction; old objects always come from previous transactions.
+The close integration between STM and the GC allows us to use this
+information effectively: young objects do not trigger the slow path at
+all, overflow objects only need to be handled by the GC part of the
+write barrier once between collections, and only the old objects need
+to be handled by the TM part once per transaction. This keeps the
+overhead of the write barrier very low.
 
@@
 regard it as a simple performance enhancement, but that is not what
-happens in reality. First, since the JIT is a tracing
-JIT\remi{explain?}  running in multiple threads, it may compile
+happens in reality. First, since the JIT is a tracing JIT\cite{cfbolz09}
+\remi{explain?}  running in multiple threads, it may compile
 different things in each run because of the non-deterministic
@@
 overhead of STM~\cite{warmhoff13,spear09} especially for a low number
-of CPU cores. They have completely different approaches and it would
-be interesting to see how well they work for our specific use
-case. \remi{enough?}
+of CPU cores. \cite{spear09} works well for some workloads, in the
+case of a dynamic language VM, we have to assume any possible
+workload.  \cite{warmhoff13} uses a compile-time approach to select
+from multiple optimised STMs. Their partitioned FastLane STM seems to
+provide good performance already on two threads using master and
+helper threads. It would be interesting to see if its design also
+performs well in our use case.
 

### 2014-06-03 talk/dls2014/paper/paper.tex
Add comment

--- 2014-06-02 talk/dls2014/paper/paper.tex
@@
 sequential consistency model~\cite{lamport79}. Another technology that
-can provide the same guarantees is transactional memory (TM).
+can provide the same guarantees is transactional memory
+(TM). \remi{cite our position paper}
 
@@
 
-TM systems can be broadly categorised as hardware based (HTM),
+TM systems come in\arigo{typo?} can be broadly categorised as hardware based (HTM),
 software based (STM), or hybrid systems (HyTM). HTM systems are limited
@@
 
-We now use these mechanisms to provide isolation for transactions.  We
-implement a \emph{Copy-On-Write (COW)} on the level of
-pages~\footnote{Conflict detection still occurs on the level of
-objects.}. Starting from the initial fully-shared configuration
+We now use these mechanisms to provide isolation for transactions.
+Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
+level of pages~\footnote{Conflict detection still occurs on the level
+of objects.}. Starting from the initial fully-shared configuration
 (figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
@@
 
-More precisely, this is done with a write barrier that detects that we are
-about to write to an object that we did not record
+More precisely, this is done by a write barrier that detects that we are
+about to write to an old (pre-transaction) object that we did not record
 in the write-set yet.  When this occurs, the slow-path of the write barrier
@@
 \item [{Commit:}] If a transaction commits, we synchronise all threads
-  so that all of them are waiting in a safe point. In the committing
+  so that all of them are waiting in a safe point.\arigo{Explain why it's
+  not completely horribly slow here} In the committing
   transaction, we go through all objects in the write set and check if
@@
 scale when using coarse-grained locking. STM, however, uses atomic
-blocks instead of a single lock to synchronise accesses to the
-shared data structures. Since atomic blocks map to transactions,
-our STM system may still be able to get a speedup on more than one
-thread by running transactions in parallel.
+blocks instead, so it may still be able to scale since they are
+implemented as simple transactions.
 
@@
 thread-switching and GIL handling (see \cite{beazley10} for a detailed
-analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
-raytrace, richards}), and behave similar to the GIL interpreters in the
-other cases. The reason again being the coarse grained locking.
-
-PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
-benchmarks to a certain degree. It scales best in the ones where
-Jython scales as well and a little less in the others. The reason
-for that is that in the former group there are no real, logical
-conflicts -- all threads do independent calculations. In the latter
-case, the threads work on a common data structure and therefore
-create much more conflicts, which limits the scalability.
-
-Looking at the average overhead from switching from GIL to STM, we see
-that it is \remi{$\approx 35.5\%$}. The maximum in richards is
-\remi{$63\%$}.
-
-\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
+analysis). We also see Jython scale when we expect it to (mandelbrot,
+raytrace, richards), and behave similar to the GIL interpreters in the
+other cases.
+
+PyPy using our STM system (pypy-stm-nojit) scales in all benchmarks to
+a certain degree. We see that the average overhead from switching from
+GIL to STM is \remi{$35.5\%$}, the maximum in richards is
+\remi{$63\%$}. pypy-stm-nojit beats pypy-nojit already on two threads;
 however, it never even beats CPython, the reference implementation of
 Python. This means that without the JIT, our performance is not
-competitive. So we will now look at how well our system works when we
-enable the JIT.
+competitive. We now look at how well our system works when we enable
+the JIT.
 
@@
   \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
-  \caption{Comparing execution time between interpreters without a JIT\label{fig:performance-nojit}}
+  \caption{Comparing runtime between interpreters without a JIT\label{fig:performance-nojit}}
 \end{figure}
@@
 see that the performance is much less stable. There is certainly more
-work required in this area. The slowdown factor for switching from GIL
-to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
-in half of the benchmarks.
-
-We see that generally, the group of non-locked benchmarks scales
-best. The other three scale barely or not at all with the number of
-threads. The reason for this is likely again the conflicts in the
-latter group. Since the JIT accelerates all code but not the STM
-overhead, we do more work per transaction. And this increases the
-likelihood of conflicts between them and therefore limits scalability
-even more than in the no-JIT benchmarks.
-
-Overall PyPy needs the JIT in order for its performance to be
-competitive. It would be interesting to see how using our STM system
-in CPython would turn out, but it is a lot of work. On its own, our
-system scales well, so we hope to also see that with the JIT in the
-future.
+work required in this area. In general, we see that the group of
+non-locked benchmarks certainly scales best. The other three scale
+barely or not at all with the number of threads. The slowdown factor
+from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
+performance in half of the benchmarks.
+
+\remi{Reason for bad scaling: acceleration of code that produces
+conflicts $-->$ more iterations $-->$ more conflicts. The overhead
+doesn't get accelerated by the JIT.}
 
@@
   \includegraphics[width=1\columnwidth]{plots/performance.pdf}
-  \caption{Comparing execution time between interpreters with JIT\label{fig:performance-jit}}
+  \caption{Comparing runtime between interpreters with JIT\label{fig:performance-jit}}
 \end{figure}
+
+Overall PyPy needs the JIT in order for its performance to be
+competitive.  It would be interesting to see how using our STM system
+in CPython would perform, but it is a lot of work. On its own, our
+system scales well so we hope to also see that with the JIT in the
+future.
 
@@
 \item Virtualizing HTM: \cite{rajwar05}
-\item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
-  detection, otherwise hardware extensions required; assumes most
-  transactions fit HTM capacities (not so true here); COW using page-faults;
-  they assume OS-level access to page-tables (maybe not inherent to their
-  design); eval on simulator; value-based confl detection;
-
- (XTM can be
+\item Page-based virtualizing HyTM: \cite{chung06} (XTM can be
   implemented either in the OS as part of the virtual memory manager or
@@
 
-% \bibitem{dan07}
-%   Dan Grossman. 2007. The transactional memory / garbage collection
-%   analogy. \emph{In Proceedings of the 22nd annual ACM SIGPLAN
-%     conference on Object-oriented programming systems and
-%     applications} (OOPSLA '07).
+\bibitem{dan07}
+  Dan Grossman. 2007. The transactional memory / garbage collection
+  analogy. \emph{In Proceedings of the 22nd annual ACM SIGPLAN
+    conference on Object-oriented programming systems and
+    applications} (OOPSLA '07).
 
@@
 
-% \bibitem{felber07}
-%   Felber, Pascal, et al. "Transactifying applications using an open
-%   compiler framework." \emph{TRANSACT}, August (2007): 4-6.
-
-% \bibitem{bill06}
-%   Bill McCloskey, Feng Zhou, David Gay, and Eric
-%   Brewer. 2006. Autolocker: synchronization inference for atomic
-%   sections. \emph{In Conference record of the 33rd ACM SIGPLAN-SIGACT
-%   symposium on Principles of programming languages (POPL '06)}. ACM,
-%   New York, NY, USA
+\bibitem{felber07}
+  Felber, Pascal, et al. "Transactifying applications using an open
+  compiler framework." \emph{TRANSACT}, August (2007): 4-6.
+
+\bibitem{bill06}
+  Bill McCloskey, Feng Zhou, David Gay, and Eric
+  Brewer. 2006. Autolocker: synchronization inference for atomic
+  sections. \emph{In Conference record of the 33rd ACM SIGPLAN-SIGACT
+  symposium on Principles of programming languages (POPL '06)}. ACM,
+  New York, NY, USA
 
@@
 
-% \bibitem{leis14}
-%   Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
-%   Hardware Transactional Memory in Main-Memory Databases."
-%   \emph{Proc. of ICDE}. 2014.
-
-% \bibitem{biased}
-%   Kenneth Russell and David Detlefs. 2006. Eliminating
-%   synchronization-related atomic operations with biased locking and
-%   bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
-%     conference on Object-oriented programing, systems, languages, and
-%     applications} (OOPSLA '06).
+\bibitem{leis14}
+  Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
+  Hardware Transactional Memory in Main-Memory Databases."
+  \emph{Proc. of ICDE}. 2014.
+
+\bibitem{biased}
+  Kenneth Russell and David Detlefs. 2006. Eliminating
+  synchronization-related atomic operations with biased locking and
+  bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
+    conference on Object-oriented programing, systems, languages, and
+    applications} (OOPSLA '06).
 

### 2014-06-03 talk/dls2012/benchmarks/sqrt/sqrt_fix16.c
Add comment

@@ -1,0 +1,14 @@
+#include <stdio.h>
+
+int main() {
+  long y = 123 << 16;
+  long x = y / 2;
+  long n = 100000000;
+  while (n>0) {
+    n -= 1;
+    x = ((x + (y << 8)/(x >> 8))) / 2;
+  }
+  printf("%f\n", ((double) x) / ((double) (1<<16)));
+  fprintf(stderr, "sqrt(Fix16):   ");
+  return 0;
+}

### 2014-06-03 talk/dls2014/paper/plots/plot_scaling.py

@@ -1,0 +1,106 @@
+#!/usr/bin/python
+
+# obtained with time on
+#   pypy-c --jit off bench_scaling.py [1-4]
+
+
+import matplotlib
+import os
+import sys
+matplotlib.use('gtkagg')
+
+from matplotlib import rc
+#rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
+## for Palatino and other serif fonts use:
+rc('font',**{'family':'serif','serif':['Palatino']})
+rc('text', usetex=True)
+
+args = None
+import matplotlib.pyplot as plt
+# import pprint - slow as hell
+
+xs = range(1,5)
+ys = [[1.73, 1.74, 1.73, 1.73, 1.74],
+      [1.75, 1.77, 1.78, 1.75, 1.75],
+      [1.8, 1.79, 1.76, 1.76, 1.79],
+      [1.82, 2.1, 1.84, 1.9, 2.13]]
+
+
+
+def plot_mems(ax):
+    import numpy as np
+    y = []
+    yerr = []
+    opt_y = [1.0] * len(xs)
+    first_time = np.mean(ys[0])
+    for x, d in zip(xs, ys):
+        normalized = map(lambda x:x/first_time, d)
+        y.append(np.mean(normalized))
+        yerr.append(np.std(normalized))
+
+    print y
+    ax.errorbar(xs, y, yerr=yerr,
+                label="pypy-stm-nojit")
+    ax.plot(xs, opt_y, label="optimal")
+    return ax.legend(loc=0)
+
+
+def main():
+    global fig
+
+    print "Draw..."
+    fig = plt.figure()
+
+    ax = fig.add_subplot(111)
+
+    ax.set_ylabel("Runtime normalized to 1 thread")
+    ax.set_xlabel("Threads")
+    ax.set_ylim(0.9, 1.3)
+    ax.set_xlim(0, 5)
+
+    legend = plot_mems(ax)
+
+
+    #axs[0].set_ylim(0, len(x))
+    #ax.set_yticks([r+0.5 for r in range(len(logs))])
+    #ax.set_yticklabels(range(1, len(logs)+1))
+    #axs[0].set_xticks([])
+
+    # def label_format(x, pos):
+    #     return "%.2f" % (abs((x - left) * 1e-6), )
+    # major_formatter = matplotlib.ticker.FuncFormatter(label_format)
+    # axs[0].xaxis.set_major_formatter(major_formatter)
+
[…]

### 2014-06-03 talk/dls2014/paper/paper.tex
capitalize Figure

--- 2014-06-03 talk/dls2014/paper/paper.tex
@@
 
-\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
+\conferenceinfo{DLS 2014}{to be supplied}
 \copyrightyear{2014}
@@
 regarded as very expressive but also very slow. In recent years, the
-introduction of just-in-time compilers (JIT) for these languages (e.g.
-PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
+introduction of just-in-time compilers (JIT) for these languages
+(e.g.\ PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
 started to change this perception by delivering
@@
 programming model was not part of the design of those languages. Thus,
-the reference implementations of e.g. Python and Ruby use a single,
+the reference implementations of e.g.\ Python and Ruby use a single,
 global interpreter lock (GIL) to serialise the execution of code in
@@
 threads. There have been several attempts at lowering the overhead
-of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
+of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
 In this paper, we describe how we manage to lower the overhead of our
@@
 To also support these properties for irreversible operations that
-cannot be undone when we abort a transaction (e.g. I/O, syscalls, and
+cannot be undone when we abort a transaction (e.g.\ I/O, syscalls, and
 non-transactional code in general), we use \emph{irrevocable} or
@@
 
-In this section, we will present the general idea of how the TM model
-is implemented. Especially the aspects of providing isolation and
-atomicity, as well as conflict detection are explained. We try to do
-this without going into too much detail about the implementation.  The
-later section \ref{sub:Low-level-Implementation} will discuss it in
-more depth.
-
-\subsubsection{Memory Segmentation}
+In this section, we will present the general idea of how the TM model is
+implemented.  The later section~\ref{sub:Low-level-Implementation} will
+discuss it with more technical details.
+
+\subsubsection{Memory Segmentation and Page Sharing}
 
@@
 partition the virtual memory of a process into $N$ segments, one per
-thread. Each segment then holds a copy of all the memory available to
-the program. Thus, each thread automatically has a private copy of
-every object that it can modify in complete isolation from other
-threads.
-
-In order to reference these objects, we need references that are valid
-in all threads and automatically point to the private copies. Since
-an object's offset inside a segment is the same in all segments, we
-can use this offset to reference objects. Because all segments are
-copies of each other, this \emph{Segment Offset ($SO$)} points to the
-private version of an object in all threads~/ segments. To then
-translate this $SO$ to a real virtual memory address when used inside a
-thread, we need to add the thread's segment start address to the
-$SO$. The result of this operation is called a \emph{Linear Address
-  (LA)}. This is illustrated in figure \ref{fig:Segment-Addressing}.
-
-x86-CPUs provide a feature called \emph{memory segmentation}. It
-performs this translation from a $SO$ to a LA directly in hardware.  We
-can use the segment register $\%gs$, which is mostly unused in current
-applications. When this register points to a thread's segment start
-address, we can instruct the CPU to perform the above translation from
-a reference of the form $\%gs{::}SO$ to the right LA on its own. This
-process is efficient enough that we can do it on every access to an
-object.
-
-In summary, we can use a single $SO$ to reference the same object in all
-threads, and it will be translated by the CPU to a LA that always
-points to the thread's private version of this object. Thereby,
-threads are fully isolated from each other. However, $N$ segments
-require $N$-times the memory and modifications on an object need to be
-propagated to all segments.
-
-\begin{figure*}[t]
-  \centering
-  \includegraphics[scale=0.8]{\string"segment addressing\string".pdf}
-  \caption{Segment Addressing\label{fig:Segment-Addressing}}
-\end{figure*}
-
-\subsubsection{Page Sharing}
-
-In order to eliminate the prohibitive memory requirements of keeping
-around $N$ segment copies, we share memory between them. The segments
-are initially allocated in a single range of virtual memory by a call
-to \lstinline!mmap()!.  As illustrated in figure
-\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
-between a range of virtual memory pages and virtual file pages. The
-virtual file pages are then mapped lazily by the kernel to real
-physical memory pages.
-
-The mapping generated by \lstinline!mmap()! is initially linear but
-can be changed arbitrarily. Especially, we can remap so that multiple
-virtual memory pages map to a single virtual file page. This is what
-we use to share memory between the segments, since we then only
-require one page of physical memory for all of them.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[scale=0.8]{\string"mmap pages\string".pdf}
-  \caption{\texttt{mmap()} Page Mapping\label{fig:mmap()-Page-Mapping}}
-\end{figure}
-
-As illustrated in figure \ref{fig:Page-Remapping}, in our initial
-configuration (I) all segments are backed by their own range of
-virtual file pages. This is the share-nothing configuration where
-all threads have private versions of all objects.
-
-We then designate segment~0 to be the \emph{sharing-segment}. No
-thread gets this segment assigned to it, it simply holds the pages
-shared between threads. So in step (II), we remap all virtual pages of
-the segments~$>0$ to the file pages of our sharing-segment. This is
-the fully-shared configuration where no threads have private versions
-of any objects.
-
-During runtime, we can then privatise single pages in segments~$>0$
-again by remapping single pages as seen in (III). All objects in that
-page now have a private version in some thread.
-
-Looking back at address translation for object references, we see now
-that this is actually a two-step process. First, $\%gs{::}SO$ gets
-translated to different linear addresses in different threads by the
-CPU. Then, depending on the current mapping of virtual pages to file
-pages, these LAs can map to a single file page in the sharing-segment,
-or to privatised file pages in the corresponding segments. This
-mapping is cost-free: it is part of the normal operations done by the
-Memory Management Unit (MMU) of the CPU on every memory access.
-
-In summary, $\%gs{::}SO$ is translated efficiently by the CPU to
-either a physical memory location which is shared between several
-threads~/ segments or to a location in memory private to the segment~/
-thread. Page sharing makes the memory segmentation model for isolation
-memory efficient again.
+thread.  Each segment then holds a complete copy of the memory logically
+seen by the program.  All threads read and write data inside their own
+segment; exchanges between segments occur only at commit-time.  So far,
+this is a model that would also work for distributed transactional
+memory.
+
+Obviously, if we are in a shared-memory setting, this model is extremely
+wasteful: at a given time, most objects are identical in all, or most,
+segments.  They differ only in segments that have modified but not yet
+committed the object.  So we use a trick of the Memory Management Unit
+(MMU) of the CPU.  At the coarse unit of one page of memory (4096
+bytes), we use remappings so that, in general, pages within segments are
+\emph{shared} with each other if they would contain a fully identical
+copy.  Conversely, when one segment needs to write to one object
+contained in a page, the complete page is first \emph{privatised}: it is
+duplicated and the copy is remapped so that it will be visible in this
+segment (only).  This process is illustrated in
+figure~\ref{fig:Page-Remapping}.
+
+The result that we get is that the virtual address space of the process
+is bloated -- by a factor $N$ -- but the real memory usage is not --
+because most pages are shared with their copies in most other segments.
 
@@
 
-\subsubsection{Isolation: Copy-On-Write}
-
-We now use these mechanisms to provide isolation for transactions.  We
-implement a \emph{Copy-On-Write (COW)} on the level of
-pages~\footnote{Conflict detection still occurs on the level of
-objects.}. Starting from the initial fully-shared configuration
-(figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
-object without other threads seeing the changes immediately, we ensure
-that all pages belonging to the object are private to our segment.
-
-More precisely, this is done with a write barrier that detects that we are
-about to write to an object that we did not record
-in the write-set yet.  When this occurs, the slow-path of the write barrier
-will also check if the page (or pages) containing the object is still
-shared, and if so, privatise it.  This is done by remapping and copying
-the page to the thread's own segment. From now on, the translation
-of $\%gs{::}SO$ in this particular thread will resolve to a private
-version of the object automatically. Note that the $SO$ used to reference
-the object does not change during that process.
-
-\subsubsection{Isolation: Barriers}
-
-The job of barriers is to ensure complete isolation between transactions
-and to register the objects in the read or write set. We insert read
-and write barriers before reading or modifying an object, except if
-we statically know an object to be readable or writable already.
+\subsubsection{Memory Segmentation on the x86}
+
+When we need to store into our objects pointers to other objects, we
+need to be careful: if we were to directly store regular addresses valid
+in the local segment, then the objects' binary content would differ
+between segments.  This would prevent any sharing.  Instead, we store
+pointers as \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start
+of the segment.  A $SO$ can then be interpreted from any segment by
+adding it to the base address of that segment.\footnote{An alternative
+model would be to simulate threads using several independent processes
+in which all segments are at the same address.  This might work well for
+demonstration purposes, but it is not really suitable for implementing
+an existing programming language: programs assume that most of the state
+(file descriptors, external libraries, ...) is shared.} The result is
+called a \emph{Linear Address (LA)}. This is illustrated in
+figure~\ref{fig:Segment-Addressing}.
+
+The x86 CPUs provide a feature which is precisely called \emph{memory
+segmentation}.  On the modern x86-64, this is enough to perform the
+addition described above very efficiently.  We use the segment register
+$\%gs$, which is mostly unused in current applications, to point to the
+thread's segment start address.\footnote{The other segment register
+$\%fs$ is typically used by the threading library to provide access to
+thread-local data.  One point of view is that we are using $\%gs$ for a
+similar purpose: thread-local data -- but a lot of it.}  The translation
+from a $SO$ to a LA is done for us by using the ``$\%gs\colon$'' prefix
+before any CPU instruction that reads or writes memory using a $SO$
+address.  This is very efficient and some compilers support it natively
+(e.g.\ clang).
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[scale=0.8]{\string"segment addressing\string".pdf}
+  \caption{Segment Addressing\label{fig:Segment-Addressing}}
+\end{figure*}
+
+\subsubsection{Conflict detection}
+
+What we described so far has no mesurable performance impact by itself,
+but is missing conflict detection to be part of a full-fledged STM
+system.  This requires adding \emph{barriers} to the program to register
+the objects into the read or write set, and adapting the commit
+protocol in consequence.
+
 \begin{description}
-\item [{Read~Barrier:}] Adds the object to the read set of the current
-  transaction. Since our two-step address translation automatically
-  resolves the reference to the private version of the object on every
-  access anyway, the read barrier does not have the job to find the
-  private version. This job is fully performed by the CPU so that
-  read barriers have very little work to do.
-\item [{Write~Barrier:}] Adds the object to the read and write set of
-  the current transaction and checks if all pages of the object are
-  private, doing COW otherwise.\\
-  Furthermore, we currently allow only one transaction modifying an
-  object at a time. To ensure this, we acquire a write lock on the object
-  and also eagerly check for a write-write conflict at this point. If
-  there is a conflict, we do some contention management to decide which
-  transaction has to wait or abort.\\
-  Eagerly detecting this kind of conflict is not inherent to our
-  system, future experiments may show that we want to lift this
-  restriction.
+
+\item [{Read~Barrier:}] A major contribution to our performance results
+  is the simplicity of the read barrier: a large program (like an
+  interpreter for a complex language) contains a number of reads that
+  typically exceeds many times the number of writes.  We managed to
+  design the read barrier so that it is done by one memory write
+  (setting a flag) into non-shared memory.  This is done with a
+  segment-local array of flags indexed by a number derived from the $SO$
+  of the object.  Unlike other STM systems, the read barrier does not
+  have to find the private version of some object -- our two-step
+  address translation automatically resolves the reference to the
+  private version on every access anyway.  This means the read barrier
+  is not constrained to execute before the actual read -- both the
+  compiler and the CPU are free to reorder them.
+
+\item [{Write~Barrier:}] This is triggered by checking a flag in the
+  header of the object.  When the flag is set, we call the slow path,
+  which adds the object into the write set and unconditionally resets
+  the flag.  We eagerly detect write-write conflicts by allowing only
+  one transaction modifying an object at a time (by acquiring a write
+  lock).\footnote{Eager write-write detection is not inherent to our
+  approach; future experiments may show that we want to lift this
+  restriction.}  This slow path is also responsible for privatising
+  pages,\footnote{The write set contains objects, not pages.} as
+  described above.  Interestingly, we are using a generational garbage
+  collector (not described here) which requires a similar write barrier
+  anyway -- so the same header flag is used for both cases.  The flag
+  means ``something needs to be done before writing to this object'',
+  and the slow path logic determines what it is.
+
+\item [{Commit:}] When we want to commit a transaction, we need to check
+  for read-write conflicts.  Such conflicts cannot be detected in the
+  minimal read barrier, but can be detected during commit: the case to
+  look for is that the committing transaction wrote to an object that
+  another transaction has already read.\footnote{Which is why the read
+  sets can be just sparsely populated arrays of flags: we never walk the
+  whole read set.} A priori, checking other transactions' read sets
+  would involve reading concurrently modified memory: a different thread
+  could be issuing a read barrier in parallel with an unprotected
+  commit.  To prevent this from occurring, we must suspend all other
+  threads at known safe points.  This is a trade-off for the simplicity
+  of our read barriers -- which is well worth it in our case, due to the
+  very high number of read barriers when compared to the number of
+  commits and the (low) total number of threads.
+
 \end{description}
 
-\subsubsection{Atomicity: Commit \& Abort}
-
-To provide atomicity for a transaction, we want to make changes globally
-visible on commit. We also need to be able to completely abort a
-transaction without a trace, like it never happened.
-\begin{description}
-\item [{Commit:}] If a transaction commits, we synchronise all threads
-  so that all of them are waiting in a safe point. In the committing
-  transaction, we go through all objects in the write set and check if
-  another transaction in a different segment read the same object.
-  Conflicts are resolved again by either the committing or the other
-  transaction waiting or aborting.\\
-  We then push all changes of modified objects in private pages to all
-  the pages in other segments, including the sharing-segment (segment
-  0).
-\item [{Abort:}] On abort the transaction will forget about all the
-  changes it has done. All objects in the write set are reset by
-  copying their previous version from the sharing-segment into the
-  private pages of the aborting transaction.\\
-  Re-sharing these pages instead of resetting the changes would not
-  only be slower because it involves a system call, it is also very
-  likely the page would get privatised again immediately after re-trying
-  the transaction.  Since this again involves a full page copy,
-  resetting should be faster than re-sharing.
-\end{description}
-
-\subsubsection{Summary}
-
-We provide isolation between transactions by privatising the pages of
-the segments belonging to the threads the transactions run in.  To
-detect when and which pages need privatisation, we use write barriers
-that trigger a COW of one or several pages. Conflicts, however, are
-detected on the level of objects; based on the concept of read and
-write sets. Barriers before reading and writing add objects to the
-corresponding set; particularly detecting write-write conflicts
-eagerly.  On commit, we resolve read-write conflicts and push
-modifications to other segments. Aborting transactions simply undo
-their changes by copying from the sharing-segment.
+Conflicts can be resolved by aborting one of the two transactions, or in
+some cases by pausing one of them until the other commits.  If a commit
+succeeds, we copy the changes to modified objects so that they become
+visible in other segments.  Note that this only includes objects that
+existed before the transaction and have been modified by it.  It does
+not includes all the new objects created by the current transaction
+(there is typically a very high turn-around rate in dynamic languages).
+Our garbage collector allocates new objects in pages that are shared in
+the first place, so that they do not need to be copied -- and moreover,
+such objects will commonly never be modified any more.
+
+Privatised pages are re-shared during major collection cycles.
 
@@
 and write barriers. (II) It also includes a \emph{garbage collector
-(GC)} that is closely integrated with the TM part (e.g. it shares the
+(GC)} that is closely integrated with the TM part (e.g.\ it shares the
 write barrier). The close integration helps in order to know more
@@
 privatise a few areas again right away. These areas are illustrated in
-figure \ref{fig:Segment-Layout} and explained here:
+Figure~\ref{fig:Segment-Layout} and explained here:
 \begin{description}[noitemsep]
@@
 to store it in private pages. The area is called \emph{read markers}
-and already mentioned in section \ref{sub:Setup}.
+and already mentioned in section~\ref{sub:Setup}.
 
@@
 read-write conflict, we do contention management to either abort us or
-the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).
+the other transaction, or to simply wait a bit (see~\ref{subsub:contentionmanagement}).
 
@@
 system with PyPy's JIT. In fact, we will disable it for all benchmarks
-except those in section \ref{sec:performance-bench}. We would like to
+except those in section~\ref{sec:performance-bench}. We would like to
 regard it as a simple performance enhancement, but that is not what
@@
 Overall, we believe that disabling it on all benchmarks except the
-performance benchmarks in section \ref{sec:performance-bench} is better
+performance benchmarks in section~\ref{sec:performance-bench} is better
 because we can minimise non-determinism. We also do not want to depend
@@
 Second, there are areas in memory private to each segment (see
-section \ref{sub:Setup}). The Nursery for example is 4~MiB in
+section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
 size. Additionally, the area used for the read markers is lazily
@@
 
-In figure \ref{fig:richards_mem} we look at the memory usage of one of
+In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
 our benchmarks called Richards\footnote{OS kernel simulation
@@
 way. Additionally, since normal PyPy uses a GIL, it does not need to
-duplicate any data structures like e.g. the Nursery for each
+duplicate any data structures like e.g.\ the Nursery for each
 thread. This, the missing optimisation, and the additional memory
@@
 
-For the results in figure \ref{fig:scaling}, we averaged over 5 runs
+For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
 and normalised the average runtimes to the time it took on a single
@@
 and PyPy with STM and with the GIL (both without the JIT). The results
-are shown in figure \ref{fig:performance-nojit}.
+are shown in Figure~\ref{fig:performance-nojit}.
 
@@
 
-The results are presented in figure \ref{fig:performance-nojit}. We
+The results are presented in Figure~\ref{fig:performance-nojit}. We
 see that the performance is much less stable. There is certainly more

### 2014-06-03 talk/dls2014/paper/paper.tex
be a bit more explicit about properties of page sharing. + a note.
Formatting only

--- 2014-06-03 talk/dls2014/paper/paper.tex
@@
 privatise a few areas again right away. These areas are illustrated in
-figure \ref{fig:Segment-Layout} and explained here:
+figure~\ref{fig:Segment-Layout} and explained here:
 \begin{description}[noitemsep]
@@
 
-In figure \ref{fig:richards_mem} we look at the memory usage of one of
+In figure~\ref{fig:richards_mem} we look at the memory usage of one of
 our benchmarks called Richards\footnote{OS kernel simulation
@@
 
-For the results in figure \ref{fig:scaling}, we averaged over 5 runs
+For the results in figure~\ref{fig:scaling}, we averaged over 5 runs
 and normalised the average runtimes to the time it took on a single
@@
 and PyPy with STM and with the GIL (both without the JIT). The results
-are shown in figure \ref{fig:performance-nojit}.
+are shown in figure~\ref{fig:performance-nojit}.
 
@@
 
-The results are presented in figure \ref{fig:performance-nojit}. We
+The results are presented in figure~\ref{fig:performance-nojit}. We
 see that the performance is much less stable. There is certainly more

### 2014-06-03 talk/dls2014/paper/paper.tex
Rewrite section 3.2.  Please review...

--- 2014-06-03 talk/dls2014/paper/paper.tex
@@
 
-\conferenceinfo{DLS 2014}{to be supplied}
+\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
 \copyrightyear{2014}
@@
 \item We introduce a new software transactional memory (STM) system
-  that performs well on low numbers of CPUs.
-\item We make novel use of the memory management capabilities of
-  off-the-shelf CPUs, e.g. memory segmentation and 64bit address space.
-\item We integrate our STM system closely with a garbage collector
-  (GC) in order to lower the overhead of STM.
-\item This new STM system is used to replace the GIL in one
-  implementation of Python and is then evaluated extensively.
+  that performs well even on low numbers of CPUs. It uses a novel
+  combination of long-existing CPU features and garbage collector (GC)
+  integration in order to keep the overhead of STM very low.
+\item This new STM system is used to replace the GIL in one implementation
+  of Python and is then evaluated extensively.
 \item We introduce atomic blocks to the Python language to provide a
@@
 privatise a few areas again right away. These areas are illustrated in
-Figure~\ref{fig:Segment-Layout} and explained here:
+figure~\ref{fig:Segment-Layout} and explained here:
 \begin{description}[noitemsep]
@@
 before we do the next minor collection or we start the next
-transaction~\footnote{We always do a minor collection during a commit.}.
-
-Note that we have three kinds of objects: \emph{young, overflow} and
-\emph{old} objects. Young and overflow objects were created in the
-same transaction; old objects always come from previous transactions.
-The close integration between STM and the GC allows us to use this
-information effectively: young objects do not trigger the slow path at
-all, overflow objects only need to be handled by the GC part of the
-write barrier once between collections, and only the old objects need
-to be handled by the TM part once per transaction. This keeps the
-overhead of the write barrier very low.
+transaction (we always do a minor collection during a commit).
 
@@
 regard it as a simple performance enhancement, but that is not what
-happens in reality. First, since the JIT is a tracing JIT\cite{cfbolz09}
-\remi{explain?}  running in multiple threads, it may compile
+happens in reality. First, since the JIT is a tracing
+JIT\remi{explain?}  running in multiple threads, it may compile
 different things in each run because of the non-deterministic
@@
 
-In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
+In figure~\ref{fig:richards_mem} we look at the memory usage of one of
 our benchmarks called Richards\footnote{OS kernel simulation
@@
 
-For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
+For the results in figure~\ref{fig:scaling}, we averaged over 5 runs
 and normalised the average runtimes to the time it took on a single
@@
 and PyPy with STM and with the GIL (both without the JIT). The results
-are shown in Figure~\ref{fig:performance-nojit}.
+are shown in figure~\ref{fig:performance-nojit}.
 
@@
 
-The results are presented in Figure~\ref{fig:performance-nojit}. We
+The results are presented in figure~\ref{fig:performance-nojit}. We
 see that the performance is much less stable. There is certainly more
-work required in this area. The slowdown factor for switching from GIL
-to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
-in half of the benchmarks.
-
-We see that generally, the group of non-locked benchmarks scales
-best. The other three scale barely or not at all with the number of
-threads. The reason for this is likely again the conflicts in the
-latter group. Since the JIT accelerates all code but not the STM
-overhead, we do more work per transaction. And this increases the
-likelihood of conflicts between them and therefore limits scalability
-even more than in the no-JIT benchmarks.
-
-Overall PyPy needs the JIT in order for its performance to be
-competitive. It would be interesting to see how using our STM system
-in CPython would turn out, but it is a lot of work. On its own, our
-system scales well, so we hope to also see that with the JIT in the
-future.
+work required in this area. In general, we see that the group of
+non-locked benchmarks certainly scales best. The other three scale
+barely or not at all with the number of threads. The slowdown factor
+from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
+performance in half of the benchmarks.
+
+\remi{Reason for bad scaling: acceleration of code that produces
+conflicts $-->$ more iterations $-->$ more conflicts. The overhead
+doesn't get accelerated by the JIT.}
 
@@
   \includegraphics[width=1\columnwidth]{plots/performance.pdf}
-  \caption{Comparing execution time between interpreters with JIT\label{fig:performance-jit}}
+  \caption{Comparing runtime between interpreters with JIT\label{fig:performance-jit}}
 \end{figure}
 
+Overall PyPy needs the JIT in order for its performance to be
+competitive.  It would be interesting to see how using our STM system
+in CPython would perform, but it is a lot of work. On its own, our
+system scales well so we hope to also see that with the JIT in the
+future.
+
 \section{Related Work}
 
-There have been several attempts at removing the GIL using TM. We
-ourselves did a few earlier attempts with more conventional STM
-systems~\cite{stmupdate13}. Because of the huge overhead we
-encountered, the result was mostly useless on $\le 4$ CPU cores.
-Since this could be considered the main operating conditions of
-dynamic languages, it was never as useful as we wanted it to be.
-
-Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
-promising results, especially the very recent \cite{odaira14} that
-examines the current generation of Intel CPUs. However, they did a
-simplification by essentially disabling the GC during the
-benchmarks. This may be valid for the hand-tuned allocations in their
-interpreter. For us the GC is essential, and because it is a moving
-GC, it unfortunately touches a lot more memory and thereby reaches the
-HTM capacity much faster.  This, and the fact that HTM's capacity
-limits do not allow the implementation of arbitrary atomic blocks, let
-us to believe that STM is the right choice at the moment.
-
-There are multiple attempts at removing the limits of HTM by means of
-\emph{virtualising} HTM~\cite{rajwar05,chung06}. Indeed this may also
-be the future direction of our research. So far, some of them depend
-on additional hardware features. Or in the case of \cite{chung06},
-they have the assumption that most transactions fit the HTM capacity
-and that virtualisation is only a fallback. Our use case, however,
-features arbitrarily long transactions, which is why this assumption
-first has to be verified in our case.
-
-From the other end, there also have been attempts at lowering the
-overhead of STM~\cite{warmhoff13,spear09} especially for a low number
-of CPU cores. \cite{spear09} works well for some workloads, in the
-case of a dynamic language VM, we have to assume any possible
-workload.  \cite{warmhoff13} uses a compile-time approach to select
-from multiple optimised STMs. Their partitioned FastLane STM seems to
-provide good performance already on two threads using master and
-helper threads. It would be interesting to see if its design also
-performs well in our use case.
-
-Similarly to us, \cite{martin09} use memory management techniques to
-provide isolation between transactions. They focus on providing strong
-atomicity between transactional and non-transactional code.  We,
-however, use inevitable transactions to provide strong atomicity.
-Additionally, while they use page fault handlers to detect conflicts,
-we use traditional write barriers. Again, there are only some
-superficial similarities between our approach and theirs.
-
-% Similar STMs:
-% \begin{itemize}
-% \item FastLane: \cite{warmhoff13}
-% \item TML: \cite{spear09}
-% \item Virtualizing HTM: \cite{rajwar05}
-% \item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
-%   detection, otherwise hardware extensions required; assumes most
-%   transactions fit HTM capacities (not so true here); COW using page-faults;
-%   they assume OS-level access to page-tables (maybe not inherent to their
-%   design); eval on simulator; value-based confl detection;
-
-%  (XTM can be
-%   implemented either in the OS as part of the virtual memory manager or
-%   between underlying TM systems and the OS, like virtual machines;
-%   Conflicts for overflowed transactions are tracked at page granularity;
-%   XTM-e allows conflict detection at cache line granu-
-%   larity, even for overflowed data in virtual memory)
-% \item using mmap(): Memory-Mapped Transactions
-% \item mem-protected conflict detection: \cite{martin09}
-% \end{itemize}
+Eliminate GIL:
+\begin{itemize}
+\item Previous attempts with HTM: \cite{nicholas06,odaira14,fuad10}
+\item Previous attempts with STM: \cite{stmupdate13}
+\end{itemize}
+
+Similar STMs:
+\begin{itemize}
+\item FastLane: \cite{warmhoff13}
+\item TML: \cite{spear09}
+\item Virtualizing HTM: \cite{rajwar05}
+\item Page-based virtualizing HyTM: \cite{chung06} (XTM can be
+  implemented either in the OS as part of the virtual memory manager or
+  between underlying TM systems and the OS, like virtual machines;
+  Conflicts for overflowed transactions are tracked at page granularity;
+  XTM-e allows conflict detection at cache line granu-
+  larity, even for overflowed data in virtual memory)
+\item using mmap(): Memory-Mapped Transactions
+\item mem-protected conflict detection: \cite{martin09}
+\end{itemize}
 

### 2014-06-03 talk/dls2014/paper/paper.tex
capitalize Section

--- 2014-06-03 talk/dls2014/paper/paper.tex
@@
 \section{Method}
+
+\cfbolz{this needs a better section title. "Transactional Memory Implementation"?}
 
@@
 to store it in private pages. The area is called \emph{read markers}
-and already mentioned in section~\ref{sub:Setup}.
+and already mentioned in Section~\ref{sub:Setup}.
 
@@
 read-write conflict, we do contention management to either abort us or
-the other transaction, or to simply wait a bit (see~\ref{subsub:contentionmanagement}).
+the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).
 
@@
 system with PyPy's JIT. In fact, we will disable it for all benchmarks
-except those in section~\ref{sec:performance-bench}. We would like to
+except those in Section~\ref{sec:performance-bench}. We would like to
 regard it as a simple performance enhancement, but that is not what
@@
 Overall, we believe that disabling it on all benchmarks except the
-performance benchmarks in section~\ref{sec:performance-bench} is better
+performance benchmarks in Section~\ref{sec:performance-bench} is better
 because we can minimise non-determinism. We also do not want to depend
@@
 Second, there are areas in memory private to each segment (see
-section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
+Section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
 size. Additionally, the area used for the read markers is lazily

### 2014-06-03 talk/dls2014/paper/paper.tex

@@
 \section{Method}
-
-\cfbolz{this needs a better section title. "Transactional Memory Implementation"?}
 
@@
 to store it in private pages. The area is called \emph{read markers}
-and already mentioned in Section~\ref{sub:Setup}.
+and already mentioned in section~\ref{sub:Setup}.
 
@@
 read-write conflict, we do contention management to either abort us or
-the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).
+the other transaction, or to simply wait a bit (see~\ref{subsub:contentionmanagement}).
 
@@
 system with PyPy's JIT. In fact, we will disable it for all benchmarks
-except those in Section~\ref{sec:performance-bench}. We would like to
+except those in section~\ref{sec:performance-bench}. We would like to
 regard it as a simple performance enhancement, but that is not what
@@
 Overall, we believe that disabling it on all benchmarks except the
-performance benchmarks in Section~\ref{sec:performance-bench} is better
+performance benchmarks in section~\ref{sec:performance-bench} is better
 because we can minimise non-determinism. We also do not want to depend
@@
 Second, there are areas in memory private to each segment (see
-Section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
+section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
 size. Additionally, the area used for the read markers is lazily
@@
 
-The results are presented in Figure~\ref{fig:performance-nojit}. We
+The results are presented in figure~\ref{fig:performance-nojit}. We
 see that the performance is much less stable. There is certainly more

### 2014-06-03 talk/dls2014/paper/paper.tex
clean up footnotes
the full stop goes before the footnote

--- 2014-06-03 talk/dls2014/paper/paper.tex
@@
  \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
-and pop objects on the shadow stack~\footnote{A stack for pointers to
+and pop objects on the shadow stack.\footnote{A stack for pointers to
   GC objects that allows for precise garbage collection. All objects
   on that stack are never seen as garbage and are thus always kept
-  alive.~\cite{fergus02}}.  Objects have to be saved using this stack around calls
+  alive.~\cite{fergus02}}  Objects have to be saved using this stack around calls
 that may cause a GC cycle to happen, and also while there is no
@@
 The type \lstinline!object_t!  is special as it causes the
-compiler~\footnote{Clang 3.5 with some patches to its address-space
+compiler\footnote{Clang 3.5 with some patches to fix bugs in its address-space
  256 feature} to make all accesses through it relative to the $\%gs$
@@
 before we do the next minor collection or we start the next
-transaction~\footnote{We always do a minor collection during a commit.}.
+transaction.\footnote{We always do a minor collection during a commit.}
 
@@
 In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
-our benchmarks called Richards\footnote{OS kernel simulation
-benchmark}. The \emph{GC managed memory} counts all memory used in the
+our benchmarks called Richards.\footnote{Richards is a OS kernel simulation
+benchmark.} The \emph{GC managed memory} counts all memory used in the
 old object space including the memory required for private pages. The

### 2014-06-03 talk/dls2014/paper/paper.tex
clean up cites and a few tweaks

@@
 by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
-a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
+a lot of overhead~\cite{cascaval08,drago11}. In~\cite{wayforward14},
 we argue that STM is still the best way forward, especially since it
@@
 \item to support \emph{inevitable} transactions, we always prefer them
-  to others since they cannot abort (similar to \cite{blundell06})
+  to others since they cannot abort (similar to~\cite{blundell06})
 \end{itemize}
@@
 regard it as a simple performance enhancement, but that is not what
-happens in reality. First, since the JIT is a tracing JIT\cite{cfbolz09}
+happens in reality. First, since the JIT is a tracing JIT~\cite{cfbolz09}
 \remi{explain?}  running in multiple threads, it may compile
@@
 of threads. They even become slower because of the overhead of
-thread-switching and GIL handling (see \cite{beazley10} for a detailed
+thread-switching and GIL handling (see~\cite{beazley10} for a detailed
 analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
@@
 Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
-promising results, especially the very recent \cite{odaira14} that
+promising results, especially the very recent~\cite{odaira14} that
 examines the current generation of Intel CPUs. However, they did a
@@
 be the future direction of our research. So far, some of them depend
-on additional hardware features. Or in the case of \cite{chung06},
+on additional hardware features. Or in the case of~\cite{chung06},
 they have the assumption that most transactions fit the HTM capacity

### 2014-06-03 talk/dls2014/paper/paper.tex
Tweaks

@@
 copy.  Conversely, when one segment needs to write to one object
-contained in a page, the complete page is first \emph{privatised}: it is
+contained in a page, the complete page is first \emph{privatised:} it is
 duplicated and the copy is remapped so that it will be visible in this
@@
 from the start of the segment by an object's reference ($SO$) divided
-by 16 (this requires objects to have a size which is dividable by 16).
+by 16 (to avoid false conflicts, all objects have a size of at least 16 bytes).
 Instead of just setting the byte to \lstinline!true!  if the

### 2014-06-03 talk/dls2014/paper/paper.tex
Wrong, we are describing the GC at least a bit

@@
   described above.  Interestingly, we are using a generational garbage
-  collector (not described here) which requires a similar write barrier
+  collector (described below) which requires a similar write barrier
   anyway -- so the same header flag is used for both cases.  The flag

### 2014-06-03 talk/dls2014/paper/paper.tex
Move the subsection 3.1, 3.2 and 3.3 into their own section, 3, 4 and 5.

@@
 
-\section{Method}
-
-\cfbolz{this needs a better section title. "Transactional Memory Implementation"?}
-
-We now take a closer look at how our TM system that we use to replace
-the GIL works, what properties it has, and how it is implemented.
-
-\subsection{Transactional Memory Model}
+\section{Transactional Memory Model}
 
@@
 
-\subsubsection{Conflict Handling}
+\subsection{Conflict Handling}
 
@@
 
-\subsubsection{Semantics}
+\subsection{Semantics}
 
@@
 
-\subsection{Implementation}
+\section{High-level overview}
 
@@
 
-\subsubsection{Memory Segmentation and Page Sharing}
+\subsection{Memory Segmentation and Page Sharing}
 
@@
 
-\subsubsection{Memory Segmentation on the x86}
+\subsection{Memory Segmentation on the x86}
 
@@
 
-\subsubsection{Conflict detection}
+\subsection{Conflict detection}
 
@@
 
-\subsection{Low-level Implementation\label{sub:Low-level-Implementation}}
+\section{Low-level Implementation\label{sub:Low-level-Implementation}}
 
@@
 
-\subsubsection{Architecture}
+\subsection{Architecture}
 
@@
 
-\subsubsection{Application Programming Interface\label{sub:Application-Programming-Interfac}}
+\subsection{Application Programming Interface\label{sub:Application-Programming-Interfac}}
 
@@
 
-\subsubsection{Setup\label{sub:Setup}}
+\subsection{Setup\label{sub:Setup}}
 
@@
 
-\subsubsection{Assigning Segments}
+\subsection{Assigning Segments}
 
@@
 
-\subsubsection{Garbage Collection}
+\subsection{Garbage Collection}
 
@@
 
-\subsubsection{Read Barrier}
+\subsection{Read Barrier}
 
@@
 
-\subsubsection{Write Barrier}
+\subsection{Write Barrier}
 
@@
 
-\subsubsection{Abort}
+\subsection{Abort}
 
@@
 
-\subsubsection{Commit}
+\subsection{Commit}
 
@@
 
-\subsubsection{Thread Synchronisation}
+\subsection{Thread Synchronisation}
 
@@
 
-\subsubsection{Contention Management\label{subsub:contentionmanagement}}
+\subsection{Contention Management\label{subsub:contentionmanagement}}
 

### 2014-06-03 talk/dls2014/paper/paper.tex
Add a paragraph 'if we're not on x86 then'

@@
 (e.g.\ clang).
+
+If support for this is not available (due to the compiler or on a
+different architecture), there are other optimizations possible.  For
+example, using gcc, a regular register can be globally reserved and
+all ``simple enough'' memory accesses will still resolve to a single
+instruction.  More complicated accesses, e.g.\ those involving array
+indexing, will need an extra explicit addition.
 

### 2014-06-03 talk/dls2014/paper/paper.tex
a few notes

@@
 STM system so that it can be seen as a viable replacement for the GIL.
+
+\cfbolz{both the introduction and the contributions need to stress the co-design of GC and STM}
 
@@
 need to be careful: if we were to directly store regular addresses valid
-in the local segment, then the objects' binary content would differ
+in the local segment, then the objects' binary content would have to differ
 between segments.  This would prevent any sharing.  Instead, we store
@@
 not includes all the new objects created by the current transaction
-(there is typically a very high turn-around rate in dynamic languages).
+(there is typically a very high rate of object churn in dynamic languages).
 Our garbage collector allocates new objects in pages that are shared in
@@
 The type \lstinline!object_t!  is special as it causes the
-compiler\footnote{Clang 3.5 with some patches to fix bugs in its address-space
- 256 feature} to make all accesses through it relative to the $\%gs$
+compiler\footnote{We use Clang 3.5 with some patches to fix bugs in its address-space
+ 256 feature.} to make all accesses through it relative to the $\%gs$
 register.  With exceptions, nearly all accesses to objects managed by
@@
 integration is useful. highlight more or move to own section?}
+\cfbolz{definitely its own section listing all the benefits}
 
@@
 
+\cfbolz{maybe it's worth explicitly pointing out here that this is one of the
+    few places where the \%gs translation must not be used}
+
 \subsection{Write Barrier}
@@
 
+\cfbolz{is \%s translation used here? I guess yes, currently it's possible to misunderstand that}
+
 \begin{lstlisting}
@@
 
+\cfbolz{put the slow path into a figure, if it straddles a page it's too hard figuring out indentation}
 The \textbf{slow path} is shown here:
@@
 
+\cfbolz{at this point I kind of had the need for a diagram of the various
+states an object can be in, and the transitions between them.}
+
 In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the

### 2014-06-03 talk/dls2014/paper/paper.tex
the term "bump-pointer" seems to be the much more common one

@@
   objects (\emph{young objects}) of the current transaction. The GC
-  uses pointer-bump allocation in this area to allocate objects in the
+  uses bump-pointer allocation in this area to allocate objects in the
   first generation.
@@
 
+\cfbolz{you need to describe when and how to re-share pages}
+
 \subsection{Thread Synchronisation}
@@
 
-This code does simple pointer-bump allocation in the nursery. If there
+This code does simple bump-pointer allocation in the nursery. If there
 is still space left in the nursery, we return

### 2014-06-03 talk/icooolps2009-dotnet/clibackend.tex
last minute fix
the term "bump-pointer" seems to be the much more common one

@@
 
-It is important to underline the difference between flow graphs and a methods:
+It is important to underline the difference between flow graphs and methods:
 the first are the logical unit of code as seen by the JIT compiler, each of
@@
 
-To implement each flexswitch, the CLI backend creates an instace of a subclass
+To implement each flexswitch, the CLI backend creates an instance of a subclass
 of \lstinline{BaseLowLevelFlexSwitch}: such an instance stores the mapping
@@
 
-The following snippet shows the special case of integer flexswitches.
+The following snippet shows the special case of integer flexswitches:
 \begin{lstlisting}[language={[Sharp]C}] 

### 2014-06-03 talk/dls2012/benchmarks/scimark/run_FFT.c

@@ -1,0 +1,27 @@
+#include <stdio.h>
+#include <assert.h>
+
+#include "Random.c"
+#include "FFT.c"
+
+int main(int ac, char **av) {
+    assert(ac==3);
+    int N = atoi(av[1]);
+    int cycles = atoi(av[2]);
+    int twoN = 2*N;
+    Random R = new_Random_seed(7);
+    double *x = RandomVector(twoN, R);
+    int i=0;
+
+    for (i=0; i<cycles; i++)
+    {
+        FFT_transform(twoN, x);     /* forward transform */
+        FFT_inverse(twoN, x);       /* backward transform */
+    }
+
+
+    fprintf(stderr, "FFT(%d,%d):    ", N, cycles);
+    return 0;
+}
+
+

### 2014-06-03 talk/dls2014/paper/paper.tex
clarifications

--- 2014-06-03 talk/dls2014/paper/paper.tex
@@
   \includegraphics[width=1\columnwidth]{\string"page remapping\string".pdf}
-  \caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
+  \caption{Page Remapping: (I) initial setting. (II) remap all pages to
     segment~0, fully shared memory configuration. (III) privatise single
@@
 want to run $N$ threads in parallel while segment~0 is designated as
-the \emph{sharing-segment} that is never assigned to a thread.
+the \emph{sharing-segment} that is never assigned to a thread. The
+sharing-segment is essentially read-only except during commit.
 
 The next step involves using \lstinline!remap_file_pages()!, a Linux
-system call, to establish the fully-shared configuration.  All pages
+system call, to establish the \emph{fully-shared configuration}.  All pages
 of segments $>0$ map to the pages of the sharing-segment.
@@
 \bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. A Way Forward
-  in Parallelising Dynamic Languages. Under review in ICOOOLPS'14.
+  in Parallelising Dynamic Languages. ICOOOLPS'14.
 

### 2014-06-03 talk/dls2014/paper/paper.tex
Footnote: remap_file_pages

@@
 The next step involves using \lstinline!remap_file_pages()!, a Linux
-system call, to establish the \emph{fully-shared configuration}.  All pages
+system call\footnote{remap\_file\_pages() can also be done with more mmap()
+calls in a roughly portable way, but the former is more efficient on Linux.},
+to establish the \emph{fully-shared configuration}.  All pages
 of segments $>0$ map to the pages of the sharing-segment.

### 2014-06-03 talk/dls2014/paper/paper.tex
Why no space before *this* section title??
more notes

@@
 
+\medskip   % why, Latex, why??
 \subsection{Setup\label{sub:Setup}}
@@
 system with PyPy's JIT. In fact, we will disable it for all benchmarks
-except those in Section~\ref{sec:performance-bench}. We would like to
+except those in Section~\ref{sec:performance-bench}.
+\cfbolz{the details about why the JIT confuses everything should go into
+Section~\ref{sec:performance-bench}. here just write "because the integration
+with the JIT is incomplete"}
+We would like to
 regard it as a simple performance enhancement, but that is not what
@@
 
+\cfbolz{can you explain how you measure the errors?}
+
 % benchmarks with: pypy-c--Ojit-d1454093dd48+-14-05-26-17:16
@@
 \subsection{Memory Requirements}
+
+\cfbolz{part of this section is wrong here. maybe in the low-level section? here you should only measure}
 
@@
 improve this in the future, but right now this is the overall picture.}
+\cfbolz{you can't kill it, otherwise a reviewer will ask. also, the pypy-no-stm numbers are meaningless as long as the options are different}
 
@@
 
+\cfbolz{why is the error so big with 4 threads?}
+
 \begin{figure}[h]
@@
 our STM system may still be able to get a speedup on more than one
-thread by running transactions in parallel.
+thread by optimistically running transactions in parallel.
 
@@
 \emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
-however, it never even beats CPython, the reference implementation of
+however, it does not consistently beat CPython, the reference implementation of
 Python. This means that without the JIT, our performance is not
@@
 
+\cfbolz{I suggest making the rest a new section}
 \paragraph{JIT benchmarks:} The speedups from enabling the JIT in
@@
 
-The results are presented in Figure~\ref{fig:performance-nojit}. We
+The results are presented in Figure~\ref{fig:performance-jit}. We
 see that the performance is much less stable. There is certainly more

### 2014-06-03 talk/icooolps2014/position-paper.tex
Add a comment

@@
 we believe that it is worthwhile to focus the STM research more
-specifically on the context shown in this paper --- for use in
+specifically onto the context shown in this paper --- for use in
 implementations of high-level languages, rather than as a tool

### 2014-06-03 talk/dls2012/benchmarks/iter/xysum2d.c

@@ -1,0 +1,23 @@
+#include <stdio.h>
+#include <math.h>
+#include <stdlib.h>
+
+double result;
+
+double sum(double *a, int w, int h) {
+  int x, y;
+  double sa = 0;
+  for (y=0; y<h; y++) for(x=0; x<w; x++) sa += a[y*w + x] + x + y;
+  return sa;
+}
+
+#define W 10000
+#define H 10000
+
+int main(int ac, char **av) {
+  double *a = malloc(W*H*sizeof(double));
+  int i, n = atoi(av[1]);
+  for (i=0; i<n; i++) result=sum(a, W, H);
+  fprintf(stderr, "xysum2d:   ");
+  return 0;
+}

### 2014-06-03 talk/stanford-ee380-2011/demo-jit.py

@@ -1,0 +1,87 @@
+import sys
+from pypy.rlib import jit
+
+
+class Object:
+    pass
+
+
+class Integer(Object):
+
+    def __init__(self, value):
+        self.value = value
+
+    def next(self):
+        return Integer(self.value + 1)
+
+    def small(self):
+        return self.value < 1000000
+
+    def as_str(self):
+        return str(self.value)
+
+
+class String(Object):
+
+    def __init__(self, str):
+        self.str = str
+
+    def next(self):
+        return String(self.str + "x")
+
+    def small(self):
+        return len(self.str) < 50
+
+    def as_str(self):
+        return self.str
+
+
+jitdriver = jit.JitDriver(greens=['pc', 'bytecode'], reds=['input'])
+
+
+def interpret(bytecode, input):
+    pc = 0
+    while True:
+        jitdriver.jit_merge_point(bytecode=bytecode, input=input, pc=pc)
+
+        nextop = bytecode[pc]
+        pc += 1
+
+        if nextop == 'n':
+            input = input.next()
+
+        elif nextop == 'l':
+            if input.small():
+                pc = 0
+            else:
+                return input.as_str()
+
+
+# ____________________________________________________________
+
+
+if __name__ == '__main__':
+    print interpret("nl", String("x"))
+    print interpret("nl", Integer(1))
+    sys.exit()
+
+
+# ____________________________________________________________
+
+
+def main(argv):
+    num = int(argv[2])
+    print interpret(argv[1], Integer(num))
+    return 0
+
+def target(*args):
+    return main, None
+
+
+# ____________________________________________________________
+
+
+from pypy.jit.codewriter.policy import JitPolicy
+
+def jitpolicy(driver):
+    return JitPolicy()

### 2014-06-04 talk/dls2014/paper/plots/plot_performance.py
first round of feedback

@@ -39,7 +39,7 @@ interps_styles = {
 }
 
 benchs = {
-    "raytrace":{
+    "raytrace (large)":{
         "pypy-stm-jit":[
             [3.91, 3.87],
             [2.53, 2.52],
@@ -53,7 +53,7 @@ benchs = {
             [4.16]
         ]},
 
-    "btree":{
+    "btree (large)":{
         "pypy-stm-jit":[
             [1.68],
             [1.3],
@@ -67,7 +67,7 @@ benchs = {
             [5.8]
         ]},
 
-    "skiplist":{
+    "skiplist (large)":{
         "pypy-stm-jit":[
             [2.9],
             [3.0],
@@ -81,7 +81,7 @@ benchs = {
             [6.58]
         ]},
 
-    "threadworms":{
+    "threadworms (large)":{
         "pypy-stm-jit":[
             [4.23,4.33],
             [3.4,3.34,3.39],
@@ -95,7 +95,7 @@ benchs = {
             [20]
         ]},
 
-    "mandelbrot":{
+    "mandelbrot (large)":{
         "pypy-stm-jit":[
             [17.87,17.88,17.88],
             [9.4,9.42,9.34],
@@ -109,7 +109,7 @@ benchs = {
             [14.1]
         ]},
 
-    "richards":{
+    "richards (large)":{
         "pypy-stm-jit":[
             [63.4],
             [33.1],

### 2014-06-04 talk/dls2014/paper/paper.tex
clarification
Add a comment
first round of feedback
minor changes before restructuring

@@
 approach that provides a \emph{direct} replacement for the GIL. It
-does not require careful placing of locks in the right spots. We will
-compare our work with Jython for evaluation.
+does not require careful placing of locks in the right spots.
 
@@
 addition described above very efficiently.  We use the segment register
-$\%gs$, which is mostly unused in current applications, to point to the
+$\%gs$, which is available to applications, to point to the
 thread's segment start address.\footnote{The other segment register
@@
 existed before the transaction and have been modified by it.  It does
-not includes all the new objects created by the current transaction
+not include all the new objects created by the current transaction
 (there is typically a very high rate of object churn in dynamic languages).
@@
 \begin{lstlisting}
-void stm_start_transaction(tl, jmpbuf)
+void stm_start_transaction()
 void stm_commit_transaction()
@@
 object_t *stm_allocate(ssize_t size_rounded)
-STM_PUSH_ROOT(tl, obj)
-STM_POP_ROOT(tl, obj)
+STM_PUSH_ROOT(obj)
+STM_POP_ROOT(obj)
 \end{lstlisting}
 
-\lstinline!stm_start_transaction()!  starts a transaction. It requires
-two arguments, the first being a thread-local data structure and the
-second a buffer for use by \lstinline!setjmp()!.
+\lstinline!stm_start_transaction()!  starts a transaction in the
+current thread. Internally, it uses \lstinline!setjmp()! to remember
+where the transaction started in case of an abort.
 \lstinline!stm_commit_transaction()! tries to commit the current
 transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
-a read or a write barrier on an object and \lstinline!stm_allocate()!
-allocates a new object with the specified size.
- \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
-and pop objects on the shadow stack.\footnote{A stack for pointers to
-  GC objects that allows for precise garbage collection. All objects
+a read or a write barrier on the object given as the argument, and
+\lstinline!stm_allocate()!  allocates a new object with the specified
+size.  \lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
+push and pop objects onto the shadow stack\footnote{A stack for pointers
+  to GC objects that allows for precise garbage collection. All objects
   on that stack are never seen as garbage and are thus always kept
-  alive.~\cite{fergus02}}  Objects have to be saved using this stack around calls
-that may cause a GC cycle to happen, and also while there is no
-transaction running. In this simplified API, only
+  alive.~\cite{fergus02}} of the current thread.  References to objects
+have to be saved using this stack around calls that may cause a GC
+cycle to happen, and also while there is no transaction
+running. Otherwise these references may not be valid any more because
+the object moved or was seen as garbage. In this simplified API, only
 \lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
@@
 The type \lstinline!object_t! is special as it causes the
-compiler\footnote{We use Clang 3.5 with some patches to fix bugs in its address-space
- 256 feature.} to make all accesses through it relative to the $\%gs$
-register.  With exceptions, nearly all accesses to objects managed by
-the TM system should use this type so that the CPU will translate the
-reference to the right version of the object.
+compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
+its ``address-space 256'' feature. Patches are available from authors
+until inclusion into the official clang.} to make all accesses through
+it relative to the $\%gs$ register.  With exceptions, nearly all
+accesses to objects managed by the TM system should use this type so
+that the CPU will translate the reference to the right version of the
+object.
 
@@
 \emph{minor collections}. These collections move the surviving objects
-out of the nursery into the old object space, which can be done
+out of a thread's nursery into the old object space, which can be done
 without stopping other threads. This is done either if the nursery has
@@
     few places where the \%gs translation must not be used}
+\arigo{wrong, it is used.  maybe describe a bit more why}
 
@@
 regard it as a simple performance enhancement, but that is not what
-happens in reality. First, since the JIT is a tracing JIT~\cite{cfbolz09}
-\remi{explain?}  running in multiple threads, it may compile
+happens in reality. First, since the JIT~\cite{cfbolz09} bases its decision what to
+compile on runtime profiling, running in multiple threads, it may compile
 different things in each run because of the non-deterministic
@@
 benchmark.} The \emph{GC managed memory} counts all memory used in the
-old object space including the memory required for private pages. The
+old object space including the memory required for private pages. Note,
+it does not measure the memory required for the segment-private areas
+like the nursery and read markers. The
 sharp drops in memory usage come from major collections that free old
@@
 
-\subsection{Overhead Breakdown}
-
-\remi{do it on a non-jit build (see reason above)}
-\remi{gs:segment prefix overhead is virtually none (maybe instruction cache)}
-\remi{update numbers in pypy/TODO}
-
-\begin{itemize}
-\item time taken by read \& write barriers
-\item time spent committing \& aborting (maybe with different numbers
-  of threads; maybe split conflict detection and obj sync on commit)
-\item time in GC
-\end{itemize}
+%%% TODO: if time permits
+% \subsection{Overhead Breakdown}
+
+% \remi{do it on a non-jit build (see reason above)}
+% \remi{gs:segment prefix overhead is virtually none (maybe instruction cache)}
+% \remi{update numbers in pypy/TODO}
+
+% \begin{itemize}
+% \item time taken by read \& write barriers
+% \item time spent committing \& aborting (maybe with different numbers
+%   of threads; maybe split conflict detection and obj sync on commit)
+% \item time in GC
+% \end{itemize}
 
@@
 
-\subsection{Performance Benchmarks\label{sec:performance-bench}}
+\subsection{Small-Scale Benchmarks\label{sec:performance-bench}}
 
@@
 
-\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
+\subsection{Non-JIT Benchmarks}
+First we run our benchmarks on four
 different interpreters: Jython (fine-grained locking), CPython (GIL),
@@
 
-\cfbolz{I suggest making the rest a new section}
-\paragraph{JIT benchmarks:} The speedups from enabling the JIT in
+\subsection{JIT Benchmarks}
+The speedups from enabling the JIT in
 these benchmarks range from $10-50\times$. This is why we had to do
@@
 
-\bibitem{stmupdate13} Armin Rigo, Remigius Meier. Update on
+\bibitem{stmupdate13} Armin Rigo, Remigius Meier. 2013. Update on
   STM. \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}

### 2014-06-04 talk/dls2014/paper/paper.tex
simply move memory requirements explanation to its own section in low-level impl

@@
 
+\subsection{Memory Requirements\label{sub:memoryreq}}
+
+There are several sources of extra memory requirements in our
+STM system. First, we need to keep track of the state an object
+is in. We do this using flags and an overflow number. Both currently
+fit in a single additional header of 4~bytes per object.
+
+Second, there are areas in memory private to each segment (see
+Section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
+size\footnote{Configurable value taken from experience with our
+non-STM reference interpreter}. Additionally, the area used for the
+read markers is lazily mapped by the kernel and only needed for
+objects in the old object space. Thus, it can be at most the total
+size of memory used by old objects divided by 16, adding up to
+$6.3\%$ to the total memory requirements.
+
+Third, since we do copy on write, if all objects were to be written to
+at the same time, all pages would need to be privatised for all
+objects in the old object space. In that case we would need the total
+amount of memory required by old objects multiplied by $N+1$
+(incl. the sharing segment). This is certainly the worst case and
+so far not observed in practice. During major collections we re-share
+privatised pages that are currently not modified by any transaction.
+
 \section{Evaluation}
@@
 
-\cfbolz{part of this section is wrong here. maybe in the low-level section? here you should only measure}
-
-There are several sources of extra memory requirements in our
-STM system. First, we need to keep track of the state an object
-is in. We do this using flags and an overflow number. Both currently
-fit in a single additional header of 4~bytes per object.
-
-Second, there are areas in memory private to each segment (see
-Section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
-size. Additionally, the area used for the read markers is lazily
-mapped by the kernel and only needed for objects in the old object
-space. Thus, it can be at most the total size of memory used by
-old objects divided by 16.
-
-Third, since we do copy on write, if all objects were to be written to
-at the same time, all pages would need to be privatised for all
-objects in the old object space. In that case we would need the total
-amount of memory required by old objects multiplied by $N+1$
-(incl. the sharing segment). During major collections we re-share the
-pages if possible.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=1\columnwidth]{plots/richards_mem.pdf}
-  \caption{Actual memory managed by the GC and the page privatisation
-    over time in Richards benchmark\label{fig:richards_mem}}
-\end{figure}
-
-In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
-our benchmarks called Richards.\footnote{Richards is a OS kernel simulation
-benchmark.} The \emph{GC managed memory} counts all memory used in the
-old object space including the memory required for private pages. Note,
-it does not measure the memory required for the segment-private areas
-like the nursery and read markers. The
-sharp drops in memory usage come from major collections that free old
-objects and re-share pages. The average memory usage stays around
-29~MiB and we see that in this benchmark we have around 1 major
-collection every second.
-
-The \emph{page privatisation}, which represents the ratio between
-private pages and shared pages, gives us an idea about how many
-private copies exist for each shared page. If the ratio is $>1.0$, it
-means that on average we have more than one private copy for each
-shared page. In the worst case described above, this number would
-reach $N$ and means we need $N\times$ the memory for the private
-pages alone. In this benchmark, we see the same spikes as the memory
-usage. These come directly from re-sharing the pages. The maximum page
-privatisation is around $0.9$ between major collections, thus the
-private pages are responsible for a $90\%$ increase in the required
-memory. Since the spikes in the GC managed memory line actually
-show increases by $~100\%$, it means that the rest comes from actual
-garbage objects that were collected.
+\remi{reference \ref{sub:memoryreq}}
+
+% \begin{figure}[h]
+%   \centering
+%   \includegraphics[width=1\columnwidth]{plots/richards_mem.pdf}
+%   \caption{Actual memory managed by the GC and the page privatisation
+%     over time in Richards benchmark\label{fig:richards_mem}}
+% \end{figure}
+
+% In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
+% our benchmarks called Richards.\footnote{Richards is a OS kernel simulation
+% benchmark.} The \emph{GC managed memory} counts all memory used in the
+% old object space including the memory required for private pages. Note,
+% it does not measure the memory required for the segment-private areas
+% like the nursery and read markers. The
+% sharp drops in memory usage come from major collections that free old
+% objects and re-share pages. The average memory usage stays around
+% 29~MiB and we see that in this benchmark we have around 1 major
+% collection every second.
+
+% The \emph{page privatisation}, which represents the ratio between
+% private pages and shared pages, gives us an idea about how many
+% private copies exist for each shared page. If the ratio is $>1.0$, it
+% means that on average we have more than one private copy for each
+% shared page. In the worst case described above, this number would
+% reach $N$ and means we need $N\times$ the memory for the private
+% pages alone. In this benchmark, we see the same spikes as the memory
+% usage. These come directly from re-sharing the pages. The maximum page
+% privatisation is around $0.9$ between major collections, thus the
+% private pages are responsible for a $90\%$ increase in the required
+% memory. Since the spikes in the GC managed memory line actually
+% show increases by $~100\%$, it means that the rest comes from actual
+% garbage objects that were collected.
+
+\remi{do this and RSS numbers for pypy-stm-nojit and normal pypy-nojit:}
 

### 2014-06-04 talk/dls2014/paper/paper.tex
move around sections and remove Jython from evaluation

@@
 \subsection{Memory Requirements\label{sub:memoryreq}}
+\remi{maybe its own ``limitations'' section? also about
+  static segmentation and limited scaling }
 
@@
 so far not observed in practice. During major collections we re-share
-privatised pages that are currently not modified by any transaction.
+privatised pages that are currently not modified by any transaction,
+thereby reclaiming some of this memory.
 
@@
 running on top of it. For comparison, we compare between normal PyPy
-using a GIL and a PyPy with STM, as well as between other Python interpreters:
+using a GIL and a PyPy with STM, as well as to CPython.
 \begin{description}
@@
   execution and it does not feature a JIT compiler.
-\item[Jython] (version 2.7b1) is an implementation of Python on top of
-  the Java Virtual Machine (JVM). Instead of a GIL, this interpreter
-  uses fine-grained locking for synchronisation. This enables true
-  parallelism when executing code on multiple threads. In addition, its
-  integration with the JVM provides it with a JIT compiler for faster
-  execution.
 \end{description}
@@
 system with PyPy's JIT. In fact, we will disable it for all benchmarks
-except those in Section~\ref{sec:performance-bench}.
-\cfbolz{the details about why the JIT confuses everything should go into
-Section~\ref{sec:performance-bench}. here just write "because the integration
-with the JIT is incomplete"}
-We would like to
-regard it as a simple performance enhancement, but that is not what
-happens in reality. First, since the JIT~\cite{cfbolz09} bases its decision what to
-compile on runtime profiling, running in multiple threads, it may compile
-different things in each run because of the non-deterministic
-thread-scheduling of the operating system (OS). Second, it is able to
-remove some allocations in some cases. Because compilation is already
-non-deterministic, so is this allocation-removal. And third, we did
-not have enough time to optimise integration with STM so that the JIT
-exposes the overhead of STM more by speeding up all the rest.
-
-Overall, we believe that disabling it on all benchmarks except the
-performance benchmarks in Section~\ref{sec:performance-bench} is better
-because we can minimise non-determinism. We also do not want to depend
-on the capabilities of the JIT in these experiments.
+except those in Section~\ref{subsec:jit-benchs}. The STM-JIT
+integration is currently still incomplete and not tested much. The
+JIT-less interpreter provides a much more consistent environment for
+the STM system, so we remove some unknown variables by disabling it.
 
@@
 
-\cfbolz{can you explain how you measure the errors?}
+For each point in the plots, we took 5 measurements and report the
+average and the standard deviation as error bars. For the JIT
+benchmarks, we first let warm up the JIT by doing a few
+additional iterations to get a single measurement.
 
@@
 % many allocations and exposes the overhead more.
+
+\subsection{Scaling}
+
+To asses how well the STM system scales on its own (without any real
+workload), we execute the following loop on 1 to 4 threads on our
+Python interpreter with STM:
+\begin{lstlisting}
+def workload():
+    i = 20000000
+    while i:
+        i -= 1
+\end{lstlisting}
+
+For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
+and normalised the average runtimes to the time it took on a single
+thread. From this we see that there is additional overhead introduced
+by each thread (\remi{$13\%$} for all 4 threads together). While this
+is not ideal, we think that \remi{$13\%$} are acceptable on four
+threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
+more iterations per second than a single thread.
+
+\remi{what we don't show is by how much this overhead is influenced
+by allocations}
+
+\cfbolz{why is the error so big with 4 threads?}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\columnwidth]{plots/scaling.pdf}
+  \caption{Scalability of the STM system\label{fig:scaling}}
+\end{figure}
+
+\subsection{Small-Scale Benchmarks\label{sec:performance-bench}}
+
+For the following sections we use a set of six small benchmarks
+available on \cite{pypybenchs}:
+
+\begin{itemize}
+\item \emph{btree} and \emph{skiplist}, which are both inserting,
+  removing, and finding elements in a data structure from multiple
+  threads
+\item \emph{threadworms}, which simulates worms walking on a grid in
+  parallel and checking for collisions with each other
+\item \emph{mandelbrot}, \emph{raytrace}, and \emph{richards}, which
+  all perform some independent computations in parallel (embarrassingly
+  parallel)
+\end{itemize}
+
+We use atomic blocks in order to easily synchronise access to shared
+data structures in the first three benchmarks. For the embarrassingly
+parallel benchmarks, we do not need any explicit synchronisation at
+all. These atomic blocks are simulated with a single global lock
+when running on top of GIL-supported interpreters. With our STM
+system, they map to transactions that will execute optimistically
+in parallel.
 
@@
 
-\subsection{Scaling}
-
-To asses how well the STM system scales on its own (without any real
-workload), we execute the following loop on 1 to 4 threads on our
-Python interpreter with STM:
-\begin{lstlisting}
-def workload():
-    i = 20000000
-    while i:
-        i -= 1
-\end{lstlisting}
-
-For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
-and normalised the average runtimes to the time it took on a single
-thread. From this we see that there is additional overhead introduced
-by each thread (\remi{$13\%$} for all 4 threads together). While this
-is not ideal, we think that \remi{$13\%$} are acceptable on four
-threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
-more iterations per second than a single thread.
-
-\remi{what we don't show is by how much this overhead is influenced
-by allocations}
-
-\cfbolz{why is the error so big with 4 threads?}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=1\columnwidth]{plots/scaling.pdf}
-  \caption{Scalability of the STM system\label{fig:scaling}}
-\end{figure}
-
-\subsection{Small-Scale Benchmarks\label{sec:performance-bench}}
-
-In this section we want to look at the general performance of our
-system. As explained above, the JIT does not simply speed up the
-performance by a certain factor. In order to better understand the
-behaviour of our system, we therefore split this section into two
-parts. We first look at how it behaves without the JIT, and then with
-the JIT. We look at six benchmarks:
-
-\begin{itemize}
-\item \emph{btree} and \emph{skiplist}, which are both inserting,
-  removing, and finding elements in a data structure
-\item \emph{threadworms}, which simulates worms walking on a grid in
-  parallel and checking for collisions with each other
-\item \emph{mandelbrot}, \emph{raytrace}, and \emph{richards}, which
-  all perform simple, independent computations in parallel
-\end{itemize}
-
-We use coarse-grained locking for the first three benchmarks and no
-locking for the last three. This is important because Jython, which
-uses fine-grained locking instead of a GIL, is only expected to scale
-with the number of threads for the latter group. It is not able to
-scale when using coarse-grained locking. STM, however, uses atomic
-blocks instead of a single lock to synchronise accesses to the
-shared data structures. Since atomic blocks map to transactions,
-our STM system may still be able to get a speedup on more than one
-thread by optimistically running transactions in parallel.
-
-% To isolate factors we look at performance w/o JIT and perf w JIT.
-% w/o JIT:
-%  - it scales
-%  - slower than reference CPython
-%  - gil scales negatively
-% w JIT:
-%  - it sometimes still scales (more work needed)
-%  - changed working set size because generally ~50x faster
-%  - competing w our own JIT on non-stm a challenge
-%  - gil scales negatively
-
 \subsection{Non-JIT Benchmarks}
-First we run our benchmarks on four
-different interpreters: Jython (fine-grained locking), CPython (GIL),
-and PyPy with STM and with the GIL (both without the JIT). The results
-are shown in Figure~\ref{fig:performance-nojit}.
-
-As expected, all interpreters with a GIL do not scale with the number
+First we run our benchmarks on three different interpreters: CPython
+(GIL), and PyPy with STM and with the GIL (both without the JIT). The
+results are shown in Figure~\ref{fig:performance-nojit}.
+
+As expected, GIL-supported interpreters do not scale with the number
 of threads. They even become slower because of the overhead of
 thread-switching and GIL handling (see~\cite{beazley10} for a detailed
-analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
-raytrace, richards}), and behave similar to the GIL interpreters in the
-other cases. The reason again being the coarse grained locking.
+analysis).
 
 PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
-benchmarks to a certain degree. It scales best in the ones where
-Jython scales as well and a little less in the others. The reason
+benchmarks to a certain degree. It scales best for the embarrassingly
+parallel ones and a little less in the others. The reason
 for that is that in the former group there are no real, logical
-conflicts -- all threads do independent calculations. In the latter
-case, the threads work on a common data structure and therefore
-create much more conflicts, which limits the scalability.
+conflicts -- all threads do independent calculations. STM simply
+replaces the GIL in those programs. In the latter group,
+the threads work on a common data structure and therefore
+create much more conflicts, which limits the scalability. Here
+we make active use of the STM-supported atomic blocks that
+provide a scalable synchronisation mechanism.
 
@@
 \emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
-however, it does not consistently beat CPython, the reference implementation of
-Python. This means that without the JIT, our performance is not
-competitive. So we will now look at how well our system works when we
-enable the JIT.
+however, it rarely beats CPython's single-thread performance.  For
+programs that need concurrency in CPython and that use threads to
+achieve this, it also makes sense to look at the overhead induced by
+the GIL on multiple threads. From this perspective, the STM
+implementation beats CPython's performance in all but two benchmarks.
+
+Since PyPy comes with a JIT to make its overhead compared to CPython
+go away, we will now look at how well STM works together with it.
 
@@
 
-\subsection{JIT Benchmarks}
-The speedups from enabling the JIT in
-these benchmarks range from $10-50\times$. This is why we had to do
-without Jython and CPython here, since they would be much further up
-in the plots. Also, in order to get more stable results, we increased
-the input size of all benchmarks to get reasonable execution times.
+\subsection{JIT Benchmarks\label{subsec:jit-benchs}}
+
+We would like to regard enabling the JIT as a simple performance
+enhancement, but that is not what happens in reality. First, since the
+JIT~\cite{cfbolz09} bases its decision of what to compile on runtime
+profiling, running in multiple threads, it may compile different
+things in each run because of the non-deterministic thread-scheduling
+of the operating system (OS). Second, it is able to remove some
+allocations in some cases. Because compilation is already
+non-deterministic, so is this allocation-removal. And third, we did
+not have enough time to optimise integration with STM so that the JIT
+exposes the overhead of STM more by speeding up all the rest.
+For these reasons, the following results have to be taken with a grain
+of salt.
+
+The speedups from enabling the JIT in these benchmarks range from
+$10-50\times$. This is why we had to do without CPython here, since it
+would be much further up in the plots. Also, in order to make jitting
+code worthwhile, we increased the input size of all benchmarks to get
+reasonable execution times.
+
+We also ran these benchmarks on Jython\footnote{version
+2.7b1~\cite{webjython}}, an implementation of Python on top of the
+Java Virtual Machine (JVM).  Instead of a GIL, this interpreter uses
+fine-grained locking for synchronisation. Even though it can use the
+JVM's JIT compiler, its performance in these benchmarks is behind
+PyPy's by a factor of \remi{$5-10\times$}. Because of that it does not
+make sense to include it in this evaluation. In the future, we would
+like to do an extensive comparison between the STM approach and
+fine-grained locking. It is out of the scope of this paper to do
+this thoroughly.
 
@@
 
-We see that generally, the group of non-locked benchmarks scales
+We see that generally, the group of embarrassingly parallel benchmarks scales
 best. The other three scale barely or not at all with the number of
@@
 latter group. Since the JIT accelerates all code but not the STM
-overhead, we do more work per transaction. And this increases the
+overhead, we do more work per transaction. This increases the
 likelihood of conflicts between them and therefore limits scalability
@@
 Overall PyPy needs the JIT in order for its performance to be
-competitive. It would be interesting to see how using our STM system
-in CPython would turn out, but it is a lot of work. On its own, our
-system scales well, so we hope to also see that with the JIT in the
-future.
+competitive with CPython's. It would be interesting to see how using
+our STM system in CPython would turn out, but it is a lot of work. On
+its own, our system scales well, so we hope to also see that with the
+JIT in the future.
 
@@
 
+%%% TODO: fix all citation to follow common format (always page numbers; no 'et al', etc)
+
 \begin{thebibliography}{}
@@
 \bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. A Way Forward
-  in Parallelising Dynamic Languages. ICOOOLPS'14.
+  in Parallelising Dynamic Languages. To appear in ICOOOLPS'14.
 
@@
 
+\bibitem{pypybenchs} PyPy benchmarks repository. 2014. Revision
+  a26f2fb58413. \url{bitbucket.org/pypy/benchmarks}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-06-05 talk/dls2014/paper/paper.tex
add a table with the memory requirements
some clarifications

@@
 \usepackage{multicol}
+\usepackage{tabularx}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
+\usepackage{float}
+\floatstyle{ruled}
+\newfloat{code}{tbp}{loa}
+\providecommand{\codename}{Listing}
+\floatname{code}{\protect\codename}
 
@@
 segment (only).  This process is illustrated in
-figure~\ref{fig:Page-Remapping}.
+Figure~\ref{fig:Page-Remapping}.
 
@@
 called a \emph{Linear Address (LA)}. This is illustrated in
-figure~\ref{fig:Segment-Addressing}.
+Figure~\ref{fig:Segment-Addressing}.
 
@@
 void stm_commit_transaction()
-void stm_read(object_t *obj)
-void stm_write(object_t *obj)
+void stm_read(object_t *SO)
+void stm_write(object_t *SO)
 object_t *stm_allocate(ssize_t size_rounded)
-STM_PUSH_ROOT(obj)
-STM_POP_ROOT(obj)
+STM_PUSH_ROOT(object_t *SO)
+STM_POP_ROOT(object_t *SO)
 \end{lstlisting}
@@
 
-The type \lstinline!object_t! is special as it causes the
+In the following sections, whenever we use $SO$, we go through the
+address translation to get to the actual contents of an object. This
+is also signified by the type \lstinline!object_t!.  This type is
+special as it causes the
 compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
@@
 
-This area can be seen as a continuous array of bytes that is indexed
-from the start of the segment by an object's reference ($SO$) divided
-by 16 (to avoid false conflicts, all objects have a size of at least 16 bytes).
+This area can be seen as a continuous, segment-local array of bytes
+that is indexed with an object's reference ($SO$) divided by 16. So
+that each object has its own read marker, all objects have a size of
+at least 16 bytes. Otherwise there would be false conflicts when
+reading from two adjacent objects that share a single marker.
 Instead of just setting the byte to \lstinline!true!  if the
@@
 
-\cfbolz{maybe it's worth explicitly pointing out here that this is one of the
-    few places where the \%gs translation must not be used}
-\arigo{wrong, it is used.  maybe describe a bit more why}
-
 \subsection{Write Barrier}
@@
 need to check for the flag \lstinline!WRITE_BARRIER!  in the object's
-header and call the slow path if it is set. This flag is set either if
-the object is old and comes from an earlier transaction, or if there
-was a minor collection which will add the flag again on all
-objects. It is never set on freshly allocated objects that still
-reside in the nursery.
-
-\cfbolz{is \%s translation used here? I guess yes, currently it's possible to misunderstand that}
+header through its $SO$ and call the slow path if it is set. This flag
+is set either if the object is old and comes from an earlier
+transaction, or if there was a minor collection which will add the
+flag again on all objects. It is never set on freshly allocated
+objects that still reside in the nursery.
 
@@
 
-\cfbolz{put the slow path into a figure, if it straddles a page it's too hard figuring out indentation}
-The \textbf{slow path} is shown here:
-
+\begin{code}[h]
 \begin{lstlisting}
@@
 
+\caption{Slow path of the write barrier\label{lst:w_slow}}
+\end{code}
+
+The \textbf{slow path} is shown here in Listing~\ref{lst:w_slow}.
 First comes the \emph{GC part}: In any case, the object will be added
@@
 object. We then try to acquire its write lock. \lstinline!write_locks!
-again is a simple global array of bytes that is indexed with the $SO$
-of the object divided by 16. If we already own the lock, we are done.
+is a simple, \textbf{global} array of bytes that is indexed with the
+$SO$ of the object divided by 16. Note, ``global'' here really means
+it is a single array with data for all segments, there is no address
+translation going on to access its elements contrary to e.g.\ the
+read markers array.  If we already own the lock, we are done.
 If someone else owns the lock, we will do a write-write contention
@@
 
+\begin{table}[h]
+  \centering
+  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill} }|p{2cm}|c|c|c|}
+    \hline
+                        & pypy-nojit  & pypy-stm-nojit & Overhead  \\
+    \hline
+    btree (small)       & 3.2 /  23.8 & 23.8 / 106     & 7.4 / 4.5 $\times$ \\
+    \hline
+    skiplist (small)    & 3.4 / 19.1  & 23.5 / 103     & 6.9 / 5.4 $\times$ \\
+    \hline
+    threadworms (small) & 13.6 / 28.6 & 30.1 / 150     & 2.2 / 5.2 $\times$ \\
+    \hline
+    mandelbrot (small)  & 3.7 / 16.6  & 21.2 / 101     & 5.7 / 6.1 $\times$ \\
+    \hline
+    raytrace (small)    & 9.0 / 67.0  & 22.9 / 141     & 2.5 / 2.1 $\times$ \\
+    \hline
+    richards (small)    & 3.7 / 16.6  & 24.4 / 116     & 6.6 / 7.0 $\times$ \\
+    \hline
+  \end{tabular*}
+  \caption{Memory requirements reported as average GC-memory / RSS in MiB}
+\end{table}
+
+\remi{We cut off the first second of run time to not include startup.
+We run on 4 threads in order to see some privatisation of pages in STM.}
+
 % \begin{figure}[h]

### 2014-06-05 talk/dls2014/paper/paper.tex
Attempt to simplify the Memory Requirements sections into just some text in Future Work.

--- 2014-06-04 talk/dls2014/paper/paper.tex
@@
 \usepackage{multicol}
+\usepackage{tabularx}
 
@@
 
-\subsection{Memory Requirements\label{sub:memoryreq}}
-\remi{maybe its own ``limitations'' section? also about
-  static segmentation and limited scaling }
-
-There are several sources of extra memory requirements in our
-STM system. First, we need to keep track of the state an object
-is in. We do this using flags and an overflow number. Both currently
-fit in a single additional header of 4~bytes per object.
-
-Second, there are areas in memory private to each segment (see
-Section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
-size\footnote{Configurable value taken from experience with our
-non-STM reference interpreter}. Additionally, the area used for the
-read markers is lazily mapped by the kernel and only needed for
-objects in the old object space. Thus, it can be at most the total
-size of memory used by old objects divided by 16, adding up to
-$6.3\%$ to the total memory requirements.
-
-Third, since we do copy on write, if all objects were to be written to
-at the same time, all pages would need to be privatised for all
-objects in the old object space. In that case we would need the total
-amount of memory required by old objects multiplied by $N+1$
-(incl. the sharing segment). This is certainly the worst case and
-so far not observed in practice. During major collections we re-share
-privatised pages that are currently not modified by any transaction,
-thereby reclaiming some of this memory.
-
 \section{Evaluation}
@@
 
-\subsection{Memory Requirements}
-
-\remi{reference \ref{sub:memoryreq}}
-
-% \begin{figure}[h]
-%   \centering
-%   \includegraphics[width=1\columnwidth]{plots/richards_mem.pdf}
-%   \caption{Actual memory managed by the GC and the page privatisation
-%     over time in Richards benchmark\label{fig:richards_mem}}
-% \end{figure}
-
-% In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
-% our benchmarks called Richards.\footnote{Richards is a OS kernel simulation
-% benchmark.} The \emph{GC managed memory} counts all memory used in the
-% old object space including the memory required for private pages. Note,
-% it does not measure the memory required for the segment-private areas
-% like the nursery and read markers. The
-% sharp drops in memory usage come from major collections that free old
-% objects and re-share pages. The average memory usage stays around
-% 29~MiB and we see that in this benchmark we have around 1 major
-% collection every second.
-
-% The \emph{page privatisation}, which represents the ratio between
-% private pages and shared pages, gives us an idea about how many
-% private copies exist for each shared page. If the ratio is $>1.0$, it
-% means that on average we have more than one private copy for each
-% shared page. In the worst case described above, this number would
-% reach $N$ and means we need $N\times$ the memory for the private
-% pages alone. In this benchmark, we see the same spikes as the memory
-% usage. These come directly from re-sharing the pages. The maximum page
-% privatisation is around $0.9$ between major collections, thus the
-% private pages are responsible for a $90\%$ increase in the required
-% memory. Since the spikes in the GC managed memory line actually
-% show increases by $~100\%$, it means that the rest comes from actual
-% garbage objects that were collected.
-
-\remi{do this and RSS numbers for pypy-stm-nojit and normal pypy-nojit:}
-
-For PyPy-STM the average memory requirement is 21~MiB and there are
-$\sim 10$ major collections during the runtime. Normal PyPy with a GIL
-grows its memory up to just 7~MiB and does not do a single major
-collection in that time. Compared to normal PyPy, we are missing a
-memory optimisation to store small objects in a more compact
-way. Additionally, since normal PyPy uses a GIL, it does not need to
-duplicate any data structures like e.g.\ the Nursery for each
-thread. This, the missing optimisation, and the additional memory
-requirements for STM explained above account for this difference.
-We expect to improve this aspect in the future; in this paper we
-want to focus first on performance.
-\remi{I don't know how much sense it makes to go deeper. We will
-improve this in the future, but right now this is the overall picture.}
-\cfbolz{you can't kill it, otherwise a reviewer will ask. also, the pypy-no-stm numbers are meaningless as long as the options are different}
-
 %%% TODO: if time permits
@@
 
+\section{Future Work}
+
+\arigo{what else than memory usage goes here?}
+
+We are still working on a number of optimizations in the
+implementation.  We want to mention notably that memory usage has not
+been optimized so far.  There are several sources of extra memory
+requirements in our STM system.  The multiple nurseries add
+4~MiB\footnote{Configurable value taken from experience with our
+  non-STM reference interpreter.} per segment.  Additionally, the area
+used for the read markers uses, per segment, up to 1/16th of the
+memory used by the old objects.  With four segments, this adds $25\%$
+to the total memory requirements.
+
+One of the most important sources of additional memory usage is the
+presence of privatised pages.  It is hard to predict and control.
+Right now, if a program creates a large number of privatised pages,
+major garbage collections will occur more frequently in the hope of
+resharing some of these pages.  More advanced techniques might be
+useful.
+
+Overall, early measures seem to show that a four-segment version of
+pypy-stm-nojit consumes between 2 and 7 times as much memory as a
+plain pypy-nojit.  The high part of this spectrum seems to occur only
+if the total memory usage is rather low (less than 20~MiB): it might
+simply mean that our approach has a high constant memory usage
+overhead.  The low part, i.e.\ a factor 2, can approximately be
+attributed to the reasons described above.  It is future work to
+investigate in more details.
+
 \section{Related Work}

### 2014-06-05 talk/pycon2010/crossinterp/talk.tex
Attempt to simplify the Memory Requirements sections into just some text in Future Work.

@@
     \item won't talk about py3k
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{My background}
+  \begin{itemize}
+    \item worked a lot with PyPy compatibility issues
+    \item helped to port twisted, django and other projects to run on PyPy
+    \item a lot of cooperation with Jython people
   \end{itemize}
@@
       a while
-    \item the single most-common problem when fixing twisted
+    \item the single most-common problem when porting twisted to PyPy
   \end{itemize}
@@
       other Pythons exactly once
-    \item cycles with \_\_del\_\_s are not collected by CPython,
+    \item cycles with {\ttfamily \_\_del\_\_}s are not collected by CPython,
       PyPy breaks them randomly instead
-    \item in PyPy and Jython \_\_del\_\_ cannot be attached to classes after creation
+    \item in PyPy and Jython {\ttfamily \_\_del\_\_} cannot be attached to classes after creation
   \end{itemize}
@@
     \item 3.x ready
-    \item faster on PyPy, too
+    \item much faster on PyPy, too
   \end{itemize}
@@
   \frametitle{{\ttfamily sys.prefix}}
+  XXX think about
   \begin{itemize}
@@
     \item for 2.x Pythons use str for bytes and unicode for text
-    \item the distinction is deeper in 3.x (str is unicode, bytes exist with
-  slightly different interface than old str)
+    \item the distinction is deeper in 3.x ({\ttfamily str} is {\ttfamily unicode}, {\ttfamily bytes} exist with
+  slightly different interface than old {\ttfamily str})
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Don't concatenate strings}
+  \begin{itemize}
+    \item use {\ttfamily "".join(...)}
+    \item if you care about performance, try this and {\ttfamily cStringIO}
   \end{itemize}

### 2014-06-05 talk/dls2014/paper/plots/plot_performance_nojit.py

@@ -1,0 +1,259 @@
+#!/usr/bin/python
+
+# benchmarks-repo at a26f2fb58413
+
+# for now: avg & stddev of the best
+
+
+
+from plot_performance import plot_speedups
+import matplotlib
+import os
+import sys
+matplotlib.use('gtkagg')
+
+from matplotlib import rc
+#rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
+## for Palatino and other serif fonts use:
+rc('font',**{'family':'serif','serif':['Palatino']})
+rc('text', usetex=True)
+
+args = None
+import matplotlib.pyplot as plt
+# import pprint - slow as hell
+
+
+interps_styles = {
+    "pypy-stm-nojit": {'fmt':'r-', 'linewidth':2},
+    "cpython": {'fmt':'b', 'dashes':(1,1)},
+    "pypy-nojit": {'fmt':'g', 'dashes':(5, 2)},
+    "jython": {'fmt':'m', 'dashes':(2, 5)},
+    "best": {'fmt':"k:"}        # only fmt allowed
+}
+
+
+#    pypy-c-paper-nojit bench.py -k5 raytrace/raytrace.py 1-4 256 256
+#    pypy-c-paper-nojit bench.py -k5 btree/btree.py 1-4 500000
+#    pypy-c-paper-nojit bench.py -k5 skiplist/skiplist.py 1-4 200000
+#    pypy-c-paper-nojit bench.py -k5 threadworms/threadworms.py 1-4 500000
+#    pypy-c-paper-nojit bench.py -k5 mandelbrot/mandelbrot.py 1-4 64 512 512
+#    pypy-c-paper-nojit multithread-richards.py 30 1-4 # report runtime
+
+benchs = {
+    "raytrace (small)":{
+        "pypy-stm-nojit":[
+            [8.3],
+            [4.33],
+            [3.74],
+            [3.08]
+        ],
+        "cpython":[
+            [2.5],
+            [2.7],
+            [2.75],
+            [2.84]
+        ],
+        # "jython":[
+        #     [2.95,2.95,2.96],
+        #     [1.65,1.68,1.54],
+        #     [1.2,1.15,1.3,1.3],
+        #     [1.09,0.9,0.97,0.99,1.03]
+        # ],
+        "pypy-nojit":[
+            [5.5,5.7,5.8],
+            [7,6.97],
+            [6.68,6.77],
+            [6.4,6.4]
+        ]},
+
[…]

### 2014-06-05 talk/dls2014/paper/paper.tex

@@
 \usepackage{tabularx}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
+\usepackage{float}
+\floatstyle{ruled}
+\newfloat{code}{tbp}{loa}
+\providecommand{\codename}{Listing}
+\floatname{code}{\protect\codename}
 

### 2014-06-05 talk/dls2014/paper/paper.tex
add a diagram for object states

@@
 out of a thread's nursery into the old object space, which can be done
-without stopping other threads. This is done either if the nursery has
+without stopping other threads. This is happens either if the nursery has
 no space left anymore or if we are committing the current
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\columnwidth]{object_states.pdf}
+  \caption{Objects states handled by the write barrier\label{fig:obj_states}}
+\end{figure}
+
 \subsection{Write Barrier}
@@
 barrier for the garbage collector and second, it supports COW and adds
-objects to the write set of the transaction.
+objects to the write set of the transaction. These two jobs are
+depicted in Figure~\ref{fig:obj_states}.
 
@@
 transaction, or if there was a minor collection which will add the
-flag again on all objects. It is never set on freshly allocated
-objects that still reside in the nursery.
+flag again on all objects and may produce overflow objects. The flag
+is never set on freshly allocated objects, because they do not need
+to be traced by the GC or handled by the STM system. This fast path
+covers all cases for the GC and for the STM system.
 
@@
 
-The \textbf{slow path} is shown here in Listing~\ref{lst:w_slow}.
+The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.
 First comes the \emph{GC part}: In any case, the object will be added
@@
 created in this transaction. In that case, we do not need to execute
-the following \emph{TM part}.  We especially do not need to privatise
+the following \emph{STM part}.  We especially do not need to privatise
 its pages since no other transaction knows about these overflow
@@
 
-For the \emph{TM part}, we first perform a read barrier on the
+Up to here, we see that all cases of the GC shown in
+Figure~\ref{fig:obj_states} are handled. What is left is to handle the
+cases of the STM system that need to make old objects writable.
+
+For the \emph{STM part}, we first perform a read barrier on the
 object. We then try to acquire its write lock. \lstinline!write_locks!
@@
 all pages belonging to it (copy-on-write).
-
-\cfbolz{at this point I kind of had the need for a diagram of the various
-states an object can be in, and the transitions between them.}
 
@@
 \arigo{what else than memory usage goes here?}
+\remi{limited scalability; static segmentation?}
 

### 2014-06-05 talk/dls2014/paper/plots/bench_scaling.py
bah, bogus micro benchmark

@@ -1,20 +1,29 @@
-import thread
+import threading
 import sys
+import time
+from atomic import hint_commit_soon, atomic
 
-
-lock = thread.allocate_lock()
-
+times = []
 def workload():
+    t = time.time()
     i = 20000000
     while i:
         i -= 1
-    lock.release()
+    times.append(time.time() - t)
+    hint_commit_soon()
 
-running = range(int(sys.argv[1]))
+running = range(int(sys.argv[1]) - 1)
 
-lock.acquire()
-for i in running[:]:
-    thread.start_new_thread(workload, ())
-lock.acquire()
+ths = []
+for i in running:
+    t = threading.Thread(target=workload)
+    ths.append(t)
+    t.start()
+
+workload()
+
+[u.join() for u in ths]
+print times
+
 print "done"
 #import os; os._exit(0)

### 2014-06-05 talk/dls2014/paper/paper.tex
bah, bogus micro benchmark

@@
 thread. From this we see that there is additional overhead introduced
-by each thread (\remi{$13\%$} for all 4 threads together). While this
-is not ideal, we think that \remi{$13\%$} are acceptable on four
-threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
+by each thread (\remi{$12.3\%$} for all 4 threads together). While this
+is not ideal, we think that \remi{$12.3\%$} are acceptable on four
+threads. In terms of throughput, 4 threads have \remi{$3.54\times$}
 more iterations per second than a single thread.
-
-\remi{what we don't show is by how much this overhead is influenced
-by allocations}
-
-\cfbolz{why is the error so big with 4 threads?}
 
@@
 used for the read markers uses, per segment, up to 1/16th of the
-memory used by the old objects.  With four segments, this adds $25\%$
-to the total memory requirements.
+memory used by the old objects.  With four segments, this can add
+as much as $25\%$ to the total memory requirements.
 
@@
 investigate in more details.
+
+Besides memory usage, we also want to mention scalability. We
+think that current desktop machines with up to 8 cores should
+be seen as the main area of applications of dynamic languages.
+This is why our evaluation was performed on an off-the-shelf
+desktop CPU with only 4 cores. In the future, we will look
+at how far this STM system can really scale.
 

### 2014-06-05 talk/dls2014/paper/paper.tex
typo and size
tweaks

@@
   \centering
-  \includegraphics[width=1\columnwidth]{object_states.pdf}
-  \caption{Objects states handled by the write barrier\label{fig:obj_states}}
+  \includegraphics[width=0.8\columnwidth]{object_states.pdf}
+  \caption{Object states handled by the write barrier\label{fig:obj_states}}
 \end{figure}
@@
 benchmarks to a certain degree. It scales best for the embarrassingly
-parallel ones and a little less in the others. The reason
-for that is that in the former group there are no real, logical
-conflicts -- all threads do independent calculations. STM simply
-replaces the GIL in those programs. In the latter group,
-the threads work on a common data structure and therefore
-create much more conflicts, which limits the scalability. Here
-we make active use of the STM-supported atomic blocks that
-provide a scalable synchronisation mechanism.
+parallel ones (\remi{$avg=2.5\times$} speedup) and a little less for
+the others (\remi{$avg=1.9\times$} speedup). The reason for that is
+that in the former group there are no real, logical conflicts -- all
+threads do independent calculations. STM simply replaces the GIL in
+those programs. In the latter group, the threads work on a common data
+structure and therefore create much more conflicts, which limits the
+scalability. Here we make active use of the STM-supported atomic
+blocks in order do synchronise access to the data structure. The hope
+is that the STM system is still able to parallelise, even if we use
+the atomic blocks in a coarse-grained way. While less than the other
+benchmarks, we indeed see some speedup going from 1 to 3 threads.
+There is no visible benefit from 3 to 4 threads, even a slight
+regression.
 
@@
 that it is \remi{$\approx 35.5\%$}. The maximum in richards is
-\remi{$63\%$}.
-
-\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
-however, it rarely beats CPython's single-thread performance.  For
+\remi{$63\%$}. In all benchmarks \emph{pypy-stm-nojit} beats
+\emph{pypy-nojit} already on two threads despite of this overhead.
+However, it rarely beats CPython's single-thread performance.  For
 programs that need concurrency in CPython and that use threads to
@@
 
+First results promising; a lot to understand yet; a lot to improve
+
+contributions:
+\begin{itemize}
+\item memory segmentation, page mapping $\rightarrow$ performance
+\item STM \& GC: high allocation rate of dynlangs; shared write barrier;
+  page privatisation and re-sharing by GC; used to synchronise threads;
+  used to classify objects (transaction owned / not)
+\item GIL replaced successfully in embarrassingly parallel benchs;
+  avg overhead of $35\%$; beating GIL at 2 threads (2.5x speedup)
+\item atomic blocks introduced as a scalable sync mechanism (1.9x speedup)
+\item JIT needs work
+\end{itemize}
+
 %% \appendix

### 2014-06-05 talk/dls2014/paper/paper.tex
tweaks

@@
 \item This new STM system is used to replace the GIL in one
-  implementation of Python and is then evaluated extensively.
+  implementation of Python and is then evaluated.
 \item We introduce atomic blocks to the Python language to provide a
@@
 
-What we described so far has no mesurable performance impact by itself,
-but is missing conflict detection to be part of a full-fledged STM
-system.  This requires adding \emph{barriers} to the program to register
-the objects into the read or write set, and adapting the commit
-protocol in consequence.
+What we described so far has no mesurable performance impact by
+itself, and it is missing conflict detection to be considered a
+full-fledged STM system.  This requires adding \emph{barriers} to the
+program to register the objects into the read or write set before
+reading or writing to them\footnote{Barriers added by an automatic and
+local program transformation~\cite{felber07}}.  Furthermore, we need
+to describe the commit protocol.
 
@@
 
-% \bibitem{felber07}
-%   Felber, Pascal, et al. "Transactifying applications using an open
-%   compiler framework." \emph{TRANSACT}, August (2007): 4-6.
+\bibitem{felber07}
+  Felber, Pascal, et al. "Transactifying applications using an open
+  compiler framework." \emph{TRANSACT}, August (2007): 4-6.
 

### 2014-06-05 talk/dls2014/paper/paper.tex
completely remove "future work" section

@@
 
-\section{Future Work}
-
-\arigo{what else than memory usage goes here?}
-\remi{limited scalability; static segmentation?}
-
-We are still working on a number of optimizations in the
-implementation.  We want to mention notably that memory usage has not
-been optimized so far.  There are several sources of extra memory
-requirements in our STM system.  The multiple nurseries add
-4~MiB\footnote{Configurable value taken from experience with our
-  non-STM reference interpreter.} per segment.  Additionally, the area
-used for the read markers uses, per segment, up to 1/16th of the
-memory used by the old objects.  With four segments, this can add
-as much as $25\%$ to the total memory requirements.
-
-One of the most important sources of additional memory usage is the
-presence of privatised pages.  It is hard to predict and control.
-Right now, if a program creates a large number of privatised pages,
-major garbage collections will occur more frequently in the hope of
-resharing some of these pages.  More advanced techniques might be
-useful.
-
-Overall, early measures seem to show that a four-segment version of
-pypy-stm-nojit consumes between 2 and 7 times as much memory as a
-plain pypy-nojit.  The high part of this spectrum seems to occur only
-if the total memory usage is rather low (less than 20~MiB): it might
-simply mean that our approach has a high constant memory usage
-overhead.  The low part, i.e.\ a factor 2, can approximately be
-attributed to the reasons described above.  It is future work to
-investigate in more details.
-
-Besides memory usage, we also want to mention scalability. We
-think that current desktop machines with up to 8 cores should
-be seen as the main area of applications of dynamic languages.
-This is why our evaluation was performed on an off-the-shelf
-desktop CPU with only 4 cores. In the future, we will look
-at how far this STM system can really scale.
-
 \section{Related Work}

### 2014-06-05 talk/dls2014/paper/paper.tex
start equalising the style of writing

@@
 
-\section{High-level overview}
+\section{High-Level Overview}
 
@@
 
-A naive approach to providing complete isolation between threads is to
-partition the virtual memory of a process into $N$ segments, one per
-thread.  Each segment then holds a complete copy of the memory logically
-seen by the program.  All threads read and write data inside their own
-segment; exchanges between segments occur only at commit-time.  So far,
-this is a model that would also work for distributed transactional
-memory.
-
-Obviously, if we are in a shared-memory setting, this model is extremely
-wasteful: at a given time, most objects are identical in all, or most,
-segments.  They differ only in segments that have modified but not yet
-committed the object.  So we use a trick of the Memory Management Unit
-(MMU) of the CPU.  At the coarse unit of one page of memory (4096
-bytes), we use remappings so that, in general, pages within segments are
-\emph{shared} with each other if they would contain a fully identical
-copy.  Conversely, when one segment needs to write to one object
-contained in a page, the complete page is first \emph{privatised:} it is
-duplicated and the copy is remapped so that it will be visible in this
-segment (only).  This process is illustrated in
-Figure~\ref{fig:Page-Remapping}.
-
-The result that we get is that the virtual address space of the process
-is bloated -- by a factor $N$ -- but the real memory usage is not --
-because most pages are shared with their copies in most other segments.
+Our STM system fundamentally builds on threads. Transactions run in
+their context, and there can be only one transaction per thread at a
+time. Transactions run in parallel because these threads can run in
+parallel. In order to isolate transactions from each other, we look
+at isolating the memory accesses from threads.
+
+A naive approach to providing this complete isolation between threads
+is to partition the virtual memory of a process into $N$ segments, one
+for each thread currently executing.  Each segment then holds a
+complete copy of the memory logically seen by the program.  All
+threads read and write data inside their own segment; exchanges
+between segments occur only when transactions commit.  So far, this is a
+model that would also work for distributed transactional memory.
+
+Clearly, if we are in a shared-memory setting, this model is extremely
+wasteful: at any given time, most objects are identical in all, or
+most, segments.  They differ only in segments that have modified but
+not yet committed these objects.  So we use a trick of the Memory
+Management Unit (MMU) of the CPU.  The MMU maps pages of virtual to
+physical memory.  By changing this page mapping, at the coarse unit of
+one page of memory (4096 bytes), we can share memory that would
+otherwise exist as identical copies in all segments.
+
+In the common case, where objects are identical in all segments, we
+can thus make these pages \emph{shared}. Conversely, when one thread~/
+segment wants to modify an object, all pages of this object need to be
+\emph{privatised} for this segment. This involves duplicating the
+pages and changing the page mapping so that the private pages are only
+accessible by this segment and full isolation is guaranteed.  This
+process is illustrated in Figure~\ref{fig:Page-Remapping}.
+
+The result is that the virtual address space of the process is bloated
+by a factor $N$ -- but the real memory usage is not, because most
+pages are shared with their identical copies in other segments.
 
@@
 full-fledged STM system.  This requires adding \emph{barriers} to the
-program to register the objects into the read or write set before
+program to register the objects in the read or write set before
 reading or writing to them\footnote{Barriers added by an automatic and
@@
   interpreter for a complex language) contains a number of reads that
-  typically exceeds many times the number of writes.  We managed to
+  typically exceeds many times the number of writes.  The described
+  approach allows the read barrier to be a single, unconditional memory
+  write into segment-local memory.
+ We managed to
   design the read barrier so that it is done by one memory write

### 2014-06-05 talk/dls2014/paper/paper.tex
adapt the style of writing

@@
 complete copy of the memory logically seen by the program.  All
-threads read and write data inside their own segment; exchanges
-between segments occur only when transactions commit.  So far, this is a
-model that would also work for distributed transactional memory.
+threads read and write data inside their own segment; and this data
+gets copied to other segments in an atomic fashion only when a
+transaction commits.  So far, this is a model that would also work for
+distributed transactional memory.
 
@@
 
-When we need to store into our objects pointers to other objects, we
-need to be careful: if we were to directly store regular addresses valid
-in the local segment, then the objects' binary content would have to differ
-between segments.  This would prevent any sharing.  Instead, we store
-pointers as \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start
-of the segment.  A $SO$ can then be interpreted from any segment by
-adding it to the base address of that segment.\footnote{An alternative
-model would be to simulate threads using several independent processes
-in which all segments are at the same address.  This might work well for
+When we store into our objects the references to other objects, we
+need to be careful: addresses that are valid in one segment are not
+valid in another, because they would break isolation by pointing to
+the copy of the object in another segment.  Consequently, we would
+need to adapt these addresses to the segment they live in, which would
+prevent completely any page sharing.  Instead, we store pointers as
+\emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start of the
+segment.  A $SO$ can then be interpreted in any segment by adding to
+it the start address of that segment.\footnote{An alternative model
+would be to simulate threads using several independent processes in
+which all segments are at the same address.  This might work well for
 demonstration purposes, but it is not really suitable for implementing
-an existing programming language: programs assume that most of the state
-(file descriptors, external libraries, ...) is shared.} The result is
-called a \emph{Linear Address (LA)}. This is illustrated in
+an existing programming language: programs assume that most of the
+state (file descriptors, external libraries, ...) is shared.} The
+result is called a \emph{Linear Address (LA)}. This is illustrated in
 Figure~\ref{fig:Segment-Addressing}.
 
-The x86 CPUs provide a feature which is precisely called \emph{memory
+The x86 CPUs provide a feature which is called \emph{memory
 segmentation}.  On the modern x86-64, this is enough to perform the
-addition described above very efficiently.  We use the segment register
-$\%gs$, which is available to applications, to point to the
-thread's segment start address.\footnote{The other segment register
-$\%fs$ is typically used by the threading library to provide access to
-thread-local data.  One point of view is that we are using $\%gs$ for a
-similar purpose: thread-local data -- but a lot of it.}  The translation
-from a $SO$ to a LA is done for us by using the ``$\%gs\colon$'' prefix
-before any CPU instruction that reads or writes memory using a $SO$
-address.  This is very efficient and some compilers support it natively
-(e.g.\ clang).
-
-If support for this is not available (due to the compiler or on a
-different architecture), there are other optimizations possible.  For
-example, using gcc, a regular register can be globally reserved and
-all ``simple enough'' memory accesses will still resolve to a single
-instruction.  More complicated accesses, e.g.\ those involving array
-indexing, will need an extra explicit addition.
+addition described above very efficiently.  We use the segment
+register $\%gs$\footnote{The other segment register $\%fs$ is
+typically used by the threading library to provide access to
+thread-local data.  One point of view is that we are using $\%gs$ for
+a similar purpose: thread-local data -- but a lot of it.}, which is
+available to applications, to point to the current thread's segment
+start address.  The translation from a $SO$ to a LA is done for us by
+using the ``$\%gs\colon$'' prefix before any CPU instruction that
+reads or writes memory using a $SO$ address.  This is very efficient
+(we can do it on every object access) and some compilers support it
+natively (e.g.\ clang).
+
+In summary, translating a $\%gs:SO$ to a physical address is a
+two-step process: First the memory segmentation feature of the CPU
+constructs a linear address. Then, this LA gets mapped by the MMU to
+some place in the physical memory. This makes the $SO$ a valid reference
+to an object in all segments.
+
+% If support for this is not available (due to the compiler or a
+% different CPU architecture), there are other optimisations possible.
+% For example, using gcc, a regular register can be globally reserved
+% and all ``simple enough'' memory accesses will still resolve to a
+% single instruction.  More complicated accesses, e.g.\ those involving
+% array indexing, will need an extra explicit addition.
 
@@
 
-What we described so far has no mesurable performance impact by
-itself, and it is missing conflict detection to be considered a
+What we described so far has no measurable performance overhead by
+itself, but it is missing conflict detection to be considered a
 full-fledged STM system.  This requires adding \emph{barriers} to the
@@
 reading or writing to them\footnote{Barriers added by an automatic and
-local program transformation~\cite{felber07}}.  Furthermore, we need
-to describe the commit protocol.
+local program transformation~\cite{felber07}}.  Furthermore, we still
+need to describe the commit protocol.
 
@@
 
-\item [{Read~Barrier:}] A major contribution to our performance results
+\item [{Read~Barrier:}] A major contribution to our performance
   is the simplicity of the read barrier: a large program (like an
@@
   typically exceeds many times the number of writes.  The described
-  approach allows the read barrier to be a single, unconditional memory
-  write into segment-local memory.
- We managed to
-  design the read barrier so that it is done by one memory write
-  (setting a flag) into non-shared memory.  This is done with a
-  segment-local array of flags indexed by a number derived from the $SO$
-  of the object.  Unlike other STM systems, the read barrier does not
-  have to find the private version of some object -- our two-step
-  address translation automatically resolves the reference to the
-  private version on every access anyway.  This means the read barrier
-  is not constrained to execute before the actual read -- both the
-  compiler and the CPU are free to reorder them.
-
-\item [{Write~Barrier:}] This is triggered by checking a flag in the
-  header of the object.  When the flag is set, we call the slow path,
-  which adds the object into the write set and unconditionally resets
-  the flag.  We eagerly detect write-write conflicts by allowing only
-  one transaction modifying an object at a time (by acquiring a write
+  approach allows the read barrier to be a single, unconditional
+  write into segment-local memory. We only need to mark an object
+  as read by the current transaction. The mark is stored in an
+  segment-local array indexed by a number derived from the $SO$
+  of the object.
+
+  Unlike other STM systems, the read barrier does not have to find the
+  private version of the object -- our two-step address translation
+  automatically resolves the reference to the private version on every
+  access anyway\footnote{Additional benefit: The read barrier is not
+    constrained to execute before the actual read -- both the compiler and
+    the CPU are free to reorder them.}.
+
+\item [{Write~Barrier:}] This barrier is triggered by checking a flag
+  on the object that signifies if we did not already execute the write
+  barrier on this object before.  When the flag is set, we call the slow
+  path, which adds the object into the write set and unconditionally
+  resets the flag.
+
+  Here, we eagerly detect write-write conflicts by allowing only one
+  transaction modifying an object at a time (by acquiring a write
   lock).\footnote{Eager write-write detection is not inherent to our
-  approach; future experiments may show that we want to lift this
-  restriction.}  This slow path is also responsible for privatising
-  pages,\footnote{The write set contains objects, not pages.} as
-  described above.  Interestingly, we are using a generational garbage
-  collector (described below) which requires a similar write barrier
-  anyway -- so the same header flag is used for both cases.  The flag
-  means ``something needs to be done before writing to this object'',
-  and the slow path logic determines what it is.
+    approach; we may lift this restriction in the future.}  Part of making
+  the object write-ready is also the privatisation of all pages the
+  object belongs to as described above.
+
+  Described below, we also need a write barrier for our generational
+  garbage collector (GC). Interestingly, we are able to share the same
+  barrier and flag for both use cases. Since the GC is an integral
+  part of many dynamic languages, we actually do not incur any additional
+  overhead here since the barriers are already there.
 
@@
   for read-write conflicts.  Such conflicts cannot be detected in the
-  minimal read barrier, but can be detected during commit: the case to
-  look for is that the committing transaction wrote to an object that
-  another transaction has already read.\footnote{Which is why the read
-  sets can be just sparsely populated arrays of flags: we never walk the
-  whole read set.} A priori, checking other transactions' read sets
-  would involve reading concurrently modified memory: a different thread
-  could be issuing a read barrier in parallel with an unprotected
-  commit.  To prevent this from occurring, we must suspend all other
-  threads at known safe points.  This is a trade-off for the simplicity
-  of our read barriers -- which is well worth it in our case, due to the
-  very high number of read barriers when compared to the number of
-  commits and the (low) total number of threads.
+  minimal read barrier, but can be detected during commit: we check
+  if any of the objects in our write set was already read by another
+  transaction. We never walk the read set, but just check if the objects
+  are marked in other segments.
+
+  Checking other transactions' read marks would involve reading
+  concurrently modified memory: another thread could be issuing a read
+  barrier in parallel with a commit.  To prevent this from occurring, we
+  must suspend all other threads at known safe points.  This is a
+  trade-off for the simplicity of our read barrier -- which is well
+  worth it in our case, due to the very high number of read barriers
+  when compared to the number of commits.
 
@@
 
-Conflicts can be resolved by aborting one of the two transactions, or in
-some cases by pausing one of them until the other commits.  If a commit
-succeeds, we copy the changes to modified objects so that they become
-visible in other segments.  Note that this only includes objects that
-existed before the transaction and have been modified by it.  It does
-not include all the new objects created by the current transaction
-(there is typically a very high rate of object churn in dynamic languages).
-Our garbage collector allocates new objects in pages that are shared in
-the first place, so that they do not need to be copied -- and moreover,
-such objects will commonly never be modified any more.
-
-Privatised pages are re-shared during major collection cycles.
+Conflicts can be resolved by aborting one of the two transactions, or
+in some cases by pausing one of them until the other commits.  If a
+commit succeeds, we copy the changes made to the objects in the write
+set so that they become visible in other segments.  Note that this
+only includes objects that existed before the transaction and have
+been modified by it.  It does not include all the new objects created
+by the current transaction, which typically outnumber modified objects
+by far in dynamic languages.  We allocate these new objects in pages
+that are still shared, so that they do not need to be copied to other
+segments at all.
 

### 2014-06-05 talk/dls2014/paper/paper.tex
introduction overhaul by trg

@@
 
+\newcommand{\comment}[1]{}
+
 \begin{document}
@@
 
-\title{Virtual Memory Assisted Transactional Memory for Dynamic Languages}
-\subtitle{DLS'14}
+\title{
+}
+%\subtitle{DLS'14}
+
+\comment{
+A Platform for Parallelism in Dynamic languages
+Parallelism for Python
+A TM Implementation for Python
+Transactional Memory as a foundation for parallel Python
+Transactional Memory as a foundation for parallelism in Python
+Parallel Python based on VM-assisted TM
+Parallel Python based on VM-assisted Transactional Memory
+A Platform for parallel Python
+A TM Platform for parallel Python
+A Transactional Memory Platform for parallel Python
+A Transactional Memory system for parallel Python
+
+Virtual Memory Assisted Transactional Memory for Dynamic Languages}
 
@@
 
-Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
-regarded as very expressive but also very slow. In recent years, the
-introduction of just-in-time compilers (JIT) for these languages
-(e.g.\ PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
-started to change this perception by delivering
-good performance that enables new applications. However, a parallel
-programming model was not part of the design of those languages. Thus,
-the reference implementations of e.g.\ Python and Ruby use a single,
-global interpreter lock (GIL) to serialise the execution of code in
-threads.
-
-While this GIL prevents any parallelism from occurring, it also
-provides some useful guarantees. Since this lock is always acquired
-while executing bytecode instructions and it may only be released
-in-between such instructions, it provides perfect isolation and
-atomicity between multiple threads for a series of
-instructions. Additionally, it provides the application with a
-sequential consistency model~\cite{lamport79}. Another technology that
-can provide the same guarantees is transactional memory (TM).
+Dynamic languages like Python, PHP, Ruby, and JavaScript receive a lot of attention but
+have not yet embraced parallelism. A parallel programming model was not part of the design
+of those languages. Thus, the reference implementations of, e.g., Python and Ruby use a
+single, global interpreter lock (GIL) to serialise the execution of code in threads. The
+use of a single global lock causes several disadvantages, and as multicore processors
+become the de-facto standard in platfroms ranging from smart phones to datacenters, the
+issue of parallel execution must be addressed by these dynamic languages.
+
+Unfortunately the GIL plays a central role in current implementations. Even
+recent just-in-time compilers (JIT) for these languages
+(e.g., PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
+are based on the GIL; Jython (an implementation of Python that is based on the JVM) is a
+notable exception.
+
+The standard model employed in these systems is fairly straightforward: there can be
+multiple threads, but each acquires the GIL, then performs a single instruction, and
+subsequently releases the lock.
+While this setup limits the benefits due to parallelism
+%% not clear we must say so TRG
+% -- parallelism is limited to I/O or other activities external to the interpreter
+the use of a GIL also provides some crucial guarantees. Since this lock is always acquired
+while executing bytecode instructions and may be released only in-between such
+instructions, the GIL ensures perfect isolation between multiple threads for a series of
+instructions, which are executed atomically.  In addition, the GIL provides the
+application with a sequential consistency model~\cite{lamport79}.
+
+A global interpreter lock assures the atomic execution of individual instructions but
+offers the programmer no help in managing concurrent data structures. Many parallel
+programs must maintain invariants involving multiple objects (or even multiple fields of a
+single object).  So to enable Python programs for parallel execution, we need a way to
+identify which blocks of data must be executed atomically and hardware or software support
+to effect this atomic execution.
+
+\emph{Atomic
+blocks}~\cite{tim03,tim05} allow dynamic languages to indicate
+the units of computation that must be executed atomically to ensure correct concurrent
+execution.  Atomic blocks map nicely to transactional memory, which provide a convenient
+mechansim to replace the GIL.
+ Using transactions to enclose
+multiple bytecode instructions, we can get the very same semantics as
+provided by the GIL while possibly executing several transactions in
+parallel. And by exposing these interpreter-level
+transactions to the application, we
+avoid the well-known problems caused by locks as applications use atomic blocks instead
+of explicit acquire/release operations on locks.
 
 There have been several attempts at replacing the GIL with
-TM~\cite{nicholas06,odaira14,fuad10}. Using transactions to enclose
-multiple bytecode instructions, we can get the very same semantics as
-the GIL while possibly executing several transactions in
-parallel. Furthermore, by exposing these interpreter-level
-transactions to the application in the form of \emph{atomic
-blocks}~\cite{tim03,tim05}, we give dynamic languages a new
-synchronisation mechanism that avoids several of the problems of locks
-as they are used now.
-
-TM systems can be broadly categorised as hardware based (HTM),
-software based (STM), or hybrid systems (HyTM). HTM systems are limited
-by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
-a lot of overhead~\cite{cascaval08,drago11}. In~\cite{wayforward14},
-we argue that STM is still the best way forward, especially since it
-supports large atomic blocks as a new way for synchronising multiple
-threads. There have been several attempts at lowering the overhead
-of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
-In this paper, we describe how we manage to lower the overhead of our
-STM system so that it can be seen as a viable replacement for the GIL.
+TM~\cite{nicholas06,odaira14,fuad10}.
+We focuse here on
+software based transactional memory (STM)
+because they provide the unique opportunity to consider the issues of atomicity,
+synchronization, and memory management (resp. garbage collection) together.
+STM systems impose no a-priori bound on the size of an atomic block and provide therefore
+a flexibility that is absent from current hardware-based TM systems~\cite{wayforward14}.
+The downside of STM systems is runtime overhead~\cite{cascaval08,drago11}, which can
+be substantial despite recent efforts to lower the overhead~\cite{warmhoff13,spear09}.
+In this paper, we describe how we manage to lower the overhead of an
+STM system so that it can serve as a viable replacement for the GIL by leveraging the
+demands of a dynamic language like Python to integrate the STM system with the garbage
+collector.
 
 \cfbolz{both the introduction and the contributions need to stress the co-design of GC and STM}
-
-Our contributions include:
+% agree trg
+
+This paper makes the following contributions:
 \begin{itemize}[noitemsep]
 \item We introduce a new software transactional memory (STM) system
-  that performs well on low numbers of CPUs.
-\item We make novel use of the memory management capabilities of
-  off-the-shelf CPUs, e.g. memory segmentation and 64bit address space.
-\item We integrate our STM system closely with a garbage collector
-  (GC) in order to lower the overhead of STM.
+  that performs well for Python programs.
+\item We integrate the  system closely with a garbage collector
+  (GC) to lower the overhead of STM.
 \item This new STM system is used to replace the GIL in one
   implementation of Python and is then evaluated.
-\item We introduce atomic blocks to the Python language to provide a
-  backwards compatible, composable synchronisation mechanism for
+\end{itemize}
+The STM described here is efficient and has low overhead and is therefore beneficial also
+for environments with a small number of threads or cores. The paper describes also
+%% can be dropped, not the core topic
+how to introduce atomic blocks into the Python language to provide a
+  backwards compatible, composable synchronization mechanism for
   threads.
-\end{itemize}
+
+%% What is really needed from this section?? Maybe the subsection on Python can be moved
+%%  to the intro -- the rest is questionable.  TRG
 

### 2014-06-06 talk/dls2014/paper/paper.tex
tweaks and implement some more feedback

@@
 
-\title{
-}
+\title{A Transactional Memory System for Parallel Python}
 %\subtitle{DLS'14}
@@
            {arigo@tunes.org}
+\authorinfo{Carl Friedrich Bolz}
+           {King's College London}
+           {http://cfbolz.de}
+\authorinfo{Thomas Gross}
+           {Department of Computer Science\\ ETH Zürich}
+           {thomas.gross@inf.ethz.ch}
 
@@
 avoid the well-known problems caused by locks as applications use atomic blocks instead
-of explicit acquire/release operations on locks.
+of explicit acquire/release operations on locks~\cite{christopher10,victor11,shan08}.
 
 There have been several attempts at replacing the GIL with
-TM~\cite{nicholas06,odaira14,fuad10}.
-We focuse here on
-software based transactional memory (STM)
+TM~\cite{nicholas06,odaira14,fuad10}.  We focus here on software based
+transactional memory (STM) systems,
 because they provide the unique opportunity to consider the issues of atomicity,
@@
 
-\subsection{Global Interpreter Lock}
-
-The GIL is a very simple synchronisation mechanism for supporting
-multithreading in an interpreter. The basic guarantee is that the GIL
-may only be released in between bytecode instructions. Thus, these
-instructions are always executed atomically and in complete isolation
-from others running in other threads. \emph{Atomicity} means that each
-instruction and its effects seem to happen at one, indivisible point
-in time. Other instructions never see inconsistent state of a
-partially executed instruction (\emph{isolation}).
-
-In addition to these guarantees, instructions are executed in a
-sequential consistency model~\cite{lamport79}. This means that
-the outcome of any execution of instructions in multiple threads is
-equal to \emph{some} sequential execution of them.
-
-\subsection{Transactional Memory}
-
-Transactional memory (TM) is a concurrency control mechanism that
-comes from database systems. Using transactions, we can group a series
-of instructions performing operations on memory and make them happen
-atomically and in complete isolation from other transactions.
-Atomicity and isolation are basic properties of transactions.
-
-If we start multiple such transactions in multiple threads, the TM
-system guarantees that the outcome of running the transactions is
-\emph{serialisable}. Meaning, the outcome is equal to some sequential
-execution of these transactions. By that, we can again provide a
-sequentially consistent model for programming in multiple threads. We
-can therefore use TM to directly replace the GIL. Instead of releasing
-and acquiring the GIL between bytecode instructions, we commit and
-start the transactions our instructions are running in.
+% \subsection{Global Interpreter Lock}
+
+% The GIL is a very simple synchronisation mechanism for supporting
+% multithreading in an interpreter. The basic guarantee is that the GIL
+% may only be released in between bytecode instructions. Thus, these
+% instructions are always executed atomically and in complete isolation
+% from others running in other threads. \emph{Atomicity} means that each
+% instruction and its effects seem to happen at one, indivisible point
+% in time. Other instructions never see inconsistent state of a
+% partially executed instruction (\emph{isolation}).
+
+% In addition to these guarantees, instructions are executed in a
+% sequential consistency model~\cite{lamport79}. This means that
+% the outcome of any execution of instructions in multiple threads is
+% equal to \emph{some} sequential execution of them.
+
+% \subsection{Transactional Memory}
+
+% Transactional memory (TM) is a concurrency control mechanism that
+% comes from database systems. Using transactions, we can group a series
+% of instructions performing operations on memory and make them happen
+% atomically and in complete isolation from other transactions.
+% Atomicity and isolation are basic properties of transactions.
+
+% If we start multiple such transactions in multiple threads, the TM
+% system guarantees that the outcome of running the transactions is
+% \emph{serialisable}. Meaning, the outcome is equal to some sequential
+% execution of these transactions. By that, we can again provide a
+% sequentially consistent model for programming in multiple threads. We
+% can therefore use TM to directly replace the GIL. Instead of releasing
+% and acquiring the GIL between bytecode instructions, we commit and
+% start the transactions our instructions are running in.
 
 \subsection{Python}
+\remi{possibly adapt to application view \& benefits (atomic blocks)}
+\remi{XXX: Also, make own section or integrate with introduction}
 
@@
 
-\subsection{Synchronisation}
-
-In Python, since the GIL is not directly exposed to the interpreter,
-applications still need to synchronise memory accesses from multiple
-threads using locks. Locks can be very hard to get
-right~\cite{christopher10,victor11,shan08}.  They are non-composable,
-have overhead, may deadlock, limit scalability, and add to the overall
-complexity of the program logic. We think that \emph{atomic
-blocks}~\cite{tim03,tim05} provide a better way for synchronisation.
-
-Atomic blocks are composable, deadlock-free, higher-level and expose
-useful atomicity and isolation guarantees to the application for a
-series of instructions. This is why we think that the introduction
-of atomic blocks to Python is a valuable contribution. Since atomicity
-is a property of transactions, TM and atomic blocks are a natural fit.
+% \subsection{Synchronisation}
+
+% In Python, since the GIL is not directly exposed to the interpreter,
+% applications still need to synchronise memory accesses from multiple
+% threads using locks. Locks can be very hard to get
+% right~\cite{christopher10,victor11,shan08}.  They are non-composable,
+% have overhead, may deadlock, limit scalability, and add to the overall
+% complexity of the program logic. We think that \emph{atomic
+% blocks}~\cite{tim03,tim05} provide a better way for synchronisation.
+
+% Atomic blocks are composable, deadlock-free, higher-level and expose
+% useful atomicity and isolation guarantees to the application for a
+% series of instructions. This is why we think that the introduction
+% of atomic blocks to Python is a valuable contribution. Since atomicity
+% is a property of transactions, TM and atomic blocks are a natural fit.
 
@@
 management that generally prefers the older transaction to the younger.
+This gives long transactions a better chance to succeed.
 
@@
 need to be careful: addresses that are valid in one segment are not
-valid in another, because they would break isolation by pointing to
+valid in another, because they would break isolation by possibly pointing to
 the copy of the object in another segment.  Consequently, we would
 need to adapt these addresses to the segment they live in, which would
-prevent completely any page sharing.  Instead, we store pointers as
+prevent completely any page sharing because then the binary contents
+differ between segments.  Instead, we store pointers as
 \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start of the
-segment.  A $SO$ can then be interpreted in any segment by adding to
+segment.  An $SO$ can then be interpreted in any segment by adding to
 it the start address of that segment.\footnote{An alternative model
@@
 available to applications, to point to the current thread's segment
-start address.  The translation from a $SO$ to a LA is done for us by
+start address.  The translation from an $SO$ to an LA is done for us by
 using the ``$\%gs\colon$'' prefix before any CPU instruction that
-reads or writes memory using a $SO$ address.  This is very efficient
-(we can do it on every object access) and some compilers support it
+reads or writes memory using an $SO$ address.  This is very efficient
+-- we can do it on every object access -- and some compilers support it
 natively (e.g.\ clang).
 
-In summary, translating a $\%gs:SO$ to a physical address is a
+In summary, translating an $\%gs:SO$ to a physical address is a
 two-step process: First the memory segmentation feature of the CPU
@@
 \item [{Read~Barrier:}] A major contribution to our performance
-  is the simplicity of the read barrier: a large program (like an
-  interpreter for a complex language) contains a number of reads that
-  typically exceeds many times the number of writes.  The described
+  is the simplicity of the read barrier: a large program, like an
+  interpreter for a complex language, contains a number of reads that
+  typically vastly exceeds the number of writes.  The described
   approach allows the read barrier to be a single, unconditional
-  write into segment-local memory. We only need to mark an object
-  as read by the current transaction. The mark is stored in an
+  write into segment-local memory. We only need to \emph{mark} an object
+  as read by the current transaction. The mark is stored in a
   segment-local array indexed by a number derived from the $SO$
@@
   automatically resolves the reference to the private version on every
-  access anyway\footnote{Additional benefit: The read barrier is not
-    constrained to execute before the actual read -- both the compiler and
-    the CPU are free to reorder them.}.
+  access anyway.
 
@@
   Here, we eagerly detect write-write conflicts by allowing only one
-  transaction modifying an object at a time (by acquiring a write
-  lock).\footnote{Eager write-write detection is not inherent to our
+  transaction modifying an object at a time using write
+  locks.\footnote{Eager write-write detection is not inherent to our
     approach; we may lift this restriction in the future.}  Part of making
   the object write-ready is also the privatisation of all pages the
-  object belongs to as described above.
+  object belongs to, as described above.
 
@@
 Conflicts can be resolved by aborting one of the two transactions, or
-in some cases by pausing one of them until the other commits.  If a
+in some cases by pausing the writing one until the reader commits.  If a
 commit succeeds, we copy the changes made to the objects in the write
@@
 (GC)} that is closely integrated with the TM part (e.g.\ it shares the
-write barrier). The close integration helps in order to know more
-about the lifetime of an object, as will be explained in the following
-sections.
+write barrier).
 
@@
 void stm_write(object_t *SO)
-object_t *stm_allocate(ssize_t size_rounded)
+object_t *stm_allocate(size_t size_rounded)
 STM_PUSH_ROOT(object_t *SO)
@@
 \lstinline!stm_commit_transaction()! tries to commit the current
-transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
+transaction, aborting it otherwise. \lstinline!stm_read()!,
+\lstinline!stm_write()!  perform
 a read or a write barrier on the object given as the argument, and
@@
   alive.~\cite{fergus02}} of the current thread.  References to objects
-have to be saved using this stack around calls that may cause a GC
+have to be saved using this stack locally around calls that may cause a GC
 cycle to happen, and also while there is no transaction
@@
 the \emph{sharing-segment} that is never assigned to a thread. The
-sharing-segment is essentially read-only except during commit.
+sharing-segment is essentially read-only, except during commit.
 
@@
 \item [{Segment-local~data:}] Some area private to the segment that
-  contains segment-local information.
-\item [{Read~markers:}] These are pages that store information about
+  contains segment-local information for bookkeeping.
+\item [{Read~markers:}] These are private pages that store information about
   which objects were read in the current transaction running in this
   segment.
-\item [{Nursery:}] This area contains all the freshly allocated
+\item [{Nursery:}] This private area contains all the freshly allocated
   objects (\emph{young objects}) of the current transaction. The GC
@@
 
-At the start of a transaction, the thread it is running in acquires a
+At the start of a transaction, the thread needs to acquire a
 segment. It may have to wait until another thread finishes its
-transaction and releases a segment. Fairness is not guaranteed yet, as
+transaction and releases its segment. Fairness is not guaranteed yet, as
 we simply assume a fair scheduling policy in the operating system when
@@
 starts a transaction. Although, we try to assign it the same segment
-again if possible.
+again if possible to maximise the effectiveness of caches and the
+page privatisation.
 
@@
 out of a thread's nursery into the old object space, which can be done
-without stopping other threads. This is happens either if the nursery has
+without stopping other threads. This happens either if the nursery has
 no space left anymore or if we are committing the current
-transaction. Consequently, all objects are old and the nursery empty
+transaction. Consequently, all objects are \emph{old} and the nursery empty
 after a transaction commits.  Furthermore, all objects in the nursery
@@
 medium-fast path inside barriers that the object still belongs to the
-current transaction. \remi{so this is where we mention how the GC-STM
-integration is useful. highlight more or move to own section?}
-\cfbolz{definitely its own section listing all the benefits}
+current transaction. Other and later transactions will treat them
+like regular \emph{old} objects.
 
@@
 kind of way and first force minor collections in all threads. The
-major goal is to free objects in the old objects space. Furthermore,
-we optimistically re-share pages that do not need to be private
-anymore.
+major goal is to free objects in the old objects space.
+
+Furthermore, we optimistically re-share pages during major collections.
+This is another example where the STM-GC cooperation makes sense. We
+regard private pages as additional memory that may be collected.
+Pages should usually stay private for some time, but at some regular
+intervals, we still want to try to re-share pages in case they actually
+stay unmodified in the future. Major collections are allowed to do
+that because all threads are synchronised, and we can re-share all
+private pages that do not currently contain an object in some
+transaction's  write set.
 
@@
 a private copy, but since the CPU performs our address translation on
-every object access efficiently, we do not need to do that in our
-barrier.
+every object access efficiently, we can avoid this extra overhead.
 
@@
 that each object has its own read marker, all objects have a size of
-at least 16 bytes. Otherwise there would be false conflicts when
+at least 16 bytes. Otherwise there could be false conflicts when
 reading from two adjacent objects that share a single marker.
@@
 transactions. The whole code for the barrier is easily optimisable for
-compilers as well as perfectly predictable for CPUs:
+compilers as well as perfectly predictable for
+CPUs\footnote{Additional benefit: The read barrier is not constrained
+to execute before the actual read -- both the compiler and the CPU are
+free to reorder or group them.}:
 
@@
 write barrier once between collections, and only the old objects need
-to be handled by the TM part once per transaction. This keeps the
+to be handled by the STM part \textbf{once per transaction}. This keeps the
 overhead of the write barrier very low.
@@
 
-For other synchronisation requirements, for example:
-\begin{itemize}[noitemsep]
-\item waiting for a segment to be released,
-\item waiting for a transaction to abort or commit,
-\item waiting for all threads to reach their safe points,
-\end{itemize}
-we use a set of condition variables to wait or signal other threads.
+% For other synchronisation requirements, for example:
+% \begin{itemize}[noitemsep]
+% \item waiting for a segment to be released,
+% \item waiting for a transaction to abort or commit,
+% \item waiting for all threads to reach their safe points,
+% \end{itemize}
+% we use a set of condition variables to wait or signal other threads.
 
@@
 \item prefer transactions that started earlier to younger transactions
+  in order to let long transactions succeed more often
 \item to support \emph{inevitable} transactions, we always prefer them
@@
 We evaluate our system in a Python interpreter called
-PyPy\footnote{www.pypy.org} (version 2.3). PyPy is an implementation of an
-interpreter for the Python language. It has a special focus on speed,
-as it provides a just-in-time (JIT) compiler to speed up applications
+PyPy~\cite{pypy} (version 2.3). PyPy is an implementation of an
+interpreter for the Python language. It has a special focus on speed
+and provides a just-in-time (JIT) compiler to speed up applications
 running on top of it. For comparison, we compare between normal PyPy
@@
 average and the standard deviation as error bars. For the JIT
-benchmarks, we first let warm up the JIT by doing a few
+benchmarks, we first let the JIT warm up by doing a few
 additional iterations to get a single measurement.
@@
 For the following sections we use a set of six small benchmarks
-available on \cite{pypybenchs}:
+available at~\cite{pypybenchs}:
 
@@
   \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
-  \caption{Comparing execution time between interpreters without a JIT\label{fig:performance-nojit}}
+  \caption{Comparing execution time between interpreters without a JIT
+    and using small input sizes\label{fig:performance-nojit}}
 \end{figure}
@@
 code worthwhile, we increased the input size of all benchmarks to get
-reasonable execution times.
+reasonable execution times (denoted with ``(large)'' in benchmark names).
 
@@
   \includegraphics[width=1\columnwidth]{plots/performance.pdf}
-  \caption{Comparing execution time between interpreters with JIT\label{fig:performance-jit}}
+  \caption{Comparing execution time between interpreters with JIT
+    and large input sizes\label{fig:performance-jit}}
 \end{figure}
@@
 HTM capacity much faster.  This, and the fact that HTM's capacity
-limits do not allow the implementation of arbitrary atomic blocks, let
+limits do not allow the implementation of arbitrary-sized atomic blocks, let
 us to believe that STM is the right choice at the moment.
@@
 of CPU cores. \cite{spear09} works well for some workloads, in the
-case of a dynamic language VM, we have to assume any possible
+case of a dynamic language VM, however, we have to assume any possible
 workload.  \cite{warmhoff13} uses a compile-time approach to select
@@
 atomicity between transactional and non-transactional code.  We,
-however, use inevitable transactions to provide strong atomicity.
-Additionally, while they use page fault handlers to detect conflicts,
-we use traditional write barriers. Again, there are only some
-superficial similarities between our approach and theirs.
+however, use inevitable transactions to provide strong atomicity
+for non-reversible operations and otherwise do not have to deal
+with non-transactional code.  Additionally, while they use page fault
+handlers to detect conflicts, we use traditional write barriers. There
+are only some superficial similarities between our approach and
+theirs.
 
@@
 \bibitem{ironpython} IronPython. \url{www.ironpython.net}
+\bibitem{pypy} PyPy Project. \url{www.pypy.org}
 

### 2014-06-06 talk/dls2014/paper/paper.tex
tweaks

@@
            {arigo@tunes.org}
-\authorinfo{Carl Friedrich Bolz}
-           {King's College London}
-           {http://cfbolz.de}
 \authorinfo{Thomas Gross}
@@
 
-Dynamic languages like Python, PHP, Ruby, and JavaScript receive a lot of attention but
-have not yet embraced parallelism. A parallel programming model was not part of the design
-of those languages. Thus, the reference implementations of, e.g., Python and Ruby use a
-single, global interpreter lock (GIL) to serialise the execution of code in threads. The
-use of a single global lock causes several disadvantages, and as multicore processors
-become the de-facto standard in platfroms ranging from smart phones to datacenters, the
-issue of parallel execution must be addressed by these dynamic languages.
-
-Unfortunately the GIL plays a central role in current implementations. Even
-recent just-in-time compilers (JIT) for these languages
-(e.g., PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
-are based on the GIL; Jython (an implementation of Python that is based on the JVM) is a
-notable exception.
-
-The standard model employed in these systems is fairly straightforward: there can be
-multiple threads, but each acquires the GIL, then performs a single instruction, and
-subsequently releases the lock.
-While this setup limits the benefits due to parallelism
-%% not clear we must say so TRG
-% -- parallelism is limited to I/O or other activities external to the interpreter
-the use of a GIL also provides some crucial guarantees. Since this lock is always acquired
-while executing bytecode instructions and may be released only in-between such
-instructions, the GIL ensures perfect isolation between multiple threads for a series of
-instructions, which are executed atomically.  In addition, the GIL provides the
-application with a sequential consistency model~\cite{lamport79}.
-
-A global interpreter lock assures the atomic execution of individual instructions but
-offers the programmer no help in managing concurrent data structures. Many parallel
-programs must maintain invariants involving multiple objects (or even multiple fields of a
-single object).  So to enable Python programs for parallel execution, we need a way to
-identify which blocks of data must be executed atomically and hardware or software support
-to effect this atomic execution.
-
-\emph{Atomic
-blocks}~\cite{tim03,tim05} allow dynamic languages to indicate
-the units of computation that must be executed atomically to ensure correct concurrent
-execution.  Atomic blocks map nicely to transactional memory, which provide a convenient
-mechansim to replace the GIL.
- Using transactions to enclose
-multiple bytecode instructions, we can get the very same semantics as
-provided by the GIL while possibly executing several transactions in
-parallel. And by exposing these interpreter-level
-transactions to the application, we
-avoid the well-known problems caused by locks as applications use atomic blocks instead
-of explicit acquire/release operations on locks~\cite{christopher10,victor11,shan08}.
+Dynamic languages like Python, PHP, Ruby, and JavaScript receive a lot
+of attention but have not yet embraced parallelism. A parallel
+programming model was not part of the design of those languages. Thus,
+the reference implementations of, e.g., Python and Ruby use a single,
+global interpreter lock (GIL) to serialise the execution of code in
+threads. The use of a single global lock causes several disadvantages,
+and as multicore processors become the de-facto standard in platfroms
+ranging from smart phones to datacenters, the issue of parallel
+execution must be addressed by these dynamic languages.
+
+Unfortunately the GIL plays a central role in current
+implementations. Even recent just-in-time compilers (JIT) for these
+languages (e.g., PyPy~\cite{cfbolz09}, V8~\cite{kevin10},
+IonMonkey~\cite{ionmonkey}) are based on the GIL; Jython (an
+implementation of Python that is based on the JVM) is a notable
+exception.
+
+The standard model employed in these systems is fairly
+straightforward: there can be multiple threads, but each acquires the
+GIL, then performs a single instruction, and subsequently releases the
+lock.  While this setup limits the benefits due to parallelism %% not
+clear we must say so TRG % -- parallelism is limited to I/O or other
+activities external to the interpreter the use of a GIL also provides
+some crucial guarantees. Since this lock is always acquired while
+executing bytecode instructions and may be released only in-between
+such instructions, the GIL ensures perfect isolation between multiple
+threads for a series of instructions, which are executed atomically.
+In addition, the GIL provides the application with a sequential
+consistency model~\cite{lamport79}.
+
+A global interpreter lock assures the atomic execution of individual
+instructions but offers the programmer no help in managing concurrent
+data structures. Many parallel programs must maintain invariants
+involving multiple objects (or even multiple fields of a single
+object).  So to enable Python programs for parallel execution, we need
+a way to identify which blocks of data must be executed atomically and
+hardware or software support to effect this atomic execution.
+
+\emph{Atomic blocks}~\cite{tim03,tim05} allow dynamic languages to
+indicate the units of computation that must be executed atomically to
+ensure correct concurrent execution.  Atomic blocks map nicely to
+transactional memory, which provide a convenient mechansim to replace
+the GIL.  Using transactions to enclose multiple bytecode
+instructions, we can get the very same semantics as provided by the
+GIL while possibly executing several transactions in parallel. And by
+exposing these interpreter-level transactions to the application, we
+avoid the well-known problems caused by locks as applications use
+atomic blocks instead of explicit acquire/release operations on
+locks~\cite{christopher10,victor11,shan08}.
 
@@
 TM~\cite{nicholas06,odaira14,fuad10}.  We focus here on software based
-transactional memory (STM) systems,
-because they provide the unique opportunity to consider the issues of atomicity,
-synchronization, and memory management (resp. garbage collection) together.
-STM systems impose no a-priori bound on the size of an atomic block and provide therefore
-a flexibility that is absent from current hardware-based TM systems~\cite{wayforward14}.
-The downside of STM systems is runtime overhead~\cite{cascaval08,drago11}, which can
-be substantial despite recent efforts to lower the overhead~\cite{warmhoff13,spear09}.
-In this paper, we describe how we manage to lower the overhead of an
-STM system so that it can serve as a viable replacement for the GIL by leveraging the
-demands of a dynamic language like Python to integrate the STM system with the garbage
-collector.
+transactional memory (STM) systems, because they provide the unique
+opportunity to consider the issues of atomicity, synchronization, and
+memory management (resp. garbage collection) together.  STM systems
+impose no a-priori bound on the size of an atomic block and provide
+therefore a flexibility that is absent from current hardware-based TM
+systems~\cite{wayforward14}.  The downside of STM systems is runtime
+overhead~\cite{cascaval08,drago11}, which can be substantial despite
+recent efforts to lower the overhead~\cite{warmhoff13,spear09}.  In
+this paper, we describe how we manage to lower the overhead of an STM
+system so that it can serve as a viable replacement for the GIL by
+leveraging the demands of a dynamic language like Python to integrate
+the STM system with the garbage collector.
 

### 2014-06-06 talk/dls2014/paper/paper.tex
add a conclusion

@@
 TM~\cite{nicholas06,odaira14,fuad10}.  We focus here on software based
-transactional memory (STM) systems, because they provide the unique
-opportunity to consider the issues of atomicity, synchronization, and
+transactional memory (STM) because that provides the unique
+opportunity to consider the issues of atomicity, synchronisation, and
 memory management (resp. garbage collection) together.  STM systems
@@
 \item We introduce a new software transactional memory (STM) system
-  that performs well for Python programs.
+  that performs well for Python virtual machines.
 \item We integrate the  system closely with a garbage collector
@@
 
-First results promising; a lot to understand yet; a lot to improve
-
-contributions:
-\begin{itemize}
-\item memory segmentation, page mapping $\rightarrow$ performance
-\item STM \& GC: high allocation rate of dynlangs; shared write barrier;
-  page privatisation and re-sharing by GC; used to synchronise threads;
-  used to classify objects (transaction owned / not)
-\item GIL replaced successfully in embarrassingly parallel benchs;
-  avg overhead of $35\%$; beating GIL at 2 threads (2.5x speedup)
-\item atomic blocks introduced as a scalable sync mechanism (1.9x speedup)
-\item JIT needs work
-\end{itemize}
+% review of paper
+We set out to resolve the issue of parallelism in Python by replacing
+the GIL with TM. In particular we deemed a close integration of STM
+and the GC best as a way to attack the issues of atomicity,
+synchronisation, and memory management in a unified way. While the
+interplay of STM and the GC gives us some important performance
+benefits, we also implemented atomic blocks backed by STM for Python
+as a scalable synchronisation mechanism.
+
+% STM needs good performance:
+Since in our approach, STM becomes a vital part of the interpreter,
+good performance is essential. We found an interesting way to combine
+the existing memory segmentation and virtual memory features of
+current CPUs to lower the overhead of STM. We want to stress the
+importance of memory segmentation to our approach, because it seems
+rather unvalued. In the move from the x86 to the x86{-}64
+architecture, they already removed 4 out of 6 segment registers.  We
+want to make the point that there are still valid and interesting uses
+for them (see also \cite{bennet10}).
+
+% results, outlook
+The early results presented here are very encouraging. STM as a simple
+GIL replacement yields an average speedup of \remi{$2.5\times$} for
+embarrassingly parallel workloads on 4 threads.  When also used for
+synchronisation in the form of atomic blocks, the average speedup
+still reached \remi{$1.9\times$}.
+
+For the overall performance compared to our baseline interpreter
+CPython, integration with our JIT compiler is necessary. Once this
+integration matures, we believe our approach to not only serve as a
+simple GIL replacement, but also as a way forward in parallelising
+languages -- making atomicity and synchronisation the language's
+instead of the programmer's problem.
+
+% First results promising; a lot to understand yet; a lot to improve
+
+% contributions:
+% \begin{itemize}
+% \item memory segmentation, page mapping $\rightarrow$ performance
+% \item STM \& GC: high allocation rate of dynlangs; shared write barrier;
+%   page privatisation and re-sharing by GC; used to synchronise threads;
+%   used to classify objects (transaction owned / not)
+% \item GIL replaced successfully in embarrassingly parallel benchs;
+%   avg overhead of $35\%$; beating GIL at 2 threads (2.5x speedup)
+% \item atomic blocks introduced as a scalable sync mechanism (1.9x speedup)
+% \item JIT needs work
+% \end{itemize}
 
@@
 
+\bibitem{bennet10}
+  Bennet Yee, David Sehr, Gregory Dardyk, J. Bradley Chen, Robert
+  Muth, Tavis Ormandy, Shiki Okasaka, Neha Narula, and Nicholas
+  Fullagar. 2010. Native Client: a sandbox for portable, untrusted x86
+  native code. \emph{Commun. ACM} 53, 1 (January 2010), 91-99.
+
 % \bibitem{leis14}

### 2014-06-06 talk/dls2014/paper/paper.tex
introduction changes

@@
 threads. The use of a single global lock causes several disadvantages,
-and as multicore processors become the de-facto standard in platfroms
-ranging from smart phones to datacenters, the issue of parallel
+and as multi-core processors become the de facto standard in platforms
+ranging from smart phones to data centres, the issue of parallel
 execution must be addressed by these dynamic languages.
@@
 GIL, then performs a single instruction, and subsequently releases the
-lock.  While this setup limits the benefits due to parallelism %% not
-clear we must say so TRG % -- parallelism is limited to I/O or other
-activities external to the interpreter the use of a GIL also provides
+lock.  While this setup limits the benefits due to parallelism,
+the use of a GIL also provides
 some crucial guarantees. Since this lock is always acquired while
@@
 ensure correct concurrent execution.  Atomic blocks map nicely to
-transactional memory, which provide a convenient mechansim to replace
+transactional memory, which in turn provides a convenient mechanism to replace
 the GIL.  Using transactions to enclose multiple bytecode
@@
 overhead~\cite{cascaval08,drago11}, which can be substantial despite
-recent efforts to lower the overhead~\cite{warmhoff13,spear09}.  In
+recent efforts to lower it~\cite{warmhoff13,spear09}.  In
 this paper, we describe how we manage to lower the overhead of an STM
@@
 
-\cfbolz{both the introduction and the contributions need to stress the co-design of GC and STM}
-% agree trg
-
 This paper makes the following contributions:
 \begin{itemize}[noitemsep]
-\item We introduce a new software transactional memory (STM) system
-  that performs well for Python virtual machines.
+\item We introduce a new STM system that performs well for dynamic
+  language interpreters.
 \item We integrate the system closely with a garbage collector
@@
 \end{itemize}
-The STM described here is efficient and has low overhead and is therefore beneficial also
-for environments with a small number of threads or cores. The paper describes also
-%% can be dropped, not the core topic
-how to introduce atomic blocks into the Python language to provide a
-  backwards compatible, composable synchronization mechanism for
-  threads.
-
-%% What is really needed from this section?? Maybe the subsection on Python can be moved
-%%  to the intro -- the rest is questionable.  TRG
-
-\section{Background}
-
-% \subsection{Global Interpreter Lock}
-
-% The GIL is a very simple synchronisation mechanism for supporting
-% multithreading in an interpreter. The basic guarantee is that the GIL
-% may only be released in between bytecode instructions. Thus, these
-% instructions are always executed atomically and in complete isolation
-% from others running in other threads. \emph{Atomicity} means that each
-% instruction and its effects seem to happen at one, indivisible point
-% in time. Other instructions never see inconsistent state of a
-% partially executed instruction (\emph{isolation}).
-
-% In addition to these guarantees, instructions are executed in a
-% sequential consistency model~\cite{lamport79}. This means that
-% the outcome of any execution of instructions in multiple threads is
-% equal to \emph{some} sequential execution of them.
-
-% \subsection{Transactional Memory}
-
-% Transactional memory (TM) is a concurrency control mechanism that
-% comes from database systems. Using transactions, we can group a series
-% of instructions performing operations on memory and make them happen
-% atomically and in complete isolation from other transactions.
-% Atomicity and isolation are basic properties of transactions.
-
-% If we start multiple such transactions in multiple threads, the TM
-% system guarantees that the outcome of running the transactions is
-% \emph{serialisable}. Meaning, the outcome is equal to some sequential
-% execution of these transactions. By that, we can again provide a
-% sequentially consistent model for programming in multiple threads. We
-% can therefore use TM to directly replace the GIL. Instead of releasing
-% and acquiring the GIL between bytecode instructions, we commit and
-% start the transactions our instructions are running in.
-
-\subsection{Python}
+The STM described here is efficient and has low overhead and is
+therefore beneficial also for environments with a small number of
+threads or cores. It is also used to implement atomic blocks into the
+Python language to provide a backwards compatible, composable
+synchronisation mechanism for threads.
+
+\subsection{Application View}
 \remi{possibly adapt to application view \& benefits (atomic blocks)}
@@
 in mind. Its reference implementation, CPython~\cite{cpython}, uses a
-GIL to synchronise instructions in multiple threads.
-Over the years, Python added multiple ways to provide concurrency and
+GIL. Over the years, Python added multiple ways to provide concurrency and
 parallelism to its applications. We want to highlight two of them,
@@
 We focus on the \emph{threading} approach. This requires us to remove
-the GIL from our interpreter in order to run code in parallel on
+the GIL from the interpreter in order to run code in parallel on
 multiple threads. One approach to this is fine-grained locking instead
 of a single global lock. Jython~\cite{webjython} and
-IronPython~\cite{ironpython} are implementations of this. It requires
-great care in order to avoid deadlocks, which is why we follow the TM
-approach that provides a \emph{direct} replacement for the GIL. It
-does not require careful placing of locks in the right spots.
-
-% \subsection{Synchronisation}
-
-% In Python, since the GIL is not directly exposed to the interpreter,
-% applications still need to synchronise memory accesses from multiple
-% threads using locks. Locks can be very hard to get
-% right~\cite{christopher10,victor11,shan08}.  They are non-composable,
-% have overhead, may deadlock, limit scalability, and add to the overall
-% complexity of the program logic. We think that \emph{atomic
-% blocks}~\cite{tim03,tim05} provide a better way for synchronisation.
-
-% Atomic blocks are composable, deadlock-free, higher-level and expose
-% useful atomicity and isolation guarantees to the application for a
-% series of instructions. This is why we think that the introduction
-% of atomic blocks to Python is a valuable contribution. Since atomicity
-% is a property of transactions, TM and atomic blocks are a natural fit.
+IronPython~\cite{ironpython} follow this approach. Fine-grained
+locking is, however, not a \emph{direct} replacement for the GIL. It
+requires multiple locks in multiple places, not only in a central
+location. We follow the direct approach of using TM instead of the
+GIL.
 
@@
 
-% First results promising; a lot to understand yet; a lot to improve
-
-% contributions:
-% \begin{itemize}
-% \item memory segmentation, page mapping $\rightarrow$ performance
-% \item STM \& GC: high allocation rate of dynlangs; shared write barrier;
-%   page privatisation and re-sharing by GC; used to synchronise threads;
-%   used to classify objects (transaction owned / not)
-% \item GIL replaced successfully in embarrassingly parallel benchs;
-%   avg overhead of $35\%$; beating GIL at 2 threads (2.5x speedup)
-% \item atomic blocks introduced as a scalable sync mechanism (1.9x speedup)
-% \item JIT needs work
-% \end{itemize}
-
 %% \appendix

### 2014-06-06 talk/dls2014/paper/paper.tex
fix references

@@
   Rajwar. "Transactional memory." \emph{Synthesis Lectures on Computer
-  Architecture 5.1} (2010): 1-263.
-
-\bibitem{guerraoui08} Guerraoui, Rachid, and Michal Kapalka. "On the
-  correctness of transactional memory." \emph{Proceedings of the 13th
-    ACM SIGPLAN Symposium on Principles and practice of parallel
-    programming.} ACM, 2008.
+  Architecture 5.1} (2010)
+
+\bibitem{guerraoui08}
+  Rachid Guerraoui and Michal Kapalka. 2008. On the correctness of
+  transactional memory. In \emph{Proceedings of the 13th ACM SIGPLAN
+    Symposium on Principles and practice of parallel programming} (PPoPP
+  '08).
 
@@
   calls within transactions." (2006).
-\bibitem{spear08} Spear, Michael F., et al. "Implementing and
-  exploiting inevitability in software transactional memory."
-  \emph{Parallel Processing, 2008}. ICPP'08. 37th International
-  Conference on. IEEE, 2008.
+
+\bibitem{spear08} Michael F. Spear and Michael Silverman and Luke
+  Daless and Maged M. Michael and Michael L. Scott. 2008. "Implementing
+  and exploiting inevitability in software transactional memory."  In
+  \emph{Proc. 37th IEEE international conference on parallel
+    processing} (ICPP '08), 59-66
 
@@
 
-\bibitem{rajwar05} Rajwar, Ravi, Maurice Herlihy, and Konrad
-  Lai. "Virtualizing transactional memory." \emph{Computer
-    Architecture}, 2005. ISCA'05. Proceedings. 32nd International
-  Symposium on. IEEE, 2005.
-
-\bibitem{chung06} Chung, JaeWoong, et al. "Tradeoffs in transactional
-  memory virtualization." \emph{ACM SIGARCH Computer Architecture
-  News}. Vol. 34. No. 5. ACM, 2006.
+\bibitem{rajwar05} Ravi Rajwar, Maurice Herlihy, and Konrad
+  Lai. 2005. Virtualizing Transactional Memory. In \emph{Proceedings of
+    the 32nd annual international symposium on Computer Architecture}
+  (ISCA '05).
+
+\bibitem{chung06} JaeWoong Chung, Chi Cao Minh, Austen McDonald,
+  Travis Skare, Hassan Chafi, Brian D. Carlstrom, Christos Kozyrakis,
+  and Kunle Olukotun. 2006. Tradeoffs in transactional memory
+  virtualization. \emph{SIGOPS Oper. Syst. Rev.} 40, 5 (October 2006),
+  371-381.
 
@@
   Mehrara. 2009. Transactional memory with strong atomicity using
-  off-the-shelf memory protection hardware. SIGPLAN Not. 44, 4 (February
-  2009), 185-196.
+  off-the-shelf memory protection hardware. \emph{SIGPLAN Not.} 44, 4
+  (February 2009), 185-196.
 
@@
 \bibitem{odaira14}
-  Odaira, Rei, Jose G. Castanos, and Hisanobu Tomari.  "Eliminating
-  global interpreter locks in Ruby through hardware transactional
-  memory."  \emph{Proceedings of the 19th ACM SIGPLAN symposium on
-    Principles and practice of parallel programming.} ACM, 2014.
+  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. Eliminating
+  global interpreter locks in ruby through hardware transactional
+  memory. In \emph{Proceedings of the 19th ACM SIGPLAN symposium on
+    Principles and practice of parallel programming} (PPoPP '14).
 
 \bibitem{warmhoff13}
-  Wamhoff, Jons-Tobias, et al. "FastLane: improving performance of
-  software transactional memory for low thread counts."
-  \emph{Proceedings of the 18th ACM SIGPLAN symposium on Principles
-    and practice of parallel programming.} ACM, 2013.
-
-\bibitem{drago11}
-  Dragojević, Aleksandar, et al. "Why STM can be more than a research
-  toy." \emph{Communications of the ACM} 54.4 (2011): 70-77.
+  Jons-Tobias Wamhoff, Christof Fetzer, Pascal Felber, Etienne Rivière,
+  and Gilles Muller. 2013. FastLane: improving performance of software
+  transactional memory for low thread counts. \emph{SIGPLAN Not.} 48, 8
+  (February 2013), 113-122.
+
+\bibitem{drago11} Aleksandar Dragojević, Pascal Felber, Vincent
+  Gramoli, and Rachid Guerraoui. 2011. Why STM can be more than a
+  research toy. \emph{Commun. ACM} 54, 4 (April 2011), 70-77.
 
 \bibitem{cascaval08}
-  Cascaval, Calin, et al. "Software transactional memory: Why is it
-  only a research toy?." \emph{Queue} 6.5 (2008): 40.
+  Calin Cascaval, Colin Blundell, Maged Michael, Harold W. Cain, Peng
+  Wu, Stefanie Chiras, and Siddhartha Chatterjee. 2008. Software
+  transactional memory: why is it only a research
+  toy?. \emph{Commun. ACM} 51, 11 (November 2008), 40-46.
 
@@
     programming systems, languages, and applications} (OOPSLA
-  '06). ACM, New York, NY, USA
+  '06).
 
@@
   processor's hardware transactional memory support. \emph{SIGARCH
-  Comput. Archit. News 38}, 5 (April 2010)
+  Comput. Archit. News 38}, 5 (April 2010), 12-19.
 
 \bibitem{felber07}
-  Felber, Pascal, et al. "Transactifying applications using an open
-  compiler framework." \emph{TRANSACT}, August (2007): 4-6.
+  Pascal Felber and Torvald Riegel and Christof Fetzer and Martin
+  Süßkraut and Ulrich Müller and Heiko Sturzrehm. 2007. Transactifying
+  applications using an open compiler framework. \emph{TRANSACT}, August
+  (2007): 4-6.
 
@@
 \bibitem{spear09}
-  Spear, Michael F., et al. "Transactional mutex locks." \emph{SIGPLAN
-    Workshop on Transactional Computing.} 2009.
+  Luke Dalessandro, Dave Dice, Michael Scott, Nir Shavit, and Michael
+  Spear. 2010. Transactional mutex locks. In \emph{Proceedings of the
+    16th international Euro-Par conference on Parallel processing: Part
+    II} (Euro-Par'10), Pasqua D'Ambra, Mario Guarracino, and Domenico
+  Talia (Eds.). Springer-Verlag, Berlin, Heidelberg, 2-13.
 
 \bibitem{lamport79}
-  Lamport, Leslie. "How to make a multiprocessor computer that
-  correctly executes multiprocess programs." \emph{Computers, IEEE
-    Transactions} on 100.9 (1979): 690-691.
+  L. Lamport. 1979. How to Make a Multiprocessor Computer That Correctly
+  Executes Multiprocess Programs. \emph{IEEE Trans. Comput.} 28, 9
+  (September 1979), 690-691.
 
@@
     the twenty-third annual ACM symposium on Parallelism in algorithms
-    and architectures} (SPAA '11). ACM, New York, NY, USA
+    and architectures} (SPAA '11).
 
@@
     conference on Object-oriented programing, systems, languages, and
-    applications} (OOPSLA '03).
-
-\bibitem{tim05}
-  Tim Harris, Simon Marlow, Simon Peyton-Jones, and Maurice
-  Herlihy. 2005. Composable memory transactions. \emph{In Proceedings
-    of the tenth ACM SIGPLAN symposium on Principles and practice of
-    parallel programming} (PPoPP '05).
+    applications} (OOPSLA '03), 388-402.
+
+\bibitem{tim05} Tim Harris, Simon Marlow, Simon Peyton Jones, and
+  Maurice Herlihy. 2008. Composable memory transactions.
+  \emph{Commun. ACM} 51, 8 (August 2008), 91-100.
 

### 2014-06-06 talk/dls2014/paper/paper.tex
revised conclusions

@@
 \begin{abstract}
-....
+\remi{missing abstract}
 \end{abstract}
@@
 \keywords
-...
+\remi{missing keywords}
 
@@
 \subsection{Application View}
-\remi{possibly adapt to application view \& benefits (atomic blocks)}
-\remi{XXX: Also, make own section or integrate with introduction}
 
@@
 We focus on the \emph{threading} approach. This requires us to remove
-the GIL from the interpreter in order to run code in parallel on
+the GIL from the interpreter to run code in parallel on
 multiple threads. One approach to this is fine-grained locking instead
@@
 
-Our TM system is fully implemented in software. However, we do exploit
+Our TM system is fully implemented in software. However, we exploit
 some more advanced features of current CPUs, particularly \emph{memory
@@
 time. Transactions run in parallel because these threads can run in
-parallel. In order to isolate transactions from each other, we look
+parallel. To isolate transactions from each other, we look
 at isolating the memory accesses from threads.
@@
 As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
-we use a \emph{shadow stack}~\cite{fergus02} in order to provide precise garbage
+we use a \emph{shadow stack}~\cite{fergus02} to provide precise garbage
 collection.  Any time we call a function that possibly triggers a
@@
 reference to it that points to a young object. We then need to trace
-it during the next minor collection in order to mark the young object
+it during the next minor collection to mark the young object
 alive and to update its reference to the new location it gets moved
@@
 read markers array.  If we already own the lock, we are done.
-If someone else owns the lock, we will do a write-write contention
-management that will abort either us or the current owner of the
+If someone else owns the lock, we perform  write-write contention
+management that aborts either us or the current owner of the
 object.  If we succeed in acquiring the lock using an atomic
@@
 \lstinline!read_markers!  in other threads~/ segments. If we detect a
-read-write conflict, we do contention management to either abort us or
+read-write conflict, we perform contention management to either abort us or
 the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).
@@
 synchronisation fast and cheap, we do not want to insert an additional
-check regularly in order to see if synchronisation is requested. We
+check regularly to see if synchronisation is requested. We
 use a trick relying on the fact that dynamic languages are usually
@@
 \lstinline!nursery_current > nursery_end!  which will trigger the slow
-path of the function to possibly perform a minor collection in order
+path of the function to possibly perform a minor collection
 to free up space in the nursery.
@@
 \item prefer transactions that started earlier to younger transactions
-  in order to let long transactions succeed more often
+  to let long transactions succeed more often
 \item to support \emph{inevitable} transactions, we always prefer them
@@
 
-We use atomic blocks in order to easily synchronise access to shared
+We use atomic blocks to easily synchronise access to shared
 data structures in the first three benchmarks. For the embarrassingly
@@
 scalability. Here we make active use of the STM-supported atomic
-blocks in order do synchronise access to the data structure. The hope
+blocks to synchronise access to the data structure. The hope
 is that the STM system is still able to parallelise, even if we use
@@
 $10-50\times$. This is why we had to do without CPython here, since it
-would be much further up in the plots. Also, in order to make jitting
+would be much further up in the plots. Also, to make jitting
 code worthwhile, we increased the input size of all benchmarks to get
@@
 
-Overall PyPy needs the JIT in order for its performance to be
+Overall PyPy needs the JIT  for its performance to be
 competitive with CPython's. It would be interesting to see how using
-our STM system in CPython would turn out, but it is a lot of work. On
-its own, our system scales well, so we hope to also see that with the
+our STM system in CPython would turn out, but an in-depth evaluation is
+beyond the scope of this paper. On
+its own, our system scales well, so we expect to also see this trend in the presence of a
 JIT in the future.
@@
 % review of paper
-We set out to resolve the issue of parallelism in Python by replacing
-the GIL with TM. In particular we deemed a close integration of STM
-and the GC best as a way to attack the issues of atomicity,
-synchronisation, and memory management in a unified way. While the
-interplay of STM and the GC gives us some important performance
-benefits, we also implemented atomic blocks backed by STM for Python
-as a scalable synchronisation mechanism.
+As multi-core processors continue to increase the number of cores per
+processor, it is important that dynamic languages like Python embrace
+parallelism.  We report here an effort to replace the Global
+Interpreter Lock (GIL) used in many Python implementations with
+transactional memory (TM). The key factor for success is a close
+integration of a software TM (STM) and the garbage collector to
+provide atomicity, synchronisation, and memory management in a unified
+way.  The STM system supports atomic blocks as a user abstraction and
+thereby provides a scalable synchronisation mechanism to Python
+applications.
 
 % STM needs good performance:
-Since in our approach, STM becomes a vital part of the interpreter,
-good performance is essential. We found an interesting way to combine
-the existing memory segmentation and virtual memory features of
-current CPUs to lower the overhead of STM. We want to stress the
-importance of memory segmentation to our approach, because it seems
-rather unvalued. In the move from the x86 to the x86{-}64
-architecture, they already removed 4 out of 6 segment registers.  We
-want to make the point that there are still valid and interesting uses
-for them (see also \cite{bennet10}).
+The solution presented here is based on a software TM system.  As the
+STM is a vital part of the interpreter, good performance is
+essential. We present an interesting way to combine the existing
+memory segmentation and virtual memory features of current CPUs to
+lower the overhead of STM. This software implementation reduces
+dramatically the cost of transactional memory and at the same time
+provides an opportunity to integrate the STM system with garbage
+collection.  This implementation leverages the processor's memory
+segmentation hardware, which has lately played only a minor role in
+our operating system.  Unfortunately, processor architects seem to
+undervalue this hardware resource -- in the evolution from the x86 to
+the x86{-}64 architecture, 4 out of 6 segment registers have been
+removed. This development potentially threatens the STM strategy
+outlined here and we want to remind the architecture community that
+this unit can support interesting implementations -- a viewpoint that
+has also been expressed by other researchers (see also
+\cite{bennet10}).
 
@@
 synchronisation in the form of atomic blocks, the average speedup
-still reached \remi{$1.9\times$}.
-
-For the overall performance compared to our baseline interpreter
-CPython, integration with our JIT compiler is necessary. Once this
-integration matures, we believe our approach to not only serve as a
-simple GIL replacement, but also as a way forward in parallelising
-languages -- making atomicity and synchronisation the language's
-instead of the programmer's problem.
+still reaches \remi{$1.9\times$}.
+
+To obtain an overall performance that is competitive with the
+best-performing Python systems (Jython, PyPy), integration of the
+STM-based approach with a JIT compiler is necessary. Once this
+integration matures, the approach outlined here serves not only as a
+simple GIL replacement but also provides a way forward towards
+parallelizing Python.
 
@@
 \acks
-...
+\remi{cfbolz, felber, marlier, fäs, mastoras}
 
@@
 
-\bibitem{beazley10} Beazley, David. "Understanding the python gil."
+\bibitem{beazley10} Beazley, David. "Understanding the Python GIL."
   \emph{PyCON Python Conference}. Atlanta, Georgia. 2010.
@@
   global interpreter locks in ruby through hardware transactional
-  memory. In \emph{Proceedings of the 19th ACM SIGPLAN symposium on
-    Principles and practice of parallel programming} (PPoPP '14).
+  memory. In \emph{Proc. 19th ACM SIGPLAN Symposium on
+    Principles and Practice of Parallel Programming} (PPoPP '14).
 
@@
 \bibitem{fuad10}
-  Fuad Tabba. 2010. Adding concurrency in python using a commercial
+  Fuad Tabba. 2010. Adding concurrency in Python using a commercial
   processor's hardware transactional memory support. \emph{SIGARCH
@@
   Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. A study of
-  transactional memory vs. locks in practice. In \emph{Proceedings of
-    the twenty-third annual ACM symposium on Parallelism in algorithms
-    and architectures} (SPAA '11).
+  transactional memory vs. locks in practice. In \emph{Proc.
+     twenty-third annual ACM Symposium on Parallelism in Algorithms
+    and Architectures} (SPAA '11).
 
@@
   Witchel. 2010. Is transactional programming actually
-  easier?. \emph{SIGPLAN} Not. 45, 5 (January 2010), 47-56.
+  easier?. \emph{Proc. 15th ACM SIGPLAN Symposium on
+    Principles and Practice of Parallel Programming} (PPoPP '10).
 

### 2014-06-06 talk/dls2014/paper/paper.tex
acks

@@
 \keywords
-\remi{missing keywords}
+transactional memory, global interpreter lock, dynamic language,
+python, virtual machine
 
@@
 
-\cfbolz{you need to describe when and how to re-share pages}
-
 \subsection{Thread Synchronisation}
@@
 \acks
-\remi{cfbolz, felber, marlier, fäs, mastoras}
+We would like to thank the following people for their valuable
+inputs and the many fruitful discussions: Carl F. Bolz, Michael Faes,
+Pascal Felber, Patrick Marlier, and Aristeidis Mastoras.
 

### 2014-06-06 talk/dls2014/paper/paper.tex
updated benchmark numbers

@@
 thread. From this we see that there is additional overhead introduced
-by each thread (\remi{$12.3\%$} for all 4 threads together). While this
-is not ideal, we think that \remi{$12.3\%$} are acceptable on four
-threads. In terms of throughput, 4 threads have \remi{$3.54\times$}
+by each thread ($12.3\%$ for all 4 threads together). While this
+is not ideal, we think that $12.3\%$ are acceptable on four
+threads. In terms of throughput, 4 threads have $3.54\times$
 more iterations per second than a single thread.
@@
 benchmarks to a certain degree. It scales best for the embarrassingly
-parallel ones (\remi{$avg=2.5\times$} speedup) and a little less for
-the others (\remi{$avg=1.9\times$} speedup). The reason for that is
+parallel ones ($avg=2.6\times$ speedup) and a little less for
+the others ($avg=2.0\times$ speedup). The reason for that is
 that in the former group there are no real, logical conflicts -- all
@@
 Looking at the average overhead from switching from GIL to STM, we see
-that it is \remi{$\approx 35.5\%$}. The maximum in richards is
-\remi{$63\%$}. In all benchmarks \emph{pypy-stm-nojit} beats
+that it is $\approx 43.3\%$. The maximum in richards is
+$71\%$. In all benchmarks \emph{pypy-stm-nojit} beats
 \emph{pypy-nojit} already on two threads despite of this overhead.
@@
 JVM's JIT compiler, its performance in these benchmarks is behind
-PyPy's by a factor of \remi{$5-10\times$}. Because of that it does not
+PyPy's by a factor of $5-10\times$. Because of that it does not
 make sense to include it in this evaluation. In the future, we would
@@
 work required in this area. The slowdown factor for switching from GIL
-to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
+to STM ranges around $1-2.4\times$, and we beat GIL performance
 in half of the benchmarks.
@@
 our STM system in CPython would turn out, but an in-depth evaluation is
-beyond the scope of this paper. On
-its own, our system scales well, so we expect to also see this trend in the presence of a
+beyond the scope of this paper. On its own, our system scales well,
+so we expect to also see this trend in the presence of a
 JIT in the future.
@@
 The early results presented here are very encouraging. STM as a simple
-GIL replacement yields an average speedup of \remi{$2.5\times$} for
+GIL replacement yields an average speedup of $2.6\times$ for
 embarrassingly parallel workloads on 4 threads.  When also used for
 synchronisation in the form of atomic blocks, the average speedup
-still reaches \remi{$1.9\times$}.
+still reaches $2.0\times$.
 

### 2014-06-06 talk/dls2014/paper/plots/plot_performance_nojit.py

@@ -42,16 +42,16 @@ interps_styles = {
 benchs = {
     "raytrace (small)":{
         "pypy-stm-nojit":[
-            [8.3],
-            [4.33],
-            [3.74],
-            [3.08]
+            [8.3, 8.04,7.99,7.91,8.03,],
+            [4.33,4.21,4.28,4.23,4.24,],
+            [3.74,3.61,3.09,4.18,3.01,],
+            [3.08,2.89,2.75,3.06,2.98,]
         ],
         "cpython":[
-            [2.5],
-            [2.7],
-            [2.75],
-            [2.84]
+            [2.5,2.55,2.51,2.54,2.52,],
+            [2.7,2.67,2.67,2.69,2.71,],
+            [2.75,2.77,2.75,2.80,2.76,],
+            [2.84,2.83,2.85,2.86,2.83,]
         ],
         # "jython":[
         #     [2.95,2.95,2.96],
@@ -60,24 +60,24 @@ benchs = {
         #     [1.09,0.9,0.97,0.99,1.03]
         # ],
         "pypy-nojit":[
-            [5.5,5.7,5.8],
-            [7,6.97],
-            [6.68,6.77],
-            [6.4,6.4]
+            [5.41,5.36,5.34,5.31,5.38],
+            [6.66,6.63,6.61,6.51,6.60],
+            [6.34,6.29,6.22,6.32,6.31],
+            [5.91,5.72,5.88,5.87,5.78]
         ]},
 
     "btree (small)":{
         "pypy-stm-nojit":[
-            [8.3],
-            [4.9],
-            [3.9,4.3],
-            [4.4,4.0,4.3]
+            [8.21,8.10,8.19,8.13,8.14],
+            [4.60,4.64,4.61,5.21,4.65],
+            [3.61,4.48,5.02,5.17,3.55],
+            [4.01,4.07,4.68,4.07,3.76]
         ],
         "cpython":[
-            [1.93],
-            [5.76],
-            [5.91],
-            [6.03]
+            [1.93,1.93,1.88,1.90,1.95],
+            [5.76,5.78,5.78,5.71,5.79],
+            [5.91,5.66,5.66,5.60,5.68],
+            [6.03,5.98,6.01,6.03,5.97]
         ],
         # "jython":[
         #     [1.76,1.84],
@@ -86,24 +86,24 @@ benchs = {
         #     [2.57,2.52,2.48]
         # ],
[…]

### 2014-06-06 talk/dls2014/paper/paper.tex
add abstract

@@
 \begin{abstract}
-\remi{missing abstract}
+
+Since some years, the popularity of dynamic languages is on the rise.
+A common trait of many of these language implementations is the use of
+a single global interpreter lock (GIL) to synchronise the interpreter
+in a multithreading scenario. Since this lock effectively serialises
+the execution, they can not make use of the increasing parallelism in
+current hardware.
+
+In this paper, we present a software transactional memory (STM) system
+specifically designed for the purpose of replacing the GIL in dynamic
+language interpreters. The key idea is a close integration of STM with
+the garbage collector to provide atomicity, synchronisation, and
+memory management in a unified way.  As a user abstraction for a
+scalable synchronisation mechanism, we implement atomic blocks by
+exposing part of the STM system to applications.  The described STM
+uses a combination of features present in current CPUs to lower its
+overhead so that it can beat the performance of a GIL supported Python
+interpreter already on 2 threads.
+
 \end{abstract}
@@
 therefore beneficial also for environments with a small number of
-threads or cores. It is also used to implement atomic blocks into the
+threads or cores. It is also used to implement atomic blocks for the
 Python language to provide a backwards compatible, composable
@@
 simple GIL replacement but also provides a way forward towards
-parallelizing Python.
+parallelising Python.
 

### 2014-06-07 talk/dls2014/paper/paper.tex
updates

@@
 in a multithreading scenario. Since this lock effectively serialises
-the execution, they can not make use of the increasing parallelism in
+the execution, applications can not make use of the increasing parallelism in
 current hardware.
@@
 language interpreters. The key idea is a close integration of STM with
-the garbage collector to provide atomicity, synchronisation, and
+the garbage collector to approach atomicity, synchronisation, and
 memory management in a unified way.  As a user abstraction for a
@@
 exposing part of the STM system to applications.  The described STM
-uses a combination of features present in current CPUs to lower its
-overhead so that it can beat the performance of a GIL supported Python
-interpreter already on 2 threads.
+uses a combination of features present in current CPUs, so that the
+parallelisation benefits outweigh the system's overhead already on 2
+threads when comparing to a GIL supported Python interpreter.
 
@@
 therefore beneficial also for environments with a small number of
-threads or cores. It is also used to implement atomic blocks for the
-Python language to provide a backwards compatible, composable
-synchronisation mechanism for threads.
+threads or cores.
 
@@
 concurrency. It is, however, limited by the GIL and thus does not
-provide parallelism. At this point we should mention that it is indeed
+provide parallelism\footnote{At this point we should mention that it is indeed
 possible to run external functions written in C instead of Python in
 parallel. Our work focuses on Python itself and ignores this aspect as
-it requires writing in a different language.
+it requires writing in a different language.}. Because the GIL is only
+seen as an implementation detail of the interpreter, applications are
+required to coordinate concurrent accesses to data structures using
+conventional methods -- locks being the most common way.
 
@@
 of the interpreter itself and runs them in separate OS processes.
-Here we actually get parallelism because we have one GIL per
+Here we actually get parallelism because there is one GIL per
 interpreter, but of course we have the overhead of multiple processes~/
@@
 We focus on the \emph{threading} approach. This requires us to remove
-the GIL from the interpreter to run code in parallel on
-multiple threads. One approach to this is fine-grained locking instead
-of a single global lock. Jython~\cite{webjython} and
+the GIL from the interpreter to run code in parallel on multiple
+threads. One approach to this is fine-grained locking instead of a
+single global lock. Jython~\cite{webjython} and
 IronPython~\cite{ironpython} follow this approach. Fine-grained
@@
 requires multiple locks in multiple places, not only in a central
-location. We follow the direct approach of using TM instead of the
-GIL.
+location. We on the other side follow the direct approach of using TM
+instead of the GIL. There, GIL acquisition and release operations
+conceptually directly map to starting and committing transactions.
+
+TM also allows us to attack the issue of application-level
+synchronisation and atomicity. We introduce atomic blocks to the
+Python language by mapping them directly to the underlying
+transactions -- providing the application with the guarantee of
+atomicity for the enclosed instructions.
 
@@
 
-Our TM system is fully implemented in software. However, we exploit
+The described TM system is fully implemented in software. However, we exploit
 some more advanced features of current CPUs, particularly \emph{memory
@@
 it cannot be classified as a hybrid TM system since it currently
-makes no use of any HTM present in the CPU.
+makes no use of any hardware TM (HTM) present in the CPU.
 
@@
 most, segments.  They differ only in segments that have modified but
-not yet committed these objects.  So we use a trick of the Memory
+not yet committed these objects.  So we make use of the Memory
 Management Unit (MMU) of the CPU.  The MMU maps pages of virtual to
@@
 The result is that the virtual address space of the process is bloated
-by a factor $N$ -- but the real memory usage is not, because most
+by a factor $N$ -- but the real, physical memory usage is not, because most
 pages are shared with their identical copies in other segments.
@@
 some place in the physical memory. This makes the $SO$ a valid reference
-to an object in all segments.
+to an object in all segments, automatically resolving either to a
+shared or a private version.
 
@@
   concurrently modified memory: another thread could be issuing a read
-  barrier in parallel with a commit.  To prevent this from occurring, we
+  barrier in parallel to a commit.  To prevent this from occurring, we
   must suspend all other threads at known safe points.  This is a
@@
 Conflicts can be resolved by aborting one of the two transactions, or
-in some cases by pausing the writing one until the reader commits.  If a
+in some cases by pausing the writing one until the reader commits --
+serialising the write after the read.  If a
 commit succeeds, we copy the changes made to the objects in the write
@@
 
-In this section, we will provide details about the actual
+In this section, we will provide more details about the actual
 implementation of the system and discuss some of the issues that we
@@
 and write barriers. (II) It also includes a \emph{garbage collector
-(GC)} that is closely integrated with the TM part (e.g.\ it shares the
-write barrier).
+(GC)} that is closely integrated with the TM part.
 
@@
 void stm_commit_transaction()
+void stm_become_inevitable()
 void stm_read(object_t *SO)
@@
 \lstinline!stm_commit_transaction()! tries to commit the current
-transaction, aborting it otherwise. \lstinline!stm_read()!,
-\lstinline!stm_write()!  perform
+transaction, aborting it otherwise. Similarly,
+\lstinline!stm_become_inevitable()!  tries to make the current
+transaction inevitable, possibly aborting the current one if not
+possible.
+ \lstinline!stm_read()!, \lstinline!stm_write()!  perform
 a read or a write barrier on the object given as the argument, and
@@
 running. Otherwise these references may not be valid any more because
-the object moved or was seen as garbage. In this simplified API, only
-\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
-require saving object references.
+the object moved or was seen as garbage. In this simplified Application
+Programming Interface (API), only \lstinline!stm_allocate()!,
+\lstinline!stm_become_inevitable()! and
+\lstinline!stm_commit_transaction()!  require saving object
+references.
 
@@
 it relative to the $\%gs$ register.  With exceptions, nearly all
-accesses to objects managed by the TM system should use this type so
+accesses to objects managed by the TM system use this type so
 that the CPU will translate the reference to the right version of the
 object.
-
 \medskip   % why, Latex, why??
+
+\subsection{Replacing the GIL}
+
+With this simple API, replacing the GIL is just a matter of starting a
+new transaction when we would usually acquire the GIL, and committing
+the current transaction when we normally release the GIL.  It would be
+wasteful to do this between every single bytecode instruction.  We
+thus depend on a simple heuristic that estimates how expensive
+aborting and retrying the current transaction is based on how much
+memory was allocated in the current transaction. If we deem it
+expensive enough, we commit and start a new transaction \emph{as soon
+as possible}. Deciding on the length of a transaction is an
+interesting challenge in and of itself.
+
+It may not be possible to immediately break the transaction, as we
+could be executing an atomic block. Atomic blocks are currently
+implemented as guaranteeing that the whole block's execution takes
+place in a single transaction. Another issue is the support of
+irreversible operations, like calls to external functions or general
+input~/ output. If we are allowed to, we simply commit the transaction
+before and start a new one after these operations. Within an atomic
+block, however, we turn the transaction inevitable so that the
+transaction cannot abort and strong isolation is guaranteed.
+
 \subsection{Setup\label{sub:Setup}}

### 2014-06-07 talk/dls2014/paper/paper.tex
more updates

@@
 
-\newcommand{\mynote}[2]{%
-  \textcolor{red}{%
-    \fbox{\bfseries\sffamily\scriptsize#1}%
-    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
-  }%
-}
-
-\newcommand\remi[1]{\mynote{Remi}{#1}}
-\newcommand\cfbolz[1]{\mynote{cfbolz}{#1}}
-\newcommand\arigo[1]{\mynote{arigo}{#1}}
-
-\newcommand{\comment}[1]{}
+% Keine "Schusterjungen"
+\clubpenalty = 10000
+% Keine "Hurenkinder"
+\widowpenalty = 10000 \displaywidowpenalty = 10000
+
+% \newcommand{\mynote}[2]{%
+%   \textcolor{red}{%
+%     \fbox{\bfseries\sffamily\scriptsize#1}%
+%     {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
+%   }%
+% }
+
+% \newcommand\remi[1]{\mynote{Remi}{#1}}
+% \newcommand\cfbolz[1]{\mynote{cfbolz}{#1}}
+% \newcommand\arigo[1]{\mynote{arigo}{#1}}
+
+% \newcommand{\comment}[1]{}
 
@@
 
-\comment{
-A Platform for Parallelism in Dynamic languages
-Parallelism for Python
-A TM Implementation for Python
-Transactional Memory as a foundation for parallel Python
-Transactional Memory as a foundation for parallelism in Python
-Parallel Python based on VM-assisted TM
-Parallel Python based on VM-assisted Transactional Memory
-A Platform for parallel Python
-A TM Platform for parallel Python
-A Transactional Memory Platform for parallel Python
-A Transactional Memory system for parallel Python
-
-Virtual Memory Assisted Transactional Memory for Dynamic Languages}
+% \comment{
+% A Platform for Parallelism in Dynamic languages
+% Parallelism for Python
+% A TM Implementation for Python
+% Transactional Memory as a foundation for parallel Python
+% Transactional Memory as a foundation for parallelism in Python
+% Parallel Python based on VM-assisted TM
+% Parallel Python based on VM-assisted Transactional Memory
+% A Platform for parallel Python
+% A TM Platform for parallel Python
+% A Transactional Memory Platform for parallel Python
+% A Transactional Memory system for parallel Python
+
+% Virtual Memory Assisted Transactional Memory for Dynamic Languages}
 
@@
 This paper makes the following contributions:
-\begin{itemize}[noitemsep]
+\vspace{-\topsep}               % XXX:HACK:TODO
+\begin{itemize}
 \item We introduce a new STM system that performs well for dynamic
@@
 \end{itemize}
+
 The STM described here is efficient and has low overhead and is
@@
 
-\subsection{Architecture}
-
-Our TM system is designed as a library that covers all aspects around
-transactions and object management. It is designed for object-oriented
-dynamic language VMs as a replacement for the GIL.
-
-The library consists of two parts: (I) It provides a simple interface
-to starting and committing transactions, as well as the required read
-and write barriers. (II) It also includes a \emph{garbage collector
-(GC)} that is closely integrated with the TM part.
-
 \subsection{Application Programming Interface\label{sub:Application-Programming-Interfac}}
 
+\begin{code}[h]
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{Application programming interface\label{lst:api}}
+\end{code}
+
+Our TM system is designed as a library that covers all aspects around
+transactions and object memory management. It is designed for
+object-oriented dynamic language VMs as a replacement for the GIL.
+The library exposes the functions in Listing~\ref{lst:api} as its
+interface for use by a dynamic language interpreter.
 
@@
 of segments $>0$ map to the pages of the sharing-segment.
-
 However, the layout of a segment is not uniform and we actually
@@
 Figure~\ref{fig:Segment-Layout} and explained here:
-\begin{description}[noitemsep]
+
+\begin{description}
 \item [{NULL~page:}] This page is unmapped and will produce a
@@
 
+Note, these areas are at offsets inside the segments known at
+compile time. This makes some checks very efficient, e.g.\ checking
+if an object resides in the nursery only requires comparing its $SO$
+to the boundaries of the nursery.
+
 \begin{figure*}[t]
@@
 we simply assume a fair scheduling policy in the operating system when
-waiting on a condition variable.
+waiting on a lock.
 
@@
 unique to the current transaction. That way we can still detect in a
-medium-fast path inside barriers that the object still belongs to the
+medium-fast path inside the write barrier that the object still belongs to the
 current transaction. Other and later transactions will treat them
@@
 \lstinline!false!  on commit and only need to do this every 255
-transactions. The whole code for the barrier is easily optimisable for
-compilers as well as perfectly predictable for
-CPUs\footnote{Additional benefit: The read barrier is not constrained
-to execute before the actual read -- both the compiler and the CPU are
-free to reorder or group them.}:
-
+transactions. The whole code for the barrier shown in
+Listing~\ref{lst:rb} is easily optimisable for compilers as well as
+perfectly predictable for CPUs\footnote{Additional benefit: The read
+barrier is not constrained to execute before the actual read -- both
+the compiler and the CPU are free to reorder or group them.}.
+
+\begin{code}[h]
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{The complete read barrier\label{lst:rb}}
+\end{code}
 
@@
 
-The job of the write barrier is twofold: first, it serves as a write
-barrier for the garbage collector and second, it supports COW and adds
-objects to the write set of the transaction. These two jobs are
-depicted in Figure~\ref{fig:obj_states}.
+The job of the write barrier is twofold: it serves as a write barrier
+for the garbage collector (GC) and the STM system. These two jobs are
+depicted in Figure~\ref{fig:obj_states}. The GC needs to keep track of
+objects that may contain references to young, movable objects. These
+objects need to be traced during minor collections to update
+references to moved objects. Young objects automatically get traced if
+the GC ascertains that they survive the collection by being reachable
+from traced objects. The STM system needs to record objects in the
+write set, acquire write locks, and privatise their pages
+(copy-on-write). It only needs to do this for old objects from
+previous transactions, as the others are not known to other
+transactions anyway.
 
@@
 transaction, or if there was a minor collection which will add the
-flag again on all objects and may produce overflow objects. The flag
-is never set on freshly allocated objects, because they do not need
-to be traced by the GC or handled by the STM system. This fast path
-covers all cases for the GC and for the STM system.
-
+flag again on all objects, including new overflow objects. The flag
+is never set on young, freshly allocated objects. This fast path
+covers all cases for the GC and for the STM system (see
+Listing~\ref{lst:w_fast}).
+
+\begin{code}[h]
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{Write barrier fast path\label{lst:w_fast}}
+\end{code}
 
@@
 to the list of objects that need tracing in the next minor collection
-(\lstinline!to_trace!).  This is necessary in case we write a
-reference to it that points to a young object. We then need to trace
-it during the next minor collection to mark the young object
-alive and to update its reference to the new location it gets moved
-to.
+(\lstinline!to_trace!). Any reference we write to this object will
+be kept alive and up-to-date during minor collections.
 
@@
 object. We then try to acquire its write lock. \lstinline!write_locks!
-is a simple, \textbf{global} array of bytes that is indexed with the
+is a simple, \emph{global} array of bytes that is indexed with the
 $SO$ of the object divided by 16. Note, ``global'' here really means
@@
 write barrier once between collections, and only the old objects need
-to be handled by the STM part \textbf{once per transaction}. This keeps the
+to be handled by the STM part \emph{once per transaction}. This keeps the
 overhead of the write barrier very low.
@@
 This is done through the function \lstinline!stm_allocate!  shown
-below:
-
+in Listing~\ref{lst:alloc_func}.
+
+\begin{code}[h]
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{Function to allocate objects\label{lst:alloc_func}}
+\end{code}
 
@@
 the problem as well as we can. The general rules are:
-\begin{itemize}[noitemsep]
+
+\begin{itemize}
 \item prefer transactions that started earlier to younger transactions
-  to let long transactions succeed more often
+  to increase the chance of long transactions succeeding
 \item to support \emph{inevitable} transactions, we always prefer them
@@
 \end{itemize}
+
 We can either simply abort a transaction to let the other one succeed,
@@
 and provides a just-in-time (JIT) compiler to speed up applications
-running on top of it. For comparison, we compare between normal PyPy
+running on top of it. We compare between normal PyPy
 using a GIL and a PyPy with STM, as well as to CPython.
@@
 For each point in the plots, we took 5 measurements and report the
-average and the standard deviation as error bars. For the JIT
-benchmarks, we first let the JIT warm up by doing a few
-additional iterations to get a single measurement.
+arithmetic mean with the standard deviation as error bars. Average
+speedup numbers are calculated using the geometric mean. For the JIT
+benchmarks, we first let the JIT warm up by doing a few additional
+iterations to get a single measurement.
 
@@
 To asses how well the STM system scales on its own (without any real
-workload), we execute the following loop on 1 to 4 threads on our
-Python interpreter with STM:
+workload), we execute the loop in Listing~\ref{lst:scaling_workload}
+on 1 to 4 threads on the PyPy interpreter with STM.
+
+\begin{code}[h]
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{Dummy workload\label{lst:scaling_workload}}
+\end{code}
 
@@
 thread. From this we see that there is additional overhead introduced
-by each thread ($12.3\%$ for all 4 threads together). While this
-is not ideal, we think that $12.3\%$ are acceptable on four
+by each thread ($12.3\%$ for all 4 threads together). Every thread
+adds some overhead because during a commit, there is one more thread
+which has to reach a safe point. Additionally, conflict detection
+during commit needs to check in all concurrently running transactions.
+
+While not ideal, we think that $12.3\%$ is acceptable on four
 threads. In terms of throughput, 4 threads have $3.54\times$
@@
 Looking at the average overhead from switching from GIL to STM, we see
-that it is $\approx 43.3\%$. The maximum in richards is
-$71\%$. In all benchmarks \emph{pypy-stm-nojit} beats
-\emph{pypy-nojit} already on two threads despite of this overhead.
-However, it rarely beats CPython's single-thread performance.  For
+that it is $\approx 43.3\%$. The maximum in richards is $71\%$. In all
+benchmarks \emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on
+two threads despite of this overhead.  The achieved speedup comparing
+STM to the GIL is between $1.14\times$ and $1.94\times$.
+
+Still, STM rarely beats CPython's single-thread performance. However, for
 programs that need concurrency in CPython and that use threads to
@@
 
-We see that generally, the group of embarrassingly parallel benchmarks scales
-best. The other three scale barely or not at all with the number of
+We see that generally, the group of embarrassingly parallel benchmarks
+scales best. (There is a notable performance stability problem in the
+\emph{richards} benchmark. This is a bug we will fix in the future.)
+The other three benchmarks scale barely or not at all with the number of
 threads. The reason for this is likely again the conflicts in the
@@
 handlers to detect conflicts, we use traditional write barriers. There
-are only some superficial similarities between our approach and
-theirs.
+are only some superficial similarities between the approach presented
+here and theirs.
 
@@
 inputs and the many fruitful discussions: Carl F. Bolz, Michael Faes,
-Pascal Felber, Patrick Marlier, and Aristeidis Mastoras.
+Pascal Felber, Maciej Fijalkowski, Patrick Marlier, and Aristeidis Mastoras.
 
@@
 % The bibliography should be embedded for final submission.
-
-%%% TODO: fix all citation to follow common format (always page numbers; no 'et al', etc)
 

### 2014-06-08 talk/dls2014/paper/paper.tex
Results of a final pass through the paper

@@
 straightforward: there can be multiple threads, but each acquires the
-GIL, then performs a single instruction, and subsequently releases the
+GIL, then performs one (or several) instruction, and subsequently releases the
 lock.  While this setup limits the benefits due to parallelism,
@@
 hardware or software support to effect this atomic execution.
-
-\emph{Atomic blocks}~\cite{tim03,tim05} allow dynamic languages to
+This is traditionally done with locks.
+
+\emph{Atomic blocks}~\cite{tim03,tim05} would allow dynamic languages to
 indicate the units of computation that must be executed atomically to
@@
 This paper makes the following contributions:
-\vspace{-\topsep}               % XXX:HACK:TODO
+%\vspace{-\topsep}               % XXX:HACK:TODO
 \begin{itemize}
@@
 We implement and evaluate our system for the Python language. Python
-is a dynamic programming language that was designed with GIL semantics
-in mind. Its reference implementation, CPython~\cite{cpython}, uses a
+is a dynamic programming language that was originally designed without
+concurrency in mind.
+Its reference implementation, CPython~\cite{cpython}, uses a
 GIL. Over the years, Python added multiple ways to provide concurrency and
@@
 parallel. Our work focuses on Python itself and ignores this aspect as
-it requires writing in a different language.}. Because the GIL is only
-seen as an implementation detail of the interpreter, applications are
+it requires writing in a different language.}.  Moreover, the GIL can
+only ensure correctness of the interpreter itself: applications are
 required to coordinate concurrent accesses to data structures using
@@
 requires multiple locks in multiple places, not only in a central
-location. We on the other side follow the direct approach of using TM
+location. We on the other hand follow the direct approach of using TM
 instead of the GIL. There, GIL acquisition and release operations
@@
 
-In summary, translating an $\%gs:SO$ to a physical address is a
+In summary, translating a $\%gs:SO$ to a physical address is a
 two-step process: First the memory segmentation feature of the CPU
@@
 
-Our TM system is designed as a library that covers all aspects around
+Our TM system is designed as a C library that covers all aspects around
 transactions and object memory management. It is designed for
@@
 the GC ascertains that they survive the collection by being reachable
-from traced objects. The STM system needs to record objects in the
+from traced objects. So far, this is how generational GCs generally
+work. But the STM system also needs to record objects in the
 write set, acquire write locks, and privatise their pages
 (copy-on-write). It only needs to do this for old objects from
-previous transactions, as the others are not known to other
+previous transactions, as any other object is not known to other
 transactions anyway.
@@
 is set either if the object is old and comes from an earlier
-transaction, or if there was a minor collection which will add the
-flag again on all objects, including new overflow objects. The flag
+transaction, or if there was a minor collection (in this case, the flag
+is added again on all objects, including new overflow objects). The flag
 is never set on young, freshly allocated objects. This fast path
@@
 
-The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.
+The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.  For
+each object, this code is called at most once per minor collection.
 First comes the \emph{GC part}: In any case, the object will be added
@@
 management that aborts either us or the current owner of the
-object.  If we succeed in acquiring the lock using an atomic
+object.  If the lock is not owned so far, and if we succeed in
+acquiring it using an atomic
 \lstinline!cmp_and_swap!, we need to add the object to the write set
-(a simple list called \lstinline!modified_old_objects!)  and privatise
-all pages belonging to it (copy-on-write).
+(a simple list called \lstinline!modified_old_objects!).  We then
+privatise the page or pages where it resides (copy-on-write).
 
@@
 all threads so that the committing one is the only one running and all
-the others are waiting in safe points. We then go through the write
+the others are waiting in safe points.\footnote{This design choice is
+important to support our very cheap barriers. It is a trade-off that
+works best if the number of concurrent threads is small.
+See section~\ref{subsub:sync} for how we mitigate the costs.}
+We then go through the write
 set (\lstinline!modified_old_objects!)  and check the corresponding
@@
 After verifying that there are no conflicts anymore, we copy all our
-changes done to the objects in the write set to all other segments,
+changes (i.e.\ objects in our write set) to all other segments,
 including the sharing-segment. This is safe since we synchronised all
@@
 private pages. At that point we also get a new
-\lstinline!overflow_number! by increasing a global one, so that it
-stays globally unique for each transaction. Increasing the
+\lstinline!overflow_number! by increasing a global counter, so that it
+stays globally unique for each transaction. Increasing the (local)
 \lstinline!read_version!  is then done at the start of a new
@@
 
-\subsection{Thread Synchronisation}
+\subsection{Thread Synchronisation\label{subsub:sync}}
 
@@
 We evaluate our system in a Python interpreter called
-PyPy~\cite{pypy} (version 2.3). PyPy is an implementation of an
+PyPy~\cite{pypy}.
+\begin{description}
+\item[PyPy] (version 2.3) is an implementation of an
   interpreter for the Python language. It has a special focus on speed
@@
   using a GIL and a PyPy with STM, as well as to CPython.
-\begin{description}
 \item[CPython] (version 2.7.6) is the reference implementation of the Python
@@
 benchmarks, we first let the JIT warm up by doing a few additional
-iterations to get a single measurement.
+iterations before the actual measurement.
 
@@
 First we run our benchmarks on three different interpreters: CPython
-(GIL), and PyPy with STM and with the GIL (both without the JIT). The
+(GIL), PyPy with STM, and PyPy with the GIL (all without the JIT). The
 results are shown in Figure~\ref{fig:performance-nojit}.
@@
 threads. The reason for this is likely again the conflicts in the
-latter group. Since the JIT accelerates all code but not the STM
-overhead, we do more work per transaction. This increases the
+latter group. Right now, because the code runs much more quickly with
+the JIT than without, it has the effect of making the transactions
+logically longer.  This increases the
 likelihood of conflicts between them and therefore limits scalability
@@
 segmentation hardware, which has lately played only a minor role in
-our operating system.  Unfortunately, processor architects seem to
+our operating systems.  Unfortunately, processor architects seem to
 undervalue this hardware resource -- in the evolution from the x86 to

### 2014-06-08 talk/dls2014/paper/paper.tex
my last changes

@@
 Since some years, the popularity of dynamic languages is on the rise.
-A common trait of many of these language implementations is the use of
+A common trait of many of these language's implementations is the use of
 a single global interpreter lock (GIL) to synchronise the interpreter
@@
 specifically designed for the purpose of replacing the GIL in dynamic
-language interpreters. The key idea is a close integration of STM with
-the garbage collector to approach atomicity, synchronisation, and
-memory management in a unified way.  As a user abstraction for a
-scalable synchronisation mechanism, we implement atomic blocks by
-exposing part of the STM system to applications.  The described STM
-uses a combination of features present in current CPUs, so that the
-parallelisation benefits outweigh the system's overhead already on 2
-threads when comparing to a GIL supported Python interpreter.
+language interpreters while keeping the semantics. The key idea is a
+close integration of STM with the garbage collector to approach
+atomicity, synchronisation, and memory management in a unified
+way. Additionally, we implement atomic blocks as a scalable
+synchronisation mechanism by exposing part of this STM system to
+applications.
+
+The described system uses a combination of features present in current
+CPUs, so that the parallelisation benefits outweigh the system's
+overhead already on 2 threads when comparing two Python interpreters
+-- one with STM and one with a GIL.
 
@@
 Unfortunately the GIL plays a central role in current
-implementations. Even recent just-in-time compilers (JIT) for these
-languages (e.g., PyPy~\cite{cfbolz09}, V8~\cite{kevin10},
-IonMonkey~\cite{ionmonkey}) are based on the GIL; Jython (an
+implementations. Even recent efforts to address the performance problems
+of dynamic languages with just-in-time compilers (JIT)
+(e.g., PyPy~\cite{cfbolz09}, V8~\cite{kevin10},
+IonMonkey~\cite{ionmonkey}) keep the GIL; Jython (an
 implementation of Python that is based on the JVM) is a notable
@@
 such instructions, the GIL ensures perfect isolation between multiple
-threads for a series of instructions, which are executed atomically.
+threads for a series of instructions -- executing them atomically.
 In addition, the GIL provides the application with a sequential
@@
 data structures. Many parallel programs must maintain invariants
-involving multiple objects (or even multiple fields of a single
+involving multiple objects (or multiple fields of a single
 object).  So to enable Python programs for parallel execution, we need
@@
 
-\emph{Atomic blocks}~\cite{tim03,tim05} would allow dynamic languages to
+\emph{Atomic blocks}~\cite{tim03,tim05} allow languages to
 indicate the units of computation that must be executed atomically to
@@
   language interpreters.
+\vspace{3mm}               % XXX:HACK:TODO
 \item We integrate the system closely with a garbage collector
@@
 Python language by mapping them directly to the underlying
-transactions -- providing the application with the guarantee of
-atomicity for the enclosed instructions.
+transactions -- providing the application with the guarantees of
+atomicity and isolation for the enclosed instructions.
 
@@
 Management Unit (MMU) of the CPU.  The MMU maps pages of virtual to
-physical memory.  By changing this page mapping, at the coarse unit of
+physical memory.  By changing this page mapping at the coarse unit of
 one page of memory (4096 bytes), we can share memory that would
@@
 
-\subsection{Memory Segmentation on the x86}
+\subsection{Memory Segmentation on x86 CPUs}
 
@@
 program to register the objects in the read or write set before
-reading or writing to them\footnote{Barriers added by an automatic and
+reading or writing to them\footnote{Barriers added by an automatic, conservative, and
 local program transformation~\cite{felber07}}.  Furthermore, we still
@@
 \lstinline!stm_become_inevitable()!  tries to make the current
-transaction inevitable, possibly aborting the current one if not
-possible.
+transaction inevitable, waiting or aborting the current one if there
+is already an inevitable transaction.
  \lstinline!stm_read()!, \lstinline!stm_write()!  perform
@@
 \lstinline!stm_allocate()!  allocates a new object with the specified
-size.  \lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
+size.
+
+\lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
 push and pop objects onto the shadow stack\footnote{A stack for pointers
@@
   alive.~\cite{fergus02}} of the current thread.  References to objects
-have to be saved using this stack locally around calls that may cause a GC
+have to be saved using this stack locally, and conservatively around calls that may cause a GC
 cycle to happen, and also while there is no transaction
 running. Otherwise these references may not be valid any more because
-the object moved or was seen as garbage. In this simplified Application
+the object moved or was discarded as garbage. In this simplified Application
 Programming Interface (API), only \lstinline!stm_allocate()!,
@@
 implemented as guaranteeing that the whole block's execution takes
-place in a single transaction. Another issue is the support of
+place in a single transaction.
+
+Another issue is the support of
 irreversible operations, like calls to external functions or general
-input~/ output. If we are allowed to, we simply commit the transaction
-before and start a new one after these operations. Within an atomic
-block, however, we turn the transaction inevitable so that the
-transaction cannot abort and strong isolation is guaranteed.
+input~/ output. If we are allowed to, we commit the transaction
+before and start a new one after these operations\footnote{The language
+specification needs to specify if this is allowed.}. Within an atomic
+block, however, the transaction is unbreakable, and we need to turn
+the transaction inevitable so that it
+cannot abort and strong isolation is guaranteed.
 
@@
 On startup, we reserve a big range of virtual memory with a call to
-\lstinline!mmap()! and partition this space into $N+1$ segments. We
-want to run $N$ threads in parallel while segment~0 is designated as
+\lstinline!mmap()! and partition this space into $N+1$ segments.
+We want to run $N$ threads in parallel while segment~0 is designated as
 the \emph{sharing-segment} that is never assigned to a thread. The
-sharing-segment is essentially read-only, except during commit.
+sharing-segment is essentially read-only (except during commit) and
+holds all pages shared between the other segments.
 
@@
 
-Note, these areas are at offsets inside the segments known at
-compile time. This makes some checks very efficient, e.g.\ checking
+Note, since the above configuration is specified at compile time,
+all these areas are at offsets inside the segments known to the
+compiler. This makes some checks very efficient, e.g.\ checking
 if an object resides in the nursery only requires comparing its $SO$
-to the boundaries of the nursery.
+to the static boundaries of the nursery.
 
@@
 Pages should usually stay private for some time, but at some regular
-intervals, we still want to try to re-share pages in case they actually
+intervals, we still want to optimistically re-share pages in case they actually
 stay unmodified in the future. Major collections are allowed to do
@@
 The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.  For
-each object, this code is called at most once per minor collection.
+each object, this code is called at most once between two minor collections.
 First comes the \emph{GC part}: In any case, the object will be added
@@
 We evaluate our system in a Python interpreter called
-PyPy~\cite{pypy}.
+PyPy and compare to CPython.
 \begin{description}
@@
   running on top of it. We compare between normal PyPy
-  using a GIL and a PyPy with STM, as well as to CPython.
+  using a GIL and a PyPy with STM. All code is available at~\cite{pypy}.
 \item[CPython] (version 2.7.6) is the reference implementation of the Python
@@
 
-For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
-and normalised the average runtimes to the time it took on a single
+For the results in Figure~\ref{fig:scaling}, we
+normalised the average runtimes to the time it took on a single
 thread. From this we see that there is additional overhead introduced
@@
 which has to reach a safe point. Additionally, conflict detection
-during commit needs to check in all concurrently running transactions.
+needs to check for conflicts in all concurrently running transactions.
 
@@
 First we run our benchmarks on three different interpreters: CPython
-(GIL), PyPy with STM, and PyPy with the GIL (all without the JIT). The
+(GIL), PyPy with STM, and PyPy with the GIL (all without a JIT). The
 results are shown in Figure~\ref{fig:performance-nojit}.
@@
 parallel ones ($avg=2.6\times$ speedup) and a little less for
-the others ($avg=2.0\times$ speedup). The reason for that is
+the others ($avg=2.0\times$ speedup). The reason for this difference is
 that in the former group there are no real, logical conflicts -- all
@@
 
-Since PyPy comes with a JIT to make its overhead compared to CPython
+Since PyPy comes with a JIT~\cite{cfbolz09} to make its overhead compared to CPython
 go away, we will now look at how well STM works together with it.
@@
 things in each run because of the non-deterministic thread-scheduling
-of the operating system (OS). Second, it is able to remove some
-allocations in some cases. Because compilation is already
-non-deterministic, so is this allocation-removal. And third, we did
+of the operating system (OS). Second, it is able to perform some
+advanced optimisations. Because compilation is already non-deterministic,
+so are these optimisations. And third, we did
 not have enough time to optimise integration with STM so that the JIT
@@
 features arbitrarily long transactions, which is why this assumption
-first has to be verified in our case.
+first has to be verified again in this setting.
 
@@
 collection.  This implementation leverages the processor's memory
-segmentation hardware, which has lately played only a minor role in
-our operating systems.  Unfortunately, processor architects seem to
+segmentation hardware, which has lately not received much attention.
+Unfortunately, processor architects seem to
 undervalue this hardware resource -- in the evolution from the x86 to
@@
 The early results presented here are very encouraging. STM as a simple
-GIL replacement yields an average speedup of $2.6\times$ for
+GIL replacement scales well and yields an average speedup of $2.6\times$ for
 embarrassingly parallel workloads on 4 threads.  When also used for
@@
 simple GIL replacement but also provides a way forward towards
-parallelising Python.
+a parallel programming model for Python.
 
@@
 We would like to thank the following people for their valuable
-inputs and the many fruitful discussions: Carl F. Bolz, Michael Faes,
+input and the many fruitful discussions: Carl F. Bolz, Michael Faes,
 Pascal Felber, Maciej Fijalkowski, Patrick Marlier, and Aristeidis Mastoras.

### 2014-06-16 talk/icooolps2014/position-paper.tex
fix some references

@@
 \title{A Way Forward in Parallelising Dynamic Languages}
-\subtitle{Position Paper, ICOOOLPS'14}
+%\subtitle{Position Paper, ICOOOLPS'14}
 
 \authorinfo{Remigius Meier}
-           {Department of Computer Science\\ ETH Zürich}
+           {Department of Computer Science\\ ETH Zürich, Switzerland}
            {remi.meier@inf.ethz.ch}
@@
 \bibitem{warmhoff13}
-  Wamhoff, Jons-Tobias, et al. "FastLane: improving performance of
-  software transactional memory for low thread counts."
-  \emph{Proceedings of the 18th ACM SIGPLAN symposium on Principles
-    and practice of parallel programming.} ACM, 2013.
+  Jons-Tobias Wamhoff, Christof Fetzer, Pascal Felber, Etienne Rivière,
+  and Gilles Muller. 2013. FastLane: improving performance of software
+  transactional memory for low thread counts. \emph{SIGPLAN Not.} 48, 8
+  (February 2013), 113-122.
 
-\bibitem{drago11}
-  Dragojević, Aleksandar, et al. "Why STM can be more than a research
-  toy." \emph{Communications of the ACM} 54.4 (2011): 70-77.
+\bibitem{drago11} Aleksandar Dragojević, Pascal Felber, Vincent
+  Gramoli, and Rachid Guerraoui. 2011. Why STM can be more than a
+  research toy. \emph{Commun. ACM} 54, 4 (April 2011), 70-77.
 
 \bibitem{cascaval08}
-  Cascaval, Calin, et al. "Software transactional memory: Why is it
-  only a research toy?." \emph{Queue} 6.5 (2008): 40.
+  Calin Cascaval, Colin Blundell, Maged Michael, Harold W. Cain, Peng
+  Wu, Stefanie Chiras, and Siddhartha Chatterjee. 2008. Software
+  transactional memory: why is it only a research
+  toy?. \emph{Commun. ACM} 51, 11 (November 2008), 40-46.
 
@@
 \bibitem{felber07}
-  Felber, Pascal, et al. "Transactifying applications using an open
-  compiler framework." \emph{TRANSACT}, August (2007): 4-6.
+  Pascal Felber and Torvald Riegel and Christof Fetzer and Martin
+  Süßkraut and Ulrich Müller and Heiko Sturzrehm. 2007. Transactifying
+  applications using an open compiler framework. \emph{TRANSACT}, August
+  (2007): 4-6.
 
@@
 \bibitem{spear09}
-  Spear, Michael F., et al. "Transactional mutex locks." \emph{SIGPLAN
-    Workshop on Transactional Computing.} 2009.
+  Luke Dalessandro, Dave Dice, Michael Scott, Nir Shavit, and Michael
+  Spear. 2010. Transactional mutex locks. In \emph{Proceedings of the
+    16th international Euro-Par conference on Parallel processing: Part
+    II} (Euro-Par'10), Pasqua D'Ambra, Mario Guarracino, and Domenico
+  Talia (Eds.). Springer-Verlag, Berlin, Heidelberg, 2-13.
 

### 2014-06-16 talk/dls2014/paper/paper.tex
fix some references

@@
 \authorinfo{Remigius Meier}
-           {Department of Computer Science\\ ETH Zürich}
+           {Department of Computer Science\\ ETH Zürich, Switzerland}
            {remi.meier@inf.ethz.ch}
@@
 \authorinfo{Thomas Gross}
-           {Department of Computer Science\\ ETH Zürich}
+           {Department of Computer Science\\ ETH Zürich, Switzerland}
            {thomas.gross@inf.ethz.ch}

### 2014-06-17 talk/icooolps2014/position-paper.tex
add copyright data
first round of integrating feedback

@@
 \usepackage{amssymb}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% listings
+\usepackage{float}
+\floatstyle{ruled}
+\newfloat{code}{tbp}{loa}
+\providecommand{\codename}{Listing}
+\floatname{code}{\protect\codename}
+
+% nice listings
+\usepackage{xcolor}
+\usepackage{newverbs}
+
+\usepackage{color}
+\definecolor{verylightgray}{rgb}{0.93,0.93,0.93}
+\definecolor{darkblue}{rgb}{0.2,0.2,0.6}
+\definecolor{commentgreen}{rgb}{0.25,0.5,0.37}
+\usepackage{letltxmacro}
+
+\usepackage{listings}
+
+\makeatletter
+\LetLtxMacro{\oldlstinline}{\lstinline}
+
+\renewcommand\lstinline[1][]{%
+  \Collectverb{\@@myverb}%
+}
+
+\def\@@myverb#1{%
+    \begingroup
+    \fboxsep=0.2em
+    \colorbox{verylightgray}{\oldlstinline|#1|}%
+    \endgroup
+}
+\makeatother
+
+\lstset{backgroundcolor={\color{verylightgray}},
+  basicstyle={\scriptsize\ttfamily},
+  commentstyle={\ttfamily\color{commentgreen}},
+  keywordstyle={\bfseries\color{darkblue}},
+  morecomment={[l]{//}},
+  tabsize=4,
+  morekeywords={foreach,in,def,type,dynamic,Int,
+    Boolean,infer,void,super,if,boolean,int,else,
+    while,do,extends,class,assert,for,switch,case,
+    private,protected,public,const,final,static,
+    interface,new,true,false,null,return}}
+\renewcommand{\lstlistingname}{Listing}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
 
@@
 
-\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
+\conferenceinfo{ICOOOLPS'14}{July 28 2014, Uppsala, Sweden}
 \copyrightyear{2014}
-%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
-\doi{nnnnnnn.nnnnnnn}
+\copyrightdata{978-1-4503-2914-9/14/07}
+\doi{2633301.2633305}
 
@@
 These requirements are not easy to meet. We argue that STM is the
-overall winner. While it has a big performance problem currently, it
+overall winner. While it currently has a big performance problem, it
 gets more points in all the other categories. We think that it is the
@@
 multithreading in an interpreter. The basic guarantee is that the GIL
-may only be released in between bytecode instructions. The interpreter
-can thus rely on complete isolation and atomicity of these
-instructions. Additionally, it provides the application with a
-sequential consistency model~\cite{lamport79}. As a consequence,
-applications can rely on certain operations to be atomic and that they
-will always be executed in the order in which they appear in the
-code. While depending on this may not always be a good idea, it is
-done in practice. A GIL-replacement should therefore uphold these
-guarantees, while preferably also be as easily implementable as a GIL
-for the interpreter. The latter can be especially important since
-many of these languages are developed and maintained by very large
-open-source communities, which are not easy to coordinate.
+may only be released in between bytecode instructions\footnote{This
+also applies to Abstract Syntax Tree (AST) interpreters, where the GIL
+may only be released between interpreting two AST nodes.}. The interpreter
+can thus rely on complete isolation and atomicity for the
+instructions' execution. Thus, accesses to data structures like
+dictionaries and lists happen atomically and do not need additional
+protection from data races when shared between threads.
+
+The GIL also provides the application with a sequential consistency
+model~\cite{lamport79}. This can be very valuable as it means less
+surprises for the programmer. For example in Table~\ref{tab:seq_cons},
+the programmer can expect the critical section to only be entered by
+one thread.  If the model allowed to buffer the writes, both threads
+may enter the critical section at the same time.
+
+\begin{table}[!ht]
+  \begin{center}
+    \begin{tabular}{|l|l|}
+      \hline
+      Thread 1 & Thread 2 \\
+      \hline
+      \multicolumn{2}{|l|}{\texttt{A = B = 0}} \\
+      \hline
+      \texttt{A = 1} & \texttt{B = 1}\\
+      \texttt{if B == 0:} & \texttt{if A == 0:}\\
+      \texttt{    critical section} & \texttt{    critical section}\\
+      \hline
+    \end{tabular}
+    \caption{Critical section with a sequential consistency model.}
+    \label{tab:seq_cons}
+  \end{center}
+\end{table}
+
+As a consequence, applications can rely on certain operations to be
+atomic and that they will always be executed in the order in which
+they appear in the code. While depending on this may not always be a
+good idea, it is done in practice. A GIL-replacement should therefore
+uphold these guarantees, while preferably also be as easily
+implementable as a GIL for the interpreter. The latter can be
+especially important since many of these languages are developed and
+maintained by very large open-source communities, which are not easy
+to coordinate.
 
@@
 \begin{description}
-\item[Performance:] How well does the approach perform compared to the
-  GIL on a single and on multiple threads?
+\item[Performance:] How much does the approach impact performance on a single
+  and how much on multiple threads? Can it make use of parallelism?
 \item[Existing applications:] How big are the changes required to
   integrate with and parallelise existing applications?
-\item[Better synchronisation:] Does the approach enable better, paralleliseable
-  synchronisation mechanisms for applications (e.g. atomic blocks)?
+\item[Better synchronisation:] Does the approach enable better,
+  paralleliseable synchronisation mechanisms for applications
+  (e.g.\ atomic blocks)?  Many synchronisation mechanisms can be built on
+  top of all solutions (e.g.\  message passing). We look for mechanisms
+  that are directly enabled by the contending approaches.
 \item[Implementation:] How difficult is it to implement the approach
@@
 interpreter. For a language like Python, one needs quite a few,
-carefully placed locks. Since there is no central location, the
+carefully placed locks -- every dictionary, list, instance, or mutable
+object in general needs a lock. Compared to e.g.\ Java, object
+attributes are backed by a dictionary. Accesses to it must be
+synchronised because the interpreter could crash otherwise.  Since
+there is no central location for all these locks, the
 complexity of the implementation is quite a bit larger compared to
@@
 changes\footnote{There are rare cases where not having atomic
-  bytecodes actually changes the semantics.}. An implementation has to
-follow the GIL semantics very closely, otherwise it may expose some
-latent data races in existing applications which are just not exposed
-with a GIL. This approach does however not provide a better parallelising
-synchronisation mechanism to the application like e.g. atomic blocks.
+  bytecodes actually changes the semantics. E.g.\ in Jython,
+  \texttt{dict1.update(dict2)} is not atomic: it first reads data from
+  \texttt{dict2} with \texttt{dict2}'s lock, and then puts it into
+  \texttt{dict1} with \texttt{dict1}'s lock. A lot can happen
+  in-between.}. An implementation has to follow the GIL semantics very
+closely, otherwise it may expose some latent data races in existing
+applications which are just not exposed with a GIL. This approach does
+however not provide a better parallelising synchronisation mechanism
+to the application like e.g. atomic blocks.
 

### 2014-06-17 talk/icooolps2014/position-paper.tex
second round of feedback

@@
       \hline
-      \texttt{A = 1} & \texttt{B = 1}\\
-      \texttt{if B == 0:} & \texttt{if A == 0:}\\
-      \texttt{    critical section} & \texttt{    critical section}\\
+      \texttt{A = 1}                & \texttt{B = 1}                \\
+      \texttt{if B == 0:}           & \texttt{if A == 0:}           \\
+      \texttt{    critical section} & \texttt{    critical section} \\
       \hline
@@
 synchronisation mechanism called \emph{atomic
-  blocks}~\cite{tim03,tim05}.
+  blocks}~\cite{tim03,tim05}. This is also suggested by
+\cite{christopher10,victor11} as an easier mechanism than locks.
 
@@
     \hline
-    Better synchronisation        & o    & o  &      & o  & ++   \\
+    Better synchronisation        & o    & o  & +    & o  & ++   \\
     \hline
@@
 \subsection{Potential Solutions}
+\label{sec:pot_solutions}
 
@@
 however not provide a better parallelising synchronisation mechanism
-to the application like e.g. atomic blocks.
-
-%% - support of atomic blocks?\\
-%% - hard to get right (deadlocks, performance, lock-granularity)\\
-%% - very hard to get right for a large language\\
-%% - hard to retro-fit, as all existing code assumes GIL semantics\\
-%% - (there are some semantic differences, right? not given perfect lock-placement, but well)
-%% ( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )
+to the application like e.g.\ atomic blocks.
 
@@
 independent processes; some web servers and some numeric computations
-do. We will consider here a slightly more general approach: the
-\emph{multiprocessing}\footnote{https://docs.python.org/2/library/multiprocessing.html}
+do.
+
+We will consider here a slightly more general approach: the
+\emph{multiprocessing}~\cite{multiprocessing}
 module of Python. In essence, it uses process-forking to provide the
 application with multiple interpreters that can run in parallel.
-Communication is then done explicitly through pipes.
+Communication is then done explicitly through pipes.\footnote{There
+are multiple alternative designs like e.g.\ actors or tuple spaces.
+Since they are similar and do not replace the GIL directly, we
+focus on the example of \emph{multiprocessing}.}
 
@@
 easy integration of external libraries without any changes.
-
-%% - often needs major restructuring of programs (explicit data exchange)\\
-%% - sometimes communication overhead is too large\\
-%% - shared memory is a problem, copies of memory are too expensive
 
@@
 Following the above argumentation for each approach, we assembled a
-general overview in Table \ref{tab:comparison}. The general picture is
-everything else than clear. It looks like HTM may be a good solution
-to replace the GIL in the near future. Current implementations are
-however far too limiting and do not provide good scaling.
+general overview in Table \ref{tab:comparison}. The points were assigned
+according to the criteria described in \ref{sec:pot_solutions}. Since
+the criteria are defined intuitively, there are no formal justifications
+for the assigned points. The reader is thus advised to take them with a
+grain of salt.
+
+The general picture is everything else than clear. It looks like HTM
+may be a good solution to replace the GIL in the near future. Current
+implementations are however far too limiting, not widely available,
+and do not provide good scaling.
 
@@
 parallelism. This is why we think that only the STM approach is a
-viable solution in the long-term. It provides the application with a
-simple memory model (sequential consistency) and a composable way to
-synchronise memory accesses using atomic blocks.
+viable solution in the long-term. It unifies both, the simple memory
+model (sequential consistency) and the synchronisation of memory accesses
+using composable atomic blocks. It is not \emph{only} a simple GIL
+replacement.
 
@@
 
-We believe that further work to reduce the overhead of STM is
+We think that further work to reduce the overhead of STM is
 very worthwhile. In fact, considering some analogies that have been
 drawn between garbage collection and transactional memory~\cite{dan07},
-we believe that it is worthwhile to focus the STM research more
+we think that it is worthwhile to focus the STM research more
 specifically onto the context shown in this paper --- for use in
@@
   The Jython Project, \url{www.jython.org}
+
+\bibitem{multiprocessing}
+  The Multiprocessing Module of Python,
+  \url{docs.python.org/2/library/multiprocessing.html}
 

### 2014-06-17 talk/icooolps2014/position-paper.tex
some adjustments

@@
   choice to use a single global interpreter lock (GIL) to synchronise
-  the interpreter in a multithreading scenario. This choice however
+  the interpreter in a multithreading scenario. This choice, however,
   makes it impossible to actually run code in parallel.
@@
 languages like Python and Ruby, their ability to make use of multiple
-cores is somewhat limited. For ease of implementation they chose to
+cores is somewhat limited. For ease of implementation, they chose to
 use a single, global interpreter lock (GIL) to synchronise the
@@
 
-These requirements are not easy to meet. We argue that STM is the
-overall winner. While it currently has a big performance problem, it
-gets more points in all the other categories. We think that it is the
-only solution that also provides a better synchronisation mechanism to
-the application in the form of parallelisable atomic blocks.
-
-%% \subsection{Issue}
-%% The issue that we want to discuss is how to efficiently support
-%% multi-core parallel execution of code in dynamic languages that were
-%% designed with GIL semantics in mind.
-
-%% Furthermore, a solution to this problem should also bring better
-%% synchronisation mechanism with it...
-
-%% (supporting (large) atomic blocks for synchronisation)
-
-%% \subsection{Our Position}
-%% Current solutions for replacing the GIL include STM, HTM, and
-%% fine-grained locking. STM is usually too slow, HTM very limited, and
-%% locking suffers from complexity that makes it hard to implement
-%% correctly. We argue that the best way forward is still STM and that
-%% its performance problem can be solved.
-
-%% Current solutions like STM, HTM, and fine-grained locking are slow, hard
-%% to implement correctly, and don't fit the specific problems of dynamic
-%% language.  STM is the best way forward but has bad performance, so we
-%% fix that.
+These requirements are not easy to meet. The author's position is that
+STM provides the best way forward. While STM currently has a big
+performance problem, it gets more points in the other categories. We
+think that it is the only solution that also provides a better
+synchronisation mechanism to the application in the form of
+parallelisable atomic blocks.  In the following section, we try to
+present a balanced view of the compared approaches.
 
@@
 
-In this section we examine the approaches and highlight their
-advantages and disadvantages.
-%% \paragraph{dynamic language VM problems}
-%% XXX:
-%% - high allocation rate (short lived objects)\\
-%% - (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)
+In this section we first explain the motivation for using a GIL and
+then examine different approaches to remove or avoid it -- highlighting
+their advantages and disadvantages.
 
@@
 can thus rely on complete isolation and atomicity for the
-instructions' execution. Thus, accesses to data structures like
+instructions' execution. Also, accesses to data structures like
 dictionaries and lists happen atomically and do not need additional
@@
 the programmer can expect the critical section to only be entered by
-one thread.  If the model allowed to buffer the writes, both threads
-may enter the critical section at the same time.
+one thread.  On the other hand, if the model allowed to buffer the
+writes, both threads may enter the critical section at the same time.
 
@@
       \hline
-      \multicolumn{2}{|l|}{\texttt{A = B = 0}}                      \\
+      \multicolumn{2}{|c|}{\texttt{A = B = 0}}                      \\
       \hline
@@
       \texttt{if B == 0:}           & \texttt{if A == 0:}           \\
-      \texttt{    critical section} & \texttt{    critical section} \\
+      \multicolumn{2}{|c|}{only one thread enters here}             \\
+      \multicolumn{2}{|c|}{(e.g.\ critical section)}             \\
       \hline
@@
 argumentation more on running code in the interpreted language in
-parallel, not the external C calls.
+parallel, not the external C code.
 
@@
 execute in parallel or even concurrently.  Potential solutions are
-preferable if they provide a good way to implement atomic blocks that
+preferable if they provide a good way to implement atomic blocks
+(or another, comparable synchronisation mechanism) that
 are also able to be executed in parallel.
@@
 
-For the discussion we define a set of criteria to evaluate the
-multiple potential solutions for removing or avoiding the GIL and its
+For the discussion, we define a set of criteria to evaluate the
+potential solutions for removing or avoiding the GIL and its
 limitations:
@@
 potential parallelism in order to keep the complexity manageable.
-The latter means that we lose a bit of performance in the
-single-threaded case compared to the GIL, which requires much less
-acquire-release operations.
+The latter means that we lose a bit of performance compared to the
+GIL, which requires much less acquire-release operations.
 
@@
 that uses fine-grained locking\footnote{The performance impact of
-fine-grained locking is milder in Java than it would be in a typical piece
+fine-grained locking is milder on the JVM than it would be in a typical piece
 of C code; see e.g.~\cite{biased}.} to correctly synchronise the
@@
 using a GIL. Integrating external, non-thread-safe libraries should
-however be very simple too. One could simply use one lock per library
+however be very simple too. One can simply use one lock per library
 to avoid this issue.
@@
 therefore parallelise existing applications, generally without any
-changes\footnote{There are rare cases where not having atomic
-  bytecodes actually changes the semantics. E.g.\ in Jython,
-  \texttt{dict1.update(dict2)} is not atomic: it first reads data from
-  \texttt{dict2} with \texttt{dict2}'s lock, and then puts it into
-  \texttt{dict1} with \texttt{dict1}'s lock. A lot can happen
-  in-between.}. An implementation has to follow the GIL semantics very
+changes. An implementation has to follow the GIL semantics very
 closely, otherwise it may expose some latent data races in existing
-applications which are just not exposed with a GIL. This approach does
-however not provide a better parallelising synchronisation mechanism
-to the application like e.g.\ atomic blocks.
+applications which are just not exposed with a GIL\footnote{There are
+  rare cases where not having atomic bytecodes actually changes the
+  semantics. E.g.\ in Jython, \texttt{dict1.update(dict2)} is not
+  atomic: it first reads data from \texttt{dict2} with \texttt{dict2}'s
+  lock, and then puts it into \texttt{dict1} with \texttt{dict1}'s
+  lock. A lot can happen in-between.}. This approach does however not
+provide a better parallelising synchronisation mechanism to the
+application and still requires explicit locking in the application.
 
@@
 to replace it. If an application can be split into completely
-independent parts that only very rarely need to share anything, or
+independent parts that only very rarely need to share something, or
 only do so via an external program like a database, then it is
@@
 The performance of HTM is pretty good as it does not introduce much
-overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
+overhead ($<40\%$~\cite{odaira14}). And it can transparently
 parallelise existing applications to some degree. The implementation
@@
 the criteria are defined intuitively, there are no formal justifications
-for the assigned points. The reader is thus advised to take them with a
-grain of salt.
+for the number of points. The reader is thus advised to take the result
+with a grain of salt.
 
@@
 better use of low-level features in existing OS kernels.  We are
-currently working on a STM system that makes use of several such
+currently working on an STM system that makes use of several such
 features like virtual memory and memory segmentation.  We further
-tailor the system to the discussed use case which gives us an
+tailor the system to the discussed use case, which gives us an
 advantage over other STM systems that are more general. With this

### 2014-06-18 talk/icooolps2014/position-paper.tex
some more adjustments

@@
 \usepackage{amssymb}
+
+% Keine "Schusterjungen"
+\clubpenalty = 10000
+% Keine "Hurenkinder"
+\widowpenalty = 10000 \displaywidowpenalty = 10000
 
@@
 \item[Better synchronisation:] Does the approach enable better,
-  paralleliseable synchronisation mechanisms for applications
+  parallelisable synchronisation mechanisms for applications
   (e.g.\ atomic blocks)?  Many synchronisation mechanisms can be built on
-  top of all solutions (e.g.\  message passing). We look for mechanisms
+  top of all solutions (e.g.\  message passing, monitors). We look for mechanisms
   that are directly enabled by the contending approaches.
@@
 limitations. Transactions can in principle be arbitrarily long.  This makes it
-possible to actually expose transactions to the application in the
-form of atomic blocks. This is the only approach that enables a better
-synchronisation mechanism than locks for applications \emph{and} still
-parallelises when using it. We think this is a very important point
-because it not only gives dynamic languages the ability to parallelise
-(already commonplace in most other languages), but also pushes
-parallel programming forward. Together with sequential consistency it
-provides a lot of simplification for parallel applications.
+possible to expose transactions to the application in the
+form of atomic blocks -- thereby attacking the issues
+of parallelisation and synchronisation in a unified way. While many
+synchronisation mechanisms can be bolted on top of any GIL replacement,
+this is the only approach that directly enables a better, parallelising
+synchronisation mechanism than locks.  We think this is a very
+important point because it not only gives dynamic languages the
+ability to parallelise (already commonplace in most other languages),
+but also pushes parallel programming forward in a way that other
+approaches cannot. Together with sequential consistency, it provides an
+environment for parallel applications that has much less surprises than
+e.g.\ Java or C\#.
 
@@
 for the number of points. The reader is thus advised to take the result
-with a grain of salt.
+with a grain of salt and form their own opinion.
 
@@
 model (sequential consistency) and the synchronisation of memory accesses
-using composable atomic blocks. It is not \emph{only} a simple GIL
+using composable atomic blocks. It is not \emph{just} a simple GIL
 replacement.
 
-Unfortunately, STM has a big performance problem. Particularly, for
+Unfortunately, STM has a big performance problem, which currently
+makes it lose this comparison. Particularly, for
 our use case there is not much static information available since we
@@
 tailor the system to the discussed use case, which gives us an
-advantage over other STM systems that are more general. With this
+advantage over other STM systems that try to be more general or
+simply focus on other use cases. With this
 approach, initial results suggest that we can keep the overhead of STM
@@
 we think that it is worthwhile to focus the STM research more
-specifically onto the context shown in this paper --- for use in
+specifically onto the context shown in this paper -- for use in
 implementations of high-level languages, rather than as a tool

### 2014-06-18 talk/icooolps2014/position-paper.tex
Two minor fixes (too late?  never mind)

@@
 also applies to Abstract Syntax Tree (AST) interpreters, where the GIL
-may only be released between interpreting two AST nodes.}. The interpreter
+may only be released between interpreting two AST nodes.  We talk about
+``bytecode instructions'' in a general way as a basic step in the
+interpreter.}.  The interpreter
 can thus rely on complete isolation and atomicity for the
@@
 \begin{description}
-\item[Performance:] How much does the approach impact performance on a single
+\item[Performance:] How much does the approach impact performance
+  on a single thread
   and how much on multiple threads? Can it make use of parallelism?

### 2014-07-29 talk/dls2014/paper/paper.tex
clarifications

@@
 
-Note, since the above configuration is specified at compile time,
+Note, since the above configuration is currently specified at compile time,
 all these areas are at offsets inside the segments known to the
@@
 if an object resides in the nursery only requires comparing its $SO$
-to the static boundaries of the nursery.
+to the static boundaries of the nursery. It is possible that we want
+some parameters to be configurable at startup or even during the
+runtime. In that case we can still use a compile-time specified
+maximum so that these checks are still efficient. E.g.\ limiting the
+maximum amount of memory available to the application statically to a few
+terabytes is fine because it corresponds to virtual memory,
+the real physical memory is assigned on-demand by the operating
+system.
 
@@
 
+Note, this works well for interpreters that really allocate things
+all the time. In the presence of an optimising just-in-time compiler,
+some loops may indeed not allocate anything and therefore still need
+the insertion of a safe-point check.
+
 % For other synchronisation requirements, for example:
@@
 We performed all benchmarks on a machine with an Intel Core i7-4770
-CPU~@3.40GHz (4 cores, 8 threads).  There are 16~GiB of memory
+CPU~@3.40GHz (4 cores, 8 threads) with disabled Hyper-Threading and
+Turbo Boost for less variation in the results.  There are 16~GiB of memory
 available and we ran them under Ubuntu 14.04 with a Linux 3.13.0

### 2014-07-29 talk/dls2014/paper/paper.tex
re-add mentioning non-x86; explain better why JIT-STM matters less at the moment; make JIT-STM result summary sound less pessimistic

@@
 natively (e.g.\ clang).
+
+On non-x86 architectures, most simple memory accesses could still be
+done efficiently if the supported addressing modes allow for the
+addition of an offset stored in some register (e.g.\ ARM). For more
+complicated accesses (e.g.\ array indexing) or if the CPU does not
+support such an addressing mode, one extra addition may be required.
 
@@
 the STM system, so we remove some unknown variables by disabling it.
+Furthermore, we think the interpreter-STM evaluation is more relevant
+at this stage as the results can be more directly applied to other
+similar interpreters. PyPy's JIT~\cite{cfbolz09}, however, is quite
+unique as it is a JIT tracing the interpreter instead of the
+interpreted language itself. This demands its own thorough
+evaluation, which is out of the scope of this paper.
 
@@
 
-Still, STM rarely beats CPython's single-thread performance. However, for
+Still, STM rarely beats CPython's \emph{single-thread} performance. However, for
 programs that need concurrency in CPython and that use threads to
@@
 
-Since PyPy comes with a JIT~\cite{cfbolz09} to make its overhead compared to CPython
-go away, we will now look at how well STM works together with it.
+Since PyPy comes with a JIT~\cite{cfbolz09} to make its overhead
+compared to CPython go away, we will now look at how well STM works
+together with it.
 
@@
 
-Overall PyPy needs the JIT  for its performance to be
-competitive with CPython's. It would be interesting to see how using
-our STM system in CPython would turn out, but an in-depth evaluation is
-beyond the scope of this paper. On its own, our system scales well,
-so we expect to also see this trend in the presence of a
-JIT in the future.
+Overall, PyPy without STM is around $2\times$ slower than CPython.
+Enabling its JIT allows it to outperform CPython by a huge margin.
+We see the same kind of speedup on PyPy with STM when enabling the
+JIT. This means that STM does not generally inhibit the JIT from
+optimising the programs execution, which is already a very important
+result on its own. We also see that for some
+benchmarks, STM is already able to give additional speedups
+compared to just the JIT-induced acceleration. This looks very
+promising and investigating when this is not the case is the next
+logical step.
 
@@
 
-To obtain an overall performance that is competitive with the
-best-performing Python systems (Jython, PyPy), integration of the
-STM-based approach with a JIT compiler is necessary. Once this
-integration matures, the approach outlined here serves not only as a
-simple GIL replacement but also provides a way forward towards
-a parallel programming model for Python.
+To generally outperform the best-performing Python systems (Jython,
+PyPy), integration of the STM-based approach with a JIT compiler is
+necessary. Our early results of this integration suggest that there is
+no inherent incompatibility between STM and PyPy's JIT.  Once the
+implementation matures, the approach outlined here serves not only as a
+simple GIL replacement but also provides a way forward towards a
+parallel programming model for Python.
 

### 2014-07-30 talk/dls2014/paper/paper.tex
some corrections and additions

@@
   barrier and flag for both use cases. Since the GC is an integral
-  part of many dynamic languages, we actually do not incur any additional
-  overhead here since the barriers are already there.
+  part of many dynamic languages, we often do not incur additional
+  overhead since the barriers are already there. However, we do need
+  some additional barriers when writing non-references into objects.
+  In that case, generational GCs do not need a barrier but the STM
+  system still does.
 
@@
 write barrier once between collections, and only the old objects need
-to be handled by the STM part \emph{once per transaction}. This keeps the
-overhead of the write barrier very low.
+to be handled by the STM part \emph{once per transaction}. The
+privatisation step in particular is done very rarely because the pages
+stay private until the next major collection where they may get
+re-shared. This keeps the overhead of the write barrier very low in
+the common cases.
 
@@
 % STM needs good performance:
-The solution presented here is based on a software TM system.  As the
+The solution presented here is based on a software TM system developed
+as an independent and reusable C library for integration in dynamic
+language interpreters. As the 
 STM is a vital part of the interpreter, good performance is
@@
   Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. Eliminating
-  global interpreter locks in ruby through hardware transactional
+  global interpreter locks in Ruby through hardware transactional
   memory. In \emph{Proc. 19th ACM SIGPLAN Symposium on

### 2014-07-30 talk/dls2014/paper/paper.tex
removing a lot of footnotes and other cleanups

@@
 concurrency. It is, however, limited by the GIL and thus does not
-provide parallelism\footnote{At this point we should mention that it is indeed
-possible to run external functions written in C instead of Python in
-parallel. Our work focuses on Python itself and ignores this aspect as
-it requires writing in a different language.}.  Moreover, the GIL can
-only ensure correctness of the interpreter itself: applications are
+provide real parallelism.
+The only code that is not necessarily restricted by the GIL in CPython
+is code written in e.g.\ C instead of Python. External functions can
+choose to run without the GIL and thus, in parallel. We ignore this
+aspect here as it requires writing in a different language than Python
+and is therefore not relevant to our work.
+
+The GIL can also
+only ensure correctness of the interpreter itself: applications using
+threads are still
 required to coordinate concurrent accesses to data structures using
@@
 differ between segments.  Instead, we store pointers as
-\emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start of the
-segment.  An $SO$ can then be interpreted in any segment by adding to
-it the start address of that segment.\footnote{An alternative model
-would be to simulate threads using several independent processes in
-which all segments are at the same address.  This might work well for
-demonstration purposes, but it is not really suitable for implementing
-an existing programming language: programs assume that most of the
-state (file descriptors, external libraries, ...) is shared.} The
+\emph{Segment Offsets (SO)}, i.e.\ offsets from the start of the
+segment.  An SO can then be interpreted in any segment by adding to
+it the start address of that segment. The
 result is called a \emph{Linear Address (LA)}. This is illustrated in
@@
 The x86 CPUs provide a feature which is called \emph{memory
-segmentation}.  On the modern x86-64, this is enough to perform the
-addition described above very efficiently.  We use the segment
-register $\%gs$\footnote{The other segment register $\%fs$ is
-typically used by the threading library to provide access to
-thread-local data.  One point of view is that we are using $\%gs$ for
-a similar purpose: thread-local data -- but a lot of it.}, which is
-available to applications, to point to the current thread's segment
-start address.  The translation from an $SO$ to an LA is done for us by
+  segmentation}.  On these CPU this is enough to perform the addition
+described above very efficiently. On the modern x86-64, there are two
+segment registers left: $\%fs$ and $\%gs$.  On Linux for example,
+$\%fs$ is already used for addressing thread-local data
+efficiently. Thus, we use the remaining segment register $\%gs$, which is
+still available to applications, to point to the current thread's segment
+start address.  The translation from an SO to an LA is done for us by
 using the ``$\%gs\colon$'' prefix before any CPU instruction that
-reads or writes memory using an $SO$ address.  This is very efficient
+reads or writes memory using an SO address.  This is very efficient
 -- we can do it on every object access -- and some compilers support it
@@
 constructs a linear address. Then, this LA gets mapped by the MMU to
-some place in the physical memory. This makes the $SO$ a valid reference
+some place in the physical memory. This makes the SO a valid reference
 to an object in all segments, automatically resolving either to a
@@
 program to register the objects in the read or write set before
-reading or writing to them\footnote{Barriers added by an automatic, conservative, and
-local program transformation~\cite{felber07}}.  Furthermore, we still
+reading or writing to them. We do this with an automatic, conservative, and
+local program transformation (cf.~\cite{felber07}).  Furthermore, we still
 need to describe the commit protocol.
@@
   as read by the current transaction. The mark is stored in a
-  segment-local array indexed by a number derived from the $SO$
+  segment-local array indexed by a number derived from the SO
   of the object.
@@
   transaction modifying an object at a time using write
-  locks.\footnote{Eager write-write detection is not inherent to our
-    approach; we may lift this restriction in the future.}  Part of making
-  the object write-ready is also the privatisation of all pages the
-  object belongs to, as described above.
+  locks\footnote{Eager write-write conflict detection is not inherent
+    to our approach; we may lift this restriction in the future if it
+    proves to be useful.}. To make the object write-ready, this is
+  also the place to privatise all pages the object belongs to, as
+  described above.
 
@@
 
-In the following sections, whenever we use $SO$, we go through the
+In the following sections, whenever we use SO, we go through the
 address translation to get to the actual contents of an object. This
@@
 irreversible operations, like calls to external functions or general
-input~/ output. If we are allowed to, we commit the transaction
-before and start a new one after these operations\footnote{The language
-specification needs to specify if this is allowed.}. Within an atomic
+input~/ output. If Python allows it, we commit the transaction before
+and start a new one after these operations (the same way as the GIL is
+released around thread-safe external functions). Within an atomic
 block, however, the transaction is unbreakable, and we need to turn
-the transaction inevitable so that it
-cannot abort and strong isolation is guaranteed.
+the transaction inevitable so that it cannot abort and strong
+isolation is guaranteed at all times.
 
@@
   $\%gs{::}SO$ translated to linear addresses will point to NULL pages
-  if $SO$ is set to \lstinline!NULL!.
+  if SO is set to \lstinline!NULL!.
 \item [{Segment-local~data:}] Some area private to the segment that
@@
   objects (\emph{young objects}) of the current transaction. The GC
-  uses bump-pointer allocation in this area to allocate objects in the
-  first generation.
+  uses bump-pointer allocation in this area to allocate objects of the
+  first generation. To make sure that only objects of the current
+  transaction reside in this space, a collection of the first
+  generation happens at the end of every transaction (see
+  section~\ref{sub:gc}).
 \item [{Old~object~space:}] These pages are the ones that are really
@@
 compiler. This makes some checks very efficient, e.g.\ checking
-if an object resides in the nursery only requires comparing its $SO$
+if an object resides in the nursery only requires comparing its SO
 to the static boundaries of the nursery. It is possible that we want
@@
 maximum amount of memory available to the application statically to a few
-terabytes is fine because it corresponds to virtual memory,
+terabytes is fine because it corresponds to virtual memory;
 the real physical memory is assigned on-demand by the operating
@@
 
-\subsection{Garbage Collection}
+\subsection{Garbage Collection\label{sub:gc}}
 
@@
 This area can be seen as a continuous, segment-local array of bytes
-that is indexed with an object's reference ($SO$) divided by 16. So
+that is indexed with an object's reference (SO) divided by 16. So
 that each object has its own read marker, all objects have a size of
@@
 Listing~\ref{lst:rb} is easily optimisable for compilers as well as
-perfectly predictable for CPUs\footnote{Additional benefit: The read
-barrier is not constrained to execute before the actual read -- both
-the compiler and the CPU are free to reorder or group them.}.
+perfectly predictable for CPUs. Additionally, the read barrier is not
+constrained to execute before the actual read -- both the compiler and
+the CPU are free to reorder or group them for efficiency.
 
@@
 need to check for the flag \lstinline!WRITE_BARRIER! in the object's
-header through its $SO$ and call the slow path if it is set. This flag
+header through its SO and call the slow path if it is set. This flag
 is set either if the object is old and comes from an earlier
@@
 is a simple, \emph{global} array of bytes that is indexed with the
-$SO$ of the object divided by 16. Note, ``global'' here really means
+SO of the object divided by 16. Note, ``global'' here really means
 it is a single array with data for all segments, there is no address
@@
 before we do the next minor collection or we start the next
-transaction.\footnote{We always do a minor collection during a commit.}
+transaction (remember: we always perform a minor collection as part of
+each commit anyway).
 
@@
 all threads so that the committing one is the only one running and all
-the others are waiting in safe points.\footnote{This design choice is
-important to support our very cheap barriers. It is a trade-off that
-works best if the number of concurrent threads is small.
-See section~\ref{subsub:sync} for how we mitigate the costs.}
+the others are waiting in safe points.
 We then go through the write
@@
 the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).
+
+Note that synchronising all threads is an important design choice we
+made to support our very cheap read barriers. Since they now consist
+only of a single unconditional write to some memory location, all
+threads need to be stopped while we detect conflicts. This is a
+trade-off that works best if the number of concurrent threads is
+small.
 
@@
 
-We also ran these benchmarks on Jython\footnote{version
-2.7b1~\cite{webjython}}, an implementation of Python on top of the
+We also ran these benchmarks on Jython (v2.7b1 from~\cite{webjython}),
+an implementation of Python on top of the
 Java Virtual Machine (JVM).  Instead of a GIL, this interpreter uses

### 2014-07-30 talk/dls2014/paper/paper.tex
explain why we only have micro benchs

@@
 
-To asses how well the STM system scales on its own (without any real
-workload), we execute the loop in Listing~\ref{lst:scaling_workload}
-on 1 to 4 threads on the PyPy interpreter with STM.
+To asses how well the STM system in combination with a Python
+interpreter scales on its own (without any real workload), we execute
+the loop in Listing~\ref{lst:scaling_workload} on 1 to 4 threads on
+the PyPy interpreter with STM and without a JIT.  There are very few
+allocations or calculations in this loop, so the main purpose of this
+benchmark is simply to check that there are no inherent conflicts in
+the interpreter when everything is thread-local. We also get some
+idea about how much overhead each additional thread introduces.
 
@@
 
-For the results in Figure~\ref{fig:scaling}, we
+The STM system detected no conflicts when running this code on 4
+threads. For the results in Figure~\ref{fig:scaling}, we
 normalised the average runtimes to the time it took on a single
-thread. From this we see that there is additional overhead introduced
+thread. From this we see that there is some additional overhead introduced
 by each thread ($12.3\%$ for all 4 threads together). Every thread
@@
 For the following sections we use a set of six small benchmarks
-available at~\cite{pypybenchs}:
+available at~\cite{pypybenchs}. There are, unsurprisingly, not
+many threaded applications written in Python that can be used
+as a benchmark. Since until now, threading was rarely used
+for performance reasons because of the GIL, we mostly collected
+small demos and wrote our own benchmarks to evaluate our system:
 

### 2014-07-30 talk/dls2014/paper/paper.tex
clarifications

@@
 
-PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
-benchmarks to a certain degree. It scales best for the embarrassingly
-parallel ones ($avg=2.6\times$ speedup) and a little less for
-the others ($avg=2.0\times$ speedup). The reason for this difference is
+On 4 cores, PyPy using our STM system (\emph{pypy-stm-nojit}) scales
+in all benchmarks to a certain degree. It scales best for the
+embarrassingly parallel ones ($avg=2.6\times$ speedup over 1 thread)
+and a little less for the others ($avg=2.0\times$ speedup over 1
+thread). The reason for this difference is
 that in the former group there are no real, logical conflicts -- all
@@
 
-Looking at the average overhead from switching from GIL to STM, we see
-that it is $\approx 43.3\%$. The maximum in richards is $71\%$. In all
-benchmarks \emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on
-two threads despite of this overhead.  The achieved speedup comparing
-STM to the GIL is between $1.14\times$ and $1.94\times$.
+Looking at the average overhead on a single thread that is induced by
+switching from GIL to STM, we see that it is $\approx 43.3\%$. The
+maximum in richards is $71\%$. In all benchmarks \emph{pypy-stm-nojit}
+beats \emph{pypy-nojit} already on two threads, despite of this
+overhead.  The achieved speedup comparing the lowest runtimes of STM
+to the single-threaded GIL execution is between $1.14\times$ and
+$1.94\times$.
 
@@
 
-Since PyPy comes with a JIT~\cite{cfbolz09} to make its overhead
-compared to CPython go away, we will now look at how well STM works
-together with it.
+Since PyPy comes with a JIT~\cite{cfbolz09} to make it vastly faster
+than CPython, we will now look at how well STM works together with it.
 
@@
 
-The speedups from enabling the JIT in these benchmarks range from
+The speedups from simply enabling the JIT in these benchmarks range from
 $10-50\times$. This is why we had to do without CPython here, since it
@@
 
-The results are presented in Figure~\ref{fig:performance-jit}. We
-see that the performance is much less stable. There is certainly more
-work required in this area. The slowdown factor for switching from GIL
-to STM ranges around $1-2.4\times$, and we beat GIL performance
-in half of the benchmarks.
+The results are presented in Figure~\ref{fig:performance-jit}. We see
+that the performance gains are much less reliable.  The slowdown
+factor for switching from GIL to STM ranges around $1-2.4\times$, and
+we beat the GIL's single-thread performance in half of the benchmarks.
 

### 2014-07-30 talk/dls2014/paper/paper.tex
be nicer to JS and update pdf

@@
 Dynamic languages like Python, PHP, Ruby, and JavaScript receive a lot
-of attention but have not yet embraced parallelism. A parallel
-programming model was not part of the design of those languages. Thus,
+of attention but only provide limited forms of parallelism. A parallel
+programming model was not part of the initial design of those languages. Thus,
 the reference implementations of, e.g., Python and Ruby use a single,
@@
 constrained to execute before the actual read -- both the compiler and
-the CPU are free to reorder or group them for efficiency.
+the CPU can reorder or group them freely between potential safe points
+for additional performance.
 

### 2014-07-31 talk/dls2014/paper/paper.tex
update benchmark numbers in paper

@@
 A common trait of many of these language's implementations is the use of
-a single global interpreter lock (GIL) to synchronise the interpreter
+a single, global interpreter lock (GIL) to synchronise the interpreter
 in a multithreading scenario. Since this lock effectively serialises
@@
 global interpreter lock (GIL) to serialise the execution of code in
-threads. The use of a single global lock causes several disadvantages,
+threads. The use of a single, global lock causes several disadvantages,
 and as multi-core processors become the de facto standard in platforms
@@
 threads. One approach to this is fine-grained locking instead of a
-single global lock. Jython~\cite{webjython} and
+single, global lock. Jython~\cite{webjython} and
 IronPython~\cite{ironpython} follow this approach. Fine-grained
@@
 kernel. The STM system was compiled with a number of segments $N=4$
-and a maximum amount of memory of 1.5~GiB (both are configurable at
+and a maximum amount of memory of 1.5~GiB (both configured at
 compile time).
@@
 thread. From this we see that there is some additional overhead introduced
-by each thread ($12.3\%$ for all 4 threads together). Every thread
+by each thread ($9.1\%$ for all 4 threads together). Every thread
 adds some overhead because during a commit, there is one more thread
@@
 
-While not ideal, we think that $12.3\%$ is acceptable on four
-threads. In terms of throughput, 4 threads have $3.54\times$
-more iterations per second than a single thread.
+While not ideal, we think that $9.1\%$ is acceptable on four
+threads. In terms of throughput, 4 threads reach $3.67\times$
+the iterations per second of a single thread.
 
@@
 
-For the following sections we use a set of six small benchmarks
+For the following sections we use a set of 8 small benchmarks
 available at~\cite{pypybenchs}. There are, unsurprisingly, not
@@
   parallel and checking for collisions with each other
-\item \emph{mandelbrot}, \emph{raytrace}, and \emph{richards}, which
-  all perform some independent computations in parallel (embarrassingly
-  parallel)
+\item \emph{miller-rabin}, \emph{mandelbrot}, \emph{raytrace},
+  \emph{richards}, and \emph{mersenne}, which all perform some mostly
+  independent computations in parallel (embarrassingly parallel).
 \end{itemize}
 
-We use atomic blocks to easily synchronise access to shared
-data structures in the first three benchmarks. For the embarrassingly
-parallel benchmarks, we do not need any explicit synchronisation at
-all. These atomic blocks are simulated with a single global lock
-when running on top of GIL-supported interpreters. With our STM
-system, they map to transactions that will execute optimistically
-in parallel.
+We use atomic blocks as the primary mechanism to synchronise access to
+shared data structures in the first three benchmarks. For the
+embarrassingly parallel benchmarks, we only need explicit
+synchronisation in some small isolated part. These atomic blocks are
+simulated with a single, global lock when running on top of
+GIL-supported interpreters. With our STM system, they map to
+transactions that will execute optimistically in parallel.
 
@@
 On 4 cores, PyPy using our STM system (\emph{pypy-stm-nojit}) scales
-in all benchmarks to a certain degree. It scales best for the
-embarrassingly parallel ones ($avg=2.6\times$ speedup over 1 thread)
-and a little less for the others ($avg=2.0\times$ speedup over 1
-thread). The reason for this difference is
-that in the former group there are no real, logical conflicts -- all
-threads do independent calculations. STM simply replaces the GIL in
-those programs. In the latter group, the threads work on a common data
-structure and therefore create much more conflicts, which limits the
-scalability. Here we make active use of the STM-supported atomic
-blocks to synchronise access to the data structure. The hope
-is that the STM system is still able to parallelise, even if we use
-the atomic blocks in a coarse-grained way. While less than the other
-benchmarks, we indeed see some speedup going from 1 to 3 threads.
-There is no visible benefit from 3 to 4 threads, even a slight
-regression.
+in all benchmarks to a certain degree. \emph{mersenne} is a bit of a
+special case, since it is easily parallelisable but a few of the
+parallel computations dominate the runtime and cannot be made faster
+easily by parallelisation. This is why the benefit of adding more
+threads than 2 is diminishing.
+
+\emph{pypy-stm-nojit} scales best for the embarrassingly parallel
+benchmarks ($avg=2.7\times$ speedup over 1 thread) and a little less
+for the others ($avg=2.5\times$ speedup over 1 thread). The reason for
+this difference is that in the former group there are no real, logical
+conflicts -- all threads do independent calculations. STM simply
+replaces the GIL in those programs. In the latter group, the threads
+work on a common data structure and therefore create much more
+conflicts, which limits the scalability. Here we make active use of
+the STM-supported atomic blocks to synchronise access to the data
+structure. The hope is that the STM system is still able to
+parallelise, even if we use the atomic blocks in a coarse-grained
+way. While less than the other benchmarks, we indeed see some speedup
+going from 1 to 4 threads.
 
 Looking at the average overhead on a single thread that is induced by
-switching from GIL to STM, we see that it is $\approx 43.3\%$. The
-maximum in richards is $71\%$. In all benchmarks \emph{pypy-stm-nojit}
+switching from GIL to STM, we see that it is $\approx 40.1\%$. The
+maximum in richards is $65.5\%$. In all benchmarks \emph{pypy-stm-nojit}
 beats \emph{pypy-nojit} already on two threads, despite of this
-overhead.  The achieved speedup comparing the lowest runtimes of STM
-to the single-threaded GIL execution is between $1.14\times$ and
-$1.94\times$.
-
-Still, STM rarely beats CPython's \emph{single-thread} performance. However, for
+overhead.  The achieved speedup comparing the fastest runtimes of STM
+to the single-threaded GIL execution of \emph{pypy-nojit} is between
+$1.11\times$ and $2.55\times$.
+
+Still, \emph{pypy-stm-nojit} barely beats CPython's \emph{single-thread}
+performance. However, for
 programs that need concurrency in CPython and that use threads to
@@
 the GIL on multiple threads. From this perspective, the STM
-implementation beats CPython's performance in all but two benchmarks.
+implementation beats CPython's performance in all benchmarks.
 
@@
 exposes the overhead of STM more by speeding up all the rest.
-For these reasons, the following results have to be taken with a grain
-of salt.
+For these reasons, the following results should be regarded as a
+preliminary outlook.
 
@@
 The results are presented in Figure~\ref{fig:performance-jit}. We see
-that the performance gains are much less reliable.  The slowdown
-factor for switching from GIL to STM ranges around $1-2.4\times$, and
-we beat the GIL's single-thread performance in half of the benchmarks.
+that the performance gains of STM are a bit less reliable with the JIT.
+The slowdown factor for switching from GIL to STM ranges around
+$1-2.5\times$, and we beat the GIL's single-thread performance in
+5 out of 8 benchmarks.
 
 We see that generally, the group of embarrassingly parallel benchmarks
-scales best. (There is a notable performance stability problem in the
-\emph{richards} benchmark. This is a bug we will fix in the future.)
-The other three benchmarks scale barely or not at all with the number of
-threads. The reason for this is likely again the conflicts in the
-latter group. Right now, because the code runs much more quickly with
-the JIT than without, it has the effect of making the transactions
-logically longer.  This increases the
-likelihood of conflicts between them and therefore limits scalability
-even more than in the no-JIT benchmarks.
-
-Overall, PyPy without STM is around $2\times$ slower than CPython.
-Enabling its JIT allows it to outperform CPython by a huge margin.
-We see the same kind of speedup on PyPy with STM when enabling the
-JIT. This means that STM does not generally inhibit the JIT from
-optimising the programs execution, which is already a very important
-result on its own. We also see that for some
-benchmarks, STM is already able to give additional speedups
-compared to just the JIT-induced acceleration. This looks very
-promising and investigating when this is not the case is the next
-logical step.
+scales best.  The other three benchmarks scale barely or not at all
+with the number of threads. The reason for this is likely again the
+conflicts in the latter group. Right now, because the code runs much
+more quickly with the JIT than without, it has the effect of making
+the transactions logically longer.  This increases the likelihood of
+conflicts between them and therefore limits scalability even more than
+in the no-JIT benchmarks.
+
+Overall, PyPy without STM is around $2\times$ slower than CPython
+(\emph{cpython} vs.\ \emph{pypy-nojit}).  Enabling its JIT allows it
+to outperform CPython by a huge margin.  We see the same kind of
+speedup on PyPy with STM when enabling the JIT. This means that STM
+does not generally inhibit the JIT from optimising the programs
+execution, which is already a very important result on its own. We
+also see that for some benchmarks, STM is already able to give
+additional speedups compared to just the JIT-induced
+acceleration. This looks very promising and improving this situation
+even more is the next step for reaching the goal of parallelising
+Python.
 
@@
 here and theirs.
-
-% Similar STMs:
-% \begin{itemize}
-% \item FastLane: \cite{warmhoff13}
-% \item TML: \cite{spear09}
-% \item Virtualizing HTM: \cite{rajwar05}
-% \item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
-%   detection, otherwise hardware extensions required; assumes most
-%   transactions fit HTM capacities (not so true here); COW using page-faults;
-%   they assume OS-level access to page-tables (maybe not inherent to their
-%   design); eval on simulator; value-based confl detection;
-
-%  (XTM can be
-%   implemented either in the OS as part of the virtual memory manager or
-%   between underlying TM systems and the OS, like virtual machines;
-%   Conflicts for overflowed transactions are tracked at page granularity;
-%   XTM-e allows conflict detection at cache line granu-
-%   larity, even for overflowed data in virtual memory)
-% \item using mmap(): Memory-Mapped Transactions
-% \item mem-protected conflict detection: \cite{martin09}
-% \end{itemize}
 
@@
 The early results presented here are very encouraging. STM as a simple
-GIL replacement scales well and yields an average speedup of $2.6\times$ for
+GIL replacement scales well and yields an average speedup of $2.7\times$ for
 embarrassingly parallel workloads on 4 threads.  When also used for
 synchronisation in the form of atomic blocks, the average speedup
-still reaches $2.0\times$.
+still reaches $2.5\times$.
 

### 2014-08-02 talk/dls2014/paper/paper.tex
fix formatting and references

@@
 
+\vspace{1cm} %% formatting hack
+
 \section{Evaluation}
@@
 \bibitem{cfbolz09} Carl Friedrich Bolz, Antonio Cuni, Maciej
-  Fijalkowski, and Armin Rigo. 2009. Tracing the meta-level: PyPy's
-  tracing JIT compiler.  \emph{In Proceedings of the 4th workshop on the
+  Fijalkowski, and Armin Rigo. 2009. ``Tracing the meta-level: PyPy's
+  tracing JIT compiler.''  \emph{Proc. 4th workshop on the
     Implementation, Compilation, Optimization of Object-Oriented Languages
-    and Programming Systems} (ICOOOLPS '09).
-
-\bibitem{kevin10} Kevin Millikin, Florian Schneider. 2010.  A New
-  Crankshaft for V8.
+    and Programming Systems} (ICOOOLPS '09), 18-25
+
+\bibitem{kevin10} Kevin Millikin, Florian Schneider. 2010.  ``A New
+  Crankshaft for V8.''
   \url{http://blog.chromium.org/2010/12/new-crankshaft-for-v8.html}
@@
 
-\bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. A Way Forward
-  in Parallelising Dynamic Languages. To appear in ICOOOLPS'14.
+\bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. ``A Way Forward
+  in Parallelising Dynamic Languages.'' In \emph{Proc. 9th International
+    Workshop on Implementation, Compilation, Optimization of Object-Oriented
+    Languages, Programs and Systems PLE} (ICOOOLPS '14)
 
@@
 
-\bibitem{beazley10} Beazley, David. "Understanding the Python GIL."
+\bibitem{beazley10} David Beazley. ``Understanding the Python GIL.''
   \emph{PyCON Python Conference}. Atlanta, Georgia. 2010.
 
-\bibitem{harris10} Harris, Tim, James Larus, and Ravi
-  Rajwar. "Transactional memory." \emph{Synthesis Lectures on Computer
-  Architecture 5.1} (2010)
+\bibitem{harris10} Tim Harris, James Larus, and Ravi
+  Rajwar. 2010. ``Transactional memory.'' \emph{Synthesis Lectures on Computer
+  Architecture 5.1}
 
 \bibitem{guerraoui08}
-  Rachid Guerraoui and Michal Kapalka. 2008. On the correctness of
-  transactional memory. In \emph{Proceedings of the 13th ACM SIGPLAN
-    Symposium on Principles and practice of parallel programming} (PPoPP
-  '08).
-
-\bibitem{blundell06} Blundell, Colin, E. Christopher Lewis, and Milo
-  Martin. "Unrestricted transactional memory: Supporting I/O and system
-  calls within transactions." (2006).
+  Rachid Guerraoui and Michal Kapalka. 2008. ``On the correctness of
+  transactional memory.'' In \emph{Proc. 13th ACM SIGPLAN Symposium on
+    Principles and practice of parallel programming} (PPoPP '08),
+  175-184
+
+\bibitem{blundell06} Colin Blundell, E. Christopher Lewis, and Milo
+  Martin. 2006. ``Unrestricted transactional memory: Supporting I/O and system
+  calls within transactions.''
 
 \bibitem{spear08} Michael F. Spear and Michael Silverman and Luke
-  Daless and Maged M. Michael and Michael L. Scott. 2008. "Implementing
-  and exploiting inevitability in software transactional memory."  In
+  Daless and Maged M. Michael and Michael L. Scott. 2008. ``Implementing
+  and exploiting inevitability in software transactional memory.''  In
   \emph{Proc. 37th IEEE international conference on parallel
@@
 
-\bibitem{fergus02} Fergus Henderson. 2002. Accurate garbage collection
-  in an uncooperative environment. \emph{In Proceedings of the 3rd
-    international symposium on Memory management} (ISMM '02).
-
-\bibitem{stmupdate13} Armin Rigo, Remigius Meier. 2013. Update on
-  STM. \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}
+\bibitem{fergus02} Fergus Henderson. 2002. ``Accurate garbage collection
+  in an uncooperative environment.'' In \emph{Proc. 3rd
+    International Symposium on Memory management} (ISMM '02), 150-156
+
+\bibitem{stmupdate13} Armin Rigo, Remigius Meier. 2013. ``Update on
+  STM.'' \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}
 
 \bibitem{rajwar05} Ravi Rajwar, Maurice Herlihy, and Konrad
-  Lai. 2005. Virtualizing Transactional Memory. In \emph{Proceedings of
-    the 32nd annual international symposium on Computer Architecture}
-  (ISCA '05).
+  Lai. 2005. ``Virtualizing Transactional Memory.'' In \emph{Proc.
+    32nd annual International Symposium on Computer Architecture}
+  (ISCA '05), 494-505
 
@@
   Travis Skare, Hassan Chafi, Brian D. Carlstrom, Christos Kozyrakis,
-  and Kunle Olukotun. 2006. Tradeoffs in transactional memory
-  virtualization. \emph{SIGOPS Oper. Syst. Rev.} 40, 5 (October 2006),
-  371-381.
+  and Kunle Olukotun. 2006. ``Tradeoffs in transactional memory
+  virtualization.'' In \emph{Proc. 12th international conference on
+    Architectural support for programming languages and operating
+    systems} (ASPLOS XII), 371-381
 
 \bibitem{martin09} Martín Abadi, Tim Harris, and Mojtaba
-  Mehrara. 2009. Transactional memory with strong atomicity using
-  off-the-shelf memory protection hardware. \emph{SIGPLAN Not.} 44, 4
-  (February 2009), 185-196.
+  Mehrara. 2009. ``Transactional memory with strong atomicity using
+  off-the-shelf memory protection hardware.'' In \emph{Proc. 14th ACM
+    SIGPLAN symposium on Principles and practice of parallel
+    programming} (PPoPP '09), 185-196
 
 \bibitem{pypybenchs} PyPy benchmarks repository. 2014. Revision
-  a26f2fb58413. \url{bitbucket.org/pypy/benchmarks}
+  fd2da4da8f33. \url{bitbucket.org/pypy/benchmarks}
 
@@
 \bibitem{odaira14}
-  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. Eliminating
+  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. ``Eliminating
   global interpreter locks in Ruby through hardware transactional
-  memory. In \emph{Proc. 19th ACM SIGPLAN Symposium on
-    Principles and Practice of Parallel Programming} (PPoPP '14).
+  memory.'' In \emph{Proc. 19th ACM SIGPLAN Symposium on
+    Principles and Practice of Parallel Programming} (PPoPP '14), 131-142
 
 \bibitem{warmhoff13}
-  Jons-Tobias Wamhoff, Christof Fetzer, Pascal Felber, Etienne Rivière,
-  and Gilles Muller. 2013. FastLane: improving performance of software
-  transactional memory for low thread counts. \emph{SIGPLAN Not.} 48, 8
-  (February 2013), 113-122.
+  Jons-Tobias Wamhoff, Christof Fetzer, Pascal Felber, Etienne
+  Rivière, and Gilles Muller. 2013. ``FastLane: improving performance
+  of software transactional memory for low thread counts.'' In
+  \emph{Proc. 18th ACM SIGPLAN symposium on Principles and practice of
+    parallel programming} (PPoPP '13), 113-122
 
 \bibitem{drago11} Aleksandar Dragojević, Pascal Felber, Vincent
-  Gramoli, and Rachid Guerraoui. 2011. Why STM can be more than a
-  research toy. \emph{Commun. ACM} 54, 4 (April 2011), 70-77.
+  Gramoli, and Rachid Guerraoui. 2011. ``Why STM can be more than a
+  research toy.'' \emph{Commun. ACM} 54, 4 (April 2011), 70-77.
 
@@
   Calin Cascaval, Colin Blundell, Maged Michael, Harold W. Cain, Peng
-  Wu, Stefanie Chiras, and Siddhartha Chatterjee. 2008. Software
+  Wu, Stefanie Chiras, and Siddhartha Chatterjee. 2008. ``Software
   transactional memory: why is it only a research
-  toy?. \emph{Commun. ACM} 51, 11 (November 2008), 40-46.
+  toy?.'' \emph{Commun. ACM} 51, 11 (November 2008), 40-46.
 
 \bibitem{nicholas06}
-  Nicholas Riley and Craig Zilles. 2006. Hardware transactional memory
-  support for lightweight dynamic language evolution. \emph{In
+  Nicholas Riley and Craig Zilles. 2006. ``Hardware transactional memory
+  support for lightweight dynamic language evolution.'' \emph{In
     Companion to the 21st ACM SIGPLAN symposium on Object-oriented
     programming systems, languages, and applications} (OOPSLA
-  '06).
+  '06), 998-1008
 
 \bibitem{fuad10}
-  Fuad Tabba. 2010. Adding concurrency in Python using a commercial
-  processor's hardware transactional memory support. \emph{SIGARCH
+  Fuad Tabba. 2010. ``Adding concurrency in Python using a commercial
+  processor's hardware transactional memory support.'' \emph{SIGARCH
   Comput. Archit. News 38}, 5 (April 2010), 12-19.
@@
   Pascal Felber and Torvald Riegel and Christof Fetzer and Martin
-  Süßkraut and Ulrich Müller and Heiko Sturzrehm. 2007. Transactifying
-  applications using an open compiler framework. \emph{TRANSACT}, August
+  Süßkraut and Ulrich Müller and Heiko Sturzrehm. 2007. ``Transactifying
+  applications using an open compiler framework.'' \emph{TRANSACT}, August
   (2007): 4-6.
@@
   Luke Dalessandro, Dave Dice, Michael Scott, Nir Shavit, and Michael
-  Spear. 2010. Transactional mutex locks. In \emph{Proceedings of the
+  Spear. 2010. ``Transactional mutex locks.'' In \emph{Proc.
     16th international Euro-Par conference on Parallel processing: Part
@@
 \bibitem{lamport79}
-  L. Lamport. 1979. How to Make a Multiprocessor Computer That Correctly
-  Executes Multiprocess Programs. \emph{IEEE Trans. Comput.} 28, 9
+  L. Lamport. 1979. ``How to Make a Multiprocessor Computer That Correctly
+  Executes Multiprocess Programs.'' \emph{IEEE Trans. Comput.} 28, 9
   (September 1979), 690-691.
@@
 \bibitem{victor11}
-  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. A study of
-  transactional memory vs. locks in practice. In \emph{Proc.
-     twenty-third annual ACM Symposium on Parallelism in Algorithms
-    and Architectures} (SPAA '11).
+  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. ``A study of
+  transactional memory vs. locks in practice.'' In \emph{Proc.
+     23rd annual ACM Symposium on Parallelism in Algorithms
+    and Architectures} (SPAA '11), 43-52
 
@@
   Christopher J. Rossbach, Owen S. Hofmann, and Emmett
-  Witchel. 2010. Is transactional programming actually
-  easier?. \emph{Proc. 15th ACM SIGPLAN Symposium on
-    Principles and Practice of Parallel Programming} (PPoPP '10).
+  Witchel. 2010. ``Is transactional programming actually
+  easier?.'' \emph{Proc. 15th ACM SIGPLAN Symposium on
+    Principles and Practice of Parallel Programming} (PPoPP '10), 47-56
 
 \bibitem{tim03}
-  Tim Harris and Keir Fraser. 2003. Language support for lightweight
-  transactions. \emph{In Proceedings of the 18th annual ACM SIGPLAN
+  Tim Harris and Keir Fraser. 2003. ``Language support for lightweight
+  transactions.'' In \emph{Proc. 18th annual ACM SIGPLAN
     conference on Object-oriented programing, systems, languages, and
@@
 \bibitem{tim05} Tim Harris, Simon Marlow, Simon Peyton Jones, and
-  Maurice Herlihy. 2008. Composable memory transactions.
+  Maurice Herlihy. 2008. ``Composable memory transactions.''
   \emph{Commun. ACM} 51, 8 (August 2008), 91-100.
@@
 \bibitem{shan08}
-  Shan Lu, Soyeon Park, Eunsoo Seo, and Yuanyuan Zhou. 2008. Learning
+  Shan Lu, Soyeon Park, Eunsoo Seo, and Yuanyuan Zhou. 2008. ``Learning
   from mistakes: a comprehensive study on real world concurrency bug
-  characteristics. \emph{SIGARCH Comput. Archit. News} 36, 1 (March 2008),
+  characteristics.'' \emph{SIGARCH Comput. Archit. News} 36, 1 (March 2008),
   329-339.
@@
   Muth, Tavis Ormandy, Shiki Okasaka, Neha Narula, and Nicholas
-  Fullagar. 2010. Native Client: a sandbox for portable, untrusted x86
-  native code. \emph{Commun. ACM} 53, 1 (January 2010), 91-99.
+  Fullagar. 2010. ``Native Client: a sandbox for portable, untrusted x86
+  native code.'' \emph{Commun. ACM} 53, 1 (January 2010), 91-99.
 
 % \bibitem{leis14}
-%   Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
-%   Hardware Transactional Memory in Main-Memory Databases."
+%   Leis, Viktor, Alfons Kemper, and Thomas Neumann. ``Exploiting
+%   Hardware Transactional Memory in Main-Memory Databases.''
 %   \emph{Proc. of ICDE}. 2014.

### 2014-08-02 talk/dls2014/paper/paper.tex
some more adjustments

@@
   (GC) to lower the overhead of STM.
-\item This new STM system is used to replace the GIL in one
-  implementation of Python and is then evaluated.
+\item This new STM system is used to replace the GIL in a Python
+  interpreter called PyPy and is then evaluated.
 \end{itemize}
@@
 
-Our TM system is designed as a C library that covers all aspects around
-transactions and object memory management. It is designed for
-object-oriented dynamic language VMs as a replacement for the GIL.
-The library exposes the functions in Listing~\ref{lst:api} as its
-interface for use by a dynamic language interpreter.
+Our TM system is designed as a C library~\cite{stmgc-c7} that covers
+all aspects around transactions and object memory management. It is
+designed for object-oriented dynamic language VMs as a replacement for
+the GIL.  The library exposes the functions in Listing~\ref{lst:api}
+as its interface for use by a dynamic language interpreter.
 
@@
 is also signified by the type \lstinline!object_t!.  This type is
-special as it causes the
-compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
-its ``address-space 256'' feature. Patches are available from authors
-until inclusion into the official clang.} to make all accesses through
-it relative to the $\%gs$ register.  With exceptions, nearly all
-accesses to objects managed by the TM system use this type so
-that the CPU will translate the reference to the right version of the
-object.
+special as it causes the compiler\footnote{We use Clang 3.5 with
+  patches to deal with bugs in its ``address-space 256''
+  feature. Patches are available from authors until inclusion into the
+  official clang~\cite{stmgc-c7}.} to make all accesses through it
+relative to the $\%gs$ register.  With exceptions, nearly all accesses
+to objects managed by the TM system use this type so that the CPU will
+translate the reference to the right version of the object.
 \medskip   % why, Latex, why??
@@
 \item to support \emph{inevitable} transactions, we always prefer them
-  to others since they cannot abort (similar to~\cite{blundell06})
+  to others since they cannot abort~\cite{blundell06}
 \end{itemize}
@@
 
+Again we see that the GIL degrades performance on multiple threads.
+On the other hand, STM gets faster with more threads, or at least
+not much slower. Hence, any application using threads for concurrency
+immediately benefits from running on STM. On 4 threads, its performance
+is better than the GIL's in 7 out of 8 benchmarks by up to $4\times$.
+
 We see that generally, the group of embarrassingly parallel benchmarks
@@
 
+\vspace{3mm} % formatting hack
 \section{Related Work}
@@
 
+\bibitem{stmgc-c7} STMGC-C7 library repository. 2014
+  \url{bitbucket.org/pypy/stmgc}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-08-20 talk/uct2014/example/demo1.py
Add a draft of UCT talk

@@ -1,0 +1,23 @@
+
+def adder(x, y):
+    return x + y
+
+def entry_point(argv):
+    if len(argv) != 3:
+        print "Wrong number of args"
+        return 1
+    try:
+        arg0 = int(argv[1])
+        arg1 = int(argv[2])
+    except ValueError:
+        print "Requires ints"
+        return 1
+    print "Added", arg0 + arg1
+    return 0
+
+def target(*args):
+    return entry_point
+
+if __name__ == '__main__':
+    import sys
+    entry_point(sys.argv)

### 2014-09-11 talk/dyla2007/dyla.tex
work on the proposal

@@
 SLang, which use the syntax and
-metaprogramming facilities of a high-level language but encode
+meta-programming facilities of a high-level language but encode
 low-level details like object layout and memory management.} language.
@@
 approaches, using various Python implementations as the main example. In
-section \ref{sect:metaprogramming} we will describe our proposed
+section \ref{sect:meta-programming} we will describe our proposed
 meta-programming approach and compare the two solutions.  We summarize
@@
 JVM which just loaded Jython consumes between 34 and 42 MB of memory
-(Linux/IA32), while on the same machine a CPython process fits in 3 to 4
+(Linux/IA32), while on the same machine a CPython process fits into 3 to 4
 MB.
@@
 \section{Meta-Programming Is Good}
-\label{sect:metaprogramming}
+\label{sect:meta-programming}
 
@@
 
-\subsection{PyPy architecture}
+\subsection{PyPy's architecture}
 
@@
 because of its size and rather intricate semantics, is a good target for
-our approach in the following sense: its previous reimplementations
+our approach in the following sense: its previous re-implementations
 (Jython for the JVM and IronPython for .NET) have each proved to be very
@@
 framework.  However, this is not the only way to obtain good GCs:
-existing GCs can also be reused.  Good candidates are the GCs written in
+existing GCs could also be reused.  Good candidates are the GCs written in
 the Jikes RVM \cite{JikesGC}.  As they are written in Java, it should be
@@
 Aside from the advantages described in section
-\ref{sect:metaprogramming}, a translation toolchain need not be
+\ref{sect:meta-programming}, a translation toolchain need not be
 standardized for inter-operability but can be tailored to the needs of

### 2014-09-11 confrest.py

@@ -1,0 +1,37 @@
+import py
+from confrest_oldpy import *
+
+class PyPyPage(Page): 
+    def fill(self):
+        super(PyPyPage, self).fill()
+        self.menubar[:] = html.div(
+            html.a("news", href="/pypy/dist/pypy/doc/news.html", class_="menu"), " ",
+            html.a("doc", href="/pypy/dist/pypy/doc/index.html", class_="menu"), " ",
+            html.a("contact", href="/pypy/dist/pypy/doc/contact.html", class_="menu"), " ", 
+            html.a("getting-started", 
+                   href="/pypy/dist/pypy/doc/getting-started.html", class_="menu"), " ",
+            html.a("issue", 
+                   href="https://codespeak.net/issue/pypy-dev/", 
+                   class_="menu"), 
+            " ", id="menubar")
+
+    def get_doclink(self, target):
+        return relpath(self.targetpath.strpath,
+                       self.project.get_docpath().join(target).strpath)
+
+class Project(Project): 
+    title = "PyPy" 
+    stylesheet = 'http://codespeak.net/pypy/dist/pypy/doc/style.css'
+    encoding = 'latin1' 
+    prefix_title = "PyPy"
+    logo = html.div(
+        html.a(
+            html.img(alt="PyPy", id="pyimg", 
+                     src="http://codespeak.net/pypy/img/py-web1.png", 
+                     height=110, width=149)))
+    Page = PyPyPage 
+    mydir = py.magic.autopath().dirpath()
+    def get_docpath(self):
+        return self.mydir
+
+

### 2014-09-11 talk/uct2014/example/interp.py
work on the proposal

@@ -46,3 +46,7 @@ def entry_point(argv):
 
 def target(*args):
     return entry_point
+
+if __name__ == '__main__':
+    import sys
+    entry_point(sys.argv)

### 2014-09-19 talk/s3-2008/talk.tex
tweak

@@
 bool_ops = [
-    (LESSTHAN, operator.lt),
-    (GREATERTHAN, operator.gt),
-    (LESSOREQUAL, operator.le),
-    (GREATEROREQUAL,operator.ge),
-    (EQUAL, operator.eq),
-    (NOTEQUAL, operator.ne)
+    (LESSTHAN, lambda x, y: x < y),
+    (GREATERTHAN, lambda x, y: x > y),
+    (LESSOREQUAL, lambda x, y: x <= y),
+    (GREATEROREQUAL,lambda x, y: x >= y),
+    (EQUAL, lambda x, y: x == y),
+    (NOTEQUAL, lambda x, y: x != y)
     ]

### 2014-09-19 talk/dls2012/benchmarks/iter/while.py

@@ -1,0 +1,176 @@
+from array import array
+
+def _sum1d(a):
+    sa = i = 0
+    while i < len(a):
+        sa += a[i]
+        i += 1
+
+def _xsum1d(a):
+    sa = i = 0
+    while i < len(a):
+        sa += a[i] + i
+        i += 1
+
+def _wsum1d(a):
+    sa = i = 0
+    while i < len(a):
+        sa += a[i] + len(a)
+        i += 1
+        
+def _sum2d(a, w, h):
+    sa = y = 0
+    while y < h:
+        x = 0
+        while x < w:
+            sa += a[y*w + x]
+            x += 1
+        y += 1
+
+def _wsum2d(a, w, h):
+    sa = y = 0
+    while y < h:
+        x = 0
+        while x < w:
+            sa += a[y*w + x] + w
+            x += 1
+        y += 1
+
+def _xsum2d(a, w, h):
+    sa = y = 0
+    while y < h:
+        x = 0
+        while x < w:
+            sa += a[y*w + x] + x
+            x += 1
+        y += 1
+    sa = 0
+
+def _whsum2d(a, w, h):
+    sa = y = 0
+    while y < h:
+        x = 0
+        while x < w:
+            sa += a[y*w + x] + w + h
+            x += 1
+        y += 1
+    sa = 0
+
+def _xysum2d(a, w, h):
+    sa = y = 0
+    while y < h:
+        x = 0
+        while x < w:
+            sa += a[y*w + x] + x + y
+            x += 1
+        y += 1
+
+def _mean1d(a):
+    sa = i = 0
+    while i < len(a):    
+        sa = (i*sa + a[i])/(i + 1.0);
+        i += 1
+        
+def _median1d(a):
+    sa = i = 0
+    while i < len(a):
+        if sa > a[i]:
+            sa -= 1.0/(i + 1.0)
+        elif sa < a[i]:
+            sa += 1.0/(i + 1.0)
+        i += 1
+
+def _ripple1d(a):
+    sa = i = 0
+    while i < len(a):
+        if sa > a[i]:
+            sa -= 0.1
+        elif sa < a[i]:
+            sa += 0.1
+        i += 1
+
+def _ripple2d(a, w, h):
+    sa = 0
+    sa = y = 0
+    while y < h:
+        x = 0
+        while x < w:
+            if sa > a[y*w + x]:
+                sa -= 0.1
[…]

### 2014-10-01 talk/badhonnef2007/badhonnef.tex
More slides

\documentclass{llncs}

\usepackage{makeidx}
\usepackage{graphicx}

\begin{document}

\pagestyle{headings}

\title{A Flexible Prolog Interpreter in Python}

%\titlerunning{XXX}     % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used

\author{Carl Friedrich Bolz}

\authorrunning{Bolz}   % abbreviated author list (for running head)

%%%% modified list of authors for the TOC (add the affiliations)
\tocauthor{Bolz (D\"usseldorf)}

\institute{ Lehrstuhl Softwaretechnik und Programmiersprachen\\
Institut f\"{u}r Informatik, Universit\"at D\"usseldorf, Germany\\
  \email{ cfbolz@gmx.de}}

\maketitle

\begin{abstract}

Efficient virtual machines for the Prolog programming language (and also for
other dynamic highlevel languages) are usually large and intricate pieces of
software implemented in a low-level language such as C that are extensively
optimized over many years and are therefore difficult to change in many
respects. Due to their inherent complexity it is hard to develop them further,
to maintain them or to use them to experiment with language extensions or new
built-ins.

Early implementation decisions, such as choice of garbage collector, are
extremely hard to change afterwards. In addition, it is also much harder to
perform advanced optimizations or transformations (such as partial evaluation)
on the virtual machine itself due to the low level of the implementation
language. Furthermore it is (by definition) impossible to port them to a non-C
platform such as the Java Virtual Machine or the .NET framework Therefore the
numerous implementations of a Prolog virtual machine integrated into one of
these environments are all new implementations which have to be maintained
themselves and therefore all have different levels of maturity and slightly
different sets of built-ins available.

Using a higher-level language than C addresses many of these problems. It
allows the language implementor to work at a higher level of abstraction, which
eases implementation, maintenance and extensions of the virtual machine and
reduces the size of the code base.  Furthermore, advanced optimizations and
transformations can be performed. To regain usable performance the
implementation needs then to be translated to a low-level language again.

The goal of the PyPy project is to support the implementation of interpreters
in the high-level language Python. Python is an object oriented dynamic
language which is becoming more and more popular for scripting use as well as
for “real” applications. PyPy provides a translation toolchain that helps to
translate interpreters written in Python to a low-level language such as C and
the .NET Intermediate Language. To make this translation work, these
interpreters have to be written in RPython, which is a proper subset of Python
chosen in such a way that it is amenable to analysis.

In this talk we give an quick overview of the PyPy project and present a Prolog
interpreter written in RPython, exploring the possibilities of implementing a
well-known declarative language in a high-level language (as opposed to C). One
of the main goals of the implementation was to keep it as simple and as
extensible as possible (even at the cost of performance). The PyPy tool suite
is used to reach a reasonable level of performance and to ensure portability to
various platforms.\footnote{This research was partially supported by the EU funded
 project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

\end{abstract}

% ---- Bibliography ----
%\begin{small}
\bibliographystyle{abbrv}
\bibliography{badhonnef}
%\end{small}

\end{document}

### 2014-10-01 talk/microsoft-jan-2011/talk.tex
More slides

--- 2011-02-07 talk/ibm-feb-2011/talk.tex
@@
 %\includegraphics[scale=0.25]{figures/hpi_logo_kl.jpg}\\
-IBM Watson Research Center, February 8th, 2011
+%IBM Watson Research Center, February 9th, 2011
+Microsoft Research, January 31st, 2011
 }
@@
   \frametitle{How do Interpreters Meet the Requirements?}
-  \includegraphics[scale=0.6]{figures/output1.pdf}
-%  \begin{enumerate}
-%      \item \alert{control flow slowed by bytecode or AST dispatch overhead}
-%      \item \alert{late binding works but slow}
-%      \item \alert{dispatching works but slow}
-%      \item \alert{no special dependencies support}
-%      \item \alert{everything is boxed}
-%      \item \alert{reified frames are easy but slow}
-%%      \item \alert{access to native libraries easy but slow}
-%  \end{enumerate}
+  \begin{enumerate}
+      \item \alert{control flow slowed by bytecode or AST dispatch overhead}
+      \item \alert{late binding works but slow}
+      \item \alert{dispatching works but slow}
+      \item \alert{no special dependencies support}
+      \item \alert{everything is boxed}
+      \item \alert{reified frames are easy but slow}
+%      \item \alert{access to native libraries easy but slow}
+  \end{enumerate}
 \end{frame}
@@
   \frametitle{How do Static Compilers Meet the Requirements?}
-  \includegraphics[scale=0.6]{figures/output2.pdf}
-%  \begin{enumerate}
-%      \item control flow works well
-%      \item \alert{late binding not improved}
-%      \item \alert{dispatching not improved}
-%      \item \alert{dependencies not improved}
-%      \item \alert{everything is boxed}
-%      \item \alert{reified frames often not supported}
-%%      \item direct access to native libraries
-%  \end{enumerate}
+  \begin{enumerate}
+      \item control flow works well
+      \item \alert{late binding not improved}
+      \item \alert{dispatching not improved}
+      \item \alert{dependencies not improved}
+      \item \alert{everything is boxed}
+      \item \alert{reified frames often not supported}
+%      \item direct access to native libraries
+  \end{enumerate}
 \end{frame}
@@
   \frametitle{How do Method-Based JIT Compilers Meet the Requirements?}
-  \includegraphics[scale=0.6]{figures/output3.pdf}
-%  \begin{enumerate}
-%      \item control flow works well
-%      \item late binding can be handled
-%      \item dispatching can be handled
-%      \item \alert{dependencies not necessarily improved}
-%      \item boxing hard to support
-%      \item \alert{reified frames hard to support}
-%%      \item \alert{access to native libraries not improved}
-%  \end{enumerate}
+  \begin{enumerate}
+      \item control flow works well
+      \item late binding can be handled
+      \item dispatching can be handled
+      \item \alert{dependencies not necessarily improved}
+      \item boxing hard to support
+      \item \alert{reified frames hard to support}
+%      \item \alert{access to native libraries not improved}
+  \end{enumerate}
 \end{frame}
@@
   \frametitle{How do Tracing JITs Meet the Requirements?}
-  \includegraphics[scale=0.6]{figures/output4.pdf}
-%  \begin{enumerate}
-%      \item control flow works rather well
-%      \item late binding can be handled
-%      \item dispatching can be handled
-%      \item dependencies improved by path splitting
-%      \item unboxing optimization much simpler
-%      \item reified frames can be implemented \\
-%      by falling back to the interpreter
-%%      \item \alert{access to native libraries not improved}
-%  \end{enumerate}
+  \begin{enumerate}
+      \item control flow works rather well
+      \item late binding can be handled
+      \item dispatching can be handled
+      \item dependencies improved by path splitting
+      \item unboxing optimization much simpler
+      \item reified frames can be implemented \\
+      by falling back to the interpreter
+%      \item \alert{access to native libraries not improved}
+  \end{enumerate}
 \end{frame}
@@
   \frametitle{How do OO VMs Meet the Requirements?}
-  \includegraphics[scale=0.6]{figures/output5.pdf}
-%  \begin{enumerate}
-%      \item control flow works well
-%      \item late binding can be handled with a lot of effort
-%      \item dispatching can be handled with a lot of effort
-%      \item \alert{dependencies not improved}
-%      \item \alert{boxing not necessarily improved}
-%      \item \alert{reified frames are inefficient}
-%%      \item access to native libraries not improved, but VM libraries help
-%  \end{enumerate}
+  \begin{enumerate}
+      \item control flow works well
+      \item late binding can be handled with a lot of effort
+      \item dispatching can be handled with a lot of effort
+      \item \alert{dependencies not improved}
+      \item \alert{boxing not necessarily improved}
+      \item \alert{reified frames are inefficient}
+%      \item access to native libraries not improved, but VM libraries help
+  \end{enumerate}
 \end{frame}

### 2014-10-01 talk/dls2012/demo/view.py

@@ -1,0 +1,6 @@
+from io import mplayer, view
+import sys
+
+for img in mplayer(sys.argv[1]):
+    view(img)
+

### 2014-10-01 talk/pycon2013/pypy_without_gil/transactions0.py

@@ -1,0 +1,13 @@
+import transaction
+
+
+
+def compute():
+    for k in range(10**7):
+        pass
+
+
+for i in range(10):
+    transaction.add(compute)
+
+transaction.run()

### 2014-10-29 planning/hg-migration/fix_usermap.py
draft

@@ -1,0 +1,30 @@
+"""
+fix_usermap <authormap> <fixups>
+:takes a name: realmail listing, fixes up a authormap (see usermap.txt)
+
+"""
+import sys
+
+fixups = {}
+with open(sys.argv[2]) as fp:
+    for line in fp:
+        try:
+            name, mail = line.split(':')
+            fixups[name.strip()] = mail.strip()
+        except ValueError:
+            pass
+            
+
+with open(sys.argv[1]) as fp:
+    for line in fp:
+        if 'codespeak.net' not in line:
+            sys.stdout.write(line)
+        else:
+            before = line.split('<')[0]
+            name = line.split('=')[0]
+            if name in fixups:
+                sys.stdout.write('%s<%s>\n'% (before, fixups[name]))
+            else:
+                sys.stdout.write(line)
+                sys.stderr.write('didnnt find %s\n'%name)
+

### 2014-10-29 talk/uct2014/example/interp.py

@@ -1,0 +1,52 @@
+
+import time
+
+class InterpreterError(Exception):
+    def __init__(self, msg):
+        self.msg = msg
+
+def interp(bytecode, start_value):
+    i = 0
+    lgt = len(bytecode)
+    accumulator = start_value
+    while i < lgt:
+        c = bytecode[i]
+        if c == 'd':
+            accumulator -= 1
+            i += 1
+        elif c == 'l':
+            arg = ord(bytecode[i + 1])
+            i += 2
+            if accumulator > 0:
+                i -= arg + 2
+        else:
+            raise InterpreterError("Unknown char %s" % (c,))
+    return accumulator
+
+def entry_point(argv):
+    if len(argv) != 3:
+        print "Wrong number of args, requires bytecode and start_value"
+        return 1
+    bytecode = argv[1]
+    try:
+        acc = int(argv[2])
+    except ValueError:
+        print "Expected int, got %s" % (argv[2],)
+        return 2
+    try:
+        t0 = time.time()
+        res = interp(bytecode, acc)
+        tk = time.time()
+        dt = tk - t0
+    except InterpreterError, e:
+        print e.msg
+        return 3
+    print "Got %d, time %f" % (res, dt)
+    return 0
+
+def target(*args):
+    return entry_point
+
+if __name__ == '__main__':
+    import sys
+    entry_point(sys.argv)

### 2014-10-29 talk/pyconpl-2014/benchmarks/fib.py
draft

@@ -1,7 +1,11 @@
 
 import time
 import numpy
-from matplotlib import pylab
+try:
+    from matplotlib import pylab
+except:
+    from embed.emb import import_mod
+    pylab = import_mod('matplotlib.pylab')
 
 def fib(n):
     if n == 0 or n == 1:
@@ -21,7 +25,7 @@ for i in xrange(1000):
 
 hist, bins = numpy.histogram(times, 20)
 #pylab.plot(bins[:-1], hist)
-pylab.ylim(ymin=0, ymax=max(times) * 1.2)
-pylab.plot(times)
+pylab.ylim(0, max(times) * 1.2)
+pylab.plot(numpy.array(times))
 #pylab.hist(hist, bins, histtype='bar')
 pylab.show()

### 2014-11-17 talk/agile2006/draftpaper_agile2006.tex
Add my own comments in three places

\documentclass[twocolumn,10pt,a4paper,english]{article}
\usepackage{latex8}
\usepackage{babel}
\usepackage{ae}
\usepackage{aeguill}
\usepackage{shortvrb}
\usepackage[latin1]{inputenc}
\usepackage{tabularx}
\usepackage{longtable}
\setlength{\extrarowheight}{2pt}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{color}
\usepackage{multirow}
\usepackage{ifthen}
\usepackage[colorlinks=false,linkcolor=black,urlcolor=black]{hyperref}
\usepackage[DIV12]{typearea}
%% generator Docutils: http://docutils.sourceforge.net/
\newlength{\admonitionwidth}
\setlength{\admonitionwidth}{0.9\textwidth}
\newlength{\docinfowidth}
\setlength{\docinfowidth}{0.9\textwidth}
\newlength{\locallinewidth}
\newcommand{\optionlistlabel}[1]{\bf #1 \hfill}
\newenvironment{optionlist}[1]
{\begin{list}{}
  {\setlength{\labelwidth}{#1}
   \setlength{\rightmargin}{1cm}
   \setlength{\leftmargin}{\rightmargin}
   \addtolength{\leftmargin}{\labelwidth}
   \addtolength{\leftmargin}{\labelsep}
   \renewcommand{\makelabel}{\optionlistlabel}}
}{\end{list}}
\newlength{\lineblockindentation}
\setlength{\lineblockindentation}{2.5em}
\newenvironment{lineblock}[1]
{\begin{list}{}
  {\setlength{\partopsep}{\parskip}
   \addtolength{\partopsep}{\baselineskip}
   \topsep0pt\itemsep0.15\baselineskip\parsep0pt
   \leftmargin#1}
 \raggedright}
{\end{list}}
% begin: floats for footnotes tweaking.
\setlength{\floatsep}{0.5em}
\setlength{\textfloatsep}{\fill}
\addtolength{\textfloatsep}{3em}
\renewcommand{\textfraction}{0.5}
\renewcommand{\topfraction}{0.5}
\renewcommand{\bottomfraction}{0.5}
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}
% end floats for footnotes
% some commands, that could be overwritten in the style file.
\newcommand{\rubric}[1]{\subsection*{~\hfill {\it #1} \hfill ~}}
\newcommand{\titlereference}[1]{\textsl{#1}}
% end of "some commands"
\title{Trouble in Paradise: the Open Source project PyPy, EU-funding and Agile practices}
\input{paper.sty}
\hypersetup{
pdftitle={Trouble in Paradise: the Open Source project PyPy, EU-funding and Agile practices}
}
\raggedbottom
\begin{document}
\maketitle

\setlength{\locallinewidth}{\linewidth}
% sectnum 
\begin{abstract}
PyPy is an Open Source project, partly funded by the European Union, employing
agile techniques evolved within the Python Community such as ``sprint-driven
development''.The project started as a grass-root F/OSS effort in late 2002 and
received EU-funding from December 2004 until November 2006.  In this paper we
will present the various influencing factors that creates the hybrid project
process that is PyPy. These influencing factors are the F/OSS Python Community
(climate of the community from which PyPy grew from), agile practices (such as
the Python community evolved technique of ``sprinting'') and the EU funding
practices (resource tracking and reporting) impacting the project. These
influencing factors laid the foundation for the custom-made project process
that makes this unique hybrid project work. The main factor for driving this
process is the skills of the team of core developers instigating the project.
PyPy, with its open and transparent communication and collaborative work style,
is again a proof that the best agile practice is the people factor.
\end{abstract}

%___________________________________________________________________________

\hypertarget{introduction}{}
\section{Introduction}

This paper presents the story and the experiences gathered so far in the PyPy
project. The challenge has been to integrate such diverse perspectives as agile
practices being used in a distributed development style in an
Open Source project that is partly funded by the European Union.

The PyPy project started as a grass-root effort among core developers in the
Python language community, aiming at building a Python implementation purely
built in Python.  It was a language implementation project that was purely
driven from a non-profit perspective which rapidly increased in complexity when
the Open Source project applied for EU-funding.

Today the project has over 4 years of intense activity as an Open Source
community (F/OSS, Free Open Source Software) effort and have completed a
successful first year of the EU-funded effort.  In order to understand how the
PyPy project manages to strike a balance between being an F/OSS community,
while also having the core work of the project being funded by the European
Union (with all the structure and process requirements that entails) we need to
look at the influencing factors of the project process.

The influencing factors of the PyPy project process are:
\begin{itemize}
\item {} 
The F/OSS/Python community factor. Different Open Source communities have
different ``flavors'' regarding culture, communication climate, structure,
decision-process etc. PyPy was born out of the Python Community and shares
some of its characteristics which have had a huge influence on the project
process so far. Aspects such as distributed/dispersed development and the
supporting practices and infrastructure in place to quality assure software
development in a collaborated manner are part of this important influencing
factor.

\item {} 
The Agile factor. Almost a subset of the F/OSS/Python community factor -
agile practices that have been in use since the inception of the project, that
in fact have been the main drivers of the project are very much inspired by
other projects in the Python community. A key aspect of this influencing
factor is the Python version of ``sprinting''. We will present this ``agile''
practice, explain and trace its origins within the Python community and
identify which of the main agile practices inspired the Python world to go
``sprinting'' - a key technique today not only for the PyPy project but also
for all main Python projects. We will also touch on aspects such as Test
Driven Development and explain how these are implemented in PyPy.

\item {} 
The EU factor. PyPy is the first F/OSS community that organized themselves
into an EU-project, receiving EU-funding. It has had a huge impact on the
project, and it is important to identify how EU project and process
requirements can be ``weaved'' into an Agile F/OSS project - a challenge
indeed. The EU-funding has been and still is an enormous opportunity; the
project would not have had the rapid progress and impact it is now having
without the funding. It is also very important to note that without striking
the right balance between the three main influencing factors (F/OSS, agile
and EU-funding) this factor might also prove to be a ``show stopper'', something
that might influence the PyPy community negatively. This is a constant
challenge - we explain here how we created our process and added a
``consortium-level'' structure to the F/OSS community and the steps we took in
order to mitigate this risk.

\end{itemize}

Our conclusion so far in the project is that we believe that the practice of
sprint-driven development, as being used in the PyPy project, makes Agile and
Distributed/dispersed work styles more possible to combine. We also believe it
is a very useful practice for creating value and ensuring quality in projects
with hybrid cultures and methodologies. It is our aim to start to show-case
this with experiences from the first funded year of PyPy. We also feel that our
experiences as a unique hybrid project would be of interest to other
communities such as the agile community and we hope that this paper will
encourage cross-community interaction and communication.

%___________________________________________________________________________

\hypertarget{project-context}{}
\subsection{Project context}

%___________________________________________________________________________

\hypertarget{objectives}{}
\subsubsection{Objectives}

Due to the dual nature of the project even such a thing as objectives can get
complicated. If you, as a potential PyPy user, would visit out main developer
server, Codespeak, you would find the following statement:

``The PyPy project aims at producing a flexible and fast Python implementation.
The guiding idea is to translate a Python-level description of the Python
language itself to lower level languages. Rumors have it that the secret goal
is being faster-than-C which is nonsense, isn't it? more...''

Codespeak housed the F/OSS project since its inception and it has been a
driving strategy to continue a look-and-feel of a purely F/OSS project in order
to not confuse developers interested in just that - PyPy as a Python
implementation.

Digging a bit deeper you can find documentation which states the objectives of
the EU-funded part of the project. The EU project´s title is ``PyPy'' which is an
``acronym'' for ``Researching a Highly Flexible and Modular Language Platform and
Implementing it by Leveraging the Open Source Python Language and Community''.
The EU-project has three objectives: one technical, one research-oriented and
one methodology-oriented objective. In short PyPy is about building an
optimized compiler and interpreter purely in Python, applying Aspect Oriented
Programming and abstract interpretation techniques.  The methodology objective
aims at showcasing the ``sprint-driven development method'' being used by PyPy
and you can view this paper as part of fulfilling the methodology objective.
For more references on sprint-driven development, please see our paper for
the XP 2006 conference \cite{key-1}.

%___________________________________________________________________________

\hypertarget{strategy}{}
\subsubsection{Strategy}

The strategy of the project is to leverage the community of PyPy and Python
through an open and transparent communication and working style. The challenge has
been to implement this strategy, not only in the  F/OSS community part of the
project, but also in the partially funded consortium structure of the project.

The F/OSS part of PyPy has no budget tied to it and no formal organizational
structure.  There are no resources tied to it other than the volunteers
participating in the development process. These are the core developers and
architects behind PyPy but also people from all over the world, driven by a
combination of professional need and personal interest in Python language
implementations such as PyPy and choosing to contribute to PyPy. There is also
a large group of students on PHD levels participating and contributing, using
their areas of interest in PyPy as thesis materials.

It is difficult to estimate the amount of people involved in PyPy but we
estimate that 300-500 people actively follow the progress of the project -
this might mean reading emails, IRC logs and documentation in some cases, asking
questions and sending bug reports in others. There are around 50 people who
have commit-rights to the source code. The core group of developers consists of
around 10 people.

The EU part of the project is organized through a consortium which consists of
8 partners: DFKI (Germany), Ab Strakt (Sweden), Logilab (France), merlinux GmbH
(Germany), tismerysoft GmbH (Germany), Change Maker (Sweden) , Impara GmbH
(Germany) and Heinrich Heine Universität Düsseldorf (Germany) and 4 physical
person partners: Laura Creighton (Sweden), Richard Emslie (UK), Eric Van Riet
Paap (Netherlands) , Niklaus Haldiman (Switzerland).

The project effort of work for the 2 years of funding consists of 14
work-packages and in total 58 deliverables which are high-level functional and
non-functional requirements that were formulated in a proposal and form the
``Description of Work'' in the contract with the European Commission. The funding
received for this effort is 1.3 million euro. Of the core group of developers
mentioned above almost all of them (10 people) are involved in some sense in
the partner companies and organizations of the PyPy consortium.

%___________________________________________________________________________

\hypertarget{history}{}
\subsubsection{History}

The ideas behind PyPy started via discussions on European mailing-lists in the
Python community late 2002 by people who had been active in the Python
community for some time; core developers interested in language implementation.
They based the discussions and ideas partly on their experiences with
developing and driving some well-known projects in the Python community. They
met up in February 2003 at a ``sprint'', a one week working meeting,to draft
their ideas and to start PyPy as an Open Source project. The sprint facilities
as well as travel, accommodation and time was financed by all participants
privately.

Mid 2003 the idea of trying to get EU-funding for the project was born. It
became clear that the project had an arbitrarily large scale and that receiving
some funding would dramatically increase the pace and seriousness of the
project - because funded developers can dedicate more of their time to the
project.  The involved developers and people stretched outside of the  Open
Source ecologies to try to gather as much information and contacts as  possible
in order to answer the question: ``Should we go for it?''  to which the answer
quickly became ``Let's see how far we get!''.

Acting on this strategy proved to be a more difficult task. The entire proposal
and negotiation process took over a year (Autumn 2003 until November 2004).A
proper description of planned work, necessary to satisfy formal requirements,
had not previously been part of the development focus and both the EU and the
parties involved had to adapt to the situation.

The contract with the EU was signed and the funded part of the project, with
its consortium structure, started 1 of December 2004. The funded part of the
project ends in November 2006.

The first funded year of the project ended with a review in Brussels in January
2006, hosted by the Commission, reviewing the 10 deliverables comprising the
work of the first year. All deliverables were accepted; based on
recommendations by the external reviewers. The consortium and the Commission is
now restructuring the volume of the remaining the deliverables (although not
changing the scope of the project).

%___________________________________________________________________________

\hypertarget{influencing-factors-the-f-oss-python-community-culture}{}
\section{Influencing factors: the F/OSS Python community culture}

%___________________________________________________________________________

\hypertarget{the-python-community}{}
\subsection{The Python community}

Python is an Open Source language, published under an OSI approved open source
license.  It was created by Guido van Rossum and is now one of the five most
used languages in the world.  Due to the nature of the language there is a
strong focus on glue and integration with other languages such as C and Java.
Typical aspects of the Python community is that it houses four large
language implementations as separate projects which communicate and discuss
their experiences and approaches with each other. This intra-extra community
focus and interest has created a collaborative atmosphere with an open and
transparent communication climate.

%___________________________________________________________________________

\hypertarget{the-pypy-community}{}
\subsection{The PyPy community}

The PyPy project is one of the four large language implementation projects in
the Python community.  PyPy grew from the architectural experiences from
previous successful Python projects and this prior community experience was
vital for the evolving of the PyPy community due to the established trust in
expert skills of the core developers starting PyPy - again the people
factor. Thus making it easier recruit people into the PyPy community.

The PyPy project inherited the focus on collaborative approaches and open
communication climate.  The strategy of using sprints, as a core technique, to
kick-start the project as well as moving the sprints to different
locations had clear effects. It encouraged participation by meeting people locally.
During the period of 2003-2004 6 sprints were arranged in various European cities.
These sprints were hosted by universities and private
individuals, participation was funded privately. The effect on the evolving community
was a stable subscriber participation on the development list of between
140-150 people. After the EU-funding and the more systematic structure of
sprinting every 6th week the amount of subscribers went from around 150 people to
over 250 people. Thus turning the prior experience of separate sprints into
a more systematic approach, sprint-driven development.

%___________________________________________________________________________

\hypertarget{supporting-infrastructure}{}
\subsection{Supporting infrastructure}

The amount of collaborative focus and open and transparent communication in an
open source community will manifest itself in the supporting infrastructure.
In PyPy version control is the primary means of providing a secure platform for
incremental development. Subversion is used, covering both
program files and documentation. Several automated mailing lists make sure
that every person involved receives instant notification on changes -
supporting peer review on all levels.

PyPy is also in some sense very much test-driven. PyPy is one of the primary
users of a separate F/OSS project called the py.test, tools for automated
testing with PyPy developers contributing improvements to py.test.  Py.test
contains a sophisticated system for running tests against the Python Standard
Library. It provides automatic selection of overriding tests in case there is
internally developed code that overrides the Python Standard Library. These are
important compliance features when the project is about language implementation.
There are also powerful mechanisms for disabling specific
tests and for producing useful traceback output. There are also automatically
generated web pages that show the status of all tests as per the latest checked
in test results. This together with a public issue-tracker covering bugs keeps the
development group focused on the quality of the code when doing
continuous integration. It greatly reduces the need
to coordinate and delegate refactoring work - this is handled in a self-organized way.

The main communication channels between developers involved in PyPy is to
discuss over IRC (Internet-Relay-Chat) - open for all who are interested.
Several mailing lists for discussions and information are also used. Web pages,
documentation, tutorials, talks and papers etc are all available on the central
developer server for everyone to access. A very useful feature that really supports
information and communication are the public email archives covering the key
mailing lists - going back to the start of 2003. As a newcomer to the project
in the fall of 2003 these public and easily accessible mailing list archives
was the primary means for me personally to get into the project. It provided answers
to who had been key people in the process, what had been crucial topics, how had
discussions and decisions been handled. It is also regularly being used with newcomers
in the development team, providing fast answers to questions and also providing context
to technical decisions being made in the past. A useful just-in-time documentation
and a unorthodox but very efficient way of disseminating information and knowledge.

Now - all this infrastructure is being used in almost all larger open source
projects and quite a few of them are much more hierarchical and non-transparent
in their communication.  In PyPy there is no hierarchy for receiving
commit-rights. A newcomer to PyPy can instantly receive an account on the
development server, with full commit rights to the version control system. The
reason why there are no constraints is that the process is very much
self-organized and managed by the developers from a social level (peer review,
code review, supporting communication channels, coaching, mentoring) and with
strong automated tools covering tests, versions and back ups covering the
technical level.

An extensive coding guide, published on codespeak.net (development website)
serves as an introduction to the set of specifications being used while coding
PyPy.  Yet another piece of the puzzle, the coding guide together with
commit-access to both source code and tests supports cohesion in the development
process. This increases readability and uniformity so that automated testing tools can
be more efficiently used. Again - simple and useful support that is much needed when
working dispersed in a larger team of developers that might never meet face to face.

In a distributed and dispersed work style these two (social and technical)
levels needs to be consistent and support each other. Discrepancies would be
immediately noticed. As stated before - the people factor is again evident.
If you wish to encourage participation and contribution there has to be trust as
well as a supportive environment in order to manage a virtual collaborative
workspace. The main feature of such a supportive infrastructure is to reduce
the cost of information since coordination and communication is even more difficult
in a dispersed environment. If this easily accessed information of the status of the software
is combined with trust (that is commit-rights) you have the basis for a proactive,
self-organized culture.

%___________________________________________________________________________

\hypertarget{supporting-practices}{}
\subsection{Supporting practices}

Another good example of the collaborative nature of the Python community is the
way in which the various projects share best practices. Core developers in the
PyPy project picked up the practice of synchronization meetings as a powerful
way of supporting distributed and dispersed development. This practice was inspired
by the experiences of development processes at Canonical.

Sync-meetings are weekly short coordination meetings between developers. These are
open to all developers active in the PyPy community, usually but not necessarily
involving aspects of EU-funded work on deliverables.  These 30 minute
IRC-meetings serve a weekly synchronization for regular discussions and
integration of ongoing work.  Meetings are prepared with an agenda sent out to
pypy-dev and minutes are distributed on pypy-dev and archived in the repository
with publicly accessible links. The work of preparing, managing and documenting the
meetings is rotated between active developers and is self-organized as well.

Sync-meetings have proved to be a very good complement to a process in which
the project sprints every 6th week. Sync-meetings keep cohesion as well as a
team focus-and-feel for the dispersed work style between sprints. There are three
crucial aspects of this practice. The first one is that our experience as well as
that of Canonical points to keeping the time for the meeting brief. The reason for
30 minutes limit is that it gives priority on what topics to choose. However
complex your development situation is you should choose not more than 3 topics,
topics that could be discussed and decided upon during these 30 minutes. Not having
this time limit would create long and tiresome IRC meetings which would affect motivation
people and also create more confusion than results.

The second aspect is that it has a fixed format in which the meeting starts with
all developers participating in the meeting presents a very short status - answering
the questions LAST (last week), NEXT (their focus for the upcoming week)
and BLOCKERS (if they are stuck and need help). This means that the group get
overview and can track progress, albeit loosely. All individuals get an opportunity to
voice their personal situation and the group jointly discuss potential blockers and how
to solve them. The fixed format help to keep focus and creates a rhythm that makes it
easier for newcomers to get into the flow of the meeting.

The third aspect is also the main challenge with the practice. In PyPy we rotate
the organization of the meetings between the developers. But because meeting
moderation has to do with experience, people skills as well as a certain overview of
the project not everyone can actually do this task. Our experience shows that it is
also very important to have core people involved in identifying topics needing to be
discussed - having the progress of the project in mind. It is very easy to choose
topics that do not need the group mind to discuss, that are too detailed, that has
too much focus on information distribution rather than discussions.

Here is what one of the new core developers of PyPy says about sync-meetings:
\begin{quote}

\emph{``Sync-meetings are useful because they enable developers to discuss and
clarify issues among themselves and to provide a common focus when working
distributedly. They are also a lightweight way to synchronize activities,
resolve blockers and to get an overview about who is currently doing what
work.''}
-- Carl Friedrich Bolz
\end{quote}

%___________________________________________________________________________

\hypertarget{influencing-factors-agile-practices-in-pypy}{}
\section{Influencing factors: agile practices in PyPy}

%___________________________________________________________________________

\hypertarget{sprints}{}
\subsection{Sprints}

PyPy first started during a one-week meeting, a ``sprint'', held at Trillke-Gut
in Hildesheim February 2003. The sprint was inspired by practices used by other
Python projects such as Zope3.  Originally the sprint methodology used in the
Python community grew from practices applied by the Zope Corporation.  Their
definition of a sprint was:  ``two-day or three-day focused development session,
in which developers pair off together in a room and focus on building a
particular subsystem''.

Tres Seaver of the Zope Community, one of the instigators of the sprint
method as it is used in the Python community says the following about how they
evolved the method during 2001 and 2002:
\begin{quote}

\emph{``The motivation was to begin using as much of XP as would fit to
accelerate Zope3 development. Given the different development culture of
the Zope community (framework-centric, distributed, no ''business user``
present), some of the XP mantras /practices couldn't be done directly.  We
decided to try to do ''as much XP as possible`` for short, highly-focused
sessions, with all active committers collocated.''}
\end{quote}

The Zope community as well as other Python projects such as PyPy have seen that
sprints generate results beyond the creation of software:
\begin{itemize}
\item {} 
It helped to evolve the community through participation and hands-on contact
with the core developers.

\item {} 
It is a live training session not only in producing code but also in the
development methods being used (the sprint method in itself, pair programming
and TDD).

\item {} 
It supports activities such as design decisions and high-level requirements
discussions, creating cohesion and understanding as well as minimizing risks
with dispersed/distributed work where these kind of activities can be really
difficult to manage.

\end{itemize}

In order to provide examples - in PyPy these discussion sessions have been done in
various different ways during sprints. The proposal for the EU and the work package
descriptions were drafted through group discussions during sprints. Later sprints in which
the process needed design decisions, developers paired up in several groups of 2-4
people per group for daily discussion sessions. These sessions ended with brief
documentation and presentations for the other groups where decisions were made.
Usually what happens is that the daily sprint planning results in a pair of core developers
choosing a topic to discuss. These discussions have even acted as tutorials for other
developers sitting in and following the discussions. Some of these discussions have also
been filmed for dissemination purposes.

The reason for naming the method sprinting, is not at all connected to Scrum and
the usage of sprints in that context, according to Tres Seaver is:
\begin{quote}

\emph{``... because of the fact that we weren't doing XP in its proper sense,
which is more like running a long-distance race, where pacing oneself is
critical.  Instead, we were pushing as fast as possible in order to
maximize the benefit of limited face time.  Of all the XP practices, the
one we gave shortest shrift to is ''no overtime``: as sprints often go far
into the evening (or even the wee hours of the morning.''}
\end{quote}

To summarize the results of this first try out of the methodology shows that
sprinting was the driver for making Zope3 a community-driven project:  Zope
Corporation in fact ceded control over the development process to the new
``Zope3 core'' group created via sprinting. Less than 2 years later the same effect
was to be found in the F/OSS project PyPy - the community evolved through
sprint driven development.

The sprint method, as well as key aspects of F/OSS supportive infrastructure
and practices was established within the project before PyPy received its
EU-funding.  Thus, in the proposal,we put much emphasis on the methodology in
itself when designing the consortium level process. Examples of how this was
done was that we created a methodology objective, specific deliverables
regarding documentation of the sprint-driven methodology as well as designing a
process during the 2 years of funding in which the project and the developers
sprints every 6th week. This was taken into account when estimating the budget,
adding costs for travel and accommodation for all partner organizations -
covering 14 sprints during the 2 years.  So far during the funded part of the
project we have organized 10 sprints  - successful sprints from both a
development, recruiting, dissemination and networking aspects.  Another
important fact is that the sprints are organized by the developers and as such
is another good example of the collaborative manner of the culture of the
project and the community.

Why did PyPy choose sprinting as a key technique?  It is a method that fits
distributed teams well because it gets the team focused around visible
challenging goals while working collaboratively (pair-programming, status
meetings, discussions etc) as well as accelerated (short increments and tasks,
``doing'' and testing instead of long startups of planning and requirement
gathering, continuous integration). This means that most of the time a sprint
is a great way of getting results and getting new people acquainted - a good
method for dissemination of knowledge and learning within the team.

Another insight, worthwhile for other more discipline based projects to ponder,
is how an agile process like sprinting is much more suited for creative work
between groups of distributed people. Traditional software development, as well
as traditional project management techniques have a tendency to hinder creativity
due to the inbuilt over-structured, segmented and control-oriented approach
which in most cases ends in less quality when results are being measured.

%___________________________________________________________________________

\hypertarget{eu-project-practices}{}
\section{EU-project practices}

%___________________________________________________________________________

\hypertarget{consortium-structure}{}
\subsection{Consortium structure}

Key requirements regarding organizational structures of a Framework Programme
6 EU-project are:
\begin{itemize}
\item {} 
a consortium of partners performing the work described in the contract

\item {} 
a project co-coordinator managing contract administration and communication
between consortium and the Commission

\item {} 
a project manager, responsible for the project reaching its goals within the
time frame and budget

\end{itemize}

The challenge was to design a project process that created a minimal amount of
changes to the structure being in use in the F/OSS project. It was especially
important to maintain an open and transparent communication, even for decision
on consortium level and such. There was a fear of the EU-project adding
``top-down'' structures on the F/OSS work and on the core developers instigating
the project.

We identified a minimalistic approach of management roles (project
co-coordinator, project manager, assistant project manager) and created a
management team to make sure that there was collaboration between these roles.
Although the responsibility rested on the different management roles and the
management team the strategy implemented was to delegate as much as possible of
the responsibilities and decision-making to the core developers.

The strategy was to keep ``conceptual integrity'' \cite{key-2} of
the vision and the idea in the hands of the core developers. As Brooks
stresses, a uniform system design is better than uncoordinated and independent
design ideas. The core developers had a clear and agreed vision - but would
they be allowed to implement it within a fixed contract work style, with new
partners involved that had not been involved in the process from the start? The
core developers were organized into a technical board, responsible for planning
and coordinating the development work between the partners, with the mandate to
make decisions. A somewhat negative result was the added workload and
responsibility on developers regarding EU related work.

This structure was the most difficult to design and implement due to the very
different nature of its purpose compared to the collaborative, self-organized
nature of the F/OSS project - again it succeeded the way it has because of the
core developers acting as an interface between the developer group and the
consortium level work - again we see the trust aspect and the people factor in
play.

Sprints were budgeted for and designed into the process, together with a
time plan for all deliverables. The project was divided into three different
phases, depending on the nature of the work flow. In that sense we could
describe the EU-part of the project as a fixed-contract style of work, but with
time plan, deliverables and work package descriptions on a high-level, not
broken down into more granular tasks.

%___________________________________________________________________________

\hypertarget{communication-and-documentation}{}
\subsection{Communication and documentation}

All documentation and supporting communication infrastructure is hosted on the
same developer server as the source code, covered by the same version control
system. Only a few repositories are not publically available (such as the resource
tracking repositories) - but they are fully accessible for everyone employed by
a partner organization.

The communication infrastructure being used on the consortium level of work
mirrors that of the development work - having years of experience on the
distributed work style. In the case of PyPY there are mailing lists as well as
IRC channels for consortium level work. We even implemented a procedure in our
internal consortium agreement to allow for virtual meetings for decisions in
the consortium. So far this have been the primary meeting form and it has
worked well as a channel for making decisions. IRC-logs and minutes support the
procedure. In some cases decisions have also been made via email on the
consortium mailing list.

Although not the primary focus it is also useful to have the regular sprints to
coordinate specific issues between some partners or between all partners, have
physical meetings and discussions between the management team and the technical
board etc.

Our funding have also resulted in the possibility to have a more unorthodox approach
to documentation - the project have experimented in filming sprints to show the
development method as well as filming talks and discussion. Our film material
is planned to be released before summer 2006.

%___________________________________________________________________________

\hypertarget{troubles-in-paradise-striking-a-balance}{}
\section{Troubles in Paradise: striking a balance}

%___________________________________________________________________________

\hypertarget{developer-driven-versus-formal-project-structure}{}
\subsection{Developer driven versus formal project structure}

The fear of a top-down, hierarchical decision process of the consortium was a
justified one.  It is interesting for us to note that now, having the
experience and more contractual overview to note that there is nothing in the
requirements from the Commission that forces a project to a more traditional
and strict project process. It is mostly a question of the experience and
customs of the partner companies. In that sense there was much flexibility in
designing a process that allowed for a developer driven process regarding not
only software development but also sprint administration and planning as well
as consortium work.

The majority of the partners with key roles in the consortium organization had
been working together since before the funding. In that sense procedures and best
practices had been tried out. The results from the first year showed that a minimalistic
path could be identified and that the important work would be to review and
adjust the process when it did not support the work any more, or new situations
arose that we had not planned for (and that did happen!).

Already year two has another look-and-feel to it on a consortium level when it
comes to how the management team and the technical board works, because of the
different nature of work in year two.

%___________________________________________________________________________

\hypertarget{agile-strategies-versus-formal-eu-contractual-requirements}{}
\subsection{Agile strategies versus formal EU-contractual requirements}

Our main agile practice, sprint-driven development, was successfully integrated
into the formal contractual requirements, allowing for the same free-floating
agile process of self organizing and decision making as existed before the
funding, but with a more systematic and documented style.

But here is also our biggest failure, the main ``un-agile'' aspect of the entire
project having a large negative effect not only on the EU-part of the project
but also on the community.  We had planned for a process that kept an
``open-door'' policy that allowed us to fund non-consortium persons from the
community to attend sprints. The budget was allocated, the procedure within the
sprint context of handling more newcomers were known to us - the main
show stopper was that the PyPy sprint funding did not fit within the customs and
practices of contracts for cost claims in the Commission.

Every time we want to encourage participation and fund non-consortium
people to participate in sprints and contribute, the procedure now is that
they have to join the consortium as a full partner.
This creates the need for a contractual amendment with the Commission which
adds administrative work to the project as well as for the persons in question.
A too blunt instrument and today we still do not have a working solution to this
problem. It is an unfortunate example on how the influencing factor of F/OSS,
agile practices and EU-funding collides and creates negative impact on the project.

%___________________________________________________________________________

\hypertarget{f-oss-community-versus-hierarchies-for-conceptual-integrity}{}
\subsection{F/OSS community versus hierarchies for ``conceptual integrity''}

There are many examples of F/OSS projects and communities that have dried out
because of the fear of the architects to allow the contributors to fully
participate and also influence features and visions of the software. In PyPy
this challenge comes twofold. Not only is there a  challenge to let the core
developers keep ``conceptual integrity'' while also allowing the community to
influence the direction and features of the software. There is also the
consortium level of work in which the core developers could fear to ``lose'' the
mandate of driving the architectural and design work. And yet another layer of
risk and complexity would be if the consortium level would ``shut the door'' on
the community, enforcing a more closed development process.

As in many cases - being risk aware is the first step to mitigation. Because the
project is so deeply rooted in the F/OSS Python community, with its specific
culture and climate, and because so many people involved in the core work on
both the development and consortium level also shared this background they took
great pains to avoid situations like this to happen. In fact, what we have
realized only some months ago is that there is another risk, already partly
having effect on the community. Because of the funding the development process
is progressing much more rapidly in PyPy than more normal F/OSS projects. The
speed and the dramatically increasing learning curve could make members more
passive because they do not have the time to follow full-time IRC discussions,
postings on mailing lists and the actual source code and increasing test suites.

This is a challenge and it is the focus of the developer group to try to
distill and package information in order to help people to better navigate the
progress.

%___________________________________________________________________________

\hypertarget{conclusion}{}
\section{Conclusion}

The important question is the following (and is also part of the methodological
objective of the project): is there enough room to manage a project and Open
Source community within the plan-driven inspired methods that are required in
EU-funded projects, while still working agile and distributed?

We believe so. The one clear dominating factor to make all this succeed is, as
always, the people factor, the CRACK performers as Boehm and Turner calls them
(``Collaborative, Representative, Authorized, Committed, Knowledgeable'')
\cite{key-3}.

The core developers of the PyPy project had the right mix of various skills in
order to succeed in setting up a hybrid environment - enabling them to work
full time on a project they strongly believed in.  The most crucial mix of
skills for making this possible was/are:
\begin{itemize}
\item {} 
\textbf{Social:} The ability to communicate open and transparent, to mentor and
tutor dispersed as well as reinventing different collaborative work styles of
the sprint method, ``manage'' groups and community as well as consortium, and
handle conflicts)

\item {} 
\textbf{Leadership abilities:} The ability to step into formal leadership roles in
technical board structures, manage sprints and sync-meetings as well as the
more informal management of the community of developers. Managing the balance
between encouraging participation but still holding true to their vision,
their ``conceptual integrity''.

\item {} 
\textbf{Ability to network:} To be open to other communities, inviting new
partners in order to create a working consortium structure in the EU-project,
curious and collaborative towards other Python implementations and other
languages and research approaches, sharing knowledge and experiences and
seeking bets practices of other projects.

\item {} 
\textbf{Entrepreneurs:} To risk the community through pursuing the idea of
EU-funding in order to fulfill the ambitious vision, managing to not only
create a consortium with innovative structures of cooperation but also to
create new companies.

\item {} 
\textbf{Technical skills:} Programming language and implementation
aspects,frameworks, mathematics, computer science - core skills for the
project. Also the ability to design, setup and effectively manage supporting
infrastructure for the development process.

\item {} 
Managing the balance between encouraging participation but still holding true
to their vision, their ``\textbf{conceptual integrity}'' This while working agile with
open and transparent communication through sprints, documentation, tutorials,
mentoring, sync-meetings. Resulting in a lively and growing the F/OSS
community around the project.

\end{itemize}

So, could it be said that for an agile software development process, especially
one that is distributed and community oriented, within a framework of
EU-funding, that it is heavily people dependent? Or to stress it even further,
sprint-driven development as a methodology does not exist and function without
an agile group of people, Crack performers. The people are the methodology in
some sense and if you wish to draw upon the experience of the PyPy team you
need to look at the supporting practices around the people in order to find
what can be duplicated and tested in another project environment. This
conclusion matches what Alistair Cockburn writes in his paper ``Characterizing
People as Non-Linear, First-Order Components in Software Development''
\cite{key-4}:
\begin{quote}

\emph{``The fundamental characteristics of ''people`` have a first-order
effect on software development, not a lower-order effect.''}
\end{quote}

If we accept this conclusion then we can also, thanks to the people, start to
get innovative regarding practices. Designing the project process based on the
specific needs of the unique project environment you are facing. In the case of
PyPy this means that we are exploring the methodology as we go along, adjusting
and fine tuning the process as well as the software.

So, when drawing from these different skills within the community of
developers, the people, in the PyPy project one possible conclusion would be
that a truly agile approach dominating the work style of an Open Source project
will increase the ability of the community to spread the strategy of agility to
other domains.

By this we mean that what started as agile practices in the development process
quickly became influencing factors when designing other project processes.
Examples of this in PyPy is how the sprint-driven development acts as a focal
point not just for the development work (co-located as well as dispersed) but
also for the formal and informal management of the project. Sprints together
with the CRACK performers was what made the community grow and evolve. It was
the foundation for a hybrid project where agile practices and EU-funding can
fit within a distributed Open Source context.

%___________________________________________________________________________

\hypertarget{acknowledgments}{}
\section{Acknowledgments}

The author would like to thank the following people who have in various ways
helped with the creation of this paper: Angela Martin, Emily Bache, Tres
Seaver, Carl Friedrich Bolz.

I would like to dedicate this paper to my dear friend and mentor of the Open
Source Python and PyPy community Holger Krekel.
% References: 
% =========== 
\begin{thebibliography}{1}

 \bibitem{key-1} Beatrice Düring, "Sprint-Driven Development: Agile methodologies
 in a Distributed Open Source Project (PyPy)", XP 2006

\bibitem{key-2} Fredrick P. Brooks, Jr, "The mythical man-month, anniversary
edition", Addison-Wesley, 1995

\bibitem{key-3} Barry Boehm,Richard Turner, "Observations on Balancing
Discipline and Agility", (drawn from the book "Balancing Agility and
Discipline: A Guide to the Perplexed", Addison Wesley, 2003)

\bibitem{key-4} Alistair Cockburn, "Characterizing People as Non-Linear, First-Order
Components in Software Development", Presented at the 4th International
Multi-Conference on Systems, Cybernetics and Informatics, Orlando,
Florida, June, 2000, \texttt{http://alistair.cockburn.us/crystal/\\
articles/cpanfocisd/\\
characterizingpeopleasnonlinear.html}

\end{thebibliography}
\end{document}

### 2014-11-17 talk/dls2012/benchmarks/iter/wsum2d.c

@@ -1,0 +1,23 @@
+#include <stdio.h>
+#include <math.h>
+#include <stdlib.h>
+
+double result;
+
+double sum(double *a, int w, int h) {
+  int x, y;
+  double sa = 0;
+  for (y=0; y<h; y++) for(x=0; x<w; x++) sa += a[y*w + x] + w;
+  return sa;
+}
+
+#define W 10000
+#define H 10000
+
+int main(int ac, char **av) {
+  double *a = malloc(W*H*sizeof(double));
+  int i, n = atoi(av[1]);
+  for (i=0; i<n; i++) result=sum(a, W, H);
+  fprintf(stderr, "wsum2d:    ");
+  return 0;
+}

### 2014-11-17 talk/ecoop2009/benchmarks/factorial.cs

@@ -1,0 +1,24 @@
+using System;
+
+class Factorial
+{
+  public static void Main(string[] args)
+  {
+    int n = Convert.ToInt32(args[0]);
+
+    long start, stop;
+    start = DateTime.UtcNow.Ticks;
+    int res = factorial(n);
+    stop = DateTime.UtcNow.Ticks;
+    double secs = (stop-start) * 1e-7;
+    Console.WriteLine("C#:            {0} ({1} seconds)", res, secs);
+  }
+
+  public static int factorial(int n)
+  {
+    int res=1;
+    for(int i=1; i<=n; i++)
+      res *= i;
+    return res;
+  }
+}

### 2015-04-16 talk/ep2008/examples/resurrections.py
template for my talk

@@ -1,0 +1,14 @@
+
+l = []
+
+class A(object):
+    def __del__(self):
+        print "called"
+        l.append(self)
+
+a = A()
+del a
+import gc
+gc.collect()
+del l[0]
+gc.collect()

### 2016-04-02 talk/icooolps2009-dotnet/intro.tex
start my talk

@@
 ones found in other projects like IronPython is that the latter compile
-code at the method granularity: if on the one hand they can exploit
-some of the knowledge gathered at runtime \davide{this contradicts what is stated in the intro} (e.g.\ the types of method
-parameters), on the other hand they can do little to optimize most of
+code at the method granularity: they can do little to optimize most of
 the operations inside, because few assumptions can be made about the
+types of the arguments and the
 global state of the program.  The PyPy JITs, on the other hand, work at

### 2016-04-02 talk/dls2012/benchmarks/iter/xsum2d.c

@@ -1,0 +1,23 @@
+#include <stdio.h>
+#include <math.h>
+#include <stdlib.h>
+
+double result;
+
+double sum(double *a, int w, int h) {
+  int x, y;
+  double sa = 0;
+  for (y=0; y<h; y++) for(x=0; x<w; x++) sa += a[y*w + x] + x;
+  return sa;
+}
+
+#define W 10000
+#define H 10000
+
+int main(int ac, char **av) {
+  double *a = malloc(W*H*sizeof(double));
+  int i, n = atoi(av[1]);
+  for (i=0; i<n; i++) result=sum(a, W, H);
+  fprintf(stderr, "xsum2d:    ");
+  return 0;
+}

### 2016-04-02 talk/dls2014/paper/plots/plot_performance_nojit.py

@@ -1,0 +1,319 @@
+#!/usr/bin/python
+
+# benchmarks-repo at a26f2fb58413
+
+# for now: avg & stddev of the best
+
+
+
+from plot_performance import plot_speedups
+import matplotlib
+import os
+import sys
+matplotlib.use('gtkagg')
+
+from matplotlib import rc
+#rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
+## for Palatino and other serif fonts use:
+rc('font',**{'family':'serif','serif':['Palatino']})
+rc('text', usetex=True)
+
+args = None
+import matplotlib.pyplot as plt
+# import pprint - slow as hell
+
+
+interps_styles = {
+    "pypy-stm-nojit": {'fmt':'r-', 'linewidth':2},
+    "cpython": {'fmt':'b', 'dashes':(1,1)},
+    "pypy-nojit": {'fmt':'g', 'dashes':(5, 2)},
+    "jython": {'fmt':'m', 'dashes':(2, 5)},
+    "best": {'fmt':"k:"}        # only fmt allowed
+}
+
+
+#    pypy-c-paper-nojit bench.py -k5 raytrace/raytrace.py 1-4 256 256
+#    pypy-c-paper-nojit bench.py -k5 btree/btree.py 1-4 500000
+#    pypy-c-paper-nojit bench.py -k5 skiplist/skiplist.py 1-4 200000
+#    pypy-c-paper-nojit bench.py -k5 threadworms/threadworms.py 1-4 500000
+#    pypy-c-paper-nojit bench.py -k5 mandelbrot/mandelbrot.py 1-4 64 512 512
+#    pypy-c-paper-nojit multithread-richards.py 30 1-4 # report runtime
+# miller_rabin: bench.py -k5 primes/primes.py 1-4 400000
+# mersenne:  bench.py -k5 mersenne/mersenne.py $i 1500
+
+from collections import OrderedDict
+benchs = OrderedDict([
+     ("btree (small)", {
+        "pypy-stm-nojit":[
+            [7.98744797707,7.95994615555,8.0069360733,8.01075410843,7.99107217789],
+            [4.4982790947,4.51797103882,4.53245615959,4.51872205734,4.49158000946],
+            [3.48803210258,3.5547580719,3.51397109032,3.51541280746,3.50235295296],
+            [3.0812599659,3.05026102066,3.07854914665,3.05468702316,3.05405092239],
+        ],
[…]

### 2016-04-02 talk/dls2014/paper/plots/plot_scaling.py

@@ -1,0 +1,108 @@
+#!/usr/bin/python
+
+# obtained with time on
+#   pypy-c-paper-nojit bench_scaling.py [1-4]
+
+
+import matplotlib
+import os
+import sys
+matplotlib.use('gtkagg')
+
+from matplotlib import rc
+#rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
+## for Palatino and other serif fonts use:
+rc('font',**{'family':'serif','serif':['Palatino']})
+rc('text', usetex=True)
+
+args = None
+import matplotlib.pyplot as plt
+# import pprint - slow as hell
+
+xs = range(1,5)
+ys = [
+    [1.77,1.75,1.75,1.74,1.75],
+    [1.75,1.74,1.75,1.76,1.75],
+    [1.80,1.92,1.87,1.88,1.84],
+    [1.90,1.92,1.94,1.88,1.92],
+]
+
+
+
+def plot_mems(ax):
+    import numpy as np
+    y = []
+    yerr = []
+    # opt_y = [1.0] * len(xs)
+    first_time = np.mean(ys[0])
+    for x, d in zip(xs, ys):
+        normalized = map(lambda x:x/first_time, d)
+        y.append(np.mean(normalized))
+        yerr.append(np.std(normalized))
+
+    print y
+    ax.errorbar(xs, y, yerr=yerr,
+                label="pypy-stm-nojit")
+    # ax.plot(xs, opt_y, label="optimal")
+    return ax.legend(loc=4)
+
+
+def main():
+    global fig
+
+    print "Draw..."
+    fig = plt.figure()
+
+    ax = fig.add_subplot(111)
+
+    ax.set_ylabel("Runtime normalized to 1 thread")
+    ax.set_xlabel("Threads")
+    ax.set_ylim(0, 1.3)
+    ax.set_xlim(0, 5)
+
+    legend = plot_mems(ax)
+
+
+    #axs[0].set_ylim(0, len(x))
+    #ax.set_yticks([r+0.5 for r in range(len(logs))])
+    #ax.set_yticklabels(range(1, len(logs)+1))
+    #axs[0].set_xticks([])
+
+    # def label_format(x, pos):
+    #     return "%.2f" % (abs((x - left) * 1e-6), )
+    # major_formatter = matplotlib.ticker.FuncFormatter(label_format)
+    # axs[0].xaxis.set_major_formatter(major_formatter)
+
[…]

### 2016-04-02 talk/pycon-italy-2015/src/decode2.py

@@ -1,0 +1,42 @@
+import sys
+import struct
+
+P1 = '\x0c\x00\x00\x00"\x00\x00\x00\x07\x00'
+P2 = '\x15\x00\x00\x00+\x00\x00\x00\x08\x00'
+
+PLIST = [P1, P2] * 2000
+
+class Field(object):
+
+    def __init__(self, fmt, offset):
+        self.fmt = fmt
+        self.offset = offset
+
+def Message(name, fields):
+    class M(object):
+        def read(self, buf, name):
+            f = getattr(self, name)
+            return struct.unpack_from(f.fmt, buf, f.offset)[0]            
+
+    for fname, f in fields.iteritems():
+        setattr(M, fname, f)
+
+    M.__name__ = name
+    return M()
+
+
+Point = Message('Point', {
+    'x': Field('l', 0),
+    'y': Field('l', 4),
+    'color': Field('i', 8)
+    })
+
+
+def main():
+    res = 0
+    for p in PLIST:
+        x = Point.read(p, 'x')
+        res += x
+    print res
+
+main()

### 2016-04-07 talk/ecoop2009/jitgen.tex
(fijal, cfbolz) work on blog post

@@
 There are cases in which having a green variable is essential for generating
-good code, but it is not possible to use the \texttt{concrete} due to an
+good code, but it is not possible to use the \texttt{concrete} hint due to an
 unsatisfiable dependency: Section~\ref{sec:promotion} introduces

### 2016-07-18 talk/rupy2009/examples/interp.py
Slides, from the Swiss Python Summit

@@ -1,0 +1,67 @@
+
+
+
+
+
+
+
+[x, y]
+# LOAD_FAST
+frame.valuestack.push(frame.locals[0])
+# LOAD_FAST
+frame.valuestack.push(frame.locals[1])
+# BINARY_ADD
+v1 = frame.valuestack.pop()
+v0 = frame.valuestack.pop()
+v2 = v0.add(v1)
+  return IntObject(self.value + other.value) 
+frame.valustack.push(v2)
+# RETURN_VALUE
+return frame.valuestack[-1]
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+[x, y]
+v0 = IntObject(x.value + y.value)
+return v0
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+[x, y]
+return x + y
+
+
+

### 2016-08-01 talk/dagstuhl2012/talk.tex
draft for jitlog blog post

--- 2011-12-19 talk/icooolps2011/talk/talk.tex
@@
 \usepackage{alltt}
+\usepackage{hyperref}
 
 \usepackage[utf8x]{inputenc}
-
-\input{pygments}
 
@@
 
-\title[Runtime Feedback in a Meta-Tracing JIT]{Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages}
+\title{Meta-Tracing in the PyPy Project}
 
 \author[Carl Friedrich Bolz et.
-al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{3} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and Samuele Pedroni\inst{3} \and Armin Rigo\inst{1}}
+al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{1} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and Samuele Pedroni\inst{3} \and Armin Rigo\inst{1} \and many~more}
 % - Give the names in the same order as the appear in the paper.
@@
 
- $^3$Open End, Göteborg, Sweden \and
+ $^3$Canonical
 }
 
-\date{ICOOOLPS 2011, July 26, 2011}
+\date{Foundations of Scription Languages, Dagstuhl, 5th January 2012}
 % - Either use conference name or its abbreviation.
@@
 \begin{frame}
-  \frametitle{Good JIT Compilers for Dynamic Languages are Hard}
+  \frametitle{Good JIT Compilers for Scripting Languages are Hard}
   \begin{itemize}
-      \item recent languages like Python, Ruby, JS have complex core semantics
+      \item recent languages like Python, Ruby, JS, PHP have complex core semantics
       \item many corner cases, even hard to interpret correctly
-      \pause
-      \item feedback of runtime information to the compiler is necessary
-      \item correct exploitation of this information in the compiler
+      \item particularly in contexts where you have limited resources (like
+      academic, Open Source)
   \end{itemize}
@@
           \item ... and the common cases efficiently
-          \item feed back and exploit runtime information
+          \item while maintaing reasonable simplicity in the implementation
       \end{enumerate}
   \end{block}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Example: Attribute Reads in Python}
+  What happens when an attribute \texttt{x.m} is read? (simplified)
+  \pause
+  \begin{itemize}
+      \item check for \texttt{x.\_\_getattribute\_\_}, if there, call it
+      \pause
+      \item look for the attribute in the object's dictionary, if it's there, return it
+      \pause
+      \item walk up the MRO and look in each class' dictionary for the attribute
+      \pause
+      \item if the attribute is found, call its \texttt{\_\_get\_\_} attribute and return the result
+      \pause
+      \item if the attribute is not found, look for \texttt{x.\_\_getattr\_\_}, if there, call it
+      \pause
+      \item raise an \texttt{AttributeError}
+  \end{itemize}
 \end{frame}
@@
   \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Granularity Problems}
+  \pause
+  \begin{block}{Problems}
       \begin{itemize}
-      \item if the tracer records bytecode, not enough information is there
-      \item many dynamic languages have bytecodes that contain complex logic
+          \item traces typically contain bytecodes
+          \item many scripting languages have bytecodes that contain complex logic
           \item need to expand the bytecode in the trace into something more explicit
-      \item this duplicates the lanuage semantics in the tracer/optimizer
+          \item this duplicates the language semantics in the tracer/optimizer
       \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Example: Attribute Reads in Python}
-  What happens when an attribute \texttt{x.m} is read? (simplified)
-  \pause
-  \begin{itemize}
-      \item check for \texttt{x.\_\_getattribute\_\_}, if there, call it
-      \pause
-      \item look for the attribute in the object's dictionary, if it's there, return it
-      \pause
-      \item walk up the MRO and look in each class' dictionary for the attribute
-      \pause
-      \item if the attribute is found, call its \texttt{\_\_get\_\_} attribute and return the result
-      \pause
-      \item if the attribute is not found, look for \texttt{x.\_\_getattr\_\_}, if there, call it
-      \pause
-      \item raise an \texttt{AttributeError}
-  \end{itemize}
-  \pause
-  all this is one bytecode
+  \end{block}
 \end{frame}
@@
   \frametitle{PyPy}
-  A general environment for implementing dynamic languages
+  A general environment for implementing scripting languages
   \pause
@@
   \pause
-  \begin{block}{PyPy's Meta-Tracing JIT}
+\end{frame}
+
+\begin{frame}
+  \frametitle{PyPy's Meta-Tracing JIT}
   \begin{itemize}
@@
       \item needs a few source-code hints (or annotations) \emph{in the interpreter}
-          \item powerful general optimizations
+      \item allows interpreter-author to express language specific type feedback
+      \item contains powerful general optimizations
+      \pause
+      \item general techniques to deal with reified frames
   \end{itemize}
-  \end{block}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{Runtime Feedback}
+  \frametitle{Language Implementations Done with PyPy}
+  \begin{itemize}
+      \item Most complete language implemented: Python
+      \item regular expression matcher of Python standard library
+      \item A reasonably complete Prolog
+      \item Converge (previous talk)
+      \item lots of experiments (Squeak, Gameboy emulator, JS, start of a PHP, Haskell, ...)
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Some Benchmarks for Python}
+  \begin{itemize}
+      \item benchmarks done using PyPy's Python interpreter
+      \item about 30'000 lines of code
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[scale=0.3]{figures/all_numbers.png}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Telco Benchmark}
+  \includegraphics[scale=0.3]{figures/telco.png}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Conclusion}
+  \begin{itemize}
+      \item writing good JITs for recent scripting languages is too hard!
+      \item only reasonable if the language is exceptionally simple
+      \item or if somebody has a lot of money
+      \item PyPy is one point in a large design space of meta-solutions
+      \item uses tracing on the level of the interpreter (meta-tracing) to get speed
+      \pause
+      \item \textbf{In a way, the exact approach is not too important: let's write more meta-tools!}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Thank you! Questions?}
+  \begin{itemize}
+      \item writing good JITs for recent scripting languages is too hard!
+      \item only reasonable if the language is exceptionally simple
+      \item or if somebody has a lot of money
+      \item PyPy is one point in a large design space of meta-solutions
+      \item uses tracing on the level of the interpreter (meta-tracing) to get speed
+      \item \textbf{In a way, the exact approach is not too important: let's write more meta-tools!}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Possible Further Slides}
+  \hyperlink{necessary-hints}{\beamergotobutton{}} Getting Meta-Tracing to Work
+
+  \hyperlink{feedback}{\beamergotobutton{}} Language-Specific Runtime Feedback
+
+  \hyperlink{optimizations}{\beamergotobutton{}} Powerful General Optimizations
+
+  \hyperlink{virtualizables}{\beamergotobutton{}} Optimizing Reified Frames
+
+  \hyperlink{which-langs}{\beamergotobutton{}} Which Languages Can Meta-Tracing be Used With?
+
+  \hyperlink{OOVM}{\beamergotobutton{}} Using OO VMs as an implementation substrate
+
+  \hyperlink{PE}{\beamergotobutton{}} Comparison with Partial Evaluation
+
+\end{frame}
+
+\begin{frame}[label=necessary-hints]
+  \frametitle{Getting Meta-Tracing to Work}
+  \begin{itemize}
+      \item Interpreter author needs add some hints to the interpreter
+      \item one hint to identify the bytecode dispatch loop
+      \item one hint to identify the jump bytecode
+      \item with these in place, meta-tracing works
+      \item but produces non-optimal code
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[label=feedback]
+  \frametitle{Language-Specific Runtime Feedback}
   Problems of Naive Meta-Tracing:
   \begin{itemize}
-      \item no runtime feedback of user-level types
+      \item user-level types are normal instances on the implementation level
+      \item thus no runtime feedback of user-level types
       \item tracer does not know about invariants in the interpreter
@@
   \pause
-  \begin{block}{Proposed Solutions}
+  \begin{block}{Solution in PyPy}
       \begin{itemize}
-          \item introduce \textit{hints} that the interpreter-author can use
+          \item introduce more hints that the interpreter-author can use
           \item hints are annotation in the interpreter
@@
           \pause
-          \item two hints presented here
-          \item one to induce runtime feedback of arbitrary information
+          \item one to induce runtime feedback of arbitrary information (typically types)
           \item the second one to influence constant folding
@@
 
-\begin{frame}
-  \frametitle{Example: Instances with Maps}
-  \includegraphics[scale=0.7]{figures/map01.pdf}
+\begin{frame}[label=optimizations]
+  \frametitle{Powerful General Optimizations}
+  \begin{itemize}
+      \item Very powerful general optimizations on traces
+      \pause
+      \begin{block}{Heap Optimizations}
+        \begin{itemize}
+            \item escape analysis/allocation removal
+            \item remove short-lived objects
+            \item gets rid of the overhead of boxing primitive types
+            \item also reduces overhead of constant heap accesses
+        \end{itemize}
+      \end{block}
+  \end{itemize}
 \end{frame}
 
-\begin{frame}
-  \frametitle{Example: Instances with Maps}
-  \includegraphics[scale=0.7]{figures/map02.pdf}
+\begin{frame}[label=virtualizables]
+  \frametitle{Optimizing Reified Frames}
+  \begin{itemize}
+      \item Common problem in scripting languages
+      \item frames are reified in the language, i.e. can be accessed via reflection
+      \item used to implement the debugger in the language itself
+      \item or for more advanced usecases (backtracking in Smalltalk)
+      \item when using a JIT, quite expensive to keep them up-to-date
+      \pause
+      \begin{block}{Solution in PyPy}
+          \begin{itemize}
+              \item General mechanism for updating reified frames lazily
+              \item use deoptimization when frame objects are accessed by the program
+              \item interpreter just needs to mark the frame class
+          \end{itemize}
+      \end{block}
+  \end{itemize}
 \end{frame}
 
-\begin{frame}
-  \frametitle{Example: Instances with Maps}
-  \includegraphics[scale=0.7]{figures/map03.pdf}
+\begin{frame}[label=which-langs]
+  \frametitle{Bonus: Which Languages Can Meta-Tracing be Used With?}
+  \begin{itemize}
+      \item To make meta-tracing useful, there needs to be some kind of runtime variability
+      \item that means it definitely works for all dynamically typed languages
+      \item ... but also for other languages with polymorphism that is not resolvable at compile time
+      \item most languages that have any kind of runtime work
+  \end{itemize}
 \end{frame}
 
-\begin{frame}[containsverbatim]
-\frametitle{Map Implementation}
-
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{indexes}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{n}{indexes}
-        \PY{o}{.}\PY{o}{.}\PY{o}{.}
-
-    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{o}{.}\PY{o}{.}\PY{o}{.}
-
-\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}\PY{p}{)}
-\end{Verbatim}
-\end{frame}
-
-\begin{frame}[plain,containsverbatim]
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}
-
-    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n}{index} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
-        \PY{k}{return} \PY{n+nb+bp}{None}
-
-    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{o}{.}\PY{o}{.}\PY{o}{.}
-\end{Verbatim}
-\end{frame}
-
-\begin{frame}[plain,containsverbatim]
-\frametitle{Trace for Code \texttt{inst.a + inst.b}}
-\begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
-# $inst_1$.getfield("a")
-$map_1$ = $inst_1$.map
-$index_1$ = Map.getindex($map_1$, "a")
-guard($index_1$ != -1)
-$storage_1$ = $inst_1$.storage
-$result_1$ = $storage_1$[$index_1$]
-|\pause|
-# $inst_1$.getfield("b")
-$map_2$ = $inst_1$.map
-$index_2$ = Map.getindex($map_2$, "b")
-guard($index_2$ != -1)
-$storage_2$ = $inst_1$.storage
-$result_2$ = $storage_2$[$index_2$]
-
-$v_1$ = $result_1$ + $result_2$
-\end{lstlisting}
-\end{frame}
-
-\begin{frame}[plain,containsverbatim]
-\frametitle{Trace for Code \texttt{inst.a + inst.b}}
-\begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
-# $inst_1$.getfield("a")
-$map_1$ = $inst_1$.map
-$index_1$ = Map.getindex($map_1$, "a")
-guard($index_1$ != -1)
-$storage_1$ = $inst_1$.storage
-$result_1$ = $storage_1$[$index_1$]
-
-# $inst_1$.getfield("b")
-$map_2$ = $inst_1$.map
-$index_2$ = Map.getindex($map_2$, "b")
-guard($index_2$ != -1)
-$storage_2$ = $inst_1$.storage
-$result_2$ = $storage_2$[$index_2$]
-
-$v_1$ = $result_1$ + $result_2$
-\end{lstlisting}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-  \frametitle{Runtime Feedback Controlled by the Interpreter Author}
+\begin{frame}[label=OOVM]
+  \frametitle{Bonus: Using OO VMs as an implementation substrate}
+  \begin{block}{Benefits}
       \begin{itemize}
-      \item give the interpreter author a way to feed back runtime values into the trace
-      \item written as \texttt{promote(x)}
-      \item captures the argument's runtime value during tracing
-      \item should be used only for variables that take few values
+          \item higher level of implementation
+          \item the VM supplies a GC and mostly a JIT
+          \item better interoperability than what the C level provides
+          \item \texttt{invokedynamic} should make it possible to get language-specific runtime feedback
       \end{itemize}
+  \end{block}
+  \pause
+  \begin{block}{Problems}
+      \begin{itemize}
+          \item can be hard to map concepts of the scripting language to
+            the host OO VM
+          \item performance is often not improved, and can be very bad, because of this
+            semantic mismatch
+          \item getting good performance needs a huge amount of tweaking
+          \item tools not really prepared to deal with people that care about
+          the shape of the generated assembler
+      \end{itemize}
+  \end{block}
   \pause
@@
 
-\begin{frame}[containsverbatim]
-  \frametitle{Tiny Example}
-  \begin{minipage}[b]{6cm}
-      \centering
-      {\noop
-      \begin{lstlisting}[mathescape,basicstyle=\ttfamily]
-def f1(x, y):
-    promote(x)
-    z = x * 2 + 1
-    return z + y
-      \end{lstlisting}
-      }
-  \end{minipage}
-  \vline
-  \hspace{0.5cm}
-  \begin{minipage}[b]{4cm}
-      {\noop
-      \begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]
-guard($x_1$ == 4)
-|{\color{gray}$v_1$ = $x_1$ * 2}|
-|{\color{gray}$z_1$ = $v_1$ + 1}|
-$v_2$ = $z_1$ + $y_1$
-return($v_2$)
-      \end{lstlisting}
-      }
-  \end{minipage}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Foldable Operations Defined by the Interpreter Author}
-  \begin{itemize}
-      \item let the interpreter author define foldable functions
-      \item those functions typically don't look foldable
-      \item otherwise there is no need for an annotation
-      \item done via a function decorator \texttt{@elidable}
-      \pause
-      \item decorated functions should be pure
-      \item or have idempotent side effects (such as a function that memoizes)
-      \item trace optimizer will remove calls to such functions with constant arguments
-  \end{itemize}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-\frametitle{Adding Hints to Maps}
-
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{indexes}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{n}{indexes}
-        \PY{o}{.}\PY{o}{.}\PY{o}{.}
-
-    \PY{n+nd}{@elidable}
-    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}
-
-    \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{o}{.}\PY{o}{.}\PY{o}{.}
-
-\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}\PY{p}{)}
-\end{Verbatim}
-\end{frame}
-
-\begin{frame}[containsverbatim]
-\frametitle{Adding Hints to Maps}
-
-\begin{Verbatim}[commandchars=\\\{\}]
-\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
-    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
-        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}
-
-    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
-        \PY{n}{promote}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{)}
-        \PY{n}{index} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
-        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
-            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
-        \PY{k}{return} \PY{n+nb+bp}{None}
-
-    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
-        \PY{o}{.}\PY{o}{.}\PY{o}{.}
-\end{Verbatim}
-\end{frame}
-
-\begin{frame}[containsverbatim,plain]
-  \frametitle{Trace with Hints for Code \texttt{inst.a + inst.b}}
-\begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
-# $inst_1$.getfield("a")
-$map_1$ = $inst_1$.map
-guard($map_1$ == 0xb74af4a8)
-|{\color{gray}$index_1$ = Map.getindex($map_1$, "a")}|
-|{\color{gray}guard($index_1$ != -1)}|
-$storage_1$ = $inst_1$.storage
-$result_1$ = $storage_1$[$index_1$]
-
-# $inst_1$.getfield("b")
-|{\color{gray}$map_2$ = $inst_1$.map|
-|{\color{gray}guard($map_2$ == 0xb74af4a8)}|
-|{\color{gray}$index_2$ = Map.getindex($map_2$, "b")}|
-|{\color{gray}guard($index_2$ != -1)}|
-|{\color{gray}$storage_2$ = $inst_1$.storage}|
-$result_2$ = $storage_2$[$index_2$]
-
-$v_1$ = $result_1$ + $result_2$
-\end{lstlisting}
-\end{frame}
-
-\begin{frame}[containsverbatim,plain]
-  \frametitle{Final Trace}
-\begin{lstlisting}[mathescape,escapechar=|,basicstyle=\ttfamily]]
-$map_1$ = $inst_1$.map
-guard($map_1$ == 0xb74af4a8)
-$storage_1$ = $inst_1$.storage
-$result_1$ = $storage_1$[$0$]
-$result_2$ = $storage_2$[$1$]
-$v_1$ = $result_1$ + $result_2$
-\end{lstlisting}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Uses of These Hints}
-  \begin{block}{\texttt{promote} lets one specialize on various things:}
-      \begin{itemize}
-          \item user-level types
-          \item shapes of instances
-          \item the current state of a classes' methods
-          \item ...
-      \end{itemize}
-  \end{block}
-  \pause
-  \begin{block}{uses of \texttt{@elidable}}
-      \begin{itemize}
-          \item define immutable fields by decorating a getter
-          \item declare arbitrary invariants
-      \end{itemize}
-
-  \end{block}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Some Benchmarks}
-  \begin{itemize}
-      \item benchmarks done using PyPy's Python interpreter
-      \item about 30'000 lines of code
-      \item 20 calls to \texttt{promote}
-      \item 10 applications of \texttt{@elidable}
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Some Benchmarks}
-  \includegraphics[scale=0.5]{figures/bench.pdf}
-\end{frame}
-
-\begin{frame}
-  \frametitle{Conclusion}
-  \begin{itemize}
-      \item meta-tracing can make the efficient implementation of complex dynamic languages easier
-      \item only requires to write a correct interpreter
-      \item two kinds of hints to be added by the interpreter author allow arbitrary runtime feedback and its exploitation
-      \item the hints are expressive enough to re-implement classical optimizations such as maps
-      \item usage of the hints leads to good speedups for object-oriented code in PyPy's Python interpreter
-  \end{itemize}
-\end{frame}
-
-\begin{frame}
+\begin{frame}[label=PE]
   \frametitle{Bonus: Comparison with Partial Evaluation}

### 2016-08-01 talk/dls2012/demo/analytics.py

@@ -1,0 +1,21 @@
+
+from reloader import ReloadHack
+from io import view
+from background import Background
+from foreground import foreground
+from detect import find_objects
+
+class Tracker(ReloadHack):
+    def __init__(self):
+        self.background = Background()
+
+    def update(self, frame):
+        self.background.update(frame)
+        fg = foreground(frame, self.background.image)
+        for box in find_objects(fg):
+            box.draw(frame)
+        view(frame)
+        #view(self.background.image)
+        #view(255 * fg)
+
+

### 2016-08-01 talk/uct2014/example/demo2.py

@@ -1,0 +1,26 @@
+
+def adder(x, y):
+    return x + y
+
+def entry_point(argv):
+    if len(argv) != 3:
+        print "Wrong number of args"
+        return 1
+    try:
+        arg0 = int(argv[1])
+        arg1 = int(argv[2])
+    except ValueError:
+        print "Requires ints"
+        return 1
+    x = float(arg1)
+    for i in xrange(arg0):
+        x = adder(x, 1.2)
+    print "Result", x
+    return 0
+
+def target(*args):
+    return entry_point
+
+if __name__ == '__main__':
+    import sys
+    entry_point(sys.argv)

### 2016-08-18 talk/dls2006/paper.tex
remove a sentence -- enough to get the document to fit onto 12 pages.
\label was at the wrong place
Fix table labels.
Add another point

@@
 \end{verbatim}
+\caption{built-in dict insertion helper.}
 \label{fig_insertclean}
-\caption{built-in dict insertion helper.}
 \end{figure}
@@
 
-%%% XXX the first sentence seems to come a bit out of the blue.
-
-We found in-process debugging sufficient and did not implement dumping
-of any intermediate step to disk.  Figure \ref{fig_screenshot}
+Figure \ref{fig_screenshot}
 shows the control flow graph
@@
 really an abstract domain.  It handles objects that are placeholders.
-
-Its lattice order (actually a join-semilattice only) is extremely
-simple, because most actual analysis is delayed to the next phase, the
-type inference engine.  The objects are either \textit{Variables},
-which are pure placeholders for entirely unknown values, or
-\textit{Constants} with a concrete Python object as value.  The order
-places Variable as the top, and keeps all \textit{Constants}
-unordered.  Thus if two different constants merge during abstract
-interpretation, we immediately widen them to Variable.
+Its lattice order (actually a join-semilattice only) is shown in
+figure \ref{flowlattice}.
+
+\begin{figure*}
+\centering
+\includegraphics[scale=0.667]{image/flowlattice.pdf}
+\caption{the lattice order of the flow object space.}
+\label{flowlattice}
+\end{figure*}
+
+This order is extremely simple, because most actual analysis is delayed
+to the next phase, the type inference engine.  The objects are either
+\textit{Variables}, which are pure placeholders for entirely unknown values,
+or \textit{Constants} with a concrete Python object as value.  The order places
+Variable as the top, and keeps all \textit{Constants} unordered.  Thus if two
+different constants merge during abstract interpretation, we immediately
+widen them to Variable.
 
@@
 
-These types are ordered to form a lattice -- we generally use its
-structure of join-semilattice only, but we will see in section
-\ref{precision} a use case for the \textit{meet}. Pbcs form a
-classical finite set-of-subsets lattice, class annotations are ordered
-by inheritance.  We have left out a number of other annotations that
-are irrelevant for the basic description of the annotator and
-straightforward to handle: $Dictionary$, $Tuple$, $Float$,
-$UnicodePoint$, $Iterator$, etc.  The complete list and order is
-described in \cite{T}.
+Figures \ref{latticeoverall} and \ref{latticedetail} show how these
+types are ordered to form a lattice -- we generally use its structure of
+join-semilattice only, but we will see in section \ref{precision}
+a use case for the \textit{meet}.
+Not shown on the figures are the list terms, which are simply
+unordered with each other, and the Pbcs, which form a
+classical finite set-of-subsets lattice.  In addition, we have left
+out a number of other annotations that are irrelevant for the basic
+description of the annotator and straightforward to handle:
+$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
+complete list is described in \cite{T}.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.667]{image/lattice2.pdf}
+\caption{the part about instances and nullable instances, assuming a
+  simple class hierarchy with only two direct subclasses of
+  \texttt{object}.}
+\label{latticedetail}
+\end{figure}
 
@@
 \centering
-\label{perfsumm}
 \begin{tabular}{|l|c|c|} \hline
@@
 \caption{summary of interpreter performance.}
+\label{perfsumm}
 \end{table*}
@@
 \centering
-\label{translationtimes}
 \begin{tabular}{|p{0.65\linewidth}|c|} \hline
@@
 \caption{dividing overall translation time by stage.}
+\label{translationtimes}
 \end{table*}
@@
 
+\bigskip
+
 \bibliographystyle{abbrv}

### 2016-08-18 talk/dls2012/benchmarks/iter/median1d.c

@@ -1,0 +1,30 @@
+#include <stdio.h>
+#include <math.h>
+#include <stdlib.h>
+
+double result;
+
+double sum(double *a, int n) {
+  int i;
+  double sa = 0;
+  for (i=0; i<n; i++)  {
+    if (sa > a[i]) {
+      sa -= 1.0/(i + 1.0);
+    } else if (sa < a[i]) {
+      sa += 1.0/(i + 1.0);
+    }
+  }
+  return sa;
+}
+
+#define N 100000000
+
+int main(int ac, char **av) {
+  double *a = malloc(N*sizeof(double));
+  int i, n = atoi(av[1]);
+  double data[] = {-1.0, 1.0};
+  for (i=0; i<N; i++) a[i] = data[i&1];
+  for (i=0; i<n; i++) result = sum(a, N);
+  fprintf(stderr, "median1d:     ");
+  return 0;
+}

### 2016-08-27 talk/pyconpl-2010/demos/jit.py
More things to do later

@@ -1,0 +1,7 @@
+
+def f():
+    i = 0
+    while i < 20000000:
+        i += 1
+
+f()

### 2016-10-08 talk/ep2007/interpreter/listserver.py
Final tweaks

@@ -1,0 +1,35 @@
+from pypy.translator.js.lib import server
+from py.__.green.server.httpserver import GreenHTTPServer
+import random
+
+class Root(server.Collection):
+    def __init__(self):
+        self.l = range(10)
+        random.shuffle(self.l)
+    
+    def index(self):
+        return repr(self.l)
+    index.exposed = True
+    
+    def getitem(self, item):
+        return str(self.l[int(item)])
+    getitem.exposed = True
+
+    def setitem(self, item, value):
+        self.l[int(item)] = int(value)
+        return ""
+    setitem.exposed = True
+
+class Handler(server.NewHandler):
+    application = Root()
+
+if __name__ == '__main__':
+    addr = ('', 8010)
+    httpd = server.create_server(server_address=addr, handler=Handler,
+                                 server=GreenHTTPServer)
+    try:
+        httpd.serve_forever()
+    except KeyboardInterrupt:
+        pass
+
+    

### 2016-10-08 talk/ep2010/pypy_demo/settings.py

@@ -1,0 +1,80 @@
+# Django settings for pypy_demo project.
+
+DEBUG = True
+TEMPLATE_DEBUG = DEBUG
+
+ADMINS = (
+    # ('Your Name', 'your_email@domain.com'),
+)
+
+MANAGERS = ADMINS
+
+DATABASE_ENGINE = ''           # 'postgresql_psycopg2', 'postgresql', 'mysql', 'sqlite3' or 'oracle'.
+DATABASE_NAME = ''             # Or path to database file if using sqlite3.
+DATABASE_USER = ''             # Not used with sqlite3.
+DATABASE_PASSWORD = ''         # Not used with sqlite3.
+DATABASE_HOST = ''             # Set to empty string for localhost. Not used with sqlite3.
+DATABASE_PORT = ''             # Set to empty string for default. Not used with sqlite3.
+
+# Local time zone for this installation. Choices can be found here:
+# http://en.wikipedia.org/wiki/List_of_tz_zones_by_name
+# although not all choices may be available on all operating systems.
+# If running in a Windows environment this must be set to the same as your
+# system time zone.
+TIME_ZONE = 'America/Chicago'
+
+# Language code for this installation. All choices can be found here:
+# http://www.i18nguy.com/unicode/language-identifiers.html
+LANGUAGE_CODE = 'en-us'
+
+SITE_ID = 1
+
+# If you set this to False, Django will make some optimizations so as not
+# to load the internationalization machinery.
+USE_I18N = True
+
+# Absolute path to the directory that holds media.
+# Example: "/home/media/media.lawrence.com/"
+MEDIA_ROOT = ''
+
+# URL that handles the media served from MEDIA_ROOT. Make sure to use a
+# trailing slash if there is a path component (optional in other cases).
+# Examples: "http://media.lawrence.com", "http://example.com/media/"
+MEDIA_URL = ''
+
+# URL prefix for admin media -- CSS, JavaScript and images. Make sure to use a
+# trailing slash.
+# Examples: "http://foo.com/media/", "/media/".
+ADMIN_MEDIA_PREFIX = '/media/'
+
+# Make this unique, and don't share it with anybody.
+SECRET_KEY = '_+n^blk6hik*s1kfylp1pm8fpu9n0#=z_9cngayme5)-**$(#9'
+
+# List of callables that know how to import templates from various sources.
+TEMPLATE_LOADERS = (
[…]

### 2016-10-08 talk/pyconza2016/revdb/demo/demo1.py
Final tweaks

@@ -8,7 +8,6 @@ def make_list():
     lst.append(lst[50])
     return lst
 
-
 lst = make_list()
 
 for x in lst:
